### 2021-02-17 index.Rmd
Create bookdown project

---
title: "Characterizing and mitigating the effects of experimental bias on differential (absolute) abundance analysis in microbiome studies"
author: "Michael R. McLaren"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [main.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: mikemc/bda-theory
description: "Mathematical theory of the effects of experimental bias on differential (relative and absolute) abundance analysis in microbiome studies."
---

# Abstract

This monograph develops a mathematical theory of the effects of protocol- and taxon-specific experimental bias on various types of differential (relative and absolute) abundance analyses in microbiome studies and uses it to formulate several mitigation strategies.

### 2021-02-17 models.Rmd
Sketch Models section

# Measurement models {#models}

Deterministic measurement error models that extends the @McLaren2019 model of metagenomics measurement to 1) describe absolute as well as relative abundance, 2) include spike-in taxa, and 3) include supplementary measurements of bulk and targeted (absolute) abundance.

<!-- TODO: go through org notes and make sure all incorporated. -->

## Actual abundances and general notation

(For now) Number of samples = J; Number of taxa = I.

$A_{ij}$ is the actual absolute abundance of taxon $i$ in sample $j$, in the target units.
Unless specified otherwise, $A$ always refers to abundance in the original sample material, not in extracted DNA.
I will assume the target units are concentration of cells per unit mass. However, other units (such as biomass per volume) may be relevant.

Let $A_{Tj}$ (or $A_{\cdot j}$??) be the total abundance in sample $j$, $A_{Tj} = \sum_i A_{ij}$.

For variables that hold matrixes or vectors, capital letters denote absolute abundances and lower case letters denote proportions.
Thus $a_{ij}$ is the matrix of actual relative abundances in terms of cells as a proportion of the total amount of cells in the sample; $a_{ij} = A_{ij} / A_{Tj}$.

Bias vectors $B^{(P)}$ are vectors of length $I$ indicating the taxon-specific bias associated with a measurement protocol $P$.
I generally assume these to be sample-independent.
An overbar denotes the average value over taxa weighted by actual abundance; 
i.e., $\bar B_j = I^{-1} \sum_i B_i a_{ij}$.

<!-- Table  -->

Measurement type        Units                              Abundance matrix                 Bias vector
----------------        -----                              ----------------                 -----------
**A**ctual abundance    [cells]                            $A$, $a$ ($I\times J$)           -
**M**etagenomics        counts                             $M$, $m$ ($I\times J$)           $B^{(M)}$
Bul**k** abundance      [cells], [DNA], or [gene copies]   $K$ ($I\times 1$ column vector)  $B^{(K)}$
**T**argeted abundance  [cells], [DNA], or [gene copies]   $T$ ($|I_T| \times J$)           $B^{(T)}$
**S**pike-in abundance  [cells], [DNA], or [gene copies]   $S$ ($|I_S| \times J$)           $B^{(S)}$

Let $I_T$ and $I_S$ be the sets of taxa with targeted abundance measurements or known spike-in abundances, respectively.
Question: Will it be more convenient for $S$ and $T$ to have the standard shape $I\times J$?

## Metagenomics measurement

$M$ is a X BY X matrix representing the (marker-gene or) metagenomics measurement; $M_{ij}$ is the sequencing count associated with taxon $i$ in sample $j$, such that $M_{ij} = 10$ indicates that 10 fragments of DNA were sequenced and assigned to taxon $i$ and sample $j$.
In the case of paired-end sequencing, a single fragment of DNA yields two reads, which may or may not be counted jointly or independently by different algorithms; here I assume that read-pairs are assigned jointly, and will use "read" and "fragment" interchangeably.
For simplicity, I ignore the possibility for index switching or other sources of cross-contamination and assume that reads assigned to sample $j$ truly arose from sample $j$.

Following the notation defined above, I let $M_{Tj} = \sum_i M_{ij}$ is the total count for sample $j$; I will sometimes refer to this number as the *sequencing depth* or *read depth* of the sample, though in reality the total count will be less than the sample's sequencing depth since often a significant fraction of reads are lost by filtering steps prior to producing the final count matrix.
I use $m_{ij} = M_{ij} / M_{Tj}$ to denote the proportion of reads assigned to taxon $i$ in sample $j$.

I assume that the composition (relative abundances) is given by the deterministic error model presented in @McLaren2019.
In this model, the bias vector $B^{(M)}$ represents the *relative* efficiencies with which various taxa are measured, and so has the same effect on measurement if we rescale it $B \to cB$ for any $c > 0$.
To remove this degree of freedom, I will suppose that $B^{(M)}_{1} = 1$, so that $B^{(M)}_i$ is the efficiency of taxon $i$ relative to the first taxon.
We may take the definition of the MWC model as being that, conditional on the total count $M_{Tj}$, the observed read count of taxon $i$ is proportional to its actual abundance times its efficiency times a taxon-independent factor $C_j$ chosen to make sample $j$'s counts sum to $M_{Tj}$,
\begin{align}
  (\#eq:M)
  M_{ij} = A_{ij} B^{(M)}_{i} C_j, && C_j = \frac{M_{Tj}}{\sum_i A_{ij} B^{(M)}_{i}}.
\end{align}
In this model, the ratios among taxa are distorted by constant factors; for two taxa $i$ and $i'$,
\begin{align}
  (\#eq:M-ratio)
  \frac{M_{ij}}{M_{i'j}} 
    = \frac{m_{ij}}{m_{i'j}} 
    = \frac{A_{ij}}{A_{i'j}} \cdot \frac{B^{(M)}_{i}}{B^{(M)}_{i'}}
\end{align}
for all samples $j$.
If one or both taxa are absent ($A_i = 0$) or undetectable ($B_i = 0$), both sides of the equation are either $0$, $\infty$, or $0/0$ (undefined).
The observed proportion of a taxon equals its actual proportion multiplied by its efficiency relative to the sample mean efficiency,
\begin{align}
  (\#eq:m)
  m_{ij} = a_{ij} \cdot \frac{B_i}{\bar B^{(M)}_j}.
\end{align}

## Spike-ins

Let the set $I_S$ denote the indexes of the spiked taxa, which I assume are also measured by the metagenomics experiment and so also correspond to rows in $M$.
The $|I_S|\times J$ matrix $S$ describes the nominal abundances with which spike-in taxa were added to each sample.
Spike-in experiments are often designed so that a taxon $i$ is added in the same concentration to all samples, which we can represent by supposing the columns of $S$ to be identical.

The bias $B^{(S)}$ should be interpreted as error in the quantification of how much spike-in was added, such that
\begin{align}
  S_{ij} = A_{ij} B^{(S)}_i.
\end{align}
I assume that this error is constant across samples.
The motivation for this error model is as follows: Suppose that we misestimated the proportion of spike-in taxon $s\in I_S$ in our original spike-in stock, by a factor $B^{(S)}_s$; now its true abundance will be off by that factor in any spike-in derived from the starting stock.
Unlike the metagenomics efficiencies, these error factors are absolute numbers.

Taxa can be added as cells (pre-extraction) or as DNA (post-extraction).
In an application where DNA spike-ins are used to make inferences about changes in absolute cell abundances, we can include the difference between the nominal abundances $S$ (the spiked DNA concentration) and the actual abundances $A$ that are expected due to factors such as DNA extraction and copy-number variation in the bias term $B^{(S)}$, so long as DNA extraction is linear (see section on saturation effects below).

## Bulk absolute abundance measurement

Consider estimating the total absolute abundance $A_{T}$ via an aggregate or broad-range measurement, such as cell counting, flow cytometry, universal 16S qPCR, or total DNA concentration (e.g. Qubit).
I let $K_j$ donote the bulk absolute abundance measurement for sample $j$ (mneumonic: "K" for the last letter in "bulk").

Bulk measurements are affected by taxon-specific bias $B^{(K)}$, where $B^{(K)}_i$ indicates how efficiently cells of taxon $i$ contribute to the bulk measurement $K$.
Unlike the metagenomics efficiencies, these bulk measurement efficiencies are absolute numbers.
For example, cells of one taxon may be more reliably counted than another; and for broad-range 16S qPCR measurement, we expect taxa to contribute in proportion to how reliably they are lysed and to their 16S copy number.
Let $B^{(K)}_i$ denote the absolute efficiency with which a cell of taxon $i$ contributes to the measurement in sample $s$, which I assume is sample-independent.
We can write the estimate as
\begin{align}
  K_j = \sum_i A_{ij} B^{(K)}_i = A_{Tj} \bar B^{(K)}_j,
\end{align}
where $bar B^{(K)}_j$ is the mean bulk-measurement efficiency in the sample.

## Targeted absolute abundance measurement

Targeted measurement of the absolute abundance of specific taxa can be made via a method such as qPCR, ddPCR, or counting CFUs selective media.
These measurements too may be subject to bias; 
I suppose that the measurement of taxon $i$ is given by
\begin{align}
  T_{ij} = A_{ij} B^{(T)}_i,
\end{align}
where $B^{(T)}_i$ is the sample-independent, taxon-specific efficiency of the targeted measurement for taxon $i$.

## Complications

These models are the simplest for these types of experiments that have taxon-specific bias.
They therefore are idealizations that may be violated by real experiments in ways that affect our conclusions.
Here I list some relevant complications that we will return to in our analysis.
For more considerations, see the Discussion of @McLaren2019.

### Abundances units and protocol stages

*Still need to find the right way to synthesize these issues.*

qPCR measurements directly estimating the concentration of a marker gene in the extracted DNA.
Efficiency factors can translate the units from DNA to cell concentration *and* account for the bias imposed by extraction and marker-gene CNV (copy-number variation).
Sometimes we may want to refer to the DNA concentrations in the sample after extraction, and in that case I will use the variable $D$, and (perhaps) use $B^{(T/D)}$ to refer to the bias of the targeted measurement vs the DNA concentrations it was applied to.
This can be useful when we are considering qPCR and also DNA spike-ins.

### Saturation in DNA extraction yields

The equation for bulk measurement reflects an idealized form of bulk measurement in which, for a fixed taxonomic composition, $K_j$ is directly proportional to $A_T$.
However, a more realistic model for qPCR may be that $K_j$ is a saturating function of $A_T$, due to factors such as enzyme consumption or saturation in the elution step creating saturation in DNA yield during extraction.
However, it is not clear what this function should depend on (input biomass, cell concentration, total DNA?) and it is not clear what bias components should be included in determining the saturation, and these are not simply $\bar B$ since that includes 16S copy-number which isn't relevant here.
Rather than try to accurately model the relationship between $K$ and $A$ in this more complicated scenario, I will use the models $K_j = f(A_{Tj})$ and/or $K_j = f(A_{Tj}\bar B^{(K)}_j)$ to get some intuition for the effect that a strong saturation effect has on our inferences when it is not accounted for.

Saturation during extraction also implications for targeted PCR measurements but not for measurements that work directly on cells; and for DNA spike-ins, but not cellular spike-ins.

### Taxonomic resolution of measurements

I suppose that the "taxa" we consider are well-defined in a special sense, such that they can be equated both across samples within a measurement type and across different measurement types.
In other words, I assume that there is a real group of organisms corresponding to "taxon $i$" with abundance $A_{i\cdot}$ that the various "taxon $i$" measurements are really measuring, such that we can, for example, directly equate the taxonomic source of the counts $M_{i\cdot}$ and the targeted measurement $T_{i\cdot}$.
Achieving this situation in practice requires careful attention to things like primer design and how the metagenomic taxonomic assignment is done, and even then it may not always be possible.

One important special case where this assumption breaks down is when measurements aggregate one or more of the *atomic taxa* (corresponding to the rows in $A$), possibly in different ways by different measurement types.
This raises at least two sorts of complications.
First, constant multplicative bias at the level of atomic taxa does not translate into constant bias at the level of aggregate taxa [@McLaren2019].
Second, measurements may target different aggregates.
For example, if our metagenomics protocol provides ASV or species-level estimates but our PCR-based targeted measurements may quantify a genus or family.
Often through sufficiently careful bioinformatics we can at least nest the metagenomics taxa within the targeted taxon, but complications remain if efficiency varies within the aggregates.

Maybe:

* Give example of effect of aggregation when efficiencies vary among the aggregated taxa; understand how in terms of the average efficiency within the aggregate.
* Write out algebraically some of the cases I'll consider later.

### 2021-02-17 _common.R

@@ -1,0 +1,19 @@
+set.seed(42)
+
+knitr::opts_chunk$set(
+  comment = "#>",
+  collapse = TRUE,
+  cache = FALSE,
+  echo = FALSE,
+  out.width = "70%",
+  fig.align = 'center',
+  fig.width = 6,
+  fig.asp = 0.618,
+  fig.show = "hold"
+)
+
+options(dplyr.print_min = 6, dplyr.print_max = 6)
+options(crayon.enabled = FALSE)
+
+# library(tidyverse)
+library(here)

### 2021-02-19 proofs.Rmd
Add proofs of regression results

# (APPENDIX) Appendix {-}

# Proofs of regression results

```{theorem, label = "regression-linearity", name = "Linearity of regression coefficients", echo = TRUE}
Consider a scalar response variable with $J$ observations, which I represent by the vector $y$, that equals a sum of $K$ component response variables $y^{(k)}$ each scaled by a non-zero factor $c^{(k)}$, 
\begin{align}
  (\#eq:y-sum)
  y = \sum_{k = 1}^K c^{(k)} y^{(k)}.
\end{align}
(We may need to assume that the $y^{(k)}$ are linearly independent.)
Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
Let $\hat \gamma$ denote the least-squares estimate of the coefficient matrix $\gamma$ in the linear regression
\begin{align}
  y = X \gamma + \epsilon.
\end{align}
Similarly, let $\hat \gamma^{(k)}$ denote the least-squares estimates for the $K$ linear regressions
\begin{align}
  y^{(k)} = X \gamma^{(k)} + \epsilon^{(k)}.
\end{align}
The least-squares coefficient estimates for $y$ are given by the sum of those of the $y^{(k)}$,
\begin{align}
  \hat \gamma = \sum_{k=1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}
```

```{proof, echo = TRUE}
The Moore-Penrose pseudoinverse of $X$ is $X^+ = (X^T X)^{-1} X^T$.
The least squares estimates are given by multiplying the matrix $X^+$ by the corresponding response vector ([Wikipedia](https://en.wikipedia.org/wiki/Ordinary_least_squares#Matrix/vector_formulation)), so that $\hat \gamma = X^+ y$ and $\hat \gamma^{(k)} = X^+ y^{(k)}$ (for $k = 1, \dots, K$).
The result then follows from the stipulation \@ref(eq:y-sum) and the linearity of matrix multiplication,
\begin{align}
  \hat \gamma 
  = X^+ y 
  = X^+ \sum_{k = 1}^K c^{(k)} y^{(k)}
  = \sum_{k = 1}^K c^{(k)} X^+ y^{(k)}
  = \sum_{k = 1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}
```

We can use \@ref(thm:regression-linearity) to describe how measurement error, such as that caused by experimental bias, affects least-squares estimates of regression coefficients.

```{theorem, label = "regression-error", echo = TRUE, name = "Error in regression coefficients"}
Suppose that $y$ is some response variable in nature, $z$ is our imperfect measurement of $y$, and $d = z - y$ the difference between the two, so that $z =  y + d$.
Let each be a vector of length $J$ describing a set of $J$ observations.
Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
Consider the linear regression equations for $y$, $d$, and $z$,
\begin{align}
  y &= X \gamma^{(y)} + \epsilon^{(y)} \\
  z &= X \gamma^{(z)} + \epsilon^{(z)} \\
  d &= X \gamma^{(d)} + \epsilon^{(d)}.
\end{align}
The relationship between the least-squares estimates mirrors that between the variables themselves,
\begin{align}
  \hat \gamma^{(z)} = \hat \gamma^{(y)} + \hat \gamma^{(d)}.
\end{align}
```

```{proof, echo = TRUE}
The result follows directly from $z =  y + d$ and \@ref(thm:regression-linearity).
```

```{corollary, label = "regression-error-simple", echo = TRUE}
For the special case of simple linear regression,
\begin{align}
  y &= \gamma_0^{(y)} + \gamma_1^{(y)} x + \epsilon^{(y)} \\
  z &= \gamma_0^{(z)} + \gamma_1^{(z)} x + \epsilon^{(z)} \\
  d &= \gamma_0^{(d)} + \gamma_1^{(d)} x + \epsilon^{(d)},
\end{align}
the estimates for the intercept and slope coefficients satisfy
\begin{align}
  \hat \gamma_0^{(z)} &= \hat \gamma_0^{(y)} + \hat \gamma_0^{(d)} \\
  \hat \gamma_1^{(z)} &= \hat \gamma_1^{(y)} + \hat \gamma_1^{(d)}.
\end{align}
```

Next, aiming to show the forms for the different transformations of interest.
It might be useful to cover the case where $d$ is split into a constant and a sample-specific part, since probably everything of interest will follow from that.

What, if any general results should go in the main text?

Also relate the expected values of $y$ and $z$ conditional on $X$?.

To figure out what I need:

1. extend the below result for log proportions to multiple regression
2. work out the derivation in terms of \@ref(thm:regression-error)
3. Write a corollary if it seems to be useful for doing all the cases (e.g. logit etc)

### Linear regression of microbiome abundances

First, consider log proportions.
In this case, the intercept coefficient is biased upward by $\log B_i$ and the other coefficients are biased downward by the corresponding coefficient of $\log \bar B$, based on the error formulas for individual samples.

TODO: write the general result (for multiple regression, assuming that an intercept term is present.

```{proposition, echo = TRUE}
Consider the linear equations
\begin{align}
  (\#eq:least-squares-formulation)
  \log a_i    &= \alpha_0 + \alpha_1 x \\
  \log \bar B &= \beta_0 + \beta_1 x   \\
  \log m_i    &= \gamma_0 + \gamma_1 x,
\end{align}
and let $\hat \alpha_0$, etc. denote the least-squares estimates for these equations supposing perfect information (i.e., that the $a_i$ and $B_i$ are known).
The coefficients for $\log m_i$ are related to those for $\log a_i$ and $\log \bar B$ through the equations
\begin{align}
  (\#eq:least-squares-estimates)
  \hat \gamma_0 &= \hat \alpha_0 + \log B_i - \hat \beta_0 \\
  \hat \gamma_1 &= \hat \alpha_1 - \hat \beta_1,
\end{align}
so that the fitted model for $\log m_i$ is given by the right-hand side of
\@ref(eq:expected-observed-abundance-log-linear-regression) with the coefficients replaced by the fitted values.
```

```{proof, echo = TRUE}
```

### 2021-02-26 absolute-abundance.Rmd
Draft initial absolute-abundance results
Covers single-sample and fold-change estimates for bulk-abundance and
reference-taxa methods.

# Differential absolute abundance {#absolute}

This section considers various ways to estimate absolute abundances of individual taxa for the purposes of estimating log fold changes (LFCs) in absolute abundance, or more generically of performing linear regression on log absolute abundance.

## Using bulk abundance measurements

The absolute abundance of taxon $i$ is estimated by multiplying its metagenomics proportion by the bulk abundance measurement,
\begin{align}
  (\#eq:A-hat-bulk)
  \hat A_{ij} &= m_{ij} K_j.
\end{align}
The error in this estimate due to the experimental bias in $M$ and $K$ is given by substituting equations \@ref(eq:m) and \@ref(eq:K) into \@ref(eq:A-hat-bulk), giving
\begin{align}
  (\#eq:error-bulk)
  \hat A_{ij} &= \frac{a_{ij} B^{(M)}_i}{\bar B^{(M)}_j} \cdot A_{Tj} \bar B^{(K)}_j
  \\&= A_{ij} \cdot \frac{B^{(M)}_i \bar B^{(K)}_j}{\bar B^{(M)}_j}.
\end{align}
The fold error in the estimated abundance of a particular taxon in a particular sample thus equals the metagenomics efficiency of that taxon times the ratio of the sample mean efficiencies of the bulk measurement to the metagenomics measurement.
Next I consider the error in the estimated fold change in $A_i$ between two samples $j$ and $j'$.
As in the case of log proportion estimates, the taxon-specific, sample-independent error term ($B^{(M)}_i$) cancels, but the taxon-independent, sample-specific error term (${\bar B^{(K)}_j}/{\bar B^{(M)}_j}$) does not, giving
\begin{align}
  (\#eq:fc-bulk)
  \frac{\hat A_{ij'}}{\hat A_{ij}}
  &= \frac{A_{ij'}}{A_{ij}} 
  \cdot \frac{\bar B^{(M)}_j \bar B^{(K)}_{j'}}{\bar B^{(M)}_{j'} \bar B^{(K)}_j}.
  \\&= \frac{A_{ij'}}{A_{ij}} 
  \cdot \frac{\bar B^{(K)}_{j'} / \bar B^{(M)}_{j'}}{\bar B^{(K)}_j / \bar B^{(M)}_j }.
\end{align}
In words, the fold error in the estimated fold change in the absolute abundance of any taxon equals the fold change in the ratio of the mean efficiency of the bulk abundance measurement to that of the metagenomics measurement.

TODO: Add regression result.

### Shared bias components

Motivation: Consider an experiment in which the community has been profiled by 16S sequencing and bulk abundance has been estimated by 16S qPCR.
The taxon-specific bias associated with each of these measurements may be highly similar; both are measuring 16S copies in the same aliquot of already-extracted DNA, which has already been affected by DNA extraction bias and 16S copy-number variation.
If the same primers are used then PCR bias might also be shared.
How does having a large amount of bias being shared between the bulk and sequencing measurement affect the error in abundance estimates in individual samples and the fold changes between samples?

In this case, the ratio of mean efficiencies, $\bar B^{(K)}_j / \bar B^{(M)}_j$, will vary much less across samples than either mean efficiency itself, such that the error term in  \@ref(eq:fc-bulk) will be small.
Therefore the estimated fold changes may be accurate despite potentially large variation in the metagenomics sample mean efficiency across samples.
Note that the estimated absolute abundance in the individual sample, \@ref(eq:error-bulk), remains inaccurate since the unknown taxon-specific efficiency remains.

TODO: Add derivation, where we formally define $B^{(M)}$ and $B^{(K)}$ in terms of shared and non-shared components, and show that the variation in the ratio of mean efficiencies reduces to that of the non-shared components.

## Using reference taxa

TODO: Consider rewriting to be in terms of the ratio of counts * the reference abundance, to emphasize that this is a ratio-based method.

### Taxa with targeted abundance measurements

First consider the case of just a single reference taxon, $r$, for which we have  targeted measurements.
We can estimate the absolute abundances of all taxa by scaling the read counts by the ratio of the abundance measurement to read count for the reference taxon,
\begin{align}
  (\#eq:A-hat-targeted)
  \hat A_{ij} &= M_{ij} \cdot \frac{T_{rj}}{M_{rj}}.
\end{align}
Accounting for systematic error in both the metagenomcs measurement \@ref(eq:M) and the targeted measurement \@ref(eq:T) gives
\begin{align}
  (\#eq:error-targeted)
  \hat A_{ij} &= A_{ij} \cdot \frac{B^{(M)}_i B^{(T)}_r}{B^{(M)}_r}.
\end{align}
Since the error in (\#eq:error-targeted) is sample-independent, it cancels out when we use this equation to compute the fold change between two samples,
\begin{align}
  (\#eq:fc-targeted)
  \frac{\hat A_{ij'}}{\hat A_{ij}} &=\frac{\hat A_{ij'}}{\hat A_{ij}}.
\end{align}
Thus our model predicts that normalizing read counts to a reference taxon, while providing systematically-distorted abundance estimates, provides accurate fold-change estimates.

### Cellular spike-ins

Spike-ins are a way to create a reference taxon whose abundance is known by design and so doesn't need to be measured.
Typically the spike-in is added in a fixed concentration to all samples; however, this is unnecessary, so long as we know the amount it was added.
In either case we can use \@ref(eq:A-hat-targeted) to estimate the abundances of all taxa from the spike-in taxa, substituting the spike-in abundance $S_{ir}$ for $T_{ir}$,
Our model predicts that normalizing read counts to a spike-in taxon gives systematically-distorted abundance estimates but accurate fold-change estimates for the same reason as in the targeted case.

### DNA spike-ins

### Taxa assumed to have a constant abundance

By this approach, we identify one or more taxa that we assume to have a constant absolute abundance across samples.
Inference proceeds as in the targeted and spike-in case, using \@ref(eq:A-hat-targeted) but with the reference abundance $T_r$ set to an arbitrary constant as we may not know the true abundance.
In this case, the effect of bias cancels in fold-change calculations for the same reasons as the targeted and spike-in cases.
Note: The reference taxon could be the host.

### Multiple reference taxa

Under the deterministic model we consider here, having multiple reference taxa in the same sample is completely redundant - we learn no new information.
In reality, having information from multiple taxa from a single sample should allow us to make more precise estimates, provided that we use them in a way so that the effects of bias do not offset any reductions in noise.
How might we leverage measurements of multiple reference taxa, $r \in R$, in a way that remains robust to bias?

First consider the situation where we can assume that all reference taxa are in each sample, as would typically be the case for a spike-in.
One possibility is to sum the metagenomic and targeted abundances from the different taxa,
\begin{align}
  (\#eq:A-hat-targeted-sum)
  \hat A_{ij} &= M_{ij} \cdot \frac{\sum_{r \in R} T_{rj}}{\sum_{r \in R} M_{rj}}.
\end{align}
To see this, let $A_{Rj} = \sum_{r \in R} A_{rj}$ be the summed actual abundances of the reference taxa. 
Similarly define $M_{Rj} = \sum_{r \in R} M_{rj}$ and $T_{Rj} = \sum_{r \in R} T_{rj}$ be the summed read counts and targeted measurements. 
Now
\begin{align}
  \hat A_{ij} &= M_{ij} \cdot \frac{T_{Rj}}{M_{Rj}}.
\end{align}
Also let $\bar B^{(M)}_{Rj} = \sum_{r \in R} B^{(M)}_{rj} A_{rj} / A_{Rj}$ be the mean metagenomics efficiency among the reference taxa and similarly define $\bar B^{(T)}_{Rj}$ as the mean targeted efficiency.
A little algebra shows that
\begin{align}
  (\#eq:T-Rj)
  T_{Rj} 
    &= A_{Rj} \cdot \sum_{r \in R} \frac{A_{rj}}{A_{Rj}} \bar B^{(T)}_{Rj}
  \\&= A_{Rj} \bar B^{(T)}_{Rj}
\end{align}
while
\begin{align}
  (\#eq:M-Rj)
  M_{Rj} &= A_{Rj} \bar B^{(M)}_{Rj} C_j.
\end{align}
And since $M_{ij} = A_{ij} B^{(M)}_{i} C_j$ \@ref(eq:M) we can write the estimated \@ref(eq:A-hat-targeted-sum) abundance of taxon $i$ as
\begin{align}
  (\#eq:error-targeted-sum)
  \hat A_{ij} &= A_{ij} \cdot \frac{B^{(M)}_i \bar B^{(T)}_{Rj}}{\bar B^{(M)}_{Rj}}.
\end{align}
The error now depends on relative abundances among the reference taxa through their mean bias values.
Therefore, fold changes estimated using this approach are not necessarily robust to bias if the relative abundances among the reference taxa vary.
Spike-in experiments are typically designed so that the reference taxa typically have the same relative abundances with respect to each other in all samples, in which case this method would remain robust to bias.

Another approach is to multiply by the geometric mean abundances of the taxa,
\begin{align}
  (\#eq:A-hat-targeted-prod)
  \hat A_{ij} &= M_{ij} \cdot \left[\frac{\prod_{r \in R} T_{rj}}{\prod_{r \in R} M_{rj}}\right]^{1/|R|}.
\end{align}
The error in this case is
\begin{align}
  (\#eq:error-targeted-prod)
  \hat A_{ij} 
  &= A_{ij} B^{(M)}_r C_j \cdot 
    \left[\frac{\prod_{r \in R} A_{rj} B^{(T)}_r} {\prod_{r \in R} A_{rj} B^{(M)}_r C_j} \right]^{1/|R|} \\
  &= A_{ij} \cdot 
    \left[\frac{\prod_{r \in R} B^{(T)}_r} {\prod_{r \in R} B^{(M)}_r} \right]^{1/|R|}.
\end{align}
The multiplicative error is now constant and will cancel in fold-change calculations even if the reference taxa vary in their relative abundances.

Thus the geometric-mean method appears to be robust to bias, but as written it requires that all reference taxa are present in all samples, whereas the summation approach is sensitive to bias but only requires at least one reference taxon to be present in the sample.

Now consider the case where different reference taxa are present (above our detection limit) in each sample, as might be the case for naturally occuring reference taxa.
We can modify the geometric-mean approach to work (so long as at least one reference is present in each sample) as follows.
From the samples where multiple references are present, we are able to measure the differnetial bias $B^{(M/T)}_r / B^{(M/T)}_{r'}$ among the references $r, r'$.
From measurement of reference $r'$, we know the value of $T_{r'j} / M_{r'j}$, which combined with the differential bias tells us what the ratio $T_{rj} / M_{rj}$ should equal, namely
\begin{align}
  \frac{T_{rj}}{M_{rj}} = \frac{T_{r'j}}{M_{r'j}} \cdot \frac{B^{(M/T)}_r}{B^{(M/T)}_{r'}}.
\end{align}
Therefore, in samples where $r$ is missing but $r'$ is present, we can replace 
$\frac{T_{rj}}{M_{rj}}$ in the product with its predicted value.
(More generally, we can use the geometric mean of the values predicted by each of the reference taxa that are present).

This imputation procedure is a simple plug-in approach to show that it is possible to use a multiplicative approach to using multiple references even when the references vary in their presence among samples. 
Real inference should ideally be done with a fully generative statistical model that includes bias as well as noise-generating processes and can be used for maximum likelihood or Bayesian inference. In such models, the imputation might be implicitly handled as part of the likelihood, and more generally the values $\frac{T_{rj}}{M_{rj}}$ for various $r$ can be naturally and automatically weighted (in terms of their influence on $\hat A_{ij}$) according to their predicted precision by the model fitting procedure.

<!-- TODO: Move the definitions and relevant intermediate results to models.Rmd and  trim the above derivations, including for the aggregates and the differential bias among multiple reference taxa. -->

## Using an equivolumetric protocol

## Computational methods

### 2021-02-26 models.Rmd
Draft initial absolute-abundance results
Covers single-sample and fold-change estimates for bulk-abundance and
reference-taxa methods.

@@
 
-Deterministic measurement error models that extends the @McLaren2019 model of metagenomics measurement to 1) describe absolute as well as relative abundance, 2) include spike-in taxa, and 3) include supplementary measurements of bulk and targeted (absolute) abundance.
+Deterministic measurement error models that extends the @mclaren2019cons model of metagenomics measurement to 1) describe absolute as well as relative abundance, 2) include spike-in taxa, and 3) include supplementary measurements of bulk and targeted (absolute) abundance.
 
@@
 $A_{ij}$ is the actual absolute abundance of taxon $i$ in sample $j$, in the target units.
+For variables that hold matrixes or vectors, capital letters denote absolute abundances and lower case letters denote proportions.
+Thus $a_{ij}$ is the matrix of actual relative abundances in terms of cells as a proportion of the total amount of cells in the sample; $a_{ij} = A_{ij} / A_{Tj}$.
+Let $A_{Tj}$ (or $A_{\cdot j}$??) be the total abundance in sample $j$, $A_{Tj} = \sum_i A_{ij}$.
+For a subset of taxa $Q \subset \{1, \dots, I\}$, define $A_{Qj} \equiv \sum_{i \in Q} A_{ij}$ to be the total abundance of the taxa in $Q$.
+
 Unless specified otherwise, $A$ always refers to abundance in the original sample material, not in extracted DNA.
@@
 
-Let $A_{Tj}$ (or $A_{\cdot j}$??) be the total abundance in sample $j$, $A_{Tj} = \sum_i A_{ij}$.
-
-For variables that hold matrixes or vectors, capital letters denote absolute abundances and lower case letters denote proportions.
-Thus $a_{ij}$ is the matrix of actual relative abundances in terms of cells as a proportion of the total amount of cells in the sample; $a_{ij} = A_{ij} / A_{Tj}$.
+To consider various addition information that can be used for absolute abundance inference, I let $B$ denote bulk abundance measurements (such as via flow cytometry or broad-range 16S qPCR), $S$ genome the known taxon abundances of a spike-in, and $T$ denote the abundances of a set of taxa made via targeted measurement.
+I let $R \subset \{1, \dots, I\}$ denote the set of _reference taxa_ whose abundance is known independently from the metagenomics measurement (up to experimental bias factors) because they were spiked at a known density, subjected to a targeted measurement, or have been determined through some other means to have a constant or known abundance across samples.
 
@@
 I generally assume these to be sample-independent.
-An overbar denotes the average value over taxa weighted by actual abundance; 
-i.e., $\bar B_j = I^{-1} \sum_i B_i a_{ij}$.
+An overbar denotes the average value over taxa weighted by actual abundance; i.e., $\bar B_j = I^{-1} \sum_i B_i a_{ij}$.
+An overbar and a subcript set of taxa indicates the mean bias within that subset: $\bar B_{Qj} = \sum_{i\in Q} B_i a_{ij} / \sum_{i\in Q} a_{ij}$.
 
@@
 Bul**k** abundance      [cells], [DNA], or [gene copies]   $K$ ($I\times 1$ column vector)  $B^{(K)}$
-**T**argeted abundance  [cells], [DNA], or [gene copies]   $T$ ($|I_T| \times J$)           $B^{(T)}$
-**S**pike-in abundance  [cells], [DNA], or [gene copies]   $S$ ($|I_S| \times J$)           $B^{(S)}$
+**T**argeted abundance  [cells], [DNA], or [gene copies]   $T$ ($I \times J$)               $B^{(T)}$
+**S**pike-in abundance  [cells], [DNA], or [gene copies]   $S$ ($I \times J$)               $B^{(S)}$
 
-Let $I_T$ and $I_S$ be the sets of taxa with targeted abundance measurements or known spike-in abundances, respectively.
-Question: Will it be more convenient for $S$ and $T$ to have the standard shape $I\times J$?
+Note: $S_{ij}$ and $T_{ij}$ are only defined for the reference taxa $i \in R$.
 
@@
 
-$M$ is a X BY X matrix representing the (marker-gene or) metagenomics measurement; $M_{ij}$ is the sequencing count associated with taxon $i$ in sample $j$, such that $M_{ij} = 10$ indicates that 10 fragments of DNA were sequenced and assigned to taxon $i$ and sample $j$.
+$M$ is an $I\times J$ matrix representing the (marker-gene or) metagenomics measurement; $M_{ij}$ is the sequencing count associated with taxon $i$ in sample $j$, such that $M_{ij} = 10$ indicates that 10 fragments of DNA were sequenced and assigned to taxon $i$ and sample $j$.
 In the case of paired-end sequencing, a single fragment of DNA yields two reads, which may or may not be counted jointly or independently by different algorithms; here I assume that read-pairs are assigned jointly, and will use "read" and "fragment" interchangeably.
@@
 
-Following the notation defined above, I let $M_{Tj} = \sum_i M_{ij}$ is the total count for sample $j$; I will sometimes refer to this number as the *sequencing depth* or *read depth* of the sample, though in reality the total count will be less than the sample's sequencing depth since often a significant fraction of reads are lost by filtering steps prior to producing the final count matrix.
+Following the notation defined above, I let $M_{Tj} = \sum_i M_{ij}$ be the total count for sample $j$; I will sometimes refer to this number as the *sequencing depth* or *read depth* of the sample, though in reality the total count will be less than the sample's sequencing depth since often a significant fraction of reads are lost by filtering steps prior to producing the final count matrix.
 I use $m_{ij} = M_{ij} / M_{Tj}$ to denote the proportion of reads assigned to taxon $i$ in sample $j$.
 
-I assume that the composition (relative abundances) is given by the deterministic error model presented in @McLaren2019.
+I assume that the composition (relative abundances) is given by the deterministic error model presented in @mclaren2019cons.
 In this model, the bias vector $B^{(M)}$ represents the *relative* efficiencies with which various taxa are measured, and so has the same effect on measurement if we rescale it $B \to cB$ for any $c > 0$.
@@
   (\#eq:M)
-  M_{ij} = A_{ij} B^{(M)}_{i} C_j, && C_j = \frac{M_{Tj}}{\sum_i A_{ij} B^{(M)}_{i}}.
+  M_{ij} = A_{ij} B^{(M)}_{i} C_j, && 
+  C_j = \frac{M_{Tj}}{\sum_i A_{ij} B^{(M)}_{i}}
+      = \frac{M_{Tj}}{A_{Tj} \bar B^{(M)}_j}.
 \end{align}
@@
 
-## Spike-ins
-
-Let the set $I_S$ denote the indexes of the spiked taxa, which I assume are also measured by the metagenomics experiment and so also correspond to rows in $M$.
-The $|I_S|\times J$ matrix $S$ describes the nominal abundances with which spike-in taxa were added to each sample.
-Spike-in experiments are often designed so that a taxon $i$ is added in the same concentration to all samples, which we can represent by supposing the columns of $S$ to be identical.
-
-The bias $B^{(S)}$ should be interpreted as error in the quantification of how much spike-in was added, such that
-\begin{align}
-  S_{ij} = A_{ij} B^{(S)}_i.
-\end{align}
-I assume that this error is constant across samples.
-The motivation for this error model is as follows: Suppose that we misestimated the proportion of spike-in taxon $s\in I_S$ in our original spike-in stock, by a factor $B^{(S)}_s$; now its true abundance will be off by that factor in any spike-in derived from the starting stock.
-Unlike the metagenomics efficiencies, these error factors are absolute numbers.
-
-Taxa can be added as cells (pre-extraction) or as DNA (post-extraction).
-In an application where DNA spike-ins are used to make inferences about changes in absolute cell abundances, we can include the difference between the nominal abundances $S$ (the spiked DNA concentration) and the actual abundances $A$ that are expected due to factors such as DNA extraction and copy-number variation in the bias term $B^{(S)}$, so long as DNA extraction is linear (see section on saturation effects below).
-
 ## Bulk absolute abundance measurement
@@
 \begin{align}
+  (\#eq:K)
   K_j = \sum_i A_{ij} B^{(K)}_i = A_{Tj} \bar B^{(K)}_j,
 \end{align}
-where $bar B^{(K)}_j$ is the mean bulk-measurement efficiency in the sample.
+where $\bar B^{(K)}_j$ is the mean bulk-measurement efficiency in the sample.
 
@@
 \begin{align}
+  (\#eq:T)
   T_{ij} = A_{ij} B^{(T)}_i,
@@
 where $B^{(T)}_i$ is the sample-independent, taxon-specific efficiency of the targeted measurement for taxon $i$.
+
+In discussing the case of multiple reference taxa, it will be useful to refer to the total targeted and metagenomics abundance and the average targeted and metagenomics efficiencies of the reference taxa.
+I do so with the notation $A_{Rj}$ and $M_{Rj}$ (total abundances) and $\bar B^{(T)}_{Rj}$ and $\bar B^{(M)}_{Rj}$ (mean efficiencies) as defined above.
+Note that
+\begin{align}
+  (\#eq:T-Rj)
+  T_{Rj} 
+    &= A_{Rj} \cdot \sum_{r \in R} \frac{A_{rj}}{A_{Rj}} \bar B^{(T)}_{Rj}
+  \\&= A_{Rj} \bar B^{(T)}_{Rj}.
+\end{align}
+
+## Spike-ins
+
+The matrix $S$ describes the nominal abundances with which spike-in taxa were added to each sample.
+Spike-in experiments are often designed so that a taxon $i$ is added in the same concentration to all samples, which we can represent by supposing the defined columns of $S$ to be identical.
+The bias $B^{(S)}$ should be interpreted as error in the quantification of how much spike-in was added, such that
+\begin{align}
+  (\#eq:S)
+  S_{ij} = A_{ij} B^{(S)}_i.
+\end{align}
+I assume that this error is constant across samples.
+The motivation for this error model is as follows: Suppose that we misestimated the proportion of spike-in taxon $s\in I_S$ in our original spike-in stock, by a factor $B^{(S)}_s$; now its true abundance will be off by that factor in any spike-in derived from the starting stock.
+Unlike the metagenomics efficiencies, these error factors are absolute numbers.
+
+Taxa can be added as cells (pre-extraction) or as DNA (post-extraction).
+In an application where DNA spike-ins are used to make inferences about changes in absolute cell abundances, we can include the difference between the nominal abundances $S$ (the spiked DNA concentration) and the actual abundances $A$ that are expected due to factors such as DNA extraction and copy-number variation in the bias term $B^{(S)}$, so long as DNA extraction is linear (see section on saturation effects below).
+
+(If we want to allow for different targeted and spike-in taxa, we can use $R_S$ and $R_T$.)
 
@@
 Here I list some relevant complications that we will return to in our analysis.
-For more considerations, see the Discussion of @McLaren2019.
+For more considerations, see the Discussion of @mclaren2019cons.
 
@@
 This can be useful when we are considering qPCR and also DNA spike-ins.
+
+Note that the "actual" units could be chosen by the researcher to be something other than cell density; e.g. biomass, or even 16S density.
+But for concreteness I'll take to be cell density, as this seems to be what is most typically meant.
 
@@
 This raises at least two sorts of complications.
-First, constant multplicative bias at the level of atomic taxa does not translate into constant bias at the level of aggregate taxa [@McLaren2019].
+First, constant multplicative bias at the level of atomic taxa does not translate into constant bias at the level of aggregate taxa [@mclaren2019cons].
 Second, measurements may target different aggregates.

### 2021-02-26 proofs.Rmd
Draft initial absolute-abundance results
Covers single-sample and fold-change estimates for bulk-abundance and
reference-taxa methods.

@@
 # Proofs of regression results
+
+## General regression
+
+Regression analysis can often be framed as seeking the regression function $r(x)$ that describes how the expected value of some response variable $Y$ varies with the a vector of covariates $X$ (@wasserman2004allo),
+\begin{align}
+  r(x) = E [Y \mid X = x].
+\end{align}
+For example, the response might be log absolute abundance of a particular taxon.
+In our case, however, we don't know $Y$, but rather a measure that is subject to random and systematic error, which I call $Z$.
+Let $D = Z - Y$ be the difference between the true response $Y$ and its measurement $Z$, so that $Z = Y + D$.
+It follows that
+\begin{align}
+  (\#eq:regression-error)
+  E [Z \mid X] = E [Y \mid X] + E [D \mid X].
+\end{align}
+
+## Linear least-squares regression
 
@@
 ```{theorem, label = "regression-error", echo = TRUE, name = "Error in regression coefficients"}
-Suppose that $y$ is some response variable in nature, $z$ is our imperfect measurement of $y$, and $d = z - y$ the difference between the two, so that $z =  y + d$.
+Suppose that $y$ is the response variable we wish to understand, $z$ is our imperfect measurement of $y$, and $d = z - y$ the difference between the two, so that $z = y + d$.
 Let each be a vector of length $J$ describing a set of $J$ observations.
@@
 \begin{align}
+  (\#eq:regression-coefficient-error)
   \hat \gamma^{(z)} = \hat \gamma^{(y)} + \hat \gamma^{(d)}.

### 2021-04-03 notebook/_posts/2020-11-12-leopold2020host-bias-sensitivity/index.Rmd
Import analyses of leopold2020host experiment
These two analyses of the leopold2020host experiment are imported as
patches from MRM's personal research notebook. The first (earlier)
analysis, 'leopold2020host-bias-sensitivity', conducts an initial
exploration of the sample mean efficiency. The second analysis,
'illustration-of-a-spurious-differential-abundance-result-due-to-bias',
compares the results of regression on calibrated and uncalibrated data
as a case study of the impact of bias on differential abundance.  This
second analysis is the source of the file
'figures/leopold2020host-example

---
title: >
  Exploring the effect of bias in Leopold and Busby (2020)
author: Michael McLaren
date: '2019-10-21'
categories:
  - bias sensitivity
  - R
  - ref:leopold2020host
description:
bibliography: _references.bib
draft: false
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    dev: svg
---

```{r, include = FALSE}
# knitr chunk options
knitr::opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

Analysis of bias in @leopold2020host.

# Setup

```{r setup}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot())

# stats helpers
library(broom)

library(metacal); packageVersion("metacal")
```

```{r load-data, include = FALSE}
## From the metacal 2.0 tutorial
# Download data from https://zenodo.org/record/3872145
data_path <- here::here("_data", "leopold2020host")
# To use a temporary directory:
# data_path <- file.path(tempdir(), "leopold2020")
if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  download.file(
    "https://zenodo.org/record/3872145/files/dleopold/Populus_priorityEffects-v1.2.zip",
    file.path(data_path, "Populus_priorityEffects-v1.2.zip")
  )
  unzip(
    file.path(data_path, "Populus_priorityEffects-v1.2.zip"), 
    exdir = data_path
  )
}
#> The microbiome data is stored in a phyloseq object,
ps <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/output/compiled/phy.rds") %>%
  readRDS

sample_data(ps) <- sample_data(ps) %>%
  transform(
    Timepoint = factor(Timepoint)
  )

mock_actual <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/data/MockCommunities.csv") %>%
  read.csv(row.names = 1) %>%
  select(-Sym4) %>%
  as("matrix") %>%
  otu_table(taxa_are_rows = FALSE) %>%
  transform_sample_counts(function(x) close_elts(1 / x))
mock_taxa <- taxa_names(mock_actual)
sam <- sample_data(ps) %>% as("data.frame") %>% as_tibble(rownames = "Sample")
tax <- tax_table(ps) %>% as("matrix") %>% as_tibble(rownames = "Taxon")
ps.mock <- ps %>% 
  subset_samples(Samp_type == "Mock") %>%
  prune_taxa(mock_taxa, .)
```

```{r estimate-bias}
ps.mock.pseudo <- ps.mock %>%
  transform_sample_counts(function(x) x + 1)
mc_fit <- estimate_bias(ps.mock.pseudo, mock_actual, boot = TRUE) %>% print
bias <- coef(mc_fit) %>% print
```

```{r}
mc_fit.summary <- summary(mc_fit)
print(mc_fit.summary)
```

```{r, fig.dim = c(5,3)}
coef_tb <- mc_fit.summary$coefficients
coef_tb %>%
  mutate(taxon = fct_reorder(taxon, estimate)) %>%
  ggplot(aes(taxon, estimate, 
      ymin = estimate / gm_se^2, ymax = estimate * gm_se^2)) +
  geom_hline(yintercept = 1, color = "grey") +
  geom_pointrange() +
  scale_y_log10() +
  coord_flip()
```

```{r calibrate}
ps.pseudo <- transform_sample_counts(ps, function(x) x + 1)
ps.pseudo.cal <- calibrate(ps.pseudo, bias) %>% print
```

# Sample mean efficiency across samples

To estimate the SMRE with formula `sum_i A_i E_i`, use the calibrated abundances.

```{r}
ps0 <- ps.pseudo.cal %>%
  prune_taxa(names(bias), .)
```

```{r}
otu <- ps0 %>% otu_table %>% orient_taxa(as = "cols") %>%
  transform_sample_counts(close_elts)
sam0 <- sample_data(ps0) %>% 
  as("data.frame") %>% 
  as_tibble(rownames = "Sample")
```

```{r}
sme <- otu %>% psmelt %>%
  left_join(bias %>% enframe("OTU", "Efficiency"), by = "OTU") %>%
  group_by(Sample) %>%
  summarize(SMRE = sum(Efficiency * Abundance)) %>%
  left_join(sam0, by = "Sample") %>%
  mutate(
    across(Timepoint, as.factor),
    across(Timepoint, fct_explicit_na, "Other"),
  )
```

```{r}
range(sme$SMRE)
range(bias)
```

```{r}
p1 <- ggplot(sme, aes(SMRE, fill = Samp_type)) +
  geom_histogram() +
  annotate("point", y = 200, x = bias, color = "black", shape = "|", size = 7) +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  expand_limits(x = range(bias)) +
  scale_fill_brewer(type = "qual") +
  labs(
    # x = "Sample mean relative efficiency (SMRE)",
    x = "Mean efficiency", 
    y = "Number of samples",
    fill = "Sample type"
  )
p1 +
  plot_annotation(
    title = "Distribution of sample mean relative efficiencies",
  )
```

```{r, eval = F}
ggsave("/tmp/leopold2020host-smre-distribution.svg", units = "in",
  width = 6, height = 3, scale = 1.5)
ggsave("/tmp/leopold2020host-smre-distribution.png", units = "in",
  width = 6, height = 3, scale = 1.5)
```

In this case, the bimodal distribution is likely driven by the high proportion
of Melampsora (pathogen) in ~half the samples.

TODO: check this hypothesis. by making a histogram of melampspora %, and a scatterplot of the melamspora vs. SMRE. Can do with facet matrix from ggforce

```{r}
mel <- otu %>%
  prune_taxa("Melampsora", .) %>%
  psmelt %>%
  select(-OTU, Melampsora = Abundance)
sme0 <- left_join(sme, mel, by = "Sample")
```

```{r}
# p2 <- ggplot(sme0, aes(SMRE, fill = Melampsora)) +
p2 <- ggplot(sme0, aes(y = Melampsora, x = SMRE, color = Samp_type)) +
  geom_point() +
  # annotate("point", y = 1, x = bias, color = "black", shape = "|", size = 7) +
  scale_x_log10() +
  # scale_y_sqrt() +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 1e-2),
    breaks = c(0, 0.01, 0.1, 0.5, 1)
    ) +
  expand_limits(x = range(bias)) +
  scale_color_brewer(type = "qual") +
  labs(
    y = "Melampsora proportion",
    # x = "Sample mean relative efficiency (SMRE)",
    x = "Mean efficiency",
    color = "Sample type"
  )
p2
```

```{r}
(p1 + ggtitle("Distribution of the mean efficiency across samples")) / 
  (p2 + ggtitle("Melampsora proportion vs. mean efficiency")) +
  plot_annotation(tag_levels = "A")
```

```{r, eval = F}
ggsave("/tmp/leopold2020host-smre-distribution.svg", units = "in",
  width = 6, height = 6, scale = 1.5)
ggsave("/tmp/leopold2020host-smre-distribution.pdf", units = "in",
  width = 6, height = 6, scale = 1.5)
ggsave("/tmp/leopold2020host-smre-distribution.png", units = "in",
  width = 6, height = 6, scale = 1.5)
```

Note, the high-Mel samples are the timepoint 2 samples, and were not used for some (all?) of his analysis due to the dominance of Mel. Let's check this out,

```{r}
sme %>%
  mutate(
    across(Timepoint, as.factor),
    across(Timepoint, fct_explicit_na, "Other"),
  ) %>%
  # filter(Samp_type == "Experiment") %>%
  ggplot(aes(SMRE, fill = Timepoint)) +
  geom_histogram() +
  annotate("point", y = 200, x = bias, color = "black", shape = "|", size = 7) +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  expand_limits(x = range(bias)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  )
```

```{r}
sme %>%
  filter(Samp_type == "Experiment") %>%
  ggplot(aes(SMRE, fill = Treatment)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  ) +
  facet_wrap(~Timepoint, ncol = 1)
```

```{r}
sme %>%
  filter(Samp_type == "Experiment", Treatment != "Negative") %>%
  ggplot(aes(SMRE, fill = Region)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  ) +
  facet_wrap(~Timepoint, ncol = 1)
```

Can see that Eastern genotypes more susceptible (more Mel and higher SME)

Are some western genotypes less susceptible?

```{r}
sme %>%
  filter(Samp_type == "Experiment", Region == "West",
    Treatment != "Negative") %>%
  ggplot(aes(SMRE, fill = Genotype)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  ) +
  facet_wrap(~Timepoint, ncol = 1)
```

Eastern?

```{r}
sme %>%
  filter(Samp_type == "Experiment", Region == "East",
    Treatment != "Negative") %>%
  ggplot(aes(SMRE, fill = Genotype)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  ) +
  facet_wrap(~Timepoint, ncol = 1)
```

### Alt plot

Might be better to just focus on the treated experimental samples; can note
that the bimodal distribution in SMRE is due to the presence of Mel in the
timepoint 2, and thus we'd mess up our AA inferences if we tried to infer the
change in AA after Mel introduction. We might also (if were were analyzing
timepoint 2 samples)

Actually, including the bias values serves the important purpose of showing how
the SME varies less than the efficiencies. For that reason it could even be
worth having the single-isolate samples, though seems unneeded.

Points

* SME varies less across samples than efficiencies vary across taxa
* SME much greater after rusting (timepoint 2) -> it will look like the growth of the
  pathogen suppressed the proportion and absolute abundance of the colonizers
  5-6X more than it actually did
* The Mel prop and hence the SME varies significantly in the timepoint 2
  samples. This may lead to systematic errors in certain inferences. E.g., if
  host genotype affects Mel and another taxon, that other taxon's value will be
  distorted (and, I think, the effect on Mel will be inflated).
  * As an example, think through the case of more resistant host genotypes
  * Might also consider the taxa found to suppress Mel; see paper

```{r}
p1.1 <- sme %>%
  filter(Samp_type == "Experiment", Treatment != "Negative") %>%
  ggplot(aes(SMRE, fill = Timepoint)) +
  geom_histogram() +
  annotate("point", y = 200, x = bias, color = "black", shape = "|", size = 7) +
  expand_limits(x = range(bias)) +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    # x = "Sample mean relative efficiency (SMRE)",
    x = "Mean efficiency", 
    y = "Number of samples"
  )
p2.1 <- sme0 %>%
  filter(Samp_type == "Experiment", Treatment != "Negative") %>%
  ggplot(aes(y = Melampsora, x = SMRE, color = Timepoint)) +
  geom_point() +
  expand_limits(x = range(bias)) +
  scale_x_log10() +
  # scale_y_sqrt() +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 1e-2),
    breaks = c(0, 0.01, 0.1, 0.5, 1)
    ) +
  scale_color_brewer(type = "qual") +
  labs(
    y = "Melampsora proportion",
    x = "Mean efficiency"
  )
```

```{r}
(p1.1 + ggtitle("Distribution of the mean efficiency across samples")) / 
  (p2.1 + ggtitle("Melampsora proportion vs. mean efficiency")) +
  plot_annotation(tag_levels = "A")
```

TODO: Still need to determien if the y axis is accurate on histogram; I think
its off when using pseudo-log. But perhaps can get away with linear?

## Check errors in FCs

We can use this distribution to pick examples of samples where bias
approximately cancels in the FC estimates (because their SMRE's are similar),
and samples where bias does not approximately cancel (because their SMRE's are
very different). Could do for mock samples (using the ground truth) or for
natural samples (using the calibrated composition). Might as well do it both
ways; not sure which would be best for a paper. The mocks have the benefit of
knowing the ground truth, while the natural samples demonstrate practical
significance. Also, we can give nice intuitive description of what we're doing
in the natural samples by saying that we're comparing samples w/ low-low,
high-high, or low-high pathogen %.

TODO ^^

### mock samples

To pick from the mock samples:

```{r}
sme0 %>%
  filter(Samp_type == "Mock") %>%
  select(Sample, SMRE, Melampsora) %>%
  arrange(SMRE) %>%
  mutate(
    log10_SMRE = log10(SMRE),
    diff = c(NA, diff(log10_SMRE))
  )
```

- Mock.3 -> Mock.5 have the biggest difference in SMRE. Notably, its still only
  a factor of 5.5 difference.
- Mock.7 -> Mock.9 have the smallest log difference.

Next: Compute the actual and observed FC in proportions, for each of these
pairs of communities. Perhaps define a function to spit out the relevant data
frame, given any pair of samples, the phyloseq object (or calibrated phyloseq
object?), and the bias vector? Actually, for the mocks, should use the
"actual" table instead of calibrated.

Use these objects,

```{r}
mock_actual
ps.mock.pseudo
```

We can use `metacal::pairwise_ratios()` to compute the FCs in abundances
between all samples + for all taxa, given a phyloseq object.

```{r}
fcs <- list(Actual = mock_actual, Observed = ps.mock.pseudo) %>%
  map(transform_sample_counts, close_elts) %>%
  map(pairwise_ratios, margin = "samples", filter = FALSE) %>%
  map_dfr(psmelt, .id = "Type") %>%
  # redo sample.1 and sample.2 since missing for the mock
  select(-sample.1, -sample.2) %>%
  rename(Pair = Sample, Fold_change = Abundance) %>%
  separate(Pair, str_c("Sample.", 1:2), sep = ":", remove = FALSE) %>%
  mutate(across(OTU, factor, levels = bias %>% sort %>% names)) %>%
  left_join(sme0 %>% select(Sample.1 = Sample, SMRE, Melampsora), 
    by = "Sample.1") %>%
  left_join(sme0 %>% select(Sample.2 = Sample, SMRE, Melampsora), 
    by = "Sample.2", suffix = c(".1", ".2"))
```

```{r}
fcs %>%
  filter(Pair %in% c("Mock.5:Mock.3", "Mock.9:Mock.7")) %>%
  ggplot(aes(y = OTU, x = Fold_change, shape = Type, color = Type)) +
  facet_grid(. ~ Pair) +
  scale_shape_manual(values = c(Actual = 3, Observed = 1)) +
  scale_color_brewer(type = "qual", palette = 6) +
  geom_vline(xintercept = 1, color = "black", size = 0.2) +
  geom_point(size = 2) +
  theme_minimal_hgrid() +
  scale_x_log10(limits = c(3e-3, 1e2), oob = scales::squish) +
  labs(x = "Fold change in proportion") +
  plot_annotation(
    title = "Actual and observed fold changes between pairs of mocks"
  )
```

```{r}
ggsave("/tmp/leopold2020-fc-mock-examples.png", units = "in",
  width = 6, height = 3, scale = 1.5)
```

Can see that on the left, the error is positive for all taxa, and of a similar
magnitude (on the order of ~6X, as predicted) for all taxa except for
Epicoccum, which dropped out during sequencing in Mock.5; and on the right, the
error is smaller and not consistently positive or negative.  Sign errors
correspond to the observed and actual fold changes being on opposing sides of
the vertical red line at 1.

todo: add the SMRE changes to the facet labels, and perhaps mark the expected

What other types of plots, which might use all pairs of samples?

- we could plot the observed vs. predicted error

- we could check that the error really is taxon independent; one way to do this
  is to calc the correlation in LFC between observed and actual

```{r}
lvls <- sme0 %>% arrange(SMRE) %>% pull(Sample)
fcs0 <- fcs %>%
  # filter(Sample.1 < Sample.2) %>%
  select(Type:Fold_change, SMRE.1:Melampsora.2) %>%
  pivot_wider(names_from = Type, values_from = Fold_change) %>%
  mutate(across(c(Sample.1, Sample.2), factor, levels = lvls))
```

```{r}
fcs0 %>%
  filter(as.integer(Sample.1) < as.integer(Sample.2)) %>%
  ggplot(aes(Actual, Observed, color = OTU)) +
  geom_abline(color = "grey") +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  facet_grid(Sample.1 ~ Sample.2) +
  scale_color_brewer(type = "qual", palette = 3)
```

Note how (especially) for Mock 1 and 6, the points follow a straight line with
a greater slope than predicted. This seems likely related to the systematic
deviation from the model that I observed.

Check that there is indeed very high correlations; remove the Epicoccum Mock-5
dropout first to avoid major distortions just from that.

```{r}
corrs <- fcs0 %>%
  filter(
    Sample.1 != Sample.2,
    !((Sample.1 == "Mock.5" | Sample.2 == "Mock.5") & OTU == "Epicoccum")
  ) %>%
  mutate(across(c(Actual, Observed), log10)) %>%
  group_by(Pair) %>%
  summarize(correlation = cor(Actual, Observed, method = "pearson"))
corrs %>% pull(correlation) %>% summary
# qplot(corrs$correlation)
```

### Experiment samples

We could just try picking a high and low Melampsora sample - perhaps from the
mode of the peaks, rather than the extreme.

```{r}
sme0 %>%
  filter(Samp_type == "Experiment") %>%
  select(Sample, SMRE, Melampsora) %>%
  arrange(SMRE) %>%
  mutate(
    log10_SMRE = log10(SMRE),
    diff = c(NA, diff(log10_SMRE))
  )
```

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2021-04-03 notebook/_posts/2021-04-26-illustration-of-a-spurious-differential-abundance-result-due-to-bias/index.Rmd
Import analyses of leopold2020host experiment
These two analyses of the leopold2020host experiment are imported as
patches from MRM's personal research notebook. The first (earlier)
analysis, 'leopold2020host-bias-sensitivity', conducts an initial
exploration of the sample mean efficiency. The second analysis,
'illustration-of-a-spurious-differential-abundance-result-due-to-bias',
compares the results of regression on calibrated and uncalibrated data
as a case study of the impact of bias on differential abundance.  This
second analysis is the source of the file
'figures/leopold2020host-example

---
title: "Illustration of a spurious differential abundance result due to bias"
description: |
  Create illustration of a spurious differential abundance using the Leopold and Busby (2020) dataset.
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - bias sensitivity
  - R
  - ref:leopold2020host
  - differential abundance
date: 2021-04-26
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

# Setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot())
# stats helpers
library(broom)

library(metacal); packageVersion("metacal")
```

```{r load-data, include = FALSE}
## From the metacal 2.0 tutorial
# Download data from https://zenodo.org/record/3872145
data_path <- here::here("_data", "leopold2020host")
# To use a temporary directory:
# data_path <- file.path(tempdir(), "leopold2020")
if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  download.file(
    "https://zenodo.org/record/3872145/files/dleopold/Populus_priorityEffects-v1.2.zip",
    file.path(data_path, "Populus_priorityEffects-v1.2.zip")
  )
  unzip(
    file.path(data_path, "Populus_priorityEffects-v1.2.zip"), 
    exdir = data_path
  )
}
#> The microbiome data is stored in a phyloseq object,
ps <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/output/compiled/phy.rds") %>%
  readRDS

sample_data(ps) <- sample_data(ps) %>%
  transform(
    Timepoint = factor(Timepoint)
  )

mock_actual <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/data/MockCommunities.csv") %>%
  read.csv(row.names = 1) %>%
  select(-Sym4) %>%
  as("matrix") %>%
  otu_table(taxa_are_rows = FALSE) %>%
  transform_sample_counts(function(x) close_elts(1 / x))
mock_taxa <- taxa_names(mock_actual)
sam <- sample_data(ps) %>% as("data.frame") %>% as_tibble(rownames = "Sample")
tax <- tax_table(ps) %>% as("matrix") %>% as_tibble(rownames = "Taxon")
ps.mock <- ps %>% 
  subset_samples(Samp_type == "Mock") %>%
  prune_taxa(mock_taxa, .)
```

```{r estimate-bias}
ps.mock.pseudo <- ps.mock %>%
  transform_sample_counts(function(x) x + 1)
mc_fit <- estimate_bias(ps.mock.pseudo, mock_actual, boot = TRUE) %>% print
bias <- coef(mc_fit) %>% print
```

```{r}
mc_fit.summary <- summary(mc_fit)
print(mc_fit.summary)
```

```{r, fig.dim = c(5,3)}
coef_tb <- mc_fit.summary$coefficients
coef_tb %>%
  mutate(taxon = fct_reorder(taxon, estimate)) %>%
  ggplot(aes(taxon, estimate, 
      ymin = estimate / gm_se^2, ymax = estimate * gm_se^2)) +
  geom_hline(yintercept = 1, color = "grey") +
  geom_pointrange() +
  scale_y_log10() +
  coord_flip()
```

```{r calibrate}
ps.pseudo <- transform_sample_counts(ps, function(x) x + 1)
ps.pseudo.cal <- calibrate(ps.pseudo, bias) %>% print
```

# Sample mean efficiency across samples

To estimate the SMRE with formula `sum_i A_i E_i`, use the calibrated abundances.

```{r}
ps0 <- ps.pseudo.cal %>%
  prune_taxa(names(bias), .)
```

```{r}
otu <- ps0 %>% otu_table %>% orient_taxa(as = "cols") %>%
  transform_sample_counts(close_elts)
sam0 <- sample_data(ps0) %>% 
  as("data.frame") %>% 
  as_tibble(rownames = "Sample")
```

```{r}
sme <- otu %>% psmelt %>%
  left_join(bias %>% enframe("OTU", "Efficiency"), by = "OTU") %>%
  group_by(Sample) %>%
  summarize(SMRE = sum(Efficiency * Abundance)) %>%
  left_join(sam0, by = "Sample") %>%
  mutate(
    across(Timepoint, as.factor),
    across(Timepoint, fct_explicit_na, "Other"),
  )
```

# Regression analysis

First let's pick a subset of samples to work with.

```{r}
sam %>%
  count(Samp_type, Timepoint)
sam %>%
  filter(Samp_type == "Experiment") %>%
  count(Treatment, Timepoint)
sam %>%
  filter(Samp_type == "Experiment", Timepoint == 2) %>%
  count(Region, Treatment)
sam %>%
  filter(Samp_type == "Experiment", Treatment != "Negative") %>%
  count(Region, Timepoint)
sam %>%
  filter(Samp_type == "Experiment", Treatment != "Negative") %>%
  with_groups(Genotype, summarize, n1 = sum(Timepoint == 1), n2 = sum(Timepoint == 2), tot = n())
```

The "Negative" treatment was not inoculated with commensals; let's drop that from the analysis.
Note that there are 2 more West genotypes adn thus more West than East samples.

## Time point

Asks, how did taxa vary in proportion with the addition of Mel?
Note, we expect the commensal taxa to typically decrease given that Mel successfully infects most plants.

Want to do linear regression on the timepoint.

starting point - observed and calibrated proportions in the experiment samples; SME in the experiment samples

```{r}
ps1.obs <- ps.pseudo %>%
  subset_samples(Samp_type == "Experiment" & Treatment != "Negative") %>%
  prune_taxa(mock_taxa, .) %>%
  orient_taxa(as = "cols") %>%
  transform_sample_counts(close_elts)
ps1.cal <- ps1.obs %>% calibrate(bias)
sam1 <- sample_data(ps1.obs) %>% as("data.frame")
sme1 <- sme %>% filter(Sample %in% sample_names(ps1.obs)) %>% rename(Mean_efficiency = SMRE)
```

Note the different numbers of samples at each timepoint
Should consider more sophisticated resgression models.

```{r}
fit.obs <- lm(log2(otu_table(ps1.obs)) ~ Timepoint, data = sam1)
fit.cal <- lm(log2(otu_table(ps1.cal)) ~ Timepoint, data = sam1)
fit.sme <- lm(log2(Mean_efficiency) ~ Timepoint, data = sme1)
```

Check that the observed changes - the calibrated changes are off by the expected constant shift,

```{r}
t(coef(fit.obs) - coef(fit.cal))
coef(fit.sme)
```

```{r}
coef(fit.obs) %>% t
coef(fit.cal) %>% t
```

The effect of bias is significant; e.g. Aureobasidium and Trichoderma appear to decrease in proportion by 16X but actually decrease by 4X, and Penicillium appears to decrease by 5.9X but only decreases by 1.5X.

Could use the case of Penicillium as our example.

```{r}
# Check expected relationship between coefficients on intercept and slope
all.equal(
  coef(fit.obs)[,"Penicillium"],
  coef(fit.cal)[,"Penicillium"] + c(log2(bias["Penicillium"]), 0) - coef(fit.sme)
)
```

```{r}
tb <- bind_rows(
  Observed = psmelt(ps1.obs),
  Calibrated = psmelt(ps1.cal),
  .id = "Type"
) %>%
  mutate(across(Type, fct_relevel, "Calibrated")) %>%
  rename(Proportion = Abundance)
```

```{r}
tb %>%
  filter(OTU == "Penicillium") %>%
  ggplot(aes(y = log2(Proportion), x = Timepoint, color = Type)) +
  geom_quasirandom(alpha = 0.4) +
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.8) +
  scale_color_brewer(type = "qual", palette = 6)
```

```{r}
tb %>%
  filter(OTU == "Penicillium") %>%
  ggplot(aes(y = log2(Proportion), x = Timepoint)) +
  geom_quasirandom(alpha = 0.4) +
  stat_summary(fun = "mean", geom = "point", size = 3) +
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.8, arrow = grid::arrow(length = unit(0.15, "inches"))
  ) +
  facet_wrap(~Type)
```

TODO: Add a panel with the sample mean efficiency

```{r}
sme1 %>%
  ggplot(aes(y = log2(Mean_efficiency), x = Timepoint)) +
  geom_quasirandom(alpha = 0.4) +
  stat_summary(fun = "mean", geom = "point", size = 3) +
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.8, arrow = grid::arrow(length = unit(0.15, "inches"))
  )
```

Question: For the purposes of illustration, might we want to find the host region or host genotype with the largest change in SME?
Ok as long as we're clear what we're doing.
Reducing the number of data points and biological variation might actually make things cleaner as well (or could use boxplots), since the point is not to dwell on all the variation, but rather just the average effect.

for paper, might be clearer to call "Corrected" instead of "Calibrated".

Can easily split graph by region or host genotype:

```{r}
tb %>%
  filter(OTU == "Penicillium") %>%
  ggplot(aes(y = log2(Proportion), x = Timepoint, color = Region)) +
  geom_quasirandom(alpha = 0.4) +
  stat_summary(fun = "mean", geom = "point", size = 3) +
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.8, arrow = grid::arrow(length = unit(0.15, "inches"))
  ) +
  facet_wrap(~Type)
```

```{r}
tb %>%
  filter(OTU == "Penicillium") %>%
  ggplot(aes(y = log2(Proportion), x = Timepoint)) +
  geom_quasirandom(alpha = 0.3) +
  #> stat_summary(fun = "mean", geom = "point", size = 3) +
  stat_summary(fun.data = mean_cl_normal, fun.args = c(conf.int = 0.95), 
    geom = "pointrange") +
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.8, arrow = grid::arrow(length = unit(0.15, "inches"))
  ) +
  facet_grid(Region~Type)
```

Note that the 95% CIs are small and hard to see.

Want to plot the sme's with the change in proportion; make sure the y-axes have the same scale.
One way to simplify this is to show the mean efficiency relative to the taxon in question;
then it will give the difference between calibrated and observed.

```{r}
sme2 <- sme1 %>%
  mutate(Mean_relative_efficiency = Mean_efficiency / bias["Penicillium"]) %>%
  rename(Proportion = Mean_relative_efficiency) %>%
  mutate(Type = "Mean relative efficiency")
tb1 <- tb %>%
  filter(OTU == "Penicillium") %>%
  bind_rows(sme2) %>%
  mutate(across(Type, fct_relevel, c("Calibrated", "Observed", "Mean relative efficiency")))
```

```{r}
lyrs <- list(
  geom_quasirandom(alpha = 0.3),
  #> stat_summary(fun = "mean", geom = "point", size = 3),
  stat_summary(fun.data = mean_cl_normal, fun.args = c(conf.int = 0.95), 
    geom = "pointrange"),
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.8, arrow = grid::arrow(length = unit(0.15, "inches"))
  )
)
tb1 %>%
  ggplot(aes(y = log2(Proportion), x = Timepoint)) +
  lyrs +
  facet_grid(Region~Type)
```

```{r}
tb1 %>%
  mutate(across(Type, fct_reorder, Proportion)) %>%
  ggplot(aes(y = log2(Proportion), x = Timepoint, color = Type)) +
  geom_quasirandom(alpha = 0.3) +
  stat_summary(fun = "mean", geom = "point", size = 3) +
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.8, arrow = grid::arrow(length = unit(0.15, "inches"))
  ) +
  scale_color_brewer(type = "qual", palette = 2) +
  guides(color = guide_legend(reverse = TRUE)) +
  facet_grid(.~Region) +
  geom_text(data = tibble(Region = "West"), x = 2.7, y = 5,
    color = "black", size = 4.5, hjust = 0, vjust = 1,
    label = "Pathogen growth causes\nmean efficiency to increase,",
  ) +
  geom_text(data = tibble(Region = "West"), x = 2.7, y = -10,
    color = "black", size = 4.5, hjust = 0, vjust = 1,
    label = "which leads to larger-than-\nactual decreases in the\nfocal taxon Penicillium.",
  ) +
  coord_cartesian(xlim = c(1, 2), clip = "off") +
  theme(
    legend.position = "top",
    plot.margin = margin(r = 2.5, unit = "in"),
  ) +
  plot_annotation(
    title = "Effect of bias on Penicillium differential abundance"
  )
```

Summary:

- Regression of log2(Proportion) of the fungus Penicillium versus timepoint; timepoints 1 and 2 are pre- and post- challenge with the pathogen Melampsora. Data is split by the region the host plants are derived from.
- Calibrated proportions = Observed proportions in the real, experimental samples after adjustment for the bias measured in mock communities
- Mean efficiency of each community is inferred by treating the calibrated proportions as the truth, and multiplying by efficiencies estimated from the mocks
- The pathogen Melampsora has a high measurement efficiency; thus once it infects the plants, the mean efficiency of the sample increases (purple points). Efficiency is here taken as relative to the focal taxon Penicillium.
- West plants tend to be more resistant to the pathogen, which likely explains why the mean efficiency doesn't increase as much in the West plants
- Penicillium is observed to decrease in log proportion, in both the East and West plants. But the calibrated measurements show that it actually slightly increases in the East plants, and has a lesser decrease in the West plants than what was observed before bias correction.
- The difference between the Calibrated and Observed data points and regression lines equals the regression line of the mean efficiency: Orange = Green + Purple; Green = Orange - Purple
- The absolute error in regression coeffiecients is the same for all taxa. I picked Penicillium for illustration since it has the smallest observed decrease, which makes the error due to bias have a particularly significant impact.
<!--  -->

```{r, eval = FALSE, echo = FALSE}
ggsave("/tmp/leopold2020-example.pdf", units = "in", width = 6, height = 4, scale = 1.4)
ggsave("/tmp/leopold2020-example.png", units = "in", width = 6, height = 4, scale = 1.4)
```

### Alt. plot

```{r}
tb2 <- tb1 %>%
  mutate(
    log2_proportion = log2(Proportion),
    across(Type, fct_relevel, c("Observed", "Calibrated", "Mean relative efficiency")),
    #> across(Type, fct_recode, Corrected = "Calibrated", Uncorrected = "Observed")
  )
```

Let's show the calibrated, observed, and mean efficiency side by side, similar to my initial explanatory plot.

Question: How to get the same y-axis range in each facet?
We need to determine what we want the min and max y values for each facet to be.
Choose so that the range in each panel equals the max data range for any one panel, and center the data range of each panel within the plot range.

```{r}
yr <- tb2 %>%
  with_groups(c(Region, Type), summarize, 
    across(log2_proportion, c(min = min, max = max), .names =  "data_{.fn}")
 ) %>%
  mutate(
    data_range = data_max - data_min,
    data_mid = (data_max + data_min) / 2,
    y_min = data_mid - max(data_range) / 2,
    y_max = data_mid + max(data_range) / 2
  ) %>%
  print
stopifnot(yr %>% {sd((.$y_max - .$y_min))} < 1e-14)
yr.long <- yr %>%
  pivot_longer(c(y_min, y_max))
```

```{r, fig.dim = 1.7 * c(6, 4)}
tb2 %>%
  ggplot(aes(y = log2_proportion, x = Timepoint)) +
  geom_quasirandom(alpha = 0.3) +
  geom_blank(data = yr.long, aes(y = value, x = 1)) +
  stat_summary(fun = "mean", geom = "point", size = 3) +
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.9, arrow = grid::arrow(length = unit(0.15, "inches")), color = "red"
  ) +
  scale_color_brewer(type = "qual", palette = 2) +
  guides(color = guide_legend(reverse = TRUE)) +
  facet_wrap(Region~Type, scales = "free_y") +
  coord_cartesian(xlim = c(1, 2), clip = "off") +
  scale_y_continuous(breaks = seq(-20, 10, by = 2)) +
  labs(y = "log2(Proportion)")
```

```{r}
text_tb <- yr %>%
  filter(Type %in% c("Calibrated", "Observed")) %>%
  mutate(
    text = case_when(
      Type == "Observed" ~ "=",
      Type == "Calibrated" ~ "–",
    ),
    y = y_min + (y_max - y_min) / 2
  )

#> Label "Calibrated" as "Actual" for purposes of illustration?
lblr <- function(labels) {
  labels %>% pull(1) %>% fct_recode(Actual = "Calibrated") %>% as.character %>% list
}

lyrs <- list(
  geom_quasirandom(alpha = 0.3),
  #> stat_summary(fun = "mean", geom = "point", size = 3),
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange"),
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    #> arrow = grid::arrow(length = unit(0.15, "inches")),
    size = 0.9, color = "red"
  ),
  scale_color_brewer(type = "qual", palette = 2),
  guides(color = guide_legend(reverse = TRUE)),
  facet_wrap(~Type, scales = "free_y", labeller = lblr),
  #> theme_minimal_hgrid(),
  scale_y_continuous(breaks = seq(-20, 10, by = 2)),
  theme(
    plot.title = element_text(face = "plain"),
    legend.position = "top",
    panel.spacing.x = unit(0.75, "in")
  ),
  labs(y = "log2(Proportion)"),
  # Corrected = Observed - Mean relative efficiency
  coord_cartesian(xlim = c(1, 2), clip = "off")
)
p1 <- tb2 %>%
  filter(Region == "East") %>%
  ggplot(aes(y = log2_proportion, x = Timepoint)) +
  lyrs +
  geom_blank(data = yr.long %>% filter(Region == "East"), aes(y = value, x = factor(1))) +
  labs(title = "Eastern genotypes: Bias causes sign error") +
  geom_text(
    data = text_tb %>% filter(Region == "East"),
    aes(y = y, label = text),
    x = 2.9,
    color = "black", size = 8, hjust = 0.5, vjust = 0.5,
  )
p2 <- tb2 %>%
  filter(Region == "West") %>%
  ggplot(aes(y = log2_proportion, x = Timepoint)) +
  lyrs +
  geom_blank(data = yr.long %>% filter(Region == "West"), aes(y = value, x = factor(1))) +
  labs(title = "Western genotypes: Bias causes magnitude error") +
  geom_text(
    data = text_tb %>% filter(Region == "West"),
    aes(y = y, label = text),
    x = 2.9,
    color = "black", size = 8, hjust = 0.5, vjust = 0.5,
  )
```

```{r, fig.dim = 1.7 * c(6, 4)}
p1 / p2 + 
  plot_annotation(
    #> title = "Effect of bias on Penicillium differential abundance",
    tag_levels = "A"
  )
```

This ^ is my favorite so far

```{r, eval = FALSE, echo = FALSE}
#> ggsave("/tmp/leopold2020-example.pdf", units = "in", width = 6, height = 4, scale = 1.7)
ggsave("/tmp/leopold2020-example.png", units = "in", width = 6, height = 4, scale = 1.7)
```

TODO:

- Change y label to log2(value) and adding "proportion" to the two proportion facets
- Change back to "Calibrated proportion" and "Observed proportion"?
- Slightly increase spacing between patchwork  panels (rows)
- Swap positions of calibrated and uncalibrated so that the equation is correct: Observed = Actual - Mean efficiency

alternate approach using 2-d facet wrap:

```{r, fig.dim = 1.5 * c(6, 3.5)}
symbol_labels <- yr %>%
  filter(Type %in% c("Calibrated", "Observed")) %>%
  mutate(
    text = case_when(
      Type == "Observed" ~ "=",
      Type == "Calibrated" ~ "–",
      #> Type == "Mean relative efficiency" & Region == "East" ~ "Eastern\ngenotypes",
      #> Type == "Mean relative efficiency" & Region == "West" ~ "Western\ngenotypes",
    ),
    y = y_min + (y_max - y_min) / 2
  )
region_labels <- yr %>%
  filter(Type == "Mean relative efficiency") %>%
  mutate(
    text = case_when(
      Region == "East" ~ "Eastern genotypes\n\nBias causes\nsign error",
      Region == "West" ~ "Western genotypes\n\nBias causes\nmagnitude error",
    ),
    y = y_min + (y_max - y_min) / 2
  )

lyrs <- list(
  geom_quasirandom(alpha = 0.3),
  #> geom_blank(data = yr, aes(y = y_min, x = factor(1))),
  #> geom_blank(data = yr, aes(y = y_max, x = factor(1))),
  geom_blank(data = yr.long, aes(y = value, x = 1)),
  stat_summary(fun = "mean", geom = "point", size = 3),
  stat_smooth(aes(x = as.integer(Timepoint)), method = "lm", geom = "line",
    size = 0.9, arrow = grid::arrow(length = unit(0.15, "inches")), color = "red"
  ),
  scale_color_brewer(type = "qual", palette = 2),
  guides(color = guide_legend(reverse = TRUE)),
  #> facet_wrap(~Type, scales = "free_y"),
  #> theme_minimal_hgrid(),
  scale_y_continuous(breaks = seq(-20, 10, by = 2)),
  labs(y = "log2(Proportion)"),
  # Calibrated = Observed - Mean relative efficiency
  geom_text(
    data = symbol_labels,
    aes(y = y, label = text),
    x = 2.9, color = "black", size = 8, hjust = 0.5, vjust = 0.5,
  ),
  geom_text(
    data = region_labels,
    aes(y = y, label = text),
    x = 2.8, color = "black", size = 4.8, hjust = 0, vjust = 0.5,
  ),
  coord_cartesian(xlim = c(1, 2), clip = "off"),
  theme(
    legend.position = "top",
    panel.spacing.x = unit(0.75, "in"),
    plot.margin = margin(r = 1.75, unit = "in")
  )
)

# labeller assuming first col is region, second is type
lblr <- function(labels) {
  labels %>% pull(2) %>% as.character %>% list
}

tb2 %>%
  ggplot(aes(y = log2_proportion, x = Timepoint)) +
  lyrs +
  facet_wrap(Region~Type, scales = "free_y", 
    labeller = lblr
    #> labeller = labeller(Region = NULL, .multi_line = FALSE),
    #> strip.position = "right"
  )
```

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2021-04-05 relative-abundance.Rmd
Draft section on analysis of relative abundances

# Differential relative abundance {#relative}

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
TeX: {extensions: ["cancel.js"]}
});
</script>

This section describes the effects of bias on common approaches to analyzing differential relative abundance.
We introduce the critical distinction between analyses based on proportions and analyses based on ratios.
Analyses based on proportions can be distorted by bias, whereas analyses of log fold changes (LFCs) in ratios of atomic taxa are invariant to bias.
If atomic taxa may be aggregated into synthetic taxa by multiplication (or by adding log abundance), and ratios formed from such aggregates; LFCs for such generalized ratios remain bias invariant.
But if atomic taxa are aggregated additively by summing their read counts—as often done to analyze higher-order taxa such as phyla—then bias invariance only applies if the aggregated taxa have the same efficiency.
The error induced by bias in (some) proportion-based methods has a simple form that suggests tractable experimental methods by which it may be estimated and corrected, which we discuss in Section \@ref(calibration).
Section \@ref(absolute) shows that methods for obtaining absolute abundances from MGS data can generally be split into those based on proportions and those based on ratios, so that the proportion-ratio dichotomy also provides a useful framework for understanding analyses of differential absolute abundance.

## Proportion-based analyses

### Log proportion

To simplify notation where possible, I will write $B$ for $B^{(M)}$ when only the metagenomics measurements are relevant.

It follows from  \@ref(eq:m) that the observed fold change in the proportion of taxon $i$ from sample $j$ to $j'$ is
\begin{align}
  (\#eq:m-fc)
  \frac{m_{ij'}}{m_{ij}}
  = \frac{a_{ij'} \cancel{B_i} / \bar B_{j'}}{a_{ij} \cancel{B_i} / \bar B_{j}}
  = \frac{a_{ij'}}{a_{ij}} \cdot \frac{\bar B_j}{\bar B_{j'}}.
\end{align}
The taxon-specific error term ($B_i$) cancels, leaving a multiplicative error equal to the inverse change in the sample mean efficiency $\bar B^{(M)}$.
Framed in terms of the log fold change, this equation becomes
\begin{align}
  (\#eq:m-lfc)
  \underbrace{ \log m_{ij'} - \log m_{ij} }_\text{observed LFC}
  &= \underbrace{ \log a_{ij'} - \log a_{ij} }_\text{true LFC}
  - \underbrace{ \left(\log \bar B_{j'} - \log \bar B_j\right) }_\text{LFC in mean efficiency},
\end{align}
so that the additive error is the LFC in mean efficiency between the two samples.

Analyzing fold changes in a regression framework typically entails considering the expectation of the logarithm of the response conditional on the value of a set of covariates $X$.
I use $E [Y \mid X]$ as shorthand for $E [Y \mid X = x]$, the expected value of a random variable $Y$ given a vector of covariate values $x$.
Applying logarithms to \@ref(eq:m) gives
\begin{align}
  (\#eq:m-log)
  \log m_i = \log a_i + \log B_i - \log \bar B
\end{align}
which, after conditioning on the covariates, becomes
\begin{align}
  (\#eq:m-log-regression)
  E [\log m_i \mid X] 
  = E [\log a_i \mid X] + \log B_i - E [\log \bar B \mid X].
\end{align}
The taxon-specific term $\log B_i$ creates a constant error that is unproblematic for differential abundance analysis, which is typically ignores the baseline abundance of the taxon.
However, the sample-specific error $\log \bar B$ can distort the inferred relationship between $X$ and the expected value of the true log proportion if its expected value also varies with $X$.

**TODO: Explain this $\gamma$ notation for the regression coefficients, or find a more recognizable notation.**

Consider the special case of the linear regression
\begin{align}
  \log a_i &= X \gamma + \epsilon,
\end{align}
where $X$ is a $J\times p$ covariate matrix and $\gamma$ is a $p\times 1$ vector of coefficients.
Suppose we knew the values of each of the terms in \@ref(eq:m-log).
Setting each term as the left-hand-side of its own corresponding regression, the least-squares coefficients are related as (\@ref(eq:regression-coefficient-error))
\begin{align}
  (\#eq:m-log-regression-coefficients)
  \hat \gamma{(\log m_i)} = \hat \gamma{(\log a_i)} + \hat \gamma{(\log B_i)} - \hat \gamma{(\log \bar B)}.
\end{align}
For the simple linear regression,
\begin{align}
  \log a_i &= \gamma_0 + \gamma_1 x + \epsilon,
\end{align}
with an intercept coefficient $\gamma_0$ and a slope coefficient $\gamma_1$, the relation \@ref(eq:m-log-regression-coefficients) implies
\begin{align}
  (\#eq:m-log-simple-regression-coefficients)
  \hat \gamma_0{(\log m_i)} &= \hat \gamma_0{(\log a_i)} + \log B_i - \hat \gamma_0{(\log \bar B)} \\
  \hat \gamma_1{(\log m_i)} &= \hat \gamma_1{(\log a_i)} - \hat \gamma_1{(\log \bar B)},
\end{align}
Since $\log B_i$ is constant, it only affects the intercept.
The slope estimate for the metagenomics proportions, $\hat \gamma_1{(\log m_i)}$, is systematically decreased from that of the true proportions, $\hat \gamma_1{(\log a_i)}$, by that for the sample mean efficiency, $\hat \gamma_1{(\log \bar B)}$.

### Log odds (logit)

The proportion of a taxon saturates at $1$ as its absolute abundance increases, while the odds, $a_i / (1 - a_i) = A_i / (A_T - A_i)$, continue to increase.
For this reason it is often more natural to perform regression on the log odds, or *logit-transformed proportion*, rather than log proportion.

The observed log odds of taxon $i$ is
\begin{align}
  \operatorname{logit} m_{ij} 
  = \log \frac{m_{ij}}{1 - m_{ij}}
  = \log \frac{M_{ij}}{M_{Tj} - M_{ij}}.
\end{align}
To understand the effects of bias on the measured log odds, it is helpful to define a variable $\bar B_{-i,j}$ to denote the mean efficiency in sample $j$ among 
taxa *other than $i$*,
\begin{align}
  (\#eq:mean-efficiency-not-i)
  \bar B_{-i,j} 
  \equiv \frac{\sum_{k \ne i} B_k A_{kj}}{\sum_{k \ne i} A_{kj}}
  = \frac{\sum_{k \ne i} B_k a_{kj}}{1 - a_{ij}}.
\end{align}
The measured proportion of non-$i$ taxa is
\begin{align}
  (\#eq:one-minus-m)
  1 - m_{ij}
  &= \sum_{k \ne i} m_{kj}
\\&= \sum_{k \ne i} a_{kj} \cdot \frac{B_k}{\bar B_j}
\\&= (1 - a_{ij}) \cdot \frac{\bar B_{-i,j}}{\bar B_j},
\end{align}
where the final expression follows from $\sum_{k\ne i} a_{kj} B_k = (1-a_{ij}) \bar B_{-i,j}$.
Intuitively, Equation \@ref(eq:one-minus-m) says that the fold error in the proportion of "not-$i$" equals the mean efficiency of the "not-$i$" part of the sample relative to the mean efficiency of the sample as a whole.
This result means that the measured odds of taxon $i$ is
\begin{align}
  (\#eq:m-odds)
  \frac{m_{ij}}{1 - m_{ij}}
  &= \frac{a_{ij}}{1 - a_{ij}} \cdot \frac{B_i}{\bar B_{-i,j}},
\end{align}
This formula resembles Equation \@ref(eq:M-ratio) for the measured ratio of two taxa $i$ and $i'$, but with taxon $i'$ corresponding to all not-$i$ taxa.
In this case, however, the efficiency of not-$i$ varies among samples as changes in their relative abundances change $\bar B_{-i, j}$.
Hence, in contrast to the ratio of two taxa, the error is not consistent across samples.

The error in the logit of taxon $i$ is 
\begin{align}
  (\#eq:m-logit)
  \operatorname{logit} m_i = \operatorname{logit} a_i + \log B_i - \log \bar B;
\end{align}
note the log rather than logit operators in the efficiency terms.
The corresponding regression equation is
\begin{align}
  (\#eq:m-logit-regression)
  E [\operatorname{logit} m_i \mid X] 
  = E [\operatorname{logit} a_i \mid X] + \log B_i - E [\log \bar B_{-i} \mid X].
\end{align}

## Ratio-based analyses

The previous section showed that differential abundance analyses based on proportions may be sensitive to bias, due to the dependence of the multiplicative error in measured proportions on the sample mean efficiency (Equation \@ref(eq:m)).
In contrast, the multiplicative error in the ratios among taxa (Equation \@ref(eq:M-ratio)) is independent of sample composition, such that a differential abundance analysis based on log ratios is invariant to bias.
The simplest such analysis is when "abundance" is equated with the log ratio of two specific taxa; however, any linear combination of log ratios may be used, which includes a variety of so-called Compositional Data Analysis (CoDA) methods that have recently been applied in microbiome DA analysis (see below).
@mclaren2019cons described the bias invariance of ratio-based methods (see their Equations 6 and 7).
Here we show how this general result applies to specific types of differential relative abundance.

We first consider the ratio a pair of taxa $i$ and $i'$.
It follows from Equation \@ref(eq:M-ratio) that the measured fold change in the ratio of taxon $i$ to taxon $i'$ from sample $j$ to sample $j'$ is
\begin{align}
  (\#eq:M-ratio-fc)
  \frac{m_{ij'} / m_{i'j'}}{m_{ij} / m_{i'j}}
  = \frac{a_{ij'} \cancel{B_i} / a_{i'j'} \cancel{B_{i'}}}
         {a_{ij} \cancel{B_i} / a_{i'j} \cancel{B_{i'}}}
  = \frac{a_{ij'} / a_{i'j'}}{a_{ij} / a_{i'j}}.
\end{align}
The error in each ratio, $B_i / B_{i'}$, is constant and so cancels completely, leaving the true fold change.
To see the effect of bias on regression of log ratios, we note that the error equation \@ref(eq:M-ratio) implies
\begin{align}
  (\#eq:M-ratio-log)
%  \log m_i - \log m_{i'} = \log a_i - \log a_{i'} + \log B_i - \log B_{i'} \\
  \log \frac{m_i}{m_{i'}} = \log \frac{a_i}{a_{i'}} + \log \frac{B_i}{B_{i'}}.
\end{align}
Taking the conditional expectation given a covariate vector $X$ gives
\begin{align}
  (\#eq:m-ratio-log-regression)
  E\left[ \log \frac{m_i}{m_{i'}} \mid X \right] 
  = E\left[ \log \frac{a_i}{a_{i'}} \mid X \right] + \log \frac{B_i}{B_{i'}}.
\end{align}
In contrast to the case of a log proportion (Equation \@ref(eq:m-log-regression)), the effect of bias on the log ratio is simply to create a constant shift and therefore does not affect differential-abundance analysis.

CoDA DA methods often draw upon generalized notion of ratio, in which the numerator and/or denominator consist of a product of powers of multiple taxa.
Such a generalized ratio is determined by length-$I$ vectors $n$ and $d$ giving the power of each taxon in the numerator and denominator, with $n_i = 0$ indicating that the taxon does not affect the term.
Denoting the actual value of a given generalized ratio in sample $j$ as $y_j$ and the observed as $z_j$, we have
\begin{align}
  (\#eq:generalized-ratio)
  \text{actual:}   \quad  y_j = \frac{\prod_i A_{ij}^{n_i}}{\prod_i A_{ij}^{d_i}} &&
  \text{observed:} \quad  z_j = \frac{\prod_i M_{ij}^{n_i}}{\prod_i M_{ij}^{d_i}}.
\end{align}
Under the MWC bias model, the observed and actual log ratios are related as
\begin{align}
  (\#eq:generalized-ratio-log)
  \log z_j &= \log y_j + \log \frac{\prod_i B_{ij}^{n_i}}{\prod_i B_{ij}^{d_i}}.
\end{align}
Hence, as with the simple ratio between two taxa, the error due to bias is independent of sample composition and so does not affect the observed variation in $\log z_j$.

Most of the log-ratio transformations that are used in CoDA DA analysis can be understood as particular applications of the generalized ratio \@ref(eq:generalized-ratio) and thus lead to bias-invariant DA results under the deterministic MWC model.
For example, the additive log-ratio (ALR) transformation consists of computing the log of the ratios obtained by dividing each taxon's abundance by that of a particularly chosen reference taxon $r$,
\begin{align}
  (\#eq:alr)
  \operatorname{alr} A_{j} = \left[ \log \frac{A_{ij}}{A_{rj}}, \dots, \log \frac{A_{Ij}}{A_{rj}} \right],
\end{align}
while the centered log-ratio (CLR) transformation instead sets the denominator to the geometric mean $g(A_j) = \left(\prod_i A_{ij}\right)^{1/I}$ of all taxa,
\begin{align}
  (\#eq:clr)
  \operatorname{clr} A_{j} = \left[ \log \frac{A_{ij}}{g(A_{j})}, \dots, \log \frac{A_{Ij}}{g(A_{j})} \right].
\end{align}
More generally, one might choose any set $R$ of reference taxa and use their geometric mean in the denominator;
the ALR and CLR transformations then correspond to taking $R=r$ and $R=I$, respectively.
@quinn2019afie call this transformation the /multiple additive log-ratio/ (MALR) transformation,
\begin{align}
  (\#eq:malr)
  \operatorname{malr} A_{j} = \left[ \log \frac{A_{ij}}{g(A_{R,j})}, \dots, \log \frac{A_{Ij}}{g(A_{R,j})} \right].
\end{align}
The actual and observed MALR-transformed abundance associated with a specific numerator taxon $i$ is
\begin{align}
  (\#eq:malr-error)
  \operatorname{malr}(M_{j})_i = \operatorname{malr}(A_{j})_i + \operatorname{malr}(B)_i;
\end{align}
hence the error due to bias is constant and does not factor into the inferred changes in a DA analysis.
Note: Unlike @quinn2019afie we do not include the "robust centered log-ratio transformation" of @martino2019anov as an example of an MALR; this transformation chooses a distinct set of reference taxa for each sample and so does not have bias invariance under the MWC model.

Other generalized log-ratios commonly used in microbiome data analysis include *balances*.
A balance is defined by two sets of taxa, $Q$ and $R$, and equals the log of the ratio of the geometric mean of $Q$ taxa to the geometric mean of $R$ taxa, multiplied by a scaling factor,
\begin{align}
  (\#eq:balance)
  \sqrt{\frac{|Q| |R|}{|Q| + |R|}} \log \frac{g(A_{Q,j})}{g(A_{R,j})}.
\end{align}
Balances have become a popular tool for identifying biomarkers or groups of taxa associated with an environmental or health condition while operating within the CoDA framework (@washburne2017phyl, @riverapinto2018bala, @quinn2020inte).
The exponential of a balance is an example of the generalized ratio in \@ref(eq:generalized-ratio) and so regression of balances is again invariant to bias.

The products in the numerators and denominators in \@ref(eq:generalized-ratio) can be seen as a way to multiplicatively aggregate abundances of different taxa, as opposed to the additive aggregation that is commonly used when relative abundances are viewed as proportions.
Multiplicative aggregation preserves the property of perturbation invariance that is the source of bias invariance under the MWC model.
In contrast, additive aggregation (known as "amalgamation" in the CoDA literature) violates perturbation invariance and can lead to regression results that depend on bias even when they are ostensibly based on the ratios of two (non-atomic) taxa, as we now explain.

## Higher-order taxa formed by additive aggregation

Microbiome researchers often combine lower-order taxa into higher-order taxa prior to conducting a DA analysis.
For instance, Amplicon Sequence Variants (ASVs), Operational Taxonomic Units (OTUs), or species-level counts may be aggregated into genus- or family-level counts by simply summing the counts within each group of lower-level taxa.
Such aggregation can increase statistical power by reducing noise, sparsity, and the number of tests conducted.
It also simplifies the task of interpretation by limiting the number of taxa considered.
In addition, some degree of uncontrolled aggregation is inevitable due to the inherent limitations of our sequencing and bioinformatics protocols to distinguish sufficiently similar organisms (@mclaren2019), which may be lumped together into an OTU or species-level feature.
Finally, though our focus is on analysis of taxa, we note that taxonomic aggregation also occurs (at least implicitly) in gene or function analyses, where counts are combined from different taxa that share the same genes or predicted functions.
Hence, to understand the effects of bias on DA analysis as it is actually practiced, we must understand the effects of such taxonomic aggregation on differential abundance.

Consider a set of atomic taxa given by the set $Q \subset \{1, \dots, I\}$ and let $A_{Qj}$ be the vector of abundances in sample $j$ of the taxa in $Q$.
The actual absolute abundance of the synthetic taxon $Q$ is simply the sum of abundances of its component taxa, $\operatorname{sum}\left(A_{Q,j}\right) = \sum_{q \in Q} A_{qj}$.
<!-- where $\operatorname{sum}$ so that -->
Similarly, the actual proportion of $Q$ is given by the sum of proportions, $\operatorname{sum}\left(a_{Q,j}\right) = \sum_{q \in Q} a_{qj} = \operatorname{sum}\left(A_{Q,j}\right) / A_{Tj}$.
The abundance of $Q$ in the metagenomics measurement is $\operatorname{sum}\left(M_{Q,j}\right)$, and its proportion $\operatorname{sum}\left(m_{Q,j}\right)$.

Such a synthetic taxon has a consistent, composition-independent measurement efficiency only if its component taxa have equal efficiencies.
Let $\bar B_{Q,j}$ denote the mean efficiency of the taxa in $Q$ in sample $j$,
\begin{align}
  (\#eq:subset-mean-efficiency)
  \bar B_{Q,j} = \frac{\sum_{q \in Q} B_q A_{qj}}{\sum_{q \in Q} A_{qj}}.
\end{align}
Unless all the taxa in $Q$ have the same efficiency, the mean efficiency of $Q$ varies with the relative abundances among these taxa.
Summing over Equation \@ref(eq:m) and performing some rearranging shows that the observed proportion of $Q$ is
\begin{align}
  (\#eq:sum-m)
  \operatorname{sum}\left(m_{Q,j}\right) 
  &= \frac{1}{\bar B_j} \sum_{q\in Q} a_{qj} B_q \sum_{q\in Q} m_{qj} \\
  &= \operatorname{sum}\left(a_{Q,j}\right) \cdot \frac{\bar B_{Q,j}}{\bar B_j}.
\end{align}
In words, the error in the proportion of the aggregate taxon $Q$ is given by the mean efficiency of taxa in $Q$ relative to the sample mean.
Hence only if the atomic taxa that make up $Q$ all have the same efficiencies or always appear in the same ratios to each other will the efficiency of $Q$ be constant across samples.
This observation was first made in @mclaren2019cons (Discussion and Appendix 1) and is here extended to its general form.

The lack of consistency in the efficiency of the synthetic taxon $Q$ complicates the form of the error in both proportion- and ratio-based DA analyses.
The equivalent of the regression equation \@ref(eq:m-log-regression) for the log proportion of $Q$ is
\begin{align}
  (\#eq:sum-m-log-regression)
  E \left[ \log \sum_{q\in Q} m_{q} \mid X \right] 
  = E \left[ \log \sum_{q\in Q} a_{q} \mid X \right] + E\left[\log \bar B_{Q} - \log \bar B \mid X\right];
\end{align}
we can no longer move the log efficiency of the focal taxon out of the expectation operator and so must consider the variation in the mean efficiency of $Q$ as well as of the sample as a whole.
To consider the effect on ratio-based analyses, we define another synthetic taxon $R\subset \{1, \dots, I\}$ and consider the ratio of $Q$ to $R$.
The observed ratio,
\begin{align}
  (\#eq:M-ratio-u-v)
  \frac{M_{Q,j}}{M_{R,j}}
  &= \frac{A_{Q,j}}{A_{R,j}} \cdot \frac{\bar B_{Q,j}}{\bar B_{R,j}},
\end{align}
has an error equal to the ratio in mean efficiency of $Q$ to $R$, both of which can vary with the relative abundances of the component taxa.
As the ratio no longer as a consistent error, it is possible for bias to lead to spurious inferences in the fold change in the ratio.

### 2021-05-04 index.Rmd
Update title, abstract, and readme

@@
 ---
-title: "Characterizing and mitigating the effects of experimental bias on differential (absolute) abundance analysis in microbiome studies"
+title: "A mathematical framework for understanding microbiome differential-abundance methods"
 author: "Michael R. McLaren"
@@
 github-repo: mikemc/bda-theory
-description: "Mathematical theory of the effects of experimental bias on differential (relative and absolute) abundance analysis in microbiome studies."
+description: "A mathematical framework for understanding (relative and absolute) microbiome differential-abundance methods"
 ---
@@
 
-This monograph develops a mathematical theory of the effects of protocol- and taxon-specific experimental bias on various types of differential (relative and absolute) abundance analyses in microbiome studies and uses it to formulate several mitigation strategies.
+This manuscript develops a mathematical framework for understanding microbiome differential-abundance analysis; 
+specifically, how taxonomic bias and other complications of real microbiome measurement can distort the results of commonly-used methods.
+This framework predicts that certain approaches to relative and absolute differential abundance are more sensitive to bias than others.
+The effects of bias can be mitigated by choosing an insensitive approach or performing a calibration procedure using targeted abundance measurements of one or more specific taxa.

### 2021-05-05 appendix-dna.Rmd
Add derivations relating to DNA measurement

# DNA measurement and spike-ins

This appendix aims to derive expressions for the error in individual samples and figure out whether the error is constant across samples (so that it cancels out LFC analysis). 
Why difference? Extraction has occured prior to the targeted measurement.

Punchline: Bias will cancel like before if and only if DNA extraction yield is perfectly proportional to input, in a particular sense I should perhaps create a name for. 
It needs to be proportional to input for a fixed composition; it doesn't have to be proportional across compositions.

## Expanded model and notation

I partition the experiment into extraction and sequencing steps,
\begin{align}
  A_{ij} 
  \overset{\text{extraction}}{\underset{F_{ij}'}\longrightarrow}
  A_{ij}'
  \overset{\text{sequencing}}{\underset{F_{ij}''}\longrightarrow}
  \to A_{ij}'' = M_{ij}.
\end{align}
where $A'_{ij}$ denotes the absolute abundances after extraction.
The abundances $A''_{ij}$ after sequencing are just the counts $M_{ij}$.
I define a matrix $F_{ij}'$ of factors equalling the fold change in abundance during the extraction step, $F_{ij}' = A_{ij}' / A_{ij}$, and similarly let $F_{ij}''$ be the fold changes during the sequencing step,
The total fold change is just the product: $F_{ij} = F'_{ij} F''_{ij} = M_{ij} / A_{ij}$.

The MWC model implies that the relative values of $F_{ij}$ to $F_{i'j}$ are independent of $j$, for both the protocol as a whole and the individual steps.
Therefore we can write the factors as product of the taxon-specific bias $B_i$ and a sample-specific scaling factor $C_j$, as in Section \@ref(models).
I use $'$s to do this for the different steps, for example $F_{ij}' = B_i' C_j'$.
The total metagenomics bias and scaling factors thus split into steps as $B_i = B_i' B_i''$ and $C_j = C_j' C_j''$.

**TODO:** Add Note about units. We could be using bp, ng, genome copies, or marker-gene copies. It shouldn't affect the conclusions, but affects the units and values of the multipliers. Though perhaps it will help to pick ng DNA or genome copies

What is $C'_j$?
\begin{align}
  C'_j = \frac{A'_{Tj}}{A_{Tj} \bar{B}'_j}.
\end{align}
Note that we have so far been defining the metagenomics efficiencies $B$ as being (arbitrarily) the efficiency relative to the first taxon, which affects the scale of $C'$.

## Targeted DNA measurement

**TODO:** explain the big picture of what I'm doing here

Let $B^{(T)}_r$ be the conversion factor associated with the targeted measurement of reference taxon $r$ from the extracted DNA, which I assume is sample-independent but can vary by taxon.
NOTE: I'm using $B$ for "bias" here but we need to keep in mind that I'm talking about the bias relative to the DNA and not the cellular abundance.
In other words, the result of the targeted DNA measurement is 
\begin{align}
  (\#eq:T-dna)
  T_{rj} 
  &= A_{rj}' B^{(T)}_r \\
  &= A_{rj} B'_r C'_j B^{(T)}_r.
\end{align}
Substituting the second expression for $T_{rj}$ in \@ref(eq:A-hat-targeted) gives an expression for the estimate of the abundance of taxon $i$ from the targeted measurement,
\begin{align}
  (\#eq:A-hat-targeted-dna)
  \hat A_{ij} 
  &= \frac{M_{ij}}{M_{rj}} \cdot T_{rj} \\
  &= \frac{A_{ij} B_i C_j}{A_{rj} B_r C_j} \cdot A_{rj} B'_r C'_j B^{(T)}_r \\
  &= A_{ij} \cdot \frac{B_i B^{(T)}_r}{B''_r} \cdot C'_j,
\end{align}
where I have replaced $B_r / B'_r$ by $B''_r$ in the denominator.
Compared to the case of cellular measurement, the efficiency ratio for the reference taxon is just of the post-extraction steps, and we also have an additional term $C_j'$ that could vary across samples.

Question: Is that the same as the mean DNA extraction efficiency varies across samples?

## DNA spike-ins

HERE: Do same thing as above, and get result that

Suppose we add the spike-in taxon $r$ to sample $j$ in amount $A'_{rj}$ with error 
$B^{(T)}_r$, such that we believe the spike-in's abundance to be $T_{rj}$ in \@ref(eq:T-dna).
As with the cellular spike-in, the abundance of taxon $i$ can be estimated by \@ref(eq:A-hat-targeted) (for the targeted cellular method);
and as with the case of a targted DNA measurement above, the resulting estimate is given by \@ref(eq:A-hat-targeted-dna).

## Implications

These results show that, for both methods, the fold variation in the error across samples equals that in the scaling factor $C_j'$.
Ideally, $C_j'$ would be constant, corresponding to the situation where doubling the cell concentration in a sample (while leaving the taxonomic composition unchanged) results in double the DNA yield.
In this case, these DNA-based methods would work just as well with respect to bias not affecting fold-change estimates as their cell-based counterparts.

Perhaps even easier to see why this is the "ideal" situation (and perhaps also a common default assumption when thinking about microbiome experiments), is to think in terms of the factors $F'_{ij} = B_i' C_j'$.
Intuitively, we would like the factor increase for a taxon to be the same across samples; this situation corresponds to $C_j = c$ being constant across samples.
(The value of $c$ depends on the experimental details but also how we've set $B_j' = 1$.)

NOTE: Also need to consider whether it is true that the relative abundances are biased by a constant amount; and remember this is am important assumption I'm making as well, and we have some evidence that it can break down.
But the motivation for how I've done it is that it seems reasonable to worry about saturation in the extraction yields that is independent of taxa once we account for variation in extraction efficiency and genome size that is simply due to the elution step.
May need to give this motivation earlier.

### 2021-05-05 appendix-regression.Rmd
Fix broken refs

--- 2021-02-26 proofs.Rmd
@@
 \end{align}
-and let $\hat \alpha_0$, etc. denote the least-squares estimates for these equations supposing perfect information (i.e., that the $a_i$ and $B_i$ are known).
+and let $\hat \alpha_0$, etc. denote the least-squares estimates for the regression coefficients supposing perfect information (i.e., that the $a_i$ and $B_i$ are known).
 The coefficients for $\log m_i$ are related to those for $\log a_i$ and $\log \bar B$ through the equations
@@
   \hat \gamma_0 &= \hat \alpha_0 + \log B_i - \hat \beta_0 \\
-  \hat \gamma_1 &= \hat \alpha_1 - \hat \beta_1,
+  \hat \gamma_1 &= \hat \alpha_1 - \hat \beta_1.
 \end{align}
-so that the fitted model for $\log m_i$ is given by the right-hand side of
-\@ref(eq:expected-observed-abundance-log-linear-regression) with the coefficients replaced by the fitted values.
 ```

### 2021-05-05 relative-abundance.Rmd
Fix broken refs

@@
 It also simplifies the task of interpretation by limiting the number of taxa considered.
-In addition, some degree of uncontrolled aggregation is inevitable due to the inherent limitations of our sequencing and bioinformatics protocols to distinguish sufficiently similar organisms (@mclaren2019), which may be lumped together into an OTU or species-level feature.
+In addition, some degree of uncontrolled aggregation is inevitable due to the inherent limitations of our sequencing and bioinformatics protocols to distinguish sufficiently similar organisms (@mclaren2019cons), which may be lumped together into an OTU or species-level feature.
 Finally, though our focus is on analysis of taxa, we note that taxonomic aggregation also occurs (at least implicitly) in gene or function analyses, where counts are combined from different taxa that share the same genes or predicted functions.

### 2021-05-10 article.Rmd
Import article from Google Docs
Also includes editing to fix formatting as well as some minor text
improvements.

```{r vaginal-example-src, cache = TRUE}
#### Vaginal microbiome example
bias <- tibble(taxon = c("Lactobacillus", "Gardnerella", "Atopobium", "Spike-in"),
  efficiency = c(4.68, 0.16, 0.285, 1)) %>%
  mutate(across(efficiency, ~. / min(.)))
# efficiency associated with the targeted measurement of Lactobacillus
targeted_efficiency <- 2
a0 <- tribble(
  ~taxon, ~timepoint, ~abundance,
  "Lactobacillus", "T1", 5,
  "Lactobacillus", "T2", 0.5,
  "Gardnerella", "T1", 2,
  "Gardnerella", "T2", 8,
  "Atopobium", "T1", 3,
  "Atopobium", "T2", 1.5,
) %>%
  # Shrink abundance to have similar scale as proportions
  mutate(across(abundance, ~ . / max(.))) %>%
  left_join(bias, by = "taxon") %>%
  with_groups(timepoint, mutate,
    total_abundance = sum(abundance),
    proportion = close_elts(abundance),
    biased_abundance = efficiency * abundance,
    biased_proportion = close_elts(biased_abundance),
    biased_count = biased_proportion * 1e3,
    abundance_estimate_bulk = biased_proportion * total_abundance,
    #> abundance_estimate_spikein = biased_count / biased_count[taxon == "Spike-in"],
    abundance_estimate_targeted = biased_count * targeted_efficiency *
      abundance[taxon == "Lactobacillus"] / biased_count[taxon == "Lactobacillus"],
  )
a1 <- a0 %>%
  select(-total_abundance) %>%
  pivot_longer(-c(taxon, efficiency, timepoint), names_to = "type")
ptb <- a1 %>%
  filter(
    type %in% c("proportion", "biased_proportion"),
  ) %>%
  mutate(
    across(type, fct_relevel, "proportion", "biased_proportion"),
    across(type, fct_recode,
      "Actual" = "proportion",
      "Observed" = "biased_proportion"),
  )

## Panels showing the error in measurement and in differential abundance
shared_layers <- list(
  geom_path(aes(group = taxon), 
    arrow = grid::arrow(length = unit(0.15, "inches"))),
  geom_point(size = 2),
  scale_color_brewer(type = "qual", palette = 1, guide = FALSE),
    # guide = guide_legend(reverse = TRUE)),
  labs(y = "Proportion", color = "Taxon"),
  scale_y_log10(),
  coord_cartesian(clip = "off"),
  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
  expand_limits(y = 1e-2),
  theme(plot.margin = unit(c(0, 1.3, 0, 0), units = "in"))
)
# How much to nudge the taxon labels and proportions
nudge.taxon <- 0.48
nudge.prop <- 0.24
# In future iterations, consider labelling the taxa in both facets
p.meas <- ptb %>%
  ggplot(aes(type, value, color = taxon)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  shared_layers +
  geom_text(data = ~filter(., type == "Actual"), 
    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
  geom_text(data = ~filter(., type == "Observed"), 
    aes(label = round(value, 2)), nudge_x = nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"), 
    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
  labs(
    x = "Type",
    title = "Measurement error at each time point"
  )
p.fc <- ptb %>%
  ggplot(aes(timepoint, value, color = taxon)) +
  facet_wrap(~type, nrow = 1, scales = "fixed") +
  shared_layers +
  geom_text(data = ~filter(., timepoint == "T1"), 
    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2"), 
    aes(label = round(value, 2)), nudge_x = nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"), 
    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
  labs(
    x = "Time point",
    title = "Actual and observed fold changes"
  )
## Panel showing the efficiencies of individual taxa and the sample means
sme <- a0 %>%
  with_groups(timepoint, summarize, mean_efficiency = sum(proportion * efficiency))
sme0 <- sme %>%
  mutate(label = str_glue("mean ({timepoint})")) %>%
  select(label, efficiency = mean_efficiency)
sme1 <- sme %>%
  mutate(taxon = "Mean", type = "Mean") %>%
  select(taxon, efficiency = mean_efficiency, timepoint, type)
bias1 <- bias %>% 
  filter(taxon != "Spike-in") %>% 
  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
  mutate(type = "Taxon")
lvls = c("Atopobium", "Gardnerella", "Lactobacillus", "Mean")
etb1 <- bind_rows(bias1, sme1) %>%
  mutate(across(taxon, factor, levels = lvls))

lvls = c("Gardnerella", "Atopobium", "Lactobacillus")
bias2 <- bias %>% 
  filter(taxon != "Spike-in") %>% 
  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
  mutate(
    type = "Taxon",
    across(taxon, factor, levels = lvls),
    x = as.integer(taxon)
  ) %>%
  left_join(sme, by = "timepoint")

clrs <- c(RColorBrewer::brewer.pal(n = 3, "Accent"), rep("#585858", 2))
p.eff <- bias2 %>%
  ggplot(aes(x = x, y = efficiency, color = taxon)) + 
  geom_point(size = 2) +
  geom_text(data = ~filter(., timepoint == "T2"),
    aes(label = taxon), x = 3.3, hjust = 0) +
  geom_segment(aes(xend = x, yend  = efficiency, y = mean_efficiency),
    arrow = grid::arrow(length = unit(0.15, "inches"))) +
  geom_segment(data = sme, 
    aes(x = 1, xend = 3, y = mean_efficiency, yend = mean_efficiency), 
    color = "#585858",
    inherit.aes = FALSE) +
  geom_text(data = sme0 %>% mutate(timepoint = "T2"),
    aes(label = label, y = efficiency), x = 3.3, hjust = 0,
    color = "#585858",
    inherit.aes = FALSE) +
  scale_color_manual(values = clrs) +
  labs(y = "Relative efficiency", x = NULL, color = "Taxon", 
    title = "Taxonomic bias of protocol"
  ) +
  scale_y_log10(limits = c(1, 100) / 2,
    breaks = etb1$efficiency, 
    labels = signif(etb1$efficiency, 2)
  ) +
  xlim(c(0.0, 4)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  coord_cartesian(clip = "off") +
  theme(
    axis.line.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none",
    plot.margin = unit(c(0, 1.0, 0, 0), units = "in")
  )
```

# Introduction

Most microbiome research aims to go beyond qualitative descriptions to make quantitative conclusions about associations between specific microbes or community states and key host or environmental properties.
But there is serious disagreement about whether the microbiome measurements made by marker-gene and metagenomic sequencing (MGS) are or can ever be truly quantitative. A primary reason is that taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
<!-- TODO: Possible references: Mallick et al., 2017 DOI 10.1186/s13059-017-1359-z; Barlow et al. 2020 https://doi.org/10.1038/s41467-020-16224-6 -->
As a result taxa can differ dramatically (e.g. 10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—making the relative abundances we observe by MGS inaccurate representations of the actual sample compositions.
Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads among taxa make this taxonomic bias a feature of shotgun metagenomic measurements as much as marker-gene measurements.
Taxonomic bias is protocol specific (@mclaren2019cons) and can even vary among batches within an experiment (@yeh2018taxo); it thus not only makes MGS measurements inaccurate, but also incomparable between studies.
Taxonomic bias thereby poses a major challenge for any seeking to draw quantitative conclusions from MGS studies.

The field's current answer to this challenge is methods standardization, prioritizing consistency over accuracy.
By treating each sample within a study in exactly the same manner, we can ensure that observed differences between samples are not simply due to methodological differences.
Yet standardization does not guarantee that these observed differences reflect the actual differences in taxonomic composition among samples.
Consider what might be the simplest and most common inference made from MGS: How does the proportion of a taxon vary between samples? An observation that *Escherichia* coli makes up 10% of the sequencing reads in a fecal sample might, thanks to bias, have arisen from a sample in which the proportion of *E. coli* is actually 50% or 1%.
Yet many hope that if we observe the proportion of *E. coli* to increase from 10% to 20% between two samples, its proportion must have actually doubled—by fixing the taxonomic bias of our measurements through standardization, we have (so the argument goes) fixed the measurement error in both samples so that it will cancel when estimate the fold change.
Unfortunately, the results of @mclaren2019cons show that reality is not so neat.
Because the taxa compete for sequencing effort within a sample, the effect that taxonomic bias has on the observed proportions depends on the taxonomic composition of the sample.
As a result, MGS measurements can lead to spurious inferences about changes across samples even when taxonomic bias is constant for each taxon across samples.

This problem posed by taxonomic bias intersects with a second, growing concern within the microbiome field: that the relative abundances measured by MGS can present a misleading view of absolute-abundance dynamics.
**Absolute abundances are more relevant for certain biological questions such as...**
<!-- TODO: Give sentence or two motivating the measurement of absolute abundances -->
This concern has led to a recent rise in methods for supplementing MGS measurements to allow inference of absolute abundances.
<!-- Cite various approaches -->
The most popular approach is to multiply the proportions from MGS by an independent measurement of total microbial abundance.
<!-- studies to cite using this approach: in environmental microbiology (Props2017, Kevorkian2018, Lloyd2020) but also increasingly in human microbiome studies (TettamantiBoshier2020,Contijoch2019,Vandeputte2017,Vieira-Silva2019). -->
Multiplying inaccurate proportions by a measurement of cell concentration results in inaccurate absolute abundances that are subject to the same limitations on the accuracy of fold changes as standard MGS proportions alone.

Here we summarize recent advances in our understanding of how taxonomic bias affects our ability to accurately determine how taxa vary in relative and absolute abundance across samples from different conditions—the inference problem known as *differential abundance (DA) analysis*.
We first explain how taxonomic bias can lead to spurious DA results and consider the biological conditions in which such spurious results are likely.
But we then describe several experimental and computational methods that can be used to avoid or correct these errors.
Collectively, these methods may provide practical solutions to mitigating the effects of taxonomic bias on DA analysis for a majority of microbiome studies.

# Bias can cause spurious results in proportion-based DA analysis

```{r vaginal-example, fig.dim = c(8, 6) * 1.5, fig.cap = '(ref:vaginal-example)', dependson = "vaginal-example-src"}
p.meas + p.eff + p.fc + plot_spacer() + 
  plot_layout(byrow = TRUE, ncol = 2, widths = c(1, 0.6)) +
  plot_annotation(tag_levels = "A")
```

(ref:vaginal-example) **Taxonomic bias can distort differential abundance results even when it is consistent for each taxon across samples.** Panel A shows the actual and observed proportions for hypothetical community samples from two time points, which differ in their relative abundance of three taxa. Panel B shows taxonomic bias in terms of the relative efficiencies of the three taxa against the mean efficiency of each sample; the difference between the taxon's efficiency and the sample's mean (vertical arrows) determines the fold error seen in Panel A. Panel C rearranges the plot from Panel A to show the actual and observed fold changes between time points. The efficiencies of individual taxa were estimated by @mclaren2019cons from mock community data from @brooks2015thet. The abundances are hypothetical but inspired by observations from the human vaginal microbiome; see main text.

We illustrate the problems posed by taxonomic bias for differential abundance using an example where the bias of an MGS protocol was explicitly measured using mock communities.
To analyze the effects of taxonomic bias on measurements of the human vaginal microbiome, @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven key vaginal species using a similar protocol to that of the Vaginal Human Microbiome Project (VaHMP).
@mclaren2019cons used this data to estimate the relative measurement efficiency of the seven species.
They found that the most efficiently measured species, *Lactobacillus iners*, had an efficiency that was 29X that of the least efficiently measured species, *Gardnerella vaginalis*, primarily due to a greater DNA extraction efficiency and 16S copy number.
A third species, *Atopobium vaginae*, had an efficiency 1.8X that of *G. vaginalis*.
A common aim of human vaginal microbiome studies, including the VaHMP, is to understand the causes and consequences of shifts in communities from being dominated by a particular *Lactobacillus* species to dominance by a non-*Lactobacillus* species such as *G. vaginalis*, which have been associated with deleterious health outcomes including an increased risk of preterm births in pregnant women.
It is therefore important to know whether the efficiency variation found by @mclaren2019cons affects the observed differential abundances between *Lactobacillus*-dominated and *Gardnerella*-dominated communities.

Figure \@ref(fig:vaginal-example) illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by *Lactobacillus* to being dominated by *Gardnerella*.
In each case, we observe a greater proportion of *Lactobacillus* than is truly present, and less of the other two species (Figure \@ref(fig:vaginal-example)A).
This positive multiplicative error arises because *Lactobacillus* is more efficiently measured than the average species in each sample.
@mclaren2019cons showed that the observed and actual proportions of a taxon $i$ in a sample $s$ are related by the equation
\begin{align}
  (\#eq:error)
  \text{observed}_{i}(s) = \text{actual}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
\end{align}
where $\text{mean efficiency}(s)=\text{actual}_j(s)*\text{efficiency}_j$ in the denominator is the average efficiency of all taxa in that sample.
Intuitively, Equation \@ref(eq:error) says that the taxon will be overrepresented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
Figure \@ref(fig:vaginal-example)B shows the efficiencies of the three taxa and the mean efficiency in each sample.
The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of *Lactobacillus*.
The error in the second sample is larger — specifically, by 6X — than the first sample because the high proportion of *Gardnerella* has reduced the sample mean efficiency (and thus increased the efficiency of *Lactobacillus* relative to the mean) by 6X.

The enrichment depends on the taxon’s efficiency relative to the sample average, rather than the efficiency itself, because the requirement that the proportions of all taxa sum to one locks the taxa in a zero-sum competition.
Notably, competition occurs (though in a less absolute form) even at the level of the raw sequencing reads: DNA extractions and PCR reactions have limited output capacity; high-concentration samples are typically subsampled to fixed, lower concentrations; and sequencing libraries are given limited sequencing effort; so that the experimental process itself typically imposes competition to be sequenced.
And it is competition that prevents the effects of taxonomic bias from canceling when we analyze how the proportions of individual taxa change across samples.

Figure \@ref(fig:vaginal-example)C shows the consequences of measurement error on an analysis of the fold change in proportions between the two samples.
In reality, *Gardnerella* changes by 4X, *Lactobacillus* by 0.1X, and *Atopobium* by 0.5X.
Yet what is observed is that *Gardnerella* changes by 32X, *Lactobacillus* by 0.6X, and *Atopobium* by 3X.
In other words, the change in each case is multiplied by 6X, causing a much larger increase in *Gardnerella* and a smaller decrease *Lactobacillus* than actual, and the appearance that *Atopobium* increased when it in fact decreased.
Why do such errors occur?
The multiplicative error in Equation \@ref(eq:error) has two parts: a taxon-specific term in the numerator that is the same in each sample, and a taxon-independent term in the denominator that depends on the composition of the sample through the mean efficiency.
When we compute the fold change in a taxon i between samples s and t, 
\begin{align}
  \frac{\text{observed}_{i}(t)}{\text{observed}_{i}(s)}=\frac{\text{actual}_{i}(t)}{\text{actual}_{i}(s)}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{i}}*\frac{\text{mean efficiency}(s)}{\text{mean efficiency}(t)}
\end{align}
the constant efficiency in the numerator of Equation \@ref(eq:error) cancels, but the varying mean efficiency does not, leaving an error that is equal to the inverse change in mean efficiency.
As a result, we are here left inferring that the fold-changes in proportions have all increased less (or decreased more) than they truly have.
In our example, the mean efficiency decreased by 6X (from 15 to 2.5) due to the shift in dominant taxon from the high-efficiency *Lactobacillus* to the low-efficiency *Gardnerella*, causing the observed fold changes to increase by 6X above the actual.
This increase distorted the magnitudes of the changes in *Lactobacillus* and *Gardnerella* (decreasing the first and increasing the second) and changed the sign of *Atopobium* from negative to positive.
Thus partial cancelling in the effect of a constant multiplicative bias may not be sufficient to protect against spurious results.

Spurious results can occur for the same basic reason in proportion-based absolute-abundance inference.
Suppose that the true total abundance in each sample can be determined experimentally.
Because the absolute abundance estimate simply multiplies the erroneous observed proportions by this total, the estimated change in absolute abundance is wrong by the same factor as the fold change in proportions.
In our Figure \@ref(fig:vaginal-example) example, if the change in total abundance between samples were 2X, then we would observed both *Lactobacillus* and *Atopobium* to have increased in absolute abundance (by 1.2X and 6X), when in fact *Lactobacillus* decreased (by 5X) and *Atopobium* remained the same.

## Effect on regression of multiple samples

We showed how fold changes of individual samples may be unreliable.
But microbiome analyses often do not interpret such individual differences, and instead look for average patterns across many samples, such as how the abundance of a taxon varies with some covariate of interest.
The covariate may be discrete, such as whether the sample is from a healthy or sick person, or it may be continuous, such as a measure of space, time, or temperature.
Often these analyses can be framed as a regression problem.
For example, we might hypothesize that the log absolute abundance of taxon $i$ changes with a variable $x$ according to the simple linear regression, 
\begin{align}
  E[\text{log abundance}_{i} \mid x] = a_{0} + a_{1}x
\end{align}
where $x$ is either continuous (e.g., sediment depth) or binary (e.g., $x=1$ for treated patients and $x=0$ for controls).
Much of the effect of bias will be absorbed by the model intercept, which is typically not of interest to the researcher.
However, variation in the mean efficiency creates additional variation (reducing statistical power to detect variation in the slope) and can also create a statistical bias in our slope estimate equal to the inverse slope of $E[\text{log mean efficiency}\mid x]$ as a function of $x$.
Therefore if the mean efficiency varies across samples, but is not associated with the covariate $x$, its effect may simply be to create larger standard errors in our slope estimate.
But the larger worry may be the systematically distorted slope ($a_1$) estimates that arise in situations where the mean efficiency changes with $x$.
Notably, the error in an absolute sense is the same for all taxa—the slope is always reduced by the same amount—but the implications differ for various taxa depending on their slopes, causing magnitude or sign errors depending how the coefficient derived from the (unknown) true abundances compares to that of the mean efficiency (Figure \@ref(fig:regression-example)).

```{r regression-example, fig.cap = '(ref:regression-example)'}
svg_path <- here::here("figures/leopold2020host-example.svg")
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  cowplot::ggdraw() + cowplot::draw_image(svg_path)
}
rm(svg_path)
```

(ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.** Details to synthesize into a caption: Regression of log2(Proportion) of the commensal fungus Penicillium versus timepoint; timepoints 1 and 2 are pre- and post-challenge with the pathogen Melampsora. Data is split by the region the host plants are derived from (Eastern and Western US). Calibrated proportions = Observed proportions in the real, experimental samples after adjustment for the bias measured in mock communities. Mean efficiency of each community is inferred by treating the calibrated proportions as the truth, and multiplying by efficiencies estimated from the mocks. The pathogen Melampsora has a high measurement efficiency; thus once it infects the plants, the mean efficiency of the sample increases (purple points). Efficiency is here taken as relative to the focal taxon Penicillium. West plants tend to be more resistant to the pathogen, which likely explains why the mean efficiency doesn't increase as much in the West plants. Penicillium is observed to decrease in log proportion, in both the East and West plants. But the calibrated measurements show that it actually slightly increases in the East plants, and has a lesser decrease in the West plants than what was observed before bias correction. The difference between the Calibrated and Observed data points and regression lines equals the regression line of the mean efficiency: Orange = Green + Purple; Green = Orange - Purple. The absolute error in regression coefficients is the same for all taxa. I picked Penicillium for illustration since it has the smallest observed decrease, which makes the error due to bias have a particularly significant impact.

# Implications for real-world inference

These observations show that the effects of taxonomic bias do in fact cause error in proportion-based differential abundance analyses.
Yet if the mean efficiency varies much less across samples than the abundances of individual taxa, this error will be negligible.
Alternatively, if variation in the mean efficiency is effectively random (not associated with regression covariates), the effect of bias may be to increase the residual variance and thus reduce our statistical power to assay differential abundance—unfortunate to be sure, but perhaps not as much as confidently inferring changes with the wrong sign or magnitude.
But there are salient biological scenarios where substantial associations between the mean efficiency and covariate can in fact lead to spurious differential abundance inference.

## Scenarios in which bias may cause spurious results

Figure \@ref(fig:vaginal-example) illustrates how a change in dominance by from a high- to low- efficiency taxon (or vice versa) can easily create spurious DA results.
Indeed, this scenario may be particularly common in vaginal microbiome studies.
Human microbiomes often dominated by *Lactobacillus*, but can become dominated by non-*Lactobacillus* species such as *Gardnerella* vaginalis and *Atopobium vaginae*, and these low-*Lactobacillus* states are associated with the disease bacterial vaginosis and, in pregnant women, an increased risk of preterm birth.
Using mock community measurements from @brooks2015thet, @mclaren2019cons showed that the efficiency of two common *Lactobacillus* species were 20-30X greater than *G. vaginalis* and *A. vaginae*.
Therefore we should expect samples dominated by *Lactobacillus* to have a much larger mean efficiency than those those dominated by *G. vaginalis* and *A. vaginae*, which would distort the observed associations of all taxa with any covariate that is also associated with *Lactobacillus* dominance.
The extraction and 16S sequencing protocol used by @brooks2015thet is similar to that used in the VaHMP MOMs-PI study also led by these researchers to investigate the associations of these taxa with preterm birth (@fettweis2019thev), demonstrating the potential for serious real-world consequences for these statistical/technical concerns.

Similar dynamics occur in a plant-fungal interactions experiment performed by @leopold2020host (Figure \@ref(fig:regression-example)).
The authors inoculated commensally-colonized trees with a fungal pathogen.
Using DNA mock communities, they showed that (excluding bias from DNA extraction) the pathogen was 10X more efficiently measured than the median commensal and 40X more efficiently measured than the lowest-efficiency commensal.
In most hosts, the pathogen increased rapidly between timepoints, driving an increase in the sample mean efficiency that in turn leads to larger decreases being observed in the proportions of commensal taxa than what is predicted using calibrated (bias-corrected) values (REF APPENDIX).
For example, the calibrated proportions indicate that the commensal *Penicillium* slightly increased in hosts from the Eastern US and slightly decreased in hosts from the Western US.
Yet the observed (uncalibrated) proportions show moderate and large decreases, respectively (Figure \@ref(fig:regression-example) above).

These examples involved scenarios where individual taxa can constitute a large proportion of the community.
But in samples from highly-diverse ecosystems such as human gut, soil, and ocean sediment, one species rarely dominates.
In these cases it might seem intuitive that the mean efficiency should be relatively stable across samples: As diversity increases, the mean efficiency effectively averages over efficiencies from a greater number of taxa and so (by the central limit theorem) will converge to a constant value—if taxon efficiencies are statistically independent of taxon abundances (REF APPENDIX).
In practice, associations between efficiencies and abundances may be common and can allow large shifts in mean efficiency to occur even in highly diverse samples.

The human gut provides a clear example of how such associations might arise.
Gut microbiomes typically have high diversity at the species level but are dominated by just a small number of phyla, and two in particular: The Bacteroidetes and the Firmicutes.
The ratio of Bacteroidetes to Firmicutes can shift substantially between individuals and has been linked to a number of host traits and health conditions.
In addition, many DNA extraction protocols more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though there can also be significant variation in extraction efficiency among species within these phyla (@mclaren2019cons).
For such protocols, we should expect Bacteroidetes-dominated samples to have substantially larger mean efficiencies than Firmicutes-dominated samples, regardless of species-level diversity.
This association of efficiency with phylum abundance will distort DA inferences at any taxonomic level if the covariate is also associated with these phyla.

This example illustrates a general mechanism by which association between the abundances and efficiencies of taxa are created by the common influence of evolutionary history.
Just as shared evolutionary history creates positive associations in ecological traits that drives positive correlations in how related taxa vary in abundance across samples, so does it create positive associations in bias-related traits that can lead to similar efficiencies among closely related taxa.
Such phylogenically-associated, bias-affecting traits include cell-wall toughness, ribosomal-operon copy number, binding affinity for a given set of primers, and representativeness in taxonomic databases.
Ecology and efficiency are both affected by the same evolutionary history, leading to positive associations between the two: A change in diet that increases the relative abundances of many Bacteroidetes species and in so doing also increases the relative abundance of many easy-to-lyse species.

Associations between measurement efficiency and relative abundance can also arise because a single trait affects both.
For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
@lloyd2020evid estimate the log fold change in the estimated absolute abundance of various taxa with sediment depth (as a proxy for time) to determine which taxa are able to persist and even grow in this difficult environment.
It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
As the relative abundance of tougher species increases with depth, the mean extraction efficiency decreases.
This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.

### Error in absolute abundance measurements

So far we have ignored error associated with our absolute abundance measurements.
In fact, these measurements can also have a tendency to measure contributions from some taxa more efficiently than others.
**TODO: State how this could create problems.
Might also note the ambiguity in what the truth/target is (e.g.
live cells, or all cells; cells, or 16S rRNA gene copies; observations in Jian, Salonen, and Korpela 2021).**
Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
Yet these biases are shared by the sequencing measurement.
Though they make qPCR measurements a bad proxy for total cell density, when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can cancel, leaving relative accurate fold changes.
(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.
Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.

## We cannot simply ignore taxonomic bias

These examples show that there is potential for bias to distort DA results in both low and high diversity settings.
We therefore need methods to control or correct for bias to ensure the validity of future results, as well as methods to probe the robustness of past results.
It is possible that spurious results are unlikely in many experimental contexts.
By applying such methods across a range of biological and experimental contexts we will deepen our understanding of when bias is unlikely to significantly distort biological findings and when additional measures are needed to mitigate its effects.

# Potential solutions

## Using control samples to calibrate relative abundances

In principle, control ("mock") communities containing representative taxa from the environment of interest can be used to directly estimate the measurement efficiencies and correct bias in the MGS measurements ("calibration").
Such an approach may be sufficient for synthetic-community experiments, where all taxa are culturable, and for relatively simple natural communities like the vaginal microbiome that are dominated by a small number of culturable taxa.
But suitable mock controls may not be feasible for most complex natural ecosystems and require significant effort to develop.

An closely-related alternative to mocks are controls derivedd from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (@mclaren2019cons).
A natural fecal standard is currently being developed by [NIST](https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material) and one has recently been made available commercially by Zymo Research ([ZymoBIOMICS Fecal Reference](https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology)).
Ensuring the stability and homogeneity of such control samples can be challenging.
Careful testing is also needed to ensure that the preparation and storage of the controls has not significantly affected the taxonomic bias relative to the experimental samples in a given application.
As it is feasible to characterize a single standard much more extensively than a typical community sample, we may be able to obtain an estimated composition we feel comfortable treating as the ground truth. 
Yet even when this is not possible, such natural standards can allow us to reconcile results across studies despite not knowing the truth.

## Ratio-based relative and absolute abundance inference

A comprehensive approach of estimating the efficiency of all taxa is needed for getting calibrated relative abundances of all taxa within individual samples.
But what if we only want calibrated DA analysis?
That is, we are comfortable with not knowing whether *E. coli* is 10% of our sample if we can confidently determine that it doubled or halved between sample conditions?
This problem is easier to solve, and one solution is to use analysis methods based on the ratios among taxa rather than the proportions of individual taxa.

A subset of methods for analyzing differential relative abundance are derived from the field of Compositional Data Analysis (CoDA).
The defining feature of CoDA methods is that they are based on fold-changes in ratios among elements (here, the taxa) (REF AITCHISON).
The use of CoDA methods in microbiome analysis has largely been motivated by concerns over the negative correlations between taxa induced by the sum-to-one constraint in taxon proportions.
But @mclaren2019cons showed that, due to their property of perturbation invariance (REF AITCHISON), the results of CoDA differential relative abundance analysis are invariant to consistent taxonomic bias.
From Equation \@ref(eq:error), the observed ratio of a taxon $i$ to a taxon $j$ in a sample $s$ is
\begin{align}
  \frac{\text{observed}_{i}}{\text{observed}_{j}}=\frac{\text{actual}_{i}}{\text{actual}_{j}}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}
\end{align}
The error is independent of the sample composition and thus cancels in any function that is a fold-change of this (or any) ratio of taxa between samples.
(Perhaps insert example from Figure \@ref(fig:vaginal-example)ABC all giving same ratio FCs.)

Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
To estimate the abundance of a focal taxon $i$, this approach multiplies the ratio of reads (or proportions) of taxon i to a reference taxon $r$ by a known or estimated abundance of the reference taxon, (while ignoring bias)
\begin{align}
  \text{estimate}_{i} = \frac{\text{observed}_{i}}{\text{observed}_{r}}*\text{abundance}_{r}
\end{align}
So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
Yet the reference taxon could also be a naturally occurring taxon whose absolute abundance we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.
Because taxa do not compete (or competition is greatly reduced) in such targeted measurements, any taxonomic bias associated with them is expected to create constant multiplicative error across samples and so not affect fold change estimates [APPENDIX].
APPENDIX describe various theoretical and experimental considerations for both approaches.

## Estimating error with targeted measurements of a small number of taxa

Because targeted absolute-abundance measurements are expected to provide (relatively) accurate  estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
We illustrate the basic idea using the @lloyd2020evid experiment described above.
Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.
Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
@lloyd2020evid found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].

## Computational approaches

What can we do for experiments that have already been conducted without mock controls, targeted control measurements, or spike-ins?
At least two purely computational approaches can still be used.

**Bias sensitivity analysis:**
A straightforward and universally available approach is to use computer simulation to determine how the results change under a range of possible sets of efficiencies, which can be randomly generated to reflect certain hypotheses about bias in the given system.
The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—not just for differential absolute abundance, but for all microbiome analyses.

**Bias-aware meta-analysis:**
When the goal is to perform a meta-analysis that combines studies that have used different protocols, the unknown measurement efficiencies of each protocol can be explicitly included as parameters of the statistical model that is used.
Unknown efficiencies can be included in "compositional" linear modeling frameworks such as ALDEx2, DivNet, and fido simply by adding a protocol-specific term to the linear model of taxon log ratios.
Thus such bias-aware meta-analyses are already technically feasible and—if bias is truly consistent within a protocol or study—may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not explicitly model bias.

### 2021-05-10 _common.R

@@ -5,15 +5,20 @@ knitr::opts_chunk$set(
   collapse = TRUE,
   cache = FALSE,
   echo = FALSE,
-  out.width = "70%",
-  fig.align = 'center',
-  fig.width = 6,
-  fig.asp = 0.618,
+  # out.width = "70%",
+  # fig.align = 'center',
+  # fig.width = 6,
+  # fig.asp = 0.618,
   fig.show = "hold"
 )
 
 options(dplyr.print_min = 6, dplyr.print_max = 6)
 options(crayon.enabled = FALSE)
 
-# library(tidyverse)
+library(tidyverse)
 library(here)
+library(cowplot)
+library(patchwork)
+theme_set(theme_cowplot(12))
+
+close_elts <- function(x) x / sum(x)

### 2021-05-10 notebook/vaginal-microbiome-example.Rmd
Import Rmd developing an example
This Rmd doc develops an hypothetical example in which bias leads to a
spurious differential abundance result. It is based on a real case of
bias measured among three species of vaginally-associated bacteria.

```{r}
set.seed(42)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  echo = FALSE
  #> out.width = "70%",
  #> fig.align = 'center',
  #> fig.width = 6,
  #> fig.asp = 0.618,
  #> fig.show = "hold"
)

#> options(dplyr.print_min = 6, dplyr.print_max = 6)
#> options(crayon.enabled = FALSE)

library(tidyverse)
library(here)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot(12))

close_elts <- function(x) x / sum(x)
```

```{r}
bias <- tibble(taxon = c("Lactobacillus", "Gardnerella", "Atopobium", "Spike-in"),
  efficiency = c(4.68, 0.16, 0.285, 1)) %>%
  mutate(across(efficiency, ~. / min(.)))
# efficiency associated with the targeted measurement of Lactobacillus
targeted_efficiency <- 2
a0 <- tribble(
  ~taxon, ~timepoint, ~abundance,
  #> "Lactobacillus", "T1", 5,
  #> "Lactobacillus", "T2", 1,
  #> "Gardnerella", "T1", 1,
  #> "Gardnerella", "T2", 10,
  #> "Atopobium", "T1", 3,
  #> "Atopobium", "T2", 1.5,
  "Lactobacillus", "T1", 5,
  "Lactobacillus", "T2", 0.5,
  "Gardnerella", "T1", 2,
  "Gardnerella", "T2", 8,
  "Atopobium", "T1", 3,
  "Atopobium", "T2", 1.5,
) %>%
  # Shrink abundance to have similar scale as proportions
  mutate(across(abundance, ~ . / max(.))) %>%
  left_join(bias, by = "taxon") %>%
  with_groups(timepoint, mutate,
    total_abundance = sum(abundance),
    proportion = close_elts(abundance),
    biased_abundance = efficiency * abundance,
    biased_proportion = close_elts(biased_abundance),
    biased_count = biased_proportion * 1e3,
    abundance_estimate_bulk = biased_proportion * total_abundance,
    #> abundance_estimate_spikein = biased_count / biased_count[taxon == "Spike-in"],
    abundance_estimate_targeted = biased_count * targeted_efficiency *
      abundance[taxon == "Lactobacillus"] / biased_count[taxon == "Lactobacillus"],
  )
a1 <- a0 %>%
  select(-total_abundance) %>%
  pivot_longer(-c(taxon, efficiency, timepoint), names_to = "type")
#> what are the fold changes?
a1 %>% 
  pivot_wider(names_from = timepoint) %>%
  mutate(fc = T2 / T1) %>%
  #> filter(str_detect(type, "proportion")) %>%
  arrange(type, desc(taxon)) %>%
  print(n = Inf)

a1 %>% 
  with_groups(c(type, timepoint), summarize, across(value, sum)) %>%
  pivot_wider(names_from = timepoint) %>%
  mutate(fc = T2 / T1)
```

For now, focus on plotting the proportions. 
Show the error in the top panel: Actual -> Observed, for each time point. 
Show the fold changes in the bottom panel: T1 -> T1, for both proportion types.

```{r}
ptb <- a1 %>%
  filter(
    type %in% c("proportion", "biased_proportion"),
  ) %>%
  mutate(
    across(type, fct_relevel, "proportion", "biased_proportion"),
    across(type, fct_recode,
      "Actual" = "proportion",
      "Observed" = "biased_proportion"),
  )

shared_layers <- list(
  geom_path(aes(group = taxon), 
    arrow = grid::arrow(length = unit(0.15, "inches"))),
  geom_point(size = 2),
  scale_color_brewer(type = "qual", palette = 1, guide = FALSE),
    # guide = guide_legend(reverse = TRUE)),
  labs(y = "Proportion", color = "Taxon"),
  scale_y_log10(),
  coord_cartesian(clip = "off"),
  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
  expand_limits(y = 1e-2),
  theme(plot.margin = unit(c(0, 1.3, 0, 0), units = "in"))
)

# How much to nudge the taxon labels and proportions
nudge.taxon <- 0.48
nudge.prop <- 0.24
# In future iterations, consider labelling the taxa in both facets

p1 <- ptb %>%
  ggplot(aes(type, value, color = taxon)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  shared_layers +
  geom_text(data = ~filter(., type == "Actual"), 
    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
  geom_text(data = ~filter(., type == "Observed"), 
    aes(label = round(value, 2)), nudge_x = nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"), 
    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
  labs(
    x = "Type",
    title = "Measurement error at each time point"
  )
p2 <- ptb %>%
  ggplot(aes(timepoint, value, color = taxon)) +
  facet_wrap(~type, nrow = 1, scales = "fixed") +
  shared_layers +
  geom_text(data = ~filter(., timepoint == "T1"), 
    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2"), 
    aes(label = round(value, 2)), nudge_x = nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"), 
    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
  labs(
    x = "Time point",
    title = "Actual and observed fold changes"
  )
p1 / p2
```

In addition, I want a plot showing the efficiencies and mean efficiencies, so that they can be related to the error.

First, compute the mean efficiency, then join with the taxon efficiencies in a table for plotting.
```{r}
sme <- a0 %>%
  with_groups(timepoint, summarize, mean_efficiency = sum(proportion * efficiency))
sme0 <- sme %>%
  mutate(label = str_glue("mean ({timepoint})")) %>%
  select(label, efficiency = mean_efficiency)
lvls = c("Atopobium", "Gardnerella", "Lactobacillus", "mean (T1)", "mean (T2)")
etb <- bind_rows(
  bias %>% filter(taxon != "Spike-in") %>% mutate(label = taxon, type = "Taxon"),
  sme0 %>% mutate(type = "Mean")
) %>%
  mutate(across(label, factor, levels = lvls))
```

Plot the individual and mean efficiencies. 
Set the vertical span to be 100 to match the other plots.
```{r}
#> RColorBrewer::display.brewer.all(type = "qual")
#> clrs <- RColorBrewer::brewer.pal(n = 8, "Accent")[c(1:3, 7:8)]
clrs <- c(RColorBrewer::brewer.pal(n = 3, "Accent"), rep("#585858", 2))
p3 <- etb %>%
  ggplot(aes(x = 1, y = efficiency, color = label)) + 
  geom_path(data = ~filter(., type == "Mean"), group = 1, 
    color = "black", size = 0.3, x = 1,
    arrow = grid::arrow(length = unit(0.15, "inches"))) +
  geom_point(aes(size = type, shape = type)) +
  geom_text(aes(label = label), nudge_x = 0.3, size = 5, hjust = 0) +
  scale_shape_manual(values = c(95, 16), guide = FALSE) +
  scale_size_manual(values = c(10, 2), guide = FALSE) +
  #> scale_color_brewer(type = "qual", palette = 1, guide = FALSE) +
  scale_color_manual(values = clrs) +
  labs(y = "Relative efficiency", x = NULL, color = "Taxon", 
    title = "Taxonomic bias of protocol"
  ) +
  scale_y_log10(limits = c(1, 100) / 2,
    breaks = etb$efficiency, 
    #> labels = round(etb$efficiency, 1)
    labels = signif(etb$efficiency, 2)
    ) +
  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
  #> expand_limits(y = 1e-2) +
  xlim(c(0.5, 3)) +
  theme(
    axis.line.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none"
  )
p3
```

Create a version split by time point,
```{r}
sme1 <- sme %>%
  mutate(taxon = "Mean", type = "Mean") %>%
  select(taxon, efficiency = mean_efficiency, timepoint, type)
bias1 <- bias %>% 
  filter(taxon != "Spike-in") %>% 
  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
  mutate(type = "Taxon")
lvls = c("Atopobium", "Gardnerella", "Lactobacillus", "Mean")
etb1 <- bind_rows(bias1, sme1) %>%
  mutate(across(taxon, factor, levels = lvls))
```
```{r}
p4 <- etb1 %>%
  ggplot(aes(x = 1, y = efficiency, color = taxon)) + 
  geom_point(aes(size = type, shape = type)) +
  geom_text(aes(label = taxon), nudge_x = 0.3, size = 5, hjust = 0) +
  scale_shape_manual(values = c(95, 16), guide = FALSE) +
  scale_size_manual(values = c(10, 2), guide = FALSE) +
  #> scale_color_brewer(type = "qual", palette = 1, guide = FALSE) +
  scale_color_manual(values = clrs) +
  labs(y = "Relative efficiency", x = NULL, color = "Taxon", 
    title = "Taxonomic bias of protocol"
  ) +
  scale_y_log10(limits = c(1, 100) / 2,
    breaks = etb1$efficiency, 
    #> labels = round(etb$efficiency, 1)
    labels = signif(etb1$efficiency, 2)
    ) +
  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
  #> expand_limits(y = 1e-2) +
  xlim(c(0.5, 2)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  theme(
    axis.line.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none"
  )
p4
```

Try a different way.
```{r}
lvls = c("Gardnerella", "Atopobium", "Lactobacillus")
bias2 <- bias %>% 
  filter(taxon != "Spike-in") %>% 
  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
  mutate(
    type = "Taxon",
    across(taxon, factor, levels = lvls),
    x = as.integer(taxon)
  ) %>%
  left_join(sme, by = "timepoint")

```
```{r}
p4 <- bias2 %>%
  #> mutate(
  #>   across(timepoint, factor, 
  #>     levels = c("T1", "T2"), 
  #>     labels = str_c("Time point T", 1:2))
  #> ) %>%
  ggplot(aes(x = x, y = efficiency, color = taxon)) + 
  geom_point(size = 2) +
  geom_text(data = ~filter(., timepoint == "T2"),
    aes(label = taxon), x = 3.3, hjust = 0) +
  geom_segment(aes(xend = x, yend  = efficiency, y = mean_efficiency),
    arrow = grid::arrow(length = unit(0.15, "inches"))) +
  geom_segment(data = sme, 
    aes(x = 1, xend = 3, y = mean_efficiency, yend = mean_efficiency), 
    color = "#585858",
    inherit.aes = FALSE) +
  #> geom_text(data = sme0 %>% filter(timepoint == "T2"), 
  geom_text(data = sme0 %>% mutate(timepoint = "T2"),
    aes(label = label, y = efficiency), x = 3.3, hjust = 0,
    color = "#585858",
    inherit.aes = FALSE) +
  scale_color_manual(values = clrs) +
  labs(y = "Relative efficiency", x = NULL, color = "Taxon", 
    title = "Taxonomic bias of protocol"
  ) +
  scale_y_log10(limits = c(1, 100) / 2,
    breaks = etb1$efficiency, 
    labels = signif(etb1$efficiency, 2)
    ) +
  xlim(c(0.0, 4)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  coord_cartesian(clip = "off") +
  theme(
    axis.line.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none",
    plot.margin = unit(c(0, 1.0, 0, 0), units = "in")
  )
p4
```

### Multipanel

```{r}
p1 + p4 + p2 + plot_spacer() + 
  plot_layout(byrow = TRUE, ncol = 2, widths = c(1, 0.6)) +
  plot_annotation(tag_levels = "A")
```

```{r}
str_c("figures/example-vaginal-microbiome.", fmts) %>%
  here %>%
  walk(ggsave, units = "in", width = 8, height = 6, scale = 1.5)
```

This is a nice start.
The difference between a taxon's efficiency and the mean is the error for that sample.
It could make sense to show this with arrows.

Next:

- Experiment with ways to make the connections even clearer between the efficiency differences and the error. 
- Try making the arrows vertical, so that they are directly connected to the fold changes in proportion and in efficiency.
- Can also split the third plot into T1 and T2.
- Might consider changing "Time point T1" to "Sample S1", to fit the primary language of the text.
- Can mention in the text that these could be samples from different women, or the same woman at different times.
- also use a shape scale for the taxa, to give a second way besides color to distinguih them

## Notes about the example

In this example, the proportion of Atopobium is observed to increase; it actually decreases.
Since the total abundance increases, the total abundance of Atopobium is also incorrectly observed to increase.

### 2021-05-11 index.Rmd
Edit frontmatter
Update authors, title, and preface to account for the article

--- 2021-02-17 index.Rmd
@@
 ---
-title: "Characterizing and mitigating the effects of experimental bias on differential (absolute) abundance analysis in microbiome studies"
-author: "Michael R. McLaren"
+title: "Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for"
+author:
+  - Michael R. McLaren^[North Carolina State University; send correspondence to m.mclaren42@gmail.com]
+  - Karen G. Lloyd^[University of Tennessee]
+  - Benjamin J. Callahan^[North Carolina State University]
 date: "`r Sys.Date()`"
@@
 link-citations: yes
+# description:
 github-repo: mikemc/bda-theory
-description: "Mathematical theory of the effects of experimental bias on differential (relative and absolute) abundance analysis in microbiome studies."
 ---
 
-# Abstract
+# Preface {-}
 
-This monograph develops a mathematical theory of the effects of protocol- and taxon-specific experimental bias on various types of differential (relative and absolute) abundance analyses in microbiome studies and uses it to formulate several mitigation strategies.
+This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.
+We describe the basic problem posed by taxonomic bias for measuring changes in the abundance of particular taxa across conditions and describe new strategies for mitigating the errors it induces.
+Analyses of both relative and absolute abundances are considered.
+In its current form, the manuscript sits somewhere between a standard scientific article and a monograph; 
+it consists of an article followed by a series of appendices which together give a comprehensive treatment of the implications of the @mclaren2019cons model of taxonomic bias for differential-abundance analysis and experimental design.
+It is licensed under a [CC BY 4.0 License](https://creativecommons.org/licenses/by/4.0/).
+See [the Zenodo record](https://doi.org/10.5281/zenodo.4552717) for how to cite the latest version.

### 2021-05-12 index.Rmd
Improve pdf formatting
- Migrate all output options to the YAML block of index.Rmd to ensure
  all options are specified correctly and don't conflict; see links
  below
- Adjust pandoc options for pdf output

https://github.com/rstudio/bookdown/issues/918
https://github.com/rstudio/bookdown/issues/921

@@
 site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: book
 bibliography: [main.bib, packages.bib]
@@
 # description:
-github-repo: mikemc/bda-theory
+# For pandoc latex processing
+documentclass: article
+geometry:
+  - top=1in
+  - bottom=1in
+  - left=1.5in
+  - right=1.5in
+output:
+  bookdown::gitbook:
+    css: style.css
+    number_sections: true
+    config:
+      sharing:
+        facebook: no
+        github: yes
+        twitter: no
+        linkedin: no
+        weibo: no
+        instapaper: no
+        vk: no
+        all: ['facebook', 'twitter', 'linkedin', 'weibo', 'instapaper']
+  bookdown::pdf_book:
+    includes:
+      in_header: latex/preamble.tex
+    keep_tex: true
+    dev: "pdf"
+    papersize: letter
+    toc_depth: 2
+    toc_unnumbered: yes
+    toc_appendix: yes
 ---

### 2021-05-12 index.Rmd
Add "in progress" note

@@
 
+```{block, type = "rmdnote", echo = TRUE}
+**This in-progress manuscript is not intended for general scientific use.**
+It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
+Please post comments or questions on the [GitHub Issues page](https://github.com/mikemc/differential-abundance-theory/issues) or [email Mike](m.mclaren42@gmail.com).
+```
+
 This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.

### 2021-06-05 article.Rmd
Revise discussion of absolute abundance in intro

@@
 This problem posed by taxonomic bias intersects with a second, growing concern within the microbiome field: that the relative abundances measured by MGS can present a misleading view of absolute-abundance dynamics.
-**Absolute abundances are more relevant for certain biological questions such as...**
-<!-- TODO: Give sentence or two motivating the measurement of absolute abundances -->
-This concern has led to a recent rise in methods for supplementing MGS measurements to allow inference of absolute abundances.
-<!-- Cite various approaches -->
-The most popular approach is to multiply the proportions from MGS by an independent measurement of total microbial abundance.
-<!-- studies to cite using this approach: in environmental microbiology (Props2017, Kevorkian2018, Lloyd2020) but also increasingly in human microbiome studies (TettamantiBoshier2020,Contijoch2019,Vandeputte2017,Vieira-Silva2019).
--->
-Multiplying inaccurate proportions by a measurement of cell concentration results in inaccurate absolute abundances that are subject to the same limitations on the accuracy of fold changes as standard MGS proportions alone.
+Without making further assumptions (which are often difficult or impossible to verify) one cannot tell whether a doubling in the proportion of a taxon reflects an increase in its abundance or a decrease in the abundances of other taxa.
+This concern has motivated the development of various experimental methods to enable researchers to convert the relative abundances measured by MGS into absolute abundances (methods to be reviewed in the Appendix in a future version).
+An increasingly popular approach in the study of both environmental and host-associated microbiomes involves making an independent measurement of total microbial abundance in each sample (@props2017abso,@kevorkian2018esti,@lloyd2020evid,@tettamantiboshier2020comp,@contijoch2019gutm,@vandeputte2017quan,@vieirasilva2019quan).
+This estimate is then multiplied by the MGS proportions to convert them to estimates of absolute abundance.
+This simple calibration procedure directly transfers the error in the MGS proportions due to taxonomic bias to the estimated absolute abundances, leading to estimates that are not only individually inaccurate but can imply inaccurate variation in abundance across samples.
 
@@
 The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of *Lactobacillus*.
-The error in the second sample is larger — specifically, by 6X — than the first sample because the high proportion of *Gardnerella* has reduced the sample mean efficiency (and thus increased the efficiency of *Lactobacillus* relative to the mean) by 6X.
+The error in the second sample is larger—specifically, by 6X—than the first sample because the high proportion of *Gardnerella* has reduced the sample mean efficiency (and thus increased the efficiency of *Lactobacillus* relative to the mean) by 6X.
 

### 2021-06-24 models.Rmd
Put non-article chapters in the appendix
Fix typo

@@
+# (APPENDIX) Appendix {-}
+
 # Measurement models {#models}
@@
 
-To consider various addition information that can be used for absolute abundance inference, I let $B$ denote bulk abundance measurements (such as via flow cytometry or broad-range 16S qPCR), $S$ genome the known taxon abundances of a spike-in, and $T$ denote the abundances of a set of taxa made via targeted measurement.
+To consider various addition information that can be used for absolute abundance inference, I let $B$ denote bulk abundance measurements (such as via flow cytometry or broad-range 16S qPCR), $S$ denote the known taxon abundances of a spike-in, and $T$ denote the abundances of a set of taxa made via targeted measurement.
 I let $R \subset \{1, \dots, I\}$ denote the set of _reference taxa_ whose abundance is known independently from the metagenomics measurement (up to experimental bias factors) because they were spiked at a known density, subjected to a targeted measurement, or have been determined through some other means to have a constant or known abundance across samples.

### 2021-07-12 article.Rmd
Import comments from Google Doc

@@
 
+- On whether MGS can be quantitative
+    - Karen Lloyd: Is there a reference that explains the disagreement? In my field, it hasn't really been explicitly addressed adequately.
+    - Michael McLaren: I'll have to look/think about this some more.
+I was trying to express a vague impression I have based on various things I've read and heard, and how community sequencing is used in practice, that implicitly suggest there is disagreement or at least a tension in the field where on the one hand there are many papers on the problems with community sequencing and on the other hand many papers (often by the same authors, and including me and Ben) using it to make quantitative inferences
+    - Karen Lloyd: You could reference Mallick et al., 2017 DOI 10.1186/s13059-017-1359-z; Or Barlow et al. 2020 https://doi.org/10.1038/s41467-020-16224-6
+
 The field's current answer to this challenge is methods standardization, prioritizing consistency over accuracy.
@@
 As a result, MGS measurements can lead to spurious inferences about changes across samples even when taxonomic bias is constant for each taxon across samples.
+
+- 'many hope' - consider make small tweaks to phrases like this to sound less critical
 
@@
 
+## Notes for revision
+
+- introduce differential abundance as the focus earlier
+- improve introduction of absolute abundance vs. relative abundance, and methods for obtaining absolute abundances
+    + might include computational methods in scope, and use the term "calibration"; and ref to the future appendix that surveys the various methods
+    + consider giving general explanation and/or specific examples of why absolute abundances are important
 # Bias can cause spurious results in proportion-based DA analysis
@@
 
+- TODO: Add main-message sentence; see Ben's note below
+
 We illustrate the problems posed by taxonomic bias for differential abundance using an example where the bias of an MGS protocol was explicitly measured using mock communities.
@@
 
+- Ben flagged that we want to make sure that it is clear from the text that this example is not just a one-off case of bias mattering, but is an illustration of a general/universal problem.
+The later discussion helps with this already; but what will help more is to start this section with a main-message sentence stating the general problem first.
+
 Figure \@ref(fig:vaginal-example) illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by *Lactobacillus* to being dominated by *Gardnerella*.
 In each case, we observe a greater proportion of *Lactobacillus* than is truly present, and less of the other two species (Figure \@ref(fig:vaginal-example)A).
-This positive multiplicative error arises because *Lactobacillus* is more efficiently measured than the average species in each sample.
+This positive multiplicative error^[what is the right terminology for multiplicative error that is greater than 1?] arises because *Lactobacillus* is more efficiently measured than the average species in each sample.
 @mclaren2019cons showed that the observed and actual proportions of a taxon $i$ in a sample $s$ are related by the equation
@@
 
+- see note on terminology for mult. error
+
 The enrichment depends on the taxon’s efficiency relative to the sample average, rather than the efficiency itself, because the requirement that the proportions of all taxa sum to one locks the taxa in a zero-sum competition.
@@
 And it is competition that prevents the effects of taxonomic bias from canceling when we analyze how the proportions of individual taxa change across samples.
+
+- on the description of competition - Ben suggested adding "for reads in the sequencer", but this addition doesn't quite make sense the way I've written this sentence, which is talking about the mathematical operation of normalization to proportions.
+But this paragraph deserves more attention, as the fact that we are imposing competition  via normalization to proportions, but the experiment also imposes competition that we can't avoid by just not normalizing, is subtle
 
@@
 Thus partial cancelling in the effect of a constant multiplicative bias may not be sufficient to protect against spurious results.
+
+- TODO: Add strike-throughs to the above equation
+- Revision should hone the selection of equations
 
@@
 Notably, the error in an absolute sense is the same for all taxa—the slope is always reduced by the same amount—but the implications differ for various taxa depending on their slopes, causing magnitude or sign errors depending how the coefficient derived from the (unknown) true abundances compares to that of the mean efficiency (Figure \@ref(fig:regression-example)).
+
+- on "(e.g., sediment depth)": Want to confirm with KL that I'm understanding this correctly.
+She notes: "Even though these values are sequential, we use them as discrete variables because the samples are only from one depth each.
+I'm not sure if this distinction matters here." To which I replied: "right....you first converted sediment depth to time, and then regressed log abundance on time to estimate growth rate?
+(new line) Here I was just using sediment depth as an example variable, but I could replace with time to more match your analysis"
+- Consider different term than 'slope' since it's not a slope in the binary case
 
@@
 
+- Note about this figure: It is currently framed as O = A - \bar B to show the effect of error on inferred DA results; it could be flipped to A = O + \bar B to show how a correction might be done.
 # Implications for real-world inference
@@
 
+- Discussion of the vaginal example in the above paragraph should be distilled in conjunction with the discussion of this example in the previous section
+
 Similar dynamics occur in a plant-fungal interactions experiment performed by @leopold2020host (Figure \@ref(fig:regression-example)).
@@
 
+- In last sentence, insert the the values of the LFCs (ideally with evaluated R code)
+
 These examples involved scenarios where individual taxa can constitute a large proportion of the community.
@@
 In practice, associations between efficiencies and abundances may be common and can allow large shifts in mean efficiency to occur even in highly diverse samples.
+
+- on the example of 'ocean sediment' in the above P: I asked KL "is it reasonable to characterize the ocean sediment communities from your 2020 paper as "high diversity"?
+In particular, species- or OTU-level Inverse Simpson values are what's relevant here.
+I'm guessing it's intermediate diversity - much less than gut and soil, but much more than the vaginal microbiome", and she replied "The total microbial diversity in marine sediments worldwide is similar to that of soil.
+I'll email you Hoshino et al., 2020 which addresses this." However I'm not sure if the high diversity KL is noting applies to individual soil samples (i.e.
+what scale)
+- Should make sure the above paragraph's claims are precise and accurate; see AW's feedback
+- "As diversity increases" specifically refers to inverse simpson diversity / q=2. Should I explicitly state that?
 
@@
 
+- mclaren2019cons measures overall extraction efficiency, not lysis efficiency; is that important for the above text?
+
 This example illustrates a general mechanism by which association between the abundances and efficiencies of taxa are created by the common influence of evolutionary history.
@@
 Ecology and efficiency are both affected by the same evolutionary history, leading to positive associations between the two: A change in diet that increases the relative abundances of many Bacteroidetes species and in so doing also increases the relative abundance of many easy-to-lyse species.
+
+- first sentence in above P: "the abundances and efficiencies" - the idea is that both are determined by genetically encoded traits
+- this is prob a tricky idea for readers and will have to take care in communicating it, and find nice refs
 
@@
 
+- Ben flags that this concept/section main idea should be highlighted more
+
 So far we have ignored error associated with our absolute abundance measurements.
@@
 Yet these biases are shared by the sequencing measurement.
-Though they make qPCR measurements a bad proxy for total cell density, when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can cancel, leaving relative accurate fold changes.
+Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can cancel, leaving relative accurate fold changes.
 (Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
-If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.
+If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.^[See discussion below]
 Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
+
+- from KL on qPCR being a bad proxy: We showed this empirically for all published measurements in marine sediments in Lloyd et al., 2013 doi:10.1128/AEM.02090-13. Apologies for the self-citation.
+- In the above sentence "If common primers are used...", KL asked a question that shows I'm not being clear enough about what I'm talking about here.
+KL: "Wouldn't you want to use taxa specific primers for qPCR and general primers for the amplicon libraries to do this analysis?
+If they're the same primer sets, then it doesn't tell you anything about the relative abundance of individual taxa within the population grabbed by the primers." my reply: "The "solutions" section talks about using targeted qPCR; this section is talking about qPCR with "universal" primers for the purpose of estimating the total absolute abundance, for use in FRAxC instead of cell counting.
+There have been quite a few papers doing this recently and 16S sequencing companies are starting to offer it as a service so I've been thinking through the implications.
+This section needs to be expanded and the point I'm making about using qPCR for FRAxC is still somewhat preliminary and something we should chat about"
 
@@
 
+- from gdoc: this distinction of ratios vs proportions needs to be spelled out more explicitly; could even use my seminar figure (or equations) to help
+- later thoughts: This distinction is so central for how I understand and explain the key ideas, should refer to it early and often
+
 A subset of methods for analyzing differential relative abundance are derived from the field of Compositional Data Analysis (CoDA).
@@
 
+- note from KL on choice of reference taxon: "It seems like you'd have to be pretty careful to pick a reference taxon that was not changing a lot between samples."
+    - this comment indicates I need to be clearer about the three reference strategies - assume constant AA; (but may not be available so can) spike-in at constant AA; (but that might not be convenient, so can) measure AA to adjust for changes in AA.
+    - my reply in the doc: "the idea is that either the reference is a taxon spiked at a fixed abundance to each sample, or it is a taxon you've measured independently with ddPCR or qPCR and so can account for its variation through the given equation." "There are also computational methods that attempt to identify taxa that don't vary (under the assumption that most taxa don't vary), which is an approach inspired from the RNAseq/gene expression literature (where it is probably a more valid assumption that most genes have consistent expression); Ben in his comments was suggesting we mention this approach too.
+"
+- On taxa not competing in targeted measures: "I think DNA-based measurements would still be affected by competition if DNA extraction yield is a strongly saturating function of input.
+But if this saturation isn't very strong then the effect should remain much less than the MGS measurement, which also has competition imposed by various normalization steps "
+- on 'considerations' noted in the last sentence: "perhaps most notable is how to deal with a given reference taxon not being present or detectable in every sample "
+
 ## Estimating error with targeted measurements of a small number of taxa
@@
 We illustrate the basic idea using the @lloyd2020evid experiment described above.
-Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.
+Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.^[see note]
 Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
@@
 In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].
+
+- 'specific microbial taxa' - I'm not sure if/how to address the taxonomic levels issue
+- Details: Table 1 indicates that the doubling time estimated for Bathyarchaeota in Core 30 was 10.1 by FRAxC and 10.3 by qPCR, which corresponds to slopes (log abundance / time) of 1/10.1 and 1/10.3.
+The difference between these slopes, 1/10.3 - 1/10.1, is an estimate of the slope of the mean efficiency and hence is the correction that should be applied to all taxa’s FRAxC estimates.
 

### 2021-07-14 article.Rmd
add ref to Rao et al
just suggesting an additional reference for the idea of spikein-then-multiply-by-relative-abundances

--- 2021-06-05 article.Rmd
@@
 This concern has motivated the development of various experimental methods to enable researchers to convert the relative abundances measured by MGS into absolute abundances (methods to be reviewed in the Appendix in a future version).
-An increasingly popular approach in the study of both environmental and host-associated microbiomes involves making an independent measurement of total microbial abundance in each sample (@props2017abso,@kevorkian2018esti,@lloyd2020evid,@tettamantiboshier2020comp,@contijoch2019gutm,@vandeputte2017quan,@vieirasilva2019quan).
+An increasingly popular approach in the study of both environmental and host-associated microbiomes involves making an independent measurement of total microbial abundance in each sample (@props2017abso,@kevorkian2018esti,@lloyd2020evid,@tettamantiboshier2020comp,@contijoch2019gutm,@vandeputte2017quan,@vieirasilva2019quan,@rao2021multi).
 This estimate is then multiplied by the MGS proportions to convert them to estimates of absolute abundance.

### 2021-07-17 article.Rmd
Prep for revision
- Edit/add notes for the revision
- Switch from rendering the vaginal example figure in article.Rmd to
  pulling the figure that is created in the notebook post
  2021-05-10-vaginal-microbiome-example

--- 2021-07-12 article.Rmd
@@
-```{r vaginal-example-src, cache = TRUE}
-#### Vaginal microbiome example
-bias <- tibble(taxon = c("Lactobacillus", "Gardnerella", "Atopobium", "Spike-in"),
-  efficiency = c(4.68, 0.16, 0.285, 1)) %>%
-  mutate(across(efficiency, ~. / min(.)))
-# efficiency associated with the targeted measurement of Lactobacillus
-targeted_efficiency <- 2
-a0 <- tribble(
-  ~taxon, ~timepoint, ~abundance,
-  "Lactobacillus", "T1", 5,
-  "Lactobacillus", "T2", 0.5,
-  "Gardnerella", "T1", 2,
-  "Gardnerella", "T2", 8,
-  "Atopobium", "T1", 3,
-  "Atopobium", "T2", 1.5,
-) %>%
-  # Shrink abundance to have similar scale as proportions
-  mutate(across(abundance, ~ . / max(.))) %>%
-  left_join(bias, by = "taxon") %>%
-  with_groups(timepoint, mutate,
-    total_abundance = sum(abundance),
-    proportion = close_elts(abundance),
-    biased_abundance = efficiency * abundance,
-    biased_proportion = close_elts(biased_abundance),
-    biased_count = biased_proportion * 1e3,
-    abundance_estimate_bulk = biased_proportion * total_abundance,
-    #> abundance_estimate_spikein = biased_count / biased_count[taxon == "Spike-in"],
-    abundance_estimate_targeted = biased_count * targeted_efficiency *
-      abundance[taxon == "Lactobacillus"] / biased_count[taxon == "Lactobacillus"],
-  )
-a1 <- a0 %>%
-  select(-total_abundance) %>%
-  pivot_longer(-c(taxon, efficiency, timepoint), names_to = "type")
-ptb <- a1 %>%
-  filter(
-    type %in% c("proportion", "biased_proportion"),
-  ) %>%
-  mutate(
-    across(type, fct_relevel, "proportion", "biased_proportion"),
-    across(type, fct_recode,
-      "Actual" = "proportion",
-      "Observed" = "biased_proportion"),
-  )
-
-## Panels showing the error in measurement and in differential abundance
-shared_layers <- list(
-  geom_path(aes(group = taxon), 
-    arrow = grid::arrow(length = unit(0.15, "inches"))),
-  geom_point(size = 2),
-  scale_color_brewer(type = "qual", palette = 1, guide = FALSE),
-    # guide = guide_legend(reverse = TRUE)),
-  labs(y = "Proportion", color = "Taxon"),
-  scale_y_log10(),
-  coord_cartesian(clip = "off"),
-  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
-  expand_limits(y = 1e-2),
-  theme(plot.margin = unit(c(0, 1.3, 0, 0), units = "in"))
-)
-# How much to nudge the taxon labels and proportions
-nudge.taxon <- 0.48
-nudge.prop <- 0.24
-# In future iterations, consider labelling the taxa in both facets
-p.meas <- ptb %>%
-  ggplot(aes(type, value, color = taxon)) +
-  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
-    labeller = as_labeller(function(x) str_c("Time point ", x))
-  ) +
-  shared_layers +
-  geom_text(data = ~filter(., type == "Actual"), 
-    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
-  geom_text(data = ~filter(., type == "Observed"), 
-    aes(label = round(value, 2)), nudge_x = nudge.prop) +
-  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"), 
-    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
-  labs(
-    x = "Type",
-    title = "Measurement error at each time point"
-  )
-p.fc <- ptb %>%
-  ggplot(aes(timepoint, value, color = taxon)) +
-  facet_wrap(~type, nrow = 1, scales = "fixed") +
-  shared_layers +
-  geom_text(data = ~filter(., timepoint == "T1"), 
-    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
-  geom_text(data = ~filter(., timepoint == "T2"), 
-    aes(label = round(value, 2)), nudge_x = nudge.prop) +
-  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"), 
-    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
-  labs(
-    x = "Time point",
-    title = "Actual and observed fold changes"
-  )
-## Panel showing the efficiencies of individual taxa and the sample means
-sme <- a0 %>%
-  with_groups(timepoint, summarize, mean_efficiency = sum(proportion * efficiency))
-sme0 <- sme %>%
-  mutate(label = str_glue("mean ({timepoint})")) %>%
-  select(label, efficiency = mean_efficiency)
-sme1 <- sme %>%
-  mutate(taxon = "Mean", type = "Mean") %>%
-  select(taxon, efficiency = mean_efficiency, timepoint, type)
-bias1 <- bias %>% 
-  filter(taxon != "Spike-in") %>% 
-  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
-  mutate(type = "Taxon")
-lvls = c("Atopobium", "Gardnerella", "Lactobacillus", "Mean")
-etb1 <- bind_rows(bias1, sme1) %>%
-  mutate(across(taxon, factor, levels = lvls))
-
-lvls = c("Gardnerella", "Atopobium", "Lactobacillus")
-bias2 <- bias %>% 
-  filter(taxon != "Spike-in") %>% 
-  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
-  mutate(
-    type = "Taxon",
-    across(taxon, factor, levels = lvls),
-    x = as.integer(taxon)
-  ) %>%
-  left_join(sme, by = "timepoint")
-
-clrs <- c(RColorBrewer::brewer.pal(n = 3, "Accent"), rep("#585858", 2))
-p.eff <- bias2 %>%
-  ggplot(aes(x = x, y = efficiency, color = taxon)) + 
-  geom_point(size = 2) +
-  geom_text(data = ~filter(., timepoint == "T2"),
-    aes(label = taxon), x = 3.3, hjust = 0) +
-  geom_segment(aes(xend = x, yend  = efficiency, y = mean_efficiency),
-    arrow = grid::arrow(length = unit(0.15, "inches"))) +
-  geom_segment(data = sme, 
-    aes(x = 1, xend = 3, y = mean_efficiency, yend = mean_efficiency), 
-    color = "#585858",
-    inherit.aes = FALSE) +
-  geom_text(data = sme0 %>% mutate(timepoint = "T2"),
-    aes(label = label, y = efficiency), x = 3.3, hjust = 0,
-    color = "#585858",
-    inherit.aes = FALSE) +
-  scale_color_manual(values = clrs) +
-  labs(y = "Relative efficiency", x = NULL, color = "Taxon", 
-    title = "Taxonomic bias of protocol"
-  ) +
-  scale_y_log10(limits = c(1, 100) / 2,
-    breaks = etb1$efficiency, 
-    labels = signif(etb1$efficiency, 2)
-  ) +
-  xlim(c(0.0, 4)) +
-  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
-    labeller = as_labeller(function(x) str_c("Time point ", x))
-  ) +
-  coord_cartesian(clip = "off") +
-  theme(
-    axis.line.x = element_blank(),
-    axis.text.x = element_blank(),
-    axis.ticks.x = element_blank(),
-    axis.title.x = element_blank(),
-    legend.position = "none",
-    plot.margin = unit(c(0, 1.0, 0, 0), units = "in")
-  )
-```
-
 # Introduction
@@
 A primary reason is that taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
-<!-- TODO: Possible references: Mallick et al., 2017 DOI 10.1186/s13059-017-1359-z; Barlow et al. 2020 https://doi.org/10.1038/s41467-020-16224-6 -->
 As a result taxa can differ dramatically (e.g.
@@
 
-- On whether MGS can be quantitative
-    - Karen Lloyd: Is there a reference that explains the disagreement? In my field, it hasn't really been explicitly addressed adequately.
-    - Michael McLaren: I'll have to look/think about this some more.
-I was trying to express a vague impression I have based on various things I've read and heard, and how community sequencing is used in practice, that implicitly suggest there is disagreement or at least a tension in the field where on the one hand there are many papers on the problems with community sequencing and on the other hand many papers (often by the same authors, and including me and Ben) using it to make quantitative inferences
-    - Karen Lloyd: You could reference Mallick et al., 2017 DOI 10.1186/s13059-017-1359-z; Or Barlow et al. 2020 https://doi.org/10.1038/s41467-020-16224-6
+- On whether MGS can be quantitative: Is there a reference that explains the disagreement?
+Perhaps this question hasn't been explicitly addressed in the literature.
+I was gesturing towards a tension where on one hand there are many papers on the problems with community sequencing and on the other hand many papers (often by the same authors, and including us) are using it to make quantitative inferences.
+KL suggests Mallick et al., 2017 DOI 10.1186/s13059-017-1359-z; Or Barlow et al.
+2020 https://doi.org/10.1038/s41467-020-16224-6 as possible references.
 
@@
 
+- Amy points to rao2021mult as another good ref for the above
+
 Here we summarize recent advances in our understanding of how taxonomic bias affects our ability to accurately determine how taxa vary in relative and absolute abundance across samples from different conditions—the inference problem known as *differential abundance (DA) analysis*.
@@
     + consider giving general explanation and/or specific examples of why absolute abundances are important
+
 # Bias can cause spurious results in proportion-based DA analysis
 
-```{r vaginal-example, fig.dim = c(8, 6) * 1.5, fig.cap = '(ref:vaginal-example)', dependson = "vaginal-example-src"}
-p.meas + p.eff + p.fc + plot_spacer() + 
-  plot_layout(byrow = TRUE, ncol = 2, widths = c(1, 0.6)) +
-  plot_annotation(tag_levels = "A")
+Comments on this version:
+
+- This version skips over the effect of bias on MGS measurement, leaving that for the intro, and goes right to the effect on FCs.
+I think this is a mistake; I'd rather walk the reader through the logic starting with the effect on one sample.
+- To accompany this, I'd like a new first figure, that illustrates the effect of competition. Essentially like Figure 2 of the eLife paper; perhaps using the vaginal mb example.
+
+```{r vaginal-example, cache = TRUE, fig.cap = '(ref:vaginal-example)'}
+svg_path <- here::here(
+  "notebook/_posts/2021-05-10-vaginal-microbiome-example",
+  "vaginal-microbiome-example_files/figure-html5",
+  "vaginal-example-proportions-timepoints-1.svg"
+)
+if (knitr::is_html_output()) {
+  knitr::include_graphics(svg_path)
+} else if (knitr::is_latex_output()) {
+  cowplot::ggdraw() + cowplot::draw_image(svg_path)
+}
+rm(svg_path)
 ```
@@
 \end{align}
-where $\text{mean efficiency}(s)=\text{actual}_j(s)*\text{efficiency}_j$ in the denominator is the average efficiency of all taxa in that sample.
-Intuitively, Equation \@ref(eq:error) says that the taxon will be overrepresented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
+where $\text{mean efficiency}(s)=\sum_{j}\text{actual}_j(s)\cdot \text{efficiency}_j$ in the denominator is the average efficiency of all taxa in that sample.
+Intuitively, Equation \@ref(eq:error) says that the taxon will be over-represented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
 Figure \@ref(fig:vaginal-example)B shows the efficiencies of the three taxa and the mean efficiency in each sample.
@@
 
-- on the description of competition - Ben suggested adding "for reads in the sequencer", but this addition doesn't quite make sense the way I've written this sentence, which is talking about the mathematical operation of normalization to proportions.
-But this paragraph deserves more attention, as the fact that we are imposing competition  via normalization to proportions, but the experiment also imposes competition that we can't avoid by just not normalizing, is subtle
+- When revising the above paragraph, pay attention to making clear that altough we are imposing competition via normalization to proportions, the experiment itself imposes competition on the absolute intermediates like DNA concentration and sequenced fragments.
 
@@
 - Consider different term than 'slope' since it's not a slope in the binary case
+- Comment from AW indicated that it was hard to follow the shift from describing the hypothesized absolute abundance regression model to describing the effect of bias on the fitted model.
+It could help to spell out this transition and make clear we're back to discussing the situation where the absolute abundance measurements are observed (biased) relative abundance measurements multiplied by some observation of the total.
+- Could be some potential for confusion in that the text notes we can get magnitude and sign errors for different taxa in the sample samples, but the figure shows a magnitude and sign error in two different sample sets on the same taxon.
+It could therefore be a better illustration to use two different taxa , both on the Eastern genotypes; We'll get a sign error for Penicillium but a magnitude error for any taxon that decreases more.
 
@@
 - Note about this figure: It is currently framed as O = A - \bar B to show the effect of error on inferred DA results; it could be flipped to A = O + \bar B to show how a correction might be done.
+
 # Implications for real-world inference
@@
 - Discussion of the vaginal example in the above paragraph should be distilled in conjunction with the discussion of this example in the previous section
+- Group-B strep / S. agalactiae is another "pathobiont" that has a low efficiency and I might as well included here for completeness. Can cite brooks2015thet and pace2021comp
 
@@
 - "As diversity increases" specifically refers to inverse simpson diversity / q=2. Should I explicitly state that?
+- "statistically independent of taxon abundances" - I need to make sure I'm being accurate/precise here.
+What do we actually know?
+Maybe simpler to say, if taxon efficiencies are not correlated with taxon abundance
 
@@
 Such phylogenically-associated, bias-affecting traits include cell-wall toughness, ribosomal-operon copy number, binding affinity for a given set of primers, and representativeness in taxonomic databases.
-Ecology and efficiency are both affected by the same evolutionary history, leading to positive associations between the two: A change in diet that increases the relative abundances of many Bacteroidetes species and in so doing also increases the relative abundance of many easy-to-lyse species.
+Ecology and efficiency are both affected by the same evolutionary history, leading to positive associations between the two: A change in diet that increases the relative abundances of many Bacteroidetes species also increases the relative abundance of many easy-to-lyse species.
 
@@
 Yet these biases are shared by the sequencing measurement.
-Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can cancel, leaving relative accurate fold changes.
+Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
 (Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
@@
 This section needs to be expanded and the point I'm making about using qPCR for FRAxC is still somewhat preliminary and something we should chat about"
+- should caveat something like 'Unfortunately, qPCR shares a limitation with DNA-based measurements: It assumes linearity of DNA extraction'
+    - this should start own paragraph on this issue. For now, can be a 2 sentence paragraph and ref the appendix.
 
@@
 - on 'considerations' noted in the last sentence: "perhaps most notable is how to deal with a given reference taxon not being present or detectable in every sample "
+- clarify that there are 2 ways to use spike-ins: Proportion-based and ratio-based method; and only the ratio-based method is bias-invariant
 
@@
 
+### Notes for revision
+
+- We need an illustration of this idea.
+Might drive home the point that multiple measurements can be used both as qualitative checks and quantitative corrections.
+Could also mention the ability to use host and/or diet reads (or computational AA approaches) as a check against the experimental AA approach.
 - 'specific microbial taxa' - I'm not sure if/how to address the taxonomic levels issue
@@
 The difference between these slopes, 1/10.3 - 1/10.1, is an estimate of the slope of the mean efficiency and hence is the correction that should be applied to all taxa’s FRAxC estimates.
+
+> Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
+In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
+Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
+
+This needs to be revised to make clear the novelty of using non-spiked taxa as a reference; could help to replace "In fact, some methods" with clearer language.
+Also make clear that some studies using reference taxa are using proportion-based methods.
+Also, cite the appendix that gives an overview of AA methods, and explicitly state distinction between proportion-based and ratio-based methods.
+
+Could help to change to:
+
+- Ratio-based analysis can also be used to overcome the effects of bias in differential absolute abundance.
+- /Make clear that the above ratio equation applies for relative and absolute observed or actual abudances (since totals cancel)/
+- Suppose we knew the true absolute abundance of a reference taxon $r$; then we might estimate the absolute abundance of taxon $i$ from
+  + Give the estimation equation
+  + Explain the effect of taxonomic bias - get correct FCs
+- It remains to determine the AA of one or more reference taxa.
+  + Some existing AA methods involve using spike ins of extraneous taxa
+    - discuss spike-in approach
+    - often stated that bias a problem for spike-ins; however we see that it only affects the baseline and not the fold changes
+      + this first part could be shunted to a discussion
+    - most/all papers on spike-ins suppose the spike-in must be added in a known fixed amount.
+    - however, the above argument indicates that it isn't required that the reference taxon is extraneous or has a fixed abundance; what is important is that we know the correct abundance up to a fixed multiplicative error
+    - hence it can be a native taxon we've measured using a targeted method; this method can be biased as long as it has a fixed multiplicative error (unlike MGS bias.)
+
+Need to consider if this is too detailed for the main text.
+Might want to make this section of the appendix particularly clear if can't put this in the main text.
+
+Key equations, to show side-by-side: Proportion-based and ratio-based AA inference.
 
@@
 Unknown efficiencies can be included in "compositional" linear modeling frameworks such as ALDEx2, DivNet, and fido simply by adding a protocol-specific term to the linear model of taxon log ratios.
-Thus such bias-aware meta-analyses are already technically feasible and—if bias is truly consistent within a protocol or study—may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not explicitly model bias.
+[*Add sentence on how this can be used for non-compositional analyses.*]
+Thus such bias-aware meta-analyses are already technically feasible and—if bias is truly consistent within a protocol or study—may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not model bias explicitly.

### 2021-07-17 notebook/_posts/2021-07-09-illustrations/illustrations.Rmd
Develop new illustrations
This doc workshops new illustrations of spurious inference in
hypothetical examples involving vaginal taxa. This version was used to
create some figures that I shared with the Callahan lab on 2021-07-16.

```{r}
set.seed(42)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  echo = FALSE
  #> out.width = "70%",
  #> fig.align = 'center',
  #> fig.width = 6,
  #> fig.asp = 0.618,
  #> fig.show = "hold"
)

#> options(dplyr.print_min = 6, dplyr.print_max = 6)
#> options(crayon.enabled = FALSE)

library(tidyverse)
library(here)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot(12))

close_elts <- function(x) x / sum(x)
```

## Example 1

Three species, with exponential growth rates of 0, <0, and >0.

```{r}
taxa <- tibble(
  taxon = c("Lactobacillus", "Gardnerella", "Atopobium", "Spike-in"),
  efficiency = c(4.68, 0.16, 0.285, 1),
  growth_rate = c(-1, 1, 0, 0) / 10
) %>%
# for now, ignore the spike-in
  filter(taxon != "Spike-in") %>%
  mutate(across(efficiency, ~. / min(.)))
```

Simulate expoential growth
```{r}
x <- taxa %>%
  expand(nesting(taxon, efficiency, growth_rate), time = seq(0, 20)) %>%
  with_groups(time, mutate,
    abundance = exp(growth_rate * time),
    total_abundance = sum(abundance),
    proportion = close_elts(abundance),
    observed_proportion = close_elts(efficiency * abundance),
    # Assuming perfect measurement of total abundance
    observed_abundance = observed_proportion * total_abundance
  )
```

```{r}
x %>%
  select(taxon:time, actual = abundance, observed = observed_abundance) %>%
  pivot_longer(c(actual, observed), names_to = "type", values_to = "abundance") %>%
  ggplot(aes(time, abundance, color = taxon)) +
  facet_wrap(~type) +
  scale_color_brewer(type = "qual", palette = 1) +
  geom_line()
```

Now let's try to show the values as stacked regions.
Strategy: Calculate ymin and ymax for each (taxon, time) value, then use `geom_ribbon()`.

We want

- ymin of first taxon = 0
- ymin of one taxon = ymax of the previous taxon
- ymax of each taxon = its ymin + its abundance
<!--  -->

We can calculate the values by just accumlating the abundances, then pull the mins and maxs.

```{r}
stack_abundance <- function(y, labels = NULL) {
  stopifnot(is.null(labels) | identical(length(y), length(labels)))
  vals <- accumulate(y, sum, .init = 0)
  ymin <- vals[seq_along(y)]
  ymax <- vals[1 + seq_along(y)]
  x <- tibble(ymin, ymax)
  if (!is.null(labels))
    x <- add_column(x, label = labels, .before = 1)
  x
}

stack_abundance(1:3)
stack_abundance(1:3, letters[1:3] %>% rev)
```

```{r}
x %>%
  select(taxon:time, actual = abundance, observed = observed_abundance) %>%
  pivot_longer(c(actual, observed), names_to = "type", values_to = "abundance") %>%
  with_groups(c(time, type), mutate,
    stack_abundance(abundance)
  ) %>%
  ggplot(aes(time, color = taxon, fill = taxon)) +
  facet_wrap(~type) +
  scale_color_brewer(type = "qual", palette = 1) +
  scale_fill_brewer(type = "qual", palette = 1) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax))
```

This actually seems to be working, though I'm not sure how reliable this method is.
E.g. this is assuming the taxa are in a consistent order.

```{r}
y <- x %>%
  select(taxon:time, actual = abundance, observed = observed_abundance) %>%
  pivot_longer(c(actual, observed), names_to = "type", values_to = "abundance") %>%
  arrange(time, type, taxon) %>%
  group_by(time, type) %>%
  nest() %>%
  mutate(stacked = map(data, ~stack_abundance(.$abundance))) %>%
  unnest(c(data, stacked))
```

```{r}
y %>%
  ggplot(aes(time, color = taxon, fill = taxon)) +
  facet_wrap(~type) +
  scale_color_brewer(type = "qual", palette = 1) +
  scale_fill_brewer(type = "qual", palette = 1) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax))
```

reverse the taxon order,

```{r}
z <- x %>%
  select(taxon:time, actual = abundance, observed = observed_abundance) %>%
  pivot_longer(c(actual, observed), names_to = "type", values_to = "abundance") %>%
  arrange(time, type, desc(taxon)) %>%
  group_by(time, type) %>%
  nest() %>%
  mutate(stacked = map(data, ~stack_abundance(.$abundance))) %>%
  unnest(c(data, stacked)) %>%
  ungroup
z %>%
  ggplot(aes(time, color = taxon, fill = taxon)) +
  facet_wrap(~type) +
  scale_color_brewer(type = "qual", palette = 1) +
  scale_fill_brewer(type = "qual", palette = 1) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax))
```

It seems like these plots don't actually do a good job of showing what is happening for the top taxon.
E.g. in this version, it looks like Atopobium is decreasing in actuality, even though it is actually fixed; and the observed increase is there but slight.

```{r}
z %>%
  filter(taxon == "Atopobium", type == "actual") %>%
  mutate(y = ymax - ymin) %>%
  count(y)
```

This suggests an intersting problem for the use of these types of plots generally.
I wonder if the problem goes away if we descretize / use stacked bars; this way we know the areas will be proportional to the abundance.

```{r}
x %>%
  select(taxon:time, actual = abundance, observed = observed_abundance) %>%
  pivot_longer(c(actual, observed), names_to = "type", values_to = "abundance") %>%
  ggplot(aes(time, abundance, color = taxon, fill = taxon)) +
  facet_wrap(~type) +
  scale_color_brewer(type = "qual", palette = 1) +
  scale_fill_brewer(type = "qual", palette = 1) +
  geom_col()
```

This visualization still suffers from the same problem - it is hard to tell that Atopobium is not really increasing.

What about the view of the proportions?

```{r}
x %>%
  select(taxon:time, actual = proportion, observed = observed_proportion) %>%
  pivot_longer(c(actual, observed), names_to = "type", values_to = "proportion") %>%
  ggplot(aes(time, proportion, color = taxon, fill = taxon)) +
  facet_wrap(~type) +
  scale_color_brewer(type = "qual", palette = 1) +
  scale_fill_brewer(type = "qual", palette = 1) +
  geom_col()
```

Perhaps it would be easier to see what's going on if we just compared two time-points, like in Figure 2 of the eLife paper.

```{r}
y <- x %>%
  filter(time %in% c(0, 10)) %>%
  mutate(timepoint = str_c("T", time)) %>%
  rename(
    "actual absolute" = abundance,
    "observed absolute" = observed_abundance,
    "actual proportion" = proportion,
    "observed proportion" = observed_proportion,
  ) %>%
  pivot_longer(c(starts_with("actual"), starts_with("observed"))) %>%
  separate(name, into = c("measurement_type", "abundance_type"), sep = " ")
```

```{r}
y %>%
  ggplot(aes(timepoint, value, color = taxon, fill = taxon)) +
  facet_grid(abundance_type~measurement_type, scales = "free_y") +
  scale_color_brewer(type = "qual", palette = 1) +
  scale_fill_brewer(type = "qual", palette = 1) +
  geom_col(width = 0.4)
```

This example is interesting; can see Lct increasing in absolute abundance despite decreasing, and Atp increasing despite saying fixed.
Probably instructive to think through why

## Example 2

Here the idea is to illustrate what happens when we increase Gard.
Let's keep Lacto and Ato fixed.

```{r}
taxa <- tibble(
  taxon = c("Lactobacillus", "Gardnerella", "Atopobium"),
  initial_density = c(2, 1, 1),
  efficiency = c(4.68, 0.16, 0.285),
  growth_rate = c(0, 1, 0) / 10
) %>%
  mutate(across(efficiency, ~. / min(.)))
x <- taxa %>%
  expand(
    nesting(taxon, initial_density, efficiency, growth_rate), 
    time = seq(0, 20)
  ) %>%
  with_groups(time, mutate,
    actual_density = initial_density * exp(growth_rate * time),
    total_density = sum(actual_density),
    actual_proportion = close_elts(actual_density),
    observed_proportion = close_elts(efficiency * actual_density),
    # Assuming perfect measurement of total abundance
    observed_density = observed_proportion * total_density
  )
xl <- x %>%
  pivot_longer(c(starts_with("actual"), starts_with("observed")), 
    values_to = "abundance") %>% 
  separate(name, into = c("type", "abundance_type"))
```

```{r}
xl %>%
  ggplot(aes(time, abundance, color = taxon)) +
  facet_grid(abundance_type~type, scales = "free_y") +
  scale_color_brewer(type = "qual", palette = 1) +
  geom_line()
```

## Example 3: Minimal AA example

Here the idea is to illustrate what happens when we increase Gard with just Lacto.

```{r}
taxa <- tibble(
  taxon = c("Lactobacillus", "Gardnerella", "Atopobium", "Spike-in"),
  initial_density = c(4, 1, 0, 0.5),
  efficiency = c(4.68, 0.16, 0.285, 1),
  growth_rate = c(0, 1, 0, 0) / 10
) %>%
  mutate(across(efficiency, ~. / min(.))) %>%
  filter(initial_density > 0)

x <- taxa %>%
  filter(taxon != "Spike-in") %>%
  expand(
    nesting(taxon, initial_density, efficiency, growth_rate), 
    time = seq(0, 20, by = 1)
  ) %>%
  with_groups(time, mutate,
    actual_density = initial_density * exp(growth_rate * time),
    total_density = sum(actual_density),
    actual_proportion = close_elts(actual_density),
    observed_proportion = close_elts(efficiency * actual_density),
    # Assuming perfect measurement of total abundance
    observed_density = observed_proportion * total_density,
    mean_efficiency = sum(actual_proportion * efficiency)
  )
xl <- x %>%
  pivot_longer(c(starts_with("actual"), starts_with("observed")), 
    values_to = "abundance") %>% 
  separate(name, into = c("type", "abundance_type"))
```

```{r}
xl %>%
  ggplot(aes(time, abundance, color = taxon)) +
  facet_grid(abundance_type~type, scales = "free_y") +
  scale_color_brewer(type = "qual", palette = 1) +
  geom_line()
```

We see Lactobacillus increasing in AA, because the growth of Gard raises the total AA while dropping the SME.

We can try plotting the total AA as well as the SME.

```{r}
# y <- x %>%
#   select(time, total_density, mean_efficiency) %>%
#   pivot_longer(
xl %>%
  filter(abundance_type == "density") %>%
  ggplot(aes(time, abundance, color = taxon)) +
  facet_wrap(~type) +
  scale_color_brewer(type = "qual", palette = 1) +
  geom_line() +
  geom_line(data = x, aes(y = total_density), color = "grey") +
  geom_line(data = x %>% mutate(type = "actual"), aes(y = mean_efficiency), color = "grey", linetype = "33")
```

Because here the mean efficiency drops faster than the total abundance grows, we see the observed abundance of Lactobacillus increase.

Does it also help to show what's happening with the relative abundances?

Here. Try showing this in the context of also showing how the QMP/AA method works, and the equations.

Logic of understanding the observed AA estimates:

Two ways:

- The actual AA estimates
- The observed proportions - Lacto dominates despite not actually dominating; hence the product with total AA shows Lacto increasing. In effect we're incorrectly assigning most of the increased cell density to Lacto
    - or: its observed prop has decreased much less than the total AA

- Compare the change in SME to the change in total abundance; use the equation for observed FC in terms of other stuff

<!--  -->

Might be better to put on log scale,
```{r}
xl %>%
  filter(abundance_type == "density") %>%
  ggplot(aes(time, abundance, color = taxon)) +
  facet_wrap(~type) +
  scale_y_log10() +
  scale_x_continuous(breaks = c(0, 10, 20)) +
  scale_color_brewer(type = "qual", palette = 1) +
  geom_line() +
  geom_line(data = x, aes(y = total_density), color = "grey") +
  geom_line(data = x %>% mutate(type = "actual"), aes(y = mean_efficiency), color = "grey", linetype = "33")
```

### Compare to using spike-in (todo)

### animated

```{r}
library(gganimate)
```

What if we show the growth of gard as an animation?
(Note, for this it might be better to have linear growth)

```{r}
xl %>%
  filter(abundance_type == "density") %>%
  ggplot(aes(type, abundance, fill = taxon)) +
  geom_col(width = 0.5) + 
  # scale_fill_brewer(type = "qual", palette = 1) +
  # Here comes the gganimate code
  transition_time(time)
# +
#   ease_aes('sine-in-out')
```

## Example 3 revisited

Taxa: Lactobacillus, Gardnerella, and a spike-in.
First show the error that arises from proportion-based inference. 
Here use just the two primary taxa to emphasize how the growth in the low-eff Gard messes up the the Lacto estimate.

```{r}
taxa <- tibble(
  taxon = c("Lactobacillus", "Gardnerella", "Atopobium", "Spike-in"),
  initial_density = c(4, 1, 0, 0.5),
  efficiency = c(4.68, 0.16, 0.285, 1),
  growth_rate = c(0, 1, 0, 0) / 10
) %>%
  mutate(across(efficiency, ~. / min(.))) %>%
  filter(initial_density > 0)
```

```{r}
x <- taxa %>%
  filter(taxon != "Spike-in") %>%
  expand(
    nesting(taxon, initial_density, efficiency, growth_rate), 
    time = seq(0, 20, by = 1)
  ) %>%
  with_groups(time, mutate,
    actual_density = initial_density * exp(growth_rate * time),
    total_density = sum(actual_density),
    actual_proportion = close_elts(actual_density),
    estimated_proportion = close_elts(efficiency * actual_density),
    # Assuming perfect measurement of total abundance
    estimated_density = estimated_proportion * total_density,
    mean_efficiency = sum(actual_proportion * efficiency)
  )
xl <- x %>%
  pivot_longer(c(starts_with("actual"), starts_with("estimated")), 
    values_to = "abundance") %>% 
  separate(name, into = c("type", "abundance_type"))
```

```{r}
clrs <- c(
  "Lactobacillus" = "#7FC97F",
  "Gardnerella" = "#BEAED4",
  "Spike-in" = "#FDC086",
  "Total" = "darkgrey",
  "Mean" = "black"
)
#
# RColorBrewer::brewer.pal(3, "Accent")
# #> [1] "#7FC97F" "#BEAED4" "#FDC086"
# RColorBrewer::display.brewer.all()
```

```{r}
# Restrict to density, and add the total as if it were another taxon.
x_dens <- xl  %>%
  filter(abundance_type == "density")
ttl <- x_dens %>%
  transmute(taxon = "Total", abundance = total_density, time, type) %>%
  distinct
x_dens <- bind_rows(x_dens, ttl) %>%
  mutate(
    label_y = case_when(
      type == "estimated" & taxon == "Total" ~ abundance * 1.2,
      type == "estimated" & taxon == "Lactobacillus" ~ abundance / 1.2,
      TRUE ~ abundance,
    )
  )

nudge_time <- 1

p_traj <- x_dens %>%
  ggplot(aes(time, abundance, color = taxon)) +
  facet_wrap(~type) +
  scale_y_log10(breaks = c(0.1, 1, 10), labels = c(0.1, 1, 10)) +
  scale_x_continuous(breaks = c(0, 10, 20)) +
  # scale_color_brewer(type = "qual", palette = 1) +
  scale_color_manual(values = clrs) +
  labs(y = "cell density") +
  theme(
    panel.spacing.x = unit(2.5, "cm"),
    legend.position = "none",
    plot.margin = unit(c(0.2, 2.3, 0.2, 0.2), units = "cm")
  ) +
  coord_cartesian(clip = "off") +
  geom_line() +
  geom_text(data = ~filter(., time == "20"), 
    aes(label = taxon, y = label_y), hjust = 0, nudge_x = nudge_time)
p_traj
```

To explain what is going on, let's consider the sample mean efficiency

```{r}
x_eff <- x %>%
  select(taxon, time, efficiency, Mean = mean_efficiency) %>%
  pivot_wider(names_from = taxon, values_from = efficiency) %>%
  pivot_longer(-time, names_to = "taxon", values_to = "efficiency")
```

```{r}
p_sme <- x_eff %>%
  ggplot(aes(time, efficiency, color = taxon)) +
  scale_y_log10() +
  # scale_y_log10(breaks = c(0.1, 1, 10), labels = c(0.1, 1, 10)) +
  scale_x_continuous(breaks = c(0, 10, 20)) +
  scale_color_manual(values = clrs) +
  theme(
    panel.spacing.x = unit(2.5, "cm"),
    legend.position = "none",
    plot.margin = unit(c(0.2, 2.3, 0.2, 0.2), units = "cm")
  ) +
  coord_cartesian(clip = "off") +
  geom_line() +
  geom_text(data = ~filter(., time == "20"), 
    aes(label = taxon), hjust = 0, nudge_x = nudge_time)
p_sme
```

Next, want to align the mean eff plot with the time axis on the actual plot.
To acheive, I might need to use distinct plots instead of facets in the first row.
Then, think of how to use the remaining quadrant - either insert the equation, or show the proportions.

Proportions at t=0 and t=20
```{r}
x_prop <- xl %>%
  filter(abundance_type == "proportion", time %in% c(0, 20)) %>%
  mutate(
    across(time, factor),
    across(taxon, as.factor),
    across(taxon, fct_rev),
  )
x_prop %>% pull(taxon) %>% levels
# x_lbl <- x_prop %>%
#   filter(type == "estimated") %>%
#   mutate(abundance = ifelse(taxon == "Lactobacillus", 0, 1))
p_prop <- x_prop %>%
  ggplot(aes(type, abundance, fill = taxon)) +
  facet_wrap(~time, labeller = label_both) +
  scale_color_manual(values = clrs[1:2]) +
  scale_fill_manual(values = clrs[1:2]) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(y = "proportion") +
  theme(
    panel.spacing.x = unit(1, "cm"),
    legend.position = "bottom",
    plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), units = "cm"),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.title.x = element_blank()
    # legend.title = element_text(6)
  ) +
  coord_cartesian(clip = "off") +
  geom_col(width = 0.6)
  # geom_text(data = x_lbl, 
  #   aes(color = taxon, label = taxon), hjust = 0, nudge_x = 0.4)
p_prop

```

```{r}
p_traj / (p_sme + p_prop) +
  plot_annotation(tag_levels = "A")
```

- save whole fig, and separate panels; add into lab meeting.

- practice how to explain it

<!--  -->

```{r}
pth <- "~/research/presentations/lab-meetings/2021-07-16-differential-abundance-manuscript/figures"
ggsave(fs::path(pth, "aa-example-traj.svg"), p_traj, units = "in", width = 6, height = 3.2)
ggsave(fs::path(pth, "aa-example-sme.svg"), p_sme, units = "in", width = 3.2, height = 3.0)
ggsave(fs::path(pth, "aa-example-prop.svg"), p_prop, units = "in", width = 3.6, height = 3.2)
```

### 2021-07-17 notebook/_posts/2021-05-10-vaginal-microbiome-example/vaginal-microbiome-example.Rmd
Update and render first vaginal example
Cleaned up the doc and updated to use the cleaner code and include the
caption from /article.Rmd. Rendered with distill_article class.
Multipanel figure appears in the Distill output in
vaginal-example-proportions-timepoints-1.svg so can be easily included
in the manuscript or other sources.

---
title: 'Hypothetical example of spurious fold changes in proportions in the vaginal microbiome'
author: Michael R. McLaren
date: '2021-05-10'
description: >
    This doc develops a hypothetical example in which bias leads to a spurious
    differential abundance result. It is based on a real case of bias measured
    among three species of vaginally-associated bacteria.
output:
  distill::distill_article:
    self_contained: false
    toc: true
    dev: svg
---

**Changelog**

- 2021-07-17: Cleaned up the doc and updated to use the cleaner code and
  include the caption from /article.Rmd

# Setup

```{r}
set.seed(42)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  echo = TRUE
)

library(tidyverse)
library(here)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot(12))

close_elts <- function(x) x / sum(x)
```

# Simulate data

```{r}
bias <- tibble(taxon = c("Lactobacillus", "Gardnerella", "Atopobium", "Spike-in"),
  efficiency = c(4.68, 0.16, 0.285, 1)) %>%
  mutate(across(efficiency, ~. / min(.)))
# efficiency associated with the targeted measurement of Lactobacillus
targeted_efficiency <- 2
a0 <- tribble(
  ~taxon, ~timepoint, ~abundance,
  "Lactobacillus", "T1", 5,
  "Lactobacillus", "T2", 0.5,
  "Gardnerella", "T1", 2,
  "Gardnerella", "T2", 8,
  "Atopobium", "T1", 3,
  "Atopobium", "T2", 1.5,
) %>%
  # Shrink abundance to have similar scale as proportions
  mutate(across(abundance, ~ . / max(.))) %>%
  left_join(bias, by = "taxon") %>%
  with_groups(timepoint, mutate,
    total_abundance = sum(abundance),
    proportion = close_elts(abundance),
    biased_abundance = efficiency * abundance,
    biased_proportion = close_elts(biased_abundance),
    biased_count = biased_proportion * 1e3,
    abundance_estimate_bulk = biased_proportion * total_abundance,
    #> abundance_estimate_spikein = biased_count / biased_count[taxon == "Spike-in"],
    abundance_estimate_targeted = biased_count * targeted_efficiency *
      abundance[taxon == "Lactobacillus"] / biased_count[taxon == "Lactobacillus"],
  )
a1 <- a0 %>%
  select(-total_abundance) %>%
  pivot_longer(-c(taxon, efficiency, timepoint), names_to = "type")
# data frame for plots
ptb <- a1 %>%
  filter(
    type %in% c("proportion", "biased_proportion"),
  ) %>%
  mutate(
    across(type, fct_relevel, "proportion", "biased_proportion"),
    across(type, fct_recode,
      "Actual" = "proportion",
      "Observed" = "biased_proportion"),
  )
```

# Plots

Panels showing the error in measurement and in differential abundance

```{r}
shared_layers <- list(
  geom_path(aes(group = taxon),
    arrow = grid::arrow(length = unit(0.15, "inches"))),
  geom_point(size = 2),
  scale_color_brewer(type = "qual", palette = 1, guide = "none"),
  labs(y = "Proportion", color = "Taxon"),
  scale_y_log10(),
  coord_cartesian(clip = "off"),
  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
  expand_limits(y = 1e-2),
  theme(plot.margin = unit(c(0, 1.3, 0, 0), units = "in"))
)
# How much to nudge the taxon labels and proportions
nudge.taxon <- 0.48
nudge.prop <- 0.24
# In future iterations, consider labelling the taxa in both facets
p.meas <- ptb %>%
  ggplot(aes(type, value, color = taxon)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed",
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  shared_layers +
  geom_text(data = ~filter(., type == "Actual"),
    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
  geom_text(data = ~filter(., type == "Observed"),
    aes(label = round(value, 2)), nudge_x = nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"),
    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
  labs(
    x = "Type",
    title = "Measurement error at each time point"
  )
p.fc <- ptb %>%
  ggplot(aes(timepoint, value, color = taxon)) +
  facet_wrap(~type, nrow = 1, scales = "fixed") +
  shared_layers +
  geom_text(data = ~filter(., timepoint == "T1"),
    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2"),
    aes(label = round(value, 2)), nudge_x = nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"),
    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
  labs(
    x = "Time point",
    title = "Actual and observed fold changes"
  )
```

Panel showing the efficiencies of individual taxa and the sample means

```{r}
# First, compute the mean efficiency, then join with the taxon efficiencies in
# a table for plotting.
sme <- a0 %>%
  with_groups(timepoint, summarize, mean_efficiency = sum(proportion * efficiency))
sme0 <- sme %>%
  mutate(label = str_glue("mean ({timepoint})")) %>%
  select(label, efficiency = mean_efficiency)
sme1 <- sme %>%
  mutate(taxon = "Mean", type = "Mean") %>%
  select(taxon, efficiency = mean_efficiency, timepoint, type)
bias1 <- bias %>%
  filter(taxon != "Spike-in") %>%
  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
  mutate(type = "Taxon")
lvls = c("Atopobium", "Gardnerella", "Lactobacillus", "Mean")
etb1 <- bind_rows(bias1, sme1) %>%
  mutate(across(taxon, factor, levels = lvls))

lvls = c("Gardnerella", "Atopobium", "Lactobacillus")
bias2 <- bias %>%
  filter(taxon != "Spike-in") %>%
  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
  mutate(
    type = "Taxon",
    across(taxon, factor, levels = lvls),
    x = as.integer(taxon)
  ) %>%
  left_join(sme, by = "timepoint")

clrs <- c(RColorBrewer::brewer.pal(n = 3, "Accent"), rep("#585858", 2))
p.eff <- bias2 %>%
  ggplot(aes(x = x, y = efficiency, color = taxon)) +
  geom_point(size = 2) +
  geom_text(data = ~filter(., timepoint == "T2"),
    aes(label = taxon), x = 3.3, hjust = 0) +
  geom_segment(aes(xend = x, yend  = efficiency, y = mean_efficiency),
    arrow = grid::arrow(length = unit(0.15, "inches"))) +
  geom_segment(data = sme,
    aes(x = 1, xend = 3, y = mean_efficiency, yend = mean_efficiency),
    color = "#585858",
    inherit.aes = FALSE) +
  geom_text(data = sme0 %>% mutate(timepoint = "T2"),
    aes(label = label, y = efficiency), x = 3.3, hjust = 0,
    color = "#585858",
    inherit.aes = FALSE) +
  scale_color_manual(values = clrs) +
  labs(y = "Relative efficiency", x = NULL, color = "Taxon",
    title = "Taxonomic bias of protocol"
  ) +
  # Set the vertical span to match the other plots
  scale_y_log10(limits = c(1, 100) / 2,
    breaks = etb1$efficiency,
    labels = signif(etb1$efficiency, 2)
  ) +
  xlim(c(0.0, 4)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed",
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  coord_cartesian(clip = "off") +
  theme(
    axis.line.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none",
    plot.margin = unit(c(0, 1.0, 0, 0), units = "in")
  )
```

Plot as a multi-panel figure

```{r vaginal-example-proportions-timepoints, fig.dim = c(8, 6) * 1.5}
p.meas + p.eff + p.fc + plot_spacer() + 
  plot_layout(byrow = TRUE, ncol = 2, widths = c(1, 0.6)) +
  plot_annotation(tag_levels = "A")
```

```{r save, include = FALSE, eval = FALSE}
fmts <- c("svg", "png", "pdf")
str_c("figures/example-vaginal-microbiome.", fmts) %>%
  here %>%
  walk(ggsave, units = "in", width = 8, height = 6, scale = 1.5)
```

Caption (copied from main article version 47ce39c): 
**Taxonomic bias can distort differential abundance results even when it is consistent for each taxon across samples.** 
Panel A shows the actual and observed proportions for hypothetical community samples from two time points, which differ in their relative abundance of three taxa.
Panel B shows taxonomic bias in terms of the relative efficiencies of the three taxa against the mean efficiency of each sample; the difference between the taxon's efficiency and the sample's mean (vertical arrows) determines the fold error seen in Panel A.
Panel C rearranges the plot from Panel A to show the actual and observed fold changes between time points.
The efficiencies of individual taxa were estimated by @mclaren2019cons from mock community data from @brooks2015thet.
The abundances are hypothetical but inspired by observations from the human vaginal microbiome; see main text.

Some ideas for improvement:

- Experiment with ways to make the connections even clearer between the efficiency differences and the error. 
- Try making the arrows vertical, so that they are directly connected to the fold changes in proportion and in efficiency.
- Can also split the third plot into T1 and T2.
- Might consider changing "Time point T1" to "Sample S1", to fit the primary language of the text.
- Can mention in the text that these could be samples from different women, or the same woman at different times.
- also use a shape scale for the taxa, to give a second way besides color to distinguish them

## Session info {.appendix}

<details><summary>Click for session info</summary>

```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```

</details>

### 2021-07-19 abundance-measurement.Rmd
Revise manuscript (in progress)
- Revise introduction
- Add section on model and error in measurement of individual samples
- Add partial outline of new summary section
- Split article into multiple Rmd files
- Swap order of solutions and implications sections
- Revise section on theory of error in DA analysis (in progress)
- Various edits to other sections

# How taxonomic bias affects abundance measurements {#abundance}

<!-- TODO: Add an illustration of the bias model in panel 1B -->

```{r fig-abundance, out.width = '65%', fig.cap = '(ref:cap-abundance)', cache = TRUE}
svg_path <- here("figures/absolute-and-relative-abundance.svg")
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  cowplot::ggdraw() + cowplot::draw_image(svg_path)
}
rm(svg_path)
```

(ref:cap-abundance) Panel A: Metagenomics output is not directly proportional to microbial cell density and so is typically interpreted as _relative abundances_, either as abundances relative to each other (ratio view) or to the total (proportion view). For illustration purposes, the effect of taxonomic bias on the metagenomics output is not shown. Panel B: The effect of taxonomic bias is to distort the relative abundances that are measured by fixed ratios.

_Thoughts for figure 1: ideas for caption: topic sentence can be something about how metagenomics measures biased relative abudances; indicate that A is showing unbiased MGS output; OR: Can I remake this figure such that it shows biased metagenomics output? Consider making label about the model, rather than abundance_

We can understand how taxonomic bias affect relative and absolute abundance measurement using a simplified model of microbiome measurement.
(Details are given in Appendix Model REF)
First, consider the MGS sequencing measurement. 
We start from a sample with a given combination of species in differing absolute abundances, which throughout we take to be in units of density of cells per unit mass.
(For simplicity we ignore issues relating to the sequencing of live versus dead cells and non-cellular DNA.)
Sequencing results in a total number of taxonomically-classified reads that, while not arbitrary, do not straightforwardly or consistently reflect the differences in total microbial density among samples. 
For this reason, these counts are typically only interpreted in relative terms  (Figure 1A)—either relative to each other (i.e., as ratios among taxa) or relative to the total (i.e., as proportions).
Protocols can be implemented to preserve more information about total abundances (See Cruz et al, and Appendix); however, in the main text we focus on the more typical situation where the total count only tells us about the precision of the measured relative abundances and we must make additional measurements or assumptions to recover information about absolute microbial densities.

The effect of taxonomic bias on the relative abundances in a sample is to perturb them by taxon-specific multiplicative factors—the taxa's _measurement efficiencies_—which quantify how efficiently each taxon is converted from cells into taxonomically classified sequencing reads (Figure 1B).
A taxon's efficiency aggregates the bias acrewed during the entire measurement process and can be thought of as a product of efficiencies at various steps in the experimental workflow and other taxon- and protocol-specific factors.
For example, _(Insert Gard and Lacto example, breaking down by extraction, PCR, and copy-number)_.
Only the relative efficiencies affect the measured relative abundances; doubling all efficiencies would not change the mesaured ratios or proportions.
Therefore we define the efficiencies as relative to a chosen reference taxon (or set of taxa); here we'll take to be that of the least-efficiently measured taxon being considered.

- TODO: Somewhere Give 1-sentence summary of the model - absolute information is lost, and measured relative abundances are perturbed by taxon- and protocol-specific factors. (In the figure caption and/or start of this section and or right here as a summary/resolution)
- Think about How what I write below depends on whether I'm taking this equation to in effect be a statement of the model, or the model is what I illustrated.
- Should define what I mean by Taxon, and address that we're making a big assumption that efficiency of a taxon is fixed (including doesn't vary from lower-level variation)
- A footnote or "Details" box can discuss what happens when there is genotypic or non-genotypic variation in efficiency within a taxon.

## Relative-abundance measurements

By this model, the ratios in the read counts of taxa, compared to the ratios among their cell densities, are multiplied by taxon- and protocol-specific factors (Figure, MWC).
\begin{align}
  (\#eq:ratio-err)
  \frac{\text{reads}_{i}(s)}{\text{reads}_{j}(s)}
  &= \frac{\text{density}_{i}(s)}{\text{density}_{j}(s)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
\end{align}
But what about the estimated proportions, estimated by dividing the taxon read counts by the total?
The estimated proportion of a taxon $i$ in a sample $s$ under this model equals (@mclaren2019cons, Box)
\begin{align}
  (\#eq:prop-err)
  \hat{\text{prop}}_{i}(s) 
  &= \text{prop}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
\end{align}
where $\text{mean efficiency}(s)\equiv \sum_{j}\text{prop}_j(s)\cdot \text{efficiency}_j$ is the average efficiency of cells in the sample.
The mean efficiency in the denominator of \@ref(eq:prop-err) was omitted in models of bias prior to that of @mclaren2019cons, but naturally arises from the requirement that the estimated proportions sum to 1 and captures the competition-to-be-measured that this sum-to-1 constraint implies ([Details](#details:relative)).
Intuitively, Equation \@ref(eq:error) says that the estimated proportion of taxon $i$ will be too large when its measurement efficiency is greater than the average efficiency of cells in the sample.

<details id="details:relative">
<summary style="font-weight:bold;font-size:120%">Details: Taxonomic bias and relative-abundance estimates</summary>
::: {.rmdnote}
<!-- ### Details: Taxonomic bias and relative-abundance estimates {-} -->

**Proportions:** The measured proportion of a taxon $i$ in a sample $s$, defined as the ratio of the taxon's density to the total density of all taxa in the sample, is
\begin{align}
  (\#eq:prop-def)
  \text{prop}_{i}(s) \equiv \frac{\text{density}_{i}(s)}{\text{total density}(s)}.
\end{align}
The standard estimate for the proportion of taxon $i$ is the proportion of its reads,
\begin{align}
  (\#eq:prop-est)
  \hat{\text{prop}}_{i}(s) \equiv \frac{\text{reads}_{i}(s)}{\text{total reads}(s)}.
\end{align}
These definitions imply that both the actual and the estimated proportions each sum to 1.
Our model stipulates that within a given sample $s$, the estimated and actual proportions are related by 
\begin{align}
  \hat{\text{prop}}_{i}(s) = c(s) \cdot \text{prop}_{i}(s) \cdot \text{efficiency}_{i}
\end{align}
for some taxon-independent coefficient $c(s)$.
The requirement that the estimated proportions sum to 1 implies that 
\begin{align}
  c(s) = \sum_j \text{prop}_{j}(s) \cdot \text{efficiency}_j = \text{mean efficiency}(s).
\end{align}

**Ratios:** The actual ratio of taxon $i$ to taxon $j$,
\begin{align}
  (\#eq:ratio-def)
  \text{density}_{i/j}(s) \equiv \frac{\text{density}_{i}(s)}{\text{density}_{j}(s)}
\end{align}
is estimated by the ratio of reads assigned to each taxon,
\begin{align}
  (\#eq:ratio-est)
  \hat{\text{density}}_{i/j}(s) \equiv \frac{\text{reads}_{i}(s)}{\text{reads}_{j}(s)} = \frac{\hat{\text{prop}}_{i}(s)}{\hat{\text{prop}}_{j}(s)}.
\end{align}
The error under multiplicative bias is
\begin{align}
  (\#eq:ratio-err)
  \hat{\text{density}}_{i/j}(s) = \frac{\text{density}_{i}(s)}{\text{density}_{j}(s)} \cdot
  \frac{\text{efficiency}_{i}(s)}{\text{efficiency}_{j}(s)}.
\end{align}

:::
</details>

<!-- begin figure -->

```{r fig-error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
svg_path <- here("figures/error-proportions.svg")
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  cowplot::ggdraw() + cowplot::draw_image(svg_path)
}
rm(svg_path)
```

(ref:cap-error-proportions) Top row: Error in proportions measured by MGS in two communities that contain different relative abundances of three taxa. Bottom row: Error in the estimated fold-change in the third taxon that is derived from these measurements.

<!-- end figure -->

This competitive aspect of proportional measurement creates a notable feature of the errors: They depend on the composition of the sample.
The same taxon, subject to the same measurement efficiency, will have a different error in its estimated proportion in different samples depending on the taxonomic composition of the sample (Figure 2A+B).
A taxon will be overestimated in samples where its efficiency is greater than that of the sample average, and underestimated in samples where its efficiency is less than the sample average (e.g. Taxon 3 in 2A vs 2B).
This difference in the consistancy of ratios versus proportions propagates to ratio- and proportion-based absolute abundance estimates and, as we describe below, is a key tool for understanding how taxonomic bias affects different DA analyses.

<!-- discussion of competition; tentative positioning -->

_The below discussion makes more sense if the goal as density/AA inference, and not proportion inference. Perhaps should split out some if it into the AA section._

Why does the sequencing outcome associated with one taxon depend on the rest of the sample?
We can think of proportional measurement as a competition among all the cells in the sample for a limited pool of 100 percentage points, for which
the measurement efficiencies determine each cell's ability to compete.
It is therefore a cell's efficiency relative to its average competitor (all the cells in the sample) that determines how many percentage points it receives.
This account highlights that it is the competition is the problem---if we were able to measure taxa/cells in a taxonomically biased but non-competitive manner, then we could get consistent errors. We cannot remove competition simply by using reads counts instead of proportions, since competition is intrinsic to sequencing experiments, due to limits in the yield during DNA extraction, PCR, and sequencing that occur in sequecing experiments.

_Perhaps say something about ratios - since cells or taxa in the sample are subject to the same average competitor, the effect of the average competitor cancels out._

Our ability to design-out these limits is limited and rarely attempted (ref Cruz for counter example).
But we might overcome competition with other types of measurements, particularly those that target individual taxa. For example, ddPCR can be applied directly to an unextracted sample to count individual cells of a given taxon, before any competition associated with the sequencing protocol has occurred. Targeted qPCR quantifies a marker (typically the 16S gene) in the extracted DNA, and so is only affected by competition during the DNA extraction step. These measurements are still subjected to taxonomic bias, but the lack or reduction in competition means that the error due to taxonomic bias is likely to be relatively consistent. For this reason, these measurements can serve as useful control measurements for correctly for the variation in the sample mean efficiency in the sequencing measurement.

## Absolute-abundance measurements

- P: Motivation for AA
However, these relative-abundance measures leave ambiguous what is happening to the absolute cell densities of taxa.
- e.g. of proportion decrease coming from different explanations
- e.g. of ratio varying having different explanations.

AA methods have generally been motivated, developed, and applied ignoring taxonomic bias and assuming the estimated proportions are correct.

The estimation procedures used by AA methods can be split into those that use the proportions estimated from the sequencing reads, and those that use ratios.

Proportion-based estimation involves the intuitive procedure of multiplying the estimated proportions by an independent estimate of the total cell density in the sample.
(The total-density estimates can be obtained by a wide variety of means, reviewed in APPENDIX.)
By this approach, the estimated density of a taxon $i$ in sample $s$ is simply
\begin{align}
  (\#eq:density-prop-est)
  \hat{\text{density}}_{i}(s) 
  = \hat{\text{prop}}_{i}(s) \cdot \hat{\text{total density}}(s).
\end{align}
This estimate equals the true density if the estimated proportions and total density are both accurate.
Accounting for the error in the estimated proportions due to taxonomic bias (by substituting in EQ-REF) and potential error in the estimated total density gives
\begin{align}
  (\#eq:density-prop-err)
  \hat{\text{density}}_{i}(s) 
  &= \text{density}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)} \cdot \text{error in }\hat{\text{total density}}(s).
\end{align}
Densities estimated from proportions are subject to the same bias-induced error as those proportions, which depends on sample composition through the sample's mean efficiency.

Error in the total density estimate is also typically ignored but may be considerable (Ref future box) and can worsen or improve the error in DA analysis; we return to its implications in the context of considering how these errors impact DA analysis.

Ratio-based density estimation embodies the fact that, without knowing the total abundance, we can regain absolute abundance information if we can determine the abundance of one or more _reference taxa_.
Given an estimate of the density of reference taxon $r$, we can estimate the density of taxon $i$ by multipling the ratio of $i$ to $r$ in the reads by the density of $r$,
\begin{align}
  \hat{\text{density}}_{i}(s) 
  &= \frac{\text{reads}_{i}(s)}{\text{reads}_{r}(s)} \cdot \hat{\text{density}}_{r}(s);
\end{align}
So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
It has also been used by methods that computationally determine a set of "housekeeping taxa" that they then assume have a constant abundance across samples.
Another potential application is to use naturally occurring taxa that we have biological reason to believe are constant, as has been argued for host and organelle-derived reads in plant and animal microbiome studies.
Moreover, EQREF makes clear there is no need for the reference taxa to be constant or known in advance, so long as we can measure its abundance: Hence the reference taxon can also be a naturally occurring taxon of varying abundance whose density we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.

The error in this estimate, accounting for the error in the estimated ratio due to taxonomic bias (EQ-REF) and potential error in the estimated density of the reference taxon, is described by
\begin{align}
  \hat{\text{density}}_{i}(s) 
  &= \text{density}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} \cdot \text{error in } \hat{\text{density}}_{r}(s)
\end{align}
Again, the properties of the corresponding relative-abundance estimate carries over, so that the error induced by taxonomic bias in the metagenomics measurement is independent of sample composition.
The reference density estimates are subject to error, but we can reasonably expect this to be more consistent across samples (see box), leaving the overall error relatively constant.

A warning note: The distinction between ratio and proportion-based estimation depends on whether an equation resembling X or Y is used for the final density estimates, and does not cleanly align with which empirical method is used to obtain absolute-abundance information. In particular, some studies that use spike-ins and most that use host or organelle reads to obtain absolute abundances do not follow this approach: Instead, these studies use the ratio of total reads of focal taxa to the reference reads as a proxy for total density to use in forming proportion-based estimates (EQREF). (For studies using host reads, the total density estimate may come from WGS sequencing and the estimated proportions from 16S sequencing.) 
Alhough there may be other advantages to this approach, the above observations suggest that it may be more susceptible to taxonomic bias in the sequencing measurement than using a ratio-based approach on the same underlying data.

### 2021-07-19 differential-abundance.Rmd
Revise manuscript (in progress)
- Revise introduction
- Add section on model and error in measurement of individual samples
- Add partial outline of new summary section
- Split article into multiple Rmd files
- Swap order of solutions and implications sections
- Revise section on theory of error in DA analysis (in progress)
- Various edits to other sections

# How taxonomic bias affects differential-abundance analysis

The current practice of ignoring bias in DA and other analyses of the differences betwen samples is sometimes justified by the idea that the error in measurements of in individual samples should not affect the measured differences between samples, so long as they have been subjected to the same taxonomic bias.
This argument is predicated on an assumption that consistent taxonomic bias creates consistent errors in the quanity of interest.
The previous section showed that while this may be true for ratio-based abundance measurements, it is not true for proportion-based measurements.
This section formally describes how the errors in the measurements of individual samples translate into errors (or lack thereof) in differential abundance analysis of (log) fold changes in proportions, ratios, and densities.

## Change between a pair of samples

Before considering the sterotypical many-sample DA analysis, it is instructive to 
consider simplest analysis of differential abundance: the estimation of fold changes in abundance between a pair of samples.
This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment.
Just as important, it helps to bridge intuitions from the single-sample results of the previous section to the many-sample case.

<!-- P: error in FC proportions -->
The composition-dependent effect of bias on fold error in proportions leads to error in fold-change estimates that is proportional to the inverse change in mean efficiency (REF Figure 2B).
From the error in individual sample \@ref(eq:prop-err), it follows that the estimated fold change in proportion of taxon $i$ from sample $s$ to sample $t$ is
\begin{align}
  (\#eq:prop-fc-err)
\underbrace{\frac{\hat{\text{prop}}_{i}(t)}{\hat{\text{prop}}_{i}(s)}} _\text{estimate}
  &= \frac
    {\text{prop}_{i}(t) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(t)}}
    {\text{prop}_{i}(s) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(s)}}
\\[0.5ex]
  &= 
  \underbrace{\frac{\text{prop}_{i}(t)}{\text{prop}_{i}(s)}}_\text{actual}
  \cdot 
  \underbrace{\left[\frac{\text{mean efficiency}(t)}{\text{mean efficiency}(s)}\right]^{-1}}_\text{error}
  .
\end{align}
The sample-independent efficiency factor of the error cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency from $s$ to $t$.
In the example in Figure 2, the mean efficiency decreases by 2.64X from 8.33 to 3.15, creating an artificial fold increase of 2.64X in the estimated fold changes.
This same error applies to all taxa, but its consequences differ depending on the taxon's actual fold change.
For Taxon 1, which increases by 2.3X, we estimate larger increase of 6.5X; For Taxon 2, which decreases with a FC of 0.15X, we see a lesser decrease of 0.44X; For taxon 3, which decreases with a FC by 0.6X, we see an _increase_ of 1.75X.
Thus for the same pair of samples we can see errors of increased magnitude, errors of decreased magnitude, and errors of changed direction (or sign) for different taxa depending on how the magnitide and direction of the true change compares to that of the mean efficiency. Finally, any taxon whose proportion is constant will appear to vary precisely as the inverse mean efficiency.

TODO: Fig 2 Add numbers for the other taxa.

<!-- P: error in proportion-based AAs (revise) -->
Taxonomic bias causes the same error in the estimated fold changes in proportion-based density estimates.
A similar calculation from the error in individual density estimates \@ref(eq:density-prop-err) shows that the error in the fold change of a taxon $i$ is given by the analogous formula to \@ref(eq:prop-fc-err), with density subsituted for proprotion.
The same types of errors (in magnitude and sign) can arise (Figure 3), now depending on how the magnitide and direction of the true change in density compares to that of the mean efficiency (Figure 3 TBD).
Maybe text:
- Alternative view of what is doing on, using the Lacto/Gard example, where a rarer but lower efficiency taxon grows. By first view, the mean efficiency decreases, causing an upward bias in the FC estimates. By the second view, the gains in total density are incorrectly attributed to the more efficiently measured taxon.
- This figure and its discussion should relate these two views. Perhaps the first mention of AA estimates should as well.
- Note: Whereas sign errors in proportion FCs require the involvement of at least three taxa (TODO: explain in footnote), sign errors in proportion-based density FCs can occur even with just two taxa.

NOTE: Having a figure and example specifically for density is not necessary for this draft.

<!-- P: error in FC ratios and ratio-based AAs -->
(In contrast to proportion-based estimates) Becuase consistent taxonomic bias creates a consistent error in ratio-based estimates, the error completely cancels, leaving accurate fold changes.
From \@ref(eq:ratio-err), the estimated fold change in the ratio of taxon $i$ to taxon $j$ from sample $s$ to sample $t$ is
\begin{align}
  (\#eq:ratio-fc-err)
  \underbrace{ \frac{{\text{reads}_{i}(t)}/{\text{reads}_{j}(t)}} {{\text{reads}_{i}(s)}/{\text{reads}_{j}(s)}} }_\text{estimate}
  &= 
  \frac{{\text{density}_{i}(t)}/{\text{density}_{j}(t)}} {{\text{density}_{i}(s)}/{\text{density}_{j}(s)}}
  \cdot
  \frac{\cancel{\text{efficiency}_{i}(t)}/\cancel{\text{efficiency}_{j}(t)}} {\cancel{\text{efficiency}_{i}(s)}/\cancel{\text{efficiency}_{j}(s)}}
  \\[0.5ex]&= 
  \underbrace{ \frac{{\text{density}_{i}(t)}/{\text{density}_{j}(t)}} {{\text{density}_{i}(s)}/{\text{density}_{j}(s)}} }_\text{actual}
  .
\end{align}
(Describe FCs in Figure 2 and Figure 3 density e.g., seeing the correct FCs.)

## Regression over many samples

Formal DA analysis is typically used to measure average patterns across many samples, such as how the abundance of a taxon varies with some covariate of interest.
The covariate may be discrete, such as whether the sample is from a healthy or sick person, or it may be continuous, such as a measure of space, time, or temperature.
Often these analyses can be framed as a regression problem.
For example, we might hypothesize that the log density of an individual taxon $i$ changes with a variable $x$ according to the simple linear regression, 
\begin{align}
  E\left[\log \text{density}_{i} \mid x \right] = a_{0} + a_{1}x
\end{align}
where $x$ is either continuous (e.g., sediment depth) or binary (e.g., $x=1$ for treated patients and $x=0$ for controls).
For a continuous covariate, these coefficients are the intercept, $a_0$, and the slope, $a_1$, of a line.
For a binary covariate, they are the average abundance in the first condition and the change in average abundance of the second condition.
How does taxonomic bias in the measurements of individual samples affect our estimate of the regression, and in particular the slope or difference term $a_1$, for log proportions, ratios, and densities?

_NOTE: Consider making the above abundance instead of density, and say that it could be proportion, ratio, or density_

### Proportion-based

P: Results for proportions and for proportion-based density, and 
- Equations for both; ref details box and SI for details and derivations
- The sample-dependent term is absorbed by the model intercept, which is typically not of interest to the researcher. - connect to canceling of efficiencies in FC result
- Error in slope equal to inverse slope in mean efficiency - connect to FC result
- Intuitive summary in terms of the correlation / covariation with the covariate 
    - When covaries - statistical bias
    - When doesn't covary - systematic errors between pairs of samples acts as random error in the regression -> reduced precision and power to detect differences.
- as w/ FCs, whether the error is significant depends on magnitude of true slope relative to slope of ME.
- in addition, depends on the SEs of the estimated slope

P: Discuss figure showing results of real and/or simulated example
- shared LFC error -> sign or (2 types of) magnitude errors for different taxa
        - if use leopold e.g., edit to show diff taxa on sample samples
- increase in residual variation and hence standard errors (NEED TO SHOW)
- old text
    - Notably, the error in an absolute sense is the same for all taxa—the slope is always reduced by the same amount—but the implications differ for various taxa depending on their slopes, causing magnitude or sign errors depending how the coefficient derived from the (unknown) true abundances compares to that of the mean efficiency (Figure \@ref(fig:regression-example)).

#### Figure

```{r regression-example, fig.cap = '(ref:regression-example)'}
svg_path <- here::here("figures/leopold2020host-example.svg")
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  cowplot::ggdraw() + cowplot::draw_image(svg_path)
}
rm(svg_path)
```

(ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.**

TODO: Add SI figure showing the distribution of mean efficiencies and the relationship with Mel that goes with this figure

Ideal figure perhaps shows how observed = actual + eff - mean eff, and individual taxa cases of sign and magnitude errors, and the distributional shift of all taxa, with the specific examples pulled from this distribution

```{r lfc-distribution, fig.cap = '(ref:lfc-distribution)', out.width = "60%"}
svg_path <- here::here("figures/illustrations/lfc-distribution.svg")
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  cowplot::ggdraw() + cowplot::draw_image(svg_path)
}
rm(svg_path)
```

(ref:lfc-distribution) **Taxonomic bias causes magnitude and sign errors in differential-abundance estimates.** In the absence of other types of errors, consistent taxonomic bias causes shift in the estimated log fold changes (LFC) or slopes in a differential-abundance analysis that is shared by all taxa, creating magnitude or sign errors depending on the true LFC.

### Ratio-based

1 short P explaining how it is no surprise that in ratio based, the error is absorbed by the intercept and the slopes remain accurate.
- (summary of previous) leads to Taxonomic bias does not affect ratio-based DA analysis
- include consideration of error in density estimate

## Caveats and other considerations

- Other error sources and violations of this idealized model
    - inconsistent error in density estimate, could be correlated or not with the covariate
    - for DNA based: non-linearity of extraction
    - contamination
- implications of these in our above results
    - proportion-based: won't get perfect error sharing across taxa
    - ratio-based: won't get perfect canceling
    - -> can limit the effectiveness of our 'solutions' below.

For this version, can just say this, and in future versions I can be more concrete as I think through more .

### Error in density estimates

In reality, there will be some error in the total density.
If this error is correlated with the mean efficiency of the sequencing measurement, it can actually reduce the error in the fold change estimate (See Box).

Possible continuations of this P:
1) Such a correlation might arise through DNA extraction bias when the total density is measurements of total abundance from DNA, such as florescent-based DNA quantification or qPCR of a universal 16S marker.
In this case, a perfectly linear but taxonomically biased extraction protocol will have a DNA yield that is proportional to the true density and the average absolute extraction efficiency of its cells.
16S concentration in the extracted DNA will be proportional to the average produce of extraction efficiency with 16S copy number.

2) work from old, and just use qPCR as a focal example.

- Could also mention live vs dead cells and cell-free DNA as another bias source, giving a ref to below article
- also mention primer bias; again can be helpful if shared. Obvious case is if both qPCR and seq primers are not getting certain taxonomic groups, that might be picked up by an alt approach like bulk DNA or flow cytometry.

Next P:
Shows subtleties and utility of thinking through an explicit model of the measurement process. 
When we do not attempt to correct for bias, DNA-based estimate of total might be better for proportion-based density estimates despite giving a poor reflection of the total density.
For example, (Lacto/Gard example) -> extraction and copy number bias gibe spurious results in total, but good results in individual taxa.
(Have a details box that does this as a worked example)
However, the non-linearity of DNA extraction may be a major worry in some cases and does not affect direct cell-density measurements.
Therefore, alternative options from the solutions section below may be preferable.

Old text to use:

Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
Yet these biases are shared by the sequencing measurement.
Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.^[See discussion below]
Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
In this case, ...

## Notes

<details><summary style="font-weight:bold;font-size:120%">Details: Taxonomic bias and differential-abundance estimates</summary>
::: {#details:differential .rmdnote}

#### Proportion-based inference {-}

#### Ratio-based inference {-}

:::
</details>

### 2021-07-19 introduction.Rmd
Revise manuscript (in progress)
- Revise introduction
- Add section on model and error in measurement of individual samples
- Add partial outline of new summary section
- Split article into multiple Rmd files
- Swap order of solutions and implications sections
- Revise section on theory of error in DA analysis (in progress)
- Various edits to other sections

# Introduction

<!-- First paragraph: DA analysis -->

The most basic question we can ask about microbial communities after which taxa are found in an ecosystem is, how do the abundances of these taxa vary across space, time, and host or environmental conditions?
Advances in sequencing technology allow us to now measure these _differential abundances_ simultaneously for 100s to 1000s of species using marker-gene and shotgun-metagenomic sequencing (jointly, MGS).
Standard MGS measurements do not preserve information about the total microbial density in our sample and thus are most often used to make observations about the relative abundaces of taxa.
But increasingly, MGS protocols are being modified or combined with additional measurements to
allow analysis of changes in absolute quantities such as cell density, which are more directly connected to ecological processes and thus may be more informative for host and environmental impact.

- (say something about how MGS-DA has become an indespensible tool with vast use in basic science and applications)
- End on with tension: Lack of reproducibility, questions over how quantitative they are, trustworthiness in critical applications, not ready for 'game time'

notes

- when mention r.a.'s for the first time: good spot to introduce the proportions versus ratios distinction, to set up contrast in the bias+DA paragraph, and approach/summary paragraph

<!-- P: Taxonomic bias -->

Primary reason for doubt: MGS measurements are taxonomically biased.
Taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
As a result taxa can differ dramatically (e.g. 10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—making the relative abundances we observe by MGS inaccurate representations of the actual sample compositions.
Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads among taxa make this taxonomic bias a feature of shotgun metagenomic measurements as much as marker-gene measurements.
Taxonomic bias is protocol specific (@mclaren2019cons) and can even vary among batches within an experiment (@yeh2018taxo); it thus not only makes MGS measurements inaccurate, but also incomparable between studies.

- might mention that bias can have an effect that is larger than important biological differences; can site lozupone2013meta and others. Perhaps put in the next paragraph. Point is that we know bias has a big effect on individual sample compositions, but does it have a big effect on differences between samples?

<!-- P: Taxonomic bias and DA -->

The dominant answer to taxonomic bias is to standardize protocols within an experiment and only compare differences between samples measured by the same protocol (and ideally even the same experimental batch). 
The reasoning goes that so long as all samples are subject to the same taxonomic bias, its effects should cancel when we compute changes in a particular taxon's abundance across samples.
However, @mclaren2019cons demonstrated that taxonomic bias can cause spurious changes in the proportion of a taxon, in which a taxon is observed to change more or less—or even in the opposite direction—than it really did.
(The leading?) A prominant method for absolute DA analysis multiplies the MGS proportions by an estimate of the total, and so is subject to the same problem.
Advancing the role of DA analysis in mb science and applications requires that we (the community) do a thorough examination of the impact that taxonomic bias on DA analysis and develop methods to correct or otherwise mitigate its effects.

Future versions:

- Introduce ratio-based DA analysis and constrast w/ proportion-based, in such a way as to motivate the body structure
- emphasize the importance of DA analysis of proportions for microbiome studies; make clear that the MWC result is based on the first accurate model of how taxonomic bias affects MGS measurements. Might want to mention others e.g. Brooks2016 on the more general point of spurious inference.

<!-- P: Our approach + main message (TODO) -->

Here we synthesize our current understanding of how taxonomic bias affects microbial differential abundance analysis and provide a roadmap for ???. 
We first explain how taxonomic bias affects relative and absolute abundance measurements, and how the errors it creates can lead to spurious differential abundance inferences.
We then consider the biological conditions in which such spurious results are likely.
Finally, we turn to new and existing experimental and computational methods that can be used to avoid or correct these errors.
Collectively, these methods may provide practical solutions to mitigating the effects of taxonomic bias on DA analysis for a majority of microbiome studies.

thoughts

- use of models; (perhaps introduce role of models in measurement? or leave for next section)
- define scope to all current AA methods; distinguish based on how affected by bias and other concerns like non-linear extraction, and suggest a new approach that is less affected by bias
    - Note that there are a variety of AA methods which make different assumptions about MGS and are differently interact with bias; we survey/taxonomize/synthesize them
- might include computational methods in scope, and use the term "calibration"; and ref to the future appendix that surveys the various methods
- need to think more about the first sentence
- Might want to hear mention the ratio and proportion divide

### 2021-07-19 real-world-implications.Rmd
Revise manuscript (in progress)
- Revise introduction
- Add section on model and error in measurement of individual samples
- Add partial outline of new summary section
- Split article into multiple Rmd files
- Swap order of solutions and implications sections
- Revise section on theory of error in DA analysis (in progress)
- Various edits to other sections

--- 2021-07-17 article.Rmd
@@
-# Introduction
-
-Most microbiome research aims to go beyond qualitative descriptions to make quantitative conclusions about associations between specific microbes or community states and key host or environmental properties.
-But there is serious disagreement about whether the microbiome measurements made by marker-gene and metagenomic sequencing (MGS) are or can ever be truly quantitative.
-A primary reason is that taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
-As a result taxa can differ dramatically (e.g.
-10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—making the relative abundances we observe by MGS inaccurate representations of the actual sample compositions.
-Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads among taxa make this taxonomic bias a feature of shotgun metagenomic measurements as much as marker-gene measurements.
-Taxonomic bias is protocol specific (@mclaren2019cons) and can even vary among batches within an experiment (@yeh2018taxo); it thus not only makes MGS measurements inaccurate, but also incomparable between studies.
-Taxonomic bias thereby poses a major challenge for any seeking to draw quantitative conclusions from MGS studies.
-
-- On whether MGS can be quantitative: Is there a reference that explains the disagreement?
-Perhaps this question hasn't been explicitly addressed in the literature.
-I was gesturing towards a tension where on one hand there are many papers on the problems with community sequencing and on the other hand many papers (often by the same authors, and including us) are using it to make quantitative inferences.
-KL suggests Mallick et al., 2017 DOI 10.1186/s13059-017-1359-z; Or Barlow et al.
-2020 https://doi.org/10.1038/s41467-020-16224-6 as possible references.
-
-The field's current answer to this challenge is methods standardization, prioritizing consistency over accuracy.
-By treating each sample within a study in exactly the same manner, we can ensure that observed differences between samples are not simply due to methodological differences.
-Yet standardization does not guarantee that these observed differences reflect the actual differences in taxonomic composition among samples.
-Consider what might be the simplest and most common inference made from MGS: How does the proportion of a taxon vary between samples?
-An observation that *Escherichia* coli makes up 10% of the sequencing reads in a fecal sample might, thanks to bias, have arisen from a sample in which the proportion of *E.
-coli* is actually 50% or 1%.
-Yet many hope that if we observe the proportion of *E.
-coli* to increase from 10% to 20% between two samples, its proportion must have actually doubled—by fixing the taxonomic bias of our measurements through standardization, we have (so the argument goes) fixed the measurement error in both samples so that it will cancel when estimate the fold change.
-Unfortunately, the results of @mclaren2019cons show that reality is not so neat.
-Because the taxa compete for sequencing effort within a sample, the effect that taxonomic bias has on the observed proportions depends on the taxonomic composition of the sample.
-As a result, MGS measurements can lead to spurious inferences about changes across samples even when taxonomic bias is constant for each taxon across samples.
-
-- 'many hope' - consider make small tweaks to phrases like this to sound less critical
-
-This problem posed by taxonomic bias intersects with a second, growing concern within the microbiome field: that the relative abundances measured by MGS can present a misleading view of absolute-abundance dynamics.
-Without making further assumptions (which are often difficult or impossible to verify) one cannot tell whether a doubling in the proportion of a taxon reflects an increase in its abundance or a decrease in the abundances of other taxa.
-This concern has motivated the development of various experimental methods to enable researchers to convert the relative abundances measured by MGS into absolute abundances (methods to be reviewed in the Appendix in a future version).
-An increasingly popular approach in the study of both environmental and host-associated microbiomes involves making an independent measurement of total microbial abundance in each sample (@props2017abso,@kevorkian2018esti,@lloyd2020evid,@tettamantiboshier2020comp,@contijoch2019gutm,@vandeputte2017quan,@vieirasilva2019quan).
-This estimate is then multiplied by the MGS proportions to convert them to estimates of absolute abundance.
-This simple calibration procedure directly transfers the error in the MGS proportions due to taxonomic bias to the estimated absolute abundances, leading to estimates that are not only individually inaccurate but can imply inaccurate variation in abundance across samples.
-
-- Amy points to rao2021mult as another good ref for the above
-
-Here we summarize recent advances in our understanding of how taxonomic bias affects our ability to accurately determine how taxa vary in relative and absolute abundance across samples from different conditions—the inference problem known as *differential abundance (DA) analysis*.
-We first explain how taxonomic bias can lead to spurious DA results and consider the biological conditions in which such spurious results are likely.
-But we then describe several experimental and computational methods that can be used to avoid or correct these errors.
-Collectively, these methods may provide practical solutions to mitigating the effects of taxonomic bias on DA analysis for a majority of microbiome studies.
-
-## Notes for revision
-
-- introduce differential abundance as the focus earlier
-- improve introduction of absolute abundance vs. relative abundance, and methods for obtaining absolute abundances
-    + might include computational methods in scope, and use the term "calibration"; and ref to the future appendix that surveys the various methods
-    + consider giving general explanation and/or specific examples of why absolute abundances are important
-
-# Bias can cause spurious results in proportion-based DA analysis
-
-Comments on this version:
-
-- This version skips over the effect of bias on MGS measurement, leaving that for the intro, and goes right to the effect on FCs.
-I think this is a mistake; I'd rather walk the reader through the logic starting with the effect on one sample.
-- To accompany this, I'd like a new first figure, that illustrates the effect of competition. Essentially like Figure 2 of the eLife paper; perhaps using the vaginal mb example.
-
-```{r vaginal-example, cache = TRUE, fig.cap = '(ref:vaginal-example)'}
-svg_path <- here::here(
-  "notebook/_posts/2021-05-10-vaginal-microbiome-example",
-  "vaginal-microbiome-example_files/figure-html5",
-  "vaginal-example-proportions-timepoints-1.svg"
-)
-if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
-} else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
-}
-rm(svg_path)
-```
-
-(ref:vaginal-example) **Taxonomic bias can distort differential abundance results even when it is consistent for each taxon across samples.** Panel A shows the actual and observed proportions for hypothetical community samples from two time points, which differ in their relative abundance of three taxa.
-Panel B shows taxonomic bias in terms of the relative efficiencies of the three taxa against the mean efficiency of each sample; the difference between the taxon's efficiency and the sample's mean (vertical arrows) determines the fold error seen in Panel A.
-Panel C rearranges the plot from Panel A to show the actual and observed fold changes between time points.
-The efficiencies of individual taxa were estimated by @mclaren2019cons from mock community data from @brooks2015thet.
-The abundances are hypothetical but inspired by observations from the human vaginal microbiome; see main text.
-
-- TODO: Add main-message sentence; see Ben's note below
-
-We illustrate the problems posed by taxonomic bias for differential abundance using an example where the bias of an MGS protocol was explicitly measured using mock communities.
-To analyze the effects of taxonomic bias on measurements of the human vaginal microbiome, @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven key vaginal species using a similar protocol to that of the Vaginal Human Microbiome Project (VaHMP).
-@mclaren2019cons used this data to estimate the relative measurement efficiency of the seven species.
-They found that the most efficiently measured species, *Lactobacillus iners*, had an efficiency that was 29X that of the least efficiently measured species, *Gardnerella vaginalis*, primarily due to a greater DNA extraction efficiency and 16S copy number.
-A third species, *Atopobium vaginae*, had an efficiency 1.8X that of *G. vaginalis*.
-A common aim of human vaginal microbiome studies, including the VaHMP, is to understand the causes and consequences of shifts in communities from being dominated by a particular *Lactobacillus* species to dominance by a non-*Lactobacillus* species such as *G.
-vaginalis*, which have been associated with deleterious health outcomes including an increased risk of preterm births in pregnant women.
-It is therefore important to know whether the efficiency variation found by @mclaren2019cons affects the observed differential abundances between *Lactobacillus*-dominated and *Gardnerella*-dominated communities.
-
-- Ben flagged that we want to make sure that it is clear from the text that this example is not just a one-off case of bias mattering, but is an illustration of a general/universal problem.
-The later discussion helps with this already; but what will help more is to start this section with a main-message sentence stating the general problem first.
-
-Figure \@ref(fig:vaginal-example) illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by *Lactobacillus* to being dominated by *Gardnerella*.
-In each case, we observe a greater proportion of *Lactobacillus* than is truly present, and less of the other two species (Figure \@ref(fig:vaginal-example)A).
-This positive multiplicative error^[what is the right terminology for multiplicative error that is greater than 1?] arises because *Lactobacillus* is more efficiently measured than the average species in each sample.
-@mclaren2019cons showed that the observed and actual proportions of a taxon $i$ in a sample $s$ are related by the equation
-\begin{align}
-  (\#eq:error)
-  \text{observed}_{i}(s) = \text{actual}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
-\end{align}
-where $\text{mean efficiency}(s)=\sum_{j}\text{actual}_j(s)\cdot \text{efficiency}_j$ in the denominator is the average efficiency of all taxa in that sample.
-Intuitively, Equation \@ref(eq:error) says that the taxon will be over-represented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
-Figure \@ref(fig:vaginal-example)B shows the efficiencies of the three taxa and the mean efficiency in each sample.
-The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of *Lactobacillus*.
-The error in the second sample is larger—specifically, by 6X—than the first sample because the high proportion of *Gardnerella* has reduced the sample mean efficiency (and thus increased the efficiency of *Lactobacillus* relative to the mean) by 6X.
-
-- see note on terminology for mult. error
-
-The enrichment depends on the taxon’s efficiency relative to the sample average, rather than the efficiency itself, because the requirement that the proportions of all taxa sum to one locks the taxa in a zero-sum competition.
-Notably, competition occurs (though in a less absolute form) even at the level of the raw sequencing reads: DNA extractions and PCR reactions have limited output capacity; high-concentration samples are typically subsampled to fixed, lower concentrations; and sequencing libraries are given limited sequencing effort; so that the experimental process itself typically imposes competition to be sequenced.
-And it is competition that prevents the effects of taxonomic bias from canceling when we analyze how the proportions of individual taxa change across samples.
-
-- When revising the above paragraph, pay attention to making clear that altough we are imposing competition via normalization to proportions, the experiment itself imposes competition on the absolute intermediates like DNA concentration and sequenced fragments.
-
-Figure \@ref(fig:vaginal-example)C shows the consequences of measurement error on an analysis of the fold change in proportions between the two samples.
-In reality, *Gardnerella* changes by 4X, *Lactobacillus* by 0.1X, and *Atopobium* by 0.5X.
-Yet what is observed is that *Gardnerella* changes by 32X, *Lactobacillus* by 0.6X, and *Atopobium* by 3X.
-In other words, the change in each case is multiplied by 6X, causing a much larger increase in *Gardnerella* and a smaller decrease *Lactobacillus* than actual, and the appearance that *Atopobium* increased when it in fact decreased.
-Why do such errors occur?
-The multiplicative error in Equation \@ref(eq:error) has two parts: a taxon-specific term in the numerator that is the same in each sample, and a taxon-independent term in the denominator that depends on the composition of the sample through the mean efficiency.
-When we compute the fold change in a taxon i between samples s and t, 
-\begin{align}
-  \frac{\text{observed}_{i}(t)}{\text{observed}_{i}(s)}=\frac{\text{actual}_{i}(t)}{\text{actual}_{i}(s)}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{i}}*\frac{\text{mean efficiency}(s)}{\text{mean efficiency}(t)}
-\end{align}
-the constant efficiency in the numerator of Equation \@ref(eq:error) cancels, but the varying mean efficiency does not, leaving an error that is equal to the inverse change in mean efficiency.
-As a result, we are here left inferring that the fold-changes in proportions have all increased less (or decreased more) than they truly have.
-In our example, the mean efficiency decreased by 6X (from 15 to 2.5) due to the shift in dominant taxon from the high-efficiency *Lactobacillus* to the low-efficiency *Gardnerella*, causing the observed fold changes to increase by 6X above the actual.
-This increase distorted the magnitudes of the changes in *Lactobacillus* and *Gardnerella* (decreasing the first and increasing the second) and changed the sign of *Atopobium* from negative to positive.
-Thus partial cancelling in the effect of a constant multiplicative bias may not be sufficient to protect against spurious results.
-
-- TODO: Add strike-throughs to the above equation
-- Revision should hone the selection of equations
-
-Spurious results can occur for the same basic reason in proportion-based absolute-abundance inference.
-Suppose that the true total abundance in each sample can be determined experimentally.
-Because the absolute abundance estimate simply multiplies the erroneous observed proportions by this total, the estimated change in absolute abundance is wrong by the same factor as the fold change in proportions.
-In our Figure \@ref(fig:vaginal-example) example, if the change in total abundance between samples were 2X, then we would observed both *Lactobacillus* and *Atopobium* to have increased in absolute abundance (by 1.2X and 6X), when in fact *Lactobacillus* decreased (by 5X) and *Atopobium* remained the same.
-
-## Effect on regression of multiple samples
-
-We showed how fold changes of individual samples may be unreliable.
-But microbiome analyses often do not interpret such individual differences, and instead look for average patterns across many samples, such as how the abundance of a taxon varies with some covariate of interest.
-The covariate may be discrete, such as whether the sample is from a healthy or sick person, or it may be continuous, such as a measure of space, time, or temperature.
-Often these analyses can be framed as a regression problem.
-For example, we might hypothesize that the log absolute abundance of taxon $i$ changes with a variable $x$ according to the simple linear regression, 
-\begin{align}
-  E[\text{log abundance}_{i} \mid x] = a_{0} + a_{1}x
-\end{align}
-where $x$ is either continuous (e.g., sediment depth) or binary (e.g., $x=1$ for treated patients and $x=0$ for controls).
-Much of the effect of bias will be absorbed by the model intercept, which is typically not of interest to the researcher.
-However, variation in the mean efficiency creates additional variation (reducing statistical power to detect variation in the slope) and can also create a statistical bias in our slope estimate equal to the inverse slope of $E[\text{log mean efficiency}\mid x]$ as a function of $x$.
-Therefore if the mean efficiency varies across samples, but is not associated with the covariate $x$, its effect may simply be to create larger standard errors in our slope estimate.
-But the larger worry may be the systematically distorted slope ($a_1$) estimates that arise in situations where the mean efficiency changes with $x$.
-Notably, the error in an absolute sense is the same for all taxa—the slope is always reduced by the same amount—but the implications differ for various taxa depending on their slopes, causing magnitude or sign errors depending how the coefficient derived from the (unknown) true abundances compares to that of the mean efficiency (Figure \@ref(fig:regression-example)).
-
-- on "(e.g., sediment depth)": Want to confirm with KL that I'm understanding this correctly.
-She notes: "Even though these values are sequential, we use them as discrete variables because the samples are only from one depth each.
-I'm not sure if this distinction matters here." To which I replied: "right....you first converted sediment depth to time, and then regressed log abundance on time to estimate growth rate?
-(new line) Here I was just using sediment depth as an example variable, but I could replace with time to more match your analysis"
-- Consider different term than 'slope' since it's not a slope in the binary case
-- Comment from AW indicated that it was hard to follow the shift from describing the hypothesized absolute abundance regression model to describing the effect of bias on the fitted model.
-It could help to spell out this transition and make clear we're back to discussing the situation where the absolute abundance measurements are observed (biased) relative abundance measurements multiplied by some observation of the total.
-- Could be some potential for confusion in that the text notes we can get magnitude and sign errors for different taxa in the sample samples, but the figure shows a magnitude and sign error in two different sample sets on the same taxon.
-It could therefore be a better illustration to use two different taxa , both on the Eastern genotypes; We'll get a sign error for Penicillium but a magnitude error for any taxon that decreases more.
-
-```{r regression-example, fig.cap = '(ref:regression-example)'}
-svg_path <- here::here("figures/leopold2020host-example.svg")
-if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
-} else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
-}
-rm(svg_path)
-```
-
-(ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.** Details to synthesize into a caption: Regression of log2(Proportion) of the commensal fungus Penicillium versus timepoint; timepoints 1 and 2 are pre- and post-challenge with the pathogen Melampsora.
-Data is split by the region the host plants are derived from (Eastern and Western US).
-Calibrated proportions = Observed proportions in the real, experimental samples after adjustment for the bias measured in mock communities.
-Mean efficiency of each community is inferred by treating the calibrated proportions as the truth, and multiplying by efficiencies estimated from the mocks.
-The pathogen Melampsora has a high measurement efficiency; thus once it infects the plants, the mean efficiency of the sample increases (purple points).
-Efficiency is here taken as relative to the focal taxon Penicillium.
-West plants tend to be more resistant to the pathogen, which likely explains why the mean efficiency doesn't increase as much in the West plants.
-Penicillium is observed to decrease in log proportion, in both the East and West plants.
-But the calibrated measurements show that it actually slightly increases in the East plants, and has a lesser decrease in the West plants than what was observed before bias correction.
-The difference between the Calibrated and Observed data points and regression lines equals the regression line of the mean efficiency: Orange = Green + Purple; Green = Orange - Purple.
-The absolute error in regression coefficients is the same for all taxa.
-I picked Penicillium for illustration since it has the smallest observed decrease, which makes the error due to bias have a particularly significant impact.
-
-- Note about this figure: It is currently framed as O = A - \bar B to show the effect of error on inferred DA results; it could be flipped to A = O + \bar B to show how a correction might be done.
-
 # Implications for real-world inference
@@
 
-## We cannot simply ignore taxonomic bias
+## Section conclusion
 
-These examples show that there is potential for bias to distort DA results in both low and high diversity settings.
-We therefore need methods to control or correct for bias to ensure the validity of future results, as well as methods to probe the robustness of past results.
-It is possible that spurious results are unlikely in many experimental contexts.
-By applying such methods across a range of biological and experimental contexts we will deepen our understanding of when bias is unlikely to significantly distort biological findings and when additional measures are needed to mitigate its effects.
+These examples show that there is potential for bias to distort DA results in both low and high diversity settings, but its effects may be negligible in many studies.
+By applying the types of methods demonstrated here to a range of biological and experimental contexts, and further developing theory, we will deepen our understanding of when bias is unlikely to significantly distort biological findings.
 
-# Potential solutions
+(prob move to the conclusion:)
+With this knowledge in hand we will be better placed to make the inevitable tradeoffs when designing new experiments (in terms of how much bias-mitigation measures to use) and also be better able to evaluate our confidence in past results.
 
-## Using control samples to calibrate relative abundances
+## Notes for revision
 
-In principle, control ("mock") communities containing representative taxa from the environment of interest can be used to directly estimate the measurement efficiencies and correct bias in the MGS measurements ("calibration").
-Such an approach may be sufficient for synthetic-community experiments, where all taxa are culturable, and for relatively simple natural communities like the vaginal microbiome that are dominated by a small number of culturable taxa.
-But suitable mock controls may not be feasible for most complex natural ecosystems and require significant effort to develop.
+Include simulation results indicating that bias might not be a problem.
 
-An closely-related alternative to mocks are controls derivedd from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (@mclaren2019cons).
-A natural fecal standard is currently being developed by [NIST](https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material) and one has recently been made available commercially by Zymo Research ([ZymoBIOMICS Fecal Reference](https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology)).
-Ensuring the stability and homogeneity of such control samples can be challenging.
-Careful testing is also needed to ensure that the preparation and storage of the controls has not significantly affected the taxonomic bias relative to the experimental samples in a given application.
-As it is feasible to characterize a single standard much more extensively than a typical community sample, we may be able to obtain an estimated composition we feel comfortable treating as the ground truth.
-Yet even when this is not possible, such natural standards can allow us to reconcile results across studies despite not knowing the truth.
-
-## Ratio-based relative and absolute abundance inference
-
-A comprehensive approach of estimating the efficiency of all taxa is needed for getting calibrated relative abundances of all taxa within individual samples.
-But what if we only want calibrated DA analysis?
-That is, we are comfortable with not knowing whether *E. coli* is 10% of our sample if we can confidently determine that it doubled or halved between sample conditions?
-This problem is easier to solve, and one solution is to use analysis methods based on the ratios among taxa rather than the proportions of individual taxa.
-
-- from gdoc: this distinction of ratios vs proportions needs to be spelled out more explicitly; could even use my seminar figure (or equations) to help
-- later thoughts: This distinction is so central for how I understand and explain the key ideas, should refer to it early and often
-
-A subset of methods for analyzing differential relative abundance are derived from the field of Compositional Data Analysis (CoDA).
-The defining feature of CoDA methods is that they are based on fold-changes in ratios among elements (here, the taxa) (REF AITCHISON).
-The use of CoDA methods in microbiome analysis has largely been motivated by concerns over the negative correlations between taxa induced by the sum-to-one constraint in taxon proportions.
-But @mclaren2019cons showed that, due to their property of perturbation invariance (REF AITCHISON), the results of CoDA differential relative abundance analysis are invariant to consistent taxonomic bias.
-From Equation \@ref(eq:error), the observed ratio of a taxon $i$ to a taxon $j$ in a sample $s$ is
-\begin{align}
-  \frac{\text{observed}_{i}}{\text{observed}_{j}}=\frac{\text{actual}_{i}}{\text{actual}_{j}}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}
-\end{align}
-The error is independent of the sample composition and thus cancels in any function that is a fold-change of this (or any) ratio of taxa between samples.
-(Perhaps insert example from Figure \@ref(fig:vaginal-example)ABC all giving same ratio FCs.)
-
-Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
-In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
-Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
-To estimate the abundance of a focal taxon $i$, this approach multiplies the ratio of reads (or proportions) of taxon i to a reference taxon $r$ by a known or estimated abundance of the reference taxon, (while ignoring bias)
-\begin{align}
-  \text{estimate}_{i} = \frac{\text{observed}_{i}}{\text{observed}_{r}}*\text{abundance}_{r}
-\end{align}
-So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
-Yet the reference taxon could also be a naturally occurring taxon whose absolute abundance we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.
-Because taxa do not compete (or competition is greatly reduced) in such targeted measurements, any taxonomic bias associated with them is expected to create constant multiplicative error across samples and so not affect fold change estimates [APPENDIX].
-APPENDIX describe various theoretical and experimental considerations for both approaches.
-
-- note from KL on choice of reference taxon: "It seems like you'd have to be pretty careful to pick a reference taxon that was not changing a lot between samples."
-    - this comment indicates I need to be clearer about the three reference strategies - assume constant AA; (but may not be available so can) spike-in at constant AA; (but that might not be convenient, so can) measure AA to adjust for changes in AA.
-    - my reply in the doc: "the idea is that either the reference is a taxon spiked at a fixed abundance to each sample, or it is a taxon you've measured independently with ddPCR or qPCR and so can account for its variation through the given equation." "There are also computational methods that attempt to identify taxa that don't vary (under the assumption that most taxa don't vary), which is an approach inspired from the RNAseq/gene expression literature (where it is probably a more valid assumption that most genes have consistent expression); Ben in his comments was suggesting we mention this approach too.
-"
-- On taxa not competing in targeted measures: "I think DNA-based measurements would still be affected by competition if DNA extraction yield is a strongly saturating function of input.
-But if this saturation isn't very strong then the effect should remain much less than the MGS measurement, which also has competition imposed by various normalization steps "
-- on 'considerations' noted in the last sentence: "perhaps most notable is how to deal with a given reference taxon not being present or detectable in every sample "
-- clarify that there are 2 ways to use spike-ins: Proportion-based and ratio-based method; and only the ratio-based method is bias-invariant
-
-## Estimating error with targeted measurements of a small number of taxa
-
-Because targeted absolute-abundance measurements are expected to provide (relatively) accurate  estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
-We illustrate the basic idea using the @lloyd2020evid experiment described above.
-Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.^[see note]
-Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
-@lloyd2020evid found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
-In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].
-
-### Notes for revision
-
-- We need an illustration of this idea.
-Might drive home the point that multiple measurements can be used both as qualitative checks and quantitative corrections.
-Could also mention the ability to use host and/or diet reads (or computational AA approaches) as a check against the experimental AA approach.
-- 'specific microbial taxa' - I'm not sure if/how to address the taxonomic levels issue
-- Details: Table 1 indicates that the doubling time estimated for Bathyarchaeota in Core 30 was 10.1 by FRAxC and 10.3 by qPCR, which corresponds to slopes (log abundance / time) of 1/10.1 and 1/10.3.
-The difference between these slopes, 1/10.3 - 1/10.1, is an estimate of the slope of the mean efficiency and hence is the correction that should be applied to all taxa’s FRAxC estimates.
-
-> Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
-In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
-Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
-
-This needs to be revised to make clear the novelty of using non-spiked taxa as a reference; could help to replace "In fact, some methods" with clearer language.
-Also make clear that some studies using reference taxa are using proportion-based methods.
-Also, cite the appendix that gives an overview of AA methods, and explicitly state distinction between proportion-based and ratio-based methods.
-
-Could help to change to:
-
-- Ratio-based analysis can also be used to overcome the effects of bias in differential absolute abundance.
-- /Make clear that the above ratio equation applies for relative and absolute observed or actual abudances (since totals cancel)/
-- Suppose we knew the true absolute abundance of a reference taxon $r$; then we might estimate the absolute abundance of taxon $i$ from
-  + Give the estimation equation
-  + Explain the effect of taxonomic bias - get correct FCs
-- It remains to determine the AA of one or more reference taxa.
-  + Some existing AA methods involve using spike ins of extraneous taxa
-    - discuss spike-in approach
-    - often stated that bias a problem for spike-ins; however we see that it only affects the baseline and not the fold changes
-      + this first part could be shunted to a discussion
-    - most/all papers on spike-ins suppose the spike-in must be added in a known fixed amount.
-    - however, the above argument indicates that it isn't required that the reference taxon is extraneous or has a fixed abundance; what is important is that we know the correct abundance up to a fixed multiplicative error
-    - hence it can be a native taxon we've measured using a targeted method; this method can be biased as long as it has a fixed multiplicative error (unlike MGS bias.)
-
-Need to consider if this is too detailed for the main text.
-Might want to make this section of the appendix particularly clear if can't put this in the main text.
-
-Key equations, to show side-by-side: Proportion-based and ratio-based AA inference.
-
-## Computational approaches
-
-What can we do for experiments that have already been conducted without mock controls, targeted control measurements, or spike-ins?
-At least two purely computational approaches can still be used.
-
-**Bias sensitivity analysis:**
-A straightforward and universally available approach is to use computer simulation to determine how the results change under a range of possible sets of efficiencies, which can be randomly generated to reflect certain hypotheses about bias in the given system.
-The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
-Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—not just for differential absolute abundance, but for all microbiome analyses.
-
-**Bias-aware meta-analysis:**
-When the goal is to perform a meta-analysis that combines studies that have used different protocols, the unknown measurement efficiencies of each protocol can be explicitly included as parameters of the statistical model that is used.
-Unknown efficiencies can be included in "compositional" linear modeling frameworks such as ALDEx2, DivNet, and fido simply by adding a protocol-specific term to the linear model of taxon log ratios.
-[*Add sentence on how this can be used for non-compositional analyses.*]
-Thus such bias-aware meta-analyses are already technically feasible and—if bias is truly consistent within a protocol or study—may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not model bias explicitly.
+Outline?

### 2021-07-19 solutions.Rmd
Revise manuscript (in progress)
- Revise introduction
- Add section on model and error in measurement of individual samples
- Add partial outline of new summary section
- Split article into multiple Rmd files
- Swap order of solutions and implications sections
- Revise section on theory of error in DA analysis (in progress)
- Various edits to other sections

--- 2021-07-17 article.Rmd
@@
-# Introduction
-
-Most microbiome research aims to go beyond qualitative descriptions to make quantitative conclusions about associations between specific microbes or community states and key host or environmental properties.
-But there is serious disagreement about whether the microbiome measurements made by marker-gene and metagenomic sequencing (MGS) are or can ever be truly quantitative.
-A primary reason is that taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
-As a result taxa can differ dramatically (e.g.
-10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—making the relative abundances we observe by MGS inaccurate representations of the actual sample compositions.
-Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads among taxa make this taxonomic bias a feature of shotgun metagenomic measurements as much as marker-gene measurements.
-Taxonomic bias is protocol specific (@mclaren2019cons) and can even vary among batches within an experiment (@yeh2018taxo); it thus not only makes MGS measurements inaccurate, but also incomparable between studies.
-Taxonomic bias thereby poses a major challenge for any seeking to draw quantitative conclusions from MGS studies.
-
-- On whether MGS can be quantitative: Is there a reference that explains the disagreement?
-Perhaps this question hasn't been explicitly addressed in the literature.
-I was gesturing towards a tension where on one hand there are many papers on the problems with community sequencing and on the other hand many papers (often by the same authors, and including us) are using it to make quantitative inferences.
-KL suggests Mallick et al., 2017 DOI 10.1186/s13059-017-1359-z; Or Barlow et al.
-2020 https://doi.org/10.1038/s41467-020-16224-6 as possible references.
-
-The field's current answer to this challenge is methods standardization, prioritizing consistency over accuracy.
-By treating each sample within a study in exactly the same manner, we can ensure that observed differences between samples are not simply due to methodological differences.
-Yet standardization does not guarantee that these observed differences reflect the actual differences in taxonomic composition among samples.
-Consider what might be the simplest and most common inference made from MGS: How does the proportion of a taxon vary between samples?
-An observation that *Escherichia* coli makes up 10% of the sequencing reads in a fecal sample might, thanks to bias, have arisen from a sample in which the proportion of *E.
-coli* is actually 50% or 1%.
-Yet many hope that if we observe the proportion of *E.
-coli* to increase from 10% to 20% between two samples, its proportion must have actually doubled—by fixing the taxonomic bias of our measurements through standardization, we have (so the argument goes) fixed the measurement error in both samples so that it will cancel when estimate the fold change.
-Unfortunately, the results of @mclaren2019cons show that reality is not so neat.
-Because the taxa compete for sequencing effort within a sample, the effect that taxonomic bias has on the observed proportions depends on the taxonomic composition of the sample.
-As a result, MGS measurements can lead to spurious inferences about changes across samples even when taxonomic bias is constant for each taxon across samples.
-
-- 'many hope' - consider make small tweaks to phrases like this to sound less critical
-
-This problem posed by taxonomic bias intersects with a second, growing concern within the microbiome field: that the relative abundances measured by MGS can present a misleading view of absolute-abundance dynamics.
-Without making further assumptions (which are often difficult or impossible to verify) one cannot tell whether a doubling in the proportion of a taxon reflects an increase in its abundance or a decrease in the abundances of other taxa.
-This concern has motivated the development of various experimental methods to enable researchers to convert the relative abundances measured by MGS into absolute abundances (methods to be reviewed in the Appendix in a future version).
-An increasingly popular approach in the study of both environmental and host-associated microbiomes involves making an independent measurement of total microbial abundance in each sample (@props2017abso,@kevorkian2018esti,@lloyd2020evid,@tettamantiboshier2020comp,@contijoch2019gutm,@vandeputte2017quan,@vieirasilva2019quan).
-This estimate is then multiplied by the MGS proportions to convert them to estimates of absolute abundance.
-This simple calibration procedure directly transfers the error in the MGS proportions due to taxonomic bias to the estimated absolute abundances, leading to estimates that are not only individually inaccurate but can imply inaccurate variation in abundance across samples.
-
-- Amy points to rao2021mult as another good ref for the above
-
-Here we summarize recent advances in our understanding of how taxonomic bias affects our ability to accurately determine how taxa vary in relative and absolute abundance across samples from different conditions—the inference problem known as *differential abundance (DA) analysis*.
-We first explain how taxonomic bias can lead to spurious DA results and consider the biological conditions in which such spurious results are likely.
-But we then describe several experimental and computational methods that can be used to avoid or correct these errors.
-Collectively, these methods may provide practical solutions to mitigating the effects of taxonomic bias on DA analysis for a majority of microbiome studies.
-
-## Notes for revision
-
-- introduce differential abundance as the focus earlier
-- improve introduction of absolute abundance vs. relative abundance, and methods for obtaining absolute abundances
-    + might include computational methods in scope, and use the term "calibration"; and ref to the future appendix that surveys the various methods
-    + consider giving general explanation and/or specific examples of why absolute abundances are important
-
-# Bias can cause spurious results in proportion-based DA analysis
-
-Comments on this version:
-
-- This version skips over the effect of bias on MGS measurement, leaving that for the intro, and goes right to the effect on FCs.
-I think this is a mistake; I'd rather walk the reader through the logic starting with the effect on one sample.
-- To accompany this, I'd like a new first figure, that illustrates the effect of competition. Essentially like Figure 2 of the eLife paper; perhaps using the vaginal mb example.
-
-```{r vaginal-example, cache = TRUE, fig.cap = '(ref:vaginal-example)'}
-svg_path <- here::here(
-  "notebook/_posts/2021-05-10-vaginal-microbiome-example",
-  "vaginal-microbiome-example_files/figure-html5",
-  "vaginal-example-proportions-timepoints-1.svg"
-)
-if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
-} else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
-}
-rm(svg_path)
-```
-
-(ref:vaginal-example) **Taxonomic bias can distort differential abundance results even when it is consistent for each taxon across samples.** Panel A shows the actual and observed proportions for hypothetical community samples from two time points, which differ in their relative abundance of three taxa.
-Panel B shows taxonomic bias in terms of the relative efficiencies of the three taxa against the mean efficiency of each sample; the difference between the taxon's efficiency and the sample's mean (vertical arrows) determines the fold error seen in Panel A.
-Panel C rearranges the plot from Panel A to show the actual and observed fold changes between time points.
-The efficiencies of individual taxa were estimated by @mclaren2019cons from mock community data from @brooks2015thet.
-The abundances are hypothetical but inspired by observations from the human vaginal microbiome; see main text.
-
-- TODO: Add main-message sentence; see Ben's note below
-
-We illustrate the problems posed by taxonomic bias for differential abundance using an example where the bias of an MGS protocol was explicitly measured using mock communities.
-To analyze the effects of taxonomic bias on measurements of the human vaginal microbiome, @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven key vaginal species using a similar protocol to that of the Vaginal Human Microbiome Project (VaHMP).
-@mclaren2019cons used this data to estimate the relative measurement efficiency of the seven species.
-They found that the most efficiently measured species, *Lactobacillus iners*, had an efficiency that was 29X that of the least efficiently measured species, *Gardnerella vaginalis*, primarily due to a greater DNA extraction efficiency and 16S copy number.
-A third species, *Atopobium vaginae*, had an efficiency 1.8X that of *G. vaginalis*.
-A common aim of human vaginal microbiome studies, including the VaHMP, is to understand the causes and consequences of shifts in communities from being dominated by a particular *Lactobacillus* species to dominance by a non-*Lactobacillus* species such as *G.
-vaginalis*, which have been associated with deleterious health outcomes including an increased risk of preterm births in pregnant women.
-It is therefore important to know whether the efficiency variation found by @mclaren2019cons affects the observed differential abundances between *Lactobacillus*-dominated and *Gardnerella*-dominated communities.
-
-- Ben flagged that we want to make sure that it is clear from the text that this example is not just a one-off case of bias mattering, but is an illustration of a general/universal problem.
-The later discussion helps with this already; but what will help more is to start this section with a main-message sentence stating the general problem first.
-
-Figure \@ref(fig:vaginal-example) illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by *Lactobacillus* to being dominated by *Gardnerella*.
-In each case, we observe a greater proportion of *Lactobacillus* than is truly present, and less of the other two species (Figure \@ref(fig:vaginal-example)A).
-This positive multiplicative error^[what is the right terminology for multiplicative error that is greater than 1?] arises because *Lactobacillus* is more efficiently measured than the average species in each sample.
-@mclaren2019cons showed that the observed and actual proportions of a taxon $i$ in a sample $s$ are related by the equation
-\begin{align}
-  (\#eq:error)
-  \text{observed}_{i}(s) = \text{actual}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
-\end{align}
-where $\text{mean efficiency}(s)=\sum_{j}\text{actual}_j(s)\cdot \text{efficiency}_j$ in the denominator is the average efficiency of all taxa in that sample.
-Intuitively, Equation \@ref(eq:error) says that the taxon will be over-represented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
-Figure \@ref(fig:vaginal-example)B shows the efficiencies of the three taxa and the mean efficiency in each sample.
-The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of *Lactobacillus*.
-The error in the second sample is larger—specifically, by 6X—than the first sample because the high proportion of *Gardnerella* has reduced the sample mean efficiency (and thus increased the efficiency of *Lactobacillus* relative to the mean) by 6X.
-
-- see note on terminology for mult. error
-
-The enrichment depends on the taxon’s efficiency relative to the sample average, rather than the efficiency itself, because the requirement that the proportions of all taxa sum to one locks the taxa in a zero-sum competition.
-Notably, competition occurs (though in a less absolute form) even at the level of the raw sequencing reads: DNA extractions and PCR reactions have limited output capacity; high-concentration samples are typically subsampled to fixed, lower concentrations; and sequencing libraries are given limited sequencing effort; so that the experimental process itself typically imposes competition to be sequenced.
-And it is competition that prevents the effects of taxonomic bias from canceling when we analyze how the proportions of individual taxa change across samples.
-
-- When revising the above paragraph, pay attention to making clear that altough we are imposing competition via normalization to proportions, the experiment itself imposes competition on the absolute intermediates like DNA concentration and sequenced fragments.
-
-Figure \@ref(fig:vaginal-example)C shows the consequences of measurement error on an analysis of the fold change in proportions between the two samples.
-In reality, *Gardnerella* changes by 4X, *Lactobacillus* by 0.1X, and *Atopobium* by 0.5X.
-Yet what is observed is that *Gardnerella* changes by 32X, *Lactobacillus* by 0.6X, and *Atopobium* by 3X.
-In other words, the change in each case is multiplied by 6X, causing a much larger increase in *Gardnerella* and a smaller decrease *Lactobacillus* than actual, and the appearance that *Atopobium* increased when it in fact decreased.
-Why do such errors occur?
-The multiplicative error in Equation \@ref(eq:error) has two parts: a taxon-specific term in the numerator that is the same in each sample, and a taxon-independent term in the denominator that depends on the composition of the sample through the mean efficiency.
-When we compute the fold change in a taxon i between samples s and t, 
-\begin{align}
-  \frac{\text{observed}_{i}(t)}{\text{observed}_{i}(s)}=\frac{\text{actual}_{i}(t)}{\text{actual}_{i}(s)}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{i}}*\frac{\text{mean efficiency}(s)}{\text{mean efficiency}(t)}
-\end{align}
-the constant efficiency in the numerator of Equation \@ref(eq:error) cancels, but the varying mean efficiency does not, leaving an error that is equal to the inverse change in mean efficiency.
-As a result, we are here left inferring that the fold-changes in proportions have all increased less (or decreased more) than they truly have.
-In our example, the mean efficiency decreased by 6X (from 15 to 2.5) due to the shift in dominant taxon from the high-efficiency *Lactobacillus* to the low-efficiency *Gardnerella*, causing the observed fold changes to increase by 6X above the actual.
-This increase distorted the magnitudes of the changes in *Lactobacillus* and *Gardnerella* (decreasing the first and increasing the second) and changed the sign of *Atopobium* from negative to positive.
-Thus partial cancelling in the effect of a constant multiplicative bias may not be sufficient to protect against spurious results.
-
-- TODO: Add strike-throughs to the above equation
-- Revision should hone the selection of equations
-
-Spurious results can occur for the same basic reason in proportion-based absolute-abundance inference.
-Suppose that the true total abundance in each sample can be determined experimentally.
-Because the absolute abundance estimate simply multiplies the erroneous observed proportions by this total, the estimated change in absolute abundance is wrong by the same factor as the fold change in proportions.
-In our Figure \@ref(fig:vaginal-example) example, if the change in total abundance between samples were 2X, then we would observed both *Lactobacillus* and *Atopobium* to have increased in absolute abundance (by 1.2X and 6X), when in fact *Lactobacillus* decreased (by 5X) and *Atopobium* remained the same.
-
-## Effect on regression of multiple samples
-
-We showed how fold changes of individual samples may be unreliable.
-But microbiome analyses often do not interpret such individual differences, and instead look for average patterns across many samples, such as how the abundance of a taxon varies with some covariate of interest.
-The covariate may be discrete, such as whether the sample is from a healthy or sick person, or it may be continuous, such as a measure of space, time, or temperature.
-Often these analyses can be framed as a regression problem.
-For example, we might hypothesize that the log absolute abundance of taxon $i$ changes with a variable $x$ according to the simple linear regression, 
-\begin{align}
-  E[\text{log abundance}_{i} \mid x] = a_{0} + a_{1}x
-\end{align}
-where $x$ is either continuous (e.g., sediment depth) or binary (e.g., $x=1$ for treated patients and $x=0$ for controls).
-Much of the effect of bias will be absorbed by the model intercept, which is typically not of interest to the researcher.
-However, variation in the mean efficiency creates additional variation (reducing statistical power to detect variation in the slope) and can also create a statistical bias in our slope estimate equal to the inverse slope of $E[\text{log mean efficiency}\mid x]$ as a function of $x$.
-Therefore if the mean efficiency varies across samples, but is not associated with the covariate $x$, its effect may simply be to create larger standard errors in our slope estimate.
-But the larger worry may be the systematically distorted slope ($a_1$) estimates that arise in situations where the mean efficiency changes with $x$.
-Notably, the error in an absolute sense is the same for all taxa—the slope is always reduced by the same amount—but the implications differ for various taxa depending on their slopes, causing magnitude or sign errors depending how the coefficient derived from the (unknown) true abundances compares to that of the mean efficiency (Figure \@ref(fig:regression-example)).
-
-- on "(e.g., sediment depth)": Want to confirm with KL that I'm understanding this correctly.
-She notes: "Even though these values are sequential, we use them as discrete variables because the samples are only from one depth each.
-I'm not sure if this distinction matters here." To which I replied: "right....you first converted sediment depth to time, and then regressed log abundance on time to estimate growth rate?
-(new line) Here I was just using sediment depth as an example variable, but I could replace with time to more match your analysis"
-- Consider different term than 'slope' since it's not a slope in the binary case
-- Comment from AW indicated that it was hard to follow the shift from describing the hypothesized absolute abundance regression model to describing the effect of bias on the fitted model.
-It could help to spell out this transition and make clear we're back to discussing the situation where the absolute abundance measurements are observed (biased) relative abundance measurements multiplied by some observation of the total.
-- Could be some potential for confusion in that the text notes we can get magnitude and sign errors for different taxa in the sample samples, but the figure shows a magnitude and sign error in two different sample sets on the same taxon.
-It could therefore be a better illustration to use two different taxa , both on the Eastern genotypes; We'll get a sign error for Penicillium but a magnitude error for any taxon that decreases more.
-
-```{r regression-example, fig.cap = '(ref:regression-example)'}
-svg_path <- here::here("figures/leopold2020host-example.svg")
-if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
-} else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
-}
-rm(svg_path)
-```
-
-(ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.** Details to synthesize into a caption: Regression of log2(Proportion) of the commensal fungus Penicillium versus timepoint; timepoints 1 and 2 are pre- and post-challenge with the pathogen Melampsora.
-Data is split by the region the host plants are derived from (Eastern and Western US).
-Calibrated proportions = Observed proportions in the real, experimental samples after adjustment for the bias measured in mock communities.
-Mean efficiency of each community is inferred by treating the calibrated proportions as the truth, and multiplying by efficiencies estimated from the mocks.
-The pathogen Melampsora has a high measurement efficiency; thus once it infects the plants, the mean efficiency of the sample increases (purple points).
-Efficiency is here taken as relative to the focal taxon Penicillium.
-West plants tend to be more resistant to the pathogen, which likely explains why the mean efficiency doesn't increase as much in the West plants.
-Penicillium is observed to decrease in log proportion, in both the East and West plants.
-But the calibrated measurements show that it actually slightly increases in the East plants, and has a lesser decrease in the West plants than what was observed before bias correction.
-The difference between the Calibrated and Observed data points and regression lines equals the regression line of the mean efficiency: Orange = Green + Purple; Green = Orange - Purple.
-The absolute error in regression coefficients is the same for all taxa.
-I picked Penicillium for illustration since it has the smallest observed decrease, which makes the error due to bias have a particularly significant impact.
-
-- Note about this figure: It is currently framed as O = A - \bar B to show the effect of error on inferred DA results; it could be flipped to A = O + \bar B to show how a correction might be done.
-
-# Implications for real-world inference
-
-These observations show that the effects of taxonomic bias do in fact cause error in proportion-based differential abundance analyses.
-Yet if the mean efficiency varies much less across samples than the abundances of individual taxa, this error will be negligible.
-Alternatively, if variation in the mean efficiency is effectively random (not associated with regression covariates), the effect of bias may be to increase the residual variance and thus reduce our statistical power to assay differential abundance—unfortunate to be sure, but perhaps not as much as confidently inferring changes with the wrong sign or magnitude.
-But there are salient biological scenarios where substantial associations between the mean efficiency and covariate can in fact lead to spurious differential abundance inference.
-
-## Scenarios in which bias may cause spurious results
-
-Figure \@ref(fig:vaginal-example) illustrates how a change in dominance by from a high- to low- efficiency taxon (or vice versa) can easily create spurious DA results.
-Indeed, this scenario may be particularly common in vaginal microbiome studies.
-Human microbiomes often dominated by *Lactobacillus*, but can become dominated by non-*Lactobacillus* species such as *Gardnerella* vaginalis and *Atopobium vaginae*, and these low-*Lactobacillus* states are associated with the disease bacterial vaginosis and, in pregnant women, an increased risk of preterm birth.
-Using mock community measurements from @brooks2015thet, @mclaren2019cons showed that the efficiency of two common *Lactobacillus* species were 20-30X greater than *G. vaginalis* and *A. vaginae*.
-Therefore we should expect samples dominated by *Lactobacillus* to have a much larger mean efficiency than those those dominated by *G.
-vaginalis* and *A.
-vaginae*, which would distort the observed associations of all taxa with any covariate that is also associated with *Lactobacillus* dominance.
-The extraction and 16S sequencing protocol used by @brooks2015thet is similar to that used in the VaHMP MOMs-PI study also led by these researchers to investigate the associations of these taxa with preterm birth (@fettweis2019thev), demonstrating the potential for serious real-world consequences for these statistical/technical concerns.
-
-- Discussion of the vaginal example in the above paragraph should be distilled in conjunction with the discussion of this example in the previous section
-- Group-B strep / S. agalactiae is another "pathobiont" that has a low efficiency and I might as well included here for completeness. Can cite brooks2015thet and pace2021comp
-
-Similar dynamics occur in a plant-fungal interactions experiment performed by @leopold2020host (Figure \@ref(fig:regression-example)).
-The authors inoculated commensally-colonized trees with a fungal pathogen.
-Using DNA mock communities, they showed that (excluding bias from DNA extraction) the pathogen was 10X more efficiently measured than the median commensal and 40X more efficiently measured than the lowest-efficiency commensal.
-In most hosts, the pathogen increased rapidly between timepoints, driving an increase in the sample mean efficiency that in turn leads to larger decreases being observed in the proportions of commensal taxa than what is predicted using calibrated (bias-corrected) values (REF APPENDIX).
-For example, the calibrated proportions indicate that the commensal *Penicillium* slightly increased in hosts from the Eastern US and slightly decreased in hosts from the Western US.
-Yet the observed (uncalibrated) proportions show moderate and large decreases, respectively (Figure \@ref(fig:regression-example) above).
-
-- In last sentence, insert the the values of the LFCs (ideally with evaluated R code)
-
-These examples involved scenarios where individual taxa can constitute a large proportion of the community.
-But in samples from highly-diverse ecosystems such as human gut, soil, and ocean sediment, one species rarely dominates.
-In these cases it might seem intuitive that the mean efficiency should be relatively stable across samples: As diversity increases, the mean efficiency effectively averages over efficiencies from a greater number of taxa and so (by the central limit theorem) will converge to a constant value—if taxon efficiencies are statistically independent of taxon abundances (REF APPENDIX).
-In practice, associations between efficiencies and abundances may be common and can allow large shifts in mean efficiency to occur even in highly diverse samples.
-
-- on the example of 'ocean sediment' in the above P: I asked KL "is it reasonable to characterize the ocean sediment communities from your 2020 paper as "high diversity"?
-In particular, species- or OTU-level Inverse Simpson values are what's relevant here.
-I'm guessing it's intermediate diversity - much less than gut and soil, but much more than the vaginal microbiome", and she replied "The total microbial diversity in marine sediments worldwide is similar to that of soil.
-I'll email you Hoshino et al., 2020 which addresses this." However I'm not sure if the high diversity KL is noting applies to individual soil samples (i.e.
-what scale)
-- Should make sure the above paragraph's claims are precise and accurate; see AW's feedback
-- "As diversity increases" specifically refers to inverse simpson diversity / q=2. Should I explicitly state that?
-- "statistically independent of taxon abundances" - I need to make sure I'm being accurate/precise here.
-What do we actually know?
-Maybe simpler to say, if taxon efficiencies are not correlated with taxon abundance
-
-The human gut provides a clear example of how such associations might arise.
-Gut microbiomes typically have high diversity at the species level but are dominated by just a small number of phyla, and two in particular: The Bacteroidetes and the Firmicutes.
-The ratio of Bacteroidetes to Firmicutes can shift substantially between individuals and has been linked to a number of host traits and health conditions.
-In addition, many DNA extraction protocols more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though there can also be significant variation in extraction efficiency among species within these phyla (@mclaren2019cons).
-For such protocols, we should expect Bacteroidetes-dominated samples to have substantially larger mean efficiencies than Firmicutes-dominated samples, regardless of species-level diversity.
-This association of efficiency with phylum abundance will distort DA inferences at any taxonomic level if the covariate is also associated with these phyla.
-
-- mclaren2019cons measures overall extraction efficiency, not lysis efficiency; is that important for the above text?
-
-This example illustrates a general mechanism by which association between the abundances and efficiencies of taxa are created by the common influence of evolutionary history.
-Just as shared evolutionary history creates positive associations in ecological traits that drives positive correlations in how related taxa vary in abundance across samples, so does it create positive associations in bias-related traits that can lead to similar efficiencies among closely related taxa.
-Such phylogenically-associated, bias-affecting traits include cell-wall toughness, ribosomal-operon copy number, binding affinity for a given set of primers, and representativeness in taxonomic databases.
-Ecology and efficiency are both affected by the same evolutionary history, leading to positive associations between the two: A change in diet that increases the relative abundances of many Bacteroidetes species also increases the relative abundance of many easy-to-lyse species.
-
-- first sentence in above P: "the abundances and efficiencies" - the idea is that both are determined by genetically encoded traits
-- this is prob a tricky idea for readers and will have to take care in communicating it, and find nice refs
-
-Associations between measurement efficiency and relative abundance can also arise because a single trait affects both.
-For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
-@lloyd2020evid estimate the log fold change in the estimated absolute abundance of various taxa with sediment depth (as a proxy for time) to determine which taxa are able to persist and even grow in this difficult environment.
-It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
-As the relative abundance of tougher species increases with depth, the mean extraction efficiency decreases.
-This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
-Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.
-
-### Error in absolute abundance measurements
-
-- Ben flags that this concept/section main idea should be highlighted more
-
-So far we have ignored error associated with our absolute abundance measurements.
-In fact, these measurements can also have a tendency to measure contributions from some taxa more efficiently than others.
-**TODO: State how this could create problems.
-Might also note the ambiguity in what the truth/target is (e.g.
-live cells, or all cells; cells, or 16S rRNA gene copies; observations in Jian, Salonen, and Korpela 2021).**
-Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
-16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
-Yet these biases are shared by the sequencing measurement.
-Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
-(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
-If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.^[See discussion below]
-Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
-
-- from KL on qPCR being a bad proxy: We showed this empirically for all published measurements in marine sediments in Lloyd et al., 2013 doi:10.1128/AEM.02090-13. Apologies for the self-citation.
-- In the above sentence "If common primers are used...", KL asked a question that shows I'm not being clear enough about what I'm talking about here.
-KL: "Wouldn't you want to use taxa specific primers for qPCR and general primers for the amplicon libraries to do this analysis?
-If they're the same primer sets, then it doesn't tell you anything about the relative abundance of individual taxa within the population grabbed by the primers." my reply: "The "solutions" section talks about using targeted qPCR; this section is talking about qPCR with "universal" primers for the purpose of estimating the total absolute abundance, for use in FRAxC instead of cell counting.
-There have been quite a few papers doing this recently and 16S sequencing companies are starting to offer it as a service so I've been thinking through the implications.
-This section needs to be expanded and the point I'm making about using qPCR for FRAxC is still somewhat preliminary and something we should chat about"
-- should caveat something like 'Unfortunately, qPCR shares a limitation with DNA-based measurements: It assumes linearity of DNA extraction'
-    - this should start own paragraph on this issue. For now, can be a 2 sentence paragraph and ref the appendix.
-
-## We cannot simply ignore taxonomic bias
-
-These examples show that there is potential for bias to distort DA results in both low and high diversity settings.
-We therefore need methods to control or correct for bias to ensure the validity of future results, as well as methods to probe the robustness of past results.
-It is possible that spurious results are unlikely in many experimental contexts.
-By applying such methods across a range of biological and experimental contexts we will deepen our understanding of when bias is unlikely to significantly distort biological findings and when additional measures are needed to mitigate its effects.
-
 # Potential solutions
 
-## Using control samples to calibrate relative abundances
+## Use control communities to correct relative abundances
 
@@
 
-An closely-related alternative to mocks are controls derivedd from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (@mclaren2019cons).
+An closely-related alternative to mocks are controls derived from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (@mclaren2019cons).
 A natural fecal standard is currently being developed by [NIST](https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material) and one has recently been made available commercially by Zymo Research ([ZymoBIOMICS Fecal Reference](https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology)).
@@
 
-## Ratio-based relative and absolute abundance inference
+_Except in synthetic community experiments, we are always likely to have taxa not in our controls.
+We can attempt to use phylogenetic prediction or other modeling of efficiencies to predict the efficiencies of other taxa; however there will be signifciant undertainty.
+Therefore our control measurements may ideally serve to inform a bias-sensitivity analysis of the sort described below._
 
-A comprehensive approach of estimating the efficiency of all taxa is needed for getting calibrated relative abundances of all taxa within individual samples.
-But what if we only want calibrated DA analysis?
-That is, we are comfortable with not knowing whether *E. coli* is 10% of our sample if we can confidently determine that it doubled or halved between sample conditions?
-This problem is easier to solve, and one solution is to use analysis methods based on the ratios among taxa rather than the proportions of individual taxa.
+For revision:
 
-- from gdoc: this distinction of ratios vs proportions needs to be spelled out more explicitly; could even use my seminar figure (or equations) to help
-- later thoughts: This distinction is so central for how I understand and explain the key ideas, should refer to it early and often
+- Discuss the vaginal mb and leopold examples
 
-A subset of methods for analyzing differential relative abundance are derived from the field of Compositional Data Analysis (CoDA).
-The defining feature of CoDA methods is that they are based on fold-changes in ratios among elements (here, the taxa) (REF AITCHISON).
-The use of CoDA methods in microbiome analysis has largely been motivated by concerns over the negative correlations between taxa induced by the sum-to-one constraint in taxon proportions.
-But @mclaren2019cons showed that, due to their property of perturbation invariance (REF AITCHISON), the results of CoDA differential relative abundance analysis are invariant to consistent taxonomic bias.
-From Equation \@ref(eq:error), the observed ratio of a taxon $i$ to a taxon $j$ in a sample $s$ is
-\begin{align}
-  \frac{\text{observed}_{i}}{\text{observed}_{j}}=\frac{\text{actual}_{i}}{\text{actual}_{j}}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}
-\end{align}
-The error is independent of the sample composition and thus cancels in any function that is a fold-change of this (or any) ratio of taxa between samples.
-(Perhaps insert example from Figure \@ref(fig:vaginal-example)ABC all giving same ratio FCs.)
+## Use control measurements to estimate error
 
-Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
-In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
-Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
-To estimate the abundance of a focal taxon $i$, this approach multiplies the ratio of reads (or proportions) of taxon i to a reference taxon $r$ by a known or estimated abundance of the reference taxon, (while ignoring bias)
-\begin{align}
-  \text{estimate}_{i} = \frac{\text{observed}_{i}}{\text{observed}_{r}}*\text{abundance}_{r}
-\end{align}
-So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
-Yet the reference taxon could also be a naturally occurring taxon whose absolute abundance we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.
-Because taxa do not compete (or competition is greatly reduced) in such targeted measurements, any taxonomic bias associated with them is expected to create constant multiplicative error across samples and so not affect fold change estimates [APPENDIX].
-APPENDIX describe various theoretical and experimental considerations for both approaches.
-
-- note from KL on choice of reference taxon: "It seems like you'd have to be pretty careful to pick a reference taxon that was not changing a lot between samples."
-    - this comment indicates I need to be clearer about the three reference strategies - assume constant AA; (but may not be available so can) spike-in at constant AA; (but that might not be convenient, so can) measure AA to adjust for changes in AA.
-    - my reply in the doc: "the idea is that either the reference is a taxon spiked at a fixed abundance to each sample, or it is a taxon you've measured independently with ddPCR or qPCR and so can account for its variation through the given equation." "There are also computational methods that attempt to identify taxa that don't vary (under the assumption that most taxa don't vary), which is an approach inspired from the RNAseq/gene expression literature (where it is probably a more valid assumption that most genes have consistent expression); Ben in his comments was suggesting we mention this approach too.
-"
-- On taxa not competing in targeted measures: "I think DNA-based measurements would still be affected by competition if DNA extraction yield is a strongly saturating function of input.
-But if this saturation isn't very strong then the effect should remain much less than the MGS measurement, which also has competition imposed by various normalization steps "
-- on 'considerations' noted in the last sentence: "perhaps most notable is how to deal with a given reference taxon not being present or detectable in every sample "
-- clarify that there are 2 ways to use spike-ins: Proportion-based and ratio-based method; and only the ratio-based method is bias-invariant
-
-## Estimating error with targeted measurements of a small number of taxa
+targeted measurements of a small number of taxa
 
@@
 
-## Computational approaches
+## Use ratio-based estimation
+
+A comprehensive approach of estimating the efficiency of all taxa is needed for getting calibrated relative abundances of all taxa within individual samples.
+But what if we only want calibrated DA analysis?
+That is, we are comfortable with not knowing whether *E. coli* is 10% of our sample if we can confidently determine that it doubled or halved between sample conditions?
+This problem is easier to solve, and one solution is to use analysis methods based on the ratios among taxa rather than the proportions of individual taxa.
+
+- from gdoc: this distinction of ratios vs proportions needs to be spelled out more explicitly; could even use my seminar figure (or equations) to help
+- later thoughts: This distinction is so central for how I understand and explain the key ideas, should refer to it early and often
+
+A subset of methods for analyzing differential relative abundance are derived from the field of Compositional Data Analysis (CoDA).
+The defining feature of CoDA methods is that they are based on fold-changes in ratios among elements (here, the taxa) (REF AITCHISON).
+The use of CoDA methods in microbiome analysis has largely been motivated by concerns over the negative correlations between taxa induced by the sum-to-one constraint in taxon proportions.
+But @mclaren2019cons showed that, due to their property of perturbation invariance (REF AITCHISON), the results of CoDA differential relative abundance analysis are invariant to consistent taxonomic bias.
+From Equation \@ref(eq:error), the observed ratio of a taxon $i$ to a taxon $j$ in a sample $s$ is
+\begin{align}
+  \frac{\text{observed}_{i}}{\text{observed}_{j}}=\frac{\text{actual}_{i}}{\text{actual}_{j}}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}
+\end{align}
+The error is independent of the sample composition and thus cancels in any function that is a fold-change of this (or any) ratio of taxa between samples.
+(Perhaps insert example from Figure \@ref(fig:vaginal-example)ABC all giving same ratio FCs.)
+
+Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
+In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
+Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
+To estimate the abundance of a focal taxon $i$, this approach multiplies the ratio of reads (or proportions) of taxon i to a reference taxon $r$ by a known or estimated abundance of the reference taxon, (while ignoring bias)
+\begin{align}
+  \text{estimate}_{i} = \frac{\text{observed}_{i}}{\text{observed}_{r}}*\text{abundance}_{r}
+\end{align}
+So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
+Yet the reference taxon could also be a naturally occurring taxon whose absolute abundance we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.
+Because taxa do not compete (or competition is greatly reduced) in such targeted measurements, any taxonomic bias associated with them is expected to create constant multiplicative error across samples and so not affect fold change estimates [APPENDIX].
+APPENDIX describe various theoretical and experimental considerations for both approaches.
+
+- note from KL on choice of reference taxon: "It seems like you'd have to be pretty careful to pick a reference taxon that was not changing a lot between samples."
+    - this comment indicates I need to be clearer about the three reference strategies - assume constant AA; (but may not be available so can) spike-in at constant AA; (but that might not be convenient, so can) measure AA to adjust for changes in AA.
+    - my reply in the doc: "the idea is that either the reference is a taxon spiked at a fixed abundance to each sample, or it is a taxon you've measured independently with ddPCR or qPCR and so can account for its variation through the given equation." "There are also computational methods that attempt to identify taxa that don't vary (under the assumption that most taxa don't vary), which is an approach inspired from the RNAseq/gene expression literature (where it is probably a more valid assumption that most genes have consistent expression); Ben in his comments was suggesting we mention this approach too.
+"
+- On taxa not competing in targeted measures: "I think DNA-based measurements would still be affected by competition if DNA extraction yield is a strongly saturating function of input.
+But if this saturation isn't very strong then the effect should remain much less than the MGS measurement, which also has competition imposed by various normalization steps "
+- on 'considerations' noted in the last sentence: "perhaps most notable is how to deal with a given reference taxon not being present or detectable in every sample "
+- clarify that there are 2 ways to use spike-ins: Proportion-based and ratio-based method; and only the ratio-based method is bias-invariant
+
+## Perform a computational sensitivity analysis
 
@@
 
-**Bias sensitivity analysis:**
 A straightforward and universally available approach is to use computer simulation to determine how the results change under a range of possible sets of efficiencies, which can be randomly generated to reflect certain hypotheses about bias in the given system.
 The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
-Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—not just for differential absolute abundance, but for all microbiome analyses.
+Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance as well as other microbiome analyses such as alpha diversity and community similarity analyses.
 
-**Bias-aware meta-analysis:**
+## Use bias-aware meta-analysis
+
 When the goal is to perform a meta-analysis that combines studies that have used different protocols, the unknown measurement efficiencies of each protocol can be explicitly included as parameters of the statistical model that is used.

### 2021-07-19 summary.Rmd
Revise manuscript (in progress)
- Revise introduction
- Add section on model and error in measurement of individual samples
- Add partial outline of new summary section
- Split article into multiple Rmd files
- Swap order of solutions and implications sections
- Revise section on theory of error in DA analysis (in progress)
- Various edits to other sections

# Background and summary of results {#summary}

outline:

- DA analysis
  + abundance: absolute abundance e.g. cell density; relative abundance (e.g. proportions, ratios); other transformations are commonly used, but we focus on these b/c.
    - absolute abundance is more directly connected to ecological processes affecting the microbe and the host and thus may be more informative for host and environmental impact.
  + (log) fold changes: Microbiome DA most often done on the log scale (b/c different organisms might have vastly different average densities, and because of the connection with the multiplicative growth process underlying ecological dynamics). DA analysis is not interested in the average abundance differences in organisms, but in the changes - hence the focus elsewhere and here on log fold changes (LFCs) of a taxon across samples.
  + (linear) regression; intercepts and slopes: (take )
- error in DA analysis
  + generally not interested in the average abudnance, but in the change across conditions; this means the LFC for pair of samples or the slope in the regression model.
  + hope that the effect of bias absorbed in the intercept, leaving the slope estimate unaffected.
- consistent vs inconsistent multiplicative (or fold) error
  + for accurate FCs
    - need consistent multiplicative (or fold) error; then the error cancels in the FC.
    - inconsistent FC, does not; this can lead to sign errors (estimating a positive LFC when truly negative, or vice versa) or magnitude errors (estimating an LFC with the correct sign but a larger or smaller absolute value than the true value)
  + for accurate regression slopes
    - b/c analyzing many samples, a randomly varying error serves to add noise without affecting the expected slope - leads to noisier and less precise slope estimates.
    - correlation of error with covariate leads to statistical bias (systematic error) in the slope estimate, leading to magnitude or sign errors in the slope estimate
- HERE HERE
- relative vs. absolute abundance
- ratios vs. proportions
- sample mean efficiency
- reference taxa
- todo: illustrate these ideas with a schematic figure

clarifications:
- Microbiome scientists are generally not really after absolute abundance in the sense of an absolute number of cells. By "absolute abundance" they generally mean abundance relative to some non-taxonomic property, such as the total volume or (wet or dry) mass of the sample.
  - Write this here for now, but might want to make a footnote.

thoughts

- use of models; (perhaps introduce role of models in measurement? or leave for next section)
- define scope to all current AA methods; distinguish based on how affected by bias and other concerns like non-linear extraction, and suggest a new approach that is less affected by bias
    - Note that there are a variety of AA methods which make different assumptions about MGS and are differently interact with bias; we survey/taxonomize/synthesize them
- might include computational methods in scope, and use the term "calibration"; and ref to the future appendix that surveys the various methods
- need to think more about the first sentence
- Might want to hear mention the ratio and proportion divide

### 2021-07-21 index.Rmd
Update syntax for block environments
Theorem (and similar) environments and custom block environments are
updated to use the new Bookdown syntax based on Pandoc's fenced syntax
for Div blocks; see the Bookdown book for more information. Resolves #7.

--- 2021-05-12 index.Rmd
@@
 
+::: {#preface-warning .rmdnote}
+
+**This in-progress manuscript is not intended for general scientific use.**
+It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
+Please post comments or questions on the [GitHub Issues page](https://github.com/mikemc/differential-abundance-theory/issues) or [email Mike](m.mclaren42@gmail.com).
+
+:::
+
 This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.

### 2021-07-26 index.Rmd
Add HTML header to load MathJax extension

@@
     number_sections: true
+    includes:
+      in_header: header.html
     config:

### 2021-08-03 introduction.Rmd
Revise introduction

@@
-# Introduction
+# Introduction {#introduction}
 
@@
 
-The most basic question we can ask about microbial communities after which taxa are found in an ecosystem is, how do the abundances of these taxa vary across space, time, and host or environmental conditions?
-Advances in sequencing technology allow us to now measure these _differential abundances_ simultaneously for 100s to 1000s of species using marker-gene and shotgun-metagenomic sequencing (jointly, MGS).
-Standard MGS measurements do not preserve information about the total microbial density in our sample and thus are most often used to make observations about the relative abundaces of taxa.
-But increasingly, MGS protocols are being modified or combined with additional measurements to
-allow analysis of changes in absolute quantities such as cell density, which are more directly connected to ecological processes and thus may be more informative for host and environmental impact.
-
-- (say something about how MGS-DA has become an indespensible tool with vast use in basic science and applications)
-- End on with tension: Lack of reproducibility, questions over how quantitative they are, trustworthiness in critical applications, not ready for 'game time'
-
-notes
-
-- when mention r.a.'s for the first time: good spot to introduce the proportions versus ratios distinction, to set up contrast in the bias+DA paragraph, and approach/summary paragraph
+The most basic question we can ask about microbial communities after which taxa are found in an ecosystem is, how do the abundances of these taxa vary in their abundance—across space, time, and host or environmental conditions?
+Advances in sequencing technology allow us to now simultaneously measure the abundances of 100s to 1000s of species using marker-gene and shotgun-metagenomic sequencing (jointly, MGS).
+Although standard MGS measurements lose information about total microbial density—and so are typically used to analyze the abundances of taxa relative to each other or their total—new studies are increasingly employing strategies to enable the analysis of cell density or other measures of "absolute abundance."
+These relative and absolute abundances serve as the basis for a _differential-abundance (DA) analysis_, in which the change in abundance of a microbial taxon across samples or conditions is used to learn about the biology of the taxon and its impact on the host and other microbes as well as to detect predictive biomarkers of host and environmental health and disease.
 
 <!-- P: Taxonomic bias -->
-
-Primary reason for doubt: MGS measurements are taxonomically biased.
-Taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
-As a result taxa can differ dramatically (e.g.
-10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—making the relative abundances we observe by MGS inaccurate representations of the actual sample compositions.
-Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads among taxa make this taxonomic bias a feature of shotgun metagenomic measurements as much as marker-gene measurements.
-Taxonomic bias is protocol specific (@mclaren2019cons) and can even vary among batches within an experiment (@yeh2018taxo); it thus not only makes MGS measurements inaccurate, but also incomparable between studies.
-
-- might mention that bias can have an effect that is larger than important biological differences; can site lozupone2013meta and others.
-Perhaps put in the next paragraph.
-Point is that we know bias has a big effect on individual sample compositions, but does it have a big effect on differences between samples?
+Although MGS-based DA analysis has been widely deployed and achieved many notable successes, it faces serious concerns over accuracy and reproducibility due to the inherent technical limitations of MGS measurements.
+In particular, MGS measurements are _taxonomically biased_: Taxa vary dramatically (e.g.
+10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—by a given MGS protocol.
+As a result, the abundance measurements obtained by MGS are inaccurate representations of the actual abundances and also tend to differ across protocols (@mclaren2019cons), studies, and even experimental batches (@yeh2018taxo).
+This bias arises from variation in how taxa respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
+Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads make taxonomic bias a feature of both shotgun and marker-gene measurements.
+The error it causes have been found to in some cases to supersede sizeable biological differences (e.g.
+@lozupone2013meta) and has plausibly caused replication failures for prominent findings such as the association of decreased Bacteroides and increased Firmicutes in stool with obesity (@finucane2014atax) and the association of certain taxa in the vagina of pregnant women with preterm birth (@callahan2017repl).
 
@@
 
-The dominant answer to taxonomic bias is to standardize protocols within an experiment and only compare differences between samples measured by the same protocol (and ideally even the same experimental batch).
-The reasoning goes that so long as all samples are subject to the same taxonomic bias, its effects should cancel when we compute changes in a particular taxon's abundance across samples.
-However, @mclaren2019cons demonstrated that taxonomic bias can cause spurious changes in the proportion of a taxon, in which a taxon is observed to change more or less—or even in the opposite direction—than it really did.
-(The leading?) A prominant method for absolute DA analysis multiplies the MGS proportions by an estimate of the total, and so is subject to the same problem.
-Advancing the role of DA analysis in mb science and applications requires that we (the community) do a thorough examination of the impact that taxonomic bias on DA analysis and develop methods to correct or otherwise mitigate its effects.
+Although many microbiome researchers are aware of taxonomic bias, it is standard practice to ignore it when performing a DA analysis and interpretting its results.
+One reason for this practice is a widespread belief that taxonomic bias does not affect the inferred _differences between samples_, so long as those samples have been subjected to the same measurement protocol.
+This belief stems from an assumption that the effect of taxonomic bias is to consistently increase or decrease the measured abundance of a given taxon in every sample, so that its effects cancels when computing change across samples.
+@kevorkian2018esti (see also @lloyd2020evid) were, to our knowledge, the first to make this argument mathematically explicit; however, their argument was based on a faulty assumption about the effect of bias on MGS measurement (@mclaren2019cons and Section \@ref(abundance)).
+@mclaren2019cons subsequently analyzed the effects of bias using a mathematical model of MGS measurement, which they validated using marker-gene and shotgun measurements of artificially-assembled (or "mock") bacterial communities.
+They found that the _relative_ measurement efficiencies among taxa are consistent across samples that vary in taxonomic composition (provided they are measured by the same protocol).
+But a bias that is consistent in terms of relative efficiencies will nevertheless cause error in the proportion of a taxon that depends on sample composition and hence can lead to spurious estimated changes across samples.
+In contrast, they showed that error in the ratio between a pair of taxa has an error that is sample-independent error and hence does in fact cancel when estimating log fold changes in the ratio.
+Although they considered measures of relative abundance (proportions and ratios), the same underlying error dynamics apply to methods for inferring cell densities from MGS, which themselves are based on MGS-derived proportions or ratios.
+Together, these results of @mclaren2019cons indicate that taxonomic bias may lead to spurious DA results, but may be more problematic for certain experimental and analytical approaches than others.
 
-Future versions:
+<!-- P: What is needed? -->
 
-- Introduce ratio-based DA analysis and constrast w/ proportion-based, in such a way as to motivate the body structure
-- emphasize the importance of DA analysis of proportions for microbiome studies; make clear that the MWC result is based on the first accurate model of how taxonomic bias affects MGS measurements.
-Might want to mention others e.g.
-Brooks2016 on the more general point of spurious inference.
+Here we address a number of crucial remaining questions about the impact of taxonomic bias on DA analysis:
 
-<!-- P: Our approach + main message (TODO) -->
+- How do these results for the changes between a pair of samples extend to the proto-typical DA analysis of a (linear) regression over many samples?
+- How does taxonomic bias affect the wide variety of abundance measures, and methods for converting relative to absolute abundance, which are currently used in DA analyses?
+- What is the impact on DA analysis of taxonomic bias and other forms of error in the supplemental measurements (such as cytometry and bulk 16S qPCR) that are commonly combined with MGS to obtain absolute abundances?
+- What is the impact of variation in efficiency within a taxon on DA analysis of the taxon?
+(e.g.
+what is the impact of variation in efficiency among Firmicutes species on the infered change in Firmicute abundance?)
+- Last but not least, for which biological scenarios is taxonomic bias likely and not likely to lead to biologically significant inferential errors?
 
-Here we synthesize our current understanding of how taxonomic bias affects microbial differential abundance analysis and provide a roadmap for ???. 
-We first explain how taxonomic bias affects relative and absolute abundance measurements, and how the errors it creates can lead to spurious differential abundance inferences.
-We then consider the biological conditions in which such spurious results are likely.
-Finally, we turn to new and existing experimental and computational methods that can be used to avoid or correct these errors.
-Collectively, these methods may provide practical solutions to mitigating the effects of taxonomic bias on DA analysis for a majority of microbiome studies.
+We address these questions through a combination of mathematical analysis, simulation, and analysis and interpretation of data from previously published experiments.
 
-thoughts
+<!-- P: Solutions -->
 
-- use of models; (perhaps introduce role of models in measurement? or leave for next section)
-- define scope to all current AA methods; distinguish based on how affected by bias and other concerns like non-linear extraction, and suggest a new approach that is less affected by bias
-    - Note that there are a variety of AA methods which make different assumptions about MGS and are differently interact with bias; we survey/taxonomize/synthesize them
-- might include computational methods in scope, and use the term "calibration"; and ref to the future appendix that surveys the various methods
-- need to think more about the first sentence
-- Might want to hear mention the ratio and proportion divide
+A second reason that taxonomic bias is generally not addressed (beyond protocol standardization) is a severe lack of proposed, tractable solutions to address it.
+We propose and demonstrate several methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be widely deployed today with only modest changes to existing experimental and analytical workflows.
+Over time, application of these methods to past and future experiments will provide crucial quantitative information about taxonomic bias and the conditions under which spurious results arise for particular DA methodologies.
+Collectively, these methods and insights can provide a practical solution to taxonomic bias in DA analysis and the confidence that is necessary to codify our statistical findings into scientific knowledge and translate them into real-world diagnostics and interventions.

### 2021-08-04 abundance-measurement.Rmd
Revise manuscript (in progress)
- Begin drafting new section on measurement model
- Make minor edits and add notes to other sections
- Add refs to .bib file
- Remove summary and equations files from the rendered manuscript

@@
 # How taxonomic bias affects abundance measurements {#abundance}
+
+## Rewrite (in progress)
+
+Behind any quantitative measurement lies a mathematical model, either implicit or explicit, that relates the readings of our instruments and the results of our statistical analyses with the biological reality we are measuring.
+This section develops a model of microbiome measurement, based on that of @mclaren2019cons, to describe how the actual absolute compositions of microbiome samples are related to estimates of relative and absolute abundance obtained by MGS.
+The next section shows how these relationships impact our measurements of differential abundance.
+
+An MGS measurement of a microbiome sample turns a collection of biological source material (cells, viruses, and extracellular DNA) into a collection of taxonomically-assigned reads (or read pairs).
+For concreteness and simplicity, we suppose that all assigned reads come from cells and that the ground-truth biophysical quantity we wish to measure is the density of cells (by volume, or by wet or dry mass) belonging to various taxa, or relative abundances directly derived from these densities.
+This particular simplification reflects our own and the literature's prokaryotic bias; however, the features of MGS measurement that we explore also apply to situations (such as studies of viruses, fungi, or environmental DNA) where cell-based measurement may be inappropriate.
+We next suppose that there are a set of atomic taxa, which are mutually distinct and within which we assume the properties that determine taxonomic bias are conserved.
+For convenience, we identify these atomic taxa with species under a given taxonomic classification system, while acknowledging that whether taxonomic bias is conserved within commonly-accepted species has yet to be investigated in any detail.
+Since DA analysis is commonly performed at a variety of taxonomic levels, we consider measurement of species and of synthetic taxa (such as a phylum or a clade in a phylogeny) that consist of multiple species.
+We assume that reads assigned to a given taxon and sample really do come from that taxon and sample, but note that taxonomic mislabeling and contamination (from the reagents, lab environment, or other samples) pose further challenges to DA analysis.
+
+- might want to note that we're mostly going to talk as if species-level taxonomic assignment is possible.
+
+Consider the MGS measurement of a particular sample.
+For a given taxon, we define the _absolute efficiency_ as the number of reads assigned to that taxon divided by the number of cells in the original sample.
+The absolute efficiency depends not only on the ability of the chosen protocol to measure this taxon, but further aspects such as how much the sequencing library was diluted and how deeply the library was then sequenced.
+We suppose that the absolute efficiencies of species can be written as a product of a species-specific, sample-independent factor and a species-independent, sample-specific factor.
+The species-specific factors reflects the differential ability of a given protocol to measure different species (i.e.
+taxonomic bias), whereas the sample-specific factor captures the fact that the reads in an MGS measurement are not directly related to the absolute number or density of cells in a sample regardless of which taxa are being measured.
+When we consider the _relative efficiencies_—or the ratios among absolute efficiencies—the sample-specific, taxon-independent factor cancels.
+Relative efficiencies quantify how much more efficiently different taxa are measured relative to each other.
+When we say efficiency unqualified, we mean efficiency relative to some fixed, arbitrary species in the denominator.
+
+For species, the relative efficiencies are sample-independent.
+In contrast, the relative efficiency of a synthetic taxon varies depending on the relative densities of its constituent species,
+\begin{align}
+  \text{efficiency}_T(a) = \frac{\sum_{t \in T} \text{density}_t(a) \cdot \text{efficiency}_t}{\sum_{t \in T} \text{density}_t(a)}
+\end{align}
+(May need to explain or derive this.)
+For example, @mclaren2019cons found that for a particular shotgun-sequencing protocol, the relative efficiency was X between two Firmicute species X and Y;
+hence the efficiency of Firmicutes could vary by X depending on which of these two species were dominant in a given sample.
+
+From efficiencies, we can determine the error in relative abundances estimated by our MGS measurement.
+
+- atomic taxa
+- synthetic taxa
+
+MGS analysis is commonly done on proportions - the density of a taxon relative to some designated taxonomic whole, such as all prokaryotes.
+
+- give equation def of proportion
+- give error in proportion, from efficiency
+
+outline
+
+- Relative-abundance measurements
+
+- Absolute-abundance measurements
+
+questions, thoughts
+
+- should I define absolute efficiency in terms of cells or density?
+- need to revise the story of the first paragraphs above.
+
+## Current working version
 
@@
 Alhough there may be other advantages to this approach, the above observations suggest that it may be more susceptible to taxonomic bias in the sequencing measurement than using a ratio-based approach on the same underlying data.
+
+**Update:** Actually, the two methods are equivalent when the total density and proportion are estimated using the same reads.
+Let $B$ be all bacteria and $H$ be the host.
+Consider using WGS reads to estimate the absolute abundance of bacterial taxon $i$,
+\begin{align}
+  \hat{\text{density}}_{i}(s) 
+  &= 
+  \underbrace{\frac{\text{reads}_{B}}{\text{reads}_{H}}}_\text{est'd bact'l density} \cdot 
+  \underbrace{\frac{\text{reads}_{i}}{\text{reads}_{B}}}_\text{est'd prop'n}
+  = \frac{\text{reads}_{i}}{\text{reads}_{H}}
+\end{align}
+So the problem/difference arises when we the proportion is estimated either by a different technology (e.g.
+when density is estimated via WGS and proportion is estimated by 16S sequencing) or the two estimates going into the above are not just ratios of reads.
+For example, ratios of bacterial to host reads could be used for the density estimate, while the proportion is estimated by Metaphlan.

### 2021-08-04 differential-abundance.Rmd
Revise manuscript (in progress)
- Begin drafting new section on measurement model
- Make minor edits and add notes to other sections
- Add refs to .bib file
- Remove summary and equations files from the rendered manuscript

@@
 # How taxonomic bias affects differential-abundance analysis
+
+_The current text generally assumes that taxa are atomic._
 
@@
 <!-- P: error in FC ratios and ratio-based AAs -->
-(In contrast to proportion-based estimates) Becuase consistent taxonomic bias creates a consistent error in ratio-based estimates, the error completely cancels, leaving accurate fold changes.
+(In contrast to proportion-based estimates) Because consistent taxonomic bias creates a consistent error in ratio-based estimates, the error completely cancels, leaving accurate fold changes.
 From \@ref(eq:ratio-err), the estimated fold change in the ratio of taxon $i$ to taxon $j$ from sample $s$ to sample $t$ is
@@
   \cdot
-  \frac{\cancel{\text{efficiency}_{i}(t)}/\cancel{\text{efficiency}_{j}(t)}} {\cancel{\text{efficiency}_{i}(s)}/\cancel{\text{efficiency}_{j}(s)}}
+  \frac{\cancel{\text{efficiency}_{i}}/\cancel{\text{efficiency}_{j}}} {\cancel{\text{efficiency}_{i}}/\cancel{\text{efficiency}_{j}}}
   \\[0.5ex]&= 
@@
 (Describe FCs in Figure 2 and Figure 3 density e.g., seeing the correct FCs.)
+(_Need to make plain in the equation that this is the estimated fold change in the density of taxon i; perhaps have notation for this._)
+
+### Synthetic or higher-level taxa
+
+_Some of these concepts would best be introduced in the previous secton on abundances.
+Alternatively, we might want to separate them into a box or appendix on multiplicative aggregation._
+
+<!-- P: Synthetic taxa intro -->
+
+Frequently, relative- and absolute-abundance estimation are used to infer fold changes in higher-level taxa.
+
+- Visual or statistical estimation of changes in phylum abundance is a ubiquitous form of analysis in bacterial microbiome studies.
+- @chng2020meta - use fold changes in the ratio of bacterial to host reads to estimate the change in bacterial biomass during mouse antibiotics experiment
+- @rao2021mult - use a spike-in to infer changes in total domain abundance for each of archaea, bacteria, and fungi.
+- @karasov2020ther - ratio of microbial to host reads to infer changes in microbial biomass
+
+We know that efficiencies vary widely within domains and also has been observed to vary within phyla and even within species.
+What are the implications of this within-taxon variation for infered fold changes?
+
+<!-- P: Synthetic taxa results -->
+
+Unfortunately, the bias invariance of ratio-based inference of species need not apply to ratios involving higher-level taxa.
+The estimated fold change in the proportion, ratio, or density of a synthetic taxon $T$ has additional multiplicative error equal to its fold change in efficiency.
+This error can be substantial when there is a substantial turnover from dominance by high- to low-efficiency species (or vice versa).
+(todo: ref relevant equation from previous section)
+
+An alternative approach to creating synthetic quantitative features was develoepd in the Compositional Data Analysis field that retains the so-called perturbation invariance property of atomic features.
+These "balances" are constructed by multiplying, rather than adding, the abundances of the atomic elements (equivalent to adding the abundances after log transformation instead of before) (@vandenboogaart2013anal).
+The changes in balances retain bias invariance (@mclaren2019cons Appendix 1, and REF Appendix).
+Balances have been used to construct candidate disease biomarkers (@riverapinto2018bala, @quinn2020inte), develop phylogenetic community distance measures (@silverman2017aphy), and look for clades associated with traits (@washburne2017phyl).
+But they face from conceptual and practice issues relating to the sparsity of microbiome data. 
+Perhaps more important, products of species abundances do not obviously map onto the ecological notion of abundance and hence do not seem to fully substitute for attempting to infer the changes in taxon abundance as conceived in the usual sense of additive aggregation.
+A benefit of products, however, is that they are also robust to the fact that we may not know the appropriate ecological conversion rate for different species; for example, cells of one clade might be twice as impactful per cell on a host or  ecosystem as cells from another clade.
+Grappling with these issues remains an important goal for future theoretical and empirical research.
 
@@
 Yet these biases are shared by the sequencing measurement.
-Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
+Though they make qPCR measurements a bad proxy for total cell density, when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
 (Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
-If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.^[See discussion below]
+If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.
 Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
 In this case, ...
-
-## Notes
-
-<details><summary style="font-weight:bold;font-size:120%">Details: Taxonomic bias and differential-abundance estimates</summary>
-::: {#details:differential .rmdnote}
-
-#### Proportion-based inference {-}
-
-#### Ratio-based inference {-}
-
-:::
-</details>

### 2021-08-04 solutions.Rmd
Revise manuscript (in progress)
- Begin drafting new section on measurement model
- Make minor edits and add notes to other sections
- Add refs to .bib file
- Remove summary and equations files from the rendered manuscript

@@
 # Potential solutions
+
+_This section needs to be revised to account for the expanded earlier theory, especially relating to targeted measurements and ratio-based methods._
 
@@
 
-_Except in synthetic community experiments, we are always likely to have taxa not in our controls.
+- Except in synthetic community experiments, we are always likely to have taxa not in our controls.
 We can attempt to use phylogenetic prediction or other modeling of efficiencies to predict the efficiencies of other taxa; however there will be signifciant undertainty.
-Therefore our control measurements may ideally serve to inform a bias-sensitivity analysis of the sort described below._
-
-For revision:
-
-- Discuss the vaginal mb and leopold examples
+Therefore our control measurements may ideally serve to inform a bias-sensitivity analysis of the sort described below.
+- Should mention the two extant cases where mocks of relevant taxa have been created: The vaginal-HMP / MOMS-PI study (@brooks2015thet and @fettweis2019thev) and the @leopold2020host plant-fungal experiment.
+Note that even in these cases, calibration hasn't been much used, partly because of a lack of workflows or understanding of when and why it is necessary.
+Ideally we can give an illustration; this section can be a good use of the @leopold2020host change-over-time result, though I'm not sure whether to discuss that here or in the 'real-world' section.
 
@@
 
-targeted measurements of a small number of taxa
-
 Because targeted absolute-abundance measurements are expected to provide (relatively) accurate estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
 We illustrate the basic idea using the @lloyd2020evid experiment described above.
-Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.^[see note]
+Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.
 Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
@@
 In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].
-
-### Notes for revision
 
@@
 
-> Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
-In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
-Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
-
-This needs to be revised to make clear the novelty of using non-spiked taxa as a reference; could help to replace "In fact, some methods" with clearer language.
-Also make clear that some studies using reference taxa are using proportion-based methods.
-Also, cite the appendix that gives an overview of AA methods, and explicitly state distinction between proportion-based and ratio-based methods.
-
-Could help to change to:
-
-- Ratio-based analysis can also be used to overcome the effects of bias in differential absolute abundance.
-- /Make clear that the above ratio equation applies for relative and absolute observed or actual abudances (since totals cancel)/
-- Suppose we knew the true absolute abundance of a reference taxon $r$; then we might estimate the absolute abundance of taxon $i$ from
-  + Give the estimation equation
-  + Explain the effect of taxonomic bias - get correct FCs
-- It remains to determine the AA of one or more reference taxa.
-  + Some existing AA methods involve using spike ins of extraneous taxa
-    - discuss spike-in approach
-    - often stated that bias a problem for spike-ins; however we see that it only affects the baseline and not the fold changes
-      + this first part could be shunted to a discussion
-    - most/all papers on spike-ins suppose the spike-in must be added in a known fixed amount.
-    - however, the above argument indicates that it isn't required that the reference taxon is extraneous or has a fixed abundance; what is important is that we know the correct abundance up to a fixed multiplicative error
-    - hence it can be a native taxon we've measured using a targeted method; this method can be biased as long as it has a fixed multiplicative error (unlike MGS bias.)
-
-Need to consider if this is too detailed for the main text.
-Might want to make this section of the appendix particularly clear if can't put this in the main text.
-
-Key equations, to show side-by-side: Proportion-based and ratio-based AA inference.
-
 ## Use ratio-based estimation
@@
 This problem is easier to solve, and one solution is to use analysis methods based on the ratios among taxa rather than the proportions of individual taxa.
-
-- from gdoc: this distinction of ratios vs proportions needs to be spelled out more explicitly; could even use my seminar figure (or equations) to help
-- later thoughts: This distinction is so central for how I understand and explain the key ideas, should refer to it early and often
 
@@
 
+- Aim to illustrate with a simulated or real example
+
 ## Use bias-aware meta-analysis
@@
 Thus such bias-aware meta-analyses are already technically feasible and—if bias is truly consistent within a protocol or study—may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not model bias explicitly.
+
+- Aim to  illustrate with a simulated or real example

### 2021-08-05 v/0f596f09880fb209cd5a6b87bff16da6d7418ea0/_main.tex
Update syntax for block environments
This commit adds the rendered manuscript associated with
0f596f09880fb209cd5a6b87bff16da6d7418ea0.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for}
\author{Michael R. McLaren\footnote{North Carolina State University; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Karen G. Lloyd\footnote{University of Tennessee} \and Benjamin J. Callahan\footnote{North Carolina State University}}
\date{2021-08-04}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for},
  pdfauthor={Michael R. McLaren; Karen G. Lloyd; Benjamin J. Callahan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{cancel}

% Define note environment based on example from https://bookdown.org/yihui/rmarkdown-cookbook/custom-blocks.html
\usepackage{color}
\usepackage{framed}
\setlength{\fboxsep}{.8em}
\newenvironment{rmdnote}{
  \definecolor{shadecolor}{RGB}{248,248,248}
  \begin{shaded}}
 {\end{shaded}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{definition}
\newtheorem{hypothesis}{Hypothesis}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{preface}{%
\section*{Preface}\label{preface}}
\addcontentsline{toc}{section}{Preface}

\leavevmode\vadjust pre{\hypertarget{preface-warning}{}}%
\textbf{This in-progress manuscript is not intended for general scientific use.}
It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
Please post comments or questions on the \href{https://github.com/mikemc/differential-abundance-theory/issues}{GitHub Issues page} or \href{m.mclaren42@gmail.com}{email Mike}.

This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.
We describe the basic problem posed by taxonomic bias for measuring changes in the abundance of particular taxa across conditions and describe new strategies for mitigating the errors it induces.
Analyses of both relative and absolute abundances are considered.
In its current form, the manuscript sits somewhere between a standard scientific article and a monograph;
it consists of an article followed by a series of appendices which together give a comprehensive treatment of the implications of the \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) model of taxonomic bias for differential-abundance analysis and experimental design.
It is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
See \href{https://doi.org/10.5281/zenodo.4552717}{the Zenodo record} for how to cite the latest version.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Most microbiome research aims to go beyond qualitative descriptions to make quantitative conclusions about associations between specific microbes or community states and key host or environmental properties.
But there is serious disagreement about whether the microbiome measurements made by marker-gene and metagenomic sequencing (MGS) are or can ever be truly quantitative. A primary reason is that taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
As a result taxa can differ dramatically (e.g.~10-1000X) in how efficiently they are measured---that is, converted from cells into taxonomically classified sequencing reads---making the relative abundances we observe by MGS inaccurate representations of the actual sample compositions.
Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads among taxa make this taxonomic bias a feature of shotgun metagenomic measurements as much as marker-gene measurements.
Taxonomic bias is protocol specific (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})) and can even vary among batches within an experiment (\protect\hyperlink{ref-yeh2018taxo}{Yeh et al.} (\protect\hyperlink{ref-yeh2018taxo}{2018})); it thus not only makes MGS measurements inaccurate, but also incomparable between studies.
Taxonomic bias thereby poses a major challenge for any seeking to draw quantitative conclusions from MGS studies.

The field's current answer to this challenge is methods standardization, prioritizing consistency over accuracy.
By treating each sample within a study in exactly the same manner, we can ensure that observed differences between samples are not simply due to methodological differences.
Yet standardization does not guarantee that these observed differences reflect the actual differences in taxonomic composition among samples.
Consider what might be the simplest and most common inference made from MGS: How does the proportion of a taxon vary between samples? An observation that \emph{Escherichia} coli makes up 10\% of the sequencing reads in a fecal sample might, thanks to bias, have arisen from a sample in which the proportion of \emph{E. coli} is actually 50\% or 1\%.
Yet many hope that if we observe the proportion of \emph{E. coli} to increase from 10\% to 20\% between two samples, its proportion must have actually doubled---by fixing the taxonomic bias of our measurements through standardization, we have (so the argument goes) fixed the measurement error in both samples so that it will cancel when estimate the fold change.
Unfortunately, the results of \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) show that reality is not so neat.
Because the taxa compete for sequencing effort within a sample, the effect that taxonomic bias has on the observed proportions depends on the taxonomic composition of the sample.
As a result, MGS measurements can lead to spurious inferences about changes across samples even when taxonomic bias is constant for each taxon across samples.

This problem posed by taxonomic bias intersects with a second, growing concern within the microbiome field: that the relative abundances measured by MGS can present a misleading view of absolute-abundance dynamics.
Without making further assumptions (which are often difficult or impossible to verify) one cannot tell whether a doubling in the proportion of a taxon reflects an increase in its abundance or a decrease in the abundances of other taxa.
This concern has motivated the development of various experimental methods to enable researchers to convert the relative abundances measured by MGS into absolute abundances (methods to be reviewed in the Appendix in a future version).
An increasingly popular approach in the study of both environmental and host-associated microbiomes involves making an independent measurement of total microbial abundance in each sample (\protect\hyperlink{ref-props2017abso}{Props et al.} (\protect\hyperlink{ref-props2017abso}{2017}),\protect\hyperlink{ref-kevorkian2018esti}{Kevorkian et al.} (\protect\hyperlink{ref-kevorkian2018esti}{2018}),\protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}),\protect\hyperlink{ref-tettamantiboshier2020comp}{Tettamanti Boshier et al.} (\protect\hyperlink{ref-tettamantiboshier2020comp}{2020}),\protect\hyperlink{ref-contijoch2019gutm}{Contijoch et al.} (\protect\hyperlink{ref-contijoch2019gutm}{2019}),\protect\hyperlink{ref-vandeputte2017quan}{Vandeputte et al.} (\protect\hyperlink{ref-vandeputte2017quan}{2017}),\protect\hyperlink{ref-vieirasilva2019quan}{Vieira-Silva et al.} (\protect\hyperlink{ref-vieirasilva2019quan}{2019})).
This estimate is then multiplied by the MGS proportions to convert them to estimates of absolute abundance.
This simple calibration procedure directly transfers the error in the MGS proportions due to taxonomic bias to the estimated absolute abundances, leading to estimates that are not only individually inaccurate but can imply inaccurate variation in abundance across samples.

Here we summarize recent advances in our understanding of how taxonomic bias affects our ability to accurately determine how taxa vary in relative and absolute abundance across samples from different conditions---the inference problem known as \emph{differential abundance (DA) analysis}.
We first explain how taxonomic bias can lead to spurious DA results and consider the biological conditions in which such spurious results are likely.
But we then describe several experimental and computational methods that can be used to avoid or correct these errors.
Collectively, these methods may provide practical solutions to mitigating the effects of taxonomic bias on DA analysis for a majority of microbiome studies.

\hypertarget{bias-can-cause-spurious-results-in-proportion-based-da-analysis}{%
\section{Bias can cause spurious results in proportion-based DA analysis}\label{bias-can-cause-spurious-results-in-proportion-based-da-analysis}}

\begin{verbatim}
#> Warning: It is deprecated to specify `guide = FALSE` to remove a guide. Please
#> use `guide = "none"` instead.

#> Warning: It is deprecated to specify `guide = FALSE` to remove a guide. Please
#> use `guide = "none"` instead.
\end{verbatim}

\begin{figure}
\centering
\includegraphics{_main_files/figure-latex/vaginal-example-1.pdf}
\caption{\label{fig:vaginal-example}\textbf{Taxonomic bias can distort differential abundance results even when it is consistent for each taxon across samples.} Panel A shows the actual and observed proportions for hypothetical community samples from two time points, which differ in their relative abundance of three taxa. Panel B shows taxonomic bias in terms of the relative efficiencies of the three taxa against the mean efficiency of each sample; the difference between the taxon's efficiency and the sample's mean (vertical arrows) determines the fold error seen in Panel A. Panel C rearranges the plot from Panel A to show the actual and observed fold changes between time points. The efficiencies of individual taxa were estimated by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) from mock community data from \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}). The abundances are hypothetical but inspired by observations from the human vaginal microbiome; see main text.}
\end{figure}

We illustrate the problems posed by taxonomic bias for differential abundance using an example where the bias of an MGS protocol was explicitly measured using mock communities.
To analyze the effects of taxonomic bias on measurements of the human vaginal microbiome, \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}) performed 16S rRNA gene sequencing of mock communities of seven key vaginal species using a similar protocol to that of the Vaginal Human Microbiome Project (VaHMP).
\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) used this data to estimate the relative measurement efficiency of the seven species.
They found that the most efficiently measured species, \emph{Lactobacillus iners}, had an efficiency that was 29X that of the least efficiently measured species, \emph{Gardnerella vaginalis}, primarily due to a greater DNA extraction efficiency and 16S copy number.
A third species, \emph{Atopobium vaginae}, had an efficiency 1.8X that of \emph{G. vaginalis}.
A common aim of human vaginal microbiome studies, including the VaHMP, is to understand the causes and consequences of shifts in communities from being dominated by a particular \emph{Lactobacillus} species to dominance by a non-\emph{Lactobacillus} species such as \emph{G. vaginalis}, which have been associated with deleterious health outcomes including an increased risk of preterm births in pregnant women.
It is therefore important to know whether the efficiency variation found by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) affects the observed differential abundances between \emph{Lactobacillus}-dominated and \emph{Gardnerella}-dominated communities.

Figure \ref{fig:vaginal-example} illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by \emph{Lactobacillus} to being dominated by \emph{Gardnerella}.
In each case, we observe a greater proportion of \emph{Lactobacillus} than is truly present, and less of the other two species (Figure \ref{fig:vaginal-example}A).
This positive multiplicative error arises because \emph{Lactobacillus} is more efficiently measured than the average species in each sample.
\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed that the observed and actual proportions of a taxon \(i\) in a sample \(s\) are related by the equation
\begin{align}
  \label{eq:error}
  \text{observed}_{i}(s) = \text{actual}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
\end{align}
where \(\text{mean efficiency}(s)=\text{actual}_j(s)*\text{efficiency}_j\) in the denominator is the average efficiency of all taxa in that sample.
Intuitively, Equation \eqref{eq:error} says that the taxon will be overrepresented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
Figure \ref{fig:vaginal-example}B shows the efficiencies of the three taxa and the mean efficiency in each sample.
The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of \emph{Lactobacillus}.
The error in the second sample is larger---specifically, by 6X---than the first sample because the high proportion of \emph{Gardnerella} has reduced the sample mean efficiency (and thus increased the efficiency of \emph{Lactobacillus} relative to the mean) by 6X.

The enrichment depends on the taxon's efficiency relative to the sample average, rather than the efficiency itself, because the requirement that the proportions of all taxa sum to one locks the taxa in a zero-sum competition.
Notably, competition occurs (though in a less absolute form) even at the level of the raw sequencing reads: DNA extractions and PCR reactions have limited output capacity; high-concentration samples are typically subsampled to fixed, lower concentrations; and sequencing libraries are given limited sequencing effort; so that the experimental process itself typically imposes competition to be sequenced.
And it is competition that prevents the effects of taxonomic bias from canceling when we analyze how the proportions of individual taxa change across samples.

Figure \ref{fig:vaginal-example}C shows the consequences of measurement error on an analysis of the fold change in proportions between the two samples.
In reality, \emph{Gardnerella} changes by 4X, \emph{Lactobacillus} by 0.1X, and \emph{Atopobium} by 0.5X.
Yet what is observed is that \emph{Gardnerella} changes by 32X, \emph{Lactobacillus} by 0.6X, and \emph{Atopobium} by 3X.
In other words, the change in each case is multiplied by 6X, causing a much larger increase in \emph{Gardnerella} and a smaller decrease \emph{Lactobacillus} than actual, and the appearance that \emph{Atopobium} increased when it in fact decreased.
Why do such errors occur?
The multiplicative error in Equation \eqref{eq:error} has two parts: a taxon-specific term in the numerator that is the same in each sample, and a taxon-independent term in the denominator that depends on the composition of the sample through the mean efficiency.
When we compute the fold change in a taxon i between samples s and t,
\begin{align}
  \frac{\text{observed}_{i}(t)}{\text{observed}_{i}(s)}=\frac{\text{actual}_{i}(t)}{\text{actual}_{i}(s)}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{i}}*\frac{\text{mean efficiency}(s)}{\text{mean efficiency}(t)}
\end{align}
the constant efficiency in the numerator of Equation \eqref{eq:error} cancels, but the varying mean efficiency does not, leaving an error that is equal to the inverse change in mean efficiency.
As a result, we are here left inferring that the fold-changes in proportions have all increased less (or decreased more) than they truly have.
In our example, the mean efficiency decreased by 6X (from 15 to 2.5) due to the shift in dominant taxon from the high-efficiency \emph{Lactobacillus} to the low-efficiency \emph{Gardnerella}, causing the observed fold changes to increase by 6X above the actual.
This increase distorted the magnitudes of the changes in \emph{Lactobacillus} and \emph{Gardnerella} (decreasing the first and increasing the second) and changed the sign of \emph{Atopobium} from negative to positive.
Thus partial cancelling in the effect of a constant multiplicative bias may not be sufficient to protect against spurious results.

Spurious results can occur for the same basic reason in proportion-based absolute-abundance inference.
Suppose that the true total abundance in each sample can be determined experimentally.
Because the absolute abundance estimate simply multiplies the erroneous observed proportions by this total, the estimated change in absolute abundance is wrong by the same factor as the fold change in proportions.
In our Figure \ref{fig:vaginal-example} example, if the change in total abundance between samples were 2X, then we would observed both \emph{Lactobacillus} and \emph{Atopobium} to have increased in absolute abundance (by 1.2X and 6X), when in fact \emph{Lactobacillus} decreased (by 5X) and \emph{Atopobium} remained the same.

\hypertarget{effect-on-regression-of-multiple-samples}{%
\subsection{Effect on regression of multiple samples}\label{effect-on-regression-of-multiple-samples}}

We showed how fold changes of individual samples may be unreliable.
But microbiome analyses often do not interpret such individual differences, and instead look for average patterns across many samples, such as how the abundance of a taxon varies with some covariate of interest.
The covariate may be discrete, such as whether the sample is from a healthy or sick person, or it may be continuous, such as a measure of space, time, or temperature.
Often these analyses can be framed as a regression problem.
For example, we might hypothesize that the log absolute abundance of taxon \(i\) changes with a variable \(x\) according to the simple linear regression,
\begin{align}
  E[\text{log abundance}_{i} \mid x] = a_{0} + a_{1}x
\end{align}
where \(x\) is either continuous (e.g., sediment depth) or binary (e.g., \(x=1\) for treated patients and \(x=0\) for controls).
Much of the effect of bias will be absorbed by the model intercept, which is typically not of interest to the researcher.
However, variation in the mean efficiency creates additional variation (reducing statistical power to detect variation in the slope) and can also create a statistical bias in our slope estimate equal to the inverse slope of \(E[\text{log mean efficiency}\mid x]\) as a function of \(x\).
Therefore if the mean efficiency varies across samples, but is not associated with the covariate \(x\), its effect may simply be to create larger standard errors in our slope estimate.
But the larger worry may be the systematically distorted slope (\(a_1\)) estimates that arise in situations where the mean efficiency changes with \(x\).
Notably, the error in an absolute sense is the same for all taxa---the slope is always reduced by the same amount---but the implications differ for various taxa depending on their slopes, causing magnitude or sign errors depending how the coefficient derived from the (unknown) true abundances compares to that of the mean efficiency (Figure \ref{fig:regression-example}).

\begin{verbatim}
#> Warning: Package `magick` is required to draw images. Image not drawn.
\end{verbatim}

\begin{figure}
\centering
\includegraphics{_main_files/figure-latex/regression-example-1.pdf}
\caption{\label{fig:regression-example}\textbf{Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.} Details to synthesize into a caption: Regression of log2(Proportion) of the commensal fungus Penicillium versus timepoint; timepoints 1 and 2 are pre- and post-challenge with the pathogen Melampsora. Data is split by the region the host plants are derived from (Eastern and Western US). Calibrated proportions = Observed proportions in the real, experimental samples after adjustment for the bias measured in mock communities. Mean efficiency of each community is inferred by treating the calibrated proportions as the truth, and multiplying by efficiencies estimated from the mocks. The pathogen Melampsora has a high measurement efficiency; thus once it infects the plants, the mean efficiency of the sample increases (purple points). Efficiency is here taken as relative to the focal taxon Penicillium. West plants tend to be more resistant to the pathogen, which likely explains why the mean efficiency doesn't increase as much in the West plants. Penicillium is observed to decrease in log proportion, in both the East and West plants. But the calibrated measurements show that it actually slightly increases in the East plants, and has a lesser decrease in the West plants than what was observed before bias correction. The difference between the Calibrated and Observed data points and regression lines equals the regression line of the mean efficiency: Orange = Green + Purple; Green = Orange - Purple. The absolute error in regression coefficients is the same for all taxa. I picked Penicillium for illustration since it has the smallest observed decrease, which makes the error due to bias have a particularly significant impact.}
\end{figure}

\hypertarget{implications-for-real-world-inference}{%
\section{Implications for real-world inference~}\label{implications-for-real-world-inference}}

These observations show that the effects of taxonomic bias do in fact cause error in proportion-based differential abundance analyses.
Yet if the mean efficiency varies much less across samples than the abundances of individual taxa, this error will be negligible.
Alternatively, if variation in the mean efficiency is effectively random (not associated with regression covariates), the effect of bias may be to increase the residual variance and thus reduce our statistical power to assay differential abundance---unfortunate to be sure, but perhaps not as much as confidently inferring changes with the wrong sign or magnitude.
But there are salient biological scenarios where substantial associations between the mean efficiency and covariate can in fact lead to spurious differential abundance inference.

\hypertarget{scenarios-in-which-bias-may-cause-spurious-results}{%
\subsection{Scenarios in which bias may cause spurious results}\label{scenarios-in-which-bias-may-cause-spurious-results}}

Figure \ref{fig:vaginal-example} illustrates how a change in dominance by from a high- to low- efficiency taxon (or vice versa) can easily create spurious DA results.
Indeed, this scenario may be particularly common in vaginal microbiome studies.
Human microbiomes often dominated by \emph{Lactobacillus}, but can become dominated by non-\emph{Lactobacillus} species such as \emph{Gardnerella} vaginalis and \emph{Atopobium vaginae}, and these low-\emph{Lactobacillus} states are associated with the disease bacterial vaginosis and, in pregnant women, an increased risk of preterm birth.
Using mock community measurements from \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}), \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed that the efficiency of two common \emph{Lactobacillus} species were 20-30X greater than \emph{G. vaginalis} and \emph{A. vaginae}.
Therefore we should expect samples dominated by \emph{Lactobacillus} to have a much larger mean efficiency than those those dominated by \emph{G. vaginalis} and \emph{A. vaginae}, which would distort the observed associations of all taxa with any covariate that is also associated with \emph{Lactobacillus} dominance.
The extraction and 16S sequencing protocol used by \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}) is similar to that used in the VaHMP MOMs-PI study also led by these researchers to investigate the associations of these taxa with preterm birth (\protect\hyperlink{ref-fettweis2019thev}{Fettweis et al.} (\protect\hyperlink{ref-fettweis2019thev}{2019})), demonstrating the potential for serious real-world consequences for these statistical/technical concerns.

Similar dynamics occur in a plant-fungal interactions experiment performed by \protect\hyperlink{ref-leopold2020host}{Leopold and Busby} (\protect\hyperlink{ref-leopold2020host}{2020}) (Figure \ref{fig:regression-example}).
The authors inoculated commensally-colonized trees with a fungal pathogen.
Using DNA mock communities, they showed that (excluding bias from DNA extraction) the pathogen was 10X more efficiently measured than the median commensal and 40X more efficiently measured than the lowest-efficiency commensal.
In most hosts, the pathogen increased rapidly between timepoints, driving an increase in the sample mean efficiency that in turn leads to larger decreases being observed in the proportions of commensal taxa than what is predicted using calibrated (bias-corrected) values (REF APPENDIX).
For example, the calibrated proportions indicate that the commensal \emph{Penicillium} slightly increased in hosts from the Eastern US and slightly decreased in hosts from the Western US.
Yet the observed (uncalibrated) proportions show moderate and large decreases, respectively (Figure \ref{fig:regression-example} above).

These examples involved scenarios where individual taxa can constitute a large proportion of the community.
But in samples from highly-diverse ecosystems such as human gut, soil, and ocean sediment, one species rarely dominates.
In these cases it might seem intuitive that the mean efficiency should be relatively stable across samples: As diversity increases, the mean efficiency effectively averages over efficiencies from a greater number of taxa and so (by the central limit theorem) will converge to a constant value---if taxon efficiencies are statistically independent of taxon abundances (REF APPENDIX).
In practice, associations between efficiencies and abundances may be common and can allow large shifts in mean efficiency to occur even in highly diverse samples.

The human gut provides a clear example of how such associations might arise.
Gut microbiomes typically have high diversity at the species level but are dominated by just a small number of phyla, and two in particular: The Bacteroidetes and the Firmicutes.
The ratio of Bacteroidetes to Firmicutes can shift substantially between individuals and has been linked to a number of host traits and health conditions.
In addition, many DNA extraction protocols more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though there can also be significant variation in extraction efficiency among species within these phyla (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})).
For such protocols, we should expect Bacteroidetes-dominated samples to have substantially larger mean efficiencies than Firmicutes-dominated samples, regardless of species-level diversity.
This association of efficiency with phylum abundance will distort DA inferences at any taxonomic level if the covariate is also associated with these phyla.

This example illustrates a general mechanism by which association between the abundances and efficiencies of taxa are created by the common influence of evolutionary history.
Just as shared evolutionary history creates positive associations in ecological traits that drives positive correlations in how related taxa vary in abundance across samples, so does it create positive associations in bias-related traits that can lead to similar efficiencies among closely related taxa.
Such phylogenically-associated, bias-affecting traits include cell-wall toughness, ribosomal-operon copy number, binding affinity for a given set of primers, and representativeness in taxonomic databases.
Ecology and efficiency are both affected by the same evolutionary history, leading to positive associations between the two: A change in diet that increases the relative abundances of many Bacteroidetes species and in so doing also increases the relative abundance of many easy-to-lyse species.

Associations between measurement efficiency and relative abundance can also arise because a single trait affects both.
For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
\protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) estimate the log fold change in the estimated absolute abundance of various taxa with sediment depth (as a proxy for time) to determine which taxa are able to persist and even grow in this difficult environment.
It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
As the relative abundance of tougher species increases with depth, the mean extraction efficiency decreases.
This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.

\hypertarget{error-in-absolute-abundance-measurements}{%
\subsubsection{Error in absolute abundance measurements}\label{error-in-absolute-abundance-measurements}}

So far we have ignored error associated with our absolute abundance measurements.
In fact, these measurements can also have a tendency to measure contributions from some taxa more efficiently than others.
\textbf{TODO: State how this could create problems.
Might also note the ambiguity in what the truth/target is (e.g.
live cells, or all cells; cells, or 16S rRNA gene copies; observations in Jian, Salonen, and Korpela 2021).}
Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
Yet these biases are shared by the sequencing measurement.
Though they make qPCR measurements a bad proxy for total cell density, when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can cancel, leaving relative accurate fold changes.
(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.
Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.

\hypertarget{we-cannot-simply-ignore-taxonomic-bias}{%
\subsection{We cannot simply ignore taxonomic bias}\label{we-cannot-simply-ignore-taxonomic-bias}}

These examples show that there is potential for bias to distort DA results in both low and high diversity settings.
We therefore need methods to control or correct for bias to ensure the validity of future results, as well as methods to probe the robustness of past results.
It is possible that spurious results are unlikely in many experimental contexts.
By applying such methods across a range of biological and experimental contexts we will deepen our understanding of when bias is unlikely to significantly distort biological findings and when additional measures are needed to mitigate its effects.

\hypertarget{potential-solutions}{%
\section{Potential solutions}\label{potential-solutions}}

\hypertarget{using-control-samples-to-calibrate-relative-abundances}{%
\subsection{Using control samples to calibrate relative abundances}\label{using-control-samples-to-calibrate-relative-abundances}}

In principle, control (``mock'') communities containing representative taxa from the environment of interest can be used to directly estimate the measurement efficiencies and correct bias in the MGS measurements (``calibration'').
Such an approach may be sufficient for synthetic-community experiments, where all taxa are culturable, and for relatively simple natural communities like the vaginal microbiome that are dominated by a small number of culturable taxa.
But suitable mock controls may not be feasible for most complex natural ecosystems and require significant effort to develop.

An closely-related alternative to mocks are controls derivedd from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})).
A natural fecal standard is currently being developed by \href{https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material}{NIST} and one has recently been made available commercially by Zymo Research (\href{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology}{ZymoBIOMICS Fecal Reference}).
Ensuring the stability and homogeneity of such control samples can be challenging.
Careful testing is also needed to ensure that the preparation and storage of the controls has not significantly affected the taxonomic bias relative to the experimental samples in a given application.
As it is feasible to characterize a single standard much more extensively than a typical community sample, we may be able to obtain an estimated composition we feel comfortable treating as the ground truth.
Yet even when this is not possible, such natural standards can allow us to reconcile results across studies despite not knowing the truth.

\hypertarget{ratio-based-relative-and-absolute-abundance-inference}{%
\subsection{Ratio-based relative and absolute abundance inference}\label{ratio-based-relative-and-absolute-abundance-inference}}

A comprehensive approach of estimating the efficiency of all taxa is needed for getting calibrated relative abundances of all taxa within individual samples.
But what if we only want calibrated DA analysis?
That is, we are comfortable with not knowing whether \emph{E. coli} is 10\% of our sample if we can confidently determine that it doubled or halved between sample conditions?
This problem is easier to solve, and one solution is to use analysis methods based on the ratios among taxa rather than the proportions of individual taxa.

A subset of methods for analyzing differential relative abundance are derived from the field of Compositional Data Analysis (CoDA).
The defining feature of CoDA methods is that they are based on fold-changes in ratios among elements (here, the taxa) (REF AITCHISON).
The use of CoDA methods in microbiome analysis has largely been motivated by concerns over the negative correlations between taxa induced by the sum-to-one constraint in taxon proportions.
But \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed that, due to their property of perturbation invariance (REF AITCHISON), the results of CoDA differential relative abundance analysis are invariant to consistent taxonomic bias.
From Equation \eqref{eq:error}, the observed ratio of a taxon \(i\) to a taxon \(j\) in a sample \(s\) is
\begin{align}
  \frac{\text{observed}_{i}}{\text{observed}_{j}}=\frac{\text{actual}_{i}}{\text{actual}_{j}}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}
\end{align}
The error is independent of the sample composition and thus cancels in any function that is a fold-change of this (or any) ratio of taxa between samples.
(Perhaps insert example from Figure \ref{fig:vaginal-example}ABC all giving same ratio FCs.)

Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
To estimate the abundance of a focal taxon \(i\), this approach multiplies the ratio of reads (or proportions) of taxon i to a reference taxon \(r\) by a known or estimated abundance of the reference taxon, (while ignoring bias)
\begin{align}
  \text{estimate}_{i} = \frac{\text{observed}_{i}}{\text{observed}_{r}}*\text{abundance}_{r}
\end{align}
So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
Yet the reference taxon could also be a naturally occurring taxon whose absolute abundance we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.
Because taxa do not compete (or competition is greatly reduced) in such targeted measurements, any taxonomic bias associated with them is expected to create constant multiplicative error across samples and so not affect fold change estimates {[}APPENDIX{]}.
APPENDIX describe various theoretical and experimental considerations for both approaches.

\hypertarget{estimating-error-with-targeted-measurements-of-a-small-number-of-taxa}{%
\subsection{Estimating error with targeted measurements of a small number of taxa}\label{estimating-error-with-targeted-measurements-of-a-small-number-of-taxa}}

Because targeted absolute-abundance measurements are expected to provide (relatively) accurate estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
We illustrate the basic idea using the \protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) experiment described above.
Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.
Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
\protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa {[}APPENDIX{]}.

\hypertarget{computational-approaches}{%
\subsection{Computational approaches}\label{computational-approaches}}

What can we do for experiments that have already been conducted without mock controls, targeted control measurements, or spike-ins?
At least two purely computational approaches can still be used.

\textbf{Bias sensitivity analysis:}
A straightforward and universally available approach is to use computer simulation to determine how the results change under a range of possible sets of efficiencies, which can be randomly generated to reflect certain hypotheses about bias in the given system.
The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results---not just for differential absolute abundance, but for all microbiome analyses.

\textbf{Bias-aware meta-analysis:}
When the goal is to perform a meta-analysis that combines studies that have used different protocols, the unknown measurement efficiencies of each protocol can be explicitly included as parameters of the statistical model that is used.
Unknown efficiencies can be included in ``compositional'' linear modeling frameworks such as ALDEx2, DivNet, and fido simply by adding a protocol-specific term to the linear model of taxon log ratios.
Thus such bias-aware meta-analyses are already technically feasible and---if bias is truly consistent within a protocol or study---may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not explicitly model bias.

\hypertarget{appendix-appendix}{%
\appendix \addcontentsline{toc}{section}{\appendixname}}

\hypertarget{models}{%
\section{Measurement models}\label{models}}

Deterministic measurement error models that extends the \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) model of metagenomics measurement to 1) describe absolute as well as relative abundance, 2) include spike-in taxa, and 3) include supplementary measurements of bulk and targeted (absolute) abundance.

\hypertarget{actual-abundances-and-general-notation}{%
\subsection{Actual abundances and general notation}\label{actual-abundances-and-general-notation}}

(For now) Number of samples = J; Number of taxa = I.

\(A_{ij}\) is the actual absolute abundance of taxon \(i\) in sample \(j\), in the target units.
For variables that hold matrixes or vectors, capital letters denote absolute abundances and lower case letters denote proportions.
Thus \(a_{ij}\) is the matrix of actual relative abundances in terms of cells as a proportion of the total amount of cells in the sample; \(a_{ij} = A_{ij} / A_{Tj}\).
Let \(A_{Tj}\) (or \(A_{\cdot j}\)??) be the total abundance in sample \(j\), \(A_{Tj} = \sum_i A_{ij}\).
For a subset of taxa \(Q \subset \{1, \dots, I\}\), define \(A_{Qj} \equiv \sum_{i \in Q} A_{ij}\) to be the total abundance of the taxa in \(Q\).

Unless specified otherwise, \(A\) always refers to abundance in the original sample material, not in extracted DNA.
I will assume the target units are concentration of cells per unit mass. However, other units (such as biomass per volume) may be relevant.

To consider various addition information that can be used for absolute abundance inference, I let \(B\) denote bulk abundance measurements (such as via flow cytometry or broad-range 16S qPCR), \(S\) denote the known taxon abundances of a spike-in, and \(T\) denote the abundances of a set of taxa made via targeted measurement.
I let \(R \subset \{1, \dots, I\}\) denote the set of \emph{reference taxa} whose abundance is known independently from the metagenomics measurement (up to experimental bias factors) because they were spiked at a known density, subjected to a targeted measurement, or have been determined through some other means to have a constant or known abundance across samples.

Bias vectors \(B^{(P)}\) are vectors of length \(I\) indicating the taxon-specific bias associated with a measurement protocol \(P\).
I generally assume these to be sample-independent.
An overbar denotes the average value over taxa weighted by actual abundance; i.e., \(\bar B_j = I^{-1} \sum_i B_i a_{ij}\).
An overbar and a subcript set of taxa indicates the mean bias within that subset: \(\bar B_{Qj} = \sum_{i\in Q} B_i a_{ij} / \sum_{i\in Q} a_{ij}\).

\begin{longtable}[]{@{}llll@{}}
\toprule
Measurement type & Units & Abundance matrix & Bias vector \\
\midrule
\endhead
\textbf{A}ctual abundance & {[}cells{]} & \(A\), \(a\) (\(I\times J\)) & - \\
\textbf{M}etagenomics & counts & \(M\), \(m\) (\(I\times J\)) & \(B^{(M)}\) \\
Bul\textbf{k} abundance & {[}cells{]}, {[}DNA{]}, or {[}gene copies{]} & \(K\) (\(I\times 1\) column vector) & \(B^{(K)}\) \\
\textbf{T}argeted abundance & {[}cells{]}, {[}DNA{]}, or {[}gene copies{]} & \(T\) (\(I \times J\)) & \(B^{(T)}\) \\
\textbf{S}pike-in abundance & {[}cells{]}, {[}DNA{]}, or {[}gene copies{]} & \(S\) (\(I \times J\)) & \(B^{(S)}\) \\
\bottomrule
\end{longtable}

Note: \(S_{ij}\) and \(T_{ij}\) are only defined for the reference taxa \(i \in R\).

\hypertarget{metagenomics-measurement}{%
\subsection{Metagenomics measurement}\label{metagenomics-measurement}}

\(M\) is an \(I\times J\) matrix representing the (marker-gene or) metagenomics measurement; \(M_{ij}\) is the sequencing count associated with taxon \(i\) in sample \(j\), such that \(M_{ij} = 10\) indicates that 10 fragments of DNA were sequenced and assigned to taxon \(i\) and sample \(j\).
In the case of paired-end sequencing, a single fragment of DNA yields two reads, which may or may not be counted jointly or independently by different algorithms; here I assume that read-pairs are assigned jointly, and will use ``read'' and ``fragment'' interchangeably.
For simplicity, I ignore the possibility for index switching or other sources of cross-contamination and assume that reads assigned to sample \(j\) truly arose from sample \(j\).

Following the notation defined above, I let \(M_{Tj} = \sum_i M_{ij}\) be the total count for sample \(j\); I will sometimes refer to this number as the \emph{sequencing depth} or \emph{read depth} of the sample, though in reality the total count will be less than the sample's sequencing depth since often a significant fraction of reads are lost by filtering steps prior to producing the final count matrix.
I use \(m_{ij} = M_{ij} / M_{Tj}\) to denote the proportion of reads assigned to taxon \(i\) in sample \(j\).

I assume that the composition (relative abundances) is given by the deterministic error model presented in \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}).
In this model, the bias vector \(B^{(M)}\) represents the \emph{relative} efficiencies with which various taxa are measured, and so has the same effect on measurement if we rescale it \(B \to cB\) for any \(c > 0\).
To remove this degree of freedom, I will suppose that \(B^{(M)}_{1} = 1\), so that \(B^{(M)}_i\) is the efficiency of taxon \(i\) relative to the first taxon.
We may take the definition of the MWC model as being that, conditional on the total count \(M_{Tj}\), the observed read count of taxon \(i\) is proportional to its actual abundance times its efficiency times a taxon-independent factor \(C_j\) chosen to make sample \(j\)'s counts sum to \(M_{Tj}\),
\begin{align}
  \label{eq:M}
  M_{ij} = A_{ij} B^{(M)}_{i} C_j, && 
  C_j = \frac{M_{Tj}}{\sum_i A_{ij} B^{(M)}_{i}}
      = \frac{M_{Tj}}{A_{Tj} \bar B^{(M)}_j}.
\end{align}
In this model, the ratios among taxa are distorted by constant factors; for two taxa \(i\) and \(i'\),
\begin{align}
  \label{eq:M-ratio}
  \frac{M_{ij}}{M_{i'j}} 
    = \frac{m_{ij}}{m_{i'j}} 
    = \frac{A_{ij}}{A_{i'j}} \cdot \frac{B^{(M)}_{i}}{B^{(M)}_{i'}}
\end{align}
for all samples \(j\).
If one or both taxa are absent (\(A_i = 0\)) or undetectable (\(B_i = 0\)), both sides of the equation are either \(0\), \(\infty\), or \(0/0\) (undefined).
The observed proportion of a taxon equals its actual proportion multiplied by its efficiency relative to the sample mean efficiency,
\begin{align}
  \label{eq:m}
  m_{ij} = a_{ij} \cdot \frac{B_i}{\bar B^{(M)}_j}.
\end{align}

\hypertarget{bulk-absolute-abundance-measurement}{%
\subsection{Bulk absolute abundance measurement}\label{bulk-absolute-abundance-measurement}}

Consider estimating the total absolute abundance \(A_{T}\) via an aggregate or broad-range measurement, such as cell counting, flow cytometry, universal 16S qPCR, or total DNA concentration (e.g.~Qubit).
I let \(K_j\) donote the bulk absolute abundance measurement for sample \(j\) (mneumonic: ``K'' for the last letter in ``bulk'').

Bulk measurements are affected by taxon-specific bias \(B^{(K)}\), where \(B^{(K)}_i\) indicates how efficiently cells of taxon \(i\) contribute to the bulk measurement \(K\).
Unlike the metagenomics efficiencies, these bulk measurement efficiencies are absolute numbers.
For example, cells of one taxon may be more reliably counted than another; and for broad-range 16S qPCR measurement, we expect taxa to contribute in proportion to how reliably they are lysed and to their 16S copy number.
Let \(B^{(K)}_i\) denote the absolute efficiency with which a cell of taxon \(i\) contributes to the measurement in sample \(s\), which I assume is sample-independent.
We can write the estimate as
\begin{align}
  \label{eq:K}
  K_j = \sum_i A_{ij} B^{(K)}_i = A_{Tj} \bar B^{(K)}_j,
\end{align}
where \(\bar B^{(K)}_j\) is the mean bulk-measurement efficiency in the sample.

\hypertarget{targeted-absolute-abundance-measurement}{%
\subsection{Targeted absolute abundance measurement}\label{targeted-absolute-abundance-measurement}}

Targeted measurement of the absolute abundance of specific taxa can be made via a method such as qPCR, ddPCR, or counting CFUs selective media.
These measurements too may be subject to bias;
I suppose that the measurement of taxon \(i\) is given by
\begin{align}
  \label{eq:T}
  T_{ij} = A_{ij} B^{(T)}_i,
\end{align}
where \(B^{(T)}_i\) is the sample-independent, taxon-specific efficiency of the targeted measurement for taxon \(i\).

In discussing the case of multiple reference taxa, it will be useful to refer to the total targeted and metagenomics abundance and the average targeted and metagenomics efficiencies of the reference taxa.
I do so with the notation \(A_{Rj}\) and \(M_{Rj}\) (total abundances) and \(\bar B^{(T)}_{Rj}\) and \(\bar B^{(M)}_{Rj}\) (mean efficiencies) as defined above.
Note that
\begin{align}
  \label{eq:T-Rj}
  T_{Rj} 
    &= A_{Rj} \cdot \sum_{r \in R} \frac{A_{rj}}{A_{Rj}} \bar B^{(T)}_{Rj}
  \\&= A_{Rj} \bar B^{(T)}_{Rj}.
\end{align}

\hypertarget{spike-ins}{%
\subsection{Spike-ins}\label{spike-ins}}

The matrix \(S\) describes the nominal abundances with which spike-in taxa were added to each sample.
Spike-in experiments are often designed so that a taxon \(i\) is added in the same concentration to all samples, which we can represent by supposing the defined columns of \(S\) to be identical.
The bias \(B^{(S)}\) should be interpreted as error in the quantification of how much spike-in was added, such that
\begin{align}
  \label{eq:S}
  S_{ij} = A_{ij} B^{(S)}_i.
\end{align}
I assume that this error is constant across samples.
The motivation for this error model is as follows: Suppose that we misestimated the proportion of spike-in taxon \(s\in I_S\) in our original spike-in stock, by a factor \(B^{(S)}_s\); now its true abundance will be off by that factor in any spike-in derived from the starting stock.
Unlike the metagenomics efficiencies, these error factors are absolute numbers.

Taxa can be added as cells (pre-extraction) or as DNA (post-extraction).
In an application where DNA spike-ins are used to make inferences about changes in absolute cell abundances, we can include the difference between the nominal abundances \(S\) (the spiked DNA concentration) and the actual abundances \(A\) that are expected due to factors such as DNA extraction and copy-number variation in the bias term \(B^{(S)}\), so long as DNA extraction is linear (see section on saturation effects below).

(If we want to allow for different targeted and spike-in taxa, we can use \(R_S\) and \(R_T\).)

\hypertarget{complications}{%
\subsection{Complications}\label{complications}}

These models are the simplest for these types of experiments that have taxon-specific bias.
They therefore are idealizations that may be violated by real experiments in ways that affect our conclusions.
Here I list some relevant complications that we will return to in our analysis.
For more considerations, see the Discussion of \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}).

\hypertarget{abundances-units-and-protocol-stages}{%
\subsubsection{Abundances units and protocol stages}\label{abundances-units-and-protocol-stages}}

\emph{Still need to find the right way to synthesize these issues.}

qPCR measurements directly estimating the concentration of a marker gene in the extracted DNA.
Efficiency factors can translate the units from DNA to cell concentration \emph{and} account for the bias imposed by extraction and marker-gene CNV (copy-number variation).
Sometimes we may want to refer to the DNA concentrations in the sample after extraction, and in that case I will use the variable \(D\), and (perhaps) use \(B^{(T/D)}\) to refer to the bias of the targeted measurement vs the DNA concentrations it was applied to.
This can be useful when we are considering qPCR and also DNA spike-ins.

Note that the ``actual'' units could be chosen by the researcher to be something other than cell density; e.g.~biomass, or even 16S density.
But for concreteness I'll take to be cell density, as this seems to be what is most typically meant.

\hypertarget{saturation-in-dna-extraction-yields}{%
\subsubsection{Saturation in DNA extraction yields}\label{saturation-in-dna-extraction-yields}}

The equation for bulk measurement reflects an idealized form of bulk measurement in which, for a fixed taxonomic composition, \(K_j\) is directly proportional to \(A_T\).
However, a more realistic model for qPCR may be that \(K_j\) is a saturating function of \(A_T\), due to factors such as enzyme consumption or saturation in the elution step creating saturation in DNA yield during extraction.
However, it is not clear what this function should depend on (input biomass, cell concentration, total DNA?) and it is not clear what bias components should be included in determining the saturation, and these are not simply \(\bar B\) since that includes 16S copy-number which isn't relevant here.
Rather than try to accurately model the relationship between \(K\) and \(A\) in this more complicated scenario, I will use the models \(K_j = f(A_{Tj})\) and/or \(K_j = f(A_{Tj}\bar B^{(K)}_j)\) to get some intuition for the effect that a strong saturation effect has on our inferences when it is not accounted for.

Saturation during extraction also implications for targeted PCR measurements but not for measurements that work directly on cells; and for DNA spike-ins, but not cellular spike-ins.

\hypertarget{taxonomic-resolution-of-measurements}{%
\subsubsection{Taxonomic resolution of measurements}\label{taxonomic-resolution-of-measurements}}

I suppose that the ``taxa'' we consider are well-defined in a special sense, such that they can be equated both across samples within a measurement type and across different measurement types.
In other words, I assume that there is a real group of organisms corresponding to ``taxon \(i\)'' with abundance \(A_{i\cdot}\) that the various ``taxon \(i\)'' measurements are really measuring, such that we can, for example, directly equate the taxonomic source of the counts \(M_{i\cdot}\) and the targeted measurement \(T_{i\cdot}\).
Achieving this situation in practice requires careful attention to things like primer design and how the metagenomic taxonomic assignment is done, and even then it may not always be possible.

One important special case where this assumption breaks down is when measurements aggregate one or more of the \emph{atomic taxa} (corresponding to the rows in \(A\)), possibly in different ways by different measurement types.
This raises at least two sorts of complications.
First, constant multplicative bias at the level of atomic taxa does not translate into constant bias at the level of aggregate taxa (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan 2019}).
Second, measurements may target different aggregates.
For example, if our metagenomics protocol provides ASV or species-level estimates but our PCR-based targeted measurements may quantify a genus or family.
Often through sufficiently careful bioinformatics we can at least nest the metagenomics taxa within the targeted taxon, but complications remain if efficiency varies within the aggregates.

Maybe:

\begin{itemize}
\tightlist
\item
  Give example of effect of aggregation when efficiencies vary among the aggregated taxa; understand how in terms of the average efficiency within the aggregate.
\item
  Write out algebraically some of the cases I'll consider later.
\end{itemize}

\hypertarget{relative}{%
\section{Differential relative abundance}\label{relative}}

This section describes the effects of bias on common approaches to analyzing differential relative abundance.
We introduce the critical distinction between analyses based on proportions and analyses based on ratios.
Analyses based on proportions can be distorted by bias, whereas analyses of log fold changes (LFCs) in ratios of atomic taxa are invariant to bias.
If atomic taxa may be aggregated into synthetic taxa by multiplication (or by adding log abundance), and ratios formed from such aggregates; LFCs for such generalized ratios remain bias invariant.
But if atomic taxa are aggregated additively by summing their read counts---as often done to analyze higher-order taxa such as phyla---then bias invariance only applies if the aggregated taxa have the same efficiency.
The error induced by bias in (some) proportion-based methods has a simple form that suggests tractable experimental methods by which it may be estimated and corrected, which we discuss in Section \ref{calibration}.
Section \ref{absolute} shows that methods for obtaining absolute abundances from MGS data can generally be split into those based on proportions and those based on ratios, so that the proportion-ratio dichotomy also provides a useful framework for understanding analyses of differential absolute abundance.

\hypertarget{proportion-based-analyses}{%
\subsection{Proportion-based analyses}\label{proportion-based-analyses}}

\hypertarget{log-proportion}{%
\subsubsection{Log proportion}\label{log-proportion}}

To simplify notation where possible, I will write \(B\) for \(B^{(M)}\) when only the metagenomics measurements are relevant.

It follows from \eqref{eq:m} that the observed fold change in the proportion of taxon \(i\) from sample \(j\) to \(j'\) is
\begin{align}
  \label{eq:m-fc}
  \frac{m_{ij'}}{m_{ij}}
  = \frac{a_{ij'} \cancel{B_i} / \bar B_{j'}}{a_{ij} \cancel{B_i} / \bar B_{j}}
  = \frac{a_{ij'}}{a_{ij}} \cdot \frac{\bar B_j}{\bar B_{j'}}.
\end{align}
The taxon-specific error term (\(B_i\)) cancels, leaving a multiplicative error equal to the inverse change in the sample mean efficiency \(\bar B^{(M)}\).
Framed in terms of the log fold change, this equation becomes
\begin{align}
  \label{eq:m-lfc}
  \underbrace{ \log m_{ij'} - \log m_{ij} }_\text{observed LFC}
  &= \underbrace{ \log a_{ij'} - \log a_{ij} }_\text{true LFC}
  - \underbrace{ \left(\log \bar B_{j'} - \log \bar B_j\right) }_\text{LFC in mean efficiency},
\end{align}
so that the additive error is the LFC in mean efficiency between the two samples.

Analyzing fold changes in a regression framework typically entails considering the expectation of the logarithm of the response conditional on the value of a set of covariates \(X\).
I use \(E [Y \mid X]\) as shorthand for \(E [Y \mid X = x]\), the expected value of a random variable \(Y\) given a vector of covariate values \(x\).
Applying logarithms to \eqref{eq:m} gives
\begin{align}
  \label{eq:m-log}
  \log m_i = \log a_i + \log B_i - \log \bar B
\end{align}
which, after conditioning on the covariates, becomes
\begin{align}
  \label{eq:m-log-regression}
  E [\log m_i \mid X] 
  = E [\log a_i \mid X] + \log B_i - E [\log \bar B \mid X].
\end{align}
The taxon-specific term \(\log B_i\) creates a constant error that is unproblematic for differential abundance analysis, which is typically ignores the baseline abundance of the taxon.
However, the sample-specific error \(\log \bar B\) can distort the inferred relationship between \(X\) and the expected value of the true log proportion if its expected value also varies with \(X\).

\textbf{TODO: Explain this \(\gamma\) notation for the regression coefficients, or find a more recognizable notation.}

Consider the special case of the linear regression
\begin{align}
  \log a_i &= X \gamma + \epsilon,
\end{align}
where \(X\) is a \(J\times p\) covariate matrix and \(\gamma\) is a \(p\times 1\) vector of coefficients.
Suppose we knew the values of each of the terms in \eqref{eq:m-log}.
Setting each term as the left-hand-side of its own corresponding regression, the least-squares coefficients are related as (\eqref{eq:regression-coefficient-error})
\begin{align}
  \label{eq:m-log-regression-coefficients}
  \hat \gamma{(\log m_i)} = \hat \gamma{(\log a_i)} + \hat \gamma{(\log B_i)} - \hat \gamma{(\log \bar B)}.
\end{align}
For the simple linear regression,
\begin{align}
  \log a_i &= \gamma_0 + \gamma_1 x + \epsilon,
\end{align}
with an intercept coefficient \(\gamma_0\) and a slope coefficient \(\gamma_1\), the relation \eqref{eq:m-log-regression-coefficients} implies
\begin{align}
  \label{eq:m-log-simple-regression-coefficients}
  \hat \gamma_0{(\log m_i)} &= \hat \gamma_0{(\log a_i)} + \log B_i - \hat \gamma_0{(\log \bar B)} \\
  \hat \gamma_1{(\log m_i)} &= \hat \gamma_1{(\log a_i)} - \hat \gamma_1{(\log \bar B)},
\end{align}
Since \(\log B_i\) is constant, it only affects the intercept.
The slope estimate for the metagenomics proportions, \(\hat \gamma_1{(\log m_i)}\), is systematically decreased from that of the true proportions, \(\hat \gamma_1{(\log a_i)}\), by that for the sample mean efficiency, \(\hat \gamma_1{(\log \bar B)}\).

\hypertarget{log-odds-logit}{%
\subsubsection{Log odds (logit)}\label{log-odds-logit}}

The proportion of a taxon saturates at \(1\) as its absolute abundance increases, while the odds, \(a_i / (1 - a_i) = A_i / (A_T - A_i)\), continue to increase.
For this reason it is often more natural to perform regression on the log odds, or \emph{logit-transformed proportion}, rather than log proportion.

The observed log odds of taxon \(i\) is
\begin{align}
  \operatorname{logit} m_{ij} 
  = \log \frac{m_{ij}}{1 - m_{ij}}
  = \log \frac{M_{ij}}{M_{Tj} - M_{ij}}.
\end{align}
To understand the effects of bias on the measured log odds, it is helpful to define a variable \(\bar B_{-i,j}\) to denote the mean efficiency in sample \(j\) among
taxa \emph{other than \(i\)},
\begin{align}
  \label{eq:mean-efficiency-not-i}
  \bar B_{-i,j} 
  \equiv \frac{\sum_{k \ne i} B_k A_{kj}}{\sum_{k \ne i} A_{kj}}
  = \frac{\sum_{k \ne i} B_k a_{kj}}{1 - a_{ij}}.
\end{align}
The measured proportion of non-\(i\) taxa is
\begin{align}
  \label{eq:one-minus-m}
  1 - m_{ij}
  &= \sum_{k \ne i} m_{kj}
\\&= \sum_{k \ne i} a_{kj} \cdot \frac{B_k}{\bar B_j}
\\&= (1 - a_{ij}) \cdot \frac{\bar B_{-i,j}}{\bar B_j},
\end{align}
where the final expression follows from \(\sum_{k\ne i} a_{kj} B_k = (1-a_{ij}) \bar B_{-i,j}\).
Intuitively, Equation \eqref{eq:one-minus-m} says that the fold error in the proportion of ``not-\(i\)'' equals the mean efficiency of the ``not-\(i\)'' part of the sample relative to the mean efficiency of the sample as a whole.
This result means that the measured odds of taxon \(i\) is
\begin{align}
  \label{eq:m-odds}
  \frac{m_{ij}}{1 - m_{ij}}
  &= \frac{a_{ij}}{1 - a_{ij}} \cdot \frac{B_i}{\bar B_{-i,j}},
\end{align}
This formula resembles Equation \eqref{eq:M-ratio} for the measured ratio of two taxa \(i\) and \(i'\), but with taxon \(i'\) corresponding to all not-\(i\) taxa.
In this case, however, the efficiency of not-\(i\) varies among samples as changes in their relative abundances change \(\bar B_{-i, j}\).
Hence, in contrast to the ratio of two taxa, the error is not consistent across samples.

The error in the logit of taxon \(i\) is
\begin{align}
  \label{eq:m-logit}
  \operatorname{logit} m_i = \operatorname{logit} a_i + \log B_i - \log \bar B;
\end{align}
note the log rather than logit operators in the efficiency terms.
The corresponding regression equation is
\begin{align}
  \label{eq:m-logit-regression}
  E [\operatorname{logit} m_i \mid X] 
  = E [\operatorname{logit} a_i \mid X] + \log B_i - E [\log \bar B_{-i} \mid X].
\end{align}

\hypertarget{ratio-based-analyses}{%
\subsection{Ratio-based analyses}\label{ratio-based-analyses}}

The previous section showed that differential abundance analyses based on proportions may be sensitive to bias, due to the dependence of the multiplicative error in measured proportions on the sample mean efficiency (Equation \eqref{eq:m}).
In contrast, the multiplicative error in the ratios among taxa (Equation \eqref{eq:M-ratio}) is independent of sample composition, such that a differential abundance analysis based on log ratios is invariant to bias.
The simplest such analysis is when ``abundance'' is equated with the log ratio of two specific taxa; however, any linear combination of log ratios may be used, which includes a variety of so-called Compositional Data Analysis (CoDA) methods that have recently been applied in microbiome DA analysis (see below).
\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) described the bias invariance of ratio-based methods (see their Equations 6 and 7).
Here we show how this general result applies to specific types of differential relative abundance.

We first consider the ratio a pair of taxa \(i\) and \(i'\).
It follows from Equation \eqref{eq:M-ratio} that the measured fold change in the ratio of taxon \(i\) to taxon \(i'\) from sample \(j\) to sample \(j'\) is
\begin{align}
  \label{eq:M-ratio-fc}
  \frac{m_{ij'} / m_{i'j'}}{m_{ij} / m_{i'j}}
  = \frac{a_{ij'} \cancel{B_i} / a_{i'j'} \cancel{B_{i'}}}
         {a_{ij} \cancel{B_i} / a_{i'j} \cancel{B_{i'}}}
  = \frac{a_{ij'} / a_{i'j'}}{a_{ij} / a_{i'j}}.
\end{align}
The error in each ratio, \(B_i / B_{i'}\), is constant and so cancels completely, leaving the true fold change.
To see the effect of bias on regression of log ratios, we note that the error equation \eqref{eq:M-ratio} implies
\begin{align}
  \label{eq:M-ratio-log}
%  \log m_i - \log m_{i'} = \log a_i - \log a_{i'} + \log B_i - \log B_{i'} \\
  \log \frac{m_i}{m_{i'}} = \log \frac{a_i}{a_{i'}} + \log \frac{B_i}{B_{i'}}.
\end{align}
Taking the conditional expectation given a covariate vector \(X\) gives
\begin{align}
  \label{eq:m-ratio-log-regression}
  E\left[ \log \frac{m_i}{m_{i'}} \mid X \right] 
  = E\left[ \log \frac{a_i}{a_{i'}} \mid X \right] + \log \frac{B_i}{B_{i'}}.
\end{align}
In contrast to the case of a log proportion (Equation \eqref{eq:m-log-regression}), the effect of bias on the log ratio is simply to create a constant shift and therefore does not affect differential-abundance analysis.

CoDA DA methods often draw upon generalized notion of ratio, in which the numerator and/or denominator consist of a product of powers of multiple taxa.
Such a generalized ratio is determined by length-\(I\) vectors \(n\) and \(d\) giving the power of each taxon in the numerator and denominator, with \(n_i = 0\) indicating that the taxon does not affect the term.
Denoting the actual value of a given generalized ratio in sample \(j\) as \(y_j\) and the observed as \(z_j\), we have
\begin{align}
  \label{eq:generalized-ratio}
  \text{actual:}   \quad  y_j = \frac{\prod_i A_{ij}^{n_i}}{\prod_i A_{ij}^{d_i}} &&
  \text{observed:} \quad  z_j = \frac{\prod_i M_{ij}^{n_i}}{\prod_i M_{ij}^{d_i}}.
\end{align}
Under the MWC bias model, the observed and actual log ratios are related as
\begin{align}
  \label{eq:generalized-ratio-log}
  \log z_j &= \log y_j + \log \frac{\prod_i B_{ij}^{n_i}}{\prod_i B_{ij}^{d_i}}.
\end{align}
Hence, as with the simple ratio between two taxa, the error due to bias is independent of sample composition and so does not affect the observed variation in \(\log z_j\).

Most of the log-ratio transformations that are used in CoDA DA analysis can be understood as particular applications of the generalized ratio \eqref{eq:generalized-ratio} and thus lead to bias-invariant DA results under the deterministic MWC model.
For example, the additive log-ratio (ALR) transformation consists of computing the log of the ratios obtained by dividing each taxon's abundance by that of a particularly chosen reference taxon \(r\),
\begin{align}
  \label{eq:alr}
  \operatorname{alr} A_{j} = \left[ \log \frac{A_{ij}}{A_{rj}}, \dots, \log \frac{A_{Ij}}{A_{rj}} \right],
\end{align}
while the centered log-ratio (CLR) transformation instead sets the denominator to the geometric mean \(g(A_j) = \left(\prod_i A_{ij}\right)^{1/I}\) of all taxa,
\begin{align}
  \label{eq:clr}
  \operatorname{clr} A_{j} = \left[ \log \frac{A_{ij}}{g(A_{j})}, \dots, \log \frac{A_{Ij}}{g(A_{j})} \right].
\end{align}
More generally, one might choose any set \(R\) of reference taxa and use their geometric mean in the denominator;
the ALR and CLR transformations then correspond to taking \(R=r\) and \(R=I\), respectively.
\protect\hyperlink{ref-quinn2019afie}{Quinn et al.} (\protect\hyperlink{ref-quinn2019afie}{2019}) call this transformation the \emph{multiple additive log-ratio} (MALR) transformation,
\begin{align}
  \label{eq:malr}
  \operatorname{malr} A_{j} = \left[ \log \frac{A_{ij}}{g(A_{R,j})}, \dots, \log \frac{A_{Ij}}{g(A_{R,j})} \right].
\end{align}
The actual and observed MALR-transformed abundance associated with a specific numerator taxon \(i\) is
\begin{align}
  \label{eq:malr-error}
  \operatorname{malr}(M_{j})_i = \operatorname{malr}(A_{j})_i + \operatorname{malr}(B)_i;
\end{align}
hence the error due to bias is constant and does not factor into the inferred changes in a DA analysis.
Note: Unlike \protect\hyperlink{ref-quinn2019afie}{Quinn et al.} (\protect\hyperlink{ref-quinn2019afie}{2019}) we do not include the ``robust centered log-ratio transformation'' of \protect\hyperlink{ref-martino2019anov}{Martino et al.} (\protect\hyperlink{ref-martino2019anov}{2019}) as an example of an MALR; this transformation chooses a distinct set of reference taxa for each sample and so does not have bias invariance under the MWC model.

Other generalized log-ratios commonly used in microbiome data analysis include \emph{balances}.
A balance is defined by two sets of taxa, \(Q\) and \(R\), and equals the log of the ratio of the geometric mean of \(Q\) taxa to the geometric mean of \(R\) taxa, multiplied by a scaling factor,
\begin{align}
  \label{eq:balance}
  \sqrt{\frac{|Q| |R|}{|Q| + |R|}} \log \frac{g(A_{Q,j})}{g(A_{R,j})}.
\end{align}
Balances have become a popular tool for identifying biomarkers or groups of taxa associated with an environmental or health condition while operating within the CoDA framework (\protect\hyperlink{ref-washburne2017phyl}{Washburne et al.} (\protect\hyperlink{ref-washburne2017phyl}{2017}), \protect\hyperlink{ref-riverapinto2018bala}{Rivera-Pinto et al.} (\protect\hyperlink{ref-riverapinto2018bala}{2018}), \protect\hyperlink{ref-quinn2020inte}{Quinn and Erb} (\protect\hyperlink{ref-quinn2020inte}{2020})).
The exponential of a balance is an example of the generalized ratio in \eqref{eq:generalized-ratio} and so regression of balances is again invariant to bias.

The products in the numerators and denominators in \eqref{eq:generalized-ratio} can be seen as a way to multiplicatively aggregate abundances of different taxa, as opposed to the additive aggregation that is commonly used when relative abundances are viewed as proportions.
Multiplicative aggregation preserves the property of perturbation invariance that is the source of bias invariance under the MWC model.
In contrast, additive aggregation (known as ``amalgamation'' in the CoDA literature) violates perturbation invariance and can lead to regression results that depend on bias even when they are ostensibly based on the ratios of two (non-atomic) taxa, as we now explain.

\hypertarget{higher-order-taxa-formed-by-additive-aggregation}{%
\subsection{Higher-order taxa formed by additive aggregation}\label{higher-order-taxa-formed-by-additive-aggregation}}

Microbiome researchers often combine lower-order taxa into higher-order taxa prior to conducting a DA analysis.
For instance, Amplicon Sequence Variants (ASVs), Operational Taxonomic Units (OTUs), or species-level counts may be aggregated into genus- or family-level counts by simply summing the counts within each group of lower-level taxa.
Such aggregation can increase statistical power by reducing noise, sparsity, and the number of tests conducted.
It also simplifies the task of interpretation by limiting the number of taxa considered.
In addition, some degree of uncontrolled aggregation is inevitable due to the inherent limitations of our sequencing and bioinformatics protocols to distinguish sufficiently similar organisms (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})), which may be lumped together into an OTU or species-level feature.
Finally, though our focus is on analysis of taxa, we note that taxonomic aggregation also occurs (at least implicitly) in gene or function analyses, where counts are combined from different taxa that share the same genes or predicted functions.
Hence, to understand the effects of bias on DA analysis as it is actually practiced, we must understand the effects of such taxonomic aggregation on differential abundance.

Consider a set of atomic taxa given by the set \(Q \subset \{1, \dots, I\}\) and let \(A_{Qj}\) be the vector of abundances in sample \(j\) of the taxa in \(Q\).
The actual absolute abundance of the synthetic taxon \(Q\) is simply the sum of abundances of its component taxa, \(\operatorname{sum}\left(A_{Q,j}\right) = \sum_{q \in Q} A_{qj}\).
Similarly, the actual proportion of \(Q\) is given by the sum of proportions, \(\operatorname{sum}\left(a_{Q,j}\right) = \sum_{q \in Q} a_{qj} = \operatorname{sum}\left(A_{Q,j}\right) / A_{Tj}\).
The abundance of \(Q\) in the metagenomics measurement is \(\operatorname{sum}\left(M_{Q,j}\right)\), and its proportion \(\operatorname{sum}\left(m_{Q,j}\right)\).

Such a synthetic taxon has a consistent, composition-independent measurement efficiency only if its component taxa have equal efficiencies.
Let \(\bar B_{Q,j}\) denote the mean efficiency of the taxa in \(Q\) in sample \(j\),
\begin{align}
  \label{eq:subset-mean-efficiency}
  \bar B_{Q,j} = \frac{\sum_{q \in Q} B_q A_{qj}}{\sum_{q \in Q} A_{qj}}.
\end{align}
Unless all the taxa in \(Q\) have the same efficiency, the mean efficiency of \(Q\) varies with the relative abundances among these taxa.
Summing over Equation \eqref{eq:m} and performing some rearranging shows that the observed proportion of \(Q\) is
\begin{align}
  \label{eq:sum-m}
  \operatorname{sum}\left(m_{Q,j}\right) 
  &= \frac{1}{\bar B_j} \sum_{q\in Q} a_{qj} B_q \sum_{q\in Q} m_{qj} \\
  &= \operatorname{sum}\left(a_{Q,j}\right) \cdot \frac{\bar B_{Q,j}}{\bar B_j}.
\end{align}
In words, the error in the proportion of the aggregate taxon \(Q\) is given by the mean efficiency of taxa in \(Q\) relative to the sample mean.
Hence only if the atomic taxa that make up \(Q\) all have the same efficiencies or always appear in the same ratios to each other will the efficiency of \(Q\) be constant across samples.
This observation was first made in \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) (Discussion and Appendix 1) and is here extended to its general form.

The lack of consistency in the efficiency of the synthetic taxon \(Q\) complicates the form of the error in both proportion- and ratio-based DA analyses.
The equivalent of the regression equation \eqref{eq:m-log-regression} for the log proportion of \(Q\) is
\begin{align}
  \label{eq:sum-m-log-regression}
  E \left[ \log \sum_{q\in Q} m_{q} \mid X \right] 
  = E \left[ \log \sum_{q\in Q} a_{q} \mid X \right] + E\left[\log \bar B_{Q} - \log \bar B \mid X\right];
\end{align}
we can no longer move the log efficiency of the focal taxon out of the expectation operator and so must consider the variation in the mean efficiency of \(Q\) as well as of the sample as a whole.
To consider the effect on ratio-based analyses, we define another synthetic taxon \(R\subset \{1, \dots, I\}\) and consider the ratio of \(Q\) to \(R\).
The observed ratio,
\begin{align}
  \label{eq:M-ratio-u-v}
  \frac{M_{Q,j}}{M_{R,j}}
  &= \frac{A_{Q,j}}{A_{R,j}} \cdot \frac{\bar B_{Q,j}}{\bar B_{R,j}},
\end{align}
has an error equal to the ratio in mean efficiency of \(Q\) to \(R\), both of which can vary with the relative abundances of the component taxa.
As the ratio no longer as a consistent error, it is possible for bias to lead to spurious inferences in the fold change in the ratio.

\hypertarget{absolute}{%
\section{Differential absolute abundance}\label{absolute}}

This section considers various ways to estimate absolute abundances of individual taxa for the purposes of estimating log fold changes (LFCs) in absolute abundance, or more generically of performing linear regression on log absolute abundance.

\hypertarget{using-bulk-abundance-measurements}{%
\subsection{Using bulk abundance measurements}\label{using-bulk-abundance-measurements}}

The absolute abundance of taxon \(i\) is estimated by multiplying its metagenomics proportion by the bulk abundance measurement,
\begin{align}
  \label{eq:A-hat-bulk}
  \hat A_{ij} &= m_{ij} K_j.
\end{align}
The error in this estimate due to the experimental bias in \(M\) and \(K\) is given by substituting equations \eqref{eq:m} and \eqref{eq:K} into \eqref{eq:A-hat-bulk}, giving
\begin{align}
  \label{eq:error-bulk}
  \hat A_{ij} &= \frac{a_{ij} B^{(M)}_i}{\bar B^{(M)}_j} \cdot A_{Tj} \bar B^{(K)}_j
  \\&= A_{ij} \cdot \frac{B^{(M)}_i \bar B^{(K)}_j}{\bar B^{(M)}_j}.
\end{align}
The fold error in the estimated abundance of a particular taxon in a particular sample thus equals the metagenomics efficiency of that taxon times the ratio of the sample mean efficiencies of the bulk measurement to the metagenomics measurement.
Next I consider the error in the estimated fold change in \(A_i\) between two samples \(j\) and \(j'\).
As in the case of log proportion estimates, the taxon-specific, sample-independent error term (\(B^{(M)}_i\)) cancels, but the taxon-independent, sample-specific error term (\({\bar B^{(K)}_j}/{\bar B^{(M)}_j}\)) does not, giving
\begin{align}
  \label{eq:fc-bulk}
  \frac{\hat A_{ij'}}{\hat A_{ij}}
  &= \frac{A_{ij'}}{A_{ij}} 
  \cdot \frac{\bar B^{(M)}_j \bar B^{(K)}_{j'}}{\bar B^{(M)}_{j'} \bar B^{(K)}_j}.
  \\&= \frac{A_{ij'}}{A_{ij}} 
  \cdot \frac{\bar B^{(K)}_{j'} / \bar B^{(M)}_{j'}}{\bar B^{(K)}_j / \bar B^{(M)}_j }.
\end{align}
In words, the fold error in the estimated fold change in the absolute abundance of any taxon equals the fold change in the ratio of the mean efficiency of the bulk abundance measurement to that of the metagenomics measurement.

TODO: Add regression result.

\hypertarget{shared-bias-components}{%
\subsubsection{Shared bias components}\label{shared-bias-components}}

Motivation: Consider an experiment in which the community has been profiled by 16S sequencing and bulk abundance has been estimated by 16S qPCR.
The taxon-specific bias associated with each of these measurements may be highly similar; both are measuring 16S copies in the same aliquot of already-extracted DNA, which has already been affected by DNA extraction bias and 16S copy-number variation.
If the same primers are used then PCR bias might also be shared.
How does having a large amount of bias being shared between the bulk and sequencing measurement affect the error in abundance estimates in individual samples and the fold changes between samples?

In this case, the ratio of mean efficiencies, \(\bar B^{(K)}_j / \bar B^{(M)}_j\), will vary much less across samples than either mean efficiency itself, such that the error term in \eqref{eq:fc-bulk} will be small.
Therefore the estimated fold changes may be accurate despite potentially large variation in the metagenomics sample mean efficiency across samples.
Note that the estimated absolute abundance in the individual sample, \eqref{eq:error-bulk}, remains inaccurate since the unknown taxon-specific efficiency remains.

TODO: Add derivation, where we formally define \(B^{(M)}\) and \(B^{(K)}\) in terms of shared and non-shared components, and show that the variation in the ratio of mean efficiencies reduces to that of the non-shared components.

\hypertarget{using-reference-taxa}{%
\subsection{Using reference taxa}\label{using-reference-taxa}}

TODO: Consider rewriting to be in terms of the ratio of counts * the reference abundance, to emphasize that this is a ratio-based method.

\hypertarget{taxa-with-targeted-abundance-measurements}{%
\subsubsection{Taxa with targeted abundance measurements}\label{taxa-with-targeted-abundance-measurements}}

First consider the case of just a single reference taxon, \(r\), for which we have targeted measurements.
We can estimate the absolute abundances of all taxa by scaling the read counts by the ratio of the abundance measurement to read count for the reference taxon,
\begin{align}
  \label{eq:A-hat-targeted}
  \hat A_{ij} &= M_{ij} \cdot \frac{T_{rj}}{M_{rj}}.
\end{align}
Accounting for systematic error in both the metagenomcs measurement \eqref{eq:M} and the targeted measurement \eqref{eq:T} gives
\begin{align}
  \label{eq:error-targeted}
  \hat A_{ij} &= A_{ij} \cdot \frac{B^{(M)}_i B^{(T)}_r}{B^{(M)}_r}.
\end{align}
Since the error in \label{eq:error-targeted} is sample-independent, it cancels out when we use this equation to compute the fold change between two samples,
\begin{align}
  \label{eq:fc-targeted}
  \frac{\hat A_{ij'}}{\hat A_{ij}} &=\frac{\hat A_{ij'}}{\hat A_{ij}}.
\end{align}
Thus our model predicts that normalizing read counts to a reference taxon, while providing systematically-distorted abundance estimates, provides accurate fold-change estimates.

\hypertarget{cellular-spike-ins}{%
\subsubsection{Cellular spike-ins}\label{cellular-spike-ins}}

Spike-ins are a way to create a reference taxon whose abundance is known by design and so doesn't need to be measured.
Typically the spike-in is added in a fixed concentration to all samples; however, this is unnecessary, so long as we know the amount it was added.
In either case we can use \eqref{eq:A-hat-targeted} to estimate the abundances of all taxa from the spike-in taxa, substituting the spike-in abundance \(S_{ir}\) for \(T_{ir}\),
Our model predicts that normalizing read counts to a spike-in taxon gives systematically-distorted abundance estimates but accurate fold-change estimates for the same reason as in the targeted case.

\hypertarget{dna-spike-ins}{%
\subsubsection{DNA spike-ins}\label{dna-spike-ins}}

\hypertarget{taxa-assumed-to-have-a-constant-abundance}{%
\subsubsection{Taxa assumed to have a constant abundance}\label{taxa-assumed-to-have-a-constant-abundance}}

By this approach, we identify one or more taxa that we assume to have a constant absolute abundance across samples.
Inference proceeds as in the targeted and spike-in case, using \eqref{eq:A-hat-targeted} but with the reference abundance \(T_r\) set to an arbitrary constant as we may not know the true abundance.
In this case, the effect of bias cancels in fold-change calculations for the same reasons as the targeted and spike-in cases.
Note: The reference taxon could be the host.

\hypertarget{multiple-reference-taxa}{%
\subsubsection{Multiple reference taxa}\label{multiple-reference-taxa}}

Under the deterministic model we consider here, having multiple reference taxa in the same sample is completely redundant - we learn no new information.
In reality, having information from multiple taxa from a single sample should allow us to make more precise estimates, provided that we use them in a way so that the effects of bias do not offset any reductions in noise.
How might we leverage measurements of multiple reference taxa, \(r \in R\), in a way that remains robust to bias?

First consider the situation where we can assume that all reference taxa are in each sample, as would typically be the case for a spike-in.
One possibility is to sum the metagenomic and targeted abundances from the different taxa,
\begin{align}
  \label{eq:A-hat-targeted-sum}
  \hat A_{ij} &= M_{ij} \cdot \frac{\sum_{r \in R} T_{rj}}{\sum_{r \in R} M_{rj}}.
\end{align}
To see this, let \(A_{Rj} = \sum_{r \in R} A_{rj}\) be the summed actual abundances of the reference taxa.
Similarly define \(M_{Rj} = \sum_{r \in R} M_{rj}\) and \(T_{Rj} = \sum_{r \in R} T_{rj}\) be the summed read counts and targeted measurements.
Now
\begin{align}
  \hat A_{ij} &= M_{ij} \cdot \frac{T_{Rj}}{M_{Rj}}.
\end{align}
Also let \(\bar B^{(M)}_{Rj} = \sum_{r \in R} B^{(M)}_{rj} A_{rj} / A_{Rj}\) be the mean metagenomics efficiency among the reference taxa and similarly define \(\bar B^{(T)}_{Rj}\) as the mean targeted efficiency.
A little algebra shows that
\begin{align}
  \label{eq:T-Rj}
  T_{Rj} 
    &= A_{Rj} \cdot \sum_{r \in R} \frac{A_{rj}}{A_{Rj}} \bar B^{(T)}_{Rj}
  \\&= A_{Rj} \bar B^{(T)}_{Rj}
\end{align}
while
\begin{align}
  \label{eq:M-Rj}
  M_{Rj} &= A_{Rj} \bar B^{(M)}_{Rj} C_j.
\end{align}
And since \(M_{ij} = A_{ij} B^{(M)}_{i} C_j\) \eqref{eq:M} we can write the estimated \eqref{eq:A-hat-targeted-sum} abundance of taxon \(i\) as
\begin{align}
  \label{eq:error-targeted-sum}
  \hat A_{ij} &= A_{ij} \cdot \frac{B^{(M)}_i \bar B^{(T)}_{Rj}}{\bar B^{(M)}_{Rj}}.
\end{align}
The error now depends on relative abundances among the reference taxa through their mean bias values.
Therefore, fold changes estimated using this approach are not necessarily robust to bias if the relative abundances among the reference taxa vary.
Spike-in experiments are typically designed so that the reference taxa typically have the same relative abundances with respect to each other in all samples, in which case this method would remain robust to bias.

Another approach is to multiply by the geometric mean abundances of the taxa,
\begin{align}
  \label{eq:A-hat-targeted-prod}
  \hat A_{ij} &= M_{ij} \cdot \left[\frac{\prod_{r \in R} T_{rj}}{\prod_{r \in R} M_{rj}}\right]^{1/|R|}.
\end{align}
The error in this case is
\begin{align}
  \label{eq:error-targeted-prod}
  \hat A_{ij} 
  &= A_{ij} B^{(M)}_r C_j \cdot 
    \left[\frac{\prod_{r \in R} A_{rj} B^{(T)}_r} {\prod_{r \in R} A_{rj} B^{(M)}_r C_j} \right]^{1/|R|} \\
  &= A_{ij} \cdot 
    \left[\frac{\prod_{r \in R} B^{(T)}_r} {\prod_{r \in R} B^{(M)}_r} \right]^{1/|R|}.
\end{align}
The multiplicative error is now constant and will cancel in fold-change calculations even if the reference taxa vary in their relative abundances.

Thus the geometric-mean method appears to be robust to bias, but as written it requires that all reference taxa are present in all samples, whereas the summation approach is sensitive to bias but only requires at least one reference taxon to be present in the sample.

Now consider the case where different reference taxa are present (above our detection limit) in each sample, as might be the case for naturally occuring reference taxa.
We can modify the geometric-mean approach to work (so long as at least one reference is present in each sample) as follows.
From the samples where multiple references are present, we are able to measure the differnetial bias \(B^{(M/T)}_r / B^{(M/T)}_{r'}\) among the references \(r, r'\).
From measurement of reference \(r'\), we know the value of \(T_{r'j} / M_{r'j}\), which combined with the differential bias tells us what the ratio \(T_{rj} / M_{rj}\) should equal, namely
\begin{align}
  \frac{T_{rj}}{M_{rj}} = \frac{T_{r'j}}{M_{r'j}} \cdot \frac{B^{(M/T)}_r}{B^{(M/T)}_{r'}}.
\end{align}
Therefore, in samples where \(r\) is missing but \(r'\) is present, we can replace
\(\frac{T_{rj}}{M_{rj}}\) in the product with its predicted value.
(More generally, we can use the geometric mean of the values predicted by each of the reference taxa that are present).

This imputation procedure is a simple plug-in approach to show that it is possible to use a multiplicative approach to using multiple references even when the references vary in their presence among samples.
Real inference should ideally be done with a fully generative statistical model that includes bias as well as noise-generating processes and can be used for maximum likelihood or Bayesian inference. In such models, the imputation might be implicitly handled as part of the likelihood, and more generally the values \(\frac{T_{rj}}{M_{rj}}\) for various \(r\) can be naturally and automatically weighted (in terms of their influence on \(\hat A_{ij}\)) according to their predicted precision by the model fitting procedure.

\hypertarget{using-an-equivolumetric-protocol}{%
\subsection{Using an equivolumetric protocol}\label{using-an-equivolumetric-protocol}}

\hypertarget{computational-methods}{%
\subsection{Computational methods}\label{computational-methods}}

\hypertarget{proofs-of-regression-results}{%
\section{Proofs of regression results}\label{proofs-of-regression-results}}

\hypertarget{general-regression}{%
\subsection{General regression}\label{general-regression}}

Regression analysis can often be framed as seeking the regression function \(r(x)\) that describes how the expected value of some response variable \(Y\) varies with the a vector of covariates \(X\) (\protect\hyperlink{ref-wasserman2004allo}{Wasserman} (\protect\hyperlink{ref-wasserman2004allo}{2004})),
\begin{align}
  r(x) = E [Y \mid X = x].
\end{align}
For example, the response might be log absolute abundance of a particular taxon.
In our case, however, we don't know \(Y\), but rather a measure that is subject to random and systematic error, which I call \(Z\).
Let \(D = Z - Y\) be the difference between the true response \(Y\) and its measurement \(Z\), so that \(Z = Y + D\).
It follows that
\begin{align}
  \label{eq:regression-error}
  E [Z \mid X] = E [Y \mid X] + E [D \mid X].
\end{align}

\hypertarget{linear-least-squares-regression}{%
\subsection{Linear least-squares regression}\label{linear-least-squares-regression}}

\begin{theorem}[Linearity of regression coefficients]
\protect\hypertarget{thm:regression-linearity}{}\label{thm:regression-linearity}

Consider a scalar response variable with \(J\) observations, which I represent by the vector \(y\), that equals a sum of \(K\) component response variables \(y^{(k)}\) each scaled by a non-zero factor \(c^{(k)}\),
\begin{align}
  \label{eq:y-sum}
  y = \sum_{k = 1}^K c^{(k)} y^{(k)}.
\end{align}
(We may need to assume that the \(y^{(k)}\) are linearly independent.)
Let \(X\) be a \(J\)-by-\(p\) matrix of covariates for the \(J\) observations, with linearly independent columns.
Let \(\hat \gamma\) denote the least-squares estimate of the coefficient matrix \(\gamma\) in the linear regression
\begin{align}
  y = X \gamma + \epsilon.
\end{align}
Similarly, let \(\hat \gamma^{(k)}\) denote the least-squares estimates for the \(K\) linear regressions
\begin{align}
  y^{(k)} = X \gamma^{(k)} + \epsilon^{(k)}.
\end{align}
The least-squares coefficient estimates for \(y\) are given by the sum of those of the \(y^{(k)}\),
\begin{align}
  \hat \gamma = \sum_{k=1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}

\end{theorem}

\begin{proof}

The Moore-Penrose pseudoinverse of \(X\) is \(X^+ = (X^T X)^{-1} X^T\).
The least squares estimates are given by multiplying the matrix \(X^+\) by the corresponding response vector (\href{https://en.wikipedia.org/wiki/Ordinary_least_squares\#Matrix/vector_formulation}{Wikipedia}), so that \(\hat \gamma = X^+ y\) and \(\hat \gamma^{(k)} = X^+ y^{(k)}\) (for \(k = 1, \dots, K\)).
The result then follows from the stipulation \eqref{eq:y-sum} and the linearity of matrix multiplication,
\begin{align}
  \hat \gamma 
  = X^+ y 
  = X^+ \sum_{k = 1}^K c^{(k)} y^{(k)}
  = \sum_{k = 1}^K c^{(k)} X^+ y^{(k)}
  = \sum_{k = 1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}

\end{proof}

We can use \ref{thm:regression-linearity} to describe how measurement error, such as that caused by experimental bias, affects least-squares estimates of regression coefficients.

\begin{theorem}[Error in regression coefficients]
\protect\hypertarget{thm:regression-error}{}\label{thm:regression-error}

Suppose that \(y\) is the response variable we wish to understand, \(z\) is our imperfect measurement of \(y\), and \(d = z - y\) the difference between the two, so that \(z = y + d\).
Let each be a vector of length \(J\) describing a set of \(J\) observations.
Let \(X\) be a \(J\)-by-\(p\) matrix of covariates for the \(J\) observations, with linearly independent columns.
Consider the linear regression equations for \(y\), \(d\), and \(z\),
\begin{align}
  y &= X \gamma^{(y)} + \epsilon^{(y)} \\
  z &= X \gamma^{(z)} + \epsilon^{(z)} \\
  d &= X \gamma^{(d)} + \epsilon^{(d)}.
\end{align}
The relationship between the least-squares estimates mirrors that between the variables themselves,
\begin{align}
  \label{eq:regression-coefficient-error}
  \hat \gamma^{(z)} = \hat \gamma^{(y)} + \hat \gamma^{(d)}.
\end{align}

\end{theorem}

\begin{proof}

The result follows directly from \(z = y + d\) and \ref{thm:regression-linearity}.

\end{proof}

\begin{corollary}
\protect\hypertarget{cor:regression-error-simple}{}\label{cor:regression-error-simple}

For the special case of simple linear regression,
\begin{align}
  y &= \gamma_0^{(y)} + \gamma_1^{(y)} x + \epsilon^{(y)} \\
  z &= \gamma_0^{(z)} + \gamma_1^{(z)} x + \epsilon^{(z)} \\
  d &= \gamma_0^{(d)} + \gamma_1^{(d)} x + \epsilon^{(d)},
\end{align}
the estimates for the intercept and slope coefficients satisfy
\begin{align}
  \hat \gamma_0^{(z)} &= \hat \gamma_0^{(y)} + \hat \gamma_0^{(d)} \\
  \hat \gamma_1^{(z)} &= \hat \gamma_1^{(y)} + \hat \gamma_1^{(d)}.
\end{align}

\end{corollary}

Next, aiming to show the forms for the different transformations of interest.
It might be useful to cover the case where \(d\) is split into a constant and a sample-specific part, since probably everything of interest will follow from that.

What, if any general results should go in the main text?

Also relate the expected values of \(y\) and \(z\) conditional on \(X\)?.

To figure out what I need:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  extend the below result for log proportions to multiple regression
\item
  work out the derivation in terms of \ref{thm:regression-error}
\item
  Write a corollary if it seems to be useful for doing all the cases (e.g.~logit etc)
\end{enumerate}

\hypertarget{linear-regression-of-microbiome-abundances}{%
\subsubsection{Linear regression of microbiome abundances}\label{linear-regression-of-microbiome-abundances}}

First, consider log proportions.
In this case, the intercept coefficient is biased upward by \(\log B_i\) and the other coefficients are biased downward by the corresponding coefficient of \(\log \bar B\), based on the error formulas for individual samples.

TODO: write the general result (for multiple regression, assuming that an intercept term is present.

\begin{proposition}
\protect\hypertarget{prp:unlabeled-div-3}{}\label{prp:unlabeled-div-3}

Consider the linear equations
\begin{align}
  \label{eq:least-squares-formulation}
  \log a_i    &= \alpha_0 + \alpha_1 x \\
  \log \bar B &= \beta_0 + \beta_1 x   \\
  \log m_i    &= \gamma_0 + \gamma_1 x,
\end{align}
and let \(\hat \alpha_0\), etc. denote the least-squares estimates for the regression coefficients supposing perfect information (i.e., that the \(a_i\) and \(B_i\) are known).
The coefficients for \(\log m_i\) are related to those for \(\log a_i\) and \(\log \bar B\) through the equations
\begin{align}
  \label{eq:least-squares-estimates}
  \hat \gamma_0 &= \hat \alpha_0 + \log B_i - \hat \beta_0 \\
  \hat \gamma_1 &= \hat \alpha_1 - \hat \beta_1.
\end{align}

\end{proposition}

\hypertarget{dna-measurement-and-spike-ins}{%
\section{DNA measurement and spike-ins}\label{dna-measurement-and-spike-ins}}

This appendix aims to derive expressions for the error in individual samples and figure out whether the error is constant across samples (so that it cancels out LFC analysis).
Why difference? Extraction has occured prior to the targeted measurement.

Punchline: Bias will cancel like before if and only if DNA extraction yield is perfectly proportional to input, in a particular sense I should perhaps create a name for.
It needs to be proportional to input for a fixed composition; it doesn't have to be proportional across compositions.

\hypertarget{expanded-model-and-notation}{%
\subsection{Expanded model and notation}\label{expanded-model-and-notation}}

I partition the experiment into extraction and sequencing steps,
\begin{align}
  A_{ij} 
  \overset{\text{extraction}}{\underset{F_{ij}'}\longrightarrow}
  A_{ij}'
  \overset{\text{sequencing}}{\underset{F_{ij}''}\longrightarrow}
  \to A_{ij}'' = M_{ij}.
\end{align}
where \(A'_{ij}\) denotes the absolute abundances after extraction.
The abundances \(A''_{ij}\) after sequencing are just the counts \(M_{ij}\).
I define a matrix \(F_{ij}'\) of factors equalling the fold change in abundance during the extraction step, \(F_{ij}' = A_{ij}' / A_{ij}\), and similarly let \(F_{ij}''\) be the fold changes during the sequencing step,
The total fold change is just the product: \(F_{ij} = F'_{ij} F''_{ij} = M_{ij} / A_{ij}\).

The MWC model implies that the relative values of \(F_{ij}\) to \(F_{i'j}\) are independent of \(j\), for both the protocol as a whole and the individual steps.
Therefore we can write the factors as product of the taxon-specific bias \(B_i\) and a sample-specific scaling factor \(C_j\), as in Section \ref{models}.
I use \('\)s to do this for the different steps, for example \(F_{ij}' = B_i' C_j'\).
The total metagenomics bias and scaling factors thus split into steps as \(B_i = B_i' B_i''\) and \(C_j = C_j' C_j''\).

\textbf{TODO:} Add Note about units. We could be using bp, ng, genome copies, or marker-gene copies. It shouldn't affect the conclusions, but affects the units and values of the multipliers. Though perhaps it will help to pick ng DNA or genome copies

What is \(C'_j\)?
\begin{align}
  C'_j = \frac{A'_{Tj}}{A_{Tj} \bar{B}'_j}.
\end{align}
Note that we have so far been defining the metagenomics efficiencies \(B\) as being (arbitrarily) the efficiency relative to the first taxon, which affects the scale of \(C'\).

\hypertarget{targeted-dna-measurement}{%
\subsection{Targeted DNA measurement}\label{targeted-dna-measurement}}

\textbf{TODO:} explain the big picture of what I'm doing here

Let \(B^{(T)}_r\) be the conversion factor associated with the targeted measurement of reference taxon \(r\) from the extracted DNA, which I assume is sample-independent but can vary by taxon.
NOTE: I'm using \(B\) for ``bias'' here but we need to keep in mind that I'm talking about the bias relative to the DNA and not the cellular abundance.
In other words, the result of the targeted DNA measurement is
\begin{align}
  \label{eq:T-dna}
  T_{rj} 
  &= A_{rj}' B^{(T)}_r \\
  &= A_{rj} B'_r C'_j B^{(T)}_r.
\end{align}
Substituting the second expression for \(T_{rj}\) in \eqref{eq:A-hat-targeted} gives an expression for the estimate of the abundance of taxon \(i\) from the targeted measurement,
\begin{align}
  \label{eq:A-hat-targeted-dna}
  \hat A_{ij} 
  &= \frac{M_{ij}}{M_{rj}} \cdot T_{rj} \\
  &= \frac{A_{ij} B_i C_j}{A_{rj} B_r C_j} \cdot A_{rj} B'_r C'_j B^{(T)}_r \\
  &= A_{ij} \cdot \frac{B_i B^{(T)}_r}{B''_r} \cdot C'_j,
\end{align}
where I have replaced \(B_r / B'_r\) by \(B''_r\) in the denominator.
Compared to the case of cellular measurement, the efficiency ratio for the reference taxon is just of the post-extraction steps, and we also have an additional term \(C_j'\) that could vary across samples.

Question: Is that the same as the mean DNA extraction efficiency varies across samples?

\hypertarget{dna-spike-ins-1}{%
\subsection{DNA spike-ins}\label{dna-spike-ins-1}}

HERE: Do same thing as above, and get result that

Suppose we add the spike-in taxon \(r\) to sample \(j\) in amount \(A'_{rj}\) with error
\(B^{(T)}_r\), such that we believe the spike-in's abundance to be \(T_{rj}\) in \eqref{eq:T-dna}.
As with the cellular spike-in, the abundance of taxon \(i\) can be estimated by \eqref{eq:A-hat-targeted} (for the targeted cellular method);
and as with the case of a targted DNA measurement above, the resulting estimate is given by \eqref{eq:A-hat-targeted-dna}.

\hypertarget{implications}{%
\subsection{Implications}\label{implications}}

These results show that, for both methods, the fold variation in the error across samples equals that in the scaling factor \(C_j'\).
Ideally, \(C_j'\) would be constant, corresponding to the situation where doubling the cell concentration in a sample (while leaving the taxonomic composition unchanged) results in double the DNA yield.
In this case, these DNA-based methods would work just as well with respect to bias not affecting fold-change estimates as their cell-based counterparts.

Perhaps even easier to see why this is the ``ideal'' situation (and perhaps also a common default assumption when thinking about microbiome experiments), is to think in terms of the factors \(F'_{ij} = B_i' C_j'\).
Intuitively, we would like the factor increase for a taxon to be the same across samples; this situation corresponds to \(C_j = c\) being constant across samples.
(The value of \(c\) depends on the experimental details but also how we've set \(B_j' = 1\).)

NOTE: Also need to consider whether it is true that the relative abundances are biased by a constant amount; and remember this is am important assumption I'm making as well, and we have some evidence that it can break down.
But the motivation for how I've done it is that it seems reasonable to worry about saturation in the extraction yields that is independent of taxa once we account for variation in extraction efficiency and genome size that is simply due to the elution step.
May need to give this motivation earlier.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-brooks2015thet}{}}%
Brooks, J Paul, David J Edwards, Michael D Harwich, Maria C Rivera, Jennifer M Fettweis, Myrna G Serrano, Robert A Reris, et al. 2015. {``{The truth about metagenomics: quantifying and counteracting bias in 16S rRNA studies}.''} \emph{BMC Microbiol.} 15 (1): 66. \url{https://doi.org/10.1186/s12866-015-0351-6}.

\leavevmode\vadjust pre{\hypertarget{ref-contijoch2019gutm}{}}%
Contijoch, Eduardo J, Graham J Britton, Chao Yang, Ilaria Mogno, Zhihua Li, Ruby Ng, Sean R Llewellyn, et al. 2019. {``{Gut microbiota density influences host physiology and is shaped by host and microbial factors}.''} \emph{Elife} 8 (January). \url{https://doi.org/10.7554/eLife.40553}.

\leavevmode\vadjust pre{\hypertarget{ref-fettweis2019thev}{}}%
Fettweis, Jennifer M., Myrna G. Serrano, Jamie Paul Brooks, David J. Edwards, Philippe H. Girerd, Hardik I. Parikh, Bernice Huang, et al. 2019. {``{The vaginal microbiome and preterm birth}.''} \emph{Nat. Med.} 25 (6): 1012--21. \url{https://doi.org/10.1038/s41591-019-0450-2}.

\leavevmode\vadjust pre{\hypertarget{ref-kevorkian2018esti}{}}%
Kevorkian, Richard, Jordan T Bird, Alexander Shumaker, and Karen G Lloyd. 2018. {``{Estimating Population Turnover Rates by Relative Quantification Methods Reveals Microbial Dynamics in Marine Sediment}.''} \emph{Appl. Environ. Microbiol.} 84 (1): e01443--17. \url{https://doi.org/10.1128/AEM.01443-17}.

\leavevmode\vadjust pre{\hypertarget{ref-leopold2020host}{}}%
Leopold, Devin R, and Posy E Busby. 2020. {``{Host Genotype and Colonist Arrival Order Jointly Govern Plant Microbiome Composition and Function}.''} \emph{Curr. Biol.} 30 (16): 3260--3266.e5. \url{https://doi.org/10.1016/j.cub.2020.06.011}.

\leavevmode\vadjust pre{\hypertarget{ref-lloyd2020evid}{}}%
Lloyd, Karen G., Jordan T. Bird, Joy Buongiorno, Emily Deas, Richard Kevorkian, Talor Noordhoek, Jacob Rosalsky, and Taylor Roy. 2020. {``{Evidence for a Growth Zone for Deep-Subsurface Microbial Clades in Near-Surface Anoxic Sediments}.''} \emph{Appl. Environ. Microbiol.} 86 (19): 1--15. \url{https://doi.org/10.1128/AEM.00877-20}.

\leavevmode\vadjust pre{\hypertarget{ref-martino2019anov}{}}%
Martino, Cameron, James T. Morton, Clarisse A. Marotz, Luke R. Thompson, Anupriya Tripathi, Rob Knight, and Karsten Zengler. 2019. {``{A Novel Sparse Compositional Technique Reveals Microbial Perturbations}.''} \emph{mSystems} 4 (1): e00016--19. \url{https://doi.org/10.1128/mSystems.00016-19}.

\leavevmode\vadjust pre{\hypertarget{ref-mclaren2019cons}{}}%
McLaren, Michael R, Amy D Willis, and Benjamin J Callahan. 2019. {``{Consistent and correctable bias in metagenomic sequencing experiments}.''} \emph{Elife} 8 (September): 46923. \url{https://doi.org/10.7554/eLife.46923}.

\leavevmode\vadjust pre{\hypertarget{ref-props2017abso}{}}%
Props, Ruben, Frederiek-Maarten Kerckhof, Peter Rubbens, Jo De Vrieze, Emma Hernandez Sanabria, Willem Waegeman, Pieter Monsieurs, Frederik Hammes, and Nico Boon. 2017. {``{Absolute quantification of microbial taxon abundances}.''} \emph{ISME J.} 11 (2): 584--87. \url{https://doi.org/10.1038/ismej.2016.117}.

\leavevmode\vadjust pre{\hypertarget{ref-quinn2020inte}{}}%
Quinn, Thomas P., and Ionas Erb. 2020. {``{Interpretable Log Contrasts for the Classification of Health Biomarkers: a New Approach to Balance Selection}.''} \emph{mSystems} 5 (2): 1--11. \url{https://doi.org/10.1128/mSystems.00230-19}.

\leavevmode\vadjust pre{\hypertarget{ref-quinn2019afie}{}}%
Quinn, Thomas P., Ionas Erb, Greg Gloor, Cedric Notredame, Mark F. Richardson, and Tamsyn M. Crowley. 2019. {``{A field guide for the compositional analysis of any-omics data}.''} \emph{Gigascience} 8 (9): 1--14. \url{https://doi.org/10.1093/gigascience/giz107}.

\leavevmode\vadjust pre{\hypertarget{ref-riverapinto2018bala}{}}%
Rivera-Pinto, J., J. J. Egozcue, V. Pawlowsky-Glahn, R. Paredes, M. Noguera-Julian, and M. L. Calle. 2018. {``{Balances: a New Perspective for Microbiome Analysis}.''} \emph{mSystems} 3 (4): e00053--18. \url{https://doi.org/10.1128/mSystems.00053-18}.

\leavevmode\vadjust pre{\hypertarget{ref-tettamantiboshier2020comp}{}}%
Tettamanti Boshier, Florencia A., Sujatha Srinivasan, Anthony Lopez, Noah G. Hoffman, Sean Proll, David N. Fredricks, and Joshua T. Schiffer. 2020. {``{Complementing 16S rRNA Gene Amplicon Sequencing with Total Bacterial Load To Infer Absolute Species Concentrations in the Vaginal Microbiome}.''} \emph{mSystems} 5 (2): 1--14. \url{https://doi.org/10.1128/mSystems.00777-19}.

\leavevmode\vadjust pre{\hypertarget{ref-vandeputte2017quan}{}}%
Vandeputte, Doris, Gunter Kathagen, Kevin D'hoe, Sara Vieira-Silva, Mireia Valles-Colomer, João Sabino, Jun Wang, et al. 2017. {``{Quantitative microbiome profiling links gut community variation to microbial load}.''} \emph{Nature} 551 (7681): 507. \url{https://doi.org/10.1038/nature24460}.

\leavevmode\vadjust pre{\hypertarget{ref-vieirasilva2019quan}{}}%
Vieira-Silva, Sara, João Sabino, Mireia Valles-Colomer, Gwen Falony, Gunter Kathagen, Clara Caenepeel, Isabelle Cleynen, Schalk van der Merwe, Séverine Vermeire, and Jeroen Raes. 2019. {``{Quantitative microbiome profiling disentangles inflammation- and bile duct obstruction-associated microbiota alterations across PSC/IBD diagnoses}.''} \emph{Nat. Microbiol.} 4 (11): 1826--31. \url{https://doi.org/10.1038/s41564-019-0483-9}.

\leavevmode\vadjust pre{\hypertarget{ref-washburne2017phyl}{}}%
Washburne, Alex D., Justin D. Silverman, Jonathan W. Leff, Dominic J. Bennett, John L. Darcy, Sayan Mukherjee, Noah Fierer, and Lawrence A. David. 2017. {``{Phylogenetic factorization of compositional data yields lineage-level associations in microbiome datasets}.''} \emph{PeerJ} 5 (February): e2969. \url{https://doi.org/10.7717/peerj.2969}.

\leavevmode\vadjust pre{\hypertarget{ref-wasserman2004allo}{}}%
Wasserman, Larry. 2004. \emph{{All of Statistics}}. Springer Texts in Statistics. New York, NY: Springer New York. \url{https://doi.org/10.1007/978-0-387-21736-9}.

\leavevmode\vadjust pre{\hypertarget{ref-yeh2018taxo}{}}%
Yeh, Yi-Chun, David M. Needham, Ella T. Sieradzki, and Jed A. Fuhrman. 2018. {``{Taxon Disappearance from Microbiome Analysis Reinforces the Value of Mock Communities as a Standard in Every Sequencing Run}.''} \emph{mSystems} 3 (3): e00023--18. \url{https://doi.org/10.1128/mSystems.00023-18}.

\end{CSLReferences}

\end{document}

### 2021-08-13 notebook/_posts/2021-08-12-bias-sensitivity-demo-in-callahan2017repl-dataset/bias-sensitivity-demo-in-callahan2017repl-dataset.Rmd
[draft] Bias sensitivity demo in callahan2017repl

---
title: "Bias sensitivity demo in callahan2017repl dataset"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - bias sensitivity
  - ref:callahan2017repl
date: 2021-08-12
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

```{r}
library(tidyverse)
library(speedyseq)
library(metacal)
```

Our starting point is the data from @callahan2017repl.
The following code chunk downloads the ASV abundance matrix, sample metadata, and taxonomy assignments for the vaginal samples.
It requires that tidyverse and phyloseq are loaded (phyloseq can be used instead of speedyseq).
Set `dl_path` to where you'd like to download the files.

```{r import_data, cold_folding = TRUE, include = FALSE}
# Download --------------------------------------------------------------------
dl_path <- here::here("notebook/_data", "callahan2017repl")
link <- "https://stacks.stanford.edu/file/druid:yb681vm1809/RepRefine_Scripts.tar.gz"
if (!dir.exists(dl_path)) {
  dir.create(dl_path)
  download.file(link, file.path(dl_path, basename(link)))
  system2("tar", args = c("-xvf", file.path(dl_path, basename(link)), "-C", dl_path))
}
load(file.path(dl_path, "RepRefine_Scripts", "input", "processed.rda"))
# This loads:
# df: the sample metadata
# st: the sequence table (in counts)
# ft: the sequence table normalized to frequencies / proportions
# tax: taxonomy assignment matrix
# 
# To confirm relationship between ft and st, run 
# all.equal(ft, st %>% apply(1, function(x) x / sum(x)) %>% t)
# 
# The taxonomy matrix has an extra column with a duplicate name of "Species"
# that is all NAs, so let's remove that
tax <- tax[, -8]
# Import into phyloseq --------------------------------------------------------
sam <- sample_data(df)
st <- otu_table(st, taxa_are_rows = FALSE)
tax <- tax_table(tax)
ps <- phyloseq(sam, st, tax)
# Check that all samples and taxa made it into the phyloseq object
stopifnot(all.equal(colnames(st), taxa_names(ps)))
stopifnot(all.equal(rownames(sam), sample_names(ps)))
# Add ASV sequences to the refseq() slot
seqs <- taxa_names(ps) %>% rlang::set_names() %>% Biostrings::DNAStringSet()
ps <- merge_phyloseq(ps, seqs)
# Rename the ASVs to simpler names (ASV1, ASV2, ...). The full sequences remain
# in `refseq(ps)`.
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
# Clean up objects
rm(tax, df, sam, st, ft, seqs)
```

Optionally, the following code chunk uses munging functions from janitor and the latest version of speedyseq to select the metadata of primary interest, standardize format of column names, and convert categorial variables to factors.
```{r}
ps1 <- ps %>% 
  select_sample_data(SubjectID, Cohort, Race, preterm, Outcome) %>%
  rename_with_sample_data(janitor::make_clean_names) %>%
  rename_with_tax_table(janitor::make_clean_names) %>%
  mutate_sample_data(
    across(c(subject_id, cohort, race, outcome), factor),
  )
```

```{r}
saveRDS(ps1, "callahan2017repl_phyloseq_clean.rds")
```

```{r}
ps1 %>% tax_table
ps1 %>% sample_data %>% glimpse
```

## Analysis of Callahan et al Figure 1

### Basic setup

We start starting from a phyloseq object of counts at the ASV level, which multiple samples per woman (`subject_id`).
The analysis of Callahan et al Figure 1 was done on genus-level average proportions within a particular woman and used one-sided Wilcoxin tests.
The following code chunks do this analysis for Gardnerella in the Stanford Cohort.

First, obtain a data frame for plotting and testing.
```{r}
x <- ps1 %>%
  filter_sample_data(cohort == "Stanford") %>%
  transform_sample_counts(~ . / sum(.)) %>%
  filter_tax_table(genus == "Gardnerella") %>%
  tax_glom("genus") %>%
  merge_samples2("subject_id", fun_otu = mean) %>%
  as_tibble
```

This code chunk uses functions from the latest speedyseq version to simplify this procedure but should make fairly transparent what is going on:
Subset to the chosen cohort, convert counts to proportions, subset to just Gardnerella and merge the Gardnerella ASVs, then average the Gardnerella proportions across samples within a subject, then produce a long-format data frame.

```{r}
# exact = FALSE will be used regardless b/c of ties
wilcox.test(.abundance ~ outcome, data = x, alternative = "greater", exact = FALSE)
```

```{r}
x %>%
  ggplot(aes(outcome, .abundance + 1e-7)) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom() +
  scale_y_log10()
```

```{r}
# regression on ranks should be similar
f <- function(x) rank(x) / length(x)
a <- lm(f(.abundance) ~ preterm, data = x)
a %>% generics::tidy(conf.int = TRUE) %>% filter(term == "pretermTRUE")
```

Encapsulate this test in a function using `broom::tidy()` to get the results in a nicer form for analysis,

```{r}
test <- function(ps, genus, cohort = "Stanford", alternative = "two.sided") {
  x <- ps %>%
    filter_sample_data(cohort == !!cohort) %>%
    transform_sample_counts(~ . / sum(.)) %>%
    # subset_taxa(genus == "Gardnerella") %>%
    filter_tax_table(genus == !!genus) %>%
    tax_glom("genus") %>%
    merge_samples2("subject_id", fun_otu = mean) %>%
    as_tibble
  wilcox.test(.abundance ~ outcome, data = x, 
    alternative = alternative, exact = FALSE) %>%
    broom::tidy()
}
test(ps1, "Gardnerella", alternative = "greater")
test(ps1, "Lactobacillus", alternative = "less")
```

### Effect of bias against Gard

Let's filter to the stanford samples and ASVs first to speed up the computations.

```{r}
ps2 <- ps1 %>% filter_sample_data(cohort == "Stanford") %>%
  filter_taxa2(~ sum(.) > 0)
test(ps2, "Gardnerella", alternative = "greater")
```

See what happens as we crank the bias up against Gard, relative to the protocol of the present study.
How biased against Gard before the result disappears?

Note: First I did the opposite accidentally, and considered what if we were biased against Gard only; that actually increased the p values, though not by much

```{r}
gard_idx <- which(tax_table(ps2) == "Gardnerella", arr.ind = TRUE)[,"row"]
gard_avs <- ps2 %>% filter_tax_table(genus == "Gardnerella") %>% taxa_names
is_gard <- seq(ntaxa(ps2)) %in% gard_idx

#> eff_draws <- tibble(gard_efficiency = 1 / 2^seq(from = 0, to = 13)) %>%
eff_draws <- tibble(gard_efficiency = 1 / 5^seq(from = 0, to = 3)) %>%
  mutate(
    .id = row_number(), 
    efficiency_vector = map(gard_efficiency, ~.x^is_gard)
  ) %>%
  print
#> eff_draws %>% pull(efficiency_vector) %>% .[[4]] %>% head(10)
```

```{r}
res <- eff_draws %>%
  mutate(
    ps = map(efficiency_vector, ~ perturb(ps2, .x)),
    test = map(ps, test, genus = "Gardnerella", alternative = "greater")
  ) %>%
  select(.id, gard_efficiency, test) %>%
  unnest(test) %>%
  print
```

How could there be no effect? Does this make sense? 
Something to keep in mind is that I'm not resampling the counts, so the fact that the frequency of Gard may be getting dropped to undetectable levels does not come into play.

Question: What is the effect of bias on ranks?

With this particular choice of bias, where all taxa are the same against Gard, the shift in the logit of Gard will be the same in all samples, which means the rank ordering of the logits and hence of the proportions of Gard will be preserved.
Thus there will be no impact on the Wilcoxon score.

However, tests against other genera would be affected.

```{r}
res.lacto <- eff_draws %>%
  mutate(
    ps = map(efficiency_vector, ~ perturb(ps2, .x)),
    test = map(ps, test, genus = "Lactobacillus", alternative = "less")
  ) %>%
  select(.id, gard_efficiency, test) %>%
  unnest(test) %>%
  print
```

### 2021-08-23 notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example.Rmd
Simulate regression example

---
title: "Simulate regression example"
description: |
  A short description of the post.
author:
  - name: Michael R. McLaren
date: 08-03-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r}
# library(here)
library(tidyverse)

library(cowplot)
library(patchwork)
theme_set(theme_cowplot(12))

library(ggdist)
library(ggbeeswarm)

library(speedyseq)
library(metacal)

set.seed(42)
```

# Simulate actual abundances

```{r}
n_species <- 10
n_samples <- 50

species <- str_c("sp", seq(n_species))
```

## Actual abundances

Log abundances are independent and follow a simple linear model,

```
log density = a_0 + a_1 * X + epsilon
```

```{r}
n_samples <- 100
x <- c(
  rep(0, n_samples / 2),
  rep(1, n_samples / 2)
)
X <- rbind(1, x)

Sigma <- diag(1, nrow = n_species)
rownames(Sigma) <- colnames(Sigma) <- species
a_0 <- MASS::mvrnorm(
  mu = rep(0, n_species), 
  Sigma = diag(1, nrow = n_species)
)
a_1 <- MASS::mvrnorm(
  mu = rep(0, n_species), 
  Sigma = diag(1, nrow = n_species)
)
a <- cbind(a_0, a_1)
names(a_0) <- names(a_1) <- rownames(a) <- species

epsilon <- MASS::mvrnorm(
  n = n_samples,
  mu = rep(0, n_species), 
  Sigma = diag(0.5, nrow = n_species)
) %>% 
  t

y <- a %*% X + epsilon
colnames(y) <- str_glue("sa{1:n_samples}")
```

```{r}
sam <- tibble(.sample = colnames(y), x) %>%
  mutate(across(x, factor))
actual <- phyloseq(
  otu_table(y, taxa_are_rows = TRUE) %>% transform_sample_counts(exp),
  sample_data(sam)
)
```

```{r}
actual %>%
  as_tibble %>%
  ggplot(aes(y = .otu, x = .abundance, fill = x)) +
  scale_x_log10() +
  stat_dots()
```

## Biased abundances

```{r}
log_efficiency <- MASS::mvrnorm(
  mu = rep(0, n_species),
  Sigma = diag(3, nrow = n_species)
) %>%
  {. - mean(.)}
efficiency <- exp(log_efficiency)
# names(efficiency) <- species
```

```{r}
log_efficiency %>% summary
max(efficiency) / min(efficiency)
efficiency %>% qplot + scale_x_log10()
```

To create an association of log mean efficiency and the covariate, I will set set it so that the species with the largest slope coefficient also has the largest efficiency.
This scenario is inspired by the leopold2020host and brooks2015thet experiments.

```{r}
idx_slope <- which.max(a_1)
idx_eff <- which.max(efficiency)
names(efficiency) <- species
names(efficiency)[idx_eff] <- species[idx_slope]
names(efficiency)[idx_slope] <- species[idx_eff]
```

The observed (i.e. estimated) proportions and densities are given by perturbing the actual densties by the efficiencies, and normalizing to proportions or to the original (correct) total.

```{r}
observed_prop <- actual %>% perturb(efficiency)
observed_dens <- actual %>% perturb(efficiency, norm = "keep")
```

## Mean efficiencies

```{r mean-efficiency}
mean_efficiency <- actual %>%
  transform_sample_counts(close_elts) %>%
  perturb(efficiency, norm = "none") %>%
  sample_sums %>%
  enframe(".sample", "mean_efficiency") %>%
  left_join(sam, by = ".sample")
```

```{r}
mean_efficiency %>%
  ggplot(aes(y = as.factor(x), x = mean_efficiency)) +
  scale_x_log10() +
  stat_dotsinterval()
```

## Regressions and summary plots

what is a nice pipeline for doing all the regressions?

```{r}
analysis_ps <- function(ps, ...) {
  mat <- otu_table(ps) %>% orient_taxa(as = "columns") %>% log2
  lm(mat ~ 1 + x, data = sample_data(ps) %>% as_tibble) %>%
    broom::tidy()
}
analysis_sme <- function(x, ...) {
  lm(log2(mean_efficiency) ~ 1 + x, data = mean_efficiency) %>%
    broom::tidy()
}

true_coeffs <- list('(Intercept)' = a_0, x1 = a_1) %>%
  map_dfr(enframe, "response", "truth", .id = "term")

df <- bind_rows(
  analysis_ps(actual) %>% add_column(type = "Actual"),
  analysis_ps(observed_dens) %>% add_column(type = "Observed"),
  analysis_sme(mean_efficiency) %>% 
    add_column(type = "Mean efficiency", response = "Mean efficiency"),
) %>%
  mutate(
    across(type, factor, levels = c("Actual", "Observed", "Mean efficiency")),
    lower = estimate - 2 * std.error,
    upper = estimate + 2 * std.error,
  ) %>%
  # Add the true coefficients
  left_join(true_coeffs, by = c("response", "term"))
```

```{r}
p1 <- df %>% 
  filter(term == "x1") %>%
  mutate(
    across(response, fct_reorder, estimate)
  ) %>%
  ggplot(aes(y = response, x = estimate, color = type)) +
  geom_vline(xintercept = 0, color = "grey") +
  geom_pointinterval(aes(xmin = lower, xmax = upper)) +
  geom_point(data = ~filter(., type == "Actual"),
    aes(x = truth),
    color = 'black', shape = '+', size = 4)
p2 <- df %>% 
  filter(term == "x1") %>%
  mutate(
    across(response, fct_reorder, estimate)
  ) %>%
  ggplot(aes(y = type, x = estimate, fill = type)) +
  geom_vline(xintercept = 0, color = "grey") +
  stat_dots()
p1 / p2 + 
  plot_layout(heights = c(1, 0.3)) &
  colorblindr::scale_color_OkabeIto() &
  colorblindr::scale_fill_OkabeIto()
  # scale_color_brewer(type = 'qual', palette = 7) &
  # scale_fill_brewer(type = 'qual', palette = 7)
```

```{r}
ggsave('/tmp/regression-example.pdf', width = 6, height = 5, units = 'in')
ggsave('/tmp/regression-example.png', width = 6, height = 5, units = 'in')
```

here

- understand why the 'actual' estimates are generally more extreme than the true lfc
<!--  -->

## Data plots of specific examples

Plots like the first ones I made for the Leopold data.

```{r}
species_to_plot <- c('sp9', 'sp1', 'sp5', 'sp7', 'sp8')
```

```{r}
```

### attempt 1

```{r}
mean_efficiency1 <- mean_efficiency %>%
  rename(.abundance = mean_efficiency) %>%
  add_column(.otu = "Mean efficiency")
data_long <- 
  bind_rows(
    Actual = actual %>% as_tibble,
    Observed = observed_dens %>% as_tibble,
    'Mean efficiency' = mean_efficiency1,
    .id = 'type'
  )
```

```{r}
species_to_plot <- c('Mean efficiency', 'sp9', 'sp1', 'sp5', 'sp7', 'sp8')

data_long %>%
  filter(.otu %in% species_to_plot) %>%
  mutate(
    across(.otu, factor, levels = species_to_plot)
  ) %>% 
  ggplot(aes(x, log2(.abundance), color = type)) +
  facet_wrap(~.otu, scales = "free_y", ncol = 1) +
  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
  stat_smooth(aes(x = as.integer(x)),
    method = "lm", geom = "line", size = 0.9
  )
```

### attempt 2

```{r}
data_long <- 
  bind_rows(
    Actual = actual %>% as_tibble,
    Observed = observed_dens %>% as_tibble,
    .id = 'type'
  )
```

```{r}
species_to_plot <- c('sp9', 'sp1', 'sp5', 'sp7', 'sp8')

p3 <- data_long %>%
  filter(.otu %in% species_to_plot) %>%
  mutate(
    across(.otu, factor, levels = species_to_plot)
  ) %>% 
  ggplot(aes(x, log2(.abundance), color = type)) +
  facet_grid(.otu~type, scales = "free_y") +
  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
  stat_smooth(aes(x = as.integer(x)),
    method = "lm", geom = "line", size = 0.9
  ) +
  theme(legend.position = "none")

p4 <- mean_efficiency %>%
  ggplot(aes(x, log2(mean_efficiency))) +
  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
  stat_smooth(aes(x = as.integer(x)),
    method = "lm", geom = "line", size = 0.9
  )
```

```{r}
p1 + p3 + p2 + p4 +
  plot_layout(ncol = 2, heights = c(1, 0.3)) &
  colorblindr::scale_color_OkabeIto() &
  colorblindr::scale_fill_OkabeIto()

```

### Testing plot types

The difference plots in 'dabestr' seem like they might be ideal; specifically the Gardner-Altman Two Group Estimation Plot

```{r}
# library(dabestr)
```

But perhaps it will be better to manually create this plot using ggdist.
Could compute the bootstrapped difference estimates, then plot them with stat_halfeye in vertical orientation at x=3.

HERE.

https://mjskay.github.io/ggdist/

```{r}
```

```{r}
data_long %>%
  filter(.otu == 'sp5', type == 'Actual') %>%
  ggplot(aes(x, log2(.abundance))) +
  geom_quasirandom(alpha = 0.3, groupOnX = TRUE)

```

### 2021-08-31 abundance-measurement.Rmd
Revise section on abundance measurement
Revise according to outline in outline.org

--- 2021-07-19 abundance-measurement.Rmd
@@
-# How taxonomic bias affects abundance measurements {#abundance}
+# How taxonomic bias affects abundance measurements {#abundance-measurement}
 
-<!-- TODO: Add an illustration of the bias model in panel 1B -->
+To understand how bias affects the measured differential abundance between samples, we first consider how it affects measurements of individual samples.
 
-```{r fig-abundance, out.width = '65%', fig.cap = '(ref:cap-abundance)', cache = TRUE}
-svg_path <- here("figures/absolute-and-relative-abundance.svg")
+Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
+This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr) ^[todo: details about this assumption].
+In this model, we suppose there are a set of *atomic taxa*, that differ in how efficiently they are measured but within which all cells behave similarly (r/v).
+For concreteness, we equate atomic taxa with species. 
+*(Note that bias may not be conserved w/in species; ref discussion of synthetic taxa later on)*.
+We ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.
+
+Our model stipulates that the assigned read count of species $i$ in sample $a$ equal equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
+\begin{align}
+  (\#eq:measurement-model)
+  \text{reads}_{i}(a)
+  = \text{density}_{i}(a) \quad \cdot
+    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
+    \cdot \quad
+    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
+\end{align}
+The species-specific factor is the *relative measurement efficiency* (or *efficiency* for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (@mclaren2019cons).
+The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.^[todo: add note about how, as defined, sequencing effort typically goes down with total density because of library normalization; also mention that it 'sequencing effort' is sometimes equated with total read count or library size, which is not what we've done here].
+Equation \@ref(eq:measurement-model) implies that the total reads from all species in the sample equal
+\begin{align}
+  (\#eq:total-reads)
+  \text{total reads}(a)
+    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
+\end{align}
+where 
+\begin{align}
+  (\#eq:mean-efficiency)
+  \text{mean efficiency}(a) 
+  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
+\end{align}
+is the average or mean efficiency of cells in the sample.
+
+## Relative abundances (proportions and ratios)
+
+The term *relative abundance* carries multiple meanings in microbiome science.
+Microbiome researchers often use the relative abundance of a species in a sample to refer to its *proportion*, or number of cells from that species divided by the total number of cells from the taxonomic domain of interest (e.g.
+Prokaryotes).
+A second view, drawn from the statistical framework of Compositional Data Analysis, equates the relative abundances of species within a sample with the *ratios* of cell densities among two or more species (e.g.
+a ratio of 5:4:2 among three species).
+By this view, it is meaningless to refer to the relative abundance of an individual species.
+<!-- (Although a proportion is also technically a ratio [of a species to a higher-level taxon], here we use 'ratio' to refer specifically to ratios among species.) -->
+Although other interpretations exist ^[e.g.
+the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most DA analysis methods (and microbiome analysis in general), for both relative and absolute abundances, and the differences between them help to clarify the issue over when and why the effects of bias cancel in an analysis.
+
+The standard estimate of the proportion of species $i$ in sample $a$ given by the ratio of its read count to the total,
+\begin{align*}
+  \hat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
+\end{align*}
+The effect of taxonomic bias on this estimate is to create a multipliative error equal to the ratio of the species' efficiency to the mean efficiency in the sample,
+\begin{align}
+  (\#eq:prop-error)
+  \hat{\text{prop}}_{i}(a)
+  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)}.
+\end{align}
+Intuitively, Equation \@ref(eq:prop-error) says that a species is measured as having a larger-than-actual proportion in direct relation to how much more more efficiently its cells are measured than the average cell in the sample (Figure \@ref(fig:error-proportions)).
+Consequently, the same species can be over or under measured in terms of its proportion.
+For instance, in samples from two hypothetical communities in Figure \@ref(fig:error-proportions), Species 3 has an efficiency of 6 and is under-estimated in Sample 1 (which has a mean efficiency of 8.33) but over-estimated in Sample 2 (which has a mean efficiency of 3.15).
+
+The estimated ratio between species $i$ and $j$, given by the ratio of read counts, 
+\begin{align*}
+  \hat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)},
+\end{align*}
+has a multiplicative error equal to the ratio in the species' efficiencies,
+\begin{align}
+  (\#eq:ratio-error)
+  \hat{\text{ratio}}_{i/j}(a)
+%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
+  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
+\end{align}
+Equation \@ref(eq:ratio-error) says that the ratio between two species is over-estimated in direct relation to how much more efficiently cells of the first species is measured relative to the second.
+In contrast to estimates of proportions, estimates of species ratios have a consistent multiplicative error across samples.
+For instance, in the example of Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
+
+- note: the error is not necessarily consistent for ratios among higher-order taxa
+- note: The proportion equation is analogous to Equation \@ref(eq:ratio-error) (ratio case), with the denominator taxon is the entire taxonomic domain and the relevant efficiency is the weighted average of all species.
+In fact, both equations can be derived from a more general equation, where the observed ratio of two (potentially non-atomic) taxa equals the true ratio times the ratios of the mean efficiencies within each taxon.
+- note: the results for ratios of higher-order taxa, and of proportions, can both be seen as stemming from a general result where the error in the ratio is the ratio of the mean efficiency within that taxon.
+- perhaps have a box for this
+
+<!-- begin figure -->
+
+```{r error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
+svg_path <- here("figures/illustrations/error-proportions.svg")
 if (knitr::is_html_output()) {
@@
 
-(ref:cap-abundance) Panel A: Metagenomics output is not directly proportional to microbial cell density and so is typically interpreted as _relative abundances_, either as abundances relative to each other (ratio view) or to the total (proportion view).
-For illustration purposes, the effect of taxonomic bias on the metagenomics output is not shown.
-Panel B: The effect of taxonomic bias is to distort the relative abundances that are measured by fixed ratios.
-
-_Thoughts for figure 1: ideas for caption: topic sentence can be something about how metagenomics measures biased relative abudances; indicate that A is showing unbiased MGS output; OR: Can I remake this figure such that it shows biased metagenomics output?
-Consider making label about the model, rather than abundance_
-
-We can understand how taxonomic bias affect relative and absolute abundance measurement using a simplified model of microbiome measurement.
-(Details are given in Appendix Model REF)
-First, consider the MGS sequencing measurement. 
-We start from a sample with a given combination of species in differing absolute abundances, which throughout we take to be in units of density of cells per unit mass.
-(For simplicity we ignore issues relating to the sequencing of live versus dead cells and non-cellular DNA.)
-Sequencing results in a total number of taxonomically-classified reads that, while not arbitrary, do not straightforwardly or consistently reflect the differences in total microbial density among samples.
-For this reason, these counts are typically only interpreted in relative terms  (Figure 1A)—either relative to each other (i.e., as ratios among taxa) or relative to the total (i.e., as proportions).
-Protocols can be implemented to preserve more information about total abundances (See Cruz et al, and Appendix); however, in the main text we focus on the more typical situation where the total count only tells us about the precision of the measured relative abundances and we must make additional measurements or assumptions to recover information about absolute microbial densities.
-
-The effect of taxonomic bias on the relative abundances in a sample is to perturb them by taxon-specific multiplicative factors—the taxa's _measurement efficiencies_—which quantify how efficiently each taxon is converted from cells into taxonomically classified sequencing reads (Figure 1B).
-A taxon's efficiency aggregates the bias acrewed during the entire measurement process and can be thought of as a product of efficiencies at various steps in the experimental workflow and other taxon- and protocol-specific factors.
-For example, _(Insert Gard and Lacto example, breaking down by extraction, PCR, and copy-number)_.
-Only the relative efficiencies affect the measured relative abundances; doubling all efficiencies would not change the mesaured ratios or proportions.
-Therefore we define the efficiencies as relative to a chosen reference taxon (or set of taxa); here we'll take to be that of the least-efficiently measured taxon being considered.
-
-- TODO: Somewhere Give 1-sentence summary of the model - absolute information is lost, and measured relative abundances are perturbed by taxon- and protocol-specific factors.
-(In the figure caption and/or start of this section and or right here as a summary/resolution)
-- Think about How what I write below depends on whether I'm taking this equation to in effect be a statement of the model, or the model is what I illustrated.
-- Should define what I mean by Taxon, and address that we're making a big assumption that efficiency of a taxon is fixed (including doesn't vary from lower-level variation)
-- A footnote or "Details" box can discuss what happens when there is genotypic or non-genotypic variation in efficiency within a taxon.
-
-## Relative-abundance measurements
-
-By this model, the ratios in the read counts of taxa, compared to the ratios among their cell densities, are multiplied by taxon- and protocol-specific factors (Figure, MWC).
-\begin{align}
-  (\#eq:ratio-err)
-  \frac{\text{reads}_{i}(s)}{\text{reads}_{j}(s)}
-  &= \frac{\text{density}_{i}(s)}{\text{density}_{j}(s)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
-\end{align}
-But what about the estimated proportions, estimated by dividing the taxon read counts by the total?
-The estimated proportion of a taxon $i$ in a sample $s$ under this model equals (@mclaren2019cons, Box)
-\begin{align}
-  (\#eq:prop-err)
-  \hat{\text{prop}}_{i}(s) 
-  &= \text{prop}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
-\end{align}
-where $\text{mean efficiency}(s)\equiv \sum_{j}\text{prop}_j(s)\cdot \text{efficiency}_j$ is the average efficiency of cells in the sample.
-The mean efficiency in the denominator of \@ref(eq:prop-err) was omitted in models of bias prior to that of @mclaren2019cons, but naturally arises from the requirement that the estimated proportions sum to 1 and captures the competition-to-be-measured that this sum-to-1 constraint implies ([Details](#details:relative)).
-Intuitively, Equation \@ref(eq:error) says that the estimated proportion of taxon $i$ will be too large when its measurement efficiency is greater than the average efficiency of cells in the sample.
-
-<details id="details:relative">
-<summary style="font-weight:bold;font-size:120%">Details: Taxonomic bias and relative-abundance estimates</summary>
-::: {.rmdnote}
-<!-- ### Details: Taxonomic bias and relative-abundance estimates {-} -->
-
-**Proportions:** The measured proportion of a taxon $i$ in a sample $s$, defined as the ratio of the taxon's density to the total density of all taxa in the sample, is
-\begin{align}
-  (\#eq:prop-def)
-  \text{prop}_{i}(s) \equiv \frac{\text{density}_{i}(s)}{\text{total density}(s)}.
-\end{align}
-The standard estimate for the proportion of taxon $i$ is the proportion of its reads,
-\begin{align}
-  (\#eq:prop-est)
-  \hat{\text{prop}}_{i}(s) \equiv \frac{\text{reads}_{i}(s)}{\text{total reads}(s)}.
-\end{align}
-These definitions imply that both the actual and the estimated proportions each sum to 1.
-Our model stipulates that within a given sample $s$, the estimated and actual proportions are related by 
-\begin{align}
-  \hat{\text{prop}}_{i}(s) = c(s) \cdot \text{prop}_{i}(s) \cdot \text{efficiency}_{i}
-\end{align}
-for some taxon-independent coefficient $c(s)$.
-The requirement that the estimated proportions sum to 1 implies that 
-\begin{align}
-  c(s) = \sum_j \text{prop}_{j}(s) \cdot \text{efficiency}_j = \text{mean efficiency}(s).
-\end{align}
-
-**Ratios:** The actual ratio of taxon $i$ to taxon $j$,
-\begin{align}
-  (\#eq:ratio-def)
-  \text{density}_{i/j}(s) \equiv \frac{\text{density}_{i}(s)}{\text{density}_{j}(s)}
-\end{align}
-is estimated by the ratio of reads assigned to each taxon,
-\begin{align}
-  (\#eq:ratio-est)
-  \hat{\text{density}}_{i/j}(s) \equiv \frac{\text{reads}_{i}(s)}{\text{reads}_{j}(s)} = \frac{\hat{\text{prop}}_{i}(s)}{\hat{\text{prop}}_{j}(s)}.
-\end{align}
-The error under multiplicative bias is
-\begin{align}
-  (\#eq:ratio-err)
-  \hat{\text{density}}_{i/j}(s) = \frac{\text{density}_{i}(s)}{\text{density}_{j}(s)} \cdot
-  \frac{\text{efficiency}_{i}(s)}{\text{efficiency}_{j}(s)}.
-\end{align}
-
-:::
-</details>
-
-<!-- begin figure -->
-
-```{r fig-error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
-svg_path <- here("figures/error-proportions.svg")
-if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
-} else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
-}
-rm(svg_path)
-```
-
-(ref:cap-error-proportions) Top row: Error in proportions measured by MGS in two communities that contain different relative abundances of three taxa.
-Bottom row: Error in the estimated fold-change in the third taxon that is derived from these measurements.
+(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species.
+Bottom row: Error in the estimated fold-change in the third species that is derived from these measurements.
 
@@
 
-This competitive aspect of proportional measurement creates a notable feature of the errors: They depend on the composition of the sample.
-The same taxon, subject to the same measurement efficiency, will have a different error in its estimated proportion in different samples depending on the taxonomic composition of the sample (Figure 2A+B).
-A taxon will be overestimated in samples where its efficiency is greater than that of the sample average, and underestimated in samples where its efficiency is less than the sample average (e.g.
-Taxon 3 in 2A vs 2B).
-This difference in the consistancy of ratios versus proportions propagates to ratio- and proportion-based absolute abundance estimates and, as we describe below, is a key tool for understanding how taxonomic bias affects different DA analyses.
+## Absolute abundances (cell densities)
 
-<!-- discussion of competition; tentative positioning -->
+Various experimental and computational methods have or can be used to supplement MGS measurements so as to convert the relative-abundance information in the read counts into estimates of (absolute) cell densities;
+however, these methods can largely be partitioned conceptually into those that use the estimated proportions or the estimated ratios provided by MGS.
+The error in density estimates due to taxonomic bias therefore follows that of the proportion and ratio estimates described above.
 
-_The below discussion makes more sense if the goal as density/AA inference, and not proportion inference. Perhaps should split out some if it into the AA section._
-
-Why does the sequencing outcome associated with one taxon depend on the rest of the sample?
-We can think of proportional measurement as a competition among all the cells in the sample for a limited pool of 100 percentage points, for which
-the measurement efficiencies determine each cell's ability to compete.
-It is therefore a cell's efficiency relative to its average competitor (all the cells in the sample) that determines how many percentage points it receives.
-This account highlights that it is the competition is the problem---if we were able to measure taxa/cells in a taxonomically biased but non-competitive manner, then we could get consistent errors.
-We cannot remove competition simply by using reads counts instead of proportions, since competition is intrinsic to sequencing experiments, due to limits in the yield during DNA extraction, PCR, and sequencing that occur in sequecing experiments.
-
-_Perhaps say something about ratios - since cells or taxa in the sample are subject to the same average competitor, the effect of the average competitor cancels out._
-
-Our ability to design-out these limits is limited and rarely attempted (ref Cruz for counter example).
-But we might overcome competition with other types of measurements, particularly those that target individual taxa.
-For example, ddPCR can be applied directly to an unextracted sample to count individual cells of a given taxon, before any competition associated with the sequencing protocol has occurred.
-Targeted qPCR quantifies a marker (typically the 16S gene) in the extracted DNA, and so is only affected by competition during the DNA extraction step.
-These measurements are still subjected to taxonomic bias, but the lack or reduction in competition means that the error due to taxonomic bias is likely to be relatively consistent.
-For this reason, these measurements can serve as useful control measurements for correctly for the variation in the sample mean efficiency in the sequencing measurement.
-
-## Absolute-abundance measurements
-
-- P: Motivation for AA
-However, these relative-abundance measures leave ambiguous what is happening to the absolute cell densities of taxa.
-- e.g. of proportion decrease coming from different explanations
-- e.g. of ratio varying having different explanations.
-
-AA methods have generally been motivated, developed, and applied ignoring taxonomic bias and assuming the estimated proportions are correct.
-
-The estimation procedures used by AA methods can be split into those that use the proportions estimated from the sequencing reads, and those that use ratios.
-
-Proportion-based estimation involves the intuitive procedure of multiplying the estimated proportions by an independent estimate of the total cell density in the sample.
-(The total-density estimates can be obtained by a wide variety of means, reviewed in APPENDIX.)
-By this approach, the estimated density of a taxon $i$ in sample $s$ is simply
+Proportion-based density estimation is based on the idea that if we know the (true) proportion of a species and the (true) total density of all species, we can multiply these numbers obtain the species' density.
+This method therefore involves multiplying the proportion of a species estimated from MGS with an estimate of total microbial density obtain through other means (e.g., through direct cell counting or qPCR of a universal marker gene) to estimate
+its density,
 \begin{align}
   (\#eq:density-prop-est)
-  \hat{\text{density}}_{i}(s) 
-  = \hat{\text{prop}}_{i}(s) \cdot \hat{\text{total density}}(s).
+  \hat{\text{density}}_{i}(a) 
+  = \hat{\text{prop}}_{i}(a) \cdot \hat{\text{total density}}(a).
 \end{align}
-This estimate equals the true density if the estimated proportions and total density are both accurate.
-Accounting for the error in the estimated proportions due to taxonomic bias (by substituting in EQ-REF) and potential error in the estimated total density gives
+The error in this proportion-based density estimate is
 \begin{align}
-  (\#eq:density-prop-err)
-  \hat{\text{density}}_{i}(s) 
-  &= \text{density}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)} \cdot \text{error in }\hat{\text{total density}}(s).
+  (\#eq:density-prop-error)
+  \hat{\text{density}}_{i}(a) 
+  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
+  \cdot \substack{\text{fold error in} \\ \hat{\text{total density}}(a)}.
 \end{align}
-Densities estimated from proportions are subject to the same bias-induced error as those proportions, which depends on sample composition through the sample's mean efficiency.
+The multiplicative error consists of two factors: the error in the estimated proportion (due to taxonomic bias in the MGS measurement), and any error in the estimated total density.
+In the case of perfect knowledge of the total density, the fold error in proportion-based density estimates varies across samples inversely with the mean efficiency.
 
-Error in the total density estimate is also typically ignored but may be considerable (Ref future box) and can worsen or improve the error in DA analysis; we return to its implications in the context of considering how these errors impact DA analysis.
+- note: right now I'm not consistently ref'ing the equations where the results come from
 
-Ratio-based density estimation embodies the fact that, without knowing the total abundance, we can regain absolute abundance information if we can determine the abundance of one or more _reference taxa_.
-Given an estimate of the density of reference taxon $r$, we can estimate the density of taxon $i$ by multipling the ratio of $i$ to $r$ in the reads by the density of $r$,
+Ratio-based density estimation follows from the fact that, in the absence of taxonomic bias, the conversion rate from reads to density is just the sequencing effort of the sample and is thus the same for all species (Equation \@ref(eq:measurement-model)).
+Knowing the density of at least one _reference_ species therefore enables us to convert from reads to density for all species.
+One approach to obtaining such reference species is to add one or more extraneous species in known (typically constant) abundance (a "spike-in") to each sample prior to sequencing (REFs).
+A second approach is to determine a set of naturally occurring species that are thought to have a constant, though typically unknown, abundance across samples; in this case we can only estimate species' densities relative to these "housekeeping species" (REFs).
+A third (seemingly untried) possibility is to measure the density of one or more naturally occurring species using a method such as targeted CFU counting, fluorescence flow cytometry or ddPCR directly on cells, or ddPCR or qPCR on extracted DNA.
+Given the estimate of the density of reference taxon $r$, we estimate the density of taxon $i$ by multipling the ratio of $i$ to $r$ in the reads by the density of $r$,
 \begin{align}
-  \hat{\text{density}}_{i}(s) 
-  &= \frac{\text{reads}_{i}(s)}{\text{reads}_{r}(s)} \cdot \hat{\text{density}}_{r}(s);
+  (\#eq:density-ratio-est)
+  \hat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \hat{\text{density}}_{r}(a)
 \end{align}
-So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
-It has also been used by methods that computationally determine a set of "housekeeping taxa" that they then assume have a constant abundance across samples.
-Another potential application is to use naturally occurring taxa that we have biological reason to believe are constant, as has been argued for host and organelle-derived reads in plant and animal microbiome studies.
-Moreover, EQREF makes clear there is no need for the reference taxa to be constant or known in advance, so long as we can measure its abundance: Hence the reference taxon can also be a naturally occurring taxon of varying abundance whose density we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.
+(Appendix ??? describes how this equation can be extended to multiple reference species.)
+For a constant reference species with unknown density, we can set $\hat{\text{density}}_{r}(a)$ equal to 1 to estimate density in units of the density of the reference species.
+The error in this ratio-based density estimate is
+\begin{align}
+  (\#eq:density-ratio-error)
+  \hat{\text{density}}_{i}(a) 
+  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
+  \cdot \substack{\text{fold error in} \\ \hat{\text{density}}_{r}(a)}.
+\end{align}
+Again, the multiplicative error consists of two factors: the error in the estimated ratio (due to taxonomic bias in the MGS measurement), and error in the estimated density of the reference taxa.
+So long as the fold error in the reference density is constant across samples, so will be the error in the estimated density of the focal species $i$.
 
-The error in this estimate, accounting for the error in the estimated ratio due to taxonomic bias (EQ-REF) and potential error in the estimated density of the reference taxon, is described by
-\begin{align}
-  \hat{\text{density}}_{i}(s) 
-  &= \text{density}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} \cdot \text{error in } \hat{\text{density}}_{r}(s)
-\end{align}
-Again, the properties of the corresponding relative-abundance estimate carries over, so that the error induced by taxonomic bias in the metagenomics measurement is independent of sample composition.
-The reference density estimates are subject to error, but we can reasonably expect this to be more consistent across samples (see box), leaving the overall error relatively constant.
-
-A warning note: The distinction between ratio and proportion-based estimation depends on whether an equation resembling X or Y is used for the final density estimates, and does not cleanly align with which empirical method is used to obtain absolute-abundance information.
-In particular, some studies that use spike-ins and most that use host or organelle reads to obtain absolute abundances do not follow this approach: Instead, these studies use the ratio of total reads of focal taxa to the reference reads as a proxy for total density to use in forming proportion-based estimates (EQREF).
-(For studies using host reads, the total density estimate may come from WGS sequencing and the estimated proportions from 16S sequencing.)
-Alhough there may be other advantages to this approach, the above observations suggest that it may be more susceptible to taxonomic bias in the sequencing measurement than using a ratio-based approach on the same underlying data.
+- note: these two approaches are not always distinct, and it seems that many/most authors are thinking of spike-ins and housekeeping taxa in terms of the proportion-based method; I need to think about how to address this point.

### 2021-09-01 differential-abundance.Rmd
Revise section on differential abundance

--- 2021-07-19 differential-abundance.Rmd
@@
-# How taxonomic bias affects differential-abundance analysis
+# How taxonomic bias affects differential-abundance analysis {#differential-abundance}
 
-The current practice of ignoring bias in DA and other analyses of the differences betwen samples is sometimes justified by the idea that the error in measurements of in individual samples should not affect the measured differences between samples, so long as they have been subjected to the same taxonomic bias.
-This argument is predicated on an assumption that consistent taxonomic bias creates consistent errors in the quanity of interest.
-The previous section showed that while this may be true for ratio-based abundance measurements, it is not true for proportion-based measurements.
-This section formally describes how the errors in the measurements of individual samples translate into errors (or lack thereof) in differential abundance analysis of (log) fold changes in proportions, ratios, and densities.
+How do the measurement errors described in the previous section impact our ability to estimate the changes in microbial abundances across samples or between different host and environmental conditions?
+Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, as such inference is ubiquitous in microbiome DA analysis and have more direct ecological interpretations (via the processes of exponential growth and death) than other DA measures.
+These measures of DA are also affected by taxonomic bias in our model in a manner that is straightforward to understand in terms of the multiplicative errors described in the previous section.
+
+- Maybe: Add a final sentence summary that the constant multiplicative error in ratio-based abundance estimates does not affect DA measurement, but the varying error induced by variation in the mean efficiency affects DA analysis on proportion-based estimates.
 
@@
 
-Before considering the sterotypical many-sample DA analysis, it is instructive to 
+Before considering the stereotypical many-sample DA analysis, it is instructive to 
 consider simplest analysis of differential abundance: the estimation of fold changes in abundance between a pair of samples.
-This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment.
-Just as important, it helps to bridge intuitions from the single-sample results of the previous section to the many-sample case.
+This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment, and conceptually bridges the single-sample results of the previous section to the many-sample case.
 
 <!-- P: error in FC proportions -->
-The composition-dependent effect of bias on fold error in proportions leads to error in fold-change estimates that is proportional to the inverse change in mean efficiency (REF Figure 2B).
-From the error in individual sample \@ref(eq:prop-err), it follows that the estimated fold change in proportion of taxon $i$ from sample $s$ to sample $t$ is
+The composition-dependent effect of bias on fold error in proportions leads to error in fold-change estimates that is proportional to the inverse change in mean efficiency.
+From the error in an individual sample \@ref(eq:prop-error), it follows that the estimated fold change in proportion of taxon $i$ from sample $a$ to sample $b$ is
 \begin{align}
-  (\#eq:prop-fc-err)
-\underbrace{\frac{\hat{\text{prop}}_{i}(t)}{\hat{\text{prop}}_{i}(s)}} _\text{estimate}
+  (\#eq:prop-fc-error)
+% \tag*{Fold change in proportion}
+\underbrace{\frac{\hat{\text{prop}}_{i}(b)}{\hat{\text{prop}}_{i}(a)}} _\text{estimated FC}
   &= \frac
-    {\text{prop}_{i}(t) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(t)}}
-    {\text{prop}_{i}(s) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(s)}}
+    {\text{prop}_{i}(b) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(b)}}
+    {\text{prop}_{i}(a) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(a)}}
 \\[0.5ex]
   &=
-  \underbrace{\frac{\text{prop}_{i}(t)}{\text{prop}_{i}(s)}}_\text{actual}
+  \underbrace{\frac{\text{prop}_{i}(b)}{\text{prop}_{i}(a)}}_\text{actual FC}
   \cdot
-  \underbrace{\left[\frac{\text{mean efficiency}(t)}{\text{mean efficiency}(s)}\right]^{-1}}_\text{error}
+  \underbrace{\left[\frac{\text{mean efficiency}(b)}{\text{mean efficiency}(a)}\right]^{-1}}_\text{fold error}
   .
@@
 The sample-independent efficiency factor of the error cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency from $s$ to $t$.
-In the example in Figure 2, the mean efficiency decreases by 2.64X from 8.33 to 3.15, creating an artificial fold increase of 2.64X in the estimated fold changes.
-This same error applies to all taxa, but its consequences differ depending on the taxon's actual fold change.
-For Taxon 1, which increases by 2.3X, we estimate larger increase of 6.5X; For Taxon 2, which decreases with a FC of 0.15X, we see a lesser decrease of 0.44X; For taxon 3, which decreases with a FC by 0.6X, we see an _increase_ of 1.75X.
-Thus for the same pair of samples we can see errors of increased magnitude, errors of decreased magnitude, and errors of changed direction (or sign) for different taxa depending on how the magnitide and direction of the true change compares to that of the mean efficiency.
-Finally, any taxon whose proportion is constant will appear to vary precisely as the inverse mean efficiency.
 
-TODO: Fig 2 Add numbers for the other taxa.
+The bottom row of Figure \@ref(fig:error-proportions) illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
+The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
+Consequently, the FC of each species is measured to be 2.6X larger than the true value.
+Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
+We can see each type of error in Figure 2.
+For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the estimated FC (actual FC: 2.3X, measured FC: 6.5X).
+For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
+For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!
 
-<!-- P: error in proportion-based AAs (revise) -->
-Taxonomic bias causes the same error in the estimated fold changes in proportion-based density estimates.
-A similar calculation from the error in individual density estimates \@ref(eq:density-prop-err) shows that the error in the fold change of a taxon $i$ is given by the analogous formula to \@ref(eq:prop-fc-err), with density subsituted for proprotion.
-The same types of errors (in magnitude and sign) can arise (Figure 3), now depending on how the magnitide and direction of the true change in density compares to that of the mean efficiency (Figure 3 TBD).
-Maybe text:
-- Alternative view of what is doing on, using the Lacto/Gard example, where a rarer but lower efficiency taxon grows.
-By first view, the mean efficiency decreases, causing an upward bias in the FC estimates.
-By the second view, the gains in total density are incorrectly attributed to the more efficiently measured taxon.
-- This figure and its discussion should relate these two views. Perhaps the first mention of AA estimates should as well.
-- Note: Whereas sign errors in proportion FCs require the involvement of at least three taxa (TODO: explain in footnote), sign errors in proportion-based density FCs can occur even with just two taxa.
+In contrast, because species ratios are distorted by a constant factor, their measured fold changes remain accurate.
+The fold error in Equation \@ref(eq:ratio-error) completely cancels when we divide the ratio measured for one sample $a$ by another sample $b$.
 
-NOTE: Having a figure and example specifically for density is not necessary for this draft.
+If other error sources remain negligible, than this dichotomy continues to apply to proportion- and ratio-based density estimates.
+The inferred FCs in proportion-based density estimates will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
+Moreover, any species with a constant density will appear to vary inversely with the mean efficiency.
+In contrast, fold changes in ratio-based density estimates remain accurate.
 
-<!-- P: error in FC ratios and ratio-based AAs -->
-(In contrast to proportion-based estimates) Becuase consistent taxonomic bias creates a consistent error in ratio-based estimates, the error completely cancels, leaving accurate fold changes.
-From \@ref(eq:ratio-err), the estimated fold change in the ratio of taxon $i$ to taxon $j$ from sample $s$ to sample $t$ is
+## Regression analysis of many samples
+
+DA analysis of many samples across host or environmental conditions can typically be framed as a regression problem, in which we analyze the relationship between a microbial _response variable_, such as log density of some focal species $i$, and one or more _covariates_, such the pH or temperature of the sampled environment or whether the sample is from a healthy or sick person.
+A substantial fraction of DA analyses use minor elaborations on the simple linear regression model, which for a response of log species density can be written
 \begin{align}
-  (\#eq:ratio-fc-err)
-  \underbrace{ \frac{{\text{reads}_{i}(t)}/{\text{reads}_{j}(t)}} {{\text{reads}_{i}(s)}/{\text{reads}_{j}(s)}} }_\text{estimate}
-  &= 
-  \frac{{\text{density}_{i}(t)}/{\text{density}_{j}(t)}} {{\text{density}_{i}(s)}/{\text{density}_{j}(s)}}
-  \cdot
-  \frac{\cancel{\text{efficiency}_{i}(t)}/\cancel{\text{efficiency}_{j}(t)}} {\cancel{\text{efficiency}_{i}(s)}/\cancel{\text{efficiency}_{j}(s)}}
-  \\[0.5ex]&= 
-  \underbrace{ \frac{{\text{density}_{i}(t)}/{\text{density}_{j}(t)}} {{\text{density}_{i}(s)}/{\text{density}_{j}(s)}} }_\text{actual}
-  .
+  (\#eq:regression)
+  \log \text{density}_i(a) = \alpha + \beta x(a) + \varepsilon_i(a).
 \end{align}
-(Describe FCs in Figure 2 and Figure 3 density e.g., seeing the correct FCs.)
+Here $x$ is a continuous covariate (e.g.
+pH) or a binary covariate (e.g.
+$x=1$ for treated patients and $x=0$ for controls), $\alpha$ and $\beta$ are regression coefficients, and $\varepsilon_i(a)$ is a mean-zero random variable that reflects the residual (unexplained) variation in the response (log density of species $i$).
+Our interest is usually in the coefficient $\beta$ (slope or average difference between conditions) that describes how the species' abundance changes with $x$, while the intercept $\alpha$ captures differences in the baseline abundance and---we hope---measurement efficiency among species.
+How does taxonomic bias under our measurement model affect estimates of $\beta$ in the simple linear regression?
 
-## Regression over many samples
+Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \@ref(eq:density-prop-est)) with error-free estimates of the total density.
+If the true log density follows the regression Equation \@ref(eq:regression), then it follows from Equation \@ref(eq:density-prop-error) the estimated log density equals 
+\begin{align}
+  (\#eq:regression-error)
+  \log \hat{\text{density}}_i(a)
+%  &= \alpha + \beta x(a) + \varepsilon_i(a) + \log \text{efficiency}_i - \log \text{mean efficiency}(a)
+  = [\alpha + \log \text{efficiency}_i] + [\beta - \log \text{mean efficiency}(a)] x(a) + \varepsilon_i(a).
+\end{align}
+This equation shows that the species-specific portion of the error affects the intercept term while the sample-specific portion (log mean efficiency) affects the slope term.
+Thus, as in the case of fold changes between two samples (Equation \@ref(eq:prop-fc-error)), it is the variation in the (log) mean efficiency across samples that we must worry about distorting our DA results.
 
-Formal DA analysis is typically used to measure average patterns across many samples, such as how the abundance of a taxon varies with some covariate of interest.
-The covariate may be discrete, such as whether the sample is from a healthy or sick person, or it may be continuous, such as a measure of space, time, or temperature.
-Often these analyses can be framed as a regression problem.
-For example, we might hypothesize that the log density of an individual taxon $i$ changes with a variable $x$ according to the simple linear regression, 
-\begin{align}
-  E\left[\log \text{density}_{i} \mid x \right] = a_{0} + a_{1}x
-\end{align}
-where $x$ is either continuous (e.g., sediment depth) or binary (e.g., $x=1$ for treated patients and $x=0$ for controls).
-For a continuous covariate, these coefficients are the intercept, $a_0$, and the slope, $a_1$, of a line.
-For a binary covariate, they are the average abundance in the first condition and the change in average abundance of the second condition.
-How does taxonomic bias in the measurements of individual samples affect our estimate of the regression, and in particular the slope or difference term $a_1$, for log proportions, ratios, and densities?
+The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$.
+Box \@ref(regression-error) mathematically describes this effect for estimation method of Ordinary Least Squares (OLS) and Maximum Likelihood Estimation (MLE) and Figure \@ref(fig:regression-example) illustrates using simulated data.
+Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
+The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with $x$ relative to that of the response (here, $\log \text{density}_{i}$) with $x$ or, equivalently, the relative magnitudes of their slopes.
+As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
+Variation in the log mean efficiency that is uncorrelated with $x$ does not systematically distort $\hat \beta$ but does affect its precision, typically leading to increased standard errors as the variation in log mean efficiency effectively acts as an additional source of noise in measured abundance (Figure \@ref(fig:regression-example) D).
+The exception is for species whose residual variation is strongly positively correlated with that of log mean efficiency (here, Species 9), which can appear to have less random variation and receive standard errors that are too small.
+Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).
 
-_NOTE: Consider making the above abundance instead of density, and say that it could be proportion, ratio, or density_
+With this understanding in place, we briefly summarize how taxonomic bias affects estimation of the simple linear model in other abundance types.
+The results for proportion-based density estimates with accurate total densities also apply to LFC analysis of proportions.
+Similar results apply to microbiome regression tools (such as corncob; @martin2020mode) that perform regression on the logit (instead of log) proportion of a species; however, 
+the mean efficiency of the entire sample must instead be replaced with the mean efficiency among all species excluding the focal species, which causes the absolute error in regression coefficients to vary somewhat across species.
+Because ratios and ratio-based densities are subject to consistent multiplicative error, in analysis of log ratios and the log densities derived from them, only the estimated intercept $\hat \alpha$ is affected by taxonomic bias, while the point estimate and standard error of the estimated slope $\hat \beta$ remain unaffected.
 
-### Proportion-based
+- Important complications to mention:
+  - Error in total density or targeted density estimates
+  - Counting error (so far ignored) can lead to missed associations of species with low measurement efficiencies.
+Example: G.
+vaginalis associations are not detected by several studies that use certain V13 primers that simply don't amplify this species well.
+This is a case where the precision in estimated associations is dramatically reduced, by a different mechanism than what is described above.
 
-P: Results for proportions and for proportion-based density, and 
-- Equations for both; ref details box and SI for details and derivations
-- The sample-dependent term is absorbed by the model intercept, which is typically not of interest to the researcher. - connect to canceling of efficiencies in FC result
-- Error in slope equal to inverse slope in mean efficiency - connect to FC result
-- Intuitive summary in terms of the correlation / covariation with the covariate 
-    - When covaries - statistical bias
-    - When doesn't covary - systematic errors between pairs of samples acts as random error in the regression -> reduced precision and power to detect differences.
-- as w/ FCs, whether the error is significant depends on magnitude of true slope relative to slope of ME.
-- in addition, depends on the SEs of the estimated slope
-
-P: Discuss figure showing results of real and/or simulated example
-- shared LFC error -> sign or (2 types of) magnitude errors for different taxa
-        - if use leopold e.g., edit to show diff taxa on sample samples
-- increase in residual variation and hence standard errors (NEED TO SHOW)
-- old text
-    - Notably, the error in an absolute sense is the same for all taxa—the slope is always reduced by the same amount—but the implications differ for various taxa depending on their slopes, causing magnitude or sign errors depending how the coefficient derived from the (unknown) true abundances compares to that of the mean efficiency (Figure \@ref(fig:regression-example)).
-
-#### Figure
+<!-- Figure: Regression example -->
 
 ```{r regression-example, fig.cap = '(ref:regression-example)'}
-svg_path <- here::here("figures/leopold2020host-example.svg")
+svg_path <- here::here(
+  "notebook/_posts/2021-08-03-simulate-regression-example/",
+  "simulate-regression-example_files/figure-html5/",
+  "main-figure-1.svg"
+)
 if (knitr::is_html_output()) {
@@
 
-(ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.**
+(ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.** This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by $x=0$ and $x=1$.
+In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFC, which drives the positive association of the log mean efficiency with the condition (shown in Panels A and B).
+This positive LFC in the log mean efficiency induces a systematic negative shift in the estimated LFCs of all species (Panels C and D).
+Panel D shows the mean LFC (points) with 95% confidence intervals (CIs), for each species estimated from either the actual or the measured densities.
+The error (difference in LFC estimates on measured and actual) equals the negative LFC of the mean efficiency (shown in Panel B).
 
-TODO: Add SI figure showing the distribution of mean efficiencies and the relationship with Mel that goes with this figure
+<!-- end Figure -->
 
-Ideal figure perhaps shows how observed = actual + eff - mean eff, and individual taxa cases of sign and magnitude errors, and the distributional shift of all taxa, with the specific examples pulled from this distribution
+::: {.rmdnote}
 
-```{r lfc-distribution, fig.cap = '(ref:lfc-distribution)', out.width = "60%"}
-svg_path <- here::here("figures/illustrations/lfc-distribution.svg")
-if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
-} else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
-}
-rm(svg_path)
-```
+## Box: Error in estimated regression coefficients and standard errors {#regression-error}
 
-(ref:lfc-distribution) **Taxonomic bias causes magnitude and sign errors in differential-abundance estimates.** In the absence of other types of errors, consistent taxonomic bias causes shift in the estimated log fold changes (LFC) or slopes in a differential-abundance analysis that is shared by all taxa, creating magnitude or sign errors depending on the true LFC.
+Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.
 
-### Ratio-based
+Consider Equation \@ref(eq:regression) for the simple linear regression of log density of a species $i$ on covariate $x$.
+Let $y$ stand for the actual log density of a focal species $i$, $z$ stand for the log density that we've measured, and $d = z - y$ equal the difference between the two, which here is the log efficiency of species $i$ minus the log mean efficiency of the sample.
+Let $s_{xy}$ denote the sample covariance between variables $x$ and $y$, $s^{2}_{x} = s_{xx}$ and $s_{x}$ denote the sample variance and standard deviation, and $r_{xy} = s_{xy}/(s_{x}s_{y})$ denote the sample correlation.
 
-1 short P explaining how it is no surprise that in ratio based, the error is absorbed by the intercept and the slopes remain accurate.
-- (summary of previous) leads to Taxonomic bias does not affect ratio-based DA analysis
-- include consideration of error in density estimate
+The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of $z$ equals the sample covariance of $z$ and $x$ divided by the sample variance in $x$ or, equivalently, the sample correlation of $z$ and $x$ multiplied by the ratio of their sample standard deviations,
+\begin{align}
+  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
+\end{align}
+From the (bi)linearity of sample covariances it follows that
+\begin{align}
+  \hat \beta_z 
+  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
+  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
+  = \hat \beta_y + \hat \beta_d,
+\end{align}
+where $\hat \beta_y$ and $\hat \beta_d$ denote the slope estimates for $y$ and $d$ (were these values to be known).
+The absolute error in the estimate $\hat \beta_{z}$ of $\hat \beta_{y}$ is therefore $\hat \beta_{d}$; it is large in a practical sense when $\hat \beta_{d}$ is large (in absolute value) compared to $\hat \beta_{y}$, which corresponds to the covariance of $d$ with $x$ being large compared to the covariance of $y$ with $x$.
 
-## Caveats and other considerations
+In our case, the covariance of $d$ equals the negative of the covariance of log mean efficiency with $x$.
+The absolute error (i.e., the _bias_ in a statistical sense) in $\hat \beta$ is equals the negative covariance of log mean efficiency scaled by the variance in $x$.
+This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
+For species that covary with $x$ more strongly than the log mean efficiency, the error will be relatively small.
+This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with $x$ compared to the log density of the focal species.
 
-- Other error sources and violations of this idealized model
-    - inconsistent error in density estimate, could be correlated or not with the covariate
-    - for DNA based: non-linearity of extraction
-    - contamination
-- implications of these in our above results
-    - proportion-based: won't get perfect error sharing across taxa
-    - ratio-based: won't get perfect canceling
-    - -> can limit the effectiveness of our 'solutions' below.
+We can similarly understand the impact of measurement error on the precision of our slope estimates.
+The OLS and MLE estimated standard error in $\hat \beta$ are both approximately
+\begin{align}
+  \hat{\text{se}}(\hat \beta)
+  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
+\end{align}
+where $s_{\hat \varepsilon}$ is the sample standard deviation of the residuals
+(@wasserman2004allo Chapter 13).
+The sample residuals of $z$, $y$, and $d$ have a similar relationship to the regression coefficient estimates,
+\begin{align}
+  \hat \varepsilon_z 
+    &\equiv z - \hat \beta_z x
+  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
+  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
+  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
+\end{align}
+(Note, here I've omitted the subscript indicating the dependence on the sample.)
+It follows that the sample variances of the residuals of $z$, $y$, and $d$ are related through
+\begin{align}
+  s^2_{\hat \varepsilon_{z}} 
+  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
+  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
+\end{align}
+The standard deviation of the $z$ residuals is increased above that of the $y$ residuals when the $d$ residuals are either uncorrelated or positively correlated with the $y$ residuals, but may be decreased when the $y$ and $d$ residuals are negatively correlated.
 
-For this version, can just say this, and in future versions I can be more concrete as I think through more .
-
-### Error in density estimates
-
-In reality, there will be some error in the total density.
-If this error is correlated with the mean efficiency of the sequencing measurement, it can actually reduce the error in the fold change estimate (See Box).
-
-Possible continuations of this P:
-1) Such a correlation might arise through DNA extraction bias when the total density is measurements of total abundance from DNA, such as florescent-based DNA quantification or qPCR of a universal 16S marker.
-In this case, a perfectly linear but taxonomically biased extraction protocol will have a DNA yield that is proportional to the true density and the average absolute extraction efficiency of its cells.
-16S concentration in the extracted DNA will be proportional to the average produce of extraction efficiency with 16S copy number.
-
-2) work from old, and just use qPCR as a focal example.
-
-- Could also mention live vs dead cells and cell-free DNA as another bias source, giving a ref to below article
-- also mention primer bias; again can be helpful if shared.
-Obvious case is if both qPCR and seq primers are not getting certain taxonomic groups, that might be picked up by an alt approach like bulk DNA or flow cytometry.
-
-Next P:
-Shows subtleties and utility of thinking through an explicit model of the measurement process. 
-When we do not attempt to correct for bias, DNA-based estimate of total might be better for proportion-based density estimates despite giving a poor reflection of the total density.
-For example, (Lacto/Gard example) -> extraction and copy number bias gibe spurious results in total, but good results in individual taxa.
-(Have a details box that does this as a worked example)
-However, the non-linearity of DNA extraction may be a major worry in some cases and does not affect direct cell-density measurements.
-Therefore, alternative options from the solutions section below may be preferable.
-
-Old text to use:
-
-Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
-16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
-Yet these biases are shared by the sequencing measurement.
-Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
-(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
-If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.^[See discussion below]
-Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
-In this case, ...
-
-## Notes
-
-<details><summary style="font-weight:bold;font-size:120%">Details: Taxonomic bias and differential-abundance estimates</summary>
-::: {#details:differential .rmdnote}
-
-#### Proportion-based inference {-}
-
-#### Ratio-based inference {-}
+In our case, the $d$ residuals equal the negative residuals of the log mean efficiency.
+It is plausible that for most species, their residual variation will have a small covariance with log mean efficiency and the net effect of variation in the mean efficiency will be to increase the estimated standard errors, as occurs with most species in Figure \@ref(fig:regression-example).
+However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \@ref(fig:regression-example).
 
 :::
-</details>
+<!-- </details> -->

### 2021-09-02 notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example.Rmd
Revise and render simulated regression example

@@
 description: |
-  A short description of the post.
+  Simulate an example of how bias impacts a regression analysis.
 author:
@@
     self_contained: false
+    toc: true
+    dev: svg
 ---
+
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(
+  comment = "#>",
+  collapse = TRUE,
+  cache = TRUE,
+  autodep = TRUE,
+  echo = TRUE
+)
+```
 
@@
 
-# Simulate actual abundances
+# Simulate data
 
@@
 
-species <- str_c("sp", seq(n_species))
+species <- str_c("Species ", seq(n_species))
 ```
 
-## Actual abundances
+## Actual densities
 
@@
 ```{r}
-n_samples <- 100
+n_samples <- 50
 x <- c(
@@
 
+For the data frame of true coefficients, convert from log-e to log-2.
+
+```{r}
+true_coeffs <- list('(Intercept)' = a_0, x1 = a_1) %>%
+  map_dfr(enframe, ".otu", "truth", .id = "term") %>%
+  mutate(across(truth, ~ . / log(2))) %>%
+  print
+```
+
 ```{r}
@@
   mutate(across(x, factor))
-actual <- phyloseq(
+actual_dens <- phyloseq(
   otu_table(y, taxa_are_rows = TRUE) %>% transform_sample_counts(exp),
@@
 ```{r}
-actual %>%
+actual_dens %>%
   as_tibble %>%
@@
 
-## Biased abundances
+## Taxonomic bias and measured densities
+
+Draw efficiencies from a log normal distribution; set relative to the geometric mean efficiency.
 
@@
 ) %>%
+  # {. - min(.)}
   {. - mean(.)}
@@
 
-The observed (i.e. estimated) proportions and densities are given by perturbing the actual densties by the efficiencies, and normalizing to proportions or to the original (correct) total.
+Data frame with all species parameters,
 
 ```{r}
-observed_prop <- actual %>% perturb(efficiency)
-observed_dens <- actual %>% perturb(efficiency, norm = "keep")
+species_params <- true_coeffs %>%
+  pivot_wider(names_from = term, values_from = truth) %>%
+  rename(intercept = '(Intercept)') %>%
+  left_join(efficiency %>% enframe('.otu', 'efficiency'), by = '.otu') %>%
+  mutate(log2_efficiency = log2(efficiency)) %>%
+  print
 ```
 
-## Mean efficiencies
+```{r}
+species_params %>%
+  ggplot(aes(x1, log2_efficiency)) +
+  geom_text(aes(label = str_extract(.otu, '[0-9]+')))
+```
+
+Question: Is it true that the first species is driving the association?
+
+The measured (i.e. estimated) proportions and densities are given by perturbing the actual densties by the efficiencies, and normalizing to proportions or to the original (correct) total.
+
+```{r}
+measured_prop <- actual_dens %>% perturb(efficiency)
+measured_dens <- actual_dens %>% perturb(efficiency, norm = "keep")
+```
+
+## Compute mean efficiencies
 
 ```{r mean-efficiency}
-mean_efficiency <- actual %>%
+mean_eff <- actual_dens %>%
   transform_sample_counts(close_elts) %>%
@@
   enframe(".sample", "mean_efficiency") %>%
-  left_join(sam, by = ".sample")
+  left_join(sam, by = ".sample") %>%
+  mutate(log2_mean_efficiency = log2(mean_efficiency))
 ```
@@
 ```{r}
-mean_efficiency %>%
+mean_eff %>%
   ggplot(aes(y = as.factor(x), x = mean_efficiency)) +
@@
 
-## Regressions and summary plots
+# Plots
 
-what is a nice pipeline for doing all the regressions?
+## Estimated and actual densities
 
 ```{r}
-analysis_ps <- function(ps, ...) {
-  mat <- otu_table(ps) %>% orient_taxa(as = "columns") %>% log2
-  lm(mat ~ 1 + x, data = sample_data(ps) %>% as_tibble) %>%
-    broom::tidy()
-}
-analysis_sme <- function(x, ...) {
-  lm(log2(mean_efficiency) ~ 1 + x, data = mean_efficiency) %>%
-    broom::tidy()
-}
+lvls <- a_1 %>% sort(decreasing = TRUE) %>% names
 
-true_coeffs <- list('(Intercept)' = a_0, x1 = a_1) %>%
-  map_dfr(enframe, "response", "truth", .id = "term")
-
-df <- bind_rows(
-  analysis_ps(actual) %>% add_column(type = "Actual"),
-  analysis_ps(observed_dens) %>% add_column(type = "Observed"),
-  analysis_sme(mean_efficiency) %>% 
-    add_column(type = "Mean efficiency", response = "Mean efficiency"),
+dens <- bind_rows(
+  Actual = actual_dens %>% as_tibble,
+  Measured = measured_dens %>% as_tibble,
+  .id = 'type'
 ) %>%
+  rename(density = .abundance) %>%
   mutate(
-    across(type, factor, levels = c("Actual", "Observed", "Mean efficiency")),
-    lower = estimate - 2 * std.error,
-    upper = estimate + 2 * std.error,
-  ) %>%
-  # Add the true coefficients
-  left_join(true_coeffs, by = c("response", "term"))
+    log2_density = log2(density),
+    across(.otu, factor, levels = lvls)
+  )
 ```
@@
 ```{r}
-p1 <- df %>% 
+p_species_all <- dens %>%
+  ggplot(aes(x, log2_density, color = type)) +
+  labs(y = "Log density ") +
+  facet_grid(.otu~type, scales = "fixed") +
+  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
+  # stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
+  stat_summary(geom = "point") +
+  stat_smooth(aes(x = as.integer(x)),
+    method = "lm", size = 0.9, fill = 'grey', se = FALSE
+  ) +
+  theme(legend.position = "none")
+p_species_all 
+```
+
+```{r}
+p_mean_eff <- mean_eff %>%
+  ggplot(aes(x, log2(mean_efficiency))) +
+  labs(y = "Log efficiency ") +
+  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
+  # stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
+  stat_summary(geom = "point") +
+  stat_smooth(aes(x = as.integer(x)),
+    method = "lm", size = 0.9, color = 'black', fill = 'grey', se = FALSE
+  )
+p_mean_eff 
+```
+
+```{r}
+p_mean_eff + p_species_all +
+  plot_layout(ncol = 1, heights = c(0.2, 1)) &
+  colorblindr::scale_color_OkabeIto() &
+  colorblindr::scale_fill_OkabeIto()
+```
+
+## Regression coefficients
+
+```{r}
+lm_fit_mean_eff <- mean_eff %>%
+  {lm(log2_mean_efficiency ~ 1 + x, .)}
+lm_results_mean_eff <- lm_fit_mean_eff %>%
+  broom::tidy(conf.int = TRUE)
+```
+
+```{r}
+lm_fits <- dens %>%
+  group_by(.otu, type) %>%
+  nest %>%
+  mutate(
+    fit = map(data, ~lm(log2_density ~ 1 + x, .))
+  )
+lm_results <- lm_fits %>%
+  mutate(
+    fit = map(fit, broom::tidy, conf.int = TRUE)
+  ) %>%
+  select(-data) %>%
+  unnest(fit) %>%
+  ungroup %>%
+  left_join(true_coeffs, by = c(".otu", "term"))
+
+lm_results_slope <- lm_results %>%
   filter(term == "x1") %>%
   mutate(
-    across(response, fct_reorder, estimate)
+    across(.otu, fct_reorder, estimate)
   ) %>%
-  ggplot(aes(y = response, x = estimate, color = type)) +
+  arrange(.otu)
+```
+
+```{r, fig.dim = c(5.5, 7)}
+# params for arrows showing error
+delta <- lm_results_mean_eff %>% filter(term == "x1") %>% pull(estimate)
+start <- lm_results_slope %>% filter(type == 'Actual', .otu == 'Species 9') %>%
+  pull(estimate)
+
+p_coef_ci <- lm_results_slope %>% 
+  ggplot(aes(y = .otu, x = estimate, color = type)) +
+  labs(x = "Log fold change", y = NULL) +
   geom_vline(xintercept = 0, color = "grey") +
-  geom_pointinterval(aes(xmin = lower, xmax = upper)) +
+  geom_pointinterval(aes(xmin = conf.low, xmax = conf.high)) +
+  theme(legend.position = 'top') +
+  guides(color = guide_legend(reverse = TRUE)) +
+  annotate(
+    geom = 'segment', color = "darkgrey",
+    arrow = grid::arrow(length = unit(0.1, "inches")), 
+    x = start, xend = start - delta, 
+    y = 10.5, yend = 10.5
+  ) +
+  coord_cartesian(clip = 'off')
+
+p_coef_ci_with_true_coef <- p_coef_ci +
   geom_point(data = ~filter(., type == "Actual"),
@@
     color = 'black', shape = '+', size = 4)
-p2 <- df %>% 
+
+p_coef_ci_mean_eff <- lm_results_mean_eff %>% 
   filter(term == "x1") %>%
-  mutate(
-    across(response, fct_reorder, estimate)
-  ) %>%
+  ggplot(aes(y = "Mean efficiency", x = estimate)) +
+  expand_limits(
+    x = c(
+      min(lm_results_slope$conf.low), 
+      max(lm_results_slope$conf.high)
+  )) +
+  labs(x = "Log fold change", y = NULL) +
+  geom_vline(xintercept = 0, color = "grey") +
+  geom_point() +
+  annotate(
+    geom = 'segment', color = "darkgrey",
+    arrow = grid::arrow(length = unit(0.1, "inches")), 
+    x = 0, xend = delta, 
+    y = 1.0, yend = 1.0
+  )
+
+p_coef_dot <- lm_results_slope %>% 
   ggplot(aes(y = type, x = estimate, fill = type)) +
+  labs(x = "Log fold change", y = "Type") +
   geom_vline(xintercept = 0, color = "grey") +
   stat_dots()
-p1 / p2 + 
-  plot_layout(heights = c(1, 0.3)) &
+
+p_coef_ci_mean_eff / p_coef_ci_with_true_coef / p_coef_dot + 
+  plot_layout(heights = c(0.2, 1, 0.3)) &
   colorblindr::scale_color_OkabeIto() &
   colorblindr::scale_fill_OkabeIto()
-  # scale_color_brewer(type = 'qual', palette = 7) &
-  # scale_fill_brewer(type = 'qual', palette = 7)
 ```
 
+The true coefficients fall within the 'Actual' CIs, confirming that our simulations worked.
+For the manuscript figure, we won't include the true data-generating coefficients, since our aim is to illustrate the relationship between coefficients estimated on the 'Actual' and 'Measured' measurements.
+
+## Manuscript figure
+
+Chose species that show the full range of qualitative behaviors in terms of the effect of taxonomic bias on the estimated slope coefficient.
+
+- Species 9: Bias causes a substantial decrease in the magnitude of the slope
+- Species 10: Bias causes a substantial decrease in the magnitude of the slope, such that the estimate is no longer distinguishable from zero.
+- Species 5: Bias causes a sign error
+- Species 7: Bias cases a small negative slope that would not have been distinguishable from zero to become significantly (in a biological and statistical sense) negative
+- Species 2: Bias cases a substantially negative slope to become more negative (increased magnitude)
+
+<!--  -->
+
 ```{r}
+species_to_plot <- str_c("Species ", c('9', '10', '5', '7', '2'))
+p_species_focal <- p_species_all
+p_species_focal$data <- p_species_focal$data %>% 
+  filter(.otu %in% species_to_plot)
+
+# Bold these species in coeffcient CI plot
+p_coef_ci1 <- p_coef_ci +
+  theme(
+    axis.text.y = element_text(
+      face = ifelse(levels(p_coef_ci$data$.otu) %in% species_to_plot, 
+        "bold", "plain")
+    )
+  )
+
+# Set range of mean eff panel to match
+rng <- p_species_focal$data %>%
+  pull(log2_density) %>%
+  {max(.) - min(.)}
+m <- p_mean_eff$data %>% pull(mean_efficiency) %>% log2 %>% mean
+p_mean_eff1 <- p_mean_eff + 
+  expand_limits(y = c(m - rng/2, m + rng/2))
+```
+
+```{r main-figure, fig.dim = c(7, 7)}
+(p_mean_eff1 + ggtitle("Mean efficiency")) +
+  (p_coef_ci_mean_eff + ggtitle("LFC in mean efficiency")) + 
+  (p_species_focal + ggtitle("Actual and measured densities\nof select species")) + 
+  (p_coef_ci1 + ggtitle("Estimated LFCs of all species") +
+    theme(legend.box.margin = margin(b = -15))
+  ) + 
+  plot_layout(ncol = 2, heights = c(0.2, 1)) +
+  plot_annotation(tag_levels = 'A') &
+  colorblindr::scale_color_OkabeIto() &
+  colorblindr::scale_fill_OkabeIto() &
+  theme(
+    plot.title = element_text(face = "plain")
+  )
+```
+
+```{r, eval = FALSE, include = FALSE}
 ggsave('/tmp/regression-example.pdf', width = 6, height = 5, units = 'in')
@@
 
-here
+## Standard errors
 
-- understand why the 'actual' estimates are generally more extreme than the true lfc
-<!--  -->
+```{r, fig.dim = c(5,5)}
+lm_results %>%
+  mutate(across(.otu, factor, levels = lm_results_slope$.otu %>% levels)) %>%
+  ggplot(aes(y = .otu, x = std.error, color = type)) +
+  facet_wrap(~term) +
+  geom_point() +
+  colorblindr::scale_color_OkabeIto()
+```
 
-## Data plots of specific examples
+Can see that the standard error is generally increased in the measured density estimates except for Species 9, whose residual variation we expect to be correlated with the log mean efficiency and hence negatively correlated with the measurement error.
 
-Plots like the first ones I made for the Leopold data.
+Let's confirm that the relationship between the SEs in the measured response, the actual response, and the mean efficiency agree with our theoretical expectation.
+To do so, we must compute the sample covariance of the residual of the actual response and the mean efficiency for each species.
 
 ```{r}
-species_to_plot <- c('sp9', 'sp1', 'sp5', 'sp7', 'sp8')
+aug_mean_eff <- lm_fit_mean_eff %>% 
+  broom::augment() %>% 
+  add_column(.sample = mean_eff$.sample)
+aug <- lm_fits %>%
+  mutate(
+    .sample = map(data, pull, '.sample'),
+    augment = map(fit, broom::augment),
+  ) %>%
+  select(-data, -fit) %>%
+  unnest(c(.sample, augment)) %>%
+  ungroup %>%
+  left_join(aug_mean_eff, by = c(".sample", "x"), suffix = c('.dens', '.mean_eff'))
 ```
@@
 ```{r}
+cov_actual_with_mean_eff <- aug %>%
+  filter(type == 'Actual') %>%
+  with_groups(.otu, summarize,
+    cov = cov(.resid.dens, .resid.mean_eff)
+  ) %>%
+  print
 ```
 
-### attempt 1
+```{r}
+std_err_mean_eff <- lm_fit_mean_eff %>% broom::tidy() %>%
+  filter(term == 'x1') %>% pull(std.error)
+std_errs <- lm_results_slope %>%
+  select(type, .otu, std.error) %>%
+  pivot_wider(names_from = type, values_from = std.error, 
+    names_prefix = "se_") %>%
+  left_join(cov_actual_with_mean_eff, by = '.otu') %>%
+  print
+```
 
 ```{r}
-mean_efficiency1 <- mean_efficiency %>%
-  rename(.abundance = mean_efficiency) %>%
-  add_column(.otu = "Mean efficiency")
-data_long <- 
-  bind_rows(
-    Actual = actual %>% as_tibble,
-    Observed = observed_dens %>% as_tibble,
-    'Mean efficiency' = mean_efficiency1,
-    .id = 'type'
+std_errs %>%
+  mutate(.otu,
+    se2_Measured = se_Measured^2,
+    pred = se_Actual^2 + std_err_mean_eff^2 -
+      cov * 2 / ((n_samples - 2) * var(x))
+      # cov * 2 * (n_samples - 1) / (n_samples * (n_samples - 2) * var(x))
   )
@@
 
-```{r}
-species_to_plot <- c('Mean efficiency', 'sp9', 'sp1', 'sp5', 'sp7', 'sp8')
+The prediction for the squared standard error from the theoretical calculation agrees with what we observe.
 
-data_long %>%
-  filter(.otu %in% species_to_plot) %>%
-  mutate(
-    across(.otu, factor, levels = species_to_plot)
-  ) %>% 
-  ggplot(aes(x, log2(.abundance), color = type)) +
-  facet_wrap(~.otu, scales = "free_y", ncol = 1) +
-  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
-  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
-  stat_smooth(aes(x = as.integer(x)),
-    method = "lm", geom = "line", size = 0.9
-  )
-```
-
-### attempt 2
-
-```{r}
-data_long <- 
-  bind_rows(
-    Actual = actual %>% as_tibble,
-    Observed = observed_dens %>% as_tibble,
-    .id = 'type'
-  )
-```
-
-```{r}
-species_to_plot <- c('sp9', 'sp1', 'sp5', 'sp7', 'sp8')
-
-p3 <- data_long %>%
-  filter(.otu %in% species_to_plot) %>%
-  mutate(
-    across(.otu, factor, levels = species_to_plot)
-  ) %>% 
-  ggplot(aes(x, log2(.abundance), color = type)) +
-  facet_grid(.otu~type, scales = "free_y") +
-  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
-  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
-  stat_smooth(aes(x = as.integer(x)),
-    method = "lm", geom = "line", size = 0.9
-  ) +
-  theme(legend.position = "none")
-
-p4 <- mean_efficiency %>%
-  ggplot(aes(x, log2(mean_efficiency))) +
-  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
-  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
-  stat_smooth(aes(x = as.integer(x)),
-    method = "lm", geom = "line", size = 0.9
-  )
-```
-
-```{r}
-p1 + p3 + p2 + p4 +
-  plot_layout(ncol = 2, heights = c(1, 0.3)) &
-  colorblindr::scale_color_OkabeIto() &
-  colorblindr::scale_fill_OkabeIto()
-
-```
-
-### Testing plot types
-
-The difference plots in 'dabestr' seem like they might be ideal; specifically the Gardner-Altman Two Group Estimation Plot
-
-```{r}
-# library(dabestr)
-```
-
-But perhaps it will be better to manually create this plot using ggdist.
-Could compute the bootstrapped difference estimates, then plot them with stat_halfeye in vertical orientation at x=3.
-
-HERE.
-
-https://mjskay.github.io/ggdist/
-
-```{r}
-```
-
-```{r}
-data_long %>%
-  filter(.otu == 'sp5', type == 'Actual') %>%
-  ggplot(aes(x, log2(.abundance))) +
-  geom_quasirandom(alpha = 0.3, groupOnX = TRUE)
-
-```
+TODO: Sort out why it is this version, and not the other, that gives agreement.
+Perhaps the standard errors being returned by R are using the MLE estimate of sigma instead of the OLS estimate?

### 2021-09-07 solutions.Rmd
Revise section on potential solutions

@@
-# Potential solutions
+# Potential solutions {#solutions}
 
-_This section needs to be revised to account for the expanded earlier theory, especially relating to targeted measurements and ratio-based methods._
+The theoretical results from Section \@ref(differential-abundance) suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.
+We present five that seem particularly promising:
+The first three are method for correcting or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.
 
-## Use control communities to correct relative abundances
+- Somewhere note that we're considering ideas beyond 'MGS protocol optimization', but that is still important - tuning protocols to reduce taxonomic bias, and to make it as consistent as possible across samples.
+(Our results confirm the importance of both of these)
 
-In principle, control ("mock") communities containing representative taxa from the environment of interest can be used to directly estimate the measurement efficiencies and correct bias in the MGS measurements ("calibration").
-Such an approach may be sufficient for synthetic-community experiments, where all taxa are culturable, and for relatively simple natural communities like the vaginal microbiome that are dominated by a small number of culturable taxa.
-But suitable mock controls may not be feasible for most complex natural ecosystems and require significant effort to develop.
+## Calibrate compositions using community controls {#calibrate-compositions}
 
-An closely-related alternative to mocks are controls derived from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (@mclaren2019cons).
-A natural fecal standard is currently being developed by [NIST](https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material) and one has recently been made available commercially by Zymo Research ([ZymoBIOMICS Fecal Reference](https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology)).
-Ensuring the stability and homogeneity of such control samples can be challenging.
-Careful testing is also needed to ensure that the preparation and storage of the controls has not significantly affected the taxonomic bias relative to the experimental samples in a given application.
-As it is feasible to characterize a single standard much more extensively than a typical community sample, we may be able to obtain an estimated composition we feel comfortable treating as the ground truth.
-Yet even when this is not possible, such natural standards can allow us to reconcile results across studies despite not knowing the truth.
+The most complete but experimentally challenging approach to mitigating taxonomic bias involves directly measuring the efficiencies of individual species using control samples (@mclaren2019cons).
+Suitable controls are any community sample whose composition (species identity and relative abundances) has been previously characterized, either by construction (in the case of constructed 'mock' communities) or by measurement with a _reference protocol_ chosen to serve as a measurement standard.
+These controls, when measured alongside the target experimental samples, can be used to estimate the relative efficiencies among the superset of species across the control samples.
+The estimated efficiencies can then be used to calibrate the measured relative abundances among the control species in the target samples to the chosen measurement standard, by simply dividing the species' read counts by their corresponding efficiencies or using a more sophisticated statistical modeling framework.
+These calibrated compositions can then be used for arbitrary downstream microbiome analyses, including DA analysis.
+Calibration of species not in the controls requires first imputing their efficiencies from the control measurements, for example using phylogenetic relationships, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
+Though such imputation is conceptually straightforward to implement, its reliability and utility remain untested.
 
-- Except in synthetic community experiments, we are always likely to have taxa not in our controls.
-We can attempt to use phylogenetic prediction or other modeling of efficiencies to predict the efficiencies of other taxa; however there will be signifciant undertainty.
-Therefore our control measurements may ideally serve to inform a bias-sensitivity analysis of the sort described below.
+- The above states the gist; can also have a paragraph with description/discussion of how this is/has been used so far, and ongoing developments and challenges. Some bullets for this follow.
+- A mock-community approach may be sufficient for synthetic-community experiments, where all taxa are culturable, and for relatively simple natural communities like the vaginal microbiome that are dominated by a small number of culturable taxa.
+- But suitable mock controls may not be feasible for most complex natural ecosystems and require significant effort to develop.
+- An closely-related alternative to mocks are controls derived from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (@mclaren2019cons).
+- A natural fecal standard is currently being developed by [NIST](https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material) and one has recently been made available commercially by Zymo Research ([ZymoBIOMICS Fecal Reference](https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology)).
+- Ensuring accurate quantification, stability, homogeneity of such controls remains a major challenge.
+- Careful testing is also needed to ensure that the preparation and storage of the controls has not significantly affected the taxonomic bias relative to the experimental samples in a given application.
+- As it is feasible to characterize a single standard much more extensively than a typical community sample, we may be able to obtain an estimated composition we feel comfortable treating as the ground truth.
+- Yet even when this is not possible, such natural standards can allow us to reconcile results across studies despite not knowing the truth.
+- Given the significant uncertainty about estimated efficiencies (stemming from all the above raised caveats), rather than as providing calibration to the 'truth', the estimated efficiencies might be best seen as priors or hypotheses to inform a bias-sensitivity analysis of the sort described below.
 - Should mention the two extant cases where mocks of relevant taxa have been created: The vaginal-HMP / MOMS-PI study (@brooks2015thet and @fettweis2019thev) and the @leopold2020host plant-fungal experiment.
@@
 
-## Use control measurements to estimate error
+## Use ratio-based abundance measurement
 
-Because targeted absolute-abundance measurements are expected to provide (relatively) accurate estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
-We illustrate the basic idea using the @lloyd2020evid experiment described above.
-Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.
-Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
-@lloyd2020evid found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
-In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa [APPENDIX].
+The DA analyses we consider don't need accurate estimates of relative abundances within individual samples---they only require accurate fold changes across samples.
+Section \@ref(differential-abundance) showed that fold changes in species ratios and ratio-based density estimates can remain accurate even when taxonomic bias distorts the values in individual samples.
+Hence a solution is to use to analyze variation in ratios and ratio-based density estimates.
+For the latter, practical options include
 
-- We need an illustration of this idea.
-Might drive home the point that multiple measurements can be used both as qualitative checks and quantitative corrections.
-Could also mention the ability to use host and/or diet reads (or computational AA approaches) as a check against the experimental AA approach.
-- 'specific microbial taxa' - I'm not sure if/how to address the taxonomic levels issue
-- Details: Table 1 indicates that the doubling time estimated for Bathyarchaeota in Core 30 was 10.1 by FRAxC and 10.3 by qPCR, which corresponds to slopes (log abundance / time) of 1/10.1 and 1/10.3.
-The difference between these slopes, 1/10.3 - 1/10.1, is an estimate of the slope of the mean efficiency and hence is the correction that should be applied to all taxa’s FRAxC estimates.
+- spike-ins of extraneous species at constant and/or known density
+- housekeeping taxa, stipulated or computationally identified as having a constant density across samples
+- targeted measurement of specific taxa
 
-## Use ratio-based estimation
+TBD what further discussion or illustration to have here.
 
-A comprehensive approach of estimating the efficiency of all taxa is needed for getting calibrated relative abundances of all taxa within individual samples.
-But what if we only want calibrated DA analysis?
-That is, we are comfortable with not knowing whether *E. coli* is 10% of our sample if we can confidently determine that it doubled or halved between sample conditions?
-This problem is easier to solve, and one solution is to use analysis methods based on the ratios among taxa rather than the proportions of individual taxa.
+### notes
 
-A subset of methods for analyzing differential relative abundance are derived from the field of Compositional Data Analysis (CoDA).
-The defining feature of CoDA methods is that they are based on fold-changes in ratios among elements (here, the taxa) (REF AITCHISON).
-The use of CoDA methods in microbiome analysis has largely been motivated by concerns over the negative correlations between taxa induced by the sum-to-one constraint in taxon proportions.
-But @mclaren2019cons showed that, due to their property of perturbation invariance (REF AITCHISON), the results of CoDA differential relative abundance analysis are invariant to consistent taxonomic bias.
-From Equation \@ref(eq:error), the observed ratio of a taxon $i$ to a taxon $j$ in a sample $s$ is
-\begin{align}
-  \frac{\text{observed}_{i}}{\text{observed}_{j}}=\frac{\text{actual}_{i}}{\text{actual}_{j}}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}
-\end{align}
-The error is independent of the sample composition and thus cancels in any function that is a fold-change of this (or any) ratio of taxa between samples.
-(Perhaps insert example from Figure \@ref(fig:vaginal-example)ABC all giving same ratio FCs.)
+For spike-ins, a crucial consideration may be whether the spike-in is added pre- or post-DNA extraction (ref appendix, Harrison).
+Similarly, we must consider whether targeted measurements are made pre extraction (e.g. CFU counting or ddPCR directly on cells) or post extraction (e.g. ddPCR or qPCR of DNA).
+If DNA yield is linear, then either approach should give reliable fold changes; but if DNA yield is non-linear (for example, saturating in higher biomass samples) then systematic errors can still arise with the post-extraction approach, though these may still be reduced compared to proportion-based inference (REF appendix).
+(Why expect to be reduced: This saturation will still be less than what would be imposed by library normalization, though I need to think about this a bit.)
+In additional, substantial random variation in the (log) DNA yield of a sample with a given biomass will reduce the precision of the estimated abundances that is possible with the post-extraction approach in a manner that cannot be circumvented by increasing the number of spike-in or targeted species.
 
-Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
-In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
-Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
-To estimate the abundance of a focal taxon $i$, this approach multiplies the ratio of reads (or proportions) of taxon i to a reference taxon $r$ by a known or estimated abundance of the reference taxon, (while ignoring bias)
-\begin{align}
-  \text{estimate}_{i} = \frac{\text{observed}_{i}}{\text{observed}_{r}}*\text{abundance}_{r}
-\end{align}
-So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
-Yet the reference taxon could also be a naturally occurring taxon whose absolute abundance we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.
-Because taxa do not compete (or competition is greatly reduced) in such targeted measurements, any taxonomic bias associated with them is expected to create constant multiplicative error across samples and so not affect fold change estimates [APPENDIX].
-APPENDIX describe various theoretical and experimental considerations for both approaches.
+Notes on implementation:
 
-- note from KL on choice of reference taxon: "It seems like you'd have to be pretty careful to pick a reference taxon that was not changing a lot between samples."
-    - this comment indicates I need to be clearer about the three reference strategies - assume constant AA; (but may not be available so can) spike-in at constant AA; (but that might not be convenient, so can) measure AA to adjust for changes in AA.
-    - my reply in the doc: "the idea is that either the reference is a taxon spiked at a fixed abundance to each sample, or it is a taxon you've measured independently with ddPCR or qPCR and so can account for its variation through the given equation." "There are also computational methods that attempt to identify taxa that don't vary (under the assumption that most taxa don't vary), which is an approach inspired from the RNAseq/gene expression literature (where it is probably a more valid assumption that most genes have consistent expression); Ben in his comments was suggesting we mention this approach too.
-"
-- On taxa not competing in targeted measures: "I think DNA-based measurements would still be affected by competition if DNA extraction yield is a strongly saturating function of input.
-But if this saturation isn't very strong then the effect should remain much less than the MGS measurement, which also has competition imposed by various normalization steps "
-- on 'considerations' noted in the last sentence: "perhaps most notable is how to deal with a given reference taxon not being present or detectable in every sample "
-- clarify that there are 2 ways to use spike-ins: Proportion-based and ratio-based method; and only the ratio-based method is bias-invariant
+- Targeted measurement 
+- Although solutions can be hacked together, ideally we'd have new statistical tools to jointly model MGS and supplemental measurements or spike-ins and housekeeping taxa.
+- These would account for the uncertainty associated with the targeted measurements, including detection limits, and could naturally allow for handling the fact that targeted taxa may not be present in every sample.
+- The appendix gives some theoretical insight into how to deal with a given reference taxon not being present or detectable in every sample
 
-## Perform a computational sensitivity analysis
+## Calibrate fold changes using measurements of targeted taxa
 
-What can we do for experiments that have already been conducted without mock controls, targeted control measurements, or spike-ins?
-At least two purely computational approaches can still be used.
+In some cases, we may still which to use a proportion-based approach to analyzing variation in density.
+For example, in some systems measuring total cell concentration may be easier than implementing a spike-in experiment or determining a set of reference species to directly measure for which at least one species is present in all samples.
+In this case, targeted measurements of one or more reference species can be used to validate and/or correct the fold changes or regression coefficients infered from the proportion-based densities.
+Recall that the error in proportion-based log fold changes (LFCs) is shared among species and is equal to the inverse LFC fold change in mean efficiency (Equations \@ref(eq:prop-fc-error) and \@ref(eq:regression-error); Figure \@ref(fig:regression-example)D).
+Consequently, an accurate LFC of one or more reference species can serve as a general validation or used to correct the LFCs of all species.
 
-A straightforward and universally available approach is to use computer simulation to determine how the results change under a range of possible sets of efficiencies, which can be randomly generated to reflect certain hypotheses about bias in the given system.
+- NOTE: I'm not sure if I really understand this example below, since I don't see how this interpretation of the Table 1 results mesh with Figure 4
+
+We illustrate this idea with an example from the study by @lloyd2020evid described in Section \@ref(implications), which estimated doubling times for various taxa from the measured log increase in cell density with burial time in marine sediments.
+The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for several bacterial and archeal taxa (CHECK).
+In Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years (CHECK) using the proportion-based densities and 10.3 years by qPCR (@lloyd2020evid Table 1 and Figure 4).
+These doubling times correspond to highly similar estimated slopes of log density per year of $1/10.1 \approx 0.99$ and $1/10.3\approx 0.97$.
+The difference between these slopes, $\approx 0.02$, estimates the slope of the mean efficiency.
+It's small value suggests little systematic variation in mean efficiency with burial time and hence little systematic bias in the estimated doubling times derived from the proportion-based density estimates.
+In this case, close agreement in the DA results across methods for a specific taxon suggests that variation in sample mean efficiency did not significantly distort the MGS-based results and hence provides a degree of validation for all taxa, not just the taxa with targeted measurements.
+(An important caveat for this example is that this analysis assumes consistent efficiencies at the higher taxonomic levels used for qPCR measurement and MGS analysis performed by @lloyd2020evid.)
+Here we compared summary statistics; however, in principle a joint statistical analysis of targeted and proportion-based density measurements would provide more statistical power to estimate and correct the effects of variation in the mean efficiency across samples and provide calibrated fold change and regression estimates for all taxa [APPENDIX].
+
+## Perform a bias sensitivity analysis
+
+What can we do for experiments that have been conducted without control communities, targeted control measurements, or spike-ins?
+One approach is to conduct a bias sensitivity analysis to probe the extent and way in which the results of a DA analysis are sensitive to different possible taxonomic biases in our experiment.
+
+A straightforward approach to bias sensitivity analysis is to use computer simulation to randomly generate efficiency vectors, each of which is used to calibrate the MGS measurements.
+These calibrated MGS measurements represent the true compositions under the hypothesized taxonomic bias. 
+By re-running our DA analysis and comparing results across these simulated datasets, we can begin to understand how our initial results might be influenced by the strength and structure of the taxonomic bias in our experiment.
+
+The simulated efficiency vectors may be generated to reflect certain hypotheses about bias in the given system.
 The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
-Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance as well as other microbiome analyses such as alpha diversity and community similarity analyses.
+Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance and microbiome analyses more generally.
 
@@
 ## Use bias-aware meta-analysis
+
+So far we have considered analysis of samples subject to the same taxonomic bias;
+however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in different groups of samples.
+This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses where the effects of bias might cancel within a study.
+
+A potential solution is to perform a bias-aware meta-analyses that jointly analyzes multiple data from studies while explicitely acknowledges the distinct, unknown taxonomic bias in each study.
+Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model "batch effects" --- amorphous differences in the measurements of each study.
+By configuring these models so that these latent parameters correspond to the study and species specific efficiencies, we can increase the biological interpretability of these models and gain ability to determine whether disageements in DA results across studies are due to differences in taxonomic bias or other differences (such as host cohort or study design).
+Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by @mclaren2019cons, which can increase improve our ability to find reproducible microbial associations in a given meta-study.
+
+### notes
 
@@
 - Aim to  illustrate with a simulated or real example
+
+- Value: If results of individual studies might be fairly robust to unknown bias, then might be able to efficiently (from a statistical POV) pool studies when they biologically agree or rule out bias (under the used assumptions) as the cause of disagreement when they don't agree.

### 2021-09-08 introduction.Rmd
Revise introduction

@@
 Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads make taxonomic bias a feature of both shotgun and marker-gene measurements.
-The error it causes have been found to in some cases to supersede sizeable biological differences (e.g.
+The error it causes have been found to in some cases to supersede sizable biological differences (e.g.
 @lozupone2013meta) and has plausibly caused replication failures for prominent findings such as the association of decreased Bacteroides and increased Firmicutes in stool with obesity (@finucane2014atax) and the association of certain taxa in the vagina of pregnant women with preterm birth (@callahan2017repl).
+
+<!-- Revise below here to outline -->
 
@@
 
-Although many microbiome researchers are aware of taxonomic bias, it is standard practice to ignore it when performing a DA analysis and interpretting its results.
-One reason for this practice is a widespread belief that taxonomic bias does not affect the inferred _differences between samples_, so long as those samples have been subjected to the same measurement protocol.
-This belief stems from an assumption that the effect of taxonomic bias is to consistently increase or decrease the measured abundance of a given taxon in every sample, so that its effects cancels when computing change across samples.
-@kevorkian2018esti (see also @lloyd2020evid) were, to our knowledge, the first to make this argument mathematically explicit; however, their argument was based on a faulty assumption about the effect of bias on MGS measurement (@mclaren2019cons and Section \@ref(abundance)).
-@mclaren2019cons subsequently analyzed the effects of bias using a mathematical model of MGS measurement, which they validated using marker-gene and shotgun measurements of artificially-assembled (or "mock") bacterial communities.
-They found that the _relative_ measurement efficiencies among taxa are consistent across samples that vary in taxonomic composition (provided they are measured by the same protocol).
-But a bias that is consistent in terms of relative efficiencies will nevertheless cause error in the proportion of a taxon that depends on sample composition and hence can lead to spurious estimated changes across samples.
-In contrast, they showed that error in the ratio between a pair of taxa has an error that is sample-independent error and hence does in fact cancel when estimating log fold changes in the ratio.
-Although they considered measures of relative abundance (proportions and ratios), the same underlying error dynamics apply to methods for inferring cell densities from MGS, which themselves are based on MGS-derived proportions or ratios.
-Together, these results of @mclaren2019cons indicate that taxonomic bias may lead to spurious DA results, but may be more problematic for certain experimental and analytical approaches than others.
+The typical approach to countering taxonomic bias in DA analysis is to standardize the measurement protocol used within a given study.
+In broad strokes, the thinking is that the measurements of samples measured by the same protocol will be affected by bias in the same way and so the inferred differences between samples (the focus of DA analysis) will be unaffected.
+For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples.
+@kevorkian2018esti and @lloyd2020evid were the first (to our knowledge) to make this argument explicit mathematically; however, @mclaren2019cons used a theoretical model (supported by measurements of defined bacterial communities) to show that consistent taxonomic bias can lead to variable fold errors in measured proportions.
+These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase).
+Yet @mclaren2019cons also found that certain changes in relative abundance---in particular, fold changes in the ratios among species---are robust to the effects of bias.
+The implications of these findings for changes in absolute abundance---which remain subject to taxonomic bias in the underlying MGS measurement---and for DA analysis across many species and many samples---as commonly done in microbiome association testing---have yet to be investigated.
 
-<!-- P: What is needed? -->
+<!-- P: Objective and summary of the present article -->
 
-Here we address a number of crucial remaining questions about the impact of taxonomic bias on DA analysis:
-
-- How do these results for the changes between a pair of samples extend to the proto-typical DA analysis of a (linear) regression over many samples?
-- How does taxonomic bias affect the wide variety of abundance measures, and methods for converting relative to absolute abundance, which are currently used in DA analyses?
-- What is the impact on DA analysis of taxonomic bias and other forms of error in the supplemental measurements (such as cytometry and bulk 16S qPCR) that are commonly combined with MGS to obtain absolute abundances?
-- What is the impact of variation in efficiency within a taxon on DA analysis of the taxon?
-(e.g.
-what is the impact of variation in efficiency among Firmicutes species on the infered change in Firmicute abundance?)
-- Last but not least, for which biological scenarios is taxonomic bias likely and not likely to lead to biologically significant inferential errors?
-
-We address these questions through a combination of mathematical analysis, simulation, and analysis and interpretation of data from previously published experiments.
-
-<!-- P: Solutions -->
-
-A second reason that taxonomic bias is generally not addressed (beyond protocol standardization) is a severe lack of proposed, tractable solutions to address it.
-We propose and demonstrate several methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be widely deployed today with only modest changes to existing experimental and analytical workflows.
-Over time, application of these methods to past and future experiments will provide crucial quantitative information about taxonomic bias and the conditions under which spurious results arise for particular DA methodologies.
-Collectively, these methods and insights can provide a practical solution to taxonomic bias in DA analysis and the confidence that is necessary to codify our statistical findings into scientific knowledge and translate them into real-world diagnostics and interventions.
+Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundance.
+Our analysis clarifies how the folk wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
+Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.
+Over time, application of these methods to past and future experiments will provide crucial quantitative information about taxonomic bias and the conditions under which spurious results arise for various DA methodologies.
+Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.

### 2021-09-09 real-world-implications.Rmd
Revise section on real-world implications

@@
-# Implications for real-world inference
+# Implications for real-world inference {#implications}
 
-These observations show that the effects of taxonomic bias do in fact cause error in proportion-based differential abundance analyses.
-Yet if the mean efficiency varies much less across samples than the abundances of individual taxa, this error will be negligible.
-Alternatively, if variation in the mean efficiency is effectively random (not associated with regression covariates), the effect of bias may be to increase the residual variance and thus reduce our statistical power to assay differential abundance—unfortunate to be sure, but perhaps not as much as confidently inferring changes with the wrong sign or magnitude.
-But there are salient biological scenarios where substantial associations between the mean efficiency and covariate can in fact lead to spurious differential abundance inference.
+Partial but not full canceling of the effect of bias in proportion-based analyses, such that possible to have spurious results in principle; but what about in practice?
+Predicting how frequently or which results are likely to be majorly affected is difficult because it depends on the joint variation in efficiencies of species and species abundance dynamics, both of which are poorly understood in most to all biological systems.
+However, through a combination of theoretical considerations, hypothetical scenarios, and real case studies we can begin to develop an understanding or the relevant considerations and build framework for improving our understanding as our knowledge of these dynamics improves.
 
-## Scenarios in which bias may cause spurious results
+## Systematic error in slope or LFC estimates
 
-Figure \@ref(fig:vaginal-example) illustrates how a change in dominance by from a high- to low- efficiency taxon (or vice versa) can easily create spurious DA results.
-Indeed, this scenario may be particularly common in vaginal microbiome studies.
-Human microbiomes often dominated by *Lactobacillus*, but can become dominated by non-*Lactobacillus* species such as *Gardnerella* vaginalis and *Atopobium vaginae*, and these low-*Lactobacillus* states are associated with the disease bacterial vaginosis and, in pregnant women, an increased risk of preterm birth.
-Using mock community measurements from @brooks2015thet, @mclaren2019cons showed that the efficiency of two common *Lactobacillus* species were 20-30X greater than *G. vaginalis* and *A. vaginae*.
-Therefore we should expect samples dominated by *Lactobacillus* to have a much larger mean efficiency than those those dominated by *G.
-vaginalis* and *A.
-vaginae*, which would distort the observed associations of all taxa with any covariate that is also associated with *Lactobacillus* dominance.
-The extraction and 16S sequencing protocol used by @brooks2015thet is similar to that used in the VaHMP MOMs-PI study also led by these researchers to investigate the associations of these taxa with preterm birth (@fettweis2019thev), demonstrating the potential for serious real-world consequences for these statistical/technical concerns.
+Recall from Section \@ref(differential-abundance) that the error in the slope or LFC in a DA regression analysis is proportional to the covariance of the log mean efficiency with the covariate.
+This covariance can be split into two components: the standard deviation of (log) mean efficiency and its correlation with the covariate.
+Hence when considering whether taxonomic bias creates large error, it can be useful to separately ask whether the mean efficiency is likely to vary across samples and whether it is likely to be correlated with a covariate of interest.
 
-- Discussion of the vaginal example in the above paragraph should be distilled in conjunction with the discussion of this example in the previous section
-- Group-B strep / S. agalactiae is another "pathobiont" that has a low efficiency and I might as well included here for completeness. Can cite brooks2015thet and pace2021comp
+One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
+@brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome with the MGS protocol developed for the Vaginal Human Microbiome Project (VaHMP).
+@mclaren2019cons later used this data to estimate the efficiencies of these seven species and develop a method to use these estimates to infer the true compositions of natural samples (see Section \@ref(#calibrate-compositions)).
+SI Analysis (MOMS-PI) uses these estimated efficiencies to correct the compositions and explore the implied variation in mean efficiency in the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) @fettweis2019thev.
+The vaginal microbiome during pregnancy is characterized by relatively low diversity, often with a single _Lactobacillus_ species often forming a majority of reads (considered to be the healthy state) but with occasional shifts to high proportions of other species, such as _Gardnerella vaginalis_, _Atopobium vaginae_, and _Streptococcus agalactiae_, which are thought to cause or indicate poor vaginal health and increased risk of preterm birth.
+The analysis of @mclaren2019cons showed that the efficiencies of two common _Lactobacillus_ species, _L.
+crispatus_ and _L.
+iners_, were 20-30X greater than that of _G.
+vaginalis_, _A.
+vaginae_, and _S.
+agalactiae_.
+Analysis of microbiome trajectories of a single patient across visits indicates that the mean efficiency typically decreases by a factor of 3-10X when the community shifts from a high to low-_Lactobacillus_ state.
+(NOTE: This is a super preliminary claim based on looking at just one woman, and is just a placeholder)
+Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
 
-Similar dynamics occur in a plant-fungal interactions experiment performed by @leopold2020host (Figure \@ref(fig:regression-example)).
-The authors inoculated commensally-colonized trees with a fungal pathogen.
-Using DNA mock communities, they showed that (excluding bias from DNA extraction) the pathogen was 10X more efficiently measured than the median commensal and 40X more efficiently measured than the lowest-efficiency commensal.
-In most hosts, the pathogen increased rapidly between timepoints, driving an increase in the sample mean efficiency that in turn leads to larger decreases being observed in the proportions of commensal taxa than what is predicted using calibrated (bias-corrected) values (REF APPENDIX).
-For example, the calibrated proportions indicate that the commensal *Penicillium* slightly increased in hosts from the Eastern US and slightly decreased in hosts from the Western US.
-Yet the observed (uncalibrated) proportions show moderate and large decreases, respectively (Figure \@ref(fig:regression-example) above).
+<!-- Can cite brooks2015thet and pace2021comp for Group-B strep / S. agalactiae as a "pathobiont".  -->
 
-- In last sentence, insert the the values of the LFCs (ideally with evaluated R code)
+A second study in which control communities make it possible to directly analyze the impact of taxonomic bias in experimental samples was performed by @leopold2020host.
+To study the interactions between a host plant and its fungal commensals and pathogen, the authors inoculated trees with a fungal synthetic community and later exposed plants to a fungal pathogen.
+The authors used ITS amplicon sequencing to measure communities before and after infection, along with mock communities which they used to estimate the species efficiencies with the method of @mclaren2019cons.
+(DNA mocks were used, so that bias due to DNA extraction is not included, but other major bias sources such as PCR and ribosomal copy-number-variation are included.) The pathogen was 10X more efficiently measured than the median commensal and 40X more efficiently measured than the lowest-efficiency commensal.
+A re-analysis of their data (SI Analysis) shows that the mean efficiency varies by ???
+in pre-infection communities and by an average of ???
+pre- and post-infection, with the latter driven by the dramatic increase in the high-efficiency pathogen.
 
-These examples involved scenarios where individual taxa can constitute a large proportion of the community.
-But in samples from highly-diverse ecosystems such as human gut, soil, and ocean sediment, one species rarely dominates.
-In these cases it might seem intuitive that the mean efficiency should be relatively stable across samples: As diversity increases, the mean efficiency effectively averages over efficiencies from a greater number of taxa and so (by the central limit theorem) will converge to a constant value—if taxon efficiencies are statistically independent of taxon abundances (REF APPENDIX).
-In practice, associations between efficiencies and abundances may be common and can allow large shifts in mean efficiency to occur even in highly diverse samples.
+These case studies indicate that one mechanism for variability in the mean efficiency is swings between very high ($\sim 1$) and low ($\ll 1$) proportions of species with particularly high or low efficiencies.
+In highly diverse ecosystems, as seen in soil and the human gut (CHECK), one species rarely dominates within a given sample and so such large swings are not possible.
+We might therefore think that the mean efficiency will remain fairly stable in these settings.
+In fact, one can show mathematically that in randomly assembled communities, the mean efficiency becomes more stable as (Inverse Simpson) diversity increases, as the mean effectively averages over a greater number of species.
+Real communities do not assemble randomly, however.
+Variance in the mean efficiency may remain high if environmental filtering leads to different groups of organisms dominating in different environments, particularly if shared phenotypic values or evolutionary history causes these groups to differ in their efficiencies as well as their ecology.
+For example, gut microbiomes typically have high diversity at the species level but are dominated by just a small number of phyla, and two in particular: The Bacteroidetes and the Firmicutes, the ratio of which varies substantially across individuals (CHECK).
+Many DNA extraction protocols have been found to more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though variation can also be substantial within phyla, @mclaren2019cons).
+Gut samples with large differences in their Bacteroidetes and Firmicutes proportions may therefore have large variation in mean efficiency even if the individual species within each phylum never reach a large proportion.
+On the other hand, competitive-exclusion dynamics can favor more stable taxonomic and/or phenotypic compositions across samples and might thereby drive a more stable mean efficiency than under random assembly
 
-- on the example of 'ocean sediment' in the above P: I asked KL "is it reasonable to characterize the ocean sediment communities from your 2020 paper as "high diversity"?
-In particular, species- or OTU-level Inverse Simpson values are what's relevant here.
-I'm guessing it's intermediate diversity - much less than gut and soil, but much more than the vaginal microbiome", and she replied "The total microbial diversity in marine sediments worldwide is similar to that of soil.
-I'll email you Hoshino et al., 2020 which addresses this." However I'm not sure if the high diversity KL is noting applies to individual soil samples (i.e.
-what scale)
-- Should make sure the above paragraph's claims are precise and accurate; see AW's feedback
-- "As diversity increases" specifically refers to inverse simpson diversity / q=2. Should I explicitly state that?
-- "statistically independent of taxon abundances" - I need to make sure I'm being accurate/precise here.
-What do we actually know?
-Maybe simpler to say, if taxon efficiencies are not correlated with taxon abundance
+Systematic error in DA results requires that the variation in log mean efficiency is correlated with the covariate.
+The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
+For example, in the vaginal microbiome a decline in _Lactobacillus_ and rise of species including _Gardnerella vaginalis_ is associated with preterm birth in pregnant women; hence the variation in mean efficiency driven by these species may also create an association of mean efficiency with preterm status.
+In the plant-fungal experiment of @leopold2020host, the increase in pathogen proportion after infection leads to a large increase in the mean efficiency over time, which systematic increases the observed decline in the proportions of commensal species (SI Analysis).
+The ratio of Bacteroidetes to Firmicutes has been linked to several host traits and health conditions, and thus the variation in efficiency between these two phyla may also drive an association of mean efficiency with these traits.
 
-The human gut provides a clear example of how such associations might arise.
-Gut microbiomes typically have high diversity at the species level but are dominated by just a small number of phyla, and two in particular: The Bacteroidetes and the Firmicutes.
-The ratio of Bacteroidetes to Firmicutes can shift substantially between individuals and has been linked to a number of host traits and health conditions.
-In addition, many DNA extraction protocols more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though there can also be significant variation in extraction efficiency among species within these phyla (@mclaren2019cons).
-For such protocols, we should expect Bacteroidetes-dominated samples to have substantially larger mean efficiencies than Firmicutes-dominated samples, regardless of species-level diversity.
-This association of efficiency with phylum abundance will distort DA inferences at any taxonomic level if the covariate is also associated with these phyla.
+<!-- General mechanisms -->
 
-- mclaren2019cons measures overall extraction efficiency, not lysis efficiency; is that important for the above text?
+Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
+One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
+A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
+Shared evolutionary history among phylogenetically-close species is expected and has been observed to create positive associations in these traits among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
+Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
+In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
+For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.
 
-This example illustrates a general mechanism by which association between the abundances and efficiencies of taxa are created by the common influence of evolutionary history.
-Just as shared evolutionary history creates positive associations in ecological traits that drives positive correlations in how related taxa vary in abundance across samples, so does it create positive associations in bias-related traits that can lead to similar efficiencies among closely related taxa.
-Such phylogenically-associated, bias-affecting traits include cell-wall toughness, ribosomal-operon copy number, binding affinity for a given set of primers, and representativeness in taxonomic databases.
-Ecology and efficiency are both affected by the same evolutionary history, leading to positive associations between the two: A change in diet that increases the relative abundances of many Bacteroidetes species also increases the relative abundance of many easy-to-lyse species.
-
-- first sentence in above P: "the abundances and efficiencies" - the idea is that both are determined by genetically encoded traits
-- this is prob a tricky idea for readers and will have to take care in communicating it, and find nice refs
-
-Associations between measurement efficiency and relative abundance can also arise because a single trait affects both.
+Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
 For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
@@
 
-### Error in absolute abundance measurements
+<!-- P: relative magnitude of species and mean efficiency changes -->
+The biological significance of the error in $\hat \beta$ depends on the relative magnitude of the LFC (or covariance with the covariate $x$) in the focal species vs in the mean efficiency.
+It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
+(This seems intuitively likely and true in examples, but I haven't analyzed it much mathematically and don't have a firm handle on this question.)
+Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
+However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for all other species.
+The experimental and computational methods discussed in Section \@ref(solutions) can mitigate taxonomic bias in these clearly-problematic cases while also helping to amass empirical evidence on the conditions in which bias is unlikely to cause major inferential errors.
 
-- Ben flags that this concept/section main idea should be highlighted more
+- TODO: Address DA analysis of aggregate (or synthetic) taxa, such as phyla, which may be prone to errors due to changes in the species-level make-up of the taxon.
 
-So far we have ignored error associated with our absolute abundance measurements.
-In fact, these measurements can also have a tendency to measure contributions from some taxa more efficiently than others.
-**TODO: State how this could create problems.
-Might also note the ambiguity in what the truth/target is (e.g.
-live cells, or all cells; cells, or 16S rRNA gene copies; observations in Jian, Salonen, and Korpela 2021).**
-Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
-16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
-Yet these biases are shared by the sequencing measurement.
-Though they make qPCR measurements a bad proxy for total cell density^[See note from KL below], when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can offset each other, leading to estimated fold changes that are relatively unaffected by taxonomic bias.
-(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
-If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.^[See discussion below]
-Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
+**Error in total-density measurements:**
 
-- from KL on qPCR being a bad proxy: We showed this empirically for all published measurements in marine sediments in Lloyd et al., 2013 doi:10.1128/AEM.02090-13. Apologies for the self-citation.
-- In the above sentence "If common primers are used...", KL asked a question that shows I'm not being clear enough about what I'm talking about here.
-KL: "Wouldn't you want to use taxa specific primers for qPCR and general primers for the amplicon libraries to do this analysis?
-If they're the same primer sets, then it doesn't tell you anything about the relative abundance of individual taxa within the population grabbed by the primers." my reply: "The "solutions" section talks about using targeted qPCR; this section is talking about qPCR with "universal" primers for the purpose of estimating the total absolute abundance, for use in FRAxC instead of cell counting.
-There have been quite a few papers doing this recently and 16S sequencing companies are starting to offer it as a service so I've been thinking through the implications.
-This section needs to be expanded and the point I'm making about using qPCR for FRAxC is still somewhat preliminary and something we should chat about"
-- should caveat something like 'Unfortunately, qPCR shares a limitation with DNA-based measurements: It assumes linearity of DNA extraction'
-    - this should start own paragraph on this issue. For now, can be a 2 sentence paragraph and ref the appendix.
+So far we have ignored error in total-density measurements that are used to convert estimated proportions to estimated densities;
+however, in reality these measurements are likely to contain substantial systematic error that can either worsen or improve the error in LFC estimates.
 
-## Section conclusion
+- Todo: Add to previous section or Appendix with the relevant theory.
 
-These examples show that there is potential for bias to distort DA results in both low and high diversity settings, but its effects may be negligible in many studies.
-By applying the types of methods demonstrated here to a range of biological and experimental contexts, and further developing theory, we will deepen our understanding of when bias is unlikely to significantly distort biological findings.
+To understand how, we can consider the question of whether bulk 16S qPCR or flow cytometry is a more appropriate method of measuring total bacterial density for use in proportion-based density measurement (@galazzo2020howt, @jian2021comm).
+16S rRNA-gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore affected by variation among species in DNA extraction efficiency, 16S copy number, and primer-binding and amplification efficiency.
+For example, @mclaren2019cons found that the product of extraction efficiency and 16S copy number for _Lactobacillus iners_ was 20X larger than _Gardnerella vaginalis_ (CHECK NUMBERS), such that we should expect samples dominated by _L.
+iners_ to result in qPCR estimates 10-20X higher than equally dense samples dominated by _G.
+vaginalis_.
+These observations suggest that 16S qPCR is a poor proxy for fold changes in total bacterial density.
+Yet, as recently suggested by @jian2021comm, this taxonomic bias in qPCR may simultaneously make it a good measure of total density for use with 16S amplicon sequencing for proportion-based density estimation.
+16S amplicon sequencing shares these same sources of bias and so is likely to have species efficiencies highly similar to those for the qPCR measurements, which we can predict to cause the mean 16S-qPCR efficiency to closely follow the mean 16S-sequencing efficiency, such that the error induced by one largely cancels the error induced by the other (Ref Appendix).
+In contrast, species-specific efficiencies in bulk cell counting are less likely to be strongly correlated with those from 16S sequencing.
+Even supposing cell counting were to provide perfectly accurate measures of total density, it may actually perform worse than qPCR for DA analysis of proportion-based density estimation precisely because the error does not mirror that in the 16S measurement.
 
-(prob move to the conclusion:)
-With this knowledge in hand we will be better placed to make the inevitable tradeoffs when designing new experiments (in terms of how much bias-mitigation measures to use) and also be better able to evaluate our confidence in past results.
+On the other hand, bulk or targeted measurement (or DNA spike-ins) that occurs post DNA-extraction is subject to other error that is not likely to counter that of the 16S-based proportion estimates.
+The yield of DNA extraction can show highly random variation and also be highly non-linear (as a function of cellular biomass) at high concentrations.
+This features create error that is unlikely to offset the variation in mean efficiency in the 16S sequencing (todo: clarify why).
+Targeted measurement of cells and cellular spike-ins, discussed in Section \@ref(solutions), can avoid this issue while also facilitating ratio-based analysis while (unlike bulk cell counting) remaining robust mean-efficiency variation.
 
-## Notes for revision
+## Loss of power and precision (stub)
 
-Include simulation results indicating that bias might not be a problem.
+Variation in mean efficiency that is not associated with covariates can still majorly hinder the goals of microbiome studies by causing reductions in the precision of LFC estimates and the power to detect true associations of a given size.
 
-Outline?
+- Illustrate with case studies?
+- Mention counting noise as an important factor that we've so far ignored but that can interact with taxonomic bias to lead to very low precision in LFC estimates.
+Low efficiency can put a species below the detection limit set by sample read depth, making it impossible to get precise LFCs.
+Study-specific efficiencies can explain why associations often do not replicate across studies.
+Example: _Gardnerella vaginalis_ associations are not detected in certain vaginal microbiome studies that used V13 primers which we now know are bad at amplifying it.
+- Conclude with a nod to the value of the solutions in Section \@ref(solutions) to address these issues: Correcting the errors caused by mean-efficiency variation can improve precision, and bias-sensitivity analysis and bias-aware meta-analysis can give a better sense of the increased uncertainty in LFC estimates due to unknown efficiencies, thereby helping us to understand the increased rate of false positives and/or the larger range of LFC estimates that are compatible with the observed data than what is implied by a bias-free model.

### 2021-09-09 abundance-measurement.Rmd
Move notes to temporary notes file

@@
 Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
-This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr) ^[todo: details about this assumption].
-In this model, we suppose there are a set of *atomic taxa*, that differ in how efficiently they are measured but within which all cells behave similarly (r/v).
+This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
+In this model, we suppose there are a set of *atomic taxa*, that differ in how efficiently they are measured but within which all cells behave similarly.
 For concreteness, we equate atomic taxa with species. 
-*(Note that bias may not be conserved w/in species; ref discussion of synthetic taxa later on)*.
 We ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.
@@
 The species-specific factor is the *relative measurement efficiency* (or *efficiency* for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (@mclaren2019cons).
-The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.^[todo: add note about how, as defined, sequencing effort typically goes down with total density because of library normalization; also mention that it 'sequencing effort' is sometimes equated with total read count or library size, which is not what we've done here].
+The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
 Equation \@ref(eq:measurement-model) implies that the total reads from all species in the sample equal
@@
 By this view, it is meaningless to refer to the relative abundance of an individual species.
-<!-- (Although a proportion is also technically a ratio [of a species to a higher-level taxon], here we use 'ratio' to refer specifically to ratios among species.) -->
-Although other interpretations exist ^[e.g.
+Although other interpretations exist^[e.g.
 the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most DA analysis methods (and microbiome analysis in general), for both relative and absolute abundances, and the differences between them help to clarify the issue over when and why the effects of bias cancel in an analysis.
@@
 For instance, in the example of Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
-
-- note: the error is not necessarily consistent for ratios among higher-order taxa
-- note: The proportion equation is analogous to Equation \@ref(eq:ratio-error) (ratio case), with the denominator taxon is the entire taxonomic domain and the relevant efficiency is the weighted average of all species.
-In fact, both equations can be derived from a more general equation, where the observed ratio of two (potentially non-atomic) taxa equals the true ratio times the ratios of the mean efficiencies within each taxon.
-- note: the results for ratios of higher-order taxa, and of proportions, can both be seen as stemming from a general result where the error in the ratio is the ratio of the mean efficiency within that taxon.
-- perhaps have a box for this
 
@@
 
-- note: right now I'm not consistently ref'ing the equations where the results come from
-
 Ratio-based density estimation follows from the fact that, in the absence of taxonomic bias, the conversion rate from reads to density is just the sequencing effort of the sample and is thus the same for all species (Equation \@ref(eq:measurement-model)).
@@
 
-- note: these two approaches are not always distinct, and it seems that many/most authors are thinking of spike-ins and housekeeping taxa in terms of the proportion-based method; I need to think about how to address this point.
+- Note: The current distinction between proportion- and ratio-based density estimation is incomplete, as the example of spike-in normalization shows: Whether the density of species $i$ is estimated by Equation \@ref(eq:density-ratio-est), or by first estimating total density from the ratio of non-spike-in to spike-in reads and then using Equation \@ref(eq:density-prop-est), the results are mathematically identical.
+Future revisions should clarify this point.

### 2021-09-09 differential-abundance.Rmd
Move notes to temporary notes file

@@
 These measures of DA are also affected by taxonomic bias in our model in a manner that is straightforward to understand in terms of the multiplicative errors described in the previous section.
-
-- Maybe: Add a final sentence summary that the constant multiplicative error in ratio-based abundance estimates does not affect DA measurement, but the varying error induced by variation in the mean efficiency affects DA analysis on proportion-based estimates.
 
@@
 Because ratios and ratio-based densities are subject to consistent multiplicative error, in analysis of log ratios and the log densities derived from them, only the estimated intercept $\hat \alpha$ is affected by taxonomic bias, while the point estimate and standard error of the estimated slope $\hat \beta$ remain unaffected.
-
-- Important complications to mention:
-  - Error in total density or targeted density estimates
-  - Counting error (so far ignored) can lead to missed associations of species with low measurement efficiencies.
-Example: G.
-vaginalis associations are not detected by several studies that use certain V13 primers that simply don't amplify this species well.
-This is a case where the precision in estimated associations is dramatically reduced, by a different mechanism than what is described above.
 

### 2021-09-09 real-world-implications.Rmd
Move notes to temporary notes file

@@
 
-- TODO: Address DA analysis of aggregate (or synthetic) taxa, such as phyla, which may be prone to errors due to changes in the species-level make-up of the taxon.
+- Todo: Address DA analysis of aggregate (or synthetic) taxa, such as phyla, which may be prone to errors due to changes in the species-level make-up of the taxon.
 

### 2021-09-09 revision-notes.Rmd
Move notes to temporary notes file

--- 2021-09-07 solutions.Rmd
@@
+# Notes from September 2021 revision
+
+# How taxonomic bias affects abundance measurements {#abundance-measurement}
+
+- Footnote: "This model is the simplest model...total cell number or density (@gloor2017micr) ^[todo: details about this assumption]."
+- Footnote: "The sample-specific factor...and total sequencing-run output.^[todo: add note about how, as defined, sequencing effort typically goes down with total density because of library normalization; also mention that it 'sequencing effort' is sometimes equated with total read count or library size, which is not what we've done here]."
+- on assumption of conserved efficinecy within speies: Note that bias may not be conserved w/in species; might add and ref discussion of synthetic taxa later on
+
+## Relative abundances (proportions and ratios)
+
+- after defining species ratios, might note: (Although a proportion is also technically a ratio [of a species to a higher-level taxon], here we use 'ratio' to refer specifically to ratios among species.)
+
+- note: the error is not necessarily consistent for ratios among higher-order taxa
+- note: The proportion equation is analogous to Equation \@ref(eq:ratio-error) (ratio case), with the denominator taxon is the entire taxonomic domain and the relevant efficiency is the weighted average of all species.
+In fact, both equations can be derived from a more general equation, where the observed ratio of two (potentially non-atomic) taxa equals the true ratio times the ratios of the mean efficiencies within each taxon.
+- note: the results for ratios of higher-order taxa, and of proportions, can both be seen as stemming from a general result where the error in the ratio is the ratio of the mean efficiency within that taxon.
+- perhaps have a box for this
+
+## Absolute abundances (cell densities)
+
+- note: right now I'm not consistently ref'ing the equations where the results come from
+- note about ratio-based density inference: see note left in the abundance-measurement.Rmd
+
+# How taxonomic bias affects differential-abundance analysis {#differential-abundance}
+
+- Maybe: Add a final sentence summary that the constant multiplicative error in ratio-based abundance estimates does not affect DA measurement, but the varying error induced by variation in the mean efficiency affects DA analysis on proportion-based estimates.
+
+## Change between a pair of samples
+
+## Regression analysis of many samples
+
+- Important complications to mention:
+  - Error in total density or targeted density estimates
+  - Counting error (so far ignored) can lead to missed associations of species with low measurement efficiencies.
+Example: G.
+vaginalis associations are not detected by several studies that use certain V13 primers that simply don't amplify this species well.
+This is a case where the precision in estimated associations is dramatically reduced, by a different mechanism than what is described above.
+
 # Potential solutions {#solutions}
-
-The theoretical results from Section \@ref(differential-abundance) suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.
-We present five that seem particularly promising:
-The first three are method for correcting or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.
 
@@
 ## Calibrate compositions using community controls {#calibrate-compositions}
-
-The most complete but experimentally challenging approach to mitigating taxonomic bias involves directly measuring the efficiencies of individual species using control samples (@mclaren2019cons).
-Suitable controls are any community sample whose composition (species identity and relative abundances) has been previously characterized, either by construction (in the case of constructed 'mock' communities) or by measurement with a _reference protocol_ chosen to serve as a measurement standard.
-These controls, when measured alongside the target experimental samples, can be used to estimate the relative efficiencies among the superset of species across the control samples.
-The estimated efficiencies can then be used to calibrate the measured relative abundances among the control species in the target samples to the chosen measurement standard, by simply dividing the species' read counts by their corresponding efficiencies or using a more sophisticated statistical modeling framework.
-These calibrated compositions can then be used for arbitrary downstream microbiome analyses, including DA analysis.
-Calibration of species not in the controls requires first imputing their efficiencies from the control measurements, for example using phylogenetic relationships, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
-Though such imputation is conceptually straightforward to implement, its reliability and utility remain untested.
 
@@
 
-The DA analyses we consider don't need accurate estimates of relative abundances within individual samples---they only require accurate fold changes across samples.
-Section \@ref(differential-abundance) showed that fold changes in species ratios and ratio-based density estimates can remain accurate even when taxonomic bias distorts the values in individual samples.
-Hence a solution is to use to analyze variation in ratios and ratio-based density estimates.
-For the latter, practical options include
-
-- spike-ins of extraneous species at constant and/or known density
-- housekeeping taxa, stipulated or computationally identified as having a constant density across samples
-- targeted measurement of specific taxa
-
-TBD what further discussion or illustration to have here.
-
-### notes
-
 For spike-ins, a crucial consideration may be whether the spike-in is added pre- or post-DNA extraction (ref appendix, Harrison).
@@
 
-In some cases, we may still which to use a proportion-based approach to analyzing variation in density.
-For example, in some systems measuring total cell concentration may be easier than implementing a spike-in experiment or determining a set of reference species to directly measure for which at least one species is present in all samples.
-In this case, targeted measurements of one or more reference species can be used to validate and/or correct the fold changes or regression coefficients infered from the proportion-based densities.
-Recall that the error in proportion-based log fold changes (LFCs) is shared among species and is equal to the inverse LFC fold change in mean efficiency (Equations \@ref(eq:prop-fc-error) and \@ref(eq:regression-error); Figure \@ref(fig:regression-example)D).
-Consequently, an accurate LFC of one or more reference species can serve as a general validation or used to correct the LFCs of all species.
-
-- NOTE: I'm not sure if I really understand this example below, since I don't see how this interpretation of the Table 1 results mesh with Figure 4
-
-We illustrate this idea with an example from the study by @lloyd2020evid described in Section \@ref(implications), which estimated doubling times for various taxa from the measured log increase in cell density with burial time in marine sediments.
-The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for several bacterial and archeal taxa (CHECK).
-In Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years (CHECK) using the proportion-based densities and 10.3 years by qPCR (@lloyd2020evid Table 1 and Figure 4).
-These doubling times correspond to highly similar estimated slopes of log density per year of $1/10.1 \approx 0.99$ and $1/10.3\approx 0.97$.
-The difference between these slopes, $\approx 0.02$, estimates the slope of the mean efficiency.
-It's small value suggests little systematic variation in mean efficiency with burial time and hence little systematic bias in the estimated doubling times derived from the proportion-based density estimates.
-In this case, close agreement in the DA results across methods for a specific taxon suggests that variation in sample mean efficiency did not significantly distort the MGS-based results and hence provides a degree of validation for all taxa, not just the taxa with targeted measurements.
-(An important caveat for this example is that this analysis assumes consistent efficiencies at the higher taxonomic levels used for qPCR measurement and MGS analysis performed by @lloyd2020evid.)
-Here we compared summary statistics; however, in principle a joint statistical analysis of targeted and proportion-based density measurements would provide more statistical power to estimate and correct the effects of variation in the mean efficiency across samples and provide calibrated fold change and regression estimates for all taxa [APPENDIX].
+- Note: I may not correctly understand the next example, since I don't yet see how this interpretation of Table 1 meshes with Figure 4
 
 ## Perform a bias sensitivity analysis
-
-What can we do for experiments that have been conducted without control communities, targeted control measurements, or spike-ins?
-One approach is to conduct a bias sensitivity analysis to probe the extent and way in which the results of a DA analysis are sensitive to different possible taxonomic biases in our experiment.
-
-A straightforward approach to bias sensitivity analysis is to use computer simulation to randomly generate efficiency vectors, each of which is used to calibrate the MGS measurements.
-These calibrated MGS measurements represent the true compositions under the hypothesized taxonomic bias. 
-By re-running our DA analysis and comparing results across these simulated datasets, we can begin to understand how our initial results might be influenced by the strength and structure of the taxonomic bias in our experiment.
-
-The simulated efficiency vectors may be generated to reflect certain hypotheses about bias in the given system.
-The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
-Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance and microbiome analyses more generally.
 
@@
 ## Use bias-aware meta-analysis
-
-So far we have considered analysis of samples subject to the same taxonomic bias;
-however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in different groups of samples.
-This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses where the effects of bias might cancel within a study.
-
-A potential solution is to perform a bias-aware meta-analyses that jointly analyzes multiple data from studies while explicitely acknowledges the distinct, unknown taxonomic bias in each study.
-Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model "batch effects" --- amorphous differences in the measurements of each study.
-By configuring these models so that these latent parameters correspond to the study and species specific efficiencies, we can increase the biological interpretability of these models and gain ability to determine whether disageements in DA results across studies are due to differences in taxonomic bias or other differences (such as host cohort or study design).
-Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by @mclaren2019cons, which can increase improve our ability to find reproducible microbial associations in a given meta-study.
-
-### notes
 

### 2021-09-09 solutions.Rmd
Move notes to temporary notes file

@@
 The first three are method for correcting or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.
-
-- Somewhere note that we're considering ideas beyond 'MGS protocol optimization', but that is still important - tuning protocols to reduce taxonomic bias, and to make it as consistent as possible across samples.
-(Our results confirm the importance of both of these)
 
@@
 
-- The above states the gist; can also have a paragraph with description/discussion of how this is/has been used so far, and ongoing developments and challenges. Some bullets for this follow.
-- A mock-community approach may be sufficient for synthetic-community experiments, where all taxa are culturable, and for relatively simple natural communities like the vaginal microbiome that are dominated by a small number of culturable taxa.
-- But suitable mock controls may not be feasible for most complex natural ecosystems and require significant effort to develop.
-- An closely-related alternative to mocks are controls derived from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (@mclaren2019cons).
-- A natural fecal standard is currently being developed by [NIST](https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material) and one has recently been made available commercially by Zymo Research ([ZymoBIOMICS Fecal Reference](https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology)).
-- Ensuring accurate quantification, stability, homogeneity of such controls remains a major challenge.
-- Careful testing is also needed to ensure that the preparation and storage of the controls has not significantly affected the taxonomic bias relative to the experimental samples in a given application.
-- As it is feasible to characterize a single standard much more extensively than a typical community sample, we may be able to obtain an estimated composition we feel comfortable treating as the ground truth.
-- Yet even when this is not possible, such natural standards can allow us to reconcile results across studies despite not knowing the truth.
-- Given the significant uncertainty about estimated efficiencies (stemming from all the above raised caveats), rather than as providing calibration to the 'truth', the estimated efficiencies might be best seen as priors or hypotheses to inform a bias-sensitivity analysis of the sort described below.
-- Should mention the two extant cases where mocks of relevant taxa have been created: The vaginal-HMP / MOMS-PI study (@brooks2015thet and @fettweis2019thev) and the @leopold2020host plant-fungal experiment.
-Note that even in these cases, calibration hasn't been much used, partly because of a lack of workflows or understanding of when and why it is necessary.
-Ideally we can give an illustration; this section can be a good use of the @leopold2020host change-over-time result, though I'm not sure whether to discuss that here or in the 'real-world' section.
+- The above states the approach; can add description/discussion of application to the MOMS-PI and @leopold2020host cases and developments and challenges in availability of control communities
 
@@
 
-### notes
-
-For spike-ins, a crucial consideration may be whether the spike-in is added pre- or post-DNA extraction (ref appendix, Harrison).
-Similarly, we must consider whether targeted measurements are made pre extraction (e.g. CFU counting or ddPCR directly on cells) or post extraction (e.g. ddPCR or qPCR of DNA).
-If DNA yield is linear, then either approach should give reliable fold changes; but if DNA yield is non-linear (for example, saturating in higher biomass samples) then systematic errors can still arise with the post-extraction approach, though these may still be reduced compared to proportion-based inference (REF appendix).
-(Why expect to be reduced: This saturation will still be less than what would be imposed by library normalization, though I need to think about this a bit.)
-In additional, substantial random variation in the (log) DNA yield of a sample with a given biomass will reduce the precision of the estimated abundances that is possible with the post-extraction approach in a manner that cannot be circumvented by increasing the number of spike-in or targeted species.
-
-Notes on implementation:
-
-- Targeted measurement 
-- Although solutions can be hacked together, ideally we'd have new statistical tools to jointly model MGS and supplemental measurements or spike-ins and housekeeping taxa.
-- These would account for the uncertainty associated with the targeted measurements, including detection limits, and could naturally allow for handling the fact that targeted taxa may not be present in every sample.
-- The appendix gives some theoretical insight into how to deal with a given reference taxon not being present or detectable in every sample
-
 ## Calibrate fold changes using measurements of targeted taxa
@@
 
-- NOTE: I'm not sure if I really understand this example below, since I don't see how this interpretation of the Table 1 results mesh with Figure 4
+- Note: I may not correctly understand the next example, since I don't yet see how this interpretation of Table 1 meshes with Figure 4
 
@@
 
-- Aim to illustrate with a simulated or real example
-
 ## Use bias-aware meta-analysis
@@
 A potential solution is to perform a bias-aware meta-analyses that jointly analyzes multiple data from studies while explicitely acknowledges the distinct, unknown taxonomic bias in each study.
-Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model "batch effects" --- amorphous differences in the measurements of each study.
+Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model "batch effects"---amorphous differences in the measurements of each study.
 By configuring these models so that these latent parameters correspond to the study and species specific efficiencies, we can increase the biological interpretability of these models and gain ability to determine whether disageements in DA results across studies are due to differences in taxonomic bias or other differences (such as host cohort or study design).
 Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by @mclaren2019cons, which can increase improve our ability to find reproducible microbial associations in a given meta-study.
-
-### notes
-
-When the goal is to perform a meta-analysis that combines studies that have used different protocols, the unknown measurement efficiencies of each protocol can be explicitly included as parameters of the statistical model that is used.
-Unknown efficiencies can be included in "compositional" linear modeling frameworks such as ALDEx2, DivNet, and fido simply by adding a protocol-specific term to the linear model of taxon log ratios.
-[*Add sentence on how this can be used for non-compositional analyses.*]
-Thus such bias-aware meta-analyses are already technically feasible and—if bias is truly consistent within a protocol or study—may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not model bias explicitly.
-
-- Aim to  illustrate with a simulated or real example
-
-- Value: If results of individual studies might be fairly robust to unknown bias, then might be able to efficiently (from a statistical POV) pool studies when they biologically agree or rule out bias (under the used assumptions) as the cause of disagreement when they don't agree.

### 2021-09-09 abundance-measurement.Rmd
Change method of rendering SVGs in PDF output
Previously, I used the cowplot R package to display the SVG images in
the PDF rendering; however, this requires the 'magick' Latex package and
does not always seem to give good results. I now use Inkscape to convert
the SVG to a PDF file in a temporary directory and import this figure
with knitr.

@@
 } else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
+    "error-proportions.pdf")
+  args <- c(
+    "--export-area-page", 
+    "--export-type=pdf", 
+    str_c("--export-filename=", pdf_path),
+    svg_path
+  )
+  system2("inkscape", args = args)
+  knitr::include_graphics(pdf_path)
 }
-rm(svg_path)
+rm(svg_path, pdf_path)
 ```

### 2021-09-09 differential-abundance.Rmd
Change method of rendering SVGs in PDF output
Previously, I used the cowplot R package to display the SVG images in
the PDF rendering; however, this requires the 'magick' Latex package and
does not always seem to give good results. I now use Inkscape to convert
the SVG to a PDF file in a temporary directory and import this figure
with knitr.

@@
   "notebook/_posts/2021-08-03-simulate-regression-example/",
-  "simulate-regression-example_files/figure-html5/",
+  "simulate-regression-example_files/figure-html5",
   "main-figure-1.svg"
@@
 } else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
+    "regression-example.pdf")
+  args <- c(
+    "--export-area-page", 
+    "--export-type=pdf", 
+    str_c("--export-filename=", pdf_path),
+    svg_path
+  )
+  system2("inkscape", args = args)
+  knitr::include_graphics(pdf_path)
 }
-rm(svg_path)
+rm(svg_path, pdf_path)
 ```

### 2021-09-15 index.Rmd
Revise title
Add git head checksum to preface

@@
 ---
-title: "Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for"
+title: "Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements"
 author:
@@
 
+```{r}
+repo <- git2r::repository(".")
+sha <- repo %>% git2r::commits(n = 1) %>% .[[1]] %>% git2r::sha()
+```
+
+_This manuscript was rendered from commit `r sha`._
+
 ::: {#preface-warning .rmdnote}

### 2021-09-15 v/6932587bd1ea9c56638c0ebe3ff7885b79b84e8c/_main.tex
Add git head checksum to preface
This commit adds the rendered manuscript associated with
6932587bd1ea9c56638c0ebe3ff7885b79b84e8c.

--- 2021-08-05 v/0f596f09880fb209cd5a6b87bff16da6d7418ea0/_main.tex
@@
 ]{article}
-\title{Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for}
+\title{Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements}
 \author{Michael R.
@@
 Callahan\footnote{North Carolina State University}}
-\date{2021-08-04}
+\date{2021-09-15}
 
@@
 \hypersetup{
-  pdftitle={Taxonomic bias can create spurious results in microbiome differential-abundance analyses unless it is properly accounted for},
+  pdftitle={Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements},
   pdfauthor={Michael R. McLaren; Karen G. Lloyd; Benjamin J. Callahan},
@@
 
+\emph{This manuscript was rendered from commit 6932587bd1ea9c56638c0ebe3ff7885b79b84e8c.}
+
 \leavevmode\vadjust pre{\hypertarget{preface-warning}{}}%
@@
 
-Most microbiome research aims to go beyond qualitative descriptions to make quantitative conclusions about associations between specific microbes or community states and key host or environmental properties.
-But there is serious disagreement about whether the microbiome measurements made by marker-gene and metagenomic sequencing (MGS) are or can ever be truly quantitative.
-A primary reason is that taxa vary in how they respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
-As a result taxa can differ dramatically (e.g.~10-1000X) in how efficiently they are measured---that is, converted from cells into taxonomically classified sequencing reads---making the relative abundances we observe by MGS inaccurate representations of the actual sample compositions.
-Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads among taxa make this taxonomic bias a feature of shotgun metagenomic measurements as much as marker-gene measurements.
-Taxonomic bias is protocol specific (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})) and can even vary among batches within an experiment (\protect\hyperlink{ref-yeh2018taxo}{Yeh et al.} (\protect\hyperlink{ref-yeh2018taxo}{2018})); it thus not only makes MGS measurements inaccurate, but also incomparable between studies.
-Taxonomic bias thereby poses a major challenge for any seeking to draw quantitative conclusions from MGS studies.
-
-The field's current answer to this challenge is methods standardization, prioritizing consistency over accuracy.
-By treating each sample within a study in exactly the same manner, we can ensure that observed differences between samples are not simply due to methodological differences.
-Yet standardization does not guarantee that these observed differences reflect the actual differences in taxonomic composition among samples.
-Consider what might be the simplest and most common inference made from MGS: How does the proportion of a taxon vary between samples?
-An observation that \emph{Escherichia} coli makes up 10\% of the sequencing reads in a fecal sample might, thanks to bias, have arisen from a sample in which the proportion of \emph{E.
-coli} is actually 50\% or 1\%.
-Yet many hope that if we observe the proportion of \emph{E.
-coli} to increase from 10\% to 20\% between two samples, its proportion must have actually doubled---by fixing the taxonomic bias of our measurements through standardization, we have (so the argument goes) fixed the measurement error in both samples so that it will cancel when estimate the fold change.
-Unfortunately, the results of \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) show that reality is not so neat.
-Because the taxa compete for sequencing effort within a sample, the effect that taxonomic bias has on the observed proportions depends on the taxonomic composition of the sample.
-As a result, MGS measurements can lead to spurious inferences about changes across samples even when taxonomic bias is constant for each taxon across samples.
-
-This problem posed by taxonomic bias intersects with a second, growing concern within the microbiome field: that the relative abundances measured by MGS can present a misleading view of absolute-abundance dynamics.
-Without making further assumptions (which are often difficult or impossible to verify) one cannot tell whether a doubling in the proportion of a taxon reflects an increase in its abundance or a decrease in the abundances of other taxa.
-This concern has motivated the development of various experimental methods to enable researchers to convert the relative abundances measured by MGS into absolute abundances (methods to be reviewed in the Appendix in a future version).
-An increasingly popular approach in the study of both environmental and host-associated microbiomes involves making an independent measurement of total microbial abundance in each sample (\protect\hyperlink{ref-props2017abso}{Props et al.} (\protect\hyperlink{ref-props2017abso}{2017}),\protect\hyperlink{ref-kevorkian2018esti}{Kevorkian et al.} (\protect\hyperlink{ref-kevorkian2018esti}{2018}),\protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}),\protect\hyperlink{ref-tettamantiboshier2020comp}{Tettamanti Boshier et al.} (\protect\hyperlink{ref-tettamantiboshier2020comp}{2020}),\protect\hyperlink{ref-contijoch2019gutm}{Contijoch et al.} (\protect\hyperlink{ref-contijoch2019gutm}{2019}),\protect\hyperlink{ref-vandeputte2017quan}{Vandeputte et al.} (\protect\hyperlink{ref-vandeputte2017quan}{2017}),\protect\hyperlink{ref-vieirasilva2019quan}{Vieira-Silva et al.} (\protect\hyperlink{ref-vieirasilva2019quan}{2019})).
-This estimate is then multiplied by the MGS proportions to convert them to estimates of absolute abundance.
-This simple calibration procedure directly transfers the error in the MGS proportions due to taxonomic bias to the estimated absolute abundances, leading to estimates that are not only individually inaccurate but can imply inaccurate variation in abundance across samples.
-
-Here we summarize recent advances in our understanding of how taxonomic bias affects our ability to accurately determine how taxa vary in relative and absolute abundance across samples from different conditions---the inference problem known as \emph{differential abundance (DA) analysis}.
-We first explain how taxonomic bias can lead to spurious DA results and consider the biological conditions in which such spurious results are likely.
-But we then describe several experimental and computational methods that can be used to avoid or correct these errors.
-Collectively, these methods may provide practical solutions to mitigating the effects of taxonomic bias on DA analysis for a majority of microbiome studies.
-
-\hypertarget{bias-can-cause-spurious-results-in-proportion-based-da-analysis}{%
-\section{Bias can cause spurious results in proportion-based DA analysis}\label{bias-can-cause-spurious-results-in-proportion-based-da-analysis}}
-
-\begin{verbatim}
-#> Warning: It is deprecated to specify `guide = FALSE` to remove a guide. Please
-#> use `guide = "none"` instead.
-
-#> Warning: It is deprecated to specify `guide = FALSE` to remove a guide. Please
-#> use `guide = "none"` instead.
-\end{verbatim}
+The most basic question we can ask about microbial communities after which taxa are found in an ecosystem is, how do the abundances of these taxa vary in their abundance---across space, time, and host or environmental conditions?
+Advances in sequencing technology allow us to now simultaneously measure the abundances of 100s to 1000s of species using marker-gene and shotgun-metagenomic sequencing (jointly, MGS).
+Although standard MGS measurements lose information about total microbial density---and so are typically used to analyze the abundances of taxa relative to each other or their total---new studies are increasingly employing strategies to enable the analysis of cell density or other measures of ``absolute abundance.''
+These relative and absolute abundances serve as the basis for a \emph{differential-abundance (DA) analysis}, in which the change in abundance of a microbial taxon across samples or conditions is used to learn about the biology of the taxon and its impact on the host and other microbes as well as to detect predictive biomarkers of host and environmental health and disease.
+
+Although MGS-based DA analysis has been widely deployed and achieved many notable successes, it faces serious concerns over accuracy and reproducibility due to the inherent technical limitations of MGS measurements.
+In particular, MGS measurements are \emph{taxonomically biased}: Taxa vary dramatically (e.g.~10-1000X) in how efficiently they are measured---that is, converted from cells into taxonomically classified sequencing reads---by a given MGS protocol.
+As a result, the abundance measurements obtained by MGS are inaccurate representations of the actual abundances and also tend to differ across protocols (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})), studies, and even experimental batches (\protect\hyperlink{ref-yeh2018taxo}{Yeh et al.} (\protect\hyperlink{ref-yeh2018taxo}{2018})).
+This bias arises from variation in how taxa respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
+Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads make taxonomic bias a feature of both shotgun and marker-gene measurements.
+The error it causes have been found to in some cases to supersede sizable biological differences (e.g.
+\protect\hyperlink{ref-lozupone2013meta}{Lozupone et al.} (\protect\hyperlink{ref-lozupone2013meta}{2013})) and has plausibly caused replication failures for prominent findings such as the association of decreased Bacteroides and increased Firmicutes in stool with obesity (\protect\hyperlink{ref-finucane2014atax}{Finucane et al.} (\protect\hyperlink{ref-finucane2014atax}{2014})) and the association of certain taxa in the vagina of pregnant women with preterm birth (\protect\hyperlink{ref-callahan2017repl}{Callahan et al.} (\protect\hyperlink{ref-callahan2017repl}{2017})).
+
+The typical approach to countering taxonomic bias in DA analysis is to standardize the measurement protocol used within a given study.
+In broad strokes, the thinking is that the measurements of samples measured by the same protocol will be affected by bias in the same way and so the inferred differences between samples (the focus of DA analysis) will be unaffected.
+For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples.
+\protect\hyperlink{ref-kevorkian2018esti}{Kevorkian et al.} (\protect\hyperlink{ref-kevorkian2018esti}{2018}) and \protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) were the first (to our knowledge) to make this argument explicit mathematically; however, \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) used a theoretical model (supported by measurements of defined bacterial communities) to show that consistent taxonomic bias can lead to variable fold errors in measured proportions.
+These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase).
+Yet \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that certain changes in relative abundance---in particular, fold changes in the ratios among species---are robust to the effects of bias.
+The implications of these findings for changes in absolute abundance---which remain subject to taxonomic bias in the underlying MGS measurement---and for DA analysis across many species and many samples---as commonly done in microbiome association testing---have yet to be investigated.
+
+Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundance.
+Our analysis clarifies how the folk wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
+Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.
+Over time, application of these methods to past and future experiments will provide crucial quantitative information about taxonomic bias and the conditions under which spurious results arise for various DA methodologies.
+Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.
+
+\hypertarget{abundance-measurement}{%
+\section{How taxonomic bias affects abundance measurements}\label{abundance-measurement}}
+
+To understand how bias affects the measured differential abundance between samples, we first consider how it affects measurements of individual samples.
+
+Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}).
+This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the \emph{compositional} nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (\protect\hyperlink{ref-gloor2017micr}{Gloor et al.} (\protect\hyperlink{ref-gloor2017micr}{2017})).
+In this model, we suppose there are a set of \emph{atomic taxa}, that differ in how efficiently they are measured but within which all cells behave similarly.
+For concreteness, we equate atomic taxa with species.
+We ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.
+
+Our model stipulates that the assigned read count of species \(i\) in sample \(a\) equal equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
+\begin{align}
+  \label{eq:measurement-model}
+  \text{reads}_{i}(a)
+  = \text{density}_{i}(a) \quad \cdot
+    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
+    \cdot \quad
+    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
+\end{align}
+The species-specific factor is the \emph{relative measurement efficiency} (or \emph{efficiency} for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+The sample-specific factor, which we call the \emph{sequencing effort} of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
+Equation \eqref{eq:measurement-model} implies that the total reads from all species in the sample equal
+\begin{align}
+  \label{eq:total-reads}
+  \text{total reads}(a)
+    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
+\end{align}
+where
+\begin{align}
+  \label{eq:mean-efficiency}
+  \text{mean efficiency}(a) 
+  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
+\end{align}
+is the average or mean efficiency of cells in the sample.
+
+\hypertarget{relative-abundances-proportions-and-ratios}{%
+\subsection{Relative abundances (proportions and ratios)}\label{relative-abundances-proportions-and-ratios}}
+
+The term \emph{relative abundance} carries multiple meanings in microbiome science.
+Microbiome researchers often use the relative abundance of a species in a sample to refer to its \emph{proportion}, or number of cells from that species divided by the total number of cells from the taxonomic domain of interest (e.g.~Prokaryotes).
+A second view, drawn from the statistical framework of Compositional Data Analysis, equates the relative abundances of species within a sample with the \emph{ratios} of cell densities among two or more species (e.g.~a ratio of 5:4:2 among three species).
+By this view, it is meaningless to refer to the relative abundance of an individual species.
+Although other interpretations exist\footnote{e.g.~the rank order within a sample, or value associated with a species after a centered-log-ratio transform}, we focus on the proportion and ratio views as they underly most DA analysis methods (and microbiome analysis in general), for both relative and absolute abundances, and the differences between them help to clarify the issue over when and why the effects of bias cancel in an analysis.
+
+The standard estimate of the proportion of species \(i\) in sample \(a\) given by the ratio of its read count to the total,
+\begin{align*}
+  \hat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
+\end{align*}
+The effect of taxonomic bias on this estimate is to create a multipliative error equal to the ratio of the species' efficiency to the mean efficiency in the sample,
+\begin{align}
+  \label{eq:prop-error}
+  \hat{\text{prop}}_{i}(a)
+  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)}.
+\end{align}
+Intuitively, Equation \eqref{eq:prop-error} says that a species is measured as having a larger-than-actual proportion in direct relation to how much more more efficiently its cells are measured than the average cell in the sample (Figure \ref{fig:error-proportions}).
+Consequently, the same species can be over or under measured in terms of its proportion.
+For instance, in samples from two hypothetical communities in Figure \ref{fig:error-proportions}, Species 3 has an efficiency of 6 and is under-estimated in Sample 1 (which has a mean efficiency of 8.33) but over-estimated in Sample 2 (which has a mean efficiency of 3.15).
+
+The estimated ratio between species \(i\) and \(j\), given by the ratio of read counts,
+\begin{align*}
+  \hat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)},
+\end{align*}
+has a multiplicative error equal to the ratio in the species' efficiencies,
+\begin{align}
+  \label{eq:ratio-error}
+  \hat{\text{ratio}}_{i/j}(a)
+%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
+  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
+\end{align}
+Equation \eqref{eq:ratio-error} says that the ratio between two species is over-estimated in direct relation to how much more efficiently cells of the first species is measured relative to the second.
+In contrast to estimates of proportions, estimates of species ratios have a consistent multiplicative error across samples.
+For instance, in the example of Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{/tmp/bookdown-differential-abundance-theory/error-proportions} \caption{\textbf{Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.} Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species.
+Bottom row: Error in the estimated fold-change in the third species that is derived from these measurements.}\label{fig:error-proportions}
+\end{figure}
+
+\hypertarget{absolute-abundances-cell-densities}{%
+\subsection{Absolute abundances (cell densities)}\label{absolute-abundances-cell-densities}}
+
+Various experimental and computational methods have or can be used to supplement MGS measurements so as to convert the relative-abundance information in the read counts into estimates of (absolute) cell densities;
+however, these methods can largely be partitioned conceptually into those that use the estimated proportions or the estimated ratios provided by MGS.
+The error in density estimates due to taxonomic bias therefore follows that of the proportion and ratio estimates described above.
+
+Proportion-based density estimation is based on the idea that if we know the (true) proportion of a species and the (true) total density of all species, we can multiply these numbers obtain the species' density.
+This method therefore involves multiplying the proportion of a species estimated from MGS with an estimate of total microbial density obtain through other means (e.g., through direct cell counting or qPCR of a universal marker gene) to estimate
+its density,
+\begin{align}
+  \label{eq:density-prop-est}
+  \hat{\text{density}}_{i}(a) 
+  = \hat{\text{prop}}_{i}(a) \cdot \hat{\text{total density}}(a).
+\end{align}
+The error in this proportion-based density estimate is
+\begin{align}
+  \label{eq:density-prop-error}
+  \hat{\text{density}}_{i}(a) 
+  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
+  \cdot \substack{\text{fold error in} \\ \hat{\text{total density}}(a)}.
+\end{align}
+The multiplicative error consists of two factors: the error in the estimated proportion (due to taxonomic bias in the MGS measurement), and any error in the estimated total density.
+In the case of perfect knowledge of the total density, the fold error in proportion-based density estimates varies across samples inversely with the mean efficiency.
+
+Ratio-based density estimation follows from the fact that, in the absence of taxonomic bias, the conversion rate from reads to density is just the sequencing effort of the sample and is thus the same for all species (Equation \eqref{eq:measurement-model}).
+Knowing the density of at least one \emph{reference} species therefore enables us to convert from reads to density for all species.
+One approach to obtaining such reference species is to add one or more extraneous species in known (typically constant) abundance (a ``spike-in'') to each sample prior to sequencing (REFs).
+A second approach is to determine a set of naturally occurring species that are thought to have a constant, though typically unknown, abundance across samples; in this case we can only estimate species' densities relative to these ``housekeeping species'' (REFs).
+A third (seemingly untried) possibility is to measure the density of one or more naturally occurring species using a method such as targeted CFU counting, fluorescence flow cytometry or ddPCR directly on cells, or ddPCR or qPCR on extracted DNA.
+Given the estimate of the density of reference taxon \(r\), we estimate the density of taxon \(i\) by multipling the ratio of \(i\) to \(r\) in the reads by the density of \(r\),
+\begin{align}
+  \label{eq:density-ratio-est}
+  \hat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \hat{\text{density}}_{r}(a)
+\end{align}
+(Appendix ??? describes how this equation can be extended to multiple reference species.)
+For a constant reference species with unknown density, we can set \(\hat{\text{density}}_{r}(a)\) equal to 1 to estimate density in units of the density of the reference species.
+The error in this ratio-based density estimate is
+\begin{align}
+  \label{eq:density-ratio-error}
+  \hat{\text{density}}_{i}(a) 
+  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
+  \cdot \substack{\text{fold error in} \\ \hat{\text{density}}_{r}(a)}.
+\end{align}
+Again, the multiplicative error consists of two factors: the error in the estimated ratio (due to taxonomic bias in the MGS measurement), and error in the estimated density of the reference taxa.
+So long as the fold error in the reference density is constant across samples, so will be the error in the estimated density of the focal species \(i\).
+
+\begin{itemize}
+\tightlist
+\item
+  Note: The current distinction between proportion- and ratio-based density estimation is incomplete, as the example of spike-in normalization shows: Whether the density of species \(i\) is estimated by Equation \eqref{eq:density-ratio-est}, or by first estimating total density from the ratio of non-spike-in to spike-in reads and then using Equation \eqref{eq:density-prop-est}, the results are mathematically identical.
+Future revisions should clarify this point.
+\end{itemize}
+
+\hypertarget{differential-abundance}{%
+\section{How taxonomic bias affects differential-abundance analysis}\label{differential-abundance}}
+
+How do the measurement errors described in the previous section impact our ability to estimate the changes in microbial abundances across samples or between different host and environmental conditions?
+Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, as such inference is ubiquitous in microbiome DA analysis and have more direct ecological interpretations (via the processes of exponential growth and death) than other DA measures.
+These measures of DA are also affected by taxonomic bias in our model in a manner that is straightforward to understand in terms of the multiplicative errors described in the previous section.
+
+\hypertarget{change-between-a-pair-of-samples}{%
+\subsection{Change between a pair of samples}\label{change-between-a-pair-of-samples}}
+
+Before considering the stereotypical many-sample DA analysis, it is instructive to
+consider simplest analysis of differential abundance: the estimation of fold changes in abundance between a pair of samples.
+This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment, and conceptually bridges the single-sample results of the previous section to the many-sample case.
+
+The composition-dependent effect of bias on fold error in proportions leads to error in fold-change estimates that is proportional to the inverse change in mean efficiency.
+From the error in an individual sample \eqref{eq:prop-error}, it follows that the estimated fold change in proportion of taxon \(i\) from sample \(a\) to sample \(b\) is
+\begin{align}
+  \label{eq:prop-fc-error}
+% \tag*{Fold change in proportion}
+\underbrace{\frac{\hat{\text{prop}}_{i}(b)}{\hat{\text{prop}}_{i}(a)}} _\text{estimated FC}
+  &= \frac
+    {\text{prop}_{i}(b) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(b)}}
+    {\text{prop}_{i}(a) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(a)}}
+\\[0.5ex]
+  &=
+  \underbrace{\frac{\text{prop}_{i}(b)}{\text{prop}_{i}(a)}}_\text{actual FC}
+  \cdot
+  \underbrace{\left[\frac{\text{mean efficiency}(b)}{\text{mean efficiency}(a)}\right]^{-1}}_\text{fold error}
+  .
+\end{align}
+The sample-independent efficiency factor of the error cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency from \(s\) to \(t\).
+
+The bottom row of Figure \ref{fig:error-proportions} illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
+The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
+Consequently, the FC of each species is measured to be 2.6X larger than the true value.
+Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
+We can see each type of error in Figure 2.
+For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the estimated FC (actual FC: 2.3X, measured FC: 6.5X).
+For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
+For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!
+
+In contrast, because species ratios are distorted by a constant factor, their measured fold changes remain accurate.
+The fold error in Equation \eqref{eq:ratio-error} completely cancels when we divide the ratio measured for one sample \(a\) by another sample \(b\).
+
+If other error sources remain negligible, than this dichotomy continues to apply to proportion- and ratio-based density estimates.
+The inferred FCs in proportion-based density estimates will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
+Moreover, any species with a constant density will appear to vary inversely with the mean efficiency.
+In contrast, fold changes in ratio-based density estimates remain accurate.
+
+\hypertarget{regression-analysis-of-many-samples}{%
+\subsection{Regression analysis of many samples}\label{regression-analysis-of-many-samples}}
+
+DA analysis of many samples across host or environmental conditions can typically be framed as a regression problem, in which we analyze the relationship between a microbial \emph{response variable}, such as log density of some focal species \(i\), and one or more \emph{covariates}, such the pH or temperature of the sampled environment or whether the sample is from a healthy or sick person.
+A substantial fraction of DA analyses use minor elaborations on the simple linear regression model, which for a response of log species density can be written
+\begin{align}
+  \label{eq:regression}
+  \log \text{density}_i(a) = \alpha + \beta x(a) + \varepsilon_i(a).
+\end{align}
+Here \(x\) is a continuous covariate (e.g.~pH) or a binary covariate (e.g.~\(x=1\) for treated patients and \(x=0\) for controls), \(\alpha\) and \(\beta\) are regression coefficients, and \(\varepsilon_i(a)\) is a mean-zero random variable that reflects the residual (unexplained) variation in the response (log density of species \(i\)).
+Our interest is usually in the coefficient \(\beta\) (slope or average difference between conditions) that describes how the species' abundance changes with \(x\), while the intercept \(\alpha\) captures differences in the baseline abundance and---we hope---measurement efficiency among species.
+How does taxonomic bias under our measurement model affect estimates of \(\beta\) in the simple linear regression?
+
+Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \eqref{eq:density-prop-est}) with error-free estimates of the total density.
+If the true log density follows the regression Equation \eqref{eq:regression}, then it follows from Equation \eqref{eq:density-prop-error} the estimated log density equals
+\begin{align}
+  \label{eq:regression-error}
+  \log \hat{\text{density}}_i(a)
+%  &= \alpha + \beta x(a) + \varepsilon_i(a) + \log \text{efficiency}_i - \log \text{mean efficiency}(a)
+  = [\alpha + \log \text{efficiency}_i] + [\beta - \log \text{mean efficiency}(a)] x(a) + \varepsilon_i(a).
+\end{align}
+This equation shows that the species-specific portion of the error affects the intercept term while the sample-specific portion (log mean efficiency) affects the slope term.
+Thus, as in the case of fold changes between two samples (Equation \eqref{eq:prop-fc-error}), it is the variation in the (log) mean efficiency across samples that we must worry about distorting our DA results.
+
+The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient \(\beta\).
+Box \ref{regression-error} mathematically describes this effect for estimation method of Ordinary Least Squares (OLS) and Maximum Likelihood Estimation (MLE) and Figure \ref{fig:regression-example} illustrates using simulated data.
+Variation in the log mean efficiency that is associated with the covariate \(x\) creates a systematic error in the estimated slope \(\hat \beta\) equal to the negative of the (scaled) covariance of log mean efficiency with \(x\).
+The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with \(x\) relative to that of the response (here, \(\log \text{density}_{i}\)) with \(x\) or, equivalently, the relative magnitudes of their slopes.
+As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \ref{fig:regression-example}), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
+Variation in the log mean efficiency that is uncorrelated with \(x\) does not systematically distort \(\hat \beta\) but does affect its precision, typically leading to increased standard errors as the variation in log mean efficiency effectively acts as an additional source of noise in measured abundance (Figure \ref{fig:regression-example} D).
+The exception is for species whose residual variation is strongly positively correlated with that of log mean efficiency (here, Species 9), which can appear to have less random variation and receive standard errors that are too small.
+Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).
+
+With this understanding in place, we briefly summarize how taxonomic bias affects estimation of the simple linear model in other abundance types.
+The results for proportion-based density estimates with accurate total densities also apply to LFC analysis of proportions.
+Similar results apply to microbiome regression tools (such as corncob; \protect\hyperlink{ref-martin2020mode}{Martin, Witten, and Willis} (\protect\hyperlink{ref-martin2020mode}{2020})) that perform regression on the logit (instead of log) proportion of a species; however,
+the mean efficiency of the entire sample must instead be replaced with the mean efficiency among all species excluding the focal species, which causes the absolute error in regression coefficients to vary somewhat across species.
+Because ratios and ratio-based densities are subject to consistent multiplicative error, in analysis of log ratios and the log densities derived from them, only the estimated intercept \(\hat \alpha\) is affected by taxonomic bias, while the point estimate and standard error of the estimated slope \(\hat \beta\) remain unaffected.
 
@@
 \centering
-\includegraphics{_main_files/figure-latex/vaginal-example-1.pdf}
-\caption{\label{fig:vaginal-example}\textbf{Taxonomic bias can distort differential abundance results even when it is consistent for each taxon across samples.} Panel A shows the actual and observed proportions for hypothetical community samples from two time points, which differ in their relative abundance of three taxa.
-Panel B shows taxonomic bias in terms of the relative efficiencies of the three taxa against the mean efficiency of each sample; the difference between the taxon's efficiency and the sample's mean (vertical arrows) determines the fold error seen in Panel A.
-Panel C rearranges the plot from Panel A to show the actual and observed fold changes between time points.
-The efficiencies of individual taxa were estimated by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) from mock community data from \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}).
-The abundances are hypothetical but inspired by observations from the human vaginal microbiome; see main text.}
+\includegraphics{/tmp/bookdown-differential-abundance-theory/regression-example.pdf}
+\caption{\label{fig:regression-example}\textbf{Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.} This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by \(x=0\) and \(x=1\).
+In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFC, which drives the positive association of the log mean efficiency with the condition (shown in Panels A and B).
+This positive LFC in the log mean efficiency induces a systematic negative shift in the estimated LFCs of all species (Panels C and D).
+Panel D shows the mean LFC (points) with 95\% confidence intervals (CIs), for each species estimated from either the actual or the measured densities.
+The error (difference in LFC estimates on measured and actual) equals the negative LFC of the mean efficiency (shown in Panel B).}
 \end{figure}
 
-We illustrate the problems posed by taxonomic bias for differential abundance using an example where the bias of an MGS protocol was explicitly measured using mock communities.
-To analyze the effects of taxonomic bias on measurements of the human vaginal microbiome, \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}) performed 16S rRNA gene sequencing of mock communities of seven key vaginal species using a similar protocol to that of the Vaginal Human Microbiome Project (VaHMP).
-\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) used this data to estimate the relative measurement efficiency of the seven species.
-They found that the most efficiently measured species, \emph{Lactobacillus iners}, had an efficiency that was 29X that of the least efficiently measured species, \emph{Gardnerella vaginalis}, primarily due to a greater DNA extraction efficiency and 16S copy number.
-A third species, \emph{Atopobium vaginae}, had an efficiency 1.8X that of \emph{G. vaginalis}.
-A common aim of human vaginal microbiome studies, including the VaHMP, is to understand the causes and consequences of shifts in communities from being dominated by a particular \emph{Lactobacillus} species to dominance by a non-\emph{Lactobacillus} species such as \emph{G.
-vaginalis}, which have been associated with deleterious health outcomes including an increased risk of preterm births in pregnant women.
-It is therefore important to know whether the efficiency variation found by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) affects the observed differential abundances between \emph{Lactobacillus}-dominated and \emph{Gardnerella}-dominated communities.
-
-Figure \ref{fig:vaginal-example} illustrates the effects of taxonomic bias on the measured proportions for hypothetical vaginal samples from a single woman at two points in time, during which the community shifts from being dominated by \emph{Lactobacillus} to being dominated by \emph{Gardnerella}.
-In each case, we observe a greater proportion of \emph{Lactobacillus} than is truly present, and less of the other two species (Figure \ref{fig:vaginal-example}A).
-This positive multiplicative error arises because \emph{Lactobacillus} is more efficiently measured than the average species in each sample.
-\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed that the observed and actual proportions of a taxon \(i\) in a sample \(s\) are related by the equation
-\begin{align}
-  \label{eq:error}
-  \text{observed}_{i}(s) = \text{actual}_{i}(s) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(s)}
-\end{align}
-where \(\text{mean efficiency}(s)=\text{actual}_j(s)*\text{efficiency}_j\) in the denominator is the average efficiency of all taxa in that sample.
-Intuitively, Equation \eqref{eq:error} says that the taxon will be overrepresented by the extent to which its measurement efficiency is greater than the average efficiency of cells in the sample.
-Figure \ref{fig:vaginal-example}B shows the efficiencies of the three taxa and the mean efficiency in each sample.
-The (multiplicative) difference between the taxon efficiencies and the mean efficiency determines the (multiplicative) error in each sample; for example, the mean efficiency in the second sample is 2.5, leading to a 29/2.5=11.6X error in the proportion of \emph{Lactobacillus}.
-The error in the second sample is larger---specifically, by 6X---than the first sample because the high proportion of \emph{Gardnerella} has reduced the sample mean efficiency (and thus increased the efficiency of \emph{Lactobacillus} relative to the mean) by 6X.
-
-The enrichment depends on the taxon's efficiency relative to the sample average, rather than the efficiency itself, because the requirement that the proportions of all taxa sum to one locks the taxa in a zero-sum competition.
-Notably, competition occurs (though in a less absolute form) even at the level of the raw sequencing reads: DNA extractions and PCR reactions have limited output capacity; high-concentration samples are typically subsampled to fixed, lower concentrations; and sequencing libraries are given limited sequencing effort; so that the experimental process itself typically imposes competition to be sequenced.
-And it is competition that prevents the effects of taxonomic bias from canceling when we analyze how the proportions of individual taxa change across samples.
-
-Figure \ref{fig:vaginal-example}C shows the consequences of measurement error on an analysis of the fold change in proportions between the two samples.
-In reality, \emph{Gardnerella} changes by 4X, \emph{Lactobacillus} by 0.1X, and \emph{Atopobium} by 0.5X.
-Yet what is observed is that \emph{Gardnerella} changes by 32X, \emph{Lactobacillus} by 0.6X, and \emph{Atopobium} by 3X.
-In other words, the change in each case is multiplied by 6X, causing a much larger increase in \emph{Gardnerella} and a smaller decrease \emph{Lactobacillus} than actual, and the appearance that \emph{Atopobium} increased when it in fact decreased.
-Why do such errors occur?
-The multiplicative error in Equation \eqref{eq:error} has two parts: a taxon-specific term in the numerator that is the same in each sample, and a taxon-independent term in the denominator that depends on the composition of the sample through the mean efficiency.
-When we compute the fold change in a taxon i between samples s and t,
-\begin{align}
-  \frac{\text{observed}_{i}(t)}{\text{observed}_{i}(s)}=\frac{\text{actual}_{i}(t)}{\text{actual}_{i}(s)}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{i}}*\frac{\text{mean efficiency}(s)}{\text{mean efficiency}(t)}
-\end{align}
-the constant efficiency in the numerator of Equation \eqref{eq:error} cancels, but the varying mean efficiency does not, leaving an error that is equal to the inverse change in mean efficiency.
-As a result, we are here left inferring that the fold-changes in proportions have all increased less (or decreased more) than they truly have.
-In our example, the mean efficiency decreased by 6X (from 15 to 2.5) due to the shift in dominant taxon from the high-efficiency \emph{Lactobacillus} to the low-efficiency \emph{Gardnerella}, causing the observed fold changes to increase by 6X above the actual.
-This increase distorted the magnitudes of the changes in \emph{Lactobacillus} and \emph{Gardnerella} (decreasing the first and increasing the second) and changed the sign of \emph{Atopobium} from negative to positive.
-Thus partial cancelling in the effect of a constant multiplicative bias may not be sufficient to protect against spurious results.
-
-Spurious results can occur for the same basic reason in proportion-based absolute-abundance inference.
-Suppose that the true total abundance in each sample can be determined experimentally.
-Because the absolute abundance estimate simply multiplies the erroneous observed proportions by this total, the estimated change in absolute abundance is wrong by the same factor as the fold change in proportions.
-In our Figure \ref{fig:vaginal-example} example, if the change in total abundance between samples were 2X, then we would observed both \emph{Lactobacillus} and \emph{Atopobium} to have increased in absolute abundance (by 1.2X and 6X), when in fact \emph{Lactobacillus} decreased (by 5X) and \emph{Atopobium} remained the same.
-
-\hypertarget{effect-on-regression-of-multiple-samples}{%
-\subsection{Effect on regression of multiple samples}\label{effect-on-regression-of-multiple-samples}}
-
-We showed how fold changes of individual samples may be unreliable.
-But microbiome analyses often do not interpret such individual differences, and instead look for average patterns across many samples, such as how the abundance of a taxon varies with some covariate of interest.
-The covariate may be discrete, such as whether the sample is from a healthy or sick person, or it may be continuous, such as a measure of space, time, or temperature.
-Often these analyses can be framed as a regression problem.
-For example, we might hypothesize that the log absolute abundance of taxon \(i\) changes with a variable \(x\) according to the simple linear regression,
-\begin{align}
-  E[\text{log abundance}_{i} \mid x] = a_{0} + a_{1}x
-\end{align}
-where \(x\) is either continuous (e.g., sediment depth) or binary (e.g., \(x=1\) for treated patients and \(x=0\) for controls).
-Much of the effect of bias will be absorbed by the model intercept, which is typically not of interest to the researcher.
-However, variation in the mean efficiency creates additional variation (reducing statistical power to detect variation in the slope) and can also create a statistical bias in our slope estimate equal to the inverse slope of \(E[\text{log mean efficiency}\mid x]\) as a function of \(x\).
-Therefore if the mean efficiency varies across samples, but is not associated with the covariate \(x\), its effect may simply be to create larger standard errors in our slope estimate.
-But the larger worry may be the systematically distorted slope (\(a_1\)) estimates that arise in situations where the mean efficiency changes with \(x\).
-Notably, the error in an absolute sense is the same for all taxa---the slope is always reduced by the same amount---but the implications differ for various taxa depending on their slopes, causing magnitude or sign errors depending how the coefficient derived from the (unknown) true abundances compares to that of the mean efficiency (Figure \ref{fig:regression-example}).
-
-\begin{verbatim}
-#> Warning: Package `magick` is required to draw images. Image not drawn.
-\end{verbatim}
-
-\begin{figure}
-\centering
-\includegraphics{_main_files/figure-latex/regression-example-1.pdf}
-\caption{\label{fig:regression-example}\textbf{Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.} Details to synthesize into a caption: Regression of log2(Proportion) of the commensal fungus Penicillium versus timepoint; timepoints 1 and 2 are pre- and post-challenge with the pathogen Melampsora.
-Data is split by the region the host plants are derived from (Eastern and Western US).
-Calibrated proportions = Observed proportions in the real, experimental samples after adjustment for the bias measured in mock communities.
-Mean efficiency of each community is inferred by treating the calibrated proportions as the truth, and multiplying by efficiencies estimated from the mocks.
-The pathogen Melampsora has a high measurement efficiency; thus once it infects the plants, the mean efficiency of the sample increases (purple points).
-Efficiency is here taken as relative to the focal taxon Penicillium.
-West plants tend to be more resistant to the pathogen, which likely explains why the mean efficiency doesn't increase as much in the West plants.
-Penicillium is observed to decrease in log proportion, in both the East and West plants.
-But the calibrated measurements show that it actually slightly increases in the East plants, and has a lesser decrease in the West plants than what was observed before bias correction.
-The difference between the Calibrated and Observed data points and regression lines equals the regression line of the mean efficiency: Orange = Green + Purple; Green = Orange - Purple.
-The absolute error in regression coefficients is the same for all taxa.
-I picked Penicillium for illustration since it has the smallest observed decrease, which makes the error due to bias have a particularly significant impact.}
-\end{figure}
-
-\hypertarget{implications-for-real-world-inference}{%
-\section{Implications for real-world inference~}\label{implications-for-real-world-inference}}
-
-These observations show that the effects of taxonomic bias do in fact cause error in proportion-based differential abundance analyses.
-Yet if the mean efficiency varies much less across samples than the abundances of individual taxa, this error will be negligible.
-Alternatively, if variation in the mean efficiency is effectively random (not associated with regression covariates), the effect of bias may be to increase the residual variance and thus reduce our statistical power to assay differential abundance---unfortunate to be sure, but perhaps not as much as confidently inferring changes with the wrong sign or magnitude.
-But there are salient biological scenarios where substantial associations between the mean efficiency and covariate can in fact lead to spurious differential abundance inference.
-
-\hypertarget{scenarios-in-which-bias-may-cause-spurious-results}{%
-\subsection{Scenarios in which bias may cause spurious results}\label{scenarios-in-which-bias-may-cause-spurious-results}}
-
-Figure \ref{fig:vaginal-example} illustrates how a change in dominance by from a high- to low- efficiency taxon (or vice versa) can easily create spurious DA results.
-Indeed, this scenario may be particularly common in vaginal microbiome studies.
-Human microbiomes often dominated by \emph{Lactobacillus}, but can become dominated by non-\emph{Lactobacillus} species such as \emph{Gardnerella} vaginalis and \emph{Atopobium vaginae}, and these low-\emph{Lactobacillus} states are associated with the disease bacterial vaginosis and, in pregnant women, an increased risk of preterm birth.
-Using mock community measurements from \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}), \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed that the efficiency of two common \emph{Lactobacillus} species were 20-30X greater than \emph{G.
-vaginalis} and \emph{A.
-vaginae}.
-Therefore we should expect samples dominated by \emph{Lactobacillus} to have a much larger mean efficiency than those those dominated by \emph{G.
-vaginalis} and \emph{A.
-vaginae}, which would distort the observed associations of all taxa with any covariate that is also associated with \emph{Lactobacillus} dominance.
-The extraction and 16S sequencing protocol used by \protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}) is similar to that used in the VaHMP MOMs-PI study also led by these researchers to investigate the associations of these taxa with preterm birth (\protect\hyperlink{ref-fettweis2019thev}{Fettweis et al.} (\protect\hyperlink{ref-fettweis2019thev}{2019})), demonstrating the potential for serious real-world consequences for these statistical/technical concerns.
-
-Similar dynamics occur in a plant-fungal interactions experiment performed by \protect\hyperlink{ref-leopold2020host}{Leopold and Busby} (\protect\hyperlink{ref-leopold2020host}{2020}) (Figure \ref{fig:regression-example}).
-The authors inoculated commensally-colonized trees with a fungal pathogen.
-Using DNA mock communities, they showed that (excluding bias from DNA extraction) the pathogen was 10X more efficiently measured than the median commensal and 40X more efficiently measured than the lowest-efficiency commensal.
-In most hosts, the pathogen increased rapidly between timepoints, driving an increase in the sample mean efficiency that in turn leads to larger decreases being observed in the proportions of commensal taxa than what is predicted using calibrated (bias-corrected) values (REF APPENDIX).
-For example, the calibrated proportions indicate that the commensal \emph{Penicillium} slightly increased in hosts from the Eastern US and slightly decreased in hosts from the Western US.
-Yet the observed (uncalibrated) proportions show moderate and large decreases, respectively (Figure \ref{fig:regression-example} above).
-
-These examples involved scenarios where individual taxa can constitute a large proportion of the community.
-But in samples from highly-diverse ecosystems such as human gut, soil, and ocean sediment, one species rarely dominates.
-In these cases it might seem intuitive that the mean efficiency should be relatively stable across samples: As diversity increases, the mean efficiency effectively averages over efficiencies from a greater number of taxa and so (by the central limit theorem) will converge to a constant value---if taxon efficiencies are statistically independent of taxon abundances (REF APPENDIX).
-In practice, associations between efficiencies and abundances may be common and can allow large shifts in mean efficiency to occur even in highly diverse samples.
-
-The human gut provides a clear example of how such associations might arise.
-Gut microbiomes typically have high diversity at the species level but are dominated by just a small number of phyla, and two in particular: The Bacteroidetes and the Firmicutes.
-The ratio of Bacteroidetes to Firmicutes can shift substantially between individuals and has been linked to a number of host traits and health conditions.
-In addition, many DNA extraction protocols more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though there can also be significant variation in extraction efficiency among species within these phyla (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-For such protocols, we should expect Bacteroidetes-dominated samples to have substantially larger mean efficiencies than Firmicutes-dominated samples, regardless of species-level diversity.
-This association of efficiency with phylum abundance will distort DA inferences at any taxonomic level if the covariate is also associated with these phyla.
-
-This example illustrates a general mechanism by which association between the abundances and efficiencies of taxa are created by the common influence of evolutionary history.
-Just as shared evolutionary history creates positive associations in ecological traits that drives positive correlations in how related taxa vary in abundance across samples, so does it create positive associations in bias-related traits that can lead to similar efficiencies among closely related taxa.
-Such phylogenically-associated, bias-affecting traits include cell-wall toughness, ribosomal-operon copy number, binding affinity for a given set of primers, and representativeness in taxonomic databases.
-Ecology and efficiency are both affected by the same evolutionary history, leading to positive associations between the two: A change in diet that increases the relative abundances of many Bacteroidetes species and in so doing also increases the relative abundance of many easy-to-lyse species.
-
-Associations between measurement efficiency and relative abundance can also arise because a single trait affects both.
+\hypertarget{regression-error}{%
+\subsection{Box: Error in estimated regression coefficients and standard errors}\label{regression-error}}
+
+Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.
+
+Consider Equation \eqref{eq:regression} for the simple linear regression of log density of a species \(i\) on covariate \(x\).
+Let \(y\) stand for the actual log density of a focal species \(i\), \(z\) stand for the log density that we've measured, and \(d = z - y\) equal the difference between the two, which here is the log efficiency of species \(i\) minus the log mean efficiency of the sample.
+Let \(s_{xy}\) denote the sample covariance between variables \(x\) and \(y\), \(s^{2}_{x} = s_{xx}\) and \(s_{x}\) denote the sample variance and standard deviation, and \(r_{xy} = s_{xy}/(s_{x}s_{y})\) denote the sample correlation.
+
+The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of \(z\) equals the sample covariance of \(z\) and \(x\) divided by the sample variance in \(x\) or, equivalently, the sample correlation of \(z\) and \(x\) multiplied by the ratio of their sample standard deviations,
+\begin{align}
+  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
+\end{align}
+From the (bi)linearity of sample covariances it follows that
+\begin{align}
+  \hat \beta_z 
+  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
+  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
+  = \hat \beta_y + \hat \beta_d,
+\end{align}
+where \(\hat \beta_y\) and \(\hat \beta_d\) denote the slope estimates for \(y\) and \(d\) (were these values to be known).
+The absolute error in the estimate \(\hat \beta_{z}\) of \(\hat \beta_{y}\) is therefore \(\hat \beta_{d}\); it is large in a practical sense when \(\hat \beta_{d}\) is large (in absolute value) compared to \(\hat \beta_{y}\), which corresponds to the covariance of \(d\) with \(x\) being large compared to the covariance of \(y\) with \(x\).
+
+In our case, the covariance of \(d\) equals the negative of the covariance of log mean efficiency with \(x\).
+The absolute error (i.e., the \emph{bias} in a statistical sense) in \(\hat \beta\) is equals the negative covariance of log mean efficiency scaled by the variance in \(x\).
+This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
+For species that covary with \(x\) more strongly than the log mean efficiency, the error will be relatively small.
+This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with \(x\) compared to the log density of the focal species.
+
+We can similarly understand the impact of measurement error on the precision of our slope estimates.
+The OLS and MLE estimated standard error in \(\hat \beta\) are both approximately
+\begin{align}
+  \hat{\text{se}}(\hat \beta)
+  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
+\end{align}
+where \(s_{\hat \varepsilon}\) is the sample standard deviation of the residuals
+(\protect\hyperlink{ref-wasserman2004allo}{Wasserman} (\protect\hyperlink{ref-wasserman2004allo}{2004}) Chapter 13).
+The sample residuals of \(z\), \(y\), and \(d\) have a similar relationship to the regression coefficient estimates,
+\begin{align}
+  \hat \varepsilon_z 
+    &\equiv z - \hat \beta_z x
+  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
+  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
+  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
+\end{align}
+(Note, here I've omitted the subscript indicating the dependence on the sample.)
+It follows that the sample variances of the residuals of \(z\), \(y\), and \(d\) are related through
+\begin{align}
+  s^2_{\hat \varepsilon_{z}} 
+  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
+  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
+\end{align}
+The standard deviation of the \(z\) residuals is increased above that of the \(y\) residuals when the \(d\) residuals are either uncorrelated or positively correlated with the \(y\) residuals, but may be decreased when the \(y\) and \(d\) residuals are negatively correlated.
+
+In our case, the \(d\) residuals equal the negative residuals of the log mean efficiency.
+It is plausible that for most species, their residual variation will have a small covariance with log mean efficiency and the net effect of variation in the mean efficiency will be to increase the estimated standard errors, as occurs with most species in Figure \ref{fig:regression-example}.
+However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \ref{fig:regression-example}.
+
+\hypertarget{implications}{%
+\section{Implications for real-world inference~}\label{implications}}
+
+Partial but not full canceling of the effect of bias in proportion-based analyses, such that possible to have spurious results in principle; but what about in practice?
+Predicting how frequently or which results are likely to be majorly affected is difficult because it depends on the joint variation in efficiencies of species and species abundance dynamics, both of which are poorly understood in most to all biological systems.
+However, through a combination of theoretical considerations, hypothetical scenarios, and real case studies we can begin to develop an understanding or the relevant considerations and build framework for improving our understanding as our knowledge of these dynamics improves.
+
+\hypertarget{systematic-error-in-slope-or-lfc-estimates}{%
+\subsection{Systematic error in slope or LFC estimates}\label{systematic-error-in-slope-or-lfc-estimates}}
+
+Recall from Section \ref{differential-abundance} that the error in the slope or LFC in a DA regression analysis is proportional to the covariance of the log mean efficiency with the covariate.
+This covariance can be split into two components: the standard deviation of (log) mean efficiency and its correlation with the covariate.
+Hence when considering whether taxonomic bias creates large error, it can be useful to separately ask whether the mean efficiency is likely to vary across samples and whether it is likely to be correlated with a covariate of interest.
+
+One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
+\protect\hyperlink{ref-brooks2015thet}{Brooks et al.} (\protect\hyperlink{ref-brooks2015thet}{2015}) performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome with the MGS protocol developed for the Vaginal Human Microbiome Project (VaHMP).
+\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) later used this data to estimate the efficiencies of these seven species and develop a method to use these estimates to infer the true compositions of natural samples (see Section @ref(\#calibrate-compositions)).
+SI Analysis (MOMS-PI) uses these estimated efficiencies to correct the compositions and explore the implied variation in mean efficiency in the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) \protect\hyperlink{ref-fettweis2019thev}{Fettweis et al.} (\protect\hyperlink{ref-fettweis2019thev}{2019}).
+The vaginal microbiome during pregnancy is characterized by relatively low diversity, often with a single \emph{Lactobacillus} species often forming a majority of reads (considered to be the healthy state) but with occasional shifts to high proportions of other species, such as \emph{Gardnerella vaginalis}, \emph{Atopobium vaginae}, and \emph{Streptococcus agalactiae}, which are thought to cause or indicate poor vaginal health and increased risk of preterm birth.
+The analysis of \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed that the efficiencies of two common \emph{Lactobacillus} species, \emph{L.
+crispatus} and \emph{L.
+iners}, were 20-30X greater than that of \emph{G.
+vaginalis}, \emph{A.
+vaginae}, and \emph{S.
+agalactiae}.
+Analysis of microbiome trajectories of a single patient across visits indicates that the mean efficiency typically decreases by a factor of 3-10X when the community shifts from a high to low-\emph{Lactobacillus} state.
+(NOTE: This is a super preliminary claim based on looking at just one woman, and is just a placeholder)
+Decreases in mean efficiency during transitions from \emph{Lactobacillus} to \emph{Gardnerella} dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify \emph{Gardnerella}.
+
+A second study in which control communities make it possible to directly analyze the impact of taxonomic bias in experimental samples was performed by \protect\hyperlink{ref-leopold2020host}{Leopold and Busby} (\protect\hyperlink{ref-leopold2020host}{2020}).
+To study the interactions between a host plant and its fungal commensals and pathogen, the authors inoculated trees with a fungal synthetic community and later exposed plants to a fungal pathogen.
+The authors used ITS amplicon sequencing to measure communities before and after infection, along with mock communities which they used to estimate the species efficiencies with the method of \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}).
+(DNA mocks were used, so that bias due to DNA extraction is not included, but other major bias sources such as PCR and ribosomal copy-number-variation are included.) The pathogen was 10X more efficiently measured than the median commensal and 40X more efficiently measured than the lowest-efficiency commensal.
+A re-analysis of their data (SI Analysis) shows that the mean efficiency varies by ???
+in pre-infection communities and by an average of ???
+pre- and post-infection, with the latter driven by the dramatic increase in the high-efficiency pathogen.
+
+These case studies indicate that one mechanism for variability in the mean efficiency is swings between very high (\(\sim 1\)) and low (\(\ll 1\)) proportions of species with particularly high or low efficiencies.
+In highly diverse ecosystems, as seen in soil and the human gut (CHECK), one species rarely dominates within a given sample and so such large swings are not possible.
+We might therefore think that the mean efficiency will remain fairly stable in these settings.
+In fact, one can show mathematically that in randomly assembled communities, the mean efficiency becomes more stable as (Inverse Simpson) diversity increases, as the mean effectively averages over a greater number of species.
+Real communities do not assemble randomly, however.
+Variance in the mean efficiency may remain high if environmental filtering leads to different groups of organisms dominating in different environments, particularly if shared phenotypic values or evolutionary history causes these groups to differ in their efficiencies as well as their ecology.
+For example, gut microbiomes typically have high diversity at the species level but are dominated by just a small number of phyla, and two in particular: The Bacteroidetes and the Firmicutes, the ratio of which varies substantially across individuals (CHECK).
+Many DNA extraction protocols have been found to more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though variation can also be substantial within phyla, \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+Gut samples with large differences in their Bacteroidetes and Firmicutes proportions may therefore have large variation in mean efficiency even if the individual species within each phylum never reach a large proportion.
+On the other hand, competitive-exclusion dynamics can favor more stable taxonomic and/or phenotypic compositions across samples and might thereby drive a more stable mean efficiency than under random assembly
+
+Systematic error in DA results requires that the variation in log mean efficiency is correlated with the covariate.
+The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
+For example, in the vaginal microbiome a decline in \emph{Lactobacillus} and rise of species including \emph{Gardnerella vaginalis} is associated with preterm birth in pregnant women; hence the variation in mean efficiency driven by these species may also create an association of mean efficiency with preterm status.
+In the plant-fungal experiment of \protect\hyperlink{ref-leopold2020host}{Leopold and Busby} (\protect\hyperlink{ref-leopold2020host}{2020}), the increase in pathogen proportion after infection leads to a large increase in the mean efficiency over time, which systematic increases the observed decline in the proportions of commensal species (SI Analysis).
+The ratio of Bacteroidetes to Firmicutes has been linked to several host traits and health conditions, and thus the variation in efficiency between these two phyla may also drive an association of mean efficiency with these traits.
+
+Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
+One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
+A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
+Shared evolutionary history among phylogenetically-close species is expected and has been observed to create positive associations in these traits among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
+Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
+In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
+For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.
+
+Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
 For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
@@
 
-\hypertarget{error-in-absolute-abundance-measurements}{%
-\subsubsection{Error in absolute abundance measurements}\label{error-in-absolute-abundance-measurements}}
-
-So far we have ignored error associated with our absolute abundance measurements.
-In fact, these measurements can also have a tendency to measure contributions from some taxa more efficiently than others.
-\textbf{TODO: State how this could create problems.
-Might also note the ambiguity in what the truth/target is (e.g.
-live cells, or all cells; cells, or 16S rRNA gene copies; observations in Jian, Salonen, and Korpela 2021).}
-Interestingly, taxonomic bias may actually make qPCR a better method for absolute quantification than cell counting for 16S rRNA gene sequencing experiments.
-16S rRNA gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore necessarily affected by three large sources of bias in 16S rRNA gene experiments: extraction, amplification, and copy-number variation.
-Yet these biases are shared by the sequencing measurement.
-Though they make qPCR measurements a bad proxy for total cell density, when used for absolute differential abundance inference the shared bias in the qPCR and sequencing measurements can cancel, leaving relative accurate fold changes.
-(Jian, Salonen, and Korpela 2021 suggests some of these ideas and APPENDIX gives a mathematical justification.)
-If common primers are used by qPCR and 16S rRNA gene sequencing, the bias due to primer mismatches and other sources of variation in amplification could even be accounted for in this manner.
-Yet ideally we would still be able to estimate and correct remaining error due to unshared bias or to mechanisms, such as saturation during DNA extraction, that could break the assumed proportionality between qPCR measurements and total abundance.
-
-\hypertarget{we-cannot-simply-ignore-taxonomic-bias}{%
-\subsection{We cannot simply ignore taxonomic bias}\label{we-cannot-simply-ignore-taxonomic-bias}}
-
-These examples show that there is potential for bias to distort DA results in both low and high diversity settings.
-We therefore need methods to control or correct for bias to ensure the validity of future results, as well as methods to probe the robustness of past results.
-It is possible that spurious results are unlikely in many experimental contexts.
-By applying such methods across a range of biological and experimental contexts we will deepen our understanding of when bias is unlikely to significantly distort biological findings and when additional measures are needed to mitigate its effects.
-
-\hypertarget{potential-solutions}{%
-\section{Potential solutions}\label{potential-solutions}}
-
-\hypertarget{using-control-samples-to-calibrate-relative-abundances}{%
-\subsection{Using control samples to calibrate relative abundances}\label{using-control-samples-to-calibrate-relative-abundances}}
-
-In principle, control (``mock'') communities containing representative taxa from the environment of interest can be used to directly estimate the measurement efficiencies and correct bias in the MGS measurements (``calibration'').
-Such an approach may be sufficient for synthetic-community experiments, where all taxa are culturable, and for relatively simple natural communities like the vaginal microbiome that are dominated by a small number of culturable taxa.
-But suitable mock controls may not be feasible for most complex natural ecosystems and require significant effort to develop.
-
-An closely-related alternative to mocks are controls derivedd from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-A natural fecal standard is currently being developed by \href{https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material}{NIST} and one has recently been made available commercially by Zymo Research (\href{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology}{ZymoBIOMICS Fecal Reference}).
-Ensuring the stability and homogeneity of such control samples can be challenging.
-Careful testing is also needed to ensure that the preparation and storage of the controls has not significantly affected the taxonomic bias relative to the experimental samples in a given application.
-As it is feasible to characterize a single standard much more extensively than a typical community sample, we may be able to obtain an estimated composition we feel comfortable treating as the ground truth.
-Yet even when this is not possible, such natural standards can allow us to reconcile results across studies despite not knowing the truth.
-
-\hypertarget{ratio-based-relative-and-absolute-abundance-inference}{%
-\subsection{Ratio-based relative and absolute abundance inference}\label{ratio-based-relative-and-absolute-abundance-inference}}
-
-A comprehensive approach of estimating the efficiency of all taxa is needed for getting calibrated relative abundances of all taxa within individual samples.
-But what if we only want calibrated DA analysis?
-That is, we are comfortable with not knowing whether \emph{E. coli} is 10\% of our sample if we can confidently determine that it doubled or halved between sample conditions?
-This problem is easier to solve, and one solution is to use analysis methods based on the ratios among taxa rather than the proportions of individual taxa.
-
-A subset of methods for analyzing differential relative abundance are derived from the field of Compositional Data Analysis (CoDA).
-The defining feature of CoDA methods is that they are based on fold-changes in ratios among elements (here, the taxa) (REF AITCHISON).
-The use of CoDA methods in microbiome analysis has largely been motivated by concerns over the negative correlations between taxa induced by the sum-to-one constraint in taxon proportions.
-But \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed that, due to their property of perturbation invariance (REF AITCHISON), the results of CoDA differential relative abundance analysis are invariant to consistent taxonomic bias.
-From Equation \eqref{eq:error}, the observed ratio of a taxon \(i\) to a taxon \(j\) in a sample \(s\) is
-\begin{align}
-  \frac{\text{observed}_{i}}{\text{observed}_{j}}=\frac{\text{actual}_{i}}{\text{actual}_{j}}*\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}
-\end{align}
-The error is independent of the sample composition and thus cancels in any function that is a fold-change of this (or any) ratio of taxa between samples.
-(Perhaps insert example from Figure \ref{fig:vaginal-example}ABC all giving same ratio FCs.)
-
-Might ratio-based analysis also be used to overcome the effects of bias in differential absolute abundance?
-In fact, some methods for determining absolute abundance are based on ratios among taxa in the MGS measurement instead of proportions.
-Rather than using a measurement of total abundance, these methods require determining the abundance of one or more reference taxa.
-To estimate the abundance of a focal taxon \(i\), this approach multiplies the ratio of reads (or proportions) of taxon i to a reference taxon \(r\) by a known or estimated abundance of the reference taxon, (while ignoring bias)
-\begin{align}
-  \text{estimate}_{i} = \frac{\text{observed}_{i}}{\text{observed}_{r}}*\text{abundance}_{r}
-\end{align}
-So far this approach has been mainly used with spike-in experiments, in which an extraneous taxon is added in a known (and typically constant) abundance to each sample so that it can serve as the reference taxon.
-Yet the reference taxon could also be a naturally occurring taxon whose absolute abundance we have estimated using a method such as ddPCR directly on cells or (q/dd)PCR on the extracted DNA.
-Because taxa do not compete (or competition is greatly reduced) in such targeted measurements, any taxonomic bias associated with them is expected to create constant multiplicative error across samples and so not affect fold change estimates {[}APPENDIX{]}.
-APPENDIX describe various theoretical and experimental considerations for both approaches.
-
-\hypertarget{estimating-error-with-targeted-measurements-of-a-small-number-of-taxa}{%
-\subsection{Estimating error with targeted measurements of a small number of taxa}\label{estimating-error-with-targeted-measurements-of-a-small-number-of-taxa}}
-
-Because targeted absolute-abundance measurements are expected to provide (relatively) accurate estimates of fold changes, they can be used to validate and even correct the DA results derived from a proportion-based absolute DA analysis.
-We illustrate the basic idea using the \protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) experiment described above.
-Absolute abundances in this study were estimated by multiplying MGS proportions by total abundance measured by cell counting; for comparison, qPCR was used to measure specific microbial taxa.
-Because the error in fold changes or a regression from the MGS measurements is taxon-independent, the difference between the two methods for these reference taxa informs us about the difference for all taxa.
-\protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) found close agreement for the specific taxa also measured by qPCR (Table 1), suggesting that variation in sample mean efficiency did not significantly distort their results.
-In principle, a joint statistical analysis of all measurements would allow inferring variation in the sample mean efficiency across samples and obtaining calibrated fold change and regression estimates for all taxa {[}APPENDIX{]}.
-
-\hypertarget{computational-approaches}{%
-\subsection{Computational approaches}\label{computational-approaches}}
-
-What can we do for experiments that have already been conducted without mock controls, targeted control measurements, or spike-ins?
-At least two purely computational approaches can still be used.
-
-\textbf{Bias sensitivity analysis:}
-A straightforward and universally available approach is to use computer simulation to determine how the results change under a range of possible sets of efficiencies, which can be randomly generated to reflect certain hypotheses about bias in the given system.
+The biological significance of the error in \(\hat \beta\) depends on the relative magnitude of the LFC (or covariance with the covariate \(x\)) in the focal species vs in the mean efficiency.
+It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
+(This seems intuitively likely and true in examples, but I haven't analyzed it much mathematically and don't have a firm handle on this question.)
+Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
+However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for all other species.
+The experimental and computational methods discussed in Section \ref{solutions} can mitigate taxonomic bias in these clearly-problematic cases while also helping to amass empirical evidence on the conditions in which bias is unlikely to cause major inferential errors.
+
+\begin{itemize}
+\tightlist
+\item
+  Todo: Address DA analysis of aggregate (or synthetic) taxa, such as phyla, which may be prone to errors due to changes in the species-level make-up of the taxon.
+\end{itemize}
+
+\textbf{Error in total-density measurements:}
+
+So far we have ignored error in total-density measurements that are used to convert estimated proportions to estimated densities;
+however, in reality these measurements are likely to contain substantial systematic error that can either worsen or improve the error in LFC estimates.
+
+\begin{itemize}
+\tightlist
+\item
+  Todo: Add to previous section or Appendix with the relevant theory.
+\end{itemize}
+
+To understand how, we can consider the question of whether bulk 16S qPCR or flow cytometry is a more appropriate method of measuring total bacterial density for use in proportion-based density measurement (\protect\hyperlink{ref-galazzo2020howt}{Galazzo et al.} (\protect\hyperlink{ref-galazzo2020howt}{2020}), \protect\hyperlink{ref-jian2021comm}{Jian, Salonen, and Korpela} (\protect\hyperlink{ref-jian2021comm}{2021})).
+16S rRNA-gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore affected by variation among species in DNA extraction efficiency, 16S copy number, and primer-binding and amplification efficiency.
+For example, \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}) found that the product of extraction efficiency and 16S copy number for \emph{Lactobacillus iners} was 20X larger than \emph{Gardnerella vaginalis} (CHECK NUMBERS), such that we should expect samples dominated by \emph{L.
+iners} to result in qPCR estimates 10-20X higher than equally dense samples dominated by \emph{G.
+vaginalis}.
+These observations suggest that 16S qPCR is a poor proxy for fold changes in total bacterial density.
+Yet, as recently suggested by \protect\hyperlink{ref-jian2021comm}{Jian, Salonen, and Korpela} (\protect\hyperlink{ref-jian2021comm}{2021}), this taxonomic bias in qPCR may simultaneously make it a good measure of total density for use with 16S amplicon sequencing for proportion-based density estimation.
+16S amplicon sequencing shares these same sources of bias and so is likely to have species efficiencies highly similar to those for the qPCR measurements, which we can predict to cause the mean 16S-qPCR efficiency to closely follow the mean 16S-sequencing efficiency, such that the error induced by one largely cancels the error induced by the other (Ref Appendix).
+In contrast, species-specific efficiencies in bulk cell counting are less likely to be strongly correlated with those from 16S sequencing.
+Even supposing cell counting were to provide perfectly accurate measures of total density, it may actually perform worse than qPCR for DA analysis of proportion-based density estimation precisely because the error does not mirror that in the 16S measurement.
+
+On the other hand, bulk or targeted measurement (or DNA spike-ins) that occurs post DNA-extraction is subject to other error that is not likely to counter that of the 16S-based proportion estimates.
+The yield of DNA extraction can show highly random variation and also be highly non-linear (as a function of cellular biomass) at high concentrations.
+This features create error that is unlikely to offset the variation in mean efficiency in the 16S sequencing (todo: clarify why).
+Targeted measurement of cells and cellular spike-ins, discussed in Section \ref{solutions}, can avoid this issue while also facilitating ratio-based analysis while (unlike bulk cell counting) remaining robust mean-efficiency variation.
+
+\hypertarget{loss-of-power-and-precision-stub}{%
+\subsection{Loss of power and precision (stub)}\label{loss-of-power-and-precision-stub}}
+
+Variation in mean efficiency that is not associated with covariates can still majorly hinder the goals of microbiome studies by causing reductions in the precision of LFC estimates and the power to detect true associations of a given size.
+
+\begin{itemize}
+\tightlist
+\item
+  Illustrate with case studies?
+\item
+  Mention counting noise as an important factor that we've so far ignored but that can interact with taxonomic bias to lead to very low precision in LFC estimates.
+Low efficiency can put a species below the detection limit set by sample read depth, making it impossible to get precise LFCs.
+Study-specific efficiencies can explain why associations often do not replicate across studies.
+Example: \emph{Gardnerella vaginalis} associations are not detected in certain vaginal microbiome studies that used V13 primers which we now know are bad at amplifying it.
+\item
+  Conclude with a nod to the value of the solutions in Section \ref{solutions} to address these issues: Correcting the errors caused by mean-efficiency variation can improve precision, and bias-sensitivity analysis and bias-aware meta-analysis can give a better sense of the increased uncertainty in LFC estimates due to unknown efficiencies, thereby helping us to understand the increased rate of false positives and/or the larger range of LFC estimates that are compatible with the observed data than what is implied by a bias-free model.
+\end{itemize}
+
+\hypertarget{solutions}{%
+\section{Potential solutions}\label{solutions}}
+
+The theoretical results from Section \ref{differential-abundance} suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.
+We present five that seem particularly promising:
+The first three are method for correcting or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.
+
+\hypertarget{calibrate-compositions}{%
+\subsection{Calibrate compositions using community controls}\label{calibrate-compositions}}
+
+The most complete but experimentally challenging approach to mitigating taxonomic bias involves directly measuring the efficiencies of individual species using control samples (\protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+Suitable controls are any community sample whose composition (species identity and relative abundances) has been previously characterized, either by construction (in the case of constructed `mock' communities) or by measurement with a \emph{reference protocol} chosen to serve as a measurement standard.
+These controls, when measured alongside the target experimental samples, can be used to estimate the relative efficiencies among the superset of species across the control samples.
+The estimated efficiencies can then be used to calibrate the measured relative abundances among the control species in the target samples to the chosen measurement standard, by simply dividing the species' read counts by their corresponding efficiencies or using a more sophisticated statistical modeling framework.
+These calibrated compositions can then be used for arbitrary downstream microbiome analyses, including DA analysis.
+Calibration of species not in the controls requires first imputing their efficiencies from the control measurements, for example using phylogenetic relationships, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
+Though such imputation is conceptually straightforward to implement, its reliability and utility remain untested.
+
+\begin{itemize}
+\tightlist
+\item
+  The above states the approach; can add description/discussion of application to the MOMS-PI and \protect\hyperlink{ref-leopold2020host}{Leopold and Busby} (\protect\hyperlink{ref-leopold2020host}{2020}) cases and developments and challenges in availability of control communities
+\end{itemize}
+
+\hypertarget{use-ratio-based-abundance-measurement}{%
+\subsection{Use ratio-based abundance measurement}\label{use-ratio-based-abundance-measurement}}
+
+The DA analyses we consider don't need accurate estimates of relative abundances within individual samples---they only require accurate fold changes across samples.
+Section \ref{differential-abundance} showed that fold changes in species ratios and ratio-based density estimates can remain accurate even when taxonomic bias distorts the values in individual samples.
+Hence a solution is to use to analyze variation in ratios and ratio-based density estimates.
+For the latter, practical options include
+
+\begin{itemize}
+\tightlist
+\item
+  spike-ins of extraneous species at constant and/or known density
+\item
+  housekeeping taxa, stipulated or computationally identified as having a constant density across samples
+\item
+  targeted measurement of specific taxa
+\end{itemize}
+
+TBD what further discussion or illustration to have here.
+
+\hypertarget{calibrate-fold-changes-using-measurements-of-targeted-taxa}{%
+\subsection{Calibrate fold changes using measurements of targeted taxa}\label{calibrate-fold-changes-using-measurements-of-targeted-taxa}}
+
+In some cases, we may still which to use a proportion-based approach to analyzing variation in density.
+For example, in some systems measuring total cell concentration may be easier than implementing a spike-in experiment or determining a set of reference species to directly measure for which at least one species is present in all samples.
+In this case, targeted measurements of one or more reference species can be used to validate and/or correct the fold changes or regression coefficients infered from the proportion-based densities.
+Recall that the error in proportion-based log fold changes (LFCs) is shared among species and is equal to the inverse LFC fold change in mean efficiency (Equations \eqref{eq:prop-fc-error} and \eqref{eq:regression-error}; Figure \ref{fig:regression-example}D).
+Consequently, an accurate LFC of one or more reference species can serve as a general validation or used to correct the LFCs of all species.
+
+\begin{itemize}
+\tightlist
+\item
+  Note: I may not correctly understand the next example, since I don't yet see how this interpretation of Table 1 meshes with Figure 4
+\end{itemize}
+
+We illustrate this idea with an example from the study by \protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) described in Section \ref{implications}, which estimated doubling times for various taxa from the measured log increase in cell density with burial time in marine sediments.
+The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for several bacterial and archeal taxa (CHECK).
+In Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years (CHECK) using the proportion-based densities and 10.3 years by qPCR (\protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}) Table 1 and Figure 4).
+These doubling times correspond to highly similar estimated slopes of log density per year of \(1/10.1 \approx 0.99\) and \(1/10.3\approx 0.97\).
+The difference between these slopes, \(\approx 0.02\), estimates the slope of the mean efficiency.
+It's small value suggests little systematic variation in mean efficiency with burial time and hence little systematic bias in the estimated doubling times derived from the proportion-based density estimates.
+In this case, close agreement in the DA results across methods for a specific taxon suggests that variation in sample mean efficiency did not significantly distort the MGS-based results and hence provides a degree of validation for all taxa, not just the taxa with targeted measurements.
+(An important caveat for this example is that this analysis assumes consistent efficiencies at the higher taxonomic levels used for qPCR measurement and MGS analysis performed by \protect\hyperlink{ref-lloyd2020evid}{Lloyd et al.} (\protect\hyperlink{ref-lloyd2020evid}{2020}).)
+Here we compared summary statistics; however, in principle a joint statistical analysis of targeted and proportion-based density measurements would provide more statistical power to estimate and correct the effects of variation in the mean efficiency across samples and provide calibrated fold change and regression estimates for all taxa {[}APPENDIX{]}.
+
+\hypertarget{perform-a-bias-sensitivity-analysis}{%
+\subsection{Perform a bias sensitivity analysis}\label{perform-a-bias-sensitivity-analysis}}
+
+What can we do for experiments that have been conducted without control communities, targeted control measurements, or spike-ins?
+One approach is to conduct a bias sensitivity analysis to probe the extent and way in which the results of a DA analysis are sensitive to different possible taxonomic biases in our experiment.
+
+A straightforward approach to bias sensitivity analysis is to use computer simulation to randomly generate efficiency vectors, each of which is used to calibrate the MGS measurements.
+These calibrated MGS measurements represent the true compositions under the hypothesized taxonomic bias.
+By re-running our DA analysis and comparing results across these simulated datasets, we can begin to understand how our initial results might be influenced by the strength and structure of the taxonomic bias in our experiment.
+
+The simulated efficiency vectors may be generated to reflect certain hypotheses about bias in the given system.
 The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
-Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results---not just for differential absolute abundance, but for all microbiome analyses.
-
-\textbf{Bias-aware meta-analysis:}
-When the goal is to perform a meta-analysis that combines studies that have used different protocols, the unknown measurement efficiencies of each protocol can be explicitly included as parameters of the statistical model that is used.
-Unknown efficiencies can be included in ``compositional'' linear modeling frameworks such as ALDEx2, DivNet, and fido simply by adding a protocol-specific term to the linear model of taxon log ratios.
-Thus such bias-aware meta-analyses are already technically feasible and---if bias is truly consistent within a protocol or study---may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not explicitly model bias.
+Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results---for differential abundance and microbiome analyses more generally.
+
+\hypertarget{use-bias-aware-meta-analysis}{%
+\subsection{Use bias-aware meta-analysis}\label{use-bias-aware-meta-analysis}}
+
+So far we have considered analysis of samples subject to the same taxonomic bias;
+however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in different groups of samples.
+This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses where the effects of bias might cancel within a study.
+
+A potential solution is to perform a bias-aware meta-analyses that jointly analyzes multiple data from studies while explicitely acknowledges the distinct, unknown taxonomic bias in each study.
+Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model ``batch effects''---amorphous differences in the measurements of each study.
+By configuring these models so that these latent parameters correspond to the study and species specific efficiencies, we can increase the biological interpretability of these models and gain ability to determine whether disageements in DA results across studies are due to differences in taxonomic bias or other differences (such as host cohort or study design).
+Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by \protect\hyperlink{ref-mclaren2019cons}{McLaren, Willis, and Callahan} (\protect\hyperlink{ref-mclaren2019cons}{2019}), which can increase improve our ability to find reproducible microbial associations in a given meta-study.
 
@@
 
-\hypertarget{implications}{%
-\subsection{Implications}\label{implications}}
+\hypertarget{implications-1}{%
+\subsection{Implications}\label{implications-1}}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-contijoch2019gutm}{}}%
-Contijoch, Eduardo J, Graham J Britton, Chao Yang, Ilaria Mogno, Zhihua Li, Ruby Ng, Sean R Llewellyn, et al.
-2019.
-{``{Gut microbiota density influences host physiology and is shaped by host and microbial factors}.''} \emph{Elife} 8 (January).
-\url{https://doi.org/10.7554/eLife.40553}.
+\leavevmode\vadjust pre{\hypertarget{ref-callahan2017repl}{}}%
+Callahan, Benjamin J, Daniel B DiGiulio, Daniela S Aliaga Goltsman, Christine L Sun, Elizabeth K Costello, Pratheepa Jeganathan, Joseph R Biggio, et al.
+2017.
+{``{Replication and refinement of a vaginal microbial signature of preterm birth in two racially distinct cohorts of US women}.''} \emph{Proc.
+Natl.
+Acad.
+Sci.
+U.
+S.
+A.} 114 (37): 9966--71.
+\url{https://doi.org/10.1073/pnas.1705899114}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-finucane2014atax}{}}%
+Finucane, Mariel M., Thomas J.
+Sharpton, Timothy J.
+Laurent, and Katherine S.
+Pollard.
+2014.
+{``{A Taxonomic Signature of Obesity in the Microbiome?
+Getting to the Guts of the Matter}.''} Edited by Markus M.
+Heimesaat.
+\emph{PLoS One} 9 (1): e84689.
+\url{https://doi.org/10.1371/journal.pone.0084689}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-galazzo2020howt}{}}%
+Galazzo, Gianluca, Niels van Best, Birke J.
+Benedikter, Kevin Janssen, Liene Bervoets, Christel Driessen, Melissa Oomen, et al.
+2020.
+{``{How to Count Our Microbes?
+The Effect of Different Quantitative Microbiome Profiling Approaches}.''} \emph{Front.
+Cell.
+Infect.
+Microbiol.} 10 (August).
+\url{https://doi.org/10.3389/fcimb.2020.00403}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gloor2017micr}{}}%
+Gloor, Gregory B., Jean M.
+Macklaim, Vera Pawlowsky-Glahn, and Juan J.
+Egozcue.
+2017.
+{``{Microbiome Datasets Are Compositional: And This Is Not Optional}.''} \emph{Front.
+Microbiol.} 8 (November): 2224.
+\url{https://doi.org/10.3389/fmicb.2017.02224}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jian2021comm}{}}%
+Jian, Ching, Anne Salonen, and Katri Korpela.
+2021.
+{``{Commentary: How to Count Our Microbes?
+The Effect of Different Quantitative Microbiome Profiling Approaches}.''} \emph{Front.
+Cell.
+Infect.
+Microbiol.} 11 (March): 10--12.
+\url{https://doi.org/10.3389/fcimb.2021.627910}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kevorkian2018esti}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lozupone2013meta}{}}%
+Lozupone, Catherine A, Jesse Stombaugh, Antonio Gonzalez, Gail Ackermann, Janet K Jansson, Jeffrey I Gordon, Doug Wendel, Yoshiki Va, and Rob Knight.
+2013.
+{``{Meta-analyses of studies of the human microbiota}.''} \emph{Genome Res.}, 1704--14.
+\url{https://doi.org/10.1101/gr.151803.112}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-martin2020mode}{}}%
+Martin, Bryan D., Daniela Witten, and Amy D.
+Willis.
+2020.
+{``{Modeling microbial abundances and dysbiosis with beta-binomial regression}.''} \emph{Ann.
+Appl.
+Stat.} 14 (1): 94--115.
+\url{https://doi.org/10.1214/19-AOAS1283}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-martino2019anov}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-props2017abso}{}}%
-Props, Ruben, Frederiek-Maarten Kerckhof, Peter Rubbens, Jo De Vrieze, Emma Hernandez Sanabria, Willem Waegeman, Pieter Monsieurs, Frederik Hammes, and Nico Boon.
-2017.
-{``{Absolute quantification of microbial taxon abundances}.''} \emph{ISME J.} 11 (2): 584--87.
-\url{https://doi.org/10.1038/ismej.2016.117}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-quinn2020inte}{}}%
@@
 2020.
-{``{Interpretable Log Contrasts for the Classification of Health Biomarkers: a New Approach to Balance Selection}.''} \emph{mSystems} 5 (2): 1--11.
+{``{Interpretable Log Contrasts for the Classification of Health Biomarkers: a New Approach to Balance Selection}.''} Edited by Robert G.
+Beiko.
+\emph{mSystems} 5 (2): 1--11.
 \url{https://doi.org/10.1128/mSystems.00230-19}.
@@
 2018.
-{``{Balances: a New Perspective for Microbiome Analysis}.''} \emph{mSystems} 3 (4): e00053--18.
+{``{Balances: a New Perspective for Microbiome Analysis}.''} Edited by Catherine Lozupone.
+\emph{mSystems} 3 (4): e00053--18.
 \url{https://doi.org/10.1128/mSystems.00053-18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tettamantiboshier2020comp}{}}%
-Tettamanti Boshier, Florencia A., Sujatha Srinivasan, Anthony Lopez, Noah G.
-Hoffman, Sean Proll, David N.
-Fredricks, and Joshua T.
-Schiffer.
-2020.
-{``{Complementing 16S rRNA Gene Amplicon Sequencing with Total Bacterial Load To Infer Absolute Species Concentrations in the Vaginal Microbiome}.''} \emph{mSystems} 5 (2): 1--14.
-\url{https://doi.org/10.1128/mSystems.00777-19}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vandeputte2017quan}{}}%
-Vandeputte, Doris, Gunter Kathagen, Kevin D'hoe, Sara Vieira-Silva, Mireia Valles-Colomer, João Sabino, Jun Wang, et al.
-2017.
-{``{Quantitative microbiome profiling links gut community variation to microbial load}.''} \emph{Nature} 551 (7681): 507.
-\url{https://doi.org/10.1038/nature24460}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vieirasilva2019quan}{}}%
-Vieira-Silva, Sara, João Sabino, Mireia Valles-Colomer, Gwen Falony, Gunter Kathagen, Clara Caenepeel, Isabelle Cleynen, Schalk van der Merwe, Séverine Vermeire, and Jeroen Raes.
-2019.
-{``{Quantitative microbiome profiling disentangles inflammation- and bile duct obstruction-associated microbiota alterations across PSC/IBD diagnoses}.''} \emph{Nat.
-Microbiol.} 4 (11): 1826--31.
-\url{https://doi.org/10.1038/s41564-019-0483-9}.
 

### 2021-10-18 notebook/_posts/2021-10-18-evaluate-robustness-example/evaluate-robustness-example.Rmd
initial commit of robustness evaluation example

---
title: "Evaluate robustness example"
description: |
  Evaluate the robustness of an estimate and inference to bias.
author:
  - name: Amy D. Willis
date: 10-18-2021
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  autodep = TRUE,
  echo = TRUE
)
```

# Robustness of an analysis to bias

We can investigate the sensitivity of any analysis to taxonomic bias by simulating possible efficiency vectors, calibrating our observed MGS measurements by these efficiencies, and rerunning our analysis on the simulated and calibrated data. Specifically, we can take the observed proportion of each species in our dataset $\hat{\text{prop}}_{i}(a)$, and "undo" the efficiencies to obtain the calibrated proportions
\begin{align}
  (\#eq:calibrate-props)
  \text{calibrated prop}_{i}(a)
  &= \left( \frac{\hat{\text{prop}}_{i}(a)}{\text{efficiency}_{i}} \right) \bigg / \sum_{i'} \left( \frac{\hat{\text{prop}}_{i'}(a)}{\text{efficiency}_{i'}} \right),
\end{align}
and the calibrated read counts
\begin{align}
  (\#eq:calibrate-reads)
  \text{reads}_{i}(a)
  &= \text{total reads}(a) \times \text{calibrated prop}_{i}(a).
\end{align}
We can then rerun our data analysis using the calibrated read counts in place of our original data. If we do this repeatedly, each time using different realizations of the efficiency vector, we can investigate how the results of our analysis change. If the change is minimal, this suggests that our results are robust to unequal detection efficiencies in the different species. The converse would suggest that we need to be careful interpreting our results.

## Example: relative abundance of Gardnerella

For example, suppose we are interested in investigating if the genus Gardnerella is more abundant in the vaginal microbiome of women who give birth preterm, and we investigate this model by running `corncob` using preterm birth as a covariate in the mean and dispersion model. We could investigate the robustness of our findings using the following structure:

```{r, eval = FALSE}
our_sim <- new_simulation("UndoGardnerella", "Undo Gardnerella relative abundance") %>%
  generate_model(undo_efficiencies,
                 ps = cal %>%
                   filter_sample_data(cohort == "Stanford") %>%
                   tax_glom("genus") %>%
                   as_tibble,
                 sig_sq_e = as.list(seq(0, 5, length.out = 6)),
                 vary_along = "sig_sq_e") %>%
  simulate_from_model(nsim = 25, index = 1:4) %>%
  run_method(list(corncob_Gardnerella),
             parallel = list(socket_names = 4, libraries = c("corncob", "speedyseq"))) %>%
  evaluate(list(betahat_fit, pvalue_fit))
```

In brief, this approach uses the library `simulator` to generate random efficiencies and "undo" them as described above, then runs `corncob` on these generated datasets before finally obtaining the estimated coefficients on preterm birth and the p-value for the hypothesis test that the coefficient on preterm birth is zero. Let's take a look at the components of the code.

First, we load the relevant libraries and our data

```{r}
# library(here)
library(tidyverse)
library(magrittr)
library(speedyseq)
library(corncob)
# devtools::install_github("jacobbien/simulator")
library(simulator)
cal <- readRDS("callahan2017repl_phyloseq_clean.rds")
```

Then we can create the function that generates random efficiencies and accordingly alters the observed counts:

```{r}
undo_efficiencies <- function(ps, sig_sq_e) {

  ps_ms <- ps %>% group_by(.sample) %>% mutate(mm = sum(.abundance)) %>% ungroup
  ps_tax <- ps %>% select(`.otu`, `kingdom`:`species`) %>% distinct %>% tax_table
  ps_sample <- ps %>% select(`.sample`, `subject_id`:`outcome`) %>% distinct %>% sample_data

  new_model(name = "eff_undo", label = sprintf("sig_sq_e = %s", sig_sq_e),
            params = list(ps_ms = ps_ms, ps_tax = ps_tax, ps_sample = ps_sample, sig_sq_e = sig_sq_e),
            simulate = function(ps_ms, ps_tax, ps_sample, sig_sq_e, nsim) {
              replicate(nsim,
                        {
                          effs_tib <- ps_ms %>%
                            select(genus) %>% distinct %>%
                            mutate("efficiency" = rnorm(nrow(.), mean = 0, sd = sqrt(sig_sq_e)) %>% exp)

                          ps_tib <- ps_ms %>%
                            full_join(effs_tib) %>%
                            group_by(.sample) %>%
                            mutate(p_ijs = (.abundance / efficiency) / sum(.abundance / efficiency)) %>%
                            mutate(w_ij = round(mm * p_ijs)) %>%
                            ungroup

                          phyloseq(ps_tib %>%
                                     select(`.otu`, `.sample`, `w_ij`) %>%
                                     pivot_wider(names_from=`.otu`, values_from=`w_ij`) %>%
                                     otu_table(taxa_are_rows = F),
                                   ps_tax,
                                   ps_sample) %>%
                            tax_glom("genus")
                        },
                        simplify=F)
            })
}
```

Next, we specify how we use our data to construct our estimates. We are using `bbdml` from the library `corncob` to model the relative abundance of ASV 8 (Gardnerella) using preterm birth as a covariate to model the mean and dispersion, but this is only one type of analysis that we might be interested in.

```{r}
corncob_Gardnerella <- new_method("corncob", "Corncob",
                               method = function(model, draw) {

                                 fit <- bbdml(formula = ASV8 ~ preterm,
                                              formula_null= ~1,
                                              phi.formula= ~ preterm,
                                              phi.formula_null= ~1,
                                              data = draw %>%
                                                merge_samples2("subject_id", fun_otu = sum))

                                 list(fit = fit,
                                      summary = summary(fit))
                               })
```

Finally, we construct methods to obtain the estimated coefficient on preterm birth in the mean relative abundance model, and its p-value.

```{r}
betahat_fit <- new_metric("betahat", "beta-hat",
 metric = function(model, out) {
   out$summary$coefficients[2,1]
 })
pvalue_fit <- new_metric("pvalue", "p-value",
                         metric = function(model, out) {
                           out$summary$coefficients[2,4]
                         })
```

Now that we have all of the pieces, we can run them together as follows. Note that we are only going to run 4 iterations for now, but we have commented out code below showing you how to run 100 draws across 4 cores:

```{r}
our_sim <- new_simulation("UndoGardnerella", "Undo Gardnerella relative abundance") %>%
  generate_model(undo_efficiencies,
                 ps = cal %>%
                   filter_sample_data(cohort == "Stanford") %>%
                   tax_glom("genus") %>%
                   as_tibble,
                 sig_sq_e = as.list(seq(0, 5, length.out = 6)),
                 vary_along = "sig_sq_e") %>%
  simulate_from_model(nsim = 4, index = 1) %>%
  # simulate_from_model(nsim = 25, index = 1:4) %>%
  run_method(list(corncob_Gardnerella)) %>%
  # run_method(list(corncob_Gardnerella),
  #            parallel = list(socket_names = 4, libraries = c("corncob", "speedyseq"))) %>%
  evaluate(list(betahat_fit, pvalue_fit))
```

Here's how the estimated coefficient on preterm birth changes under different draws from a distribution given by $\text{efficiency}_{i}} \overset{iid}{\sim} LogNormal(0, \sigma^2_e)$, and how the p-value for testing the null hypothesis that the coefficient on preterm birth is zero:

```{r}
ev_df <- our_sim %>% evals %>% as.data.frame
model_df <- our_sim %>% model %>% as.data.frame
ev_with_model_params <- dplyr::right_join(model_df, ev_df, by = c("name" = "Model")) %>% tibble
f_names <- list('p-value' = "p-value",
                'beta-hat' = expression(hat(beta)))
f_labeller <- function(variable, value){return(f_names[value])}
ev_with_model_params %>%
  pivot_longer(cols = betahat:pvalue, names_to="eval") %>%
  mutate(eval = ifelse(eval == "pvalue", "p-value", "beta-hat")) %>%
  ggplot(aes(x = sig_sq_e, group = sig_sq_e, y = value)) +
  geom_boxplot() +
  geom_abline(aes(slope = slope, intercept = intercept), color = "blue",
              tibble(eval = c("p-value", "beta-hat"), slope = c(0,0), intercept = c(0.05, 0))) +
  theme_bw() +
  xlab(expression(sigma[e]^2)) +
  ggtitle("Corncob results for testing\nH0: Mean Gardnerella relative abundance is equal for preterm & full-term") +
  facet_wrap(~eval, scales="free_y", nrow = 2, labeller = f_labeller) +
  ylab("")
```

While the estimate of the coefficient does not change on average, we see more variation in its estimate as the variation in the efficiencies increases. For some realizations from a model with large variation in the efficiencies, we even see a change in the sign of $\hat{\beta}$ from positive to negative. Furthermore, we see a corresponding increase in the range of p-values, suggesting that more variation in the efficiencies can lead to either deattenuated estimates (larger p-values) or more precise non-zero estimates (smaller p-values).

A major advantage of this approach is that any distribution of efficiency vectors can be investigated. For example, we could specifically investigate how Gardnerella being low- or high-efficiency impacts our results. We could also investigate more sophisticated distributions for the efficiencies, such as correlated efficiencies across phylogeny.

### 2021-10-19 index.Rmd
Add authors
Add Jacob Nearing and Amy Willis as authors

@@
   - Michael R. McLaren^[North Carolina State University; send correspondence to m.mclaren42@gmail.com]
+  - Jacob T. Nearing^[Dalhousie University]
+  - Amy D. Willis^[University of Washington]
   - Karen G. Lloyd^[University of Tennessee]
@@
 sha <- repo %>% git2r::commits(n = 1) %>% .[[1]] %>% git2r::sha()
+status <- repo %>% git2r::status()
+```
+
+```{r}
+if (length(status$unstaged) > 0)
+  print(status$unstaged)
 ```

### 2021-10-19 abundance-measurement.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

--- 2021-09-09 abundance-measurement.Rmd
@@
 
+## Model of MGS measurement
+
 Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
 This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
-In this model, we suppose there are a set of *atomic taxa*, that differ in how efficiently they are measured but within which all cells behave similarly.
-For concreteness, we equate atomic taxa with species. 
-We ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.
-
-Our model stipulates that the assigned read count of species $i$ in sample $a$ equal equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
+We consider a set of microbial species that are measured by a specific MGS protocol.
+The protocol assigns each sequencing read to at most one sample andsample and  species, and we ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.
+Our model stipulates that the assigned read count of species $i$ in sample $a$ equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
 \begin{align}
@@
 Although other interpretations exist^[e.g.
-the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most DA analysis methods (and microbiome analysis in general), for both relative and absolute abundances, and the differences between them help to clarify the issue over when and why the effects of bias cancel in an analysis.
+the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most relative  and absolute DA methods and the differences between them clarify when and why the effects of bias cancel in a DA analysis.
 
-The standard estimate of the proportion of species $i$ in sample $a$ given by the ratio of its read count to the total,
-\begin{align*}
-  \hat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
-\end{align*}
-The effect of taxonomic bias on this estimate is to create a multipliative error equal to the ratio of the species' efficiency to the mean efficiency in the sample,
+The standard measurement of the proportion of species $i$ in sample $a$ is given by the ratio of its read count to the total,
+\begin{align}
+  (\#eq:prop-meas)
+  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
+\end{align}
+From Equations \@ref(eq:measurement-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas), it follows that
 \begin{align}
   (\#eq:prop-error)
-  \hat{\text{prop}}_{i}(a)
-  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)}.
+  \widehat{\text{prop}}_{i}(a)
+  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)},
 \end{align}
-Intuitively, Equation \@ref(eq:prop-error) says that a species is measured as having a larger-than-actual proportion in direct relation to how much more more efficiently its cells are measured than the average cell in the sample (Figure \@ref(fig:error-proportions)).
-Consequently, the same species can be over or under measured in terms of its proportion.
-For instance, in samples from two hypothetical communities in Figure \@ref(fig:error-proportions), Species 3 has an efficiency of 6 and is under-estimated in Sample 1 (which has a mean efficiency of 8.33) but over-estimated in Sample 2 (which has a mean efficiency of 3.15).
+where $\text{prop}_{i}(a) = \text{density}_{i}(a) / \text{total density}(a)$ is the actual proportion.
+Equation \@ref(eq:prop-error) says that taxonomic bias creates a multiplicative error equal to the ratio of the species' efficiency to the mean efficiency in the sample.
+Consequently, the same species can be over- or under-measured depending on the composition of the given sample (Figure \@ref(fig:error-proportions)).
+For instance, in samples from two hypothetical communities in Figure \@ref(fig:error-proportions), Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).
 
-The estimated ratio between species $i$ and $j$, given by the ratio of read counts, 
-\begin{align*}
-  \hat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)},
-\end{align*}
-has a multiplicative error equal to the ratio in the species' efficiencies,
+The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
+\begin{align}
+  (\#eq:ratio-meas)
+  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
+\end{align}
+From Equations \@ref(eq:measurement-model) and \@ref(eq:ratio-meas), it follows that 
 \begin{align}
   (\#eq:ratio-error)
-  \hat{\text{ratio}}_{i/j}(a)
+  \widehat{\text{ratio}}_{i/j}(a)
 %  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
@@
 \end{align}
-Equation \@ref(eq:ratio-error) says that the ratio between two species is over-estimated in direct relation to how much more efficiently cells of the first species is measured relative to the second.
-In contrast to estimates of proportions, estimates of species ratios have a consistent multiplicative error across samples.
+The fold error in the ratio between two species simply equals the ratio in their efficiencies and is therefore constant across samples.
 For instance, in the example of Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
@@
 } else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
+    "error-proportions.pdf")
+  args <- c(
+    "--export-area-page", 
+    "--export-type=pdf", 
+    str_c("--export-filename=", pdf_path),
+    svg_path
+  )
+  system2("inkscape", args = args)
+  knitr::include_graphics(pdf_path)
+  rm(pdf_path)
 }
@@
 (ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species.
-Bottom row: Error in the estimated fold-change in the third species that is derived from these measurements.
+Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.
 
@@
 
-Various experimental and computational methods have or can be used to supplement MGS measurements so as to convert the relative-abundance information in the read counts into estimates of (absolute) cell densities;
-however, these methods can largely be partitioned conceptually into those that use the estimated proportions or the estimated ratios provided by MGS.
-The error in density estimates due to taxonomic bias therefore follows that of the proportion and ratio estimates described above.
+A range of experimental and computational methods can be used to convert the relative-abundance information in the read counts into measurements of (absolute) cell densities;
+however, these methods can largely be partitioned into those that use the measured proportions or the measured ratios provided by MGS.
+The error in density measurements due to taxonomic bias therefore follows that of the proportion and ratio measurements described above.
 
-Proportion-based density estimation is based on the idea that if we know the (true) proportion of a species and the (true) total density of all species, we can multiply these numbers obtain the species' density.
-This method therefore involves multiplying the proportion of a species estimated from MGS with an estimate of total microbial density obtain through other means (e.g., through direct cell counting or qPCR of a universal marker gene) to estimate
-its density,
+Proportion-based density measurement is based on the idea that if we know the (true) proportion of a species and the (true) total density of all species, we can multiply these numbers obtain the species' density.
+This method therefore involves multiplying the proportion of a species measured from MGS with a measurement of total microbial density obtain through other means (e.g., through direct cell counting or qPCR of a universal marker gene) to measure its density,
 \begin{align}
-  (\#eq:density-prop-est)
-  \hat{\text{density}}_{i}(a) 
-  = \hat{\text{prop}}_{i}(a) \cdot \hat{\text{total density}}(a).
+  (\#eq:density-prop-meas)
+  \widehat{\text{density}}_{i}(a) 
+  = \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{total density}}(a).
 \end{align}
-The error in this proportion-based density estimate is
+Substituting the expression \@ref(eq:prop-error) for the measured proportion and allowing for error in the total density measurement, leads to the expression
 \begin{align}
   (\#eq:density-prop-error)
-  \hat{\text{density}}_{i}(a) 
+  \widehat{\text{density}}_{i}(a) 
   = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
-  \cdot \substack{\text{fold error in} \\ \hat{\text{total density}}(a)}.
+  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{total density}}(a) \end{array}.
 \end{align}
@@
 \begin{align}
-  (\#eq:density-ratio-est)
-  \hat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \hat{\text{density}}_{r}(a)
+  (\#eq:density-ratio-meas)
+  \widehat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \widehat{\text{density}}_{r}(a)
 \end{align}
 (Appendix ??? describes how this equation can be extended to multiple reference species.)
-For a constant reference species with unknown density, we can set $\hat{\text{density}}_{r}(a)$ equal to 1 to estimate density in units of the density of the reference species.
+For a constant reference species with unknown density, we can set $\widehat{\text{density}}_{r}(a)$ equal to 1 to estimate density in units of the density of the reference species.
 The error in this ratio-based density estimate is
@@
   (\#eq:density-ratio-error)
-  \hat{\text{density}}_{i}(a) 
+  \widehat{\text{density}}_{i}(a) 
   = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
-  \cdot \substack{\text{fold error in} \\ \hat{\text{density}}_{r}(a)}.
+  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{density}_r}(a) \end{array}.
 \end{align}
@@
 
-- Note: The current distinction between proportion- and ratio-based density estimation is incomplete, as the example of spike-in normalization shows: Whether the density of species $i$ is estimated by Equation \@ref(eq:density-ratio-est), or by first estimating total density from the ratio of non-spike-in to spike-in reads and then using Equation \@ref(eq:density-prop-est), the results are mathematically identical.
+- Note: The current distinction between proportion- and ratio-based density estimation is incomplete, as the example of spike-in normalization shows: Whether the density of species $i$ is estimated by Equation \@ref(eq:density-ratio-meas), or by first estimating total density from the ratio of non-spike-in to spike-in reads and then using Equation \@ref(eq:density-prop-meas), the results are mathematically identical.
 Future revisions should clarify this point.

### 2021-10-19 differential-abundance.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

--- 2021-09-09 differential-abundance.Rmd
@@
 How do the measurement errors described in the previous section impact our ability to estimate the changes in microbial abundances across samples or between different host and environmental conditions?
-Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, as such inference is ubiquitous in microbiome DA analysis and have more direct ecological interpretations (via the processes of exponential growth and death) than other DA measures.
-These measures of DA are also affected by taxonomic bias in our model in a manner that is straightforward to understand in terms of the multiplicative errors described in the previous section.
+Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, which have direct ecological interpretations (via the processes of exponential growth and death) and are ubiquitous in microbiome DA analysis.
 
@@
 Before considering the stereotypical many-sample DA analysis, it is instructive to 
-consider simplest analysis of differential abundance: the estimation of fold changes in abundance between a pair of samples.
+consider simplest analysis of differential abundance: the measurement of fold changes in abundance between a pair of samples.
 This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment, and conceptually bridges the single-sample results of the previous section to the many-sample case.
@@
 <!-- P: error in FC proportions -->
-The composition-dependent effect of bias on fold error in proportions leads to error in fold-change estimates that is proportional to the inverse change in mean efficiency.
-From the error in an individual sample \@ref(eq:prop-error), it follows that the estimated fold change in proportion of taxon $i$ from sample $a$ to sample $b$ is
+The composition-dependent effect of bias on fold error in proportions leads to error in fold-change measurements that is proportional to the inverse change in mean efficiency.
+From the error in an individual sample (Equation \@ref(eq:prop-error)), it follows that the measured fold change in proportion of taxon $i$ from sample $a$ to sample $b$ is
 \begin{align}
@@
 % \tag*{Fold change in proportion}
-\underbrace{\frac{\hat{\text{prop}}_{i}(b)}{\hat{\text{prop}}_{i}(a)}} _\text{estimated FC}
+\underbrace{\frac{\widehat{\text{prop}}_{i}(b)}{\widehat{\text{prop}}_{i}(a)}} _\text{measured FC}
   &= \frac
@@
 We can see each type of error in Figure 2.
-For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the estimated FC (actual FC: 2.3X, measured FC: 6.5X).
+For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FC (actual FC: 2.3X, measured FC: 6.5X).
 For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
@@
 
-If other error sources remain negligible, than this dichotomy continues to apply to proportion- and ratio-based density estimates.
-The inferred FCs in proportion-based density estimates will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
+If other error sources remain negligible (or constant across samples), than this dichotomy continues to apply to proportion- and ratio-based density measurements.
+The inferred FCs in proportion-based density measurements will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
 Moreover, any species with a constant density will appear to vary inversely with the mean efficiency.
-In contrast, fold changes in ratio-based density estimates remain accurate.
+In contrast, fold changes in ratio-based density measurements will remain accurate.
 
@@
 DA analysis of many samples across host or environmental conditions can typically be framed as a regression problem, in which we analyze the relationship between a microbial _response variable_, such as log density of some focal species $i$, and one or more _covariates_, such the pH or temperature of the sampled environment or whether the sample is from a healthy or sick person.
-A substantial fraction of DA analyses use minor elaborations on the simple linear regression model, which for a response of log species density can be written
+A large fraction of DA analyses use elaborations on the simple linear regression model, which for a response of log species density can be written
 \begin{align}
@@
 
-Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \@ref(eq:density-prop-est)) with error-free estimates of the total density.
+Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \@ref(eq:density-prop-meas)) with error-free estimates of the total density.
 If the true log density follows the regression Equation \@ref(eq:regression), then it follows from Equation \@ref(eq:density-prop-error) the estimated log density equals 
@@
   (\#eq:regression-error)
-  \log \hat{\text{density}}_i(a)
+  \log \widehat{\text{density}}_i(a)
 %  &= \alpha + \beta x(a) + \varepsilon_i(a) + \log \text{efficiency}_i - \log \text{mean efficiency}(a)
@@
 The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$.
-Box \@ref(regression-error) mathematically describes this effect for estimation method of Ordinary Least Squares (OLS) and Maximum Likelihood Estimation (MLE) and Figure \@ref(fig:regression-example) illustrates using simulated data.
+Box \@ref(regression-error) mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \@ref(fig:regression-example) illustrates using simulated data.
 Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
@@
 } else if (knitr::is_latex_output()) {
-  cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
+    "regression-example.pdf")
+  args <- c(
+    "--export-area-page", 
+    "--export-type=pdf", 
+    str_c("--export-filename=", pdf_path),
+    svg_path
+  )
+  system2("inkscape", args = args)
+  knitr::include_graphics(pdf_path)
+  rm(pdf_path)
 }

### 2021-10-19 introduction.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

@@
 <!-- First paragraph: DA analysis -->
-
-The most basic question we can ask about microbial communities after which taxa are found in an ecosystem is, how do the abundances of these taxa vary in their abundance—across space, time, and host or environmental conditions?
-Advances in sequencing technology allow us to now simultaneously measure the abundances of 100s to 1000s of species using marker-gene and shotgun-metagenomic sequencing (jointly, MGS).
+One of the most basic questions we can ask about microbial communities is: How do different microbial taxa vary in abundance—across space, time, and host or environmental conditions?
+Advances in sequencing technology allow us to simultaneously measure the abundances of 100s to 1000s of species using marker-gene and shotgun metagenomic sequencing (jointly, MGS).
 Although standard MGS measurements lose information about total microbial density—and so are typically used to analyze the abundances of taxa relative to each other or their total—new studies are increasingly employing strategies to enable the analysis of cell density or other measures of "absolute abundance."
@@
 In particular, MGS measurements are _taxonomically biased_: Taxa vary dramatically (e.g.
-10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—by a given MGS protocol.
-As a result, the abundance measurements obtained by MGS are inaccurate representations of the actual abundances and also tend to differ across protocols (@mclaren2019cons), studies, and even experimental batches (@yeh2018taxo).
+10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—by a given MGS protocol (@mclaren2019cons).
+As a result, the abundance measurements obtained by MGS are inaccurate representations of the actual abundances and also tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo,@mclaren2019cons).
 This bias arises from variation in how taxa respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
@@
 In broad strokes, the thinking is that the measurements of samples measured by the same protocol will be affected by bias in the same way and so the inferred differences between samples (the focus of DA analysis) will be unaffected.
-For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples.
-@kevorkian2018esti and @lloyd2020evid were the first (to our knowledge) to make this argument explicit mathematically; however, @mclaren2019cons used a theoretical model (supported by measurements of defined bacterial communities) to show that consistent taxonomic bias can lead to variable fold errors in measured proportions.
+For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples (@kevorkian2018esti,@lloyd2020evid).
+However, @mclaren2019cons used a theoretical model (supported by measurements of defined bacterial communities) to show that consistent taxonomic bias can lead to variable fold errors in measured proportions.
 These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase).
-Yet @mclaren2019cons also found that certain changes in relative abundance---in particular, fold changes in the ratios among species---are robust to the effects of bias.
+Yet @mclaren2019cons also found that the fold error in the ratios among species was constant, making fold changes in ratios robust to bias.
 The implications of these findings for changes in absolute abundance---which remain subject to taxonomic bias in the underlying MGS measurement---and for DA analysis across many species and many samples---as commonly done in microbiome association testing---have yet to be investigated.
@@
 Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.
-Over time, application of these methods to past and future experiments will provide crucial quantitative information about taxonomic bias and the conditions under which spurious results arise for various DA methodologies.
+Over time, application of these methods to past and future experiments will provide crucial quantitative information about the conditions under which taxonomic bias creates spurious results for various DA methodologies.
 Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.

### 2021-10-19 real-world-implications.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

@@
 
-Partial but not full canceling of the effect of bias in proportion-based analyses, such that possible to have spurious results in principle; but what about in practice?
-Predicting how frequently or which results are likely to be majorly affected is difficult because it depends on the joint variation in efficiencies of species and species abundance dynamics, both of which are poorly understood in most to all biological systems.
-However, through a combination of theoretical considerations, hypothetical scenarios, and real case studies we can begin to develop an understanding or the relevant considerations and build framework for improving our understanding as our knowledge of these dynamics improves.
+Are biologically significant errors in DA analyses likely in practice?
+We can begin to answer this question using a combination of theoretical analysis, consideration of hypothetical scenarios, and analysis of case studies for which some information about taxonomic bias is available.
 
@@
 SI Analysis (MOMS-PI) uses these estimated efficiencies to correct the compositions and explore the implied variation in mean efficiency in the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) @fettweis2019thev.
-The vaginal microbiome during pregnancy is characterized by relatively low diversity, often with a single _Lactobacillus_ species often forming a majority of reads (considered to be the healthy state) but with occasional shifts to high proportions of other species, such as _Gardnerella vaginalis_, _Atopobium vaginae_, and _Streptococcus agalactiae_, which are thought to cause or indicate poor vaginal health and increased risk of preterm birth.
+The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single _Lactobacillus_ species often forming a majority of reads (considered to be the healthy state) but with occasional shifts to high proportions of other species, such as _Gardnerella vaginalis_, _Atopobium vaginae_, and _Streptococcus agalactiae_, which are thought to cause or indicate poor vaginal health and increased risk of preterm birth.
 The analysis of @mclaren2019cons showed that the efficiencies of two common _Lactobacillus_ species, _L.
@@
 A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
-Shared evolutionary history among phylogenetically-close species is expected and has been observed to create positive associations in these traits among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
+Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
 Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
@@
 It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
-As the relative abundance of tougher species increases with depth, the mean extraction efficiency decreases.
+As the relative abundance of tougher species increases with depth, the mean extraction efficiency might therefore decrease.
 This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
@@
 It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
-(This seems intuitively likely and true in examples, but I haven't analyzed it much mathematically and don't have a firm handle on this question.)
+_(This seems intuitively likely and true in examples, but I haven't analyzed it much mathematically and don't have a firm handle on this question.)_
 Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
-However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for all other species.
+However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for other species.
 The experimental and computational methods discussed in Section \@ref(solutions) can mitigate taxonomic bias in these clearly-problematic cases while also helping to amass empirical evidence on the conditions in which bias is unlikely to cause major inferential errors.

### 2021-10-19 solutions.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

@@
 We present five that seem particularly promising:
-The first three are method for correcting or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.
+The first three are methods to correct or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.
 
@@
 The most complete but experimentally challenging approach to mitigating taxonomic bias involves directly measuring the efficiencies of individual species using control samples (@mclaren2019cons).
-Suitable controls are any community sample whose composition (species identity and relative abundances) has been previously characterized, either by construction (in the case of constructed 'mock' communities) or by measurement with a _reference protocol_ chosen to serve as a measurement standard.
+Suitable controls are any community sample whose composition (species identities and relative abundances) has been previously characterized, either by construction (in the case of constructed 'mock' communities) or by measurement with a _reference protocol_ chosen to serve as a measurement standard.
 These controls, when measured alongside the target experimental samples, can be used to estimate the relative efficiencies among the superset of species across the control samples.
-The estimated efficiencies can then be used to calibrate the measured relative abundances among the control species in the target samples to the chosen measurement standard, by simply dividing the species' read counts by their corresponding efficiencies or using a more sophisticated statistical modeling framework.
+The estimated efficiencies can then be used to calibrate the measured relative abundances among the control species in the target samples to the chosen measurement standard, by simply dividing the species' read counts by their corresponding efficiencies (@mclaren2019cons) or using a more sophisticated statistical modeling framework.
 These calibrated compositions can then be used for arbitrary downstream microbiome analyses, including DA analysis.
 Calibration of species not in the controls requires first imputing their efficiencies from the control measurements, for example using phylogenetic relationships, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
-Though such imputation is conceptually straightforward to implement, its reliability and utility remain untested.
+Though such imputation is conceptually straightforward, its accuracy and ultimate utility remain untested.
 
 - The above states the approach; can add description/discussion of application to the MOMS-PI and @leopold2020host cases and developments and challenges in availability of control communities
+- 'more sophisticated modeling framework' - cite @williamson2021amul and https://microbiomemeasurement.org/posts/bias-estimation-from-mocks-with-fido/
 

### 2021-10-25 notebook/_posts/2021-10-25-brooks2015thet-calibration/brooks2015thet-calibration.Rmd
Illustrate calibration in the Brooks dataset

---
title: "Illustrate calibration types in the mock communities from Brooks et al (2015)"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:brooks2015thet
date: 2021-10-25
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300
)
```

# Setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
theme_set(theme_cowplot())
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
```

```{r}
colors_brooks <- c(
  "Atopobium_vaginae" = "#009E73",
  "Gardnerella_vaginalis" = "#56B4E9",
  "Lactobacillus_crispatus" = "#D55E00",
  "Lactobacillus_iners" = "#505050",
  "Prevotella_bivia" = "#0072B2",
  "Sneathia_amnii" = "#CC79A7",
  "Streptococcus_agalactiae" = "#E69F00")

scale_y_custom <- scale_y_continuous(
    trans = 'log10',
    breaks = trans_breaks('log10', function(x) 10^x),
    labels = trans_format('log10', math_format(10^.x))
  )
```

```{r}
# Load data from the cellular mock communities of Brooks et al 2015
dr <- system.file("extdata", package = "metacal")
list.files(dr)
actual <- file.path(dr, "brooks2015-actual.csv") |>
  read.csv(row.names = "Sample") |>
  otu_table(taxa_are_rows = FALSE)
observed <- file.path(dr, "brooks2015-observed.csv") |>
  read.csv(row.names = "Sample") |>
  subset(select = - Other) |>
  otu_table(taxa_are_rows = FALSE)

# Estimate bias with bootstrapping for error estimation
mc_fit <- estimate_bias(observed, actual, boot = TRUE)
control_species <- mc_fit %>% coef %>% names
summary(mc_fit)
```

are the single-species samples present?

```{r}
actual %>% 
  as_tibble %>%
  filter(.abundance == 1) %>%
  count(.otu)
```

Yes.

# Calibration

## Calibration from community control

Demonstrate that calibration using a single mock community can create a substantial improvement.

```{r}
smpls <- actual %>%
  as_tibble %>%
  with_groups(.sample, summarize, n_species = sum(.abundance > 0)) %>%
  filter(n_species == 7) %>%
  pull(.sample)
```

```{r}
control_samples <- smpls[1]
bias_from_control <- estimate_bias(
  observed %>% prune_samples(control_samples, .),
  actual %>% prune_samples(control_samples, .),
  ) %>%
  coef
calibrated <- calibrate(observed, bias_from_control)
```

## Calibration using a reference species

One way to do this is to get the correction for each pair of samples, then join it with the table from previous.
Another is to make the calibrated OTU table, then redo the pairwise ratio stuff.

Basic Idea: Multiply the ratio of focal to ref species by the true proportion of the ref.
One way to achieve this with phyloseq objects is to construct the correction matrix, then multiply this by the observed proportions

```{r}
ref_species <- 'Lactobacillus_crispatus'

ref_actual <- actual %>% prune_taxa(ref_species, .) %>% c
ref_observed <- observed %>% prune_taxa(ref_species, .) %>% c

correction_matrix <- matrix(
  ref_actual / ref_observed, 
  nrow = nsamples(observed), ncol= ntaxa(observed),
  byrow = FALSE
)

calibrated_ref <- otu_table(observed * correction_matrix, taxa_are_rows = FALSE)
```

## Plot comparing calibrated and uncalibrated fold changes

```{r}
x <- list(
  Actual = actual, 
  Uncalibrated = observed %>% transform_sample_counts(close_elts),
  Calibrated_community = calibrated,
  Calibrated_reference = calibrated_ref
  ) %>%
  map(pairwise_ratios, margin = 'samples', filter = FALSE) %>%
  map_dfr(as_tibble, .id = 'type')
```

```{r}
lvls <- mc_fit %>% coef %>% sort %>% names

brooks2015_fcs <- x %>%
  pivot_wider(names_from = type, values_from = .abundance) %>%
  pivot_longer(contains('calibrated'), 
    names_to = 'type', values_to = 'Measured'
  ) %>%
  separate(.sample, c('sample1', 'sample2'), sep = ':') %>%
  # Filter to remove cases where Actual is 0 in one or both samples, or where
  # the numerator and denominator sample are equal
  filter(is.finite(Actual), Actual != 0, sample1 != sample2) %>%
  mutate(
    across(type, fct_relevel, 'Uncalibrated'),
    across(
      type, fct_recode,
      'Calibrated (community)' = 'Calibrated_community',
      'Calibrated (reference)' = 'Calibrated_reference'
      ),
    across(.otu, factor, levels = rev(lvls)),
    # across(.otu, fct_relabel, adjust_species_names)
  )
```

Check on number of rows in these data frames,

```{r}
# expected in 'x'
n1 <- observed %>% nsamples %>% print
n2 <- observed %>% ntaxa
all.equal(n1 * n1 * n2 * 4, x %>% nrow)
x %>% nrow
brooks2015_fcs %>% nrow
```

Let's plot just the sample-pairs with L. crispatus present, so that the same samples are shown in all panels.

```{r brooks2015thet_fc_calibration, fig.dim = c(7, 5.5), dev = 'png'}
# Find subset of pairs to plot
sample_pairs <- brooks2015_fcs %>%
  filter(is.finite(Actual), Actual > 0, is.finite(Measured), Measured > 0) %>%
  select(sample1, sample2) %>%
  unite(pair, sample1, sample2) %>%
  distinct %>%
  pull(pair)

# Species labels
species_label <- function(species) {
  species %>%
    str_replace('(?<=^[A-Z])[a-z]+_', '. ')
}
species_label(lvls)

p <- brooks2015_fcs %>%
  unite(pair, sample1, sample2) %>%
  filter(pair %in% sample_pairs, is.finite(Actual), Actual > 0, is.finite(Measured), Measured > 0) %>%
  ggplot(aes(Actual, Measured, color = .otu)) +
  scale_color_manual(
    values = colors_brooks, 
    labels = species_label(colors_brooks %>% names)
  ) +
  facet_wrap(~type) +
  scale_x_log10() +
  scale_y_custom +
  geom_abline(color = 'grey', size = 1) +
  geom_point(position = position_jitter(), alpha = 0.5) +
  coord_fixed() +
  labs(x = 'Actual fold change', y = 'Measured fold change', color = 'Species') +
  theme(legend.position = 'bottom')
p
```

### 2021-10-25 notebook/_posts/2021-10-18-evaluate-robustness-example/evaluate-robustness-example.Rmd
change output from html to distill
Edit and render analysis
- Add reference and information about the source data
- Add 'Data setup' section with code to download and clean the source
  data
- Add session info
- Adjust the appearance of the summary plot
- Make minor text edits
- Run analysis with 100 iterations and render output to
  Distill-formatted html page

@@
   - name: Amy D. Willis
+categories:
+  - bias sensitivity
+  - ref:callahan2017repl
 date: 10-18-2021
-output: html_document
+bibliography: ../../../main.bib
+output:
+  distill::distill_article:
+    self_contained: false
+    toc: true
+    dev: svg
 ---
@@
 \begin{align}
-  (\#eq:calibrate-props)
   \text{calibrated prop}_{i}(a)
@@
 \begin{align}
-  (\#eq:calibrate-reads)
   \text{reads}_{i}(a)
@@
 
-## Example: relative abundance of Gardnerella
+## Example: relative abundance of _Gardnerella_
 
-For example, suppose we are interested in investigating if the genus Gardnerella is more abundant in the vaginal microbiome of women who give birth preterm, and we investigate this model by running `corncob` using preterm birth as a covariate in the mean and dispersion model.
+We illustrate this method using vaginal microbiome data from pregnant women published in a study by @callahan2017repl seeking to replicate associations of specific genera (_Lactobacillus_, _Gardnerella_, and _Ureaplasma_) with preterm birth in two cohorts.
+For example, suppose we are interested in investigating if the genus _Gardnerella_ is more abundant in the vaginal microbiome of women who give birth preterm, and we investigate this model by running `corncob` using preterm birth as a covariate in the mean and dispersion model.
 We could investigate the robustness of our findings using the following structure:
@@
 
-First, we load the relevant libraries and our data
+First, we load the relevant libraries and our data (see [Data setup]),
 
@@
 library(simulator)
-cal <- readRDS("callahan2017repl_phyloseq_clean.rds")
+cal <- readRDS(
+  here::here("notebook/_data", "callahan2017repl", "callahan2017repl_phyloseq_clean.rds")
+)
 ```
@@
 Next, we specify how we use our data to construct our estimates.
-We are using `bbdml` from the library `corncob` to model the relative abundance of ASV 8 (Gardnerella) using preterm birth as a covariate to model the mean and dispersion, but this is only one type of analysis that we might be interested in.
+We are using `bbdml` from the library `corncob` to model the relative abundance of ASV 8 (_Gardnerella_) using preterm birth as a covariate to model the mean and dispersion, but this is only one type of analysis that we might be interested in.
 
@@
 Now that we have all of the pieces, we can run them together as follows.
-Note that we are only going to run 4 iterations for now, but we have commented out code below showing you how to run 100 draws across 4 cores:
+The below code runs 100 simulations split across 4 cores; 
+for testing, the commented code runs just 4 iterations.
 
@@
                  vary_along = "sig_sq_e") %>%
-  simulate_from_model(nsim = 4, index = 1) %>%
-  # simulate_from_model(nsim = 25, index = 1:4) %>%
-  run_method(list(corncob_Gardnerella)) %>%
-  # run_method(list(corncob_Gardnerella),
-  #            parallel = list(socket_names = 4, libraries = c("corncob", "speedyseq"))) %>%
+  # simulate_from_model(nsim = 4, index = 1) %>%
+  # run_method(list(corncob_Gardnerella)) %>%
+  simulate_from_model(nsim = 25, index = 1:4) %>%
+  run_method(list(corncob_Gardnerella),
+             parallel = list(socket_names = 4, libraries = c("corncob", "speedyseq"))) %>%
   evaluate(list(betahat_fit, pvalue_fit))
@@
 
-Here's how the estimated coefficient on preterm birth changes under different draws from a distribution given by $\text{efficiency}_{i}} \overset{iid}{\sim} LogNormal(0, \sigma^2_e)$, and how the p-value for testing the null hypothesis that the coefficient on preterm birth is zero:
+Here's how the estimated coefficient on preterm birth changes under different draws from a distribution given by $\text{efficiency}_{i} \overset{iid}{\sim} \text{LogNormal}(0, \sigma^2_e)$, and how the p-value for testing the null hypothesis that the coefficient on preterm birth is zero:
 
@@
 f_labeller <- function(variable, value){return(f_names[value])}
+```
+
+```{r summary_plot, fig.dim = c(6,5) * 1.2}
 ev_with_model_params %>%
@@
               tibble(eval = c("p-value", "beta-hat"), slope = c(0,0), intercept = c(0.05, 0))) +
-  theme_bw() +
   xlab(expression(sigma[e]^2)) +
-  ggtitle("Corncob results for testing\nH0: Mean Gardnerella relative abundance is equal for preterm & full-term") +
+  # ggtitle("Corncob results for testing\nH0: Mean Gardnerella relative abundance is equal for preterm & full-term") +
   facet_wrap(~eval, scales="free_y", nrow = 2, labeller = f_labeller) +
-  ylab("")
+  ylab("") +
+  cowplot::theme_cowplot()
 ```
@@
 A major advantage of this approach is that any distribution of efficiency vectors can be investigated.
-For example, we could specifically investigate how Gardnerella being low- or high-efficiency impacts our results.
+For example, we could specifically investigate how _Gardnerella_ being low- or high-efficiency impacts our results.
 We could also investigate more sophisticated distributions for the efficiencies, such as correlated efficiencies across phylogeny.
+
+# Data setup {.appendix}
+
+The following code was used to download and prepare the data from @callahan2017repl for analysis; it is not evaluated when this document is knit.
+The first code chunk downloads the ASV abundance matrix, sample metadata, and taxonomy assignments for the vaginal samples.
+
+```{r import_data, eval = FALSE}
+library(tidyverse)
+library(speedyseq)
+
+# Download --------------------------------------------------------------------
+dl_path <- here::here("notebook/_data", "callahan2017repl")
+link <- "https://stacks.stanford.edu/file/druid:yb681vm1809/RepRefine_Scripts.tar.gz"
+if (!dir.exists(dl_path)) {
+  dir.create(dl_path)
+  download.file(link, file.path(dl_path, basename(link)))
+  system2("tar", args = c("-xvf", file.path(dl_path, basename(link)), "-C", dl_path))
+}
+load(file.path(dl_path, "RepRefine_Scripts", "input", "processed.rda"))
+# This loads:
+# df: the sample metadata
+# st: the sequence table (in counts)
+# ft: the sequence table normalized to frequencies / proportions
+# tax: taxonomy assignment matrix
+# 
+# To confirm relationship between ft and st, run 
+# all.equal(ft, st %>% apply(1, function(x) x / sum(x)) %>% t)
+# 
+# The taxonomy matrix has an extra column with a duplicate name of "Species"
+# that is all NAs, so let's remove that
+tax <- tax[, -8]
+# Import into phyloseq --------------------------------------------------------
+sam <- sample_data(df)
+st <- otu_table(st, taxa_are_rows = FALSE)
+tax <- tax_table(tax)
+ps <- phyloseq(sam, st, tax)
+# Check that all samples and taxa made it into the phyloseq object
+stopifnot(all.equal(colnames(st), taxa_names(ps)))
+stopifnot(all.equal(rownames(sam), sample_names(ps)))
+# Add ASV sequences to the refseq() slot
+seqs <- taxa_names(ps) %>% rlang::set_names() %>% Biostrings::DNAStringSet()
+ps <- merge_phyloseq(ps, seqs)
+# Rename the ASVs to simpler names (ASV1, ASV2, ...). The full sequences remain
+# in `refseq(ps)`.
+taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
+# Clean up objects
+rm(tax, df, sam, st, ft, seqs)
+```
+
+Next, we use munging functions from janitor and speedyseq to select the metadata of primary interest, standardize format of column names, and convert categorical variables to factors.
+
+```{r, eval = FALSE}
+ps1 <- ps %>% 
+  select_sample_data(SubjectID, Cohort, Race, preterm, Outcome) %>%
+  rename_with_sample_data(janitor::make_clean_names) %>%
+  rename_with_tax_table(janitor::make_clean_names) %>%
+  mutate_sample_data(
+    across(c(subject_id, cohort, race, outcome), factor),
+  )
+```
+
+Finally, we save the phyloseq object for use in the analysis.
+
+```{r, eval = FALSE}
+saveRDS(ps1, file.path(dl_path, "callahan2017repl_phyloseq_clean.rds"))
+rm(ps, ps1)
+```
+
+# Session info {.appendix}
+
+<details><summary>Click for session info</summary>
+```{r, R.options = list(width = 83)}
+sessioninfo::session_info()
+```
+</details>

### 2021-10-27 solutions.Rmd
Add figures to solutions section

@@
 The first three are methods to correct or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.
+
+<!-- begin figure -->
+
+```{r calibration-example, fig.cap = '(ref:cap-calibration-example)'}
+fig_path <- here::here(
+  "notebook/_posts/2021-10-25-brooks2015thet-calibration",
+  "brooks2015thet-calibration_files/figure-html5/brooks2015thet_fc_calibration-1.png"
+)
+knitr::include_graphics(fig_path)
+rm(fig_path)
+```
+
+(ref:cap-calibration-example) **Fold changes can be calibrated using community controls or reference species.** The figure compares the performance of three methods for measuring fold changes in absolute cell density in cellular mock communities of 7 vaginal species, which were constructed and measured via 16S sequencing by @brooks2015thet.
+The 'Uncalibrated' fold changes are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which here is be known to be constant by construction).
+The 'Calibrated (community)' measurements are computed from abundance measurements where the proportions are first corrected for the taxonomic bias that was estimated from a single sample that contained all 7 species.
+The 'Calibrated (reference)' measurements are computed from abundances measured with the reference-species method, with _Lactobacillus crispatus_ used as the reference; that is, the true abundance of _L.
+crispatus_ is treated as known and used to infer the abundance of the remaining 6 species.
+Only samples that contain _L.
+crispatus_ are included.
+
+<!-- end figure -->
 
@@
 Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by @mclaren2019cons, which can increase improve our ability to find reproducible microbial associations in a given meta-study.
+
+## Supplemental figures
+
+<!-- begin figure -->
+
+```{r sensitivity-example, fig.cap = '(ref:cap-sensitivity-example)'}
+svg_path <- here::here(
+  "notebook/_posts/2021-10-18-evaluate-robustness-example/",
+  "evaluate-robustness-example_files/figure-html5/summary_plot-1.svg"
+)
+if (knitr::is_html_output()) {
+  knitr::include_graphics(svg_path)
+} else if (knitr::is_latex_output()) {
+  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
+    "sensitivity-example.pdf")
+  args <- c(
+    "--export-area-page", 
+    "--export-type=pdf", 
+    str_c("--export-filename=", pdf_path),
+    svg_path
+  )
+  system2("inkscape", args = args)
+  knitr::include_graphics(pdf_path)
+  rm(pdf_path)
+}
+rm(svg_path)
+```
+
+(ref:cap-sensitivity-example) **Example of a bias sensivity analysis.**
+
+<!-- end figure -->

### 2021-10-27 theory/derivations-regression.Rmd
Split original theory out from the manuscript
The original theoretical calculations, which have been put in the
appendix, have largely become redundant and/or superfluous to the
manuscript. In order to improve repo organization and streamline the
rendered manuscript, this commit moves these files to a new theory/
folder and removes them from the rendered manuscript. The theory files
remain structured as a separate Bookdown article.

--- 2021-05-05 appendix-regression.Rmd
@@
-# (APPENDIX) Appendix {-}
-
 # Proofs of regression results
@@
 
-```{theorem, label = "regression-linearity", name = "Linearity of regression coefficients", echo = TRUE}
+::: {.theorem #regression-linearity name="Linearity of regression coefficients"}
 Consider a scalar response variable with $J$ observations, which I represent by the vector $y$, that equals a sum of $K$ component response variables $y^{(k)}$ each scaled by a non-zero factor $c^{(k)}$,
@@
 \end{align}
-```
+:::
 
-```{proof, echo = TRUE}
+::: {.proof}
 The Moore-Penrose pseudoinverse of $X$ is $X^+ = (X^T X)^{-1} X^T$.
@@
 \end{align}
-```
+:::
 
@@
 
-```{theorem, label = "regression-error", echo = TRUE, name = "Error in regression coefficients"}
+::: {.theorem #regression-error name="Error in regression coefficients"}
 Suppose that $y$ is the response variable we wish to understand, $z$ is our imperfect measurement of $y$, and $d = z - y$ the difference between the two, so that $z = y + d$.
@@
 \end{align}
-```
+:::
 
-```{proof, echo = TRUE}
+::: {.proof}
 The result follows directly from $z =  y + d$ and \@ref(thm:regression-linearity).
-```
+:::
 
-```{corollary, label = "regression-error-simple", echo = TRUE}
+::: {.corollary #regression-error-simple}
 For the special case of simple linear regression,
@@
 
-```{proposition, echo = TRUE}
+::: {.proposition}
 Consider the linear equations
@@
 \end{align}
-```
-
-```{proof, echo = TRUE}
-```
+:::

### 2021-10-27 theory/index.Rmd
Split original theory out from the manuscript
The original theoretical calculations, which have been put in the
appendix, have largely become redundant and/or superfluous to the
manuscript. In order to improve repo organization and streamline the
rendered manuscript, this commit moves these files to a new theory/
folder and removes them from the rendered manuscript. The theory files
remain structured as a separate Bookdown article.

---
title: "Theory for taxonomic bias and differential abundance"
author:
  - Michael R. McLaren^[North Carolina State University; send correspondence to m.mclaren42@gmail.com]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
bibliography: [../main.bib, ../packages.bib]
biblio-style: apalike
link-citations: yes
# description:
# For pandoc latex processing
documentclass: article
geometry:
  - top=1in
  - bottom=1in
  - left=1.5in
  - right=1.5in
output:
  bookdown::gitbook:
    css: ../style.css
    number_sections: true
    includes:
      in_header: ../header.html
    config:
      sharing:
        facebook: no
        github: yes
        twitter: no
        linkedin: no
        weibo: no
        instapaper: no
        vk: no
        all: ['facebook', 'twitter', 'linkedin', 'weibo', 'instapaper']
  bookdown::pdf_book:
    includes:
      in_header: ../latex/preamble.tex
    keep_tex: true
    dev: "pdf"
    papersize: letter
    toc_depth: 2
    toc_unnumbered: yes
    toc_appendix: yes
---

# Preface {-}

```{r}
repo <- git2r::repository(".")
sha <- repo %>% git2r::commits(n = 1) %>% .[[1]] %>% git2r::sha()
status <- repo %>% git2r::status()
```

```{r}
if (length(status$unstaged) > 0)
  print(status$unstaged)
```

_This manuscript was rendered from commit `r sha`._

::: {#preface-warning .rmdnote}

**This in-progress manuscript is not intended for general scientific use.**
It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
Please post comments or questions on the [GitHub Issues page](https://github.com/mikemc/differential-abundance-theory/issues) or [email Mike](m.mclaren42@gmail.com).

:::

This Bookdown document contains theoretical calculations which supported the development of the manuscript.

### 2021-10-27 notebook/_posts/2021-10-27-illustrate-the-problem/illustrate-the-problem.Rmd
Illustrate the basic problem
Rmd notebook post illustrating the fact that the error in proportions is
not necessarily consistent, and hence the fold changes are not
necessarily accurate. This post looks at an example in the Brooks
dataset, but also briefly explores possible illustrations using the
Leopold dataset, including a look at ratios.

---
title: "Illustrate the basic problem"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:brooks2015thet
  - ref:leopold2020host
date: 2021-10-27
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300
)
```

# R setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
theme_set(theme_cowplot())
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
```

# Brooks et al

## Setup

```{r}
colors_brooks <- c(
  "Atopobium_vaginae" = "#009E73",
  "Gardnerella_vaginalis" = "#56B4E9",
  "Lactobacillus_crispatus" = "#D55E00",
  "Lactobacillus_iners" = "#505050",
  "Prevotella_bivia" = "#0072B2",
  "Sneathia_amnii" = "#CC79A7",
  "Streptococcus_agalactiae" = "#E69F00")

scale_y_custom <- scale_y_continuous(
    trans = 'log10',
    breaks = trans_breaks('log10', function(x) 10^x),
    labels = trans_format('log10', math_format(10^.x))
  )
```

Load data from the cellular mock communities of Brooks et al 2015 from metacal,

```{r}
dr <- system.file("extdata", package = "metacal")
list.files(dr)
actual <- file.path(dr, "brooks2015-actual.csv") |>
  read.csv(row.names = "Sample") |>
  otu_table(taxa_are_rows = FALSE)
observed <- file.path(dr, "brooks2015-observed.csv") |>
  read.csv(row.names = "Sample") |>
  subset(select = - Other) |>
  otu_table(taxa_are_rows = FALSE)
```

Note that the single-species samples present.

```{r}
actual %>% 
  as_tibble %>%
  filter(.abundance == 1) %>%
  count(.otu)
```

## Plots

Idea is to focus on a single species and show that the error in log proportions is inconsistent, and therefore there is error in the fold changes.
_L. crispatus_ is a species MM and BC commonly use to illustrate the inconsistent error, as since it has an intermediate efficiency, we see its error varying in sign/direction.
However, the errors in fold changes don't depend on this aspect, but only the variation in mean efficiency, and is therefore the same for all species.

Data frame for comparing measured and actual proportions:

```{r}
brooks_prop <- list(
  Actual = actual, 
  Measured = observed
) %>%
  map(transform_sample_counts, close_elts) %>%
  map_dfr(as_tibble, .id = 'type') %>%
  pivot_wider(names_from = type, values_from = .abundance)

brooks_fc <- list(
  Actual = actual, 
  Measured = observed
) %>%
  map(transform_sample_counts, close_elts) %>%
  map(pairwise_ratios, margin = 'samples', filter = FALSE) %>%
  map_dfr(as_tibble, .id = 'type') %>%
  pivot_wider(names_from = type, values_from = .abundance) %>%
  separate(.sample, c('sample1', 'sample2'), sep = ':') %>%
  # Filter to remove cases where Actual is 0 in one or both samples, or where
  # the numerator and denominator sample are equal
  filter(is.finite(Actual), Actual != 0, sample1 != sample2)
```

```{r}
set.seed(42)
species <- 'Lactobacillus_crispatus'

samples <- actual %>%
  as_tibble %>%
  filter(.otu == species, .abundance > 0) %>%
  pull(.sample)

p1 <- brooks_prop %>%
  filter(.otu == species, Actual > 0) %>%
  ggplot(aes(Actual, Measured, color = .otu)) +
  scale_color_manual(values = colors_brooks) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(color = 'grey', size = 1) +
  geom_point(position = position_jitter(width = 0.02, height = 0.02), alpha = 0.5) +
  # geom_quasirandom() +
  coord_fixed() +
  expand_limits(x = 0.1, y = 0.1) +
  labs(x = 'Actual proportion', y = 'Measured proportion', color = 'Species') +
  theme(legend.position = 'none')
p2 <- brooks_fc %>%
  filter(.otu == species, sample1 %in% samples, sample2 %in% samples) %>%
  ggplot(aes(Actual, Measured, color = .otu)) +
  scale_color_manual(values = colors_brooks) +
  scale_x_log10() +
  # scale_y_custom +
  scale_y_log10() +
  geom_abline(color = 'grey', size = 1) +
  geom_point(position = position_jitter(width = 0.02, height = 0.02), alpha = 0.5) +
  # geom_quasirandom() +
  coord_fixed() +
  labs(x = 'Actual fold change', y = 'Measured fold change', color = 'Species') +
  theme(legend.position = 'none')
# p2
```

```{r brooks_lcrisp_prop_and_fc, fig.dim = c(7, 3.7), dev = 'svg'}
(p1 + labs(x = 'Actual', y = 'Measured', title = 'Proportion')) + 
  (p2 + labs(x = 'Actual', y = 'Measured', title = 'Fold change')) + 
  plot_annotation(tag_levels = 'A')
```

# Leopold and Busby 2020

This dataset is for DNA mocks rather than cellular mocks; however the magnitude of bias is similar.
This dataset has a couple advantages for illustration purposes:

- Each species has a wider range of actual proportions than in the Brooks dataset.
- The ratios between species varies, whereas in the Brooks dataset it is always 1:1.
<!--  -->

## Setup

```{r leopold2020host-data, include = TRUE}
## From the metacal 2.0 tutorial
# Download data from https://zenodo.org/record/3872145
data_path <- here::here("notebook/_data", "leopold2020host")
# To use a temporary directory:
# data_path <- file.path(tempdir(), "leopold2020")
if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  download.file(
    "https://zenodo.org/record/3872145/files/dleopold/Populus_priorityEffects-v1.2.zip",
    file.path(data_path, "Populus_priorityEffects-v1.2.zip")
  )
  unzip(
    file.path(data_path, "Populus_priorityEffects-v1.2.zip"), 
    exdir = data_path
  )
}
#> The microbiome data is stored in a phyloseq object,
ps <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/output/compiled/phy.rds") %>%
  readRDS

sample_data(ps) <- sample_data(ps) %>%
  transform(
    Timepoint = factor(Timepoint)
  )

mock_actual <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/data/MockCommunities.csv") %>%
  read.csv(row.names = 1) %>%
  select(-Sym4) %>%
  as("matrix") %>%
  otu_table(taxa_are_rows = FALSE) %>%
  transform_sample_counts(function(x) close_elts(1 / x))
mock_taxa <- taxa_names(mock_actual)
sam <- sample_data(ps) %>% as("data.frame") %>% as_tibble(rownames = "Sample")
tax <- tax_table(ps) %>% as("matrix") %>% as_tibble(rownames = "Taxon")
ps.mock <- ps %>% 
  subset_samples(Samp_type == "Mock") %>%
  prune_taxa(mock_taxa, .)
leopold_actual <- mock_actual
leopold_observed <- ps.mock %>% otu_table
rm(mock_actual, mock_taxa, sam, tax, ps.mock)
```

## Plots

```{r}
leopold_prop <- list(
  Actual = leopold_actual, 
  Measured = leopold_observed
) %>%
  map(transform_sample_counts, close_elts) %>%
  map_dfr(as_tibble, .id = 'type') %>%
  pivot_wider(names_from = type, values_from = .abundance)

leopold_fc <- list(
  Actual = leopold_actual, 
  Measured = leopold_observed
) %>%
  map(transform_sample_counts, close_elts) %>%
  map(pairwise_ratios, margin = 'samples', filter = FALSE) %>%
  map_dfr(as_tibble, .id = 'type') %>%
  pivot_wider(names_from = type, values_from = .abundance) %>%
  separate(.sample, c('sample1', 'sample2'), sep = ':') %>%
  # Filter to remove cases where Actual is 0 in one or both samples, or where
  # the numerator and denominator sample are equal
  filter(is.finite(Actual), Actual != 0, sample1 != sample2)
```

Note, in this case all species are (nominally) in all samples, but that Epicoccum is observed in 0 reads in one sample.

```{r}
species <- 'Cladosporium'
# species <- 'Alternaria'
# species <- 'Trichoderma'

p1 <- leopold_prop %>%
  filter(.otu == species, Actual > 0) %>%
  ggplot(aes(Actual, Measured, color = .otu)) +
  # scale_color_manual(values = colors_brooks) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(color = 'grey', size = 1) +
  geom_point(position = position_jitter(width = 0.02, height = 0.02), alpha = 0.5) +
  # geom_quasirandom() +
  coord_fixed() +
  expand_limits(x = 0.1, y = 0.1) +
  labs(x = 'Actual proportion', y = 'Measured proportion', color = 'Species') +
  theme(legend.position = 'none')
p2 <- leopold_fc %>%
  filter(.otu == species) %>%
  ggplot(aes(Actual, Measured, color = .otu)) +
  # scale_color_manual(values = colors_brooks) +
  scale_x_log10() +
  # scale_y_custom +
  scale_y_log10() +
  geom_abline(color = 'grey', size = 1) +
  geom_point(position = position_jitter(width = 0.02, height = 0.02), alpha = 0.5) +
  # geom_quasirandom() +
  coord_fixed() +
  labs(x = 'Actual fold change', y = 'Measured fold change', color = 'Species') +
  theme(legend.position = 'none')
```

```{r, fig.dim = c(7, 4), dev = 'svg'}
p1 + p2
```

Same but for the ratios between two species:

```{r}
x <- list(
  Actual = leopold_actual, 
  Measured = leopold_observed
) %>%
  map(transform_sample_counts, close_elts) %>%
  map(pairwise_ratios, margin = 'taxa', filter = FALSE)

leopold_ratio <- x %>%
  map_dfr(as_tibble, .id = 'type') %>%
  pivot_wider(names_from = type, values_from = .abundance) %>%
  separate(.otu, c('otu1', 'otu2'), sep = ':') %>%
  filter(otu1 != otu2)

leopold_ratio_fc <- x %>%
  map(pairwise_ratios, margin = 'samples', filter = FALSE) %>%
  map_dfr(as_tibble, .id = 'type') %>%
  pivot_wider(names_from = type, values_from = .abundance) %>%
  separate(.otu, c('otu1', 'otu2'), sep = ':') %>%
  separate(.sample, c('sample1', 'sample2'), sep = ':') %>%
  # Filter to remove cases where Actual is 0 in one or both samples, or where
  # the numerator and denominator sample are equal
  filter(otu1 != otu2, sample1 != sample2)
```

```{r}
species1 <- 'Cladosporium'
species2 <- 'Fusarium'
# species2 <- 'Melampsora'
# species <- 'Trichoderma'

p3 <- leopold_ratio %>%
  filter(otu1 == species1, otu2 == species2) %>%
  ggplot(aes(Actual, Measured)) +
  # scale_color_manual(values = colors_brooks) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(color = 'grey', size = 1) +
  geom_point(position = position_jitter(width = 0.02, height = 0.02), alpha = 0.5) +
  # geom_quasirandom() +
  coord_fixed() +
  expand_limits(x = 0.1, y = 0.1) +
  labs(x = 'Actual ratio', y = 'Measured ratio', color = 'Species') +
  theme(legend.position = 'none')
p4 <- leopold_ratio_fc %>%
  filter(otu1 == species1, otu2 == species2) %>%
  ggplot(aes(Actual, Measured)) +
  # scale_color_manual(values = colors_brooks) +
  scale_x_log10() +
  # scale_y_custom +
  scale_y_log10() +
  geom_abline(color = 'grey', size = 1) +
  geom_point(position = position_jitter(width = 0.02, height = 0.02), alpha = 0.5) +
  # geom_quasirandom() +
  coord_fixed() +
  labs(x = 'Actual fold change', y = 'Measured fold change', color = 'Species') +
  theme(legend.position = 'none')
```

```{r, fig.dim = c(7, 7)}
(p1 + p2) / (p3 + p4)
```

Note, the ratio-view isn't always _obviously_ better; e.g. when Mel is involved; perhaps because of noise?

To make things more directly comparable, it might be best to use faceting and to fix the axes to have the same span.

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2021-10-27 introduction.Rmd
Add 'basic problem' figure to introduction

@@
 For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples (@kevorkian2018esti,@lloyd2020evid).
-However, @mclaren2019cons used a theoretical model (supported by measurements of defined bacterial communities) to show that consistent taxonomic bias can lead to variable fold errors in measured proportions.
+However, @mclaren2019cons used theoretical arguments and sequencing measurements of defined bacterial (or 'mock') communities to show that consistent taxonomic bias can lead to variable fold errors in measured proportions (Figure \@ref(fig:basic-problem)).
 These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase).
@@
 Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.
+
+<!-- begin figure -->
+
+```{r basic-problem, fig.cap = '(ref:cap-basic-problem)'}
+svg_path <- here::here(
+  "notebook/_posts/2021-10-27-illustrate-the-problem/illustrate-the-problem_files", "figure-html5/brooks_lcrisp_prop_and_fc-1.svg"
+)
+if (knitr::is_html_output()) {
+  knitr::include_graphics(svg_path)
+} else if (knitr::is_latex_output()) {
+  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
+  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
+    "error-proportions.pdf")
+  args <- c(
+    "--export-area-page", 
+    "--export-type=pdf", 
+    str_c("--export-filename=", pdf_path),
+    svg_path
+  )
+  system2("inkscape", args = args)
+  knitr::include_graphics(pdf_path)
+  rm(pdf_path)
+}
+rm(svg_path)
+```
+
+(ref:cap-basic-problem) **Mock community experiments show that taxonomic bias can distort the measured fold change in an individual species' proportion across samples.** The figure shows the measured vs.
+actual proportions for a single bacterial species, _Lactobacillus crispatus_, in a set of bacterial cellular mock communities, and the resulting fold changes between community samples.
+The inconsistent error in the measured proportions of individual samples (Panel A) leads to inaccurate measurements of fold changes (Panel B).
+Mock communities were constructed and measured with 16S sequencing by @brooks2015thet.
+The data was re-analyzed by @mclaren2019cons, who showed that despite the inconsistency of the errors in Panel A, taxonomic bias acted consistently across samples.
+
+<!-- end figure -->

### 2021-11-02 notebook/_posts/2021-11-01-momspi-summary/momspi-summary.Rmd
Add MOMS-PI summary analysis

---
title: "MOMS-PI: Key dynamics"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:brooks2015thet
  - ref:fettweis2019thev
date: 2021-11-01
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300
)
```

This document is part of a series of analyses of the impact of taxonomic bias on vaginal community dynamics in MOMSPI dataset (@fettweis2019thev), using the bias estimated from the @brooks2015thet mock community data.
The purpose of this document is to demonstrate the key dynamics we want to highlight in the manuscript.

# Setup

## R setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
```

```{r ggplot}
theme_set(theme_cowplot())

colors_brooks <- c(
  "Atopobium_vaginae" = "#009E73",
  "Gardnerella_vaginalis" = "#56B4E9",
  "Lactobacillus_crispatus" = "#D55E00",
  "Lactobacillus_iners" = "#505050",
  "Prevotella_bivia" = "#0072B2",
  "Sneathia_amnii" = "#CC79A7",
  "Streptococcus_agalactiae" = "#E69F00")

scale_y_custom <- scale_y_continuous(
    trans = 'log10',
    breaks = trans_breaks('log10', function(x) 10^x),
    labels = trans_format('log10', math_format(10^.x))
  )
```

## Estimate bias from @brooks2015thet

We can estimate the efficiencies for the control species used in the cellular mocks of @brooks2015thet by running the example code in the docs for `metacal::estimate_bias()`. 
This code uses the observed and actual species-level abundance tables which were produced by @mclaren2019cons from the SI files of @brooks2015thet and are included in the metacal package.

```{r}
dr <- system.file("extdata", package = "metacal")
list.files(dr)
actual <- file.path(dr, "brooks2015-actual.csv") |>
  read.csv(row.names = "Sample") |>
  as("matrix")
observed <- file.path(dr, "brooks2015-observed.csv") |>
  read.csv(row.names = "Sample") |>
  subset(select = - Other) |>
  as("matrix")

# Estimate bias with bootstrapping for error estimation
mc_fit <- estimate_bias(observed, actual, margin = 1, boot = TRUE)
summary(mc_fit)

rm(actual, observed, dr)
```

```{r}
control_species <- mc_fit %>% coef %>% names
control_genera <- control_species %>% str_extract('^[^_]+')
```

## Load the MOMSPI data

TODO: replace this code chunk with one that pulls data from the github repo into `_data/`

Load the MOMSPI Stirrups profiles into phyloseq

```{r}
path_momspi <- '~/research/momspi'

otu <- path(path_momspi, "output", "stirrups-profiles", "abundance-matrix.csv.bz2") %>%
  read_csv(
    col_types = cols(.default = col_double(), sample_name = col_character())
  ) %>%
  otu_table(taxa_are_rows = FALSE)
sam <- path(path_momspi, "output", "stirrups-profiles", "sample-data.csv.bz2") %>%
  read_csv(col_types = "ccccccic") %>%
  mutate(across(host_visit_number, factor, ordered = TRUE)) %>%
  sample_data
tax <- path(path_momspi, "output", "stirrups-profiles", "taxonomy.csv.bz2") %>%
  read_csv(col_types = cols(.default = col_character())) %>%
  tax_table %>%
  mutate_tax_table(
    species = case_when(!is.na(genus) ~ .otu)
  )
momspi_raw <- phyloseq(otu, sam, tax) %>%
  mutate_tax_table(across(.otu, str_replace, 
      "(?<=Lactobacillus_crispatus)_cluster", "")) %>%
  mutate_sample_data(., sample_sum = sample_sums(.))
taxa_names(momspi_raw) %>% str_subset("crispatus")
stopifnot(all(control_species %in% taxa_names(momspi_raw)))
```

L crisp type 1 is just a few reads so we can ignore it (check)

```{r}
p1 <- momspi_raw %>% sample_sums %>% qplot + scale_x_log10() +
  labs(x = 'sample sum')
p2 <- momspi_raw %>% taxa_sums %>% qplot + scale_x_log10() +
  labs(x = 'otu sum')
p1 + p2
```

Let's filter samples with fewer than 1000 reads (also requirement used by the authors)

```{r}
momspi <- momspi_raw %>% 
  filter_sample_data(sample_sum >= 1e3)
```

Note, we might want to be conservative in terms of composition shifts by making the min prop of species 1e-3.
However this might add up to a large effect if we apply it to all species.
Could instead filter to 1e4 reads and threshold at 1e-4, or 3e-4

Let's see if we can filter the super-low-count species without affecting any remaining samples too much:

```{r}
x <- (momspi %>% filter_taxa2(~sum(.) >= 2e2) %>% sample_sums) / (momspi %>% sample_sums)
summary(x)
```

Dropping OTUs with less than 200 reads has a negligible impact on the samples, so let's do that.

```{r}
momspi <- momspi %>% filter_taxa2(~sum(.) >= 2e2)
```

# Calibrate the MOMSPI measurements

Owing to the large sample size of the mock community experiment, the standard errors in the estimated bias for the control species are quite small, and
our uncertainty in the true bias of the MOMSPI study is much larger than what these standard errors imply.
Therefore I will use the point estimate of the bias for the control species as if it were known exactly, but we should consider this to just be a hypothesis of what the bias truly is in the MOMSPI study.
This hypothetical nature applies even more strongly when we extrapolate to other species.

Since don't have a way to directly estimate the efficiencies of the other OTUs, we'll impute them as follows.

1. Set the efficiencies of the 7 control species to the directly estimated values
2. Compute efficiencies for the 6 control genera to the efficiency of the control species (if only one) or the geometric mean efficiency (if more than one; here this is just _Lactobacillus_)
3. Use these genera-level efficiencies for the efficiencies of OTUs that are not control species but are in the same genus as one
4. For other OTUs, use the geometric mean of the genera efficiencies; this is similar to using the mean of the 7 control species but gives _Lactobacillus_ equal weight to the other genera.

<!--  -->

```{r}
bias_species <- coef(mc_fit) %>% 
  enframe("species", "efficiency") %>%
  print
bias_genus <- bias_species %>%
  mutate(genus = str_extract(species, "^[^_]+"), .before = 1) %>%
  with_groups(genus, summarize, across(efficiency, gm_mean)) %>%
  print
# Match on genus or species, depending on which is available; then set others
# to average genus efficiency
bias_all <- tax_table(momspi) %>% as_tibble %>%
  left_join(bias_species, by = "species") %>%
  left_join(bias_genus, by = "genus") %>%
  mutate(
    efficiency = case_when(
      !is.na(efficiency.x) ~ efficiency.x,
      !is.na(efficiency.y) ~ efficiency.y,
      TRUE ~ gm_mean(bias_genus$efficiency)
    )
  ) %>%
  select(-efficiency.x, -efficiency.y) %>%
  # standardize to L. iners, the most efficiently measured
  mutate(
    across(efficiency, ~ . / max(.))
  ) %>%
  glimpse
bias_all_vec <- bias_all %>% select(.otu, efficiency) %>% deframe 
```

## Observed and calibrated proportions

Compute observed and calibrated proportions with all taxa.

```{r}
momspi_obs <- momspi %>% 
  transform_sample_counts(close_elts)
# note, calibrate() automatically subsets to just the focal taxa if we haven't
# already
momspi_cal <- momspi_obs %>% calibrate(bias_all_vec)
```

Note, I'm not using pseudo-props so the zeros are staying zero.
TODO: Reconsider this choice.

## Sample summary statistics

- Mean efficiencies
- Most abundant species
- Proportion of Lactobacillus
<!--  -->

To compute the mean efficiencies, we can use `metacal::perturb()` to multiply efficiencies by the actual proportions for each sample (using `norm = "none"` to avoid renormalizing the results), then summing up the values from each sample with `phyloseq::sample_sums()`.
It's also useful to get the mean efficiency among just the control species.

```{r}
tmp1 <- momspi_cal %>%
  perturb(bias_all_vec, norm = "none") %>% 
  sample_sums %>%
  enframe('.sample', 'mean_efficiency')
tmp2 <- momspi_cal %>%
  filter_tax_table(.otu %in% control_species) %>%
  transform_sample_counts(close_elts) %>%
  perturb(bias_species %>% deframe, norm = "none") %>% 
  sample_sums %>%
  enframe('.sample', 'mean_efficiency_controls')
mean_eff <- left_join(tmp1, tmp2, by = '.sample') %>%
  left_join(sample_data(momspi_cal) %>% as_tibble, by = '.sample')
```

For each sample, let's get the most abundant species in the calibrated and uncalibrated profiles.

```{r}
get_top_species <- function(ps) {
  x <- ps %>%
    otu_table %>%
    as_tibble(pivot = TRUE) %>%
    with_groups(.sample, slice_max, .abundance, n = 1) %>%
    # add count to check for ties
    add_count(.sample)
  if (anyDuplicated(x$.sample))
    stop('Ties present')
  x %>% select(.sample, .otu, .abundance)
}
top_species <- list(
  Uncalibrated = momspi_obs,
  Calibrated = momspi_cal
  ) %>%
  map_dfr(get_top_species, .id = 'type')
```

Which species tend to be dominant?

```{r}
top_species %>%
  filter(type == 'Uncalibrated') %>%
  count(.otu) %>%
  arrange(desc(n)) %>%
  print(n=20)
```

```{r}
top_species %>%
  filter(type == 'Calibrated') %>%
  count(.otu) %>%
  arrange(desc(n)) %>%
  print(n=20)
```

The most important species/OTU that we do not have direct bias information on at the species or genus level is Lachnospiraceae BVAB1.

The proportion of Lactobacillus in each sample - will simply do all control genera.

```{r}
genus_prop <- momspi_cal %>%
  subset_taxa(genus %in% control_genera) %>%
  tax_glom('genus') %>%
  as_tibble %>%
  select(.sample, genus, .abundance) %>%
  pivot_wider(names_from = genus, values_from = .abundance)
```

Let's join the most abundant species and key genera proportions to the mean efficiecies for further analysis.
For plotting, we'll create a version of the top taxa that only colors the taxa that are most often dominant.

```{r}
tmp <- top_species %>% 
  select(-.abundance) %>% 
  pivot_wider(names_from = type, values_from = .otu, 
    names_glue = 'top_{str_to_lower(type)}')
mean_eff1 <- mean_eff %>%
  left_join(genus_prop, by = '.sample') %>%
  left_join(tmp, by = '.sample') %>%
  mutate(
    top_plot = fct_lump_n(top_calibrated, 7)
  )
rm(tmp)
```

# Visual analysis of mean efficiency distribution and role of key species

We can make a plot similar to that used for the Leopold et al analysis.

```{r}
# First let's look at the distribution of mean efficiencies
p1 <- mean_eff1 %>%
  ggplot(aes(mean_efficiency, fill = top_plot)) +
  geom_histogram() +
  # annotate("point", y = 200, x = bias, color = "black", shape = "|", size = 7) +
  scale_x_log10() +
  expand_limits(x = range(bias_all_vec)) +
  scale_fill_brewer(type = "qual") +
  labs(
    # x = "Sample mean relative efficiency (SMRE)",
    x = "Mean efficiency", 
    y = "Number of samples",
    fill = 'Most abundant species'
  )
# Next let's plot Lactobacillus proportion against mean efficiency, as another
  # look at the relationship between Lacto and mean efficiency.
p2 <- mean_eff1 %>%
  ggplot(aes(mean_efficiency, Lactobacillus, color = top_plot)) +
  geom_point() +
  scale_x_log10() +
  expand_limits(x = range(bias_all_vec)) +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 1e-2),
    breaks = c(0, 0.01, 0.03, 0.1, 0.3, 1)
    ) +
  scale_color_brewer(type = "qual") +
  labs(
    y = "Lactobacillus proportion",
    x = "Mean efficiency",
    color = 'Most abundant species'
  )
```

```{r momspi-mean-efficiency-dist-and-lacto, fig.dim = c(6,5) *1.6}
# The yellow makes the L crispatus points a bit hard to see, so let's swap it's
# color with the rarer L. gasseri in the final figure.
colors_top <- RColorBrewer::brewer.pal(8, 'Accent')[c(1,2,3,5,4,6,7,8)] 
(p1 + ggtitle("Distribution of sample mean efficiencies")) / 
  (p2 + 
    ggtitle("Lactobacillus proportion vs. mean efficiency") + 
    theme(legend.position = 'none')
  ) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = 'collect') &
  scale_fill_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  ) &
  scale_color_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  )
# The x-axis is very close but not completely aligned; I think I need to
# manually set the x limits to get perfect alignment, but that leads to some
# samples being dropped in the histogram plot; some investigation is needed
# to go that route.
```

This plot shows that Lactobacillus dominance drives a bimodal distribution in the mean efficiencies, and that Gardnerella dominance drives the low end.
Note that the 'other' and 'BVAB1' dominated samples are lumped around the middle because of our assumption in imputing the efficiencies.

For the caption/methods: Note that the most abundant species is that inferred from the calibrated data; could just say that the calibrated data is treated as the actual composition.

The Lacto proportion doesn't seem that much more illuminating than the histogram colored by dominant species, so let's make a version with just that for the manuscript.

```{r momspi-mean-efficiency-dist, fig.dim = c(7,3.5)}
colors_top <- RColorBrewer::brewer.pal(8, 'Accent')[c(1,2,3,5,4,6,7,8)] 
p1 + 
#  ggtitle("Distribution of sample mean efficiencies") +
  scale_fill_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  ) +
  scale_color_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  )
```

# Fold changes in mean efficiency

How often do shifts in the mean efficiency occur within a single subject?
How often do shifts from Lacto to non-Lacto dominance occur within a single subject?
How often do shifts from Lacto to Gardnerella dominance occur within a single subject?

Let's start by computing the pairwise changes in the sample mean efficiency between consecutive samples within women, and between the first samples of different women (first samples only to limit the number of sample pairs).

Since we may be missing some host visits, let's create a new host visit number that has no gaps:

```{r}
mean_eff2 <- mean_eff1 %>%
  with_groups(host_subject_id, mutate, 
    host_visit_rank = rank(host_visit_number))
```

```{r}
tb <- crossing(
  sample.1 = mean_eff2$.sample, 
  sample.2 = mean_eff2$.sample
  ) %>%
  left_join(mean_eff2, by = c(sample.1 = '.sample')) %>%
  left_join(mean_eff2, by = c(sample.2 = '.sample'), suffix = c('.1', '.2')) %>%
  mutate(
    comparison_type = case_when(
      host_subject_id.1 == host_subject_id.2 & host_visit_rank.2 == host_visit_rank.1 + 1 ~ 
        'Between consecutive visits',
      host_subject_id.1 != host_subject_id.2 & host_visit_rank.2 == 1 & host_visit_rank.1 == 1 ~
        "Between subjects' first visits"
    )
  ) %>%
  filter(!is.na(comparison_type)) %>%
  mutate(
    switch_class = case_when(
      top_plot.1 == top_plot.2 ~ 'No change',
      str_detect(top_plot.1, 'Gardnerella') & str_detect(top_plot.2, 'Lactobacillus') ~ 'Gard. to Lacto.',
      str_detect(top_plot.1, 'Lactobacillus') & str_detect(top_plot.2, 'Gardnerella') ~ 'Lacto. to Gard.',
      str_detect(top_plot.1, 'Lactobacillus') & str_detect(top_plot.2, 'Lactobacillus') ~ 'Lacto. to Lacto.',
      TRUE ~ 'Other',
    )
  )

# tb_within <- tb %>%
#   filter(host_subject_id.1 == host_subject_id.2, host_visit_rank.2 == host_visit_rank.1 + 1)
# tb_between <- tb %>%
#   filter(host_subject_id.1 != host_subject_id.2, host_visit_rank.2 == 1, host_visit_rank.1 == 1)
```

Let's look at the distribution of FCs in mean efficiency, coloring by whether there is a shift in dominant species.

```{r momspi-mean-efficiency-fcs, fig.dim = c(6, 7.0)}
tb %>%
  ggplot(aes(mean_efficiency.2 / mean_efficiency.1, fill = switch_class)) +
  facet_wrap(~comparison_type, ncol = 1, scales = 'free_y') +
  geom_histogram() +
  scale_x_log10() +
  # expand_limits(x = range(bias_all_vec)) +
  labs(
    x = "Fold change in mean efficiency", 
    y = "Number of sample pairs",
    fill = 'Change in most\nabundant species'
  ) +
  scale_fill_brewer(type = 'qual', palette = 3)
```

Unsurprisingly, shifts in the mean efficiency tend to be larger between different subjects.
Note, the between-subject panel is perfectly symmetric, but the consecutive-visits panel is not.

```{r}
tb %>%
  count(comparison_type, switch_class) %>%
  with_groups(comparison_type, mutate, prop = close_elts(n)) %>%
  knitr::kable(digits = 3)
```

# (TODO) Example community trajectory

Pick a woman with multiple visits and a large change in mean efficiency

```{r}
tb %>%
  filter(host_subject_id.1 == host_subject_id.2) %>%
  mutate(fc = mean_efficiency.2 / mean_efficiency.1) %>%
  arrange(desc(abs(log(fc)))) %>%
  select(fc, host_subject_id.1, top_plot.1, top_plot.2, host_visit_rank.1)
```

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2021-11-06 real-world-implications.Rmd
Revise Solutions section
Besides revising existing sections on calibration, sensitivity analysis,
and meta-analysis, I added a section on ratio-based relative DA analysis
and a section on complementary total-density and sequencing
measurements. I deleted the part of the Implications section on error in
total-density measurement, as this material is now covered by the
'complementary measurements' solution section.

@@
 
-**Error in total-density measurements:**
-
-So far we have ignored error in total-density measurements that are used to convert estimated proportions to estimated densities;
-however, in reality these measurements are likely to contain substantial systematic error that can either worsen or improve the error in LFC estimates.
-
-- Todo: Add to previous section or Appendix with the relevant theory.
-
-To understand how, we can consider the question of whether bulk 16S qPCR or flow cytometry is a more appropriate method of measuring total bacterial density for use in proportion-based density measurement (@galazzo2020howt, @jian2021comm).
-16S rRNA-gene qPCR measures the concentration of 16S rRNA gene copies in the extracted DNA and is therefore affected by variation among species in DNA extraction efficiency, 16S copy number, and primer-binding and amplification efficiency.
-For example, @mclaren2019cons found that the product of extraction efficiency and 16S copy number for _Lactobacillus iners_ was 20X larger than _Gardnerella vaginalis_ (CHECK NUMBERS), such that we should expect samples dominated by _L.
-iners_ to result in qPCR estimates 10-20X higher than equally dense samples dominated by _G.
-vaginalis_.
-These observations suggest that 16S qPCR is a poor proxy for fold changes in total bacterial density.
-Yet, as recently suggested by @jian2021comm, this taxonomic bias in qPCR may simultaneously make it a good measure of total density for use with 16S amplicon sequencing for proportion-based density estimation.
-16S amplicon sequencing shares these same sources of bias and so is likely to have species efficiencies highly similar to those for the qPCR measurements, which we can predict to cause the mean 16S-qPCR efficiency to closely follow the mean 16S-sequencing efficiency, such that the error induced by one largely cancels the error induced by the other (Ref Appendix).
-In contrast, species-specific efficiencies in bulk cell counting are less likely to be strongly correlated with those from 16S sequencing.
-Even supposing cell counting were to provide perfectly accurate measures of total density, it may actually perform worse than qPCR for DA analysis of proportion-based density estimation precisely because the error does not mirror that in the 16S measurement.
-
-On the other hand, bulk or targeted measurement (or DNA spike-ins) that occurs post DNA-extraction is subject to other error that is not likely to counter that of the 16S-based proportion estimates.
-The yield of DNA extraction can show highly random variation and also be highly non-linear (as a function of cellular biomass) at high concentrations.
-This features create error that is unlikely to offset the variation in mean efficiency in the 16S sequencing (todo: clarify why).
-Targeted measurement of cells and cellular spike-ins, discussed in Section \@ref(solutions), can avoid this issue while also facilitating ratio-based analysis while (unlike bulk cell counting) remaining robust mean-efficiency variation.
-
 ## Loss of power and precision (stub)

### 2021-11-06 solutions.Rmd
Revise Solutions section
Besides revising existing sections on calibration, sensitivity analysis,
and meta-analysis, I added a section on ratio-based relative DA analysis
and a section on complementary total-density and sequencing
measurements. I deleted the part of the Implications section on error in
total-density measurement, as this material is now covered by the
'complementary measurements' solution section.

@@
 The theoretical results from Section \@ref(differential-abundance) suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.
-We present five that seem particularly promising:
-The first three are methods to correct or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.
+
+## Analyses of fold changes in species ratios
+
+Since the fold changes in species ratios are unaffected by consistent taxonomic bias, a natural approach to countering bias in relative DA analysis is to ratio- rather than proportion-based methods.
+A variety of such methods have been developed for microbiome DA analysis that build on earlier work from the field of Compositional Data Analysis (CoDA).
+These methods analyze log-fold changes in the ratios among a pair of species or, more generally, some product of exponential power of multiple species, which likewise are bias-invariant.
+
+Besides a greater robustness to taxonomic bias, such ratio-based analyses have other advantages.
+First, analysis of ratios avoids other standard criticisms leveled at proportion-based analysis—namely that the change in proportion of a species is affected by change in the density of all other species and the set of species used in the denominator of the proportion calculation (@gloor2017micr).
+Second, ratios may be more informative than absolute density in some cases, such as for swabs or other sample types for which the absolute density in the sample may be an unreliable indicator of the absolute density of the sampled ecosystem.
+
+Ratios come with their limitations, however.
+First, there are a vast number of species ratios (and derivatives) to potentially consider, none of which provide direct information about absolute densities, which can make choosing and interpreting a ratio-based DA analysis challenging.
+Second, ratio measurements are impacted by noise in both the numerator and denominator and so can be noisier than proportion measurements, particularly when the read counts in the denominator are small.
+A related problem is that, due to the discrete nature of biological organisms and the reads that are generated during sequencing, species in the denominator are commonly observed to have 0 reads; the assumptions that made by different methods about the meaning of these 0 counts can have a major impact on DA results.
+Finally, bias invariance at the level of ratios among species does not extend to ratios of additive aggregates of species, such as between bacterial phyla, unless further conditions are met (@mclaren2019cons).
+
+## Calibration using community controls {#calibrate-compositions}
+
+Measurement of appropriate control samples along with the primary experimental samples can make it possible to directly measure and remove the effect of bias before or while conducting downstream analysis.
+Under the assumption that taxonomic bias acts consistently across samples, it can be directly estimated for the species in one or more community control samples—samples whose species identities and relative abundances are known either by construction of by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
+The estimated relative efficiencies among the control species can then be used to correct the effect of bias on the relative abundances (ratios) among these species in a set of target (non-control) samples measured with the same protocol (ideally in the same experiment and sequencing run) @mclaren2019cons.
+Calibration can be extended to species not in the controls by making additional assumptions that allow one to impute the missing efficiencies; such imputation might employ phylogenetic relationships, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
+
+The calibrated compositions can be used for arbitrary downstream microbiome analyses, including relative and absolute DA analysis.
+To demonstrate the potential for calibration to improve fold-change measurements, we estimated bias from one sample in the mock community data from @brooks2015thet that contained all 7 species and used it to calibrate the compositions of all samples (Figure \@ref(fig:calibration-example)).
+We then estimated species' densities using the uncalibrated and calibrated proportions using the community normalization method, treating the total density as known and fixed.
+Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions and densities (Figure \@ref(fig:calibration-example)).
+To demonstrate a more practical application, we used the bias estimated from the full set of mock communities as the basis for calibrating the vaginal community time series measurements from the MOMS-PI study (@fettweis2019thev) that were made using the same 16S sequencing protocol.
+We imputed the efficiencies of other species using the taxonomic relationship to the control species (Methods).
+Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in [implications].
 
@@
 
-## Calibrate compositions using community controls {#calibrate-compositions}
+## Calibration using reference species
 
-The most complete but experimentally challenging approach to mitigating taxonomic bias involves directly measuring the efficiencies of individual species using control samples (@mclaren2019cons).
-Suitable controls are any community sample whose composition (species identities and relative abundances) has been previously characterized, either by construction (in the case of constructed 'mock' communities) or by measurement with a _reference protocol_ chosen to serve as a measurement standard.
-These controls, when measured alongside the target experimental samples, can be used to estimate the relative efficiencies among the superset of species across the control samples.
-The estimated efficiencies can then be used to calibrate the measured relative abundances among the control species in the target samples to the chosen measurement standard, by simply dividing the species' read counts by their corresponding efficiencies (@mclaren2019cons) or using a more sophisticated statistical modeling framework.
-These calibrated compositions can then be used for arbitrary downstream microbiome analyses, including DA analysis.
-Calibration of species not in the controls requires first imputing their efficiencies from the control measurements, for example using phylogenetic relationships, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
-Though such imputation is conceptually straightforward, its accuracy and ultimate utility remain untested.
+Obtaining control communities that span the taxa of interest is often not practical.
+Moreover, there is no guarantee that bias is consistent across the control and primary samples.
+The results of the previous sections suggest an alternative approach to calibration that overcomes these problems for the specific purpose of species-level DA analysis, requiring only that there are one or more reference species whose abundance can be determined, up to a constant error factor, by background knowledge, experimental design, or independent measurement.
 
-- The above states the approach; can add description/discussion of application to the MOMS-PI and @leopold2020host cases and developments and challenges in availability of control communities
-- 'more sophisticated modeling framework' - cite @williamson2021amul and https://microbiomemeasurement.org/posts/bias-estimation-from-mocks-with-fido/
+Reference-species calibration allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from (taxonomically biased, relative abundance) metagenomics measurements.
+Section X described the reference-species approach to density measurement and showed that such measurements have a constant fold error, provided that the error in the density of the reference species is constant.
+Three categories of reference species of practical import are described in Box X: housekeeping species assumed to have constant abundance, spike-in species added at constant abundance, and species whose abundance has been measured with a targeted method like qPCR.
+To demonstrate the ability for reference calibration to improve fold changes, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
+Doing so improved the resulting measurements of fold changes in density for all species (Figure \@ref(fig:calibration-example)).
 
-## Use ratio-based abundance measurement
+## Choosing complementary total-density and metagenomics measurements
 
-The DA analyses we consider don't need accurate estimates of relative abundances within individual samples---they only require accurate fold changes across samples.
-Section \@ref(differential-abundance) showed that fold changes in species ratios and ratio-based density estimates can remain accurate even when taxonomic bias distorts the values in individual samples.
-Hence a solution is to use to analyze variation in ratios and ratio-based density estimates.
-For the latter, practical options include
+A variety of methods for measuring total community density have been paired with both amplicon and shotgun sequencing data for the community-normalization method to measure absolute densities for absolute DA analysis, with little consensus on the pros and cons of the various methods.
+Although these methods differ in many aspects, a notable aspect is the sources of taxonomic bias inherent in each method.
+Just as species differ in how efficiently they are measured by MGS, so can they differ in how efficiently they contribute to a total density measurement.
+Consider the popular method of assessing bacterial abundance with total-16S qPCR measurement.
+Even an ideal qPCR protocol that perfectly amplifies all species remains a biased measurement of cell density, since the total 16S copies in the extracted DNA contributed by each species will be proportional to its (species-specific) lysis efficiency and 16S copy-number.
+16S qPCR is commonly paired with 16S amplicon sequencing, with which these sources of bias are shared, perhaps along with variation in primer binding and amplification efficiency.
+Methods like using flow cytometry that directly measures cell density lacks these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
+By extending the analysis of Section X to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(appendix-total-density)).
+This finding suggests that methods of total-density measurement that are more accurate as measures of total cell density may actually perform worse than less accurate methods for the purposes of absolute DA inference.
 
-- spike-ins of extraneous species at constant and/or known density
-- housekeeping taxa, stipulated or computationally identified as having a constant density across samples
-- targeted measurement of specific taxa
+We illustrate with a hypothetical example of two vaginal communities of equal density, but which are dominated either by _Lactobacillus iners_ or _Gardnerella vaginalis_.
+Based on our earlier analysis in @mclaren2019cons, we estimate that cells of _L. iners_ 20X more 16S copies than _G. vaginalis_ from DNA extraction (CHECK NUMBERS).
+Ignoring other sources of bias, we expect relative efficiencies of _L. iners_:_G. vaginalis_ of 20 for both qPCR and 16S sequencing measurements.
+Consequently, the _G. vaginalis_ community will yield a substantially lower qPCR measurement than the _L. iners_ community and we would incorrectly infer a decrease in total abundance.
+At the same time, this decrease in the mean efficiency of the sequencing measurement causes artificially high fold changes in the proportions of all species (ref previous section).
+When using qPCR and 16S sequencing measurements for community-based density estimation, these two opposing errors cancel, yielding accurate fold changes in species densities.
 
-TBD what further discussion or illustration to have here.
+More generally, changes in the mean efficiency of the total density measurement offset those in the sequencing measurement when comparing log density of species across samples.
+If the (log) efficiencies of the total and sequencing measurement are positively correlated across species, then the (log) mean efficiencies will tend to be positively correlated across samples, leading to a reduction in error in DA analysis relative to total-density methods whose efficiencies are uncorrelated with those of the sequencing measurement.
+These observations suggest that qPCR of a marker gene may be the ideal pairing for amplicon sequencing measurements, despite being a poor measure of changes in total cell density.
+Similarly, bulk DNA quantification may be an ideal pairing for shotgun sequencing, making it possible to account for variation in lysis efficiency and genome length.
+These pairings even make it possible to account for error caused by variation among samples in the fraction of unclassified reads, which can form a major fraction of amplicon and shotgun data.
+Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (REF Appendix).
 
-## Calibrate fold changes using measurements of targeted taxa
+## Bias-sensitivity analysis and bias-aware meta-analysis
 
-In some cases, we may still which to use a proportion-based approach to analyzing variation in density.
-For example, in some systems measuring total cell concentration may be easier than implementing a spike-in experiment or determining a set of reference species to directly measure for which at least one species is present in all samples.
-In this case, targeted measurements of one or more reference species can be used to validate and/or correct the fold changes or regression coefficients infered from the proportion-based densities.
-Recall that the error in proportion-based log fold changes (LFCs) is shared among species and is equal to the inverse LFC fold change in mean efficiency (Equations \@ref(eq:prop-fc-error) and \@ref(eq:regression-error); Figure \@ref(fig:regression-example)D).
-Consequently, an accurate LFC of one or more reference species can serve as a general validation or used to correct the LFCs of all species.
+For experiments that have been conducted without calibration controls, it is still possible to investigate the likelihood that DA results can be explained by taxonomic bias by performing a computational bias-sensitivity analysis.
 
-- Note: I may not correctly understand the next example, since I don't yet see how this interpretation of Table 1 meshes with Figure 4
+A bias-sensitivity analysis examines how sensitive the results of a given DA analysis are to the working assumption about the bias of a given protocol.
+This working assumption is typically is that there is no bias, but could also be bias measured in a previous experiment or predicted from taxonomic features like 16S copy number.
+A straightforward approach to bias-sensitivity analysis is to use computer simulation to re-analyze the MGS data across a set of simulated taxonomic biases.
+First, multiple sets of efficiency vectors are randomly generated, each representing a possible quantification of taxonomic bias for the study.
+Second, each efficiency vector is used to calibrate the original MGS measurements; each set of calibrated measurements represents the true community compositions under the given hypothesized bias. 
+Finally, the DA analysis is re-run on each set of calibrated measurements and the distribution of results is compared graphically or with summary statistics.
+A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool 'corncob' are shown in \@ref(fig:sensitivity-example).
+The simulated efficiency vectors may be generated to different hypotheses about bias in the given system, such as its overall magnitude and correlation among phylogenetically-related species.
+Thus the utility of such simulations will increase the more we learn about the properties of taxonomic bias in different systems.
 
-We illustrate this idea with an example from the study by @lloyd2020evid described in Section \@ref(implications), which estimated doubling times for various taxa from the measured log increase in cell density with burial time in marine sediments.
-The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for several bacterial and archeal taxa (CHECK).
-In Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years (CHECK) using the proportion-based densities and 10.3 years by qPCR (@lloyd2020evid Table 1 and Figure 4).
-These doubling times correspond to highly similar estimated slopes of log density per year of $1/10.1 \approx 0.99$ and $1/10.3\approx 0.97$.
-The difference between these slopes, $\approx 0.02$, estimates the slope of the mean efficiency.
-It's small value suggests little systematic variation in mean efficiency with burial time and hence little systematic bias in the estimated doubling times derived from the proportion-based density estimates.
-In this case, close agreement in the DA results across methods for a specific taxon suggests that variation in sample mean efficiency did not significantly distort the MGS-based results and hence provides a degree of validation for all taxa, not just the taxa with targeted measurements.
-(An important caveat for this example is that this analysis assumes consistent efficiencies at the higher taxonomic levels used for qPCR measurement and MGS analysis performed by @lloyd2020evid.)
-Here we compared summary statistics; however, in principle a joint statistical analysis of targeted and proportion-based density measurements would provide more statistical power to estimate and correct the effects of variation in the mean efficiency across samples and provide calibrated fold change and regression estimates for all taxa [APPENDIX].
+An alternative approach to bias-sensitivity analysis would be to directly include the (unknown) taxonomic bias of the protocol in the statistical model used for DA analysis, as described for bias-aware meta-analysis below.
+Further development of tools and workflows for performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance and microbiome analyses more generally.
 
-## Perform a bias sensitivity analysis
-
-What can we do for experiments that have been conducted without control communities, targeted control measurements, or spike-ins?
-One approach is to conduct a bias sensitivity analysis to probe the extent and way in which the results of a DA analysis are sensitive to different possible taxonomic biases in our experiment.
-
-A straightforward approach to bias sensitivity analysis is to use computer simulation to randomly generate efficiency vectors, each of which is used to calibrate the MGS measurements.
-These calibrated MGS measurements represent the true compositions under the hypothesized taxonomic bias. 
-By re-running our DA analysis and comparing results across these simulated datasets, we can begin to understand how our initial results might be influenced by the strength and structure of the taxonomic bias in our experiment.
-
-The simulated efficiency vectors may be generated to reflect certain hypotheses about bias in the given system.
-The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
-Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance and microbiome analyses more generally.
-
-## Use bias-aware meta-analysis
+## Bias-aware meta-analysis
 
 So far we have considered analysis of samples subject to the same taxonomic bias;
-however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in different groups of samples.
-This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses where the effects of bias might cancel within a study.
+however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in samples from different studies.
+This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses.
+A potential solution is to perform a bias-aware meta-analysis that that explicitly accounts for the possibility of distinct taxonomic bias across studies.
 
-A potential solution is to perform a bias-aware meta-analyses that jointly analyzes multiple data from studies while explicitely acknowledges the distinct, unknown taxonomic bias in each study.
-Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model "batch effects"---amorphous differences in the measurements of each study.
-By configuring these models so that these latent parameters correspond to the study and species specific efficiencies, we can increase the biological interpretability of these models and gain ability to determine whether disageements in DA results across studies are due to differences in taxonomic bias or other differences (such as host cohort or study design).
-Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by @mclaren2019cons, which can increase improve our ability to find reproducible microbial associations in a given meta-study.
+Parametric microbiome meta-analysis models include sets of species-specific latent parameters to model "batch effects"—amorphous differences in the measurements of each study.
+By configuring these models so that these latent parameters correspond to study-specific species efficiencies, we can improve their biological interpretability and gain the ability to directly assess whether disagreements in DA results across studies can be explained by differences in taxonomic bias.
+To the extent that taxonomic bias truly is consistent within studies, meta-analyses that directly account for it can have greater statistical power to extract the truly consistent and inconsistent microbial patterns across studies.
 
@@
 
-(ref:cap-sensitivity-example) **Example of a bias sensivity analysis.**
+(ref:cap-sensitivity-example) **A bias-sensivity analysis can be performed to examine how sensitive the results of a DA analysis are to assumptions about taxonomic bias in community measurements.** The figure shows the results of a bias-sensitivity analysis used to study the effect of bias on the association of _Gardnerella vaginalis_ and preterm birth that was investigated by @callahan2017repl.
+100 random efficiency vectors were drawn at 6 different bias strengths (quantified by the variance in log efficiency, $\sigma_{e}^{2}$).
+Each efficiency vector was used to calibrate the MGS profiles and perform a DA association test of _G.
+vaginalis_ versus the host's preterm birth outcome; regression coefficients $\hat \beta$ indicate the increase of average logit proportion of  _G.
+vaginalis_ in women who experienced preterm birth.
 

### 2021-11-08 appendix-regression.Rmd
Move regression results into appendix

--- 2021-02-19 proofs.Rmd
@@
-# (APPENDIX) Appendix {-}
+# Error in estimated regression coefficients and standard errors {#appendix-regression}
 
-# Proofs of regression results
+Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.
 
-```{theorem, label = "regression-linearity", name = "Linearity of regression coefficients", echo = TRUE}
-Consider a scalar response variable with $J$ observations, which I represent by the vector $y$, that equals a sum of $K$ component response variables $y^{(k)}$ each scaled by a non-zero factor $c^{(k)}$,
+Consider Equation \@ref(eq:regression) for the simple linear regression of log density of a species $i$ on covariate $x$.
+Let $y$ stand for the actual log density of a focal species $i$, $z$ stand for the log density that we've measured, and $d = z - y$ equal the difference between the two, which here is the log efficiency of species $i$ minus the log mean efficiency of the sample.
+Let $s_{xy}$ denote the sample covariance between variables $x$ and $y$, $s^{2}_{x} = s_{xx}$ and $s_{x}$ denote the sample variance and standard deviation, and $r_{xy} = s_{xy}/(s_{x}s_{y})$ denote the sample correlation.
+
+The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of $z$ equals the sample covariance of $z$ and $x$ divided by the sample variance in $x$ or, equivalently, the sample correlation of $z$ and $x$ multiplied by the ratio of their sample standard deviations,
 \begin{align}
-  (\#eq:y-sum)
-  y = \sum_{k = 1}^K c^{(k)} y^{(k)}.
+  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
 \end{align}
-(We may need to assume that the $y^{(k)}$ are linearly independent.)
-Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
-Let $\hat \gamma$ denote the least-squares estimate of the coefficient matrix $\gamma$ in the linear regression
+From the (bi)linearity of sample covariances it follows that
 \begin{align}
-  y = X \gamma + \epsilon.
+  \hat \beta_z 
+  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
+  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
+  = \hat \beta_y + \hat \beta_d,
 \end{align}
-Similarly, let $\hat \gamma^{(k)}$ denote the least-squares estimates for the $K$ linear regressions
+where $\hat \beta_y$ and $\hat \beta_d$ denote the slope estimates for $y$ and $d$ (were these values to be known).
+The absolute error in the estimate $\hat \beta_{z}$ of $\hat \beta_{y}$ is therefore $\hat \beta_{d}$; it is large in a practical sense when $\hat \beta_{d}$ is large (in absolute value) compared to $\hat \beta_{y}$, which corresponds to the covariance of $d$ with $x$ being large compared to the covariance of $y$ with $x$.
+
+In our case, the covariance of $d$ equals the negative of the covariance of log mean efficiency with $x$.
+The absolute error in $\hat \beta$ equals the negative covariance of log mean efficiency scaled by the variance in $x$.
+This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
+For species that covary with $x$ more strongly than the log mean efficiency, the error will be relatively small.
+This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with $x$ compared to the log density of the focal species.
+
+We can similarly understand the impact of measurement error on the precision of our slope estimates.
+The OLS and MLE estimated standard error in $\hat \beta$ are both approximately
 \begin{align}
-  y^{(k)} = X \gamma^{(k)} + \epsilon^{(k)}.
+  \hat{\text{se}}(\hat \beta)
+  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
 \end{align}
-The least-squares coefficient estimates for $y$ are given by the sum of those of the $y^{(k)}$,
+where $s_{\hat \varepsilon}$ is the sample standard deviation of the residuals
+(@wasserman2004allo Chapter 13).
+The sample residuals of $z$, $y$, and $d$ have a similar relationship to the regression coefficient estimates,
 \begin{align}
-  \hat \gamma = \sum_{k=1}^K c^{(k)} \hat \gamma^{(k)}.
+  \hat \varepsilon_z 
+    &\equiv z - \hat \beta_z x
+  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
+  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
+  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
 \end{align}
-```
+(Note, here I've omitted the subscript indicating the dependence on the sample.)
+It follows that the sample variances of the residuals of $z$, $y$, and $d$ are related through
+\begin{align}
+  s^2_{\hat \varepsilon_{z}} 
+  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
+  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
+\end{align}
+The standard deviation of the $z$ residuals is increased above that of the $y$ residuals when the $d$ residuals are either uncorrelated or positively correlated with the $y$ residuals, but may be decreased when the $y$ and $d$ residuals are negatively correlated.
 
-```{proof, echo = TRUE}
-The Moore-Penrose pseudoinverse of $X$ is $X^+ = (X^T X)^{-1} X^T$.
-The least squares estimates are given by multiplying the matrix $X^+$ by the corresponding response vector ([Wikipedia](https://en.wikipedia.org/wiki/Ordinary_least_squares#Matrix/vector_formulation)), so that $\hat \gamma = X^+ y$ and $\hat \gamma^{(k)} = X^+ y^{(k)}$ (for $k = 1, \dots, K$).
-The result then follows from the stipulation \@ref(eq:y-sum) and the linearity of matrix multiplication,
-\begin{align}
-  \hat \gamma 
-  = X^+ y 
-  = X^+ \sum_{k = 1}^K c^{(k)} y^{(k)}
-  = \sum_{k = 1}^K c^{(k)} X^+ y^{(k)}
-  = \sum_{k = 1}^K c^{(k)} \hat \gamma^{(k)}.
-\end{align}
-```
-
-We can use \@ref(thm:regression-linearity) to describe how measurement error, such as that caused by experimental bias, affects least-squares estimates of regression coefficients.
-
-```{theorem, label = "regression-error", echo = TRUE, name = "Error in regression coefficients"}
-Suppose that $y$ is some response variable in nature, $z$ is our imperfect measurement of $y$, and $d = z - y$ the difference between the two, so that $z =  y + d$.
-Let each be a vector of length $J$ describing a set of $J$ observations.
-Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
-Consider the linear regression equations for $y$, $d$, and $z$,
-\begin{align}
-  y &= X \gamma^{(y)} + \epsilon^{(y)} \\
-  z &= X \gamma^{(z)} + \epsilon^{(z)} \\
-  d &= X \gamma^{(d)} + \epsilon^{(d)}.
-\end{align}
-The relationship between the least-squares estimates mirrors that between the variables themselves,
-\begin{align}
-  \hat \gamma^{(z)} = \hat \gamma^{(y)} + \hat \gamma^{(d)}.
-\end{align}
-```
-
-```{proof, echo = TRUE}
-The result follows directly from $z =  y + d$ and \@ref(thm:regression-linearity).
-```
-
-```{corollary, label = "regression-error-simple", echo = TRUE}
-For the special case of simple linear regression,
-\begin{align}
-  y &= \gamma_0^{(y)} + \gamma_1^{(y)} x + \epsilon^{(y)} \\
-  z &= \gamma_0^{(z)} + \gamma_1^{(z)} x + \epsilon^{(z)} \\
-  d &= \gamma_0^{(d)} + \gamma_1^{(d)} x + \epsilon^{(d)},
-\end{align}
-the estimates for the intercept and slope coefficients satisfy
-\begin{align}
-  \hat \gamma_0^{(z)} &= \hat \gamma_0^{(y)} + \hat \gamma_0^{(d)} \\
-  \hat \gamma_1^{(z)} &= \hat \gamma_1^{(y)} + \hat \gamma_1^{(d)}.
-\end{align}
-```
-
-Next, aiming to show the forms for the different transformations of interest.
-It might be useful to cover the case where $d$ is split into a constant and a sample-specific part, since probably everything of interest will follow from that.
-
-What, if any general results should go in the main text?
-
-Also relate the expected values of $y$ and $z$ conditional on $X$?.
-
-To figure out what I need:
-
-1. extend the below result for log proportions to multiple regression
-2. work out the derivation in terms of \@ref(thm:regression-error)
-3. Write a corollary if it seems to be useful for doing all the cases (e.g. logit etc)
-
-### Linear regression of microbiome abundances
-
-First, consider log proportions.
-In this case, the intercept coefficient is biased upward by $\log B_i$ and the other coefficients are biased downward by the corresponding coefficient of $\log \bar B$, based on the error formulas for individual samples.
-
-TODO: write the general result (for multiple regression, assuming that an intercept term is present.
-
-```{proposition, echo = TRUE}
-Consider the linear equations
-\begin{align}
-  (\#eq:least-squares-formulation)
-  \log a_i    &= \alpha_0 + \alpha_1 x \\
-  \log \bar B &= \beta_0 + \beta_1 x   \\
-  \log m_i    &= \gamma_0 + \gamma_1 x,
-\end{align}
-and let $\hat \alpha_0$, etc. denote the least-squares estimates for these equations supposing perfect information (i.e., that the $a_i$ and $B_i$ are known).
-The coefficients for $\log m_i$ are related to those for $\log a_i$ and $\log \bar B$ through the equations
-\begin{align}
-  (\#eq:least-squares-estimates)
-  \hat \gamma_0 &= \hat \alpha_0 + \log B_i - \hat \beta_0 \\
-  \hat \gamma_1 &= \hat \alpha_1 - \hat \beta_1,
-\end{align}
-so that the fitted model for $\log m_i$ is given by the right-hand side of
-\@ref(eq:expected-observed-abundance-log-linear-regression) with the coefficients replaced by the fitted values.
-```
-
-```{proof, echo = TRUE}
-```
+In our case, the $d$ residuals equal the negative residuals of the log mean efficiency.
+It is plausible that for most species, their residual variation will have a small covariance with log mean efficiency and the net effect of variation in the mean efficiency will be to increase the estimated standard errors, as occurs with most species in Figure \@ref(fig:regression-example).
+However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \@ref(fig:regression-example).

### 2021-11-08 differential-abundance.Rmd
Fix figure reference
Move regression results into appendix

@@
 Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
-We can see each type of error in Figure 2.
+We can see each type of error in Figure \@ref(fig:error-proportions).
 For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FC (actual FC: 2.3X, measured FC: 6.5X).
@@
 The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$.
-Box \@ref(regression-error) mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \@ref(fig:regression-example) illustrates using simulated data.
+Appendix \@ref(appendix-regression) mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \@ref(fig:regression-example) illustrates using simulated data.
 Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
@@
 <!-- end Figure -->
-
-::: {.rmdnote}
-
-## Box: Error in estimated regression coefficients and standard errors {#regression-error}
-
-Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.
-
-Consider Equation \@ref(eq:regression) for the simple linear regression of log density of a species $i$ on covariate $x$.
-Let $y$ stand for the actual log density of a focal species $i$, $z$ stand for the log density that we've measured, and $d = z - y$ equal the difference between the two, which here is the log efficiency of species $i$ minus the log mean efficiency of the sample.
-Let $s_{xy}$ denote the sample covariance between variables $x$ and $y$, $s^{2}_{x} = s_{xx}$ and $s_{x}$ denote the sample variance and standard deviation, and $r_{xy} = s_{xy}/(s_{x}s_{y})$ denote the sample correlation.
-
-The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of $z$ equals the sample covariance of $z$ and $x$ divided by the sample variance in $x$ or, equivalently, the sample correlation of $z$ and $x$ multiplied by the ratio of their sample standard deviations,
-\begin{align}
-  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
-\end{align}
-From the (bi)linearity of sample covariances it follows that
-\begin{align}
-  \hat \beta_z 
-  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
-  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
-  = \hat \beta_y + \hat \beta_d,
-\end{align}
-where $\hat \beta_y$ and $\hat \beta_d$ denote the slope estimates for $y$ and $d$ (were these values to be known).
-The absolute error in the estimate $\hat \beta_{z}$ of $\hat \beta_{y}$ is therefore $\hat \beta_{d}$; it is large in a practical sense when $\hat \beta_{d}$ is large (in absolute value) compared to $\hat \beta_{y}$, which corresponds to the covariance of $d$ with $x$ being large compared to the covariance of $y$ with $x$.
-
-In our case, the covariance of $d$ equals the negative of the covariance of log mean efficiency with $x$.
-The absolute error (i.e., the _bias_ in a statistical sense) in $\hat \beta$ is equals the negative covariance of log mean efficiency scaled by the variance in $x$.
-This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
-For species that covary with $x$ more strongly than the log mean efficiency, the error will be relatively small.
-This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with $x$ compared to the log density of the focal species.
-
-We can similarly understand the impact of measurement error on the precision of our slope estimates.
-The OLS and MLE estimated standard error in $\hat \beta$ are both approximately
-\begin{align}
-  \hat{\text{se}}(\hat \beta)
-  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
-\end{align}
-where $s_{\hat \varepsilon}$ is the sample standard deviation of the residuals
-(@wasserman2004allo Chapter 13).
-The sample residuals of $z$, $y$, and $d$ have a similar relationship to the regression coefficient estimates,
-\begin{align}
-  \hat \varepsilon_z 
-    &\equiv z - \hat \beta_z x
-  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
-  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
-  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
-\end{align}
-(Note, here I've omitted the subscript indicating the dependence on the sample.)
-It follows that the sample variances of the residuals of $z$, $y$, and $d$ are related through
-\begin{align}
-  s^2_{\hat \varepsilon_{z}} 
-  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
-  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
-\end{align}
-The standard deviation of the $z$ residuals is increased above that of the $y$ residuals when the $d$ residuals are either uncorrelated or positively correlated with the $y$ residuals, but may be decreased when the $y$ and $d$ residuals are negatively correlated.
-
-In our case, the $d$ residuals equal the negative residuals of the log mean efficiency.
-It is plausible that for most species, their residual variation will have a small covariance with log mean efficiency and the net effect of variation in the mean efficiency will be to increase the estimated standard errors, as occurs with most species in Figure \@ref(fig:regression-example).
-However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \@ref(fig:regression-example).
-
-:::
-<!-- </details> -->

### 2021-11-08 appendix.Rmd
Create new appendix

# (APPENDIX) Appendix {-}

# Details of MGS model {#model-details}

- Give the model in terms of arbitrary taxa
- Consider effect of variation in efficiency within a taxon
- Discuss assumption about efficiencies being constant at the species level

# Methods for assaying absolute densities {#review-absolute-methods}

# Taxonomic bias, counting noise, and zeros

- Motivation/intro P

Simplest model: deterministic bias model + multinomial sampling.
Number of reads assigned to a species as multinomial with a given total read count for the sample, and a the biased proportions as input.
(Can treat this as a heuristic model, or can give an actual justification as to why this makes sense even for bioinformatic bias.)
\begin{align}
\text{reads}_i(a)
  &= \text{Multinomial}\left( \text{total reads}(a), \text{prop}_{i}(a) \times \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} \right).
\end{align}
The expected read count (conditional on read total) is
\begin{align}
E\left[\text{reads}_i(a)\right] = \text{total reads}(a) \times \text{prop}_{i}(a) \times \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)}.
\end{align}

(Note, might be simpler to use a Poisson model, especially for overdispersion case.)

Zero counts are likely when the expected read count is $\lesssim 1$. 
The expected read count is lower and hence the likelihood of zeros is much higher for species with efficiencies substantially below the sample mean.
Whereas in the absence of taxonomic bias, an observation of zero is taken as evidence that the actual proportion of a species is similar or less than the reciprocal of the total read count, with taxonomic bias, it instead implies only that
\begin{align}
  \text{prop}_{i}(a)
  \lesssim \frac{1}{ \text{total reads}(a) } 
  \times \frac{\text{mean efficiency}(a)}{ \text{efficiency}_{i}}.
\end{align}

The likelihood of a zero for a given taxon varies with the mean efficiency of a sample.
For example, when the vaginal microbiome shifts from Lactobacillus to Gardnerella dominance and the mean efficiency drops by 10X in figure x, it becomes ...
This becomes problematic for presence/absense analysis.
But it is also an issue for analysis of log proportions, log ratios, or log densities, in a manner that depends on how the DA method handles zeros.
A simple and common approach is to ignore the uncertainty from the counting process, and opt to replace zeros with some defined, small value (either in the counts or in the proportions).
(TODO: implications)
Some statistical packages (e.g. corncob, fido) take a more sophisticated approach and directly account for the uncertainty by modeling the counting process.
For these methods, bias against an organism can cause reduced precision, in a sample-dependent manner.

- Maybe: implications of overdispersion / multiplicative noise prior to counting noise.
- Note, the above is for regression analysis, but looking at two-sample case is also illustrative. Can see a case where having a zero or very small proportion (if overdispersed) makes it very hard to 
- Consider case of a zero observation; how to interpret, in the light of taxonomic bias?
- Limit of zero efficiency -> a value of zero reveals no information about the taxon.
- Maybe: Implications for presence/absense analysis

# Taxonomic bias in total-density measurements {#appendix-total-density}

A noted in the main text, we should expect species-specific efficiencies for total density estimates.
State the model: Assume constant absolute measurement efficiencies, $\text{efficiency}^{\text{tot}}_i$, which denote the absolute efficiency with which a cell of species $i$ contributes to the measurement of total density in sample $a$.
This implies consistent relative efficiencies and that density measurement is linear, in the sense that a doubling of the actual density when species composition is fixed always leads to a doubling in the measured density.
Absolute because we're not considering a compositional measurement; the scale of the species' efficiencies does impact the measured density, though not the fold variation across samples.

How do these species specific efficiencies impact the total density measurement?
We can write the measured total density in terms of the species' efficiencies as
\begin{align}
  \widehat{\text{total density}}(a) 
  &= \sum_i \text{density}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
\\&= \text{total density}(a) \cdot \text{mean efficiency}^{\text{tot}}(a)
\end{align}
where
\begin{align}
  (\#eq:app-total-mean-efficiency)
  \text{mean efficiency}^{\text{tot}}(a) 
  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}^{\text{tot}}_j}{\text{total density}(a)}
\end{align}
is the mean efficiency in the sample with respect to the total-density measurement.
The error in the total density depends on the species composition through the mean efficiency of the sample; hence spurious changes in measured density can occur simply due to shifts from high- to low-efficiency species (or vice versa).

How does this error affect the measured densities of individual species when the total is used for normalization?
Substituting the mean efficiency of the total measurement for the error term in Equation \@ref(eq:density-prop-error) gives
\begin{align}
  (\#eq:app-density-prop-error)
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{mean efficiency}^{\text{tot}}(a)}{\text{mean efficiency}(a)}.
\end{align}
<!-- \begin{align} -->
<!--   \newcommand{\allspecies}{\boldsymbol{\cdot}} -->
<!--   \widehat{\text{density}}_{i}(a)  -->
<!--   = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}_{\allspecies}^{\text{tot}}(a)}{\text{efficiency}_{\allspecies}(a)}. -->
<!-- \end{align} -->
Note, having more similar efficiencies between MGS and total measurements does not necessarily improve the accuracy of the individual sample density estimates (possible misconception of other article?), since even if they perfectly cancel the species-specific error remains.
Yet it does improve fold change estimates.
Equation \@ref(eq:app-density-prop-error) shows that the variation in the fold error across samples is driven by the variation in the ratio of the mean efficiency of the total measurement and that of the sequencing measurement.
To the extent that these two are positively associated (specifically, their logarithms are positively linearly correlated), then the variation across samples will be lower than if a perfectly accurate total density were used.

Applications of these ideas:

- example: 16S sequencing may be better paired with 16S qPCR than flow cytometry
- example: shotgun sequencing may be best paired with bulk DNA measurement (depending on bioinformatics protocol); can even handle a large fraction of unassigned reads _if_ total reads are used in the denominator of the proportion calculation (rather than assigned reads)
- example: estimating total density from a spike-in; mathematically reduces to the reference-species measurement; here the error completely offsets.

Extra assumptions in this section:
- Linearity of DNA extraction
- Thinking that log efficiencies between MGS and total are positively correlated comes from the idea that we can treat bias from different steps and/or sources as approximately independent

# Taxonomic bias in spike-ins and targeted measurements

- Distinction between cellular vs DNA and  pre- vs post-extraction spike-ins and targeted measurements
- Pull from Section 5 of the theory notebook

### 2021-11-08 real-world-implications.Rmd
Revise Implications section

@@
 @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome with the MGS protocol developed for the Vaginal Human Microbiome Project (VaHMP).
-@mclaren2019cons later used this data to estimate the efficiencies of these seven species and develop a method to use these estimates to infer the true compositions of natural samples (see Section \@ref(#calibrate-compositions)).
+@mclaren2019cons later used this data to estimate the efficiencies of these seven species and develop a method to use these estimates to infer the true compositions of natural samples (see Section \@ref(calibrate-compositions)).
 SI Analysis (MOMS-PI) uses these estimated efficiencies to correct the compositions and explore the implied variation in mean efficiency in the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) @fettweis2019thev.
-The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single _Lactobacillus_ species often forming a majority of reads (considered to be the healthy state) but with occasional shifts to high proportions of other species, such as _Gardnerella vaginalis_, _Atopobium vaginae_, and _Streptococcus agalactiae_, which are thought to cause or indicate poor vaginal health and increased risk of preterm birth.
-The analysis of @mclaren2019cons showed that the efficiencies of two common _Lactobacillus_ species, _L.
-crispatus_ and _L.
-iners_, were 20-30X greater than that of _G.
-vaginalis_, _A.
-vaginae_, and _S.
-agalactiae_.
-Analysis of microbiome trajectories of a single patient across visits indicates that the mean efficiency typically decreases by a factor of 3-10X when the community shifts from a high to low-_Lactobacillus_ state.
-(NOTE: This is a super preliminary claim based on looking at just one woman, and is just a placeholder)
+The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
+The three species that most often dominate the calibrated profiles are _Gardnerella vaginalis_, _Lactobacillus iners_, and _Lactobacillus crispatus_.
+Our mock-community analysis indicate _G. vaginalis_ has the lowest efficiency of the seven species, roughly 20X lower than _L. crispatus_ and 30X lower than _L. iners_ (the most efficiently measured).
+As a result, the mean efficiency varies substantially across samples, with samples dominated by a _Lactobacillus_ species typically having an efficiency that is 3-20X greater than samples dominated by _Gardnerella_ (SI figure).
+Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure).
+Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure).
 Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
-
-<!-- Can cite brooks2015thet and pace2021comp for Group-B strep / S. agalactiae as a "pathobiont".  -->
 
@@
 The authors used ITS amplicon sequencing to measure communities before and after infection, along with mock communities which they used to estimate the species efficiencies with the method of @mclaren2019cons.
-(DNA mocks were used, so that bias due to DNA extraction is not included, but other major bias sources such as PCR and ribosomal copy-number-variation are included.) The pathogen was 10X more efficiently measured than the median commensal and 40X more efficiently measured than the lowest-efficiency commensal.
-A re-analysis of their data (SI Analysis) shows that the mean efficiency varies by ???
-in pre-infection communities and by an average of ???
-pre- and post-infection, with the latter driven by the dramatic increase in the high-efficiency pathogen.
+(DNA mocks were used, so that bias due to DNA extraction is not included, but other major bias sources such as PCR and ribosomal copy-number-variation are included.) 
+SI Analysis X reanalyses this experiment to consider the impact that bias has on different possible DA analyses.
+There was a 13X difference between the most and least efficiently measured commensals, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
+Despite the 13X variation among commensals, the mean efficiency is remarkably stable across communities prior to pathogen colonization.
+Consequently, taxonomic bias creates little systematic error in proportion-based DA analysis on these samples.
+However, the high efficiency of the pathogen typically led to a roughly 6X increase in mean efficiency in post-colonization samples.
+An analysis of how commenals responded to calibration that used uncalibrated proportions without first discarding the pathogen reads would therefore tend to suggest that the commensals had substantially smaller fold changes than they actually did.
 
-These case studies indicate that one mechanism for variability in the mean efficiency is swings between very high ($\sim 1$) and low ($\ll 1$) proportions of species with particularly high or low efficiencies.
-In highly diverse ecosystems, as seen in soil and the human gut (CHECK), one species rarely dominates within a given sample and so such large swings are not possible.
-We might therefore think that the mean efficiency will remain fairly stable in these settings.
-In fact, one can show mathematically that in randomly assembled communities, the mean efficiency becomes more stable as (Inverse Simpson) diversity increases, as the mean effectively averages over a greater number of species.
-Real communities do not assemble randomly, however.
-Variance in the mean efficiency may remain high if environmental filtering leads to different groups of organisms dominating in different environments, particularly if shared phenotypic values or evolutionary history causes these groups to differ in their efficiencies as well as their ecology.
-For example, gut microbiomes typically have high diversity at the species level but are dominated by just a small number of phyla, and two in particular: The Bacteroidetes and the Firmicutes, the ratio of which varies substantially across individuals (CHECK).
-Many DNA extraction protocols have been found to more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though variation can also be substantial within phyla, @mclaren2019cons).
-Gut samples with large differences in their Bacteroidetes and Firmicutes proportions may therefore have large variation in mean efficiency even if the individual species within each phylum never reach a large proportion.
-On the other hand, competitive-exclusion dynamics can favor more stable taxonomic and/or phenotypic compositions across samples and might thereby drive a more stable mean efficiency than under random assembly
+It is clear that an individual species with an unusually high or low efficiency can drive large changes in the mean efficiency when it shifts from a very large ($\sim 1$) to very small proportion ($\ll 1$) of the sample.
+Consideration of dynamics in the human gut illustrates another a way in which large shifts in higher-level taxa might create similar shifts, due to the fact that efficiencies do not vary independently among closely related species.
+Gut microbiomes are dominated by two phyla: The Bacteroidetes and the Firmicutes, the ratio of which varies substantially across individuals.
+Within a sample, there are usually multiple species contributing to the abundance of each phylum, with the median Inverse Simpson diversity being around 4 for each phylum in the HMP2 IBD study.
+Thus changes from Bacteroidetes to Firmicutes dominance are unlikely to be driven by a single species.
+However, if Firmicutes species tend to have systematically different efficiencies than Bacteroidetes species, then changes in phylum dominance can be expected to drive substantial changes in the mean efficiency even if many species contribute.
+Indeed, several studies have found DNA extraction protocols to more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though substantial variation within phyla also occurs, @mclaren2019cons).
 
@@
 In the plant-fungal experiment of @leopold2020host, the increase in pathogen proportion after infection leads to a large increase in the mean efficiency over time, which systematic increases the observed decline in the proportions of commensal species (SI Analysis).
-The ratio of Bacteroidetes to Firmicutes has been linked to several host traits and health conditions, and thus the variation in efficiency between these two phyla may also drive an association of mean efficiency with these traits.
+The Bacteroidetes-to-Firmicutes ratio has also been linked to host conditions.
+Yet in the HMP2 IBD dataset, the large variation in this ratio is largely independent of disease status, suggesting that the primary effect of any variation in mean efficiency driven by these phyla would be to increase noise rather than create systematic error in DA analysis.
 
@@
 This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
+Additional measurements by @lloyd2020evid make it possible to investigate this possibility explicitly.
+The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for particular taxa.
+For instance, in Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years using the proportion-based densities and 10.3 years by qPCR (@lloyd2020evid Table 1 and Figure 4).
+These doubling times correspond to highly similar estimated slopes of log density per year of $1/10.1 \approx 0.99$ and $1/10.3\approx 0.97$.
+The small difference between these slopes suggests that there is little correlation of log mean efficiency with burial time and hence little systematic error in the estimated doubling times derived from the proportion-based density estimates.
+
 Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.
@@
 It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
-_(This seems intuitively likely and true in examples, but I haven't analyzed it much mathematically and don't have a firm handle on this question.)_
 Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
@@
 
-- Todo: Address DA analysis of aggregate (or synthetic) taxa, such as phyla, which may be prone to errors due to changes in the species-level make-up of the taxon.
+## Loss of precision and power
 
-## Loss of power and precision (stub)
+Another way in which taxonomic bias hinders DA analysis is by increasing the noise in abundance measurements, leading to a reduction in the precision of LFC estimates and thus the power to detect true associations.
 
-Variation in mean efficiency that is not associated with covariates can still majorly hinder the goals of microbiome studies by causing reductions in the precision of LFC estimates and the power to detect true associations of a given size.
+There are two distinct mechanisms by which such reductions occur.
+The first was described in Section \@ref(differential-abundance): Variation in the mean efficiency that is random (in the sense of being unassociated with the covariates in a regression analysis) tends to increase the standard errors in estimated regression coefficients.
+As noted above, this effect may be the dominant one in cases like the HMP2 IBD dataset where large variation in dominant phylum is independent of disease status.
+Correcting for the variation in mean efficiency using the calibration methods described in the next section can reverse this effect, thereby increasing the precision as well reducing the systematic error in DA inference.
 
-- Illustrate with case studies?
-- Mention counting noise as an important factor that we've so far ignored but that can interact with taxonomic bias to lead to very low precision in LFC estimates.
-Low efficiency can put a species below the detection limit set by sample read depth, making it impossible to get precise LFCs.
-Study-specific efficiencies can explain why associations often do not replicate across studies.
-Example: _Gardnerella vaginalis_ associations are not detected in certain vaginal microbiome studies that used V13 primers which we now know are bad at amplifying it.
-- Conclude with a nod to the value of the solutions in Section \@ref(solutions) to address these issues: Correcting the errors caused by mean-efficiency variation can improve precision, and bias-sensitivity analysis and bias-aware meta-analysis can give a better sense of the increased uncertainty in LFC estimates due to unknown efficiencies, thereby helping us to understand the increased rate of false positives and/or the larger range of LFC estimates that are compatible with the observed data than what is implied by a bias-free model.
+The second mechanism involves the counting noise that stems from the random sampling of reads in the sequencing experiment. 
+This counting noise is expected to be the dominant source of noise for species whose expected proportion _after accounting for taxonomic bias_ is $\lesssim 1 / \text{total reads}$ for a given sample.
+Having an atypically low efficiency makes a species much more likely to be near this threshold and hence to receive highly uncertain measurements.
+Such imprecise measurements can make it near impossible to make biologically meaningful conclusions.
+For example, in the experiment of @leopold2020host described above, the median read counts in samples taken after pathogen colonization for the 5 commensals with the lowest efficiencies were 0, 0, 0, 1, and 11.
+These species' efficiencies ranged from 40X-10X lower than the pathogen's efficiency; thus, in the absence of bias the counts may have been much higher and it might have been possible to make much more precise conclusions about how their relative abundances changed during pathogen invasion.
+The literature on the vaginal microbiome during pregnancy provides us with another example: Disagreement among studies over whether _Gardnerella_ is positively associated with preterm birth seems to be in part driven by the fact that several studies used 16S primers that are extremely poor at amplifying _Gardnerella_ and hence lack the precision necessary to find associations even were they to exist.
+Unfortunately, even perfect knowledge of the measurement efficiencies will not allow us to overcome the low precision caused by small expected read counts; however, the methods described in the section provide principled ways to assess whether the low counts are due to taxonomic bias rather than truly low frequencies in a species' abundance.

### 2021-11-09 abundance-measurement.Rmd
Revise section on abundance measurement

@@
 We consider a set of microbial species that are measured by a specific MGS protocol.
-The protocol assigns each sequencing read to at most one sample andsample and  species, and we ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.
-Our model stipulates that the assigned read count of species $i$ in sample $a$ equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
+The protocol assigns each sequencing read to at most one sample and species and we ignore the possibility of contamination or false-positive taxonomic assignments---we assume that reads assigned to a given species and sample really do come from that species and sample.
+Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the random error due to the sampling of sequencing reads and aspects of the sequencing process.
+Our model stipulates that the assigned read count of species $i$ in sample $a$ equals its cell density multiplied by a species-specific factor and a sample-specific factor,
 \begin{align}
@@
 The term *relative abundance* carries multiple meanings in microbiome science.
-Microbiome researchers often use the relative abundance of a species in a sample to refer to its *proportion*, or number of cells from that species divided by the total number of cells from the taxonomic domain of interest (e.g.
-Prokaryotes).
-A second view, drawn from the statistical framework of Compositional Data Analysis, equates the relative abundances of species within a sample with the *ratios* of cell densities among two or more species (e.g.
+Often, relative abundance refers to the number of cells of a given species divided by the total number of cells from the taxonomic domain of interest (e.g. Prokaryotes).
+We use the term *proportion* for this type of relative abundance.
+A second view, drawn from the statistical framework of Compositional Data Analysis (CoDA), equates the relative abundances in a sample with the *ratios* among two or more species (e.g.
 a ratio of 5:4:2 among three species).
-By this view, it is meaningless to refer to the relative abundance of an individual species.
+By this view, it is meaningless to refer to the relative abundance of a species independently of one or more others.
+Confusingly, many microbiome papers use 'relative abundance' to mean proportion despite analysing these 'relative abundances' with CoDA-inspired ratio-based analyses.
+Additionally, as we will see, DA analyses based on proportions and ratios are differently affected by bias.
+For these reasons, we distinguish between proportions and ratios throughout.
 Although other interpretations exist^[e.g.
@@
 
-A range of experimental and computational methods can be used to convert the relative-abundance information in the read counts into measurements of (absolute) cell densities;
-however, these methods can largely be partitioned into those that use the measured proportions or the measured ratios provided by MGS.
-The error in density measurements due to taxonomic bias therefore follows that of the proportion and ratio measurements described above.
+A wide range of experimental methods exist for converting the relative-abundance information in MGS measurements into measurements of (absolute) cell densities;
+however, these methods can be generally partitioned into two types, based on whether the absolute-abundance information derives from knowing the density of the total community or knowing the density of one or more individual species.
 
-Proportion-based density measurement is based on the idea that if we know the (true) proportion of a species and the (true) total density of all species, we can multiply these numbers obtain the species' density.
-This method therefore involves multiplying the proportion of a species measured from MGS with a measurement of total microbial density obtain through other means (e.g., through direct cell counting or qPCR of a universal marker gene) to measure its density,
+**Methods using the density of the total community:**
+The total microbial density of a community can be assayed by a variety of methods (reviewed in Appendix \@ref(review-absolute-methods)).
+Once the total density is known, the density of individual species can be estimated by multiplying the total density by the species' proportions (as measured by MGS),
 \begin{align}
@@
   \widehat{\text{density}}_{i}(a) 
-  = \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{total density}}(a).
+  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{total density}}(a)
+\\&= \text{reads}_{i}(a) \cdot \frac{\widehat{\text{total density}}(a)}{\text{total reads}(a)}.
 \end{align}
-Substituting the expression \@ref(eq:prop-error) for the measured proportion and allowing for error in the total density measurement, leads to the expression
+We can rewrite the species' density measurements to account for the effect of taxonomic bias on the MGS proportions (Equation \@ref(eq:prop-error)) and any error in the total density measurement, 
 \begin{align}
@@
   = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
+%  \cdot \frac{\widehat{\text{total density}}(a)}{\text{total density}(a)}.
   \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{total density}}(a) \end{array}.
 \end{align}
-The multiplicative error consists of two factors: the error in the estimated proportion (due to taxonomic bias in the MGS measurement), and any error in the estimated total density.
-In the case of perfect knowledge of the total density, the fold error in proportion-based density estimates varies across samples inversely with the mean efficiency.
+The variable fold error that bias creates in the measured proportions directly transfers to the density measurements.
 
-Ratio-based density estimation follows from the fact that, in the absence of taxonomic bias, the conversion rate from reads to density is just the sequencing effort of the sample and is thus the same for all species (Equation \@ref(eq:measurement-model)).
-Knowing the density of at least one _reference_ species therefore enables us to convert from reads to density for all species.
-One approach to obtaining such reference species is to add one or more extraneous species in known (typically constant) abundance (a "spike-in") to each sample prior to sequencing (REFs).
-A second approach is to determine a set of naturally occurring species that are thought to have a constant, though typically unknown, abundance across samples; in this case we can only estimate species' densities relative to these "housekeeping species" (REFs).
-A third (seemingly untried) possibility is to measure the density of one or more naturally occurring species using a method such as targeted CFU counting, fluorescence flow cytometry or ddPCR directly on cells, or ddPCR or qPCR on extracted DNA.
-Given the estimate of the density of reference taxon $r$, we estimate the density of taxon $i$ by multipling the ratio of $i$ to $r$ in the reads by the density of $r$,
+Taxonomic bias can also create systematic error in the total-density measurements (Appendix \@ref(appendix-total-density)).
+For now, we suppose that the total density can be measured accurately; 
+later, Section \@ref(solutions) describes how experiments might be designed such that taxonomic bias in the total density measurement offsets that in the MGS proportions, leading to a (more) consistent fold error in the measured densities.
+
+**Methods using the density of one or more reference species:**
+The density of one or more species can be directly measured using targeted measurement methods like species-specific qPCR; alternatively, the densities of one or more species may be thought to be constant across samples either from background knowledge or because they have been spiked in at a fixed density (these situations are reviewed in Appendix \@ref(review-absolute-methods)).
+These _reference species_ with known or constant density can be used to convert read counts to densities for all species.
+For instance, given the density of a reference species $r$, we can estimate the density of a focal species $i$ by 
 \begin{align}
   (\#eq:density-ratio-meas)
-  \widehat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \widehat{\text{density}}_{r}(a)
+%  \widehat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \widehat{\text{density}}_{r}(a).
+  \widehat{\text{density}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{density}}_{r}(a)}{\text{reads}_{r}(a)}.
 \end{align}
-(Appendix ??? describes how this equation can be extended to multiple reference species.)
-For a constant reference species with unknown density, we can set $\widehat{\text{density}}_{r}(a)$ equal to 1 to estimate density in units of the density of the reference species.
-The error in this ratio-based density estimate is
+Alternatively, if we only know that the reference species is constant, we can set $\widehat{\text{density}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) equal to 1 to estimate the density in unknown but fixed units, which is sufficient for multiplicative DA analysis.
+(Appendix \@ref(model-details) considers possible extensions to multiple reference species.)
+It follows from Equation \@ref(eq:ratio-error) that the error caused by taxonomic bias in this density measurement is
 \begin{align}
@@
 \end{align}
-Again, the multiplicative error consists of two factors: the error in the estimated ratio (due to taxonomic bias in the MGS measurement), and error in the estimated density of the reference taxa.
-So long as the fold error in the reference density is constant across samples, so will be the error in the estimated density of the focal species $i$.
+Here, it is the constant error in the measured ratio of species $i$ to $r$ that propagates into the density measurement.
+There will generally be systematic error in the density of the reference species; however, if the systematic fold error is constant across samples, so will be that of the densities of other species.
 
-- Note: The current distinction between proportion- and ratio-based density estimation is incomplete, as the example of spike-in normalization shows: Whether the density of species $i$ is estimated by Equation \@ref(eq:density-ratio-meas), or by first estimating total density from the ratio of non-spike-in to spike-in reads and then using Equation \@ref(eq:density-prop-meas), the results are mathematically identical.
-Future revisions should clarify this point.
+**Differences between the two approaches:**
+Equations \@ref(eq:density-prop-error) and \@ref(eq:density-ratio-error) show that the manner in which taxonomic bias in the sequencing measurement impacts the two approaches mirrors that of proportions and ratios.
+If the fold error in the total-density or reference species measurement is negligible (or at least constant) then the fold errors in species densities by the total-density approach vary across samples with the mean efficiency, whereas the fold errors by the reference-species approach are constant.
+The fundamental determinant of whether constant error is obtained is not which of these two equations is ultimately used, but rather whether densities of individual species, which (we assume) have constant efficiency, or the total community, which has varying efficiency, provide the absolute-abundance information.
+Studies using spike-ins as constant reference species will often, instead of using Equation \@ref(eq:density-ratio-meas), first use the spike-in species to estimate the total density, and then estimate the density of focal species with Equation \@ref(eq:density-prop-meas).
+If the calculation is done carefully, however, this method still yields constant fold errors so long as the species efficiencies (including those of the spike-in) are constant across samples (Appendix \@ref(appendix-total-density)).

### 2021-11-09 abundance-measurement.Rmd
Fix figure rendering in pdf output
The calls to `knitr::include_graphics()` were failing in pdf output
because these calls have to be included in a top level R statement. This
commit fixes this issue; it also removes code redundancy, switches to
relative paths (the recommended method for bookdown), and puts the pdf
images in a new folder contained within the project directoty.

@@
 ```{r error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
-svg_path <- here("figures/illustrations/error-proportions.svg")
+svg_path <- "figures/illustrations/error-proportions.svg"
 if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
+  fig_path <- svg_path
 } else if (knitr::is_latex_output()) {
-  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
-  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
-    "error-proportions.pdf")
-  args <- c(
-    "--export-area-page", 
-    "--export-type=pdf", 
-    str_c("--export-filename=", pdf_path),
-    svg_path
-  )
-  system2("inkscape", args = args)
-  knitr::include_graphics(pdf_path)
-  rm(pdf_path)
+  fig_path <- svg_to_pdf(svg_path)
 }
-rm(svg_path)
+knitr::include_graphics(fig_path)
+rm(svg_path, fig_path)
 ```

### 2021-11-09 differential-abundance.Rmd
Fix figure rendering in pdf output
The calls to `knitr::include_graphics()` were failing in pdf output
because these calls have to be included in a top level R statement. This
commit fixes this issue; it also removes code redundancy, switches to
relative paths (the recommended method for bookdown), and puts the pdf
images in a new folder contained within the project directoty.

@@
 if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
+  fig_path <- svg_path
 } else if (knitr::is_latex_output()) {
-  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
-  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
-    "regression-example.pdf")
-  args <- c(
-    "--export-area-page", 
-    "--export-type=pdf", 
-    str_c("--export-filename=", pdf_path),
-    svg_path
-  )
-  system2("inkscape", args = args)
-  knitr::include_graphics(pdf_path)
-  rm(pdf_path)
+  fig_path <- svg_to_pdf(svg_path)
 }
-rm(svg_path)
+knitr::include_graphics(fig_path)
+rm(svg_path, fig_path)
 ```

### 2021-11-09 introduction.Rmd
Fix figure rendering in pdf output
The calls to `knitr::include_graphics()` were failing in pdf output
because these calls have to be included in a top level R statement. This
commit fixes this issue; it also removes code redundancy, switches to
relative paths (the recommended method for bookdown), and puts the pdf
images in a new folder contained within the project directoty.

@@
 ```{r basic-problem, fig.cap = '(ref:cap-basic-problem)'}
-svg_path <- here::here(
-  "notebook/_posts/2021-10-27-illustrate-the-problem/illustrate-the-problem_files", "figure-html5/brooks_lcrisp_prop_and_fc-1.svg"
+# NOTE: call to include_graphics() has to be in top level expression
+svg_path <- fs::path(
+  "notebook/_posts/2021-10-27-illustrate-the-problem/illustrate-the-problem_files", 
+  "figure-html5/brooks_lcrisp_prop_and_fc-1.svg"
 )
 if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
+  fig_path <- svg_path
 } else if (knitr::is_latex_output()) {
-  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
-  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
-    "error-proportions.pdf")
-  args <- c(
-    "--export-area-page", 
-    "--export-type=pdf", 
-    str_c("--export-filename=", pdf_path),
-    svg_path
-  )
-  system2("inkscape", args = args)
-  knitr::include_graphics(pdf_path)
-  rm(pdf_path)
+  fig_path <- svg_to_pdf(svg_path)
 }
-rm(svg_path)
+knitr::include_graphics(fig_path)
+rm(svg_path, fig_path)
 ```

### 2021-11-09 solutions.Rmd
Fix figure rendering in pdf output
The calls to `knitr::include_graphics()` were failing in pdf output
because these calls have to be included in a top level R statement. This
commit fixes this issue; it also removes code redundancy, switches to
relative paths (the recommended method for bookdown), and puts the pdf
images in a new folder contained within the project directoty.

@@
 
-```{r calibration-example, fig.cap = '(ref:cap-calibration-example)'}
-fig_path <- here::here(
+```{r calibration-example, fig.cap = '(ref:cap-calibration-example)', out.width = '100%'}
+fig_path <- fs::path(
   "notebook/_posts/2021-10-25-brooks2015thet-calibration",
@@
 if (knitr::is_html_output()) {
-  knitr::include_graphics(svg_path)
+  fig_path <- svg_path
 } else if (knitr::is_latex_output()) {
-  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
-  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
-    "sensitivity-example.pdf")
-  args <- c(
-    "--export-area-page", 
-    "--export-type=pdf", 
-    str_c("--export-filename=", pdf_path),
-    svg_path
-  )
-  system2("inkscape", args = args)
-  knitr::include_graphics(pdf_path)
-  rm(pdf_path)
+  fig_path <- svg_to_pdf(svg_path)
 }
-rm(svg_path)
+knitr::include_graphics(fig_path)
+rm(svg_path, fig_path)
 ```

### 2021-11-09 _common.R

@@ -22,3 +22,20 @@ library(patchwork)
 theme_set(theme_cowplot(12))
 
 close_elts <- function(x) x / sum(x)
+
+# Create a pdf from an svg with inkscape
+# svg_path should be relative to project root
+svg_to_pdf <- function(svg_path) {
+  fig_path <- fs::path(
+    'figures', 'export-pdf', 
+    fs::path_ext_set(svg_path, 'pdf')
+  )
+  args <- c(
+    "--export-area-page", 
+    "--export-type=pdf", 
+    str_c("--export-filename=", fig_path),
+    svg_path
+  )
+  system2("inkscape", args = args)
+  fig_path
+}

### 2021-11-10 abundance-measurement.Rmd
Make minor text edits

@@
 
-Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
+Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
 This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
-We consider a set of microbial species that are measured by a specific MGS protocol.
-The protocol assigns each sequencing read to at most one sample and species and we ignore the possibility of contamination or false-positive taxonomic assignments---we assume that reads assigned to a given species and sample really do come from that species and sample.
-Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the random error due to the sampling of sequencing reads and aspects of the sequencing process.
+We consider a set of microbial species that are measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes samples to assign reads to various microbial species.
+We assume that assigned reads have been correctly assigned to the given species and sample.
+Thus we ignore the possibility of contamination or false-positive taxonomic assignment, though we allow that many reads may not receive a taxonomic assignment and so ultimately be discarded.
+Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'noise' or random error due to the random sampling of sequencing reads and other aspects of the MGS process.
+
 Our model stipulates that the assigned read count of species $i$ in sample $a$ equals its cell density multiplied by a species-specific factor and a sample-specific factor,
@@
 
-The term *relative abundance* carries multiple meanings in microbiome science.
-Often, relative abundance refers to the number of cells of a given species divided by the total number of cells from the taxonomic domain of interest (e.g. Prokaryotes).
-We use the term *proportion* for this type of relative abundance.
-A second view, drawn from the statistical framework of Compositional Data Analysis (CoDA), equates the relative abundances in a sample with the *ratios* among two or more species (e.g.
-a ratio of 5:4:2 among three species).
-By this view, it is meaningless to refer to the relative abundance of a species independently of one or more others.
-Confusingly, many microbiome papers use 'relative abundance' to mean proportion despite analysing these 'relative abundances' with CoDA-inspired ratio-based analyses.
-Additionally, as we will see, DA analyses based on proportions and ratios are differently affected by bias.
-For these reasons, we distinguish between proportions and ratios throughout.
-Although other interpretations exist^[e.g.
-the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most relative and absolute DA methods and the differences between them clarify when and why the effects of bias cancel in a DA analysis.
+We distinguish between two types of species-level *relative abundances*---abundances in which the absolute scale is not intrinsically meaningful.
+The first type consists of the *ratios* among the abundances of two or more species, such as a ratio of 5:4:2 among three species.
+The second type consists of the *proportion* of an individual species, equal to its abundance divided by that of the chosen root taxon (e.g. Prokaryotes).
+Species' ratios and proportions constitute distinct views of relative abundance and hence what it means for relative abundances to change across samples.
+Although proportions are dominant in microbiome DA analysis, researchers increasingly adopt the ratio view via analyses adapted from the statistical framework of Compositional Data Analysis (CoDA).
+The different manner in which taxonomic bias affects ratios vs.
+proportions (@mclaren2019cons) forms the basis in subsequent sections for understanding the extent to which its effects cancel in DA analysis.
 
@@
 <!-- end figure -->
+
+**Difference between species proportions and species ratios:**
+Consistent taxonomic bias at the species level creates a consistent fold error in the ratios among species, but a varying fold error in the proportions of individual species.
+But a proportion is, after all, just a ratio between two taxa in which the denominator taxon is the entire taxonomic domain under study.
+The difference in behavior from that of ratios between species arises because the efficiency of higher-level taxa, which consist of the aggregate of many species, varies across samples with the changing relative abudances among its constituent species.
 

### 2021-11-10 differential-abundance.Rmd
Make minor text edits

@@
 
-If other error sources remain negligible (or constant across samples), than this dichotomy continues to apply to proportion- and ratio-based density measurements.
+If other error sources remain negligible (or constant across samples), then this dichotomy continues to apply to proportion- and ratio-based density measurements.
 The inferred FCs in proportion-based density measurements will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.

### 2021-11-10 introduction.Rmd
Make minor text edits

@@
 One of the most basic questions we can ask about microbial communities is: How do different microbial taxa vary in abundance—across space, time, and host or environmental conditions?
-Advances in sequencing technology allow us to simultaneously measure the abundances of 100s to 1000s of species using marker-gene and shotgun metagenomic sequencing (jointly, MGS).
-Although standard MGS measurements lose information about total microbial density—and so are typically used to analyze the abundances of taxa relative to each other or their total—new studies are increasingly employing strategies to enable the analysis of cell density or other measures of "absolute abundance."
-These relative and absolute abundances serve as the basis for a _differential-abundance (DA) analysis_, in which the change in abundance of a microbial taxon across samples or conditions is used to learn about the biology of the taxon and its impact on the host and other microbes as well as to detect predictive biomarkers of host and environmental health and disease.
+Marker-gene and shotgun metagenomic sequencing (jointly, MGS) can be used to measure the abundances of thousands of species simultaneously, making it possible to ask this question on a community-wide scale.
+In these _differential-abundance (DA) analyses_, the change in abundance of a microbial taxon across samples or conditions is used to infer ecological dynamics or find microbes that are associated with specific host diseases or environmental conditions.
+Although standard MGS measurements lose information about total microbial density—and so are typically used to analyze the abundances of taxa relative to each other—new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density.
+In its various forms, DA analysis remain one of, if not the most, common forms of analyses applied to MGS data to elucide the inner workings of microbiomes and their relationships to host and environmental health.
 
 <!-- P: Taxonomic bias -->
-Although MGS-based DA analysis has been widely deployed and achieved many notable successes, it faces serious concerns over accuracy and reproducibility due to the inherent technical limitations of MGS measurements.
-In particular, MGS measurements are _taxonomically biased_: Taxa vary dramatically (e.g.
+Yet these DA analysis are built on a fundamentally flawed foundation.
+MGS measurements are _taxonomically biased_: Microbial species vary dramatically (e.g.
 10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—by a given MGS protocol (@mclaren2019cons).
-As a result, the abundance measurements obtained by MGS are inaccurate representations of the actual abundances and also tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo,@mclaren2019cons).
-This bias arises from variation in how taxa respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
-Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads make taxonomic bias a feature of both shotgun and marker-gene measurements.
-The error it causes have been found to in some cases to supersede sizable biological differences (e.g.
-@lozupone2013meta) and has plausibly caused replication failures for prominent findings such as the association of decreased Bacteroides and increased Firmicutes in stool with obesity (@finucane2014atax) and the association of certain taxa in the vagina of pregnant women with preterm birth (@callahan2017repl).
-
-<!-- Revise below here to outline -->
+This bias arises from variation in how species respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
+Although often associated with features specific to marker-gene sequencing—the variation among species in marker copy numbers and in primer-binding and amplification efficiencies—the existence of large variation in DNA extraction efficiencies and in the ability to correctly classify reads make taxonomic bias a universal feature of both marker-gene and shotgun measurements.
+As a result of taxonomic bias, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo,@mclaren2019cons).
+These errors have been found in some cases to supersede sizable biological differences (e.g.
+@lozupone2013meta) and have may have contributed to failed replications of prominent findings in the human microbiome literature, such as the associations of Bacteroides and Firmicutes in stool with obesity (@finucane2014atax) and the associations of certain species in the vagina of pregnant women with preterm birth (@callahan2017repl).
 
 <!-- P: Taxonomic bias and DA -->
-
-The typical approach to countering taxonomic bias in DA analysis is to standardize the measurement protocol used within a given study.
-In broad strokes, the thinking is that the measurements of samples measured by the same protocol will be affected by bias in the same way and so the inferred differences between samples (the focus of DA analysis) will be unaffected.
-For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples (@kevorkian2018esti,@lloyd2020evid).
-However, @mclaren2019cons used theoretical arguments and sequencing measurements of defined bacterial (or 'mock') communities to show that consistent taxonomic bias can lead to variable fold errors in measured proportions (Figure \@ref(fig:basic-problem)).
-These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase).
-Yet @mclaren2019cons also found that the fold error in the ratios among species was constant, making fold changes in ratios robust to bias.
-The implications of these findings for changes in absolute abundance---which remain subject to taxonomic bias in the underlying MGS measurement---and for DA analysis across many species and many samples---as commonly done in microbiome association testing---have yet to be investigated.
+The extent to which taxonomic bias has impacted the DA results in the scientific literature is unknown.
+The typical approach taken to counter taxonomic bias is to standardize the measurement protocol used within a given study, with the (often tacit) assumption being that samples measured by the same protocol will be affected by bias in the same way and so the measured differences between samples will be unaffected.
+For example, if taxonomic bias were to cause the measured proportion of a species to consistently be 10X too high, we would still be able to accurately infer the fold change in its proportion across samples (@kevorkian2018esti,@lloyd2020evid).
+Unfortunately, mathematical arguments and analysis of experiments with artificially constructed ('mock') communities demonstrate that this assumption is not always warranted: 
+Consistent taxonomic bias can lead to variable fold errors in species' proportions (Figure \@ref(fig:basic-problem), @mclaren2019cons).
+These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase) (@mclaren2019cons).
+Yet @mclaren2019cons also found that certain types of DA analysis---those based on fold changes in the ratios among species---where robust to bias.
+The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as typically done in microbiome association testing, have yet to be investigated.
 

### 2021-11-10 real-world-implications.Rmd
Make minor text edits

@@
 One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
-@brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome with the MGS protocol developed for the Vaginal Human Microbiome Project (VaHMP).
-@mclaren2019cons later used this data to estimate the efficiencies of these seven species and develop a method to use these estimates to infer the true compositions of natural samples (see Section \@ref(calibrate-compositions)).
-SI Analysis (MOMS-PI) uses these estimated efficiencies to correct the compositions and explore the implied variation in mean efficiency in the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) @fettweis2019thev.
+To characterize bias in a 16S sequencing protocol developed for the Vaginal Human Microbiome Project (VaHMP), @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome.
+This protocol was later used for the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study (@fettweis2019thev).
+We used the method for measuring and correcting bias developed by @mclaren2019cons to measure bias from the mock communities and analyze its potential impact on DA analyses that use the MOMSPI measurements (SI Analysis MOMS-PI).
 The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
@@
 (DNA mocks were used, so that bias due to DNA extraction is not included, but other major bias sources such as PCR and ribosomal copy-number-variation are included.) 
-SI Analysis X reanalyses this experiment to consider the impact that bias has on different possible DA analyses.
-There was a 13X difference between the most and least efficiently measured commensals, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
-Despite the 13X variation among commensals, the mean efficiency is remarkably stable across communities prior to pathogen colonization.
-Consequently, taxonomic bias creates little systematic error in proportion-based DA analysis on these samples.
-However, the high efficiency of the pathogen typically led to a roughly 6X increase in mean efficiency in post-colonization samples.
-An analysis of how commenals responded to calibration that used uncalibrated proportions without first discarding the pathogen reads would therefore tend to suggest that the commensals had substantially smaller fold changes than they actually did.
+We reanalyzed this experiment to consider the impact that bias has on different possible DA analyses (SI Analysis Fungi).
+As noted by @leopold2020host, there was a 13X difference between the most and least efficiently measured commensals, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
+Yet despite this 13X variation among commensals, the mean efficiency is remarkably stable across communities prior to pathogen colonization.
+Consequently, taxonomic bias creates little systematic error in proportion-based DA analysis of these pre-colonization samples.
+The high efficiency of the pathogen did, however, typically lead to a roughly 6X increase in the mean efficiency of post-colonization samples.
+Hence DA analysis of how commensals responded to the pathogen did not first discard pathogen reads would therefore tend to suggest that the commensals had substantially smaller fold changes than they actually did, unless bias were accounted for.
 

### 2021-11-10 solutions.Rmd
Make minor text edits

@@
 These methods analyze log-fold changes in the ratios among a pair of species or, more generally, some product of exponential power of multiple species, which likewise are bias-invariant.
-
 Besides a greater robustness to taxonomic bias, such ratio-based analyses have other advantages.
@@
 
-Measurement of appropriate control samples along with the primary experimental samples can make it possible to directly measure and remove the effect of bias before or while conducting downstream analysis.
-Under the assumption that taxonomic bias acts consistently across samples, it can be directly estimated for the species in one or more community control samples—samples whose species identities and relative abundances are known either by construction of by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
-The estimated relative efficiencies among the control species can then be used to correct the effect of bias on the relative abundances (ratios) among these species in a set of target (non-control) samples measured with the same protocol (ideally in the same experiment and sequencing run) @mclaren2019cons.
-Calibration can be extended to species not in the controls by making additional assumptions that allow one to impute the missing efficiencies; such imputation might employ phylogenetic relationships, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
+Measurement of community calibration controls along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis.
+Community calibration controls are samples whose species identities and relative abundances are known either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
+MGS measurement of one or more control communities can be used to directly measure the relative efficiencies among the superset of species in the controls (@mclaren2019cons).
+The measured relative efficiencies can then be used to calibrate (remove the effect of bias from) the relative abundances (ratios) of the control species in a set of primary (non-control) experimental samples that were measured with the same protocol (ideally in the same experiment and sequencing run) (@mclaren2019cons).
+Calibration can be extended to species not in the controls using statistical methods to impute (predict) the unobserved efficiencies using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
 
@@
 Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions and densities (Figure \@ref(fig:calibration-example)).
-To demonstrate a more practical application, we used the bias estimated from the full set of mock communities as the basis for calibrating the vaginal community time series measurements from the MOMS-PI study (@fettweis2019thev) that were made using the same 16S sequencing protocol.
-We imputed the efficiencies of other species using the taxonomic relationship to the control species (Methods).
-Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in [implications].
+To demonstrate a practical application, we used the bias estimated from the full set of mock communities as the basis for calibrating the vaginal community time series measurements from the MOMS-PI study (@fettweis2019thev) that were made using the same 16S sequencing protocol.
+We imputed the efficiencies of other species using taxonomic relationships (Methods).
+Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \@ref(implications).
 
@@
 Obtaining control communities that span the taxa of interest is often not practical.
-Moreover, there is no guarantee that bias is consistent across the control and primary samples.
-The results of the previous sections suggest an alternative approach to calibration that overcomes these problems for the specific purpose of species-level DA analysis, requiring only that there are one or more reference species whose abundance can be determined, up to a constant error factor, by background knowledge, experimental design, or independent measurement.
+Moreover, the bias that is measured from controls may differ from that in the primary samples due to differences in cell state or sample preparation.
+These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-density measurement that was described in Section \@ref(differential-abundance).
+This *reference-species calibration* allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from taxonomically biased, relative metagenomics measurements.
 
-Reference-species calibration allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from (taxonomically biased, relative abundance) metagenomics measurements.
-Section X described the reference-species approach to density measurement and showed that such measurements have a constant fold error, provided that the error in the density of the reference species is constant.
-Three categories of reference species of practical import are described in Box X: housekeeping species assumed to have constant abundance, spike-in species added at constant abundance, and species whose abundance has been measured with a targeted method like qPCR.
+Recall that densities estimated by the reference-species approach have a constant fold error, provided that the fold error in the reference species is also constant (Equation \@ref(eq:density-ratio-error)).
+Three categories of reference species are reviewed for this purpose in Appendix \@ref(review-absolute-methods), including why they might be expected to produce constant fold errors: housekeeping species assumed to have constant abundance, spike-in species added at known or constant abundance, and species whose abundance has been measured with a targeted method like species-specific qPCR.
+
 To demonstrate the ability for reference calibration to improve fold changes, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
@@
 
-A variety of methods for measuring total community density have been paired with both amplicon and shotgun sequencing data for the community-normalization method to measure absolute densities for absolute DA analysis, with little consensus on the pros and cons of the various methods.
-Although these methods differ in many aspects, a notable aspect is the sources of taxonomic bias inherent in each method.
-Just as species differ in how efficiently they are measured by MGS, so can they differ in how efficiently they contribute to a total density measurement.
+A variety of methods for measuring total community density have been paired with either amplicon or shotgun sequencing data for the purpose of conducting absolute DA analysis using the total-density approach described in Section \@ref(differential-abundance).
+Each of these methods is subject to its own taxonomic bias: Cells from different species are likely to contribute more or less efficiently to the total abundance measurement.
 Consider the popular method of assessing bacterial abundance with total-16S qPCR measurement.
@@
 Methods like using flow cytometry that directly measures cell density lacks these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
-By extending the analysis of Section X to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(appendix-total-density)).
+By extending the analysis of Section \@ref(differential-abundance) to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(appendix-total-density)).
 This finding suggests that methods of total-density measurement that are more accurate as measures of total cell density may actually perform worse than less accurate methods for the purposes of absolute DA inference.
@@
 We illustrate with a hypothetical example of two vaginal communities of equal density, but which are dominated either by _Lactobacillus iners_ or _Gardnerella vaginalis_.
-Based on our earlier analysis in @mclaren2019cons, we estimate that cells of _L. iners_ 20X more 16S copies than _G. vaginalis_ from DNA extraction (CHECK NUMBERS).
-Ignoring other sources of bias, we expect relative efficiencies of _L. iners_:_G. vaginalis_ of 20 for both qPCR and 16S sequencing measurements.
+Based on our earlier analysis in @mclaren2019cons, we estimate that _L. iners_ yields 10X more 16S copies per cell than _G. vaginalis_ in extracted DNA.
+Ignoring other sources of bias, we expect relative efficiencies of _L. iners_:_G. vaginalis_ of 10 for both qPCR and 16S sequencing measurements.
 Consequently, the _G. vaginalis_ community will yield a substantially lower qPCR measurement than the _L. iners_ community and we would incorrectly infer a decrease in total abundance.
-At the same time, this decrease in the mean efficiency of the sequencing measurement causes artificially high fold changes in the proportions of all species (ref previous section).
+At the same time, this decrease in the mean efficiency of the sequencing measurement would cause artificially high fold changes in the proportions of all species (Section \@ref(differential-abundance)).
 When using qPCR and 16S sequencing measurements for community-based density estimation, these two opposing errors cancel, yielding accurate fold changes in species densities.
@@
 These pairings even make it possible to account for error caused by variation among samples in the fraction of unclassified reads, which can form a major fraction of amplicon and shotgun data.
-Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (REF Appendix).
+Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (Appendix \@ref(appendix-total-density)).
 

### 2021-11-10 real-world-implications.Rmd
Add and rearrange figures
- Add 1 main and 1 supplemental MOMSPI figure
- Move supplemental MOMSPI bias-sensitivity-demo figures to a new
  supplemental figures appendix

@@
 Our mock-community analysis indicate _G. vaginalis_ has the lowest efficiency of the seven species, roughly 20X lower than _L. crispatus_ and 30X lower than _L. iners_ (the most efficiently measured).
-As a result, the mean efficiency varies substantially across samples, with samples dominated by a _Lactobacillus_ species typically having an efficiency that is 3-20X greater than samples dominated by _Gardnerella_ (SI figure).
-Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure).
-Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure).
+As a result, the mean efficiency varies substantially across samples, with samples dominated by a _Lactobacillus_ species typically having an efficiency that is 3-20X greater than samples dominated by _Gardnerella_ (Figure \@ref(fig:momspi-mean-efficiency-dist)).
+Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
+Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure TODO).
 Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
+
+<!-- begin figure -->
+
+```{r momspi-mean-efficiency-dist, fig.cap = '(ref:cap-momspi-mean-efficiency-dist)'}
+svg_path <- fs::path(
+  "notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files", 
+  "figure-html5/momspi-mean-efficiency-dist-1.svg"
+)
+if (knitr::is_html_output()) {
+  fig_path <- svg_path
+} else if (knitr::is_latex_output()) {
+  fig_path <- svg_to_pdf(svg_path)
+}
+knitr::include_graphics(fig_path)
+rm(svg_path, fig_path)
+```
+
+(ref:cap-momspi-mean-efficiency-dist) **The mean efficiency in vaginal samples from the MOMS-PI study varies with the most abundant species.**
+
+<!-- end figure -->
 

### 2021-11-10 solutions.Rmd
Add and rearrange figures
- Add 1 main and 1 supplemental MOMSPI figure
- Move supplemental MOMSPI bias-sensitivity-demo figures to a new
  supplemental figures appendix

@@
 Finally, the DA analysis is re-run on each set of calibrated measurements and the distribution of results is compared graphically or with summary statistics.
-A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool 'corncob' are shown in \@ref(fig:sensitivity-example).
+A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool 'corncob' are shown in SI Figure \@ref(fig:sensitivity-example).
 The simulated efficiency vectors may be generated to different hypotheses about bias in the given system, such as its overall magnitude and correlation among phylogenetically-related species.
@@
 To the extent that taxonomic bias truly is consistent within studies, meta-analyses that directly account for it can have greater statistical power to extract the truly consistent and inconsistent microbial patterns across studies.
-
-## Supplemental figures
-
-<!-- begin figure -->
-
-```{r sensitivity-example, fig.cap = '(ref:cap-sensitivity-example)'}
-svg_path <- here::here(
-  "notebook/_posts/2021-10-18-evaluate-robustness-example/",
-  "evaluate-robustness-example_files/figure-html5/summary_plot-1.svg"
-)
-if (knitr::is_html_output()) {
-  fig_path <- svg_path
-} else if (knitr::is_latex_output()) {
-  fig_path <- svg_to_pdf(svg_path)
-}
-knitr::include_graphics(fig_path)
-rm(svg_path, fig_path)
-```
-
-(ref:cap-sensitivity-example) **A bias-sensivity analysis can be performed to examine how sensitive the results of a DA analysis are to assumptions about taxonomic bias in community measurements.** The figure shows the results of a bias-sensitivity analysis used to study the effect of bias on the association of _Gardnerella vaginalis_ and preterm birth that was investigated by @callahan2017repl.
-100 random efficiency vectors were drawn at 6 different bias strengths (quantified by the variance in log efficiency, $\sigma_{e}^{2}$).
-Each efficiency vector was used to calibrate the MGS profiles and perform a DA association test of _G.
-vaginalis_ versus the host's preterm birth outcome; regression coefficients $\hat \beta$ indicate the increase of average logit proportion of  _G.
-vaginalis_ in women who experienced preterm birth.
-
-<!-- end figure -->

### 2021-11-10 supplemental-figures.Rmd
Add and rearrange figures
- Add 1 main and 1 supplemental MOMSPI figure
- Move supplemental MOMSPI bias-sensitivity-demo figures to a new
  supplemental figures appendix

# Supplemental figures

<!-- Implications: MOMSPI -->

```{r momspi-mean-efficiency-fcs, fig.cap = '(ref:cap-momspi-mean-efficiency-fcs)'}
svg_path <- fs::path(
  "notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files", 
  "figure-html5/momspi-mean-efficiency-fcs-1.svg"
)
if (knitr::is_html_output()) {
  fig_path <- svg_path
} else if (knitr::is_latex_output()) {
  fig_path <- svg_to_pdf(svg_path)
}
knitr::include_graphics(fig_path)
rm(svg_path, fig_path)
```

(ref:cap-momspi-mean-efficiency-fcs) **Fold changes in the mean efficiency within and between women in the MOMS-PI study.**

<br><br>

<!-- Solutions -->

```{r sensitivity-example, fig.cap = '(ref:cap-sensitivity-example)'}
svg_path <- fs::path(
  "notebook/_posts/2021-10-18-evaluate-robustness-example/",
  "evaluate-robustness-example_files/figure-html5/summary_plot-1.svg"
)
if (knitr::is_html_output()) {
  fig_path <- svg_path
} else if (knitr::is_latex_output()) {
  fig_path <- svg_to_pdf(svg_path)
}
knitr::include_graphics(fig_path)
rm(svg_path, fig_path)
```

(ref:cap-sensitivity-example) **A bias-sensivity analysis can be performed to examine how sensitive the results of a DA analysis are to assumptions about taxonomic bias in community measurements.** The figure shows the results of a bias-sensitivity analysis used to study the effect of bias on the association of _Gardnerella vaginalis_ and preterm birth that was investigated by @callahan2017repl. 100 random efficiency vectors were drawn at 6 different bias strengths (quantified by the variance in log efficiency, $\sigma_{e}^{2}$). Each efficiency vector was used to calibrate the MGS profiles and perform a DA association test of _G. vaginalis_ versus the host's preterm birth outcome; regression coefficients $\hat \beta$ indicate the increase of average logit proportion of  _G. vaginalis_ in women who experienced preterm birth.

### 2021-11-10 notebook/_posts/2021-11-09-explore-gut-data/explore-gut-data.Rmd
Add preliminary analysis of gut microbiome data
Quick analysis to look at variation in Bacteroidetes and Firmicutes in a
gut microbiome dataset

---
title: "Explore gut datasets"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - R
date: 2021-11-10
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: svg
---

```{r}
library(here)
library(tidyverse)
library(speedyseq)
library(metacal)

library(cowplot)
theme_set(theme_cowplot())
```

```{r}
# curatedMetagenomicData("HMP")
tse <- curatedMetagenomicData::curatedMetagenomicData(
  "HMP_2019_ibdmdb.relative_abundance", 
  dryrun = FALSE
)
stopifnot(identical(length(tse), 1L))
tse <- tse[[1]]
ibd_ps <- mia::makePhyloseqFromTreeSummarizedExperiment(
  tse,
  abund_values = "relative_abundance"
) %>%
  rename_with_sample_data(janitor::make_clean_names) %>%
  rename_with_tax_table(janitor::make_clean_names)
sam <- ibd_ps %>% sample_data %>% as_tibble
```

```{r}
```

## metadata

```{r}
sam %>% glimpse
sam %>% dplyr::count(body_site)
sam %>% dplyr::count(dna_extraction_kit)
sam %>% dplyr::count(location)
sam %>% dplyr::count(subject_id) %>% dplyr::count(n)
sam %>% dplyr::count(study_condition, disease, disease_subtype)
```

Most subjects have many samples.

## what is dominance like?

```{r}
ps <- ibd_ps %>% transform_sample_counts(close_elts)
x <- ps %>% as_tibble
```

```{r}
x %>%
  group_by(.sample) %>%
  slice_max(.abundance, n = 1) %>%
  ggplot(aes(.abundance)) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
x %>%
  group_by(.sample) %>%
  slice_max(.abundance, n = 1) %>%
  pull(.abundance) %>%
  summary
```

Note that even in the gut, the most abundant species may often be dominant.

## B:F ratio

```{r}
ps.phy <- ps %>% tax_glom('phylum')
x.phy <- ps.phy %>% as_tibble
```

```{r}
x.phy %>%
  group_by(.sample) %>%
  slice_max(.abundance, n = 1) %>%
  pull(.abundance) %>%
  summary
```

```{r}
x.phy %>%
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  ggplot(aes(.abundance + 1e-4, fill = disease_subtype)) +
  facet_wrap(~phylum, ncol = 1) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
x.phy %>%
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  mutate(
    across(.abundance, ~ . + 1e-4),
    ratio_f = .abundance / .abundance[phylum == 'Firmicutes']
  ) %>%
  filter(phylum == 'Bacteroidetes') %>%
  ggplot(aes(ratio_f, fill = disease_subtype)) +
  scale_x_log10(breaks = 10^(-3:4)) +
  facet_wrap(~disease, ncol = 1) +
  geom_histogram() +
  theme_minimal_vgrid()
```

There is a lot of variation, but most if it is generally not associated with disease.
This suggests that in this case, the main impact of variation in mean efficinecy from B:F ratio may be to create noise.

(note, I am ignoring structure in the data such as replicate sampling)

how correlated are species within a phylum?
what is the typical within-phylum diversity?

Question: is the HMP2 gut protocol the same as HMP1?

What is the inverse simpson diversity within phyla?

```{r}
div <- x %>% 
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  with_groups(c(.sample, phylum), mutate,
    across(.abundance, close_elts)
  ) %>%
  with_groups(c(.sample, phylum), summarize,
    diversity = 1/sum(.abundance^2)
  )
```

```{r}
div %>%
  ggplot(aes(diversity)) +
  scale_x_log10() +
  facet_wrap(~phylum, ncol = 1) +
  geom_histogram() +
  theme_minimal_vgrid()
```

```{r}
div %>%
  with_groups(phylum, summarize,
    qs = quantile(diversity, c(0.25, 0.5, 0.75), na.rm = TRUE), 
    prob = c(0.25, 0.5, 0.75)
  )
```

### 2021-11-12 notebook/_posts/2021-10-09-momspi-patient-exploration-and-summary/MOMS_patient_and_summary.Rmd
Import MOMS-PI fold-change analysis
This commit imports an earlier analysis by Jacob Nearing and Michael
McLaren of the impact of calibration on the fold changes in species
proportions in the MOMSPI profiles.

---
title: "MOMS Patient Exploration and Overall Summary"
author: "Jacob T. Nearing"
date: "07/10/2021"
output: 
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "chaos"
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.dim=c(6,6))
```

# Set up

Load in required libaries for the analysis
```{r, message=F, echo=F}

#work in the github repo
setwd("~/GitHub_Repos/MOMSPI/momspi/")
library(here)

library(tidyverse)

#Use cowplot theme for ggplots
library(cowplot)
theme_set(theme_cowplot())

#used for mult panel figures
library(patchwork)

#used to calibrate microbe abundances using mock communities and multiplicative bias.
library(metacal)

#improved implementation of some phyloseq functions
library(speedyseq)

library(kableExtra)
```

# Load in mock data

This code section is copied from MM's original analysis
```{r}
#Load in otu data
otu <- here("output", "stirrups-profiles", "abundance-matrix.csv.bz2") %>%
  read_csv(
    col_types = cols(.default = col_double(), sample_name = col_character())
  ) %>%
  otu_table(taxa_are_rows = FALSE)

#load in sample data
sam <- here("output", "stirrups-profiles", "sample-data.csv.bz2") %>%
  read_csv(col_types = "ccccccic") %>%
  mutate(across(host_visit_number, factor, ordered = TRUE)) %>%
  sample_data

#load in taxa profiles
tax <- here("output", "stirrups-profiles", "taxonomy.csv.bz2") %>%
  read_csv(col_types = cols(.default = col_character())) %>%
  tax_table %>%
  mutate_tax_table(
    species = case_when(!is.na(genus) ~ .otu)
  )

ps <- phyloseq(otu, sam, tax)

taxa_names(ps) <- taxa_names(ps) %>% 
  str_replace("(?<=Lactobacillus_crispatus)_cluster", "")
taxa_names(ps) %>% str_subset("crispatus")
```

# Calibrate data
```{r}
#grab data files from metacal package 
dr <- system.file("extdata", package = "metacal")

#table contains the expected abundances of the mock communties from
# brooks et al., 2015
actual <- file.path(dr, "brooks2015-actual.csv") %>%
  read.csv(row.names = "Sample") %>%
  as("matrix")

#table contains the observed abundances from 16S sequencing above mock communties
# data from brooks et al., 2015
observed <- file.path(dr, "brooks2015-observed.csv") %>%
  read.csv(row.names = "Sample") %>%
  subset(select = - Other) %>%
  as("matrix")

# Estimate bias with bootstrapping for error estimation
mc_fit <- estimate_bias(observed, actual, margin = 1, boot = TRUE)
summary(mc_fit)

#to keep our workspace clean we will remove the tables we don;t need.
rm(actual, observed, dr)

control_species <- mc_fit %>% coef %>% names
stopifnot(all(control_species %in% taxa_names(ps)))
```

We'll just use the point estimate, ignoring the uncertainty, since the standard errors are so small.

We don't have a way to directly estimate the efficiencies of the other OTUs, so we'll impute them as follows.

1. Set the efficiencies of the 7 control species to the directly estimated values
2. Compute efficiencies for the 6 control genera to the efficiency of the control species (if only one) or the geometric mean efficiency (if more than one; here this is just _Lactobacillus_)
3. Use these genera-level efficiencies for the efficiencies of OTUs that are not control species but are in the same genus as one
4. For other OTUs, use the geometric mean of the genera efficiencies; this is similar to using the mean of the 7 control species but gives _Lactobacillus_ equal weight to the other genera.

```{r}
#bias of all 
bias_species <- coef(mc_fit) %>% 
  enframe("species", "efficiency") %>%
  print

bias_genus <- bias_species %>%
  mutate(genus = str_extract(species, "^[^_]+"), .before = 1) %>%
  with_groups(genus, summarize, across(efficiency, gm_mean)) %>%
  print

# Match on genus or species, depending on which is available; then set others
# to average genus efficiency
bias_all <- tax_table(ps) %>% as_tibble %>%
  left_join(bias_species, by = "species") %>%
  left_join(bias_genus, by = "genus") %>%
  mutate(
    efficiency = case_when(
      !is.na(efficiency.x) ~ efficiency.x,
      !is.na(efficiency.y) ~ efficiency.y,
      TRUE ~ gm_mean(bias_genus$efficiency)
    )
  ) %>%
  select(-efficiency.x, -efficiency.y) %>%
  # standardize to L. iners, the most efficiently measured
  mutate(
    across(efficiency, ~ . / max(.))
  ) %>%
  glimpse

bias_all_vec <- bias_all %>% select(.otu, efficiency) %>% deframe 
```

```{r}
ps.obs <- ps %>% 
  # prune_taxa(control_species, .) %>%
  transform_sample_counts(close_elts)

# note, calibrate() automatically subsets to just the focal taxa if we haven't
# already
ps.cal <- ps.obs %>% calibrate(bias_all_vec)

mean_eff <- perturb(ps.cal, bias_all_vec, norm = "none")%>% 
  sample_sums %>%
  enframe('.sample', 'mean_efficiency') %>%
  left_join(sam %>% as_tibble, by = '.sample')
```

# Calculate fold-change between subsequent visits

```{r}

#takes in as input 2 phyloseq objects (obs and calibrated phyloseq objects)
#the subject of interest (i.e. patient ID)
#whether you want to prune to only the taxa in the mock community (i.e. control species)
#a vector that contains the names of all the cntrl species (required when pruning)
get_taxa_fold_change <- function(cal, obs, subject, prune=F, cntrl_spec=control_species){

  #if we want to only keep taxa in the mocks prune set to T
  if(prune){

    #borrow Mike's code from above
    temp <- list(Calibrated = cal, Observed = obs) %>%
    map(filter_sample_data, host_subject_id == subject) %>%
    #remove taxa that are not control taxa
    map(~prune_taxa(cntrl_spec, .)) %>%
    #change to relative abundances are they not already in proportions though?
    #ohhh we need to recalculate the relative abundances as we removed some taxa!

    map(transform_sample_counts, fun = close_elts) %>%
    map(as_tibble)

  }else{
    temp <- list(Calibrated = cal, Observed = obs) %>%
    map(filter_sample_data, host_subject_id == subject) %>%
    map(as_tibble)
  }

  #set the type of overvation
  temp[["Calibrated"]]$Type <- "Calibrated"
  temp[["Observed"]]$Type <- "Observed"

  #list that will contain the returned info.
  ret_list_cal <- list()

  taxa_names <- unique(temp$Calibrated$.otu)
  #go through list and create a new dataframe of each taxa arrange by the their visit number
  for(i in taxa_names){
    ret_list_cal[[i]] <- filter(temp$Calibrated, .otu==i) %>% arrange(host_visit_number)

  }
  #okay for each of these taxa we now need to calculate the fold change in abundance between row i and row i + 1
  # i.e. each subsequent visit
  for(i in taxa_names){
    #loop through each taxa abundance for that sample
    for(j in 1:nrow(ret_list_cal[[i]])){
      #initalize the column as -1 (fold change value should never be negative)
      #if its the first sample we will just mark at is "START"
      if(j==1){
        ret_list_cal[[i]][j,"fold_change_from_last"] <- NA
        ret_list_cal[[i]][j,"change"] <- "START"
        next()
      }
      #get the previous sample index
      last_index <- j-1
      #divide current index abundance by last abundance to get FC change
      fc_change <- ret_list_cal[[i]][j,".abundance"]/ret_list_cal[[i]][last_index,".abundance"]
      ret_list_cal[[i]][j,"fold_change_from_last"] <- fc_change

      #sometimes NaN is retunred as abundance when all abundances are 0 during reclosing to RA
      #this is only an issue when pruning to control taxa.
      if(is.na(ret_list_cal[[i]][j, ".abundance"]) | is.na(ret_list_cal[[i]][last_index, ".abundance"])){
        ret_list_cal[[i]][j, "change"] <- "FAIL"
        next()
      }
      #set what the change was (went up down or stayed same between visits)
      if(ret_list_cal[[i]][j,".abundance"] > ret_list_cal[[i]][last_index, ".abundance"]){
        ret_list_cal[[i]][j,"change"] <- "UP"
      }
      if(ret_list_cal[[i]][j,".abundance"] < ret_list_cal[[i]][last_index, ".abundance"]){
        ret_list_cal[[i]][j,"change"] <- "DOWN"
      }
      if(ret_list_cal[[i]][j,".abundance"] == ret_list_cal[[i]][last_index, ".abundance"]){
        ret_list_cal[[i]][j,"change"] <- "SAME"
      }

    }

  }

  #same analysis as above but for observed values.
  ret_list_obs <- list()

  taxa_names <- unique(temp$Observed$.otu)
  for(i in taxa_names){
    ret_list_obs[[i]] <- filter(temp$Observed, .otu==i) %>% arrange(host_visit_number)

  }
  #okay for each of these taxa we now need to calculate the fold change in abundance between row i and row i + 1
  for(i in taxa_names){
    for(j in 1:nrow(ret_list_obs[[i]])){
      #initalize the column as -1 (fold change value should never be negative)
      if(j==1){
        ret_list_obs[[i]][j,"fold_change_from_last"] <- NA
        ret_list_obs[[i]][j,"change"] <- "START"
        next()
      }
      last_index <- j-1
      #divide current index abundance by last abundance to get FC change
      fc_change <- ret_list_obs[[i]][j,".abundance"]/ret_list_obs[[i]][last_index,".abundance"]
      ret_list_obs[[i]][j,"fold_change_from_last"] <- fc_change

      #sometimes NaN is retunred as abundance when all abundances are 0 during reclosing to RA
      if(is.na(ret_list_obs[[i]][j, ".abundance"]) | is.na(ret_list_obs[[i]][last_index, ".abundance"])){
        ret_list_obs[[i]][j, "change"] <- "FAIL"
        next()
      }

      if(ret_list_obs[[i]][j,".abundance"] > ret_list_obs[[i]][last_index, ".abundance"]){
        ret_list_obs[[i]][j,"change"] <- "UP"
      }
      if(ret_list_obs[[i]][j,".abundance"] < ret_list_obs[[i]][last_index, ".abundance"]){
        ret_list_obs[[i]][j,"change"] <- "DOWN"
      }
      if(ret_list_obs[[i]][j,".abundance"] == ret_list_obs[[i]][last_index, ".abundance"]){
        ret_list_obs[[i]][j,"change"] <- "SAME"
      }
    }

  }

  return(list(ret_list_cal,ret_list_obs))

}

### Function that goes through the list that is returned from get_taxa_fold_change and finds taxa/samples that have a sign change between obs and calibrated
## for now we will return a DF with two rows for each occurance of a fold change difference between calibrated and observed.
find_sign_change <- function(taxa_list){

  taxa_names <- names(taxa_list[[1]])

  found_hit=F
  ret_df <- NULL
  #go through each taxa
  for(taxa in taxa_names){
    #1 is calibrated list
    for(j in 1:nrow(taxa_list[[1]][[taxa]])){
      if(taxa_list[[1]][[taxa]][j,"change"] != taxa_list[[2]][[taxa]][j,"change"]){
        message("found diff")
        if(found_hit==F){
          ret_df <- rbind(taxa_list[[1]][[taxa]][j,], taxa_list[[2]][[taxa]][j,])

          val1 <- ret_df[1,"fold_change_from_last"]/ret_df[2,"fold_change_from_last"]
          val2 <- ret_df[2,"fold_change_from_last"]/ret_df[1,"fold_change_from_last"]
          vals <- c(val1,val2)

          fc_diff <- vals[which.max(vals)]
          message(fc_diff)
          ret_df$FC_Diff <- fc_diff

          found_hit=T
        }else{
          temp_df <- rbind(taxa_list[[1]][[taxa]][j,], taxa_list[[2]][[taxa]][j,])

          val1 <- temp_df[1,"fold_change_from_last"]/temp_df[2,"fold_change_from_last"]
          val2 <- temp_df[2,"fold_change_from_last"]/temp_df[1,"fold_change_from_last"]
          vals <- c(val1,val2)

          fc_diff <- vals[which.max(vals)]
          temp_df$FC_Diff <- fc_diff

          ret_df <- rbind(ret_df, temp_df)
        }

      }

    }
  }
  return(ret_df)

}

Get_total_fc <- function(taxa_list){

  taxa_names <- names(taxa_list[[1]])
  found_hit=F
  ret_df <- NULL
  #go through each taxa
  for(taxa in taxa_names){
    #1 is calibrated list
    for(j in 1:nrow(taxa_list[[1]][[taxa]])){
        if(found_hit==F){
          ret_df <- rbind(taxa_list[[1]][[taxa]][j,], taxa_list[[2]][[taxa]][j,])
          found_hit=T
        }else{
          temp_df <- rbind(taxa_list[[1]][[taxa]][j,], taxa_list[[2]][[taxa]][j,])
          ret_df <- rbind(ret_df, temp_df)
        }

      }

    }

  return(ret_df)
}

```

## Generate FC change values
For the purposes of this analysis we will only focus on the values that also include the imputed bias for non-control taxa

This takes awhile to run so we will save the results as an RDS file. 
```{r, eval=F}
#get a list of subjects in the dataset
Subject_names <- unique(sample_data(ps.cal)$host_subject_id)

#remove subjects with only 1 samples
one_sample_subjects <- names(which(table(sample_data(ps.cal)$host_subject_id) ==1 ))
Subject_names <- Subject_names[! Subject_names %in% one_sample_subjects]

#list to contain FC_changes
nonpruned_fc_changes <- list()
#list to contain samples/taxa that there is FC dif between observed and calibrated
nonpruned_sign_dif <- list()

#Go through each subject and calulate fold changes for every taxa
for(i in Subject_names){
  message(i)
  nonpruned_fc_changes[[i]] <- get_taxa_fold_change(cal=ps.cal, obs = ps.obs, subject=i,
                                                 prune=F, cntrl_spec = control_species)
  nonpruned_sign_dif[[i]] <- find_sign_change(nonpruned_fc_changes[[i]])
}

#bind all sign_difs for each patient into a single dataframe.
all_together_nonprune_dif <- do.call(rbind, nonpruned_sign_dif)
#save it
saveRDS(all_together_nonprune_dif, "~/GitHub_Repos/MOMSPI/momspi/data/FC_data/FC_changes_dif_nonprune.RDS")

#save the fc_changes
saveRDS(nonpruned_fc_changes, "~/GitHub_Repos/MOMSPI/momspi/data/FC_data/FC_changes_all_nonprune.RDS")

### To make this simplier lets only look at control species

all_together_control <- all_together[which(all_together$control==T),]
which.max(all_together_control$FC_Diff)
```

## Explore FC differences of control taxa

While we imputed the bias for all taxon lets just focus on the taxa that were included in the mock communnities.

```{r}
all_together_dif <- readRDS("~/GitHub_Repos/MOMSPI/momspi/data/FC_data/FC_changes_dif_nonprune.RDS")
all_together_dif$FC_Diff <- unlist(all_together_dif$FC_Diff)

#only keep taxa that were included in the mock
all_together_dif_control <- all_together_dif[which(all_together_dif$control==T),]

#lets see what the maximum FC difference is
max_diff_index <- which.max(all_together_dif_control$FC_Diff)

#list of metadata to look at
metadata_of_int <- c(".sample", "host_subject_id", ".otu", "host_visit_number", "FC_Diff")

all_together_dif_control[max_diff_index,metadata_of_int]
#Patient EP229454 has the largest fold change difference between the different individuals

#fairly large FC change for this individual as well mostly in Prevotella bivia
all_together_dif_control[which(all_together_dif_control$host_subject_id=="EP972345"), metadata_of_int]

#lets do the same as above but filter on FC_diff being larger than 5.
all_together_dif_control_large <- all_together_dif_control[which(all_together_dif_control$FC_Diff>=5),]
tail(sort(table(all_together_dif_control_large$host_subject_id)))

```

Patient EP229454 has the largest FC difference between calibrated and observed individuals and it is for the taxon Atopobium_vaginae.

Patient EP972345 is also a good example of a patient with a large FC difference between the calibrated and observed individuals.

Patient EP587032 has multiple instances of fairly large FC differences between the calibrated and observed individuals

We will inspect each of these patients individually

```{r}

x1 <- list(Calibrated = ps.cal, Observed = ps.obs) %>%
  map(~prune_taxa(control_species, .)) %>%
  map_dfr(as_tibble, .id = "type") %>%
  mutate(otu_abbrev = .otu %>% str_replace("_", " ") %>% abbreviate)
```

# Patient EP229454

```{r}
EP229454_lines <-  x1 %>%
  filter(host_subject_id == "EP229454") %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_grid(otu_abbrev~., scales = 'fixed') +
  geom_line(aes(group = type)) +
  ylab("Log10 relative abundance") +
  xlab("Visit Number") +
  labs(color="Observation Type")

EP229454_me <- mean_eff %>%
  filter(host_subject_id == "EP154266") %>%
  ggplot(aes(host_visit_number, mean_efficiency)) +
  #scale_y_log10() +
  scale_y_log10(limits = c(1e-2, 1e1)) +
  geom_line(aes(group = 1)) +
  ylab("Log10 Mean Efficiency") +
  xlab("Visit Number")

EP229454_lines / EP229454_me + plot_layout(heights = c(1, 0.1))
```

We can see an example of a sign change between the two Observation types from visits 1 - 2 in Atopobium vaginae

# Patient EP972345
```{r}
EP972345_lines <- x1 %>%
  filter(host_subject_id == "EP972345") %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_grid(otu_abbrev~., scales = 'fixed') +
  geom_line(aes(group = type)) +
  ylab("Log10 relative abundance") +
  xlab("Visit Number") +
  labs(color="Observation Types")

EP972345_me <- mean_eff %>%
  filter(host_subject_id == "EP972345") %>%
  ggplot(aes(host_visit_number, mean_efficiency)) +
  #scale_y_log10() +
  scale_y_log10(limits = c(1e-2, 1e1)) +
  geom_line(aes(group = 1)) +
  ylab("Log10 Mean Efficiency") +
  xlab("Visit Number")

EP972345_lines / EP972345_me + plot_layout(heights = c(1, 0.1))
```

Here we see that there is a sign change between the observation types from visits 2-4. This is accompanied by a decrease in mean effiency due to an increase in the abundance of Gardnerella vaginalis

# Patient EP587032
```{r}
EP587032_lines <- x1 %>%
  filter(host_subject_id == "EP587032") %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_grid(otu_abbrev~., scales = 'fixed') +
  geom_line(aes(group = type)) +
  ylab("Log10 relative abundance") +
  xlab("Visit Number") +
  labs(color="Observation Types")

EP587032_me <- mean_eff %>%
  filter(host_subject_id == "EP587032") %>%
  ggplot(aes(host_visit_number, mean_efficiency)) +
  #scale_y_log10() +
  scale_y_log10(limits = c(1e-2, 1e1)) +
  geom_line(aes(group = 1)) +
  ylab("Log10 Mean Efficiency") +
  xlab("Visit Number")

EP587032_lines / EP587032_me + plot_layout(heights = c(1, 0.1))
```

This patient is a very interesting example as there is sign changes in almost all of the taxa being measured. This is also accompanied by a decrease in the mean effiency due to an increase in Streptococcus aglactiae.

# Overall fold changes

```{r, eval=F}
#get a list of subjects in the dataset
Subject_names <- unique(sample_data(ps.cal)$host_subject_id)

#remove subjects with only 1 samples
one_sample_subjects <- names(which(table(sample_data(ps.cal)$host_subject_id) ==1 ))
Subject_names <- Subject_names[! Subject_names %in% one_sample_subjects]

nonpruned_fc_changes <- readRDS("~/GitHub_Repos/MOMSPI/momspi/data/FC_data/FC_changes_all_nonprune.RDS")
All_fc_change <- list()
for(i in Subject_names){
  message(i)
  All_fc_change[[i]] <- Get_total_fc(nonpruned_fc_changes[[i]])

}

View(All_fc_change)
ALL_fc_change_df <- do.call(rbind, All_fc_change)
saveRDS(ALL_fc_change_df, "~/GitHub_Repos/MOMSPI/momspi/data/FC_data/FC_changes_all_nonprune_df.RDS")
```

## All Fold Change Comparisons

```{r}
All_fc_change_df <- readRDS("~/GitHub_Repos/MOMSPI/momspi/data/FC_data/FC_changes_all_nonprune_df.RDS")

### First lets contrict to just control taxa
All_fc_change_df <- All_fc_change_df %>% filter(.otu %in% control_species)
### remove start samples
All_fc_change_df <- All_fc_change_df %>% filter(change != "START")

#remove when FC was the same usually this is when FC is NaN
#this is not a completely "fair" way to filter the data most likely...
All_fc_change_df <- All_fc_change_df %>% filter(change != "SAME")

split_dfs <- split(All_fc_change_df, All_fc_change_df$Type)

identical(split_dfs[[1]]$.otu, split_dfs[[2]]$.otu)
identical(split_dfs[[1]]$.sample, split_dfs[[2]]$.sample)

split_dfs[[1]]$obs_val <- split_dfs[[2]]$fold_change_from_last
split_dfs[[1]]$obs_change <- split_dfs[[2]]$change

test_plot <- ggplot(split_dfs[[1]], aes(y=log(fold_change_from_last), x=log(obs_val), color=.otu)) + geom_point() +
  xlab("log(Observed Fold Changes)") + ylab("log(Calibrated Fold Changes)") + geom_abline(intercept=0, slope=1)
test_plot
```

## Lets examine these results looking at specific taxa {.tabset}
```{r, fig.height=20, fig.width=5, out.height="200%", out.width="150%"}

taxa_split_dfs <- split(split_dfs[[1]], split_dfs[[1]]$.otu)

fc_change_perc <- c()
for(i in control_species){

  agg_occ <- table(taxa_split_dfs[[i]]$change, taxa_split_dfs[[i]]$obs_change)
  total_dis <- sum(agg_occ[1,2], agg_occ[2,1])
  fc_change_perc[i] <- total_dis/dim(taxa_split_dfs[[i]])[1]*100
}

fc_change_perc

agreeance_plots <- list()
for(i in control_species){

  agreeance_plots[[i]] <- ggplot(taxa_split_dfs[[i]], aes(y=log(fold_change_from_last), x=log(obs_val), color=.otu)) + geom_point() +
  xlab("log(Observed Fold Changes)") + ylab("log(Calibrated Fold Changes)") + geom_abline(intercept=0, slope=1)
}
```

### Atopobium vaginae

```{r}
agreeance_plots[1]
```

### Gardnerella vaginalis

```{r}
agreeance_plots[2]
```

### Lactobacillus crispatus

```{r}
agreeance_plots[3]
```

### Lactobacillus iners

```{r}
agreeance_plots[4]
```

### Prevotella bivia

```{r}
agreeance_plots[5]
```

### Sneathia amnii

```{r}
agreeance_plots[6]
```

### Steptococcus agalactiae

```{r}
agreeance_plots[7]
```

### Percent occurences of fold change differences

```{r}
#table showing total occurences of when things don't agree
agreeance_occurances <- table(split_dfs[[1]]$change, split_dfs[[1]]$obs_change)

total_disagreeance <- sum(agreeance_occurances[1,2],agreeance_occurances[2,1])
total_disagreeance/dim(split_dfs[[1]])[1]*100

### in total it looks like it occurs roughly 5% of the time regardless of taxa

## Note this is after filtering out start samples and samples that didn't have a fold change from their last visit.
```

### 2021-11-12 notebook/_posts/2021-10-09-momspi-patient-exploration-and-summary/momspi-patient-exploration-and-summary.Rmd
Import MOMS-PI fold-change analysis
This commit imports an earlier analysis by Jacob Nearing and Michael
McLaren of the impact of calibration on the fold changes in species
proportions in the MOMSPI profiles.

---
title: "MOMS-PI patient exploration and analysis of fold changes"
description: 
author:
  - name: Jacob T. Nearing
  - name: Michael R. McLaren
date: 2021-10-09
categories:
  - ref:brooks2015thet
  - ref:fettweis2019thev
# bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE
)
```

This is a link post to an analysis of the MOMS-PI vaginal microbiome profiles.
The analysis computes, for each species, the fold changes between subsequent visits within individual women.
It then plots species trajectories for several women, with and without calibration, and visualizes and computes some summary statistics for the observed versus actual fold changes for the species in the mock community.

[Link to analysis](MOMS_patient_and_summary.html)

### 2021-11-16 appendix.Rmd
Draft model and experimental-methods appendices
- Appendix with details of the model and additional theoretical results
  related to abundance measurement and fold changes
- Appendix reviewing different experimental methods for absolute
  abundance in the light of our results

@@
 
-# Details of MGS model {#model-details}
+# Model details {#model-details}
 
-- Give the model in terms of arbitrary taxa
-- Consider effect of variation in efficiency within a taxon
-- Discuss assumption about efficiencies being constant at the species level
+This appendix provides a more detailed description of our deterministic model of the microbiome measurement process, which it uses to derive some of the additional claims of the main text.
 
-# Methods for assaying absolute densities {#review-absolute-methods}
+## MGS measurement
 
-# Taxonomic bias, counting noise, and zeros
+This section expands the description of the model in the main text to more explicitly lay out our assumptions and to consider the effect of variation in measurement efficiencies within a taxonomic group on its MGS measurement.
 
-- Motivation/intro P
+A taxon can be any group of organisms.
+For any taxon $I$ and sample $a$, we define the absolute efficiency of that taxon and that sample as the number of reads assigned to that taxon divided by the number of cells of that taxon in the given  sample.
+We assume that these assigned reads really do come from that particular taxon and sample (though this assumption is often violated in practice).
 
-Simplest model: deterministic bias model + multinomial sampling.
-Number of reads assigned to a species as multinomial with a given total read count for the sample, and a the biased proportions as input.
-(Can treat this as a heuristic model, or can give an actual justification as to why this makes sense even for bioinformatic bias.)
+The efficiency of a taxon $I$ is the average number of reads produced by each of its constituent cells.
+This number varies due to
+
+1. Genetic variation among cells
+1. Non-genetic variation among cells
+1. Randomness in the experimental process
+
+Variation encoded in the genome of cells determines properties that affect its efficiency, such as how easy it is to lyse, how easily PCR primers will bind to their target locus, how many copies of a marker gene are in the genome, and whether its sequencing reads can be taxonomically assigned.
+This variation created by genetically encoded traits is what we refer to as 'taxonomic bias' and is the primary focus of this article.
+In principle, even a single nucleotide change (e.g. in a primer-binding site) could affect the efficiency of a cell.
+Nevertheless, cells will more similar genomes will tend to have more similar genetically-encoded efficiencies.
+For the purposes of our analysis, we suppose that cells within the same species have approximately identical genetically-encoded efficiencies.
+The impact of violations in this assumption can be understood using the approach we describe below for taxa above the species level.
+More generally, our results can be read as applying to the finest taxonomic unit that is reported by the given MGS method.
+
+Genetically identical cells can differ in cell state such that their efficiencies vary significantly.
+An extreme example is the effect of sporulation: Spores are generally harder to lyse than vegetative cells, and @vaninsberghe2020diar found that the efficiency of a common DNA extraction protocol was 800X on vegetative cells vs spores of a strain of _Clostridium difficile_.
+Physiological differences might also arise simply due where they are in the cell cycle.
+To the extent that the cells of a given genotype tend to have a similar distribution of cell states across samples, the efficiency of the genotype will remain stable.
+We assume this stability throughout our analysis.
+Violations in this assumption can again be understood using an approach similar to that we use to considering taxa above the species level.
+
+The final number of reads also varies due to idiosyncratic events that we typically think of (and model) as 'random', such as the precise handling of a sample during pipetting or loading of a sequencing flow cell.
+Such variation may or may not be dwarfed by the 'random' variation in the sample composition (i.e., the variation not explained by the covariates in a regression analysis).
+For simplicity, we ignore this random variation when considering the impact of taxonomic bias on MGS measurement.
+The effect of random variation in regression analysis is addressed in Appendix
+\@ref(appendix-regression) and the effect of random counting error is discussed in Section \@ref(implications).
+
+The absolute efficiency of a taxon can also vary simply because more effort is put towards sequencing that sample, either intentionally or unintentionally.
+The sequencing effort might depend on the taxonomic composition and bias of a sample.
+For example, a sample with higher output DNA will often be diluted more prior to amplification or sequencing.
+The critical assumption we make regarding sequencing effort is that it affects all species in the sample equally and hence doesn't affect their relative abundances.
+
+With these assumptions in place, we can partition the multiplicative difference between the reads assigned to a taxon and its density in the source sample into a taxon-specific factor and a sample-specific factor.
+There is an extra degree of freedom, which we settle by equating the taxon-specific factors with relative measurement efficiencies and arbitrarily choose the relative efficiency of a particular species to equal 1.
+
+Our model for the read counts of a species $i$ in sample $a$ can thus be expressed mathematically as
 \begin{align}
-\text{reads}_i(a)
-  &= \text{Multinomial}\left( \text{total reads}(a), \text{prop}_{i}(a) \times \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} \right).
+  \text{reads}_{i}(a)
+  = \text{density}_{i}(a) \cdot \text{efficiency}_{i} \cdot \text{effort}(a).
 \end{align}
-The expected read count (conditional on read total) is
+Note that the species' efficiencies are properties of the species $i$ but not the sample $a$.
+Similarly, let $\text{density}_{I}(a)$, $\text{reads}_{I}(a)$, and $\text{efficiency}_{I}(a)$ denote the density, read count, and relative efficiency of an arbitrary group of species $I$ in sample $a$.
+The read count for taxon $I$ in sample $a$ can be written
 \begin{align}
-E\left[\text{reads}_i(a)\right] = \text{total reads}(a) \times \text{prop}_{i}(a) \times \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)}.
+  (\#eq:measurement-model-general)
+  \text{reads}_{I}(a)
+  = \text{density}_{I}(a) \cdot \text{efficiency}_{I}(a) \cdot \text{effort}(a).
+\end{align}
+The efficiency of taxon $I$ in sample $a$ equals the weighted average of its constituent species,
+\begin{align}
+  (\#eq:efficiency-general)
+  \text{efficiency}_I(a) 
+  \equiv \frac{\sum_{i\in I}\text{density}_i(a)\cdot \text{efficiency}_i}{\text{density}_I(a)}.
+\end{align}
+The ratio between the read counts of two taxa $I$ and $J$ is
+\begin{align}
+  (\#eq:ratio-error-general)
+  \widehat{\text{ratio}}_{I/J}(a)
+  &= \frac{\text{density}_{I}(a)}{\text{density}_{J}(a)} \cdot \frac{\text{efficiency}_{I}(a)}{\text{efficiency}_{J}(a)}.
+\end{align}
+Equation \@ref(eq:prop-error) for the error in a species' proportion and Equation \@ref(eq:ratio-error) for the error in the ratio between two species both follow directly from this more general expression.
+
+### Revised main text equations
+
+Let $S$ be the set of all species.
+Then
+\begin{align}
+  \text{reads}_S(a)
+    = \text{density}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
+\end{align}
+where 
+\begin{align}
+  \text{efficiency}_S(a) 
+  \equiv \frac{\sum_{j\in S}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{density}_S(a)}
+\end{align}
+Also,
+\begin{align}
+  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
+\end{align}
+and
+\begin{align}
+  \widehat{\text{prop}}_{i}(a)
+  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}.
 \end{align}
 
-(Note, might be simpler to use a Poisson model, especially for overdispersion case.)
+## Taxonomic bias in total-density measurement {#total-density-bias}
 
-Zero counts are likely when the expected read count is $\lesssim 1$. 
-The expected read count is lower and hence the likelihood of zeros is much higher for species with efficiencies substantially below the sample mean.
-Whereas in the absence of taxonomic bias, an observation of zero is taken as evidence that the actual proportion of a species is similar or less than the reciprocal of the total read count, with taxonomic bias, it instead implies only that
+To understand the impact of taxonomic bias in the total-density measurement on community normalization, we model error in total-density measurements similarly to that for MGS measurements.
+Let $\widehat{\text{density}}_{S}(a)$ be the measurement of the density of all species $S$.
+We suppose that this number is the sum of the (unobserved) contributions from each species.
+The contribution of a species $i$ is proportional to its actual density $\text{density}_{i}(a)$ and its _absolute efficiency for the total-density measurement_ $\text{efficiency}_{i}^{\text{tot}}(a)$.
+The contributions may also be multiplied by a common sample-specific error factor to account for non-linearity of extraction and/or random error, which we ignore for now.
+The measured total density therefore equals
 \begin{align}
-  \text{prop}_{i}(a)
-  \lesssim \frac{1}{ \text{total reads}(a) } 
-  \times \frac{\text{mean efficiency}(a)}{ \text{efficiency}_{i}}.
-\end{align}
-
-The likelihood of a zero for a given taxon varies with the mean efficiency of a sample.
-For example, when the vaginal microbiome shifts from Lactobacillus to Gardnerella dominance and the mean efficiency drops by 10X in figure x, it becomes ...
-This becomes problematic for presence/absense analysis.
-But it is also an issue for analysis of log proportions, log ratios, or log densities, in a manner that depends on how the DA method handles zeros.
-A simple and common approach is to ignore the uncertainty from the counting process, and opt to replace zeros with some defined, small value (either in the counts or in the proportions).
-(TODO: implications)
-Some statistical packages (e.g. corncob, fido) take a more sophisticated approach and directly account for the uncertainty by modeling the counting process.
-For these methods, bias against an organism can cause reduced precision, in a sample-dependent manner.
-
-- Maybe: implications of overdispersion / multiplicative noise prior to counting noise.
-- Note, the above is for regression analysis, but looking at two-sample case is also illustrative. Can see a case where having a zero or very small proportion (if overdispersed) makes it very hard to 
-- Consider case of a zero observation; how to interpret, in the light of taxonomic bias?
-- Limit of zero efficiency -> a value of zero reveals no information about the taxon.
-- Maybe: Implications for presence/absense analysis
-
-# Taxonomic bias in total-density measurements {#appendix-total-density}
-
-A noted in the main text, we should expect species-specific efficiencies for total density estimates.
-State the model: Assume constant absolute measurement efficiencies, $\text{efficiency}^{\text{tot}}_i$, which denote the absolute efficiency with which a cell of species $i$ contributes to the measurement of total density in sample $a$.
-This implies consistent relative efficiencies and that density measurement is linear, in the sense that a doubling of the actual density when species composition is fixed always leads to a doubling in the measured density.
-Absolute because we're not considering a compositional measurement; the scale of the species' efficiencies does impact the measured density, though not the fold variation across samples.
-
-How do these species specific efficiencies impact the total density measurement?
-We can write the measured total density in terms of the species' efficiencies as
-\begin{align}
-  \widehat{\text{total density}}(a) 
-  &= \sum_i \text{density}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
-\\&= \text{total density}(a) \cdot \text{mean efficiency}^{\text{tot}}(a)
+  \widehat{\text{density}}_S(a) 
+  &= \sum_{i\in S} \text{density}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
+\\&= \text{density}_S(a) \cdot \text{efficiency}^{\text{tot}}_S(a)
 \end{align}
@@
   (\#eq:app-total-mean-efficiency)
-  \text{mean efficiency}^{\text{tot}}(a) 
-  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}^{\text{tot}}_j}{\text{total density}(a)}
+  \text{efficiency}^{\text{tot}}_S(a) 
+  \equiv \frac{\sum_{i \in S}\text{density}_j(a)\cdot \text{efficiency}^{\text{tot}}_i}{\text{density}_S(a)}
 \end{align}
 is the mean efficiency in the sample with respect to the total-density measurement.
+
 The error in the total density depends on the species composition through the mean efficiency of the sample; hence spurious changes in measured density can occur simply due to shifts from high- to low-efficiency species (or vice versa).
@@
   \widehat{\text{density}}_{i}(a) 
-  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{mean efficiency}^{\text{tot}}(a)}{\text{mean efficiency}(a)}.
+  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}.
 \end{align}
-<!-- \begin{align} -->
-<!--   \newcommand{\allspecies}{\boldsymbol{\cdot}} -->
-<!--   \widehat{\text{density}}_{i}(a)  -->
-<!--   = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}_{\allspecies}^{\text{tot}}(a)}{\text{efficiency}_{\allspecies}(a)}. -->
-<!-- \end{align} -->
-Note, having more similar efficiencies between MGS and total measurements does not necessarily improve the accuracy of the individual sample density estimates (possible misconception of other article?), since even if they perfectly cancel the species-specific error remains.
-Yet it does improve fold change estimates.
-Equation \@ref(eq:app-density-prop-error) shows that the variation in the fold error across samples is driven by the variation in the ratio of the mean efficiency of the total measurement and that of the sequencing measurement.
-To the extent that these two are positively associated (specifically, their logarithms are positively linearly correlated), then the variation across samples will be lower than if a perfectly accurate total density were used.
+The mean efficiency of the total measurement appears in the numerator, while the mean efficiency of the MGS measurement is in the denominator. 
+To the extent that these two are positively associated across samples (specifically, their logarithms are positively linearly correlated), then the variation in one will tend to be offset by the variation in the other.
+In this case the (geometric) variation in the ratio will be reduced, and the fold error in the species density measurement will tend to be more consistent than if the total density measurement was taxonomically unbiased (i.e.
+$\text{efficiency}^\text{tot}_S(a) = 1$ for all samples).
+The error in species densities for individual samples may not be reduced; but the error in fold changes and regression analysis across samples will be.
 
-Applications of these ideas:
+The extent to which the log mean efficiency of the MGS measurement and the total measurement are correlated depends on the species efficiencies as well as the changes in species composition across samples, making it difficult to make general statements about this effect.
+We can get a sense for the effect by considering the correlation between log efficiencies across species between the two measurement types.
+Suppose that distribution over species of log efficiencies of the MGS measurement has variance $\sigma^{2}$, the log efficiencies for the total density measurement has variance $\sigma_{\text{tot}}^{2}$, and the correlation between the two is $\rho$.
+The variance in the difference $\log \text{efficiency}_{i}^{\text{tot}} - \log \text{efficiency}_{i}$ is thus $\sigma^{2} + \sigma_{\text{tot}}^{2} - \rho \sigma \sigma_{\text{tot}}$.
+All else equal, the smaller this variance, the smaller we expect the variation in the error factor $\text{efficiency}^{\text{tot}}_S(a)/ \text{efficiency}_S(a)$.
 
-- example: 16S sequencing may be better paired with 16S qPCR than flow cytometry
-- example: shotgun sequencing may be best paired with bulk DNA measurement (depending on bioinformatics protocol); can even handle a large fraction of unassigned reads _if_ total reads are used in the denominator of the proportion calculation (rather than assigned reads)
-- example: estimating total density from a spike-in; mathematically reduces to the reference-species measurement; here the error completely offsets.
+**Unclassified reads:** It is common that a given MGS protocol generates sequencing reads from particular taxa that the bioinformatics portion entirely discards, due to an inability to taxonomically assign the reads to the chosen taxonomic units.
+For instance, 16S data that is analyzed with closed-reference OTU assignment and shotgun data analyzed by mapping to a database of reference genomes will be unable to assign sequences that are less than a chosen similarity cutoff (e.g.
+97%) from the reference sequences.
+These taxa may form an appreciable fraction of the community and thus lead to a large fraction of the total sequenced reads for a given sample may remain unassigned.
 
-Extra assumptions in this section:
-- Linearity of DNA extraction
-- Thinking that log efficiencies between MGS and total are positively correlated comes from the idea that we can treat bias from different steps and/or sources as approximately independent
+So far, we have defined the efficiency of the MGS protocol in terms of taxonomically classified reads and so have treated such taxa as having an MGS efficiency of 0.
+If such species are included in the total density measurement, then we have a mismatch between the two measurement types.
+However, we may be able to remove this mismatch by altering the equation we use to form species density measurements.
+In particular, we can compute the proportion of the focal species from the ratio of its assigned reads to the total sequenced reads, rather than to the total set of assigned reads.
+This approach raises additional questions around how to treat reads from assignable species that are filtered during quality control steps.
 
-# Taxonomic bias in spike-ins and targeted measurements
+Suppose that the set of species that can be classified by our bioinformatics pipeline, $S$, is a proper subset of a larger set of species $S'$ that are present and yield sequencing reads in our sample.
+Suppose that there is no additional taxonomic bias in our MGS measurement; that is, all species in $S$ have an efficiency of 1, and the species in $S' \setminus S$ have an efficiency of 0.
+Moreover, suppose that all reads from the $S$ species are classified; that is, there is no loss of reads due to routine QC filtering.
+Further suppose that we are capable of perfectly measuring the total density of all cells from the full set of species $S'$; that is, $\text{efficiency}^{\text{tot}}_{i} = 1$ for all species $i \in S'$.
 
-- Distinction between cellular vs DNA and  pre- vs post-extraction spike-ins and targeted measurements
-- Pull from Section 5 of the theory notebook
+In this case, we can overcome the mismatch between MGS and total measurement when we use them to measure species densities, by using the total reads rather than the assigned reads to compute the species proportions.
+That is, instead of 
+\begin{align}
+  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)},
+\end{align}
+we take
+\begin{align}
+  \widehat{\text{prop}}_{i}(a) 
+  &= \frac{\text{reads}_i(a)}{\text{total reads}(a)}
+\\&= \frac{\text{reads}_i(a)}{\text{reads}_{S'}(a)}.
+\end{align}
+By accounting for the contribution of unknown species when computing proportions, we are able to resolve the mismatch and obtain accurate species densities.
+
+## Using reference species for total-density normalization {#total-density-ref}
+
+Constant reference species are sometimes used to measure total density of $S$ by the ratio of $S$ reads to $R$ reads.
+For example, a study of *Arabidopsis* microbiomes used the ratio of bacterial to host reads in shotgun sequencing as a proxy for total bacterial density, which they then used for total-community normalization of 16S amplicon sequencing measurements (@karasov2020ther, @regalado2019comb).
+@chng2020meta similarly used the ratio of bacterial to host or diet reads in shotgun sequencing of mouse fecal samples as a proxy for total bacterial density (though they did not use this measurement for community normalization).
+@smets2016amet similarly used the ratio of non-spike-in to spike-in reads to estimate total density.
+
+What is the impact of taxonomic bias on these total density estimates and the species densities derived from them?
+The measured density of $S$ is
+\begin{align}
+  \widehat{\text{density}}_{S}(a) 
+  &= \frac{\text{reads}_S(a)}{\text{reads}_{R}(a)}
+\\&= \frac{\text{density}_S(a)}{\text{density}_{R}} \cdot \frac{\text{efficiency}_S(a)}{\text{efficiency}_{R}}
+\\&= \text{density}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{constant}.
+\end{align}
+Hence variation in the mean efficiency among the species $S$ across samples creates a variable fold error in the density measurement, similar to direct measurements of total density.
+
+The impact of this variable error for species density measurement via Equation \@ref(eq:density-prop-meas) depends on whether the species proportions are derived from the same or a different MGS measurement.
+
+If proportions from the same MGS measurement are used, then the measured species densities are
+\begin{align}
+  \widehat{\text{density}}_{i}(a) 
+  &= \frac{\text{reads}_i(a)}{\text{reads}_{S}(a)} \cdot \widehat{\text{density}}_{S}(a)
+\\&= \frac{\text{reads}_{i}(a)}{\text{reads}_{R}(a)}.
+\end{align}
+where the second line follows by substitution of the previous equation.
+The variable error in the total density cancels with that in the proportion, yielding a constant fold error.
+In fact, the measurement we obtain is exactly the same as that from reference normalization (Equation \@ref(eq:density-ratio-meas)).
+
+The situation differs when the total density and community composition are estimated using different sequencing and/or bioinformatic methods, as in the plant microbiome example above.
+In this case, the mean efficiency of the total density measurement will not equal that in the proportions, and the more general case of total-density normalization discussed in Appendix \@ref(total-density-bias) applies.
+
+Although we only consider constant reference species, similar behavior occurs if for varying reference species assayed by targeted measurements.
+
+## Linearity of extraction
+
+Some popular absolute-abundance methods use post-extraction DNA (or RNA) density as a proxy for cell density in the original sample; in particular, using florescence-based total DNA quantification or using qPCR or ddPCR to quantify a marker-gene.
+In addition, some have suggested normalization of species to spike-ins of extraneous DNA added after extraction.
+Both of these approaches presuppose that DNA extraction is linear in the sense that the DNA concentration is proportional to the cell concentration in the original sample (possibly after correction by known dilution or concentration factors).
+Deviations from linearity can occur due to random and systematic variation in DNA yields.
+Here we consider the impact of these deviations on species density measurements.
+
+One source of systematic variation in yield is taxonomic bias.
+We already explored the impact of taxonomic bias in extraction.
+
+NOTE: This might not be a correct way to talk about this; should perhaps make a more explicit model to verify this argument.
+
+To understand the impact of non-linearity after bias has been accounted for, let $C(a)$ be a sample-specific factor that accounts for DNA extraction non-linearity after controlling for bias.
+From Equation \@ref(eq:density-prop-error), we have the more general equation
+\begin{align}
+  (\#eq:app-density-prop-error)
+  \widehat{\text{density}}_{i}(a) 
+  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)} \cdot C(a).
+\end{align}
+$C(a)$ might fluctuate randomly among samples, or might vary systematically --- for example, if DNA yield saturates at high concentrations then $C(a)$ will be smaller in higher-concentration samples.
+If DNA extraction is not linear, then measures of cell density might give more accurate fold changes even if they share less bias with the MGS measurement.
+
+Post-extraction spike-ins and targeted measurements face a similar problem.
+These perfectly control for bias in fold change estimation _if_ DNA extraction is linear.
+However, if not then the fold error in $\widehat{\text{density}_r}(a)$ for a species with a targeted measurement will vary with $C(a)$ and cause error in fold changes.
+The same problem applies to DNA spike-ins added after DNA extraction (although different accounting is needed since $\text{density}_r(a)$ no longer has meaning).
+
+# Review of experimental methods for obtaining absolute densities {#review-absolute-methods}
+
+There are many experimental techniques to be able to add absolute-density information to MGS measurements.
+Here we review the experimental techniques; the next section considers the implications for systematic error.
+
+_NOTE: Right now I don't consistently address why various targeted methods might be expected to produce constant fold errors. In revision, seek to connect each method with the relevant theory._
+
+## Measurement of total cell density
+
+Total cell density in the original sample can be directly measured by cell counting, either via microscopy (@kevorkian2018esti, @lloyd2020evid) or flow cytometry (@props2017abso, @vandeputte2017quan).
+Total cell density or biomass can also be measured via properties assumed to be proportional to cell density, such as fluorescence (as in fluorescence spectroscopy, @wang2021curr), components of microbial cell membranes (as in PLFA analysis, @smets2016amet), and the rate of microbial respiration (as in SIR method, @smets2016amet).
+So far, it is primarily the cell counting methods that have been used for species-density measurement (rather than simply for the total density), by multiplying the estimated total density by the MGS proportions.
+
+## Measurement of total DNA density post extraction
+
+It is also common to use density of bulk DNA or a marker gene as a proxy for total community density.
+Marker-gene density is typically measured with qPCR or ddPCR using 'universal primers' that target the marker gene of interest (typically the 16S gene for bacterial microbiome experiments) (@tettamantiboshier2020comp, @jian2020quan, @galazzo2020howt).
+Bulk DNA density can be measured using fluorescence-based DNA quantification assays (@contijoch2019gutm; @korpela2018inte).
+In either case, the DNA density is measured after DNA extraction and so is affected by taxonomic bias in the extraction process, such as variation in lysis efficiency among species.
+Other sources of bias that affect the DNA density measurement include variation in marker-gene copy number (for marker-gene density) and variation in genome size (for bulk DNA density).
+
+The measured DNA density is typically used as a direct proxy for cell density in the original sample.
+In particular, it is assumed that a doubling of cell density in the original sample leads to a doubling of DNA density in the extraction (possibly after adjustment for known dilution factors).
+This _linearity assumption_ may be violated for several reasons.
+First, because of taxonomic bias.
+For example, samples dominated by easy-to-lyse species will give more DNA per cell than samples dominated by hard-to-lyse species.
+Second, systematic non-linearity may occur in the DNA yield as a function of input, even if species composition is held fixed. 
+For example, DNA yield may saturate at high sample inputs.
+Third, the DNA yield may vary apparently randomly due to subtle differences in sample chemistry or handling during the experiment.
+
+## Equivolumetric protocol
+
+A large part of the reason that there is not a direct correspondence between total density in the sample and total reads sequenced is that MGS experiments are typically intentionally designed to yield a similar number of sequencing reads from each sample, regardless of total density.
+@cruz2021equi propose instead designing the MGS experiment so as to make total reads proportional total density.
+The 'equivolumetric protocol' they develop represents a first attempt in this direction. 
+In their protocol, total reads is a saturating function of total density; this function can be measured with a calibration experiment, and the calibration curve used to predict the total density in the source sample.
+This total density estimate is then used to scale the read counts to estimate species densities in a manner equivalent to the total-community density method.
+
+## Housekeeping species
+
+We use _housekeeping species_ (by analogy with housekeeping genes used for normalization in RNAseq experiments) to refer to species whose density is assumed to be constant, either in the MGS sample or in the source ecosystem it is derived from.
+
+Housekeeping species can sometimes be identified from prior scientific knowledge.
+Several studies that have employed shotgun sequencing of host-associated microbiomes have use the plant or animal host for this purpose.
+A study of *Arabidopsis* microbiomes used the ratio of bacterial to host reads in shotgun sequencing as a proxy for total bacterial density, which they then used for total-community normalization of 16S amplicon sequencing measurements (@karasov2020ther, @regalado2019comb).
+@chng2020meta similarly used the ratio of bacterial to host reads in shotgun sequencing of mouse fecal samples as a proxy for total bacterial density (though they did not use this measurement for community normalization).
+They also use reads from dietary plants for the same purpose.
+@wallace2021thed used shotgun sequencing to study the virome of _Drosophila_, and normalized virus reads to _Drosophila_ reads to measure viral abundance per fly.
+Organelle reads can also be used. 
+@diener2021nonr use mitochondria reads in 16S sequencing of mouse fecal pellets to assess total microbial load, though in a qualitative fashion (as mitochondrial reads were only non-zero at very low bacterial densities induced by antibiotics).
+
+In some cases, there may also be microbes or viruses thought to have stable densities.
+A recent example is that the most abundant DNA virus in human feces, crAssphage, and the most abundant RNA virus in human feces, Pepper Mild Mottle Virus, have been treated as stable reference species in wastewater monitoring for SARS-CoV-2.
+Although primarily used in the context of qPCR measurements, these viruses could also be used as references in RNA and DNA shotgun sequencing experiments.
+
+Housekeeping species may not be known _a priori_; to address this case, several methods have been put forward to computationally identify unchanging microbes species directly from MGS measurements.
+These studies are often focused on mammalian gut bacterial communities.
+It is perhaps unreasonable to expect bacterial species to be unchanging across hosts, but weaker assumptions can be made to develop normalization methods for the MGS measurements with a similar spirit to reference-based normalization.
+These studies have instead developed normalization methods based on assumptions such as that most species do not change between any pair of samples (@david2014host) or that the mean (log) abundance between two sample conditions is unchanged for at least some species (@mandal2015anal, @kumar2018anal).
+
+When housekeeping species are sequenced along with the primary MGS measurement, they can be used to obtain species densities via reference-normalization (Equation \@ref(eq:density-ratio-meas). 
+In this case, the only relevant taxonomic bias is that of the primary MGS measurement; if it is constant then it will cancel in fold-change calculations.
+Because the density of the housekeeping species is unknown, we can consider either that the density of focal species has a constant error, or is in units of the housekeeping species.
+Non-constant error might arise if the species is treated as constant when it is in fact not.
+
+Housekeeping species have also been used to estimate total community density by $\text{reads}_{S} / \text{reads}_{R}$.
+This estimate has been used to study variation in total density across samples, or for total-density normalization.
+
+## Spike-ins
+
+Spike-in methods differ by the biological spike-in material: 
+Cellular spike-ins can be added prior to DNA extraction, and DNA spike-ins can be added prior to or following DNA extraction.
+In either case, a variety of methods can be used to actually leverage the spike-ins for absolute density analysis.
+
+**Cellular spike-ins:**
+Cellular spike-ins are added to the sample prior to DNA extraction.
+Some sample processing has typically occurred prior to spiking, for the purposes of storage (e.g. a freeze thaw cycle) and homogenization.
+We should expect there to be taxonomic bias between the spike-in species and those naturally in the sample due to genetic differences and because of physiological differences induced by the sample processing prior to spiking and the experimental procedure used to grow and prepare the spike-in cells.
+Our analysis acknowledges this bias, but assumes that it is consistent across samples.
+The nominal density of the spike-in species added to each sample is subject to random and systematic fold error; but systematic fold error that is shared across samples will induce a constant fold error in species densities and so not impact DA analysis.
+For instance, if source stock is actually 1.5X higher concentration than thought, the true spike-in concentration will be 1.5X greater than nominal in all samples and not pose a problem for accurate DA inference beyond leading to a greater than intended sequencing effort being expended on the spike-in.
+
+Example studies include @stammler2016adju, @ji2019quan, and @rao2021mult.
+
+**DNA spike-ins:** 
+Another possibility is to add DNA spike-ins, derived from natural or artificial sequence.
+DNA spike-ins can be added the samples before DNA extraction (e.g. @smets2016amet, @tkacz2018abso, @zemb2020abso) or after DNA extraction (e.g. @hardwick2018synt, @tkacz2018abso).
+Adding spike-ins prior to extraction is thought to be preferable as it makes it possible to detect and correct for variation in DNA extraction yield among samples
+(@tkacz2018abso, @zemb2020abso, @harrison2021theq).
+Below, we consider the distinction between pre- and post-extraction spike-ins in the light of taxonomic bias coupled with other sources of variation in extraction efficiency.
+
+**How spike-ins are used:**
+Like housekeeping species, spike-ins have been used in a variety of ways to analyze absolute abundances.
+Let $R$ (for reference) be the spike-in species and $S$ be the native species.
+@smets2016amet used the ratio of $S$ to $R$ reads as an estimate of total density,  which they were interested in for its own sake, though one could imagine then also using this total density estimate in community normalization (Equation \@ref(eq:density-prop-meas)).
+@zemb2020abso used the spike-ins to measure total density from the ratio of $S$ to $R$ qPCR abundance estimates and then used Equation \@ref(eq:density-prop-meas).
+@stammler2016adju and others used the ratio-based method of Equation \@ref(eq:density-ratio-meas).
+
+## Targeted measurements
+
+A variety of methods exist for targeted measurement of absolute density of a specific species (or higher-order taxon).
+The most common approach is to use qPCR or ddPCR to measure the concentration of a marker gene in the extracted DNA, using primers scoped to the target taxon.
+This approach is therefore subject to sources of taxonomic bias including  extraction, marker-gene copy number, and primer-binding.
+It is also subject to non-species-specific variation in extraction yields unless these are otherwise controlled for.
+It is also possible to directly measure cell density using methods.
+Some species can be measured by CFU counting after plating on selective media (REFs), and ddPCR has been used to direct measure cells (@dreo2014opti, @morella2018rapi) and viruses (@pavsic2016digi, @morella2018rapi) without first performing an extraction.
+Species-specific florescent probes also make it possible to measure individual species via microscopy or flow cytometry (REFs).

### 2021-11-16 abundance-measurement.Rmd
Edit abundance-measurement section

@@
 
-We distinguish between two types of species-level *relative abundances*---abundances in which the absolute scale is not intrinsically meaningful.
+We distinguish between two types of species-level *relative abundances*.
 The first type consists of the *ratios* among the abundances of two or more species, such as a ratio of 5:4:2 among three species.
-The second type consists of the *proportion* of an individual species, equal to its abundance divided by that of the chosen root taxon (e.g. Prokaryotes).
-Species' ratios and proportions constitute distinct views of relative abundance and hence what it means for relative abundances to change across samples.
-Although proportions are dominant in microbiome DA analysis, researchers increasingly adopt the ratio view via analyses adapted from the statistical framework of Compositional Data Analysis (CoDA).
+The second type consists of the *proportion* of an individual species, equal to its abundance divided by that of a particular root taxon (e.g. Prokaryotes).
+DA analyses of proportions consider how the proportion of a species changes indepently of other species, whereas DA analyses of ratios consider the conserted change in relative abundance among two or more species.
+Although proportion-based analysis is dominant, microbiome researchers increasingly use ratio-based analyses adapted from the statistical framework of Compositional Data Analysis (CoDA).
 The different manner in which taxonomic bias affects ratios vs.
@@
 But a proportion is, after all, just a ratio between two taxa in which the denominator taxon is the entire taxonomic domain under study.
-The difference in behavior from that of ratios between species arises because the efficiency of higher-level taxa, which consist of the aggregate of many species, varies across samples with the changing relative abudances among its constituent species.
+The difference in behavior from that of ratios between species arises because the efficiency of higher-level taxa, which consist of the aggregate of many species, varies across samples with the changing relative abundances among its constituent species.
 

### 2021-11-20 abundance-measurement.Rmd
Make minor text fixes
Fix section references

@@
   (\#eq:prop-meas)
-  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
+  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads}(a)}.
 \end{align}
@@
 
-Taxonomic bias can also create systematic error in the total-density measurements (Appendix \@ref(appendix-total-density)).
+Taxonomic bias can also create systematic error in the total-density measurements (Appendix \@ref(total-density-bias)).
 For now, we suppose that the total density can be measured accurately; 
@@
 
+<!-- NOTE: the above doesn't quite make sense for DNA spike-ins -->
+
 **Differences between the two approaches:**
@@
 Studies using spike-ins as constant reference species will often, instead of using Equation \@ref(eq:density-ratio-meas), first use the spike-in species to estimate the total density, and then estimate the density of focal species with Equation \@ref(eq:density-prop-meas).
-If the calculation is done carefully, however, this method still yields constant fold errors so long as the species efficiencies (including those of the spike-in) are constant across samples (Appendix \@ref(appendix-total-density)).
+If the calculation is done carefully, however, this method still yields constant fold errors so long as the species efficiencies (including those of the spike-in) are constant across samples (Appendix \@ref(total-density-ref)).

### 2021-11-20 solutions.Rmd
Make minor text fixes
Fix section references

@@
 Methods like using flow cytometry that directly measures cell density lacks these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
-By extending the analysis of Section \@ref(differential-abundance) to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(appendix-total-density)).
+By extending the analysis of Section \@ref(differential-abundance) to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(total-density-bias)).
 This finding suggests that methods of total-density measurement that are more accurate as measures of total cell density may actually perform worse than less accurate methods for the purposes of absolute DA inference.
@@
 These pairings even make it possible to account for error caused by variation among samples in the fraction of unclassified reads, which can form a major fraction of amplicon and shotgun data.
-Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (Appendix \@ref(appendix-total-density)).
+Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (Appendix \@ref(total-density-bias)).
 
-## Bias-sensitivity analysis and bias-aware meta-analysis
+## Bias-sensitivity analysis
 

### 2021-11-29 v/8f81dd5293e7ef5bc3621bbbec2f5d7a41b6560d/_main.tex
Fix section references
This commit adds the rendered manuscript associated with
8f81dd5293e7ef5bc3621bbbec2f5d7a41b6560d

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements},
  pdfauthor={Michael R. McLaren; Jacob T. Nearing; Amy D. Willis; Karen G. Lloyd; Benjamin J. Callahan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{cancel}

% Define note environment based on example from https://bookdown.org/yihui/rmarkdown-cookbook/custom-blocks.html
\usepackage{color}
\usepackage{framed}
\setlength{\fboxsep}{.8em}
\newenvironment{rmdnote}{
  \definecolor{shadecolor}{RGB}{248,248,248}
  \begin{shaded}}
 {\end{shaded}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements}
\author{Michael R. McLaren\footnote{North Carolina State University; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Jacob T. Nearing\footnote{Dalhousie University} \and Amy D. Willis\footnote{University of Washington} \and Karen G. Lloyd\footnote{University of Tennessee} \and Benjamin J. Callahan\footnote{North Carolina State University}}
\date{2021-11-29}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{preface}{%
\section*{Preface}\label{preface}}
\addcontentsline{toc}{section}{Preface}

\begin{verbatim}
#> working directory clean
\end{verbatim}

\emph{This manuscript was rendered from commit 8f81dd5293e7ef5bc3621bbbec2f5d7a41b6560d.}

\leavevmode\vadjust pre{\hypertarget{preface-warning}{}}%
\textbf{This in-progress manuscript is not intended for general scientific use.}
It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
Please post comments or questions on the \href{https://github.com/mikemc/differential-abundance-theory/issues}{GitHub Issues page} or \href{m.mclaren42@gmail.com}{email Mike}.

This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.
We describe the basic problem posed by taxonomic bias for measuring changes in the abundance of particular taxa across conditions and describe new strategies for mitigating the errors it induces.
Analyses of both relative and absolute abundances are considered.
In its current form, the manuscript sits somewhere between a standard scientific article and a monograph;
it consists of an article followed by a series of appendices which together give a comprehensive treatment of the implications of the McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) model of taxonomic bias for differential-abundance analysis and experimental design.
It is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
See \href{https://doi.org/10.5281/zenodo.4552717}{the Zenodo record} for how to cite the latest version.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

One of the most basic questions we can ask about microbial communities is: How do different microbial taxa vary in abundance---across space, time, and host or environmental conditions?
Marker-gene and shotgun metagenomic sequencing (jointly, MGS) can be used to measure the abundances of thousands of species simultaneously, making it possible to ask this question on a community-wide scale.
In these \emph{differential-abundance (DA) analyses}, the change in abundance of a microbial taxon across samples or conditions is used to infer ecological dynamics or find microbes that are associated with specific host diseases or environmental conditions.
Although standard MGS measurements lose information about total microbial density---and so are typically used to analyze the abundances of taxa relative to each other---new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density.
In its various forms, DA analysis remain one of, if not the most, common forms of analyses applied to MGS data to elucide the inner workings of microbiomes and their relationships to host and environmental health.

Yet these DA analysis are built on a fundamentally flawed foundation.
MGS measurements are \emph{taxonomically biased}: Microbial species vary dramatically (e.g.~10-1000X) in how efficiently they are measured---that is, converted from cells into taxonomically classified sequencing reads---by a given MGS protocol (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
This bias arises from variation in how species respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
Although often associated with features specific to marker-gene sequencing---the variation among species in marker copy numbers and in primer-binding and amplification efficiencies---the existence of large variation in DNA extraction efficiencies and in the ability to correctly classify reads make taxonomic bias a universal feature of both marker-gene and shotgun measurements.
As a result of taxonomic bias, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (Yeh et al. (\protect\hyperlink{ref-yeh2018taxo}{2018}),McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
These errors have been found in some cases to supersede sizable biological differences (e.g. Lozupone et al. (\protect\hyperlink{ref-lozupone2013meta}{2013})) and have may have contributed to failed replications of prominent findings in the human microbiome literature, such as the associations of Bacteroides and Firmicutes in stool with obesity (Finucane et al. (\protect\hyperlink{ref-finucane2014atax}{2014})) and the associations of certain species in the vagina of pregnant women with preterm birth (Callahan et al. (\protect\hyperlink{ref-callahan2017repl}{2017})).

The extent to which taxonomic bias has impacted the DA results in the scientific literature is unknown.
The typical approach taken to counter taxonomic bias is to standardize the measurement protocol used within a given study, with the (often tacit) assumption being that samples measured by the same protocol will be affected by bias in the same way and so the measured differences between samples will be unaffected.
For example, if taxonomic bias were to cause the measured proportion of a species to consistently be 10X too high, we would still be able to accurately infer the fold change in its proportion across samples (Kevorkian et al. (\protect\hyperlink{ref-kevorkian2018esti}{2018}),Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020})).
Unfortunately, mathematical arguments and analysis of experiments with artificially constructed (`mock') communities demonstrate that this assumption is not always warranted:
Consistent taxonomic bias can lead to variable fold errors in species' proportions (Figure \ref{fig:basic-problem}, McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase) (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
Yet McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that certain types of DA analysis---those based on fold changes in the ratios among species---where robust to bias.
The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as typically done in microbiome association testing, have yet to be investigated.

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundance.
Our analysis clarifies how the folk wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.
Over time, application of these methods to past and future experiments will provide crucial quantitative information about the conditions under which taxonomic bias creates spurious results for various DA methodologies.
Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2021-10-27-illustrate-the-problem/illustrate-the-problem_files/figure-html5/brooks_lcrisp_prop_and_fc-1.pdf}
\caption{\label{fig:basic-problem}\textbf{Mock community experiments show that taxonomic bias can distort the measured fold change in an individual species' proportion across samples.} The figure shows the measured vs.~actual proportions for a single bacterial species, \emph{Lactobacillus crispatus}, in a set of bacterial cellular mock communities, and the resulting fold changes between community samples. The inconsistent error in the measured proportions of individual samples (Panel A) leads to inaccurate measurements of fold changes (Panel B). Mock communities were constructed and measured with 16S sequencing by Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}). The data was re-analyzed by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}), who showed that despite the inconsistency of the errors in Panel A, taxonomic bias acted consistently across samples.}
\end{figure}

\hypertarget{abundance-measurement}{%
\section{How taxonomic bias affects abundance measurements}\label{abundance-measurement}}

To understand how bias affects the measured differential abundance between samples, we first consider how it affects measurements of individual samples.

\hypertarget{model-of-mgs-measurement}{%
\subsection{Model of MGS measurement}\label{model-of-mgs-measurement}}

Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the \emph{compositional} nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (Gloor et al. (\protect\hyperlink{ref-gloor2017micr}{2017})).
We consider a set of microbial species that are measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes samples to assign reads to various microbial species.
We assume that assigned reads have been correctly assigned to the given species and sample.
Thus we ignore the possibility of contamination or false-positive taxonomic assignment, though we allow that many reads may not receive a taxonomic assignment and so ultimately be discarded.
Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the `noise' or random error due to the random sampling of sequencing reads and other aspects of the MGS process.

Our model stipulates that the assigned read count of species \(i\) in sample \(a\) equals its cell density multiplied by a species-specific factor and a sample-specific factor,
\begin{align}
  \label{eq:measurement-model}
  \text{reads}_{i}(a)
  = \text{density}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor is the \emph{relative measurement efficiency} (or \emph{efficiency} for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
The sample-specific factor, which we call the \emph{sequencing effort} of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
Equation \eqref{eq:measurement-model} implies that the total reads from all species in the sample equal
\begin{align}
  \label{eq:total-reads}
  \text{total reads}(a)
    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
\end{align}
where
\begin{align}
  \label{eq:mean-efficiency}
  \text{mean efficiency}(a) 
  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
\end{align}
is the average or mean efficiency of cells in the sample.

\hypertarget{relative-abundances-proportions-and-ratios}{%
\subsection{Relative abundances (proportions and ratios)}\label{relative-abundances-proportions-and-ratios}}

We distinguish between two types of species-level \emph{relative abundances}.
The first type consists of the \emph{ratios} among the abundances of two or more species, such as a ratio of 5:4:2 among three species.
The second type consists of the \emph{proportion} of an individual species, equal to its abundance divided by that of a particular root taxon (e.g.~Prokaryotes).
DA analyses of proportions consider how the proportion of a species changes indepently of other species, whereas DA analyses of ratios consider the conserted change in relative abundance among two or more species.
Although proportion-based analysis is dominant, microbiome researchers increasingly use ratio-based analyses adapted from the statistical framework of Compositional Data Analysis (CoDA).
The different manner in which taxonomic bias affects ratios vs.~proportions (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})) forms the basis in subsequent sections for understanding the extent to which its effects cancel in DA analysis.

The standard measurement of the proportion of species \(i\) in sample \(a\) is given by the ratio of its read count to the total,
\begin{align}
  \label{eq:prop-meas}
  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads}(a)}.
\end{align}
From Equations \eqref{eq:measurement-model}, \eqref{eq:total-reads}, and \eqref{eq:prop-meas}, it follows that
\begin{align}
  \label{eq:prop-error}
  \widehat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)},
\end{align}
where \(\text{prop}_{i}(a) = \text{density}_{i}(a) / \text{total density}(a)\) is the actual proportion.
Equation \eqref{eq:prop-error} says that taxonomic bias creates a multiplicative error equal to the ratio of the species' efficiency to the mean efficiency in the sample.
Consequently, the same species can be over- or under-measured depending on the composition of the given sample (Figure \ref{fig:error-proportions}).
For instance, in samples from two hypothetical communities in Figure \ref{fig:error-proportions}, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).

The measured ratio between species \(i\) and \(j\) is given by the ratio of their read counts,
\begin{align}
  \label{eq:ratio-meas}
  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
\end{align}
From Equations \eqref{eq:measurement-model} and \eqref{eq:ratio-meas}, it follows that
\begin{align}
  \label{eq:ratio-error}
  \widehat{\text{ratio}}_{i/j}(a)
%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
\end{align}
The fold error in the ratio between two species simply equals the ratio in their efficiencies and is therefore constant across samples.
For instance, in the example of Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/export-pdf/figures/illustrations/error-proportions} \caption{\textbf{Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.} Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species. Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.}\label{fig:error-proportions}
\end{figure}

\textbf{Difference between species proportions and species ratios:}
Consistent taxonomic bias at the species level creates a consistent fold error in the ratios among species, but a varying fold error in the proportions of individual species.
But a proportion is, after all, just a ratio between two taxa in which the denominator taxon is the entire taxonomic domain under study.
The difference in behavior from that of ratios between species arises because the efficiency of higher-level taxa, which consist of the aggregate of many species, varies across samples with the changing relative abundances among its constituent species.

\hypertarget{absolute-abundances-cell-densities}{%
\subsection{Absolute abundances (cell densities)}\label{absolute-abundances-cell-densities}}

A wide range of experimental methods exist for converting the relative-abundance information in MGS measurements into measurements of (absolute) cell densities;
however, these methods can be generally partitioned into two types, based on whether the absolute-abundance information derives from knowing the density of the total community or knowing the density of one or more individual species.

\textbf{Methods using the density of the total community:}
The total microbial density of a community can be assayed by a variety of methods (reviewed in Appendix \ref{review-absolute-methods}).
Once the total density is known, the density of individual species can be estimated by multiplying the total density by the species' proportions (as measured by MGS),
\begin{align}
  \label{eq:density-prop-meas}
  \widehat{\text{density}}_{i}(a) 
  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{total density}}(a)
\\&= \text{reads}_{i}(a) \cdot \frac{\widehat{\text{total density}}(a)}{\text{total reads}(a)}.
\end{align}
We can rewrite the species' density measurements to account for the effect of taxonomic bias on the MGS proportions (Equation \eqref{eq:prop-error}) and any error in the total density measurement,
\begin{align}
  \label{eq:density-prop-error}
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
%  \cdot \frac{\widehat{\text{total density}}(a)}{\text{total density}(a)}.
  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{total density}}(a) \end{array}.
\end{align}
The variable fold error that bias creates in the measured proportions directly transfers to the density measurements.

Taxonomic bias can also create systematic error in the total-density measurements (Appendix \ref{total-density-bias}).
For now, we suppose that the total density can be measured accurately;
later, Section \ref{solutions} describes how experiments might be designed such that taxonomic bias in the total density measurement offsets that in the MGS proportions, leading to a (more) consistent fold error in the measured densities.

\textbf{Methods using the density of one or more reference species:}
The density of one or more species can be directly measured using targeted measurement methods like species-specific qPCR; alternatively, the densities of one or more species may be thought to be constant across samples either from background knowledge or because they have been spiked in at a fixed density (these situations are reviewed in Appendix \ref{review-absolute-methods}).
These \emph{reference species} with known or constant density can be used to convert read counts to densities for all species.
For instance, given the density of a reference species \(r\), we can estimate the density of a focal species \(i\) by
\begin{align}
  \label{eq:density-ratio-meas}
%  \widehat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \widehat{\text{density}}_{r}(a).
  \widehat{\text{density}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{density}}_{r}(a)}{\text{reads}_{r}(a)}.
\end{align}
Alternatively, if we only know that the reference species is constant, we can set \(\widehat{\text{density}}_{r}(a)\) in Equation \eqref{eq:density-ratio-meas} equal to 1 to estimate the density in unknown but fixed units, which is sufficient for multiplicative DA analysis.
(Appendix \ref{model-details} considers possible extensions to multiple reference species.)
It follows from Equation \eqref{eq:ratio-error} that the error caused by taxonomic bias in this density measurement is
\begin{align}
  \label{eq:density-ratio-error}
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{density}_r}(a) \end{array}.
\end{align}
Here, it is the constant error in the measured ratio of species \(i\) to \(r\) that propagates into the density measurement.
There will generally be systematic error in the density of the reference species; however, if the systematic fold error is constant across samples, so will be that of the densities of other species.

\textbf{Differences between the two approaches:}
Equations \eqref{eq:density-prop-error} and \eqref{eq:density-ratio-error} show that the manner in which taxonomic bias in the sequencing measurement impacts the two approaches mirrors that of proportions and ratios.
If the fold error in the total-density or reference species measurement is negligible (or at least constant) then the fold errors in species densities by the total-density approach vary across samples with the mean efficiency, whereas the fold errors by the reference-species approach are constant.
The fundamental determinant of whether constant error is obtained is not which of these two equations is ultimately used, but rather whether densities of individual species, which (we assume) have constant efficiency, or the total community, which has varying efficiency, provide the absolute-abundance information.
Studies using spike-ins as constant reference species will often, instead of using Equation \eqref{eq:density-ratio-meas}, first use the spike-in species to estimate the total density, and then estimate the density of focal species with Equation \eqref{eq:density-prop-meas}.
If the calculation is done carefully, however, this method still yields constant fold errors so long as the species efficiencies (including those of the spike-in) are constant across samples (Appendix \ref{total-density-ref}).

\hypertarget{differential-abundance}{%
\section{How taxonomic bias affects differential-abundance analysis}\label{differential-abundance}}

How do the measurement errors described in the previous section impact our ability to estimate the changes in microbial abundances across samples or between different host and environmental conditions?
Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, which have direct ecological interpretations (via the processes of exponential growth and death) and are ubiquitous in microbiome DA analysis.

\hypertarget{change-between-a-pair-of-samples}{%
\subsection{Change between a pair of samples}\label{change-between-a-pair-of-samples}}

Before considering the stereotypical many-sample DA analysis, it is instructive to
consider simplest analysis of differential abundance: the measurement of fold changes in abundance between a pair of samples.
This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment, and conceptually bridges the single-sample results of the previous section to the many-sample case.

The composition-dependent effect of bias on fold error in proportions leads to error in fold-change measurements that is proportional to the inverse change in mean efficiency.
From the error in an individual sample (Equation \eqref{eq:prop-error}), it follows that the measured fold change in proportion of taxon \(i\) from sample \(a\) to sample \(b\) is
\begin{align}
  \label{eq:prop-fc-error}
% \tag*{Fold change in proportion}
\underbrace{\frac{\widehat{\text{prop}}_{i}(b)}{\widehat{\text{prop}}_{i}(a)}} _\text{measured FC}
  &= \frac
    {\text{prop}_{i}(b) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(b)}}
    {\text{prop}_{i}(a) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(a)}}
\\[0.5ex]
  &=
  \underbrace{\frac{\text{prop}_{i}(b)}{\text{prop}_{i}(a)}}_\text{actual FC}
  \cdot
  \underbrace{\left[\frac{\text{mean efficiency}(b)}{\text{mean efficiency}(a)}\right]^{-1}}_\text{fold error}
  .
\end{align}
The sample-independent efficiency factor of the error cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency from \(s\) to \(t\).

The bottom row of Figure \ref{fig:error-proportions} illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
Consequently, the FC of each species is measured to be 2.6X larger than the true value.
Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
We can see each type of error in Figure \ref{fig:error-proportions}.
For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FC (actual FC: 2.3X, measured FC: 6.5X).
For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!

In contrast, because species ratios are distorted by a constant factor, their measured fold changes remain accurate.
The fold error in Equation \eqref{eq:ratio-error} completely cancels when we divide the ratio measured for one sample \(a\) by another sample \(b\).

If other error sources remain negligible (or constant across samples), then this dichotomy continues to apply to proportion- and ratio-based density measurements.
The inferred FCs in proportion-based density measurements will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
Moreover, any species with a constant density will appear to vary inversely with the mean efficiency.
In contrast, fold changes in ratio-based density measurements will remain accurate.

\hypertarget{regression-analysis-of-many-samples}{%
\subsection{Regression analysis of many samples}\label{regression-analysis-of-many-samples}}

DA analysis of many samples across host or environmental conditions can typically be framed as a regression problem, in which we analyze the relationship between a microbial \emph{response variable}, such as log density of some focal species \(i\), and one or more \emph{covariates}, such the pH or temperature of the sampled environment or whether the sample is from a healthy or sick person.
A large fraction of DA analyses use elaborations on the simple linear regression model, which for a response of log species density can be written
\begin{align}
  \label{eq:regression}
  \log \text{density}_i(a) = \alpha + \beta x(a) + \varepsilon_i(a).
\end{align}
Here \(x\) is a continuous covariate (e.g.~pH) or a binary covariate (e.g.~\(x=1\) for treated patients and \(x=0\) for controls), \(\alpha\) and \(\beta\) are regression coefficients, and \(\varepsilon_i(a)\) is a mean-zero random variable that reflects the residual (unexplained) variation in the response (log density of species \(i\)).
Our interest is usually in the coefficient \(\beta\) (slope or average difference between conditions) that describes how the species' abundance changes with \(x\), while the intercept \(\alpha\) captures differences in the baseline abundance and---we hope---measurement efficiency among species.
How does taxonomic bias under our measurement model affect estimates of \(\beta\) in the simple linear regression?

Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \eqref{eq:density-prop-meas}) with error-free estimates of the total density.
If the true log density follows the regression Equation \eqref{eq:regression}, then it follows from Equation \eqref{eq:density-prop-error} the estimated log density equals
\begin{align}
  \label{eq:regression-error}
  \log \widehat{\text{density}}_i(a)
%  &= \alpha + \beta x(a) + \varepsilon_i(a) + \log \text{efficiency}_i - \log \text{mean efficiency}(a)
  = [\alpha + \log \text{efficiency}_i] + [\beta - \log \text{mean efficiency}(a)] x(a) + \varepsilon_i(a).
\end{align}
This equation shows that the species-specific portion of the error affects the intercept term while the sample-specific portion (log mean efficiency) affects the slope term.
Thus, as in the case of fold changes between two samples (Equation \eqref{eq:prop-fc-error}), it is the variation in the (log) mean efficiency across samples that we must worry about distorting our DA results.

The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient \(\beta\).
Appendix \ref{appendix-regression} mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \ref{fig:regression-example} illustrates using simulated data.
Variation in the log mean efficiency that is associated with the covariate \(x\) creates a systematic error in the estimated slope \(\hat \beta\) equal to the negative of the (scaled) covariance of log mean efficiency with \(x\).
The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with \(x\) relative to that of the response (here, \(\log \text{density}_{i}\)) with \(x\) or, equivalently, the relative magnitudes of their slopes.
As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \ref{fig:regression-example}), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
Variation in the log mean efficiency that is uncorrelated with \(x\) does not systematically distort \(\hat \beta\) but does affect its precision, typically leading to increased standard errors as the variation in log mean efficiency effectively acts as an additional source of noise in measured abundance (Figure \ref{fig:regression-example} D).
The exception is for species whose residual variation is strongly positively correlated with that of log mean efficiency (here, Species 9), which can appear to have less random variation and receive standard errors that are too small.
Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).

With this understanding in place, we briefly summarize how taxonomic bias affects estimation of the simple linear model in other abundance types.
The results for proportion-based density estimates with accurate total densities also apply to LFC analysis of proportions.
Similar results apply to microbiome regression tools (such as corncob; Martin, Witten, and Willis (\protect\hyperlink{ref-martin2020mode}{2020})) that perform regression on the logit (instead of log) proportion of a species; however,
the mean efficiency of the entire sample must instead be replaced with the mean efficiency among all species excluding the focal species, which causes the absolute error in regression coefficients to vary somewhat across species.
Because ratios and ratio-based densities are subject to consistent multiplicative error, in analysis of log ratios and the log densities derived from them, only the estimated intercept \(\hat \alpha\) is affected by taxonomic bias, while the point estimate and standard error of the estimated slope \(\hat \beta\) remain unaffected.

\begin{figure}
\centering
\includegraphics{figures/export-pdf/home/michael/research/differential-abundance-theory/notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example_files/figure-html5/main-figure-1.pdf}
\caption{\label{fig:regression-example}\textbf{Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.} This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by \(x=0\) and \(x=1\). In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFC, which drives the positive association of the log mean efficiency with the condition (shown in Panels A and B). This positive LFC in the log mean efficiency induces a systematic negative shift in the estimated LFCs of all species (Panels C and D). Panel D shows the mean LFC (points) with 95\% confidence intervals (CIs), for each species estimated from either the actual or the measured densities. The error (difference in LFC estimates on measured and actual) equals the negative LFC of the mean efficiency (shown in Panel B).}
\end{figure}

\hypertarget{implications}{%
\section{Implications for real-world inference~}\label{implications}}

Are biologically significant errors in DA analyses likely in practice?
We can begin to answer this question using a combination of theoretical analysis, consideration of hypothetical scenarios, and analysis of case studies for which some information about taxonomic bias is available.

\hypertarget{systematic-error-in-slope-or-lfc-estimates}{%
\subsection{Systematic error in slope or LFC estimates}\label{systematic-error-in-slope-or-lfc-estimates}}

Recall from Section \ref{differential-abundance} that the error in the slope or LFC in a DA regression analysis is proportional to the covariance of the log mean efficiency with the covariate.
This covariance can be split into two components: the standard deviation of (log) mean efficiency and its correlation with the covariate.
Hence when considering whether taxonomic bias creates large error, it can be useful to separately ask whether the mean efficiency is likely to vary across samples and whether it is likely to be correlated with a covariate of interest.

One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
To characterize bias in a 16S sequencing protocol developed for the Vaginal Human Microbiome Project (VaHMP), Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}) performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome.
This protocol was later used for the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study (Fettweis et al. (\protect\hyperlink{ref-fettweis2019thev}{2019})).
We used the method for measuring and correcting bias developed by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to measure bias from the mock communities and analyze its potential impact on DA analyses that use the MOMSPI measurements (SI Analysis MOMS-PI).
The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
The three species that most often dominate the calibrated profiles are \emph{Gardnerella vaginalis}, \emph{Lactobacillus iners}, and \emph{Lactobacillus crispatus}.
Our mock-community analysis indicate \emph{G. vaginalis} has the lowest efficiency of the seven species, roughly 20X lower than \emph{L. crispatus} and 30X lower than \emph{L. iners} (the most efficiently measured).
As a result, the mean efficiency varies substantially across samples, with samples dominated by a \emph{Lactobacillus} species typically having an efficiency that is 3-20X greater than samples dominated by \emph{Gardnerella} (Figure \ref{fig:momspi-mean-efficiency-dist}).
Shifts between \emph{Lactobacillus} and \emph{Gardnerella} dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \ref{fig:momspi-mean-efficiency-fcs}).
Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure TODO).
Decreases in mean efficiency during transitions from \emph{Lactobacillus} to \emph{Gardnerella} dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify \emph{Gardnerella}.

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files/figure-html5/momspi-mean-efficiency-dist-1.pdf}
\caption{\label{fig:momspi-mean-efficiency-dist}\textbf{The mean efficiency in vaginal samples from the MOMS-PI study varies with the most abundant species.}}
\end{figure}

A second study in which control communities make it possible to directly analyze the impact of taxonomic bias in experimental samples was performed by Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}).
To study the interactions between a host plant and its fungal commensals and pathogen, the authors inoculated trees with a fungal synthetic community and later exposed plants to a fungal pathogen.
The authors used ITS amplicon sequencing to measure communities before and after infection, along with mock communities which they used to estimate the species efficiencies with the method of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
(DNA mocks were used, so that bias due to DNA extraction is not included, but other major bias sources such as PCR and ribosomal copy-number-variation are included.)
We reanalyzed this experiment to consider the impact that bias has on different possible DA analyses (SI Analysis Fungi).
As noted by Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}), there was a 13X difference between the most and least efficiently measured commensals, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
Yet despite this 13X variation among commensals, the mean efficiency is remarkably stable across communities prior to pathogen colonization.
Consequently, taxonomic bias creates little systematic error in proportion-based DA analysis of these pre-colonization samples.
The high efficiency of the pathogen did, however, typically lead to a roughly 6X increase in the mean efficiency of post-colonization samples.
Hence DA analysis of how commensals responded to the pathogen did not first discard pathogen reads would therefore tend to suggest that the commensals had substantially smaller fold changes than they actually did, unless bias were accounted for.

It is clear that an individual species with an unusually high or low efficiency can drive large changes in the mean efficiency when it shifts from a very large (\(\sim 1\)) to very small proportion (\(\ll 1\)) of the sample.
Consideration of dynamics in the human gut illustrates another a way in which large shifts in higher-level taxa might create similar shifts, due to the fact that efficiencies do not vary independently among closely related species.
Gut microbiomes are dominated by two phyla: The Bacteroidetes and the Firmicutes, the ratio of which varies substantially across individuals.
Within a sample, there are usually multiple species contributing to the abundance of each phylum, with the median Inverse Simpson diversity being around 4 for each phylum in the HMP2 IBD study.
Thus changes from Bacteroidetes to Firmicutes dominance are unlikely to be driven by a single species.
However, if Firmicutes species tend to have systematically different efficiencies than Bacteroidetes species, then changes in phylum dominance can be expected to drive substantial changes in the mean efficiency even if many species contribute.
Indeed, several studies have found DNA extraction protocols to more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though substantial variation within phyla also occurs, McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).

Systematic error in DA results requires that the variation in log mean efficiency is correlated with the covariate.
The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
For example, in the vaginal microbiome a decline in \emph{Lactobacillus} and rise of species including \emph{Gardnerella vaginalis} is associated with preterm birth in pregnant women; hence the variation in mean efficiency driven by these species may also create an association of mean efficiency with preterm status.
In the plant-fungal experiment of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}), the increase in pathogen proportion after infection leads to a large increase in the mean efficiency over time, which systematic increases the observed decline in the proportions of commensal species (SI Analysis).
The Bacteroidetes-to-Firmicutes ratio has also been linked to host conditions.
Yet in the HMP2 IBD dataset, the large variation in this ratio is largely independent of disease status, suggesting that the primary effect of any variation in mean efficiency driven by these phyla would be to increase noise rather than create systematic error in DA analysis.

Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.

Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020}) estimate the log fold change in the estimated absolute abundance of various taxa with sediment depth (as a proxy for time) to determine which taxa are able to persist and even grow in this difficult environment.
It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
As the relative abundance of tougher species increases with depth, the mean extraction efficiency might therefore decrease.
This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
Additional measurements by Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020}) make it possible to investigate this possibility explicitly.
The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for particular taxa.
For instance, in Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years using the proportion-based densities and 10.3 years by qPCR (Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020}) Table 1 and Figure 4).
These doubling times correspond to highly similar estimated slopes of log density per year of \(1/10.1 \approx 0.99\) and \(1/10.3\approx 0.97\).
The small difference between these slopes suggests that there is little correlation of log mean efficiency with burial time and hence little systematic error in the estimated doubling times derived from the proportion-based density estimates.

Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.

The biological significance of the error in \(\hat \beta\) depends on the relative magnitude of the LFC (or covariance with the covariate \(x\)) in the focal species vs in the mean efficiency.
It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for other species.
The experimental and computational methods discussed in Section \ref{solutions} can mitigate taxonomic bias in these clearly-problematic cases while also helping to amass empirical evidence on the conditions in which bias is unlikely to cause major inferential errors.

\hypertarget{loss-of-precision-and-power}{%
\subsection{Loss of precision and power}\label{loss-of-precision-and-power}}

Another way in which taxonomic bias hinders DA analysis is by increasing the noise in abundance measurements, leading to a reduction in the precision of LFC estimates and thus the power to detect true associations.

There are two distinct mechanisms by which such reductions occur.
The first was described in Section \ref{differential-abundance}: Variation in the mean efficiency that is random (in the sense of being unassociated with the covariates in a regression analysis) tends to increase the standard errors in estimated regression coefficients.
As noted above, this effect may be the dominant one in cases like the HMP2 IBD dataset where large variation in dominant phylum is independent of disease status.
Correcting for the variation in mean efficiency using the calibration methods described in the next section can reverse this effect, thereby increasing the precision as well reducing the systematic error in DA inference.

The second mechanism involves the counting noise that stems from the random sampling of reads in the sequencing experiment.
This counting noise is expected to be the dominant source of noise for species whose expected proportion \emph{after accounting for taxonomic bias} is \(\lesssim 1 / \text{total reads}\) for a given sample.
Having an atypically low efficiency makes a species much more likely to be near this threshold and hence to receive highly uncertain measurements.
Such imprecise measurements can make it near impossible to make biologically meaningful conclusions.
For example, in the experiment of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) described above, the median read counts in samples taken after pathogen colonization for the 5 commensals with the lowest efficiencies were 0, 0, 0, 1, and 11.
These species' efficiencies ranged from 40X-10X lower than the pathogen's efficiency; thus, in the absence of bias the counts may have been much higher and it might have been possible to make much more precise conclusions about how their relative abundances changed during pathogen invasion.
The literature on the vaginal microbiome during pregnancy provides us with another example: Disagreement among studies over whether \emph{Gardnerella} is positively associated with preterm birth seems to be in part driven by the fact that several studies used 16S primers that are extremely poor at amplifying \emph{Gardnerella} and hence lack the precision necessary to find associations even were they to exist.
Unfortunately, even perfect knowledge of the measurement efficiencies will not allow us to overcome the low precision caused by small expected read counts; however, the methods described in the section provide principled ways to assess whether the low counts are due to taxonomic bias rather than truly low frequencies in a species' abundance.

\hypertarget{solutions}{%
\section{Potential solutions}\label{solutions}}

The theoretical results from Section \ref{differential-abundance} suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.

\hypertarget{analyses-of-fold-changes-in-species-ratios}{%
\subsection{Analyses of fold changes in species ratios}\label{analyses-of-fold-changes-in-species-ratios}}

Since the fold changes in species ratios are unaffected by consistent taxonomic bias, a natural approach to countering bias in relative DA analysis is to ratio- rather than proportion-based methods.
A variety of such methods have been developed for microbiome DA analysis that build on earlier work from the field of Compositional Data Analysis (CoDA).
These methods analyze log-fold changes in the ratios among a pair of species or, more generally, some product of exponential power of multiple species, which likewise are bias-invariant.
Besides a greater robustness to taxonomic bias, such ratio-based analyses have other advantages.
First, analysis of ratios avoids other standard criticisms leveled at proportion-based analysis---namely that the change in proportion of a species is affected by change in the density of all other species and the set of species used in the denominator of the proportion calculation (Gloor et al. (\protect\hyperlink{ref-gloor2017micr}{2017})).
Second, ratios may be more informative than absolute density in some cases, such as for swabs or other sample types for which the absolute density in the sample may be an unreliable indicator of the absolute density of the sampled ecosystem.

Ratios come with their limitations, however.
First, there are a vast number of species ratios (and derivatives) to potentially consider, none of which provide direct information about absolute densities, which can make choosing and interpreting a ratio-based DA analysis challenging.
Second, ratio measurements are impacted by noise in both the numerator and denominator and so can be noisier than proportion measurements, particularly when the read counts in the denominator are small.
A related problem is that, due to the discrete nature of biological organisms and the reads that are generated during sequencing, species in the denominator are commonly observed to have 0 reads; the assumptions that made by different methods about the meaning of these 0 counts can have a major impact on DA results.
Finally, bias invariance at the level of ratios among species does not extend to ratios of additive aggregates of species, such as between bacterial phyla, unless further conditions are met (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).

\hypertarget{calibrate-compositions}{%
\subsection{Calibration using community controls}\label{calibrate-compositions}}

Measurement of community calibration controls along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis.
Community calibration controls are samples whose species identities and relative abundances are known either by construction or by characterization with a chosen `gold standard' or \emph{reference protocol} (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
MGS measurement of one or more control communities can be used to directly measure the relative efficiencies among the superset of species in the controls (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
The measured relative efficiencies can then be used to calibrate (remove the effect of bias from) the relative abundances (ratios) of the control species in a set of primary (non-control) experimental samples that were measured with the same protocol (ideally in the same experiment and sequencing run) (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
Calibration can be extended to species not in the controls using statistical methods to impute (predict) the unobserved efficiencies using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).

The calibrated compositions can be used for arbitrary downstream microbiome analyses, including relative and absolute DA analysis.
To demonstrate the potential for calibration to improve fold-change measurements, we estimated bias from one sample in the mock community data from Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}) that contained all 7 species and used it to calibrate the compositions of all samples (Figure \ref{fig:calibration-example}).
We then estimated species' densities using the uncalibrated and calibrated proportions using the community normalization method, treating the total density as known and fixed.
Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions and densities (Figure \ref{fig:calibration-example}).
To demonstrate a practical application, we used the bias estimated from the full set of mock communities as the basis for calibrating the vaginal community time series measurements from the MOMS-PI study (Fettweis et al. (\protect\hyperlink{ref-fettweis2019thev}{2019})) that were made using the same 16S sequencing protocol.
We imputed the efficiencies of other species using taxonomic relationships (Methods).
Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \ref{implications}.

\begin{figure}
\includegraphics[width=1\linewidth]{notebook/_posts/2021-10-25-brooks2015thet-calibration/brooks2015thet-calibration_files/figure-html5/brooks2015thet_fc_calibration-1} \caption{\textbf{Fold changes can be calibrated using community controls or reference species.} The figure compares the performance of three methods for measuring fold changes in absolute cell density in cellular mock communities of 7 vaginal species, which were constructed and measured via 16S sequencing by Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}). The `Uncalibrated' fold changes are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which here is be known to be constant by construction). The `Calibrated (community)' measurements are computed from abundance measurements where the proportions are first corrected for the taxonomic bias that was estimated from a single sample that contained all 7 species. The `Calibrated (reference)' measurements are computed from abundances measured with the reference-species method, with \emph{Lactobacillus crispatus} used as the reference; that is, the true abundance of \emph{L. crispatus} is treated as known and used to infer the abundance of the remaining 6 species. Only samples that contain \emph{L. crispatus} are included.}\label{fig:calibration-example}
\end{figure}

\hypertarget{calibration-using-reference-species}{%
\subsection{Calibration using reference species}\label{calibration-using-reference-species}}

Obtaining control communities that span the taxa of interest is often not practical.
Moreover, the bias that is measured from controls may differ from that in the primary samples due to differences in cell state or sample preparation.
These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-density measurement that was described in Section \ref{differential-abundance}.
This \emph{reference-species calibration} allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from taxonomically biased, relative metagenomics measurements.

Recall that densities estimated by the reference-species approach have a constant fold error, provided that the fold error in the reference species is also constant (Equation \eqref{eq:density-ratio-error}).
Three categories of reference species are reviewed for this purpose in Appendix \ref{review-absolute-methods}, including why they might be expected to produce constant fold errors: housekeeping species assumed to have constant abundance, spike-in species added at known or constant abundance, and species whose abundance has been measured with a targeted method like species-specific qPCR.

To demonstrate the ability for reference calibration to improve fold changes, we treated the abundance of one species (\emph{Lactobacillus crispatus}) in the Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}) mock community data as known, and used it to calibrate the abundances of all species.
Doing so improved the resulting measurements of fold changes in density for all species (Figure \ref{fig:calibration-example}).

\hypertarget{choosing-complementary-total-density-and-metagenomics-measurements}{%
\subsection{Choosing complementary total-density and metagenomics measurements}\label{choosing-complementary-total-density-and-metagenomics-measurements}}

A variety of methods for measuring total community density have been paired with either amplicon or shotgun sequencing data for the purpose of conducting absolute DA analysis using the total-density approach described in Section \ref{differential-abundance}.
Each of these methods is subject to its own taxonomic bias: Cells from different species are likely to contribute more or less efficiently to the total abundance measurement.
Consider the popular method of assessing bacterial abundance with total-16S qPCR measurement.
Even an ideal qPCR protocol that perfectly amplifies all species remains a biased measurement of cell density, since the total 16S copies in the extracted DNA contributed by each species will be proportional to its (species-specific) lysis efficiency and 16S copy-number.
16S qPCR is commonly paired with 16S amplicon sequencing, with which these sources of bias are shared, perhaps along with variation in primer binding and amplification efficiency.
Methods like using flow cytometry that directly measures cell density lacks these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
By extending the analysis of Section \ref{differential-abundance} to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \ref{total-density-bias}).
This finding suggests that methods of total-density measurement that are more accurate as measures of total cell density may actually perform worse than less accurate methods for the purposes of absolute DA inference.

We illustrate with a hypothetical example of two vaginal communities of equal density, but which are dominated either by \emph{Lactobacillus iners} or \emph{Gardnerella vaginalis}.
Based on our earlier analysis in McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}), we estimate that \emph{L. iners} yields 10X more 16S copies per cell than \emph{G. vaginalis} in extracted DNA.
Ignoring other sources of bias, we expect relative efficiencies of \emph{L. iners}:\emph{G. vaginalis} of 10 for both qPCR and 16S sequencing measurements.
Consequently, the \emph{G. vaginalis} community will yield a substantially lower qPCR measurement than the \emph{L. iners} community and we would incorrectly infer a decrease in total abundance.
At the same time, this decrease in the mean efficiency of the sequencing measurement would cause artificially high fold changes in the proportions of all species (Section \ref{differential-abundance}).
When using qPCR and 16S sequencing measurements for community-based density estimation, these two opposing errors cancel, yielding accurate fold changes in species densities.

More generally, changes in the mean efficiency of the total density measurement offset those in the sequencing measurement when comparing log density of species across samples.
If the (log) efficiencies of the total and sequencing measurement are positively correlated across species, then the (log) mean efficiencies will tend to be positively correlated across samples, leading to a reduction in error in DA analysis relative to total-density methods whose efficiencies are uncorrelated with those of the sequencing measurement.
These observations suggest that qPCR of a marker gene may be the ideal pairing for amplicon sequencing measurements, despite being a poor measure of changes in total cell density.
Similarly, bulk DNA quantification may be an ideal pairing for shotgun sequencing, making it possible to account for variation in lysis efficiency and genome length.
These pairings even make it possible to account for error caused by variation among samples in the fraction of unclassified reads, which can form a major fraction of amplicon and shotgun data.
Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (Appendix \ref{total-density-bias}).

\hypertarget{bias-sensitivity-analysis}{%
\subsection{Bias-sensitivity analysis}\label{bias-sensitivity-analysis}}

For experiments that have been conducted without calibration controls, it is still possible to investigate the likelihood that DA results can be explained by taxonomic bias by performing a computational bias-sensitivity analysis.

A bias-sensitivity analysis examines how sensitive the results of a given DA analysis are to the working assumption about the bias of a given protocol.
This working assumption is typically is that there is no bias, but could also be bias measured in a previous experiment or predicted from taxonomic features like 16S copy number.
A straightforward approach to bias-sensitivity analysis is to use computer simulation to re-analyze the MGS data across a set of simulated taxonomic biases.
First, multiple sets of efficiency vectors are randomly generated, each representing a possible quantification of taxonomic bias for the study.
Second, each efficiency vector is used to calibrate the original MGS measurements; each set of calibrated measurements represents the true community compositions under the given hypothesized bias.
Finally, the DA analysis is re-run on each set of calibrated measurements and the distribution of results is compared graphically or with summary statistics.
A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool `corncob' are shown in SI Figure \ref{fig:sensitivity-example}.
The simulated efficiency vectors may be generated to different hypotheses about bias in the given system, such as its overall magnitude and correlation among phylogenetically-related species.
Thus the utility of such simulations will increase the more we learn about the properties of taxonomic bias in different systems.

An alternative approach to bias-sensitivity analysis would be to directly include the (unknown) taxonomic bias of the protocol in the statistical model used for DA analysis, as described for bias-aware meta-analysis below.
Further development of tools and workflows for performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results---for differential abundance and microbiome analyses more generally.

\hypertarget{bias-aware-meta-analysis}{%
\subsection{Bias-aware meta-analysis}\label{bias-aware-meta-analysis}}

So far we have considered analysis of samples subject to the same taxonomic bias;
however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in samples from different studies.
This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses.
A potential solution is to perform a bias-aware meta-analysis that that explicitly accounts for the possibility of distinct taxonomic bias across studies.

Parametric microbiome meta-analysis models include sets of species-specific latent parameters to model ``batch effects''---amorphous differences in the measurements of each study.
By configuring these models so that these latent parameters correspond to study-specific species efficiencies, we can improve their biological interpretability and gain the ability to directly assess whether disagreements in DA results across studies can be explained by differences in taxonomic bias.
To the extent that taxonomic bias truly is consistent within studies, meta-analyses that directly account for it can have greater statistical power to extract the truly consistent and inconsistent microbial patterns across studies.

\hypertarget{appendix-appendix}{%
\appendix \addcontentsline{toc}{section}{\appendixname}}

\hypertarget{model-details}{%
\section{Model details}\label{model-details}}

This appendix provides a more detailed description of our deterministic model of the microbiome measurement process, which it uses to derive some of the additional claims of the main text.

\hypertarget{mgs-measurement}{%
\subsection{MGS measurement}\label{mgs-measurement}}

This section expands the description of the model in the main text to more explicitly lay out our assumptions and to consider the effect of variation in measurement efficiencies within a taxonomic group on its MGS measurement.

A taxon can be any group of organisms.
For any taxon \(I\) and sample \(a\), we define the absolute efficiency of that taxon and that sample as the number of reads assigned to that taxon divided by the number of cells of that taxon in the given sample.
We assume that these assigned reads really do come from that particular taxon and sample (though this assumption is often violated in practice).

The efficiency of a taxon \(I\) is the average number of reads produced by each of its constituent cells.
This number varies due to

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Genetic variation among cells
\item
  Non-genetic variation among cells
\item
  Randomness in the experimental process
\end{enumerate}

Variation encoded in the genome of cells determines properties that affect its efficiency, such as how easy it is to lyse, how easily PCR primers will bind to their target locus, how many copies of a marker gene are in the genome, and whether its sequencing reads can be taxonomically assigned.
This variation created by genetically encoded traits is what we refer to as `taxonomic bias' and is the primary focus of this article.
In principle, even a single nucleotide change (e.g.~in a primer-binding site) could affect the efficiency of a cell.
Nevertheless, cells will more similar genomes will tend to have more similar genetically-encoded efficiencies.
For the purposes of our analysis, we suppose that cells within the same species have approximately identical genetically-encoded efficiencies.
The impact of violations in this assumption can be understood using the approach we describe below for taxa above the species level.
More generally, our results can be read as applying to the finest taxonomic unit that is reported by the given MGS method.

Genetically identical cells can differ in cell state such that their efficiencies vary significantly.
An extreme example is the effect of sporulation: Spores are generally harder to lyse than vegetative cells, and VanInsberghe et al. (\protect\hyperlink{ref-vaninsberghe2020diar}{2020}) found that the efficiency of a common DNA extraction protocol was 800X on vegetative cells vs spores of a strain of \emph{Clostridium difficile}.
Physiological differences might also arise simply due where they are in the cell cycle.
To the extent that the cells of a given genotype tend to have a similar distribution of cell states across samples, the efficiency of the genotype will remain stable.
We assume this stability throughout our analysis.
Violations in this assumption can again be understood using an approach similar to that we use to considering taxa above the species level.

The final number of reads also varies due to idiosyncratic events that we typically think of (and model) as `random', such as the precise handling of a sample during pipetting or loading of a sequencing flow cell.
Such variation may or may not be dwarfed by the `random' variation in the sample composition (i.e., the variation not explained by the covariates in a regression analysis).
For simplicity, we ignore this random variation when considering the impact of taxonomic bias on MGS measurement.
The effect of random variation in regression analysis is addressed in Appendix
\ref{appendix-regression} and the effect of random counting error is discussed in Section \ref{implications}.

The absolute efficiency of a taxon can also vary simply because more effort is put towards sequencing that sample, either intentionally or unintentionally.
The sequencing effort might depend on the taxonomic composition and bias of a sample.
For example, a sample with higher output DNA will often be diluted more prior to amplification or sequencing.
The critical assumption we make regarding sequencing effort is that it affects all species in the sample equally and hence doesn't affect their relative abundances.

With these assumptions in place, we can partition the multiplicative difference between the reads assigned to a taxon and its density in the source sample into a taxon-specific factor and a sample-specific factor.
There is an extra degree of freedom, which we settle by equating the taxon-specific factors with relative measurement efficiencies and arbitrarily choose the relative efficiency of a particular species to equal 1.

Our model for the read counts of a species \(i\) in sample \(a\) can thus be expressed mathematically as
\begin{align}
  \text{reads}_{i}(a)
  = \text{density}_{i}(a) \cdot \text{efficiency}_{i} \cdot \text{effort}(a).
\end{align}
Note that the species' efficiencies are properties of the species \(i\) but not the sample \(a\).
Similarly, let \(\text{density}_{I}(a)\), \(\text{reads}_{I}(a)\), and \(\text{efficiency}_{I}(a)\) denote the density, read count, and relative efficiency of an arbitrary group of species \(I\) in sample \(a\).
The read count for taxon \(I\) in sample \(a\) can be written
\begin{align}
  \label{eq:measurement-model-general}
  \text{reads}_{I}(a)
  = \text{density}_{I}(a) \cdot \text{efficiency}_{I}(a) \cdot \text{effort}(a).
\end{align}
The efficiency of taxon \(I\) in sample \(a\) equals the weighted average of its constituent species,
\begin{align}
  \label{eq:efficiency-general}
  \text{efficiency}_I(a) 
  \equiv \frac{\sum_{i\in I}\text{density}_i(a)\cdot \text{efficiency}_i}{\text{density}_I(a)}.
\end{align}
The ratio between the read counts of two taxa \(I\) and \(J\) is
\begin{align}
  \label{eq:ratio-error-general}
  \widehat{\text{ratio}}_{I/J}(a)
  &= \frac{\text{density}_{I}(a)}{\text{density}_{J}(a)} \cdot \frac{\text{efficiency}_{I}(a)}{\text{efficiency}_{J}(a)}.
\end{align}
Equation \eqref{eq:prop-error} for the error in a species' proportion and Equation \eqref{eq:ratio-error} for the error in the ratio between two species both follow directly from this more general expression.

\hypertarget{revised-main-text-equations}{%
\subsubsection{Revised main text equations}\label{revised-main-text-equations}}

Let \(S\) be the set of all species.
Then
\begin{align}
  \text{reads}_S(a)
    = \text{density}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
\end{align}
where
\begin{align}
  \text{efficiency}_S(a) 
  \equiv \frac{\sum_{j\in S}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{density}_S(a)}
\end{align}
Also,
\begin{align}
  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
\end{align}
and
\begin{align}
  \widehat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}.
\end{align}

\hypertarget{total-density-bias}{%
\subsection{Taxonomic bias in total-density measurement}\label{total-density-bias}}

To understand the impact of taxonomic bias in the total-density measurement on community normalization, we model error in total-density measurements similarly to that for MGS measurements.
Let \(\widehat{\text{density}}_{S}(a)\) be the measurement of the density of all species \(S\).
We suppose that this number is the sum of the (unobserved) contributions from each species.
The contribution of a species \(i\) is proportional to its actual density \(\text{density}_{i}(a)\) and its \emph{absolute efficiency for the total-density measurement} \(\text{efficiency}_{i}^{\text{tot}}(a)\).
The contributions may also be multiplied by a common sample-specific error factor to account for non-linearity of extraction and/or random error, which we ignore for now.
The measured total density therefore equals
\begin{align}
  \widehat{\text{density}}_S(a) 
  &= \sum_{i\in S} \text{density}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
\\&= \text{density}_S(a) \cdot \text{efficiency}^{\text{tot}}_S(a)
\end{align}
where
\begin{align}
  \label{eq:app-total-mean-efficiency}
  \text{efficiency}^{\text{tot}}_S(a) 
  \equiv \frac{\sum_{i \in S}\text{density}_j(a)\cdot \text{efficiency}^{\text{tot}}_i}{\text{density}_S(a)}
\end{align}
is the mean efficiency in the sample with respect to the total-density measurement.

The error in the total density depends on the species composition through the mean efficiency of the sample; hence spurious changes in measured density can occur simply due to shifts from high- to low-efficiency species (or vice versa).

How does this error affect the measured densities of individual species when the total is used for normalization?
Substituting the mean efficiency of the total measurement for the error term in Equation \eqref{eq:density-prop-error} gives
\begin{align}
  \label{eq:app-density-prop-error}
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}.
\end{align}
The mean efficiency of the total measurement appears in the numerator, while the mean efficiency of the MGS measurement is in the denominator.
To the extent that these two are positively associated across samples (specifically, their logarithms are positively linearly correlated), then the variation in one will tend to be offset by the variation in the other.
In this case the (geometric) variation in the ratio will be reduced, and the fold error in the species density measurement will tend to be more consistent than if the total density measurement was taxonomically unbiased (i.e.~\(\text{efficiency}^\text{tot}_S(a) = 1\) for all samples).
The error in species densities for individual samples may not be reduced; but the error in fold changes and regression analysis across samples will be.

The extent to which the log mean efficiency of the MGS measurement and the total measurement are correlated depends on the species efficiencies as well as the changes in species composition across samples, making it difficult to make general statements about this effect.
We can get a sense for the effect by considering the correlation between log efficiencies across species between the two measurement types.
Suppose that distribution over species of log efficiencies of the MGS measurement has variance \(\sigma^{2}\), the log efficiencies for the total density measurement has variance \(\sigma_{\text{tot}}^{2}\), and the correlation between the two is \(\rho\).
The variance in the difference \(\log \text{efficiency}_{i}^{\text{tot}} - \log \text{efficiency}_{i}\) is thus \(\sigma^{2} + \sigma_{\text{tot}}^{2} - \rho \sigma \sigma_{\text{tot}}\).
All else equal, the smaller this variance, the smaller we expect the variation in the error factor \(\text{efficiency}^{\text{tot}}_S(a)/ \text{efficiency}_S(a)\).

\textbf{Unclassified reads:} It is common that a given MGS protocol generates sequencing reads from particular taxa that the bioinformatics portion entirely discards, due to an inability to taxonomically assign the reads to the chosen taxonomic units.
For instance, 16S data that is analyzed with closed-reference OTU assignment and shotgun data analyzed by mapping to a database of reference genomes will be unable to assign sequences that are less than a chosen similarity cutoff (e.g.~97\%) from the reference sequences.
These taxa may form an appreciable fraction of the community and thus lead to a large fraction of the total sequenced reads for a given sample may remain unassigned.

So far, we have defined the efficiency of the MGS protocol in terms of taxonomically classified reads and so have treated such taxa as having an MGS efficiency of 0.
If such species are included in the total density measurement, then we have a mismatch between the two measurement types.
However, we may be able to remove this mismatch by altering the equation we use to form species density measurements.
In particular, we can compute the proportion of the focal species from the ratio of its assigned reads to the total sequenced reads, rather than to the total set of assigned reads.
This approach raises additional questions around how to treat reads from assignable species that are filtered during quality control steps.

Suppose that the set of species that can be classified by our bioinformatics pipeline, \(S\), is a proper subset of a larger set of species \(S'\) that are present and yield sequencing reads in our sample.
Suppose that there is no additional taxonomic bias in our MGS measurement; that is, all species in \(S\) have an efficiency of 1, and the species in \(S' \setminus S\) have an efficiency of 0.
Moreover, suppose that all reads from the \(S\) species are classified; that is, there is no loss of reads due to routine QC filtering.
Further suppose that we are capable of perfectly measuring the total density of all cells from the full set of species \(S'\); that is, \(\text{efficiency}^{\text{tot}}_{i} = 1\) for all species \(i \in S'\).

In this case, we can overcome the mismatch between MGS and total measurement when we use them to measure species densities, by using the total reads rather than the assigned reads to compute the species proportions.
That is, instead of
\begin{align}
  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)},
\end{align}
we take
\begin{align}
  \widehat{\text{prop}}_{i}(a) 
  &= \frac{\text{reads}_i(a)}{\text{total reads}(a)}
\\&= \frac{\text{reads}_i(a)}{\text{reads}_{S'}(a)}.
\end{align}
By accounting for the contribution of unknown species when computing proportions, we are able to resolve the mismatch and obtain accurate species densities.

\hypertarget{total-density-ref}{%
\subsection{Using reference species for total-density normalization}\label{total-density-ref}}

Constant reference species are sometimes used to measure total density of \(S\) by the ratio of \(S\) reads to \(R\) reads.
For example, a study of \emph{Arabidopsis} microbiomes used the ratio of bacterial to host reads in shotgun sequencing as a proxy for total bacterial density, which they then used for total-community normalization of 16S amplicon sequencing measurements (Karasov et al. (\protect\hyperlink{ref-karasov2020ther}{2020}), Regalado et al. (\protect\hyperlink{ref-regalado2019comb}{2020})).
Chng et al. (\protect\hyperlink{ref-chng2020meta}{2020}) similarly used the ratio of bacterial to host or diet reads in shotgun sequencing of mouse fecal samples as a proxy for total bacterial density (though they did not use this measurement for community normalization).
Smets et al. (\protect\hyperlink{ref-smets2016amet}{2016}) similarly used the ratio of non-spike-in to spike-in reads to estimate total density.

What is the impact of taxonomic bias on these total density estimates and the species densities derived from them?
The measured density of \(S\) is
\begin{align}
  \widehat{\text{density}}_{S}(a) 
  &= \frac{\text{reads}_S(a)}{\text{reads}_{R}(a)}
\\&= \frac{\text{density}_S(a)}{\text{density}_{R}} \cdot \frac{\text{efficiency}_S(a)}{\text{efficiency}_{R}}
\\&= \text{density}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{constant}.
\end{align}
Hence variation in the mean efficiency among the species \(S\) across samples creates a variable fold error in the density measurement, similar to direct measurements of total density.

The impact of this variable error for species density measurement via Equation \eqref{eq:density-prop-meas} depends on whether the species proportions are derived from the same or a different MGS measurement.

If proportions from the same MGS measurement are used, then the measured species densities are
\begin{align}
  \widehat{\text{density}}_{i}(a) 
  &= \frac{\text{reads}_i(a)}{\text{reads}_{S}(a)} \cdot \widehat{\text{density}}_{S}(a)
\\&= \frac{\text{reads}_{i}(a)}{\text{reads}_{R}(a)}.
\end{align}
where the second line follows by substitution of the previous equation.
The variable error in the total density cancels with that in the proportion, yielding a constant fold error.
In fact, the measurement we obtain is exactly the same as that from reference normalization (Equation \eqref{eq:density-ratio-meas}).

The situation differs when the total density and community composition are estimated using different sequencing and/or bioinformatic methods, as in the plant microbiome example above.
In this case, the mean efficiency of the total density measurement will not equal that in the proportions, and the more general case of total-density normalization discussed in Appendix \ref{total-density-bias} applies.

Although we only consider constant reference species, similar behavior occurs if for varying reference species assayed by targeted measurements.

\hypertarget{linearity-of-extraction}{%
\subsection{Linearity of extraction}\label{linearity-of-extraction}}

Some popular absolute-abundance methods use post-extraction DNA (or RNA) density as a proxy for cell density in the original sample; in particular, using florescence-based total DNA quantification or using qPCR or ddPCR to quantify a marker-gene.
In addition, some have suggested normalization of species to spike-ins of extraneous DNA added after extraction.
Both of these approaches presuppose that DNA extraction is linear in the sense that the DNA concentration is proportional to the cell concentration in the original sample (possibly after correction by known dilution or concentration factors).
Deviations from linearity can occur due to random and systematic variation in DNA yields.
Here we consider the impact of these deviations on species density measurements.

One source of systematic variation in yield is taxonomic bias.
We already explored the impact of taxonomic bias in extraction.

NOTE: This might not be a correct way to talk about this; should perhaps make a more explicit model to verify this argument.

To understand the impact of non-linearity after bias has been accounted for, let \(C(a)\) be a sample-specific factor that accounts for DNA extraction non-linearity after controlling for bias.
From Equation \eqref{eq:density-prop-error}, we have the more general equation
\begin{align}
  \label{eq:app-density-prop-error}
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)} \cdot C(a).
\end{align}
\(C(a)\) might fluctuate randomly among samples, or might vary systematically --- for example, if DNA yield saturates at high concentrations then \(C(a)\) will be smaller in higher-concentration samples.
If DNA extraction is not linear, then measures of cell density might give more accurate fold changes even if they share less bias with the MGS measurement.

Post-extraction spike-ins and targeted measurements face a similar problem.
These perfectly control for bias in fold change estimation \emph{if} DNA extraction is linear.
However, if not then the fold error in \(\widehat{\text{density}_r}(a)\) for a species with a targeted measurement will vary with \(C(a)\) and cause error in fold changes.
The same problem applies to DNA spike-ins added after DNA extraction (although different accounting is needed since \(\text{density}_r(a)\) no longer has meaning).

\hypertarget{review-absolute-methods}{%
\section{Review of experimental methods for obtaining absolute densities}\label{review-absolute-methods}}

There are many experimental techniques to be able to add absolute-density information to MGS measurements.
Here we review the experimental techniques; the next section considers the implications for systematic error.

\emph{NOTE: Right now I don't consistently address why various targeted methods might be expected to produce constant fold errors. In revision, seek to connect each method with the relevant theory.}

\hypertarget{measurement-of-total-cell-density}{%
\subsection{Measurement of total cell density}\label{measurement-of-total-cell-density}}

Total cell density in the original sample can be directly measured by cell counting, either via microscopy (Kevorkian et al. (\protect\hyperlink{ref-kevorkian2018esti}{2018}), Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020})) or flow cytometry (Props et al. (\protect\hyperlink{ref-props2017abso}{2017}), Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017})).
Total cell density or biomass can also be measured via properties assumed to be proportional to cell density, such as fluorescence (as in fluorescence spectroscopy, Wang et al. (\protect\hyperlink{ref-wang2021curr}{2021})), components of microbial cell membranes (as in PLFA analysis, Smets et al. (\protect\hyperlink{ref-smets2016amet}{2016})), and the rate of microbial respiration (as in SIR method, Smets et al. (\protect\hyperlink{ref-smets2016amet}{2016})).
So far, it is primarily the cell counting methods that have been used for species-density measurement (rather than simply for the total density), by multiplying the estimated total density by the MGS proportions.

\hypertarget{measurement-of-total-dna-density-post-extraction}{%
\subsection{Measurement of total DNA density post extraction}\label{measurement-of-total-dna-density-post-extraction}}

It is also common to use density of bulk DNA or a marker gene as a proxy for total community density.
Marker-gene density is typically measured with qPCR or ddPCR using `universal primers' that target the marker gene of interest (typically the 16S gene for bacterial microbiome experiments) (Tettamanti Boshier et al. (\protect\hyperlink{ref-tettamantiboshier2020comp}{2020}), Jian et al. (\protect\hyperlink{ref-jian2020quan}{2020}), Galazzo et al. (\protect\hyperlink{ref-galazzo2020howt}{2020})).
Bulk DNA density can be measured using fluorescence-based DNA quantification assays (Contijoch et al. (\protect\hyperlink{ref-contijoch2019gutm}{2019}); Korpela et al. (\protect\hyperlink{ref-korpela2018inte}{2018})).
In either case, the DNA density is measured after DNA extraction and so is affected by taxonomic bias in the extraction process, such as variation in lysis efficiency among species.
Other sources of bias that affect the DNA density measurement include variation in marker-gene copy number (for marker-gene density) and variation in genome size (for bulk DNA density).

The measured DNA density is typically used as a direct proxy for cell density in the original sample.
In particular, it is assumed that a doubling of cell density in the original sample leads to a doubling of DNA density in the extraction (possibly after adjustment for known dilution factors).
This \emph{linearity assumption} may be violated for several reasons.
First, because of taxonomic bias.
For example, samples dominated by easy-to-lyse species will give more DNA per cell than samples dominated by hard-to-lyse species.
Second, systematic non-linearity may occur in the DNA yield as a function of input, even if species composition is held fixed.
For example, DNA yield may saturate at high sample inputs.
Third, the DNA yield may vary apparently randomly due to subtle differences in sample chemistry or handling during the experiment.

\hypertarget{equivolumetric-protocol}{%
\subsection{Equivolumetric protocol}\label{equivolumetric-protocol}}

A large part of the reason that there is not a direct correspondence between total density in the sample and total reads sequenced is that MGS experiments are typically intentionally designed to yield a similar number of sequencing reads from each sample, regardless of total density.
Cruz, Christoff, and Oliveira (\protect\hyperlink{ref-cruz2021equi}{2021}) propose instead designing the MGS experiment so as to make total reads proportional total density.
The `equivolumetric protocol' they develop represents a first attempt in this direction.
In their protocol, total reads is a saturating function of total density; this function can be measured with a calibration experiment, and the calibration curve used to predict the total density in the source sample.
This total density estimate is then used to scale the read counts to estimate species densities in a manner equivalent to the total-community density method.

\hypertarget{housekeeping-species}{%
\subsection{Housekeeping species}\label{housekeeping-species}}

We use \emph{housekeeping species} (by analogy with housekeeping genes used for normalization in RNAseq experiments) to refer to species whose density is assumed to be constant, either in the MGS sample or in the source ecosystem it is derived from.

Housekeeping species can sometimes be identified from prior scientific knowledge.
Several studies that have employed shotgun sequencing of host-associated microbiomes have use the plant or animal host for this purpose.
A study of \emph{Arabidopsis} microbiomes used the ratio of bacterial to host reads in shotgun sequencing as a proxy for total bacterial density, which they then used for total-community normalization of 16S amplicon sequencing measurements (Karasov et al. (\protect\hyperlink{ref-karasov2020ther}{2020}), Regalado et al. (\protect\hyperlink{ref-regalado2019comb}{2020})).
Chng et al. (\protect\hyperlink{ref-chng2020meta}{2020}) similarly used the ratio of bacterial to host reads in shotgun sequencing of mouse fecal samples as a proxy for total bacterial density (though they did not use this measurement for community normalization).
They also use reads from dietary plants for the same purpose.
Wallace et al. (\protect\hyperlink{ref-wallace2021thed}{2021}) used shotgun sequencing to study the virome of \emph{Drosophila}, and normalized virus reads to \emph{Drosophila} reads to measure viral abundance per fly.
Organelle reads can also be used.
Diener et al. (\protect\hyperlink{ref-diener2021nonr}{2021}) use mitochondria reads in 16S sequencing of mouse fecal pellets to assess total microbial load, though in a qualitative fashion (as mitochondrial reads were only non-zero at very low bacterial densities induced by antibiotics).

In some cases, there may also be microbes or viruses thought to have stable densities.
A recent example is that the most abundant DNA virus in human feces, crAssphage, and the most abundant RNA virus in human feces, Pepper Mild Mottle Virus, have been treated as stable reference species in wastewater monitoring for SARS-CoV-2.
Although primarily used in the context of qPCR measurements, these viruses could also be used as references in RNA and DNA shotgun sequencing experiments.

Housekeeping species may not be known \emph{a priori}; to address this case, several methods have been put forward to computationally identify unchanging microbes species directly from MGS measurements.
These studies are often focused on mammalian gut bacterial communities.
It is perhaps unreasonable to expect bacterial species to be unchanging across hosts, but weaker assumptions can be made to develop normalization methods for the MGS measurements with a similar spirit to reference-based normalization.
These studies have instead developed normalization methods based on assumptions such as that most species do not change between any pair of samples (David et al. (\protect\hyperlink{ref-david2014host}{2014})) or that the mean (log) abundance between two sample conditions is unchanged for at least some species (Mandal et al. (\protect\hyperlink{ref-mandal2015anal}{2015}), Kumar et al. (\protect\hyperlink{ref-kumar2018anal}{2018})).

When housekeeping species are sequenced along with the primary MGS measurement, they can be used to obtain species densities via reference-normalization (Equation \eqref{eq:density-ratio-meas}.
In this case, the only relevant taxonomic bias is that of the primary MGS measurement; if it is constant then it will cancel in fold-change calculations.
Because the density of the housekeeping species is unknown, we can consider either that the density of focal species has a constant error, or is in units of the housekeeping species.
Non-constant error might arise if the species is treated as constant when it is in fact not.

Housekeeping species have also been used to estimate total community density by \(\text{reads}_{S} / \text{reads}_{R}\).
This estimate has been used to study variation in total density across samples, or for total-density normalization.

\hypertarget{spike-ins}{%
\subsection{Spike-ins}\label{spike-ins}}

Spike-in methods differ by the biological spike-in material:
Cellular spike-ins can be added prior to DNA extraction, and DNA spike-ins can be added prior to or following DNA extraction.
In either case, a variety of methods can be used to actually leverage the spike-ins for absolute density analysis.

\textbf{Cellular spike-ins:}
Cellular spike-ins are added to the sample prior to DNA extraction.
Some sample processing has typically occurred prior to spiking, for the purposes of storage (e.g.~a freeze thaw cycle) and homogenization.
We should expect there to be taxonomic bias between the spike-in species and those naturally in the sample due to genetic differences and because of physiological differences induced by the sample processing prior to spiking and the experimental procedure used to grow and prepare the spike-in cells.
Our analysis acknowledges this bias, but assumes that it is consistent across samples.
The nominal density of the spike-in species added to each sample is subject to random and systematic fold error; but systematic fold error that is shared across samples will induce a constant fold error in species densities and so not impact DA analysis.
For instance, if source stock is actually 1.5X higher concentration than thought, the true spike-in concentration will be 1.5X greater than nominal in all samples and not pose a problem for accurate DA inference beyond leading to a greater than intended sequencing effort being expended on the spike-in.

Example studies include Stämmler et al. (\protect\hyperlink{ref-stammler2016adju}{2016}), Ji et al. (\protect\hyperlink{ref-ji2019quan}{2019}), and Rao et al. (\protect\hyperlink{ref-rao2021mult}{2021}).

\textbf{DNA spike-ins:}
Another possibility is to add DNA spike-ins, derived from natural or artificial sequence.
DNA spike-ins can be added the samples before DNA extraction (e.g. Smets et al. (\protect\hyperlink{ref-smets2016amet}{2016}), Tkacz, Hortala, and Poole (\protect\hyperlink{ref-tkacz2018abso}{2018}), Zemb et al. (\protect\hyperlink{ref-zemb2020abso}{2020})) or after DNA extraction (e.g. Hardwick et al. (\protect\hyperlink{ref-hardwick2018synt}{2018}), Tkacz, Hortala, and Poole (\protect\hyperlink{ref-tkacz2018abso}{2018})).
Adding spike-ins prior to extraction is thought to be preferable as it makes it possible to detect and correct for variation in DNA extraction yield among samples
(Tkacz, Hortala, and Poole (\protect\hyperlink{ref-tkacz2018abso}{2018}), Zemb et al. (\protect\hyperlink{ref-zemb2020abso}{2020}), Harrison et al. (\protect\hyperlink{ref-harrison2021theq}{2021})).
Below, we consider the distinction between pre- and post-extraction spike-ins in the light of taxonomic bias coupled with other sources of variation in extraction efficiency.

\textbf{How spike-ins are used:}
Like housekeeping species, spike-ins have been used in a variety of ways to analyze absolute abundances.
Let \(R\) (for reference) be the spike-in species and \(S\) be the native species.
Smets et al. (\protect\hyperlink{ref-smets2016amet}{2016}) used the ratio of \(S\) to \(R\) reads as an estimate of total density, which they were interested in for its own sake, though one could imagine then also using this total density estimate in community normalization (Equation \eqref{eq:density-prop-meas}).
Zemb et al. (\protect\hyperlink{ref-zemb2020abso}{2020}) used the spike-ins to measure total density from the ratio of \(S\) to \(R\) qPCR abundance estimates and then used Equation \eqref{eq:density-prop-meas}.
Stämmler et al. (\protect\hyperlink{ref-stammler2016adju}{2016}) and others used the ratio-based method of Equation \eqref{eq:density-ratio-meas}.

\hypertarget{targeted-measurements}{%
\subsection{Targeted measurements}\label{targeted-measurements}}

A variety of methods exist for targeted measurement of absolute density of a specific species (or higher-order taxon).
The most common approach is to use qPCR or ddPCR to measure the concentration of a marker gene in the extracted DNA, using primers scoped to the target taxon.
This approach is therefore subject to sources of taxonomic bias including extraction, marker-gene copy number, and primer-binding.
It is also subject to non-species-specific variation in extraction yields unless these are otherwise controlled for.
It is also possible to directly measure cell density using methods.
Some species can be measured by CFU counting after plating on selective media (REFs), and ddPCR has been used to direct measure cells (Dreo et al. (\protect\hyperlink{ref-dreo2014opti}{2014}), Morella et al. (\protect\hyperlink{ref-morella2018rapi}{2018})) and viruses (Pavšič, Žel, and Milavec (\protect\hyperlink{ref-pavsic2016digi}{2016}), Morella et al. (\protect\hyperlink{ref-morella2018rapi}{2018})) without first performing an extraction.
Species-specific florescent probes also make it possible to measure individual species via microscopy or flow cytometry (REFs).

\hypertarget{appendix-regression}{%
\section{Error in estimated regression coefficients and standard errors}\label{appendix-regression}}

Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.

Consider Equation \eqref{eq:regression} for the simple linear regression of log density of a species \(i\) on covariate \(x\).
Let \(y\) stand for the actual log density of a focal species \(i\), \(z\) stand for the log density that we've measured, and \(d = z - y\) equal the difference between the two, which here is the log efficiency of species \(i\) minus the log mean efficiency of the sample.
Let \(s_{xy}\) denote the sample covariance between variables \(x\) and \(y\), \(s^{2}_{x} = s_{xx}\) and \(s_{x}\) denote the sample variance and standard deviation, and \(r_{xy} = s_{xy}/(s_{x}s_{y})\) denote the sample correlation.

The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of \(z\) equals the sample covariance of \(z\) and \(x\) divided by the sample variance in \(x\) or, equivalently, the sample correlation of \(z\) and \(x\) multiplied by the ratio of their sample standard deviations,
\begin{align}
  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
\end{align}
From the (bi)linearity of sample covariances it follows that
\begin{align}
  \hat \beta_z 
  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
  = \hat \beta_y + \hat \beta_d,
\end{align}
where \(\hat \beta_y\) and \(\hat \beta_d\) denote the slope estimates for \(y\) and \(d\) (were these values to be known).
The absolute error in the estimate \(\hat \beta_{z}\) of \(\hat \beta_{y}\) is therefore \(\hat \beta_{d}\); it is large in a practical sense when \(\hat \beta_{d}\) is large (in absolute value) compared to \(\hat \beta_{y}\), which corresponds to the covariance of \(d\) with \(x\) being large compared to the covariance of \(y\) with \(x\).

In our case, the covariance of \(d\) equals the negative of the covariance of log mean efficiency with \(x\).
The absolute error in \(\hat \beta\) equals the negative covariance of log mean efficiency scaled by the variance in \(x\).
This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
For species that covary with \(x\) more strongly than the log mean efficiency, the error will be relatively small.
This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with \(x\) compared to the log density of the focal species.

We can similarly understand the impact of measurement error on the precision of our slope estimates.
The OLS and MLE estimated standard error in \(\hat \beta\) are both approximately
\begin{align}
  \hat{\text{se}}(\hat \beta)
  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
\end{align}
where \(s_{\hat \varepsilon}\) is the sample standard deviation of the residuals
(Wasserman (\protect\hyperlink{ref-wasserman2004allo}{2004}) Chapter 13).
The sample residuals of \(z\), \(y\), and \(d\) have a similar relationship to the regression coefficient estimates,
\begin{align}
  \hat \varepsilon_z 
    &\equiv z - \hat \beta_z x
  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
\end{align}
(Note, here I've omitted the subscript indicating the dependence on the sample.)
It follows that the sample variances of the residuals of \(z\), \(y\), and \(d\) are related through
\begin{align}
  s^2_{\hat \varepsilon_{z}} 
  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
\end{align}
The standard deviation of the \(z\) residuals is increased above that of the \(y\) residuals when the \(d\) residuals are either uncorrelated or positively correlated with the \(y\) residuals, but may be decreased when the \(y\) and \(d\) residuals are negatively correlated.

In our case, the \(d\) residuals equal the negative residuals of the log mean efficiency.
It is plausible that for most species, their residual variation will have a small covariance with log mean efficiency and the net effect of variation in the mean efficiency will be to increase the estimated standard errors, as occurs with most species in Figure \ref{fig:regression-example}.
However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \ref{fig:regression-example}.

\hypertarget{supplemental-figures}{%
\section{Supplemental figures}\label{supplemental-figures}}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files/figure-html5/momspi-mean-efficiency-fcs-1.pdf}
\caption{\label{fig:momspi-mean-efficiency-fcs}\textbf{Fold changes in the mean efficiency within and between women in the MOMS-PI study.}}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2021-10-18-evaluate-robustness-example/evaluate-robustness-example_files/figure-html5/summary_plot-1.pdf}
\caption{\label{fig:sensitivity-example}\textbf{A bias-sensivity analysis can be performed to examine how sensitive the results of a DA analysis are to assumptions about taxonomic bias in community measurements.} The figure shows the results of a bias-sensitivity analysis used to study the effect of bias on the association of \emph{Gardnerella vaginalis} and preterm birth that was investigated by Callahan et al. (\protect\hyperlink{ref-callahan2017repl}{2017}). 100 random efficiency vectors were drawn at 6 different bias strengths (quantified by the variance in log efficiency, \(\sigma_{e}^{2}\)). Each efficiency vector was used to calibrate the MGS profiles and perform a DA association test of \emph{G. vaginalis} versus the host's preterm birth outcome; regression coefficients \(\hat \beta\) indicate the increase of average logit proportion of \emph{G. vaginalis} in women who experienced preterm birth.}
\end{figure}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-brooks2015thet}{}}%
Brooks, J Paul, David J Edwards, Michael D Harwich, Maria C Rivera, Jennifer M Fettweis, Myrna G Serrano, Robert A Reris, et al. 2015. {``{The truth about metagenomics: quantifying and counteracting bias in 16S rRNA studies}.''} \emph{BMC Microbiol.} BioMed Central. \url{https://doi.org/10.1186/s12866-015-0351-6}.

\leavevmode\vadjust pre{\hypertarget{ref-callahan2017repl}{}}%
Callahan, Benjamin J, Daniel B DiGiulio, Daniela S Aliaga Goltsman, Christine L Sun, Elizabeth K Costello, Pratheepa Jeganathan, Joseph R Biggio, et al. 2017. {``{Replication and refinement of a vaginal microbial signature of preterm birth in two racially distinct cohorts of US women}.''} \emph{Proc. Natl. Acad. Sci. U. S. A.} 114 (37): 9966--71. \url{https://doi.org/10.1073/pnas.1705899114}.

\leavevmode\vadjust pre{\hypertarget{ref-chng2020meta}{}}%
Chng, Kern Rei, Tarini Shankar Ghosh, Yi Han Tan, Tannistha Nandi, Ivor Russel Lee, Amanda Hui Qi Ng, Chenhao Li, et al. 2020. {``{Metagenome-wide association analysis identifies microbial determinants of post-antibiotic ecological recovery in the gut}.''} \emph{Nat. Ecol. Evol.} 4 (9): 1256--67. \url{https://doi.org/10.1038/s41559-020-1236-0}.

\leavevmode\vadjust pre{\hypertarget{ref-contijoch2019gutm}{}}%
Contijoch, Eduardo J, Graham J Britton, Chao Yang, Ilaria Mogno, Zhihua Li, Ruby Ng, Sean R Llewellyn, et al. 2019. {``{Gut microbiota density influences host physiology and is shaped by host and microbial factors}.''} \emph{Elife} 8 (January). \url{https://doi.org/10.7554/eLife.40553}.

\leavevmode\vadjust pre{\hypertarget{ref-cruz2021equi}{}}%
Cruz, Giuliano Netto Flores, Ana Paula Christoff, and Luiz Felipe Valter de Oliveira. 2021. {``{Equivolumetric Protocol Generates Library Sizes Proportional to Total Microbial Load in 16S Amplicon Sequencing}.''} \emph{Front. Microbiol.} 12 (February): 1--16. \url{https://doi.org/10.3389/fmicb.2021.638231}.

\leavevmode\vadjust pre{\hypertarget{ref-david2014host}{}}%
David, Lawrence A, Arne C Materna, Jonathan Friedman, Maria I Campos-Baptista, Matthew C Blackburn, Allison Perrotta, Susan E Erdman, and Eric J Alm. 2014. {``{Host lifestyle affects human microbiota on daily timescales}.''} \emph{Genome Biol.} 15 (7): R89. \url{https://doi.org/10.1186/gb-2014-15-7-r89}.

\leavevmode\vadjust pre{\hypertarget{ref-diener2021nonr}{}}%
Diener, Christian, Anna C. H. Hoge, Sean M. Kearney, Ulrike Kusebauch, Sushmita Patwardhan, Robert L. Moritz, Susan E. Erdman, and Sean M. Gibbons. 2021. {``{Non-responder phenotype reveals apparent microbiome-wide antibiotic tolerance in the murine gut}.''} \emph{Commun. Biol.} 4 (1). \url{https://doi.org/10.1038/s42003-021-01841-8}.

\leavevmode\vadjust pre{\hypertarget{ref-dreo2014opti}{}}%
Dreo, Tanja, Manca Pirc, Živa Ramšak, Jernej Pavšič, Mojca Milavec, Jana Žel, and Kristina Gruden. 2014. {``{Optimising droplet digital PCR analysis approaches for detection and quantification of bacteria: a case study of fire blight and potato brown rot}.''} \emph{Anal. Bioanal. Chem.} 406 (26): 6513--28. \url{https://doi.org/10.1007/s00216-014-8084-1}.

\leavevmode\vadjust pre{\hypertarget{ref-fettweis2019thev}{}}%
Fettweis, Jennifer M., Myrna G. Serrano, Jamie Paul Brooks, David J. Edwards, Philippe H. Girerd, Hardik I. Parikh, Bernice Huang, et al. 2019. {``{The vaginal microbiome and preterm birth}.''} \emph{Nat. Med.} 25 (6): 1012--21. \url{https://doi.org/10.1038/s41591-019-0450-2}.

\leavevmode\vadjust pre{\hypertarget{ref-finucane2014atax}{}}%
Finucane, Mariel M., Thomas J. Sharpton, Timothy J. Laurent, and Katherine S. Pollard. 2014. {``{A Taxonomic Signature of Obesity in the Microbiome? Getting to the Guts of the Matter}.''} Edited by Markus M. Heimesaat. \emph{PLoS One} 9 (1): e84689. \url{https://doi.org/10.1371/journal.pone.0084689}.

\leavevmode\vadjust pre{\hypertarget{ref-galazzo2020howt}{}}%
Galazzo, Gianluca, Niels van Best, Birke J. Benedikter, Kevin Janssen, Liene Bervoets, Christel Driessen, Melissa Oomen, et al. 2020. {``{How to Count Our Microbes? The Effect of Different Quantitative Microbiome Profiling Approaches}.''} \emph{Front. Cell. Infect. Microbiol.} 10 (August). \url{https://doi.org/10.3389/fcimb.2020.00403}.

\leavevmode\vadjust pre{\hypertarget{ref-gloor2017micr}{}}%
Gloor, Gregory B., Jean M. Macklaim, Vera Pawlowsky-Glahn, and Juan J. Egozcue. 2017. {``{Microbiome Datasets Are Compositional: And This Is Not Optional}.''} \emph{Front. Microbiol.} 8 (November): 2224. \url{https://doi.org/10.3389/fmicb.2017.02224}.

\leavevmode\vadjust pre{\hypertarget{ref-hardwick2018synt}{}}%
Hardwick, Simon A., Wendy Y. Chen, Ted Wong, Bindu S. Kanakamedala, Ira W. Deveson, Sarah E. Ongley, Nadia S. Santini, et al. 2018. {``{Synthetic microbe communities provide internal reference standards for metagenome sequencing and analysis}.''} \emph{Nat. Commun.} 9 (1): 3096. \url{https://doi.org/10.1038/s41467-018-05555-0}.

\leavevmode\vadjust pre{\hypertarget{ref-harrison2021theq}{}}%
Harrison, Joshua G., W. John Calder, Bryan Shuman, and C. Alex Buerkle. 2021. {``{The quest for absolute abundance: The use of internal standards for DNA‐based community ecology}.''} \emph{Mol. Ecol. Resour.} 21 (1): 30--43. \url{https://doi.org/10.1111/1755-0998.13247}.

\leavevmode\vadjust pre{\hypertarget{ref-ji2019quan}{}}%
Ji, Brian W., Ravi U. Sheth, Purushottam D. Dixit, Yiming Huang, Andrew Kaufman, Harris H. Wang, and Dennis Vitkup. 2019. {``{Quantifying spatiotemporal variability and noise in absolute microbiota abundances using replicate sampling}.''} \emph{Nat. Methods} 16 (8): 731--36. \url{https://doi.org/10.1038/s41592-019-0467-y}.

\leavevmode\vadjust pre{\hypertarget{ref-jian2020quan}{}}%
Jian, Ching, Panu Luukkonen, Hannele Yki-Järvinen, Anne Salonen, and Katri Korpela. 2020. {``{Quantitative PCR provides a simple and accessible method for quantitative microbiota profiling}.''} Edited by Ivone Vaz-Moreira. \emph{PLoS One} 15 (1): e0227285. \url{https://doi.org/10.1371/journal.pone.0227285}.

\leavevmode\vadjust pre{\hypertarget{ref-karasov2020ther}{}}%
Karasov, Talia L., Manuela Neumann, Alejandra Duque-Jaramillo, Sonja Kersten, Ilja Bezrukov, Birgit Schröppel, Efthymia Symeonidi, et al. 2020. {``{The relationship between microbial population size and disease in the Arabidopsis thaliana phyllosphere}.''} \emph{bioRxiv}. \url{https://doi.org/10.1101/828814}.

\leavevmode\vadjust pre{\hypertarget{ref-kevorkian2018esti}{}}%
Kevorkian, Richard, Jordan T Bird, Alexander Shumaker, and Karen G Lloyd. 2018. {``{Estimating Population Turnover Rates by Relative Quantification Methods Reveals Microbial Dynamics in Marine Sediment}.''} \emph{Appl. Environ. Microbiol.} 84 (1): e01443--17. \url{https://doi.org/10.1128/AEM.01443-17}.

\leavevmode\vadjust pre{\hypertarget{ref-korpela2018inte}{}}%
Korpela, Katri, Elin W. Blakstad, Sissel J. Moltu, Kenneth Strømmen, Britt Nakstad, Arild E. Rønnestad, Kristin Brække, Per O. Iversen, Christian A. Drevon, and Willem de Vos. 2018. {``{Intestinal microbiota development and gestational age in preterm neonates}.''} \emph{Sci. Rep.} 8 (1): 1--9. \url{https://doi.org/10.1038/s41598-018-20827-x}.

\leavevmode\vadjust pre{\hypertarget{ref-kumar2018anal}{}}%
Kumar, M. Senthil, Eric V. Slud, Kwame Okrah, Stephanie C. Hicks, Sridhar Hannenhalli, and Héctor Corrada Bravo. 2018. {``{Analysis and correction of compositional bias in sparse sequencing count data}.''} \emph{BMC Genomics} 19 (1): 799. \url{https://doi.org/10.1186/s12864-018-5160-5}.

\leavevmode\vadjust pre{\hypertarget{ref-leopold2020host}{}}%
Leopold, Devin R, and Posy E Busby. 2020. {``{Host Genotype and Colonist Arrival Order Jointly Govern Plant Microbiome Composition and Function}.''} \emph{Curr. Biol.} 30 (16): 3260--3266.e5. \url{https://doi.org/10.1016/j.cub.2020.06.011}.

\leavevmode\vadjust pre{\hypertarget{ref-lloyd2020evid}{}}%
Lloyd, Karen G., Jordan T. Bird, Joy Buongiorno, Emily Deas, Richard Kevorkian, Talor Noordhoek, Jacob Rosalsky, and Taylor Roy. 2020. {``{Evidence for a Growth Zone for Deep-Subsurface Microbial Clades in Near-Surface Anoxic Sediments}.''} \emph{Appl. Environ. Microbiol.} 86 (19): 1--15. \url{https://doi.org/10.1128/AEM.00877-20}.

\leavevmode\vadjust pre{\hypertarget{ref-lozupone2013meta}{}}%
Lozupone, Catherine A, Jesse Stombaugh, Antonio Gonzalez, Gail Ackermann, Janet K Jansson, Jeffrey I Gordon, Doug Wendel, Yoshiki Va, and Rob Knight. 2013. {``{Meta-analyses of studies of the human microbiota}.''} \emph{Genome Res.}, 1704--14. \url{https://doi.org/10.1101/gr.151803.112}.

\leavevmode\vadjust pre{\hypertarget{ref-mandal2015anal}{}}%
Mandal, Siddhartha, Will Van Treuren, Richard A. White, Merete Eggesbø, Rob Knight, and Shyamal D. Peddada. 2015. {``{Analysis of composition of microbiomes: a novel method for studying microbial composition}.''} \emph{Microb. Ecol. Heal. Dis.} 26 (1): 27663. \url{https://doi.org/10.3402/mehd.v26.27663}.

\leavevmode\vadjust pre{\hypertarget{ref-martin2020mode}{}}%
Martin, Bryan D., Daniela Witten, and Amy D. Willis. 2020. {``{Modeling microbial abundances and dysbiosis with beta-binomial regression}.''} \emph{Ann. Appl. Stat.} 14 (1): 94--115. \url{https://doi.org/10.1214/19-AOAS1283}.

\leavevmode\vadjust pre{\hypertarget{ref-mclaren2019cons}{}}%
McLaren, Michael R, Amy D Willis, and Benjamin J Callahan. 2019. {``{Consistent and correctable bias in metagenomic sequencing experiments}.''} \emph{Elife} 8 (September): 46923. \url{https://doi.org/10.7554/eLife.46923}.

\leavevmode\vadjust pre{\hypertarget{ref-morella2018rapi}{}}%
Morella, Norma M., Shangyang Christopher Yang, Catherine A. Hernandez, and Britt Koskella. 2018. {``{Rapid quantification of bacteriophages and their bacterial hosts in vitro and in vivo using droplet digital PCR}.''} \emph{J. Virol. Methods} 259 (May): 18--24. \url{https://doi.org/10.1016/j.jviromet.2018.05.007}.

\leavevmode\vadjust pre{\hypertarget{ref-pavsic2016digi}{}}%
Pavšič, Jernej, Jana Žel, and Mojca Milavec. 2016. {``{Digital PCR for direct quantification of viruses without DNA extraction}.''} \emph{Anal. Bioanal. Chem.} 408 (1): 67--75. \url{https://doi.org/10.1007/s00216-015-9109-0}.

\leavevmode\vadjust pre{\hypertarget{ref-props2017abso}{}}%
Props, Ruben, Frederiek-Maarten Kerckhof, Peter Rubbens, Jo De Vrieze, Emma Hernandez Sanabria, Willem Waegeman, Pieter Monsieurs, Frederik Hammes, and Nico Boon. 2017. {``{Absolute quantification of microbial taxon abundances}.''} \emph{ISME J.} 11 (2): 584--87. \url{https://doi.org/10.1038/ismej.2016.117}.

\leavevmode\vadjust pre{\hypertarget{ref-rao2021mult}{}}%
Rao, Chitong, Katharine Z. Coyte, Wayne Bainter, Raif S. Geha, Camilia R. Martin, and Seth Rakoff-Nahoum. 2021. {``{Multi-kingdom ecological drivers of microbiota assembly in preterm infants}.''} \emph{Nature} 591 (7851): 633--38. \url{https://doi.org/10.1038/s41586-021-03241-8}.

\leavevmode\vadjust pre{\hypertarget{ref-regalado2019comb}{}}%
Regalado, Julian, Derek S. Lundberg, Oliver Deusch, Sonja Kersten, Talia Karasov, Karin Poersch, Gautam Shirsekar, and Detlef Weigel. 2020. {``{Combining whole-genome shotgun sequencing and rRNA gene amplicon analyses to improve detection of microbe--microbe interaction networks in plant leaves}.''} \emph{ISME J.}, May, 823492. \url{https://doi.org/10.1038/s41396-020-0665-8}.

\leavevmode\vadjust pre{\hypertarget{ref-smets2016amet}{}}%
Smets, Wenke, Jonathan W. Leff, Mark A. Bradford, Rebecca L. McCulley, Sarah Lebeer, and Noah Fierer. 2016. {``{A method for simultaneous measurement of soil bacterial abundances and community composition via 16S rRNA gene sequencing}.''} \emph{Soil Biol. Biochem.} 96: 145--51. \url{https://doi.org/10.1016/j.soilbio.2016.02.003}.

\leavevmode\vadjust pre{\hypertarget{ref-stammler2016adju}{}}%
Stämmler, Frank, Joachim Gläsner, Andreas Hiergeist, Ernst Holler, Daniela Weber, Peter J. Oefner, André Gessner, and Rainer Spang. 2016. {``{Adjusting microbiome profiles for differences in microbial load by spike-in bacteria}.''} \emph{Microbiome} 4 (1): 28. \url{https://doi.org/10.1186/s40168-016-0175-0}.

\leavevmode\vadjust pre{\hypertarget{ref-tettamantiboshier2020comp}{}}%
Tettamanti Boshier, Florencia A., Sujatha Srinivasan, Anthony Lopez, Noah G. Hoffman, Sean Proll, David N. Fredricks, and Joshua T. Schiffer. 2020. {``{Complementing 16S rRNA Gene Amplicon Sequencing with Total Bacterial Load To Infer Absolute Species Concentrations in the Vaginal Microbiome}.''} \emph{mSystems} 5 (2): 1--14. \url{https://doi.org/10.1128/mSystems.00777-19}.

\leavevmode\vadjust pre{\hypertarget{ref-tkacz2018abso}{}}%
Tkacz, Andrzej, Marion Hortala, and Philip S. Poole. 2018. {``{Absolute quantitation of microbiota abundance in environmental samples}.''} \emph{Microbiome} 6 (1): 110. \url{https://doi.org/10.1186/s40168-018-0491-7}.

\leavevmode\vadjust pre{\hypertarget{ref-vandeputte2017quan}{}}%
Vandeputte, Doris, Gunter Kathagen, Kevin D'hoe, Sara Vieira-Silva, Mireia Valles-Colomer, João Sabino, Jun Wang, et al. 2017. {``{Quantitative microbiome profiling links gut community variation to microbial load}.''} \emph{Nature} 551 (7681): 507. \url{https://doi.org/10.1038/nature24460}.

\leavevmode\vadjust pre{\hypertarget{ref-vaninsberghe2020diar}{}}%
VanInsberghe, David, Joseph A. Elsherbini, Bernard Varian, Theofilos Poutahidis, Susan Erdman, and Martin F. Polz. 2020. {``{Diarrhoeal events can trigger long-term Clostridium difficile colonization with recurrent blooms}.''} \emph{Nat. Microbiol.} 5 (4): 642--50. \url{https://doi.org/10.1038/s41564-020-0668-2}.

\leavevmode\vadjust pre{\hypertarget{ref-wallace2021thed}{}}%
Wallace, Megan A., Kelsey A. Coffman, Clément Gilbert, Sanjana Ravindran, Gregory F. Albery, Jessica Abbott, Eliza Argyridou, et al. 2021. {``{The discovery, distribution, and diversity of DNA viruses associated with Drosophila melanogaster in Europe}.''} \emph{Virus Evol.} 7 (1): 1--23. \url{https://doi.org/10.1093/ve/veab031}.

\leavevmode\vadjust pre{\hypertarget{ref-wang2021curr}{}}%
Wang, Xiaofan, Samantha Howe, Feilong Deng, and Jiangchao Zhao. 2021. {``{Current Applications of Absolute Bacterial Quantification in Microbiome Studies and Decision-Making Regarding Different Biological Questions}.''} \emph{Microorganisms} 9 (9): 1797. \url{https://doi.org/10.3390/microorganisms9091797}.

\leavevmode\vadjust pre{\hypertarget{ref-wasserman2004allo}{}}%
Wasserman, Larry. 2004. \emph{{All of Statistics}}. Springer Texts in Statistics. New York, NY: Springer New York. \url{https://doi.org/10.1007/978-0-387-21736-9}.

\leavevmode\vadjust pre{\hypertarget{ref-yeh2018taxo}{}}%
Yeh, Yi-Chun, David M. Needham, Ella T. Sieradzki, and Jed A. Fuhrman. 2018. {``{Taxon Disappearance from Microbiome Analysis Reinforces the Value of Mock Communities as a Standard in Every Sequencing Run}.''} \emph{mSystems} 3 (3): e00023--18. \url{https://doi.org/10.1128/mSystems.00023-18}.

\leavevmode\vadjust pre{\hypertarget{ref-zemb2020abso}{}}%
Zemb, Olivier, Caroline S Achard, Jerome Hamelin, Marie‐Léa De Almeida, Béatrice Gabinaud, Laurent Cauquil, Lisanne M. G. Verschuren, and Jean-jacques Godon. 2020. {``{Absolute quantitation of microbes using 16S rRNA gene metabarcoding: A rapid normalization of relative abundances by quantitative PCR targeting a 16S rRNA gene spike‐in standard}.''} \emph{Microbiologyopen} 9 (3): 1--21. \url{https://doi.org/10.1002/mbo3.977}.

\end{CSLReferences}

\end{document}

### 2021-12-02 appendix.Rmd
Rename implications section

@@
 For simplicity, we ignore this random variation when considering the impact of taxonomic bias on MGS measurement.
-The effect of random variation in regression analysis is addressed in Appendix
-\@ref(appendix-regression) and the effect of random counting error is discussed in Section \@ref(implications).
+The effect of random variation in regression analysis is addressed in Appendix \@ref(appendix-regression) and the effect of random counting error is discussed in Section \@ref(case-studies).
 

### 2021-12-02 solutions.Rmd
Rename implications section

@@
 We imputed the efficiencies of other species using taxonomic relationships (Methods).
-Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \@ref(implications).
+Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \@ref(case-studies).
 

### 2021-12-08 case-studies.Rmd
Rename implications section
Draft case studies

--- 2021-11-10 real-world-implications.Rmd
@@
-# Implications for real-world inference {#implications}
+# Case studies  {#case-studies}
 
-Are biologically significant errors in DA analyses likely in practice?
-We can begin to answer this question using a combination of theoretical analysis, consideration of hypothetical scenarios, and analysis of case studies for which some information about taxonomic bias is available.
+To better understand the potential impact of taxonomic bias on DA analysis in practice, we conducted several case studies spanning a range of biological scenarios and sequencing technologies.
 
-## Systematic error in slope or LFC estimates
+## Foliar fungi experiment
 
-Recall from Section \@ref(differential-abundance) that the error in the slope or LFC in a DA regression analysis is proportional to the covariance of the log mean efficiency with the covariate.
-This covariance can be split into two components: the standard deviation of (log) mean efficiency and its correlation with the covariate.
-Hence when considering whether taxonomic bias creates large error, it can be useful to separately ask whether the mean efficiency is likely to vary across samples and whether it is likely to be correlated with a covariate of interest.
+To study the interactions between a host plant and its fungal symbionts and pathogen, @leopold2020host inoculated plants with 8 fungal species and, after allowing these fungal microbiomes to establish, exposed plants to a fungal pathogen.
+To investigate the role of colonization order and host genotype, plants varied in source location and in which symbiont was allowed to colonize first.
+The authors used ITS amplicon sequencing to measure communities before and after pathogen infection, along with mock communities with known genome concentrations of DNA from the 9 species.
+These DNA mock communities allowed the authors to directly measure the bias due to PCR and ribosomal copy-number-variation (CNV) following the method of @mclaren2019cons and correct for it when analyzing the primary experimental samples.
+The authors found a 13X difference between the most and least efficiently measured symbiont, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
+We reanalyzed this experiment to understand the impact that bias might have had on the study's analyses were it not accounted for (SI Analysis Fungi).
+
+@leopold2020host used the ITS sequencing profiles of the pre-infection samples to understand the impact of host genotype and an experimental treatment---which species was allowed to colonize first---on the community composition after all species were later allowed to colonize.
+The authors used linear regression of log species proportions, using negative-binomial regression to account for random read-sampling error, to ask whether host genotype, treatment, and their interaction significantly impacted overall log species proportions and to quantify the advantage of being allowed to colonize first (priority effects).
+Bias was accounted for via the use of species- and sample-specific offset terms in the regression model (SI / appendix).
+In principle, bias could significantly impact the results of such regressions through variation in the log mean efficiency among samples.
+Yet bias has a negligible impact on the study's analyses, with nearly identical DA results being obtained regardless of whether bias correction is used (Ref SI, which will run the analyses from the paper with and without bias correction).
+To understand why, we compared the variation in log species proportions across samples to that in log mean efficiency (Figure X).
+Individual species show substantial variation in log proportion (give numbers); nevertheless, the log mean efficiency hardly varies (number; perhaps ref a figure).
+This difference is due to the communities always being dominated by just three species, whose efficiencies happen to be the highest among the commensal species and only vary by 3X (and just 1.5X between the two most dominant).
+The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions $\ll 1$, even after bias correction) and hence don't induce significant variation in the mean efficiency.
+Due to the stable mean efficiency across samples, bias primarily impacts the intercept terms and has a negligible impact on the estimates and standard errors of slope (DA) terms in the regressions.
+
+@leopold2020host restricted their analysis to ITS profiles form the pre-infection samples, as reads from post-infection samples were dominated by the pathogen.
+<!-- (due to a combination of pathogen growth and, presumably, its much greater efficiency). -->
+However, we can use the post-infection samples to ask what the impact of the pathogen's much greater efficiency would have were community sequencing used to understand the impact of pathogen invasion on the already-colonized commensal species without accounting for bias.
+n particular, suppose the ITS sequencing data were coupled with an accurate measurement of total genome copy number using Equation \@ref(eq:density-prop-meas) to predict absolute genome copies of each species.
+The high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples; this increase would tend to lead to a concommittant 6X decrease in the fold changes in genome copies.
+
+TODO:
+
+- Give e.g. for the lowest eff species, Penicillium being ~6X too low.
+- Check that correlations aren't an important part of the story as well
+- Clarify that could still impact standard error of rare low eff species due to zeros
+- Analysis: re-run DL analysis with and without calibrated offsets
+
+## Vaginal microbiomes of pregant women
+
+(Has not been updated)
 
@@
 
-A second study in which control communities make it possible to directly analyze the impact of taxonomic bias in experimental samples was performed by @leopold2020host.
-To study the interactions between a host plant and its fungal commensals and pathogen, the authors inoculated trees with a fungal synthetic community and later exposed plants to a fungal pathogen.
-The authors used ITS amplicon sequencing to measure communities before and after infection, along with mock communities which they used to estimate the species efficiencies with the method of @mclaren2019cons.
-(DNA mocks were used, so that bias due to DNA extraction is not included, but other major bias sources such as PCR and ribosomal copy-number-variation are included.) 
-We reanalyzed this experiment to consider the impact that bias has on different possible DA analyses (SI Analysis Fungi).
-As noted by @leopold2020host, there was a 13X difference between the most and least efficiently measured commensals, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
-Yet despite this 13X variation among commensals, the mean efficiency is remarkably stable across communities prior to pathogen colonization.
-Consequently, taxonomic bias creates little systematic error in proportion-based DA analysis of these pre-colonization samples.
-The high efficiency of the pathogen did, however, typically lead to a roughly 6X increase in the mean efficiency of post-colonization samples.
-Hence DA analysis of how commensals responded to the pathogen did not first discard pathogen reads would therefore tend to suggest that the commensals had substantially smaller fold changes than they actually did, unless bias were accounted for.
+Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate.
+The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
+For example, in the vaginal microbiome a decline in _Lactobacillus_ and rise of species including _Gardnerella vaginalis_ is associated with preterm birth in pregnant women; hence the variation in mean efficiency driven by these species may also create an association of mean efficiency with preterm status.
+
+## Gut microbiome
+
+(Has not been updated)
 
@@
 
-Systematic error in DA results requires that the variation in log mean efficiency is correlated with the covariate.
+Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate.
 The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
-For example, in the vaginal microbiome a decline in _Lactobacillus_ and rise of species including _Gardnerella vaginalis_ is associated with preterm birth in pregnant women; hence the variation in mean efficiency driven by these species may also create an association of mean efficiency with preterm status.
-In the plant-fungal experiment of @leopold2020host, the increase in pathogen proportion after infection leads to a large increase in the mean efficiency over time, which systematic increases the observed decline in the proportions of commensal species (SI Analysis).
 The Bacteroidetes-to-Firmicutes ratio has also been linked to host conditions.
@@
 
-<!-- General mechanisms -->
+## Marine sediment growers
 
-Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
-One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
-A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
-Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
-Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
-In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
-For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.
+As sediment is slowly buried, the cencentration of oxygen and nutrients decreases, and cell density decreases; @lloyd2020evid asked whether some taxa are nevertheless able to continue to grow and increase in cell density.
+To measure density with depth, they used community 16S sequencing coupled with measurmeents of total community density using CARD-FISH, which they combined using community-normalization (Equation X).
+(The authors refer to these absolute density measurements as FRAxC measurements.)
+The authors assessed whether density of bacterial and archaeal families increased with depth, using linear regression on log cell density to infer the exponential growth (or decay) rate over the first 3 cm below the bioirrigation layer (corresponding to $\sim 8$ years of burial time).
+(Per time after converting depth -> time)
+The authors did this for two cores, and checked the inferences from these sequencing-based measurements against targeted measurements (qPCR) of 16S concentration for particular taxa.
+They further checked w/ sequencing-based growth-rate inference from two laboratory incubations.
+
+The authors attempted to explictly account for bias by arguing that a consistent taxonomic bias would cancel in the observed log-fold-changes that determine the inferred exponential growth rates (ref kevorkian2018 as well); however, their calculations neglected the impact of changes in mean efficiency which do not cancel and may lead to spurious conclusions (previous section).
+We were therefore interested in whether the unaccounted for error caused by bias might have led to spurious inferences about which species were increasing with depth.
+Supposing the CARD-FISH measurements accurately measured total density, our results from the previous section indicate that a systematic error in growth rates would arise if the log mean efficiency systematically changed with depth.
+Diversity is sufficiently high that this systematic error is unlikely to be driven by just a few species, as in the vaginal MB example and leopold pathogen case.
+But it is possible that the covariate (burial time) selects directly on a species efficiency as well as its ability to grow. 
+In particular, we considered that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
+In this case, we would expect the the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that are just persisting or even slowly dying off would be mistakenly inferred to have positive growth.
+
+We can test the hypothesis that systematic variation in log mean efficiency with depth distorts inferred growth rates using the qPCR measurements of two clades also collected by @lloyd2020evid.
+qPCR measurements of specific clades are also subject to taxonomic bias, but not the compositional constraint of sequencing, or at least are expected to experience it to a much weaker extent (Section \@ref(review-absolute-methods)).
+Hence we expect the systematic fold error in qPCR measurements to be approximately constant across samples and for targeted qPCR measurements to yield relatively accurate estimates of growth rates.
+Comparing qPCR to FRAxC growth rates on the same taxa allows us to estimate the systematic error in the FRAxC growth rates, equal to the rate of change of log mean efficiency with depth.
+Because the same error applies to all taxa, these targeted comparisons inform us about the impact of bias on the growth rates of all taxa.
+
+We compared the growth rates for taxa with both FRAxC and qPCR values for the two sediment cores, as reported in Table 1.
+The first soil core included qPCR measurements of a single archaeal clade, _Bathyarchaeota_, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of 0.099/yr by FRAxC and 0.097/yr by qPCR).
+The second soil core included qPCR measurements of _Bathyarchaeota_ and a second clade, _Thermoprofundales_/MBG-D.
+In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by 0.012/yr for _Bathyarchaeota_ (0.112/yr by FRAxC and 0.100/yr by qPCR) and by 0.086/yr for _Thermoprofundales_/MBG-D (0.294/yr by FRAxC and 0.208/yr by qPCR).
+A low number of experimental samples and noise in both the FRAxC and qPCR measurements place signficant uncertainty in these measurements; however, the fact that FRAxC-derived growth rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreases with depth in a manner that systematically biases FRAxC-derived rates to higher values.
+The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr is signficant for this low nutrient environment, where growth rates of the identified positive growers ranged between 0.04/yr and 0.5/yr.
+Overall, the comparison between FRAxC and qPCR measurements gives support to the study conclusions, but suggest that species at the lower end of this range of positive FRAxC-derived rates may in fact be merely persisting or even slowly declining in abundance.
+
+Caveats that might be instructive to mention
+
+- qPCR and seq analysis are both at ~family level; this is prob not a big deal when it comes to inferring the systematic varation in mean efficiency though, since qPCR and 16S seq share a lot of bias
+- relic dna
+- possibility of non-genetic variation -> change in efficiency with depth. This could still lead to a change in mean efficiency, but would offset the effect for these particular species.
+- possible error/bias in CARD-FISH
+
+### old text
 
@@
 The small difference between these slopes suggests that there is little correlation of log mean efficiency with burial time and hence little systematic error in the estimated doubling times derived from the proportion-based density estimates.
+
+Cut from previous version
+
+In Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years (CHECK) using the proportion-based densities and 10.3 years by qPCR (Lloyd et al.
+(2020) Table 1 and Figure 4).
+These doubling times correspond to highly similar estimated slopes of log density per year of 1 / 10.1 ≈ 0.99 and 1 / 10.3 ≈ 0.97 .
+The difference between these slopes, ≈ 0.02, estimates the slope of the mean efficiency.
+It’s small value suggests little systematic variation in mean efficiency with burial time and hence little systematic bias in the estimated doubling times derived from the proportion-based density estimates.
+In this case, close agreement in the DA results across methods for a specific taxon suggests that variation in sample mean efficiency did not significantly distort the MGS-based results and hence provides a degree of validation for all taxa, not just the taxa with targeted measurements.
+(An important caveat for this example is that this analysis assumes consistent efficiencies at the higher taxonomic levels used for qPCR measurement and MGS analysis performed by Lloyd et al.
+(2020).)
+
+Here we compared summary statistics; however, in principle a joint statistical analysis of targeted and proportion-based density measurements would provide more statistical power to estimate and correct the effects of variation in the mean efficiency across samples and provide calibrated fold change and regression estimates for all taxa [APPENDIX]
+
+## conclusions
+
+Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
+One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
+A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
+Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
+Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
+In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
+For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.
+
+Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.
+
+- Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
+
+# Old
+
+## Systematic error in slope or LFC estimates
+
+Recall from Section \@ref(differential-abundance) that the error in the slope or LFC in a DA regression analysis is proportional to the covariance of the log mean efficiency with the covariate.
+This covariance can be split into two components: the standard deviation of (log) mean efficiency and its correlation with the covariate.
+Hence when considering whether taxonomic bias creates large error, it can be useful to separately ask whether the mean efficiency is likely to vary across samples and whether it is likely to be correlated with a covariate of interest.
+
+<!-- General mechanisms -->
+
+Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
+One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
+A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
+Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
+Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
+In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
+For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.
 

### 2021-12-14 notebook/_posts/2022-01-06-leopold2020host-t1-samples/leopold2020host-t1-samples.Rmd
Analyze leopold2020host T1 samples

---
title: "Explore variation among T1 samples in Leopold and Busby (2020)"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - R
  - ref:leopold2020host
  - differential abundance
date: 2022-01-06
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    toc: true
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

Currently this file analyses variation in the T1 samples.
We may want to include the T2 analysis as well.
The purpose is to understand how variation in the mean efficiency might impact DA analyses based on proportions.

# Setup

```{r libraries}
library(here)
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
# ggplot helpers
library(ggbeeswarm)
library(ggdist)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot())
# stats helpers
library(broom)

library(metacal)

library(rstanarm)
options(mc.cores = 4)
```

```{r load-data, code_folding = TRUE}
## From the metacal 2.0 tutorial
# Download data from https://zenodo.org/record/3872145
data_path <- here::here("notebook/_data", "leopold2020host")
# To use a temporary directory:
# data_path <- file.path(tempdir(), "leopold2020")
if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  download.file(
    "https://zenodo.org/record/3872145/files/dleopold/Populus_priorityEffects-v1.2.zip",
    file.path(data_path, "Populus_priorityEffects-v1.2.zip")
  )
  unzip(
    file.path(data_path, "Populus_priorityEffects-v1.2.zip"), 
    exdir = data_path
  )
}

mock_actual <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/data/MockCommunities.csv") %>%
  read.csv(row.names = 1) %>%
  select(-Sym4) %>%
  as("matrix") %>%
  otu_table(taxa_are_rows = FALSE) %>%
  transform_sample_counts(function(x) close_elts(1 / x))
mock_taxa <- taxa_names(mock_actual)
commensal_taxa <- setdiff(mock_taxa, 'Melampsora')

# Rust severity
rust <- file.path(data_path, 'dleopold-Populus_priorityEffects-8594f7c',
  'data/rust_measurements.csv'
) %>%
  read_csv %>%
  select(-1)
# summarize following approach from leopold 
rust_summary <- rust %>%
  janitor::clean_names() %>%
  rename(plant_id = samp_id) %>%
  with_groups(plant_id, summarize, 
    # across(ends_with('cm2'), mean)
    rust_pct = sum(rust_cm2)/sum(leaf_cm2),
    lesion_pct = sum(lesion_cm2) / sum(leaf_cm2),
    leaf_area = sum(leaf_cm2),
    leaf_n = n()
  )

#> The microbiome data is stored in a phyloseq object,
ps <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/output/compiled/phy.rds") %>%
  readRDS %>%
  filter_tax_table(.otu %in% mock_taxa) %>%
  mutate_sample_data(., 
    Timepoint = factor(Timepoint),
    sample_sum = sample_sums(.)
  ) %>%
  mutate_sample_data(
    plant_id = str_extract(.sample, 'G[0-9]+\\.T[0-9]+\\.R[0-9]+')
  ) %>%
  left_join_sample_data(
    rust_summary, by = 'plant_id'
  ) %>%
  rename_with_sample_data(janitor::make_clean_names)

sam <- sample_data(ps) %>% as("data.frame") %>% as_tibble(rownames = "Sample")
tax <- tax_table(ps) %>% as("matrix") %>% as_tibble(rownames = "Taxon")
ps_mock <- ps %>% 
  subset_samples(samp_type == "Mock") %>%
  prune_taxa(mock_taxa, .)
```

```{r}
```

Check sample sums,

```{r}
sam %>%
  as_tibble %>%
  ggplot(aes(sample_sum, fill = samp_type)) +
  scale_x_log10() +
  geom_histogram() +
  scale_fill_brewer(type = 'qual')
```

```{r}
sam %>%
  filter(sample_sum < 3e3) %>%
  select(Sample, timepoint, treatment, sample_sum)
```

The sample with less than 3000 reads is in the 'Negative' treatment group and so will be ignored in our analysis of the T1 samples.

## Bias estimation

We estimate bias using `metacal::estimate_bias()`; see the [metacal tutorial](https://mikemc.github.io/metacal/articles/tutorial.html) for more information.
@leopold2020host identified the sample Mock.5 as an outlier sample and exluded it when estimating bias.
One taxon in this sample, Epicoccum, was undetected, and including it leads to a larger standard error especially for this taxon (as seen in the metacal tutorial).
We could improve our bias estimates by setting this taxon's nomial abundance to 0 in this sample; however, let's just drop this sample to maintain consistent bias estimates with those used in the original study.

```{r estimate-bias}
control_samples <- ps_mock %>% sample_names %>% setdiff('Mock.5')
mc_fit <- estimate_bias(
  ps_mock %>% prune_samples(control_samples, .), 
  mock_actual %>% prune_samples(control_samples, .), 
  boot = TRUE
)
bias <- coef(mc_fit) %>% print(digits = 2)
```

```{r, fig.dim = c(5,3)}
mc_fit_summary <- summary(mc_fit)
coef_tb <- mc_fit_summary$coefficients
coef_tb %>%
  mutate(taxon = fct_reorder(taxon, estimate)) %>%
  ggplot(aes(taxon, estimate, 
      ymin = estimate / gm_se^2, ymax = estimate * gm_se^2)) +
  geom_hline(yintercept = 1, color = "grey") +
  geom_pointrange() +
  scale_y_log10() +
  coord_flip()
```

# T1 samples

From our previous analyses, we know that there is little impact of bias correction on the primary analyses done on these samples, and also that there is fairly little variation in the mean efficiency in the T1 samples.
Here we further explore the variation in both the species proportions and the mean effiency to more fully understand why bias has little impact on DA analysis.

NOTES

- observation to interrogate: small variation in mean efficiency. Why? is it because the proportions don't change much? or because they change in ways that offset each other (e.g. a high eff sp and a low eff species increase)?
- DA analysis: effect of treatment, and effect of host region and/or genotype; do a single multiple regression with all of these, and what happens to mean efficiency. then compare to coefficients of individual species.
<!--  -->

## Setup

Here we subset to the T1 experimental samples and the commensal taxa.
The analysis of T1 samples by @leopold2020host first subsets to just the 8 commensal (non-pathogenic) inoculum taxa, and we follow suit here.
This step excludes Melampsora reads; since these samples are collected prior to pathogen inoculation, these reads are likely from cross contamination.
It also excludes OTUs not identified with the 9 inoculum taxa.
In addition, we add the sample sums and mean efficiency to the sample data, after subsetting to the commensals.

```{r}
# Bias among the commensals; for convenience, set relative to max
bias_t1 <- bias[names(bias) != 'Melampsora']
bias_t1 <- bias_t1 / max(bias_t1)
most_efficient_taxon <- which.max(bias_t1) %>% names
stopifnot(most_efficient_taxon == 'Fusarium')

ps_t1 <- ps %>%
  filter_sample_data(
    samp_type == "Experiment", 
    timepoint == 1, 
    treatment != 'Negative'
  ) %>%
  filter_tax_table(.otu %in% commensal_taxa) %>%
  mutate_sample_data(., 
    sample_sum = sample_sums(.),
    mean_efficiency = transform_sample_counts(., close_elts) %>%
      perturb(bias_t1, norm = 'none') %>% 
      sample_sums,
  )
sam_t1 <- ps_t1 %>% sample_data %>% as_tibble
```

A small fraction (`r ps_t1 %>% otu_table %>% as('matrix') %>% {mean(. == 0)} %>% signif(digits = 3)`) of the observed counts are zero. 
For calculations that ignore the counting error, we want to ensure there are only positive abundances.
We will replace the zeros using `zCompositions::cmultRepl()` following @leopold2020host.

```{r}
stopifnot(taxa_are_rows(ps_t1) == FALSE)
otu_new <- ps_t1 %>% 
  otu_table %>% 
  as('matrix') %>%
  zCompositions::cmultRepl(
    method = 'GBM', 
    output = 'p-counts', 
    suppress.print = TRUE
  )
ps_t1_pos <- ps_t1
otu_table(ps_t1_pos) <- otu_table(otu_new, taxa_are_rows = FALSE)
rm(otu_new)
```

TODO: consider using the approach from the MOMSPI dataset instead
TODO: recompute the mean efficiency with the imputed phyloseq object (shouldn't impact results)

Let's create a data frame with the observed and calibrated proportions of each species, with their efficiencies added in.

```{r}
props <- ps_t1_pos %>%
  as_tibble %>%
  left_join(bias_t1 %>% enframe('.otu', 'efficiency'),
    by = '.otu'
  ) %>%
  with_groups(.sample, mutate, 
    prop_obs = close_elts(.abundance),
    prop_cal = close_elts(.abundance / efficiency),
  ) %>%
  mutate(
    .otu_fct = fct_reorder(.otu, efficiency)
  )
```

## Variation in mean efficiency

Let's examine the variation in mean efficiency

```{r, fig.dim = c(7,4)}
lyrs <- list(
  geom_histogram(),
  scale_x_log10(),
  expand_limits(x = 1.2),
  scale_fill_brewer(type = "qual"),
  labs(
    x = "Mean efficiency", 
    y = "Number of samples",
    # fill = ''
  )
)

p1 <- sam_t1 %>%
  ggplot(aes(mean_efficiency, fill = region)) +
  lyrs
p2 <- sam_t1 %>%
  ggplot(aes(mean_efficiency, fill = treatment)) +
  lyrs
p1 / p2
```

The relative efficiency has a geometric range of `r bias_t1 %>% gm_range %>% signif(3)`-fold and a geometric standard deviation of `r bias_t1 %>% gm_sd %>% signif(3)`-fold.
But the mean efficiency's geometric range is only `r sam_t1$mean_efficiency %>% gm_range %>% signif(3)`-fold and its geometric standard deviation is just `r sam_t1$mean_efficiency %>% gm_sd %>% signif(3)`-fold.

## Variation in proportions vs mean efficiency

First let's create some summary statistics of the calibrated proportions, to compare to those of the mean efficiency.

```{r}
props %>%
  with_groups(.otu_fct, summarize,
    across(prop_cal, 
      list(gm_mean = gm_mean, gm_sd = gm_sd, gm_range = gm_range),
      .names = '{.fn}'
    )
  ) %>%
  arrange(desc(.otu_fct)) %>%
  mutate(across(-.otu_fct, signif, 3)) %>%
  knitr::kable()
# TODO: adjust formatting to make sig fig's clearer; consider using formatC
```

Now let's create the main figure showing the varation in taxa proportions against that in the mean efficiency, with the individual taxa efficiencies also shown for more context.

Caption: **In the pre-infection samples from @leopold2020host, multiplicative variation in taxa proportions is much larger than that in the mean efficiency.**
Panel A shows the distribution of the proportions of each commensal isolate (denoted by its genus) across all samples collected prior to pathogen inoculation;
Panel C shows the distribution of the (estimated) sample mean efficiency across these same samples on the same scale; 
and Panel B shows the efficiency of each taxon estimated from DNA mock communities as point estimates and 90% bootstrap percentile confidence intervals.
Efficiencies are shown relative to the most efficiently measured taxon (Fusarium).

```{r}
rng <- props %>% pull(prop_cal) %>% range
## species proportions
p_props <- props %>%
  ggplot(aes(y = .otu_fct, x = prop_cal)) +
  scale_x_log10() +
  stat_slab() +
  labs(x = 'Proportion', y = NULL)
## mean efficiency
p_me <- sam_t1 %>%
  ggplot(aes(y = 'Mean efficiency', x = mean_efficiency)) +
  scale_x_log10() +
  expand_limits(x = rng) +
  stat_slab() +
  labs(x = 'Efficiency relative to Fusarium', y = NULL)
# p_me / p_props + plot_layout(heights = c(0.1, 1))
## Estimated efficiency with bootstrap CIs
x <- mc_fit$bootreps %>%
  data.table::as.data.table(keep.rownames = '.id') %>%
  select(.id, all_of(commensal_taxa)) %>%
  mutate(across(-.id, ~. / Fusarium)) %>%
  pivot_longer(-.id, names_to = 'taxon', values_to = 'efficiency') 
p_eff <- x %>%
  mutate(taxon = fct_reorder(taxon, efficiency)) %>%
  ggplot(aes(y = taxon, x = efficiency)) +
  stat_pointinterval(.width = 0.9) +
  scale_x_log10() +
  labs(x = str_glue('Efficiency relative to Fusarium'), y = NULL)
```

```{r variation-in-proportions-and-mean-efficiency, fig.dim = c(8.5, 6.5)}
p_props + p_eff + p_me + plot_spacer() + 
  plot_layout(heights = c(1, 0.15), widths = c(1, 0.8), ncol = 2) +
  plot_annotation(tag_levels = 'A')
```

TODO: consider a different zero-replacement strategy that would perhaps not create bimodal distributions in Trichoderma and Penicillium. This would let us show a zoomed in scale. Could also consider plotting the efficiencies on the same scale, to show that the mean efficiency is less, or add annotated lines marking the efficiencies as I did with one leopold plot previously.
TODO: see if I can get the y axis ticks to line up better (bonus)

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2021-12-17 case-studies.Rmd
Revise vaginal case study

@@
 n particular, suppose the ITS sequencing data were coupled with an accurate measurement of total genome copy number using Equation \@ref(eq:density-prop-meas) to predict absolute genome copies of each species.
-The high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples; this increase would tend to lead to a concommittant 6X decrease in the fold changes in genome copies.
+The high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples; this increase would tend to lead to a concomitant 6X decrease in the fold changes in genome copies.
 
@@
 
-- Give e.g. for the lowest eff species, Penicillium being ~6X too low.
-- Check that correlations aren't an important part of the story as well
-- Clarify that could still impact standard error of rare low eff species due to zeros
+- Give numerical e.g. for the lowest eff species, Penicillium
+- Check that strength of correlation with covariates isn't an important part of the story as well
 - Analysis: re-run DL analysis with and without calibrated offsets
@@
 
-(Has not been updated)
+### Outline
 
-One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
-To characterize bias in a 16S sequencing protocol developed for the Vaginal Human Microbiome Project (VaHMP), @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome.
-This protocol was later used for the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study (@fettweis2019thev).
-We used the method for measuring and correcting bias developed by @mclaren2019cons to measure bias from the mock communities and analyze its potential impact on DA analyses that use the MOMSPI measurements (SI Analysis MOMS-PI).
-The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
-The three species that most often dominate the calibrated profiles are _Gardnerella vaginalis_, _Lactobacillus iners_, and _Lactobacillus crispatus_.
-Our mock-community analysis indicate _G. vaginalis_ has the lowest efficiency of the seven species, roughly 20X lower than _L. crispatus_ and 30X lower than _L. iners_ (the most efficiently measured).
-As a result, the mean efficiency varies substantially across samples, with samples dominated by a _Lactobacillus_ species typically having an efficiency that is 3-20X greater than samples dominated by _Gardnerella_ (Figure \@ref(fig:momspi-mean-efficiency-dist)).
-Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
+The vaginal microbiomes of pregnant women have been a source of intensive study due to its perceived connection with the health of both mother and her developing child.
+Many MGS studies have found associations of specific microbial species and community characteristics with rates of urinary tract and sexually-transmitted infections, bacterial vaginosis (BV), and preterm births.
+Yet these associations and the baseline, apparently healthy, community state vary across studies of different populations and using different MGS methods.
+A number of studies have experimentally probed the taxonomic bias of different MGS protocols on vaginally-associated species (@yuan2012eval,@brooks2015thet,@gill2016eval,@graspeuntner2018sele,todo ref on taxonomic resolution and also sampling methods).
+These studies have revealed substantial bias among protocols, including variation in extraction efficiency by 10-40X and an inability for certain primer sets to amplify highly-prevalent species.
+Taxonomic bias has been proposed as a potential explanation for discrepancies across studies (@callahan2017repl, Others), there has so far been little quantitative analysis of this potential explanation.
+Here we re-analyze data from published studies to examine the potential impact of taxonomic bias on DA analysis of vaginal communities of pregnant women.
+
+- P on brooks, momspi studies + calibration
+  - GC
+  - Intro the studies
+  - intro ability to use calibration to look at impact of bias - perhaps don't explain as such
+  - how we did it given limited species composition of the mock
+
+We first examine what the taxonomic bias observed in control experiments by @brooks2015thet might imply for DA analysis of vaginal microbiomes of pregnant women reported by @fettweis2019thev.
+As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, @fettweis2019thev collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were measured by amplicon sequencing of the 16S V1-V3 hypervariable region to yield species-level taxonomic profiles.
+The effect of taxonomic bias in the 16S protocol had previously been investigated by @brooks2015thet via the construction and measurement of cellular mock communities of seven vaginally-relevant bacterial species.
+We reanalyzed these data in @mclaren2019cons using a new approach that makes it possible to measure the relative efficiencies among species from mock community measurements.
+Once relative efficiencies of all species are estimated, it is possible to quantitatively examine the impact of bias and to _calibrate_, or correct the effect of bias, in samples with unknown composition (@mclaren2019cons).
+To account for species not in the mock communities, we used a simple method to impute the relative efficiencies of remaining species using their taxonomic relationship to the mock species.
+
+maybe say more about this
+- Such imputation carries with it substantial uncertainty; however, 
+- These species collectively constitute a substantial majority of taxomically assigned reads in the MOMSPI 16S profiles (X for species level, and Y when expand to genus).
+- analysis of sensitivty to uncertainty can be done and shows not a very big impact.
+
+Efficiency varied subsantially among control species in the @brooks2015thet experiment, with an approximately 30X advantage of the highest efficiency _L.
+iners_ over the lowest efficiency _G.
+vaginalis_.
+To assess the impact of bias on the observed fold changes in species proportions, we used our imputed efficiencies to consider how the mean efficiency varied across women.
+The mean efficiency varies substantially across samples (SI Figure \@ref(fig:momspi-mean-efficiency-fcs; SUMMARY STATS), with greater variation between samples from different women than samples from the same woman (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
+Vaginal microbiomes are often dominated by a single species often forming a vast majority of calibrated or uncalibrated read counts, with _Lactobacillus crispatus_, _Lactobacillus iners_, _G.
+vaginalis_, and BVAB1 being the most common (CHECK).
+Consequently, the mean efficiency tends to track that of most dominant species, with _Lactobacillus_-dominated samples typically having an efficiency that is 3-20X greater _Gardnerella_-dominated samples (Figure \@ref(fig:momspi-mean-efficiency-dist)).
+Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)), and typically result in substantial fold changes in mean efficiency.
 Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure TODO).
-Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
+In contrast, bias-correction only affects the baseline value of species ratios and not the fold changes.
+
+TODO:
+- consider whether the error is significant, which depends on the species FC and not just the mean efficiency FC.
+    - would be good to somehow visualize or summarize this.
+    - should give a better interpretation/assessment of the bottom line here.
+- to show that variation within women is lower, compute standard deviation of log mena efficiency within women and average this (perhaps weighted by n)
+- create 1-2 trajectories to show
+- also show ratios
+- writing/main issue: Consider how often the error matters, in FCs, and make trajectories, - make sure can support the idea that it usually doesn't matter, if that is true.
 
@@
 <!-- end figure -->
+
+- P: Simulated DA analysis
+    - GC: We next sought to see how ...
+    - Methods
+    - Results
+    - Discussion/caveats
+    - notes: could do with fido as well, to show that the results are unaffected.
+
+We next considered whether the variation in mean efficiency could cause systematic error in a DA analysis of species proportions.
+Due to privacy protections, we did not have access to subject variables that could be used as covariates in a DA analysis.
+We therefore performed a DA analysis of species proportion versus community alpha diversity.
+A low-diversity, _Lactobacillus_-dominated microbiome is frequently cited as the typical healthy microbiome and high alpha diversity is considered an indication of bacterial vaginosis and has been repeatedly found to be postively associated with risk of preterm birth.
+(Though notably diverse and low-_Lactobacillus_ microbiomes are also common in healthy women and even typical in certain populations, including segments of the MOMSPI cohort.)
+We split the samples into groups with low and high Shannon diversity in the observed (uncorrected) community composition, leaving out samples with medium diversity, and performed a DA analysis of species proportion versus diversity group.
+We evaluated several types of DA analyses: The simple linear regression model applied to log proportion (Section \@ref()); negative-binomial (or gamma-poisson) regression, which models expected log proportion with a linear model while also accounting for the random, discrete nature of sequencing read counts; and a rank-regression method that approximates the commonly-used non-parametric Wilcoxin test (TODO).
+We performed each DA method with and without correcting for bias on a set of X species passing a minimum abundance prevalence criterion and compared the effect of bias on estimated regression coefficients for individual species.
+The mean efficiency tended to be larger in the high-diversity group (presumably due to a lower proportion of Lactobacillus, CHECK); unsurprisingly, the effect of bias was therefore always to lower the regression coefficient.
+For each method, bias caused significant magnitude and sign errors for X-Y species, including creating sign errors for X, Y, and Z species respectively.
+The absolute error due to bias is the same for all species in the simple linear regression (see previous section), but for the other two methods varied substantially among species (though was always positive, corresponding to the decrease in log mean efficiency in the high-diversity group).
+The DA results also differed substantially across methods.
+Together, these variations meant that whether bias had a significant impact on a particular species was often method dependent.
+
+- ?? P: Real application: Callahan et al, Gard bias - maybe in future
+    - what bias could explain discrepancies?
+
+MOVE: Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
+
+### old text
+
+(Has not been updated)
+
+One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
+To characterize bias in a 16S sequencing protocol developed for the Vaginal Human Microbiome Project (VaHMP), @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome.
+This protocol was later used for the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study (@fettweis2019thev).
+We used the method for measuring and correcting bias developed by @mclaren2019cons to measure bias from the mock communities and analyze its potential impact on DA analyses that use the MOMSPI measurements (SI Analysis MOMS-PI).
+
+The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
+The three species that most often dominate the calibrated profiles are _Gardnerella vaginalis_, _Lactobacillus iners_, and _Lactobacillus crispatus_.
+Our mock-community analysis indicate _G. vaginalis_ has the lowest efficiency of the seven species, roughly 20X lower than _L. crispatus_ and 30X lower than _L. iners_ (the most efficiently measured).
+As a result, the mean efficiency varies substantially across samples, with samples dominated by a _Lactobacillus_ species typically having an efficiency that is 3-20X greater than samples dominated by _Gardnerella_ (Figure \@ref(fig:momspi-mean-efficiency-dist)).
+Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
+Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure TODO).
+Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
 
@@
 
-As sediment is slowly buried, the cencentration of oxygen and nutrients decreases, and cell density decreases; @lloyd2020evid asked whether some taxa are nevertheless able to continue to grow and increase in cell density.
-To measure density with depth, they used community 16S sequencing coupled with measurmeents of total community density using CARD-FISH, which they combined using community-normalization (Equation X).
+As sediment is slowly buried, the concentration of oxygen and nutrients decreases, and cell density decreases; @lloyd2020evid asked whether some taxa are nevertheless able to continue to grow and increase in cell density.
+To measure density with depth, they used community 16S sequencing coupled with measurements of total community density using CARD-FISH, which they combined using community-normalization (Equation X).
 (The authors refer to these absolute density measurements as FRAxC measurements.)
@@
 
-The authors attempted to explictly account for bias by arguing that a consistent taxonomic bias would cancel in the observed log-fold-changes that determine the inferred exponential growth rates (ref kevorkian2018 as well); however, their calculations neglected the impact of changes in mean efficiency which do not cancel and may lead to spurious conclusions (previous section).
+The authors attempted to explicitly account for bias by arguing that a consistent taxonomic bias would cancel in the observed log-fold-changes that determine the inferred exponential growth rates (see also @kevorkian2018esti); however, their calculations neglected the impact of changes in mean efficiency which do not cancel and may lead to spurious conclusions (previous section).
 We were therefore interested in whether the unaccounted for error caused by bias might have led to spurious inferences about which species were increasing with depth.
 Supposing the CARD-FISH measurements accurately measured total density, our results from the previous section indicate that a systematic error in growth rates would arise if the log mean efficiency systematically changed with depth.
-Diversity is sufficiently high that this systematic error is unlikely to be driven by just a few species, as in the vaginal MB example and leopold pathogen case.
+Diversity is sufficiently high that this systematic error is unlikely to be driven by just a few species, as in the vaginal microbiome and fungal case studies.
 But it is possible that the covariate (burial time) selects directly on a species efficiency as well as its ability to grow. 
 In particular, we considered that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
-In this case, we would expect the the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that are just persisting or even slowly dying off would be mistakenly inferred to have positive growth.
+In this case, we would expect the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that are just persisting or even slowly dying off would be mistakenly inferred to have positive growth.
 
@@
 A low number of experimental samples and noise in both the FRAxC and qPCR measurements place signficant uncertainty in these measurements; however, the fact that FRAxC-derived growth rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreases with depth in a manner that systematically biases FRAxC-derived rates to higher values.
-The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr is signficant for this low nutrient environment, where growth rates of the identified positive growers ranged between 0.04/yr and 0.5/yr.
+The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr is significant for this low nutrient environment, where growth rates of the identified positive growers ranged between 0.04/yr and 0.5/yr.
 Overall, the comparison between FRAxC and qPCR measurements gives support to the study conclusions, but suggest that species at the lower end of this range of positive FRAxC-derived rates may in fact be merely persisting or even slowly declining in abundance.
@@
 
-- qPCR and seq analysis are both at ~family level; this is prob not a big deal when it comes to inferring the systematic varation in mean efficiency though, since qPCR and 16S seq share a lot of bias
+- qPCR and seq analysis are both at ~family level; this is prob not a big deal when it comes to inferring the systematic variation in mean efficiency though, since qPCR and 16S seq share a lot of bias
 - relic dna
@@
 
-### old text
-
-Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
-For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
-@lloyd2020evid estimate the log fold change in the estimated absolute abundance of various taxa with sediment depth (as a proxy for time) to determine which taxa are able to persist and even grow in this difficult environment.
-It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
-As the relative abundance of tougher species increases with depth, the mean extraction efficiency might therefore decrease.
-This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
-Additional measurements by @lloyd2020evid make it possible to investigate this possibility explicitly.
-The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for particular taxa.
-For instance, in Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years using the proportion-based densities and 10.3 years by qPCR (@lloyd2020evid Table 1 and Figure 4).
-These doubling times correspond to highly similar estimated slopes of log density per year of $1/10.1 \approx 0.99$ and $1/10.3\approx 0.97$.
-The small difference between these slopes suggests that there is little correlation of log mean efficiency with burial time and hence little systematic error in the estimated doubling times derived from the proportion-based density estimates.
-
-Cut from previous version
-
-In Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years (CHECK) using the proportion-based densities and 10.3 years by qPCR (Lloyd et al.
-(2020) Table 1 and Figure 4).
-These doubling times correspond to highly similar estimated slopes of log density per year of 1 / 10.1 ≈ 0.99 and 1 / 10.3 ≈ 0.97 .
-The difference between these slopes, ≈ 0.02, estimates the slope of the mean efficiency.
-It’s small value suggests little systematic variation in mean efficiency with burial time and hence little systematic bias in the estimated doubling times derived from the proportion-based density estimates.
-In this case, close agreement in the DA results across methods for a specific taxon suggests that variation in sample mean efficiency did not significantly distort the MGS-based results and hence provides a degree of validation for all taxa, not just the taxa with targeted measurements.
-(An important caveat for this example is that this analysis assumes consistent efficiencies at the higher taxonomic levels used for qPCR measurement and MGS analysis performed by Lloyd et al.
-(2020).)
-
-Here we compared summary statistics; however, in principle a joint statistical analysis of targeted and proportion-based density measurements would provide more statistical power to estimate and correct the effects of variation in the mean efficiency across samples and provide calibrated fold change and regression estimates for all taxa [APPENDIX]
-
-## conclusions
-
-Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
-One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
-A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
-Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
-Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
-In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
-For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.
-
-Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.
-
-- Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
-
-# Old
-
-## Systematic error in slope or LFC estimates
-
-Recall from Section \@ref(differential-abundance) that the error in the slope or LFC in a DA regression analysis is proportional to the covariance of the log mean efficiency with the covariate.
-This covariance can be split into two components: the standard deviation of (log) mean efficiency and its correlation with the covariate.
-Hence when considering whether taxonomic bias creates large error, it can be useful to separately ask whether the mean efficiency is likely to vary across samples and whether it is likely to be correlated with a covariate of interest.
-
-<!-- General mechanisms -->
-
-Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
-One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
-A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
-Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
-Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
-In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
-For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.
-
-Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.
-
-<!-- P: relative magnitude of species and mean efficiency changes -->
-The biological significance of the error in $\hat \beta$ depends on the relative magnitude of the LFC (or covariance with the covariate $x$) in the focal species vs in the mean efficiency.
-It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
-Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
-However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for other species.
-The experimental and computational methods discussed in Section \@ref(solutions) can mitigate taxonomic bias in these clearly-problematic cases while also helping to amass empirical evidence on the conditions in which bias is unlikely to cause major inferential errors.
-
-## Loss of precision and power
-
-Another way in which taxonomic bias hinders DA analysis is by increasing the noise in abundance measurements, leading to a reduction in the precision of LFC estimates and thus the power to detect true associations.
-
-There are two distinct mechanisms by which such reductions occur.
-The first was described in Section \@ref(differential-abundance): Variation in the mean efficiency that is random (in the sense of being unassociated with the covariates in a regression analysis) tends to increase the standard errors in estimated regression coefficients.
-As noted above, this effect may be the dominant one in cases like the HMP2 IBD dataset where large variation in dominant phylum is independent of disease status.
-Correcting for the variation in mean efficiency using the calibration methods described in the next section can reverse this effect, thereby increasing the precision as well reducing the systematic error in DA inference.
-
-The second mechanism involves the counting noise that stems from the random sampling of reads in the sequencing experiment. 
-This counting noise is expected to be the dominant source of noise for species whose expected proportion _after accounting for taxonomic bias_ is $\lesssim 1 / \text{total reads}$ for a given sample.
-Having an atypically low efficiency makes a species much more likely to be near this threshold and hence to receive highly uncertain measurements.
-Such imprecise measurements can make it near impossible to make biologically meaningful conclusions.
-For example, in the experiment of @leopold2020host described above, the median read counts in samples taken after pathogen colonization for the 5 commensals with the lowest efficiencies were 0, 0, 0, 1, and 11.
-These species' efficiencies ranged from 40X-10X lower than the pathogen's efficiency; thus, in the absence of bias the counts may have been much higher and it might have been possible to make much more precise conclusions about how their relative abundances changed during pathogen invasion.
-The literature on the vaginal microbiome during pregnancy provides us with another example: Disagreement among studies over whether _Gardnerella_ is positively associated with preterm birth seems to be in part driven by the fact that several studies used 16S primers that are extremely poor at amplifying _Gardnerella_ and hence lack the precision necessary to find associations even were they to exist.
-Unfortunately, even perfect knowledge of the measurement efficiencies will not allow us to overcome the low precision caused by small expected read counts; however, the methods described in the section provide principled ways to assess whether the low counts are due to taxonomic bias rather than truly low frequencies in a species' abundance.
+## Conclusions

### 2021-12-17 notebook/_posts/2021-11-01-momspi-summary/momspi-summary.Rmd
Add MOMSPI regression analysis

@@
   enframe('.sample', 'mean_efficiency_controls')
-mean_eff <- left_join(tmp1, tmp2, by = '.sample') %>%
-  left_join(sample_data(momspi_cal) %>% as_tibble, by = '.sample')
+mean_eff <- left_join(tmp1, tmp2, by = '.sample')
 ```
@@
 mean_eff1 <- mean_eff %>%
+  left_join(sample_data(momspi_cal) %>% as_tibble, by = '.sample') %>%
   left_join(genus_prop, by = '.sample') %>%
@@
 
+# Testing
+
+how many subjects?
+
+```{r}
+cts <- momspi %>% sample_data %>% as_tibble %>% count(host_subject_id)
+nrow(cts)
+#> [1] 577
+cts %>% pull(n) %>% summary
+#>    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+#>    1.00    2.00    3.00    3.39    4.00    9.00 
+cts %>%
+  ggplot(aes(x = n)) +
+  geom_histogram(binwidth = 1)
+  # geom_dotplot(binwidth = 1)
+```
+
+577 subjects
+
 # Session info {.appendix}

### 2021-12-17 notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity.Rmd
Add MOMSPI regression analysis

---
title: "MOMS-PI: Regression on diversity"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:brooks2015thet
  - ref:fettweis2019thev
date: 2021-12-15
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300
)
```

# Setup

## R setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
```

```{r ggplot}
theme_set(theme_cowplot())

colors_brooks <- c(
  "Atopobium_vaginae" = "#009E73",
  "Gardnerella_vaginalis" = "#56B4E9",
  "Lactobacillus_crispatus" = "#D55E00",
  "Lactobacillus_iners" = "#505050",
  "Prevotella_bivia" = "#0072B2",
  "Sneathia_amnii" = "#CC79A7",
  "Streptococcus_agalactiae" = "#E69F00")

scale_y_custom <- scale_y_continuous(
    trans = 'log10',
    breaks = trans_breaks('log10', function(x) 10^x),
    labels = trans_format('log10', math_format(10^.x))
  )
```

# SETUP TODO

# Regression analysis

NOTE: Consider saving output so can be used in other notebook files

## Classify samples into high and low diversity groups

TODO: Consider possible better way to do calibration

Compute Shannon and Inverse-Simpson diversity indices and add to the sample metadata.
`Phyloseq::estimate_richness()` will complain about the lack of singletons, but we can ignore this since we aren't using species richness estimators.
I will exponentiate the Shannon index to convert it to the q1 Hill number so that (like Inverse Simpson) it has the scale of number of species.

```{r}
div <- list(
  'obs' = momspi_obs,
  'cal' = momspi_cal
) %>%
  map(estimate_richness, measures = c('Shannon', 'InvSimpson')) %>%
  map_dfr(as_tibble, rownames = '.sample', .id = 'type') %>%
  transmute(
    .sample, type,
    div_q1 = exp(Shannon),
    div_q2 = InvSimpson,
  ) %>%
  pivot_longer(starts_with('div_'), names_to = 'index', values_to = 'diversity')
div_wide <- div %>%
  pivot_wider(values_from = diversity, names_from = c('index', 'type')) 
```

```{r}
div_wide %>%
  mutate(across(starts_with('div'), log2)) %>%
  ggplot() +
  geom_point(aes(x = .panel_x, y = .panel_y), alpha = 0.5) +
  ggforce::geom_autodensity() +
  geom_bin2d(aes(x = .panel_x, y = .panel_y)) +
  ggforce::facet_matrix(vars(starts_with('div')), layer.diag = 2, layer.lower = 3)
```

The observed (Log) order-1 and order-2 diversity show a tight correlation with each other, as do the calibrated diversities, while the correlation is much lower when comparing observed and calibrated numbers.
The reason is likely that calibration will tend to increase the diversity of samples that are observed to be dominated by a high-efficiency taxon (like L. iners) and decrease the diversity of samples dominated by a low efficiency tason (like G. vaginalis).

Shannon diversity of the observed profiles is what is most often considered in vaginal microbiome studies, so let's use this metric to partition the samples.

```{r}
div_wide %>%
  ggplot(aes(x = div_q1_obs)) +
  scale_x_log10() +
  annotate('rect', xmin = 0, xmax = 1.5, ymin = -Inf, ymax = Inf, 
    fill = 'blue', alpha = 0.3) +
  annotate('rect', xmin = 1.5, xmax = 3, ymin = -Inf, ymax = Inf, 
    fill = 'grey', alpha = 0.3) +
  annotate('rect', xmin = 3, xmax = Inf, ymin = -Inf, ymax = Inf, 
    fill = 'green', alpha = 0.3) +
  geom_histogram() +
  geom_vline(xintercept = c(1.5, 3), color = 'black')
```

I'll pick the first mode as the 'Low diversity' group, treat the second mode as 'Medium', and the tail as 'High'.

```{r}
div_wide <- div_wide %>%
  mutate(
    div_group = case_when(
      div_q1_obs < 1.5 ~ 'Low',
      div_q1_obs > 3 ~ 'High',
      TRUE ~ 'Medium'
    ) %>% factor(levels = c('Low', 'Medium', 'High'))
  )
```

## Understanding the diversity-efficiency relationship

```{r}
mean_eff3 <- mean_eff2 %>%
  left_join(
    div,
    # div %>% pivot_wider(names_from = type, values_from = c(Shannon, InvSimpson)),
    by = '.sample'
  )
```

```{r}
mean_eff3 %>%
  ggplot(aes(diversity, mean_efficiency, color = top_plot)) +
  facet_grid(type~index) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point(alpha = 0.6) +
  stat_smooth(aes(group = 'none'), color = 'black') +
  scale_color_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  )
```

```{r}
sam_div <- momspi_div %>%
  sample_data %>%
  as_tibble
```

```{r}
sam_div %>%
  ggplot(aes(y = div_group, x = mean_efficiency)) +
  scale_x_log10() +
  ggdist::stat_slab()
```

We can see that the mean efficiency is 3-10X lower in the High diversity relative to the low diversity group.

```{r}
lm(log(mean_efficiency) ~ div_group, data = sam_div) %>% summary
```

## Regression setup

To simplify interpretation of some of the regression models, let's rarefy samples to 10K reads (we'll be dropping some samples).

```{r}
momspi_rare <- momspi %>% 
  rarefy_even_depth(
    sample.size = 1e4, 
    rngseed = 42,
    replace = TRUE, # for speed, might want to change
    trimOTUs = FALSE
  )
```

Add the diversity grouping information and the offset terms for the Gamma-Poisson (Negative Binomial) regression.

```{r}
momspi_div <- momspi_rare %>%
  left_join_sample_data(mean_eff, by = '.sample') %>%
  left_join_sample_data(div_wide, by = '.sample') %>%
  mutate_sample_data(
    gp_offset_obs = log(sample_sum),
    gp_offset_cal = log(sample_sum) - log(mean_efficiency),
  )
```

We'll use the criterion from Fettweis et al to pick the taxa to test.
Note, we may not get the same exact taxa as them since we're using a larger set of women.
Note, the sample sums used in the offset were calculated for the full species set, so we are still analzying log proportions with respect to the full species set.

```{r}
taxa_to_test <- momspi %>%
  transform_sample_counts(close_elts) %>%
  filter_taxa2(~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
  taxa_names %>%
  print
```

## Linear regression on log proportion

Do this first to confirm as expected from theory.

```{r}
stat_test <- function(ps) {
  ps %>%
    transform_sample_counts(close_elts) %>%
    filter_sample_data(div_group %in% c('Low', 'High')) %>%
    prune_taxa(taxa_to_test, .) %>%
    as_tibble %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data,
        ~lm(log(.abundance) ~ div_group, data = .x)
      ),
    ) %>%
    select(-data)
  # %>% unnest(fit)
}
res_div_lm <- list(
  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.3),
  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.3) %>% calibrate(bias_all_vec)
) %>%
  map_dfr(stat_test, .id = 'type')
res_div_lm_tidy <- res_div_lm %>%
  mutate(across(fit, map, broom.mixed::tidy)) %>%
  unnest(fit) %>% 
  filter(term != '(Intercept)')
```

```{r}
# x should be the outout of a tidier
plot_estimates <- function(x) {
  err <- x %>%
    select(name = type, .otu, value = estimate) %>%
    pivot_wider() %>%
    mutate(
      # error = Observed - Calibrated, # will be same for everything
      # error_abs = abs(error),
      # # error_rel = abs(error / Calibrated), # only makes sense if not a sign error
      error_sign = sign(Observed) == sign(Calibrated),
      sign = case_when(
        sign(Observed) > 0 & sign(Calibrated) > 0 ~ 'Positive',
        sign(Observed) < 0 & sign(Calibrated) < 0 ~ 'Negative',
        TRUE ~ 'Depends',
      )
    ) %>%
    select(.otu, error_sign, sign)

  p1 <- x %>%
    left_join(err, by = '.otu') %>%
    ggplot(aes(y = type, x = estimate, fill = sign)) +
    scale_fill_brewer(type = 'qual', palette = 2) +
    geom_vline(xintercept = 0, color = 'grey') +
    ggdist::geom_dots()
  p2 <- x %>%
    mutate(
      .otu_fct = fct_reorder(.otu, estimate, .fun = min)
    ) %>%
    ggplot(aes(y = .otu_fct, x = estimate, color = type)) +
    scale_color_brewer(type = 'qual', palette = 7) +
    geom_vline(xintercept = 0, color = 'grey') +
    geom_pointrange(
      aes(
        xmin = estimate - 2*std.error,
        xmax = estimate + 2*std.error
      ))
  p1 / p2 +
    plot_layout(ncol = 1, heights = c(0.2, 1))
}
plot_estimates(res_div_lm_tidy)
```

Consider doing bootstrap CIs.

```{r, eval = F}
lm_boot <- function(.data, ...) {
  fit <- lm(log(.abundance) ~ div_group, data = .data)
  car::Boot(fit, ...)
}
x <- momspi_div %>% transform_sample_counts(~. + 0.3) %>% 
  filter_sample_data(div_group %in% c('Low', 'High')) %>%
  filter_tax_table(.otu == 'Gardnerella_vaginalis') %>% 
  as_tibble
fit <- lm(log(.abundance) ~ div_group, data = x)
bt <- car::Boot(fit)
bt %>% broom::tidy()
broom::tidy(bt, conf.int = TRUE)
confint(bt, type = 'perc') %>% broom::tidy()
```

## Regression on rank

Similar to Wilcoxin test when there are no ties; not sure about with ties.

```{r}
stat_test <- function(ps) {
  ps %>%
    transform_sample_counts(close_elts) %>%
    filter_sample_data(div_group %in% c('Low', 'High')) %>%
    prune_taxa(taxa_to_test, .) %>%
    as_tibble %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data,
        ~lm(
          rank(.abundance, ties.method = 'average')/n() ~ div_group, 
          data = .x
        )
      ),
    ) %>%
    select(-data)
  # %>% unnest(fit)
}
res_div_rank <- list(
  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.3),
  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.3) %>% calibrate(bias_all_vec)
) %>%
  map_dfr(stat_test, .id = 'type')
res_div_rank_tidy <- res_div_rank %>%
  mutate(across(fit, map, broom.mixed::tidy)) %>%
  unnest(fit) %>% 
  filter(term != '(Intercept)')
```

```{r}
plot_estimates(res_div_rank_tidy)
```

## Gamma-Poisson regression

```{r}
library(rstanarm)
options(mc.cores = parallel::detectCores())
```

HERE. Might want to add the (lo) species efficiency to the offset, though it should make absolutely no difference to the fitted models.

HERE. Might make sense to use flat priors or MLE inference, to disable regularization, so that the results are most directly comparable to our prediction that the effect of calibration is to shift the coefficients for each species by approximately the same amount.

```{r}
stat_test <- function(ps) {
  ps %>%
    prune_taxa(taxa_to_test, .) %>%
    filter_sample_data(div_group %in% c('Low', 'High')) %>%
    as_tibble %>%
    # left_join(bias_all %>% select(.otu, efficiency), by = '.otu') %>%
    # mutate(offset = offset + log(efficiency)) %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data,
        ~stan_glm(
          .abundance ~ div_group + offset(offset), 
          data = .x, seed = 42,
          family = neg_binomial_2, 
          algorithm = 'sampling'
          # algorithm = 'optimizing'
          # algorithm = 'meanfield'
        )
      ),
    ) %>%
    select(-data)
  # %>% unnest(fit)
}
res_div_gp <- list(
  'Observed' = momspi_div %>% mutate_sample_data(offset = gp_offset_obs),
  'Calibrated' = momspi_div %>% mutate_sample_data(offset = gp_offset_cal)
) %>%
  map_dfr(stat_test, .id = 'type')
# saveRDS(res_div_gp, 'res_div_gp-meanfield.rds')
# saveRDS(res_div_gp, 'res_div_gp.rds')
res_div_gp_tidy <- res_div_gp %>%
  mutate(across(fit, map, broom.mixed::tidy)) %>%
  unnest(fit) %>% 
  filter(term != '(Intercept)')
```

```{r}
plot_estimates(res_div_gp_tidy)
```

interpretation:

- The prop of most species is greater in the high (observed) diversity samples, calibrated or not, because the low-diversity samples tend to be dominated by Lactobacillus (especially cripatus and iners).
- Because of the sum-to-one constraint, this might just be a consequence of Lacto's having lower abundance - aka a compositional effect
  - consider talking about this in the case study.
- This compositional effect is compounded by bias, since Lactos also have high efficiency.
- hence the observed increases are even larger than the bias-corrected (calibrated) ones.
- interesting to see how some taxa have a much smaller impact from calibration (Gard and BVAB1 especiaully).
  - Perhaps this has to do with these species often being dominant?
    - this means that there are conflicting asssociative forces acting on them - high diversity = low lactobacillus = higher abundance, but also places a cap on how high in abundance these species can be. but this argument doesn't explain why our results wouldn't still apply
  - Are the priors having an impact? What if we flatten the prior on the coefficient?
  - Is there something about working on counts that is causing the effect?
  - I can probably learn something by checking the fit on Gard.
<!--  -->

### flat(er) priors

```{r}
stat_test <- function(ps) {
  ps %>%
    prune_taxa(taxa_to_test, .) %>%
    filter_sample_data(div_group %in% c('Low', 'High')) %>%
    as_tibble %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data,
        ~stan_glm(
          .abundance ~ div_group + offset(offset), 
          data = .x, seed = 42,
          family = neg_binomial_2, 
          # prior_intercept = NULL,
          prior = normal(0, 100),
          # algorithm = 'sampling'
          # algorithm = 'optimizing'
          algorithm = 'meanfield'
        )
      ),
    ) %>%
    select(-data)
  # %>% unnest(fit)
}
res_div_gp_flat <- list(
  'Observed' = momspi_div %>% mutate_sample_data(offset = gp_offset_obs),
  'Calibrated' = momspi_div %>% mutate_sample_data(offset = gp_offset_cal)
) %>%
  map_dfr(stat_test, .id = 'type')
saveRDS(res_div_gp_flat, 'res_div_gp_flat-meanfield.rds')
# saveRDS(res_div_gp_flat, 'res_div_gp.rds')
```

```{r}
res_div_gp_flat_tidy <- res_div_gp_flat %>%
  mutate(across(fit, map, broom.mixed::tidy)) %>%
  unnest(fit) %>% 
  filter(term != '(Intercept)')
```

```{r}
p1 <- res_div_gp_flat_tidy %>%
  ggplot(aes(y = type, x = estimate)) +
  ggdist::geom_dots()
p2 <- res_div_gp_flat_tidy %>%
  mutate(
    .otu_fct = fct_reorder(.otu, estimate, .fun = min)
  ) %>%
  ggplot(aes(y = .otu_fct, x = estimate, color = type)) +
  geom_vline(xintercept = 0, color = 'grey') +
  geom_pointrange(
    aes(
      xmin = estimate - 2*std.error,
      xmax = estimate + 2*std.error
    ))
p1 / p2 +
  plot_layout(ncol = 1, heights = c(0.2, 1))
```

Widening the priors seems to have no impact

### checking on gard

```{r}
fit <- res_div_gp %>%
  filter(.otu == 'Gardnerella_vaginalis', type == 'Observed') %>%
  pull(fit) %>% .[[1]]
fit %>% summary
```

```{r, eval = F}
x <- fit$data %>% as_tibble
y <- x %>% pull(.abundance)
# launch_shinystan(fit, ppd = TRUE)
y_rep <- posterior_predict(fit)
dim(y_rep)
```

Let's look at the log proportion.
To make the distributions more directly comparaible, I'll add the species efficiency in the calibrated part.

```{r}
eff <- bias_all_vec['Gardnerella_vaginalis']
x1 <- x %>%
  select(-offset) %>%
  mutate(gp_offset_cal = gp_offset_cal + log(eff)) %>%
  pivot_longer(starts_with('gp_offset'), names_to = 'offset_type', values_to = 'offset') %>%
  mutate(across(offset_type, str_extract, pattern = '(obs|cal)'))
```
```{r}
x1 %>%
  ggplot(aes(log(.abundance + 0.3) - offset, fill = div_group)) +
  facet_wrap(~offset_type, ncol = 1) +
  geom_histogram()
```

```{r}
x1 %>%
  mutate(log_prop = log(.abundance + 0.3) - offset) %>%
  ggplot(aes(y = div_group, x = log_prop)) +
  facet_wrap(~offset_type) +
  ggdist::stat_halfeye()
```

```{r}
x1 %>%
  mutate(log_prop = log(.abundance + 0.3) - offset) %>%
  with_groups(c(offset_type, div_group), summarize, across(log_prop, mean))
```

```{r}
res_div_gp_tidy %>% filter(.otu == 'Gardnerella_vaginalis')
```

The estimated coefficients are smaller than the differences in the means.

Can we understand why there is relatively little impact of calibration?

Should also compare for a taxon where there is a larger impact.

C

### 2021-12-17 notebook/_code/momspi-setup.R

@@ -1,0 +1,108 @@
+
+# Setup
+
+## R setup
+
+# ```{r libraries}
+# Tools for microbiome data
+library(speedyseq)
+# Tools for general purpose data manipulation and plotting
+library(tidyverse)
+library(fs)
+
+# library(metacal); packageVersion("metacal")
+
+colors_brooks <- c(
+  "Atopobium_vaginae" = "#009E73",
+  "Gardnerella_vaginalis" = "#56B4E9",
+  "Lactobacillus_crispatus" = "#D55E00",
+  "Lactobacillus_iners" = "#505050",
+  "Prevotella_bivia" = "#0072B2",
+  "Sneathia_amnii" = "#CC79A7",
+  "Streptococcus_agalactiae" = "#E69F00"
+)
+
+## Estimate bias from @brooks2015thet
+
+# We can estimate the efficiencies for the control species used in the cellular mocks of @brooks2015thet by running the example code in the docs for `metacal::estimate_bias()`. 
+# This code uses the observed and actual species-level abundance tables which were produced by @mclaren2019cons from the SI files of @brooks2015thet and are included in the metacal package.
+
+dr <- system.file("extdata", package = "metacal")
+list.files(dr)
+actual <- file.path(dr, "brooks2015-actual.csv") |>
+  read.csv(row.names = "Sample") |>
+  as("matrix")
+observed <- file.path(dr, "brooks2015-observed.csv") |>
+  read.csv(row.names = "Sample") |>
+  subset(select = - Other) |>
+  as("matrix")
+
+# Estimate bias with bootstrapping for error estimation
+mc_fit <- metacal::estimate_bias(observed, actual, margin = 1, boot = TRUE)
+summary(mc_fit)
+
+rm(actual, observed, dr)
+
+control_species <- mc_fit %>% coef %>% names
+control_genera <- control_species %>% str_extract('^[^_]+')
+
+## Load the MOMSPI data
+
+# TODO: replace this code chunk with one that pulls data from the github repo into `_data/`
+
+# Load the MOMSPI Stirrups profiles into phyloseq
+
+path_momspi <- '~/research/momspi'
+
+otu <- path(path_momspi, "output", "stirrups-profiles", "abundance-matrix.csv.bz2") %>%
+  read_csv(
+    col_types = cols(.default = col_double(), sample_name = col_character())
[…]

### 2021-12-18 case-studies.Rmd
Revise case studies
Fungal case study: Expand paragraph considering the analysis of absolute
abundance in response to pathogen invasion.

Vaginal case study: Minor edits / text clean up

@@
 To investigate the role of colonization order and host genotype, plants varied in source location and in which symbiont was allowed to colonize first.
-The authors used ITS amplicon sequencing to measure communities before and after pathogen infection, along with mock communities with known genome concentrations of DNA from the 9 species.
-These DNA mock communities allowed the authors to directly measure the bias due to PCR and ribosomal copy-number-variation (CNV) following the method of @mclaren2019cons and correct for it when analyzing the primary experimental samples.
+The authors used ITS amplicon sequencing to measure communities before and after pathogen infection, along with mock communities with quantified genome concentrations of DNA from the 9 species.
+These DNA mock communities allowed the authors to directly measure the bias due to the sequencing workflow following DNA extraction and correct for it when analyzing the primary experimental samples following the method of @mclaren2019cons.
+Although extraction is not accounted for, considerable bias might still be expected due to the substantial ribosomal copy-number-variation (CNV) in fungi (CITE) and preferential PCR binding and amplification (CITE).
 The authors found a 13X difference between the most and least efficiently measured symbiont, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
@@
 
-@leopold2020host restricted their analysis to ITS profiles form the pre-infection samples, as reads from post-infection samples were dominated by the pathogen.
-<!-- (due to a combination of pathogen growth and, presumably, its much greater efficiency). -->
-However, we can use the post-infection samples to ask what the impact of the pathogen's much greater efficiency would have were community sequencing used to understand the impact of pathogen invasion on the already-colonized commensal species without accounting for bias.
-n particular, suppose the ITS sequencing data were coupled with an accurate measurement of total genome copy number using Equation \@ref(eq:density-prop-meas) to predict absolute genome copies of each species.
-The high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples; this increase would tend to lead to a concomitant 6X decrease in the fold changes in genome copies.
+Although not analyzed in the study, the authors also sequenced the fungal communities of plants after pathogen colonization.
+Comparison of the pre- and post-colonization samples suggests a simple, biologically relevant scenario where bias could confound attempts to infer microbial dynamics.
+A natural question to ask in such an experiment is, how do various species in the existing community respond to pathogen invasion?
+Relative abundances alone offer limited information, as they are unable to account for the potential sharp increase in the pathogen abundance post-infection, making such an experiment a natural target for absolute-abundance methods.
+One possibility would be to measure total genome concentration using qPCR of the single-copy gene that was used to quantify concentrations for the mock communities.
+Genome concentration could then be multiplied by ITS sequencing proportions to infer absolute abundances for DA analysis.
+Suppose the genome-concentration measurement is 100% accurate.
+In this case, it would be likely be important to account for bias in the ITS measurement.
+The reason is that the high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples (REF SI/FIG), which would tend to cause a concomitant 6X decrease in the infered fold changes in abundance in response to colonization for each species.
+Alternatively, one might measure total ITS concentration via qPCR, possibly using the same primers used for ITS sequencing.
+In this case, bias correction using the measured efficiencies is unnecessary and would be harmful if not also done for the qPCR measurement.
+The reason is that much (if not most) of the bias in the sequencing measurement is shared by the ITS qPCR measurement, including that due to extraction, CNV, and PCR amplification.
+Appendix X shows that this shared bias in fact leads to more accurate inferences of fold changes in the genome concentrations of individual species, even while likely giving inaccurate changes in total genome concentration.
+In summary, accurate inference of species abundances following pathogen colonization requires accounting for the increase in mean efficiency among samples; however, measuring total community abundance with a method (here, ITS qPCR) that is proportional to the mean efficiency naturally does so without needing to explicitly measure or correct for species efficiencies.
+Similarly, the (ratio-based) reference-species approaches discussed in the previous section might instead be used.
+
+<!-- The post-colonization samples were not analyzed by @leopold2020host in part because the reads from post-colonization samples tended to be dominated by the pathogen, giving little ability to infer the relative abundance of the rarer and lower efficiency species.
+-->
+<!-- This was presumably due to high pathogen growth and aided by the pathogen's post-extraction efficiency being substantially larger than the other species; -->
+<!-- the high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples (REF SI/FIG). -->
 
@@
 
-- Give numerical e.g. for the lowest eff species, Penicillium
+- Give numerical e.g. for the lowest eff species, Penicillium, whose prop is typically 6X too low but the magnitude of error in FCs is generally smaller
 - Check that strength of correlation with covariates isn't an important part of the story as well
@@
     - should give a better interpretation/assessment of the bottom line here.
-- to show that variation within women is lower, compute standard deviation of log mena efficiency within women and average this (perhaps weighted by n)
+    - one way: discuss the variation in log mean efficiency versus that in species proportions.
+See a contrast to the Leopold situation; here the variation is lower but similar enough to potentially matter
+- to show that variation within women is lower, compute standard deviation of log mean efficiency within women and average this (perhaps weighted by n)
 - create 1-2 trajectories to show
@@
 (Though notably diverse and low-_Lactobacillus_ microbiomes are also common in healthy women and even typical in certain populations, including segments of the MOMSPI cohort.)
-We split the samples into groups with low and high Shannon diversity in the observed (uncorrected) community composition, leaving out samples with medium diversity, and performed a DA analysis of species proportion versus diversity group.
+We split the samples into classes with low and high Shannon diversity in the observed (uncorrected) community composition, leaving out samples with medium diversity, and performed a DA analysis of species proportion versus diversity class.
 We evaluated several types of DA analyses: The simple linear regression model applied to log proportion (Section \@ref()); negative-binomial (or gamma-poisson) regression, which models expected log proportion with a linear model while also accounting for the random, discrete nature of sequencing read counts; and a rank-regression method that approximates the commonly-used non-parametric Wilcoxin test (TODO).
 We performed each DA method with and without correcting for bias on a set of X species passing a minimum abundance prevalence criterion and compared the effect of bias on estimated regression coefficients for individual species.
-The mean efficiency tended to be larger in the high-diversity group (presumably due to a lower proportion of Lactobacillus, CHECK); unsurprisingly, the effect of bias was therefore always to lower the regression coefficient.
+The mean efficiency tended to be larger in the high-diversity class (presumably due to a lower proportion of Lactobacillus, CHECK); unsurprisingly, the effect of bias was therefore always to lower the regression coefficient.
 For each method, bias caused significant magnitude and sign errors for X-Y species, including creating sign errors for X, Y, and Z species respectively.
-The absolute error due to bias is the same for all species in the simple linear regression (see previous section), but for the other two methods varied substantially among species (though was always positive, corresponding to the decrease in log mean efficiency in the high-diversity group).
+The absolute error due to bias is the same for all species in the simple linear regression (see previous section), but for the other two methods varied substantially among species (though was always positive, corresponding to the decrease in log mean efficiency in the high-diversity class).
 The DA results also differed substantially across methods.
@@
 MOVE: Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
-
-### old text
-
-(Has not been updated)
-
-One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
-To characterize bias in a 16S sequencing protocol developed for the Vaginal Human Microbiome Project (VaHMP), @brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome.
-This protocol was later used for the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study (@fettweis2019thev).
-We used the method for measuring and correcting bias developed by @mclaren2019cons to measure bias from the mock communities and analyze its potential impact on DA analyses that use the MOMSPI measurements (SI Analysis MOMS-PI).
-
-The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
-The three species that most often dominate the calibrated profiles are _Gardnerella vaginalis_, _Lactobacillus iners_, and _Lactobacillus crispatus_.
-Our mock-community analysis indicate _G. vaginalis_ has the lowest efficiency of the seven species, roughly 20X lower than _L. crispatus_ and 30X lower than _L. iners_ (the most efficiently measured).
-As a result, the mean efficiency varies substantially across samples, with samples dominated by a _Lactobacillus_ species typically having an efficiency that is 3-20X greater than samples dominated by _Gardnerella_ (Figure \@ref(fig:momspi-mean-efficiency-dist)).
-Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
-Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure TODO).
-Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
-
-Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate.
-The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
-For example, in the vaginal microbiome a decline in _Lactobacillus_ and rise of species including _Gardnerella vaginalis_ is associated with preterm birth in pregnant women; hence the variation in mean efficiency driven by these species may also create an association of mean efficiency with preterm status.
 
@@
 ## Conclusions
+
+Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate.
+The MOMSPI regression example illustrates a case where the mean efficiency is highly variable and also correlated with the covariate (diversity class).
+The Leopold T1 example illustrates a case where the mean efficiency simply doesn't vary significantly across samples, and hence there is no room for bias to affect regression coefficients.

### 2022-01-06 notebook/_posts/2022-01-05-leopold2020host-priority-effects/leopold2020host-priority-effects.Rmd
Run leopold2020host priority effects analysis
Run priority effects analysis from leopold2020host with and without bias
correction, to show that bias correction has little impact.

---
title: "Perform Leopold and Busby (2020) priority-effects analysis with and without bias correction"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:leopold2020host
  - differential abundance
  - bias sensitivity
date: 2022-01-05
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    toc: true
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE
  # autodep = TRUE,
  # cache.comments = FALSE
)
```

This Rmd determines whether bias correction has an impact on the analysis of priority effects in @leopold2020host.

# Run analyses with and without bias correction

```{r libraries}
library(here)

set.seed(42)
```

https://github.com/dleopold/Populus_priorityEffects

This script follows the analysis of `code/priorityEffects.R`

Start by running the setup of libraries and loading the phyloseq object and bias estimate.
The code is verbatim from the @leopold2020host code except for the changes in directory necessary for the calls to `source()` and file reading.

```{r lp-setup}
# Analysis of the relative advantage of preemptive colonization.

library(tidyverse)
library(magrittr)
library(phyloseq)
library(foreach)
library(doMC)
library(ggthemes)

this_dir <- getwd()
setwd(here('notebook/_data/leopold2020host/dleopold-Populus_priorityEffects-8594f7c/'))

source("code/Rfunctions.R")
source("code/colors.R")

# register cores for parallel processing
registerDoMC(parallel::detectCores())

# load phyloseq data
(phy <- loadPhyloseq())

# identify focal taxa for plotting 
focalTaxa <- unique(sample_data(phy)$Treatment)

# Extract OTU data, convert to proportions and long format
bias <- read.csv("output/tabs/bias.csv")

setwd(this_dir)
```

Next, we wrap the analyses of the original study within a function `analyze_priority_effects()` to facilitate running the same analyses on the calibrated and uncalibrated profiles.
The code in the function is copied verbatim from the source in `code/priorityEffects.R` except that the call to `unbias()` in the very first line is skipped so that we can control whether correction is done.
Click the toggle to expand the code chunk.

```{r, code_folding = TRUE}
analyze_priority_effects <- function(phy, nboots = 10000) { 
  df <- phy %>%
    transform_sample_counts(function(x){x/sum(x)}) %>%
    otu_table %>% data.frame %>%
    dplyr::select(all_of(focalTaxa)) %>%
    bind_cols(sample_data(phy) %>% data.frame %>% 
                dplyr::select(Region,Genotype,Treatment)) %>%
    pivot_longer(all_of(focalTaxa),names_to="Taxa",values_to="proportion") %>%
    mutate(Focal=ifelse(Taxa==Treatment,T,F),
           Region=ifelse(Region=="East","E","W") %>%
             factor(levels=c("W","E")))

  # Define a function to calculate the estimated priority effect strength for each species on each host genotype as the log-ratio of the proportional abundance when arriving early vs not arriving early
  getPEs <- function(df) {
    df %>% group_by(Focal,Region,Taxa,Genotype,Treatment) %>%
      summarize_all(gm_mean) %>% ungroup %>%
      group_by(Region,Genotype,Taxa) %>%
      summarise(PE=log(mean(proportion[Focal])/mean(proportion[!Focal])))
  }
  meanPEs <- getPEs(df) 
  meanPEs$Genotype %<>% factor(.,levels=unique(.))

  #####################
  ### Bootstrap CIs ###
  #####################

  # Get bootstrap confidence intervals on point estimates of priority effects for each species on each genotype
  # nboots <- 10000
  bootPEs <- foreach(i=1:nboots, .combine=bind_rows) %dopar% {
    df %>% group_by(Region, Taxa, Genotype,Focal,Treatment) %>%
      sample_frac(replace=T) %>%
      getPEs() %>% mutate(bootID=i)
  }
  bootPEs$Genotype %<>% factor(.,levels=unique(bootPEs$Genotype))

  # Get bias corrected and accelerated confidence intervals
  bootPE.ci <- bootPEs %>% 
    group_by(Genotype,Taxa) %>%
    summarize(LCI=coxed::bca(PE)[1],
              UCI=coxed::bca(PE)[2]) %>% full_join(meanPEs)

  ###################
  ### Region test ###
  ###################

  # First get the mean priority effect for each species on eastern and western genotypes
  regionPEs <- meanPEs %>% group_by(Region,Taxa) %>%
    summarize(meanPE=mean(PE),
              tstat=t.test(PE,mu=0,alternative="greater") %>% .$statistic)

  # Run bootstrapped t-test using 100 bootstraps at the region level for each of the previously generated genotype-level bootstraps
  bootRegion <- foreach(i=1:100,.combine=bind_rows) %dopar% {
    bootPEs %>% 
      group_by(Region,Taxa,bootID) %>%
      sample_frac(replace=T) %>%
      left_join(regionPEs,by=c("Region","Taxa")) %>%
      mutate(center=PE-meanPE) %>%
      summarize(mu.boot=mean(PE),
                tstat.obs=mean(tstat),
                tstat.boot=ifelse(var(center)==0,ifelse(mean(PE)>mean(meanPE),Inf,-Inf),
                                  t.test(center,mu=0) %>% .$statistic))}
  # calculate bootstrapped p-values (one-tailed test for significant positive priority effects)
  bootRegionPvals <- bootRegion %>% drop_na %>%
    mutate(test = tstat.boot+1>tstat.obs+1) %>% 
    summarize(pval=mean(as.numeric(test))) %>%
    mutate(stars=gtools:::stars.pval(pval) %>% gsub(".","+",fixed = T,.)) 
  regionSig <- bootRegion %>% summarize(mu.boot.max=max(mu.boot)) %>%
    left_join(bootRegionPvals)

  ############
  ### PLOT ###
  ############

  #' Define a function to plot the regional priority effects bootstrap results 
  get_inset <- function(df,inset.ymin,inset.ymax){
    ggplot(df, 
           aes(x=Region,y=mu.boot)) +
      geom_violin(scale="width",fill="grey75",color="grey75") + 
      geom_text(data=df %>% dplyr::select(Region,Taxa,stars,mu.boot.max) %>%
                  slice(1),
                aes(label=stars,y=mu.boot.max),size=5)+
      geom_hline(yintercept = 0,alpha=0.4,linetype="dotted")+
      coord_cartesian(clip='off')+
      ylim(c(inset.ymin,inset.ymax))+
      theme_few()+
      theme(strip.text = element_blank(),
            axis.title = element_blank(),
            panel.border = element_blank(),
            axis.line = element_line(color="black",size=0.2))
  }

  # create inset plots as a list
  insets <- bootRegion %>% 
    left_join(dplyr::select(regionSig,Region,Taxa,stars,mu.boot.max)) %>%
    split(f = .$Taxa) %>%
    purrr::map(~annotation_custom2(
      grob = ggplotGrob(get_inset(.,min(bootRegion$mu.boot),max(bootRegion$mu.boot))), 
      data = data.frame(Taxa=unique(.$Taxa)),
      ymin = 2.4, ymax=Inf, xmin=4, xmax=Inf)
    )

  # make full plot of bootstrapped priority effects results
  plt <- ggplot(bootPE.ci,aes(x=Genotype,y=PE))+
    geom_pointrange(aes(ymin=LCI,ymax=UCI,fill=Genotype,shape=Genotype),
                    size=0.5,stroke=0.35,fatten=5.5)+
    scale_shape_manual("Host genotype", values=c(rep(21,7),rep(23,5)))+
    scale_fill_manual("Host genotype",values=pal.genotype)+
    geom_hline(yintercept = 0,alpha=0.6,linetype="dotted")+
    scale_x_discrete(breaks = levels(bootPEs$Genotype),
                     limits = c(levels(bootPEs$Genotype)[1:7], "skip1",
                                levels(bootPEs$Genotype)[8:12]),
                     expand=expansion(0.075))+
    ylab("Strength of priority effect")+
    labs(tag="Eastern              Western")+
    coord_cartesian(clip='off')+
    ylim(c(-1,3.5))+
    facet_wrap(~Taxa,scales="free_x",nrow=1)+
    guides(fill = guide_legend(override.aes = list(size=0.8)))+
    theme_few()+
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size=14,face="italic"),
          axis.title.y = element_text(size=14),
          legend.title = element_text(size=14),
          legend.text = element_text(size=12),
          legend.margin=margin(6,6,6,12),
          plot.tag = element_text(angle=90,hjust=0),
          plot.tag.position = c(0.848,0.115))+
    insets
  # ggsave("output/figs/Fig.3.pdf",width=24,height=10,units="cm")

  return(
    list(
      meanPEs = meanPEs, 
      bootPEs = bootPEs,
      bootPE.ci = bootPE.ci,
      regionPEs = regionPEs,
      bootRegion = bootRegion,
      bootRegionPvals = bootRegionPvals,
      plot = plt
    )
  )
}
```

Next, we run the analysis on the phyloseq objects with and without bias correction (calibration).
@leopold2020host use a custom function `unbias()` for bias correction, which also implements a zero-replacement procedure.
In order to keep the calibrated and uncalibrated analyses comparable, we should apply the same zero-replacement procedure in both cases.
We do this by applying `unbias()` with the supplied bias vector set to all 1s, which corresponds to no bias.

```{r}
bias_list <- list(
  'Corrected' = bias, 
  'Uncorrected' = bias %>% mutate(Bhat = 1)
)
phy_list <- bias_list %>% map(~unbias(phy, .x))
```

The only difference between the two phyloseq objects should be a perturbation by the bias vector, which we can check by using `metacal::calibrate()`.

```{r}
tmp <- metacal::perturb(
  phy_list$Corrected, 
  bias %>% filter(Taxon != 'Melampsora') %>% select(Taxon, Bhat) %>% deframe,
  norm = 'keep'
)
list(phy_list$Uncorrected, tmp) %>%
  map(otu_table) %>%
  map(transform_sample_counts, function(x) x / sum(x)) %>%
  {all.equal(.[[1]], .[[2]])}
```

The sample totals are affected by `unbias()` such that they differ between the corrected and uncorrected objects,

```{r}
phy_list$Corrected %>% sample_sums %>% head(3)
tmp %>% sample_sums %>% head(3)
phy_list$Uncorrected %>% sample_sums %>% head(3)
```

But since the analysis function converts the counts to proportions in the first step, this difference can't impact the results.

Finally, we run the analysis on both the corrected and uncorrected phyloseq objects.
The call is wrapped in `xfun::cache_rds()` to cache the results; the hash is based on the inputs and the analysis function's source code.

```{r analyses, message = FALSE, results = 'hide'}
fun_src <- attr(analyze_priority_effects, 'srcref') %>% as('character')
res <- xfun::cache_rds({
  phy_list %>% map(analyze_priority_effects)
}, hash = list(fun_src, phy_list, focalTaxa),
  clean = FALSE
)
```

# Compare results

Next we compare the results of the analysis with and without bias correction.

```{r}
library(cowplot)
library(patchwork)
theme_set(theme_cowplot())
```

```{r}
res %>% names
res[[1]] %>% names
```

First, let's compare the main text figure with and without correction,

```{r, fig.dim = c(10, 8), layout="l-body-outset"}
map2(map(res, 'plot'), names(res), ~.x + ggtitle(.y)) %>%
  {.[[1]]  / .[[2]]} &
  theme(plot.title.position = 'plot')
```

The results are nearly indistinguishable; bias correction has no significant impact.

Next, let's compare the point estimates of the priority effects,

```{r, fig.dim = c(6, 6)}
x <- res %>%
  map_dfr('meanPEs', .id = 'type') %>%
  pivot_wider(names_from = type, values_from = PE)
mse <- x %>%
  ungroup %>%
  summarize(
    mse = mean((Corrected - Uncorrected)^2)
  ) %>% .[[1]] %>% signif(3)
x %>%
  ggplot(aes(Uncorrected, Corrected, color = Region)) +
  coord_fixed() +
  geom_abline(color = 'darkred') +
  geom_point() +
  labs(title = 'Effect of bias correction on priority-effect estimate', 
    subtitle = str_glue('Mean squared difference: {mse}')
  ) +
  theme(plot.title.position = 'plot') +
  scale_color_brewer(type = 'qual', palette = 2)
```

There is very little impact on the point estimates.
Now, let's compare the bootstrap confidence intervals with and without correction,

```{r, fig.dim = c(8, 8)}
res %>%
  map_dfr('bootPE.ci', .id = 'type') %>%
  ggplot(aes(x = Genotype, y = PE, color = type, ymin = LCI, ymax =UCI)) +
  facet_wrap(~Taxa, scales = "fixed", ncol = 1) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  scale_color_brewer(type = 'qual') +
  theme(
    plot.title.position = 'plot',
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(title = 'Effect of bias correction on priority-effect CIs')
```

This fig clearly shows the equivalence of results with uncorrected and corrected taxonomic profiles.

Let's do similar for the regional PEs.

```{r, fig.dim = c(7, 4.5)}
x <- res %>%
  map_dfr('regionPEs', .id = 'type') %>%
  pivot_longer(c(meanPE, tstat)) %>%
  pivot_wider(names_from = type)
mse <- x %>%
  ungroup %>%
  filter(name == 'meanPE') %>%
  summarize(
    mse = mean((Corrected - Uncorrected)^2)
  ) %>% .[[1]] %>% signif(3)
x %>%
  ggplot(aes(Uncorrected, Corrected)) +
  facet_wrap(~name, scales = 'free') +
  # coord_fixed() + # doesn't work with free scales
  geom_abline(color = 'darkred') +
  geom_point() +
  labs(title = 'Effect of bias correction on regional priority-effect estimate', 
    subtitle = str_glue('Mean squared difference in meanPE: {mse}')
  ) +
  theme(plot.title.position = 'plot')
```

Again there is almost no effect, except that the largest t statistic is somewhat larger before bias correction.

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2022-01-06 notebook/_posts/2022-01-06-leopold2020host-original-regression-analysis/leopold2020host-original-regression-analysis.Rmd
Run leopold2020host regression analysis
Run regression analysis from leopold2020host with and without bias
correction, to show that bias correction has little impact.

---
title: "Perform Leopold and Busby (2020) regression analysis with and without bias correction"
description: |
  We perform the regression analysis of Leopold and Busby (2020) with and without bias correction, finding that there is negligible impact of bias correction on the results.
author:
  - name: Michael R. McLaren
    url: {}
date: 2022-01-06
bibliography: ../../../main.bib
categories:
  - ref:leopold2020host
  - differential abundance
  - bias sensitivity
output:
  distill::distill_article:
    self_contained: false
    toc: true
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE
  # autodep = TRUE,
  # cache.comments = FALSE
)
```

This document asks whether bias correction has a significant impact on the regression analysis of @leopold2020host.

# Run analyses with and without bias correction

```{r libraries}
library(here)

set.seed(42)
```

This script follows the analysis of `code/jsdModels.R` in https://github.com/dleopold/Populus_priorityEffects.

We start by running the setup of libraries and loading the phyloseq object and bias estimate.
The code is modified from `jsdModels.R` to adjust working directories for reading files.

```{r lp-setup}
library(tidyverse)
library(magrittr)
library(phyloseq)
library(mvabund)
library(gt)

this_dir <- getwd()
setwd(here('notebook/_data/leopold2020host/dleopold-Populus_priorityEffects-8594f7c/'))

source("code/Rfunctions.R")

# load phyloseq data
(phy <- loadPhyloseq())

# load bias correction factors - estimated from mock communities
bias <- read.csv("output/tabs/bias.csv")

setwd(this_dir)
```

## Compute offsets with and without bias correction

Bias is accounted for in the original regression analysis through the use of taxon and sample-specific offsets, stored in a matrix `effort`.
These offsets also account for the variation in read depth across samples.
To perform the analysis with and without bias correction, we therefore compute two versions of the offsets, using the estimated bias vector and a dummy bias vector where the efficiencies are all set to 1.

```{r}
bias_list <- list(
  'Corrected' = bias, 
  'Uncorrected' = bias %>% mutate(Bhat = 1)
)

# 'effort' (offset) is calculated following jsdModels.R
effort_fun <- function(bias, phy) {
  (sample_sums(unbias(phy, bias)) %*% 
    t(bias$Bhat[match(taxa_names(phy), bias$Taxon)])
  ) %>% log 
}

effort_list <- bias_list %>%
  map(effort_fun, phy = phy)
```

The offset calculation is done as in `jsdModels.R`.
In section 'Quantification and statistical analysis'), @leopold2020host explain

> We accounted for 2 sources of unequal sampling effort, variable sampling depth and the species-specific sequencing biases, by including an offset term ($effort$) for each species $i$ in each sample $j$, in the form: $effort_{ij} = \log(bias_{i} \times depth_{j})$, where $bias_i$ is the sequencing bias correction factor for species $i$, estimated from our mock communities (see above, Estimating Sequencing Bias), and $depth_j$ is the total sum of all species in sample $j$, after dividing each by their species-specific sequencing bias correction factor.

The call to `sample_sums(unbias(phy, bias))` returns the $depth_{j}$ terms as described in this paragraph.

To understand why this calculation yields the correct offset, consider that under our model (and in our notation) we have
\begin{align}
  \log \text{reads}_i(a)
    &= \log \text{prop}_i (a) +
    \underbrace{\log \left[\frac{\text{efficiency}_i}{\text{efficiency}_S(a)} \cdot \text{reads}_S(a) \right]}_{\text{offset}}.
\end{align}
It can be shown with some algebra that
\begin{align}
  \frac{\text{reads}_S(a)}{\text{efficiency}_S(a)} 
  = \sum_i \frac{\text{reads}_i(a)}{\text{efficiency}_i}.
\end{align}
The left-hand side is the ratio of total reads to mean efficiency of the sample $a$, and the right-hand side is the $depth_{a}$ of sample $a$ defined by @leopold2020host.
It follows that the 'effort' terms computed by @leopold2020host are indeed the offsets implied by our model.

## Run analysis with each set of offsets

We'll wrap the analysis from `jsdModels.R` in a function to call with the corrected and uncorrected 'effort' offsets.
Toggle to see the code.

```{r, code_folding = TRUE}
analyze_jsd_models <- function(effort, phy, nBoot=4999) { 
  ##########################
  ### Fit genotype model ###
  ##########################

  # Make mvabund object 
  mvDat <- otu_table(phy) %>% data.frame  %>% mvabund

  # Fit joint-species model for genotype effect
  mv.full <- manyglm(mvDat ~ Genotype*Treatment, 
                     offset=effort, 
                     family="negative.binomial",
                     data=data.frame(sample_data(phy)))

  # Check model assumptions
  #plot(mv.full)
  #meanvar.plot(mvDat~sample_data(phy)$Treatment)
  #meanvar.plot(mvDat~sample_data(phy)$Genotype)

  # Test with anova.manyglm 
  # Using unstructured correlation matrix and wald tests.  
  # Including univariate test with adjustment for multiple testing.  
  mv.anova <- anova(mv.full, nBoot=nBoot, p.uni="adjusted", cor.type="shrink", test="wald")
  #> saveRDS(mv.anova, "output/rds/mv.genotype.rds")

  ########################
  ### Fit region model ###
  ########################

  #' # Fit joint-species model for genotype effect
  mv.region <- manyglm(mvDat ~ Region*Treatment, 
                       offset=effort, 
                       family="negative.binomial",
                       data=data.frame(sample_data(phy)))

  #' ## Check model assumptions
  #plot(mv.region)
  #meanvar.plot(mvDat~sample_data(phy)$Region)

  #' ## Test with anova.manyglm 
  #+ cache=T, results='asis'
  mv.region.anova <- anova(mv.region, nBoot=nBoot, p.uni="adjusted", cor.type="shrink", test="wald")
  #> saveRDS(mv.region.anova, "output/rds/mv.region.rds")

  mv.results <- bind_cols(
    mv.anova$table %>% 
      rownames_to_column() %>%
      filter(rowname!='(Intercept)') %>%
      mutate(rowname = gsub("Genotype","Host",rowname)),
    mv.region.anova$table %>% 
      rownames_to_column() %>%
      filter(rowname!='(Intercept)') %>%
      select(-rowname),
    # NOTE: [MRM] I set the name repair function to match the original behavior
    .name_repair = function(x) make.unique(x, sep = '')
  ) %>%
    gt(rowname_col = "rowname") %>%
      tab_spanner(
        label = "Genotype",
        columns = vars(Res.Df, Df.diff, wald, 'Pr(>wald)')
      ) %>%
      tab_spanner(
        label = "Ecotype",
        columns = vars(Res.Df1, Df.diff1, wald1, 'Pr(>wald)1')
      ) %>%
      fmt_number(c(4,8),
                 decimals = 1) %>%
      fmt(c(5,9),
          fns = function(x) {
            ifelse(x>=0.001,round(x,3),"< 0.001")
          }) %>%
      cols_label('Pr(>wald)'=md("*P*-value"),
                 wald=md("Wald-χ<sup>2<sup>"),
                 Res.Df=md("Df.resid"),
                 Df.diff="Df",
                 'Pr(>wald)1'=md("*P*-value"),
                 wald1=md("Wald-χ<sup>2<sup>"),
                 Res.Df1=md("Df.resid"),
                 Df.diff1="Df") %>%
      cols_move_to_start(3) %>%
      cols_move(7,5) %>%
      cols_align("center")
  #> gtsave(mv.results,"output/figs/jsdModels.png")

  list(
    mv.full = mv.full,
    mv.anova = mv.anova,
    mv.region = mv.region,
    mv.region.anova = mv.region.anova,
    mv.results = mv.results
  )
}
```

Finally, we run the analysis.
The call is wrapped in `xfun::cache_rds()` to cache the results; the hash is based on the inputs and the analysis function's source code.

```{r analyses, message = FALSE, warning = FALSE, results = 'hide'}
fun_src <- attr(analyze_jsd_models, 'srcref') %>% as('character')
res <- xfun::cache_rds({
  effort_list %>% map(analyze_jsd_models, phy = phy)
}, hash = list(fun_src, effort_list, phy),
  clean = FALSE
)
```

# Compare results

Next we compare the results of the analysis with and without bias correction.

```{r}
library(cowplot)
library(patchwork)
theme_set(theme_cowplot())
```

## ANOVA summary tables

First, let's compare the results summary tables.

**With bias correction**

```{r, echo = FALSE}
res$Corrected$mv.results
```

**Without bias correction**

```{r, echo = FALSE}
res$Uncorrected$mv.results
```

## Full model

**With bias correction**

```{r}
res[['Corrected']][['mv.full']]
```

**Without bias correction**

```{r}
res[['Uncorrected']][['mv.full']]
```

Let's compare the estimated coefficients for the various response variables.

```{r}
tidy_manyglm <- function(x) {
  x %>% coef %>%
    as_tibble(rownames = 'term') %>%
    pivot_longer(-term, names_to = 'response', values_to = 'estimate')
}
```

```{r, fig.dim = c(6, 6.5)}
coef_ests <- res %>%
  map('mv.full') %>%
  map_dfr(tidy_manyglm, .id = 'type')
x <- coef_ests %>%
  filter(term != '(Intercept)') %>%
  pivot_wider(names_from = type, values_from = estimate)
mse <- x %>%
  summarize(
    mse = mean((Corrected - Uncorrected)^2)
  ) %>% .[[1]] %>% signif(2)
x %>%
  ggplot(aes(Uncorrected, Corrected)) +
  coord_fixed() +
  geom_abline(color = 'darkred') +
  geom_point() +
  labs(title = 'Non-intercept coefficients, full model', 
    subtitle = str_glue('Mean squared difference: {mse}')
  ) +
  theme(plot.title.position = 'plot') +
  scale_color_brewer(type = 'qual', palette = 2)
```

## Region model

**With bias correction**

```{r}
res[['Corrected']][['mv.region']]
```

**Without bias correction**

```{r}
res[['Uncorrected']][['mv.region']]
```

Let's compare the estimated coefficients for the various response variables.

```{r, fig.dim = c(6, 6.5)}
coef_ests <- res %>%
  map('mv.region') %>%
  map_dfr(tidy_manyglm, .id = 'type')
x <- coef_ests %>%
  filter(term != '(Intercept)') %>%
  pivot_wider(names_from = type, values_from = estimate)
mse <- x %>%
  summarize(
    mse = mean((Corrected - Uncorrected)^2)
  ) %>% .[[1]] %>% signif(2)
x %>%
  ggplot(aes(Uncorrected, Corrected)) +
  coord_fixed() +
  geom_abline(color = 'darkred') +
  geom_point() +
  labs(title = 'Non-intercept coefficients, region model', 
    subtitle = str_glue('Mean squared difference: {mse}')
  ) +
  theme(plot.title.position = 'plot') +
  scale_color_brewer(type = 'qual', palette = 2)
```

Bias correction has little impact on the estimated coefficients for the non-intercept terms for either model.

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2022-01-07 supplemental-figures.Rmd
Add supplemental figure for fungal case study
Add supplemental figure showing variation in the proportions and mean
efficiency in the leopold2020host T1 samples.
Also, add a new function `include_svg()` to reduce code redundancy when
including svg figures.

@@
 # Supplemental figures
+
+<!-- Case studies: leopold2020host -->
+
+```{r leopold2020host-variation, fig.cap = '(ref:cap-leopold2020host-variation)'}
+fs::path(
+  'notebook/_posts/2022-01-06-leopold2020host-t1-samples',
+  'leopold2020host-t1-samples_files/figure-html5',
+  'variation-in-proportions-and-mean-efficiency-1.svg'
+  ) %>%
+  include_svg
+```
+
+(ref:cap-leopold2020host-variation) **In the pre-infection samples from @leopold2020host, multiplicative variation in taxa proportions is much larger than that in the mean efficiency.** Panel A shows the distribution of the proportions of each commensal isolate (denoted by its genus) across all samples collected prior to pathogen inoculation; Panel C shows the distribution of the (estimated) sample mean efficiency across these same samples on the same scale; and Panel B shows the efficiency of each taxon estimated from DNA mock communities as point estimates and 90% bootstrap percentile confidence intervals.
+Efficiencies are shown relative to the most efficiently measured taxon (_Fusarium_).
 
@@
 ```{r momspi-mean-efficiency-fcs, fig.cap = '(ref:cap-momspi-mean-efficiency-fcs)'}
-svg_path <- fs::path(
+fs::path(
   "notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files", 
   "figure-html5/momspi-mean-efficiency-fcs-1.svg"
-)
-if (knitr::is_html_output()) {
-  fig_path <- svg_path
-} else if (knitr::is_latex_output()) {
-  fig_path <- svg_to_pdf(svg_path)
-}
-knitr::include_graphics(fig_path)
-rm(svg_path, fig_path)
+) %>%
+  include_svg
 ```
@@
 ```{r sensitivity-example, fig.cap = '(ref:cap-sensitivity-example)'}
-svg_path <- fs::path(
+fs::path(
   "notebook/_posts/2021-10-18-evaluate-robustness-example/",
   "evaluate-robustness-example_files/figure-html5/summary_plot-1.svg"
-)
-if (knitr::is_html_output()) {
-  fig_path <- svg_path
-} else if (knitr::is_latex_output()) {
-  fig_path <- svg_to_pdf(svg_path)
-}
-knitr::include_graphics(fig_path)
-rm(svg_path, fig_path)
+) %>%
+  include_svg
 ```

### 2022-01-07 _common.R

@@ -39,3 +39,13 @@ svg_to_pdf <- function(svg_path) {
   system2("inkscape", args = args)
   fig_path
 }
+
+# include an svg file as a figure
+include_svg <- function(svg_path) {
+  if (knitr::is_html_output()) {
+    fig_path <- svg_path
+  } else if (knitr::is_latex_output()) {
+    fig_path <- svg_to_pdf(svg_path)
+  }
+  knitr::include_graphics(fig_path)
+}

### 2022-01-07 case-studies.Rmd
Fill in case-study details

@@
 
-To study the interactions between a host plant and its fungal symbionts and pathogen, @leopold2020host inoculated plants with 8 fungal species and, after allowing these fungal microbiomes to establish, exposed plants to a fungal pathogen.
-To investigate the role of colonization order and host genotype, plants varied in source location and in which symbiont was allowed to colonize first.
+Measurement of control communities along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis (@mclaren2019cons).
+Gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since, unlike most natural ecosystems, it is possible to assemble 'mock communities' containing all species in known relative abundances.
+
+@leopold2020host are the first to use mock-based calibration in a gnotobiotic microbiome experiment.
+In a study of host-commensal-pathogen interactions, @leopold2020host inoculated plants with 8 commensal fungal species and subsequently exposed plants to a fungal pathogen.
+To investigate the joint effect of colonization order and host genotype, plants varied in source location and in which commensal colonized first.
 The authors used ITS amplicon sequencing to measure communities before and after pathogen infection, along with mock communities with quantified genome concentrations of DNA from the 9 species.
 These DNA mock communities allowed the authors to directly measure the bias due to the sequencing workflow following DNA extraction and correct for it when analyzing the primary experimental samples following the method of @mclaren2019cons.
-Although extraction is not accounted for, considerable bias might still be expected due to the substantial ribosomal copy-number-variation (CNV) in fungi (CITE) and preferential PCR binding and amplification (CITE).
-The authors found a 13X difference between the most and least efficiently measured symbiont, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
-We reanalyzed this experiment to understand the impact that bias might have had on the study's analyses were it not accounted for (SI Analysis Fungi).
+Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification (CITE) and the substantial ribosomal copy-number-variation (CNV) in fungi (@lofgren2019geno).
+The authors found a 13X difference between the most and least efficiently measured commensal, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
+We reanalyzed this experiment to understand the impact that bias might have had on the study's analyses were it not accounted for.
 
-@leopold2020host used the ITS sequencing profiles of the pre-infection samples to understand the impact of host genotype and an experimental treatment---which species was allowed to colonize first---on the community composition after all species were later allowed to colonize.
-The authors used linear regression of log species proportions, using negative-binomial regression to account for random read-sampling error, to ask whether host genotype, treatment, and their interaction significantly impacted overall log species proportions and to quantify the advantage of being allowed to colonize first (priority effects).
-Bias was accounted for via the use of species- and sample-specific offset terms in the regression model (SI / appendix).
-In principle, bias could significantly impact the results of such regressions through variation in the log mean efficiency among samples.
-Yet bias has a negligible impact on the study's analyses, with nearly identical DA results being obtained regardless of whether bias correction is used (Ref SI, which will run the analyses from the paper with and without bias correction).
-To understand why, we compared the variation in log species proportions across samples to that in log mean efficiency (Figure X).
-Individual species show substantial variation in log proportion (give numbers); nevertheless, the log mean efficiency hardly varies (number; perhaps ref a figure).
-This difference is due to the communities always being dominated by just three species, whose efficiencies happen to be the highest among the commensal species and only vary by 3X (and just 1.5X between the two most dominant).
-The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions $\ll 1$, even after bias correction) and hence don't induce significant variation in the mean efficiency.
-Due to the stable mean efficiency across samples, bias primarily impacts the intercept terms and has a negligible impact on the estimates and standard errors of slope (DA) terms in the regressions.
+@leopold2020host used the ITS sequencing profiles of the pre-infection samples to understand the impact of host genotype and an experimental treatment---which species was allowed to colonize first---on the commensal community composition prior to infection.
+Two DA analyses were conducted to address these questions.
+First, the authors used linear regression of log species proportion (via negative-binomial regression to account for random read-sampling error) to ask whether host genotype, treatment, and their interaction significantly impacted overall community composition.
+Second, the authors quantified the strength of priority effects---the advantage of being allowed to colonize first---in different host genotypes by the LFC in the proportion of a species when allowed to colonize first versus later with the majority of commensals.
 
-Although not analyzed in the study, the authors also sequenced the fungal communities of plants after pathogen colonization.
-Comparison of the pre- and post-colonization samples suggests a simple, biologically relevant scenario where bias could confound attempts to infer microbial dynamics.
-A natural question to ask in such an experiment is, how do various species in the existing community respond to pathogen invasion?
-Relative abundances alone offer limited information, as they are unable to account for the potential sharp increase in the pathogen abundance post-infection, making such an experiment a natural target for absolute-abundance methods.
-One possibility would be to measure total genome concentration using qPCR of the single-copy gene that was used to quantify concentrations for the mock communities.
-Genome concentration could then be multiplied by ITS sequencing proportions to infer absolute abundances for DA analysis.
-Suppose the genome-concentration measurement is 100% accurate.
-In this case, it would be likely be important to account for bias in the ITS measurement.
-The reason is that the high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples (REF SI/FIG), which would tend to cause a concomitant 6X decrease in the infered fold changes in abundance in response to colonization for each species.
-Alternatively, one might measure total ITS concentration via qPCR, possibly using the same primers used for ITS sequencing.
-In this case, bias correction using the measured efficiencies is unnecessary and would be harmful if not also done for the qPCR measurement.
-The reason is that much (if not most) of the bias in the sequencing measurement is shared by the ITS qPCR measurement, including that due to extraction, CNV, and PCR amplification.
-Appendix X shows that this shared bias in fact leads to more accurate inferences of fold changes in the genome concentrations of individual species, even while likely giving inaccurate changes in total genome concentration.
-In summary, accurate inference of species abundances following pathogen colonization requires accounting for the increase in mean efficiency among samples; however, measuring total community abundance with a method (here, ITS qPCR) that is proportional to the mean efficiency naturally does so without needing to explicitly measure or correct for species efficiencies.
-Similarly, the (ratio-based) reference-species approaches discussed in the previous section might instead be used.
+Since these analysis are both based on proportions, bias could in principle impact their results.
+We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for^[The two sets of analyses can be seen at the following links: [negative-binomial regression analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html); [priority-effects analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html)].
+To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
+The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.57X and a geometric standard deviation of just 1.06X (Figure \@ref(fig:leopold2020host-variation)).
+<!-- Explanation -->
+Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
+This answer relates to the fact that the mean efficiency is a weighted _arithmetic_ average of species proprotions and so is insensitive to large multiplicative variation in low-proportion species.
+The pre-infection samples in the @leopold2020host study are always dominated by the three species with the highest efficiencies, whose efficiencies vary only by 3X (and just 1.5X between the two most dominant).
+The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions $\ll 1$, even after bias correction) and hence don't significantly affect the mean efficiency.
+<!-- Conclusion -->
+Because the log mean efficiency varies much less than the log proportions for each taxon, there is little room for it to distort the results of DA analyses based on variation in log proportions.
 
-<!-- The post-colonization samples were not analyzed by @leopold2020host in part because the reads from post-colonization samples tended to be dominated by the pathogen, giving little ability to infer the relative abundance of the rarer and lower efficiency species.
+<!-- Notes: -->
+<!-- - Consider whether to use log-transformed abundances and efficiencies throughout, and standard (non-geometric) ranges and standard deviations -->
+<!-- - Consider connecting back to the previous section and reminding of the condition for the variation in mean efficiency to impact regression analyses -->
+<!-- - Consider comparing the correlations among taxa with covariates; or at least looking at that of the mean efficiency, since correlation is also an important piece of the requirement -->
+<!-- - The priority-effect analysis isn't done as a regression, though it could have been (but would have been a slightly different calculation).
+I might want to explain this more here and in the notebook post.
 -->
-<!-- This was presumably due to high pathogen growth and aided by the pathogen's post-extraction efficiency being substantially larger than the other species; -->
-<!-- the high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples (REF SI/FIG). -->
+<!-- - I don't consider/mention whether there are actually 'significant' DA results; that would perhaps be a useful piece of the story.
+Or: I should perhaps say what the result was of each of the two analyses, so it is more concrete what results are not changing.
+-->
+<!-- - Consider expanding on the impact of bias, using Penicillium as an example: It's proportion is generally ~6X lower without correction, but its changes across samples are not affected by correction -->
 
-TODO:
+We additionally considered the potential for bias to impact analysis of how the abundances of commensal taxa changed across experimental timepoints in response to infection.
+First, we considered regression analysis where log proportion is the response variable.
+The observation of tissue lesions and a high observed proportion of the pathogen in the ITS profiles indicates that the pathogen grew to high abundance in most plants, so that we expect the proportion of commensals to typically decrease due to the sum-to-one constraint.
+As expected, regression analysis shows negative LFCs after infection regardless of whether bias is accounted for; however, LFCs are substantially lower (TODO) without bias correction.
+Because the pathogen tends to dominate ITS profiles in post-infection samples, the mean efficiency is much higher in post-infection samples, leading to artificially low LFCs for all species when bias is not accounted for.
 
-- Give numerical e.g. for the lowest eff species, Penicillium, whose prop is typically 6X too low but the magnitude of error in FCs is generally smaller
-- Check that strength of correlation with covariates isn't an important part of the story as well
-- Analysis: re-run DL analysis with and without calibrated offsets
+We then considered the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response.
+The actual experiment does not allow absolute-abundance measurement; nevertheless, it is illustrative to consider hypothetical scenarios in which absolute genome concentration were measured using one of the methods described in Section \@ref(abundance-measurement).
+In particular, we consider the total-community normalization method (Equation \@ref(eq:density-prop-meas)), for which shifts in MGS mean efficiency can lead to inaccurate LFCs.
+We consider two scenarios.
+_Scenario 1:_ Total genome concentration in each sample is perfectly known and used to measure the concentrations of individual taxa via Equation \@ref(eq:density-prop-meas).
+In this case, bias in the MGS measurements would create absolute errors in the LFCs for genome concentration identical to those for proportions.
+The scientific error, however, may be much worse.
+Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where taxa that persist or even increase appear to decrease.
+_Scenario 2:_ Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-community normalization method.
+In this case, we should expect the increase in mean efficiency to have little to no impact on LFCs, since it would be offset by the increase in mean efficiency of the total-concentration measurement.
+Moreover, since the total ITS concentration and ITS sequencing are both performed on the extracted DNA, we can also expect the effect of taxonomic bias caused by DNA sequencing to be offset.
+
+TODO: Write a concluding statement about how even in cases where the mean efficiency shifts,  the impact of bias in a proportion-based analysis depends on how exactly the total measurement is made and used (and an appropriate design can account for bias without the need for mock measurements).
+
+<!-- cut text -->
+<!-- - Genome concentration was chosen as the measure of abundance, due to the tendency for fungi to grow in forms where cells are not easily demarcated. -->
+<!-- - The original study focused on how host genotype and the pre-infection community affected disease severity after challenge with a pathogen, and did not analyze the impact of infection on commensals.
+-->
+<!-- - The post-colonization samples were not analyzed by @leopold2020host in part because the reads from post-colonization samples tended to be dominated by the pathogen, giving little ability to infer the relative abundance of the rarer and lower efficiency species.
+-->
+<!-- - This was presumably due to high pathogen growth and aided by the pathogen's post-extraction efficiency being substantially larger than the other species; the high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples (REF SI/FIG).
+-->
 
 ## Vaginal microbiomes of pregant women
-
-### Outline
 
@@
 We first examine what the taxonomic bias observed in control experiments by @brooks2015thet might imply for DA analysis of vaginal microbiomes of pregnant women reported by @fettweis2019thev.
-As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, @fettweis2019thev collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were measured by amplicon sequencing of the 16S V1-V3 hypervariable region to yield species-level taxonomic profiles.
-The effect of taxonomic bias in the 16S protocol had previously been investigated by @brooks2015thet via the construction and measurement of cellular mock communities of seven vaginally-relevant bacterial species.
-We reanalyzed these data in @mclaren2019cons using a new approach that makes it possible to measure the relative efficiencies among species from mock community measurements.
-Once relative efficiencies of all species are estimated, it is possible to quantitatively examine the impact of bias and to _calibrate_, or correct the effect of bias, in samples with unknown composition (@mclaren2019cons).
-To account for species not in the mock communities, we used a simple method to impute the relative efficiencies of remaining species using their taxonomic relationship to the mock species.
+As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, @fettweis2019thev collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were measured by amplicon sequencing of the 16S V1-V3 hypervariable region to yield species-level bacterial taxonomic profiles.
+DA analyses of the vaginal microbiome, including those of @fettweis2019thev, are most commonly based on species proportions, creating opportunity for bias to have an impact on their results.
+Previously, the authors evaluated taxonomic bias of this MGS protocol via the construction and measurement of cellular mock communities of seven vaginal bacterial species (@brooks2015thet), but did not directly account for this bias in @fettweis2019thev.
+We used the control measurements from @brooks2015thet to infer the bias of the MOMS-PI 16S protocol and investigate its potential impact on DA analyses of vaginal profiles from the MOMS-PI study.
+To estimate the relative efficiencies among all species, we directly estimated the efficiencies in the control samples following the method of @mclaren2019cons and imputed the remaining efficiencies based on taxonomic relationships.
+We then used these imputed efficiencies to calibrate the measured taxonomic profiles from the MOMS-PI study and compare DA analyses with and without calibration.
+Such imputed efficiencies carry considerable uncertainty, which we ignore to simplify our presentation. 
+Notably, while this uncertainty can majorly impact abundances in individual samples, it has a much weaker impact on LFCs across samples (SI analysis).
 
-maybe say more about this
-- Such imputation carries with it substantial uncertainty; however, 
-- These species collectively constitute a substantial majority of taxomically assigned reads in the MOMSPI 16S profiles (X for species level, and Y when expand to genus).
-- analysis of sensitivty to uncertainty can be done and shows not a very big impact.
+To assess the potential for bias to affect fold changes in species proportions, we first used the imputed efficiencies to consider how the mean efficiency varied across women.
+In contrast to the fungal experiment, the mean efficiency varies substantially across vaginal samples, primarily driven by variation in which species is most abundant (Figure \@ref(fig:momspi-mean-efficiency-dist).
+The four most common top species in the observed and calibrated profiles were _Gardnerella vaginalis_, _Lactobacillus iners_, _Lactobacillus crispatus_, and _Lachnospiraceae BVAB1_.
+All of these except _Lachnospiraceae BVAB1_ where included in the mock-community experiment, from which we estimated that _L.
+iners_ has an efficiency of 2X relative to _L.
+crispatus_ and 30X relative to _G.
+vaginalis_, with _L.
+iners_ and _G.
+vaginalis_ having the maximum and minimum efficiencies among the 7 control species.
 
-Efficiency varied subsantially among control species in the @brooks2015thet experiment, with an approximately 30X advantage of the highest efficiency _L.
-iners_ over the lowest efficiency _G.
-vaginalis_.
-To assess the impact of bias on the observed fold changes in species proportions, we used our imputed efficiencies to consider how the mean efficiency varied across women.
-The mean efficiency varies substantially across samples (SI Figure \@ref(fig:momspi-mean-efficiency-fcs; SUMMARY STATS), with greater variation between samples from different women than samples from the same woman (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
-Vaginal microbiomes are often dominated by a single species often forming a vast majority of calibrated or uncalibrated read counts, with _Lactobacillus crispatus_, _Lactobacillus iners_, _G.
-vaginalis_, and BVAB1 being the most common (CHECK).
+HERE
+- finish explaining how the variation is driven by changes in these taxa, esp G and L 
+- add that the importance of these taxa, and the tendency for one species to domiannt, is a gneral well-suported feature of vaginal mb studies, though the identities varies by population and mgs protocol.
+- segue into further considerations - what is the impact of this variation?
+  - trajectories
+    - impact of big shifts 
+  - DA analysis
+
+(TODO: confirm true that often these species dominate. Also check the efficiency numbers)
+
 Consequently, the mean efficiency tends to track that of most dominant species, with _Lactobacillus_-dominated samples typically having an efficiency that is 3-20X greater _Gardnerella_-dominated samples (Figure \@ref(fig:momspi-mean-efficiency-dist)).
+
 Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)), and typically result in substantial fold changes in mean efficiency.
@@
 
-## Marine sediment growers
+## Growth in marine sediments
 

### 2022-01-08 notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study.Rmd
Add infection analysis to fungal case study
Add infection analysis; update date, title, and description; make minor
overall improvements to analysis.

---
title: "Analysis of Leopold and Busby (2020) for manuscript case study"
description: This document performs analysis of the data from @leopold2020host to support the fungal case study in the manuscript.
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - R
  - ref:leopold2020host
  - differential abundance
date: 2022-01-08
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    toc: true
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

# Setup

```{r libraries}
library(here)
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
# ggplot helpers
library(ggbeeswarm)
library(ggdist)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot())
# stats helpers
library(broom)

library(metacal)

# library(rstanarm)
# options(mc.cores = 4)
```

```{r}
# color pallets from leopold2020host
pal_treatment <- c("#2F4F4F", "#E69F00", "#56B4E9", "#009E73", "#F0E442")
pal_region <- c("#A2CD5A","#6E8B3D")
pal_genotype <- c("#800000FF", "#767676FF", "#CC8214FF", "#616530FF",
  "#0F425CFF", "#9A5324FF", "#3E3E23FF", "#D6D6CEFF", "#FFB547FF", "#ADB17DFF",
  "#5B8FA8FF", "#D49464FF")
```

```{r load-data, code_folding = TRUE}
## From the metacal 2.0 tutorial
# Download data from https://zenodo.org/record/3872145
data_path <- here::here("notebook/_data", "leopold2020host")
# To use a temporary directory:
# data_path <- file.path(tempdir(), "leopold2020")
if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  download.file(
    "https://zenodo.org/record/3872145/files/dleopold/Populus_priorityEffects-v1.2.zip",
    file.path(data_path, "Populus_priorityEffects-v1.2.zip")
  )
  unzip(
    file.path(data_path, "Populus_priorityEffects-v1.2.zip"), 
    exdir = data_path
  )
}

mock_actual <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/data/MockCommunities.csv") %>%
  read.csv(row.names = 1) %>%
  select(-Sym4) %>%
  as("matrix") %>%
  otu_table(taxa_are_rows = FALSE) %>%
  transform_sample_counts(function(x) close_elts(1 / x))
mock_taxa <- taxa_names(mock_actual)
commensal_taxa <- setdiff(mock_taxa, 'Melampsora')

# Rust severity
rust <- file.path(data_path, 'dleopold-Populus_priorityEffects-8594f7c',
  'data/rust_measurements.csv'
) %>%
  read_csv %>%
  select(-1)
# summarize following approach from leopold 
rust_summary <- rust %>%
  janitor::clean_names() %>%
  rename(plant_id = samp_id) %>%
  with_groups(plant_id, summarize, 
    # across(ends_with('cm2'), mean)
    rust_pct = sum(rust_cm2)/sum(leaf_cm2),
    lesion_pct = sum(lesion_cm2) / sum(leaf_cm2),
    leaf_area = sum(leaf_cm2),
    leaf_n = n()
  )

#> The microbiome data is stored in a phyloseq object,
ps_raw <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/output/compiled/phy.rds") %>%
  readRDS %>%
  filter_tax_table(.otu %in% mock_taxa) %>%
  mutate_sample_data(., 
    Timepoint = factor(Timepoint),
    sample_sum = sample_sums(.)
  ) %>%
  mutate_sample_data(
    plant_id = str_extract(.sample, 'G[0-9]+\\.T[0-9]+\\.R[0-9]+')
  ) %>%
  left_join_sample_data(
    rust_summary, by = 'plant_id'
  ) %>%
  rename_with_sample_data(janitor::make_clean_names)

ps_mock <- ps_raw %>% 
  subset_samples(samp_type == "Mock") %>%
  prune_taxa(mock_taxa, .)
```

Check sample sums,

```{r}
sam <- ps_raw %>% sample_data %>% as_tibble
sam %>%
  as_tibble %>%
  ggplot(aes(sample_sum, fill = samp_type)) +
  scale_x_log10() +
  geom_histogram() +
  scale_fill_brewer(type = 'qual')
```

```{r}
sam %>%
  filter(sample_sum < 3e3) %>%
  select(.sample, timepoint, treatment, sample_sum)
```

The sample with less than 3000 reads is in the 'Negative' treatment group and so will be ignored in our analysis of the T1 samples.

Let's check the distribution of proportions of the mock taxa across samples.

```{r}
ps_raw %>% 
  filter_sample_data(samp_type == "Experiment", treatment != 'Negative') %>%
  filter_tax_table(.otu %in% mock_taxa) %>%
  transform_sample_counts(close_elts) %>%
  as_tibble %>%
  ggplot(aes(y = .otu, x = .abundance + 1e-5)) +
  facet_wrap(~timepoint) +
  scale_x_log10() +
  stat_slab()
```

T1 samples have reads of Melampsora, which may be contamination.
For our analysis, let's subset to the mock taxa and just the experimental samples, and set Melampsora in T1 to zero.

```{r}
otu_new <- ps_raw %>% 
  filter_tax_table(.otu %in% mock_taxa) %>%
  filter_sample_data(samp_type == "Experiment", treatment != 'Negative') %>%
  as_tibble %>%
  mutate(.abundance = ifelse(
      timepoint == 1 & .otu == 'Melampsora', 0, .abundance
  )) %>%
  select(.sample, .otu, .abundance) %>%
  pivot_wider(names_from = .otu, values_from = .abundance) %>%
  otu_table(taxa_are_rows = FALSE)
ps <- ps_raw
otu_table(ps) <- otu_new
rm(otu_new)
```

Confirm that Mel is zero'd in the t1 samples,

```{r}
ps %>% 
  prune_taxa('Melampsora', .) %>%
  as_tibble %>%
  with_groups(timepoint, summarize,
    across(.abundance, list(median = median, perc_pos = ~mean(. > 0) ))
  )
```

## Bias estimation

We estimate bias using `metacal::estimate_bias()`; see the [metacal tutorial](https://mikemc.github.io/metacal/articles/tutorial.html) for more information.
@leopold2020host identified the sample Mock.5 as an outlier sample and excluded it when estimating bias.
One taxon in this sample, Epicoccum, was undetected, and including it leads to a larger standard error especially for this taxon (as seen in the metacal tutorial).
We could improve our bias estimates by setting this taxon's nominal abundance to 0 in this sample; however, let's just drop this sample to maintain consistent bias estimates with those used in the original study.

```{r estimate-bias}
control_samples <- ps_mock %>% sample_names %>% setdiff('Mock.5')
mc_fit <- estimate_bias(
  ps_mock %>% prune_samples(control_samples, .), 
  mock_actual %>% prune_samples(control_samples, .), 
  boot = TRUE
)
bias <- coef(mc_fit) %>% print(digits = 2)
```

```{r, fig.dim = c(5,3)}
mc_fit_summary <- summary(mc_fit)
coef_tb <- mc_fit_summary$coefficients
coef_tb %>%
  mutate(taxon = fct_reorder(taxon, estimate)) %>%
  ggplot(aes(taxon, estimate, 
      ymin = estimate / gm_se^2, ymax = estimate * gm_se^2)) +
  geom_hline(yintercept = 1, color = "grey") +
  geom_pointrange() +
  scale_y_log10() +
  coord_flip()
```

# T1 samples

From our previous analyses, we know that there is little impact of bias correction on the primary analyses done on these samples, and also that there is fairly little variation in the mean efficiency in the T1 samples.
Here we further explore the variation in both the species proportions and the mean efficiency to more fully understand why bias has little impact on DA analysis.

## Setup

Here we subset to the T1 experimental samples and the commensal taxa.
The analysis of T1 samples by @leopold2020host first subsets to just the 8 commensal (non-pathogenic) inoculum taxa, and we follow suit here.
It excludes Melampsora and OTUs not identified with the 9 inoculum taxa.
In addition, we add the sample sums and mean efficiency to the sample data, after subsetting to the commensals.

```{r}
# Bias among the commensals; for convenience, set relative to max
bias_t1 <- bias[names(bias) != 'Melampsora']
bias_t1 <- bias_t1 / max(bias_t1)
most_efficient_taxon <- which.max(bias_t1) %>% names
stopifnot(most_efficient_taxon == 'Fusarium')

ps_t1 <- ps %>%
  filter_sample_data(timepoint == 1) %>%
  filter_tax_table(.otu %in% commensal_taxa) %>%
  mutate_sample_data(., 
    sample_sum = sample_sums(.),
    mean_efficiency = transform_sample_counts(., close_elts) %>%
      perturb(bias_t1, norm = 'none') %>% 
      sample_sums,
  )
sam_t1 <- ps_t1 %>% sample_data %>% as_tibble
```

A small fraction (`r ps_t1 %>% otu_table %>% as('matrix') %>% {mean(. == 0)} %>% signif(digits = 3)`) of the observed counts are zero. 
For calculations that ignore the counting error, we want to ensure there are only positive abundances.
We will replace zeros using `zCompositions::cmultRepl()` following @leopold2020host, then recompute the mean efficiency (which should only change very slightly).

```{r}
stopifnot(taxa_are_rows(ps_t1) == FALSE)
otu_new <- ps_t1 %>% 
  otu_table %>% 
  as('matrix') %>%
  zCompositions::cmultRepl(
    method = 'GBM', 
    output = 'p-counts', 
    suppress.print = TRUE
  )
ps_t1_repl <- ps_t1
otu_table(ps_t1_repl) <- otu_table(otu_new, taxa_are_rows = FALSE)
rm(otu_new)
ps_t1_repl <- ps_t1_repl %>%
  mutate_sample_data(., 
    mean_efficiency_initial = mean_efficiency,
    mean_efficiency = transform_sample_counts(., close_elts) %>%
      perturb(bias_t1, norm = 'none') %>% 
      sample_sums,
  )
```

Let's create a data frame with the observed and calibrated proportions of each species, with their efficiencies added in.

```{r}
props <- ps_t1_repl %>%
  as_tibble %>%
  left_join(bias_t1 %>% enframe('.otu', 'efficiency'),
    by = '.otu'
  ) %>%
  with_groups(.sample, mutate, 
    prop_obs = close_elts(.abundance),
    prop_cal = close_elts(.abundance / efficiency),
  ) %>%
  mutate(
    .otu_fct = fct_reorder(.otu, efficiency)
  )
```

## Variation in mean efficiency

Let's examine the variation in mean efficiency

```{r, fig.dim = c(7,4)}
lyrs <- list(
  geom_histogram(),
  scale_x_log10(),
  expand_limits(x = 1.2),
  scale_fill_brewer(type = "qual"),
  labs(
    x = "Mean efficiency", 
    y = "Number of samples",
    # fill = ''
  )
)

p1 <- sam_t1 %>%
  ggplot(aes(mean_efficiency, fill = region)) +
  lyrs
p2 <- sam_t1 %>%
  ggplot(aes(mean_efficiency, fill = treatment)) +
  lyrs
p1 / p2
```

The relative efficiency has a geometric range of `r bias_t1 %>% gm_range %>% signif(3)`-fold and a geometric standard deviation of `r bias_t1 %>% gm_sd %>% signif(3)`-fold.
But the mean efficiency's geometric range is only `r sam_t1$mean_efficiency %>% gm_range %>% signif(3)`-fold and its geometric standard deviation is just `r sam_t1$mean_efficiency %>% gm_sd %>% signif(3)`-fold.

## Variation in proportions vs mean efficiency

First let's create some summary statistics of the calibrated proportions, to compare to those of the mean efficiency.

```{r}
props %>%
  with_groups(.otu_fct, summarize,
    across(prop_cal, 
      list(gm_mean = gm_mean, gm_sd = gm_sd, gm_range = gm_range),
      .names = '{.fn}'
    )
  ) %>%
  arrange(desc(.otu_fct)) %>%
  mutate(across(-.otu_fct, signif, 3)) %>%
  knitr::kable()
```

Now let's create the main figure showing the variation in taxa proportions against that in the mean efficiency, with the individual taxa efficiencies also shown for more context.

Caption: **In the pre-infection samples from @leopold2020host, multiplicative variation in taxa proportions is much larger than that in the mean efficiency.**
Panel A shows the distribution of the proportions of each commensal isolate (denoted by its genus) across all samples collected prior to pathogen inoculation;
Panel C shows the distribution of the (estimated) sample mean efficiency across these same samples on the same scale; 
and Panel B shows the efficiency of each taxon estimated from DNA mock communities as point estimates and 90% bootstrap percentile confidence intervals.
Efficiencies are shown relative to the most efficiently measured taxon (Fusarium).

```{r}
rng <- props %>% pull(prop_cal) %>% range
## species proportions
p_props <- props %>%
  ggplot(aes(y = .otu_fct, x = prop_cal)) +
  scale_x_log10() +
  stat_slab() +
  labs(x = 'Proportion', y = NULL)
## mean efficiency
p_me <- sam_t1 %>%
  ggplot(aes(y = 'Mean efficiency', x = mean_efficiency)) +
  scale_x_log10() +
  expand_limits(x = rng) +
  stat_slab() +
  labs(x = 'Efficiency relative to Fusarium', y = NULL)
# p_me / p_props + plot_layout(heights = c(0.1, 1))
## Estimated efficiency with bootstrap CIs
x <- mc_fit$bootreps %>%
  data.table::as.data.table(keep.rownames = '.id') %>%
  select(.id, all_of(commensal_taxa)) %>%
  mutate(across(-.id, ~. / Fusarium)) %>%
  pivot_longer(-.id, names_to = 'taxon', values_to = 'efficiency') 
p_eff <- x %>%
  mutate(taxon = fct_reorder(taxon, efficiency)) %>%
  ggplot(aes(y = taxon, x = efficiency)) +
  stat_pointinterval(.width = 0.9) +
  scale_x_log10() +
  labs(x = str_glue('Efficiency relative to Fusarium'), y = NULL)
```

```{r variation-in-proportions-and-mean-efficiency, fig.dim = c(8.5, 6.5), preview = TRUE}
p_props + p_eff + p_me + plot_spacer() + 
  plot_layout(heights = c(1, 0.15), widths = c(1, 0.8), ncol = 2) +
  plot_annotation(tag_levels = 'A')
```

# Change in commensals due to infection

Goal: Do a linear regression analysis of change in log proportion following pathogen colonization (from T1 to T2).

Note, I am keeping in the pathogen Melampsora for this analysis.
The increase in Mel proportion post infection is a major driver of the impact of bias, since it causes the mean efficiency to increase.
If we instead considered the proportions among just the commensals, then we would get different results.
The results here, with Mel included, are the most informative for what would happen if we were performing absolute-abundance analysis using total-community normalization.

## Setup

For doing simple linear regression, we must replace zeros with positive values.
I will use `zCompositions::cmultRepl()` as in @leopold2020host for simplicity and consistency with our other analyses in this case study.
I will recompute the mean efficiency with these new compositions; however, 
aero replacement will have a minimal impact on the mean efficiency.

```{r}
bias_inf <- bias/bias['Fusarium']
ps_inf <- ps
stopifnot(taxa_are_rows(ps_inf) == FALSE)
otu_new <- ps_inf %>% 
  otu_table %>% 
  as('matrix') %>%
  zCompositions::cmultRepl(
    method = 'GBM', 
    output = 'p-counts', 
    suppress.print = TRUE
  )
ps_inf_repl <- ps_inf
otu_table(ps_inf_repl) <- otu_table(otu_new, taxa_are_rows = FALSE)
rm(otu_new)
ps_inf_repl <- ps_inf_repl %>%
  mutate_sample_data(., 
    mean_efficiency = transform_sample_counts(., close_elts) %>%
      perturb(bias_inf, norm = 'none') %>% 
      sample_sums,
  )
```

## Variation in mean efficiency

From our previous explorations, we know that the mean efficiency tends to increase post infection due to the higher proportion of Melampsora.

Let's see this again,

```{r}
mel <- ps_inf %>% 
  calibrate(bias = bias_inf) %>%
  transform_sample_counts(close_elts) %>%
  prune_taxa("Melampsora", .) %>%
  otu_table %>%
  as_tibble %>%
  select(.sample, melampsora_prop = .abundance)
sam <- ps_inf_repl %>% sample_data %>% as_tibble %>%
  left_join(mel, by = '.sample')
p_hist <- sam %>%
  ggplot(aes(fill = timepoint, x = mean_efficiency)) +
  scale_x_log10() +
  geom_histogram() +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  )
p_mel <- sam %>%
  ggplot(aes(y = melampsora_prop, x = mean_efficiency, color = timepoint)) +
  geom_point() +
  scale_x_log10() +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 1e-2),
    breaks = c(0, 0.01, 0.1, 0.5, 1)
    ) +
  scale_color_brewer(type = "qual") +
  labs(
    y = "Pathogen proportion",
    x = "Mean efficiency",
    color = 'Timepoint'
  )
```

```{r infection-mean-efficiency-dist, fig.dim = c(6, 6)}
p_hist / p_mel +
  plot_annotation(tag_levels = "A")
```

Let's visualize the pre and post infection mean eff's by host genotype.

```{r}
ps_inf_repl %>%
  sample_data %>%
  as_tibble %>%
  ggplot(aes(y = genotype, x = mean_efficiency, 
      color = timepoint)) +
  scale_x_log10() +
  geom_boxplot() +
  geom_quasirandom(groupOnX = FALSE) +
  scale_color_brewer(type = 'qual', palette = 3)
```

The mean efficiency generally increases by a similar factor in each genotype.

Let's measure the average fold increase in each genotype.

```{r}
mean_eff_reg <- ps_inf_repl %>%
  sample_data %>%
  as_tibble %>%
  group_by(genotype) %>%
  nest %>%
  mutate(
    fit = map(data, ~lm(log(mean_efficiency) ~ timepoint, data = .x))
  ) %>%
  select(-data) %>% 
  mutate(across(starts_with('fit'), map, tidy)) %>%
  unnest(fit)
```

```{r}
x <- mean_eff_reg %>%
  filter(term == 'timepoint2') %>%
  mutate(across(c(estimate, std.error), exp))
x %>% select(genotype:std.error) %>% arrange(genotype) %>% 
  knitr::kable(digits = 3)
x %>%
  ggplot(aes(y = genotype, x = estimate)) +
  expand_limits(x = c(1, 6)) +
  scale_x_log10() +
  geom_vline(xintercept = 1, color = 'grey') +
  geom_point() +
  geom_linerange(
    aes(
      xmin = estimate * std.error^2,
      xmax = estimate / std.error^2
    )
  )
```

The FC in mean efficiency ranges from `r min(x$estimate) %>% signif(3)` to `r max(x$estimate) %>% signif(3)`.

## Simple linear regression of log proportion

```{r}
x <- ps_inf_repl %>% 
  as_tibble %>%
  left_join(bias_inf %>% enframe('.otu', 'efficiency'), by = '.otu')
```

Run a simple linear regression with lm on log proportion,

```{r}
x_reg <- x %>%
  with_groups(.sample, mutate,
    log_prop = log(close_elts(.abundance))
  ) %>%
  filter(.otu %in% commensal_taxa) %>%
  mutate(
    offset_obs = 0,
    offset_cal = log(efficiency / mean_efficiency),
  ) %>%
  group_by(.otu, genotype) %>%
  nest() %>%
  mutate(
    fit_obs = map(data, 
      ~lm(
        log_prop ~ timepoint + offset(offset_obs), 
        data = .x,
      )),
    fit_cal = map(data, 
      ~lm(
        log_prop ~ timepoint + offset(offset_cal), 
        data = .x,
      )),
  )
```

```{r}
z <- x_reg %>%
  select(-data) %>% 
  mutate(across(starts_with('fit'), map, tidy, conf.int = TRUE)) %>%
  pivot_longer(starts_with('fit'), names_to = 'type', values_to = 'fit', 
    names_prefix = 'fit_') %>%
  unnest(fit) %>%
  ungroup %>%
  left_join(enframe(bias, '.otu', 'efficiency'), by = '.otu')
```

```{r}
z1 <- z %>%
  filter(term == 'timepoint2') %>%
  select(.otu, genotype, type, estimate, std.error) %>%
  pivot_wider(names_from = type, values_from = c(estimate, std.error)) %>%
  glimpse
```

Check that the corrections are the same for each taxon within a genotype, but differ by genotype.

```{r, fig.dim = c(6,6)}
z1 %>%
  ggplot(aes(y = .otu, x = estimate_cal - estimate_obs, color = .otu)) +
  scale_color_manual(values = pal_genotype, guide = NULL) +
  facet_wrap(~genotype) +
  stat_pointinterval()
```

The corrections are the same within genotypes, as expected, and vary to a small degree between genotypes.
They tend to be larger for Eastern genotypes; this matches with the finding that Eastern genotypes are more susceptible to rust, and so we should expect the Melampsora proportions to be larger.

Let's visualize the impact of bias correction on the coefficients, in a couple ways.
First, let's plot the uncorrected against the corrected estimates.

```{r, fig.dim = c(10, 8), layout="l-body-outset"}
z1 %>%
  ggplot(aes(estimate_cal, estimate_obs, color = .otu)) +
  scale_color_manual(values = pal_genotype) +
  facet_wrap(~genotype, ncol = 4) +
  theme_minimal_grid() +
  theme() +
  coord_fixed() +
  scale_x_continuous(breaks = seq(-6, 2, 2)) +
  labs(
    x = 'Estimate with bias correction',
    y = 'Estimate without bias correction',
    color = 'Taxon'
  ) +
  geom_hline(yintercept = 0, color = 'black') +
  geom_vline(xintercept = 0, color = 'black') +
  geom_abline(color = 'darkred') +
  geom_linerange(
    aes(
      xmin = estimate_cal - 2 * std.error_cal,
      xmax = estimate_cal + 2 * std.error_cal,
    )
  ) +
  geom_linerange(
    aes(
      ymin = estimate_obs - 2 * std.error_obs,
      ymax = estimate_obs + 2 * std.error_obs
    )
  ) +
  geom_point(size = 2)
```

This figure makes it clear how the constant shift maintains a perfect correlation between estimates.
But it is perhaps easier to see the biological impact on the results if we plot the two estimate types in a straight line,

```{r infection-lfc-estimates, fig.dim = c(8, 8)}
z %>%
  filter(term == 'timepoint2') %>%
  mutate(
    across(.otu, fct_reorder, estimate),
    across(type, fct_recode, Corrected = 'cal', Uncorrected = 'obs')
  ) %>%
  ggplot(aes(y =.otu, color = type,
      x = estimate, 
      xmin = estimate - 2 * std.error,
      xmax = estimate + 2 * std.error
  )) +
  # theme_minimal_hgrid() +
  theme(
    axis.line.y = element_blank()
  ) +
  colorblindr::scale_color_OkabeIto() +
  facet_wrap(~genotype, ncol = 4) +
  labs(
    x = 'Estimated LFC',
    y = 'Taxon',
    color = 'Type'
  ) +
  geom_vline(xintercept = 0, size = 0.5, color = 'grey') +
  geom_pointrange(fatten = 2)
```

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2022-01-08 supplemental-figures.Rmd
Add supplemental figures for fungal case study
Add two additional supplemental figures for the fungal case study, for
the analysis of how the commensals respond to infection.

@@
 fs::path(
-  'notebook/_posts/2022-01-06-leopold2020host-t1-samples',
-  'leopold2020host-t1-samples_files/figure-html5',
+  'notebook/_posts/2022-01-08-leopold2020host-case-study',
+  'leopold2020host-case-study_files/figure-html5',
   'variation-in-proportions-and-mean-efficiency-1.svg'
@@
 Efficiencies are shown relative to the most efficiently measured taxon (_Fusarium_).
+
+```{r leopold2020host-infection-mean-efficiency-dist, fig.cap = '(ref:cap-leopold2020host-infection-mean-efficiency-dist)'}
+fs::path(
+  'notebook/_posts/2022-01-08-leopold2020host-case-study',
+  'leopold2020host-case-study_files/figure-html5',
+  'infection-mean-efficiency-dist-1.svg'
+  ) %>%
+  include_svg
+```
+
+(ref:cap-leopold2020host-infection-mean-efficiency-dist) **The mean efficiency tends to increase after infection due to the high proportion of the pathogen.**
+
+```{r leopold2020host-infection-lfc, fig.cap = '(ref:cap-leopold2020host-infection-lfc)'}
+fs::path(
+  'notebook/_posts/2022-01-08-leopold2020host-case-study',
+  'leopold2020host-case-study_files/figure-html5',
+  'infection-lfc-estimates-1.svg'
+  ) %>%
+  include_svg
+```
+
+(ref:cap-leopold2020host-infection-lfc) **Bias correction increases the estimated increase in log proportion in response to infection for commensal taxa across all host genotypes.** Shown are the estimated log fold change (LFC) and 95% confidence intervals from simple linear regression of log (base e) proportion against experimental timepoint for commensal taxa.
+Negative values indicate that the proportion of the taxon decreased on average in response to infection, which we expect due to the increase in pathogen abundance and the sum-to-one constraint.
+Bias leads to artificially low estimates, as the pathogen drives an increase in mean efficiency.
+In Western genotypes, several taxa whose proportion is observed to decrease without correction are instead found to remain stable or even increase when bias is corrected.
 

### 2022-01-08 case-studies.Rmd
Revise fungal case study's infection analysis

@@
 Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
-This answer relates to the fact that the mean efficiency is a weighted _arithmetic_ average of species proprotions and so is insensitive to large multiplicative variation in low-proportion species.
+This answer relates to the fact that the mean efficiency is a weighted _arithmetic_ average of species proportions and so is insensitive to large multiplicative variation in low-proportion species.
 The pre-infection samples in the @leopold2020host study are always dominated by the three species with the highest efficiencies, whose efficiencies vary only by 3X (and just 1.5X between the two most dominant).
@@
 
-We additionally considered the potential for bias to impact analysis of how the abundances of commensal taxa changed across experimental timepoints in response to infection.
-First, we considered regression analysis where log proportion is the response variable.
-The observation of tissue lesions and a high observed proportion of the pathogen in the ITS profiles indicates that the pathogen grew to high abundance in most plants, so that we expect the proportion of commensals to typically decrease due to the sum-to-one constraint.
-As expected, regression analysis shows negative LFCs after infection regardless of whether bias is accounted for; however, LFCs are substantially lower (TODO) without bias correction.
-Because the pathogen tends to dominate ITS profiles in post-infection samples, the mean efficiency is much higher in post-infection samples, leading to artificially low LFCs for all species when bias is not accounted for.
+We additionally considered the potential for bias to impact analysis of how commensal taxa responded to infection.
+In particular, it is interesting to consider whether any commensals are able to increase in absolute concentration in response to infection.
+The experiment performed by @leopold2020host does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration
+measured using the total-community normalization method (Equation \@ref(eq:density-prop-meas)).
+It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
+The pathogen is absent in pre-infection samples but tends to dominate the community post-infection (Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)), leading to an average increase in mean efficiency ranging from 3.5X to 5X across host genotypes and an expectation that commensals will typically decrease in proportion due to the sum-to-one constraint.
+Without bias correction, the log proportion of commensals decrease in each host genotype, often substantially (log-e fold changes less than -2); however, bias leads to overly negative LFCs due to the higher mean efficiency in post-infection samples.
+Bias correction substantially lessens the magnitude of LFC estimates, in some cases leading to estimates that are approximately 0 or slightly positive (Figure \@ref(fig:leopold2020host-infection-lfc)).
 
-We then considered the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response.
-The actual experiment does not allow absolute-abundance measurement; nevertheless, it is illustrative to consider hypothetical scenarios in which absolute genome concentration were measured using one of the methods described in Section \@ref(abundance-measurement).
-In particular, we consider the total-community normalization method (Equation \@ref(eq:density-prop-meas)), for which shifts in MGS mean efficiency can lead to inaccurate LFCs.
+Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-community normalization method (Equation \@ref(eq:density-prop-meas)).
 We consider two scenarios.
@@
 Moreover, since the total ITS concentration and ITS sequencing are both performed on the extracted DNA, we can also expect the effect of taxonomic bias caused by DNA sequencing to be offset.
-
-TODO: Write a concluding statement about how even in cases where the mean efficiency shifts,  the impact of bias in a proportion-based analysis depends on how exactly the total measurement is made and used (and an appropriate design can account for bias without the need for mock measurements).
+Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-community measurement is made prior to normalization.
 

### 2022-01-10 notebook/_posts/2022-01-08-lloyd2020evi-case-study/lloyd2020evi-case-study.Rmd
Perform calculations for marine case study

---
title: "Analysis of Lloyd et al (2020) for manuscript case study"
description: Perform calculations for the Lloyd et al (2020) case study.
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:lloyd2020evid
  - differential abundance
date: 2022-01-08
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    toc: true
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

# Setup

Load libraries,

```{r libraries}
library(here)
library(tidyverse)
# library(tidyxl)
library(unpivotr)
```

Our source data is Table 1 from @lloyd2020evid, which I've manually input into a CSV file.
This table has the form of a 'pivot table' with nested headers; we can properly read it in and create a 'tidy' version table using the unpivotr package, following [this example](https://nacnudus.github.io/spreadsheet-munging-strategies/pivot-simple.html).
First we'll read it in with one row per spreadsheet cell,

```{r}
all_cells <- here('notebook/_data/lloyd2020evid/table1.csv') %>%
  read_csv(col_names = FALSE) %>%
  as_cells %>%
  # filter(!is.na(hr), chr != 'Clade') %>%
  glimpse
```

Confirm that this data matches Table 1,

```{r}
all_cells %>% rectify %>% knitr::kable()
```

Note, the values of the table are doubling times.

Now, munge the data into tidy form, with one row per measurement.
Additionally, compute the doubling rate as the reciprocal of the doubling time.

```{r}
# abbreviations: exp for experiment, meas for measurement
x <- all_cells %>%
  behead("up-left", exp_type) %>%
  behead("up-left", exp_name) %>%
  behead("up", meas_type) %>%
  behead("left", clade) %>%
  select(clade, starts_with('exp'), meas_type, doubling_time = chr) %>%
  mutate(
    across(doubling_time, as.numeric),
    doubling_rate = 1 / doubling_time,
  ) %>%
  arrange(clade, exp_type, exp_name, meas_type) %>%
  glimpse
dir.create('_output')
saveRDS(x, '_output/table1-tidy.rds')
```

# Analysis

Let's compute the error in doubling rate, assuming that the qPCR rate is accurate.

```{r}
y <- x %>%
  select(-doubling_time) %>%
  pivot_wider(names_from = meas_type, values_from = doubling_rate) %>%
  mutate(error = FRAxC - qPCR)
```

```{r}
y %>% 
  filter(!is.na(qPCR)) %>%
  select(-exp_type) %>% 
  knitr::kable(digits = 4)
```

**Core 30**

```{r}
y %>% 
  # filter(str_detect(exp_type, 'WOR')) %>%
  filter(exp_name == 'Core 30') %>%
  select(-exp_type, -exp_name) %>% 
  knitr::kable(digits = 4)
```

**Core 32**

```{r}
y %>% 
  filter(exp_name == 'Core 32') %>%
  select(-exp_type, -exp_name) %>% 
  knitr::kable(digits = 4)
```

Compute the calculations for the manuscript text,

```{r, include = TRUE}
bathy_core30_fraxc <- x %>%
  filter(exp_name == 'Core 30', str_detect(clade, 'Bathyarchaeota'), 
    meas_type == 'FRAxC') %>%
  pull(doubling_rate)
bathy_core30_qpcr <- x %>%
  filter(exp_name == 'Core 30', str_detect(clade, 'Bathyarchaeota'), 
    meas_type == 'qPCR') %>%
  pull(doubling_rate)
bathy_core32_fraxc <- x %>%
  filter(exp_name == 'Core 32', str_detect(clade, 'Bathyarchaeota'), 
    meas_type == 'FRAxC') %>%
  pull(doubling_rate)
bathy_core32_qpcr <- x %>%
  filter(exp_name == 'Core 32', str_detect(clade, 'Bathyarchaeota'), 
    meas_type == 'qPCR') %>%
  pull(doubling_rate)
mbgd_core32_fraxc <- x %>%
  filter(exp_name == 'Core 32', str_detect(clade, 'MBG-D'), 
    meas_type == 'FRAxC') %>%
  pull(doubling_rate)
mbgd_core32_qpcr <- x %>%
  filter(exp_name == 'Core 32', str_detect(clade, 'MBG-D'), 
    meas_type == 'qPCR') %>%
  pull(doubling_rate)
```

**Draft manuscript text:**

Our aim is to compare the growth rates for taxa with both FRAxC and qPCR values for the two sediment cores, as reported in Table 1.
The first soil core included qPCR measurements of a single archaeal clade, _Bathyarchaeota_, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of `r round(bathy_core30_fraxc, 3)`/yr by FRAxC and `r round(bathy_core30_qpcr, 3)`/yr by qPCR).
The second soil core included qPCR measurements of _Bathyarchaeota_ and a second clade, _Thermoprofundales_/MBG-D.
In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by `r round(bathy_core32_fraxc - bathy_core32_qpcr, 3)`/yr for _Bathyarchaeota_ (`r round(bathy_core32_fraxc, 3)`/yr by FRAxC and `r round(bathy_core32_qpcr, 3)`/yr by qPCR) and by `r round(mbgd_core32_fraxc - mbgd_core32_qpcr, 3)`/yr for _Thermoprofundales_/MBG-D (`r round(mbgd_core32_fraxc, 3)`/yr by FRAxC and `r round(mbgd_core32_qpcr, 3)`/yr by qPCR).
A low number of experimental samples and noise in both the FRAxC and qPCR measurements place significant uncertainty in these measurements; however, the fact that FRAxC-derived growth rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreases with depth in a manner that systematically biases FRAxC-derived rates to higher values.
The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr suggests an error large enough to impact results for some taxa classified as positive growers by @lloyd2020evid, for which growth rates ranged between 0.04/yr and 0.5/yr.
Overall, the comparison between FRAxC and qPCR measurements gives support to the study conclusions, but suggest that species at the lower end of this range of positive FRAxC-derived rates may in fact be merely persisting or even slowly declining in abundance.

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>

### 2022-01-10 case-studies.Rmd
Revise marine-sediment case study
Polish overall text. Also, calculations are now performed directly in
the section Rmd using the output from the corresponding notebook post.

@@
 
-## Growth in marine sediments
+## Microbial growth in marine sediments
 
-As sediment is slowly buried, the concentration of oxygen and nutrients decreases, and cell density decreases; @lloyd2020evid asked whether some taxa are nevertheless able to continue to grow and increase in cell density.
-To measure density with depth, they used community 16S sequencing coupled with measurements of total community density using CARD-FISH, which they combined using community-normalization (Equation X).
-(The authors refer to these absolute density measurements as FRAxC measurements.)
-The authors assessed whether density of bacterial and archaeal families increased with depth, using linear regression on log cell density to infer the exponential growth (or decay) rate over the first 3 cm below the bioirrigation layer (corresponding to $\sim 8$ years of burial time).
-(Per time after converting depth -> time)
-The authors did this for two cores, and checked the inferences from these sequencing-based measurements against targeted measurements (qPCR) of 16S concentration for particular taxa.
-They further checked w/ sequencing-based growth-rate inference from two laboratory incubations.
+Our route for mean efficiency to become associated with the covariate is if the covariate quantifies a biological process that preferentially selects for a microbial trait that also tends to increase or decrease measurement efficiency.
+We illustrate this potential mechanism with a study of microbial growth in marine sediments.
 
-The authors attempted to explicitly account for bias by arguing that a consistent taxonomic bias would cancel in the observed log-fold-changes that determine the inferred exponential growth rates (see also @kevorkian2018esti); however, their calculations neglected the impact of changes in mean efficiency which do not cancel and may lead to spurious conclusions (previous section).
-We were therefore interested in whether the unaccounted for error caused by bias might have led to spurious inferences about which species were increasing with depth.
-Supposing the CARD-FISH measurements accurately measured total density, our results from the previous section indicate that a systematic error in growth rates would arise if the log mean efficiency systematically changed with depth.
-Diversity is sufficiently high that this systematic error is unlikely to be driven by just a few species, as in the vaginal microbiome and fungal case studies.
-But it is possible that the covariate (burial time) selects directly on a species efficiency as well as its ability to grow. 
-In particular, we considered that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
-In this case, we would expect the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that are just persisting or even slowly dying off would be mistakenly inferred to have positive growth.
+The surface layers of marine sediments harbor diverse and abundant microbiomes.
+Total cell density and species richness decrease with depth as resources are consumed in older, deeper sediment layers; however, some taxa are able to grow and increase in density as they are slowly buried.
+@lloyd2020evid performed a systematic assessment of growth rates of bacterial and archaeal taxa over a depth of 10 cm (corresponding to ~40 years of burial time) in sediment of the White Oak River estuary.
+To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-community normalization method (Equation \@ref(eq:density-prop-meas)), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy (CARD-FISH).
+The authors refer to these absolute densities as FRAxC measurements, for 'fraction of 16S reads times total cell counts'.
+The FRAxC measurements were used to infer growth rate from the slope of a simple linear regression of log cell density against burial time over the first 3 cm below the bioirrigation layer (corresponding to $\sim 8$ years of burial).
+To validate the inference of positive growers, the authors compared the growth rates from FRAxC-based inference from two sediment cores, qPCR measurements in these cores for a few reference taxa, and FRAxC-based inference in two replicate laboratory incubation experiments.
+
+Taxonomic bias could lead to systematic error in FRAxC-derived growth rates if sample mean efficiency tends to systematically vary with burial time (or equivalently, depth).
+One possibility is that microbes with tougher cell walls tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
+In this case, we would expect the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that decay sufficiently slowly would be mistakenly inferred to have positive growth.
+
+```{r}
+x <- here(
+  'notebook/_posts/2022-01-08-lloyd2020evi-case-study/_output/table1-tidy.rds'
+) %>% readRDS
+bathy_core30_fraxc <- x %>%
+  filter(exp_name == 'Core 30', str_detect(clade, 'Bathyarchaeota'), 
+    meas_type == 'FRAxC') %>%
+  pull(doubling_rate)
+bathy_core30_qpcr <- x %>%
+  filter(exp_name == 'Core 30', str_detect(clade, 'Bathyarchaeota'), 
+    meas_type == 'qPCR') %>%
+  pull(doubling_rate)
+bathy_core32_fraxc <- x %>%
+  filter(exp_name == 'Core 32', str_detect(clade, 'Bathyarchaeota'), 
+    meas_type == 'FRAxC') %>%
+  pull(doubling_rate)
+bathy_core32_qpcr <- x %>%
+  filter(exp_name == 'Core 32', str_detect(clade, 'Bathyarchaeota'), 
+    meas_type == 'qPCR') %>%
+  pull(doubling_rate)
+mbgd_core32_fraxc <- x %>%
+  filter(exp_name == 'Core 32', str_detect(clade, 'MBG-D'), 
+    meas_type == 'FRAxC') %>%
+  pull(doubling_rate)
+mbgd_core32_qpcr <- x %>%
+  filter(exp_name == 'Core 32', str_detect(clade, 'MBG-D'), 
+    meas_type == 'qPCR') %>%
+  pull(doubling_rate)
+```
 
 We can test the hypothesis that systematic variation in log mean efficiency with depth distorts inferred growth rates using the qPCR measurements of two clades also collected by @lloyd2020evid.
-qPCR measurements of specific clades are also subject to taxonomic bias, but not the compositional constraint of sequencing, or at least are expected to experience it to a much weaker extent (Section \@ref(review-absolute-methods)).
-Hence we expect the systematic fold error in qPCR measurements to be approximately constant across samples and for targeted qPCR measurements to yield relatively accurate estimates of growth rates.
-Comparing qPCR to FRAxC growth rates on the same taxa allows us to estimate the systematic error in the FRAxC growth rates, equal to the rate of change of log mean efficiency with depth.
-Because the same error applies to all taxa, these targeted comparisons inform us about the impact of bias on the growth rates of all taxa.
-
-We compared the growth rates for taxa with both FRAxC and qPCR values for the two sediment cores, as reported in Table 1.
-The first soil core included qPCR measurements of a single archaeal clade, _Bathyarchaeota_, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of 0.099/yr by FRAxC and 0.097/yr by qPCR).
+In Section \@ref(review-absolute-methods), we argue that taxonomic bias in targeted qPCR measurements is expected to create a roughly constant fold error and yield fold changes in absolute density that are relatively unaffected by changes in mean efficiency.
+Thus comparing qPCR to FRAxC growth rates on the same reference taxa allows us to estimate the systematic error in FRAxC growth rates.
+Because the systematic error in regression slopes due to bias is the same for each taxon (Equation \@ref(eq:regression-error)), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
+<!-- Our aim is to compare the growth rates for taxa with both FRAxC and qPCR values for the two sediment cores, as reported in Table 1. -->
+The first soil core included qPCR measurements of a single archaeal clade, _Bathyarchaeota_, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of `r round(bathy_core30_fraxc, 3)`/yr by FRAxC and `r round(bathy_core30_qpcr, 3)`/yr by qPCR).
 The second soil core included qPCR measurements of _Bathyarchaeota_ and a second clade, _Thermoprofundales_/MBG-D.
-In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by 0.012/yr for _Bathyarchaeota_ (0.112/yr by FRAxC and 0.100/yr by qPCR) and by 0.086/yr for _Thermoprofundales_/MBG-D (0.294/yr by FRAxC and 0.208/yr by qPCR).
-A low number of experimental samples and noise in both the FRAxC and qPCR measurements place signficant uncertainty in these measurements; however, the fact that FRAxC-derived growth rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreases with depth in a manner that systematically biases FRAxC-derived rates to higher values.
-The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr is significant for this low nutrient environment, where growth rates of the identified positive growers ranged between 0.04/yr and 0.5/yr.
+In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by `r round(bathy_core32_fraxc - bathy_core32_qpcr, 3)`/yr for _Bathyarchaeota_ (`r round(bathy_core32_fraxc, 3)`/yr by FRAxC and `r round(bathy_core32_qpcr, 3)`/yr by qPCR) and by `r round(mbgd_core32_fraxc - mbgd_core32_qpcr, 3)`/yr for _Thermoprofundales_/MBG-D (`r round(mbgd_core32_fraxc, 3)`/yr by FRAxC and `r round(mbgd_core32_qpcr, 3)`/yr by qPCR).
+A low number of experimental samples and noise in both the FRAxC and qPCR measurements place significant uncertainty in these measurements; however, the fact that FRAxC-derived growth rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreases with depth in a manner that systematically biases FRAxC-derived rates to higher values.
+The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr suggests an error large enough to impact results for some taxa classified as positive growers, whose FRAxC growth rates ranged between 0.04/yr and 0.5/yr.
 Overall, the comparison between FRAxC and qPCR measurements gives support to the study conclusions, but suggest that species at the lower end of this range of positive FRAxC-derived rates may in fact be merely persisting or even slowly declining in abundance.
 
-Caveats that might be instructive to mention
+<!-- notes/questions -->
 
-- qPCR and seq analysis are both at ~family level; this is prob not a big deal when it comes to inferring the systematic variation in mean efficiency though, since qPCR and 16S seq share a lot of bias
-- relic dna
-- possibility of non-genetic variation -> change in efficiency with depth. This could still lead to a change in mean efficiency, but would offset the effect for these particular species.
-- possible error/bias in CARD-FISH
+<!-- - also bring in the incubation experiments as providing further evidence? -->
+<!-- - discuss that kl et al attempted to account for bias, incorrected concluded consistent bias couldn't mess up results? -->
+<!-- - discuss other mech's that could impact mean eff, which might even push in other directions? -->
+
+<!-- Caveats that might be instructive to mention -->
+
+<!-- - qPCR and seq analysis are both at ~family level; this is prob not a big deal when it comes to inferring the systematic variation in mean efficiency though, since qPCR and 16S seq share a lot of bias -->
+<!-- - relic DNA -->
+<!-- - possibility of non-genetic variation -> change in efficiency with depth. This could still lead to a change in mean efficiency, but would offset the effect for these particular species. -->
+<!-- - possible error/bias in CARD-FISH -->
 

### 2022-01-10 notebook/_posts/2021-12-15-momspi-regression-diversity/_cut.Rmd
Revise MOMSPI regression analysis
- Removed rank regression and Gamma-Poisson regression (kept in
  _cut.Rmd) until have time to better understand these
- Adjusted zero-replacement for the lm regression
- General refactoring for clarity
- Improved plots

--- 2021-12-17 notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity.Rmd
@@
----
-title: "MOMS-PI: Regression on diversity"
-description: |
-author:
-  - name: Michael R. McLaren
-    url: {}
-categories:
-  - ref:brooks2015thet
-  - ref:fettweis2019thev
-date: 2021-12-15
-draft: false
-bibliography: ../../../main.bib
-output:
-  distill::distill_article:
-    self_contained: false
-    dev: svg
-    toc: true
----
+# Regression - new
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = TRUE,
-  cache = TRUE,
-  autodep = TRUE,
-  cache.comments = FALSE,
-  dpi = 300
-)
-```
-
-# Setup
-
-## R setup
-
-```{r libraries}
-# Tools for microbiome data
-library(speedyseq)
-# Tools for general purpose data manipulation and plotting
-library(tidyverse)
-library(fs)
-# ggplot helpers
-library(ggbeeswarm)
-library(cowplot)
-library(patchwork)
-library(scales)
-# stats helpers
-# library(broom)
-
-library(metacal); packageVersion("metacal")
-```
-
-```{r ggplot}
-theme_set(theme_cowplot())
-
-colors_brooks <- c(
-  "Atopobium_vaginae" = "#009E73",
-  "Gardnerella_vaginalis" = "#56B4E9",
-  "Lactobacillus_crispatus" = "#D55E00",
-  "Lactobacillus_iners" = "#505050",
-  "Prevotella_bivia" = "#0072B2",
-  "Sneathia_amnii" = "#CC79A7",
-  "Streptococcus_agalactiae" = "#E69F00")
-
-scale_y_custom <- scale_y_continuous(
-    trans = 'log10',
-    breaks = trans_breaks('log10', function(x) 10^x),
-    labels = trans_format('log10', math_format(10^.x))
-  )
-```
-
-# SETUP TODO
-
-# Regression analysis
-
-NOTE: Consider saving output so can be used in other notebook files
-
-## Classify samples into high and low diversity groups
-
-TODO: Consider possible better way to do calibration
-
-Compute Shannon and Inverse-Simpson diversity indices and add to the sample metadata.
-`Phyloseq::estimate_richness()` will complain about the lack of singletons, but we can ignore this since we aren't using species richness estimators.
-I will exponentiate the Shannon index to convert it to the q1 Hill number so that (like Inverse Simpson) it has the scale of number of species.
+Let's create a data frame with the abundances and offsets for the different regressions.
+For the analyses that do not model the counting process, we'll adjust the proportions to the Dirichlet posterior mean, with the prior set to the average proportions across all samples.
 
 ```{r}
-div <- list(
-  'obs' = momspi_obs,
-  'cal' = momspi_cal
-) %>%
-  map(estimate_richness, measures = c('Shannon', 'InvSimpson')) %>%
-  map_dfr(as_tibble, rownames = '.sample', .id = 'type') %>%
-  transmute(
-    .sample, type,
-    div_q1 = exp(Shannon),
-    div_q2 = InvSimpson,
-  ) %>%
-  pivot_longer(starts_with('div_'), names_to = 'index', values_to = 'diversity')
-div_wide <- div %>%
-  pivot_wider(values_from = diversity, names_from = c('index', 'type')) 
+taxa_mean_prop <- momspi_div %>%
+  otu_table %>%
+  transform_sample_counts(close_elts) %>%
+  orient_taxa(as = 'cols') %>%
+  as('matrix') %>%
+  colMeans
+stopifnot(sum(taxa_mean_prop) == 1)
+prior_vec <- taxa_mean_prop * ntaxa(momspi)
+stopifnot(identical(length(prior_vec), ntaxa(momspi)))
+# Note the need for the seq_along trick to get phyloseq to allow this
+# adjustment
+momspi_div_adj <- momspi_div %>% 
+  transform_sample_counts(~ prior_vec[seq_along(.x)] + .x) %>%
+  transform_sample_counts(close_elts) %>%
+  otu_table %>%
+  as_tibble %>%
+  rename(prop_adj = .abundance)
 ```
@@
 ```{r}
-div_wide %>%
-  mutate(across(starts_with('div'), log2)) %>%
-  ggplot() +
-  geom_point(aes(x = .panel_x, y = .panel_y), alpha = 0.5) +
-  ggforce::geom_autodensity() +
-  geom_bin2d(aes(x = .panel_x, y = .panel_y)) +
-  ggforce::facet_matrix(vars(starts_with('div')), layer.diag = 2, layer.lower = 3)
-```
+x <- momspi_div %>%
+  as_tibble %>%
+  rename(count = .abundance) %>%
+  with_groups(.sample, mutate,
+    prop_raw = close_elts(count),
+  ) %>%
+  filter(.otu %in% taxa_to_test) %>%
+  left_join(by = c('.sample', '.otu'), momspi_div_adj) %>%
+  left_join(bias_all %>% select(.otu, efficiency), by = '.otu') %>%
+  mutate(
+    # Offset for log link when bias is ignored
+    offset_obs = log(sample_sum),
+    # Offset for log link accounting for bias
+    offset_cal = log(sample_sum) + log(efficiency / mean_efficiency),
+  ) %>%
+  # Calibrated proportions
+  with_groups(.sample, mutate,
+    across(starts_with('prop_'), .names = '{.col}_cal', 
+      ~close_elts(. / efficiency))
+  )
+all.equal(x$prop_raw_cal * x$sample_sum / exp(x$offset_cal), , x$prop_raw)
 
-The observed (Log) order-1 and order-2 diversity show a tight correlation with each other, as do the calibrated diversities, while the correlation is much lower when comparing observed and calibrated numbers.
-The reason is likely that calibration will tend to increase the diversity of samples that are observed to be dominated by a high-efficiency taxon (like L.
-iners) and decrease the diversity of samples dominated by a low efficiency tason (like G.
-vaginalis).
-
-Shannon diversity of the observed profiles is what is most often considered in vaginal microbiome studies, so let's use this metric to partition the samples.
-
-```{r}
-div_wide %>%
-  ggplot(aes(x = div_q1_obs)) +
-  scale_x_log10() +
-  annotate('rect', xmin = 0, xmax = 1.5, ymin = -Inf, ymax = Inf, 
-    fill = 'blue', alpha = 0.3) +
-  annotate('rect', xmin = 1.5, xmax = 3, ymin = -Inf, ymax = Inf, 
-    fill = 'grey', alpha = 0.3) +
-  annotate('rect', xmin = 3, xmax = Inf, ymin = -Inf, ymax = Inf, 
-    fill = 'green', alpha = 0.3) +
-  geom_histogram() +
-  geom_vline(xintercept = c(1.5, 3), color = 'black')
-```
-
-I'll pick the first mode as the 'Low diversity' group, treat the second mode as 'Medium', and the tail as 'High'.
-
-```{r}
-div_wide <- div_wide %>%
-  mutate(
-    div_group = case_when(
-      div_q1_obs < 1.5 ~ 'Low',
-      div_q1_obs > 3 ~ 'High',
-      TRUE ~ 'Medium'
-    ) %>% factor(levels = c('Low', 'Medium', 'High'))
-  )
-```
-
-## Understanding the diversity-efficiency relationship
-
-```{r}
-mean_eff3 <- mean_eff2 %>%
-  left_join(
-    div,
-    # div %>% pivot_wider(names_from = type, values_from = c(Shannon, InvSimpson)),
-    by = '.sample'
-  )
-```
-
-```{r}
-mean_eff3 %>%
-  ggplot(aes(diversity, mean_efficiency, color = top_plot)) +
-  facet_grid(type~index) +
-  scale_x_log10() +
-  scale_y_log10() +
-  geom_point(alpha = 0.6) +
-  stat_smooth(aes(group = 'none'), color = 'black') +
-  scale_color_manual(
-    values = colors_top,
-    labels = function(x) str_replace_all(x, '_', ' ')
-  )
-```
-
-```{r}
-sam_div <- momspi_div %>%
-  sample_data %>%
-  as_tibble
-```
-
-```{r}
-sam_div %>%
-  ggplot(aes(y = div_group, x = mean_efficiency)) +
-  scale_x_log10() +
-  ggdist::stat_slab()
-```
-
-We can see that the mean efficiency is 3-10X lower in the High diversity relative to the low diversity group.
-
-```{r}
-lm(log(mean_efficiency) ~ div_group, data = sam_div) %>% summary
-```
-
-## Regression setup
-
-To simplify interpretation of some of the regression models, let's rarefy samples to 10K reads (we'll be dropping some samples).
-
-```{r}
-momspi_rare <- momspi %>% 
-  rarefy_even_depth(
-    sample.size = 1e4, 
-    rngseed = 42,
-    replace = TRUE, # for speed, might want to change
-    trimOTUs = FALSE
-  )
-```
-
-Add the diversity grouping information and the offset terms for the Gamma-Poisson (Negative Binomial) regression.
-
-```{r}
-momspi_div <- momspi_rare %>%
-  left_join_sample_data(mean_eff, by = '.sample') %>%
-  left_join_sample_data(div_wide, by = '.sample') %>%
-  mutate_sample_data(
-    gp_offset_obs = log(sample_sum),
-    gp_offset_cal = log(sample_sum) - log(mean_efficiency),
-  )
-```
-
-We'll use the criterion from Fettweis et al to pick the taxa to test.
-Note, we may not get the same exact taxa as them since we're using a larger set of women.
-Note, the sample sums used in the offset were calculated for the full species set, so we are still analzying log proportions with respect to the full species set.
-
-```{r}
-taxa_to_test <- momspi %>%
-  transform_sample_counts(close_elts) %>%
-  filter_taxa2(~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
-  taxa_names %>%
-  print
-```
-
-## Linear regression on log proportion
-
-Do this first to confirm as expected from theory.
-
-```{r}
-stat_test <- function(ps) {
-  ps %>%
-    transform_sample_counts(close_elts) %>%
-    filter_sample_data(div_group %in% c('Low', 'High')) %>%
-    prune_taxa(taxa_to_test, .) %>%
-    as_tibble %>%
-    with_groups(.otu, nest) %>%
-    mutate(
-      fit = map(data,
-        ~lm(log(.abundance) ~ div_group, data = .x)
-      ),
-    ) %>%
-    select(-data)
-  # %>% unnest(fit)
-}
-res_div_lm <- list(
-  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.3),
-  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.3) %>% calibrate(bias_all_vec)
-) %>%
-  map_dfr(stat_test, .id = 'type')
-res_div_lm_tidy <- res_div_lm %>%
-  mutate(across(fit, map, broom.mixed::tidy)) %>%
-  unnest(fit) %>% 
-  filter(term != '(Intercept)')
-```
-
-```{r}
-# x should be the outout of a tidier
-plot_estimates <- function(x) {
-  err <- x %>%
-    select(name = type, .otu, value = estimate) %>%
-    pivot_wider() %>%
-    mutate(
-      # error = Observed - Calibrated, # will be same for everything
-      # error_abs = abs(error),
-      # # error_rel = abs(error / Calibrated), # only makes sense if not a sign error
-      error_sign = sign(Observed) == sign(Calibrated),
-      sign = case_when(
-        sign(Observed) > 0 & sign(Calibrated) > 0 ~ 'Positive',
-        sign(Observed) < 0 & sign(Calibrated) < 0 ~ 'Negative',
-        TRUE ~ 'Depends',
-      )
-    ) %>%
-    select(.otu, error_sign, sign)
-
-  p1 <- x %>%
-    left_join(err, by = '.otu') %>%
-    ggplot(aes(y = type, x = estimate, fill = sign)) +
-    scale_fill_brewer(type = 'qual', palette = 2) +
-    geom_vline(xintercept = 0, color = 'grey') +
-    ggdist::geom_dots()
-  p2 <- x %>%
-    mutate(
-      .otu_fct = fct_reorder(.otu, estimate, .fun = min)
-    ) %>%
-    ggplot(aes(y = .otu_fct, x = estimate, color = type)) +
-    scale_color_brewer(type = 'qual', palette = 7) +
-    geom_vline(xintercept = 0, color = 'grey') +
-    geom_pointrange(
-      aes(
-        xmin = estimate - 2*std.error,
-        xmax = estimate + 2*std.error
-      ))
-  p1 / p2 +
-    plot_layout(ncol = 1, heights = c(0.2, 1))
-}
-plot_estimates(res_div_lm_tidy)
-```
-
-Consider doing bootstrap CIs.
-
-```{r, eval = F}
-lm_boot <- function(.data, ...) {
-  fit <- lm(log(.abundance) ~ div_group, data = .data)
-  car::Boot(fit, ...)
-}
-x <- momspi_div %>% transform_sample_counts(~. + 0.3) %>% 
-  filter_sample_data(div_group %in% c('Low', 'High')) %>%
-  filter_tax_table(.otu == 'Gardnerella_vaginalis') %>% 
-  as_tibble
-fit <- lm(log(.abundance) ~ div_group, data = x)
-bt <- car::Boot(fit)
-bt %>% broom::tidy()
-broom::tidy(bt, conf.int = TRUE)
-confint(bt, type = 'perc') %>% broom::tidy()
+stopifnot()
 ```
@@
 
-Similar to Wilcoxin test when there are no ties; not sure about with ties.
+Similar to Wilcoxin test; approximately equivalent when there are no ties.
 
@@
 res_div_rank <- list(
-  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.3),
-  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.3) %>% calibrate(bias_all_vec)
+  'Observed' = momspi_div_prop_adj,
+  'Calibrated' = momspi_div_prop_adj %>% calibrate(bias_all_vec)
 ) %>%
@@
 
+NOTE: I think the results where quite different when zeros are replaced with the same value for each taxon (????)
+
+Try the same thing without zero replacement,
+
+```{r}
+stat_test <- function(ps) {
+  ps %>%
+    transform_sample_counts(close_elts) %>%
+    filter_sample_data(div_group %in% c('Low', 'High')) %>%
+    prune_taxa(taxa_to_test, .) %>%
+    as_tibble %>%
+    with_groups(.otu, nest) %>%
+    mutate(
+      fit = map(data,
+        ~lm(
+          rank(.abundance, ties.method = 'average')/n() ~ div_group, 
+          data = .x
+        )
+      ),
+    ) %>%
+    select(-data)
+  # %>% unnest(fit)
+}
+res_div_rank <- list(
+  'Observed' = momspi_div,
+  'Calibrated' = momspi_div %>% calibrate(bias_all_vec)
+) %>%
+  map_dfr(stat_test, .id = 'type')
+res_div_rank_tidy <- res_div_rank %>%
+  mutate(across(fit, map, broom.mixed::tidy)) %>%
+  unnest(fit) %>% 
+  filter(term != '(Intercept)')
+```
+
+```{r}
+plot_estimates(res_div_rank_tidy)
+```
+
+This is very mysterious. Needs more investigation before we report these results.
+
 ## Gamma-Poisson regression
+
+HERE. Don't have time to mess with this for now. Set eval to false for now.
 

### 2022-01-10 notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity.Rmd
Revise MOMSPI regression analysis
- Removed rank regression and Gamma-Poisson regression (kept in
  _cut.Rmd) until have time to better understand these
- Adjusted zero-replacement for the lm regression
- General refactoring for clarity
- Improved plots

@@
 title: "MOMS-PI: Regression on diversity"
-description: |
+description: Analyze the impact of bias on a DA analysis of relative abundances versus diversity.
 author:
@@
   echo = TRUE,
-  cache = TRUE,
+  cache = FALSE,
   autodep = TRUE,
@@
 library(fs)
+library(here)
 # ggplot helpers
@@
 
-colors_brooks <- c(
-  "Atopobium_vaginae" = "#009E73",
-  "Gardnerella_vaginalis" = "#56B4E9",
-  "Lactobacillus_crispatus" = "#D55E00",
-  "Lactobacillus_iners" = "#505050",
-  "Prevotella_bivia" = "#0072B2",
-  "Sneathia_amnii" = "#CC79A7",
-  "Streptococcus_agalactiae" = "#E69F00")
-
 scale_y_custom <- scale_y_continuous(
@@
 
-# SETUP TODO
+## MOMS-PI setup
+
+Load the MOMS-PI profiles and estimate bias,
+
+```{r}
+source(here('notebook/_code/momspi-setup.R'))
+```
+
+To simplify interpretation of some of the regression models, let's rarefy samples to 10K reads (we'll be dropping some samples that are below this limit).
+
+```{r}
+momspi_rare <- xfun::cache_rds({
+  momspi %>% 
+    rarefy_even_depth(
+      sample.size = 1e4, 
+      rngseed = 42,
+      replace = FALSE,
+      trimOTUs = FALSE
+    )
+  }, hash = momspi
+)
+# Make sure no taxa have 0 counts, in case this causes issues later on
+stopifnot(all(momspi_rare %>% taxa_sums > 0))
+```
+
+Let's also add the new sample sums and estimated mean efficiency and
+
+```{r}
+mean_eff <- momspi_rare %>%
+ calibrate(bias_all_vec, norm = 'close') %>%
+ perturb(bias_all_vec, norm = 'none') %>% 
+ sample_sums %>%
+ enframe('.sample', 'mean_efficiency')
+
+momspi_rare <- momspi_rare %>% 
+  mutate_sample_data(., 
+    sample_sum_original = sample_sum,
+    sample_sum = sample_sums(.),
+  ) %>%
+  left_join_sample_data(mean_eff, by = '.sample')
+```
+
+Let's also add the top species in each sample,
+
+```{r}
+get_top_species <- function(ps) {
+  x <- ps %>%
+    transform_sample_counts(close_elts) %>%
+    otu_table %>%
+    as_tibble(pivot = TRUE) %>%
+    with_groups(.sample, slice_max, .abundance, n = 1)
+  if (anyDuplicated(x$.sample))
+    stop('Ties present')
+  x %>% select(.sample, top_species = .otu)
+}
+top_species <- list(
+  'obs' = momspi_rare,
+  'cal' = momspi_rare %>% calibrate(bias_all_vec)
+  ) %>%
+  map_dfr(get_top_species, .id = 'type')
+top_species_wide <- top_species %>%
+  pivot_wider(values_from = top_species, names_from = type, 
+    names_prefix = 'top_species_') 
+momspi_rare <- momspi_rare %>%
+  left_join_sample_data(top_species_wide, by = '.sample')
+```
+
+```{r, include = FALSE, eval = FALSE}
+# check that see expected difference between obs and cal
+# top_species_wide  %>% count(top_species_obs, top_species_cal) %>% arrange(-n) %>% print(n=30)
+top_species_wide  %>% count(top_species_obs) %>% arrange(-n) %>% print(n=10)
+top_species_wide  %>% count(top_species_cal) %>% arrange(-n) %>% print(n=10)
+```
 
@@
 
-TODO: Consider possible better way to do calibration
-
 Compute Shannon and Inverse-Simpson diversity indices and add to the sample metadata.
@@
 div <- list(
-  'obs' = momspi_obs,
-  'cal' = momspi_cal
+  'obs' = momspi_rare,
+  'cal' = momspi_rare %>% calibrate(bias_all_vec)
 ) %>%
@@
 
-```{r}
-div_wide %>%
+```{r, fig.dim = c(7, 7)}
+p <- div_wide %>%
   mutate(across(starts_with('div'), log2)) %>%
@@
   ggforce::facet_matrix(vars(starts_with('div')), layer.diag = 2, layer.lower = 3)
+ggrastr::rasterize(p, dpi = 300)
 ```
@@
 
-## Understanding the diversity-efficiency relationship
-
-```{r}
-mean_eff3 <- mean_eff2 %>%
-  left_join(
-    div,
-    # div %>% pivot_wider(names_from = type, values_from = c(Shannon, InvSimpson)),
-    by = '.sample'
-  )
-```
-
-```{r}
-mean_eff3 %>%
-  ggplot(aes(diversity, mean_efficiency, color = top_plot)) +
-  facet_grid(type~index) +
-  scale_x_log10() +
-  scale_y_log10() +
-  geom_point(alpha = 0.6) +
-  stat_smooth(aes(group = 'none'), color = 'black') +
-  scale_color_manual(
-    values = colors_top,
-    labels = function(x) str_replace_all(x, '_', ' ')
-  )
-```
-
-```{r}
-sam_div <- momspi_div %>%
-  sample_data %>%
-  as_tibble
-```
-
-```{r}
-sam_div %>%
-  ggplot(aes(y = div_group, x = mean_efficiency)) +
-  scale_x_log10() +
-  ggdist::stat_slab()
-```
-
-We can see that the mean efficiency is 3-10X lower in the High diversity relative to the low diversity group.
-
-```{r}
-lm(log(mean_efficiency) ~ div_group, data = sam_div) %>% summary
-```
-
 ## Regression setup
 
-To simplify interpretation of some of the regression models, let's rarefy samples to 10K reads (we'll be dropping some samples).
-
-```{r}
-momspi_rare <- momspi %>% 
-  rarefy_even_depth(
-    sample.size = 1e4, 
-    rngseed = 42,
-    replace = TRUE, # for speed, might want to change
-    trimOTUs = FALSE
-  )
-```
-
-Add the diversity grouping information and the offset terms for the Gamma-Poisson (Negative Binomial) regression.
+Add the diversity grouping information, subset to the Low and High samples, and compute the offsets for the Gamma-Poisson model
 
@@
 momspi_div <- momspi_rare %>%
-  left_join_sample_data(mean_eff, by = '.sample') %>%
   left_join_sample_data(div_wide, by = '.sample') %>%
+  filter_sample_data(div_group %in% c('Low', 'High')) %>%
   mutate_sample_data(
@@
 We'll use the criterion from Fettweis et al to pick the taxa to test.
-Note, we may not get the same exact taxa as them since we're using a larger set of women.
-Note, the sample sums used in the offset were calculated for the full species set, so we are still analzying log proportions with respect to the full species set.
+Note, we may not get the same exact taxa as them since we're using a larger set of women and ignoring the medium-diversity samples.
+Note, the log proportions used in the regression analysis below are with respect to the full species set.
 
 ```{r}
-taxa_to_test <- momspi %>%
+taxa_to_test <- momspi_div %>%
   transform_sample_counts(close_elts) %>%
@@
 
-Do this first to confirm as expected from theory.
+The simple linear regression is the fastest and simplest approach, and the easiest to understand using results from the manuscript.
+
+First, we need to ensure there are no zeros.
+I'll try two approaches; the first is to use a pseudocount of 0.5.
 
 ```{r}
-stat_test <- function(ps) {
+stat_test <- function(ps, taxa) {
   ps %>%
+    filter_sample_data(div_group %in% c('Low', 'High')) %>%
     transform_sample_counts(close_elts) %>%
-    filter_sample_data(div_group %in% c('Low', 'High')) %>%
-    prune_taxa(taxa_to_test, .) %>%
+    prune_taxa(taxa, .) %>%
     as_tibble %>%
@@
     select(-data)
-  # %>% unnest(fit)
 }
 res_div_lm <- list(
-  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.3),
-  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.3) %>% calibrate(bias_all_vec)
+  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.5),
+  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.5) %>%
+    calibrate(bias_all_vec)
 ) %>%
-  map_dfr(stat_test, .id = 'type')
+  map_dfr(stat_test, taxa = taxa_to_test, .id = 'type')
 res_div_lm_tidy <- res_div_lm %>%
@@
     mutate(
-      # error = Observed - Calibrated, # will be same for everything
-      # error_abs = abs(error),
-      # # error_rel = abs(error / Calibrated), # only makes sense if not a sign error
+      error = Observed - Calibrated,
       error_sign = sign(Observed) == sign(Calibrated),
@@
     ggplot(aes(y = type, x = estimate, fill = sign)) +
-    scale_fill_brewer(type = 'qual', palette = 2) +
+    scale_fill_brewer(type = 'qual', palette = 2,
+      guide = guide_legend(reverse = TRUE)) +
     geom_vline(xintercept = 0, color = 'grey') +
-    ggdist::geom_dots()
+    ggdist::geom_dots() +
+    labs(fill = 'Sign of estimate', x = 'Estimated LFC', y = 'Meas. type')
   p2 <- x %>%
     mutate(
-      .otu_fct = fct_reorder(.otu, estimate, .fun = min)
+      .otu_fct = fct_reorder(.otu, estimate, .fun = min) %>%
+        fct_relabel(function(x) str_replace_all(x, '_', ' '))
     ) %>%
     ggplot(aes(y = .otu_fct, x = estimate, color = type)) +
-    scale_color_brewer(type = 'qual', palette = 7) +
+    # scale_color_brewer(type = 'qual', palette = 7) +
+    colorblindr::scale_color_OkabeIto() +
     geom_vline(xintercept = 0, color = 'grey') +
-    geom_pointrange(
+    geom_linerange(size = 1,
       aes(
@@
         xmax = estimate + 2*std.error
-      ))
+      )) +
+    geom_point(size = 1.5) +
+    labs(y = 'Species-level OTU', x = 'Estimated LFC', color = 'Meas. type')
   p1 / p2 +
-    plot_layout(ncol = 1, heights = c(0.2, 1))
+    plot_layout(ncol = 1, heights = c(0.3, 1)) +
+    plot_annotation(tag_levels = 'A') &
+    theme_minimal_hgrid()
 }
+```
+
+```{r regression-lm, fig.dim = c(7, 7) * 1.1}
 plot_estimates(res_div_lm_tidy)
@@
 
-Consider doing bootstrap CIs.
+The data is unlikely to meet the assumptions generally considered necessary for simple linear regression.
+In particular, the standard errors (and confidence intervals) are likely to be much too small.
+Bootstrapped intervals could be used to address this issue, if we are interested in understanding the impact of bias on confidence intervals in a more realistic scenario.
 
-```{r, eval = F}
+```{r, include = FALSE, eval = FALSE}
+# A simple approach to improving the validity of the results could be to
+# generate bootstrap CIs.
 lm_boot <- function(.data, ...) {
@@
 
-## Regression on rank
+### Alternate zero-replacement strategy
 
-Similar to Wilcoxin test when there are no ties; not sure about with ties.
+Let's also try adjust the proportions to the Dirichlet posterior mean, with the prior set to the average proportions across all samples.
+This amounts to a pseudocount of `ntaxa` times the mean (observed) proportion of the species, and so adjusts for differences in mean proportion among species.
 
 ```{r}
-stat_test <- function(ps) {
-  ps %>%
-    transform_sample_counts(close_elts) %>%
-    filter_sample_data(div_group %in% c('Low', 'High')) %>%
-    prune_taxa(taxa_to_test, .) %>%
-    as_tibble %>%
-    with_groups(.otu, nest) %>%
-    mutate(
-      fit = map(data,
-        ~lm(
-          rank(.abundance, ties.method = 'average')/n() ~ div_group, 
-          data = .x
-        )
-      ),
+momspi_div_alt <- momspi_div %>% 
+  adjust_dirchlet
+```
+
+```{r regression-lm-alt, fig.dim = c(7, 7) * 1.1}
+res_div_alt_lm <- list(
+  'Observed' = momspi_div_alt,
+  'Calibrated' = momspi_div_alt %>% calibrate(bias_all_vec)
 ) %>%
-    select(-data)
-  # %>% unnest(fit)
-}
-res_div_rank <- list(
-  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.3),
-  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.3) %>% calibrate(bias_all_vec)
-) %>%
-  map_dfr(stat_test, .id = 'type')
-res_div_rank_tidy <- res_div_rank %>%
+  map_dfr(stat_test, taxa = taxa_to_test, .id = 'type')
+res_div_alt_lm_tidy <- res_div_alt_lm %>%
   mutate(across(fit, map, broom.mixed::tidy)) %>%
@@
   filter(term != '(Intercept)')
+
+plot_estimates(res_div_alt_lm_tidy)
+```
+
+## Interpretion
+
+Check the diversity-efficiency relationship
+
+```{r}
+sam <- momspi_div %>% sample_data %>% as_tibble
+sam %>%
+  ggplot(aes(y = div_group, x = mean_efficiency)) +
+  scale_x_log10() +
+  ggdist::stat_slab()
+```
+
+From this plot, we can see that the mean efficiency is 3-10X lower in the High diversity relative to the low diversity group.
+
+The results of a linear regression analysis give us the systematic error in LFCs for the simple linear regression,
+
+```{r}
+mean_eff_lm <- lm(log(mean_efficiency) ~ div_group,
+  data = sam
+)
+mean_eff_lm %>% summary
+mean_eff_lfc <- mean_eff_lm %>% broom::tidy() %>% 
+  filter(term == 'div_groupHigh') %>% pull(estimate)
 ```
@@
 ```{r}
-plot_estimates(res_div_rank_tidy)
+res_div_lm_tidy %>%
+  select(name = type, .otu, value = estimate) %>%
+  pivot_wider() %>%
+  mutate(
+    error = Observed - Calibrated,
+  ) %>%
+  pull(error) %>%
+  summary
+mean_eff_lfc 
 ```
 
-## Gamma-Poisson regression
+Can see that the errors are the same for all species, and nearly equal to the negative LFC in mean efficiency.
+A small difference is expected from the addition of pseudocounts for the species regression.
+
+To understand why the mean efficiency is lower, can try a scatterplot of mean efficiency against diversity with samples colored by most abundant species.
 
 ```{r}
-library(rstanarm)
-options(mc.cores = parallel::detectCores())
+div1 <- div %>%
+  left_join(
+    momspi_rare %>% sample_data %>% as_tibble,
+    by = '.sample'
+  ) %>%
+  mutate(
+    top_species_plot = fct_lump_n(top_species_cal, 7)
+  )
 ```
 
-HERE. Might want to add the (lo) species efficiency to the offset, though it should make absolutely no difference to the fitted models.
-
-HERE.
-Might make sense to use flat priors or MLE inference, to disable regularization, so that the results are most directly comparable to our prediction that the effect of calibration is to shift the coefficients for each species by approximately the same amount.
-
-```{r}
-stat_test <- function(ps) {
-  ps %>%
-    prune_taxa(taxa_to_test, .) %>%
-    filter_sample_data(div_group %in% c('Low', 'High')) %>%
-    as_tibble %>%
-    # left_join(bias_all %>% select(.otu, efficiency), by = '.otu') %>%
-    # mutate(offset = offset + log(efficiency)) %>%
-    with_groups(.otu, nest) %>%
-    mutate(
-      fit = map(data,
-        ~stan_glm(
-          .abundance ~ div_group + offset(offset), 
-          data = .x, seed = 42,
-          family = neg_binomial_2, 
-          algorithm = 'sampling'
-          # algorithm = 'optimizing'
-          # algorithm = 'meanfield'
+```{r, fig.dim = c(8, 6)}
+colors_top <- RColorBrewer::brewer.pal(8, 'Accent')[c(1,2,3,5,4,6,7,8)] 
+div1 %>%
+  ggplot(aes(diversity, mean_efficiency, color = top_species_plot)) +
+  facet_grid(type~index) +
+  scale_x_log10() +
+  scale_y_log10() +
+  ggrastr::rasterize(geom_point(alpha = 0.6), dpi = 300) +
+  stat_smooth(aes(group = 'none'), color = 'black') +
+  labs(color = 'Top species', x = 'Diversity', y = 'Mean efficiency') +
+  scale_color_manual(
+    values = colors_top,
+    labels = function(x) str_replace_all(x, '_', ' ')
   )
-      ),
-    ) %>%
-    select(-data)
-  # %>% unnest(fit)
-}
-res_div_gp <- list(
-  'Observed' = momspi_div %>% mutate_sample_data(offset = gp_offset_obs),
-  'Calibrated' = momspi_div %>% mutate_sample_data(offset = gp_offset_cal)
-) %>%
-  map_dfr(stat_test, .id = 'type')
-# saveRDS(res_div_gp, 'res_div_gp-meanfield.rds')
-# saveRDS(res_div_gp, 'res_div_gp.rds')
-res_div_gp_tidy <- res_div_gp %>%
-  mutate(across(fit, map, broom.mixed::tidy)) %>%
-  unnest(fit) %>% 
-  filter(term != '(Intercept)')
 ```
-
-```{r}
-plot_estimates(res_div_gp_tidy)
-```
-
-interpretation:
-
-- The prop of most species is greater in the high (observed) diversity samples, calibrated or not, because the low-diversity samples tend to be dominated by Lactobacillus (especially cripatus and iners).
-- Because of the sum-to-one constraint, this might just be a consequence of Lacto's having lower abundance - aka a compositional effect
-  - consider talking about this in the case study.
-- This compositional effect is compounded by bias, since Lactos also have high efficiency.
-- hence the observed increases are even larger than the bias-corrected (calibrated) ones.
-- interesting to see how some taxa have a much smaller impact from calibration (Gard and BVAB1 especiaully).
-  - Perhaps this has to do with these species often being dominant?
-    - this means that there are conflicting asssociative forces acting on them - high diversity = low lactobacillus = higher abundance, but also places a cap on how high in abundance these species can be.
-but this argument doesn't explain why our results wouldn't still apply
-  - Are the priors having an impact? What if we flatten the prior on the coefficient?
-  - Is there something about working on counts that is causing the effect?
-  - I can probably learn something by checking the fit on Gard.
-<!--  -->
-
-### flat(er) priors
-
-```{r}
-stat_test <- function(ps) {
-  ps %>%
-    prune_taxa(taxa_to_test, .) %>%
-    filter_sample_data(div_group %in% c('Low', 'High')) %>%
-    as_tibble %>%
-    with_groups(.otu, nest) %>%
-    mutate(
-      fit = map(data,
-        ~stan_glm(
-          .abundance ~ div_group + offset(offset), 
-          data = .x, seed = 42,
-          family = neg_binomial_2, 
-          # prior_intercept = NULL,
-          prior = normal(0, 100),
-          # algorithm = 'sampling'
-          # algorithm = 'optimizing'
-          algorithm = 'meanfield'
-        )
-      ),
-    ) %>%
-    select(-data)
-  # %>% unnest(fit)
-}
-res_div_gp_flat <- list(
-  'Observed' = momspi_div %>% mutate_sample_data(offset = gp_offset_obs),
-  'Calibrated' = momspi_div %>% mutate_sample_data(offset = gp_offset_cal)
-) %>%
-  map_dfr(stat_test, .id = 'type')
-saveRDS(res_div_gp_flat, 'res_div_gp_flat-meanfield.rds')
-# saveRDS(res_div_gp_flat, 'res_div_gp.rds')
-```
-
-```{r}
-res_div_gp_flat_tidy <- res_div_gp_flat %>%
-  mutate(across(fit, map, broom.mixed::tidy)) %>%
-  unnest(fit) %>% 
-  filter(term != '(Intercept)')
-```
-
-```{r}
-p1 <- res_div_gp_flat_tidy %>%
-  ggplot(aes(y = type, x = estimate)) +
-  ggdist::geom_dots()
-p2 <- res_div_gp_flat_tidy %>%
-  mutate(
-    .otu_fct = fct_reorder(.otu, estimate, .fun = min)
-  ) %>%
-  ggplot(aes(y = .otu_fct, x = estimate, color = type)) +
-  geom_vline(xintercept = 0, color = 'grey') +
-  geom_pointrange(
-    aes(
-      xmin = estimate - 2*std.error,
-      xmax = estimate + 2*std.error
-    ))
-p1 / p2 +
-  plot_layout(ncol = 1, heights = c(0.2, 1))
-```
-
-Widening the priors seems to have no impact
-
-### checking on gard
-
-```{r}
-fit <- res_div_gp %>%
-  filter(.otu == 'Gardnerella_vaginalis', type == 'Observed') %>%
-  pull(fit) %>% .[[1]]
-fit %>% summary
-```
-
-```{r, eval = F}
-x <- fit$data %>% as_tibble
-y <- x %>% pull(.abundance)
-# launch_shinystan(fit, ppd = TRUE)
-y_rep <- posterior_predict(fit)
-dim(y_rep)
-```
-
-Let's look at the log proportion.
-To make the distributions more directly comparaible, I'll add the species efficiency in the calibrated part.
-
-```{r}
-eff <- bias_all_vec['Gardnerella_vaginalis']
-x1 <- x %>%
-  select(-offset) %>%
-  mutate(gp_offset_cal = gp_offset_cal + log(eff)) %>%
-  pivot_longer(starts_with('gp_offset'), names_to = 'offset_type', values_to = 'offset') %>%
-  mutate(across(offset_type, str_extract, pattern = '(obs|cal)'))
-```
-```{r}
-x1 %>%
-  ggplot(aes(log(.abundance + 0.3) - offset, fill = div_group)) +
-  facet_wrap(~offset_type, ncol = 1) +
-  geom_histogram()
-```
-
-```{r}
-x1 %>%
-  mutate(log_prop = log(.abundance + 0.3) - offset) %>%
-  ggplot(aes(y = div_group, x = log_prop)) +
-  facet_wrap(~offset_type) +
-  ggdist::stat_halfeye()
-```
-
-```{r}
-x1 %>%
-  mutate(log_prop = log(.abundance + 0.3) - offset) %>%
-  with_groups(c(offset_type, div_group), summarize, across(log_prop, mean))
-```
-
-```{r}
-res_div_gp_tidy %>% filter(.otu == 'Gardnerella_vaginalis')
-```
-
-The estimated coefficients are smaller than the differences in the means.
-
-Can we understand why there is relatively little impact of calibration?
-
-Should also compare for a taxon where there is a larger impact.
-
-C

### 2022-01-10 notebook/_code/momspi-setup.R

@@ -106,3 +106,21 @@ bias_all <- tax_table(momspi) %>% as_tibble %>%
     across(efficiency, ~ . / max(.))
   )
 bias_all_vec <- bias_all %>% select(.otu, efficiency) %>% deframe 
+
+
+## Function for replacing zeros
+adjust_dirchlet <- function(ps) {
+  taxa_mean_prop <- ps %>%
+    otu_table %>%
+    transform_sample_counts(close_elts) %>%
+    orient_taxa(as = 'cols') %>%
+    as('matrix') %>%
+    colMeans
+  stopifnot(sum(taxa_mean_prop) == 1)
+  prior_vec <- taxa_mean_prop * ntaxa(ps)
+  stopifnot(identical(length(prior_vec), ntaxa(ps)))
+  # Note the need for the seq_along trick to get phyloseq to allow this
+  # adjustment
+  ps %>% 
+    transform_sample_counts(~ prior_vec[seq_along(.x)] + .x)
+}

### 2022-01-11 notebook/_posts/2021-11-01-momspi-summary/momspi-summary.Rmd
Revise MOMSPI key-dynamics analysis
- Replace zeros prior to calibration and downstream analysis
- Create plots of example community trajectories
- Perform additional checks and calculations

@@
 
+Finally, let's impute zeros using the Dirichlet posterior mean, with the prior respecting the average proportions across samples.
+
+```{r}
+taxa_mean_prop <- momspi %>%
+  otu_table %>%
+  transform_sample_counts(close_elts) %>%
+  orient_taxa(as = 'cols') %>%
+  as('matrix') %>%
+  colMeans
+stopifnot(sum(taxa_mean_prop) == 1)
+prior_vec <- taxa_mean_prop * ntaxa(momspi)
+stopifnot(identical(length(prior_vec), ntaxa(momspi)))
+# Note the need for the seq_along trick to get phyloseq to allow this
+# adjustment
+momspi <- momspi %>% transform_sample_counts(~ prior_vec[seq_along(.x)] + .x)
+```
+
+## Checks
+
+how many subjects?
+
+```{r}
+cts <- momspi %>% sample_data %>% as_tibble %>% count(host_subject_id)
+nrow(cts)
+```
+
+how does the number of visits vary by subject?
+
+```{r}
+cts %>% pull(n) %>% summary
+cts %>%
+  ggplot(aes(x = n)) +
+  geom_histogram(binwidth = 1)
+```
+
 # Calibrate the MOMSPI measurements
@@
 ```
-
-Note, I'm not using pseudo-props so the zeros are staying zero.
-TODO: Reconsider this choice.
 
@@
 
+# Variation in mean efficiency and individual species
+
+```{r}
+focal_taxa <- momspi %>%
+  transform_sample_counts(close_elts) %>%
+  filter_taxa2(~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
+  taxa_names %>%
+  print
+```
+
+```{r}
+x <- momspi_cal %>%
+  transform_sample_counts(close_elts) %>%
+  filter_tax_table(.otu %in% focal_taxa) %>%
+  as_tibble
+```
+
+```{r}
+taxa_summary <- x %>%
+  mutate(across(.abundance, log)) %>%
+  with_groups(.otu, summarize,
+    across(.abundance, 
+      .fns = c(mean = mean, median = median, sd = sd, var = var),
+      .names = '{.fn}'
+    )
+  )
+taxa_summary %>% arrange(-mean) %>% print(n=Inf)
+```
+
+```{r}
+me_summary <- mean_eff1 %>%
+  mutate(across(mean_efficiency, log)) %>%
+  summarize(
+    across(mean_efficiency,
+      .fns = c(sd = sd, var = var),
+      .names = '{.fn}'
+    )
+  ) %>%
+  print
+```
+
+The mean efficiency has a smaller dispersion than individual species, but it is of a similar magnitude, so there is still potential for it to have a substantial impact.
+
+What about variation within women?
+
+```{r}
+taxa_summary_within <- x %>%
+  mutate(across(.abundance, log)) %>%
+  with_groups(c(.otu, host_subject_id), summarize,
+    across(.abundance,
+      .fns = c(mean = mean, median = median, sd = sd, var = var),
+      .names = '{.fn}'
+    ),
+    n = n()
+  ) %>%
+  filter(n >= 3) %>%
+  with_groups(.otu, summarize,
+    sd_mean = sd(mean),
+    mean_sd = mean(sd)
+  )
+taxa_summary_within %>% print(n=Inf)
+```
+
+```{r}
+me_summary_within <- mean_eff1 %>%
+  mutate(across(mean_efficiency, log)) %>%
+  with_groups(host_subject_id, summarize,
+    across(mean_efficiency,
+      .fns = c(mean = mean, sd = sd, var = var),
+      .names = '{.fn}'
+    ),
+    n = n()
+  ) %>%
+  filter(n >= 3) %>%
+  summarize(
+    sd_mean = sd(mean),
+    mean_sd = mean(sd)
+  ) %>%
+  print
+```
+
+Note, a better way to do this would perhaps be to use anova / linear modeling, or whatever method from those would properly weight the number of samples from each woman.
+
+what about the variation between the first samples from each woman?
+
+```{r}
+taxa_summary_first <- x %>%
+  with_groups(host_subject_id, slice_min, host_visit_number, n = 1) %>%
+  mutate(across(.abundance, log)) %>%
+  with_groups(.otu, summarize,
+    across(.abundance,
+      .fns = c(mean = mean, median = median, sd = sd, var = var),
+      .names = '{.fn}'
+    )
+  )
+taxa_summary_first %>% arrange(-mean) %>% print(n=Inf)
+# taxa_summary %>% arrange(-mean) %>% print(n=Inf)
+```
+
+These sd's are similar to the overall sds, which makes sense given most pairwise comparisons are between women.
+
+```{r}
+me_summary_first <- mean_eff1 %>%
+  with_groups(host_subject_id, slice_min, host_visit_number, n = 1) %>%
+  mutate(across(mean_efficiency, log)) %>%
+  summarize(
+    across(mean_efficiency,
+      .fns = c(sd = sd, var = var),
+      .names = '{.fn}'
+    )
+  ) %>%
+  print
+me_summary
+```
+
+This is somewhat larger, but again similar.
+
+Tentative: 
+We see that the (log-e) mean efficiency sd is around 0.9, while the sd of individual taxa ranges from 2-3.5 (between women and overall).
+Within women, the mean efficiency sd and that of individual taxa are both smaller, such that the relative importance of each seems similar.
+
+Visualize, for the first sample from each woman:
+
+```{r}
+bind_rows(
+  x, 
+  mean_eff1 %>% mutate(.otu = 'Mean efficiency', .abundance = mean_efficiency)
+) %>%
+  with_groups(c(.otu, host_subject_id), slice_min, host_visit_number, n = 1) %>%
+  mutate(
+    across(.abundance, log),
+    across(.otu, fct_reorder, .abundance)
+  ) %>%
+  ggplot(aes(y = .otu, x = .abundance)) +
+  ggdist::stat_slab()
+```
+
 # Fold changes in mean efficiency
@@
 
-# (TODO) Example community trajectory
+# Example community trajectory
 
-Pick a woman with multiple visits and a large change in mean efficiency
+Pick a woman with multiple visits and a large change in mean efficiency to illustrate the potential impact of bias on species trajectories.
+We expect large changes in efficiency to be drive by changes in dominance from Gard to Lactobacillus (or vice versa).
+
+To pick a woman, let's first examine the distribution of geometric standard deviations of mean efficiency across subjects, and pick a subject at the upper end who has multiple visits.
 
 ```{r}
-tb %>%
-  filter(host_subject_id.1 == host_subject_id.2) %>%
-  mutate(fc = mean_efficiency.2 / mean_efficiency.1) %>%
-  arrange(desc(abs(log(fc)))) %>%
-  select(fc, host_subject_id.1, top_plot.1, top_plot.2, host_visit_rank.1)
+me <- mean_eff1 %>%
+  mutate(across(mean_efficiency, log)) %>%
+  with_groups(host_subject_id, summarize,
+    across(mean_efficiency,
+      .fns = c(mean = mean, sd = sd, var = var),
+      .names = '{.fn}'
+    ),
+    n = n()
+  ) %>%
+  filter(n >= 3)
 ```
 
-# Testing
+```{r}
+me %>% slice_max(sd, n = 10)
+```
 
-how many subjects?
+Let's pick subject 'EP280575', who has the most visits within the top 10.
 
 ```{r}
-cts <- momspi %>% sample_data %>% as_tibble %>% count(host_subject_id)
-nrow(cts)
-#> [1] 577
-cts %>% pull(n) %>% summary
-#>    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-#>    1.00    2.00    3.00    3.39    4.00    9.00 
-cts %>%
-  ggplot(aes(x = n)) +
-  geom_histogram(binwidth = 1)
-  # geom_dotplot(binwidth = 1)
+sid <- 'EP280575'
+ps_obs <- momspi %>%
+  filter_sample_data(host_subject_id == sid) %>%
+  transform_sample_counts(close_elts) %>%
+  filter_taxa2(~max(.) > 1e-3)
+ps_cal <- momspi_cal %>%
+  filter_sample_data(host_subject_id == sid) %>%
+  transform_sample_counts(close_elts) %>%
+  prune_taxa(taxa_names(ps_obs), .)
+x <- list(Calibrated = ps_cal, Observed = ps_obs) %>%
+  map_dfr(as_tibble, .id = "type") %>%
+  mutate(across(.otu, fct_relabel, function(x) str_replace_all(x, '_', ' ')))
 ```
 
-577 subjects
+```{r, fig.dim = c(6, 15)}
+taxa_to_plot <- x %>% 
+  filter(type == 'Calibrated') %>%
+  with_groups(.otu, summarize, 
+    med = median(log10(.abundance)),
+    max = max(log10(.abundance)),
+  ) %>%
+  filter(max > -2.5) %>%
+  arrange(-max) %>%
+  pull(.otu) %>%
+  print
+x %>%
+  filter(.otu %in% taxa_to_plot) %>%
+  mutate(.otu = factor(.otu, taxa_to_plot)) %>%
+  ggplot(aes(host_visit_number, .abundance, color = type)) +
+  scale_y_log10() +
+  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
+  geom_line(aes(group = type)) +
+  labs(y = 'Proportion', title = 'Species proportions') +
+  scale_color_brewer(type = 'qual', palette = 6) +
+  theme(legend.position = 'top')
+```
+
+For the final plot, I'll pick the most abundant 3 taxa (which are driving the mean efficiency) and two more taxa that show the different behaviors.
+
+```{r trajectories-within-subject-high-var, fig.dim = c(6, 9)}
+taxa_to_plot <- c(
+  'Gardnerella_vaginalis', 'Lachnospiraceae_BVAB1', 'Lactobacillus_iners',
+  'Ureaplasma_cluster23', 'Megasphaera_OTU70_type1') %>%
+  str_replace_all('_', ' ')
+p_sp <- x %>%
+  filter(.otu %in% taxa_to_plot) %>%
+  mutate(.otu = factor(.otu, taxa_to_plot)) %>%
+  filter(as.integer(.otu) <= 6) %>%
+  ggplot(aes(host_visit_number, .abundance, color = type)) +
+  scale_y_log10() +
+  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
+  geom_line(aes(group = type)) +
+  labs(y = 'Proportion', title = 'Species proportions') +
+  scale_color_brewer(type = 'qual', palette = 6) +
+  theme(legend.position = 'top')
+p_me <- mean_eff1 %>%
+  filter(host_subject_id == sid) %>%
+  ggplot(aes(host_visit_number, mean_efficiency)) +
+  scale_y_log10() +
+  labs(y = 'Efficiency', title = 'Mean efficiency') +
+  geom_line(aes(group = 1))
+p_me / p_sp &
+  annotate('rect', xmin = 3, xmax = 4, ymin = Inf, ymax = 0, alpha = 0.1) &
+  labs(x = 'Host visit') &
+  plot_layout(heights = c(0.1, 1)) &
+  plot_annotation(tag_levels = 'A')
+```
+
+For comparison, let's also look at a long trajectory where there is no major change in mean efficiency.
+
+```{r}
+me %>% slice_min(sd, n = 10)
+```
+
+```{r}
+sid <- 'EP694646'
+
+ps_obs <- momspi %>%
+  filter_sample_data(host_subject_id == sid) %>%
+  transform_sample_counts(close_elts) %>%
+  filter_taxa2(~max(.) > 1e-3)
+ps_cal <- momspi_cal %>%
+  filter_sample_data(host_subject_id == sid) %>%
+  transform_sample_counts(close_elts) %>%
+  prune_taxa(taxa_names(ps_obs), .)
+x <- list(Calibrated = ps_cal, Observed = ps_obs) %>%
+  map_dfr(as_tibble, .id = "type") %>%
+  mutate(across(.otu, fct_relabel, function(x) str_replace_all(x, '_', ' ')))
+ntaxa(ps_obs)
+```
+
+```{r trajectories-within-subject-low-var, fig.dim = c(6, 9)}
+p_sp <- x %>%
+  mutate(.otu = fct_reorder(.otu, -.abundance)) %>%
+  ggplot(aes(host_visit_number, .abundance, color = type)) +
+  scale_y_log10() +
+  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
+  geom_line(aes(group = type)) +
+  labs(y = 'Proportion', title = 'Species proportions') +
+  scale_color_brewer(type = 'qual', palette = 6) +
+  theme(legend.position = 'top')
+p_me <- mean_eff1 %>%
+  filter(host_subject_id == sid) %>%
+  ggplot(aes(host_visit_number, mean_efficiency)) +
+  scale_y_log10() +
+  labs(y = 'Efficiency', title = 'Mean efficiency') +
+  geom_line(aes(group = 1))
+p_me / p_sp &
+  labs(x = 'Host visit') &
+  plot_layout(heights = c(0.1, 1)) &
+  plot_annotation(tag_levels = 'A')
+```
+
+Note that the variation in mean efficiency is very small.
 

### 2022-01-11 case-studies.Rmd
Revise vaginal-microbiome case study
Revise text, update regression analysis, and add SI figures of
trajectory and regression analysis.

@@
 To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
-The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.57X and a geometric standard deviation of just 1.06X (Figure \@ref(fig:leopold2020host-variation)).
+The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.57X and a geometric standard deviation of just 1.06X (SI Figure \@ref(fig:leopold2020host-variation)).
 <!-- Explanation -->
@@
 It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
-The pathogen is absent in pre-infection samples but tends to dominate the community post-infection (Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)), leading to an average increase in mean efficiency ranging from 3.5X to 5X across host genotypes and an expectation that commensals will typically decrease in proportion due to the sum-to-one constraint.
+The pathogen is absent in pre-infection samples but tends to dominate the community post-infection (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)), leading to an average increase in mean efficiency ranging from 3.5X to 5X across host genotypes and an expectation that commensals will typically decrease in proportion due to the sum-to-one constraint.
 Without bias correction, the log proportion of commensals decrease in each host genotype, often substantially (log-e fold changes less than -2); however, bias leads to overly negative LFCs due to the higher mean efficiency in post-infection samples.
-Bias correction substantially lessens the magnitude of LFC estimates, in some cases leading to estimates that are approximately 0 or slightly positive (Figure \@ref(fig:leopold2020host-infection-lfc)).
+Bias correction substantially lessens the magnitude of LFC estimates, in some cases leading to estimates that are approximately 0 or slightly positive (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
 
@@
 
-The vaginal microbiomes of pregnant women have been a source of intensive study due to its perceived connection with the health of both mother and her developing child.
-Many MGS studies have found associations of specific microbial species and community characteristics with rates of urinary tract and sexually-transmitted infections, bacterial vaginosis (BV), and preterm births.
-Yet these associations and the baseline, apparently healthy, community state vary across studies of different populations and using different MGS methods.
-A number of studies have experimentally probed the taxonomic bias of different MGS protocols on vaginally-associated species (@yuan2012eval,@brooks2015thet,@gill2016eval,@graspeuntner2018sele,todo ref on taxonomic resolution and also sampling methods).
-These studies have revealed substantial bias among protocols, including variation in extraction efficiency by 10-40X and an inability for certain primer sets to amplify highly-prevalent species.
-Taxonomic bias has been proposed as a potential explanation for discrepancies across studies (@callahan2017repl, Others), there has so far been little quantitative analysis of this potential explanation.
-Here we re-analyze data from published studies to examine the potential impact of taxonomic bias on DA analysis of vaginal communities of pregnant women.
+The vaginal microbiome during pregnancy has been a source of intensive study due to its apparent connection with the health of both mother and her developing child.
+Many MGS studies have found associations of specific microbial species and community characteristics with rates of urinary tract and sexually-transmitted infections, bacterial vaginosis (BV), and preterm birth.
+Yet these associations vary across studies of different populations and using different MGS methods.
+DA analyses of the vaginal microbiome are commonly based on proportions, creating an opportunity for taxonomic bias to impact results.
+A number of studies have experimentally demonstrated substantial taxonomic bias among MGS protocols and individual steps (such as extraction and PCR amplification) in vaginal samples or _in vitro_ samples of vaginally-associated species (@yuan2012eval,@brooks2015thet,@gill2016eval,@graspeuntner2018sele).
+<!-- TODO: perhaps add ref on taxonomic resolution and also sampling methods -->
+Bias has been proposed as a potential explanation for discrepancies across studies (@callahan2017repl, Others?), but there has so far been little quantitative analysis of this possibility.
+Here we use empirical bias measurements from control samples to investigate the role of bias in proportion-based DA analyses of vaginal microbiomes from a recent large-cohort study of pregnant women.
 
-- P on brooks, momspi studies + calibration
-  - GC
-  - Intro the studies
-  - intro ability to use calibration to look at impact of bias - perhaps don't explain as such
-  - how we did it given limited species composition of the mock
+As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, @fettweis2019thev collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were measured by amplicon sequencing of the 16S V1-V3 hypervariable region to yield species-level bacterial taxonomic profiles.
+Taxonomic bias of this MGS protocol was previously investigated by @brooks2015thet and @mclaren2019cons, using measurements by @brooks2015thet of cellular mock communities of seven common, clinically-relevant vaginal bacterial species.
+<!-- TODO: check on clinical relevance -->
+Of these, _Lactobacillus iners_ had the highest efficiency, which was nearly 30X larger than that of the species with the lowest efficiency, _Gardnerella vaginalis_.
+A second _Lactobacillus_ species, _L. crispatus_, had an efficiency that was approximately 2X less than _L. iners_ and 15X greater than _G. vaginalis_.
+These species, along with the unculturable _Lachnospiraceae BVAB1_, are the most common top (most-abundant) species (SI analysis) and can reach high proportions in individual samples, indicating that shifts between them might drive large changes in the mean efficiency which might in turn impact DA results.
 
-We first examine what the taxonomic bias observed in control experiments by @brooks2015thet might imply for DA analysis of vaginal microbiomes of pregnant women reported by @fettweis2019thev.
-As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, @fettweis2019thev collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were measured by amplicon sequencing of the 16S V1-V3 hypervariable region to yield species-level bacterial taxonomic profiles.
-DA analyses of the vaginal microbiome, including those of @fettweis2019thev, are most commonly based on species proportions, creating opportunity for bias to have an impact on their results.
-Previously, the authors evaluated taxonomic bias of this MGS protocol via the construction and measurement of cellular mock communities of seven vaginal bacterial species (@brooks2015thet), but did not directly account for this bias in @fettweis2019thev.
-We used the control measurements from @brooks2015thet to infer the bias of the MOMS-PI 16S protocol and investigate its potential impact on DA analyses of vaginal profiles from the MOMS-PI study.
-To estimate the relative efficiencies among all species, we directly estimated the efficiencies in the control samples following the method of @mclaren2019cons and imputed the remaining efficiencies based on taxonomic relationships.
-We then used these imputed efficiencies to calibrate the measured taxonomic profiles from the MOMS-PI study and compare DA analyses with and without calibration.
-Such imputed efficiencies carry considerable uncertainty, which we ignore to simplify our presentation. 
-Notably, while this uncertainty can majorly impact abundances in individual samples, it has a much weaker impact on LFCs across samples (SI analysis).
+We sought to assess this possibility using a joint analysis of the control measurements from @brooks2015thet and the microbiome measurements from the MOMS-PI study.
+To obtain the taxonomic bias among all species identified in the MOMS-PI measurements, we used the taxonomic relationships with the seven control species to impute the efficiencies for the remaining species.
+We used these imputed efficiencies to calibrate (correct the effect of bias in) the MOMS-PI measurements, examine variation in the mean efficiency varied across samples, and compare the results of a DA analysis with and without bias correction.
 
-To assess the potential for bias to affect fold changes in species proportions, we first used the imputed efficiencies to consider how the mean efficiency varied across women.
-In contrast to the fungal experiment, the mean efficiency varies substantially across vaginal samples, primarily driven by variation in which species is most abundant (Figure \@ref(fig:momspi-mean-efficiency-dist).
-The four most common top species in the observed and calibrated profiles were _Gardnerella vaginalis_, _Lactobacillus iners_, _Lactobacillus crispatus_, and _Lachnospiraceae BVAB1_.
-All of these except _Lachnospiraceae BVAB1_ where included in the mock-community experiment, from which we estimated that _L.
-iners_ has an efficiency of 2X relative to _L.
-crispatus_ and 30X relative to _G.
-vaginalis_, with _L.
-iners_ and _G.
-vaginalis_ having the maximum and minimum efficiencies among the 7 control species.
+<!-- Such imputed efficiencies carry considerable uncertainty, which we ignore to simplify our presentation.  -->
+<!-- Notably, while this uncertainty can majorly impact abundances in individual samples, it has a much weaker impact on LFCs across samples (SI analysis TODO). -->
 
-HERE
-- finish explaining how the variation is driven by changes in these taxa, esp G and L 
-- add that the importance of these taxa, and the tendency for one species to domiannt, is a gneral well-suported feature of vaginal mb studies, though the identities varies by population and mgs protocol.
-- segue into further considerations - what is the impact of this variation?
-  - trajectories
-    - impact of big shifts 
-  - DA analysis
-
-(TODO: confirm true that often these species dominate. Also check the efficiency numbers)
-
-Consequently, the mean efficiency tends to track that of most dominant species, with _Lactobacillus_-dominated samples typically having an efficiency that is 3-20X greater _Gardnerella_-dominated samples (Figure \@ref(fig:momspi-mean-efficiency-dist)).
-
-Shifts between _Lactobacillus_ and _Gardnerella_ dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)), and typically result in substantial fold changes in mean efficiency.
-Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure TODO).
-In contrast, bias-correction only affects the baseline value of species ratios and not the fold changes.
-
-TODO:
-- consider whether the error is significant, which depends on the species FC and not just the mean efficiency FC.
-    - would be good to somehow visualize or summarize this.
-    - should give a better interpretation/assessment of the bottom line here.
-    - one way: discuss the variation in log mean efficiency versus that in species proportions.
-See a contrast to the Leopold situation; here the variation is lower but similar enough to potentially matter
-- to show that variation within women is lower, compute standard deviation of log mean efficiency within women and average this (perhaps weighted by n)
-- create 1-2 trajectories to show
-- also show ratios
-- writing/main issue: Consider how often the error matters, in FCs, and make trajectories, - make sure can support the idea that it usually doesn't matter, if that is true.
+The mean efficiency varies substantially across vaginal samples (Figure \@ref(fig:momspi-mean-efficiency-dist).
+This variation appears to be primarily driven by variation in which species is most abundant, samples in which a _Lactobacillus_ species is most abundant (after calibration) typically have an efficiency that is 3-20X greater than samples in which _G.
+vaginalis_ is most abundant.
+Shifts between _Lactobacillus_-dominance and _Gardnerella_-dominance are common in between-women comparisons and occasionally occur between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
+These shifts typically result in substantial fold changes in mean efficiency (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)) and can cause spurious fold changes in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).
 
@@
 
-- P: Simulated DA analysis
-    - GC: We next sought to see how ...
-    - Methods
-    - Results
-    - Discussion/caveats
-    - notes: could do with fido as well, to show that the results are unaffected.
+We next considered whether the observed variation in mean efficiency could cause systematic error in a DA analysis.
+In particular, we hypothesized that DA analysis of species proportions versus a covariate that is associated with _Lactobacillus_ and/or _Gardnerella_ would be particularly prone to spurious results.
+Patient metadata was not available due to privacy restrictions; we therefore sought a clinically relevant covariate to use in a regression analysis that could be determined directly from the microbiome profiles and that we _a priori_ expected to be associated with the proportions of _Lactobacillus_ and _Gardnerella_.
+Alpha diversity metrics such as species richness, the Shannon index, and the (Inverse) Simpson index have been repeatedly found to be strongly positively associated with bacterial vaginosis (BV; @srinivasan2012bact, @cartwright2018mult).
+@cartwright2018mult found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100% and specificity of 85.1%.
+In addition, it is commonly observed that samples from women with and without BV that are dominated by _Lactobacillus spp._ tend to have higher diversity, whereas samples dominated by _Gardnerella_ tend to have lower diversity.
+We therefore chose to perform a DA analysis of species proportion versus alpha diversity, hypothesizing that _Lactobacillus_ and _Gardnerella_ would drive a negative association of mean efficiency with diversity and thereby distort DA estimates for all species.
+We split samples into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
+We then estimated the LFC in proportion from low- to high-diversity samples for a 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles.
+As expected, the mean efficiency was higher in low-diversity samples due to a larger fraction of samples dominated by _Lactobacillus_ and a lower fraction with samples dominated by _Gardnerella_.
+The decline in mean efficiency in the high diversity group caused a concomitant increase in the LFCs of all species when bias was not accounted for.
+This error due to bias resulted in serious magnitude errors in nearly all species and sign errors in over one third, with errors of both types seen for several clinically relevant species.
 
-We next considered whether the variation in mean efficiency could cause systematic error in a DA analysis of species proportions.
-Due to privacy protections, we did not have access to subject variables that could be used as covariates in a DA analysis.
-We therefore performed a DA analysis of species proportion versus community alpha diversity.
-A low-diversity, _Lactobacillus_-dominated microbiome is frequently cited as the typical healthy microbiome and high alpha diversity is considered an indication of bacterial vaginosis and has been repeatedly found to be postively associated with risk of preterm birth.
-(Though notably diverse and low-_Lactobacillus_ microbiomes are also common in healthy women and even typical in certain populations, including segments of the MOMSPI cohort.)
-We split the samples into classes with low and high Shannon diversity in the observed (uncorrected) community composition, leaving out samples with medium diversity, and performed a DA analysis of species proportion versus diversity class.
-We evaluated several types of DA analyses: The simple linear regression model applied to log proportion (Section \@ref()); negative-binomial (or gamma-poisson) regression, which models expected log proportion with a linear model while also accounting for the random, discrete nature of sequencing read counts; and a rank-regression method that approximates the commonly-used non-parametric Wilcoxin test (TODO).
-We performed each DA method with and without correcting for bias on a set of X species passing a minimum abundance prevalence criterion and compared the effect of bias on estimated regression coefficients for individual species.
-The mean efficiency tended to be larger in the high-diversity class (presumably due to a lower proportion of Lactobacillus, CHECK); unsurprisingly, the effect of bias was therefore always to lower the regression coefficient.
-For each method, bias caused significant magnitude and sign errors for X-Y species, including creating sign errors for X, Y, and Z species respectively.
-The absolute error due to bias is the same for all species in the simple linear regression (see previous section), but for the other two methods varied substantially among species (though was always positive, corresponding to the decrease in log mean efficiency in the high-diversity class).
-The DA results also differed substantially across methods.
-Together, these variations meant that whether bias had a significant impact on a particular species was often method dependent.
+### Notes
 
-- ?? P: Real application: Callahan et al, Gard bias - maybe in future
-    - what bias could explain discrepancies?
-
-MOVE: Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
+- It might be cleaner story-wise to redo the diversity partitioning using the threshold identified by @cartwright2018mult
+- For the DA analysis of proportions vs.
+diversity, I tried other types of DA analysis methods (Gamma-Poisson regression and regression on ranks, similar to Wilcoxon test) to understand how the impact of bias differs across methods.
+In each case bias has a big effect but the species whose estimates are most affected varies.
+I would need to revise and verify this analysis before including its results.
+One reason for doing so would be to show that our qualitative finding about how bias distorts results applies beyond the basic linear regression model.
+- Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
 
@@
 
-Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate.
-The MOMSPI regression example illustrates a case where the mean efficiency is highly variable and also correlated with the covariate (diversity class).
-The Leopold T1 example illustrates a case where the mean efficiency simply doesn't vary significantly across samples, and hence there is no room for bias to affect regression coefficients.
+<!-- Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate. -->
+<!-- The MOMSPI regression example illustrates a case where the mean efficiency is highly variable and also correlated with the covariate (diversity class). -->
+<!-- The Leopold T1 example illustrates a case where the mean efficiency simply doesn't vary significantly across samples, and hence there is no room for bias to affect regression coefficients. -->

### 2022-01-11 supplemental-figures.Rmd
Revise vaginal-microbiome case study
Revise text, update regression analysis, and add SI figures of
trajectory and regression analysis.

@@
 
+<br><br>
+
 ```{r leopold2020host-infection-mean-efficiency-dist, fig.cap = '(ref:cap-leopold2020host-infection-mean-efficiency-dist)'}
@@
 (ref:cap-leopold2020host-infection-mean-efficiency-dist) **The mean efficiency tends to increase after infection due to the high proportion of the pathogen.**
+
+<br><br>
 
@@
 
-<!-- Implications: MOMSPI -->
+<br><br>
+
+<!-- Case study: MOMSPI -->
 
@@
 (ref:cap-momspi-mean-efficiency-fcs) **Fold changes in the mean efficiency within and between women in the MOMS-PI study.**
+
+<br><br>
+
+```{r momspi-trajectory, fig.cap = '(ref:cap-momspi-trajectory)', out.width = '75%'}
+fs::path(
+  "notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files", 
+  "figure-html5/trajectories-within-subject-high-var-1.svg"
+) %>%
+  include_svg
+```
+
+(ref:cap-momspi-trajectory) **In vaginal microbiome measurements, shifts between _Lactobacillus_ and _Gardnerella_ dominance can drive spurious fold changes in other, lower-abundance species.** The figure shows species proportions and mean efficiency trajectories over consecutive clinical visits for a subject in the MOMS-PI study whose microbiome samples showed substantial variation in mean efficiency.
+The subject's samples are dominated by _Gardnerella vaginalis_ and _Lachnospiraceae BVAB1_ during the first three visits before transitioning to being dominated by _Lactobacillus iners_ between visits 3 and 4.
+This transition drives a sharp increase in the mean efficiency, which significantly distorts the fold changes in the observed (uncalibrated) microbiome measurements for species with less dramatic fold changes.
+Two exemplar species are shown to illustrate the magnitude (_Ureaplasma cluster 23_) and sign (_Megasphaera OTU70 type1_) errors that can arise in this situation.
+
+<br><br>
+
+```{r momspi-regression, fig.cap = '(ref:cap-momspi-regression)'}
+fs::path(
+  "notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity_files", 
+  "figure-html5/regression-lm-alt-1.svg"
+) %>%
+  include_svg
+```
+
+(ref:cap-momspi-regression) **Bias distorts log fold changes (LFCs) in species proportions in a regression analysis of vaginal microbiome samples from the MOMS-PI study.** Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
+The LFC in proportion from low- to high-diversity samples was estimated for 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles following a simple zero-replacement procedure.
+Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95% confidence intervals for each species.
+The difference between the calibrated and observed estimate for each species equals the negative LFC in mean efficiency.
 

### 2022-01-11 v/3019b9c9c1fa3b13645d60ebb08913530bda476a/_main.tex
Make minor edits to appendix
This commit adds the rendered manuscript associated with
3019b9c9c1fa3b13645d60ebb08913530bda476a

--- 2021-11-29 v/8f81dd5293e7ef5bc3621bbbec2f5d7a41b6560d/_main.tex
@@
 Callahan\footnote{North Carolina State University}}
-\date{2021-11-29}
+\date{2022-01-11}
 
@@
 
-\emph{This manuscript was rendered from commit 8f81dd5293e7ef5bc3621bbbec2f5d7a41b6560d.}
+\emph{This manuscript was rendered from commit 3019b9c9c1fa3b13645d60ebb08913530bda476a.}
 
@@
 
-\hypertarget{implications}{%
-\section{Implications for real-world inference~}\label{implications}}
-
-Are biologically significant errors in DA analyses likely in practice?
-We can begin to answer this question using a combination of theoretical analysis, consideration of hypothetical scenarios, and analysis of case studies for which some information about taxonomic bias is available.
-
-\hypertarget{systematic-error-in-slope-or-lfc-estimates}{%
-\subsection{Systematic error in slope or LFC estimates}\label{systematic-error-in-slope-or-lfc-estimates}}
-
-Recall from Section \ref{differential-abundance} that the error in the slope or LFC in a DA regression analysis is proportional to the covariance of the log mean efficiency with the covariate.
-This covariance can be split into two components: the standard deviation of (log) mean efficiency and its correlation with the covariate.
-Hence when considering whether taxonomic bias creates large error, it can be useful to separately ask whether the mean efficiency is likely to vary across samples and whether it is likely to be correlated with a covariate of interest.
-
-One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
-To characterize bias in a 16S sequencing protocol developed for the Vaginal Human Microbiome Project (VaHMP), Brooks et al.
-(\protect\hyperlink{ref-brooks2015thet}{2015}) performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome.
-This protocol was later used for the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study (Fettweis et al. (\protect\hyperlink{ref-fettweis2019thev}{2019})).
-We used the method for measuring and correcting bias developed by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to measure bias from the mock communities and analyze its potential impact on DA analyses that use the MOMSPI measurements (SI Analysis MOMS-PI).
-The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single species often forming a vast majority of calibrated or uncalibrated read counts.
-The three species that most often dominate the calibrated profiles are \emph{Gardnerella vaginalis}, \emph{Lactobacillus iners}, and \emph{Lactobacillus crispatus}.
-Our mock-community analysis indicate \emph{G.
-vaginalis} has the lowest efficiency of the seven species, roughly 20X lower than \emph{L.
-crispatus} and 30X lower than \emph{L.
-iners} (the most efficiently measured).
-As a result, the mean efficiency varies substantially across samples, with samples dominated by a \emph{Lactobacillus} species typically having an efficiency that is 3-20X greater than samples dominated by \emph{Gardnerella} (Figure \ref{fig:momspi-mean-efficiency-dist}).
-Shifts between \emph{Lactobacillus} and \emph{Gardnerella} dominance are common in between-women comparisons, and also occasionally occur between subsequent sampling points in individual women (SI Figure \ref{fig:momspi-mean-efficiency-fcs}).
-Examination of the multiplicative trajectories in the proportions of individual species indicates that such transitions can cause spurious fold changes in lower-abundance species (SI Figure TODO).
-Decreases in mean efficiency during transitions from \emph{Lactobacillus} to \emph{Gardnerella} dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify \emph{Gardnerella}.
+\hypertarget{case-studies}{%
+\section{Case studies ~}\label{case-studies}}
+
+To better understand the potential impact of taxonomic bias on DA analysis in practice, we conducted several case studies spanning a range of biological scenarios and sequencing technologies.
+
+\hypertarget{foliar-fungi-experiment}{%
+\subsection{Foliar fungi experiment}\label{foliar-fungi-experiment}}
+
+Measurement of control communities along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+Gnotobiotic community experiments are well suited to this form of \emph{calibration via community controls} since, unlike most natural ecosystems, it is possible to assemble `mock communities' containing all species in known relative abundances.
+
+Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) are the first to use mock-based calibration in a gnotobiotic microbiome experiment.
+In a study of host-commensal-pathogen interactions, Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) inoculated plants with 8 commensal fungal species and subsequently exposed plants to a fungal pathogen.
+To investigate the joint effect of colonization order and host genotype, plants varied in source location and in which commensal colonized first.
+The authors used ITS amplicon sequencing to measure communities before and after pathogen infection, along with mock communities with quantified genome concentrations of DNA from the 9 species.
+These DNA mock communities allowed the authors to directly measure the bias due to the sequencing workflow following DNA extraction and correct for it when analyzing the primary experimental samples following the method of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
+Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification (CITE) and the substantial ribosomal copy-number-variation (CNV) in fungi (Lofgren et al.
+(\protect\hyperlink{ref-lofgren2019geno}{2019})).
+The authors found a 13X difference between the most and least efficiently measured commensal, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
+We reanalyzed this experiment to understand the impact that bias might have had on the study's analyses were it not accounted for.
+
+Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) used the ITS sequencing profiles of the pre-infection samples to understand the impact of host genotype and an experimental treatment---which species was allowed to colonize first---on the commensal community composition prior to infection.
+Two DA analyses were conducted to address these questions.
+First, the authors used linear regression of log species proportion (via negative-binomial regression to account for random read-sampling error) to ask whether host genotype, treatment, and their interaction significantly impacted overall community composition.
+Second, the authors quantified the strength of priority effects---the advantage of being allowed to colonize first---in different host genotypes by the LFC in the proportion of a species when allowed to colonize first versus later with the majority of commensals.
+
+Since these analysis are both based on proportions, bias could in principle impact their results.
+We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for\footnote{The two sets of analyses can be seen at the following links: \href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html}{negative-binomial regression analysis}; \href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html}{priority-effects analysis}}.
+To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
+The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.57X and a geometric standard deviation of just 1.06X (SI Figure \ref{fig:leopold2020host-variation}).
+Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
+This answer relates to the fact that the mean efficiency is a weighted \emph{arithmetic} average of species proportions and so is insensitive to large multiplicative variation in low-proportion species.
+The pre-infection samples in the Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) study are always dominated by the three species with the highest efficiencies, whose efficiencies vary only by 3X (and just 1.5X between the two most dominant).
+The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions \(\ll 1\), even after bias correction) and hence don't significantly affect the mean efficiency.
+Because the log mean efficiency varies much less than the log proportions for each taxon, there is little room for it to distort the results of DA analyses based on variation in log proportions.
+
+We additionally considered the potential for bias to impact analysis of how commensal taxa responded to infection.
+In particular, it is interesting to consider whether any commensals are able to increase in absolute concentration in response to infection.
+The experiment performed by Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration
+measured using the total-community normalization method (Equation \eqref{eq:density-prop-meas}).
+It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
+The pathogen is absent in pre-infection samples but tends to dominate the community post-infection (SI Figure \ref{fig:leopold2020host-infection-mean-efficiency-dist}), leading to an average increase in mean efficiency ranging from 3.5X to 5X across host genotypes and an expectation that commensals will typically decrease in proportion due to the sum-to-one constraint.
+Without bias correction, the log proportion of commensals decrease in each host genotype, often substantially (log-e fold changes less than -2); however, bias leads to overly negative LFCs due to the higher mean efficiency in post-infection samples.
+Bias correction substantially lessens the magnitude of LFC estimates, in some cases leading to estimates that are approximately 0 or slightly positive (SI Figure \ref{fig:leopold2020host-infection-lfc}).
+
+Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-community normalization method (Equation \eqref{eq:density-prop-meas}).
+We consider two scenarios.
+\emph{Scenario 1:} Total genome concentration in each sample is perfectly known and used to measure the concentrations of individual taxa via Equation \eqref{eq:density-prop-meas}.
+In this case, bias in the MGS measurements would create absolute errors in the LFCs for genome concentration identical to those for proportions.
+The scientific error, however, may be much worse.
+Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where taxa that persist or even increase appear to decrease.
+\emph{Scenario 2:} Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-community normalization method.
+In this case, we should expect the increase in mean efficiency to have little to no impact on LFCs, since it would be offset by the increase in mean efficiency of the total-concentration measurement.
+Moreover, since the total ITS concentration and ITS sequencing are both performed on the extracted DNA, we can also expect the effect of taxonomic bias caused by DNA sequencing to be offset.
+Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-community measurement is made prior to normalization.
+
+\hypertarget{vaginal-microbiomes-of-pregant-women}{%
+\subsection{Vaginal microbiomes of pregant women}\label{vaginal-microbiomes-of-pregant-women}}
+
+The vaginal microbiome during pregnancy has been a source of intensive study due to its apparent connection with the health of both mother and her developing child.
+Many MGS studies have found associations of specific microbial species and community characteristics with rates of urinary tract and sexually-transmitted infections, bacterial vaginosis (BV), and preterm birth.
+Yet these associations vary across studies of different populations and using different MGS methods.
+DA analyses of the vaginal microbiome are commonly based on proportions, creating an opportunity for taxonomic bias to impact results.
+A number of studies have experimentally demonstrated substantial taxonomic bias among MGS protocols and individual steps (such as extraction and PCR amplification) in vaginal samples or \emph{in vitro} samples of vaginally-associated species (Yuan et al.
+(\protect\hyperlink{ref-yuan2012eval}{2012}),Brooks et al.
+(\protect\hyperlink{ref-brooks2015thet}{2015}),Gill et al.
+(\protect\hyperlink{ref-gill2016eval}{2016}),Graspeuntner et al.
+(\protect\hyperlink{ref-graspeuntner2018sele}{2018})).
+Bias has been proposed as a potential explanation for discrepancies across studies (Callahan et al.
+(\protect\hyperlink{ref-callahan2017repl}{2017}), Others?), but there has so far been little quantitative analysis of this possibility.
+Here we use empirical bias measurements from control samples to investigate the role of bias in proportion-based DA analyses of vaginal microbiomes from a recent large-cohort study of pregnant women.
+
+As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, Fettweis et al.
+(\protect\hyperlink{ref-fettweis2019thev}{2019}) collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were measured by amplicon sequencing of the 16S V1-V3 hypervariable region to yield species-level bacterial taxonomic profiles.
+Taxonomic bias of this MGS protocol was previously investigated by Brooks et al.
+(\protect\hyperlink{ref-brooks2015thet}{2015}) and McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}), using measurements by Brooks et al.
+(\protect\hyperlink{ref-brooks2015thet}{2015}) of cellular mock communities of seven common, clinically-relevant vaginal bacterial species.
+Of these, \emph{Lactobacillus iners} had the highest efficiency, which was nearly 30X larger than that of the species with the lowest efficiency, \emph{Gardnerella vaginalis}.
+A second \emph{Lactobacillus} species, \emph{L. crispatus}, had an efficiency that was approximately 2X less than \emph{L. iners} and 15X greater than \emph{G. vaginalis}.
+These species, along with the unculturable \emph{Lachnospiraceae BVAB1}, are the most common top (most-abundant) species (SI analysis) and can reach high proportions in individual samples, indicating that shifts between them might drive large changes in the mean efficiency which might in turn impact DA results.
+
+We sought to assess this possibility using a joint analysis of the control measurements from Brooks et al.
+(\protect\hyperlink{ref-brooks2015thet}{2015}) and the microbiome measurements from the MOMS-PI study.
+To obtain the taxonomic bias among all species identified in the MOMS-PI measurements, we used the taxonomic relationships with the seven control species to impute the efficiencies for the remaining species.
+We used these imputed efficiencies to calibrate (correct the effect of bias in) the MOMS-PI measurements, examine variation in the mean efficiency varied across samples, and compare the results of a DA analysis with and without bias correction.
+
+The mean efficiency varies substantially across vaginal samples (Figure \ref{fig:momspi-mean-efficiency-dist}.
+This variation appears to be primarily driven by variation in which species is most abundant, samples in which a \emph{Lactobacillus} species is most abundant (after calibration) typically have an efficiency that is 3-20X greater than samples in which \emph{G.
+vaginalis} is most abundant.
+Shifts between \emph{Lactobacillus}-dominance and \emph{Gardnerella}-dominance are common in between-women comparisons and occasionally occur between consecutive visits in individual women (SI Figure \ref{fig:momspi-mean-efficiency-fcs}).
+These shifts typically result in substantial fold changes in mean efficiency (SI Figure \ref{fig:momspi-mean-efficiency-fcs}) and can cause spurious fold changes in the trajectories of lower-abundance species (SI Figure \ref{fig:momspi-trajectory}).
 
@@
 
-A second study in which control communities make it possible to directly analyze the impact of taxonomic bias in experimental samples was performed by Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}).
-To study the interactions between a host plant and its fungal commensals and pathogen, the authors inoculated trees with a fungal synthetic community and later exposed plants to a fungal pathogen.
-The authors used ITS amplicon sequencing to measure communities before and after infection, along with mock communities which they used to estimate the species efficiencies with the method of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
-(DNA mocks were used, so that bias due to DNA extraction is not included, but other major bias sources such as PCR and ribosomal copy-number-variation are included.)
-We reanalyzed this experiment to consider the impact that bias has on different possible DA analyses (SI Analysis Fungi).
-As noted by Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}), there was a 13X difference between the most and least efficiently measured commensals, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
-Yet despite this 13X variation among commensals, the mean efficiency is remarkably stable across communities prior to pathogen colonization.
-Consequently, taxonomic bias creates little systematic error in proportion-based DA analysis of these pre-colonization samples.
-The high efficiency of the pathogen did, however, typically lead to a roughly 6X increase in the mean efficiency of post-colonization samples.
-Hence DA analysis of how commensals responded to the pathogen did not first discard pathogen reads would therefore tend to suggest that the commensals had substantially smaller fold changes than they actually did, unless bias were accounted for.
+We next considered whether the observed variation in mean efficiency could cause systematic error in a DA analysis.
+In particular, we hypothesized that DA analysis of species proportions versus a covariate that is associated with \emph{Lactobacillus} and/or \emph{Gardnerella} would be particularly prone to spurious results.
+Patient metadata was not available due to privacy restrictions; we therefore sought a clinically relevant covariate to use in a regression analysis that could be determined directly from the microbiome profiles and that we \emph{a priori} expected to be associated with the proportions of \emph{Lactobacillus} and \emph{Gardnerella}.
+Alpha diversity metrics such as species richness, the Shannon index, and the (Inverse) Simpson index have been repeatedly found to be strongly positively associated with bacterial vaginosis (BV; Srinivasan et al.
+(\protect\hyperlink{ref-srinivasan2012bact}{2012}), Cartwright et al.
+(\protect\hyperlink{ref-cartwright2018mult}{2018})).
+Cartwright et al.
+(\protect\hyperlink{ref-cartwright2018mult}{2018}) found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100\% and specificity of 85.1\%.
+In addition, it is commonly observed that samples from women with and without BV that are dominated by \emph{Lactobacillus spp.} tend to have higher diversity, whereas samples dominated by \emph{Gardnerella} tend to have lower diversity.
+We therefore chose to perform a DA analysis of species proportion versus alpha diversity, hypothesizing that \emph{Lactobacillus} and \emph{Gardnerella} would drive a negative association of mean efficiency with diversity and thereby distort DA estimates for all species.
+We split samples into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
+We then estimated the LFC in proportion from low- to high-diversity samples for a 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles.
+As expected, the mean efficiency was higher in low-diversity samples due to a larger fraction of samples dominated by \emph{Lactobacillus} and a lower fraction with samples dominated by \emph{Gardnerella}.
+The decline in mean efficiency in the high diversity group caused a concomitant increase in the LFCs of all species when bias was not accounted for.
+This error due to bias resulted in serious magnitude errors in nearly all species and sign errors in over one third, with errors of both types seen for several clinically relevant species.
+
+\hypertarget{notes}{%
+\subsubsection{Notes}\label{notes}}
+
+\begin{itemize}
+\tightlist
+\item
+  It might be cleaner story-wise to redo the diversity partitioning using the threshold identified by Cartwright et al. (\protect\hyperlink{ref-cartwright2018mult}{2018})
+\item
+  For the DA analysis of proportions vs.~diversity, I tried other types of DA analysis methods (Gamma-Poisson regression and regression on ranks, similar to Wilcoxon test) to understand how the impact of bias differs across methods.
+In each case bias has a big effect but the species whose estimates are most affected varies.
+I would need to revise and verify this analysis before including its results.
+One reason for doing so would be to show that our qualitative finding about how bias distorts results applies beyond the basic linear regression model.
+\item
+  Decreases in mean efficiency during transitions from \emph{Lactobacillus} to \emph{Gardnerella} dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify \emph{Gardnerella}.
+\end{itemize}
+
+\hypertarget{gut-microbiome}{%
+\subsection{Gut microbiome}\label{gut-microbiome}}
+
+(Has not been updated)
 
@@
 
-Systematic error in DA results requires that the variation in log mean efficiency is correlated with the covariate.
+Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate.
 The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
-For example, in the vaginal microbiome a decline in \emph{Lactobacillus} and rise of species including \emph{Gardnerella vaginalis} is associated with preterm birth in pregnant women; hence the variation in mean efficiency driven by these species may also create an association of mean efficiency with preterm status.
-In the plant-fungal experiment of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}), the increase in pathogen proportion after infection leads to a large increase in the mean efficiency over time, which systematic increases the observed decline in the proportions of commensal species (SI Analysis).
 The Bacteroidetes-to-Firmicutes ratio has also been linked to host conditions.
@@
 
-Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
-One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
-A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
-Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
-Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
-In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
-For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.
-
-Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
-For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
+\hypertarget{microbial-growth-in-marine-sediments}{%
+\subsection{Microbial growth in marine sediments}\label{microbial-growth-in-marine-sediments}}
+
+Our route for mean efficiency to become associated with the covariate is if the covariate quantifies a biological process that preferentially selects for a microbial trait that also tends to increase or decrease measurement efficiency.
+We illustrate this potential mechanism with a study of microbial growth in marine sediments.
+
+The surface layers of marine sediments harbor diverse and abundant microbiomes.
+Total cell density and species richness decrease with depth as resources are consumed in older, deeper sediment layers; however, some taxa are able to grow and increase in density as they are slowly buried.
 Lloyd et al.
-(\protect\hyperlink{ref-lloyd2020evid}{2020}) estimate the log fold change in the estimated absolute abundance of various taxa with sediment depth (as a proxy for time) to determine which taxa are able to persist and even grow in this difficult environment.
-It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
-As the relative abundance of tougher species increases with depth, the mean extraction efficiency might therefore decrease.
-This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
-Additional measurements by Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020}) make it possible to investigate this possibility explicitly.
-The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for particular taxa.
-For instance, in Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years using the proportion-based densities and 10.3 years by qPCR (Lloyd et al.
-(\protect\hyperlink{ref-lloyd2020evid}{2020}) Table 1 and Figure 4).
-These doubling times correspond to highly similar estimated slopes of log density per year of \(1/10.1 \approx 0.99\) and \(1/10.3\approx 0.97\).
-The small difference between these slopes suggests that there is little correlation of log mean efficiency with burial time and hence little systematic error in the estimated doubling times derived from the proportion-based density estimates.
-
-Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.
-
-The biological significance of the error in \(\hat \beta\) depends on the relative magnitude of the LFC (or covariance with the covariate \(x\)) in the focal species vs in the mean efficiency.
-It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
-Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
-However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for other species.
-The experimental and computational methods discussed in Section \ref{solutions} can mitigate taxonomic bias in these clearly-problematic cases while also helping to amass empirical evidence on the conditions in which bias is unlikely to cause major inferential errors.
-
-\hypertarget{loss-of-precision-and-power}{%
-\subsection{Loss of precision and power}\label{loss-of-precision-and-power}}
-
-Another way in which taxonomic bias hinders DA analysis is by increasing the noise in abundance measurements, leading to a reduction in the precision of LFC estimates and thus the power to detect true associations.
-
-There are two distinct mechanisms by which such reductions occur.
-The first was described in Section \ref{differential-abundance}: Variation in the mean efficiency that is random (in the sense of being unassociated with the covariates in a regression analysis) tends to increase the standard errors in estimated regression coefficients.
-As noted above, this effect may be the dominant one in cases like the HMP2 IBD dataset where large variation in dominant phylum is independent of disease status.
-Correcting for the variation in mean efficiency using the calibration methods described in the next section can reverse this effect, thereby increasing the precision as well reducing the systematic error in DA inference.
-
-The second mechanism involves the counting noise that stems from the random sampling of reads in the sequencing experiment.
-This counting noise is expected to be the dominant source of noise for species whose expected proportion \emph{after accounting for taxonomic bias} is \(\lesssim 1 / \text{total reads}\) for a given sample.
-Having an atypically low efficiency makes a species much more likely to be near this threshold and hence to receive highly uncertain measurements.
-Such imprecise measurements can make it near impossible to make biologically meaningful conclusions.
-For example, in the experiment of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) described above, the median read counts in samples taken after pathogen colonization for the 5 commensals with the lowest efficiencies were 0, 0, 0, 1, and 11.
-These species' efficiencies ranged from 40X-10X lower than the pathogen's efficiency; thus, in the absence of bias the counts may have been much higher and it might have been possible to make much more precise conclusions about how their relative abundances changed during pathogen invasion.
-The literature on the vaginal microbiome during pregnancy provides us with another example: Disagreement among studies over whether \emph{Gardnerella} is positively associated with preterm birth seems to be in part driven by the fact that several studies used 16S primers that are extremely poor at amplifying \emph{Gardnerella} and hence lack the precision necessary to find associations even were they to exist.
-Unfortunately, even perfect knowledge of the measurement efficiencies will not allow us to overcome the low precision caused by small expected read counts; however, the methods described in the section provide principled ways to assess whether the low counts are due to taxonomic bias rather than truly low frequencies in a species' abundance.
+(\protect\hyperlink{ref-lloyd2020evid}{2020}) performed a systematic assessment of growth rates of bacterial and archaeal taxa over a depth of 10 cm (corresponding to \textasciitilde40 years of burial time) in sediment of the White Oak River estuary.
+To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-community normalization method (Equation \eqref{eq:density-prop-meas}), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy (CARD-FISH).
+The authors refer to these absolute densities as FRAxC measurements, for `fraction of 16S reads times total cell counts'.
+The FRAxC measurements were used to infer growth rate from the slope of a simple linear regression of log cell density against burial time over the first 3 cm below the bioirrigation layer (corresponding to \(\sim 8\) years of burial).
+To validate the inference of positive growers, the authors compared the growth rates from FRAxC-based inference from two sediment cores, qPCR measurements in these cores for a few reference taxa, and FRAxC-based inference in two replicate laboratory incubation experiments.
+
+Taxonomic bias could lead to systematic error in FRAxC-derived growth rates if sample mean efficiency tends to systematically vary with burial time (or equivalently, depth).
+One possibility is that microbes with tougher cell walls tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
+In this case, we would expect the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that decay sufficiently slowly would be mistakenly inferred to have positive growth.
+
+We can test the hypothesis that systematic variation in log mean efficiency with depth distorts inferred growth rates using the qPCR measurements of two clades also collected by Lloyd et al.
+(\protect\hyperlink{ref-lloyd2020evid}{2020}).
+In Section \ref{review-absolute-methods}, we argue that taxonomic bias in targeted qPCR measurements is expected to create a roughly constant fold error and yield fold changes in absolute density that are relatively unaffected by changes in mean efficiency.
+Thus comparing qPCR to FRAxC growth rates on the same reference taxa allows us to estimate the systematic error in FRAxC growth rates.
+Because the systematic error in regression slopes due to bias is the same for each taxon (Equation \eqref{eq:regression-error}), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
+The first soil core included qPCR measurements of a single archaeal clade, \emph{Bathyarchaeota}, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of 0.099/yr by FRAxC and 0.097/yr by qPCR).
+The second soil core included qPCR measurements of \emph{Bathyarchaeota} and a second clade, \emph{Thermoprofundales}/MBG-D.
+In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by 0.012/yr for \emph{Bathyarchaeota} (0.112/yr by FRAxC and 0.1/yr by qPCR) and by 0.086/yr for \emph{Thermoprofundales}/MBG-D (0.294/yr by FRAxC and 0.208/yr by qPCR).
+A low number of experimental samples and noise in both the FRAxC and qPCR measurements place significant uncertainty in these measurements; however, the fact that FRAxC-derived growth rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreases with depth in a manner that systematically biases FRAxC-derived rates to higher values.
+The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr suggests an error large enough to impact results for some taxa classified as positive growers, whose FRAxC growth rates ranged between 0.04/yr and 0.5/yr.
+Overall, the comparison between FRAxC and qPCR measurements gives support to the study conclusions, but suggest that species at the lower end of this range of positive FRAxC-derived rates may in fact be merely persisting or even slowly declining in abundance.
+
+\hypertarget{conclusions}{%
+\subsection{Conclusions}\label{conclusions}}
 
@@
 We imputed the efficiencies of other species using taxonomic relationships (Methods).
-Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \ref{implications}.
+Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \ref{case-studies}.
 
@@
 For simplicity, we ignore this random variation when considering the impact of taxonomic bias on MGS measurement.
-The effect of random variation in regression analysis is addressed in Appendix
-\ref{appendix-regression} and the effect of random counting error is discussed in Section \ref{implications}.
+The effect of random variation in regression analysis is addressed in Appendix \ref{appendix-regression} and the effect of random counting error is discussed in Section \ref{case-studies}.
 
@@
 \begin{align}
-  \label{eq:app-density-prop-error}
+  \label{eq:density-prop-error-with-total-error}
   \widehat{\text{density}}_{i}(a) 
@@
 \begin{align}
-  \label{eq:app-density-prop-error}
+  \label{eq:density-prop-error-linearity}
   \widehat{\text{density}}_{i}(a) 
@@
 
+TODO: Argue that these methods may yield constant fold errors.
+
 \hypertarget{appendix-regression}{%
@@
 \begin{align}
-  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
+  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_z}{s_x}.
 \end{align}
@@
 \centering
+\includegraphics{figures/export-pdf/notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study_files/figure-html5/variation-in-proportions-and-mean-efficiency-1.pdf}
+\caption{\label{fig:leopold2020host-variation}\textbf{In the pre-infection samples from Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}), multiplicative variation in taxa proportions is much larger than that in the mean efficiency.} Panel A shows the distribution of the proportions of each commensal isolate (denoted by its genus) across all samples collected prior to pathogen inoculation; Panel C shows the distribution of the (estimated) sample mean efficiency across these same samples on the same scale; and Panel B shows the efficiency of each taxon estimated from DNA mock communities as point estimates and 90\% bootstrap percentile confidence intervals.
+Efficiencies are shown relative to the most efficiently measured taxon (\emph{Fusarium}).}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/export-pdf/notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study_files/figure-html5/infection-mean-efficiency-dist-1.pdf}
+\caption{\label{fig:leopold2020host-infection-mean-efficiency-dist}\textbf{The mean efficiency tends to increase after infection due to the high proportion of the pathogen.}}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/export-pdf/notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study_files/figure-html5/infection-lfc-estimates-1.pdf}
+\caption{\label{fig:leopold2020host-infection-lfc}\textbf{Bias correction increases the estimated increase in log proportion in response to infection for commensal taxa across all host genotypes.} Shown are the estimated log fold change (LFC) and 95\% confidence intervals from simple linear regression of log (base e) proportion against experimental timepoint for commensal taxa.
+Negative values indicate that the proportion of the taxon decreased on average in response to infection, which we expect due to the increase in pathogen abundance and the sum-to-one constraint.
+Bias leads to artificially low estimates, as the pathogen drives an increase in mean efficiency.
+In Western genotypes, several taxa whose proportion is observed to decrease without correction are instead found to remain stable or even increase when bias is corrected.}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics{figures/export-pdf/notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files/figure-html5/momspi-mean-efficiency-fcs-1.pdf}
 \caption{\label{fig:momspi-mean-efficiency-fcs}\textbf{Fold changes in the mean efficiency within and between women in the MOMS-PI study.}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.75\linewidth]{figures/export-pdf/notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files/figure-html5/trajectories-within-subject-high-var-1} \caption{\textbf{In vaginal microbiome measurements, shifts between \emph{Lactobacillus} and \emph{Gardnerella} dominance can drive spurious fold changes in other, lower-abundance species.} The figure shows species proportions and mean efficiency trajectories over consecutive clinical visits for a subject in the MOMS-PI study whose microbiome samples showed substantial variation in mean efficiency.
+The subject's samples are dominated by \emph{Gardnerella vaginalis} and \emph{Lachnospiraceae BVAB1} during the first three visits before transitioning to being dominated by \emph{Lactobacillus iners} between visits 3 and 4.
+This transition drives a sharp increase in the mean efficiency, which significantly distorts the fold changes in the observed (uncalibrated) microbiome measurements for species with less dramatic fold changes.
+Two exemplar species are shown to illustrate the magnitude (\emph{Ureaplasma cluster 23}) and sign (\emph{Megasphaera OTU70 type1}) errors that can arise in this situation.}\label{fig:momspi-trajectory}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/export-pdf/notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity_files/figure-html5/regression-lm-alt-1.pdf}
+\caption{\label{fig:momspi-regression}\textbf{Bias distorts log fold changes (LFCs) in species proportions in a regression analysis of vaginal microbiome samples from the MOMS-PI study.} Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
+The LFC in proportion from low- to high-diversity samples was estimated for 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles following a simple zero-replacement procedure.
+Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95\% confidence intervals for each species.
+The difference between the calibrated and observed estimate for each species equals the negative LFC in mean efficiency.}
 \end{figure}
@@
 \url{https://doi.org/10.1073/pnas.1705899114}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cartwright2018mult}{}}%
+Cartwright, Charles P., Amanda J.
+Pherson, Ayla B.
+Harris, Matthew S.
+Clancey, and Melinda B.
+Nye.
+2018.
+{``{Multicenter study establishing the clinical validity of a nucleic-acid amplification--based assay for the diagnosis of bacterial vaginosis}.''} \emph{Diagn.
+Microbiol.
+Infect.
+Dis.} 92 (3): 173--78.
+\url{https://doi.org/10.1016/j.diagmicrobio.2018.05.022}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-gill2016eval}{}}%
+Gill, Christina, Janneke H.
+H.
+M.
+van de Wijgert, Frances Blow, and Alistair C.
+Darby.
+2016.
+{``{Evaluation of Lysis Methods for the Extraction of Bacterial DNA for Analysis of the Vaginal Microbiota}.''} Edited by Peter E.
+Larsen.
+\emph{PLoS One} 11 (9): e0163148.
+\url{https://doi.org/10.1371/journal.pone.0163148}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-gloor2017micr}{}}%
@@
 \url{https://doi.org/10.3389/fmicb.2017.02224}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-graspeuntner2018sele}{}}%
+Graspeuntner, Simon, Nathalie Loeper, Sven Künzel, John F.
+Baines, and Jan Rupp.
+2018.
+{``{Selection of validated hypervariable regions is crucial in 16S-based microbiota studies of the female genital tract}.''} \emph{Sci.
+Rep.} 8 (1): 9678.
+\url{https://doi.org/10.1038/s41598-018-27757-8}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lofgren2019geno}{}}%
+Lofgren, Lotus A., Jessie K.
+Uehling, Sara Branco, Thomas D.
+Bruns, Francis Martin, and Peter G.
+Kennedy.
+2019.
+{``{Genome‐based estimates of fungal rDNA copy number variation across phylogenetic scales and ecological lifestyles}.''} \emph{Mol.
+Ecol.} 28 (4): 721--30.
+\url{https://doi.org/10.1111/mec.14995}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-lozupone2013meta}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-srinivasan2012bact}{}}%
+Srinivasan, Sujatha, Noah G.
+Hoffman, Martin T.
+Morgan, Frederick A.
+Matsen, Tina L.
+Fiedler, Robert W.
+Hall, Frederick J.
+Ross, et al.
+2012.
+{``{Bacterial Communities in Women with Bacterial Vaginosis: High Resolution Phylogenetic Analyses Reveal Relationships of Microbiota to Clinical Criteria}.''} Edited by Adam J.
+Ratner.
+\emph{PLoS One} 7 (6): e37818.
+\url{https://doi.org/10.1371/journal.pone.0037818}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-stammler2016adju}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-yuan2012eval}{}}%
+Yuan, Sanqing, Dora B.
+Cohen, Jacques Ravel, Zaid Abdo, and Larry J.
+Forney.
+2012.
+{``{Evaluation of Methods for the Extraction and Purification of DNA from the Human Microbiome}.''} Edited by Jack Anthony Gilbert.
+\emph{PLoS One} 7 (3): e33865.
+\url{https://doi.org/10.1371/journal.pone.0033865}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-zemb2020abso}{}}%

### 2022-01-13 case-studies.Rmd
Fix error in leopold2020host analysis
The mean efficiency was calculated incorrectly in the case study
analysis, leading to errors in the plots of mean efficiency and in the
offsets used for bias correction in the linear regression analysis. This
commit fixes the calculation and regenerates the figures and regression
analysis, and updates the main text and supplemental-figure caption to
match the new results.

@@
 To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
-The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.57X and a geometric standard deviation of just 1.06X (SI Figure \@ref(fig:leopold2020host-variation)).
+The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.62X and a geometric standard deviation of just 1.05X (SI Figure \@ref(fig:leopold2020host-variation)).
 <!-- Explanation -->
@@
 In particular, it is interesting to consider whether any commensals are able to increase in absolute concentration in response to infection.
-The experiment performed by @leopold2020host does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration
-measured using the total-community normalization method (Equation \@ref(eq:density-prop-meas)).
+The experiment performed by @leopold2020host does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration measured using the total-community normalization method (Equation \@ref(eq:density-prop-meas)).
 It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
-The pathogen is absent in pre-infection samples but tends to dominate the community post-infection (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)), leading to an average increase in mean efficiency ranging from 3.5X to 5X across host genotypes and an expectation that commensals will typically decrease in proportion due to the sum-to-one constraint.
-Without bias correction, the log proportion of commensals decrease in each host genotype, often substantially (log-e fold changes less than -2); however, bias leads to overly negative LFCs due to the higher mean efficiency in post-infection samples.
-Bias correction substantially lessens the magnitude of LFC estimates, in some cases leading to estimates that are approximately 0 or slightly positive (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
+The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, raising the mean efficiency of post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
+Across different host genotypes, the average increase in mean efficiency ranges from 2.5X to 5.2X. 
+We used simple linear regression to measure the average change in log proportion of each commensal for each host genotype, with and without bias correction.
+Regardless of whether bias is accounted for, the log proportion of each commensal decreased, which is unsurprising given the pathogen's growth and the sum-to-one constraint of species proportions.
+Without bias correction, however, the LFCs were lower by an amount corresponding to the inverse change in log mean efficiency.
+The magnitude errors created were substantial for a large fraction of commensal-host pairs; in several cases, the LFC estimates without bias correction were clearly negative, whereas the estimates with bias correction were close to 0.
 
@@
 The scientific error, however, may be much worse.
-Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where taxa that persist or even increase appear to decrease.
+Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where some taxa that persist or increase in abundance instead appear to decrease.
 _Scenario 2:_ Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-community normalization method.
@@
 
-## Vaginal microbiomes of pregant women
+## Vaginal microbiomes of pregnant women
 

### 2022-01-13 notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study.Rmd
Fix error in leopold2020host analysis
The mean efficiency was calculated incorrectly in the case study
analysis, leading to errors in the plots of mean efficiency and in the
offsets used for bias correction in the linear regression analysis. This
commit fixes the calculation and regenerates the figures and regression
analysis, and updates the main text and supplemental-figure caption to
match the new results.

@@
     sample_sum = sample_sums(.),
-    mean_efficiency = transform_sample_counts(., close_elts) %>%
+    mean_efficiency = calibrate(., bias_t1, norm = 'close') %>%
       perturb(bias_t1, norm = 'none') %>% 
@@
     mean_efficiency_initial = mean_efficiency,
-    mean_efficiency = transform_sample_counts(., close_elts) %>%
+    mean_efficiency = calibrate(., bias_t1, norm = 'close') %>%
       perturb(bias_t1, norm = 'none') %>% 
@@
   mutate_sample_data(., 
-    mean_efficiency = transform_sample_counts(., close_elts) %>%
+    mean_efficiency = calibrate(., bias_inf, norm = 'close') %>%
       perturb(bias_inf, norm = 'none') %>% 
@@
     x = "Mean efficiency", 
-    y = "Number of samples"
+    y = "Number of samples",
+    fill = 'Timepoint'
   )

### 2022-01-13 supplemental-figures.Rmd
Fix error in leopold2020host analysis
The mean efficiency was calculated incorrectly in the case study
analysis, leading to errors in the plots of mean efficiency and in the
offsets used for bias correction in the linear regression analysis. This
commit fixes the calculation and regenerates the figures and regression
analysis, and updates the main text and supplemental-figure caption to
match the new results.

@@
 (ref:cap-leopold2020host-infection-lfc) **Bias correction increases the estimated increase in log proportion in response to infection for commensal taxa across all host genotypes.** Shown are the estimated log fold change (LFC) and 95% confidence intervals from simple linear regression of log (base e) proportion against experimental timepoint for commensal taxa.
-Negative values indicate that the proportion of the taxon decreased on average in response to infection, which we expect due to the increase in pathogen abundance and the sum-to-one constraint.
-Bias leads to artificially low estimates, as the pathogen drives an increase in mean efficiency.
-In Western genotypes, several taxa whose proportion is observed to decrease without correction are instead found to remain stable or even increase when bias is corrected.
+Negative values indicate that the proportion of the taxon decreased on average in response to infection, which we expect due to an increase in pathogen abundance and the sum-to-one constraint of proportions.
+Bias leads to artificially low estimates, as the increased pathogen proportion drives an increase in mean efficiency.
 

### 2022-01-13 case-studies.Rmd
Draft conclusions for case-studies section

@@
 
-## Gut microbiome
-
-(Has not been updated)
-
-It is clear that an individual species with an unusually high or low efficiency can drive large changes in the mean efficiency when it shifts from a very large ($\sim 1$) to very small proportion ($\ll 1$) of the sample.
-Consideration of dynamics in the human gut illustrates another a way in which large shifts in higher-level taxa might create similar shifts, due to the fact that efficiencies do not vary independently among closely related species.
-Gut microbiomes are dominated by two phyla: The Bacteroidetes and the Firmicutes, the ratio of which varies substantially across individuals.
-Within a sample, there are usually multiple species contributing to the abundance of each phylum, with the median Inverse Simpson diversity being around 4 for each phylum in the HMP2 IBD study.
-Thus changes from Bacteroidetes to Firmicutes dominance are unlikely to be driven by a single species.
-However, if Firmicutes species tend to have systematically different efficiencies than Bacteroidetes species, then changes in phylum dominance can be expected to drive substantial changes in the mean efficiency even if many species contribute.
-Indeed, several studies have found DNA extraction protocols to more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though substantial variation within phyla also occurs, @mclaren2019cons).
-
-Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate.
-The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
-The Bacteroidetes-to-Firmicutes ratio has also been linked to host conditions.
-Yet in the HMP2 IBD dataset, the large variation in this ratio is largely independent of disease status, suggesting that the primary effect of any variation in mean efficiency driven by these phyla would be to increase noise rather than create systematic error in DA analysis.
-
 ## Microbial growth in marine sediments
@@
 
-## Conclusions
+## Summary and conclusions
 
-<!-- Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate. -->
-<!-- The MOMSPI regression example illustrates a case where the mean efficiency is highly variable and also correlated with the covariate (diversity class). -->
-<!-- The Leopold T1 example illustrates a case where the mean efficiency simply doesn't vary significantly across samples, and hence there is no room for bias to affect regression coefficients. -->
+The impact of bias can depend on protocol, biological system, and type of DA analysis being done.
+Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of section X help suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.
+
+<!-- P: Canceling of consistent bias does occur to an important degree -->
+
+First, the error caused by consistent taxonomic bias _can mostly cancel_ in cross-sample comparisons and so not impact DA analyses of fold changes in species proportions.
+Our theoretical results from Section @ref(differential-abundance) indicate that when the mean efficiency is roughly constant across samples, the error in proportions cancels in fold change calculations and is absorbed by the intercept term in regression models.
+We observed this scenario in the analysis of pre-infection fungal microbiome samples of @leopold2020host and when analyzing the trajectories of the vaginal microbiomes of women when the dominant species remained constant^[This point could use more support in the MOMSPI case study], and a stable mean efficiency within marine soil cores is plausible and consistent with (though unable to be fully determined by) the different growth rate estimates in the @lloyd2020evid experiment.
+Section \@ref(differential-abundance) showed that absolute DA analysis using total-community normalization are suspectible to errors as with DA analysis of proportions; yet our analysis of two (hypothetical) approaches to analyzing absolute changes in response to infection in the fungal microbiome experiment indicates that the increasingly common approach of pairing marker-gene sequencing with qPCR of the total marker density can largely mitigate this effect, since here what matters is the variation in the ratio of mean efficiencies of the MGS measurement to the total-density measurement and this ratio may remain roughly constant if bias is largely shared by the two measurements.^[This is a simple-to-state and important point that should perhaps be made in the previous theoretical section.
+More empirical evidence could be given by re-analysis of @tettamantiboshier2020comp or similar studies that perform qPCR of total-community 16S and of specific species.
+Here I expect to find a fairly good agreement in the LFCs from total-community normalization and species-specific qPCR, since much of the bias is likely shared by 16S sequencing and qPCR.
+However, there are various complications to interpretting these comparisons.]
+
+Yet in other cases, the mean efficiency can vary substantially and create substantial error in fold changes between pairs of samples.
+We saw examples when comparing fungal microbiomes pre- and post-infection in the foliar-fungi experiment and comparing vaginal microbiomes with different dominant species in the MOMS-PI experiment.
+The impact of these errors on results of DA regression analysis across many samples depends on whether the mean efficiency varies systematically with the covariate of interest.
+<!-- P: but bias can still cause substantial systematic error -->
+In these two examples, systematic variation of the mean efficiency arose as high-efficiency species tended to dominate samples in one of the sample conditions (post-infection foliar samples or low-diversity vaginal samples).
+Thus one way for signficant error in DA regression results to arise is when a small number of (or just one) species that have particularly high or low mean efficiencies, form a large proportion of the community in a substantial fraction of samples, and are associated with the covariate of interest.
+<!-- This dynamic was responsible for the errors observed in the DA analysis of pre- and post-infection fungal microbiomes from @leopold2020host and in the vaginal microbiome profiles of the MOMS-PI study.
+-->
+
+For many experimental systems, however, there may not be any one species that frequently forms a large fraction of the community.
+In these systems, systematic variation of the mean efficiency can still arise through the collective change of many species that are associated with the covariate of interest.
+We described a hypothesized scenario in the marine-sediment case study in which increased burial time (the covariate) selects for lysis-resistant (and so lower efficiency) species.
+As another example, treatment with a certain antibiotic might selectively kill easier-to-lyze Gram negative species (or harder-to-lyze Gram positives) and thereby decreasing (or increasing) the mean efficiency in samples collected from a host post-treatment.
+Such situations can arise whenever there is a microbial trait that is associated with both measurement efficiency and the biological processes of intersest.
+An example besides cell-wall structure is ribosomal copy number, which increases a species' efficiency in ribosomal amplicon measurements and is positively associated with metabolic rate.
+<!-- Even when there is not a single trait that affects efficiency and the focal biological process,  -->
+A more generic mechanism by which mean-efficiency associations might arise is from the evolutionary relationships among species.
+Species with more recent common ancestry are expected to be more similar across a wide range of heritable traits, which include traits that affect measurement efficiency (such as cell wall structure, genome size, ribosomal copy number, PCR binding sequence) as well as traits that affect the biological processes under study.
+For example, differences in the phylum-level composition of samples from two conditions might be driven by many species that show phylum-level conservation in a relevant biological trait.
+If these species also show phylum-level conservation in (potentially different) traits that affect measurement efficiency, an association of mean efficiency with the condition can arise.
+
+<!-- P: Bias can also reduce the precision and power of DA analyses -->
+
+So long as the mean efficiency does not vary systematically with the regression covariates, the error in fold changes serve primarily to reduce the precision and power of regression analyses without systematically distorting estimates and so does not necessarily lead to invalid inferences.
+But the reduction in precision could be substantial, particularly given the small sample sizes common in many microbiome studies, and thereby substantially limit the ability to draw meaningful conclusions from a study^[I suspect this non-systematic variation in the mean efficiency is the more typical situation, but we currently lack an example of it in the case studies.
+We can find a semi-real example by taking real microbiome data from a case-control-type design and simulating a strong degree of bias.
+I did this with the HMP2 IBD study, but thinnk it might be better to find a different example where there are some clear DA results.].
+
+<!-- Conclusion: Need solutions -->
+
+These observations provide reasons to both worry and hope.
+It seems likely that in many experiments the mean efficiency is consistent or is at least not associated with the covariate, so that DA inferences remain valid.
+Yet it is not obvious _a priori_ in which studies this condition holds, and there are plausible mechanisms that can create problematic associations of the mean efficiency even in ecosystems with high species diversity.
+Thus while we should not discount the large set of existing DA results, we should seek ways to better assess the robustness of results from previous studies and to measure and correct the error caused by bias in future ones.
+
+<!-- ## cut from conclusions -->
+
+<!-- ### bias can push species below the detection limit -->
+
+<!-- - Another mechanism, not considered here, by which bias can dramatically reduce precision, is by pushing speices below the sequencing detection limit. -->
+<!-- - e.g. of primers that don't detect Gard; problems with estimates of rare and/or low eff taxa on leopold -->
+<!-- - this mechanism can interact with variation in the mean efficiency - e.g. of fungal samples after Mel increases -->
+
+<!-- ### Other cases where bias can matter that we didn't consider -->
+
+<!-- Optional for this draft; may also want to leave for final 'conclusions' / discussion -->
+
+<!-- - DA analyses that aren't based on fold changes - don't get canceling, and might have bigger impact -->
+<!-- - taxonomic aggregation - e.g. of Lactobacillus in vaginal; (mabye: Firmicutes in gut) - can get a varying efficiency within the taxon -->
+<!-- - ?? zero replacement, interpretation -->

### 2022-01-14 notebook/_posts/2021-11-09-explore-gut-data/explore-gut-data.Rmd
Add calculations to HMP2 analysis

@@
 Most subjects have many samples.
+
+In the HMP2 paper, they classified some samples as dysbiotic.
+We don't have that info here.
 
@@
 ```
+
+## Enterotypes?
+
+Work of others suggests we might be able to identify just a handul of species/genera/groups that tend to be dominant.
+
+```{r}
+
+```
+
+what taxa tend to reach dominance/very high abundance?
+
+```{r}
+ps.gen <- ps %>% tax_glom('genus')
+```
+
+```{r}
+ps.gen %>%
+  transform_sample_counts(close_elts) %>%
+  as_tibble %>%
+  filter(.abundance > 0.3) %>%
+  dplyr::count(phylum, family, genus) %>%
+  arrange(-n) %>%
+  print(n=20)
+```
+
+```{r}
+ps %>%
+  tax_glom('family') %>%
+  transform_sample_counts(close_elts) %>%
+  as_tibble %>%
+  filter(.abundance > 0.3) %>%
+  dplyr::count(phylum, family) %>%
+  arrange(-n) %>%
+  print(n=20)
+```
+
+This suggests that when we are considering phylogenetic conservation for the purposes of bias effect, then conservation within firmicutes may not be as important as that within specific sub-groups like Lachnospiraceae and Ruminococcaceae.
+
+Note, the original HMP protocol was found to be biased in favor of Bacteroidetes.

### 2022-01-14 introduction.Rmd
Minor text edits

@@
 In these _differential-abundance (DA) analyses_, the change in abundance of a microbial taxon across samples or conditions is used to infer ecological dynamics or find microbes that are associated with specific host diseases or environmental conditions.
-Although standard MGS measurements lose information about total microbial density—and so are typically used to analyze the abundances of taxa relative to each other—new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density.
+Standard MGS measurements lose information about total microbial density and so are typically used to analyze the abundances of taxa relative to each other.
+But new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density.
 In its various forms, DA analysis remain one of, if not the most, common forms of analyses applied to MGS data to elucide the inner workings of microbiomes and their relationships to host and environmental health.
@@
 Although often associated with features specific to marker-gene sequencing—the variation among species in marker copy numbers and in primer-binding and amplification efficiencies—the existence of large variation in DNA extraction efficiencies and in the ability to correctly classify reads make taxonomic bias a universal feature of both marker-gene and shotgun measurements.
-As a result of taxonomic bias, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo,@mclaren2019cons).
+As a result of taxonomic bias, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo, @mclaren2019cons).
 These errors have been found in some cases to supersede sizable biological differences (e.g.
@@
 The typical approach taken to counter taxonomic bias is to standardize the measurement protocol used within a given study, with the (often tacit) assumption being that samples measured by the same protocol will be affected by bias in the same way and so the measured differences between samples will be unaffected.
-For example, if taxonomic bias were to cause the measured proportion of a species to consistently be 10X too high, we would still be able to accurately infer the fold change in its proportion across samples (@kevorkian2018esti,@lloyd2020evid).
+For example, if taxonomic bias were to cause the measured proportion of a species to consistently be 10X too high, we would still be able to accurately infer the fold change in its proportion across samples (@kevorkian2018esti, @lloyd2020evid).
 Unfortunately, mathematical arguments and analysis of experiments with artificially constructed ('mock') communities demonstrate that this assumption is not always warranted: 
@@
 Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundance.
-Our analysis clarifies how the folk wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
+Our analysis clarifies how the received wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
 Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.

### 2022-01-14 solutions.Rmd
Minor text edits

@@
 
-More generally, changes in the mean efficiency of the total density measurement offset those in the sequencing measurement when comparing log density of species across samples.
+More generally, changes in the mean efficiency of the total-density measurement can offset those in the sequencing measurement when comparing log density of species across samples.
 If the (log) efficiencies of the total and sequencing measurement are positively correlated across species, then the (log) mean efficiencies will tend to be positively correlated across samples, leading to a reduction in error in DA analysis relative to total-density methods whose efficiencies are uncorrelated with those of the sequencing measurement.

### 2022-01-14 abundance-measurement.Rmd
Revise treatment of main theoretical results
This commit revises the language, notation, and explanation of the main
theoretical results about the effect of bias. Changes include
- using 'abundance' instead of 'density' throughout
- updating equations and text to use the new notation and language
  develeped in the appendix, in which the set S of all species used in
  the denominator of a proportion calculation is explicitly defined, and
  the mean efficiency is written as the efficiency of the multi-species
  set S.
- clarifying what is meant by 'absolute abudance'
- revising the abundance-me

--- 2021-11-16 abundance-measurement.Rmd
@@
 
-To understand how bias affects the measured differential abundance between samples, we first consider how it affects measurements of individual samples.
-
-## Model of MGS measurement
+We begin by considering how taxonomic bias affects the relative- and absolute-abundance measurements which serve as the input to microbiome DA analyses.
 
 Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
-This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
-We consider a set of microbial species that are measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes samples to assign reads to various microbial species.
-We assume that assigned reads have been correctly assigned to the given species and sample.
-Thus we ignore the possibility of contamination or false-positive taxonomic assignment, though we allow that many reads may not receive a taxonomic assignment and so ultimately be discarded.
-Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'noise' or random error due to the random sampling of sequencing reads and other aspects of the MGS process.
+This model is the simplest that respects the multiplicative nature of taxonomic bias and the *compositional* nature of MGS measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
+We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes to taxonomically assign reads to a set of microbial species $S$.
+Various forms of taxonomic assignment are possible; for simplicity, we suppose that reads are assigned to the species level, with reads that cannot be uniquely assigned being discarded.
+We ignore the possibility that reads are incorrectly assigned to the wrong species or sample.
+Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
 
-Our model stipulates that the assigned read count of species $i$ in sample $a$ equals its cell density multiplied by a species-specific factor and a sample-specific factor,
+In our model, the assigned read count of a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
 \begin{align}
@@
   \text{reads}_{i}(a)
-  = \text{density}_{i}(a) \quad \cdot
+  = \text{abun}_{i}(a) \quad \cdot
     \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
     \cdot \quad
-    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
+    \underbrace{\text{effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
 \end{align}
-The species-specific factor is the *relative measurement efficiency* (or *efficiency* for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (@mclaren2019cons).
-The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
-Equation \@ref(eq:measurement-model) implies that the total reads from all species in the sample equal
+The species-specific factor, $\text{efficiency}_{i}$, equals the *relative measurement efficiency* (or simply *efficiency*) of the species---how much more easily that species is measured (converted from cells to assigned reads) relative to a arbitrary fixed reference species (@mclaren2019cons).
+We assume the efficiencies of particular species are consistent across samples.
+The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
+The sample-specific factor, $\text{effort}(a)$, we call the *sequencing effort* for that sample; it captures the variation in the total number of assigned reads due to experimental features such as library normalization and total sequencing-run output.
+Equation \@ref(eq:measurement-model) implies that the total number of assigned reads in sample $a$ equals
 \begin{align}
   (\#eq:total-reads)
-  \text{total reads}(a)
-    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
+  \text{reads}_S(a)
+    = \text{abun}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
 \end{align}
-where 
+where $\text{abun}_{S}(a) \equiv \sum_{j\in S}\text{abun}_j(a)$ is the total abundance of species in $S$ and 
 \begin{align}
   (\#eq:mean-efficiency)
-  \text{mean efficiency}(a) 
-  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
+  \text{efficiency}_S(a) 
+    \equiv \frac{\sum_{j\in S}\text{abun}_j(a) \cdot \text{efficiency}_j}{\text{abun}_S(a)}
 \end{align}
-is the average or mean efficiency of cells in the sample.
+is the _mean efficiency_ over all species in $S$.
+<!-- The subscript $S$ serves as a reminder that many reads may not be assigned---for example, reads that fail to pass quality control or that come from the host or other taxonomic domains than what is being studied.
+-->
 
-## Relative abundances (proportions and ratios)
+## Relative abundance (proportions and ratios)
 
-We distinguish between two types of species-level *relative abundances*.
-The first type consists of the *ratios* among the abundances of two or more species, such as a ratio of 5:4:2 among three species.
-The second type consists of the *proportion* of an individual species, equal to its abundance divided by that of a particular root taxon (e.g. Prokaryotes).
-DA analyses of proportions consider how the proportion of a species changes indepently of other species, whereas DA analyses of ratios consider the conserted change in relative abundance among two or more species.
-Although proportion-based analysis is dominant, microbiome researchers increasingly use ratio-based analyses adapted from the statistical framework of Compositional Data Analysis (CoDA).
-The different manner in which taxonomic bias affects ratios vs.
-proportions (@mclaren2019cons) forms the basis in subsequent sections for understanding the extent to which its effects cancel in DA analysis.
+We distinguish between two types of species-level *relative abundances* within a sample.
+The *proportion* of species $i$ in sample $a$ equals to its abundance relative to the total abundance of all species in $S$,
+\begin{align}
+  (\#eq:prop)
+  \text{prop}_{i}(a) 
+    &\equiv \frac{\text{abun}_i(a)}{\text{abun}_S(a)}.
+%  \\&= \frac{\text{abun}_i(a)}{\sum_{i \in S}\text{abun}_i(a)}.
+\end{align}
+The *ratio* between two species $i$ and $j$ equals the abundance of $i$ relative to that of $j$,
+\begin{align}
+  (\#eq:ratio)
+  \text{ratio}_{i/j}(a) = \frac{\text{abun}_i(a)}{\text{abun}_j(a)}.
+\end{align}
+Proportions and ratios each form the basis for popular DA methods;
+ratio-based methods are commonly referred to as Compositional Data Analysis (CoDA) methods.
 
-The standard measurement of the proportion of species $i$ in sample $a$ is given by the ratio of its read count to the total,
+Taxonomic bias has a different effect on species proportions versus species ratios.
+The proportion of a species is typically measured by its proportion of assigned reads,
 \begin{align}
   (\#eq:prop-meas)
-  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
+  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
 \end{align}
-From Equations \@ref(eq:measurement-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas), it follows that
+We can rewrite the right-hand side (using Equations \@ref(eq:measurement-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas)) to find
 \begin{align}
@@
   \widehat{\text{prop}}_{i}(a)
-  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)},
+  &= \text{prop}_{i}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}}_{\substack{\text{variable} \\ \text{fold error}}}.
 \end{align}
-where $\text{prop}_{i}(a) = \text{density}_{i}(a) / \text{total density}(a)$ is the actual proportion.
-Equation \@ref(eq:prop-error) says that taxonomic bias creates a multiplicative error equal to the ratio of the species' efficiency to the mean efficiency in the sample.
-Consequently, the same species can be over- or under-measured depending on the composition of the given sample (Figure \@ref(fig:error-proportions)).
-For instance, in samples from two hypothetical communities in Figure \@ref(fig:error-proportions), Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).
+Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency relative to the mean efficiency in the sample.
+Consequently, the species's proportion is measured as too high in samples that are dominated by species with lower efficiencies, and measured as too low in samples that are dominated by species with higher efficiencies.
+This phenomenon is illustrated in two hypothetical communities in Figure \@ref(fig:error-proportions). 
+Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.
 
@@
   \widehat{\text{ratio}}_{i/j}(a)
-%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
-  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
+%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
+  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
 \end{align}
-The fold error in the ratio between two species simply equals the ratio in their efficiencies and is therefore constant across samples.
-For instance, in the example of Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
+Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
+For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
 
@@
 
-(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species.
+(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species.
 Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.
+Species' proportions may be measured as too high or too low depending on sample composition.
+For instance, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).
 
@@
 
-**Difference between species proportions and species ratios:**
-Consistent taxonomic bias at the species level creates a consistent fold error in the ratios among species, but a varying fold error in the proportions of individual species.
-But a proportion is, after all, just a ratio between two taxa in which the denominator taxon is the entire taxonomic domain under study.
-The difference in behavior from that of ratios between species arises because the efficiency of higher-level taxa, which consist of the aggregate of many species, varies across samples with the changing relative abundances among its constituent species.
+The fact that the multiplicative error in proportions varies while that in ratios is constant forms the basis for understanding the effect of bias on different methods for measuring absolute abundances from MGS and the extent to which bias cancels in different DA analyses.
 
-## Absolute abundances (cell densities)
+## Absolute abundance
 
-A wide range of experimental methods exist for converting the relative-abundance information in MGS measurements into measurements of (absolute) cell densities;
-however, these methods can be generally partitioned into two types, based on whether the absolute-abundance information derives from knowing the density of the total community or knowing the density of one or more individual species.
+Researchers often would like to know how the _absolute abundance_ of a species changes.
+In this context, _absolute_ means not relative to other species; typically, the abundance of interest is relative to non-microbial entities such the volume, mass, or amount of host DNA in the sample.
+A wide range of experimental methods have been used to convert the relative abundances from MGS measurements into measurements of absolute abundances.
+Yet there has so far been little consideration as to how the resulting measurements are affected by taxonomic bias in the MGS measurements or in the supplemental measurements, such as flow cytometry and qPCR, that they often involve.
+Here and in Appendix \@ref(review-absolute-methods) we describe how various methods for measuring absolute abundance are predicted to be affected by taxonomic bias, with an eye towards determining whether the measurement error is variable or constant.
 
-**Methods using the density of the total community:**
-The total microbial density of a community can be assayed by a variety of methods (reviewed in Appendix \@ref(review-absolute-methods)).
-Once the total density is known, the density of individual species can be estimated by multiplying the total density by the species' proportions (as measured by MGS),
+Our discussion supposes the 'abundance' of interest is cell number, though it applies to other quantities, such as biomass and genome copy number, that may be more relevant in different biological contexts.
+
+We consider two general classes of methods: Those in which absolute-abundance information is derived from measurement of the aggregate abundance of the total community, and those in which it drives from targeted measurement (or prior knowledge) of one or more particular species.
+
+**Normalization by total abundance:**
+Measurements of species absolute abundances are often obtained by making a (non-MGS) measurement of total community abundance, equating it with the aggregate abundance of the species $S$ measured by MGS, and multiplying this total by the proportions from MGS,
 \begin{align}
   (\#eq:density-prop-meas)
-  \widehat{\text{density}}_{i}(a) 
-  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{total density}}(a)
-\\&= \text{reads}_{i}(a) \cdot \frac{\widehat{\text{total density}}(a)}{\text{total reads}(a)}.
+  \widehat{\text{abun}}_{i}(a) 
+  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{abun}}_S(a).
 \end{align}
-We can rewrite the species' density measurements to account for the effect of taxonomic bias on the MGS proportions (Equation \@ref(eq:prop-error)) and any error in the total density measurement, 
+This measurement is affected by taxonomic bias in the MGS measurement as well as in the total-abundance measurement.
+For example, measurement of total community abundance by 16S qPCR is affected by variation among species in extraction efficiency, 16S copy number, and PCR binding and amplification bias.
+If the species-level efficiencies of the total-abundance measurement are constant across samples and we neglect other sources, then we can express the total-abundance measurement as
+\begin{align}
+  (\#eq:total-density-error)
+  \widehat{\text{abun}}_S(a) 
+  &= \sum_{i\in S} \text{abun}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
+\\&= \text{abun}_S(a) \cdot \text{efficiency}^{\text{tot}}_S(a),
+\end{align}
+where $\text{efficiency}_{i}^{\text{tot}}(a)$ is the absolute measurement efficiency of species $i$ for the total-abundance measurement and $\text{efficiency}^{\text{tot}}_S(a)$ is the mean efficiency of the total-abundance measurement in the sample.
+<!-- Note, we have assumed that the total-abundance and MGS measurements are sensitive to the same species -->
+The equations \@ref(eq:total-density-error) and \@ref(eq:prop-error) for the error in total-abundance and proportion measurements imply that the species abundance measurement \@ref(eq:density-prop-meas) has error given by
 \begin{align}
   (\#eq:density-prop-error)
-  \widehat{\text{density}}_{i}(a) 
-  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
-%  \cdot \frac{\widehat{\text{total density}}(a)}{\text{total density}(a)}.
-  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{total density}}(a) \end{array}.
+  \widehat{\text{abun}}_{i}(a) 
+  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}.
 \end{align}
-The variable fold error that bias creates in the measured proportions directly transfers to the density measurements.
 
-Taxonomic bias can also create systematic error in the total-density measurements (Appendix \@ref(appendix-total-density)).
-For now, we suppose that the total density can be measured accurately; 
-later, Section \@ref(solutions) describes how experiments might be designed such that taxonomic bias in the total density measurement offsets that in the MGS proportions, leading to a (more) consistent fold error in the measured densities.
+Equation \@ref(eq:density-prop-error) indicates that the multiplicative error in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
+As in the case of proportions (Equation \@ref()), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
+On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to a relatively stable error.
+We discuss how this possibility might apply to real experimental workflows below.
 
-**Methods using the density of one or more reference species:**
-The density of one or more species can be directly measured using targeted measurement methods like species-specific qPCR; alternatively, the densities of one or more species may be thought to be constant across samples either from background knowledge or because they have been spiked in at a fixed density (these situations are reviewed in Appendix \@ref(review-absolute-methods)).
-These _reference species_ with known or constant density can be used to convert read counts to densities for all species.
-For instance, given the density of a reference species $r$, we can estimate the density of a focal species $i$ by 
+**Normalization by one or more reference species:**
+Suppose we had a measurement of the (absolute) abundance of a _reference species_ $r$.
+In the absence of taxonomic bias, all species are expected to have the same ratio of reads to abundance in a sample (Equation \@ref(eq:measurement-model)).
+Thus the abundance-to-reads ratio for species $r$ can serve as a conversion factor allowing us to obtain the abundance of an arbitrary species $i$,
 \begin{align}
   (\#eq:density-ratio-meas)
-%  \widehat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \widehat{\text{density}}_{r}(a).
-  \widehat{\text{density}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{density}}_{r}(a)}{\text{reads}_{r}(a)}.
+  \widehat{\text{abun}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{abun}}_{r}(a)}{\text{reads}_{r}(a)}.
 \end{align}
-Alternatively, if we only know that the reference species is constant, we can set $\widehat{\text{density}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) equal to 1 to estimate the density in unknown but fixed units, which is sufficient for multiplicative DA analysis.
-(Appendix \@ref(model-details) considers possible extensions to multiple reference species.)
-It follows from Equation \@ref(eq:ratio-error) that the error caused by taxonomic bias in this density measurement is
+The abundance of one or more reference species can be directly measured using targeted measurement methods like species-specific qPCR and used with Equation 
+\@ref(eq:density-ratio-meas) to obtain absolute abundance for all species.
+To our knowledge, this approach has not previously been suggested.
+<!-- (Appendix \@ref(model-details) considers possible extensions to multiple reference species.) -->
+Instead, reference-based measurement has been used in the context of spike-in experiments or normalization to a species (such as the host) which is treated as having a constant abundance.
+In a spike-in experiment, the abundance of reference species are added in known (up to experimental error) abundances to each sample.
+When normalizing to a species that is treated as having a constant but unknown abundance, we set $\widehat{\text{abun}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) equal to 1 and interpret the abundances as having fixed but unknown units, which is sufficient for multiplicative DA analysis.
+
+The error in the abundance measurement \@ref(eq:density-ratio-meas) due to taxonomic bias in the MGS measurement is given by
 \begin{align}
   (\#eq:density-ratio-error)
-  \widehat{\text{density}}_{i}(a) 
-  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
-  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{density}_r}(a) \end{array}.
+  \widehat{\text{abun}}_{i}(a) 
+  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
+  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) \end{array}.
 \end{align}
-Here, it is the constant error in the measured ratio of species $i$ to $r$ that propagates into the density measurement.
-There will generally be systematic error in the density of the reference species; however, if the systematic fold error is constant across samples, so will be that of the densities of other species.
+The constant error in the measured ratio of species $i$ to $r$ (see Equation \@ref(eq:ratio-error)) propagates to the abundance measurement.
+There will generally also be systematic error in the abundance of the reference species; however, if the systematic fold error is constant across samples, so will be that of the abundances of other species.
 
-**Differences between the two approaches:**
-Equations \@ref(eq:density-prop-error) and \@ref(eq:density-ratio-error) show that the manner in which taxonomic bias in the sequencing measurement impacts the two approaches mirrors that of proportions and ratios.
-If the fold error in the total-density or reference species measurement is negligible (or at least constant) then the fold errors in species densities by the total-density approach vary across samples with the mean efficiency, whereas the fold errors by the reference-species approach are constant.
-The fundamental determinant of whether constant error is obtained is not which of these two equations is ultimately used, but rather whether densities of individual species, which (we assume) have constant efficiency, or the total community, which has varying efficiency, provide the absolute-abundance information.
-Studies using spike-ins as constant reference species will often, instead of using Equation \@ref(eq:density-ratio-meas), first use the spike-in species to estimate the total density, and then estimate the density of focal species with Equation \@ref(eq:density-prop-meas).
-If the calculation is done carefully, however, this method still yields constant fold errors so long as the species efficiencies (including those of the spike-in) are constant across samples (Appendix \@ref(appendix-total-density)).
+Spike-ins are instead sometimes used to measure absolute abundances using Equation \@ref(eq:density-prop-meas).
+In this case, an intermediate step is taken in which the total community abundance is first measured by the ratio of non-spike-in to spike-in reads.
+<!-- CHECK -->
+If done correctly, this calculation yields results that are identical to directly applying Equation \@ref(eq:density-ratio-meas) (Appendix \@ref(total-density-ref)).
+
+**Difference between the two approaches:**
+Reference-species normalization yields constant fold errors because it is based on species-level read counts and abundances and we assume that efficiencies are constant at the species level.
+In contrast, total-abundance normalization is based on aggregates of species (for the calculation of proportion and the total-abundance measurement) and so depends on mean efficiencies, which can vary across samples.

### 2022-01-14 case-studies.Rmd
Revise treatment of main theoretical results
This commit revises the language, notation, and explanation of the main
theoretical results about the effect of bias. Changes include
- using 'abundance' instead of 'density' throughout
- updating equations and text to use the new notation and language
  develeped in the appendix, in which the set S of all species used in
  the denominator of a proportion calculation is explicitly defined, and
  the mean efficiency is written as the efficiency of the multi-species
  set S.
- clarifying what is meant by 'absolute abudance'
- revising the abundance-me

@@
 These DNA mock communities allowed the authors to directly measure the bias due to the sequencing workflow following DNA extraction and correct for it when analyzing the primary experimental samples following the method of @mclaren2019cons.
-Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification (CITE) and the substantial ribosomal copy-number-variation (CNV) in fungi (@lofgren2019geno).
+Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification and the substantial ribosomal copy-number-variation (CNV) in fungi (@lofgren2019geno).
+<!-- TODO: add citations -->
 The authors found a 13X difference between the most and least efficiently measured commensal, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
@@
 In particular, it is interesting to consider whether any commensals are able to increase in absolute concentration in response to infection.
-The experiment performed by @leopold2020host does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration measured using the total-community normalization method (Equation \@ref(eq:density-prop-meas)).
+The experiment performed by @leopold2020host does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration measured using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
 It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
@@
 
-Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-community normalization method (Equation \@ref(eq:density-prop-meas)).
+Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
 We consider two scenarios.
@@
 Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where some taxa that persist or increase in abundance instead appear to decrease.
-_Scenario 2:_ Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-community normalization method.
+_Scenario 2:_ Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-abundance normalization method.
 In this case, we should expect the increase in mean efficiency to have little to no impact on LFCs, since it would be offset by the increase in mean efficiency of the total-concentration measurement.
 Moreover, since the total ITS concentration and ITS sequencing are both performed on the extracted DNA, we can also expect the effect of taxonomic bias caused by DNA sequencing to be offset.
-Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-community measurement is made prior to normalization.
+Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-abundance measurement is made prior to normalization.
 
@@
 @lloyd2020evid performed a systematic assessment of growth rates of bacterial and archaeal taxa over a depth of 10 cm (corresponding to ~40 years of burial time) in sediment of the White Oak River estuary.
-To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-community normalization method (Equation \@ref(eq:density-prop-meas)), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy (CARD-FISH).
+To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy (CARD-FISH).
 The authors refer to these absolute densities as FRAxC measurements, for 'fraction of 16S reads times total cell counts'.
@@
 Thus comparing qPCR to FRAxC growth rates on the same reference taxa allows us to estimate the systematic error in FRAxC growth rates.
-Because the systematic error in regression slopes due to bias is the same for each taxon (Equation \@ref(eq:regression-error)), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
+Because the systematic error in regression slopes due to bias is the same for each taxon (Section \@ref(differential-abundance)), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
 <!-- Our aim is to compare the growth rates for taxa with both FRAxC and qPCR values for the two sediment cores, as reported in Table 1. -->
@@
 First, the error caused by consistent taxonomic bias _can mostly cancel_ in cross-sample comparisons and so not impact DA analyses of fold changes in species proportions.
-Our theoretical results from Section @ref(differential-abundance) indicate that when the mean efficiency is roughly constant across samples, the error in proportions cancels in fold change calculations and is absorbed by the intercept term in regression models.
+Our theoretical results from Section \@ref(differential-abundance) indicate that when the mean efficiency is roughly constant across samples, the error in proportions cancels in fold change calculations and is absorbed by the intercept term in regression models.
 We observed this scenario in the analysis of pre-infection fungal microbiome samples of @leopold2020host and when analyzing the trajectories of the vaginal microbiomes of women when the dominant species remained constant^[This point could use more support in the MOMSPI case study], and a stable mean efficiency within marine soil cores is plausible and consistent with (though unable to be fully determined by) the different growth rate estimates in the @lloyd2020evid experiment.
-Section \@ref(differential-abundance) showed that absolute DA analysis using total-community normalization are suspectible to errors as with DA analysis of proportions; yet our analysis of two (hypothetical) approaches to analyzing absolute changes in response to infection in the fungal microbiome experiment indicates that the increasingly common approach of pairing marker-gene sequencing with qPCR of the total marker density can largely mitigate this effect, since here what matters is the variation in the ratio of mean efficiencies of the MGS measurement to the total-density measurement and this ratio may remain roughly constant if bias is largely shared by the two measurements.^[This is a simple-to-state and important point that should perhaps be made in the previous theoretical section.
-More empirical evidence could be given by re-analysis of @tettamantiboshier2020comp or similar studies that perform qPCR of total-community 16S and of specific species.
-Here I expect to find a fairly good agreement in the LFCs from total-community normalization and species-specific qPCR, since much of the bias is likely shared by 16S sequencing and qPCR.
-However, there are various complications to interpretting these comparisons.]
+Section \@ref(differential-abundance) showed that absolute DA analysis using total-abundance normalization are suspectible to errors as with DA analysis of proportions; yet our analysis of two (hypothetical) approaches to analyzing absolute changes in response to infection in the fungal microbiome experiment indicates that the increasingly common approach of pairing marker-gene sequencing with qPCR of the total marker density can largely mitigate this effect, since here what matters is the variation in the ratio of mean efficiencies of the MGS measurement to the total-density measurement and this ratio may remain roughly constant if bias is largely shared by the two measurements.
 

### 2022-01-14 differential-abundance.Rmd
Revise treatment of main theoretical results
This commit revises the language, notation, and explanation of the main
theoretical results about the effect of bias. Changes include
- using 'abundance' instead of 'density' throughout
- updating equations and text to use the new notation and language
  develeped in the appendix, in which the set S of all species used in
  the denominator of a proportion calculation is explicitly defined, and
  the mean efficiency is written as the efficiency of the multi-species
  set S.
- clarifying what is meant by 'absolute abudance'
- revising the abundance-me

@@
-# How taxonomic bias affects differential-abundance analysis {#differential-abundance}
+# Taxonomic bias can cause errors in differential-abundance results {#differential-abundance}
 
-How do the measurement errors described in the previous section impact our ability to estimate the changes in microbial abundances across samples or between different host and environmental conditions?
-Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, which have direct ecological interpretations (via the processes of exponential growth and death) and are ubiquitous in microbiome DA analysis.
+How do the errors caused by taxonomic bias in the abundances measured for individual samples impact DA analysis?
+<!-- the changes in abundance that we estimate between samples or different host and environmental conditions? -->
+Although there are many ways to quantify the changes in abundance that form the basis of a DA analysis, we focus on DA analyses of the (log) fold changes in proportions, ratios, and absolute abundance; these multiplicative difference measures are common (though not ubiquitous) in extant DA analyses and have more direct ecological interpretations (via the processes of exponential growth and decay) than non-multiplicative measures.
 
-## Change between a pair of samples
+## Fold changes between a pair of samples
 
-Before considering the stereotypical many-sample DA analysis, it is instructive to 
-consider simplest analysis of differential abundance: the measurement of fold changes in abundance between a pair of samples.
-This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment, and conceptually bridges the single-sample results of the previous section to the many-sample case.
+The building blocks of a DA analysis are the fold changes (FCs) in abundance between individual pairs of samples.
 
-<!-- P: error in FC proportions -->
-The composition-dependent effect of bias on fold error in proportions leads to error in fold-change measurements that is proportional to the inverse change in mean efficiency.
-From the error in an individual sample (Equation \@ref(eq:prop-error)), it follows that the measured fold change in proportion of taxon $i$ from sample $a$ to sample $b$ is
+<!-- P: Proportions and ratios -->
+The impact of bias on the measured FCs in species proportions and ratios follows directly from the results of Section \@ref(abundance-measurement) for the error in individual-sample measurements.
+From Equation \@ref(eq:prop-error), it follows that the measured FC in the proportion of species $i$ from sample $a$ to sample $b$ is
 \begin{align}
@@
   &= \frac
-    {\text{prop}_{i}(b) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(b)}}
-    {\text{prop}_{i}(a) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(a)}}
+    {\text{prop}_{i}(b) \cdot \cancel{\text{efficiency}_{i}} / {\text{efficiency}_S(b)}}
+    {\text{prop}_{i}(a) \cdot \cancel{\text{efficiency}_{i}} / {\text{efficiency}_S(a)}}
 \\[0.5ex]
@@
   \cdot
-  \underbrace{\left[\frac{\text{mean efficiency}(b)}{\text{mean efficiency}(a)}\right]^{-1}}_\text{fold error}
+  \underbrace{\left[\frac{\text{efficiency}_S(b)}{\text{efficiency}_S(a)}\right]^{-1}}_\text{fold error}
   .
 \end{align}
-The sample-independent efficiency factor of the error cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency from $s$ to $t$.
+The sample-independent efficiency factor cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency.
+In contrast, when we use Equation \@ref(eq:ratio-error) to compute the error in the FC in the ratio between two species $i$ and $j$, we find that the constant error $\text{efficiency}_{i} / $\text{efficiency}_{j}$ exactly cancels, so that the measured FCs remain accurate regardless of whether the mean efficiency varies.
 
-The bottom row of Figure \@ref(fig:error-proportions) illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
-The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
-Consequently, the FC of each species is measured to be 2.6X larger than the true value.
-Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
+Figure \@ref(fig:error-proportions) (bottom row) illustrates these two different behaviors for the error in the FCs of proportions versus ratios when the mean efficiency varies between samples.
+Here the mean efficiency decreases by a factor of 2.6 (FC of 0.4X) from Sample 1 to Sample 2, which causes the FC of the proportion of each species to be measured as 2.6X larger than its true value.
+Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction; we refer to the latter as a _sign error_ since in reference to the change in sign in the corresponding log fold change (LFC).
 We can see each type of error in Figure \@ref(fig:error-proportions).
@@
 For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
-For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!
+For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase (a sign error).
+In contrast, the fold error in Equation \@ref(eq:ratio-error) completely cancels when we divide the ratio measured for one sample $a$ by another sample $b$.
+<!-- **TODO: Note that errors can still arise for higher-level taxa due to variation in efficiency within the taxon.** -->
 
-In contrast, because species ratios are distorted by a constant factor, their measured fold changes remain accurate.
-The fold error in Equation \@ref(eq:ratio-error) completely cancels when we divide the ratio measured for one sample $a$ by another sample $b$.
-
-If other error sources remain negligible (or constant across samples), then this dichotomy continues to apply to proportion- and ratio-based density measurements.
-The inferred FCs in proportion-based density measurements will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
-Moreover, any species with a constant density will appear to vary inversely with the mean efficiency.
-In contrast, fold changes in ratio-based density measurements will remain accurate.
+These differences in the FCs of proportions versus ratios are mirrored when we compare FCs in absolute abundance made with normalization to a total-abundance measurement versus a reference-species measurement.
+FCs computed from total-abundance normalization are subject to error when the ratio of the two mean efficiencies (that for the MGS measurement and for the total-abundance measurement vary across samples).
+In contrast, FCs computed from reference-species normalization are not subject to error, so long as the (fold) error in the assumed or measured abundance of the reference species is constant.
 
@@
 
-DA analysis of many samples across host or environmental conditions can typically be framed as a regression problem, in which we analyze the relationship between a microbial _response variable_, such as log density of some focal species $i$, and one or more _covariates_, such the pH or temperature of the sampled environment or whether the sample is from a healthy or sick person.
-A large fraction of DA analyses use elaborations on the simple linear regression model, which for a response of log species density can be written
+<!-- TODO: Write a transition. -->
+<!-- TODO: consider doing log proportion for simplicity, and using the vaginal mb example for the figure -->
+
+DA analysis of multiple samples can be framed as a regression problem in which we analyze the relationship between a microbial _response variable_, such as the log abundance of species $i$, with one or more _covariates_, such as the pH of the sampled environment or whether the sample is from a healthy or sick person.
+The simplest regression analysis uses the simple linear regression model,
 \begin{align}
   (\#eq:regression)
-  \log \text{density}_i(a) = \alpha + \beta x(a) + \varepsilon_i(a).
+  \log \widehat{\text{abun}}_i(a) = \alpha_i + \beta_i x(a) + \varepsilon_i(a),
 \end{align}
-Here $x$ is a continuous covariate (e.g.
+where $x$ is a continuous covariate (e.g.
 pH) or a binary covariate (e.g.
-$x=1$ for treated patients and $x=0$ for controls), $\alpha$ and $\beta$ are regression coefficients, and $\varepsilon_i(a)$ is a mean-zero random variable that reflects the residual (unexplained) variation in the response (log density of species $i$).
-Our interest is usually in the coefficient $\beta$ (slope or average difference between conditions) that describes how the species' abundance changes with $x$, while the intercept $\alpha$ captures differences in the baseline abundance and---we hope---measurement efficiency among species.
-How does taxonomic bias under our measurement model affect estimates of $\beta$ in the simple linear regression?
+$x=1$ for treated patients and $x=0$ for controls), $\alpha_i$ and $\beta_i$ are regression coefficients, and $\varepsilon_i(a)$ is a mean-zero random variable that reflects the residual (unexplained) variation in the response.
+In a DA analysis, we are interested in the slope coefficient $\beta$, which describes how the species' abundance changes with $x$.
+We are generally uninterested in the intercept coefficient, $\alpha$, which captures differences in the baseline abundance among species and---it is usually hoped---study- and species-specific systematic error such as that created by taxonomic bias.
 
-Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \@ref(eq:density-prop-meas)) with error-free estimates of the total density.
-If the true log density follows the regression Equation \@ref(eq:regression), then it follows from Equation \@ref(eq:density-prop-error) the estimated log density equals 
+How does taxonomic bias in fact impact estimates of the slope coefficient, $\beta$?
+Appendix \@ref(appendix-regression) derives the effect of bias on the estimated coefficients for the simple linear regression model under estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE).
+The result has a simple interpretation that provides intuition for what to expect in more complicated generalized linear regression models used in popular DA methods. 
+Consider the regression \@ref(eq:regression) fit to the measured log abundance of species $i$ across all samples.
+The measurement error in the response variable,
 \begin{align}
-  (\#eq:regression-error)
-  \log \widehat{\text{density}}_i(a)
-%  &= \alpha + \beta x(a) + \varepsilon_i(a) + \log \text{efficiency}_i - \log \text{mean efficiency}(a)
-  = [\alpha + \log \text{efficiency}_i] + [\beta - \log \text{mean efficiency}(a)] x(a) + \varepsilon_i(a).
+  \text{error}_{i}(a) \equiv 
+  \log \widehat{\text{abun}}_i(a) - \log \text{abun}_i(a)
+  = \log \frac{\widehat{\text{abun}}_i(a)}{\text{abun}_i(a)},
 \end{align}
-This equation shows that the species-specific portion of the error affects the intercept term while the sample-specific portion (log mean efficiency) affects the slope term.
-Thus, as in the case of fold changes between two samples (Equation \@ref(eq:prop-fc-error)), it is the variation in the (log) mean efficiency across samples that we must worry about distorting our DA results.
+can be thought of as having three components:
 
-The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$.
-Appendix \@ref(appendix-regression) mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \@ref(fig:regression-example) illustrates using simulated data.
+- Error that is constant across samples affects the intercept coefficient, $\alpha_i$.
+- Error that varies systematically with $x$ affects the slope coefficient, $\beta_i$.
+- Error that varies and is uncorrelated with $x$ affects the residual variation, $\varepsilon_{i}(s)$.
+
+We saw in Section \@ref(abundance-measurement) that the error from taxonomic bias has a species-specific, sample-independent component that is constant across samples; this component causes a species-specific systematic error in the estimated intercept, $\alpha_{i}$.
+The error also has a sample-specific, species-independent component. 
+The portion that is correlated with the covariate $x$ creates a systematic error in the slope coefficient $\beta_{i}$ that is the same for every species and opposite in sign to how the error varies with $x$;
+the portion that is uncorrelated with $x$ affects the precision (standard errors) of the slope estimates in a species-specific manor.
+
+<!-- The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$. -->
+
+Figure \@ref(fig:regression-example) illustrates with a simulated data for the case of log abundance measured using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)), assuming that total community abundance is measured perfectly accurately.
+In this case, the variable component of the error is due entirely to variation in the log mean efficiency of the MGS measurement.
 Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
-The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with $x$ relative to that of the response (here, $\log \text{density}_{i}$) with $x$ or, equivalently, the relative magnitudes of their slopes.
+The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with $x$ relative to that of the response (here, $\log \text{abun}_{i}$) with $x$ or, equivalently, the relative magnitudes of their slopes.
 As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
@@
 
-With this understanding in place, we briefly summarize how taxonomic bias affects estimation of the simple linear model in other abundance types.
-The results for proportion-based density estimates with accurate total densities also apply to LFC analysis of proportions.
-Similar results apply to microbiome regression tools (such as corncob; @martin2020mode) that perform regression on the logit (instead of log) proportion of a species; however, 
-the mean efficiency of the entire sample must instead be replaced with the mean efficiency among all species excluding the focal species, which causes the absolute error in regression coefficients to vary somewhat across species.
-Because ratios and ratio-based densities are subject to consistent multiplicative error, in analysis of log ratios and the log densities derived from them, only the estimated intercept $\hat \alpha$ is affected by taxonomic bias, while the point estimate and standard error of the estimated slope $\hat \beta$ remain unaffected.
+Overall, we find that variation in measurement error created by variation in the mean efficiency of the MGS measurement can negatively affect certain DA analyses by creating systematic errors in the slope coefficient estimates and/or reducing their precision.
+Similar effects occur for regressions of log proportions and of log abundances derived by total-abundance normalization when the total abundance is accurately measured; however, as noted in Section \@ref(abundance-measurement), total-abundance measurements that are taxonomically biased in a complementary manner can mitigate these effects.
+Regression analyses of log-ratios and log abundances derived from reference-species normalization remain unaffected---the errors are constant across samples and so only impact the intercept coefficients.
+
+<!-- Similar results apply to microbiome regression tools (such as corncob; @martin2020mode) that perform regression on the logit (instead of log) proportion of a species;  -->
+<!-- however, the mean efficiency of the entire sample must instead be replaced with the mean efficiency among all species excluding the focal species, which causes the absolute error in regression coefficients to vary somewhat across species.
+-->
 

### 2022-01-14 solutions.Rmd
Revise treatment of main theoretical results
This commit revises the language, notation, and explanation of the main
theoretical results about the effect of bias. Changes include
- using 'abundance' instead of 'density' throughout
- updating equations and text to use the new notation and language
  develeped in the appendix, in which the set S of all species used in
  the denominator of a proportion calculation is explicitly defined, and
  the mean efficiency is written as the efficiency of the multi-species
  set S.
- clarifying what is meant by 'absolute abudance'
- revising the abundance-me

@@
 
-## Analyses of fold changes in species ratios
+## Use ratio-based relative-abundance analyses
 
-Since the fold changes in species ratios are unaffected by consistent taxonomic bias, a natural approach to countering bias in relative DA analysis is to ratio- rather than proportion-based methods.
+Since bias creates a consistent fold error in species ratios, a natural approach to countering bias when analyzing relative abundances is to use DA methods that are based on fold changes in ratios.
 A variety of such methods have been developed for microbiome DA analysis that build on earlier work from the field of Compositional Data Analysis (CoDA).
@@
 To demonstrate the potential for calibration to improve fold-change measurements, we estimated bias from one sample in the mock community data from @brooks2015thet that contained all 7 species and used it to calibrate the compositions of all samples (Figure \@ref(fig:calibration-example)).
-We then estimated species' densities using the uncalibrated and calibrated proportions using the community normalization method, treating the total density as known and fixed.
+We then estimated species' densities using the uncalibrated and calibrated proportions using total-abundance normalization, treating the total density as known and fixed.
 Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions and densities (Figure \@ref(fig:calibration-example)).

### 2022-01-16 abundance-measurement.Rmd
Add support for Word export
The following changes now allow the document to be successfully exported
to Word (.docx) format. Several problems remain in the output document;
many equations don't render properly and equation and figure numbering
and referencing do not work correctly.

- Update method of including images to be consistent in all figure chunks
- Adjust the `include_svg()` to have an else clause to handle Word export
- Add 'word' option to Makefile

@@
 ```{r error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
-svg_path <- "figures/illustrations/error-proportions.svg"
-if (knitr::is_html_output()) {
-  fig_path <- svg_path
-} else if (knitr::is_latex_output()) {
-  fig_path <- svg_to_pdf(svg_path)
-}
-knitr::include_graphics(fig_path)
-rm(svg_path, fig_path)
+"figures/illustrations/error-proportions.svg" %>%
+  include_svg
 ```

### 2022-01-16 case-studies.Rmd
Add support for Word export
The following changes now allow the document to be successfully exported
to Word (.docx) format. Several problems remain in the output document;
many equations don't render properly and equation and figure numbering
and referencing do not work correctly.

- Update method of including images to be consistent in all figure chunks
- Adjust the `include_svg()` to have an else clause to handle Word export
- Add 'word' option to Makefile

@@
 ```{r momspi-mean-efficiency-dist, fig.cap = '(ref:cap-momspi-mean-efficiency-dist)'}
-svg_path <- fs::path(
+fs::path(
   "notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files", 
   "figure-html5/momspi-mean-efficiency-dist-1.svg"
-)
-if (knitr::is_html_output()) {
-  fig_path <- svg_path
-} else if (knitr::is_latex_output()) {
-  fig_path <- svg_to_pdf(svg_path)
-}
-knitr::include_graphics(fig_path)
-rm(svg_path, fig_path)
+  ) %>%
+  include_svg
 ```

### 2022-01-16 differential-abundance.Rmd
Add support for Word export
The following changes now allow the document to be successfully exported
to Word (.docx) format. Several problems remain in the output document;
many equations don't render properly and equation and figure numbering
and referencing do not work correctly.

- Update method of including images to be consistent in all figure chunks
- Adjust the `include_svg()` to have an else clause to handle Word export
- Add 'word' option to Makefile

@@
 ```{r regression-example, fig.cap = '(ref:regression-example)'}
-svg_path <- here::here(
+here::here(
   "notebook/_posts/2021-08-03-simulate-regression-example/",
@@
   "main-figure-1.svg"
-)
-if (knitr::is_html_output()) {
-  fig_path <- svg_path
-} else if (knitr::is_latex_output()) {
-  fig_path <- svg_to_pdf(svg_path)
-}
-knitr::include_graphics(fig_path)
-rm(svg_path, fig_path)
+  ) %>%
+  include_svg
 ```

### 2022-01-16 introduction.Rmd
Add support for Word export
The following changes now allow the document to be successfully exported
to Word (.docx) format. Several problems remain in the output document;
many equations don't render properly and equation and figure numbering
and referencing do not work correctly.

- Update method of including images to be consistent in all figure chunks
- Adjust the `include_svg()` to have an else clause to handle Word export
- Add 'word' option to Makefile

@@
 # NOTE: call to include_graphics() has to be in top level expression
-svg_path <- fs::path(
+fs::path(
   "notebook/_posts/2021-10-27-illustrate-the-problem/illustrate-the-problem_files", 
   "figure-html5/brooks_lcrisp_prop_and_fc-1.svg"
-)
-if (knitr::is_html_output()) {
-  fig_path <- svg_path
-} else if (knitr::is_latex_output()) {
-  fig_path <- svg_to_pdf(svg_path)
-}
-knitr::include_graphics(fig_path)
-rm(svg_path, fig_path)
+  ) %>%
+  include_svg
 ```

### 2022-01-16 solutions.Rmd
Add support for Word export
The following changes now allow the document to be successfully exported
to Word (.docx) format. Several problems remain in the output document;
many equations don't render properly and equation and figure numbering
and referencing do not work correctly.

- Update method of including images to be consistent in all figure chunks
- Adjust the `include_svg()` to have an else clause to handle Word export
- Add 'word' option to Makefile

@@
 ```{r calibration-example, fig.cap = '(ref:cap-calibration-example)', out.width = '100%'}
-fig_path <- fs::path(
+fs::path(
   "notebook/_posts/2021-10-25-brooks2015thet-calibration",
   "brooks2015thet-calibration_files/figure-html5/brooks2015thet_fc_calibration-1.png"
-)
-knitr::include_graphics(fig_path)
-rm(fig_path)
+) %>%
+  knitr::include_graphics()
 ```

### 2022-01-16 _common.R

@@ -46,6 +46,9 @@ include_svg <- function(svg_path) {
     fig_path <- svg_path
   } else if (knitr::is_latex_output()) {
     fig_path <- svg_to_pdf(svg_path)
+  } else {
+    # For word docs
+    fig_path <- svg_path
   }
   knitr::include_graphics(fig_path)
 }

### 2022-01-16 abundance-measurement.Rmd
Fix typos

@@
 Equation \@ref(eq:density-prop-error) indicates that the multiplicative error in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
-As in the case of proportions (Equation \@ref()), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
+As in the case of proportions (Equation \@ref(eq:prop-error)), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
 On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to a relatively stable error.

### 2022-01-16 case-studies.Rmd
Fix typos

@@
 We observed this scenario in the analysis of pre-infection fungal microbiome samples of @leopold2020host and when analyzing the trajectories of the vaginal microbiomes of women when the dominant species remained constant^[This point could use more support in the MOMSPI case study], and a stable mean efficiency within marine soil cores is plausible and consistent with (though unable to be fully determined by) the different growth rate estimates in the @lloyd2020evid experiment.
-Section \@ref(differential-abundance) showed that absolute DA analysis using total-abundance normalization are suspectible to errors as with DA analysis of proportions; yet our analysis of two (hypothetical) approaches to analyzing absolute changes in response to infection in the fungal microbiome experiment indicates that the increasingly common approach of pairing marker-gene sequencing with qPCR of the total marker density can largely mitigate this effect, since here what matters is the variation in the ratio of mean efficiencies of the MGS measurement to the total-density measurement and this ratio may remain roughly constant if bias is largely shared by the two measurements.
+Section \@ref(differential-abundance) showed that absolute DA analysis using total-abundance normalization are susceptible to errors as with DA analysis of proportions; yet our analysis of two (hypothetical) approaches to analyzing absolute changes in response to infection in the fungal microbiome experiment indicates that the increasingly common approach of pairing marker-gene sequencing with qPCR of the total marker density can largely mitigate this effect, since here what matters is the variation in the ratio of mean efficiencies of the MGS measurement to the total-density measurement and this ratio may remain roughly constant if bias is largely shared by the two measurements.
 
@@
 In these two examples, systematic variation of the mean efficiency arose as high-efficiency species tended to dominate samples in one of the sample conditions (post-infection foliar samples or low-diversity vaginal samples).
-Thus one way for signficant error in DA regression results to arise is when a small number of (or just one) species that have particularly high or low mean efficiencies, form a large proportion of the community in a substantial fraction of samples, and are associated with the covariate of interest.
+Thus one way for significant error in DA regression results to arise is when a small number of (or just one) species that have particularly high or low mean efficiencies, form a large proportion of the community in a substantial fraction of samples, and are associated with the covariate of interest.
 <!-- This dynamic was responsible for the errors observed in the DA analysis of pre- and post-infection fungal microbiomes from @leopold2020host and in the vaginal microbiome profiles of the MOMS-PI study.
@@
 We described a hypothesized scenario in the marine-sediment case study in which increased burial time (the covariate) selects for lysis-resistant (and so lower efficiency) species.
-As another example, treatment with a certain antibiotic might selectively kill easier-to-lyze Gram negative species (or harder-to-lyze Gram positives) and thereby decreasing (or increasing) the mean efficiency in samples collected from a host post-treatment.
-Such situations can arise whenever there is a microbial trait that is associated with both measurement efficiency and the biological processes of intersest.
+As another example, treatment with a certain antibiotic might selectively kill easier-to-lyze Gram negative species (or harder-to-lyse Gram positives) and thereby decreasing (or increasing) the mean efficiency in samples collected from a host post-treatment.
+Such situations can arise whenever there is a microbial trait that is associated with both measurement efficiency and the biological processes of interest.
 An example besides cell-wall structure is ribosomal copy number, which increases a species' efficiency in ribosomal amplicon measurements and is positively associated with metabolic rate.

### 2022-01-16 differential-abundance.Rmd
Fix typos

@@
 The sample-independent efficiency factor cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency.
-In contrast, when we use Equation \@ref(eq:ratio-error) to compute the error in the FC in the ratio between two species $i$ and $j$, we find that the constant error $\text{efficiency}_{i} / $\text{efficiency}_{j}$ exactly cancels, so that the measured FCs remain accurate regardless of whether the mean efficiency varies.
+In contrast, when we use Equation \@ref(eq:ratio-error) to compute the error in the FC in the ratio between two species $i$ and $j$, we find that the constant error $\text{efficiency}_{i} / \text{efficiency}_{j}$ exactly cancels, so that the measured FCs remain accurate regardless of whether the mean efficiency varies.
 

### 2022-01-16 introduction.Rmd
Fix typos

@@
 But new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density.
-In its various forms, DA analysis remain one of, if not the most, common forms of analyses applied to MGS data to elucide the inner workings of microbiomes and their relationships to host and environmental health.
+In its various forms, DA analysis remain one of, if not the most, common forms of analyses applied to MGS data to elucidate the inner workings of microbiomes and their relationships to host and environmental health.
 

### 2022-01-16 conclusion.Rmd
Add placeholder for conclusion section

# Conclusion

_This section is a placeholder for a short (2-3 paragraph) conclusion. We are planning to keep 'discussion' to the earlier sections and appendix, at least for initial submission._

### 2022-01-16 v/414f6e78cc9b29c8236c8c3e996b68cd4d733b86/_main.tex
Add placeholder for conclusion section
This commit adds the rendered manuscript associated with
414f6e78cc9b29c8236c8c3e996b68cd4d733b86

--- 2022-01-11 v/3019b9c9c1fa3b13645d60ebb08913530bda476a/_main.tex
@@
 Callahan\footnote{North Carolina State University}}
-\date{2022-01-11}
+\date{2022-01-16}
 
@@
 
-\emph{This manuscript was rendered from commit 3019b9c9c1fa3b13645d60ebb08913530bda476a.}
+\emph{This manuscript was rendered from commit 414f6e78cc9b29c8236c8c3e996b68cd4d733b86.}
 
@@
 In these \emph{differential-abundance (DA) analyses}, the change in abundance of a microbial taxon across samples or conditions is used to infer ecological dynamics or find microbes that are associated with specific host diseases or environmental conditions.
-Although standard MGS measurements lose information about total microbial density---and so are typically used to analyze the abundances of taxa relative to each other---new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density.
-In its various forms, DA analysis remain one of, if not the most, common forms of analyses applied to MGS data to elucide the inner workings of microbiomes and their relationships to host and environmental health.
+Standard MGS measurements lose information about total microbial density and so are typically used to analyze the abundances of taxa relative to each other.
+But new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density.
+In its various forms, DA analysis remain one of, if not the most, common forms of analyses applied to MGS data to elucidate the inner workings of microbiomes and their relationships to host and environmental health.
 
@@
 As a result of taxonomic bias, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (Yeh et al.
-(\protect\hyperlink{ref-yeh2018taxo}{2018}),McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+(\protect\hyperlink{ref-yeh2018taxo}{2018}), McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
 These errors have been found in some cases to supersede sizable biological differences (e.g.
@@
 For example, if taxonomic bias were to cause the measured proportion of a species to consistently be 10X too high, we would still be able to accurately infer the fold change in its proportion across samples (Kevorkian et al.
-(\protect\hyperlink{ref-kevorkian2018esti}{2018}),Lloyd et al.
+(\protect\hyperlink{ref-kevorkian2018esti}{2018}), Lloyd et al.
 (\protect\hyperlink{ref-lloyd2020evid}{2020})).
@@
 
-To understand how bias affects the measured differential abundance between samples, we first consider how it affects measurements of individual samples.
-
-\hypertarget{model-of-mgs-measurement}{%
-\subsection{Model of MGS measurement}\label{model-of-mgs-measurement}}
+We begin by considering how taxonomic bias affects the relative- and absolute-abundance measurements which serve as the input to microbiome DA analyses.
 
 Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
-This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the \emph{compositional} nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (Gloor et al.
+This model is the simplest that respects the multiplicative nature of taxonomic bias and the \emph{compositional} nature of MGS measurements, in which the total read count for a sample is unrelated to its total cell number or density (Gloor et al.
 (\protect\hyperlink{ref-gloor2017micr}{2017})).
-We consider a set of microbial species that are measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes samples to assign reads to various microbial species.
-We assume that assigned reads have been correctly assigned to the given species and sample.
-Thus we ignore the possibility of contamination or false-positive taxonomic assignment, though we allow that many reads may not receive a taxonomic assignment and so ultimately be discarded.
-Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the `noise' or random error due to the random sampling of sequencing reads and other aspects of the MGS process.
-
-Our model stipulates that the assigned read count of species \(i\) in sample \(a\) equals its cell density multiplied by a species-specific factor and a sample-specific factor,
+We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes to taxonomically assign reads to a set of microbial species \(S\).
+Various forms of taxonomic assignment are possible; for simplicity, we suppose that reads are assigned to the species level, with reads that cannot be uniquely assigned being discarded.
+We ignore the possibility that reads are incorrectly assigned to the wrong species or sample.
+Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the `random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
+
+In our model, the assigned read count of a species \(i\) in a sample \(a\) equals its abundance multiplied by species-specific and sample-specific factors,
 \begin{align}
@@
   \text{reads}_{i}(a)
-  = \text{density}_{i}(a) \quad \cdot
+  = \text{abun}_{i}(a) \quad \cdot
     \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
     \cdot \quad
-    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
-\end{align}
-The species-specific factor is the \emph{relative measurement efficiency} (or \emph{efficiency} for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-The sample-specific factor, which we call the \emph{sequencing effort} of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
-Equation \eqref{eq:measurement-model} implies that the total reads from all species in the sample equal
+    \underbrace{\text{effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
+\end{align}
+The species-specific factor, \(\text{efficiency}_{i}\), equals the \emph{relative measurement efficiency} (or simply \emph{efficiency}) of the species---how much more easily that species is measured (converted from cells to assigned reads) relative to a arbitrary fixed reference species (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+We assume the efficiencies of particular species are consistent across samples.
+The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
+The sample-specific factor, \(\text{effort}(a)\), we call the \emph{sequencing effort} for that sample; it captures the variation in the total number of assigned reads due to experimental features such as library normalization and total sequencing-run output.
+Equation \eqref{eq:measurement-model} implies that the total number of assigned reads in sample \(a\) equals
 \begin{align}
   \label{eq:total-reads}
-  \text{total reads}(a)
-    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
-\end{align}
-where
+  \text{reads}_S(a)
+    = \text{abun}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
+\end{align}
+where \(\text{abun}_{S}(a) \equiv \sum_{j\in S}\text{abun}_j(a)\) is the total abundance of species in \(S\) and
 \begin{align}
   \label{eq:mean-efficiency}
-  \text{mean efficiency}(a) 
-  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
-\end{align}
-is the average or mean efficiency of cells in the sample.
-
-\hypertarget{relative-abundances-proportions-and-ratios}{%
-\subsection{Relative abundances (proportions and ratios)}\label{relative-abundances-proportions-and-ratios}}
-
-We distinguish between two types of species-level \emph{relative abundances}.
-The first type consists of the \emph{ratios} among the abundances of two or more species, such as a ratio of 5:4:2 among three species.
-The second type consists of the \emph{proportion} of an individual species, equal to its abundance divided by that of a particular root taxon (e.g.~Prokaryotes).
-DA analyses of proportions consider how the proportion of a species changes indepently of other species, whereas DA analyses of ratios consider the conserted change in relative abundance among two or more species.
-Although proportion-based analysis is dominant, microbiome researchers increasingly use ratio-based analyses adapted from the statistical framework of Compositional Data Analysis (CoDA).
-The different manner in which taxonomic bias affects ratios vs.~proportions (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})) forms the basis in subsequent sections for understanding the extent to which its effects cancel in DA analysis.
-
-The standard measurement of the proportion of species \(i\) in sample \(a\) is given by the ratio of its read count to the total,
+  \text{efficiency}_S(a) 
+    \equiv \frac{\sum_{j\in S}\text{abun}_j(a) \cdot \text{efficiency}_j}{\text{abun}_S(a)}
+\end{align}
+is the \emph{mean efficiency} over all species in \(S\).
+
+\hypertarget{relative-abundance-proportions-and-ratios}{%
+\subsection{Relative abundance (proportions and ratios)}\label{relative-abundance-proportions-and-ratios}}
+
+We distinguish between two types of species-level \emph{relative abundances} within a sample.
+The \emph{proportion} of species \(i\) in sample \(a\) equals to its abundance relative to the total abundance of all species in \(S\),
+\begin{align}
+  \label{eq:prop}
+  \text{prop}_{i}(a) 
+    &\equiv \frac{\text{abun}_i(a)}{\text{abun}_S(a)}.
+%  \\&= \frac{\text{abun}_i(a)}{\sum_{i \in S}\text{abun}_i(a)}.
+\end{align}
+The \emph{ratio} between two species \(i\) and \(j\) equals the abundance of \(i\) relative to that of \(j\),
+\begin{align}
+  \label{eq:ratio}
+  \text{ratio}_{i/j}(a) = \frac{\text{abun}_i(a)}{\text{abun}_j(a)}.
+\end{align}
+Proportions and ratios each form the basis for popular DA methods;
+ratio-based methods are commonly referred to as Compositional Data Analysis (CoDA) methods.
+
+Taxonomic bias has a different effect on species proportions versus species ratios.
+The proportion of a species is typically measured by its proportion of assigned reads,
 \begin{align}
   \label{eq:prop-meas}
-  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads}(a)}.
-\end{align}
-From Equations \eqref{eq:measurement-model}, \eqref{eq:total-reads}, and \eqref{eq:prop-meas}, it follows that
+  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
+\end{align}
+We can rewrite the right-hand side (using Equations \eqref{eq:measurement-model}, \eqref{eq:total-reads}, and \eqref{eq:prop-meas}) to find
 \begin{align}
@@
   \widehat{\text{prop}}_{i}(a)
-  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)},
-\end{align}
-where \(\text{prop}_{i}(a) = \text{density}_{i}(a) / \text{total density}(a)\) is the actual proportion.
-Equation \eqref{eq:prop-error} says that taxonomic bias creates a multiplicative error equal to the ratio of the species' efficiency to the mean efficiency in the sample.
-Consequently, the same species can be over- or under-measured depending on the composition of the given sample (Figure \ref{fig:error-proportions}).
-For instance, in samples from two hypothetical communities in Figure \ref{fig:error-proportions}, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).
+  &= \text{prop}_{i}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}}_{\substack{\text{variable} \\ \text{fold error}}}.
+\end{align}
+Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency relative to the mean efficiency in the sample.
+Consequently, the species's proportion is measured as too high in samples that are dominated by species with lower efficiencies, and measured as too low in samples that are dominated by species with higher efficiencies.
+This phenomenon is illustrated in two hypothetical communities in Figure \ref{fig:error-proportions}.
+Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.
 
@@
   \widehat{\text{ratio}}_{i/j}(a)
-%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
-  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
-\end{align}
-The fold error in the ratio between two species simply equals the ratio in their efficiencies and is therefore constant across samples.
-For instance, in the example of Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
+%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
+  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
+\end{align}
+Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
+For instance, in Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/export-pdf/figures/illustrations/error-proportions} \caption{\textbf{Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.} Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species.
-Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.}\label{fig:error-proportions}
+\includegraphics[width=0.9\linewidth]{figures/export-pdf/figures/illustrations/error-proportions} \caption{\textbf{Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.} Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species.
+Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.
+Species' proportions may be measured as too high or too low depending on sample composition.
+For instance, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).}\label{fig:error-proportions}
 \end{figure}
 
-\textbf{Difference between species proportions and species ratios:}
-Consistent taxonomic bias at the species level creates a consistent fold error in the ratios among species, but a varying fold error in the proportions of individual species.
-But a proportion is, after all, just a ratio between two taxa in which the denominator taxon is the entire taxonomic domain under study.
-The difference in behavior from that of ratios between species arises because the efficiency of higher-level taxa, which consist of the aggregate of many species, varies across samples with the changing relative abundances among its constituent species.
-
-\hypertarget{absolute-abundances-cell-densities}{%
-\subsection{Absolute abundances (cell densities)}\label{absolute-abundances-cell-densities}}
-
-A wide range of experimental methods exist for converting the relative-abundance information in MGS measurements into measurements of (absolute) cell densities;
-however, these methods can be generally partitioned into two types, based on whether the absolute-abundance information derives from knowing the density of the total community or knowing the density of one or more individual species.
-
-\textbf{Methods using the density of the total community:}
-The total microbial density of a community can be assayed by a variety of methods (reviewed in Appendix \ref{review-absolute-methods}).
-Once the total density is known, the density of individual species can be estimated by multiplying the total density by the species' proportions (as measured by MGS),
+The fact that the multiplicative error in proportions varies while that in ratios is constant forms the basis for understanding the effect of bias on different methods for measuring absolute abundances from MGS and the extent to which bias cancels in different DA analyses.
+
+\hypertarget{absolute-abundance}{%
+\subsection{Absolute abundance}\label{absolute-abundance}}
+
+Researchers often would like to know how the \emph{absolute abundance} of a species changes.
+In this context, \emph{absolute} means not relative to other species; typically, the abundance of interest is relative to non-microbial entities such the volume, mass, or amount of host DNA in the sample.
+A wide range of experimental methods have been used to convert the relative abundances from MGS measurements into measurements of absolute abundances.
+Yet there has so far been little consideration as to how the resulting measurements are affected by taxonomic bias in the MGS measurements or in the supplemental measurements, such as flow cytometry and qPCR, that they often involve.
+Here and in Appendix \ref{review-absolute-methods} we describe how various methods for measuring absolute abundance are predicted to be affected by taxonomic bias, with an eye towards determining whether the measurement error is variable or constant.
+
+Our discussion supposes the `abundance' of interest is cell number, though it applies to other quantities, such as biomass and genome copy number, that may be more relevant in different biological contexts.
+
+We consider two general classes of methods: Those in which absolute-abundance information is derived from measurement of the aggregate abundance of the total community, and those in which it drives from targeted measurement (or prior knowledge) of one or more particular species.
+
+\textbf{Normalization by total abundance:}
+Measurements of species absolute abundances are often obtained by making a (non-MGS) measurement of total community abundance, equating it with the aggregate abundance of the species \(S\) measured by MGS, and multiplying this total by the proportions from MGS,
 \begin{align}
   \label{eq:density-prop-meas}
-  \widehat{\text{density}}_{i}(a) 
-  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{total density}}(a)
-\\&= \text{reads}_{i}(a) \cdot \frac{\widehat{\text{total density}}(a)}{\text{total reads}(a)}.
-\end{align}
-We can rewrite the species' density measurements to account for the effect of taxonomic bias on the MGS proportions (Equation \eqref{eq:prop-error}) and any error in the total density measurement,
+  \widehat{\text{abun}}_{i}(a) 
+  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{abun}}_S(a).
+\end{align}
+This measurement is affected by taxonomic bias in the MGS measurement as well as in the total-abundance measurement.
+For example, measurement of total community abundance by 16S qPCR is affected by variation among species in extraction efficiency, 16S copy number, and PCR binding and amplification bias.
+If the species-level efficiencies of the total-abundance measurement are constant across samples and we neglect other sources, then we can express the total-abundance measurement as
+\begin{align}
+  \label{eq:total-density-error}
+  \widehat{\text{abun}}_S(a) 
+  &= \sum_{i\in S} \text{abun}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
+\\&= \text{abun}_S(a) \cdot \text{efficiency}^{\text{tot}}_S(a),
+\end{align}
+where \(\text{efficiency}_{i}^{\text{tot}}(a)\) is the absolute measurement efficiency of species \(i\) for the total-abundance measurement and \(\text{efficiency}^{\text{tot}}_S(a)\) is the mean efficiency of the total-abundance measurement in the sample.
+The equations \eqref{eq:total-density-error} and \eqref{eq:prop-error} for the error in total-abundance and proportion measurements imply that the species abundance measurement \eqref{eq:density-prop-meas} has error given by
 \begin{align}
   \label{eq:density-prop-error}
-  \widehat{\text{density}}_{i}(a) 
-  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
-%  \cdot \frac{\widehat{\text{total density}}(a)}{\text{total density}(a)}.
-  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{total density}}(a) \end{array}.
-\end{align}
-The variable fold error that bias creates in the measured proportions directly transfers to the density measurements.
-
-Taxonomic bias can also create systematic error in the total-density measurements (Appendix \ref{total-density-bias}).
-For now, we suppose that the total density can be measured accurately;
-later, Section \ref{solutions} describes how experiments might be designed such that taxonomic bias in the total density measurement offsets that in the MGS proportions, leading to a (more) consistent fold error in the measured densities.
-
-\textbf{Methods using the density of one or more reference species:}
-The density of one or more species can be directly measured using targeted measurement methods like species-specific qPCR; alternatively, the densities of one or more species may be thought to be constant across samples either from background knowledge or because they have been spiked in at a fixed density (these situations are reviewed in Appendix \ref{review-absolute-methods}).
-These \emph{reference species} with known or constant density can be used to convert read counts to densities for all species.
-For instance, given the density of a reference species \(r\), we can estimate the density of a focal species \(i\) by
+  \widehat{\text{abun}}_{i}(a) 
+  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}.
+\end{align}
+
+Equation \eqref{eq:density-prop-error} indicates that the multiplicative error in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
+As in the case of proportions (Equation \eqref{eq:prop-error}), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
+On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to a relatively stable error.
+We discuss how this possibility might apply to real experimental workflows below.
+
+\textbf{Normalization by one or more reference species:}
+Suppose we had a measurement of the (absolute) abundance of a \emph{reference species} \(r\).
+In the absence of taxonomic bias, all species are expected to have the same ratio of reads to abundance in a sample (Equation \eqref{eq:measurement-model}).
+Thus the abundance-to-reads ratio for species \(r\) can serve as a conversion factor allowing us to obtain the abundance of an arbitrary species \(i\),
 \begin{align}
   \label{eq:density-ratio-meas}
-%  \widehat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \widehat{\text{density}}_{r}(a).
-  \widehat{\text{density}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{density}}_{r}(a)}{\text{reads}_{r}(a)}.
-\end{align}
-Alternatively, if we only know that the reference species is constant, we can set \(\widehat{\text{density}}_{r}(a)\) in Equation \eqref{eq:density-ratio-meas} equal to 1 to estimate the density in unknown but fixed units, which is sufficient for multiplicative DA analysis.
-(Appendix \ref{model-details} considers possible extensions to multiple reference species.)
-It follows from Equation \eqref{eq:ratio-error} that the error caused by taxonomic bias in this density measurement is
+  \widehat{\text{abun}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{abun}}_{r}(a)}{\text{reads}_{r}(a)}.
+\end{align}
+The abundance of one or more reference species can be directly measured using targeted measurement methods like species-specific qPCR and used with Equation
+\eqref{eq:density-ratio-meas} to obtain absolute abundance for all species.
+To our knowledge, this approach has not previously been suggested.
+Instead, reference-based measurement has been used in the context of spike-in experiments or normalization to a species (such as the host) which is treated as having a constant abundance.
+In a spike-in experiment, the abundance of reference species are added in known (up to experimental error) abundances to each sample.
+When normalizing to a species that is treated as having a constant but unknown abundance, we set \(\widehat{\text{abun}}_{r}(a)\) in Equation \eqref{eq:density-ratio-meas} equal to 1 and interpret the abundances as having fixed but unknown units, which is sufficient for multiplicative DA analysis.
+
+The error in the abundance measurement \eqref{eq:density-ratio-meas} due to taxonomic bias in the MGS measurement is given by
 \begin{align}
   \label{eq:density-ratio-error}
-  \widehat{\text{density}}_{i}(a) 
-  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
-  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{density}_r}(a) \end{array}.
-\end{align}
-Here, it is the constant error in the measured ratio of species \(i\) to \(r\) that propagates into the density measurement.
-There will generally be systematic error in the density of the reference species; however, if the systematic fold error is constant across samples, so will be that of the densities of other species.
-
-\textbf{Differences between the two approaches:}
-Equations \eqref{eq:density-prop-error} and \eqref{eq:density-ratio-error} show that the manner in which taxonomic bias in the sequencing measurement impacts the two approaches mirrors that of proportions and ratios.
-If the fold error in the total-density or reference species measurement is negligible (or at least constant) then the fold errors in species densities by the total-density approach vary across samples with the mean efficiency, whereas the fold errors by the reference-species approach are constant.
-The fundamental determinant of whether constant error is obtained is not which of these two equations is ultimately used, but rather whether densities of individual species, which (we assume) have constant efficiency, or the total community, which has varying efficiency, provide the absolute-abundance information.
-Studies using spike-ins as constant reference species will often, instead of using Equation \eqref{eq:density-ratio-meas}, first use the spike-in species to estimate the total density, and then estimate the density of focal species with Equation \eqref{eq:density-prop-meas}.
-If the calculation is done carefully, however, this method still yields constant fold errors so long as the species efficiencies (including those of the spike-in) are constant across samples (Appendix \ref{total-density-ref}).
+  \widehat{\text{abun}}_{i}(a) 
+  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
+  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) \end{array}.
+\end{align}
+The constant error in the measured ratio of species \(i\) to \(r\) (see Equation \eqref{eq:ratio-error}) propagates to the abundance measurement.
+There will generally also be systematic error in the abundance of the reference species; however, if the systematic fold error is constant across samples, so will be that of the abundances of other species.
+
+Spike-ins are instead sometimes used to measure absolute abundances using Equation \eqref{eq:density-prop-meas}.
+In this case, an intermediate step is taken in which the total community abundance is first measured by the ratio of non-spike-in to spike-in reads.
+If done correctly, this calculation yields results that are identical to directly applying Equation \eqref{eq:density-ratio-meas} (Appendix \ref{total-density-ref}).
+
+\textbf{Difference between the two approaches:}
+Reference-species normalization yields constant fold errors because it is based on species-level read counts and abundances and we assume that efficiencies are constant at the species level.
+In contrast, total-abundance normalization is based on aggregates of species (for the calculation of proportion and the total-abundance measurement) and so depends on mean efficiencies, which can vary across samples.
 
 \hypertarget{differential-abundance}{%
-\section{How taxonomic bias affects differential-abundance analysis}\label{differential-abundance}}
-
-How do the measurement errors described in the previous section impact our ability to estimate the changes in microbial abundances across samples or between different host and environmental conditions?
-Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, which have direct ecological interpretations (via the processes of exponential growth and death) and are ubiquitous in microbiome DA analysis.
-
-\hypertarget{change-between-a-pair-of-samples}{%
-\subsection{Change between a pair of samples}\label{change-between-a-pair-of-samples}}
-
-Before considering the stereotypical many-sample DA analysis, it is instructive to
-consider simplest analysis of differential abundance: the measurement of fold changes in abundance between a pair of samples.
-This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment, and conceptually bridges the single-sample results of the previous section to the many-sample case.
-
-The composition-dependent effect of bias on fold error in proportions leads to error in fold-change measurements that is proportional to the inverse change in mean efficiency.
-From the error in an individual sample (Equation \eqref{eq:prop-error}), it follows that the measured fold change in proportion of taxon \(i\) from sample \(a\) to sample \(b\) is
+\section{Taxonomic bias can cause errors in differential-abundance results}\label{differential-abundance}}
+
+How do the errors caused by taxonomic bias in the abundances measured for individual samples impact DA analysis?
+Although there are many ways to quantify the changes in abundance that form the basis of a DA analysis, we focus on DA analyses of the (log) fold changes in proportions, ratios, and absolute abundance; these multiplicative difference measures are common (though not ubiquitous) in extant DA analyses and have more direct ecological interpretations (via the processes of exponential growth and decay) than non-multiplicative measures.
+
+\hypertarget{fold-changes-between-a-pair-of-samples}{%
+\subsection{Fold changes between a pair of samples}\label{fold-changes-between-a-pair-of-samples}}
+
+The building blocks of a DA analysis are the fold changes (FCs) in abundance between individual pairs of samples.
+
+The impact of bias on the measured FCs in species proportions and ratios follows directly from the results of Section \ref{abundance-measurement} for the error in individual-sample measurements.
+From Equation \eqref{eq:prop-error}, it follows that the measured FC in the proportion of species \(i\) from sample \(a\) to sample \(b\) is
 \begin{align}
@@
   \cdot
-  \underbrace{\left[\frac{\text{mean efficiency}(b)}{\text{mean efficiency}(a)}\right]^{-1}}_\text{fold error}
+  \underbrace{\left[\frac{\text{efficiency}_S(b)}{\text{efficiency}_S(a)}\right]^{-1}}_\text{fold error}
   .
 \end{align}
-The sample-independent efficiency factor of the error cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency from \(s\) to \(t\).
-
-The bottom row of Figure \ref{fig:error-proportions} illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
-The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
-Consequently, the FC of each species is measured to be 2.6X larger than the true value.
-Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
+The sample-independent efficiency factor cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency.
+In contrast, when we use Equation \eqref{eq:ratio-error} to compute the error in the FC in the ratio between two species \(i\) and \(j\), we find that the constant error \(\text{efficiency}_{i} / \text{efficiency}_{j}\) exactly cancels, so that the measured FCs remain accurate regardless of whether the mean efficiency varies.
+
+Figure \ref{fig:error-proportions} (bottom row) illustrates these two different behaviors for the error in the FCs of proportions versus ratios when the mean efficiency varies between samples.
+Here the mean efficiency decreases by a factor of 2.6 (FC of 0.4X) from Sample 1 to Sample 2, which causes the FC of the proportion of each species to be measured as 2.6X larger than its true value.
+Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction; we refer to the latter as a \emph{sign error} since in reference to the change in sign in the corresponding log fold change (LFC).
 We can see each type of error in Figure \ref{fig:error-proportions}.
@@
 For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
-For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!
-
-In contrast, because species ratios are distorted by a constant factor, their measured fold changes remain accurate.
-The fold error in Equation \eqref{eq:ratio-error} completely cancels when we divide the ratio measured for one sample \(a\) by another sample \(b\).
-
-If other error sources remain negligible (or constant across samples), then this dichotomy continues to apply to proportion- and ratio-based density measurements.
-The inferred FCs in proportion-based density measurements will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
-Moreover, any species with a constant density will appear to vary inversely with the mean efficiency.
-In contrast, fold changes in ratio-based density measurements will remain accurate.
+For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase (a sign error).
+In contrast, the fold error in Equation \eqref{eq:ratio-error} completely cancels when we divide the ratio measured for one sample \(a\) by another sample \(b\).
+
+These differences in the FCs of proportions versus ratios are mirrored when we compare FCs in absolute abundance made with normalization to a total-abundance measurement versus a reference-species measurement.
+FCs computed from total-abundance normalization are subject to error when the ratio of the two mean efficiencies (that for the MGS measurement and for the total-abundance measurement vary across samples).
+In contrast, FCs computed from reference-species normalization are not subject to error, so long as the (fold) error in the assumed or measured abundance of the reference species is constant.
 
@@
 
-DA analysis of many samples across host or environmental conditions can typically be framed as a regression problem, in which we analyze the relationship between a microbial \emph{response variable}, such as log density of some focal species \(i\), and one or more \emph{covariates}, such the pH or temperature of the sampled environment or whether the sample is from a healthy or sick person.
-A large fraction of DA analyses use elaborations on the simple linear regression model, which for a response of log species density can be written
+DA analysis of multiple samples can be framed as a regression problem in which we analyze the relationship between a microbial \emph{response variable}, such as the log abundance of species \(i\), with one or more \emph{covariates}, such as the pH of the sampled environment or whether the sample is from a healthy or sick person.
+The simplest regression analysis uses the simple linear regression model,
 \begin{align}
   \label{eq:regression}
-  \log \text{density}_i(a) = \alpha + \beta x(a) + \varepsilon_i(a).
-\end{align}
-Here \(x\) is a continuous covariate (e.g.~pH) or a binary covariate (e.g.~\(x=1\) for treated patients and \(x=0\) for controls), \(\alpha\) and \(\beta\) are regression coefficients, and \(\varepsilon_i(a)\) is a mean-zero random variable that reflects the residual (unexplained) variation in the response (log density of species \(i\)).
-Our interest is usually in the coefficient \(\beta\) (slope or average difference between conditions) that describes how the species' abundance changes with \(x\), while the intercept \(\alpha\) captures differences in the baseline abundance and---we hope---measurement efficiency among species.
-How does taxonomic bias under our measurement model affect estimates of \(\beta\) in the simple linear regression?
-
-Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \eqref{eq:density-prop-meas}) with error-free estimates of the total density.
-If the true log density follows the regression Equation \eqref{eq:regression}, then it follows from Equation \eqref{eq:density-prop-error} the estimated log density equals
-\begin{align}
-  \label{eq:regression-error}
-  \log \widehat{\text{density}}_i(a)
-%  &= \alpha + \beta x(a) + \varepsilon_i(a) + \log \text{efficiency}_i - \log \text{mean efficiency}(a)
-  = [\alpha + \log \text{efficiency}_i] + [\beta - \log \text{mean efficiency}(a)] x(a) + \varepsilon_i(a).
-\end{align}
-This equation shows that the species-specific portion of the error affects the intercept term while the sample-specific portion (log mean efficiency) affects the slope term.
-Thus, as in the case of fold changes between two samples (Equation \eqref{eq:prop-fc-error}), it is the variation in the (log) mean efficiency across samples that we must worry about distorting our DA results.
-
-The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient \(\beta\).
-Appendix \ref{appendix-regression} mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \ref{fig:regression-example} illustrates using simulated data.
+  \log \widehat{\text{abun}}_i(a) = \alpha_i + \beta_i x(a) + \varepsilon_i(a),
+\end{align}
+where \(x\) is a continuous covariate (e.g.~pH) or a binary covariate (e.g.~\(x=1\) for treated patients and \(x=0\) for controls), \(\alpha_i\) and \(\beta_i\) are regression coefficients, and \(\varepsilon_i(a)\) is a mean-zero random variable that reflects the residual (unexplained) variation in the response.
+In a DA analysis, we are interested in the slope coefficient \(\beta\), which describes how the species' abundance changes with \(x\).
+We are generally uninterested in the intercept coefficient, \(\alpha\), which captures differences in the baseline abundance among species and---it is usually hoped---study- and species-specific systematic error such as that created by taxonomic bias.
+
+How does taxonomic bias in fact impact estimates of the slope coefficient, \(\beta\)?
+Appendix \ref{appendix-regression} derives the effect of bias on the estimated coefficients for the simple linear regression model under estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE).
+The result has a simple interpretation that provides intuition for what to expect in more complicated generalized linear regression models used in popular DA methods.
+Consider the regression \eqref{eq:regression} fit to the measured log abundance of species \(i\) across all samples.
+The measurement error in the response variable,
+\begin{align}
+  \text{error}_{i}(a) \equiv 
+  \log \widehat{\text{abun}}_i(a) - \log \text{abun}_i(a)
+  = \log \frac{\widehat{\text{abun}}_i(a)}{\text{abun}_i(a)},
+\end{align}
+can be thought of as having three components:
+
+\begin{itemize}
+\tightlist
+\item
+  Error that is constant across samples affects the intercept coefficient, \(\alpha_i\).
+\item
+  Error that varies systematically with \(x\) affects the slope coefficient, \(\beta_i\).
+\item
+  Error that varies and is uncorrelated with \(x\) affects the residual variation, \(\varepsilon_{i}(s)\).
+\end{itemize}
+
+We saw in Section \ref{abundance-measurement} that the error from taxonomic bias has a species-specific, sample-independent component that is constant across samples; this component causes a species-specific systematic error in the estimated intercept, \(\alpha_{i}\).
+The error also has a sample-specific, species-independent component.
+The portion that is correlated with the covariate \(x\) creates a systematic error in the slope coefficient \(\beta_{i}\) that is the same for every species and opposite in sign to how the error varies with \(x\);
+the portion that is uncorrelated with \(x\) affects the precision (standard errors) of the slope estimates in a species-specific manor.
+
+Figure \ref{fig:regression-example} illustrates with a simulated data for the case of log abundance measured using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}), assuming that total community abundance is measured perfectly accurately.
+In this case, the variable component of the error is due entirely to variation in the log mean efficiency of the MGS measurement.
 Variation in the log mean efficiency that is associated with the covariate \(x\) creates a systematic error in the estimated slope \(\hat \beta\) equal to the negative of the (scaled) covariance of log mean efficiency with \(x\).
-The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with \(x\) relative to that of the response (here, \(\log \text{density}_{i}\)) with \(x\) or, equivalently, the relative magnitudes of their slopes.
+The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with \(x\) relative to that of the response (here, \(\log \text{abun}_{i}\)) with \(x\) or, equivalently, the relative magnitudes of their slopes.
 As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \ref{fig:regression-example}), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
@@
 
-With this understanding in place, we briefly summarize how taxonomic bias affects estimation of the simple linear model in other abundance types.
-The results for proportion-based density estimates with accurate total densities also apply to LFC analysis of proportions.
-Similar results apply to microbiome regression tools (such as corncob; Martin, Witten, and Willis (\protect\hyperlink{ref-martin2020mode}{2020})) that perform regression on the logit (instead of log) proportion of a species; however,
-the mean efficiency of the entire sample must instead be replaced with the mean efficiency among all species excluding the focal species, which causes the absolute error in regression coefficients to vary somewhat across species.
-Because ratios and ratio-based densities are subject to consistent multiplicative error, in analysis of log ratios and the log densities derived from them, only the estimated intercept \(\hat \alpha\) is affected by taxonomic bias, while the point estimate and standard error of the estimated slope \(\hat \beta\) remain unaffected.
+Overall, we find that variation in measurement error created by variation in the mean efficiency of the MGS measurement can negatively affect certain DA analyses by creating systematic errors in the slope coefficient estimates and/or reducing their precision.
+Similar effects occur for regressions of log proportions and of log abundances derived by total-abundance normalization when the total abundance is accurately measured; however, as noted in Section \ref{abundance-measurement}, total-abundance measurements that are taxonomically biased in a complementary manner can mitigate these effects.
+Regression analyses of log-ratios and log abundances derived from reference-species normalization remain unaffected---the errors are constant across samples and so only impact the intercept coefficients.
 
@@
 These DNA mock communities allowed the authors to directly measure the bias due to the sequencing workflow following DNA extraction and correct for it when analyzing the primary experimental samples following the method of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
-Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification (CITE) and the substantial ribosomal copy-number-variation (CNV) in fungi (Lofgren et al.
+Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification and the substantial ribosomal copy-number-variation (CNV) in fungi (Lofgren et al.
 (\protect\hyperlink{ref-lofgren2019geno}{2019})).
@@
 To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
-The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.57X and a geometric standard deviation of just 1.06X (SI Figure \ref{fig:leopold2020host-variation}).
+The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.62X and a geometric standard deviation of just 1.05X (SI Figure \ref{fig:leopold2020host-variation}).
 Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
@@
 In particular, it is interesting to consider whether any commensals are able to increase in absolute concentration in response to infection.
-The experiment performed by Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration
-measured using the total-community normalization method (Equation \eqref{eq:density-prop-meas}).
+The experiment performed by Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration measured using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}).
 It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
-The pathogen is absent in pre-infection samples but tends to dominate the community post-infection (SI Figure \ref{fig:leopold2020host-infection-mean-efficiency-dist}), leading to an average increase in mean efficiency ranging from 3.5X to 5X across host genotypes and an expectation that commensals will typically decrease in proportion due to the sum-to-one constraint.
-Without bias correction, the log proportion of commensals decrease in each host genotype, often substantially (log-e fold changes less than -2); however, bias leads to overly negative LFCs due to the higher mean efficiency in post-infection samples.
-Bias correction substantially lessens the magnitude of LFC estimates, in some cases leading to estimates that are approximately 0 or slightly positive (SI Figure \ref{fig:leopold2020host-infection-lfc}).
-
-Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-community normalization method (Equation \eqref{eq:density-prop-meas}).
+The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, raising the mean efficiency of post-infection samples (SI Figure \ref{fig:leopold2020host-infection-mean-efficiency-dist}).
+Across different host genotypes, the average increase in mean efficiency ranges from 2.5X to 5.2X.
+We used simple linear regression to measure the average change in log proportion of each commensal for each host genotype, with and without bias correction.
+Regardless of whether bias is accounted for, the log proportion of each commensal decreased, which is unsurprising given the pathogen's growth and the sum-to-one constraint of species proportions.
+Without bias correction, however, the LFCs were lower by an amount corresponding to the inverse change in log mean efficiency.
+The magnitude errors created were substantial for a large fraction of commensal-host pairs; in several cases, the LFC estimates without bias correction were clearly negative, whereas the estimates with bias correction were close to 0.
+
+Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}).
 We consider two scenarios.
@@
 The scientific error, however, may be much worse.
-Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where taxa that persist or even increase appear to decrease.
-\emph{Scenario 2:} Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-community normalization method.
+Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where some taxa that persist or increase in abundance instead appear to decrease.
+\emph{Scenario 2:} Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-abundance normalization method.
 In this case, we should expect the increase in mean efficiency to have little to no impact on LFCs, since it would be offset by the increase in mean efficiency of the total-concentration measurement.
 Moreover, since the total ITS concentration and ITS sequencing are both performed on the extracted DNA, we can also expect the effect of taxonomic bias caused by DNA sequencing to be offset.
-Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-community measurement is made prior to normalization.
-
-\hypertarget{vaginal-microbiomes-of-pregant-women}{%
-\subsection{Vaginal microbiomes of pregant women}\label{vaginal-microbiomes-of-pregant-women}}
+Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-abundance measurement is made prior to normalization.
+
+\hypertarget{vaginal-microbiomes-of-pregnant-women}{%
+\subsection{Vaginal microbiomes of pregnant women}\label{vaginal-microbiomes-of-pregnant-women}}
 
@@
 
-\hypertarget{gut-microbiome}{%
-\subsection{Gut microbiome}\label{gut-microbiome}}
-
-(Has not been updated)
-
-It is clear that an individual species with an unusually high or low efficiency can drive large changes in the mean efficiency when it shifts from a very large (\(\sim 1\)) to very small proportion (\(\ll 1\)) of the sample.
-Consideration of dynamics in the human gut illustrates another a way in which large shifts in higher-level taxa might create similar shifts, due to the fact that efficiencies do not vary independently among closely related species.
-Gut microbiomes are dominated by two phyla: The Bacteroidetes and the Firmicutes, the ratio of which varies substantially across individuals.
-Within a sample, there are usually multiple species contributing to the abundance of each phylum, with the median Inverse Simpson diversity being around 4 for each phylum in the HMP2 IBD study.
-Thus changes from Bacteroidetes to Firmicutes dominance are unlikely to be driven by a single species.
-However, if Firmicutes species tend to have systematically different efficiencies than Bacteroidetes species, then changes in phylum dominance can be expected to drive substantial changes in the mean efficiency even if many species contribute.
-Indeed, several studies have found DNA extraction protocols to more efficiently lyse Gram-negative Bacteroidetes species than Gram-positive Firmicutes species (though substantial variation within phyla also occurs, McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-
-Systematic error in DA regression results requires that the variation in log mean efficiency is correlated with the covariate.
-The examples above each suggest plausible biologically-significant scenarios in which such correlations might arise.
-The Bacteroidetes-to-Firmicutes ratio has also been linked to host conditions.
-Yet in the HMP2 IBD dataset, the large variation in this ratio is largely independent of disease status, suggesting that the primary effect of any variation in mean efficiency driven by these phyla would be to increase noise rather than create systematic error in DA analysis.
-
 \hypertarget{microbial-growth-in-marine-sediments}{%
@@
 (\protect\hyperlink{ref-lloyd2020evid}{2020}) performed a systematic assessment of growth rates of bacterial and archaeal taxa over a depth of 10 cm (corresponding to \textasciitilde40 years of burial time) in sediment of the White Oak River estuary.
-To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-community normalization method (Equation \eqref{eq:density-prop-meas}), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy (CARD-FISH).
+To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy (CARD-FISH).
 The authors refer to these absolute densities as FRAxC measurements, for `fraction of 16S reads times total cell counts'.
@@
 Thus comparing qPCR to FRAxC growth rates on the same reference taxa allows us to estimate the systematic error in FRAxC growth rates.
-Because the systematic error in regression slopes due to bias is the same for each taxon (Equation \eqref{eq:regression-error}), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
+Because the systematic error in regression slopes due to bias is the same for each taxon (Section \ref{differential-abundance}), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
 The first soil core included qPCR measurements of a single archaeal clade, \emph{Bathyarchaeota}, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of 0.099/yr by FRAxC and 0.097/yr by qPCR).
@@
 
-\hypertarget{conclusions}{%
-\subsection{Conclusions}\label{conclusions}}
+\hypertarget{summary-and-conclusions}{%
+\subsection{Summary and conclusions}\label{summary-and-conclusions}}
+
+The impact of bias can depend on protocol, biological system, and type of DA analysis being done.
+Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of section X help suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.
+
+First, the error caused by consistent taxonomic bias \emph{can mostly cancel} in cross-sample comparisons and so not impact DA analyses of fold changes in species proportions.
+Our theoretical results from Section \ref{differential-abundance} indicate that when the mean efficiency is roughly constant across samples, the error in proportions cancels in fold change calculations and is absorbed by the intercept term in regression models.
+We observed this scenario in the analysis of pre-infection fungal microbiome samples of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) and when analyzing the trajectories of the vaginal microbiomes of women when the dominant species remained constant\footnote{This point could use more support in the MOMSPI case study}, and a stable mean efficiency within marine soil cores is plausible and consistent with (though unable to be fully determined by) the different growth rate estimates in the Lloyd et al.
+(\protect\hyperlink{ref-lloyd2020evid}{2020}) experiment.
+Section \ref{differential-abundance} showed that absolute DA analysis using total-abundance normalization are susceptible to errors as with DA analysis of proportions; yet our analysis of two (hypothetical) approaches to analyzing absolute changes in response to infection in the fungal microbiome experiment indicates that the increasingly common approach of pairing marker-gene sequencing with qPCR of the total marker density can largely mitigate this effect, since here what matters is the variation in the ratio of mean efficiencies of the MGS measurement to the total-density measurement and this ratio may remain roughly constant if bias is largely shared by the two measurements.
+
+Yet in other cases, the mean efficiency can vary substantially and create substantial error in fold changes between pairs of samples.
+We saw examples when comparing fungal microbiomes pre- and post-infection in the foliar-fungi experiment and comparing vaginal microbiomes with different dominant species in the MOMS-PI experiment.
+The impact of these errors on results of DA regression analysis across many samples depends on whether the mean efficiency varies systematically with the covariate of interest.
+In these two examples, systematic variation of the mean efficiency arose as high-efficiency species tended to dominate samples in one of the sample conditions (post-infection foliar samples or low-diversity vaginal samples).
+Thus one way for significant error in DA regression results to arise is when a small number of (or just one) species that have particularly high or low mean efficiencies, form a large proportion of the community in a substantial fraction of samples, and are associated with the covariate of interest.
+
+For many experimental systems, however, there may not be any one species that frequently forms a large fraction of the community.
+In these systems, systematic variation of the mean efficiency can still arise through the collective change of many species that are associated with the covariate of interest.
+We described a hypothesized scenario in the marine-sediment case study in which increased burial time (the covariate) selects for lysis-resistant (and so lower efficiency) species.
+As another example, treatment with a certain antibiotic might selectively kill easier-to-lyze Gram negative species (or harder-to-lyse Gram positives) and thereby decreasing (or increasing) the mean efficiency in samples collected from a host post-treatment.
+Such situations can arise whenever there is a microbial trait that is associated with both measurement efficiency and the biological processes of interest.
+An example besides cell-wall structure is ribosomal copy number, which increases a species' efficiency in ribosomal amplicon measurements and is positively associated with metabolic rate.
+A more generic mechanism by which mean-efficiency associations might arise is from the evolutionary relationships among species.
+Species with more recent common ancestry are expected to be more similar across a wide range of heritable traits, which include traits that affect measurement efficiency (such as cell wall structure, genome size, ribosomal copy number, PCR binding sequence) as well as traits that affect the biological processes under study.
+For example, differences in the phylum-level composition of samples from two conditions might be driven by many species that show phylum-level conservation in a relevant biological trait.
+If these species also show phylum-level conservation in (potentially different) traits that affect measurement efficiency, an association of mean efficiency with the condition can arise.
+
+So long as the mean efficiency does not vary systematically with the regression covariates, the error in fold changes serve primarily to reduce the precision and power of regression analyses without systematically distorting estimates and so does not necessarily lead to invalid inferences.
+But the reduction in precision could be substantial, particularly given the small sample sizes common in many microbiome studies, and thereby substantially limit the ability to draw meaningful conclusions from a study\footnote{I suspect this non-systematic variation in the mean efficiency is the more typical situation, but we currently lack an example of it in the case studies.
+We can find a semi-real example by taking real microbiome data from a case-control-type design and simulating a strong degree of bias.
+I did this with the HMP2 IBD study, but thinnk it might be better to find a different example where there are some clear DA results.}.
+
+These observations provide reasons to both worry and hope.
+It seems likely that in many experiments the mean efficiency is consistent or is at least not associated with the covariate, so that DA inferences remain valid.
+Yet it is not obvious \emph{a priori} in which studies this condition holds, and there are plausible mechanisms that can create problematic associations of the mean efficiency even in ecosystems with high species diversity.
+Thus while we should not discount the large set of existing DA results, we should seek ways to better assess the robustness of results from previous studies and to measure and correct the error caused by bias in future ones.
 
@@
 
-\hypertarget{analyses-of-fold-changes-in-species-ratios}{%
-\subsection{Analyses of fold changes in species ratios}\label{analyses-of-fold-changes-in-species-ratios}}
-
-Since the fold changes in species ratios are unaffected by consistent taxonomic bias, a natural approach to countering bias in relative DA analysis is to ratio- rather than proportion-based methods.
+\hypertarget{use-ratio-based-relative-abundance-analyses}{%
+\subsection{Use ratio-based relative-abundance analyses}\label{use-ratio-based-relative-abundance-analyses}}
+
+Since bias creates a consistent fold error in species ratios, a natural approach to countering bias when analyzing relative abundances is to use DA methods that are based on fold changes in ratios.
 A variety of such methods have been developed for microbiome DA analysis that build on earlier work from the field of Compositional Data Analysis (CoDA).
@@
 (\protect\hyperlink{ref-brooks2015thet}{2015}) that contained all 7 species and used it to calibrate the compositions of all samples (Figure \ref{fig:calibration-example}).
-We then estimated species' densities using the uncalibrated and calibrated proportions using the community normalization method, treating the total density as known and fixed.
+We then estimated species' densities using the uncalibrated and calibrated proportions using total-abundance normalization, treating the total density as known and fixed.
 Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions and densities (Figure \ref{fig:calibration-example}).
@@
 To the extent that taxonomic bias truly is consistent within studies, meta-analyses that directly account for it can have greater statistical power to extract the truly consistent and inconsistent microbial patterns across studies.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+\emph{This section is a placeholder for a short (2-3 paragraph) conclusion. We are planning to keep `discussion' to the earlier sections and appendix, at least for initial submission.}
 
@@
 Equation \eqref{eq:prop-error} for the error in a species' proportion and Equation \eqref{eq:ratio-error} for the error in the ratio between two species both follow directly from this more general expression.
-
-\hypertarget{revised-main-text-equations}{%
-\subsubsection{Revised main text equations}\label{revised-main-text-equations}}
-
-Let \(S\) be the set of all species.
-Then
-\begin{align}
-  \text{reads}_S(a)
-    = \text{density}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
-\end{align}
-where
-\begin{align}
-  \text{efficiency}_S(a) 
-  \equiv \frac{\sum_{j\in S}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{density}_S(a)}
-\end{align}
-Also,
-\begin{align}
-  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
-\end{align}
-and
-\begin{align}
-  \widehat{\text{prop}}_{i}(a)
-  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}.
-\end{align}
 
@@
 \caption{\label{fig:leopold2020host-infection-lfc}\textbf{Bias correction increases the estimated increase in log proportion in response to infection for commensal taxa across all host genotypes.} Shown are the estimated log fold change (LFC) and 95\% confidence intervals from simple linear regression of log (base e) proportion against experimental timepoint for commensal taxa.
-Negative values indicate that the proportion of the taxon decreased on average in response to infection, which we expect due to the increase in pathogen abundance and the sum-to-one constraint.
-Bias leads to artificially low estimates, as the pathogen drives an increase in mean efficiency.
-In Western genotypes, several taxa whose proportion is observed to decrease without correction are instead found to remain stable or even increase when bias is corrected.}
+Negative values indicate that the proportion of the taxon decreased on average in response to infection, which we expect due to an increase in pathogen abundance and the sum-to-one constraint of proportions.
+Bias leads to artificially low estimates, as the increased pathogen proportion drives an increase in mean efficiency.}
 \end{figure}
@@
 \url{https://doi.org/10.3402/mehd.v26.27663}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-martin2020mode}{}}%
-Martin, Bryan D., Daniela Witten, and Amy D.
-Willis.
-2020.
-{``{Modeling microbial abundances and dysbiosis with beta-binomial regression}.''} \emph{Ann.
-Appl.
-Stat.} 14 (1): 94--115.
-\url{https://doi.org/10.1214/19-AOAS1283}.
 

### 2022-01-19 conclusion.Rmd
Draft Conclusion

@@
 
-_This section is a placeholder for a short (2-3 paragraph) conclusion. We are planning to keep 'discussion' to the earlier sections and appendix, at least for initial submission._
+Our results show that even if taxonomic bias is consistent across samples at the taxonomic level being analyzed, it can create error in DA estimates that can lead to scientifically erroneous conclusions.
+In particular, analyses based on fold-changes in proportions or in absolute abundances derived from total-abundance normalization (sometimes called 'Quantitative Microbiome Profiling') are both capable of leading to spurious results.
+<!-- Perhaps state that we showed this theoretically, in mock communities, and in real case studies -->
+The error, however, varies by experimental context.
+It is negligible if the mean efficiency is stable across samples, and may be unimportant if it is not correlated with the focal covariate in a regression analysis.
+The use of reference-taxon measurement and/or community calibration controls provides a way to validate and correct DA results.
+In addition, bias-aware sensitivity and meta-analyses make it possible to rigorously account for bias even in the absence of such controls.
+Over time, use of these methods will lead to a collective understanding of the experimental contexts in which these errors are likely.
+
+We also showed that other DA methods are nearly or entirely invariant to consistent bias.
+Analyses based on fold changes in ratios, including analysis of absolute abundance via reference-species normalization, are invariant to consistent bias in the MGS measurement.
+In addition, careful pairing of total-abundance and MGS method may remove most of the bias-driven error.
+Important questions remain about the true bias-invariance of these methods, including the extent to which bias is consistent across samples at the taxonomic resolution of the given MGS method and the interaction of bias with analytical assumptions such as how zeros are interpreted and the possibility of taxonomic misassignment and cross-sample contamination.
+Nevertheless, these are straightforward, immediately implementable methods to addressing bias in fold-change analysis.
+
+<!-- Placeholder for potential paragraph on limitations and open questions -->
+
+In summary, our results show that protocol standardization is insufficient to nullify the effect of taxonomic bias in DA results, but that there are a range of practical solutions to mitigate it for an important subset of experimental questions and contexts.
+The theoretical framework we present can be further used to rigorously analyze the effect of bias on other variations of DA analysis and of microbiome analyses more generally.
+With this work, we look to a future in which microbiome researchers routinely choose an appropriate combination of experimental and data-analytic methods capable of answering their fundamental question while accounting for taxonomic bias and the other inherent limitations of MGS measurements---and in this way, gaining the confidence needed to codify statistical findings into true scientific knowledge.

### 2022-01-19 introduction.Rmd
Draft abstract

@@
 # Introduction {#introduction}
+
+::: {#abstract .rmdnote}
+
+**Abstract:**
+To assess how the species in microbial communities vary across conditions, researchers often measure the relative or absolute abundances of all species simultaneously using marker-gene and metagenomic sequencing.
+These measurements are taxonomically biased, with some species measured many times more efficiently than others.
+Measurement error caused by taxonomic bias is generally ignored in such _differential abundance (DA) analyses_, likely due to a combination of 1) a widespread belief that protocol standardization is sufficient to yield accurate differences between samples and 2) a lack of practical solutions for addressing it.
+Using theoretical arguments and analysis of real and simulated experiments, we show that taxonomic bias can create error in DA results that is sufficient to lead to scientifically erroneous conclusions.
+This error, however, varies by experimental context and analysis method.
+We present six approaches to addressing the error, each suited to different experimental questions and systems.
+These approaches provide practical solutions to mitigating the effect of taxonomic bias for a large fraction of DA questions.
+If adopted, they may lead to notable improvements in the reproducibility and biological interpretability of the statistical findings of microbiome studies.
+
+:::
 

### 2022-01-19 index.Rmd
Adjust git file tracking
Add folders to ignore list and set up index.Rmd so that untracked,
unstaged, and staged files are output in the Preface.
Revise title
Update preface and readme

@@
 ---
-title: "Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements"
+title: "Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements"
 author:
@@
 ```{r}
-if (length(status$unstaged) > 0)
-  print(status$unstaged)
+print(status)
 ```
@@
 Analyses of both relative and absolute abundances are considered.
-In its current form, the manuscript sits somewhere between a standard scientific article and a monograph; 
-it consists of an article followed by a series of appendices which together give a comprehensive treatment of the implications of the @mclaren2019cons model of taxonomic bias for differential-abundance analysis and experimental design.
-It is licensed under a [CC BY 4.0 License](https://creativecommons.org/licenses/by/4.0/).
+The manuscript is licensed under a [CC BY 4.0 License](https://creativecommons.org/licenses/by/4.0/).
 See [the Zenodo record](https://doi.org/10.5281/zenodo.4552717) for how to cite the latest version.

### 2022-01-24 appendix-regression.Rmd
Fix typo
Revise regression appendix

@@
-# Error in estimated regression coefficients and standard errors {#appendix-regression}
+# Taxonomic bias and linear regression {#appendix-regression}
 
-Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.
+This section derives the theoretical results of Section \@ref(differential-abundance) for the effect of taxonomic bias on regression-based DA analyses.
+As in Section \@ref(differential-abundance), we restrict our attention to DA analyses that can be expressed in terms of the simple linear regression model in which the response is log absolute abundance or log proportion of individual species or the log ratio of a pair of species.
 
-Consider Equation \@ref(eq:regression) for the simple linear regression of log density of a species $i$ on covariate $x$.
-Let $y$ stand for the actual log density of a focal species $i$, $z$ stand for the log density that we've measured, and $d = z - y$ equal the difference between the two, which here is the log efficiency of species $i$ minus the log mean efficiency of the sample.
-Let $s_{xy}$ denote the sample covariance between variables $x$ and $y$, $s^{2}_{x} = s_{xx}$ and $s_{x}$ denote the sample variance and standard deviation, and $r_{xy} = s_{xy}/(s_{x}s_{y})$ denote the sample correlation.
+## Review of simple linear regression
 
-The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of $z$ equals the sample covariance of $z$ and $x$ divided by the sample variance in $x$ or, equivalently, the sample correlation of $z$ and $x$ multiplied by the ratio of their sample standard deviations,
+Table: (\#tab:statistical-notation) Statistical notation used in this section.
+
+| Notation | Mathematical definition | Description |
+|:--|:----|:----|
+| $x$, $y$, $z$, $d$ |                                                     | Random variables, defined in text |
+| $x(a)$             |                                                     | Value of $x$ in sample $a$ |
+| $\bar x$           | $\frac{1}{n}\sum_a x(a)$                            | Sample mean of $x$ |
+| $s(x,y)$           | $\frac{1}{n}\sum_a (x(a) - \bar x) (y(a) - \bar y)$ | Sample covariance between $x$ and $y$ |
+| $s^{2}(x)$         | $s(x,x)$                                            | Sample variance of $x$ |
+| $s(x)$             | $\sqrt{s^{2}(x)}$                                   | Sample standard deviation of $x$ |
+| $r(x,y)$           | $\frac{s(x,y)}{s(x) s(y)}$                          | Sample correlation between $x$ and $y$ |
+| $\alpha$, $\beta$, $\sigma$                            | | Parameters of the regression model |
+| $\hat \alpha$, $\hat \beta$, $\hat \sigma$             | | Parameter estimates |
+| $\epsilon(a)$      | $y(a) - (\alpha + \beta x(a))$      | Error for sample $a$ |
+| $\hat \epsilon(a)$ | $y(a) - (\hat \alpha + \hat \beta x(a))$ | Residual for sample $a$ |
+| $\text{se}(\hat \beta)$, $\hat{\text{se}}(\hat \beta)$ | | Standard error of $\hat \beta$ and its estimate |
+
+We begin by reviewing definitions, notation, and results for the analysis of the simple linear regression model.
+Our presentation follows @wasserman2004allo Chapter 13 with additional interpretation of the estimated regression coefficients.
+<!-- to support our subsequent consideration of measurement error caused by bias. -->
+Statistical notation is defined in Table \@ref(tab:statistical-notation).
+
+Consider a response variable $y$ and a covariate $x$.
+Following @wasserman2004allo, we define the _simple linear regression model_ by
 \begin{align}
-  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_y}{s_x}.
+  (\#eq:lm)
+  y(a) &= \alpha + \beta x(a) + \varepsilon(a),
 \end{align}
-From the (bi)linearity of sample covariances it follows that
+where $E[\varepsilon(a) \mid x(a)] = 0$ and $V[\varepsilon(a) \mid x(a)] = \sigma^2$.
+That is, conditional on the value of $x$ in a sample $a$, the response $y$ is given by $\alpha + \beta x$ with a random error $\varepsilon$ that has an expected value of $0$ and a constant variance $\sigma^2$.
+
+<!-- Note that we distinguish between the statistical model being fit to the data, and the actual data generating process. -->
+<!-- These results for the regression model estimates hold regardless of whether the simple linear model is a good one for the data. -->
+
+Given data $(y, x)$, we _fit_ the model by finding estimates for the parameters $\alpha$, $\beta$, and $\sigma^2$ that best reflect the data under the assumption that the model is correct and according to our chosen criterion that defines 'best'.
+Here we consider the least-squares estimates for the coefficients $\alpha$ and $\beta$, which are also the maximum likelihood estimates if we further assume that the errors $\varepsilon$ are normally distributed.
+<!-- Defining the residuals as $\hat \varepsilon(a) = y(a) - (\hat \alpha + \hat \beta x(a))$,  -->
+The least-squares estimates of $\alpha$ and $\beta$ are the values that minimize the residual sum of squares, $\text{RSS} = \sum_{a} \hat \varepsilon(a)$.
+These estimates can be conveniently expressed in terms of the sample means and covariances or correlations of $x$ and $y$,
 \begin{align}
-  \hat \beta_z 
-  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
-  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
-  = \hat \beta_y + \hat \beta_d,
+  (\#eq:lm-hat)
+  \hat \alpha &= \bar y - \hat \beta \bar x \\
+  \hat \beta &= \frac{s(y,x)}{s^2(x)} = r(y,x) \cdot \frac{s(y)}{s(x)}.
 \end{align}
-where $\hat \beta_y$ and $\hat \beta_d$ denote the slope estimates for $y$ and $d$ (were these values to be known).
-The absolute error in the estimate $\hat \beta_{z}$ of $\hat \beta_{y}$ is therefore $\hat \beta_{d}$; it is large in a practical sense when $\hat \beta_{d}$ is large (in absolute value) compared to $\hat \beta_{y}$, which corresponds to the covariance of $d$ with $x$ being large compared to the covariance of $y$ with $x$.
+<!-- (@wasserman2004allo Chapter 13), -->
+The estimate for the slope coefficient, $\hat \beta$, equals the ratio of the sample covariance of $y$ with $x$ relative to the variance in $x$ or, equivalently, the correlation of $y$ with $x$ multiplied by the ratio of their standard deviations.
+The estimate for the intercept, $\hat \alpha$, is the value such that the sample means follow the regression relationship, $\bar y = \hat \alpha + \hat \beta \bar x$.
+<!-- We are most interested in the estimate of the slope coefficient, $\hat \beta$ -->
+The standard unbiased and maximum likelihood estimates for the variance $\sigma^{2}$ are both approximately given by the sample variance of the residuals,
+\begin{align}
+  (\#eq:lm-var)
+  \hat{\sigma}^2(\hat \beta) \approx s^2(\hat \varepsilon).
+\end{align}
 
-In our case, the covariance of $d$ equals the negative of the covariance of log mean efficiency with $x$.
-The absolute error in $\hat \beta$ equals the negative covariance of log mean efficiency scaled by the variance in $x$.
-This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
-For species that covary with $x$ more strongly than the log mean efficiency, the error will be relatively small.
-This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with $x$ compared to the log density of the focal species.
+We are most interested in the estimated slope coefficient, $\hat \beta$.
+Besides the point estimate indicating our best guess at the value, we also wish to understand the uncertainty or precision of the estimate. 
+This uncertainty is quantified by the standard error, $\text{se}(\hat \beta)$, estimated by
+\begin{align}
+  (\#eq:lm-se)
+  \hat{\text{se}}(\hat \beta)
+  = \frac{\hat \sigma}{s(x) \sqrt{n}}
+  \approx \frac{s(\hat \varepsilon)}{s(x) \sqrt{n}},
+\end{align}
+<!-- (@wasserman2004allo Equation 13.13). -->
+Approximate 95% confidence intervals for $\beta$ are given by $\hat \beta \pm 2\; \hat{\text{se}}(\hat \beta)$.
 
-We can similarly understand the impact of measurement error on the precision of our slope estimates.
-The OLS and MLE estimated standard error in $\hat \beta$ are both approximately
+## Measurement error in the response
+
+Now consider a researcher who would like to regress a biological quantity $y$ (the response) on a second quantity $x$ (the covariate).
+But instead of $y$, they only know a proxy $z$, which is related to $y$ via the difference $d = z - y$.
+For instance, $y$ might be the log abundance of a species, $z$ is the abundance that is measured via MGS, and $x$ is a numerical quantity like pH or a boolean variable indicting case versus control.
+Lacking a direct measurement of $y$, the researcher instead regresses $z$ on $x$ and interprets the fitted model as informing them about the relationship between $y$ and $x$.
+We wish to understand the accuracy of their conclusions.
+
+To understand how the measurement error in $y$ impacts the researcher's regression analysis, consider the three linear regression equations for $y$, $z$, and $d$,
 \begin{align}
-  \hat{\text{se}}(\hat \beta)
-  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
+  (\#eq:lm-y)
+  y(a) &= \alpha_y + \beta_y x(a) + \varepsilon_y(a),
 \end{align}
-where $s_{\hat \varepsilon}$ is the sample standard deviation of the residuals
-(@wasserman2004allo Chapter 13).
-The sample residuals of $z$, $y$, and $d$ have a similar relationship to the regression coefficient estimates,
 \begin{align}
-  \hat \varepsilon_z 
-    &\equiv z - \hat \beta_z x
-  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
-  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
-  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
+  (\#eq:lm-z)
+  z(a) &= \alpha_z + \beta_z x(a) + \varepsilon_z(a),
 \end{align}
-(Note, here I've omitted the subscript indicating the dependence on the sample.)
-It follows that the sample variances of the residuals of $z$, $y$, and $d$ are related through
 \begin{align}
-  s^2_{\hat \varepsilon_{z}} 
-  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
-  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
+  (\#eq:lm-d)
+  d(a) &= \alpha_d + \beta_d x(a) + \varepsilon_d(a).
 \end{align}
-The standard deviation of the $z$ residuals is increased above that of the $y$ residuals when the $d$ residuals are either uncorrelated or positively correlated with the $y$ residuals, but may be decreased when the $y$ and $d$ residuals are negatively correlated.
+The researcher would like to fit the model for $y$ (Equation \@ref(eq:lm-y)), but instead can only fit the model for $z$ (Equation \@ref(eq:lm-z)).
+The two are related via the Equation \@ref(eq:lm-z) for $d$.
+It is helpful to imagine, for the moment, that $y$ and $d$ are known and so we can fit all three models \@ref(eq:lm-y), \@ref(eq:lm-z), and \@ref(eq:lm-d).
+From Equation \@ref(eq:lm-hat) and the linearity of sample means and covariances it follows that 
+\begin{align}
+  (\#eq:lm-hat-rel)
+  \hat \alpha_z &= \hat \alpha_y + \hat \alpha_d \\
+  \hat \beta_z &= \hat \beta_y + \hat \beta_d.
+\end{align}
+That is, the estimated coefficients for $y$, $z$, and $d$ mirror the relationship between the variables themselves,  $z = y + d$.
+Consequently, we see that the measurement error $d$ creates absolute errors in the regression coefficients $\hat \alpha$ and $\hat \beta$ equal to
+\begin{align}
+  (\#eq:lm-hat-err)
+  \text{abs. error}(\hat \alpha) &\equiv \hat \alpha_z - \hat \alpha_y = \hat \alpha_d \\
+  \text{abs. error}(\hat \beta) &\equiv \hat \beta_z - \hat \beta_y = \hat \beta_d.
+\end{align}
+These absolute errors correspond to the _statistical bias_ in the coefficient estimates.
+Expressed in terms of sample means and covariances, these errors are
+\begin{align}
+  (\#eq:lm-hat-err-1)
+  \text{abs. error}(\hat \alpha) &= \bar d - \hat \beta_d \bar x \\
+  \text{abs. error}(\hat \beta)  &= \frac{s(d,x)}{s^2(x)} = r(d,x) \cdot \frac{s(d)}{s(x)}.
+\end{align}
 
+We are mainly interested in the estimated slope coefficient, $\hat \beta$.
+We see that the absolute error is large when the covariance of $d$ with $x$ is large.
+This error is large in a practical sense when $\hat \beta_{d}$ is large (in magnitude) relative to $\hat \beta_y$ (Equation \@ref(eq:lm-hat)), which occurs when $s(d,x)$ is large relative to $s(y,x)$.
+<!-- Thus the relative error in $\hat \beta$ is large when the covariance of $d$ with $x$ is large relative to the covariance of $y$ with $x$. -->
+A sign error occurs when $s(d,x)$ is larger in magnitude than $s(y,x)$ and of opposite sign.
+
+We can also consider how the measurement error affects the residual variation and the standard error of the slope estimate.
+The residuals of $y$, $z$, and $d$ are related through $\hat \varepsilon_z = \hat \varepsilon_y + \hat \varepsilon_d$.
+It follows that the sample variance of the $z$ residuals equal
+\begin{align}
+  s^2(\hat \varepsilon_{z}) 
+  = s^2(\hat \varepsilon_y + \hat \varepsilon_{d})
+  = s^2(\hat \varepsilon_y) + s^2(\hat \varepsilon_{d}) + 2 s(\hat \varepsilon_y, \hat \varepsilon_{d}).
+\end{align}
+The variance of the $z$ residuals is increased above that of the $y$ residuals when the $d$ residuals are either uncorrelated or positively correlated with the $y$ residuals, but may be decreased when the $y$ and $d$ residuals are negatively correlated.
+An increased residual variance will lead to a larger estimate for $\sigma$ as well as larger standard errors in the slope coefficient $\beta$.
+
+## Specific application to taxonomic bias
+
+These general results can be used to understand how taxonomic bias affects DA analyses that can be expressed as linear regression of log (relative or absolute) abundance.
+
+First, we illustrate for the particular example of absolute DA analysis, where species absolute abundance is measured with the total-abundance normalization method and we assume that the total abundances have been accurately measured.
+Consider the regression of log abundance of species $i$ on a covariate $x$.
+In this case, $y(a)= \log \text{abun}_{i}(a)$ is the actual log abundance of species $i$ and $z(a)=\log \widehat{\text{abun}}_i(a)$ is the log abundance we've measured.
+From Equation \@ref(eq:density-prop-error), the measurement error $d(a)$ due to taxonomic bias in the MGS measurement is
+\begin{align}
+  d(a) 
+  \equiv z(a) - y(a) 
+  = \log \text{efficiency}_i - \log \text{efficiency}_S(a).
+\end{align}
+The absolute error in $\hat \beta$ equals the scaled covariance of $d$ with $x$ (Equation \@ref(eq:lm-hat-err-1)).
+<!-- The first term, $\log \text{efficiency}_i$, is constant across samples; thus it affects $\hat \alpha_d$ (via its effect on $\bar d$) but does not affect $\hat \beta_d$. -->
+In the above expression for $d(a)$, only the second term, $- \log \text{efficiency}_S(a)$, varies across samples and thus affects the covariance, leading to
+\begin{align}
+  \text{abs. error}(\hat \beta)
+    &= - \frac{s[\log \text{efficiency}_S, x]}{s^2(x)}.
+%    = - \frac{r[\log \text{efficiency}_S, x] \cdot s[\log \text{efficiency}_S]}{s(x)}, \\
+\end{align}
+Thus the absolute error in $\hat \beta$ is the negative of the scaled covariance of the log mean efficiency.
+
+Although the absolute error is the same for each species, its practical significance varies.
+Recall from Equation \@ref(eq:lm-hat) that the correct value for $\hat \beta$ (i.e., the estimate without measurement error) is
+\begin{align}
+  \hat \beta &= \frac{s[\log \text{abun}, x]}{s^2(x)}.
+\end{align}
+For species whose log abundance covaries with $x$ much more than the log mean efficiency, the error will be relatively small.
+This situation can occur either because the mean efficiency varies relatively little across samples or because its variation is relatively uncorrelated with $x$.
+For species whose log abundance covaries with $x$ similar or less than the log mean efficiency, the error will be significant.
+A sign error occurs when the species covariance is of _equal sign and smaller in magnitude_ than the covariance of the log mean efficiency.
+
+We can also consider how the standard errors in the slope estimate are affected by bias.
 In our case, the $d$ residuals equal the negative residuals of the log mean efficiency.
@@
 However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \@ref(fig:regression-example).
+
+Now, suppose we were instead performing a DA analysis of a response whose fold error is constant across samples, such as the log ratio of two species.
+For the log ratio of species $i$ and $j$, the error is $d(a) = \log \text{efficiency}_i / \text{efficiency}_j$ and is constant across samples.
+Thus the covariance of $d$ with $x$ is 0 and taxonomic bias causes no error the slope estimate $\hat \beta$, only the intercept estimate $\hat \alpha$.

### 2022-02-03 notebook/_posts/2021-12-15-momspi-regression-diversity/_cut.Rmd
Expand MOMSPI regression analysis
This commit expands the regression analysis, in order to understand the
effect of the particular type of regression and how zeros are handled on
the results. In particular, it now includes a rank-based analysis and
Gamma Poisson regression, and compares the effects of replacing zeros
before or after calibration.

@@
 C
+
+## Wilcoxon - trying to understand dependence on zero replacement
+
+What is going on here?
+
+First, why does calibration not have much effect when zeros are not replaced?
+Hypothesis: The DA results are being driven by presence/absence, which is not affected by calibration.
+
+```{r}
+ps <- ps_set %>% filter(type == 'Observed', zero_replacement == 'None') %>%
+  pull(ps) %>% .[[1]]
+z <- ps %>%
+  prune_taxa(taxa_to_test, .) %>%
+  as_tibble %>%
+  with_groups(.otu, mutate, rank = rank(.abundance) / n())
+```
+
+here - how to test my hypothesis?
+
+see if get similar results in terms of average change in rank whether I look at P/A or not.
+focus on ranks / estimates? or p-values?
+
+thinking about estimates, I predict that a good approximation of the estimate is one-half of the difference in the prevalence in the high vs low diversity group.
+
+```{r}
+prev_wide <- z %>%
+  with_groups(c(.otu, div_group), summarize, prev = mean(.abundance > 0)) %>%
+  pivot_wider(names_from = div_group, values_from = prev)
+```
+
+```{r}
+tmp <- res_div_rank_tidy %>%
+  filter(type == 'Observed', test == 'lm', zero_replacement == 'None') %>%
+  left_join(prev_wide, by = '.otu')
+tmp %>%
+  ggplot(aes(estimate, (High - Low) / 2)) +
+  theme_minimal_grid() +
+  coord_fixed() +
+  geom_abline() +
+  geom_point()
+```
+
+there is a decently good concordance, but not sure if/how to say whether differential prevalence is responsible for most of the difference in rank between groups.
+
+visual check
+
+```{r}
+z %>%
+  ggplot(aes(div_group, rank)) +
+  facet_wrap(~.otu) +
+  scale_y_log10() +
+  geom_quasirandom()
+```
+
+Probs not the best plot, but does seem to indicate that the rank difference between present and absent is much bigger than the difference in average rank among present between the two groups.
+
+**Why does calibration have a big effect on the point estimates when zeros are first replaced with positive values?**
+We know there is a systematic change in the mean efficiency across groups.
+Once all the zeros for a taxon are set to a common value, this common value will (after calibration) become equal to the common value times the eff / mean eff, and hence will follow the (inverse) pattern of the mean eff.
+So the systematic shift in the mean eff will lead to a systematic diff in the ranks of these previously-zero entries.
+
+SHOW for one taxon
+
+NOTE: if this is what is going on, we perhaps can't take the result that calibration has a big effect too seriously. 
+We have a lot of uncertainty about what the 'observed' proportions should be in each sample. 
+If we consider two samples, where the mean efficiency differs by 10X, and the taxon was a zero, then the deterministic zero imputation will say with certainty that (after calibration) the taxon is lower in the sample with the higher mean efficiency.
+But if we have uncertainty greater than 10X in the pre-calibration proportion, we'll have uncertainty after calibration about which sample has the higher value, perhaps to nearly the maximum amount (of a coin flip).
+
+### maybe
+
+**Intuition check: Let's check that the computed mean efficiency doesn't much depend on the zero-replacement strategy.**
+
+# Cut from GP analysis
+
+this is when I hadn't fixed my bug and was like WTF
+
+### Gardnerella vaginalis and Lachnospiraceae BVAB1
+
+Let's see what's going on with the two taxa for which calibration actually leads to higher LFC estimates.
+These taxa are unique in that they have low efficiencies and are frequently the most abundant taxon in a sample.
+(All other common top species are Lactobacillus species and so have high efficiencies).
+
+```{r}
+x <- momspi_div %>%
+  as_tibble %>%
+  left_join(bias %>% enframe('.otu', 'efficiency'), by = '.otu') %>%
+  rename(observed = .abundance) %>%
+  with_groups(.sample, mutate,
+    calibrated = close_elts(observed / efficiency) * sample_sum
+  ) %>%
+  filter(.otu %in% taxa_to_test)
+# taxa_to_plot <- c('Atopobium_vaginae', 'Ureaplasma_cluster23',
+#   'Gardnerella_vaginalis', 'Lachnospiraceae_BVAB1')
+# x1 <- x %>%
+#   filter(.otu %in% taxa_to_plot) %>%
+#   mutate(across(.otu, factor, levels = taxa_to_plot))
+
+gard <- x %>%
+  filter(.otu == 'Gardnerella_vaginalis')
+```
+
+Note how I have to renormalize to keep the calibrated counts summing to 10K.
+Am I sure that the GP respects this? (I think it does by virtue of the offset being relative to the mean efficiency.
+
+```{r}
+gard %>%
+  pivot_longer(c(observed, calibrated), names_to = 'type') %>%
+  ggplot(aes(y = div_group, x = value, color = type)) +
+  facet_wrap(~type, ncol = 1) +
+  scale_x_continuous(
+    trans = pseudo_log_trans(sigma = 10),
+    breaks = c(0, 10, 100, 1e3, 1e4)
+    ) +
+  geom_quasirandom(groupOnX = FALSE)
+```
+
+We can see that possibility for Gard to increase is capped in the calibrated data.
+This observation suggests the reason that the effect is smaller with calibration has to do with the cap on proportions being above 1 (or the rarefied counts being above 10K).
+Why would this problem only become apparent in the GP regression and not the linear model?
+
+I suspect it has something to do with zeros.
+In the lm analysis above (with deterministic zero replacement), calibration adjusts the log-proportion is adjusted by the efficiency (relative to the mean) regardless of whether the observation is a zero or a large count.
+
+Perhaps in the GP analysis, the imprecision of the zeros as measurements of log proportion ends up leading to smaller log-scale calibration adjustments in the latent log-proportion measurement.
+
+Consider a zero observation of Gard in a Lacto-dominated sample, with a mean efficiency that is 10X that of Gard.
+There will be a range of values for the latent log porportion that are consistent with the fitted model, say from 1e-6 to 1e-3.
+After calibration, the range will be shifted upwards, but perhaps it will also be wider---perhaps the upper bound increases by 10X, but the lower bound doesn't chan ge much (very uncertain about this).
+I may need to determine what impact the prior on the intercept has, since when I used the default prior I did not see this unexpected behavior.
+
+What would happen if we computed the change in the log abundance (like lm), but zeros were set to a pseudocount after calibration, rather than before?
+
+```{r}
+gard %>%
+  pivot_longer(c(observed, calibrated), names_to = 'type') %>%
+  with_groups(c(type, div_group), summarize,
+    mean = log(value + 0.3) %>% mean) %>%
+  pivot_wider(names_from = div_group, values_from = mean) %>%
+  mutate(diff = High - Low)
+```
+
+Here, the calibrated LFC is smaller, as in the original lm analysis.
+However, the difference is a bit smaller than the negative LFC of the mean efficiency.
+
+Let's investigate the predicted counts, with and without calibration.
+
+```{r}
+fits <- res_gp %>% filter(.otu == 'Gardnerella_vaginalis') %>% 
+  select(type, fit) %>%
+  deframe()
+cfs <- fits %>% map(coef)
+```
+
+the predicted values from the two fits, before we've adjusted for bias, are
+
+```{r}
+# Low diversity
+cfs %>% map(~exp(.[1]) %>% unname)
+# High diversity
+cfs %>% map(~exp(.[1] + .[2]) %>% unname)
+```
+
+The `Calibrated` model predicts values larger than the 10K read depth; however, that is because this is the value before the sample-specific offsets are accounted for, and these will generally push the value lower.
+
+My guess as to what is going on is that there is a 'compositional bias' that is being relaxed by calibration, which pushes in the direction of beta-hat being higher, and this effect overpowers the push of the mean efficiency difference.
+Can I understand this effect more?
+
+Also, can I understand why it doesn't affect the lm method?
+First, check to see what the predicted values are.
+
+```{r}
+cfs_lm <- res_lm %>% filter(.otu == 'Gardnerella_vaginalis') %>% 
+  select(type, fit) %>%
+  deframe() %>%
+  map(coef)
+# Low diversity
+cfs_lm %>% map(~exp(.[1] + log(1e4)) %>% unname)
+# High diversity
+cfs_lm %>% map(~exp(.[1] + .[2] + log(1e4)) %>% unname)
+```
+
+We can see that the predicted values are lower and don't overshoot the read depth.
+
+But is this really a fair comparison? What if I had used offsets for calibration?
+
+### fitted values (not sure if useful)
+
+```{r}
+gard_fitted <- res_gp %>% filter(.otu == 'Gardnerella_vaginalis') %>% 
+  mutate(
+    data = map(fit, 'data'),
+    fitted = map(fit, fitted)
+  ) %>%
+  select(-fit) %>%
+  unnest(c(data, fitted))
+```
+
+```{r}
+gard_fitted %>%
+  ggplot(aes(fitted, .abundance, color = div_group)) +
+  facet_wrap(~type) +
+  geom_point()
+```
+
+### explore
+
+```{r}
+x <- momspi_div %>%
+  prune_taxa(taxa_to_test, .) %>%
+  as_tibble %>%
+  left_join(bias %>% enframe('.otu', 'efficiency'), by = '.otu')
+taxa_to_plot <- c('Atopobium_vaginae', 'Ureaplasma_cluster23',
+  'Gardnerella_vaginalis', 'Lachnospiraceae_BVAB1')
+x1 <- x %>%
+  filter(.otu %in% taxa_to_plot) %>%
+  mutate(across(.otu, factor, levels = taxa_to_plot))
+```
+
+```{r}
+x2 <- x1 %>%
+  mutate(
+    observed = .abundance,
+    calibrated = .abundance / (efficiency / mean_efficiency)
+  )
+```
+
+```{r}
+x2 %>%
+  pivot_longer(c(observed, calibrated), names_to = 'type') %>%
+  ggplot(aes(y = div_group, x = value, color = type)) +
+  facet_grid(.otu~type) +
+  scale_x_continuous(trans = pseudo_log_trans(sigma = 10)) +
+  geom_quasirandom(groupOnX = FALSE)
+```
+
+```{r}
+x1 %>%
+  ggplot(aes(x = 0.3 + .abundance, y = div_group)) +
+  scale_x_log10() +
+  facet_wrap(~.otu) +
+  stat_slab(slab_type = 'histogram')
+```
+
+```{r}
+x1 %>%
+  ggplot(aes(x = 0.3 + .abundance, y = .otu, fill = div_group)) +
+  scale_x_log10() +
+  stat_slab(slab_type = 'histogram', alpha = 0.5)
+```
+
+One possibility is that the difference has to do with the fraction of zeros; however, there doesn't seem to be an obvious difference between the two pairs of taxa.
+
+Or: It has to do with bimodality more generally
+
+Another possibility: It has to do with taxa who reach large proportions and so are impacted by the compositional effect.
+Gv and LB both reach near the upper limit of possible counts in the high-div group, placing a bound on how high the effect can be, but this is not the case for the other two taxa.
+
+Let's dig into L BVAB1 a bit more.
+
+```{r}
+library(tidybayes)
+
+drws <- res_gp %>% 
+  filter(.otu == 'Lachnospiraceae_BVAB1') %>%
+  mutate(
+    # draws = map(fit, gather_draws, `(Intercept)`, div_groupHigh)
+    draws = map(fit, gather_draws, `(Intercept)`, div_groupHigh, reciprocal_dispersion)
+  ) %>%
+  select(-fit) %>%
+  unnest(draws) %>%
+  group_by(type, .variable)
+drws %>% median_qi(.value)
+drws %>% filter(.variable != 'reciprocal_dispersion') %>% 
+  mutate(across(.value, exp)) %>% median_qi(.value)
+```
+
+Note, I think we have to interpret the intercept for the calibrated model somewhat differently, because of the offset.
+
+Looking at just the Observed, we have 177 * 5.64 being approximately 1000, or 0.1 of the total reads.
+
+The LFC from low to high diversity is around 1.6, corresponding to a FC of around 5.
+
+```{r}
+x2 <- x1 %>%
+  filter(.otu == 'Lachnospiraceae_BVAB1') %>%
+  mutate(
+    observed = .abundance,
+    calibrated = .abundance * efficiency / mean_efficiency
+  )
+x2 %>%
+  ggplot(aes(observed, calibrated, color = div_group)) +
+  scale_x_continuous(trans = pseudo_log_trans(sigma = 10)) +
+  scale_y_continuous(trans = pseudo_log_trans(sigma = 10)) +
+  geom_point()
+```
+
+```{r}
+x2 %>%
+  pivot_longer(c(observed, calibrated), names_to = 'type') %>%
+  ggplot(aes(y = div_group, x = value, color = type)) +
+  facet_wrap(~type) +
+  scale_x_continuous(trans = pseudo_log_trans(sigma = 10)) +
+  geom_quasirandom(groupOnX = FALSE)
+```
+
+TODO: use a density plot so that we account for different numbers of samples per group (or down sample)
+
+The data show a bimodal distribution in both high and low diversity groups, but particularly in the high diversity group.
+The 
+, with a small number of low-div, high LB samples, but these might be too few to be much affecting the mean LFC.
+
+Compare the impact of calibration to the change in the mean efficiency
+
+### Ureaplasma
+
+Let's look at a case where there is a 'big' effect
+
+```{r}
+drws <- res_gp %>% 
+  filter(.otu == 'Ureaplasma_cluster23') %>%
+  mutate(
+    # draws = map(fit, gather_draws, `(Intercept)`, div_groupHigh)
+    draws = map(fit, gather_draws, `(Intercept)`, div_groupHigh, reciprocal_dispersion)
+  ) %>%
+  select(-fit) %>%
+  unnest(draws) %>%
+  group_by(type, .variable)
+drws %>% median_qi(.value)
+drws %>% filter(.variable != 'reciprocal_dispersion') %>% 
+  mutate(across(.value, exp)) %>% median_qi(.value)
+```
+
+```{r}
+x2 <- x1 %>%
+  filter(.otu == 'Ureaplasma_cluster23') %>%
+  mutate(
+    observed = .abundance,
+    calibrated = .abundance * efficiency / mean_efficiency
+  )
+```
+
+```{r}
+x2 %>%
+  pivot_longer(c(observed, calibrated), names_to = 'type') %>%
+  ggplot(aes(y = div_group, x = 0.01 + value)) +
+  facet_wrap(~type) +
+  scale_x_log10() +
+  # scale_x_log10(breaks = 10^seq(-1, 4)) +
+  stat_halfeye()
+```
+
+### MCMC diagnostics
+
+Consider doing for a good subset of the fits to check.
+
+```{r, eval = F}
+f <- res_gp %>% slice(1) %>% pull(fit) %>% .[[1]]
+plot(f, plotfun = 'mcmc_dens_overlay', pars = 'div_groupHigh')
+plot(f, plotfun = 'mcmc_trace', pars = 'div_groupHigh')
+```
+
+```{r, eval = F}
+f <- res_gp %>% filter(.otu == 'Gardnerella_vaginalis') %>% pull(fit)
+f %>% map(summary)
+plot(f[[1]], plotfun = 'mcmc_dens_overlay', pars = 'div_groupHigh')
+plot(f[[2]], plotfun = 'mcmc_dens_overlay', pars = 'div_groupHigh')
+```

### 2022-02-03 notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity.Rmd
Expand MOMSPI regression analysis
This commit expands the regression analysis, in order to understand the
effect of the particular type of regression and how zeros are handled on
the results. In particular, it now includes a rank-based analysis and
Gamma Poisson regression, and compares the effects of replacing zeros
before or after calibration.

@@
   dpi = 300
+  #> cache.path = '_momspi-regression-diversity_cache/'
 )
@@
 library(ggbeeswarm)
+library(ggforce)
+library(ggdist)
 library(cowplot)
@@
 source(here('notebook/_code/momspi-setup.R'))
+bias <- bias_all_vec
 ```
@@
     )
-  }, hash = momspi
-)
-# Make sure no taxa have 0 counts, in case this causes issues later on
+}, dir = '_cache/', file = 'momspi_rare', hash = list(momspi))
+# Make sure no taxa have 0 reads, in case this causes issues later on
 stopifnot(all(momspi_rare %>% taxa_sums > 0))
@@
 ```{r}
-mean_eff <- momspi_rare %>%
- calibrate(bias_all_vec, norm = 'close') %>%
- perturb(bias_all_vec, norm = 'none') %>% 
- sample_sums %>%
- enframe('.sample', 'mean_efficiency')
-
 momspi_rare <- momspi_rare %>% 
@@
     sample_sum = sample_sums(.),
-  ) %>%
-  left_join_sample_data(mean_eff, by = '.sample')
+    mean_efficiency = mean_efficiency(., bias, type = 'observed')
+  )
 ```
@@
   'obs' = momspi_rare,
-  'cal' = momspi_rare %>% calibrate(bias_all_vec)
+  'cal' = momspi_rare %>% calibrate(bias)
   ) %>%
@@
 
-NOTE: Consider saving output so can be used in other notebook files
-
 ## Classify samples into high and low diversity groups
@@
   'obs' = momspi_rare,
-  'cal' = momspi_rare %>% calibrate(bias_all_vec)
+  'cal' = momspi_rare %>% calibrate(bias)
 ) %>%
@@
   ggplot() +
-  geom_point(aes(x = .panel_x, y = .panel_y), alpha = 0.5) +
-  ggforce::geom_autodensity() +
+  geom_autopoint(aes(x = .panel_x, y = .panel_y), alpha = 0.5) +
+  geom_autodensity() +
   geom_bin2d(aes(x = .panel_x, y = .panel_y)) +
-  ggforce::facet_matrix(vars(starts_with('div')), layer.diag = 2, layer.lower = 3)
+  facet_matrix(vars(starts_with('div')), layer.diag = 2, layer.lower = 3)
 ggrastr::rasterize(p, dpi = 300)
@@
   left_join_sample_data(div_wide, by = '.sample') %>%
-  filter_sample_data(div_group %in% c('Low', 'High')) %>%
-  mutate_sample_data(
-    gp_offset_obs = log(sample_sum),
-    gp_offset_cal = log(sample_sum) - log(mean_efficiency),
-  )
+  filter_sample_data(div_group %in% c('Low', 'High'))
 ```
@@
   print
+```
+
+What is the fraction of zeros for the tested species?
+
+```{r}
+momspi_div %>%
+  prune_taxa(taxa_to_test, .) %>%
+  as_tibble %>%
+  with_groups(.otu, summarize,
+    prev = mean(.abundance > 0),
+    frac_zero = mean(.abundance == 0),
+  ) %>%
+  arrange(prev) %>%
+  knitr::kable(digits = 2)
 ```
@@
 }
-res_div_lm <- list(
+res_lm <- list(
   'Observed' = momspi_div %>% transform_sample_counts(~. + 0.5),
   'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.5) %>%
-    calibrate(bias_all_vec)
+    calibrate(bias)
 ) %>%
   map_dfr(stat_test, taxa = taxa_to_test, .id = 'type')
-res_div_lm_tidy <- res_div_lm %>%
-  mutate(across(fit, map, broom.mixed::tidy)) %>%
+res_lm_tidy <- res_lm %>%
+  mutate(across(fit, map, broom::tidy)) %>%
   unnest(fit) %>% 
@@
     mutate(
-      .otu_fct = fct_reorder(.otu, estimate, .fun = min) %>%
+      .otu_fct = fct_reorder(.otu, estimate, .fun = max) %>%
         fct_relabel(function(x) str_replace_all(x, '_', ' '))
@@
 ```{r regression-lm, fig.dim = c(7, 7) * 1.1}
-plot_estimates(res_div_lm_tidy)
+plot_estimates(res_lm_tidy)
 ```
@@
 ```{r regression-lm-alt, fig.dim = c(7, 7) * 1.1}
-res_div_alt_lm <- list(
+res_alt_lm <- list(
   'Observed' = momspi_div_alt,
-  'Calibrated' = momspi_div_alt %>% calibrate(bias_all_vec)
+  'Calibrated' = momspi_div_alt %>% calibrate(bias)
 ) %>%
   map_dfr(stat_test, taxa = taxa_to_test, .id = 'type')
-res_div_alt_lm_tidy <- res_div_alt_lm %>%
-  mutate(across(fit, map, broom.mixed::tidy)) %>%
+res_alt_lm_tidy <- res_alt_lm %>%
+  mutate(across(fit, map, broom::tidy)) %>%
   unnest(fit) %>% 
@@
 
-plot_estimates(res_div_alt_lm_tidy)
+plot_estimates(res_alt_lm_tidy)
 ```
 
-## Interpretion
+### Interpretion
 
@@
 ```{r}
-res_div_lm_tidy %>%
+res_lm_tidy %>%
   select(name = type, .otu, value = estimate) %>%
@@
 ```
+
+## Rank-based analysis
+
+Wilcoxon test and Spearman rho are popular 'non-parametric' approaches to differential abundance.
+Wilcoxon is used when the covariate is binary (case vs control) and Spearman is used when the covariate is continuous.
+Both are approximately equivalent (in terms of p value) to performing a linear regression on rank-transformed data.
+
+For the binary-variable case here, we can use a Wilcoxon test; however, I'll also use a linear model to get confidence intervals.
+
+Zeros: The rank-based tests don't require log transformation and so are ok with zeros. 
+Therefore we have at least three options:
+
+1. leave the zeros
+2. use a 0.5 pseudocount
+3. use the Dirichlet-posterior pseudocount
+<!-- -->
+
+Based on previous experimentation and vague intuition, I suspect that we will see a big difference in the impact of calibration on #1 versus the other two approaches, but only a modest difference between #2 and #3.
+
+Note, the outcome of the rank-based tests on the 'Observed' (uncalibrated) profiles will not depend on the zero replacement strategies, since all zero entries will be set to the same small value and thus still tied for the lowest value (hence giving the same rank).
+But the 'Calibrated' profiles will differ, since the calibrated proportion of a taxon depends on the pre-calibration proportions of all taxa, which will vary from sample to sample.
+
+```{r}
+ps_set <- list(
+  'None' = momspi_div,
+  'Strategy 1' = momspi_div %>% transform_sample_counts(~. + 0.5),
+  'Strategy 2' = momspi_div_alt 
+) %>%
+  map(transform_sample_counts, close_elts) %>%
+  map(filter_sample_data, div_group %in% c('Low', 'High')) %>%
+  enframe('zero_replacement', 'Observed') %>%
+  mutate(Calibrated = map(Observed, calibrate, bias)) %>%
+  pivot_longer(c(Observed, Calibrated), names_to = 'type', values_to = 'ps')
+```
+
+```{r}
+stat_test <- function(ps, taxa, test, ...) {
+  ps %>%
+    prune_taxa(taxa, .) %>%
+    as_tibble %>%
+    with_groups(.otu, nest) %>%
+    mutate(
+      fit = map(data, function(.data) test(data = .data, ...))
+    ) %>%
+    select(-data)
+}
+#> a = stat_test(momspi_div, taxa_to_test, wilcox.test, .abundance ~ div_group)
+#> a %>% slice(1) %>% pull(fit)
+res_rank <- xfun::cache_rds({
+  ps_set %>%
+    mutate(
+      wilcox = map(ps, stat_test, taxa_to_test, wilcox.test, 
+        .abundance ~ div_group),
+      lm = map(ps, stat_test, taxa_to_test, lm, 
+        rank(.abundance)/length(.abundance) ~ div_group),
+    ) %>%
+    pivot_longer(c(wilcox, lm), names_to = 'test', values_to = 'fit') %>%
+    select(-ps) %>%
+    unnest(fit)
+}, dir = '_cache/', file = 'res_rank', 
+hash = list(ps_set, stat_test, taxa_to_test))
+
+res_rank_tidy <- res_rank %>% 
+  mutate(across(fit, map, broom::tidy, conf.int = TRUE)) %>%
+  unnest(fit) %>% 
+  filter(term != '(Intercept)' | is.na(term))
+```
+
+Check agreement between p values -
+
+```{r}
+res_rank_tidy %>%
+  select(zero_replacement:.otu, p.value) %>%
+  pivot_wider(names_from = test, values_from = p.value) %>%
+  ggplot(aes(wilcox, lm)) +
+  scale_x_log10() +
+  scale_y_log10() +
+  coord_fixed() +
+  facet_wrap(zero_replacement ~ type, scales = 'fixed') +
+  geom_abline(color = 'darkgrey') +
+  geom_point()
+```
+
+```{r}
+res_rank_tidy %>%
+  select(zero_replacement:.otu, p.value) %>%
+  pivot_wider(names_from = test, values_from = p.value) %>%
+  ggplot(aes(wilcox, lm)) +
+  scale_x_sqrt() +
+  scale_y_sqrt() +
+  coord_fixed() +
+  facet_wrap(zero_replacement ~ type, scales = 'fixed') +
+  geom_abline(color = 'darkgrey') +
+  geom_point()
+```
+
+See good agreement except at extremely low p values, which we have here due to the very large sample size.
+
+Plot the confidence intervals for the lm test, for the three zero-replacement strategies.
+Sort taxa by 'max' effectively sorts based on the 'Observed' value, which does not depend on the zero-replacement strategy.
+
+```{r regression-rank, fig.dim = c(7, 14) * 1.1}
+res_rank_tidy %>%
+  filter(test == 'lm') %>%
+  mutate(
+    .otu_fct = fct_reorder(.otu, estimate, .fun = max) %>%
+      fct_relabel(function(x) str_replace_all(x, '_', ' '))
+  ) %>%
+  ggplot(aes(y = .otu_fct, x = estimate, color = type)) +
+  facet_wrap(~zero_replacement, ncol = 1) +
+  colorblindr::scale_color_OkabeIto() +
+  geom_vline(xintercept = 0, color = 'grey') +
+  geom_linerange(size = 1,
+    aes(
+      xmin = estimate - 2*std.error,
+      xmax = estimate + 2*std.error
+    )) +
+  geom_point(size = 1.5) +
+  labs(y = 'Species-level OTU', x = 'Change in rank quantile', color = 'Meas. type')
+```
+
+The outcome is as expected - there is a large difference in the effect of calibration depending on whether zero-replacement is used, with a much smaller difference between the two replacement strategies.
+In particular, there is very little effect of calibration when zeros are left as zeros, and a fairly dramatic effect when zeros are set to small positive values.
+Even without zero replacement, however, we see substantial inferential differences for Ureaplasma_cluster23, Corynebacterium_cluster45, and Lactobacillus_gasseri_cluster.
+
+Let's confirm that the Wilcoxon and lm p-values agree for these taxa,
+
+```{r}
+res_rank_tidy %>%
+  filter(zero_replacement == 'None',
+    .otu %in% c('Ureaplasma_cluster23', 'Corynebacterium_cluster45',
+      'Lactobacillus_gasseri_cluster'),
+    ) %>%
+  select(type, test, .otu, p.value) %>%
+  pivot_wider(names_from = test, values_from = p.value)
+```
+
+**Why does calibration have a big effect on the estimates when zeros are first replaced with positive values?**
+There is a relatively straightforward explanation that should make us cautious about interpretting the big effect of calibration too literally.
+Recall that there is a systematic change in the mean efficiency across groups.
+Consider a taxon with a lot of zero values (which is essentially all of them).
+After zero replacement, all of the zeros are set to the same positive proportion, which will be the smallest possible for that taxon.
+Without calibration, these samples will receive the same rank (ties are resolved to the average rank, so the rank will be the number of zeros divided by 2).
+But with calibration, they will all have different proportions, given by the common value multiplied by the efficiency relative to the sample mean efficiency, and and hence will follow the inverse variation in mean efficiency.
+The mean efficiency systematically varies (lower in high-diversity samples), and so the ranks from the previously-zero entries will be lower in the high-diversity samples.
+So the systematic shift in the mean efficiency leads in a rather direct way to a systematic difference in the ranks of these previously-zero entries, where the ranks are lower in the high-diversity group.
+If there are a lot of zeros, this effect will be large and could create a big shift towards lower estimates.
+
+The same thing should be occuring in the analysis of log proportions using `lm()` above---the same argument applies, but with the log instead of the rank transform being applied.
+
+We be cautious about taking the big effect of calibration as a sign that bias has a big effect on the original results.
+We have a lot of uncertainty about what the 'observed' proportions should be in each sample with a zero (or near-zero) value.
+If we consider two samples, where the mean efficiency differs by 10X, and the taxon was a zero, then the deterministic zero imputation will say with certainty that (after calibration) the taxon is lower by 10X in the sample with the higher mean efficiency.
+But if we have uncertainty greater than 10X in the pre-calibration proportion, we'll have uncertainty after calibration about which sample has the higher value, perhaps to nearly the maximum amount (of a coin flip).
+I suspect that an analysis that accounts for this uncertainty would find a smaller expected impact of calibration.
+
+**Why does calibration typically have a minor impact when zeros are not replaced?**
+My leading hypothesis is that the answer has to do with the multimodal distribution of (log) proportions of most taxa.
+Many of these taxa are completely absent (zero) in a large fraction of samples.
+Let's consider a taxon that is zero or has a large proportion, and is zero in a substantial fraction (say over 20%) of samples.
+I suspect that for these taxa the average-rank differences are being driven by differential prevalence (fraction of non-zero samples).
+Differential prevalence is not affected by calibration when we do not replace zeros.
+
+Other taxa may have multiple modes, such as zero, small positive proportion, and large positive proportion; others might be generally present but with small and large modes.
+I suspect we could get a similar non-impact of calibration in this latter case, which would not depend on zero replacement.
+
+**Intuition check: Confirm that the mean efficiency doesn't depend on the zero-replacement strategy.**
+
+```{r}
+add_mean <- function(x, bias) {
+  sample_data(x)$mean_efficiency <- x %>%
+    transform_sample_counts(close_elts) %>%
+    perturb(bias, norm = 'none') %>% 
+    sample_sums
+  x
+}
+
+x <- ps_set %>%
+  filter(type == 'Calibrated') %>%
+  mutate(
+    across(ps, map, sample_data),
+    across(ps, map, as_tibble),
+  ) %>%
+  unnest(ps) %>%
+  pivot_wider(names_from = zero_replacement, values_from = .mean_efficiency)
+```
+
+```{r, eval = F, echo = F}
+x %>%
+  select(None, `Strategy 1`, `Strategy 2`) %>%
+  mutate(across(everything(), log10)) %>%
+  ggplot() +
+  geom_autopoint() +
+  geom_autodensity() +
+  facet_matrix(vars(None, `Strategy 1`, `Strategy 2`), 
+    layer.lower = FALSE, layer.diag = 2)
+```
+
+```{r}
+x %>%
+  select(None, `Strategy 1`, `Strategy 2`) %>%
+  mutate(across(everything(), log10)) %>%
+  cor
+```
+
+As expected, the mean efficiency is unaffected by zero replacement.
+
+## Gamma-Poisson regression
+
+I will try a third approach that circumvents the need for zero replacement.
+Count-based GLMs naturally account for the uncertainty associated with small or zero count observations.
+Gamma-Poisson (or negative binomial) regression with a suitably-chosen offset provides a count-based GLM where the response of the latent linear model is the log proportion.
+Thus we can use it to analyze the effect of sample condition on the average log proportion, treating log proportion as a truly continuous, always positive variable, while fitting the model to the discrete counts (including zeros).
+
+For the final regression analysis, we might consider skipping the rarefy step, which is not needed when doing GP regression.
+
+I'll use the rstanarm package to do the regression. 
+See https://mc-stan.org/rstanarm/reference/neg_binomial_2.html and https://mc-stan.org/docs/2_20/functions-reference/nbalt.html for information about the function and model used.
+
+```{r}
+library(rstanarm)
+options(mc.cores = parallel::detectCores())
+```
+
+Since we have subsampled to a constant 10K reads in all samples, the offset for the 'Observed' case is constant, and rstanarm will throw an error if we use a constant offset.
+Therefore we will not use an offset in this case, but rather adjust the prior mean on the intercept to account for the offset due to read depth.
+Let's check the (geometric) mean abundances across taxa to make sure they fall within this prior,
+
+```{r}
+momspi_div %>%
+  prune_taxa(taxa_to_test, .) %>%
+  as_tibble %>%
+  with_groups(.otu, summarize, across(.abundance, ~gm_mean(. + 0.3))) %>%
+  pull(.abundance) %>%
+  summary
+```
+
+I will set the mean to `log(20) = 3`, or a geometric-mean on the count scale of 20 corresponding, to a proportion of `0.002`, and set the scale to 3.
+Then the central 66% interval will span from counts of around 1 to 400.
+Since we have a larger number of samples, the prior should not have a very large effect on the posterior as long as it is fairly wide.
+
+```{r}
+tbl_gp <- momspi_div %>%
+  prune_taxa(taxa_to_test, .) %>%
+  as_tibble %>%
+  left_join(bias %>% enframe('.otu', 'efficiency'), by = '.otu') %>%
+  with_groups(.otu, nest)
+custom_prior_intercept <- normal(location = log(20), scale = 3)
+```
+
+```{r, echo = F, eval = F}
+res_gp_test <- tbl_gp %>%
+  slice(1) %>%
+  mutate(
+    Observed = map(data, 
+      ~stan_glm(data = .x, iter = 100,
+        .abundance ~ div_group,
+        prior_intercept = custom_prior_intercept,
+        family = neg_binomial_2, seed = 42, algorithm = 'sampling'
+      )
+    ),
+    Calibrated = map(data, 
+      ~stan_glm(data = .x, iter = 100,
+        .abundance ~ div_group + offset(log(efficiency/mean_efficiency)),
+        prior_intercept = custom_prior_intercept,
+        family = neg_binomial_2, seed = 42, algorithm = 'sampling'
+      )
+    ),
+  ) %>%
+  select(-data) %>%
+  pivot_longer(c(Observed, Calibrated), names_to = 'type', values_to = 'fit')
+# check the intercept priors
+res_gp_test
+res_gp_test %>%
+  pull(fit) %>%
+  map(prior_summary)
+```
+
+```{r, eval = T}
+res_gp <- xfun::cache_rds({
+  tbl_gp %>%
+    mutate(
+      Observed = map(data, 
+        ~stan_glm(data = .x, 
+          .abundance ~ div_group,
+          prior_intercept = custom_prior_intercept,
+          family = neg_binomial_2, seed = 42, algorithm = 'sampling'
+        )
+      ),
+      Calibrated = map(data, 
+        ~stan_glm(data = .x, 
+          .abundance ~ div_group + offset(log(efficiency/mean_efficiency)),
+          prior_intercept = custom_prior_intercept,
+          family = neg_binomial_2, seed = 42, algorithm = 'sampling'
+        )
+      ),
+    ) %>%
+    select(-data) %>%
+    pivot_longer(c(Observed, Calibrated), names_to = 'type', values_to = 'fit')
+}, dir = '_cache/', file = 'res_gp', 
+hash = list(tbl_gp, custom_prior_intercept))
+
+res_gp_tidy <- res_gp %>%
+  mutate(across(fit, map, broom.mixed::tidy)) %>%
+  unnest(fit) %>% 
+  filter(term != '(Intercept)')
+```
+
+```{r regression-gp, fig.dim = c(7, 7) * 1.1}
+plot_estimates(res_gp_tidy)
+```
+
+Let's compare the impact of calibration on the slope point estimates to the LFC in the mean efficiency,
+
+```{r}
+diffs <- res_gp_tidy %>%
+  pivot_wider(names_from = type, values_from = c(estimate, std.error)) %>%
+  # pivot_wider(names_from = type, values_from = estimate) %>%
+  mutate(difference = estimate_Calibrated - estimate_Observed)
+
+diffs %>% pull(difference) %>% summary
+```
+
+```{r regression-gp-diffs}
+diffs %>%
+  ggplot(aes(difference)) +
+  geom_dots() +
+  geom_vline(xintercept = mean_eff_lfc) +
+  expand_limits(x = 0)
+```
+
+The mode is negative aligns with the negative LFC of the mean efficiency.
+But 2 to 4 taxa have substantially less negative differences, and two taxa (Gardnerella_vaginalis and Lachnospiraceae_BVAB1) have positive differences.
+
+# Sanity checks
+
+### Offsets with `lm()`
+
+Check that using offsets gives us the same result as using calibrated profiles.
+Note, I'm updating the mean efficiency and sample sums after adding the pseudocounts.
+This should tend to push the mean efficiency torwards more moderate values and so may reduce the differene between groups.
+
+```{r}
+res_lm_offset <- momspi_div %>%
+  transform_sample_counts(~. + 0.5) %>%
+  mutate_sample_data(., 
+    mean_efficiency = mean_efficiency(., bias, type = 'observed'),
+    sample_sum = sample_sums(.)
+  ) %>%
+  #> prune_taxa(taxa_to_test, .) %>%
+  as_tibble %>%
+  left_join(bias %>% enframe('.otu', 'efficiency'), by = '.otu') %>%
+  with_groups(.sample, mutate, 
+    prop = close_elts(.abundance),
+    prop_cal = close_elts(.abundance / efficiency),
+  ) %>%
+  filter(.otu %in% taxa_to_test) %>%
+  with_groups(.otu, nest) %>%
+  mutate(
+    Observed = map(data, 
+      ~lm(data = .x, log(prop) ~ div_group)
+    ),
+    Calibrated = map(data, 
+      ~lm(data = .x, log(prop_cal) ~ div_group)
+    ),
+    Observed_offset = map(data, 
+      ~lm(data = .x, log(.abundance) ~ div_group + offset(log(sample_sum)))
+    ),
+    Calibrated_offset = map(data, 
+      ~lm(data = .x, 
+        log(.abundance) ~ div_group + offset(log(sample_sum * efficiency /
+            mean_efficiency)),
+      )
+    ),
+  )
+res_lm_offset_tidy <- res_lm_offset %>%
+  select(-data) %>%
+  pivot_longer(c(starts_with('Observed'), starts_with('Calibrated')), 
+    names_to = 'type', values_to = 'fit') %>%
+  mutate(across(fit, map, broom::tidy)) %>%
+  unnest(fit) %>% 
+  filter(term != '(Intercept)')
+```
+
+```{r, eval = F, echo = F}
+plot_estimates(res_lm_offset_tidy)
+```
+
+```{r}
+res_lm_offset_tidy %>%
+  select(.otu, type, estimate) %>%
+  pivot_wider(names_from = type, values_from = estimate) %>%
+  as_matrix(rownames = .otu) %>%
+  cor
+```
+
+### Zeros with `lm()`
+
+See what happens if use pseudocounts after calibration, instead of before.
+
+```{r}
+stat_test <- function(ps, taxa) {
+  ps %>%
+    filter_sample_data(div_group %in% c('Low', 'High')) %>%
+    transform_sample_counts(close_elts) %>%
+    prune_taxa(taxa, .) %>%
+    as_tibble %>%
+    with_groups(.otu, nest) %>%
+    mutate(
+      fit = map(data,
+        ~lm(log(.abundance) ~ div_group, data = .x)
+      ),
+    ) %>%
+    select(-data)
+}
+res_lm_after <- list(
+  'Observed' = momspi_div,
+  'Calibrated' = momspi_div %>% calibrate(bias, norm = 'keep')
+) %>%
+  map(transform_sample_counts, ~. + 0.5) %>%
+  map_dfr(stat_test, taxa = taxa_to_test, .id = 'type')
+res_lm_after_tidy <- res_lm_after %>%
+  mutate(across(fit, map, broom::tidy)) %>%
+  unnest(fit) %>% 
+  filter(term != '(Intercept)')
+```
+
+```{r regression-lm-after, fig.dim = c(7, 7) * 1.1}
+plot_estimates(res_lm_after_tidy)
+```
+
+As expected, see smaller effects of calibration for many taxa.
+Interesting to see that some taxa, including L iners and Ureaplasma, where there remains a large effect.
+It remains that the calibrated estimates are always lower; I guess that it remains a mathematical fact that the directional change must be the same for all taxa.
+
+# Session info {.appendix}
+
+<details><summary>Click for session info</summary>
+```{r, R.options = list(width = 83)}
+sessioninfo::session_info()
+```
+</details>

### 2022-02-03 case-studies.Rmd
Revise text on MOMSPI regression analysis
Update main text and SI figure to refer to the gamma-Poisson regression
analysis, plus minor fixes.

@@
 Since these analysis are both based on proportions, bias could in principle impact their results.
-We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for^[The two sets of analyses can be seen at the following links: [negative-binomial regression analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html); [priority-effects analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html)].
+We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for.^[The two sets of analyses can be seen at the following links: [negative-binomial regression analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html); [priority-effects analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html)]
 To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
@@
 These shifts typically result in substantial fold changes in mean efficiency (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)) and can cause spurious fold changes in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).
+Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance would be more extreme for vaginal microbiome primer sets that fail to amplify _Gardnerella_ (@graspeuntner2018sele).
 
@@
 Patient metadata was not available due to privacy restrictions; we therefore sought a clinically relevant covariate to use in a regression analysis that could be determined directly from the microbiome profiles and that we _a priori_ expected to be associated with the proportions of _Lactobacillus_ and _Gardnerella_.
-Alpha diversity metrics such as species richness, the Shannon index, and the (Inverse) Simpson index have been repeatedly found to be strongly positively associated with bacterial vaginosis (BV; @srinivasan2012bact, @cartwright2018mult).
-@cartwright2018mult found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100% and specificity of 85.1%.
+Alpha diversity metrics such as species richness, the Shannon index, and the (Inverse) Simpson index have been repeatedly found to be strongly positively associated with bacterial vaginosis (BV; @srinivasan2012bact, @cartwright2018mult).^[@cartwright2018mult found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100% and specificity of 85.1%.]
 In addition, it is commonly observed that samples from women with and without BV that are dominated by _Lactobacillus spp._ tend to have higher diversity, whereas samples dominated by _Gardnerella_ tend to have lower diversity.
@@
 We split samples into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
-We then estimated the LFC in proportion from low- to high-diversity samples for a 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles.
-As expected, the mean efficiency was higher in low-diversity samples due to a larger fraction of samples dominated by _Lactobacillus_ and a lower fraction with samples dominated by _Gardnerella_.
-The decline in mean efficiency in the high diversity group caused a concomitant increase in the LFCs of all species when bias was not accounted for.
-This error due to bias resulted in serious magnitude errors in nearly all species and sign errors in over one third, with errors of both types seen for several clinically relevant species.
+We then estimated the LFC in proportion from low- to high-diversity samples for 30 species passing a basic prevalence threshold, with and without bias correction.
+A large fraction of entries in the species count matrix were small counts or zeros.
+We used therefore used gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions while accounting for uncertainty from the random sampling of reads.
+We accounted for bias by including an offset term in the linear model equal to the log ratio of the species' efficiency to the sample mean efficiency.
 
-### Notes
-
-- It might be cleaner story-wise to redo the diversity partitioning using the threshold identified by @cartwright2018mult
-- For the DA analysis of proportions vs.
-diversity, I tried other types of DA analysis methods (Gamma-Poisson regression and regression on ranks, similar to Wilcoxon test) to understand how the impact of bias differs across methods.
-In each case bias has a big effect but the species whose estimates are most affected varies.
-I would need to revise and verify this analysis before including its results.
-One reason for doing so would be to show that our qualitative finding about how bias distorts results applies beyond the basic linear regression model.
-- Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.
+The mean efficiency was on average lower in high-diversity samples due to a reduction in _Lactobacillus_ dominance (LFC of -1.6, or a 4.8X reduction). 
+Our results from the previous section therefore suggest that accounting for bias should reduce the estimated LFC for each species by about this same amount.
+We found that, as expected, bias correction reduced the LFC of each species, though the effect was variable and typically somewhat smaller.
+Bias correction led to substantial magnitude changes for several species and sign changes in 3 species; however, the relative difference between the corrected and uncorrected LFCs was modest for the 19/30 species with the largest LFCs of over 2 (corresponding to a FC of 7.4X) even after correction.
+These large LFCs occur in species that rarely if ever reach high proportions and so are likely driven by the sum-to-one constraint: The high diversity communities are those not dominated by _Lactobacillus_ spp.
+(or to a lesser extent _G.
+vaginalis_ and BVAB1).
+Therefore it is not clear how well it will generalize to regressions on other covariates.
 
@@
 The impact of bias can depend on protocol, biological system, and type of DA analysis being done.
-Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of section X help suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.
+Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of Section \@ref(differential-abundance) suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.
 

### 2022-02-03 supplemental-figures.Rmd
Revise text on MOMSPI regression analysis
Update main text and SI figure to refer to the gamma-Poisson regression
analysis, plus minor fixes.

@@
   "notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity_files", 
-  "figure-html5/regression-lm-alt-1.svg"
+  "figure-html5/regression-gp-1.svg"
 ) %>%
@@
 
-(ref:cap-momspi-regression) **Bias distorts log fold changes (LFCs) in species proportions in a regression analysis of vaginal microbiome samples from the MOMS-PI study.** Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
-The LFC in proportion from low- to high-diversity samples was estimated for 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles following a simple zero-replacement procedure.
-Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95% confidence intervals for each species.
-The difference between the calibrated and observed estimate for each species equals the negative LFC in mean efficiency.
+(ref:cap-momspi-regression) **Bias inflates estimated log fold changes (LFCs) in species proportions with diversity in vaginal microbiome samples from the MOMS-PI study.** Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
+The LFC in proportion from low- to high-diversity samples was estimated for 30 common species using gamma-Poisson regression with and without bias correction.
+Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95% Bayesian credible intervals for each species.
 

### 2022-02-03 case-studies.Rmd
Draft gut-microbiome case study

@@
 
+## Human gut microbiome
+
+The vaginal microbiome is characterised by low species diversity within and across samples, so that variation in efficiency among a few dominant species could drive large variation in the mean efficiency among samples.
+The human gut microbiome has substantially greater diversity both within and across samples, which we should expect to result in lower variation in mean efficiency.
+The reason is that the mean efficiency is a weighted average of efficiencies of individual species, and so will tend to vary less the more species that are averaged over.
+Here we use publically available gut microbiome profiles to consider how the distinct dynamics of the gut microbiome impact the variation in the mean efficiency and its potential to distort DA results.
+
+First we considered variation in healthy subjects and asked how the distinct structures of vaginal and gut communities impacted mean efficiency variation.
+We considered vaginal and stool samples from the original Human Microbiome Project (@huttenhower2012stru); these and other results in this case study are based on species-level bacterial relative-abundances produced by Metaphlan3 and made available through the curatedMetagenomicData R package (CITE).
+We simulated taxonomic bias with a geometric standard deviation of 3, with and without phylogenetic conservation, 
+by sampling from a multivariate normal distribution, with efficiencies either IID or with a covariance equal to the phylogenetic covariance matrix to simulate bias as an evolving trait.
+We then compared the distribution of mean efficiency across samples.
+We also considered how species-level diversity patterns could help explain this variation.
+
+As expected, the mean efficiency varied substantially less in the higher-diversity gut microbiome samples than in the vaginal microbiome samples (median geometric standard deviation of X versus Y); yet it remained non-negligible.
+There are several potential reasons why.
+First, although richness is typically around 100 bacterial species in the Metaphlan profiles, the order-2 effective number of species ($^2D$; equivalent to the Inverse Simpson diversity index) is typically only around 10 and often lower.
+This number is more relevant when we consider the reduction in variation when computing a weighted average over species (Appendix; standard deviation in mean efficiency scales as $1 / \sqrt(^2D)$  when efficiencies are IID).
+A order-2 diversity that is much lower than species richness arises from the skewed distribution of abundances within a gut microbiome sample, with the most abundant 1-2 species often forming a majority of reads.
+It is these high-proportion species, captured by the order-2 diversity, that ultimately drive the variation in mean efficiency, not the large number of low-proportion species that drive the richness.
+A separate but related reason is that, although the total number of species across gut microbiomes is large, a small number of species form the vast majority of reads and tend to be the most abundant in individual samples.
+The overall effect is similar, though less extreme, as that of the dominance of a few key species in the vaginal microbiome: Variation in the efficiencies among these species has an outsized impact on the mean efficiency, again leading to a weakening of the averaging affect when we consider variation across samples.
+
+A notable aspect of gut microbiome variation, shared with with the vaginal microbiome, is that the most abundant species tend to come from a small set of groups.
+In the HMP1 gut data, these are various Bacteroides and Prevotella species; in other gut datasets, Gram positives such as X and Y are also common.
+In patients with diarrhea or infections, we may see dominance of Enterobactericeae like E. coli and Klebsiella spp.
+More generally, there is phylogenetic conservation at the level of ecological traits.
+There is also reports of phylogenetic conservation at the level of the species efficiency for a given protocol.
+An implication is that we may have less 'effective diversity' from the perspective of variation in the mean effiency.
+
+(Although this was not a big effect in the simulation)
+
+Might we see correlations between mean efficency and covariates?
+One way is for a small number of taxonomic groups to cause major shifts, and also have conserved efficiencies.
+An example occurs in an dataset we examined of accurate diarrhaea in children from Dhakka.
+Stool from children experiencing accute diarrhea were overwhelmingly dominated by Preoteobacteria, whereas stool from healthy controls were dominated by a roughly even mixture of Actinobacteria, Firmicutes, and Bacteroides.
+The Proteobacteria species in the diarrheaic stool were dominated by E. coli and to a lesser extent Klebsiella spp, all in the Enterobactericeae family.
+Hence a Gram+/- bias, a bias favoring Enterobactericeae, or a bias favoring E. coli would all be capable of driving systematic variation in the mean efficiency in this study.
+
+On the other hand, in other datasets we examined, the variation in phylum proportions that occured was primarily not associated with the clinical covariate.
+In these cases, a Gram or phylum-level bias might primary add noise rather than systematic error, depending on the sample size of the study and the magnitude of true effects.
+
+(Perhaps put this in the conclusion.)
+There is another reason to consider that the effect of bias should not be dismissed in the higher-diversity natural ecosystems like the gut.
+So far we have only considered the variation in mean efficiency; however, it is the variation in mean efficiency relative to the true effect sizes that determines the signficance of the error.
+In conditions where the mean efficiency varies less, we should also expect lower true effect sizes, so it isn't at all obvious that we should be less worried.
+
 ## Microbial growth in marine sediments

### 2022-02-03 introduction.Rmd
Edit abstract

@@
 Measurement error caused by taxonomic bias is generally ignored in such _differential abundance (DA) analyses_, likely due to a combination of 1) a widespread belief that protocol standardization is sufficient to yield accurate differences between samples and 2) a lack of practical solutions for addressing it.
-Using theoretical arguments and analysis of real and simulated experiments, we show that taxonomic bias can create error in DA results that is sufficient to lead to scientifically erroneous conclusions.
-This error, however, varies by experimental context and analysis method.
+We use theoretical arguments and analysis of real and simulated experiments to analyze the impact of taxonomic bias on relative and absolute DA analyses.
+We show that taxonomic bias does in fact create inferential errors for a class of popular DA methods; however, whether these errors are biologically significant depends on experimental context.
 We present six approaches to addressing the error, each suited to different experimental questions and systems.

### 2022-02-04 case-studies.Rmd
Revise fungal case study
Revised based on BJC's edits and comments on 2022-02-01.

@@
 
-Measurement of control communities along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis (@mclaren2019cons).
-Gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since, unlike most natural ecosystems, it is possible to assemble 'mock communities' containing all species in known relative abundances.
+The taxonomic bias species present in 'mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (@mclaren2019cons).
+In practice it is difficult to construct control communities that span the many species present in complex natural communities.
+However, gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since, unlike most natural ecosystems, it is possible to assemble mock communities containing all species in known relative abundances.
 
-@leopold2020host are the first to use mock-based calibration in a gnotobiotic microbiome experiment.
 In a study of host-commensal-pathogen interactions, @leopold2020host inoculated plants with 8 commensal fungal species and subsequently exposed plants to a fungal pathogen.
-To investigate the joint effect of colonization order and host genotype, plants varied in source location and in which commensal colonized first.
-The authors used ITS amplicon sequencing to measure communities before and after pathogen infection, along with mock communities with quantified genome concentrations of DNA from the 9 species.
-These DNA mock communities allowed the authors to directly measure the bias due to the sequencing workflow following DNA extraction and correct for it when analyzing the primary experimental samples following the method of @mclaren2019cons.
-Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification and the substantial ribosomal copy-number-variation (CNV) in fungi (@lofgren2019geno).
-<!-- TODO: add citations -->
+The authors used ITS amplicon sequencing to measure communities before and after pathogen infection.
+Motivated by the substantial ribosomal copy-number variation (CNV) in fungi (@lofgren2019geno), the authors also performed control measurements of mock communities that they constructed from quantified genomic DNA of the 9 species in the experiment; these controls were used to measure taxonomic bias with the method of @mclaren2019cons.
 The authors found a 13X difference between the most and least efficiently measured commensal, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
-We reanalyzed this experiment to understand the impact that bias might have had on the study's analyses were it not accounted for.
 
-@leopold2020host used the ITS sequencing profiles of the pre-infection samples to understand the impact of host genotype and an experimental treatment---which species was allowed to colonize first---on the commensal community composition prior to infection.
-Two DA analyses were conducted to address these questions.
-First, the authors used linear regression of log species proportion (via negative-binomial regression to account for random read-sampling error) to ask whether host genotype, treatment, and their interaction significantly impacted overall community composition.
-Second, the authors quantified the strength of priority effects---the advantage of being allowed to colonize first---in different host genotypes by the LFC in the proportion of a species when allowed to colonize first versus later with the majority of commensals.
+@leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gain by a species from being allowed to colonize first.
+Both analyses were based on fold changes in species proportions and so in principle were sensitive to taxonomic bias.
+To ensure the results were accurate, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.
 
-Since these analysis are both based on proportions, bias could in principle impact their results.
-We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for.^[The two sets of analyses can be seen at the following links: [negative-binomial regression analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html); [priority-effects analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html)]
-To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
-The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.62X and a geometric standard deviation of just 1.05X (SI Figure \@ref(fig:leopold2020host-variation)).
-<!-- Explanation -->
-Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
-This answer relates to the fact that the mean efficiency is a weighted _arithmetic_ average of species proportions and so is insensitive to large multiplicative variation in low-proportion species.
-The pre-infection samples in the @leopold2020host study are always dominated by the three species with the highest efficiencies, whose efficiencies vary only by 3X (and just 1.5X between the two most dominant).
-The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions $\ll 1$, even after bias correction) and hence don't significantly affect the mean efficiency.
-<!-- Conclusion -->
-Because the log mean efficiency varies much less than the log proportions for each taxon, there is little room for it to distort the results of DA analyses based on variation in log proportions.
+Calibration had negligible impact on the results (personal communication with Devin Leopold and confirmed by our own reanalysis).
+To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \@ref(fig:leopold2020host-variation)).
+Despite the 13X variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62X and a geometric standard deviation of 1.05X.
+This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
+But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3X and by just 1.5X between the two most dominant.
+The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
+Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual taxa, it had a negligible impact on the inferred fold changes and the DA analyses based on them.
 
-<!-- Notes: -->
-<!-- - Consider whether to use log-transformed abundances and efficiencies throughout, and standard (non-geometric) ranges and standard deviations -->
-<!-- - Consider connecting back to the previous section and reminding of the condition for the variation in mean efficiency to impact regression analyses -->
-<!-- - Consider comparing the correlations among taxa with covariates; or at least looking at that of the mean efficiency, since correlation is also an important piece of the requirement -->
-<!-- - The priority-effect analysis isn't done as a regression, though it could have been (but would have been a slightly different calculation).
-I might want to explain this more here and in the notebook post.
--->
-<!-- - I don't consider/mention whether there are actually 'significant' DA results; that would perhaps be a useful piece of the story.
-Or: I should perhaps say what the result was of each of the two analyses, so it is more concrete what results are not changing.
--->
-<!-- - Consider expanding on the impact of bias, using Penicillium as an example: It's proportion is generally ~6X lower without correction, but its changes across samples are not affected by correction -->
+We performed an additional DA analysis on the data from @leopold2020host to investigate whether any commensals increased in absolute concentration in response to infection.
+The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
+Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5X to 5.2X.
+We used simple linear regression to measure the LFC in commensal species proportions pre- to post-infection, with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
+All commensals tended to decrease in proportion post-infection, which is expected given the pathogen’s growth to most abundant community member and the sum-to-one constraint of species proportions.
+However, when taxonomic bias was ignored the magnitude of the post-infection decrease was overestimated, by an amount corresponding to the inverse change in log mean efficiency.
+This magnitude error was substantial for the commensal-host pairs with the weakest changes; in several cases, LFC estimates with 95% confidence intervals below 0 became indistinguishable from 0 after calibration.
 
-We additionally considered the potential for bias to impact analysis of how commensal taxa responded to infection.
-In particular, it is interesting to consider whether any commensals are able to increase in absolute concentration in response to infection.
-The experiment performed by @leopold2020host does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration measured using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
-It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
-The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, raising the mean efficiency of post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
-Across different host genotypes, the average increase in mean efficiency ranges from 2.5X to 5.2X. 
-We used simple linear regression to measure the average change in log proportion of each commensal for each host genotype, with and without bias correction.
-Regardless of whether bias is accounted for, the log proportion of each commensal decreased, which is unsurprising given the pathogen's growth and the sum-to-one constraint of species proportions.
-Without bias correction, however, the LFCs were lower by an amount corresponding to the inverse change in log mean efficiency.
-The magnitude errors created were substantial for a large fraction of commensal-host pairs; in several cases, the LFC estimates without bias correction were clearly negative, whereas the estimates with bias correction were close to 0.
-
-Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
-We consider two scenarios.
-_Scenario 1:_ Total genome concentration in each sample is perfectly known and used to measure the concentrations of individual taxa via Equation \@ref(eq:density-prop-meas).
-In this case, bias in the MGS measurements would create absolute errors in the LFCs for genome concentration identical to those for proportions.
-The scientific error, however, may be much worse.
-Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where some taxa that persist or increase in abundance instead appear to decrease.
-_Scenario 2:_ Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-abundance normalization method.
-In this case, we should expect the increase in mean efficiency to have little to no impact on LFCs, since it would be offset by the increase in mean efficiency of the total-concentration measurement.
-Moreover, since the total ITS concentration and ITS sequencing are both performed on the extracted DNA, we can also expect the effect of taxonomic bias caused by DNA sequencing to be offset.
-Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-abundance measurement is made prior to normalization.
-
-<!-- cut text -->
-<!-- - Genome concentration was chosen as the measure of abundance, due to the tendency for fungi to grow in forms where cells are not easily demarcated. -->
-<!-- - The original study focused on how host genotype and the pre-infection community affected disease severity after challenge with a pathogen, and did not analyze the impact of infection on commensals.
--->
-<!-- - The post-colonization samples were not analyzed by @leopold2020host in part because the reads from post-colonization samples tended to be dominated by the pathogen, giving little ability to infer the relative abundance of the rarer and lower efficiency species.
--->
-<!-- - This was presumably due to high pathogen growth and aided by the pathogen's post-extraction efficiency being substantially larger than the other species; the high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples (REF SI/FIG).
--->
+Although @leopold2020host did not include absolute-abundance measurements, we can consider the impact taxonomic bias would have on an absolute DA analysis in a simple scenario in which total genome concentration of each pre- and post-infection sample is perfectly known.
+The total genomic concentrations of each sample, combined with the species proportions measured by MGS, can be combined into a measurement of species concentrations using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
+In this case, the bias in the MGS measurements will create absolute errors in the estimated LFCs of genome concentration of equal magnitude to the errors in the LFC estimates for proportions.
+The scientific error, however, might become worse.
+Suppose that the species whose calibrated proportions appeared to hold steady or slightly decrease actually increased in absolute abundance by around 2X.
+As bias shifts our estimates downwards by 2.5X to 5.2X, we would instead conclude that these taxa decreased.
 

### 2022-02-06 notebook/_posts/2022-01-20-vieirasilva2019quan/vieirasilva2019quan.Rmd
Add gut microbiome analyses

---
title: "Explore Vieira-Silva et al (2019) QMP profiles"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - vieirasilva2019quan
date: 2022-01-20
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300
)
```

# Setup

## R setup

```{r libraries}
library(here)
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
```

```{r ggplot}
theme_set(theme_cowplot())
```

## Data setup

```{r}
data_path <- here('notebook/_data/vieirasilva2019quan')
dir_create(data_path)
# Supplementary tables from the nature website; contains the metadata
fn <- path(data_path, 'supplementary-data-1.xlsx')
if (!file_exists(fn)) {
  download.file(
    'https://static-content.springer.com/esm/art%3A10.1038%2Fs41564-019-0483-9/MediaObjects/41564_2019_483_MOESM3_ESM.xlsx',
    fn
  )
  file_chmod(fn, '-w')
}
# QMP proiles from the Raes lab website
fn <- path(data_path, 'qmp-matrix-genus.tsv')
if (!file_exists(fn)) {
  download.file('http://raeslab.org/software/QMP2/QMP.matrix.tsv', fn)
  file_chmod(fn, '-w')
}
```

Note, there is an extra value in the BMI column of Table S1 (Excel spreadsheet).

```{r}
otu <- path(data_path, 'qmp-matrix-genus.tsv') %>%
  read.table() %>%
  as('matrix') %>%
  otu_table(taxa_are_rows = FALSE)
sam <- path(data_path, 'supplementary-data-1.xlsx') %>%
  readxl::read_excel(sheet = 'Table S1', range = 'A2:L174', na = "NA")
original_column_names <- names(sam)
ps <- phyloseq(
  otu, 
  sam %>% janitor::clean_names() %>% sample_data
)
stopifnot(setequal(sample_names(ps), sample_names(otu)))
sam <- ps %>% sample_data %>% as_tibble %>% glimpse
```

### Taxonomy

how can I get taxonomy and/or phylogeny for the genus identifiers?

First, create a file with the RDP 16 taxonomy from the DADA2 training file

```
zcat rdp_train_set_16.fa.gz | grep '^>' | sort | uniq -u | sed -e 's/^>//' -e 's/;$//' - > rdp_train_set_16_headers.txt
```

```{r}
cns <- c('domain', 'phylum', 'class', 'order', 'family', 'genus')
cts <- 'cccccc'
# x <- path('~/data/16s-taxonomy-databases/dada2/rdp-trainset-16',
#   'rdp_train_set_16_headers.txt') %>%
#   read_delim(delim = ';', col_names = cns, col_types = cts)
x <- path('~/data/16s-taxonomy-databases/dada2/rdp-trainset-16',
  'rdp_train_set_16_headers.txt') %>%
  read_lines %>%
  enframe %>%
  select(-name) %>%
  separate(value, into = cns, sep = ';', fill = 'right') %>%
  glimpse
```

```{r}
genera <- taxa_names(ps) %>% 
  str_subset('unclassified', negate = TRUE)
genera %in% x$genus
setdiff(genera, x$genus)

unclassified_taxa <- taxa_names(ps) %>% 
  str_subset('unclassified') %>%
  str_extract('(?<=unclassified_).+')
unclassified_taxa %in% x$family
unclassified_taxa %in% x$order
unclassified_taxa %in% x$phylum
unclassified_taxa %in% x$domain
```

try again

```{r}
rnks <- c('domain', 'phylum', 'class', 'order', 'family', 'genus')
x <- path('~/data/16s-taxonomy-databases/rdp/', 
  'RDPClassifier_16S_trainsetNo16_rawtrainingdata',
  'trainset16_022016_headers.txt') %>%
  read_tsv(col_names = c('id', 'path'))
y <- x %>%
  separate(path, c('root', rnks), sep = ';', fill = 'right')

```

ok, giving up..

hmm - check to see if can get from the rrndb tables?
https://rrndb.umms.med.umich.edu/static/download/

# Explore

I think that the QMP profiles are such that the sum within a sample should equal the reported cell counts per gram in the sample data; let's check.

```{r}
all.equal(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
```

This small difference may be down to rounding error, since it seems that the entires in the QMP matrix are rounded to integers.

```{r}
otu %>% c %>% {all.equal(., round(.))}
sam$average_faecal_cell_count_cells_g %>% {all.equal(., round(.))}
```

There are `r nsamples(ps)`, with the following disease breakdown:

```{r}
sam %>% count(diagnosis)
```

Can we recover counts?
We can't recover the raw counts without using the CNV information.
But setting that aside for now, let's confirm that we understand the normalization that was used -

```{r}
x <- sam %>%
  add_column(sample_sum = sample_sums(ps))
```
```{r}
x %>% pull(sample_sum) %>% summary
x %>% transmute(sample_sum / average_faecal_cell_count_cells_g) %>% pull(1) %>% summary
```

(we already knew this - the samples sum up to the cell density).

# Reproduce paper analysis

Let's try to reproduce the results from the paper.

> Applying QMP on the combined PSC/IBD/mHC data set, we identified 11 genera as significantly correlat-ing  with  faecal  calprotectin  concentrations  (Spearman  correlation,  n= 149, FDR < 0.05; Fig. 3a and Supplementary Table 6). Stool mois-ture content had 45 associated taxa (Spearman correlation, n= 143, FDR <    0.05;  Supplementary  Table  6).  Combining  both  variables,  only  6  out  of  11  positive(+)  or  negative(−)  calprotectin  associations  remained significant when controlling for faecal water content vari-ation: Anaerostipes(−), Escherichia(+), Fusobacterium(+), Gemmiger(−), Streptococcus(+)  and  Veillonella(+)  (nested  linear  model  on  QMP  rank  abundances,  n= 133,  FDR < 0.05;  Fig. 3a  and  Supplementary  Table  6).

First, let's look for taxa associated with faecal calprotectin concentrations

```{r}
x <- ps %>% as_tibble
```

```{r}
y <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  as_tibble %>%
  with_groups(.otu, nest) %>%
  mutate(
    spearman_test = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman', # exact = FALSE, continuity = TRUE,
        )),
    spearman_test = map(spearman_test, broom::tidy)
  ) %>%
  select(-data) %>%
  unnest(spearman_test) %>%
  mutate(p.bh = p.adjust(p.value, method = 'BH'))
```

```{r}
y %>% arrange(p.bh) %>% print(n=20)
```

```{r}
study_res <- path(data_path, 'supplementary-data-1.xlsx') %>%
  readxl::read_excel(sheet = 'Table S6', na = "NA")
study_res %>%
  select(Genera, starts_with('Calprotectin') & ends_with('QMP'))
y %>% arrange(p.bh) %>% filter(!str_detect(.otu, 'unclassified')) %>%
  select(.otu, estimate, p.value, p.bh) %>%
  print(n=15)
```

These estimates match, but the p values (and their ranking) differs.

figure this out; then try to recapitulate using rank regression.

Note, a decrease FDR-corrected p values can be expected if the FDR correction was done _after_ discarding the OTUs without genus classifications.

can check the code at https://github.com/raeslab/QMP/

```{r}
study_res %>%
  select(Genera, starts_with('Calprotectin') & ends_with('QMP')) %>%
  arrange(4) %>%
  print(n=15)
y %>% 
  arrange(p.bh) %>% 
  filter(!str_detect(.otu, 'unclassified')) %>%
  mutate(
    p.bh = p.adjust(p.value, method = 'BH'),
    across(starts_with('p.'), round, digits = 4)
  ) %>%
  select(.otu, estimate, p.value, p.bh) %>%
  print(n=15)
```

ahhh - ok their's is sorted differently, which makes sense since it has many tests in one table. and the p values need to be rounded to 4 digits. then it seems like we're good.

next:

- also do for moisture, and perhaps RMP, to check these results as well
- confirm I can get similar result with linear model on ranks
<!--  -->

visually inspect

```{r}
top_hits <- y %>% 
  filter(!str_detect(.otu, 'unclassified')) %>%
  slice_min(p.bh, n = 9)
x <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  prune_taxa(top_hits$.otu, .) %>%
  as_tibble

x %>%
  mutate(across(.otu, factor, levels = top_hits$.otu)) %>%
  with_groups(.otu, mutate,
    across(c(faecal_calprotectin_mg_g, .abundance), rank),
  ) %>%
  # mutate(across(c(faecal_calprotectin_mg_g, .abundance), rank, ties.method = 'average')) %>%
  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
  facet_wrap(~.otu) +
  geom_point() +
  stat_smooth(method = 'lm')
```

It seems like zeros or below-detection-limit values are largely responsible for the observed results.

```{r}
x %>%
  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~.otu) +
  geom_point()
```

### spearman versus linear modeling

NOTE: The results differ when I first omit the samples with missing calprotectin measurements, or not.
perhaps because the ranks of the abundances differ?

```{r}
z <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  as_tibble %>%
  with_groups(.sample, mutate, proportion = close_elts(.abundance)) %>%
  with_groups(.otu, mutate, 
    across(c(.abundance, proportion, faecal_calprotectin_mg_g), 
      rank, ties.method = 'average',
      .names = '{.col}_rank')
  ) %>%
  with_groups(.otu, nest) %>%
  mutate(
    spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman', na.action = na.omit)),
    lm = map(data, 
      ~lm(data = .x, .abundance_rank ~ faecal_calprotectin_mg_g_rank,
        na.action = na.omit)),
  ) %>%
  select(-data) %>%
  pivot_longer(c(spearman, lm), names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
z1 <- z %>% filter(test == 'spearman' | term != '(Intercept)')
```

NOTE: The p values are nearly identical as long as I first drop the samples with missing data.
But not if I don't.

```{r}
z1 %>%
  select(.otu, test, p.value) %>%
  pivot_wider(names_from = test, values_from = p.value) %>%
  ggplot(aes(spearman, lm)) +
  geom_abline(color = 'darkgrey') +
  scale_x_log10() +
  scale_y_log10() +
  geom_point()
```

```{r}
z1 %>%
  select(.otu, test, estimate) %>%
  pivot_wider(names_from = test, values_from = estimate) %>%
  ggplot(aes(spearman, lm)) +
  geom_abline(color = 'darkgrey') +
  # scale_x_log10() +
  # scale_y_log10() +
  geom_point()
```

The estimates are related but differ - I suspect this has to do with ties in the abundance (0s) and in the calprotectin (min values / detection limit). Perhaps the problematic taxa are very low prevalence?

### try all tests

```{r}
y <- ps %>%
  # filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  # prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
  as_tibble %>%
  with_groups(.sample, mutate, proportion = close_elts(.abundance)) %>%
  with_groups(.otu, nest) %>%
  mutate(
    calprotectin_qmp_spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman')),
    calprotectin_rmp_spearman = map(data, 
      ~cor.test(data = .x,
        ~ proportion + faecal_calprotectin_mg_g, 
        method = 'spearman')),
    moisture_qmp_spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + moisture_content_percent, 
        method = 'spearman')),
    moisture_rmp_spearman = map(data, 
      ~cor.test(data = .x,
        ~ proportion + moisture_content_percent, 
        method = 'spearman')),
  ) %>%
  select(-data) %>%
  pivot_longer(contains('spearman'), names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
```

TODO: check that samples with NAs are handled in the spearman tests

```{r}
y.w <- y %>%
  select(.otu, test, estimate, p.bh) %>%
  pivot_wider(names_from = c(test), values_from = c(estimate, p.bh), 
    names_glue = '{test}_{.value}', names_sort = TRUE) %>%
  select(.otu, sort(tidyselect::peek_vars())) %>%
  mutate(across(where(is.numeric), round, 4)) %>%
  arrange(calprotectin_qmp_spearman_p.bh)
```

Note, I'm not sure what the RMP values used in the original study were.
From the manuscript, it's not obvious if the unclassified taxa were removed before the proportions were computed, or if copy number correction was used.

### Check if patterns are driven by presence/absence

```{r}
y <- ps %>%
  # filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  # prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
  as_tibble %>%
  with_groups(.sample, mutate, 
    proportion = close_elts(.abundance),
    present = (.abundance > 0) * 1
  ) %>%
  with_groups(.otu, nest) %>%
  mutate(
    spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman')),
    spearman_pa = map(data, 
      ~cor.test(data = .x,
        ~ present + faecal_calprotectin_mg_g, 
        method = 'spearman')),
  ) %>%
  select(-data) %>%
  pivot_longer(contains('spearman'), names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
```

```{r}
y.w <- y %>%
  select(.otu, test, estimate, p.bh) %>%
  pivot_wider(names_from = c(test), values_from = c(estimate, p.bh), 
    names_glue = '{test}_{.value}', names_sort = TRUE) %>%
  select(.otu, sort(tidyselect::peek_vars())) %>%
  mutate(across(where(is.numeric), round, 4)) %>%
  arrange(spearman_p.bh)
```

```{r}
y.w %>%
  ggplot(aes(spearman_estimate, spearman_pa_estimate)) +
  geom_point()
```

For some taxa (like Veillonella, EscherichiaShigella) the two estimates are very similar;for others like Streptococcus they are quite different.
I suspect that simulated bias will have a larger impact on Streptococcus.

# Bias sensitivity analysis

```{r, eval = F}
# Perturb by bias vector - treat ps as actual
ps.per <- perturb(ps, bias, norm = 'keep')
# Calibrate by bias vector - treat ps as observed
ps.cal <- calibrate(ps, bias, norm = 'keep')
```

Maintain the sample sums, but perturb the relative abundances.

```{r}
n_taxa <- ntaxa(ps)
# Standard deviation for the distribution of log efficiencies
sigma.log_eff <- 3
# helper function to draw a single random bias vector
rbias <- function() exp(rnorm(n_taxa, 0, sigma.log_eff))
# helper function to calibrate the phyloseq object by the bias vector
my_cal <- function(bias, ps) {
  ps %>%
    calibrate(bias, norm = 'keep')
}

set.seed(42)
# Number of replicates
R <- 3e2
# data frame with R random bias vectors
reps <- seq(R) %>%
  set_names %>%
  map(~rbias()) %>%
  enframe("id", "bias") %>%
  mutate(
    actual = map(bias, my_cal, ps = ps)
  )
```

```{r}
my_stat <- function(ps, otu) {
  data <- ps %>%
    prune_taxa(otu, .) %>%
    as_tibble
  spearman_test <- cor.test(data = data,
    ~ .abundance + faecal_calprotectin_mg_g, 
    method = 'spearman', exact = FALSE
  )
  spearman_test %>% broom::tidy()
}

#> focal_otu <- 'Veillonella'
original <- my_stat(ps, focal_otu)
reps_veill <- reps %>%
  mutate(
    .otu = 'Veillonella',
    fit = map2(actual, .otu, my_stat)
  ) %>%
  unnest(fit)
reps_strep <- reps %>%
  mutate(
    .otu = 'Streptococcus',
    fit = map2(actual, .otu, my_stat)
  ) %>%
  unnest(fit)
```

```{r}
plot_bias_sens <- function(reps) {
  est_dist <- ggplot(reps, aes(estimate)) +
    geom_histogram() +
    geom_vline(xintercept = original$estimate, color = "darkred") +
    geom_vline(xintercept = 0, color = "grey") +
    labs(title = "Distribution of point estimates", x = "Estimate", y = "Count")
  pval_dist <- ggplot(reps, aes(p.value)) +
    geom_histogram() +
    geom_vline(xintercept = original$p.value, color = "darkred") +
    geom_vline(xintercept = 0.05, color = "grey") +
    scale_x_log10() +
    labs(title = "Distribution of p-values", x = "p-value", y = "Count")
  est_dist / pval_dist
}
# ci_reps <- ggplot(bias_reps, aes(x = estimate, y = as.integer(id))) +
#   geom_point(size = 0.8) +
#   geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
#   geom_point(data = original, aes(y = R*1.05), 
#     size = 0.8, color = "darkred") +
#   geom_errorbarh(data = original, 
#     aes(xmin = conf.low, xmax = conf.high, y = R*1.05),
#     color = "darkred") +
#   geom_vline(xintercept = 0, color = "grey") +
#   theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
#   labs(title = "Replicate 95% CIs", x = "Estimate", 
#     y = "Replicate")
# plot_grid(
#     plot_grid(est_dist, pval_dist, ncol = 1, labels = c("A", "B")),
#     ci_reps,
#     labels = c("", "C")
#     )
```

```{r}
plot_bias_sens(reps_veill)
```

```{r}
plot_bias_sens(reps_strep)
```

Example where the bias sensitivity analysis shows little effect of varying the hypothesized bias.

We can probably understand this finding because the original result is driven by the presence/absence pattern, and we have treated zeros as absolute, so that presence/absence is maintained.
If we relaxed this assumption, then I suspect that we would see much more variation in the estimates.
Much of this variation, however, would come from relaxing our assumption about the zeros, rather than relaxing our assumption that there is no bias.

Challenge with this data is that we don't have the read counts, so don't know how to replace zeros with resonable values.

## saved plotting stuff

Mean efficiency distribution:

```{r}
x <- reps %>%
  slice(1) %>%
  pull(actual) %>%
  .[[1]] %>%
  sample_data %>%
  as_tibble

x %>%
  ggplot(aes(x = mean_efficiency, fill = faecal_calprotectin_mg_g > 50)) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
x %>%
  ggplot(aes(x = faecal_calprotectin_mg_g, y = mean_efficiency, color = diagnosis)) +
  scale_y_log10() +
  scale_x_log10() +
  geom_point() +
  stat_smooth(aes(group = 1), method = 'lm')
```

Now, redo the analysis with and without bias

what assumptions to make about zeros?

### 2022-02-06 notebook/_posts/2022-01-25-explore-gut-data-2/_cut.Rmd
Add gut microbiome analyses

### CHECKs

```{r}
dataset <- 'KieserS_2018'
tse <- curatedMetagenomicData(
  str_c(dataset, ".relative_abundance"),
  dryrun = FALSE
)
stopifnot(identical(length(tse), 1L))
nm <- names(tse)
tse <- tse[[1]]
ps <- mia::makePhyloseqFromTreeSummarizedExperiment(
  tse,
  abund_values = "relative_abundance"
) %>%
  rename_with_sample_data(janitor::make_clean_names) %>%
  rename_with_tax_table(janitor::make_clean_names) %>%
  transform_sample_counts(~. / sum(.))
all.equal(
  tse %>% rownames %>% str_subset('Streptococcus'),
  ps %>% taxa_names %>% str_subset('Streptococcus')
)
```

### 2022-02-06 notebook/_posts/2022-01-25-explore-gut-data-2/explore-gut-data-2.Rmd
Add gut microbiome analyses

---
title: "Explore gut datasets, 2"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - R
date: 2022-01-25
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: svg
---

Purpose: Examine phylum-level variation w/ clinical condition in different studies.

```{r}
library(here)
library(tidyverse)
library(speedyseq)
library(metacal)

library(cowplot)
theme_set(theme_cowplot())
library(ggbeeswarm)
```

# Studies

```{r}
fs::dir_ls(here('notebook/_data/gut'))

study <- c(
  'HMP_2019_ibdmdb',
  # 'HMP_2019_t2d',
  'KieserS_2018',
  # 'QinJ_2012',
  'RaymondF_2016',
  'ThomasAM_2018b'
)
tbl <- tibble(study = study) %>%
  mutate(
    file = str_glue('2021-03-31.{study}.relative_abundance.phyloseq.rds'),
    path = here('notebook/_data/gut/', file),
    ps = map(path, readRDS)
  )
```

Aim: For each study, get a tibble with the processed phyloseq object and the relevant clinical variable.
The clinical var will usually be `study_condition`, but need to check, and also set the factor levels.

Note, processing that will be applied to all ps objects will be done after.

Thomas et al colorectal cancer

```{r}
ps <- tbl %>% filter(study == 'ThomasAM_2018b') %>% pull(ps) %>% .[[1]]
sam <- ps %>% sample_data %>% as_tibble
sam %>% glimpse
sam %>% dplyr::count(body_site, study_condition, disease)
sam %>% dplyr::count(age_category)
sam %>% dplyr::count(gender)
sam %>% dplyr::count(antibiotics_current_use)
```

HERE

```{r}
process_ps <- function(ps) {
  ps %>%
    filter_tax_table(kingdom == 'Bacteria') %>%
    transform_sample_counts(close_elts) %>%
    mutate_tax_table(.otu = species)
}
tbl <- tbl %>%
  mutate(
    ps_proc = map(ps, process_ps)
  )

```

- prep each dataset
- write some functions to do phylum level plots, compute variation, and compute effects on the phyla
- perhaps do the same for mean eff supposing a G+/G- model.

NOTE: We might want to do this case by case; or do the bias both ways (since some datasets may have a high B:F ratio, others low)

## helper functions

# Acute diarrhoea - Kieser et al 2018

```{r}
ps <- tbl %>% filter(study == 'KieserS_2018') %>% pull(ps_proc) %>% .[[1]]
sam <- ps %>% sample_data %>% as_tibble
sam %>% glimpse
sam %>% dplyr::count(body_site, age_category, study_condition, disease)
sam %>% dplyr::count(study_condition, uncurated_metadata)
```

Let's check the phylum proportions in healthy and sick kids,

```{r}
ps_phy <- ps %>% tax_glom('phylum') %>%
  mutate_tax_table(.otu = phylum)
x_phy <- ps_phy %>% as_tibble
```

```{r}
ps_phy %>%
  filter_taxa2(~median(.) > 0) %>%
  as_tibble %>%
  with_groups(c(phylum, study_condition), summarize,
    across(.abundance, median)) %>%
  pivot_wider(names_from = study_condition, values_from = .abundance)
```

```{r}
ps_phy %>%
  filter_taxa2(~median(.) > 0) %>%
  as_tibble %>%
  ggplot(aes(y = study_condition, x = 1e-5 + .abundance, color =
      study_condition)) +
  facet_wrap(~phylum, ncol = 1) +
  scale_color_brewer(type = 'qual') +
  scale_x_log10() +
  geom_quasirandom(groupOnX = FALSE)
```

Healthy kids have high proportions of the G+ Actinobacteria and Firmicutes, as well as Bacteroidetes; sick are dominated by Proteobacteria.

This finding is consistent with some of the observations of the paper.

After reading the article, I learned that
- many-most of the children were identified as having organisms that are considered gut pathogens / known causes of diarrhoea, which are all in Proteobacteria (E. coli, Vibrio, Campylobacter, Shigella, Aeromonas)
- these samples are a random subset of samples that were sequenced by 16S sequencing.
- 16S qPCR measurements were made to gain some absolute abundance info
- The authors report a fairly large increase in Streptococcus proportion, which is in Firmicutes. This finding does not seem to mesh with what I'm seeing in the shotgun profiles
- The authors report various other DA results

There is still important questions and interesting things I might look into, such as the disagreement between 16s and shotgun results. However, a relatively straightforward story could already be suggested.

Observations
- large phylum+ level differences, and in particular large differences in the G+ vs G- taxa, with a high fraction of controls being the G+ Actino and Firm, and a high fraction of cases being the G- Proteobacteria.
- Proteobacteria is known to contain many gut pathogens
<!--  -->

Things to look into
- the alpha diversity within phyla within patients, to see the effective number of species (and genera) within phyla
- What fraction of comm's within patients by most abundant 1 species - want to understand whether mean eff var would be driven by just one species
  - Either way, I think I can make the point that phylum-level conservation is important when we do a cross-species analysis, since it won't always be the same species (various Proteobacteria), I'm guessing
- Specifically dig into the Proteobacteria - see what species/genera are contributing to the Prot fraction in each individual esp. with diarrhea to see if true that it's not just one species, but various, but all within the phylum
- Fraction of comms made up of G+ versus G- phyla
  - perhaps use this as a proxy for mean efficiency, rather than bothering to simulate
<!--  -->

### Proteobacteria

Question: Which Proteobacteria species are driving the high Proteobacteria proportion in sick kids?

```{r}
prot_spp <- ps %>%
  filter_tax_table(phylum == 'Proteobacteria') %>%
  filter_taxa2(~max(.) > 1e-2) %>%
  as_tibble
prot_genus <- ps %>%
  filter_tax_table(phylum == 'Proteobacteria') %>%
  tax_glom('genus') %>%
  as_tibble
```

```{r}
prot_genus %>%
  mutate(across(genus, fct_reorder, .abundance)) %>%
  ggplot(aes(y = genus, x = 1e-5 + .abundance, color = study_condition)) +
  theme_minimal_hgrid() +
  geom_vline(xintercept = c(1e-2, 0.1, 1), color = 'grey') +
  scale_color_brewer(type = 'qual') +
  scale_x_log10() +
  geom_quasirandom(groupOnX = FALSE)
```

```{r}
prot_spp %>%
  mutate(across(.otu, fct_reorder, .abundance)) %>%
  ggplot(aes(y = .otu, x = 1e-5 + .abundance, color = study_condition)) +
  theme_minimal_hgrid() +
  geom_vline(xintercept = c(1e-2, 0.1, 1), color = 'grey') +
  scale_color_brewer(type = 'qual') +
  scale_x_log10() +
  geom_quasirandom(groupOnX = FALSE)
```

We can see that E. coli and Klebsiella spp. are the drivers of the high Proteobacteria proportions, with Vibrio in one patient also being substantial (10%).
Escherichia and Klebsiella are both in the family Enterobacteriaceae.

### Streptococcus

```{r}
strep <- ps %>%
  filter_tax_table(genus == 'Streptococcus') %>%
  tax_glom('genus') %>%
  as_tibble
strep_spp <- ps %>%
  filter_tax_table(genus == 'Streptococcus') %>%
  filter_taxa2(~max(.) > 3e-3) %>%
  as_tibble
```

```{r}
strep %>%
  ggplot(aes(y = study_condition, x = 1e-5 + .abundance, color =
      study_condition)) +
  scale_color_brewer(type = 'qual') +
  scale_x_log10() +
  geom_quasirandom(groupOnX = FALSE)
```

```{r}
strep_spp %>%
  ggplot(aes(y = study_condition, x = 1e-5 + .abundance, color =
      study_condition)) +
  facet_wrap(~species, ncol = 1) +
  scale_color_brewer(type = 'qual') +
  scale_x_log10() +
  geom_quasirandom(groupOnX = FALSE)
```

```{r}
ps %>%
  filter_tax_table(family == "Streptococcaceae") %>%
  tax_glom('family') %>%
  as_tibble %>%
  ggplot(aes(y = study_condition, x = 1e-5 + .abundance, color =
      study_condition)) +
  scale_color_brewer(type = 'qual') +
  scale_x_log10() +
  geom_quasirandom(groupOnX = FALSE)
```

We also don't see the result at the family level.

In short, the Metaphlan3 profiles do not support the Streptococcus associations reported in the study, which were found using 16S sequencing and shotgun sequencing (analyzed with Metaphlan2).

## Check for any clades with large differences between conditions

```{r}
check_diffs <- function(ps, taxrank, fn = mean) {
  ps %>% tax_glom(taxrank) %>% as_tibble %>%
    with_groups(c(phylum:species, study_condition), 
      summarize, across(.abundance, fn)) %>%
    pivot_wider(names_from = study_condition, values_from = .abundance) %>%
    mutate(diff = acute_diarrhoea - control) %>%
    arrange(desc(abs(diff)))
}
```

```{r}
check_diffs(ps, 'species', mean) %>% slice(1:4)
check_diffs(ps, 'genus', mean) %>% slice(1:4)
check_diffs(ps, 'family', mean) %>% slice(1:4)
check_diffs(ps, 'order', mean) %>% slice(1:4)
check_diffs(ps, 'class', mean) %>% slice(1:4)
check_diffs(ps, 'phylum', mean) %>% slice(1:4)
```

# Antibiotics - Raymond et al 2016

```{r}
ps <- tbl %>% filter(study == 'RaymondF_2016') %>% pull(ps_proc) %>% .[[1]]
sam <- ps %>% sample_data %>% as_tibble
sam %>% glimpse
sam %>% count(body_site, age_category, study_condition, disease)
sam %>% count(study_condition, antibiotics_family)
sam %>% count(study_condition, days_from_first_collection)
```

Note, the 'study_condition' varies by timepoint for the same subject, so isn't useful for filtering or coloring the subjects who received the antibiotic.
We therefore need to create a new variable.

```{r}
sam <- sam %>%
  with_groups(subject_id, mutate,
    subject_group = case_when(
      'yes' %in% antibiotics_current_use ~ 'antibiotics',
      TRUE ~ 'control'
    ),
    days_idx = rank(days_from_first_collection) %>% as.factor
  )
sample_data(ps) <- sample_data(sam)
sam %>% count(subject_group, days_from_first_collection, days_idx)
```

Let's check the phylum proportions,

```{r}
ps_phy <- ps %>% tax_glom('phylum') %>%
  mutate_tax_table(.otu = phylum)
x_phy <- ps_phy %>% as_tibble
focal_phyla <- c('Actinobacteria', 'Bacteroidetes', 'Firmicutes',
  'Proteobacteria')
```

```{r}
x_phy %>%
  filter(phylum %in% focal_phyla, subject_group == 'antibiotics') %>%
  with_groups(c(phylum, study_condition), summarize,
    across(.abundance, median)) %>%
  pivot_wider(names_from = study_condition, values_from = .abundance)
```

```{r}
x_phy %>%
  filter(phylum %in% focal_phyla, subject_group == 'antibiotics') %>%
  ggplot(aes(y = study_condition, x = 1e-5 + .abundance, color =
      study_condition)) +
  facet_wrap(~phylum, ncol = 1) +
  scale_color_brewer(type = 'qual') +
  scale_x_log10() +
  geom_quasirandom(groupOnX = FALSE)
```

```{r}
x_phy %>%
  filter(phylum %in% focal_phyla, subject_group == 'antibiotics') %>%
  ggplot(aes(days_idx, 1e-5 + .abundance, color = phylum)) +
  facet_wrap(~phylum, scales = 'free_y', ncol = 1) +
  scale_y_log10() +
  scale_color_brewer(type = 'qual', palette = 2) +
  geom_line(aes(group = interaction(subject_id, phylum))) +
  geom_point()
```

There seems to be some increase in G- phyla with rx administration, though not a huge effect.
I wonder if the effect would seem larger if we supposed a lysis bias favoring G-?

```{r}
bias <- ps_phy %>% tax_table %>% as_tibble %>%
  mutate(
    efficiency = case_when(
      phylum %in% c('Actinobacteria', 'Firmicutes') ~ 1/10,
      TRUE ~ 1
    )
  ) %>%
  select(.otu, efficiency) %>%
  deframe
ps_phy_cal <- ps_phy %>% calibrate(bias)
```

```{r}
ps_phy_cal %>% 
  as_tibble %>%
  filter(phylum %in% focal_phyla, subject_group == 'antibiotics') %>%
  ggplot(aes(days_idx, .abundance, color = phylum)) +
  facet_wrap(~phylum, scales = 'free_y', ncol = 1) +
  scale_y_log10() +
  scale_color_brewer(type = 'qual', palette = 2) +
  geom_line(aes(group = interaction(subject_id, phylum))) +
  geom_point()
```

```{r}
ps_phy_cal %>% 
  sample_data %>%
  as_tibble %>%
  ggplot(aes(days_idx, .mean_efficiency)) +
  scale_y_log10() +
  expand_limits(y = c(0.1, 1)) + 
  scale_color_brewer(type = 'qual', palette = 2) +
  stat_summary(color = 'darkred', position = position_nudge(x = -0.1)) +
  geom_line(aes(group = subject_id), alpha = 0.5) +
  geom_point(alpha = 0.5)
```

We see sizable variation in the mean efficiency, but just a small change in the mean during and after treatment.

Note, the variation would be larger if we enforced a difference between the average efficiency for the phylum but also allowed for variation within a phylum.

## Check for any clades with large differences between conditions

```{r}
check_diffs <- function(ps, taxrank, fn = mean) {
  ps %>% tax_glom(taxrank) %>% as_tibble %>%
    with_groups(c(phylum:species, study_condition), 
      summarize, across(.abundance, fn)) %>%
    pivot_wider(names_from = study_condition, values_from = .abundance) %>%
    mutate(diff = cephalosporins - control) %>%
    arrange(desc(abs(diff)))
}
```

```{r}
check_diffs(ps, 'species', mean) %>% slice(1:2)
check_diffs(ps, 'genus', mean) %>% slice(1:2)
check_diffs(ps, 'family', mean) %>% slice(1:2)
check_diffs(ps, 'order', mean) %>% slice(1:2)
check_diffs(ps, 'class', mean) %>% slice(1:2)
check_diffs(ps, 'phylum', mean) %>% slice(1:2)
```

# Colorectal cancer

```{r}
ps <- tbl %>% filter(study == 'ThomasAM_2018b') %>% pull(ps_proc) %>% .[[1]]
sam <- ps %>% sample_data %>% as_tibble
sam %>% glimpse
sam %>% count(study_condition, disease)
sam %>% count(body_site, age_category, study_condition)
```

Let's check the phylum proportions,

```{r}
ps_phy <- ps %>% tax_glom('phylum') %>%
  mutate_tax_table(.otu = phylum)
x_phy <- ps_phy %>% as_tibble
focal_phyla <- c('Actinobacteria', 'Bacteroidetes', 'Firmicutes',
  'Proteobacteria')
```

```{r}
x_phy %>%
  filter(phylum %in% focal_phyla) %>%
  with_groups(c(phylum, study_condition), summarize,
    across(.abundance, median)) %>%
  pivot_wider(names_from = study_condition, values_from = .abundance)
```

```{r}
x_phy %>%
  filter(phylum %in% focal_phyla) %>%
  ggplot(aes(y = study_condition, x = 1e-5 + .abundance, color =
      age_category)) +
  facet_wrap(~phylum, ncol = 1) +
  scale_color_brewer(type = 'qual') +
  scale_x_log10() +
  geom_quasirandom(groupOnX = FALSE)
```

Can see that the B:F ratio is somewhat lower in cases.

```{r}
bias <- ps_phy %>% tax_table %>% as_tibble %>%
  mutate(
    efficiency = case_when(
      phylum %in% c('Actinobacteria', 'Firmicutes') ~ 1/10,
      TRUE ~ 1
    )
  ) %>%
  select(.otu, efficiency) %>%
  deframe
ps_phy_cal <- ps_phy %>% calibrate(bias)
sam_cal <- ps_phy_cal %>% 
  sample_data %>%
  as_tibble
```

```{r}
sam_cal %>%
  ggplot(aes(study_condition, .mean_efficiency)) +
  scale_y_log10() +
  expand_limits(y = c(0.1, 0.3)) + 
  scale_color_brewer(type = 'qual', palette = 2) +
  stat_summary(color = 'darkred', position = position_nudge(x = -0.1)) +
  geom_point(alpha = 0.5)
```

```{r}
sam_cal %>%
  lm(data = ., log(.mean_efficiency) ~ study_condition) %>%
  summary
```

## Check for any clades with large differences between conditions

```{r}
check_diffs <- function(ps, taxrank, fn = mean) {
  ps %>% tax_glom(taxrank) %>% as_tibble %>%
    with_groups(c(phylum:species, study_condition), 
      summarize, across(.abundance, fn)) %>%
    pivot_wider(names_from = study_condition, values_from = .abundance) %>%
    mutate(diff = CRC - control) %>%
    arrange(desc(abs(diff)))
}
```

```{r}
check_diffs(ps, 'species', mean) %>% slice(1:5)
check_diffs(ps, 'genus', mean) %>% slice(1:5)
check_diffs(ps, 'family', mean) %>% slice(1:5)
check_diffs(ps, 'order', mean) %>% slice(1:5)
check_diffs(ps, 'class', mean) %>% slice(1:5)
check_diffs(ps, 'phylum', mean) %>% slice(1:5)
```

# Finding available datasets

note, first need to set x to the master sample data table

```{r}
x %>% dplyr::count(treatment) %>% print(n=Inf)
x %>% dplyr::count(antibiotics_current_use)
x %>% dplyr::count(antibiotics_family) %>% print(n=Inf)
```

studies that involve an antibiotics treatment?

```{r}
x %>% 
  dplyr::filter(antibiotics_family == 'cephalosporins') %>%
  dplyr::count(study_name)
x %>% 
  dplyr::filter(study_name == 'RaymondF_2016') %>%
  dplyr::count(antibiotics_current_use, antibiotics_family)
x %>% 
  dplyr::filter(study_name == 'RaymondF_2016')
```

## Diarrhoea

looking at study conditions shows there are two studies with accute diarrhoea

```{r}
ps1 <- curatedMetagenomicData::curatedMetagenomicData(
  "DavidLA_2015.relative_abundance", 
  dryrun = FALSE
) %>% 
  .[[1]] %>%
  mia::makePhyloseqFromTreeSummarizedExperiment(
    abund_values = "relative_abundance"
  ) %>%
  rename_with_sample_data(janitor::make_clean_names) %>%
  rename_with_tax_table(janitor::make_clean_names)

# ps1 %>% sample_data %>% as_tibble %>% glimpse
sam1 <- ps1 %>% sample_data %>% as_tibble
sam1 %>% dplyr::count(study_condition)
sam1 %>% dplyr::count(age_category)
sam1 %>% dplyr::count(body_subsite)
sam1 %>% dplyr::count(age_category, study_condition)
sam1 %>% dplyr::count(age_category, study_condition, body_subsite)
```

# CUT

## Phylum proportions by condition

```{r}
x <- ps %>%
  transform_sample_counts(close_elts) %>%
  tax_glom('phylum') %>%
  as_tibble
```

```{r}
x %>%
  with_groups(phylum, summarize, across(.abundance, median))
```

```{r}
focal_phyla <- c('Actinobacteria', 'Bacteroidetes', 'Firmicutes',
  'Proteobacteria')
x %>%
  filter(phylum %in% focal_phyla) %>%
  ggplot(aes(.abundance, fill = disease)) +
  facet_wrap(~phylum, scales = 'fixed', ncol = 1) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
focal_phyla <- c('Actinobacteria', 'Bacteroidetes', 'Firmicutes',
  'Proteobacteria')
x %>%
  filter(phylum %in% focal_phyla) %>%
  # ggplot(aes(y = .abundance, x = disease)) +
  ggplot(aes(y = .abundance, x = (disease != 'healthy') * 1L)) +
  facet_wrap(~phylum, scales = 'free_y', ncol = 1) +
  scale_y_log10() +
  geom_quasirandom() +
  stat_smooth(method = 'lm')
```

```{r}
y <- x %>%
  filter(phylum %in% focal_phyla) %>%
  with_groups(phylum, nest) %>%
  mutate(
    fit = map(data, ~lm(data = .x, log(.abundance) ~ disease)),
    fit = map(fit, broom::tidy)
  ) %>%
  select(-data) %>%
  unnest(fit)
y %>%
  filter(term != '(Intercept)')
```

## Bias

Let's come up with a way to simulate bias vectors with a particular property.

### 2022-02-06 notebook/_posts/2022-01-30-hmp-bias/hmp.Rmd
Add gut microbiome analyses

# Setup

## code to create ps object (not run)

```{r, eval = FALSE}
# It's helpful to load the Bioconductor cruft first so that it doesn't mask the
# functions we actually want to use.
HMP16SData::V13()
library(speedyseq)

ps_v13 <- HMP16SData::V13() %>% HMP16SData::as_phyloseq()
ps_v35 <- HMP16SData::V35() %>% HMP16SData::as_phyloseq()

ps_v13 %>% sample_data %>% glimpse
ps_v35 %>% sample_data %>% glimpse

# To merge, we first need to drop the phylogenetic tree, and adjust the sample
# names.
intersect(
  sample_names(ps_v13),
  sample_names(ps_v35)
) %>% length

ps_v13 <- ps_v13 %>% 
  mutate_sample_data(
    region = 'V13', 
    specimen_id = .sample,
    .sample = str_glue('X{specimen_id}_{region}')
  )
ps_v35 <- ps_v35 %>% 
  mutate_sample_data(
    region = 'V35', 
    specimen_id = .sample,
    .sample = str_glue('X{specimen_id}_{region}')
  )

ps_merge <- merge_phyloseq(
  ps_v13 %>% {.@phy_tree <- NULL; .}, 
  ps_v35 %>% {.@phy_tree <- NULL; .}
)

# let's clean up names and aggregate to the genus level,

ps <- ps_merge %>%
  rename_with_sample_data(janitor::make_clean_names) %>%
  rename_with_tax_table(janitor::make_clean_names) %>%
  tax_glom('genus')
ps
rm(ps_merge, ps_v13, ps_v35)
dir.create('_output')
saveRDS(ps, '_output/hmp-16s-genus.rds')
```

```{r}
```

## start here

```{r}
library(here)
library(tidyverse)
library(speedyseq)
library(metacal)

library(cowplot)
theme_set(theme_cowplot())
library(ggbeeswarm)
```

```{r}
ps <- readRDS('_output/hmp-16s-genus.rds') %>%
  mutate_sample_data(., sample_sum = sample_sums(.))
ps
sam <- ps %>% sample_data %>% as_tibble
```

```{r}
sam %>% count(specimen_id) %>% count(n)
control_specimens <- sam %>% count(specimen_id) %>% filter(n>1) %>%
  pull(specimen_id)
```

To estimate bias, we'll work with the specimens with both regions.

TODO: Filter samples to a smaller more workable set, perhaps just gut.

```{r}
sam %>% 
  filter(specimen_id %in% control_specimens) %>%
  count(hmp_body_site)
```

TODO: Filter genera to high prevalence ones.

```{r}
ps1 <- ps %>% 
  filter_sample_data(
    hmp_body_subsite == 'Stool', 
    specimen_id %in% control_specimens
  ) %>%
  filter_taxa2(~sum(.) > 10)
```

```{r}
ps1 %>% sample_sums %>% sort %>% summary
ps1 %>% sample_sums %>% sort %>% head(20)
```

```{r}
ps2 <- ps1 %>% filter_sample_data(sample_sum >= 3000)
control_specimens_filt <- ps2 %>% sample_data %>% as_tibble %>% 
  count(specimen_id) %>% filter(n>1) %>%
  pull(specimen_id)
```

```{r}
x <- ps2 %>%
  transform_sample_counts(close_elts) %>%
  as_tibble %>%
  with_groups(.otu, summarize, 
    across(.abundance, .names = '{.fn}',
      .fns = list(mean = mean, median = median, max = max)
    )
  ) %>%
  arrange(max)
x %>% print(n=10)
x %>% count(max > 1e-2)
focal_taxa <- x %>% filter(max > 1e-2) %>% pull(.otu)
```

```{r}
ps_fit <- ps2 %>%
  prune_taxa(focal_taxa, .)
```

```{r}
ps_list <- list('V13', 'V35') %>% set_names %>%
  map(~filter_sample_data(ps_fit, region == .x)) %>%
  map(filter_sample_data, 
    specimen_id %in% control_specimens_filt
  ) %>%
  map(mutate_sample_data, .sample = str_c('X', specimen_id)) %>%
  map(transform_sample_counts, ~. + 0.3)
bias_fit <- estimate_bias(
  ps_list$V13,
  ps_list$V35,
  boot = TRUE
)
bias <- bias_fit %>% coef
```

```{r}
tax <- ps_fit %>% tax_table %>% as_tibble %>%
  left_join(summary(bias_fit)$coefficients, by = c('.otu' = 'taxon'))
```

```{r}
tax %>%
  ggplot(aes(estimate, fill = phylum)) +
  scale_x_log10() +
  scale_fill_brewer(type = 'qual') +
  geom_histogram()
```

```{r}
tax %>%
  mutate(.otu = fct_reorder(.otu, estimate)) %>%
  ggplot(aes(.otu, estimate, color = phylum,
      ymin = estimate / gm_se^2, ymax = estimate * gm_se^2)) +
  scale_color_brewer(type = 'qual', palette = 2) +
  geom_hline(yintercept = 1, color = "grey") +
  geom_pointrange() +
  scale_y_log10() +
  coord_flip()
```

The CIs are pretty tight, but I wonder how well the model really does?

### check model fit

```{r}
props <- ps_list %>%
  map(transform_sample_counts, close_elts) %>%
  map_dfr(as_tibble)
```

```{r}
observed.fitted <- fitted(mc_fit) %>% otu_table(taxa_are_rows = FALSE)
props.fitted <- bind_rows(
  props, 
  as_tibble(observed.fitted) %>% add_column(type = "Fitted")
)
```

### finding better taxa names

```{r}
tax <- ps2 %>% tax_table %>% as_tibble
tax %>% count(genus) %>% filter(n>1)
tax %>% count(str_c(family, genus, sep = '_')) %>% filter(n>1)
tax %>% filter(is.na(family))
```

### 2022-02-06 notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/_cut.Rmd
Add gut microbiome analyses

# HMP2 IBD study

```{r}
here('notebook/_data/gut') %>% fs::dir_ls()
ps <- here('notebook/_data/gut',
  '2021-03-31.HMP_2019_ibdmdb.relative_abundance.phyloseq.rds'
) %>% readRDS
sam <- ps %>% sample_data %>% as_tibble
tax <- ps %>% tax_table %>% as_tibble %>%
  add_column(sum = taxa_sums(ps))
```

```{r}
sam %>% glimpse
sam %>% dplyr::count(body_site)
sam %>% dplyr::count(dna_extraction_kit)
sam %>% dplyr::count(location)
sam %>% dplyr::count(subject_id) %>% dplyr::count(n)
sam %>% dplyr::count(study_condition, disease, disease_subtype)
```

Most subjects have many samples.

In the HMP2 paper, they classified some samples as dysbiotic, but I dont' think we have that info here.

Let's check out the tax table, and consider filtering some taxa.

```{r}
ps %>%
  tax_glom('phylum') %>%
  as_tibble %>%
  with_groups(c(kingdom, phylum), summarize,
    across(.abundance, list(median = median, mean = mean, max = max, sum = sum),
      .names = '{.fn}'
    ))
```

```{r}
tax %>% pull(sum) %>% sort %>% head
```

## Most abundant species

What is the proportion of the most abundant species in the sample?

```{r}
x <- ps %>% as_tibble
```

```{r}
most_abundant <- x %>%
  with_groups(.sample, slice_max, .abundance, n = 1)
```

```{r}
most_abundant %>%
  ggplot(aes(.abundance)) +
  scale_x_log10() +
  geom_histogram() +
  labs(x = 'Proportion')
```

```{r}
most_abundant %>%
  pull(.abundance) %>%
  summary
```

The most abundant species often forms a large fraction of the community.

Which species are most often most abundant?

```{r}
most_abundant %>%
  count(phylum, species) %>%
  arrange(-n) %>%
  print(n=20)
```

A relatively small set of species constitute the most abundant in most samples, and these are often from the same genera (Bacteroides, Prevotella, Roseburia) or phylum.
These observations suggest there is room for substantial variation in the mean efficiency across samples.

Let's compare to alpha diversity (effective number of species).

```{r}
div <- ps %>% as_tibble %>%
  with_groups(.sample, summarize,
    diversity_q1 = vegan::diversity(.abundance, index = 'shannon') %>% exp,
    diversity_q2 = vegan::diversity(.abundance, index = 'invsimpson')
    #> diversity_q2 = 1/sum(.abundance^2)
  ) %>%
  left_join(most_abundant)
```

```{r}
```

```{r}
div %>%
  ggplot(aes(diversity_q2)) +
  scale_x_log10() +
  geom_histogram() +
  labs(x = 'Diversity (q=2)')
```

```{r}
div %>%
  ggplot(aes(x = .abundance, y = diversity_q2)) +
  scale_x_log10() +
  scale_y_log10(limits = c(1, 30)) +
  geom_line(aes(y = 1/.abundance^2), color = 'grey') +
  geom_point(alpha = 0.5) +
  labs(x = 'Largest proportion', y = 'Diversity (q=2)')
```

Can see that there isn't a crazy amount of diversity - still typically less than 10.

## Variation in phylum proportions

If phyla vary substantially across samples, then any phylum-specific bias would drive variation in the mean efficiency.

I'll restrict plots to just the four main phyla.

```{r}
focal_phyla <- c('Actinobacteria', 'Bacteroidetes', 'Firmicutes',
  'Proteobacteria')
ps_phy <- ps %>% 
  filter_tax_table(kingdom %in% 'Bacteria') %>%
  tax_glom('phylum') %>%
  mutate_tax_table(.otu = phylum) %>%
  transform_sample_counts(close_elts) %>%
  mutate_tax_table(
    gram = case_when(
      phylum %in% c('Actinobacteria', 'Firmicutes') ~ 'Positive',
      TRUE ~ 'Negative'
    )
  )
x_phy <- ps_phy %>% as_tibble
```

```{r}
x_phy %>%
  filter(phylum %in% focal_phyla) %>%
  ggplot(aes(.abundance, fill = disease)) +
  facet_wrap(~phylum, scales = 'fixed', ncol = 1) +
  scale_x_log10() +
  geom_histogram()
```

There is substantial log-scale variation in all phyla, but the Bacteroidetes and Firmicutes are generally much more abundant than the other two.

```{r}
x_phy %>%
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  ggplot(aes(.abundance + 1e-4, fill = disease_subtype)) +
  facet_wrap(~phylum, ncol = 1) +
  scale_x_log10() +
  geom_histogram()
```

What about the B:F ratio?

```{r}
x_phy %>%
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  mutate(
    across(.abundance, ~ . + 1e-4),
    ratio_f = .abundance / .abundance[phylum == 'Firmicutes']
  ) %>%
  filter(phylum == 'Bacteroidetes') %>%
  ggplot(aes(ratio_f, fill = age_category)) +
  scale_x_log10(breaks = 10^(-3:4)) +
  facet_wrap(~disease, ncol = 1) +
  geom_histogram() +
  theme_minimal_vgrid()
```

There is a lot of variation, but most if it is generally not associated with disease.
This suggests that in this case, the main impact of variation in mean efficinecy from B:F ratio may be to create noise.

(note, I am ignoring structure in the data such as replicate sampling)

how correlated are species within a phylum?
what is the typical within-phylum diversity?

### Variation in mean efficiency due to phyla

Suppose that these profiles are biased to favor Gram negative phyla.
We might expect this because of the high Bacteroidetes / Firmicutes ratio and previous findings that the HMP1 protocol had this issue.
(Question: is the HMP2 gut protocol the same as HMP1?)

Suppose a relatively modest bias of 10X.
What variation in the mean efficiency would we see?

```{r}
bias <- ps_phy %>% tax_table %>% as_tibble %>%
  mutate(
    efficiency = case_when(
      gram == 'Positive' ~ 1/10,
      TRUE ~ 1
    )
  ) %>%
  select(.otu, efficiency) %>%
  deframe
ps_phy_cal <- ps_phy %>% calibrate(bias)
```

```{r}
ps_phy_cal %>% 
  sample_data %>%
  as_tibble %>%
  ggplot(aes(.mean_efficiency)) +
  scale_x_log10() +
  geom_histogram()
```

We see sizable variation in the mean efficiency.

```{r}
bias %>% gm_sd
ps_phy_cal %>% sample_data %>% as_tibble %>% pull(.mean_efficiency) %>% gm_sd
```

Though the geometric standard deviation is modest.
Interestingly, if we increase the efficiency difference up from 10X to 100X, the geometric sd only increases from 1.7 to 2.3.
I think this is because the bias being sufficiently strong implies that the (calibrated) communities are dominated by Gram positive and the mean efficiency stabilises at whatever the G+ efficiency is.

## Diversity within phyla

What is the effective number of speices within phyla?
For q=2 (Inverse Simpson index),

```{r}
div <- x %>% 
  filter(phylum %in% focal_phyla) %>%
  with_groups(c(.sample, phylum), mutate,
    across(.abundance, close_elts)
  ) %>%
  with_groups(c(.sample, phylum), summarize,
    diversity = 1/sum(.abundance^2)
  )
```

```{r}
div %>%
  ggplot(aes(diversity)) +
  scale_x_log10() +
  facet_wrap(~phylum, ncol = 1) +
  geom_histogram() +
  theme_minimal_vgrid()
```

```{r}
div %>%
  with_groups(phylum, summarize,
    qs = quantile(diversity, c(0.25, 0.5, 0.75), na.rm = TRUE), 
    prob = c(0.25, 0.5, 0.75)
  ) %>%
  pivot_wider(names_from = prob, values_from = qs)
```

We can see that there is typically diversity within phyla within an individual sample.

## Phylogenetic conservation

Question: How does the variation in mean efficiency across samples vary, if we increase phylogenetic variation in bias?

More generally, what does the variation in mean efficiency look like if we allow it to vary across species?
It may vary more than in the simple G+/G- case done above.

idea:

- fix the variation among species efficiencies, but vary a parameter that determines the phylogenetic conservation
    - Simpler: Do a sim where efficiencies are iid, and a sim where they come from the vcv matrix
- for each value of that parameter, generate 100 bias vectors, and compute the stnadard deviation in log mean efficiency.

Question: How does the variation in mean efficiency scale with the variation in simulated efficiencies? I would guess it would directly scale.

We can also consider a model where there is a correlation at the Gram-type level.

### Setup

To reduce the number of samples, let's grab just a couple samples per person.

Also subset to just bacteria - make sure to remove the sample that doesn't have bacterial taxa.

And drop some low prev taxa.

```{r}
```

```{r}
smpls <- sam %>%
  with_groups(subject_id, slice_min, visit_number, n = 3) %>%
  pull(.sample)

ps_focal <- ps %>%
  filter_tax_table(kingdom == 'Bacteria') %>%
  filter_taxa2(~sum(.>0) >= 10) %>%
  filter_sample_data(.sample %in% smpls) %>%
  transform_sample_counts(close_elts) %>%
  prune_samples(!is.na(sample_sums(.)), .) %>% # to drop samples w/o bacteria
  mutate_tax_table(.otu = species) %>%
  mutate_tax_table(
    gram = case_when(
      phylum %in% c('Actinobacteria', 'Firmicutes') ~ 'Positive',
      TRUE ~ 'Negative'
    )
  )

tree <- ps_focal %>% phy_tree 
phy_vcv <- ape::vcv(tree)
```

```{r}
diff_type <- ps_focal %>%
  tax_table %>%
  as_tibble %>%
  pull(gram) %>%
  {. == 'Positive'} %>%
  dist() %>%
  as('matrix')
same_type <- diff_type %>%
  {1 - .}
gram_vcv <- (same_type * 0.8) + (diff_type * -0.8)
diag(gram_vcv) <- 1
#> gram_vcv <- matrix(0, nrow = ntaxa(ps_focal), ncol = ntaxa(ps_focal))
rownames(gram_vcv) <- colnames(gram_vcv) <- taxa_names(ps_focal)
```

Idea: Fix the variance in species efficiencies, and simulate either as IID or with the given covariance.
The simplest way to do so is to sample the bias vector, then scale to the desired (geometric) standard deviation.

```{r}
# Returns a matrix with taxa as columns
sample_bias <- function(n, gsd, vcv, taxa_as = 'cols') {
  stopifnot(isSymmetric(vcv))
  stopifnot(n > 1)
  n_taxa <- nrow(vcv)
  gsd_log <- log(gsd)
  bias <- MASS::mvrnorm(n, mu = rep(0, n_taxa), Sigma = vcv) %>%
    t %>%
    scale %>%
    {. * gsd_log} %>%
    exp
  if (taxa_as == 'cols')
    bias <- t(bias)
  bias
}
#> x <- sample_bias(3, gsd = 3, phy_vcv)
#> x %>% apply(1, gm_sd) %>% head
#> x[1,] %>% qplot + scale_x_log10()
#> sample_bias(3, gsd = 3, phy_vcv, taxa_as = 'rows') %>% head
```

Now, generate R bias vectors of each type, and calibrate the data by them.

```{r}
set.seed(42)
R <- 300
gsd <- 3
reps <- list(
  iid = diag(ntaxa(ps_focal)),
  vcv = ps_focal %>% phy_tree %>% ape::vcv(),
  gram = gram_vcv
) %>%
  map_dfr(.id = 'sim_type',
    ~sample_bias(n = R, gsd = gsd, vcv = .x) %>%
      split(., row(.)) %>%
      enframe(".idx", "bias")
  ) %>%
  mutate(
    ps = map(bias, ~calibrate(ps_focal, .x))
  )
  # TODO: cache this, since it is a bit slow (due to large number of samples)
```

now, for each phyloseq object, compute the variation in the log mean efficiency, and its variation with disease state.

```{r}
reps_me <- reps %>%
  mutate(
    ps_sam = map(ps, ~sample_data(.) %>% as_tibble)
  ) %>%
  select(-bias, -ps) %>%
  unnest(ps_sam)
```

```{r}
reps_me_summ <- reps_me %>%
  with_groups(c(sim_type, .idx), summarize,
    across(.mean_efficiency, gm_sd)
  )
reps_me_summ %>%
  ggplot(aes(y = sim_type, x = .mean_efficiency)) +
  #> theme_minimal_hgrid() +
  scale_x_log10() +
  #> expand_limits(x = c(1, 3)) +
  geom_vline(xintercept = c(1,3), color = 'grey') +
  geom_boxplot(outlier.shape = NA) +
  geom_quasirandom(groupOnX = FALSE, alpha = 0.5) +
  labs(x = 'gm. std. of mean efficiency')
  # stat_slab(slab_type = 'histogram', breaks = 20)
```

As expected, the GSD's are larger under the VCV model; however, the differences is fairly small.
In particular, the median GSD isn't much larger, though the increase in the upper tail is a bit larger.
Perhaps the explanation has to do with our observation that the dominant species already often forms a large fraction of the community, and the bulk of the mean efficiency variation comes from the dominant few species swapping around, rather than of higher-level taxa.

Or perhaps it has something to do with my enforcing a constant gm-std across all taxa, and the structure of the phylogeny.

## Enterotypes? TODO

Work of others suggests we might be able to identify just a handul of species/genera/groups that tend to be dominant.

```{r}

```

what taxa tend to reach dominance/very high abundance?

```{r}
ps.gen <- ps %>% tax_glom('genus')
```

```{r}
ps.gen %>%
  transform_sample_counts(close_elts) %>%
  as_tibble %>%
  filter(.abundance > 0.3) %>%
  dplyr::count(phylum, family, genus) %>%
  arrange(-n) %>%
  print(n=20)
```

```{r}
ps %>%
  tax_glom('family') %>%
  transform_sample_counts(close_elts) %>%
  as_tibble %>%
  filter(.abundance > 0.3) %>%
  dplyr::count(phylum, family) %>%
  arrange(-n) %>%
  print(n=20)
```

This suggests that when we are considering phylogenetic conservation for the purposes of bias effect, then conservation within firmicutes may not be as important as that within specific sub-groups like Lachnospiraceae and Ruminococcaceae.

Note, the original HMP protocol was found to be biased in favor of Bacteroidetes.

# MOVE Thomas  et al

> Public metagenomic cohorts of patients with CRC, adenomas and controls. We downloaded five public fecal shotgun CRC datasets covering samples from six different countries, totaling 313patients with CRC, 143 adenomas and 308 controls (Table 1), and now available in curatedMetagenomicData27. We manually curated metadata tables for the public cohorts according to the curatedMetagenomicData27 Rpackage grammatical rules. The metadata table includes ten fields (sampleID, subjectID, body_site, country, sequencing_platform, PMID, number_reads, number_bases, minimum_read_length, median_read_length) that are mandatory for all datasets, in addition to other fields that are dataset-specific.

Check to see if I can get these, and use them for the analysis.

### 2022-02-06 notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/hmp-stool-vagina-comparison.Rmd
Add gut microbiome analyses

---
title: "Compare impact of bias in HMP stool and vagina samples"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - R
date: 2022-01-30
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: svg
---

```{r}
library(here)
library(tidyverse)
library(speedyseq)
library(metacal)

library(cowplot)
theme_set(theme_cowplot())
library(ggbeeswarm)
library(ggdist)
library(ggridges)
```

HERE - refactor to match the manuscript, and add the analysis of variation in species proportions relative to mean efficiency, and also look at the bimodality/presence/absence nature of species variation.

# HMP1

```{r}
#> here('notebook/_data/gut') %>% fs::dir_ls()
ps <- here('notebook/_data/gut',
  '2021-03-31.HMP_2012.relative_abundance.phyloseq.rds'
) %>% readRDS %>%
  filter_tax_table(kingdom == 'Bacteria') %>%
  prune_samples(sample_sums(.) > 0, .) %>% # to drop samples w/o bacteria %>%
  transform_sample_counts(close_elts) %>%
  mutate_tax_table(.otu = species)
# most abundant species and diversity
x <- ps %>% as_tibble
most_abundant <- x %>%
  with_groups(.sample, slice_max, .abundance, n = 1) %>%
  select(.sample, top_species = species, top_species_prop = .abundance)
fns <- list(
  #> richness = ~sum(.x > 0),
  diversity_q0 = ~sum(.x > 0),
  diversity_q1 = ~vegan::diversity(.x, index = 'shannon') %>% exp,
  diversity_q2 = ~vegan::diversity(.x, index = 'invsimpson')
) %>%
  map(as_mapper)
mat <- ps %>% otu_table %>% orient_taxa(as = 'cols') %>% as('matrix')
div <- fns %>%
  map_dfc(~apply(mat, 1, .x))
div1 <- bind_cols(most_abundant, div)
#  join w/ ps object
ps <- ps %>%
  left_join_sample_data(div1, by = '.sample')
rm(most_abundant, div, div1, x)
```

Let's see how things vary by body site

```{r}
sam <- ps %>% sample_data %>% as_tibble
sam %>% count(body_site)
sam %>% count(body_site, body_subsite)
```

## diversity

```{r}
sam %>%
  pivot_longer(starts_with('diversity'), names_to = 'index') %>%
  ggplot(aes(y = body_site, x = value)) +
  scale_x_log10() +
  facet_grid(.~index, scales = 'free', space = 'free') +
  geom_boxplot()
```

```{r}
sam %>%
  ggplot(aes(top_species_prop)) +
  facet_wrap(~body_site, ncol = 1) +
  scale_x_log10() +
  geom_histogram() +
  labs(x = 'Proportion')
```

```{r}
sam %>%
  with_groups(body_site, count, top_species) %>%
  with_groups(body_site, mutate, frac = close_elts(n)) %>%
  with_groups(body_site, slice_max, n, n = 5) %>%
  print(n=Inf)
```

```{r}
focal_subsites <- c('stool', 'posterior_fornix')
sam %>%
  filter(body_subsite %in% focal_subsites) %>%
  with_groups(body_site, count, top_species) %>%
  with_groups(body_site, mutate, frac = close_elts(n)) %>%
  with_groups(body_site, slice_max, n, n = 10) %>%
  print(n=Inf)
```

What about total diversity across samples (gamma diversity)?
Check that this is much larger in the gut than the vagina.

```{r}

```

### gamma diversity // diversity partition

Can use the hillR package to compute alpha, beta, and gamma diversity as Hill numbers (effective number of species) for q = 0 (richness), q = 1 (exponential of Shannon entropy), and q = 2 (Inverse Simpson index).

Note: q=0 gamma diversity is will be fairly sensitive to increasing the number of samples, and so we'd expect it to be larger for stool simply due to the larger number of samples, unless we subsample.

q=0 is also sensitive to read depth, which we don't really have a way to account for here.

#### testing

```{r}
get_site_ps <- function(body_site, ps, n ) {
  filter_sample_data(ps, body_site == !!body_site) %>% 
    filter_taxa2(~sum(.) > 0) %>%
    orient_taxa(as = 'cols')
}
div <- crossing(q = c(0, 1, 2), body_site = c('stool', 'vagina')) %>%
  mutate(
    ps = map(body_site, get_site_ps, ps = ps),
    otu = map(ps, otu_table),
    div = map2(otu, q, hillR::hill_taxa_parti),
    n = map_int(ps, nsamples)
  ) %>%
  select(-q, -otu) %>%
  unnest(div)
div %>% select(body_site, q, starts_with('TD'), n) %>% knitr::kable(digits = 1)
```

Can see that total (gamma) diversity and average alpha diversity are much larger in the gut than in hthe vagina, as expected, for each index.

The average q=2 alpha diversity in the stool is in fact quite low (only 5.4) - check on that, since it doesn't quite match what I remember from the figures.

Let's try the same thing with phylogeny diversity

```{r}
div_phy <- crossing(q = c(0, 1, 2), body_site = c('stool', 'vagina')) %>%
  mutate(
    ps = map(body_site, get_site_ps, ps = ps),
    comm = map(ps, otu_table),
    tree = map(ps, phy_tree),
    n = map_int(ps, nsamples)
  ) %>%
  rowwise %>%
  mutate(
    hillR::hill_phylo_parti(comm, tree, q)
  ) %>%
  select(-ps, -comm, -tree) %>%
  ungroup
div_phy %>% select(body_site, q, starts_with('PD'), n) %>% knitr::kable(digits = 1)
```

NOTES/TODO

- check the def of gamma diversity; does it control for the difference in the number of samples from each body-site type? I think not, and so gut will be larger simply b/c of more samples
- consider filtering down to a final set of samples; E.g. Perhaps just the one vaginal sample type, and some quick QC filtering; Though this shouldn't make a huge difference.
- filter taxa? or treat all as true positives

I also need to read about these diversity metrics.

#### better

GOAL: For an equal number of gut and vaginal samples, compute taxa and phylo gamma/alpha/beta diversity.
Do for q=0,1,2, though we can't really trust the q=0 numbers.

First, get equal numbers of stool and vaginal (posterior fornix) samples.

```{r}
sam %>% count(body_site, body_subsite)
ps_list <- list(
  stool = ps %>% filter_sample_data(body_subsite == 'stool'),
  vagina = ps %>% filter_sample_data(body_subsite == 'posterior_fornix')
) %>%
  # subsample to equal number of samples
  map(filter_sample_data, row_number() <= 62) %>%
  map(filter_taxa2, ~sum(.) > 0) %>%
  map(orient_taxa, as = 'cols')
```

```{r}
# Function that, given a phyloseq object, computes phylogenetic and species
# diversity measures for the requested q values
hill <- function(ps, q) {
  ps <- ps %>% filter_taxa2(~sum(.) > 0) %>% orient_taxa(as = 'cols')
  suppressMessages(
    tibble(q = !!q) %>%
      mutate(
        div_taxa = map(q, ~hillR::hill_taxa_parti(otu_table(ps), .x)),
        div_phylo = map(q, ~hillR::hill_phylo_parti(otu_table(ps), phy_tree(ps), .x)),
      ) %>%
      unnest(c(div_taxa, div_phylo), names_repair = 'universal') %>%
      select(q = q...1, starts_with(c('TD', 'PD')))
  )
  }
div <- ps_list %>%
  map_dfr(hill, q = c(0, 1, 2), .id = 'body_site')
```

```{r}
div %>% knitr::kable(digits = 1)
```

```{r}
div %>%
  pivot_longer(starts_with(c('TD', 'PD'))) %>%
  pivot_wider(names_from = body_site)
```

## Bias simulation

TODO: MAYBE: change so that we filter to the ecosystem-specific taxa before scaling the mean efficiencies
This mainly matters for the phylogenetic sim.

```{r}
focal_subsites <- c('stool', 'posterior_fornix')
ps_sim <- ps %>%
  filter_sample_data(body_subsite %in% focal_subsites) %>%
  filter_taxa2(~sum(.>0) >= 2) %>%
  prune_samples(sample_sums(.) > 0, .) %>% # to drop samples w/o bacteria %>%
  transform_sample_counts(close_elts) %>%
  mutate_tax_table(
    gram = case_when(
      phylum %in% c('Actinobacteria', 'Firmicutes') ~ 'Positive',
      TRUE ~ 'Negative'
    )
  )
```

Now, generate R bias vectors of each type, and calibrate the data by them.

```{r}
# Returns a matrix with taxa as columns
sample_bias <- function(n, gsd, vcv, taxa_as = 'cols') {
  stopifnot(isSymmetric(vcv))
  stopifnot(n > 1)
  n_taxa <- nrow(vcv)
  gsd_log <- log(gsd)
  bias <- MASS::mvrnorm(n, mu = rep(0, n_taxa), Sigma = vcv) %>%
    t %>%
    scale %>%
    {. * gsd_log} %>%
    exp
  if (taxa_as == 'cols')
    bias <- t(bias)
  bias
}
#> x <- sample_bias(3, gsd = 3, phy_vcv)
#> x %>% apply(1, gm_sd) %>% head
#> x[1,] %>% qplot + scale_x_log10()
#> sample_bias(3, gsd = 3, phy_vcv, taxa_as = 'rows') %>% head
```

```{r, eval = F}
set.seed(42)
R <- 300
gsd <- 3
reps <- list(
  iid = diag(ntaxa(ps_sim)),
  vcv = ps_sim %>% phy_tree %>% ape::vcv()
  #> gram = gram_vcv
) %>%
  map_dfr(.id = 'sim_type',
    ~sample_bias(n = R, gsd = gsd, vcv = .x) %>%
      split(., row(.)) %>%
      enframe(".idx", "bias")
  ) %>%
  mutate(
    ps = map(bias, ~calibrate(ps_sim, .x))
  )
  # TODO: cache this, since it is a bit slow (due to large number of samples)
reps_me <- reps %>%
  mutate(
    ps_sam = map(ps, ~sample_data(.) %>% as_tibble)
  ) %>%
  select(-bias, -ps) %>%
  unnest(ps_sam)
reps_me_summ <- reps_me %>%
  with_groups(c(sim_type, body_site, .idx), summarize,
    across(.mean_efficiency, gm_sd)
  )
```

NOTE: The var in mean eff and var in species props might be more directly comparable if we treat the observed proportions as the truth, rather than observed.

now, for each phyloseq object, compute the variation in the log mean efficiency

```{r, eval = T}
set.seed(42)
R <- 300
gsd_sim <- 3
reps <- list(
  iid = diag(ntaxa(ps_sim)),
  vcv = ps_sim %>% phy_tree %>% ape::vcv()
  #> gram = gram_vcv
) %>%
  map_dfr(.id = 'sim_type',
    ~sample_bias(n = R, gsd = gsd_sim, vcv = .x) %>%
      split(., row(.)) %>%
      enframe(".idx", "bias")
  )
reps_me <- reps %>%
  mutate(
    mean_efficiency = map(bias, ~mean_efficiency(ps_sim, .x, type = 'actual')),
    across(mean_efficiency, map, enframe, '.sample', 'mean_efficiency')
  ) %>%
  select(-bias) %>%
  unnest(mean_efficiency) %>%
  left_join(ps_sim %>% sample_data %>% as_tibble, by = '.sample')
reps_me_summ <- reps_me %>%
  with_groups(c(sim_type, body_site, .idx), summarize,
    across(mean_efficiency, gm_sd)
  )
```

```{r}
reps_me_summ %>%
  ggplot(aes(y = sim_type, x = mean_efficiency)) +
  facet_wrap(~body_site, ncol = 1) +
  #> theme_minimal_hgrid() +
  scale_x_log10() +
  #> expand_limits(x = c(1, 3)) +
  geom_vline(xintercept = c(1,3), color = 'grey') +
  geom_boxplot(outlier.shape = NA) +
  geom_quasirandom(groupOnX = FALSE, alpha = 0.5) +
  labs(x = 'gm. std. of mean efficiency')
  # stat_slab(slab_type = 'histogram', breaks = 20)
```

Interesting - note that forcing a gm-sd of 3 across all species can allow for a larger variation in the mean efficiency in some cases, in the vaginal communities.
, actually leads to a wider variation among vaginal species.
This makes some sense: We expect the variation in the efficiency of vaginal communities to be driven by a just a few species, and the there is a non-negligible chance that they end up with a much higher/lower mean efficiency than typical.

NOTE: I think the mean efficiency should scale with whatever we set the GSD to.
So we might want to set it to 1 and label the units as being per GSD.

When we compare this variation to that in the species proportions, the scale/value of GSD will matter.
However, it doesn't if our main purpose is comparing to the vaginal MB.
And if we label in units of GSD, it will be easy to read off from the graph what the correct mean eff GSD is for a given GSD.

How much more is the var in the vaginal comms versus stool?

```{r}
reps_me_summ %>%
  ggplot(aes(x = mean_efficiency, y = sim_type, color = body_site)) +
  scale_x_log10() +
  stat_summary() +
  expand_limits(x = 1)
```

```{r}
reps_me_summ %>%
  with_groups(c(body_site, sim_type), summarize, 
    across(mean_efficiency, gm_mean)) %>%
  pivot_wider(names_from = body_site, values_from = mean_efficiency) %>%
  mutate(vagina/stool)
```

## Variation in species

What else matters? Perhaps the variation in a species, relative to the variation in the mean efficiency.

For sufficiently prevalent taxa, look at the variation.

check: what do the min positive abundances look like for various species?

```{r}
x <- ps %>%
  filter_sample_data(body_subsite %in% c('stool', 'posterior_fornix')) %>%
  filter_taxa2(~ sum(.) > 0) %>%
  as_tibble
```
```{r}
x %>%
  filter(.abundance > 0) %>%
  with_groups(c(.otu, body_site), summarize, 
    min_pos = min(.abundance)) %>%
  ggplot(aes(x = min_pos, y = body_site, fill = body_site)) +
  scale_x_log10() +
  geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
```

how about by sample?

```{r}
x %>%
  filter(.abundance > 0) %>%
  with_groups(c(.sample, body_site), summarize, 
    min_pos = min(.abundance)) %>%
  ggplot(aes(x = min_pos, y = body_site, fill = body_site)) +
  scale_x_log10() +
  geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
```

get a filtered ps object for each type

```{r}
ps_list <- list(
  stool = ps %>% filter_sample_data(body_subsite == 'stool'),
  vagina = ps %>% filter_sample_data(body_subsite == 'posterior_fornix')
) %>%
  # TODO: consider filtering criterion
  map(filter_taxa2, ~ sum(. > 0) >= 2) %>%
  # map(filter_taxa2, ~ mean(. > 0) >= 0.05) %>%
  # map(filter_taxa2, ~ mean(. >= 1e-4) >= 0.10) %>%
  # map(filter_taxa2, ~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
  map(orient_taxa, as = 'cols')
ps_list %>% map_int(nsamples)
ps_list %>% map_int(ntaxa)
```

### presence/absence

```{r, eval = F}
x <- ps %>%
  filter_sample_data(body_subsite %in% c('stool', 'posterior_fornix')) %>%
  filter_taxa2(~ sum(.) > 0) %>%
  as_tibble
```

```{r, eval = T}
x <- ps_list %>%
  map_dfr(as_tibble)

min_vals <- c(0, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1)

prev <- x %>%
  with_groups(c(.otu, body_site), summarize,
    min = min_vals,
    prevalence = map_dbl(min_vals, ~mean(.abundance > .x))
  )
```
```{r}
prev %>%
  ggplot(aes(x = prevalence, y = as.factor(min), 
      fill = body_site, color = body_site)) +
  geom_density_ridges(alpha = 0.3)
```

```{r}
prev %>%
  ggplot(aes(x = prevalence, y = as.factor(min), 
      fill = body_site, color = body_site)) +
  geom_density_ridges(stat = 'binline', alpha = 0.3)
```

It's a bit hard to know what threshold to use to pick a relevant set of taxa.

```{r}
prev %>%
  filter(min < 1e-3) %>%
  ggplot(aes(x = prevalence, y = body_site,
      fill = body_site, color = body_site)) +
  facet_wrap(~min) +
  geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
```

### multiplicative variation

```{r}
ps_list_filt <- ps_list %>%
  map(filter_taxa2, ~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15)
ps_list_filt %>% map_int(ntaxa)
```

now, purpose is to get the geometric variation in species.

Problem: I expect the pseudocount will have a massive impact here, so should check different values.

```{r}
x <- ps_list_filt %>%
  map_dfr(as_tibble)

min_vals <- c(1e-6, 1e-5, 1e-4, 1e-3)

gsd <- x %>%
  with_groups(c(.otu, body_site), summarize,
    min = min_vals,
    gsd = map_dbl(min_vals, ~gm_sd(.abundance + .x))
  )
```

```{r}
gsd %>%
  ggplot(aes(x = gsd, y = as.factor(min), 
      fill = body_site, color = body_site)) +
  # facet_wrap(~body_site) +
  scale_x_log10() +
  expand_limits(x = 1) +
  geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
```

Note, the GSDs tend to be larger in the vaginal microbiome, consistent with our expectation based on its lower alpha diversity.

What is the difference between gut and vaginal GSDs?

```{r}
gsd %>%
  ggplot(aes(x = gsd, y = as.factor(min), 
      fill = body_site, color = body_site)) +
  # facet_wrap(~body_site) +
  scale_x_log10() +
  stat_summary() +
  expand_limits(x = 1)
```

```{r}
gsd %>%
  with_groups(c(body_site, min), summarize, across(gsd, gm_mean)) %>%
  pivot_wider(names_from = body_site, values_from = gsd) %>%
  mutate(vagina/stool)
```

The difference is not very dependent on the min abundance threshold.

### visualize

```{r, eval = F}
ps_list[[2]] %>%
  as_tibble %>%
  ggplot(aes(y = species, x = 1e-5 + .abundance)) +
  scale_x_log10() +
  # stat_slab(slab_type = 'histogram')
  stat_dots()
```

```{r, eval = F}
ps_list[[2]] %>%
  as_tibble %>%
  ggplot(aes(x = 1e-5 + .abundance)) +
  facet_wrap(~species, ncol = 1) +
  scale_y_log10() +
  scale_x_log10() +
  # stat_slab(slab_type = 'histogram')
  geom_histogram()
```

```{r, eval = F}
ps_list1[[1]] %>%
  as_tibble %>%
  ggplot(aes(x = 1e-5 + .abundance)) +
  facet_wrap(~species, ncol = 3) +
  # scale_y_log10() +
  scale_x_log10() +
  # stat_slab(slab_type = 'histogram')
  geom_histogram()
```

```{r}

ps_list1[[1]] %>%
  as_tibble %>%
```

HERE

## Gut: Variation in phylum proportions

If phyla vary substantially across samples, then any phylum-specific bias would drive variation in the mean efficiency.

I'll restrict plots to just the four main phyla.

```{r}
focal_phyla <- c('Actinobacteria', 'Bacteroidetes', 'Firmicutes',
  'Proteobacteria')
ps_phy <- ps %>% 
  subset_samples(body_subsite == 'stool') %>%
  tax_glom('phylum') %>%
  mutate_tax_table(.otu = phylum) %>%
  transform_sample_counts(close_elts) %>%
  mutate_tax_table(
    gram = case_when(
      phylum %in% c('Actinobacteria', 'Firmicutes') ~ 'Positive',
      TRUE ~ 'Negative'
    )
  )
x_phy <- ps_phy %>% as_tibble
```

```{r}
x_phy %>%
  filter(phylum %in% focal_phyla) %>%
  ggplot(aes(.abundance + 1e-5)) +
  facet_wrap(~phylum, scales = 'fixed', ncol = 1) +
  scale_x_log10() +
  geom_histogram()
```

There is substantial log-scale variation in all phyla, but the Bacteroidetes and Firmicutes are generally much more abundant than the other two.

What about the B:F ratio?

```{r}
x_phy %>%
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  mutate(
    across(.abundance, ~ . + 1e-4),
    ratio = .abundance / .abundance[phylum == 'Firmicutes']
  ) %>%
  filter(phylum == 'Bacteroidetes') %>%
  ggplot(aes(ratio)) +
  scale_x_log10(breaks = 10^(-3:4)) +
  geom_histogram() +
  theme_minimal_vgrid()
```

Suppose that these profiles are biased to favor Gram negative phyla.

```{r}
bias <- ps_phy %>% tax_table %>% as_tibble %>%
  mutate(
    efficiency = case_when(
      gram == 'Positive' ~ 1/10,
      TRUE ~ 1
    )
  ) %>%
  select(.otu, efficiency) %>%
  deframe
ps_phy_cal <- ps_phy %>% calibrate(bias)
```

```{r}
ps_phy_cal %>% 
  sample_data %>%
  as_tibble %>%
  ggplot(aes(.mean_efficiency)) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
bias %>% gm_sd
ps_phy_cal %>% sample_data %>% as_tibble %>% pull(.mean_efficiency) %>% gm_sd
```

Consider adjusting so that the GSD is 3, like in other simulations.
(Though should then stick to the same taxa).

### 2022-02-06 case-studies.Rmd
Revise gut microbiome case study

@@
 
-## Human gut microbiome
+## Human gut microbiomes
 
-The vaginal microbiome is characterised by low species diversity within and across samples, so that variation in efficiency among a few dominant species could drive large variation in the mean efficiency among samples.
-The human gut microbiome has substantially greater diversity both within and across samples, which we should expect to result in lower variation in mean efficiency.
-The reason is that the mean efficiency is a weighted average of efficiencies of individual species, and so will tend to vary less the more species that are averaged over.
-Here we use publically available gut microbiome profiles to consider how the distinct dynamics of the gut microbiome impact the variation in the mean efficiency and its potential to distort DA results.
+An overwhelming majority of applications of DA analysis are in studies of the human gut microbiome. (True?)
+Many are microbial association tests in a case-control design---looking for taxa that are differentially abundant in one group of subjects versus another.
+It is also increasingly common to perform time series experiments to elucidate dynamics of microbial communities within a host and their interaction with dynamic host variables.
+Although interest in other microorganisms and viruses is growing, so far these are mostly of bacteria.
+Were therefore asked whether there are features of human gut bacterial microbiomes that we might expect to make DA analyses more or less robust to taxonomic bias.
 
-First we considered variation in healthy subjects and asked how the distinct structures of vaginal and gut communities impacted mean efficiency variation.
-We considered vaginal and stool samples from the original Human Microbiome Project (@huttenhower2012stru); these and other results in this case study are based on species-level bacterial relative-abundances produced by Metaphlan3 and made available through the curatedMetagenomicData R package (CITE).
-We simulated taxonomic bias with a geometric standard deviation of 3, with and without phylogenetic conservation, 
-by sampling from a multivariate normal distribution, with efficiencies either IID or with a covariance equal to the phylogenetic covariance matrix to simulate bias as an evolving trait.
-We then compared the distribution of mean efficiency across samples.
-We also considered how species-level diversity patterns could help explain this variation.
+We first considered whether the greater diversity of the human gut microbiome as compared to the human vaginal microbiome might limit the impact of bias on species-level DA analyses.
+Intuitively, we expect the mean efficiency to vary less across samples from a given ecosystem type as alpha diversity increases, since it effectively averages over a larger number of species.
+Indeed, as shown in Appendix X, in randomly assembled communities the additive variance in the mean efficiency is inversely proportional to the order-2 _true diversity_ or _effective number of species_ in the community, which is equivalent to the popular Inverse Simpson alpha diversity index.
+But real communities do not assemble randomly.
+Moreover, it is the multiplicative variation in the mean efficiency relative to that in the species proportions that determines the impact bias has on proportion-based DA analyses; it is possible that the variation in individual species proportions is reduced to an even greater degree than that in the mean efficiency.
+Thus it is not obvious whether the gut microbiome's higher alpha diversity will effectively insulate its DA analyses from the effects of bias.
 
-As expected, the mean efficiency varied substantially less in the higher-diversity gut microbiome samples than in the vaginal microbiome samples (median geometric standard deviation of X versus Y); yet it remained non-negligible.
-There are several potential reasons why.
-First, although richness is typically around 100 bacterial species in the Metaphlan profiles, the order-2 effective number of species ($^2D$; equivalent to the Inverse Simpson diversity index) is typically only around 10 and often lower.
-This number is more relevant when we consider the reduction in variation when computing a weighted average over species (Appendix; standard deviation in mean efficiency scales as $1 / \sqrt(^2D)$  when efficiencies are IID).
+To investigate this question, we analyzed species-level bacterial profiles from vaginal and stool samples from healthy subjects that had been subjected to shotgun sequencing as part of the Human Microbiome Project (@huttenhower2012stru).[^We obtained MetaPhlAn3 species-level bacterial relative abundances from the curatedMetagenomicData R package.]
+Consistent with previous findings, the stool microbiomes had substantially greater diversity than the vaginal microbiomes by several measures, including: within individuals (alpha diversity) and in total across many individuals (gamma diversity); at the species level and by measures that account for phylogenetic divergence; and for diversity measures spanning a range of sensitivity to low abundance species as quantified by Hill's $q$, including richness ($q=0$), Shannon entropy ($q=1$), and the Inverse Simpson index ($q=2$).
+To understand how this greater diversity might modulate the effect of bias, we used computer simulation to assess the variation in the mean efficiency under a range of plausible taxonomic biases, with and without phylogenetic conservation of the efficiency trait.
+_In each simulation replicate, the species efficiencies were sampled from a multivariate log-normal distribution.
+Phylogenetic independence was achieved by setting the covariance matrix equal to the identity matrix, so that the efficiencies were IID.
+Phylogenetic conservation was achieved by setting the covariance matrix equal to the phylogenetic covariance matrix, which amounts to Brownian evolution of the log efficiency trait._
+As expected, the mean efficiency varied less in the higher-diversity gut microbiome samples than in the vaginal microbiome samples; the geometric standard deviation (GSD) of the mean efficiency in stool samples was on average 1.5X less than that in vaginal samples regardless of whether bias was phylogenetically conserved.
+This difference is not extreme and leaves substantial variation in the mean efficiency in gut samples.
+We suggest several possible explanations.
+First, although richness is typically around 100 bacterial species in the MetaPhlAn3 gut profiles, the more relevant order-2 effective number of species ($^2D$; equivalent to the Inverse Simpson diversity index) is typically only around 10 and often lower.
 A order-2 diversity that is much lower than species richness arises from the skewed distribution of abundances within a gut microbiome sample, with the most abundant 1-2 species often forming a majority of reads.
 It is these high-proportion species, captured by the order-2 diversity, that ultimately drive the variation in mean efficiency, not the large number of low-proportion species that drive the richness.
-A separate but related reason is that, although the total number of species across gut microbiomes is large, a small number of species form the vast majority of reads and tend to be the most abundant in individual samples.
+A second, related reason is that, like in the vaginal microbiome, a relatively small set of species tend to be the most abundant across individuals.
 The overall effect is similar, though less extreme, as that of the dominance of a few key species in the vaginal microbiome: Variation in the efficiencies among these species has an outsized impact on the mean efficiency, again leading to a weakening of the averaging affect when we consider variation across samples.
 
-A notable aspect of gut microbiome variation, shared with with the vaginal microbiome, is that the most abundant species tend to come from a small set of groups.
-In the HMP1 gut data, these are various Bacteroides and Prevotella species; in other gut datasets, Gram positives such as X and Y are also common.
-In patients with diarrhea or infections, we may see dominance of Enterobactericeae like E. coli and Klebsiella spp.
-More generally, there is phylogenetic conservation at the level of ecological traits.
-There is also reports of phylogenetic conservation at the level of the species efficiency for a given protocol.
-An implication is that we may have less 'effective diversity' from the perspective of variation in the mean effiency.
+We next compared the variation in species proportions in the two ecotypes.
+In assessing multiplicative variation, one must grapple with the sparsity of human microbiome data.
+Across both ecotypes, most species have low prevalence, appearing in just a small minority of samples.
+One way to handle these zero observations is to replace them with a small positive value, roughly representing a minimum detection limit.
+Although the multiplicative variation is highly sensitive to this value, we found that the relative difference between gut and vaginal samples was robust to values spanning $10^{-6}$ to $10^{-3}$.
+The average GSD of species in gut samples was around 1.8X lower than that of species in vaginal samples, regardless of zero-replacement value.
+Thus the variation in species proportions was lower in the gut samples by a similar or greater degree than the variation in mean efficiency, suggesting that bias may be just as or more problematic for inferring fold changes in species proportions.
 
-(Although this was not a big effect in the simulation)
+We sought to further understand the implications of the sparsity of gut microbiome for the effect of bias on DA analyses in the context of a real DA analysis.
+@vieirasilva2019quan analyzed variation in absolute abundance of genera in stool samples from patients with primary sclerosing cholangitis and/or inflammatory bowel disease.
+Absolute abundances of genera were obtained via the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)) with proportions measured from 16S sequencing and total abundance measured from flow cytometry.
+The authors the rank-based Spearman correlation to quantify the associations in absolute abundance and fecal calprotectin concentration, a biomarker of intestinal inflammation.
+[This will be described in Section \@ref(differential-abundance): Spearman correlation is equivalent to doing a regression on the rank-transformed data and is another form of DA analysis; changes in ranks of proportions are affected by bias, and changes in ranks of ratios are not.
+So rank-based methods (Wilcoxon for boolean covariate and Spearman for continuous covariate) have the same issue as LFC methods when it comes to analyzing proportions instead of ratios.]
+As described in Section \@ref(differential-abundance), taxonomic bias in the MGS measurement has a similar effect on analyses of absolute abundances from the total-abundance normalization as it does no analyses of the raw proportions.
+Thus we can consider whether bias might affect the Spearman correlations in the study.
+When we re-analyzed the Spearman correlations, we found that they were often driven by presence-absence of the given genus, rather than continuous variation in abundance, as evidenced by the fact that for many species, replacing the absolute genus abundances with 0 or 1 indicating presence or absence often did not affect the Spearman correlation.
+For these taxa, we can expect bias to have much less impact on the reported correlation than for taxa for which the correlation was driven by continuous variation.
+The reason is that systematic variation of the mean efficiency with the covariate (which was, in contrast, log-normally distributed) will tend to systematically shift the ranks of observations in a taxon with a continuously varying abundance, but not one where the rank correlation is driven by the proportion of non-zero observations.
+We confirmed this expectation by investigating the sensitivity of Spearman correlations for the top associations that were driven either by presence/absence variation or continuous variation, and found that the correlation was much more sensitive to bias in the second case.
 
-Might we see correlations between mean efficency and covariates?
-One way is for a small number of taxonomic groups to cause major shifts, and also have conserved efficiencies.
-An example occurs in an dataset we examined of accurate diarrhaea in children from Dhakka.
-Stool from children experiencing accute diarrhea were overwhelmingly dominated by Preoteobacteria, whereas stool from healthy controls were dominated by a roughly even mixture of Actinobacteria, Firmicutes, and Bacteroides.
-The Proteobacteria species in the diarrheaic stool were dominated by E. coli and to a lesser extent Klebsiella spp, all in the Enterobactericeae family.
-Hence a Gram+/- bias, a bias favoring Enterobactericeae, or a bias favoring E. coli would all be capable of driving systematic variation in the mean efficiency in this study.
-
-On the other hand, in other datasets we examined, the variation in phylum proportions that occured was primarily not associated with the clinical covariate.
-In these cases, a Gram or phylum-level bias might primary add noise rather than systematic error, depending on the sample size of the study and the magnitude of true effects.
-
-(Perhaps put this in the conclusion.)
-There is another reason to consider that the effect of bias should not be dismissed in the higher-diversity natural ecosystems like the gut.
-So far we have only considered the variation in mean efficiency; however, it is the variation in mean efficiency relative to the true effect sizes that determines the signficance of the error.
-In conditions where the mean efficiency varies less, we should also expect lower true effect sizes, so it isn't at all obvious that we should be less worried.
+<!-- The impact of bias in an association study depends on whether the mean efficiency is associated with the clinical covariate. -->
+_We might consider adding: A paragraph giving examples where there are large taxonomic shifts associated with a clinical covariate, which therefore have the potential to drive systematic variation in the mean efficiency.
+The primary positive example I have looked at is a study of children with acute diarrhea, in which stool from healthy controls has a microbiome with roughly equal amounts of Actinobacteria (Gram positive), Firmicutes (Gram positive), and Bacteroidetes (Gram negative), and Proteobacteria (Gram negative), but the diarrheic stool is dominated by Proteobacteria and specifically E.
+coli and to a lesser extent Klebsiella (both in Enterobacteriaceae, which is known to be causally associated with acute diarrhea generally).
+Any particular bias favoring E. coli or Enterobacteriaceae would have the potential to distort results, which is something we could simulate and visualize._
 

### 2022-02-06 differential-abundance.Rmd
Change differential-abundance section title

@@
-# Taxonomic bias can cause errors in differential-abundance results {#differential-abundance}
+# DA methods vary in their robustness to bias {#differential-abundance}
+
+<!-- # Differential-abundance methods vary in their robustness to taxonomic bias  -->
+
+<!-- TODO: Change the GC to be about how there are various methods; state the scope that we consider; and state the punchline that they vary, with proportion-based being more sensitive. -->
 

### 2022-02-07 conclusion.Rmd
Revise conclusion
Improve previous text and add text on future directions.

--- 2022-01-16 conclusion.Rmd
@@
 
-_This section is a placeholder for a short (2-3 paragraph) conclusion. We are planning to keep 'discussion' to the earlier sections and appendix, at least for initial submission._
+It is commonly thought that analyses of the differences between samples are robust to the taxonomic bias inherent in MGS measurement as long as all samples have been subjected to the same MGS workflow.
+But our results show that consistent taxonomic bias is still capable of creating scientific errors in common forms of DA analysis.
+In particular, we showed mathematically that bias affects DA analysis methods based on proportions or in absolute abundances derived from them (using total-abundance normalization), due to variation in the mean efficiency across samples of varying taxonomic composition.
+<!-- This is assuming consistent bias at the taxonomic level of analysis; things could be worse if we're analyzing higher taxonomic units than at which bias is conserved. -->
+<!-- Perhaps state that we showed this theoretically, in mock communities, and in real case studies -->
+The error, however, varies by experimental context and may often be negligible in a practical sense.
+<!-- It is negligible if the mean efficiency is stable across samples, and may be unimportant if it is not correlated with the focal covariate in a regression analysis. -->
+Moreover, it can be quantified and corrected through the use of reference-taxon measurements and/or community calibration controls.
+In addition, bias-aware sensitivity and meta-analyses make it possible to rigorously account for bias even in the absence of such controls.
+Applications of these methods will gradually improve our understanding of the experimental contexts in which variation in the mean efficiency is likely to be problematic and should be accounted for.
+
+We also showed that other DA methods are more robust---perhaps even entirely invariant---to consistent taxonomic bias.
+In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
+In addition, careful pairing of total-abundance and MGS measurement methods may remove most of the bias-driven error in absolute abundances from total-abundance normalization.
+
+Important open questions and future research directions remain.
+<!-- Taxonomic aggregation and non-FC analyses are left out because I'm assuming they will be addressed in the manuscript -->
+Key open questions include 1) determining the extent to which bias is consistent across samples at the species level for a given MGS method and sample type; 2) assessing the validity of post-extraction absolute-abundance controls as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how our assumptions about underlying community dynamics, and in particular what information zero counts provide, affect bias sensitivity.
+In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, better statistical tools are needed that are capable of including taxonomic bias and uncertainty in the MGS and supplemental non-MGS measurements.
+Finally, more concrete experimental recommendations and user-friendly statistical workflows are needed for implementing the solutions we propose in various experimental contexts.
+
+Our theoretical framework and example analyses provide a foundation for addressing these open questions and developing a practical experimental protocols and statistical tools.
+We look to a future in which microbiome researchers routinely choose an appropriate combination of experimental and data-analytic methods capable of answering their fundamental question while accounting for taxonomic bias and the other inherent limitations of MGS measurements---and in this way, gaining the confidence needed to codify statistical findings into true scientific knowledge.

### 2022-02-07 notebook/_posts/2020-11-12-leopold2020host-bias-sensitivity/index.Rmd
Update post metadata
Fixed date in YAML to match folder date; removed superfluous 'R' tag

@@
 author: Michael McLaren
-date: '2019-10-21'
+date: '2020-11-10'
 categories:
   - bias sensitivity
-  - R
   - ref:leopold2020host
@@
 # Download data from https://zenodo.org/record/3872145
-data_path <- here::here("_data", "leopold2020host")
+data_path <- here::here("notebook/_data", "leopold2020host")
 # To use a temporary directory:

### 2022-02-08 notebook/case-studies.Rmd
Create summary page and listings for case studies
Add listing pages for each case study, which contain all the posts for
that case study; and a 'Case studies' page that will provide a summary
with links of all analyses for each case study, but for now just links
to the listing pages (or the single post for the marine case study).

---
title: "Case studies"
description: |
  Overview of case-study analyses.
author:
  - name: Michael R. McLaren
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Foliar fungi experiment

[All posts](case-study-fungi.html)

# Vaginal microbiomes of pregnant women

[All posts](case-study-momspi.html)

# Human gut microbiomes

[All posts](case-study-gut.html)

# Microbial growth in marine sediments

[Case study analysis](posts/2022-01-08-lloyd2020evi-case-study/index.html)

### 2022-02-09 notebook/_posts/2022-01-20-vieirasilva2019quan/_cut.Rmd
Revise and render vieirasilva2019quan analysis
Analysis of bias sensitivity of Spearman correlations on QMP profiles
from vieirasilva2019quan.

--- 2022-02-06 notebook/_posts/2022-01-20-vieirasilva2019quan/vieirasilva2019quan.Rmd
@@
----
-title: "Explore Vieira-Silva et al (2019) QMP profiles"
-description: |
-author:
-  - name: Michael R. McLaren
-    url: {}
-categories:
-  - vieirasilva2019quan
-date: 2022-01-20
-draft: false
-bibliography: ../../../main.bib
-output:
-  distill::distill_article:
-    self_contained: false
-    dev: svg
-    toc: true
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = TRUE,
-  cache = TRUE,
-  autodep = TRUE,
-  cache.comments = FALSE,
-  dpi = 300
-)
-```
-
-# Setup
-
-## R setup
-
-```{r libraries}
-library(here)
-# Tools for microbiome data
-library(speedyseq)
-# Tools for general purpose data manipulation and plotting
-library(tidyverse)
-library(fs)
-# ggplot helpers
-library(ggbeeswarm)
-library(cowplot)
-library(patchwork)
-library(scales)
-# stats helpers
-# library(broom)
-
-library(metacal); packageVersion("metacal")
-```
-
-```{r ggplot}
-theme_set(theme_cowplot())
-```
-
-## Data setup
-
-```{r}
-data_path <- here('notebook/_data/vieirasilva2019quan')
-dir_create(data_path)
-# Supplementary tables from the nature website; contains the metadata
-fn <- path(data_path, 'supplementary-data-1.xlsx')
-if (!file_exists(fn)) {
-  download.file(
-    'https://static-content.springer.com/esm/art%3A10.1038%2Fs41564-019-0483-9/MediaObjects/41564_2019_483_MOESM3_ESM.xlsx',
-    fn
-  )
-  file_chmod(fn, '-w')
-}
-# QMP proiles from the Raes lab website
-fn <- path(data_path, 'qmp-matrix-genus.tsv')
-if (!file_exists(fn)) {
-  download.file('http://raeslab.org/software/QMP2/QMP.matrix.tsv', fn)
-  file_chmod(fn, '-w')
-}
-```
-
-Note, there is an extra value in the BMI column of Table S1 (Excel spreadsheet).
-
-```{r}
-otu <- path(data_path, 'qmp-matrix-genus.tsv') %>%
-  read.table() %>%
-  as('matrix') %>%
-  otu_table(taxa_are_rows = FALSE)
-sam <- path(data_path, 'supplementary-data-1.xlsx') %>%
-  readxl::read_excel(sheet = 'Table S1', range = 'A2:L174', na = "NA")
-original_column_names <- names(sam)
-ps <- phyloseq(
-  otu, 
-  sam %>% janitor::clean_names() %>% sample_data
-)
-stopifnot(setequal(sample_names(ps), sample_names(otu)))
-sam <- ps %>% sample_data %>% as_tibble %>% glimpse
-```
-
 ### Taxonomy
@@
 https://rrndb.umms.med.umich.edu/static/download/
-
-# Explore
-
-I think that the QMP profiles are such that the sum within a sample should equal the reported cell counts per gram in the sample data; let's check.
-
-```{r}
-all.equal(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
-```
-
-This small difference may be down to rounding error, since it seems that the entires in the QMP matrix are rounded to integers.
-
-```{r}
-otu %>% c %>% {all.equal(., round(.))}
-sam$average_faecal_cell_count_cells_g %>% {all.equal(., round(.))}
-```
-
-There are `r nsamples(ps)`, with the following disease breakdown:
-
-```{r}
-sam %>% count(diagnosis)
-```
-
-Can we recover counts?
-We can't recover the raw counts without using the CNV information.
-But setting that aside for now, let's confirm that we understand the normalization that was used -
-
-```{r}
-x <- sam %>%
-  add_column(sample_sum = sample_sums(ps))
-```
-```{r}
-x %>% pull(sample_sum) %>% summary
-x %>% transmute(sample_sum / average_faecal_cell_count_cells_g) %>% pull(1) %>% summary
-```
-
-(we already knew this - the samples sum up to the cell density).
-
-# Reproduce paper analysis
-
-Let's try to reproduce the results from the paper.
-
-> Applying QMP on the combined PSC/IBD/mHC data set, we identified 11 genera as significantly correlat-ing  with  faecal  calprotectin  concentrations  (Spearman  correlation,  n= 149, FDR < 0.05; Fig.
-3a and Supplementary Table 6).
-Stool mois-ture content had 45 associated taxa (Spearman correlation, n= 143, FDR <    0.05;  Supplementary  Table  6).
-Combining  both  variables,  only  6  out  of  11  positive(+)  or  negative(−)  calprotectin  associations  remained significant when controlling for faecal water content vari-ation: Anaerostipes(−), Escherichia(+), Fusobacterium(+), Gemmiger(−), Streptococcus(+)  and  Veillonella(+)  (nested  linear  model  on  QMP  rank  abundances,  n= 133,  FDR < 0.05;  Fig.
-3a  and  Supplementary  Table  6).
-
-First, let's look for taxa associated with faecal calprotectin concentrations
-
-```{r}
-x <- ps %>% as_tibble
-```
-
-```{r}
-y <- ps %>%
-  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
-  as_tibble %>%
-  with_groups(.otu, nest) %>%
-  mutate(
-    spearman_test = map(data, 
-      ~cor.test(data = .x,
-        ~ .abundance + faecal_calprotectin_mg_g, 
-        method = 'spearman', # exact = FALSE, continuity = TRUE,
-        )),
-    spearman_test = map(spearman_test, broom::tidy)
-  ) %>%
-  select(-data) %>%
-  unnest(spearman_test) %>%
-  mutate(p.bh = p.adjust(p.value, method = 'BH'))
-```
-
-```{r}
-y %>% arrange(p.bh) %>% print(n=20)
-```
-
-```{r}
-study_res <- path(data_path, 'supplementary-data-1.xlsx') %>%
-  readxl::read_excel(sheet = 'Table S6', na = "NA")
-study_res %>%
-  select(Genera, starts_with('Calprotectin') & ends_with('QMP'))
-y %>% arrange(p.bh) %>% filter(!str_detect(.otu, 'unclassified')) %>%
-  select(.otu, estimate, p.value, p.bh) %>%
-  print(n=15)
-```
-
-These estimates match, but the p values (and their ranking) differs.
-
-figure this out; then try to recapitulate using rank regression.
-
-Note, a decrease FDR-corrected p values can be expected if the FDR correction was done _after_ discarding the OTUs without genus classifications.
-
-can check the code at https://github.com/raeslab/QMP/
-
-```{r}
-study_res %>%
-  select(Genera, starts_with('Calprotectin') & ends_with('QMP')) %>%
-  arrange(4) %>%
-  print(n=15)
-y %>% 
-  arrange(p.bh) %>% 
-  filter(!str_detect(.otu, 'unclassified')) %>%
-  mutate(
-    p.bh = p.adjust(p.value, method = 'BH'),
-    across(starts_with('p.'), round, digits = 4)
-  ) %>%
-  select(.otu, estimate, p.value, p.bh) %>%
-  print(n=15)
-```
-
-ahhh - ok their's is sorted differently, which makes sense since it has many tests in one table. and the p values need to be rounded to 4 digits. then it seems like we're good.
-
-next:
-
-- also do for moisture, and perhaps RMP, to check these results as well
-- confirm I can get similar result with linear model on ranks
-<!--  -->
-
-visually inspect
-
-```{r}
-top_hits <- y %>% 
-  filter(!str_detect(.otu, 'unclassified')) %>%
-  slice_min(p.bh, n = 9)
-x <- ps %>%
-  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
-  prune_taxa(top_hits$.otu, .) %>%
-  as_tibble
-
-x %>%
-  mutate(across(.otu, factor, levels = top_hits$.otu)) %>%
-  with_groups(.otu, mutate,
-    across(c(faecal_calprotectin_mg_g, .abundance), rank),
-  ) %>%
-  # mutate(across(c(faecal_calprotectin_mg_g, .abundance), rank, ties.method = 'average')) %>%
-  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
-  facet_wrap(~.otu) +
-  geom_point() +
-  stat_smooth(method = 'lm')
-```
-
-It seems like zeros or below-detection-limit values are largely responsible for the observed results.
-
-```{r}
-x %>%
-  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
-  scale_x_log10() +
-  scale_y_log10() +
-  facet_wrap(~.otu) +
-  geom_point()
-```
 
@@
 
-### Check if patterns are driven by presence/absence
+# saved plotting stuff
+
+Mean efficiency distribution:
 
 ```{r}
-y <- ps %>%
-  # filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
-  # prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
-  as_tibble %>%
-  with_groups(.sample, mutate, 
-    proportion = close_elts(.abundance),
-    present = (.abundance > 0) * 1
-  ) %>%
-  with_groups(.otu, nest) %>%
-  mutate(
-    spearman = map(data, 
-      ~cor.test(data = .x,
-        ~ .abundance + faecal_calprotectin_mg_g, 
-        method = 'spearman')),
-    spearman_pa = map(data, 
-      ~cor.test(data = .x,
-        ~ present + faecal_calprotectin_mg_g, 
-        method = 'spearman')),
-  ) %>%
-  select(-data) %>%
-  pivot_longer(contains('spearman'), names_to = 'test', values_to = 'fit') %>%
-  mutate(across(fit, map, broom::tidy)) %>%
-  unnest(fit) %>%
-  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
+x <- reps %>%
+  slice(1) %>%
+  pull(actual) %>%
+  .[[1]] %>%
+  sample_data %>%
+  as_tibble
+
+x %>%
+  ggplot(aes(x = mean_efficiency, fill = faecal_calprotectin> 50)) +
+  scale_x_log10() +
+  geom_histogram()
 ```
@@
 ```{r}
-y.w <- y %>%
-  select(.otu, test, estimate, p.bh) %>%
-  pivot_wider(names_from = c(test), values_from = c(estimate, p.bh), 
-    names_glue = '{test}_{.value}', names_sort = TRUE) %>%
-  select(.otu, sort(tidyselect::peek_vars())) %>%
-  mutate(across(where(is.numeric), round, 4)) %>%
-  arrange(spearman_p.bh)
+x %>%
+  ggplot(aes(x = faecal_calprotectin, y = mean_efficiency, color = diagnosis)) +
+  scale_y_log10() +
+  scale_x_log10() +
+  geom_point() +
+  stat_smooth(aes(group = 1), method = 'lm')
 ```
 
+# Checks and notes
+
 ```{r}
-y.w %>%
-  ggplot(aes(spearman_estimate, spearman_pa_estimate)) +
-  geom_point()
+# check na handling
+x = tibble(a = c(1,2,3,NA,4,5), b = c(6,5,4,3,2,NA))
+cor.test(~a + b, data = x, method = 'spearman') %>% broom::tidy()
+cor.test(~a + b, data = x %>% drop_na, method = 'spearman') %>% broom::tidy()
 ```
 
-For some taxa (like Veillonella, EscherichiaShigella) the two estimates are very similar;for others like Streptococcus they are quite different.
-I suspect that simulated bias will have a larger impact on Streptococcus.
+Note, I'm not sure what the RMP values used in the original study were.
+From the manuscript, it's not obvious if the unclassified taxa were removed before the proportions were computed, or if copy number correction was used.
 
-# Bias sensitivity analysis
-
-```{r, eval = F}
-# Perturb by bias vector - treat ps as actual
-ps.per <- perturb(ps, bias, norm = 'keep')
-# Calibrate by bias vector - treat ps as observed
-ps.cal <- calibrate(ps, bias, norm = 'keep')
-```
+## Original bias sens analysis
 
@@
   spearman_test <- cor.test(data = data,
-    ~ .abundance + faecal_calprotectin_mg_g, 
+    ~ .abundance + faecal_calprotectin, 
     method = 'spearman', exact = FALSE
@@
 ```
-
-Example where the bias sensitivity analysis shows little effect of varying the hypothesized bias.
-
-We can probably understand this finding because the original result is driven by the presence/absence pattern, and we have treated zeros as absolute, so that presence/absence is maintained.
-If we relaxed this assumption, then I suspect that we would see much more variation in the estimates.
-Much of this variation, however, would come from relaxing our assumption about the zeros, rather than relaxing our assumption that there is no bias.
-
-Challenge with this data is that we don't have the read counts, so don't know how to replace zeros with resonable values.
-
-## saved plotting stuff
-
-Mean efficiency distribution:
-
-```{r}
-x <- reps %>%
-  slice(1) %>%
-  pull(actual) %>%
-  .[[1]] %>%
-  sample_data %>%
-  as_tibble
-
-x %>%
-  ggplot(aes(x = mean_efficiency, fill = faecal_calprotectin_mg_g > 50)) +
-  scale_x_log10() +
-  geom_histogram()
-```
-
-```{r}
-x %>%
-  ggplot(aes(x = faecal_calprotectin_mg_g, y = mean_efficiency, color = diagnosis)) +
-  scale_y_log10() +
-  scale_x_log10() +
-  geom_point() +
-  stat_smooth(aes(group = 1), method = 'lm')
-```
-
-Now, redo the analysis with and without bias
-
-what assumptions to make about zeros?

### 2022-02-09 notebook/_posts/2022-01-20-vieirasilva2019quan/vieirasilva2019quan.Rmd
Revise and render vieirasilva2019quan analysis
Analysis of bias sensitivity of Spearman correlations on QMP profiles
from vieirasilva2019quan.

@@
 ---
-title: "Explore Vieira-Silva et al (2019) QMP profiles"
+title: "Explore bias sensitivity of Vieira-Silva et al (2019) QMP analysis"
 description: |
@@
 categories:
-  - vieirasilva2019quan
+  - ref:vieirasilva2019quan
+  - bias sensitivity
+  - absolute abundance
 date: 2022-01-20
@@
 
+This document reanalyzes the 'Quantitative Microbiome Profiles' of @vieirasilva2019quan.
+
 # Setup
@@
 library(ggbeeswarm)
+library(ggridges)
 library(cowplot)
@@
 ## Data setup
+
+The genus-level 'Quantitative Microbiome Profiles' from @vieirasilva2019quan are available for download at http://raeslab.org/software/QMP2/, and the sample metadata is available in the supplemental content at the journal site.
 
@@
 
-Note, there is an extra value in the BMI column of Table S1 (Excel spreadsheet).
+I'll read in the metadata and QMP profiles into a phyloseq object.
+The metadata is Table S1 of the Excel spreadsheet.
+There is an extra value in the BMI column of Table S1 that will be dropped in the import step below.
 
@@
   sam %>% janitor::clean_names() %>% sample_data
+) %>%
+  rename_sample_data(
+   faecal_calprotectin = faecal_calprotectin_mg_g,
+   moisture_content = moisture_content_percent
   )
@@
 
-### Taxonomy
+# Explore data
 
-how can I get taxonomy and/or phylogeny for the genus identifiers?
-
-First, create a file with the RDP 16 taxonomy from the DADA2 training file
-
-```
-zcat rdp_train_set_16.fa.gz | grep '^>' | sort | uniq -u | sed -e 's/^>//' -e 's/;$//' - > rdp_train_set_16_headers.txt
-```
-
-```{r}
-cns <- c('domain', 'phylum', 'class', 'order', 'family', 'genus')
-cts <- 'cccccc'
-# x <- path('~/data/16s-taxonomy-databases/dada2/rdp-trainset-16',
-#   'rdp_train_set_16_headers.txt') %>%
-#   read_delim(delim = ';', col_names = cns, col_types = cts)
-x <- path('~/data/16s-taxonomy-databases/dada2/rdp-trainset-16',
-  'rdp_train_set_16_headers.txt') %>%
-  read_lines %>%
-  enframe %>%
-  select(-name) %>%
-  separate(value, into = cns, sep = ';', fill = 'right') %>%
-  glimpse
-```
-
-```{r}
-genera <- taxa_names(ps) %>% 
-  str_subset('unclassified', negate = TRUE)
-genera %in% x$genus
-setdiff(genera, x$genus)
-
-unclassified_taxa <- taxa_names(ps) %>% 
-  str_subset('unclassified') %>%
-  str_extract('(?<=unclassified_).+')
-unclassified_taxa %in% x$family
-unclassified_taxa %in% x$order
-unclassified_taxa %in% x$phylum
-unclassified_taxa %in% x$domain
-```
-
-try again
-
-```{r}
-rnks <- c('domain', 'phylum', 'class', 'order', 'family', 'genus')
-x <- path('~/data/16s-taxonomy-databases/rdp/', 
-  'RDPClassifier_16S_trainsetNo16_rawtrainingdata',
-  'trainset16_022016_headers.txt') %>%
-  read_tsv(col_names = c('id', 'path'))
-y <- x %>%
-  separate(path, c('root', rnks), sep = ';', fill = 'right')
-
-```
-
-ok, giving up..
-
-hmm - check to see if can get from the rrndb tables?
-https://rrndb.umms.med.umich.edu/static/download/
-
-# Explore
-
-I think that the QMP profiles are such that the sum within a sample should equal the reported cell counts per gram in the sample data; let's check.
-
-```{r}
-all.equal(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
-```
-
-This small difference may be down to rounding error, since it seems that the entires in the QMP matrix are rounded to integers.
-
-```{r}
-otu %>% c %>% {all.equal(., round(.))}
-sam$average_faecal_cell_count_cells_g %>% {all.equal(., round(.))}
-```
-
-There are `r nsamples(ps)`, with the following disease breakdown:
+There are `r nsamples(ps)`, with the following counts by disease status,
 
@@
 
-Can we recover counts?
-We can't recover the raw counts without using the CNV information.
-But setting that aside for now, let's confirm that we understand the normalization that was used -
+The QMP profiles are such that the sum within a sample should equal the reported cell counts per gram in the sample data; let's check.
 
 ```{r}
-x <- sam %>%
-  add_column(sample_sum = sample_sums(ps))
-```
-```{r}
-x %>% pull(sample_sum) %>% summary
-x %>% transmute(sample_sum / average_faecal_cell_count_cells_g) %>% pull(1) %>% summary
+all.equal(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
+cor(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
 ```
 
-(we already knew this - the samples sum up to the cell density).
+This small difference may be down to rounding error, since apparently the entires in the QMP matrix have been rounded to integers:
 
-# Reproduce paper analysis
+```{r}
+otu %>% c %>% {all.equal(., round(.))}
+```
+
+It would be nice to have read counts; however, I don't think it's possible to obtain read counts from the available information here.
+Perhaps it would be possible to reverse the QMP calculation if we had the copy-number-variation information.
+
+Since we lack read-count information, we will ignore the sequencing-count noise, which was also done in the original paper.
+
+Our primary covariates of interest are Faecal calprotectin and moisture content.
+
+```{r}
+sam %>% select(faecal_calprotectin, moisture_content) %>% summary
+```
+
+```{r}
+sam %>%
+  ggplot(aes(x = faecal_calprotectin)) +
+  scale_x_log10() +
+  geom_histogram()
+```
+
+Note that a large fraction of calprotectin measurements are precisely at 30, suggesting this is a value assigned to samples that were below the detection limit of the assay.
+Also, the values vary over multiple orders of magnitude.
+
+```{r}
+sam %>%
+  ggplot(aes(x = moisture_content)) +
+  geom_histogram()
+```
+
+In contrast, the moisture varies relatively continuously
+
+# Reproduce QMP Spearman analysis
 
 Let's try to reproduce the results from the paper.
+We'll focus on the Spearman correlations of the genera absolute abundances with faecal calprotectin and stool moisture content.
+The relevant manuscript text is
 
@@
 
-First, let's look for taxa associated with faecal calprotectin concentrations
-
-```{r}
-x <- ps %>% as_tibble
-```
-
-```{r}
-y <- ps %>%
-  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
-  as_tibble %>%
-  with_groups(.otu, nest) %>%
-  mutate(
-    spearman_test = map(data, 
-      ~cor.test(data = .x,
-        ~ .abundance + faecal_calprotectin_mg_g, 
-        method = 'spearman', # exact = FALSE, continuity = TRUE,
-        )),
-    spearman_test = map(spearman_test, broom::tidy)
-  ) %>%
-  select(-data) %>%
-  unnest(spearman_test) %>%
-  mutate(p.bh = p.adjust(p.value, method = 'BH'))
-```
-
-```{r}
-y %>% arrange(p.bh) %>% print(n=20)
-```
+A summary table with the Spearman results are Table S6 of the supplementary data Excel sheet,
 
@@
   readxl::read_excel(sheet = 'Table S6', na = "NA")
-study_res %>%
-  select(Genera, starts_with('Calprotectin') & ends_with('QMP'))
-y %>% arrange(p.bh) %>% filter(!str_detect(.otu, 'unclassified')) %>%
-  select(.otu, estimate, p.value, p.bh) %>%
-  print(n=15)
-```
-
-These estimates match, but the p values (and their ranking) differs.
-
-figure this out; then try to recapitulate using rank regression.
-
-Note, a decrease FDR-corrected p values can be expected if the FDR correction was done _after_ discarding the OTUs without genus classifications.
-
-can check the code at https://github.com/raeslab/QMP/
-
-```{r}
-study_res %>%
-  select(Genera, starts_with('Calprotectin') & ends_with('QMP')) %>%
-  arrange(4) %>%
-  print(n=15)
-y %>% 
-  arrange(p.bh) %>% 
-  filter(!str_detect(.otu, 'unclassified')) %>%
-  mutate(
-    p.bh = p.adjust(p.value, method = 'BH'),
-    across(starts_with('p.'), round, digits = 4)
-  ) %>%
-  select(.otu, estimate, p.value, p.bh) %>%
-  print(n=15)
-```
-
-ahhh - ok their's is sorted differently, which makes sense since it has many tests in one table. and the p values need to be rounded to 4 digits. then it seems like we're good.
-
-next:
-
-- also do for moisture, and perhaps RMP, to check these results as well
-- confirm I can get similar result with linear model on ranks
-<!--  -->
-
-visually inspect
-
-```{r}
-top_hits <- y %>% 
-  filter(!str_detect(.otu, 'unclassified')) %>%
-  slice_min(p.bh, n = 9)
-x <- ps %>%
-  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
-  prune_taxa(top_hits$.otu, .) %>%
-  as_tibble
-
-x %>%
-  mutate(across(.otu, factor, levels = top_hits$.otu)) %>%
-  with_groups(.otu, mutate,
-    across(c(faecal_calprotectin_mg_g, .abundance), rank),
-  ) %>%
-  # mutate(across(c(faecal_calprotectin_mg_g, .abundance), rank, ties.method = 'average')) %>%
-  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
-  facet_wrap(~.otu) +
-  geom_point() +
-  stat_smooth(method = 'lm')
-```
-
-It seems like zeros or below-detection-limit values are largely responsible for the observed results.
-
-```{r}
-x %>%
-  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
-  scale_x_log10() +
-  scale_y_log10() +
-  facet_wrap(~.otu) +
-  geom_point()
-```
-
-### spearman versus linear modeling
-
-NOTE: The results differ when I first omit the samples with missing calprotectin measurements, or not.
-perhaps because the ranks of the abundances differ?
-
-```{r}
-z <- ps %>%
-  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
-  as_tibble %>%
-  with_groups(.sample, mutate, proportion = close_elts(.abundance)) %>%
-  with_groups(.otu, mutate, 
-    across(c(.abundance, proportion, faecal_calprotectin_mg_g), 
-      rank, ties.method = 'average',
-      .names = '{.col}_rank')
-  ) %>%
-  with_groups(.otu, nest) %>%
-  mutate(
-    spearman = map(data, 
-      ~cor.test(data = .x,
-        ~ .abundance + faecal_calprotectin_mg_g, 
-        method = 'spearman', na.action = na.omit)),
-    lm = map(data, 
-      ~lm(data = .x, .abundance_rank ~ faecal_calprotectin_mg_g_rank,
-        na.action = na.omit)),
-  ) %>%
-  select(-data) %>%
-  pivot_longer(c(spearman, lm), names_to = 'test', values_to = 'fit') %>%
-  mutate(across(fit, map, broom::tidy)) %>%
-  unnest(fit) %>%
-  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
-z1 <- z %>% filter(test == 'spearman' | term != '(Intercept)')
-```
-
-NOTE: The p values are nearly identical as long as I first drop the samples with missing data.
-But not if I don't.
-
-```{r}
-z1 %>%
-  select(.otu, test, p.value) %>%
-  pivot_wider(names_from = test, values_from = p.value) %>%
-  ggplot(aes(spearman, lm)) +
-  geom_abline(color = 'darkgrey') +
-  scale_x_log10() +
-  scale_y_log10() +
-  geom_point()
 ```
@@
 ```{r}
-z1 %>%
-  select(.otu, test, estimate) %>%
-  pivot_wider(names_from = test, values_from = estimate) %>%
-  ggplot(aes(spearman, lm)) +
-  geom_abline(color = 'darkgrey') +
-  # scale_x_log10() +
-  # scale_y_log10() +
-  geom_point()
-```
-
-The estimates are related but differ - I suspect this has to do with ties in the abundance (0s) and in the calprotectin (min values / detection limit).
-Perhaps the problematic taxa are very low prevalence?
-
-### try all tests
-
-```{r}
-y <- ps %>%
-  # filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
-  # prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
+qmp_spearman <- ps %>%
+  prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
   as_tibble %>%
-  with_groups(.sample, mutate, proportion = close_elts(.abundance)) %>%
   with_groups(.otu, nest) %>%
   mutate(
-    calprotectin_qmp_spearman = map(data, 
+    qmp_calprotectin_spearman = map(data, 
       ~cor.test(data = .x,
-        ~ .abundance + faecal_calprotectin_mg_g, 
-        method = 'spearman')),
-    calprotectin_rmp_spearman = map(data, 
+        ~ .abundance + faecal_calprotectin, 
+        method = 'spearman', exact = FALSE, continuity = TRUE)),
+    qmp_moisture_spearman = map(data, 
       ~cor.test(data = .x,
-        ~ proportion + faecal_calprotectin_mg_g, 
-        method = 'spearman')),
-    moisture_qmp_spearman = map(data, 
-      ~cor.test(data = .x,
-        ~ .abundance + moisture_content_percent, 
-        method = 'spearman')),
-    moisture_rmp_spearman = map(data, 
-      ~cor.test(data = .x,
-        ~ proportion + moisture_content_percent, 
-        method = 'spearman')),
+        ~ .abundance + moisture_content, 
+        method = 'spearman', exact = FALSE, continuity = TRUE)),
   ) %>%
@@
 
-TODO: check that samples with NAs are handled in the spearman tests
+Compare Calprotectin results,
 
 ```{r}
-y.w <- y %>%
-  select(.otu, test, estimate, p.bh) %>%
-  pivot_wider(names_from = c(test), values_from = c(estimate, p.bh), 
-    names_glue = '{test}_{.value}', names_sort = TRUE) %>%
-  select(.otu, sort(tidyselect::peek_vars())) %>%
-  mutate(across(where(is.numeric), round, 4)) %>%
-  arrange(calprotectin_qmp_spearman_p.bh)
+study_res %>%
+  select(Genera, starts_with('Calprotectin') & ends_with('QMP')) %>%
+  arrange(`Calprotectin  Spearman.P QMP`) %>%
+  slice_head(n=12)
+qmp_spearman %>%
+  filter(test == 'qmp_calprotectin_spearman') %>%
+  arrange(p.value) %>% 
+  select(.otu, estimate, p.value, p.bh) %>%
+  slice_head(n=12)
 ```
 
-Note, I'm not sure what the RMP values used in the original study were.
-From the manuscript, it's not obvious if the unclassified taxa were removed before the proportions were computed, or if copy number correction was used.
+The results are similar but not identical.
 
-### Check if patterns are driven by presence/absence
+Compare moisture results,
 
 ```{r}
-y <- ps %>%
-  # filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
-  # prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
+study_res %>%
+  select(Genera, starts_with('Moisture') & ends_with('QMP')) %>%
+  arrange(`Moisture Spearman.P QMP`) %>%
+  slice_head(n=12)
+qmp_spearman %>%
+  filter(test == 'qmp_moisture_spearman') %>%
+  arrange(p.value) %>% 
+  select(.otu, estimate, p.value, p.bh) %>%
+  slice_head(n=12)
+```
+
+Here the estimate and p-values seem identical up to rounding, but the FDR-corrected p values differ somewhat.
+
+Note, I filtered the OTUs that lack genus classifications prior to doing the testing, which I suspect is what was done in the original study.
+If these OTUs are kept, the FDR-corrected p-values are larger, due to the larger number of tests.
+But the study results have smaller p-values and only list the OTUs with genus names in the table.
+
+## Visually inspect variation for top hits
+
+### Calprotectin
+
+```{r}
+top_hits <- qmp_spearman %>% 
+  filter(test == 'qmp_calprotectin_spearman') %>%
+  slice_min(p.bh, n = 9)
+x <- ps %>%
+  filter_sample_data(!is.na(faecal_calprotectin)) %>%
+  prune_taxa(top_hits$.otu, .) %>%
   as_tibble %>%
-  with_groups(.sample, mutate, 
-    proportion = close_elts(.abundance),
-    present = (.abundance > 0) * 1
+  mutate(across(.otu, factor, levels = top_hits$.otu))
+```
+
+To visualize the spearman correlation, plot and perform a linear regression on the rank-transformed values,
+
+```{r, fig.dim = c(8,8)}
+x %>%
+  with_groups(.otu, mutate,
+    across(c(faecal_calprotectin, .abundance), rank),
+  ) %>%
+  ggplot(aes(faecal_calprotectin, .abundance)) +
+  facet_wrap(~.otu) +
+  coord_fixed() +
+  geom_point() +
+  stat_smooth(method = 'lm') +
+  labs(y = 'rank(abundance)', x = 'rank(faecal calprotectin)')
+```
+
+Many of the top hits have relatively low pravalence, and it appears that the correlation is being driven by differential prevalence in the two groups of samples where calprotectin has its min value, or is above its min value.
+For these taxa, we might expect bias to have less impact on the associations.
+
+### Moisture
+
+```{r}
+top_hits <- qmp_spearman %>% 
+  filter(test == 'qmp_moisture_spearman') %>%
+  slice_min(p.bh, n = 9)
+x <- ps %>%
+  filter_sample_data(!is.na(moisture_content)) %>%
+  prune_taxa(top_hits$.otu, .) %>%
+  as_tibble %>%
+  mutate(across(.otu, factor, levels = top_hits$.otu))
+```
+
+```{r, fig.dim = c(8,8)}
+x %>%
+  with_groups(.otu, mutate,
+    across(c(moisture_content, .abundance), rank),
+  ) %>%
+  ggplot(aes(moisture_content, .abundance)) +
+  facet_wrap(~.otu) +
+  coord_fixed() +
+  geom_point() +
+  stat_smooth(method = 'lm') +
+  labs(y = 'rank(abundance)', x = 'rank(moisture content)')
+```
+
+Here we instead see that top taxa as being more prevalent.
+
+## Check if hits are driven by differential prevalence
+
+One way we can do this is to also do Spearman tests after converting abundances to presence/absence (0 if absent, 1 if present).
+If the original result is driven by prevalence, then the estimates should be similar to when the QMP abundance was used.
+
+```{r}
+qmp_presence <- ps %>%
+  prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
+  as_tibble %>%
+  mutate(
+    presence = (.abundance > 0) * 1
   ) %>%
@@
   mutate(
-    spearman = map(data, 
+    qmp_calprotectin = map(data, 
       ~cor.test(data = .x,
-        ~ .abundance + faecal_calprotectin_mg_g, 
-        method = 'spearman')),
-    spearman_pa = map(data, 
+        ~ .abundance + faecal_calprotectin, 
+        method = 'spearman', exact = FALSE, continuity = TRUE)),
+    presence_calprotectin = map(data, 
       ~cor.test(data = .x,
-        ~ present + faecal_calprotectin_mg_g, 
-        method = 'spearman')),
+        ~ presence + faecal_calprotectin, 
+        method = 'spearman', exact = FALSE, continuity = TRUE)),
+    # cond.qmp_calprotectin = map(data,
+    #   ~cor.test(data = .x %>% filter(.abundance > 0),
+    #     ~ .abundance + faecal_calprotectin, 
+    #     method = 'spearman', exact = FALSE, continuity = TRUE)),
+    qmp_moisture = map(data, 
+      ~cor.test(data = .x,
+        ~ .abundance + moisture_content, 
+        method = 'spearman', exact = FALSE, continuity = TRUE)),
+    presence_moisture = map(data, 
+      ~cor.test(data = .x,
+        ~ presence + moisture_content, 
+        method = 'spearman', exact = FALSE, continuity = TRUE)),
+    # cond.qmp_moisture = map(data,
+    #   ~cor.test(data = .x %>% filter(.abundance > 0),
+    #     ~ presence + moisture_content, 
+    #     method = 'spearman', exact = FALSE, continuity = TRUE)),
   ) %>%
   select(-data) %>%
-  pivot_longer(contains('spearman'), names_to = 'test', values_to = 'fit') %>%
+  pivot_longer(contains(c('qmp', 'presence')), 
+    names_to = 'test', values_to = 'fit') %>%
   mutate(across(fit, map, broom::tidy)) %>%
@@
 ```{r}
-y.w <- y %>%
-  select(.otu, test, estimate, p.bh) %>%
-  pivot_wider(names_from = c(test), values_from = c(estimate, p.bh), 
-    names_glue = '{test}_{.value}', names_sort = TRUE) %>%
-  select(.otu, sort(tidyselect::peek_vars())) %>%
-  mutate(across(where(is.numeric), round, 4)) %>%
-  arrange(spearman_p.bh)
+qmp_presence_wide <- qmp_presence %>%
+  select(.otu, test, estimate, p.value, p.bh) %>%
+  separate(test, into = c('response', 'covariate'), sep = '_') %>%
+  pivot_wider(names_from = response, values_from = c(estimate, p.value, p.bh), 
+    names_glue = '{response}_{.value}', names_sort = TRUE) %>%
+  select(.otu, covariate, starts_with('qmp'), starts_with('presence'))
+#> select(.otu, sort(tidyselect::peek_vars()))
 ```
@@
 ```{r}
-y.w %>%
-  ggplot(aes(spearman_estimate, spearman_pa_estimate)) +
+p1 <- qmp_presence_wide %>%
+  ggplot(aes(presence_estimate, qmp_estimate)) +
+  facet_wrap(~covariate) +
+  coord_fixed() +
   geom_point()
+p1
 ```
 
+```{r}
+p2 <- qmp_presence_wide %>%
+  ggplot(aes(presence_p.value, qmp_p.value)) +
+  facet_wrap(~covariate) +
+  coord_fixed() +
+  scale_x_log10() +
+  scale_y_log10() +
+  geom_point()
+p2
+```
+
+Overall, can see that most of the QMP correlations are highly similar to the presence correlations.
+
+```{r, echo = FALSE}
+print_table <- function(x, covariate) {
+  x %>%
+    filter(covariate == !!covariate) %>%
+    select(-covariate) %>%
+    rename(genus = .otu) %>%
+    slice_min(qmp_p.value, n = 10) %>%
+    mutate(across(where(is.numeric), signif, 2)) %>%
+    knitr::kable()
+}
+```
+
+**Faecal calprotectin**
+
+```{r, echo = FALSE}
+print_table(qmp_presence_wide, 'calprotectin')
+```
+
+**Moisture content**
+
+```{r, echo = FALSE}
+print_table(qmp_presence_wide, 'moisture')
+```
+
+Consider the calprotectin results.
 For some taxa (like Veillonella, EscherichiaShigella) the two estimates are very similar;for others like Streptococcus they are quite different.
-I suspect that simulated bias will have a larger impact on Streptococcus.
 
@@
 
-```{r, eval = F}
-# Perturb by bias vector - treat ps as actual
-ps.per <- perturb(ps, bias, norm = 'keep')
-# Calibrate by bias vector - treat ps as observed
-ps.cal <- calibrate(ps, bias, norm = 'keep')
-```
+Hypothesis: Correlations that are driven by differential prevalence will be less sensitive to bias than correlations that are not.
 
-Maintain the sample sums, but perturb the relative abundances.
+For the purposes of this analysis, zero observations will be treated as true zeros, which I expect to reduce the apparent impact of bias perturbation or calibration on the DA results relative to if we assumed zeros as indicating small positive abundances.
+
+## Analysis
+
+In order to perform a bias sensitivity analysis on the QMP profiles, we will calibrate (inverse-perturb) the relative abundances by the sampled bias vector, while keeping the total sum for the same constant, so that the total abundance of the same remains given by the fecal cell count.
 
 ```{r}
-n_taxa <- ntaxa(ps)
-# Standard deviation for the distribution of log efficiencies
-sigma.log_eff <- 3
-# helper function to draw a single random bias vector
-rbias <- function() exp(rnorm(n_taxa, 0, sigma.log_eff))
-# helper function to calibrate the phyloseq object by the bias vector
-my_cal <- function(bias, ps) {
-  ps %>%
-    calibrate(bias, norm = 'keep')
+# Returns a matrix with taxa as columns
+sample_bias <- function(n, gsd, vcv, taxa_as = 'cols') {
+  stopifnot(isSymmetric(vcv))
+  stopifnot(n > 1)
+  n_taxa <- nrow(vcv)
+  gsd_log <- log(gsd)
+  bias <- MASS::mvrnorm(n, mu = rep(0, n_taxa), Sigma = vcv) %>%
+    t %>%
+    scale %>%
+    {. * gsd_log} %>%
+    exp
+  if (taxa_as == 'cols')
+    bias <- t(bias)
+  bias
 }
 
-set.seed(42)
-# Number of replicates
-R <- 3e2
-# data frame with R random bias vectors
-reps <- seq(R) %>%
-  set_names %>%
-  map(~rbias()) %>%
-  enframe("id", "bias") %>%
+my_stat <- function(ps, otus) {
+  ps %>%
+    prune_taxa(otus, .) %>%
+    as_tibble %>%
+    with_groups(.otu, nest) %>%
     mutate(
-    actual = map(bias, my_cal, ps = ps)
-  )
+      fit = map(data, ~cor.test(data = .x,
+          ~ .abundance + faecal_calprotectin, 
+          method = 'spearman', exact = FALSE, continuity = TRUE)),
+      fit = map(fit, broom::tidy)
+    ) %>%
+    select(-data) %>%
+    unnest(fit)
+}
+#> my_stat(ps, 'Veillonella')
 ```
@@
 ```{r}
-my_stat <- function(ps, otu) {
-  data <- ps %>%
-    prune_taxa(otu, .) %>%
-    as_tibble
-  spearman_test <- cor.test(data = data,
-    ~ .abundance + faecal_calprotectin_mg_g, 
-    method = 'spearman', exact = FALSE
-  )
-  spearman_test %>% broom::tidy()
-}
-
-#> focal_otu <- 'Veillonella'
-original <- my_stat(ps, focal_otu)
-reps_veill <- reps %>%
-  mutate(
-    .otu = 'Veillonella',
-    fit = map2(actual, .otu, my_stat)
-  ) %>%
-  unnest(fit)
-reps_strep <- reps %>%
-  mutate(
-    .otu = 'Streptococcus',
-    fit = map2(actual, .otu, my_stat)
-  ) %>%
-  unnest(fit)
+set.seed(42)
+R <- 300
+gsd <- 3
+vcv <- diag(ntaxa(ps))
+reps <- sample_bias(n = R, gsd = gsd, vcv = vcv) %>%
+  split(., row(.)) %>%
+  enframe(".idx", "bias")
 ```
@@
 ```{r}
-plot_bias_sens <- function(reps) {
+taxa_to_test <- c(
+  #  low prevalence
+  'Veillonella', 'EscherichiaShigella', 'Fusobacterium',
+  # high prevalence
+  'Streptococcus', 'Anaerostipes', 'Fusicatenibacter'
+  )
+reps_fit <- xfun::cache_rds({
+  reps %>%
+    mutate(
+      fits = map(bias, 
+        ~my_stat(
+          ps = calibrate(ps, .x, norm = 'keep'),
+          otus = taxa_to_test
+        ))
+    ) %>%
+    unnest(fits)
+}, dir = '_cache/', file = 'reps_fit', 
+hash = list(reps, taxa_to_test))
+```
+
+```{r}
+plot_bias_sens <- function(reps, original, .otu) {
   est_dist <- ggplot(reps, aes(estimate)) +
@@
     labs(title = "Distribution of p-values", x = "p-value", y = "Count")
-  est_dist / pval_dist
+  est_dist / pval_dist +
+    plot_annotation(title = .otu)
 }
-# ci_reps <- ggplot(bias_reps, aes(x = estimate, y = as.integer(id))) +
-#   geom_point(size = 0.8) +
-#   geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
-#   geom_point(data = original, aes(y = R*1.05), 
-#     size = 0.8, color = "darkred") +
-#   geom_errorbarh(data = original, 
-#     aes(xmin = conf.low, xmax = conf.high, y = R*1.05),
-#     color = "darkred") +
-#   geom_vline(xintercept = 0, color = "grey") +
-#   theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
-#   labs(title = "Replicate 95% CIs", x = "Estimate", 
-#     y = "Replicate")
-# plot_grid(
-#     plot_grid(est_dist, pval_dist, ncol = 1, labels = c("A", "B")),
-#     ci_reps,
-#     labels = c("", "C")
-#     )
+
+original <- qmp_presence %>%
+  filter(.otu %in% taxa_to_test, test == 'qmp_calprotectin') %>%
+  with_groups(.otu, nest) %>%
+  rename(original = data)
+
+reps_plots <- reps_fit %>%
+  with_groups(.otu, nest) %>%
+  rename(reps = data) %>%
+  left_join(original, by = ".otu") %>%
+  mutate(.,
+    plot = pmap(., plot_bias_sens)
+  ) %>%
+  select(.otu, plot) %>%
+  deframe
 ```
 
-```{r}
-plot_bias_sens(reps_veill)
+## Top hits with low prevalence
+
+```{r, echo = F}
+fig.dim_sens <- c(6,6)
 ```
 
-```{r}
-plot_bias_sens(reps_strep)
+```{r, echo = FALSE, fig.dim = fig.dim_sens}
+reps_plots$Veillonella
 ```
 
-Example where the bias sensitivity analysis shows little effect of varying the hypothesized bias.
-
-We can probably understand this finding because the original result is driven by the presence/absence pattern, and we have treated zeros as absolute, so that presence/absence is maintained.
-If we relaxed this assumption, then I suspect that we would see much more variation in the estimates.
-Much of this variation, however, would come from relaxing our assumption about the zeros, rather than relaxing our assumption that there is no bias.
-
-Challenge with this data is that we don't have the read counts, so don't know how to replace zeros with resonable values.
-
-## saved plotting stuff
-
-Mean efficiency distribution:
-
-```{r}
-x <- reps %>%
-  slice(1) %>%
-  pull(actual) %>%
-  .[[1]] %>%
-  sample_data %>%
-  as_tibble
-
-x %>%
-  ggplot(aes(x = mean_efficiency, fill = faecal_calprotectin_mg_g > 50)) +
-  scale_x_log10() +
-  geom_histogram()
+```{r, echo = FALSE, fig.dim = fig.dim_sens}
+reps_plots$EscherichiaShigella
 ```
 
-```{r}
-x %>%
-  ggplot(aes(x = faecal_calprotectin_mg_g, y = mean_efficiency, color = diagnosis)) +
-  scale_y_log10() +
-  scale_x_log10() +
-  geom_point() +
-  stat_smooth(aes(group = 1), method = 'lm')
+```{r, echo = FALSE, fig.dim = fig.dim_sens}
+reps_plots$Fusobacterium
 ```
 
-Now, redo the analysis with and without bias
+## Top hits with high prevalence
 
-what assumptions to make about zeros?
+```{r, echo = FALSE, fig.dim = fig.dim_sens}
+reps_plots$Streptococcus
+```
+
+```{r, echo = FALSE, fig.dim = fig.dim_sens}
+reps_plots$Anaerostipes
+```
+
+```{r, echo = FALSE, fig.dim = fig.dim_sens}
+reps_plots$Fusicatenibacter
+```
+
+## Summary
+
+It isn't clear from these plots that there is a difference in bias sensitivity between the two sets.

### 2022-02-09 abundance-measurement.Rmd
Make edits from KL
Edits from Karen Lloyd on the manuscript draft rendered on 2022-01-16
from 414f6e78cc9b29c8236c8c3e996b68cd4d733b86.

@@
 This model is the simplest that respects the multiplicative nature of taxonomic bias and the *compositional* nature of MGS measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
-We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes to taxonomically assign reads to a set of microbial species $S$.
+We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and taxonomically assigns reads to a set of microbial species $S$.
 Various forms of taxonomic assignment are possible; for simplicity, we suppose that reads are assigned to the species level, with reads that cannot be uniquely assigned being discarded.

### 2022-02-09 case-studies.Rmd
Make edits from KL
Edits from Karen Lloyd on the manuscript draft rendered on 2022-01-16
from 414f6e78cc9b29c8236c8c3e996b68cd4d733b86.

@@
 @lloyd2020evid performed a systematic assessment of growth rates of bacterial and archaeal taxa over a depth of 10 cm (corresponding to ~40 years of burial time) in sediment of the White Oak River estuary.
-To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy (CARD-FISH).
+To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy.
 The authors refer to these absolute densities as FRAxC measurements, for 'fraction of 16S reads times total cell counts'.
 The FRAxC measurements were used to infer growth rate from the slope of a simple linear regression of log cell density against burial time over the first 3 cm below the bioirrigation layer (corresponding to $\sim 8$ years of burial).
-To validate the inference of positive growers, the authors compared the growth rates from FRAxC-based inference from two sediment cores, qPCR measurements in these cores for a few reference taxa, and FRAxC-based inference in two replicate laboratory incubation experiments.
+To validate the inference of positive growers, the authors compared the growth rates from FRAxC-based inference from two sediment cores, qPCR measurements in these cores for a few reference taxa, and FRAxC-based inference in two replicate two-year laboratory incubation experiments.
 
 Taxonomic bias could lead to systematic error in FRAxC-derived growth rates if sample mean efficiency tends to systematically vary with burial time (or equivalently, depth).
-One possibility is that microbes with tougher cell walls tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
+One possibility is that microbes with tougher cell walls may persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
 In this case, we would expect the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that decay sufficiently slowly would be mistakenly inferred to have positive growth.
@@
 We described a hypothesized scenario in the marine-sediment case study in which increased burial time (the covariate) selects for lysis-resistant (and so lower efficiency) species.
-As another example, treatment with a certain antibiotic might selectively kill easier-to-lyze Gram negative species (or harder-to-lyse Gram positives) and thereby decreasing (or increasing) the mean efficiency in samples collected from a host post-treatment.
+As another example, treatment with a certain antibiotic might selectively kill easier-to-lyse Gram negative species (or harder-to-lyse Gram positives) and thereby decrease (or increase) the mean efficiency in samples collected from a host post-treatment.
 Such situations can arise whenever there is a microbial trait that is associated with both measurement efficiency and the biological processes of interest.
@@
 
-So long as the mean efficiency does not vary systematically with the regression covariates, the error in fold changes serve primarily to reduce the precision and power of regression analyses without systematically distorting estimates and so does not necessarily lead to invalid inferences.
+So long as the mean efficiency does not vary systematically with the regression covariates, the measurement error from bias primarily reduces the precision and power of regression analyses without systematically distorting estimates and so does not necessarily lead to invalid inferences.
 But the reduction in precision could be substantial, particularly given the small sample sizes common in many microbiome studies, and thereby substantially limit the ability to draw meaningful conclusions from a study^[I suspect this non-systematic variation in the mean efficiency is the more typical situation, but we currently lack an example of it in the case studies.

### 2022-02-09 introduction.Rmd
Make edits from KL
Edits from Karen Lloyd on the manuscript draft rendered on 2022-01-16
from 414f6e78cc9b29c8236c8c3e996b68cd4d733b86.

@@
 These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase) (@mclaren2019cons).
-Yet @mclaren2019cons also found that certain types of DA analysis---those based on fold changes in the ratios among species---where robust to bias.
-The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as typically done in microbiome association testing, have yet to be investigated.
+Yet @mclaren2019cons also found that certain types of DA analysis---those based on fold changes in the ratios among species---were robust to bias.
+The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.
 

### 2022-02-09 solutions.Rmd
Make edits from KL
Edits from Karen Lloyd on the manuscript draft rendered on 2022-01-16
from 414f6e78cc9b29c8236c8c3e996b68cd4d733b86.

@@
 Even an ideal qPCR protocol that perfectly amplifies all species remains a biased measurement of cell density, since the total 16S copies in the extracted DNA contributed by each species will be proportional to its (species-specific) lysis efficiency and 16S copy-number.
+Empirical evidence supports this, as qPCR measurements of total 16S copies are rarely equivalent to total cell counts, even when multiple copy numbers are taken into account ()@lloyd2013meta).
 16S qPCR is commonly paired with 16S amplicon sequencing, with which these sources of bias are shared, perhaps along with variation in primer binding and amplification efficiency.
-Methods like using flow cytometry that directly measures cell density lacks these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
+Methods like flow cytometry that directly measure cell density lack these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
 By extending the analysis of Section \@ref(differential-abundance) to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(total-density-bias)).

### 2022-02-10 solutions.Rmd
Revise solutions section
Changes to the subsections on ratio-based analyses and reference-species
calibration.

@@
 A variety of such methods have been developed for microbiome DA analysis that build on earlier work from the field of Compositional Data Analysis (CoDA).
-These methods analyze log-fold changes in the ratios among a pair of species or, more generally, some product of exponential power of multiple species, which likewise are bias-invariant.
-Besides a greater robustness to taxonomic bias, such ratio-based analyses have other advantages.
-First, analysis of ratios avoids other standard criticisms leveled at proportion-based analysis—namely that the change in proportion of a species is affected by change in the density of all other species and the set of species used in the denominator of the proportion calculation (@gloor2017micr).
-Second, ratios may be more informative than absolute density in some cases, such as for swabs or other sample types for which the absolute density in the sample may be an unreliable indicator of the absolute density of the sampled ecosystem.
+These methods analyze (log) fold changes in the ratio between species or, more generally, between the products of multiple species.
+Taking the product or geometric mean of multiple species provides a approach to aggregating species into a higher-level taxonomic units such that the ratios between units maintain proportional errors, in contrast to the standard approach of additively aggregating species.
+Besides a greater robustness to taxonomic bias, analyses of ratios have other advantages over analyses of proportions and absolute abundances.
+Analyses of ratios avoid the standard criticism leveled at proportion-based analysis—namely that the changes in the proportion of one species depend in a non-transparent way on the changes in abundance of all other species (@gloor2017micr), including unidentified contaminants.
+In many experiments, the total abundance in a sample may be a poor indicator of the total abundance of the sampled ecosystem; unless this relationship can be characterized and accounted for, changes in ratios may be more informative than changes in absolute abundance for the underlying biology of interest.
 
@@
 A related problem is that, due to the discrete nature of biological organisms and the reads that are generated during sequencing, species in the denominator are commonly observed to have 0 reads; the assumptions that made by different methods about the meaning of these 0 counts can have a major impact on DA results.
-Finally, bias invariance at the level of ratios among species does not extend to ratios of additive aggregates of species, such as between bacterial phyla, unless further conditions are met (@mclaren2019cons).
+Finally, bias invariance at the level of ratios among species does not extend to ratios between additive aggregates of species, such as between bacterial phyla, unless further conditions are met (@mclaren2019cons).
+Multiplicative aggregates of species provides a bias-invariant alternative that has proven useful in biomarker discovery but are harder to interpret.
 
@@
 Moreover, the bias that is measured from controls may differ from that in the primary samples due to differences in cell state or sample preparation.
-These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-density measurement that was described in Section \@ref(differential-abundance).
+These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-abundance measurement that was described in Section \@ref(differential-abundance).
 This *reference-species calibration* allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from taxonomically biased, relative metagenomics measurements.
-
-Recall that densities estimated by the reference-species approach have a constant fold error, provided that the fold error in the reference species is also constant (Equation \@ref(eq:density-ratio-error)).
-Three categories of reference species are reviewed for this purpose in Appendix \@ref(review-absolute-methods), including why they might be expected to produce constant fold errors: housekeeping species assumed to have constant abundance, spike-in species added at known or constant abundance, and species whose abundance has been measured with a targeted method like species-specific qPCR.
 
@@
 Doing so improved the resulting measurements of fold changes in density for all species (Figure \@ref(fig:calibration-example)).
+
+Two categories of reference species have so far been used: Housekeeping species assumed to have constant abundance and spike-in species added at known (and typically constant) abundance.
+The implications of bias for these methods hadn't been previously investigated.
+Our results show that, provided that taxonomic bias among all species (including the housekeeping or spike-in species) is consistent across samples, then the fold error in species abundances is constant and does not impact DA analysis of log- or rank-transformed species abundances.
+These methods have downsides: Housekeeping species are not always available/known, and spike-ins have the downside of needing to determine species not present in the samples and an appropriate concentration in which to add them so that they form a moderate fraction of reads across samples with potentially widely varying produce of mean efficiency and total abundance.
+
+Section \@ref(abundance-measurement) presents an additional type of reference species: naturally-occurring species whose abundance has been directly measured.
+For most natural ecosystems, there is unlikely to be a single species or genus that is found in all samples.
+Sample coverage can be increased by measuring multiple species with complementary presence-absence patterns (Appendix [TODO]).
+Coverage can also be increased by using higher-order taxa as references---for example, performing qPCR measurements targeted to a particular family rather than a particular species.
+Higher-order taxa necessarily have higher prevalence than species, though may be more prone to errors from a varying efficiency across samples.
+Targeted measurements of taxa that only span a subset of samples can still provide useful validation for methods.
+
+<!-- TODO: Add paragraph: Pre- versus post- extraction; linearity of DNA extraction. -->
+<!-- Applies to spike-ins and targeted measurements. -->
+
+Housekeeping taxa and spike-ins have also been used to measure changes in total-community abundance.
+Our results show that the bias invariance does not apply for this purpose---the measured fold change in total abundance have an error equal to the inverse change in the sample mean efficiency.
+A similar problem arises when attempting to infer changes in abundance of any higher-order taxon, unless measurement efficiency is conserved among its species.
+<!-- If the purpose is to assess changes in the total community abundance rather than individual species, we recommend researchers either directly measure total abundance or, if using reference species, then explicitly acknowledge that their results are contingent on constant or at least non-systematically varying mean efficiency.
+-->
+<!-- In the second case, confidence in the results can be bolstered by a bias-sensitivity analysis. -->
 
@@
 To the extent that taxonomic bias truly is consistent within studies, meta-analyses that directly account for it can have greater statistical power to extract the truly consistent and inconsistent microbial patterns across studies.
+
+## [TODO] Summary

### 2022-02-10 abundance-measurement.Rmd
Revise abundance-measurement section
Revise the presentation of the model, the relative-abundance subsection,
and the intro and subsection titles of the absolute-abundance
subsection.

@@
 
-We begin by considering how taxonomic bias affects the relative- and absolute-abundance measurements which serve as the input to microbiome DA analyses.
+## Model of MGS measurement
 
@@
 This model is the simplest that respects the multiplicative nature of taxonomic bias and the *compositional* nature of MGS measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
+The model as described by @mclaren2019cons considers only relative abundances; here we present a straightforward extension that allows us to also consider absolute abundances.
+
+<!-- TODO: Is S the set of species that could be in the sample, or the set that can be mapped to? Also, what about the host H? -->
+<!-- This issue ties in with our general notation issue, and the problem of unassigned reads. -->
+
 We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and taxonomically assigns reads to a set of microbial species $S$.
-Various forms of taxonomic assignment are possible; for simplicity, we suppose that reads are assigned to the species level, with reads that cannot be uniquely assigned being discarded.
-We ignore the possibility that reads are incorrectly assigned to the wrong species or sample.
-Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
+We make several simplifying assumptions to facilitate our analysis and presentation.
+First, we consider only species-level assignment, and suppose that reads that cannot be uniquely assigned to a single species in $S$ are discarded.
+Second, we ignore the possibility that reads are misassigned to either the wrong species or sample.
+Third, we suppose that taxonomic bias acts consistently across samples at the level of species---that is, a given species is always measured more efficiently than another to the same degree, regardless of variation in changes in sub-species-level composition or the sample matrix.
+Finally, unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
+These assumptions, though unrealistic descriptions of most MGS experiments, serve our purpose of clearly demonstrating when and why consistent taxonomic bias creates errors in DA analysis.
+The results we derive from them form the basis for understanding the impact of bias in more complex scenarios.
 
-In our model, the assigned read count of a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
+Our model describes the mathematical relationship between the read counts obtained by MGS and the true absolute abundances of each species in a sample.
+For concreteness, we take _absolute abundance_, or simply _abundance_, to refer to cell concentration, or number of cells per unit volume, in a sample.
+Thus we take 'absolute' simply to mean not relative to other species; however, our results also apply were 'absolute' taken to mean all cells in a sample or in a sampled ecosystem.
+Similarly, our results apply to abundance units other than cells, such as biomass and genome copy number.
+Our model stipulates that the assigned read count of a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
 \begin{align}
@@
 
+We first review results from @mclaren2019cons on how bias affects measurements of relative abundance, which are essential to understanding both relative and absolute DA analysis.
 We distinguish between two types of species-level *relative abundances* within a sample.
-The *proportion* of species $i$ in sample $a$ equals to its abundance relative to the total abundance of all species in $S$,
+The *proportion* of species $i$ in sample $a$ equals its abundance relative to the total abundance of all species in $S$,
 \begin{align}
@@
 \end{align}
-Proportions and ratios each form the basis for popular DA methods;
-ratio-based methods are commonly referred to as Compositional Data Analysis (CoDA) methods.
+Proportions and ratios each form the basis for popular relative DA methods.
 
-Taxonomic bias has a different effect on species proportions versus species ratios.
 The proportion of a species is typically measured by its proportion of assigned reads,
@@
 
-The fact that the multiplicative error in proportions varies while that in ratios is constant forms the basis for understanding the effect of bias on different methods for measuring absolute abundances from MGS and the extent to which bias cancels in different DA analyses.
+Thus taxonomic bias creates variable fold errors in species proportions, but constant fold errors in species ratios.
+Consequently, a perfect cancellation of errors can only occur in ratio-based DA analysis.
 
@@
 
-Researchers often would like to know how the _absolute abundance_ of a species changes.
-In this context, _absolute_ means not relative to other species; typically, the abundance of interest is relative to non-microbial entities such the volume, mass, or amount of host DNA in the sample.
-A wide range of experimental methods have been used to convert the relative abundances from MGS measurements into measurements of absolute abundances.
-Yet there has so far been little consideration as to how the resulting measurements are affected by taxonomic bias in the MGS measurements or in the supplemental measurements, such as flow cytometry and qPCR, that they often involve.
-Here and in Appendix \@ref(review-absolute-methods) we describe how various methods for measuring absolute abundance are predicted to be affected by taxonomic bias, with an eye towards determining whether the measurement error is variable or constant.
+Various methods can be used to convert the relative abundances from MGS into absolute abundances.
+These methods are subject to taxonomic bias in the MGS measurements and, in some cases, in supplemental non-MGS measurements used for the conversion.
+Here we describe how various methods for measuring absolute abundance are predicted to be affected by taxonomic bias, with an eye towards determining whether the measurement error is variable or constant.
+We consider two general classes of methods: Those in which absolute-abundance information is derived from measurement of the aggregate abundance of the total community, and those in which it is derived from targeted measurement (or prior knowledge) of one or more particular species.
 
-Our discussion supposes the 'abundance' of interest is cell number, though it applies to other quantities, such as biomass and genome copy number, that may be more relevant in different biological contexts.
+### Total-community normalization
 
-We consider two general classes of methods: Those in which absolute-abundance information is derived from measurement of the aggregate abundance of the total community, and those in which it drives from targeted measurement (or prior knowledge) of one or more particular species.
-
-**Normalization by total abundance:**
 Measurements of species absolute abundances are often obtained by making a (non-MGS) measurement of total community abundance, equating it with the aggregate abundance of the species $S$ measured by MGS, and multiplying this total by the proportions from MGS,
@@
 
-**Normalization by one or more reference species:**
+### Reference-species normalization
+
 Suppose we had a measurement of the (absolute) abundance of a _reference species_ $r$.

### 2022-02-10 abundance-measurement.Rmd
Revise abundance-measurement section
Add paragraph on higher-order taxa to the relative-abundance subsection

@@
 
+**Higher-order taxa:** How does bias affect the proportions and ratios of higher-order taxa, such as a genera, phyla, or even all organisms?
+More generally, we can identify an higher-order taxon $I$ with a set of species, $\{i \in I\}$.
+The abundance and reads of a taxon $I$ in a sample $a$ equal the sums over its constituent species, $\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)$ and $\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)$.
+The efficiency of taxon $I$ equals the weighted average of its constituent species,
+\begin{align}
+  (\#eq:efficiency-general)
+  \text{efficiency}_I(a) 
+    \equiv \frac{\sum_{i\in I}\text{abun}_i(a)\cdot \text{efficiency}_i}{\text{abun}_I(a)}.
+\end{align}
+Thus the efficiency of a higher-level taxon may vary across samples (@mclaren2019cons).
+An exception is if the constituent species have the same efficiency or always appear in the same abundances relative to each other, in which case the efficiency of the higher-order taxon would stay constant.
+As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
+The efficiency of the phylum would then be $\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5$ in Sample 1 and $\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1$ in Sample 2.
+Equations \@ref(eq:prop-error) and \@ref(eq:ratio-error) continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent taxon efficiency defined by Equation \@ref(eq:efficiency-general) is used.
+In this way, we can see that both proportions and ratios among higher-order taxa may have inconsistent fold errors.
+
 ## Absolute abundance

### 2022-02-12 appendix-notes.Rmd
Add theory notes
- Add intuition for 'effort' term
- Add general expression for absolute-abundance measurements

# Theory notes

## Effective experimental effort: Intuition

The effective experimental effort per cell for sample $a$ is
\begin{align}
  (\#eq:effort)
  \text{effort}_S(a)
    &= \frac{\text{reads}_S(a)}{\text{abun}_S(a) \cdot \text{efficiency}_S(a)} \\
    &= \frac{\sum_{i \in S}\text{reads}_i(a)}{\sum_{i \in S} \text{abun}_i(a) \cdot \text{efficiency}_i(a)};
\end{align}
the second form follows from the definition of the mean efficiency.
The effort increases if more sequencing performed on the sample, leading to greater number of reads for all species; and decreases if a larger sample input volume or concentration is used, leading to fewer reads per cell.
Less intuitively, it also decreases with the mean efficiency of the sample.
To help interpret the effort light of this last behavior, we consider how the effort varies across a few illustrative scenarios.

**Scenario 1:**
Consider Sample 1 in Figure \@ref(fig:error-proportions).
Imagine that sequencing reads can be perfectly assigned to the three species; the taxonomic bias is entirely due to steps prior to read assignment, such as DNA extraction and PCR.
Supposing that the sample was sequenced to a read depth of 100, the three species have read counts of $(4, 72, 24)$.
In addition, suppose the total cell concentration in the sample equals 1 in some particular units.
Since the mean efficiency is $(1 + 18 + 6) / 3 = 8.\bar 3$, the effort is $100 / (1 \cdot 8.\bar 3) = 12$.

**Scenario 2:**
Now consider a situation identical to Scenario 1, except that the bioinformatics pipeline is unable to assign reads from Species 3.
In this case, the _assigned_ read counts will be $(4, 72, 0)$, for a total of $76$, though the total number of sequenced reads remains 100.
Meanwhile, the efficiency of species 3 drops to 0 and the mean efficiency of the sample drops to $(1 + 18 + 0) / 3 = 6.\bar 3$.
The effort is therefore $76 / (1 \cdot 6.\bar 3) = 12$, the same as in Scenario 1.
<!-- The decline in total reads is exactly offset by the decline in the mean efficiency. -->
<!-- This example illustrates a general pattern: The effort is invariant to removing from or adding to species' assigned reads. TODO: Make precise, and prove -->

**Scenario 3:**
Now consider a situation identical to Scenario 1, except that Species 3 fails to amplify during the PCR step in a marker-gene sequencing experiment.
But thanks to careful experimental normalization during library preparation, an equal amount of DNA is prepared in the library as in Scenario 1, leading to 100 sequenced reads from the sample.
The measured relative abundances are identical to in Scenario 2, but the read counts are larger, $(\approx 5.3, \approx 94.7)$.
(As in Section \@ref(abundance-measurement), we ignore the fact that read counts are discrete.)
The mean efficiency is also the same ($6.\bar 3$) as in Scenario 2.
The effort is therefore $100 / (1 \cdot 6.\bar 3) \approx 15.8$, larger than in Scenario 1 and Scenario 2.

**Scenario 4:**
Now consider a situation identical to Scenario 1, but supposing Species 3 to be a reference species, such as the host or a spike-in.
Does the effort depend on whether we include the reference species in the species set $S$?
As we saw in Scenario 1, the effort associated with all three species is 12.
For the two non-reference species alone, the total abundance is $2/3$, the total read count is $76$, and the mean efficiency is $(1 + 18) / 2 = 9.5$.
The effort is therefore $76 /(\tfrac{2}{3} \cdot 9.5) = 12$, the same as in Scenario 1 and Scenario 2.
<!-- Compared to Scenario 2, the change in abundance and mean efficiency offset -->

TODO:

- Make sure this result is general, and not just happening when the three species are all equally abundant.
- consider moving secenario 4 to 2 or 3, leaving the scenario where effort changes to last.
- note that for proving the general results, it may be best to go back to main model equation, where we can see that if the read count, abundance, and efficiency for at least one species is unchanged, than the effort is unchanged.
  - in scenario 3, the effort must go up since taxa 1 and 2 are unchanged but get more reads.
  - if we add or remove species, without changing the read count of other species, then the effort must stay the same.
  - this fact that effort is independent of the taxonomic scope S seems useful and convenient and intuitive.

**Summary:**
The behavior of the effective experimental effort across these scenarios makes intuitive sense if we think of it as specifically describing the 'wet lab' components of the MGS measurement, up to and including sequencing but excluding bioinformatic analysis of the sequencing data.
The taxonomic scope $S$ is an arbitrary decision made implicitly or explicitly by the researcher, which does not change the wet portion of the experiment; thus   $\text{effort}(a)$ is independent of $S$.
In addition, sequencing reads are routinely excluded during bioinformatic analysis, and such filtering does not change $\text{effort}(a)$, instead reducing the read counts by reducing the species' efficiencies.
On the other hand, if species drop out prior to sequencing due to low extraction efficiency and/or low PCR efficiency, then achieving the same total read count requires that a greater sample volume or concentration be passed on to subsequent steps.
These adjustments are reflected by a commensurate increase in $\text{effort}(a)$.

## Absolute-abundance measurement

Q: How to motivate this 'general' form? 
It comes from ignoring bias.

Make clear that R doesn't need to be in S. Perhaps want different sets - a superset of all species, and an set of species that each measurement method considers?

---

Consider this general form for measuring of the absolute abundance of a taxon $I$ from a reference species $R$,
\begin{align}
  \widehat{\text{abun}}_{I}(a) 
  &= \text{reads}_{I}(a) \cdot \frac{\widehat{\text{abun}}_{R}(a)}{\text{reads}_{R}(a)}.
\end{align}
When $R = S$, this measurement amounts to total-community normalization; when $R=r$, it amounts to reference-species normalization.
(This form does not encompass using the geometric average of multiple species.)
Under our model, we can rewrite the right-hand-side to obtain
\begin{align}
  \widehat{\text{abun}}_{I}(a) 
  &= \text{abun}_{I}(a) \cdot \frac{\text{efficiency}_{I}(a)}{\text{efficiency}_{R}(a)}
  \cdot \text{FE}\left[\widehat{\text{abun}_R}(a)\right],
\end{align}
where $\text{FE}\left[\widehat{\text{abun}_R}(a)\right] = \widehat{\text{abun}_R}(a) / \text{abun}_R(a)$ is the fold error in the reference abundance measurement.

From here, we can derive the other results.

Methods differ based on the choice of $R$ and in how the abundance of $R$ is measured.

### 2022-02-12 abundance-measurement.Rmd
Revise abundance-measurement section

@@
 # How taxonomic bias affects abundance measurements {#abundance-measurement}
+
+Taxonomic bias is fundamentally multiplicative.
+Species vary in how easily their cells are lysed, and the number of lysed cells for a given species following DNA extraction equals the number of input cells multiplied by the probability of lysis per cell, or lysis efficiency, of the species.
+Similarly, prokaryotic species vary in their number of 16S genes, such that the numbers of 16S sequencing reads for various species in a 16S amplicon experiment is expected to be proportional to the numbers of 16S genes in their genomes.
+Nevertheless, the nature of MGS experiments are such that consistent taxonomic bias creates non-proportional error in common forms of relative and absolute abundance measurements---error that varies across samples and hence distort cross-sample comparisons.
+This section reviews the theoretical results of @mclaren2019cons for how taxonomic bias affects MGS measurements of relative abundance and extends them to include measurements of absolute abundance.
+We show that only some of the measurement methods currently in use have proportional errors, making these methods intrinsically more robust to bias than the others which do not.
 
@@
 Finally, unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
-These assumptions, though unrealistic descriptions of most MGS experiments, serve our purpose of clearly demonstrating when and why consistent taxonomic bias creates errors in DA analysis.
-The results we derive from them form the basis for understanding the impact of bias in more complex scenarios.
+These assumptions, though unrealistic descriptions of most MGS experiments, serve the purpose of clearly demonstrating when and why consistent taxonomic bias creates errors in DA analysis.
+<!-- The results we derive from them form the basis for understanding the impact of bias in more complex scenarios. -->
 
@@
 The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
-The sample-specific factor, $\text{effort}(a)$, we call the *sequencing effort* for that sample; it captures the variation in the total number of assigned reads due to experimental features such as library normalization and total sequencing-run output.
+The sample-specific factor, $\text{effort}(a)$, quantifies the *effective experimental effort* per unit abundance in measuring the sample.
+<!-- It accounts for variation in sample input volume and total sequencing run output, as well as the mean species efficiency within the sample. -->
 Equation \@ref(eq:measurement-model) implies that the total number of assigned reads in sample $a$ equals
@@
 \end{align}
-is the _mean efficiency_ over all species in $S$.
+is the _mean efficiency_ over all species in $S$ weighted by their abundance.
 <!-- The subscript $S$ serves as a reminder that many reads may not be assigned---for example, reads that fail to pass quality control or that come from the host or other taxonomic domains than what is being studied.
@@
 
-## Relative abundance (proportions and ratios)
+## Relative abundance (proportions and ratios) {#relative-abundance}
 
-We first review results from @mclaren2019cons on how bias affects measurements of relative abundance, which are essential to understanding both relative and absolute DA analysis.
 We distinguish between two types of species-level *relative abundances* within a sample.
@@
 Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.
+A demonstration in artificially constructed ('mock') communities of bacterial cells is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
 
@@
 For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
+A demonstration in artificially constructed ('mock') communities of bacterial cells is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
+
+Thus taxonomic bias creates variable fold errors in species proportions, but constant fold errors in species ratios.
+Consequently, a perfect cancellation of errors can only occur in ratio-based DA analysis.
 
@@
 <!-- end figure -->
-
-Thus taxonomic bias creates variable fold errors in species proportions, but constant fold errors in species ratios.
-Consequently, a perfect cancellation of errors can only occur in ratio-based DA analysis.
 
@@
 
-## Absolute abundance
+## Absolute abundance {#absolute-abundance}
 
 Various methods can be used to convert the relative abundances from MGS into absolute abundances.
-These methods are subject to taxonomic bias in the MGS measurements and, in some cases, in supplemental non-MGS measurements used for the conversion.
-Here we describe how various methods for measuring absolute abundance are predicted to be affected by taxonomic bias, with an eye towards determining whether the measurement error is variable or constant.
-We consider two general classes of methods: Those in which absolute-abundance information is derived from measurement of the aggregate abundance of the total community, and those in which it is derived from targeted measurement (or prior knowledge) of one or more particular species.
+These methods are subject to taxonomic bias in the MGS measurements and in any supplemental, non-MGS measurements that may be used for the conversion.
+Appendix [REF] develops a general framework to understand how bias affects absolute-abundance measurements for species and higher-order taxa across these methods.
+Here we present our main results as they apply to currently-used methods.
 
-### Total-community normalization
+### Methods that measure total-community abundance
 
-Measurements of species absolute abundances are often obtained by making a (non-MGS) measurement of total community abundance, equating it with the aggregate abundance of the species $S$ measured by MGS, and multiplying this total by the proportions from MGS,
+The definition \@ref(eq:prop) for the proportion of a species indicates that we can convert proportions to absolute abundances were we to know the absolute abundance of the total community.
+Given our measured proportions from MGS and a measure of total abundance from a supplemental, non-MGS measurement, $\widehat{\text{abun}}_{S}(a)$, one can measure the abundances of individual species by
 \begin{align}
@@
 \end{align}
-This measurement is affected by taxonomic bias in the MGS measurement as well as in the total-abundance measurement.
-For example, measurement of total community abundance by 16S qPCR is affected by variation among species in extraction efficiency, 16S copy number, and PCR binding and amplification bias.
-If the species-level efficiencies of the total-abundance measurement are constant across samples and we neglect other sources, then we can express the total-abundance measurement as
+Total-abundance measurements that have recently used for this purpose include counting cells with microscopy (@lloyd2020evid) or flow cytometry (@props2017abso,@vandeputte2017quan,@galazzo2020howt), measuring the concentration of a marker-gene with qPCR or ddPCR (@zhang2017soil,@barlow2020aqau, @galazzo2020howt,@tettamantiboshier2020comp), and measuring bulk DNA concentration with a florescence-based DNA quantification method (@contijoch2019gutm).
+
+Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
+Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
+The marker-gene concentration measured by qPCR is affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency.
+We can easily understand the impact of taxonomic bias on the total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
+Suppose that each species has an _absolute efficiency_ for the total-abundance measurement that is constant across samples.
+Neglecting other sources of random and systematic error, the total-abundance measurement equals
 \begin{align}
@@
 <!-- Note, we have assumed that the total-abundance and MGS measurements are sensitive to the same species -->
-The equations \@ref(eq:total-density-error) and \@ref(eq:prop-error) for the error in total-abundance and proportion measurements imply that the species abundance measurement \@ref(eq:density-prop-meas) has error given by
+
+The species abundance measurements are affected by taxonomic bias in both the MGS and total-abundance measurement.
+We can compute this effect by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
 \begin{align}
@@
 \end{align}
-
 Equation \@ref(eq:density-prop-error) indicates that the multiplicative error in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
 As in the case of proportions (Equation \@ref(eq:prop-error)), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
-On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to a relatively stable error.
-We discuss how this possibility might apply to real experimental workflows below.
+On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to more stable errors.
+We discuss how this possibility might be exploited in real experimental workflows in Section \@ref(solutions).
 
-### Reference-species normalization
+### Methods that determine or measure a reference species
 
-Suppose we had a measurement of the (absolute) abundance of a _reference species_ $r$.
-In the absence of taxonomic bias, all species are expected to have the same ratio of reads to abundance in a sample (Equation \@ref(eq:measurement-model)).
-Thus the abundance-to-reads ratio for species $r$ can serve as a conversion factor allowing us to obtain the abundance of an arbitrary species $i$,
+The definition \@ref(eq:ratio) for the ratio between two species suggests another method to convert relative to absolute abundances: 
+By knowing the ratio of species $i$ to a _reference species_ $r$ and the abundance of species $r$, we can determine the abundance of species $i$.
+Thus given an MGS measurement and a measurement $\widehat{\text{abun}}_{r}(a)$ of the abundance of species $r$, we can measure the abundance of an arbitrary species $i$ as
 \begin{align}
@@
 \end{align}
-The abundance of one or more reference species can be directly measured using targeted measurement methods like species-specific qPCR and used with Equation 
-\@ref(eq:density-ratio-meas) to obtain absolute abundance for all species.
-To our knowledge, this approach has not previously been suggested.
-<!-- (Appendix \@ref(model-details) considers possible extensions to multiple reference species.) -->
-Instead, reference-based measurement has been used in the context of spike-in experiments or normalization to a species (such as the host) which is treated as having a constant abundance.
-In a spike-in experiment, the abundance of reference species are added in known (up to experimental error) abundances to each sample.
-When normalizing to a species that is treated as having a constant but unknown abundance, we set $\widehat{\text{abun}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) equal to 1 and interpret the abundances as having fixed but unknown units, which is sufficient for multiplicative DA analysis.
+Here we consider some practical applications of this method and how their measurements are affected by bias in the MGS measurement and error in $\widehat{\text{abun}}_{r}(a)$.
+Appendix [REF] describes the general case where there is a set $R$ of multiple reference species; here we restrict our attention to a single reference species $r$.
 
-The error in the abundance measurement \@ref(eq:density-ratio-meas) due to taxonomic bias in the MGS measurement is given by
+This approach has so far been solely or primarily used with reference species that have been determined to have an abundance that is constant across samples.
+In a spike-in experiment, researchers deliberately add (or 'spike in') one or more reference species at a known, constant abundance.
+For a single spike-in species $r$, the abundances of other species can be estimated by Equation \@ref(eq:density-ratio-meas).
+In the absence of spike-ins, researchers may be able to determine one or more native species whose abundances are thought to be constant across samples; we refer to such species as _housekeeping species_ by analogy with the housekeeping genes used for absolute-abundance conversion in studies of gene expression.
+Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference.
+A related approach involves computationally identifying species that are constant between pairs of samples (@david2014host) or between sample conditions (@mandal2015anal, @kumar2018anal) under added statistical assumptions about the existence of such species.
+The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set $\widehat{\text{abun}}_{r}(a)$ to 1 in Equation \@ref(eq:density-ratio-meas).
+The resulting abundance measurements have unknown but fixed units, which is sufficient for measuring fold changes across samples.
+
+Equation \@ref(eq:density-ratio-meas) makes clear that the reference species need not be constant so long as we can directly measure it.
+This observation suggests a new approach to obtaining species absolute abundances, by combining MGS measurements of the community with targeted measurements of the absolute abundance of one or more individual species.
+Common targeted measurements include using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA, using primers tailored to a particular species.
+Other methods that are capable of measuring cells (rather than extracted DNA) include applying ddPCR directly to cells, performing CFU counting on selective media, and using flow cytometry with species-specific florescent probes.
+The measurements from any of these methods could be used for $\widehat{\text{abun}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) to obtain the absolute abundance for all species.
+Appendix [REF] describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
+
+Despite the diversity of these methods, each can be represented by Equation (\#eq:density-ratio-meas), making it straightforward to understand the impact of taxonomic bias in the MGS measurement.
+It follows from Equation \@ref(eq:ratio-error) that the measured abundance of species $i$ is 
 \begin{align}
@@
   = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
-  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) \end{array}.
+  \cdot \underbrace{\frac{\widehat{\text{abun}_r}(a)}{\text{abun}_r(a)}}_{\substack{\text{fold error in} \\ \text{ref. species}}}.
+  % \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) 
 \end{align}
-The constant error in the measured ratio of species $i$ to $r$ (see Equation \@ref(eq:ratio-error)) propagates to the abundance measurement.
-There will generally also be systematic error in the abundance of the reference species; however, if the systematic fold error is constant across samples, so will be that of the abundances of other species.
+The error thus consists of two terms: the ratio of the efficiency between species $i$ and species $r$, and any error in the abundance of the reference species.
+The ratio of species efficiencies is constant by assumption; thus the total error is constant if the error in the reference abundance is.
+Note, if a higher-order taxon is used as a reference---for example, using targeted measurements of a genus or family---then the efficiency of the reference taxon may vary across samples, leading to non-proportional errors.
 
-Spike-ins are instead sometimes used to measure absolute abundances using Equation \@ref(eq:density-prop-meas).
-In this case, an intermediate step is taken in which the total community abundance is first measured by the ratio of non-spike-in to spike-in reads.
-<!-- CHECK -->
-If done correctly, this calculation yields results that are identical to directly applying Equation \@ref(eq:density-ratio-meas) (Appendix \@ref(total-density-ref)).
+In practice, researchers commonly use a less-direct but equivalent approach to measure species abundances from a reference such as a spike-in.
+Set $S$ to be the species of interest excluding the spike-in species $r$.
+In this approach, one measures the total abundance of the community $S$ from the ratio of non-spike-in to spike-in reads,
+\begin{align}
+  (\#eq:total-density-spike)
+  \widehat{\text{abun}}_S(a) 
+  &= \frac{\text{reads}_S(a) \cdot \widehat{\text{abun}}_R(a)}{\text{reads}_r(a)}.
+\end{align}
+Then, one uses this measurement $\widehat{\text{abun}}_{S}(a)$ along with the MGS proportions in Equation \@ref(eq:density-prop-meas) to determine the abundance of individual species.
+This calculation yields measurements that are identical to those from directly applying Equation \@ref(eq:density-ratio-meas) (Appendix \@ref(total-density-ref)).
+This observation might seem paradoxical: Equation \@ref(eq:density-ratio-error) indicates proportional error for the direct approach, whereas \@ref(eq:density-prop-error) suggests non-proportional errors for the indirect approach.
+This apparent contradiction is resolved by the fact the measurement $\widehat{\text{abun}}_{S}(a)$ by this method has errors that are proportional to $\text{efficiency}_S(a)$ and thus exactly offset the proportionality with $1/\text{efficiency}_S(a)$ in the MGS proportions, such that the measured species abundances have constant fold errors.
 
-**Difference between the two approaches:**
-Reference-species normalization yields constant fold errors because it is based on species-level read counts and abundances and we assume that efficiencies are constant at the species level.
-In contrast, total-abundance normalization is based on aggregates of species (for the calculation of proportion and the total-abundance measurement) and so depends on mean efficiencies, which can vary across samples.
+## Summary
+
+Taxonomic bias in MGS measurements makes all MGS-based measures of relative and absolute abundance inaccurate.
+For some measures---but not all---consistent bias at the species level creates proportional errors in species abundances.
+Of two types of relative abundance, the ratio between a pair of species has proportional errors whereas the proportion of an individual species does not. 
+Instead, the multiplicative error in a species' proportion varies with the mean efficiency across samples.
+Methods for measuring species absolute abundance that involve measuring the abundance of the total community likewise experience non-proportional errors which vary with the sample mean efficiency; 
+a notable exception is if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement.
+Absolute-abundance methods that instead involve measuring or spiking individual reference species yield proportional errors, provided that the abundance of the reference species can be determined up to a proportional error.
+The efficiency of higher-order taxa may vary across samples even if the efficiencies of their constituent species are constant; as a result, all standard abundance measures can have non-proportional errors at taxonomic resolutions higher than that at which bias is conserved.
+
+<!-- Possibly:  -->
+<!-- - taxonomic aggregates need not have proportional errors -->
+<!-- - the fundmental mathematical explanation spanning these results is that the sum of multiple values each having proportional error need not itself have proportional error. -->

### 2022-02-14 introduction.Rmd
Revise introduction
- Edit the end of the paragraph introducing taxonomic bias
- Revise paragraph on taxonomic bias and DA analysis
- Remove figure

@@
 Although often associated with features specific to marker-gene sequencing—the variation among species in marker copy numbers and in primer-binding and amplification efficiencies—the existence of large variation in DNA extraction efficiencies and in the ability to correctly classify reads make taxonomic bias a universal feature of both marker-gene and shotgun measurements.
-As a result of taxonomic bias, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo, @mclaren2019cons).
-These errors have been found in some cases to supersede sizable biological differences (e.g.
-@lozupone2013meta) and have may have contributed to failed replications of prominent findings in the human microbiome literature, such as the associations of Bacteroides and Firmicutes in stool with obesity (@finucane2014atax) and the associations of certain species in the vagina of pregnant women with preterm birth (@callahan2017repl).
+As a result, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo, @mclaren2019cons).
+These errors can supersede sizable biological differences (e.g.
+@lozupone2013meta) and may have contributed to failed replications of prominent findings such as the associations of Bacteroides and Firmicutes in stool with obesity (@finucane2014atax) and the associations of species in the vaginas of pregnant women with preterm birth (@callahan2017repl).
 
 <!-- P: Taxonomic bias and DA -->
-The extent to which taxonomic bias has impacted the DA results in the scientific literature is unknown.
-The typical approach taken to counter taxonomic bias is to standardize the measurement protocol used within a given study, with the (often tacit) assumption being that samples measured by the same protocol will be affected by bias in the same way and so the measured differences between samples will be unaffected.
-For example, if taxonomic bias were to cause the measured proportion of a species to consistently be 10X too high, we would still be able to accurately infer the fold change in its proportion across samples (@kevorkian2018esti, @lloyd2020evid).
-Unfortunately, mathematical arguments and analysis of experiments with artificially constructed ('mock') communities demonstrate that this assumption is not always warranted: 
-Consistent taxonomic bias can lead to variable fold errors in species' proportions (Figure \@ref(fig:basic-problem), @mclaren2019cons).
-These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase) (@mclaren2019cons).
-Yet @mclaren2019cons also found that certain types of DA analysis---those based on fold changes in the ratios among species---were robust to bias.
+The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
+Statistical analyses are then conducted with the (often tacit) assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
+This argument is at least intuitively plausible for DA analyses based on multiplicative or fold changes in a taxon's abundance.
+If bias caused the abundance of a species to be consistently measured as 10X too high---what is known as _proportional error_---then we would still correctly measure its multiplicative variation across samples.
+<!-- (@kevorkian2018esti, @lloyd2020evid). -->
+However, @mclaren2019cons showed mathematically and with MGS measurements of artificially constructed ('mock') communities that consistent taxonomic bias can create non-proportional errors that can majorly distort cross-sample comparisons.
+In particular, they showed that a species' proportion---the most common measure of relative abundance---has non-proportional errors which affect the inferred changes between samples and can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
+Yet @mclaren2019cons also found that other abundance measures---those based on the ratios among species---have proportional errors and may lead to more robust DA analyses.
 The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.
@@
 Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.
-
-<!-- begin figure -->
-
-```{r basic-problem, fig.cap = '(ref:cap-basic-problem)'}
-# NOTE: call to include_graphics() has to be in top level expression
-fs::path(
-  "notebook/_posts/2021-10-27-illustrate-the-problem/illustrate-the-problem_files", 
-  "figure-html5/brooks_lcrisp_prop_and_fc-1.svg"
-  ) %>%
-  include_svg
-```
-
-(ref:cap-basic-problem) **Mock community experiments show that taxonomic bias can distort the measured fold change in an individual species' proportion across samples.** The figure shows the measured vs.
-actual proportions for a single bacterial species, _Lactobacillus crispatus_, in a set of bacterial cellular mock communities, and the resulting fold changes between community samples.
-The inconsistent error in the measured proportions of individual samples (Panel A) leads to inaccurate measurements of fold changes (Panel B).
-Mock communities were constructed and measured with 16S sequencing by @brooks2015thet.
-The data was re-analyzed by @mclaren2019cons, who showed that despite the inconsistency of the errors in Panel A, taxonomic bias acted consistently across samples.
-
-<!-- end figure -->

### 2022-02-15 differential-abundance.Rmd
Revise differential-abundance section
Revise section intro and subsection on fold changes between pairs of
samples. Add placeholder for subsection on rank-based analyses.

@@
 
-How do the errors caused by taxonomic bias in the abundances measured for individual samples impact DA analysis?
+This section turns to how these errors in individual sample measurements described in the previous section affect the results of cross-sample comparisons and DA analyses of many samples.
 <!-- the changes in abundance that we estimate between samples or different host and environmental conditions? -->
-Although there are many ways to quantify the changes in abundance that form the basis of a DA analysis, we focus on DA analyses of the (log) fold changes in proportions, ratios, and absolute abundance; these multiplicative difference measures are common (though not ubiquitous) in extant DA analyses and have more direct ecological interpretations (via the processes of exponential growth and decay) than non-multiplicative measures.
+Microbiome researchers employ many measures for the change in species' abundance across samples.
+We focus on analyses of multiplicative or (log) fold changes, which are common across study types and have more direct ecological interpretations than other measures (via the processes of exponential growth and decay).
+In addition, we briefly consider non-parametric rank-based analyses that are common in microbiome-wide association studies.
 
@@
 
-The building blocks of a DA analysis are the fold changes (FCs) in abundance between individual pairs of samples.
+The building blocks of a multiplicative DA analysis are the fold changes (FCs) in abundance between individual pairs of samples.
+Intuitively, only abundance measurements that have proportional errors will have FCs that are completely robust to bias.
 
 <!-- P: Proportions and ratios -->
-The impact of bias on the measured FCs in species proportions and ratios follows directly from the results of Section \@ref(abundance-measurement) for the error in individual-sample measurements.
-From Equation \@ref(eq:prop-error), it follows that the measured FC in the proportion of species $i$ from sample $a$ to sample $b$ is
+The phenomenon by which non-proportional errors due to bias distort measured FCs can be seen most simply for species proportions.
+It follows from Equation \@ref(eq:prop-error) that the measured FC in the proportion of a species $i$ from samples $a$ to $b$ is
 \begin{align}
   (\#eq:prop-fc-error)
-% \tag*{Fold change in proportion}
 \underbrace{\frac{\widehat{\text{prop}}_{i}(b)}{\widehat{\text{prop}}_{i}(a)}} _\text{measured FC}
@@
 \end{align}
-The sample-independent efficiency factor cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency.
-In contrast, when we use Equation \@ref(eq:ratio-error) to compute the error in the FC in the ratio between two species $i$ and $j$, we find that the constant error $\text{efficiency}_{i} / \text{efficiency}_{j}$ exactly cancels, so that the measured FCs remain accurate regardless of whether the mean efficiency varies.
+The sample-independent efficiency factor cancels, but the sample-dependent mean efficiency does not, leaving an error in the measured FC equal to the inverse change in mean efficiency.
+In contrast, the ratio between species $i$ and $j$ has proportional error and so its FC is unaffected by bias.
+From Equation \@ref(eq:ratio-error),
+\begin{align}
+  (\#eq:ratio-fc-error)
+\underbrace{\frac{\widehat{\text{ratio}}_{i/j}(b)}{\widehat{\text{ratio}}_{i/j}(a)}} _\text{measured FC}
+  &= \frac
+    {\text{ratio}_{i/j}(b) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
+    {\text{ratio}_{i/j}(a) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
+\\[0.5ex]
+  &=
+  \underbrace{\frac{\text{ratio}_{i/j}(b)}{\text{ratio}_{i/j}(a)}}_\text{actual FC}
+  ;
+\end{align}
+the constant error, equal to the ratio between the species' efficiencies, completely cancels, leaving an accurately measured FC.
 
@@
 Here the mean efficiency decreases by a factor of 2.6 (FC of 0.4X) from Sample 1 to Sample 2, which causes the FC of the proportion of each species to be measured as 2.6X larger than its true value.
-Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction; we refer to the latter as a _sign error_ since in reference to the change in sign in the corresponding log fold change (LFC).
+Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction; we refer to the latter as a _sign error_ in reference to the corresponding log fold change (LFC).
 We can see each type of error in Figure \@ref(fig:error-proportions).
@@
 
-These differences in the FCs of proportions versus ratios are mirrored when we compare FCs in absolute abundance made with normalization to a total-abundance measurement versus a reference-species measurement.
-FCs computed from total-abundance normalization are subject to error when the ratio of the two mean efficiencies (that for the MGS measurement and for the total-abundance measurement vary across samples).
-In contrast, FCs computed from reference-species normalization are not subject to error, so long as the (fold) error in the assumed or measured abundance of the reference species is constant.
+Absolute-abundance measurements for which bias creates non-proportional errors are subject to the same limitation as proportions.
+In particular, abundance measured by normalization of MGS proportions to a total-abundance measurement will yield errors in FCs given by the inverse change in the mean efficiency, unless this change is offset by error in the total-abundance measurement itself.
+<!-- (see Section \@ref(@absolute-abundance)). -->
+On the other hand, abundances measured by normalization to one or more reference species are capable of having proportional errors that cancel in FC calculations.
 
@@
 <!-- end Figure -->
+
+### [TODO] Rank-based analyses

### 2022-02-15 case-studies.Rmd
Make clarifying edits to case studies

@@
 
-@leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gain by a species from being allowed to colonize first.
+@leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gained by a species from being allowed to colonize first.
 Both analyses were based on fold changes in species proportions and so in principle were sensitive to taxonomic bias.
-To ensure the results were accurate, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.
+To improve accuracy, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.
 
-Calibration had negligible impact on the results (personal communication with Devin Leopold and confirmed by our own reanalysis).
+We repeated the two DA analyses of @leopold2020host with and without calibration and found that the results did not meaningfully differ.
 To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \@ref(fig:leopold2020host-variation)).
@@
 
-We sought to further understand the implications of the sparsity of gut microbiome for the effect of bias on DA analyses in the context of a real DA analysis.
+We sought to further understand the implications of the sparsity of gut microbiome for the effect of bias on DA analyses in the context of a real DA analysis.^[Some of the results in this paragraph seem to not hold up to more careful investigation.
+More generally, this paragraph is quite speculative and may get cut from subsequent versions.]
 @vieirasilva2019quan analyzed variation in absolute abundance of genera in stool samples from patients with primary sclerosing cholangitis and/or inflammatory bowel disease.
@@
 
-## Summary and conclusions
+## Summary
 

### 2022-02-15 v/7d91e999ba30d25a6d392259ff11b2447181a0f9/_main.tex
Make clarifying edits to case studies
This commit adds the rendered manuscript associated with
7d91e999ba30d25a6d392259ff11b2447181a0f9

--- 2022-01-16 v/414f6e78cc9b29c8236c8c3e996b68cd4d733b86/_main.tex
@@
 \hypersetup{
-  pdftitle={Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements},
+  pdftitle={Differential abundance analysis of taxonomically-biased microbiome measurements},
   pdfauthor={Michael R. McLaren; Jacob T. Nearing; Amy D. Willis; Karen G. Lloyd; Benjamin J. Callahan},
@@
 
-\title{Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements}
+\title{Differential abundance analysis of taxonomically-biased microbiome measurements}
 \author{Michael R.
@@
 Callahan\footnote{North Carolina State University}}
-\date{2022-01-16}
+\date{2022-02-15}
 
@@
 
-\emph{This manuscript was rendered from commit 414f6e78cc9b29c8236c8c3e996b68cd4d733b86.}
+\emph{This manuscript was rendered from commit 7d91e999ba30d25a6d392259ff11b2447181a0f9.}
 
@@
 Analyses of both relative and absolute abundances are considered.
-In its current form, the manuscript sits somewhere between a standard scientific article and a monograph;
-it consists of an article followed by a series of appendices which together give a comprehensive treatment of the implications of the McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) model of taxonomic bias for differential-abundance analysis and experimental design.
-It is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
+The manuscript is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
 See \href{https://doi.org/10.5281/zenodo.4552717}{the Zenodo record} for how to cite the latest version.
@@
 \section{Introduction}\label{introduction}}
+
+\leavevmode\vadjust pre{\hypertarget{abstract}{}}%
+\textbf{Abstract:}
+To assess how the species in microbial communities vary across conditions, researchers often measure the relative or absolute abundances of all species simultaneously using marker-gene and metagenomic sequencing.
+These measurements are taxonomically biased, with some species measured many times more efficiently than others.
+Measurement error caused by taxonomic bias is generally ignored in such \emph{differential abundance (DA) analyses}, likely due to a combination of 1) a widespread belief that protocol standardization is sufficient to yield accurate differences between samples and 2) a lack of practical solutions for addressing it.
+We use theoretical arguments and analysis of real and simulated experiments to analyze the impact of taxonomic bias on relative and absolute DA analyses.
+We show that taxonomic bias does in fact create inferential errors for a class of popular DA methods; however, whether these errors are biologically significant depends on experimental context.
+We present six approaches to addressing the error, each suited to different experimental questions and systems.
+These approaches provide practical solutions to mitigating the effect of taxonomic bias for a large fraction of DA questions.
+If adopted, they may lead to notable improvements in the reproducibility and biological interpretability of the statistical findings of microbiome studies.
 
@@
 Although often associated with features specific to marker-gene sequencing---the variation among species in marker copy numbers and in primer-binding and amplification efficiencies---the existence of large variation in DNA extraction efficiencies and in the ability to correctly classify reads make taxonomic bias a universal feature of both marker-gene and shotgun measurements.
-As a result of taxonomic bias, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (Yeh et al.
+As a result, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (Yeh et al.
 (\protect\hyperlink{ref-yeh2018taxo}{2018}), McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-These errors have been found in some cases to supersede sizable biological differences (e.g.
+These errors can supersede sizable biological differences (e.g.
 Lozupone et al.
-(\protect\hyperlink{ref-lozupone2013meta}{2013})) and have may have contributed to failed replications of prominent findings in the human microbiome literature, such as the associations of Bacteroides and Firmicutes in stool with obesity (Finucane et al.
-(\protect\hyperlink{ref-finucane2014atax}{2014})) and the associations of certain species in the vagina of pregnant women with preterm birth (Callahan et al.
+(\protect\hyperlink{ref-lozupone2013meta}{2013})) and may have contributed to failed replications of prominent findings such as the associations of Bacteroides and Firmicutes in stool with obesity (Finucane et al.
+(\protect\hyperlink{ref-finucane2014atax}{2014})) and the associations of species in the vaginas of pregnant women with preterm birth (Callahan et al.
 (\protect\hyperlink{ref-callahan2017repl}{2017})).
 
-The extent to which taxonomic bias has impacted the DA results in the scientific literature is unknown.
-The typical approach taken to counter taxonomic bias is to standardize the measurement protocol used within a given study, with the (often tacit) assumption being that samples measured by the same protocol will be affected by bias in the same way and so the measured differences between samples will be unaffected.
-For example, if taxonomic bias were to cause the measured proportion of a species to consistently be 10X too high, we would still be able to accurately infer the fold change in its proportion across samples (Kevorkian et al.
-(\protect\hyperlink{ref-kevorkian2018esti}{2018}), Lloyd et al.
-(\protect\hyperlink{ref-lloyd2020evid}{2020})).
-Unfortunately, mathematical arguments and analysis of experiments with artificially constructed (`mock') communities demonstrate that this assumption is not always warranted:
-Consistent taxonomic bias can lead to variable fold errors in species' proportions (Figure \ref{fig:basic-problem}, McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase) (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-Yet McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that certain types of DA analysis---those based on fold changes in the ratios among species---where robust to bias.
-The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as typically done in microbiome association testing, have yet to be investigated.
+The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
+Statistical analyses are then conducted with the (often tacit) assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
+This argument is at least intuitively plausible for DA analyses based on multiplicative or fold changes in a taxon's abundance.
+If bias caused the abundance of a species to be consistently measured as 10X too high---what is known as \emph{proportional error}---then we would still correctly measure its multiplicative variation across samples.
+However, McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed mathematically and with MGS measurements of artificially constructed (`mock') communities that consistent taxonomic bias can create non-proportional errors that can majorly distort cross-sample comparisons.
+In particular, they showed that a species' proportion---the most common measure of relative abundance---has non-proportional errors which affect the inferred changes between samples and can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
+Yet McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that other abundance measures---those based on the ratios among species---have proportional errors and may lead to more robust DA analyses.
+The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.
 
@@
 
-\begin{figure}
-\centering
-\includegraphics{figures/export-pdf/notebook/_posts/2021-10-27-illustrate-the-problem/illustrate-the-problem_files/figure-html5/brooks_lcrisp_prop_and_fc-1.pdf}
-\caption{\label{fig:basic-problem}\textbf{Mock community experiments show that taxonomic bias can distort the measured fold change in an individual species' proportion across samples.} The figure shows the measured vs.~actual proportions for a single bacterial species, \emph{Lactobacillus crispatus}, in a set of bacterial cellular mock communities, and the resulting fold changes between community samples.
-The inconsistent error in the measured proportions of individual samples (Panel A) leads to inaccurate measurements of fold changes (Panel B).
-Mock communities were constructed and measured with 16S sequencing by Brooks et al.
-(\protect\hyperlink{ref-brooks2015thet}{2015}).
-The data was re-analyzed by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}), who showed that despite the inconsistency of the errors in Panel A, taxonomic bias acted consistently across samples.}
-\end{figure}
-
 \hypertarget{abundance-measurement}{%
@@
 
-We begin by considering how taxonomic bias affects the relative- and absolute-abundance measurements which serve as the input to microbiome DA analyses.
+Taxonomic bias is fundamentally multiplicative.
+Species vary in how easily their cells are lysed, and the number of lysed cells for a given species following DNA extraction equals the number of input cells multiplied by the probability of lysis per cell, or lysis efficiency, of the species.
+Similarly, prokaryotic species vary in their number of 16S genes, such that the numbers of 16S sequencing reads for various species in a 16S amplicon experiment is expected to be proportional to the numbers of 16S genes in their genomes.
+Nevertheless, the nature of MGS experiments are such that consistent taxonomic bias creates non-proportional error in common forms of relative and absolute abundance measurements---error that varies across samples and hence distort cross-sample comparisons.
+This section reviews the theoretical results of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) for how taxonomic bias affects MGS measurements of relative abundance and extends them to include measurements of absolute abundance.
+We show that only some of the measurement methods currently in use have proportional errors, making these methods intrinsically more robust to bias than the others which do not.
+
+\hypertarget{model-of-mgs-measurement}{%
+\subsection{Model of MGS measurement}\label{model-of-mgs-measurement}}
 
@@
 (\protect\hyperlink{ref-gloor2017micr}{2017})).
-We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes to taxonomically assign reads to a set of microbial species \(S\).
-Various forms of taxonomic assignment are possible; for simplicity, we suppose that reads are assigned to the species level, with reads that cannot be uniquely assigned being discarded.
-We ignore the possibility that reads are incorrectly assigned to the wrong species or sample.
-Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the `random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
-
-In our model, the assigned read count of a species \(i\) in a sample \(a\) equals its abundance multiplied by species-specific and sample-specific factors,
+The model as described by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) considers only relative abundances; here we present a straightforward extension that allows us to also consider absolute abundances.
+
+We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and taxonomically assigns reads to a set of microbial species \(S\).
+We make several simplifying assumptions to facilitate our analysis and presentation.
+First, we consider only species-level assignment, and suppose that reads that cannot be uniquely assigned to a single species in \(S\) are discarded.
+Second, we ignore the possibility that reads are misassigned to either the wrong species or sample.
+Third, we suppose that taxonomic bias acts consistently across samples at the level of species---that is, a given species is always measured more efficiently than another to the same degree, regardless of variation in changes in sub-species-level composition or the sample matrix.
+Finally, unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the `random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
+These assumptions, though unrealistic descriptions of most MGS experiments, serve the purpose of clearly demonstrating when and why consistent taxonomic bias creates errors in DA analysis.
+
+Our model describes the mathematical relationship between the read counts obtained by MGS and the true absolute abundances of each species in a sample.
+For concreteness, we take \emph{absolute abundance}, or simply \emph{abundance}, to refer to cell concentration, or number of cells per unit volume, in a sample.
+Thus we take `absolute' simply to mean not relative to other species; however, our results also apply were `absolute' taken to mean all cells in a sample or in a sampled ecosystem.
+Similarly, our results apply to abundance units other than cells, such as biomass and genome copy number.
+Our model stipulates that the assigned read count of a species \(i\) in a sample \(a\) equals its abundance multiplied by species-specific and sample-specific factors,
 \begin{align}
@@
 The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
-The sample-specific factor, \(\text{effort}(a)\), we call the \emph{sequencing effort} for that sample; it captures the variation in the total number of assigned reads due to experimental features such as library normalization and total sequencing-run output.
+The sample-specific factor, \(\text{effort}(a)\), quantifies the \emph{effective experimental effort} per unit abundance in measuring the sample.
 Equation \eqref{eq:measurement-model} implies that the total number of assigned reads in sample \(a\) equals
@@
 \end{align}
-is the \emph{mean efficiency} over all species in \(S\).
-
-\hypertarget{relative-abundance-proportions-and-ratios}{%
-\subsection{Relative abundance (proportions and ratios)}\label{relative-abundance-proportions-and-ratios}}
+is the \emph{mean efficiency} over all species in \(S\) weighted by their abundance.
+
+\hypertarget{relative-abundance}{%
+\subsection{Relative abundance (proportions and ratios)}\label{relative-abundance}}
 
 We distinguish between two types of species-level \emph{relative abundances} within a sample.
-The \emph{proportion} of species \(i\) in sample \(a\) equals to its abundance relative to the total abundance of all species in \(S\),
+The \emph{proportion} of species \(i\) in sample \(a\) equals its abundance relative to the total abundance of all species in \(S\),
 \begin{align}
@@
 \end{align}
-Proportions and ratios each form the basis for popular DA methods;
-ratio-based methods are commonly referred to as Compositional Data Analysis (CoDA) methods.
-
-Taxonomic bias has a different effect on species proportions versus species ratios.
+Proportions and ratios each form the basis for popular relative DA methods.
+
 The proportion of a species is typically measured by its proportion of assigned reads,
@@
 Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.
+A demonstration in artificially constructed (`mock') communities of bacterial cells is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3C} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
 
@@
 For instance, in Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
+A demonstration in artificially constructed (`mock') communities of bacterial cells is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3D} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
+
+Thus taxonomic bias creates variable fold errors in species proportions, but constant fold errors in species ratios.
+Consequently, a perfect cancellation of errors can only occur in ratio-based DA analysis.
 
@@
 
-The fact that the multiplicative error in proportions varies while that in ratios is constant forms the basis for understanding the effect of bias on different methods for measuring absolute abundances from MGS and the extent to which bias cancels in different DA analyses.
+\textbf{Higher-order taxa:} How does bias affect the proportions and ratios of higher-order taxa, such as a genera, phyla, or even all organisms?
+More generally, we can identify an higher-order taxon \(I\) with a set of species, \(\{i \in I\}\).
+The abundance and reads of a taxon \(I\) in a sample \(a\) equal the sums over its constituent species, \(\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)\) and \(\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)\).
+The efficiency of taxon \(I\) equals the weighted average of its constituent species,
+\begin{align}
+  \label{eq:efficiency-general}
+  \text{efficiency}_I(a) 
+    \equiv \frac{\sum_{i\in I}\text{abun}_i(a)\cdot \text{efficiency}_i}{\text{abun}_I(a)}.
+\end{align}
+Thus the efficiency of a higher-level taxon may vary across samples (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+An exception is if the constituent species have the same efficiency or always appear in the same abundances relative to each other, in which case the efficiency of the higher-order taxon would stay constant.
+As an example, suppose that Species 1 and Species 2 in Figure \ref{fig:error-proportions} were in the same phylum.
+The efficiency of the phylum would then be \(\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5\) in Sample 1 and \(\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1\) in Sample 2.
+Equations \eqref{eq:prop-error} and \eqref{eq:ratio-error} continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent taxon efficiency defined by Equation \eqref{eq:efficiency-general} is used.
+In this way, we can see that both proportions and ratios among higher-order taxa may have inconsistent fold errors.
 
@@
 
-Researchers often would like to know how the \emph{absolute abundance} of a species changes.
-In this context, \emph{absolute} means not relative to other species; typically, the abundance of interest is relative to non-microbial entities such the volume, mass, or amount of host DNA in the sample.
-A wide range of experimental methods have been used to convert the relative abundances from MGS measurements into measurements of absolute abundances.
-Yet there has so far been little consideration as to how the resulting measurements are affected by taxonomic bias in the MGS measurements or in the supplemental measurements, such as flow cytometry and qPCR, that they often involve.
-Here and in Appendix \ref{review-absolute-methods} we describe how various methods for measuring absolute abundance are predicted to be affected by taxonomic bias, with an eye towards determining whether the measurement error is variable or constant.
-
-Our discussion supposes the `abundance' of interest is cell number, though it applies to other quantities, such as biomass and genome copy number, that may be more relevant in different biological contexts.
-
-We consider two general classes of methods: Those in which absolute-abundance information is derived from measurement of the aggregate abundance of the total community, and those in which it drives from targeted measurement (or prior knowledge) of one or more particular species.
-
-\textbf{Normalization by total abundance:}
-Measurements of species absolute abundances are often obtained by making a (non-MGS) measurement of total community abundance, equating it with the aggregate abundance of the species \(S\) measured by MGS, and multiplying this total by the proportions from MGS,
+Various methods can be used to convert the relative abundances from MGS into absolute abundances.
+These methods are subject to taxonomic bias in the MGS measurements and in any supplemental, non-MGS measurements that may be used for the conversion.
+Appendix {[}REF{]} develops a general framework to understand how bias affects absolute-abundance measurements for species and higher-order taxa across these methods.
+Here we present our main results as they apply to currently-used methods.
+
+\hypertarget{methods-that-measure-total-community-abundance}{%
+\subsubsection{Methods that measure total-community abundance}\label{methods-that-measure-total-community-abundance}}
+
+The definition \eqref{eq:prop} for the proportion of a species indicates that we can convert proportions to absolute abundances were we to know the absolute abundance of the total community.
+Given our measured proportions from MGS and a measure of total abundance from a supplemental, non-MGS measurement, \(\widehat{\text{abun}}_{S}(a)\), one can measure the abundances of individual species by
 \begin{align}
@@
 \end{align}
-This measurement is affected by taxonomic bias in the MGS measurement as well as in the total-abundance measurement.
-For example, measurement of total community abundance by 16S qPCR is affected by variation among species in extraction efficiency, 16S copy number, and PCR binding and amplification bias.
-If the species-level efficiencies of the total-abundance measurement are constant across samples and we neglect other sources, then we can express the total-abundance measurement as
+Total-abundance measurements that have recently used for this purpose include counting cells with microscopy (Lloyd et al.
+(\protect\hyperlink{ref-lloyd2020evid}{2020})) or flow cytometry (Props et al.
+(\protect\hyperlink{ref-props2017abso}{2017}),Vandeputte et al.
+(\protect\hyperlink{ref-vandeputte2017quan}{2017}),Galazzo et al.
+(\protect\hyperlink{ref-galazzo2020howt}{2020})), measuring the concentration of a marker-gene with qPCR or ddPCR (Zhang et al.
+(\protect\hyperlink{ref-zhang2017soil}{2017}),Barlow, Bogatyrev, and Ismagilov (\protect\hyperlink{ref-barlow2020aqau}{2020}), Galazzo et al.
+(\protect\hyperlink{ref-galazzo2020howt}{2020}),Tettamanti Boshier et al.
+(\protect\hyperlink{ref-tettamantiboshier2020comp}{2020})), and measuring bulk DNA concentration with a florescence-based DNA quantification method (Contijoch et al.
+(\protect\hyperlink{ref-contijoch2019gutm}{2019})).
+
+Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
+Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
+The marker-gene concentration measured by qPCR is affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency.
+We can easily understand the impact of taxonomic bias on the total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
+Suppose that each species has an \emph{absolute efficiency} for the total-abundance measurement that is constant across samples.
+Neglecting other sources of random and systematic error, the total-abundance measurement equals
 \begin{align}
@@
 where \(\text{efficiency}_{i}^{\text{tot}}(a)\) is the absolute measurement efficiency of species \(i\) for the total-abundance measurement and \(\text{efficiency}^{\text{tot}}_S(a)\) is the mean efficiency of the total-abundance measurement in the sample.
-The equations \eqref{eq:total-density-error} and \eqref{eq:prop-error} for the error in total-abundance and proportion measurements imply that the species abundance measurement \eqref{eq:density-prop-meas} has error given by
+
+The species abundance measurements are affected by taxonomic bias in both the MGS and total-abundance measurement.
+We can compute this effect by substituting Equations \eqref{eq:prop-error} and \eqref{eq:total-density-error} into Equation \eqref{eq:density-prop-meas}, yielding
 \begin{align}
@@
 \end{align}
-
 Equation \eqref{eq:density-prop-error} indicates that the multiplicative error in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
 As in the case of proportions (Equation \eqref{eq:prop-error}), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
-On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to a relatively stable error.
-We discuss how this possibility might apply to real experimental workflows below.
-
-\textbf{Normalization by one or more reference species:}
-Suppose we had a measurement of the (absolute) abundance of a \emph{reference species} \(r\).
-In the absence of taxonomic bias, all species are expected to have the same ratio of reads to abundance in a sample (Equation \eqref{eq:measurement-model}).
-Thus the abundance-to-reads ratio for species \(r\) can serve as a conversion factor allowing us to obtain the abundance of an arbitrary species \(i\),
+On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to more stable errors.
+We discuss how this possibility might be exploited in real experimental workflows in Section \ref{solutions}.
+
+\hypertarget{methods-that-determine-or-measure-a-reference-species}{%
+\subsubsection{Methods that determine or measure a reference species}\label{methods-that-determine-or-measure-a-reference-species}}
+
+The definition \eqref{eq:ratio} for the ratio between two species suggests another method to convert relative to absolute abundances:
+By knowing the ratio of species \(i\) to a \emph{reference species} \(r\) and the abundance of species \(r\), we can determine the abundance of species \(i\).
+Thus given an MGS measurement and a measurement \(\widehat{\text{abun}}_{r}(a)\) of the abundance of species \(r\), we can measure the abundance of an arbitrary species \(i\) as
 \begin{align}
@@
 \end{align}
-The abundance of one or more reference species can be directly measured using targeted measurement methods like species-specific qPCR and used with Equation
-\eqref{eq:density-ratio-meas} to obtain absolute abundance for all species.
-To our knowledge, this approach has not previously been suggested.
-Instead, reference-based measurement has been used in the context of spike-in experiments or normalization to a species (such as the host) which is treated as having a constant abundance.
-In a spike-in experiment, the abundance of reference species are added in known (up to experimental error) abundances to each sample.
-When normalizing to a species that is treated as having a constant but unknown abundance, we set \(\widehat{\text{abun}}_{r}(a)\) in Equation \eqref{eq:density-ratio-meas} equal to 1 and interpret the abundances as having fixed but unknown units, which is sufficient for multiplicative DA analysis.
-
-The error in the abundance measurement \eqref{eq:density-ratio-meas} due to taxonomic bias in the MGS measurement is given by
+Here we consider some practical applications of this method and how their measurements are affected by bias in the MGS measurement and error in \(\widehat{\text{abun}}_{r}(a)\).
+Appendix {[}REF{]} describes the general case where there is a set \(R\) of multiple reference species; here we restrict our attention to a single reference species \(r\).
+
+This approach has so far been solely or primarily used with reference species that have been determined to have an abundance that is constant across samples.
+In a spike-in experiment, researchers deliberately add (or `spike in') one or more reference species at a known, constant abundance.
+For a single spike-in species \(r\), the abundances of other species can be estimated by Equation \eqref{eq:density-ratio-meas}.
+In the absence of spike-ins, researchers may be able to determine one or more native species whose abundances are thought to be constant across samples; we refer to such species as \emph{housekeeping species} by analogy with the housekeeping genes used for absolute-abundance conversion in studies of gene expression.
+Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference.
+A related approach involves computationally identifying species that are constant between pairs of samples (David et al.
+(\protect\hyperlink{ref-david2014host}{2014})) or between sample conditions (Mandal et al.
+(\protect\hyperlink{ref-mandal2015anal}{2015}), Kumar et al.
+(\protect\hyperlink{ref-kumar2018anal}{2018})) under added statistical assumptions about the existence of such species.
+The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set \(\widehat{\text{abun}}_{r}(a)\) to 1 in Equation \eqref{eq:density-ratio-meas}.
+The resulting abundance measurements have unknown but fixed units, which is sufficient for measuring fold changes across samples.
+
+Equation \eqref{eq:density-ratio-meas} makes clear that the reference species need not be constant so long as we can directly measure it.
+This observation suggests a new approach to obtaining species absolute abundances, by combining MGS measurements of the community with targeted measurements of the absolute abundance of one or more individual species.
+Common targeted measurements include using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA, using primers tailored to a particular species.
+Other methods that are capable of measuring cells (rather than extracted DNA) include applying ddPCR directly to cells, performing CFU counting on selective media, and using flow cytometry with species-specific florescent probes.
+The measurements from any of these methods could be used for \(\widehat{\text{abun}}_{r}(a)\) in Equation \eqref{eq:density-ratio-meas} to obtain the absolute abundance for all species.
+Appendix {[}REF{]} describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
+
+Despite the diversity of these methods, each can be represented by Equation \label{eq:density-ratio-meas}, making it straightforward to understand the impact of taxonomic bias in the MGS measurement.
+It follows from Equation \eqref{eq:ratio-error} that the measured abundance of species \(i\) is
 \begin{align}
@@
   = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
-  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) \end{array}.
-\end{align}
-The constant error in the measured ratio of species \(i\) to \(r\) (see Equation \eqref{eq:ratio-error}) propagates to the abundance measurement.
-There will generally also be systematic error in the abundance of the reference species; however, if the systematic fold error is constant across samples, so will be that of the abundances of other species.
-
-Spike-ins are instead sometimes used to measure absolute abundances using Equation \eqref{eq:density-prop-meas}.
-In this case, an intermediate step is taken in which the total community abundance is first measured by the ratio of non-spike-in to spike-in reads.
-If done correctly, this calculation yields results that are identical to directly applying Equation \eqref{eq:density-ratio-meas} (Appendix \ref{total-density-ref}).
-
-\textbf{Difference between the two approaches:}
-Reference-species normalization yields constant fold errors because it is based on species-level read counts and abundances and we assume that efficiencies are constant at the species level.
-In contrast, total-abundance normalization is based on aggregates of species (for the calculation of proportion and the total-abundance measurement) and so depends on mean efficiencies, which can vary across samples.
+  \cdot \underbrace{\frac{\widehat{\text{abun}_r}(a)}{\text{abun}_r(a)}}_{\substack{\text{fold error in} \\ \text{ref. species}}}.
+  % \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) 
+\end{align}
+The error thus consists of two terms: the ratio of the efficiency between species \(i\) and species \(r\), and any error in the abundance of the reference species.
+The ratio of species efficiencies is constant by assumption; thus the total error is constant if the error in the reference abundance is.
+Note, if a higher-order taxon is used as a reference---for example, using targeted measurements of a genus or family---then the efficiency of the reference taxon may vary across samples, leading to non-proportional errors.
+
+In practice, researchers commonly use a less-direct but equivalent approach to measure species abundances from a reference such as a spike-in.
+Set \(S\) to be the species of interest excluding the spike-in species \(r\).
+In this approach, one measures the total abundance of the community \(S\) from the ratio of non-spike-in to spike-in reads,
+\begin{align}
+  \label{eq:total-density-spike}
+  \widehat{\text{abun}}_S(a) 
+  &= \frac{\text{reads}_S(a) \cdot \widehat{\text{abun}}_R(a)}{\text{reads}_r(a)}.
+\end{align}
+Then, one uses this measurement \(\widehat{\text{abun}}_{S}(a)\) along with the MGS proportions in Equation \eqref{eq:density-prop-meas} to determine the abundance of individual species.
+This calculation yields measurements that are identical to those from directly applying Equation \eqref{eq:density-ratio-meas} (Appendix \ref{total-density-ref}).
+This observation might seem paradoxical: Equation \eqref{eq:density-ratio-error} indicates proportional error for the direct approach, whereas \eqref{eq:density-prop-error} suggests non-proportional errors for the indirect approach.
+This apparent contradiction is resolved by the fact the measurement \(\widehat{\text{abun}}_{S}(a)\) by this method has errors that are proportional to \(\text{efficiency}_S(a)\) and thus exactly offset the proportionality with \(1/\text{efficiency}_S(a)\) in the MGS proportions, such that the measured species abundances have constant fold errors.
+
+\hypertarget{summary}{%
+\subsection{Summary}\label{summary}}
+
+Taxonomic bias in MGS measurements makes all MGS-based measures of relative and absolute abundance inaccurate.
+For some measures---but not all---consistent bias at the species level creates proportional errors in species abundances.
+Of two types of relative abundance, the ratio between a pair of species has proportional errors whereas the proportion of an individual species does not.
+Instead, the multiplicative error in a species' proportion varies with the mean efficiency across samples.
+Methods for measuring species absolute abundance that involve measuring the abundance of the total community likewise experience non-proportional errors which vary with the sample mean efficiency;
+a notable exception is if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement.
+Absolute-abundance methods that instead involve measuring or spiking individual reference species yield proportional errors, provided that the abundance of the reference species can be determined up to a proportional error.
+The efficiency of higher-order taxa may vary across samples even if the efficiencies of their constituent species are constant; as a result, all standard abundance measures can have non-proportional errors at taxonomic resolutions higher than that at which bias is conserved.
 
 \hypertarget{differential-abundance}{%
-\section{Taxonomic bias can cause errors in differential-abundance results}\label{differential-abundance}}
-
-How do the errors caused by taxonomic bias in the abundances measured for individual samples impact DA analysis?
-Although there are many ways to quantify the changes in abundance that form the basis of a DA analysis, we focus on DA analyses of the (log) fold changes in proportions, ratios, and absolute abundance; these multiplicative difference measures are common (though not ubiquitous) in extant DA analyses and have more direct ecological interpretations (via the processes of exponential growth and decay) than non-multiplicative measures.
+\section{DA methods vary in their robustness to bias}\label{differential-abundance}}
+
+This section turns to how these errors in individual sample measurements described in the previous section affect the results of cross-sample comparisons and DA analyses of many samples.
+Microbiome researchers employ many measures for the change in species' abundance across samples.
+We focus on analyses of multiplicative or (log) fold changes, which are common across study types and have more direct ecological interpretations than other measures (via the processes of exponential growth and decay).
+In addition, we briefly consider non-parametric rank-based analyses that are common in microbiome-wide association studies.
 
@@
 
-The building blocks of a DA analysis are the fold changes (FCs) in abundance between individual pairs of samples.
-
-The impact of bias on the measured FCs in species proportions and ratios follows directly from the results of Section \ref{abundance-measurement} for the error in individual-sample measurements.
-From Equation \eqref{eq:prop-error}, it follows that the measured FC in the proportion of species \(i\) from sample \(a\) to sample \(b\) is
+The building blocks of a multiplicative DA analysis are the fold changes (FCs) in abundance between individual pairs of samples.
+Intuitively, only abundance measurements that have proportional errors will have FCs that are completely robust to bias.
+
+The phenomenon by which non-proportional errors due to bias distort measured FCs can be seen most simply for species proportions.
+It follows from Equation \eqref{eq:prop-error} that the measured FC in the proportion of a species \(i\) from samples \(a\) to \(b\) is
 \begin{align}
   \label{eq:prop-fc-error}
-% \tag*{Fold change in proportion}
 \underbrace{\frac{\widehat{\text{prop}}_{i}(b)}{\widehat{\text{prop}}_{i}(a)}} _\text{measured FC}
@@
 \end{align}
-The sample-independent efficiency factor cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency.
-In contrast, when we use Equation \eqref{eq:ratio-error} to compute the error in the FC in the ratio between two species \(i\) and \(j\), we find that the constant error \(\text{efficiency}_{i} / \text{efficiency}_{j}\) exactly cancels, so that the measured FCs remain accurate regardless of whether the mean efficiency varies.
+The sample-independent efficiency factor cancels, but the sample-dependent mean efficiency does not, leaving an error in the measured FC equal to the inverse change in mean efficiency.
+In contrast, the ratio between species \(i\) and \(j\) has proportional error and so its FC is unaffected by bias.
+From Equation \eqref{eq:ratio-error},
+\begin{align}
+  \label{eq:ratio-fc-error}
+\underbrace{\frac{\widehat{\text{ratio}}_{i/j}(b)}{\widehat{\text{ratio}}_{i/j}(a)}} _\text{measured FC}
+  &= \frac
+    {\text{ratio}_{i/j}(b) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
+    {\text{ratio}_{i/j}(a) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
+\\[0.5ex]
+  &=
+  \underbrace{\frac{\text{ratio}_{i/j}(b)}{\text{ratio}_{i/j}(a)}}_\text{actual FC}
+  ;
+\end{align}
+the constant error, equal to the ratio between the species' efficiencies, completely cancels, leaving an accurately measured FC.
 
@@
 Here the mean efficiency decreases by a factor of 2.6 (FC of 0.4X) from Sample 1 to Sample 2, which causes the FC of the proportion of each species to be measured as 2.6X larger than its true value.
-Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction; we refer to the latter as a \emph{sign error} since in reference to the change in sign in the corresponding log fold change (LFC).
+Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction; we refer to the latter as a \emph{sign error} in reference to the corresponding log fold change (LFC).
 We can see each type of error in Figure \ref{fig:error-proportions}.
@@
 
-These differences in the FCs of proportions versus ratios are mirrored when we compare FCs in absolute abundance made with normalization to a total-abundance measurement versus a reference-species measurement.
-FCs computed from total-abundance normalization are subject to error when the ratio of the two mean efficiencies (that for the MGS measurement and for the total-abundance measurement vary across samples).
-In contrast, FCs computed from reference-species normalization are not subject to error, so long as the (fold) error in the assumed or measured abundance of the reference species is constant.
+Absolute-abundance measurements for which bias creates non-proportional errors are subject to the same limitation as proportions.
+In particular, abundance measured by normalization of MGS proportions to a total-abundance measurement will yield errors in FCs given by the inverse change in the mean efficiency, unless this change is offset by error in the total-abundance measurement itself.
+On the other hand, abundances measured by normalization to one or more reference species are capable of having proportional errors that cancel in FC calculations.
 
@@
 
+\hypertarget{todo-rank-based-analyses}{%
+\subsubsection{{[}TODO{]} Rank-based analyses}\label{todo-rank-based-analyses}}
+
 \hypertarget{case-studies}{%
@@
 
-Measurement of control communities along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-Gnotobiotic community experiments are well suited to this form of \emph{calibration via community controls} since, unlike most natural ecosystems, it is possible to assemble `mock communities' containing all species in known relative abundances.
-
-Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) are the first to use mock-based calibration in a gnotobiotic microbiome experiment.
+The taxonomic bias species present in `mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+In practice it is difficult to construct control communities that span the many species present in complex natural communities.
+However, gnotobiotic community experiments are well suited to this form of \emph{calibration via community controls} since, unlike most natural ecosystems, it is possible to assemble mock communities containing all species in known relative abundances.
+
 In a study of host-commensal-pathogen interactions, Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) inoculated plants with 8 commensal fungal species and subsequently exposed plants to a fungal pathogen.
-To investigate the joint effect of colonization order and host genotype, plants varied in source location and in which commensal colonized first.
-The authors used ITS amplicon sequencing to measure communities before and after pathogen infection, along with mock communities with quantified genome concentrations of DNA from the 9 species.
-These DNA mock communities allowed the authors to directly measure the bias due to the sequencing workflow following DNA extraction and correct for it when analyzing the primary experimental samples following the method of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
-Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification and the substantial ribosomal copy-number-variation (CNV) in fungi (Lofgren et al.
-(\protect\hyperlink{ref-lofgren2019geno}{2019})).
+The authors used ITS amplicon sequencing to measure communities before and after pathogen infection.
+Motivated by the substantial ribosomal copy-number variation (CNV) in fungi (Lofgren et al.
+(\protect\hyperlink{ref-lofgren2019geno}{2019})), the authors also performed control measurements of mock communities that they constructed from quantified genomic DNA of the 9 species in the experiment; these controls were used to measure taxonomic bias with the method of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
 The authors found a 13X difference between the most and least efficiently measured commensal, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
-We reanalyzed this experiment to understand the impact that bias might have had on the study's analyses were it not accounted for.
-
-Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) used the ITS sequencing profiles of the pre-infection samples to understand the impact of host genotype and an experimental treatment---which species was allowed to colonize first---on the commensal community composition prior to infection.
-Two DA analyses were conducted to address these questions.
-First, the authors used linear regression of log species proportion (via negative-binomial regression to account for random read-sampling error) to ask whether host genotype, treatment, and their interaction significantly impacted overall community composition.
-Second, the authors quantified the strength of priority effects---the advantage of being allowed to colonize first---in different host genotypes by the LFC in the proportion of a species when allowed to colonize first versus later with the majority of commensals.
-
-Since these analysis are both based on proportions, bias could in principle impact their results.
-We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for\footnote{The two sets of analyses can be seen at the following links: \href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html}{negative-binomial regression analysis}; \href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html}{priority-effects analysis}}.
-To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
-The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.62X and a geometric standard deviation of just 1.05X (SI Figure \ref{fig:leopold2020host-variation}).
-Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
-This answer relates to the fact that the mean efficiency is a weighted \emph{arithmetic} average of species proportions and so is insensitive to large multiplicative variation in low-proportion species.
-The pre-infection samples in the Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) study are always dominated by the three species with the highest efficiencies, whose efficiencies vary only by 3X (and just 1.5X between the two most dominant).
-The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions \(\ll 1\), even after bias correction) and hence don't significantly affect the mean efficiency.
-Because the log mean efficiency varies much less than the log proportions for each taxon, there is little room for it to distort the results of DA analyses based on variation in log proportions.
-
-We additionally considered the potential for bias to impact analysis of how commensal taxa responded to infection.
-In particular, it is interesting to consider whether any commensals are able to increase in absolute concentration in response to infection.
-The experiment performed by Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration measured using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}).
-It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
-The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, raising the mean efficiency of post-infection samples (SI Figure \ref{fig:leopold2020host-infection-mean-efficiency-dist}).
-Across different host genotypes, the average increase in mean efficiency ranges from 2.5X to 5.2X.
-We used simple linear regression to measure the average change in log proportion of each commensal for each host genotype, with and without bias correction.
-Regardless of whether bias is accounted for, the log proportion of each commensal decreased, which is unsurprising given the pathogen's growth and the sum-to-one constraint of species proportions.
-Without bias correction, however, the LFCs were lower by an amount corresponding to the inverse change in log mean efficiency.
-The magnitude errors created were substantial for a large fraction of commensal-host pairs; in several cases, the LFC estimates without bias correction were clearly negative, whereas the estimates with bias correction were close to 0.
-
-Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}).
-We consider two scenarios.
-\emph{Scenario 1:} Total genome concentration in each sample is perfectly known and used to measure the concentrations of individual taxa via Equation \eqref{eq:density-prop-meas}.
-In this case, bias in the MGS measurements would create absolute errors in the LFCs for genome concentration identical to those for proportions.
-The scientific error, however, may be much worse.
-Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where some taxa that persist or increase in abundance instead appear to decrease.
-\emph{Scenario 2:} Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-abundance normalization method.
-In this case, we should expect the increase in mean efficiency to have little to no impact on LFCs, since it would be offset by the increase in mean efficiency of the total-concentration measurement.
-Moreover, since the total ITS concentration and ITS sequencing are both performed on the extracted DNA, we can also expect the effect of taxonomic bias caused by DNA sequencing to be offset.
-Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-abundance measurement is made prior to normalization.
+
+Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of `priority effects'---the advantage gained by a species from being allowed to colonize first.
+Both analyses were based on fold changes in species proportions and so in principle were sensitive to taxonomic bias.
+To improve accuracy, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.
+
+We repeated the two DA analyses of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) with and without calibration and found that the results did not meaningfully differ.
+To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \ref{fig:leopold2020host-variation}).
+Despite the 13X variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \ref{fig:leopold2020host-variation}C), having a geometric range of 1.62X and a geometric standard deviation of 1.05X.
+This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \ref{fig:leopold2020host-variation}A).
+But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3X and by just 1.5X between the two most dominant.
+The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
+Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual taxa, it had a negligible impact on the inferred fold changes and the DA analyses based on them.
+
+We performed an additional DA analysis on the data from Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) to investigate whether any commensals increased in absolute concentration in response to infection.
+The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \ref{fig:leopold2020host-infection-mean-efficiency-dist}).
+Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5X to 5.2X.
+We used simple linear regression to measure the LFC in commensal species proportions pre- to post-infection, with and without calibration (SI Figure \ref{fig:leopold2020host-infection-lfc}).
+All commensals tended to decrease in proportion post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint of species proportions.
+However, when taxonomic bias was ignored the magnitude of the post-infection decrease was overestimated, by an amount corresponding to the inverse change in log mean efficiency.
+This magnitude error was substantial for the commensal-host pairs with the weakest changes; in several cases, LFC estimates with 95\% confidence intervals below 0 became indistinguishable from 0 after calibration.
+
+Although Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) did not include absolute-abundance measurements, we can consider the impact taxonomic bias would have on an absolute DA analysis in a simple scenario in which total genome concentration of each pre- and post-infection sample is perfectly known.
+The total genomic concentrations of each sample, combined with the species proportions measured by MGS, can be combined into a measurement of species concentrations using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}).
+In this case, the bias in the MGS measurements will create absolute errors in the estimated LFCs of genome concentration of equal magnitude to the errors in the LFC estimates for proportions.
+The scientific error, however, might become worse.
+Suppose that the species whose calibrated proportions appeared to hold steady or slightly decrease actually increased in absolute abundance by around 2X.
+As bias shifts our estimates downwards by 2.5X to 5.2X, we would instead conclude that these taxa decreased.
 
@@
 These shifts typically result in substantial fold changes in mean efficiency (SI Figure \ref{fig:momspi-mean-efficiency-fcs}) and can cause spurious fold changes in the trajectories of lower-abundance species (SI Figure \ref{fig:momspi-trajectory}).
+Decreases in mean efficiency during transitions from \emph{Lactobacillus} to \emph{Gardnerella} dominance would be more extreme for vaginal microbiome primer sets that fail to amplify \emph{Gardnerella} (Graspeuntner et al.
+(\protect\hyperlink{ref-graspeuntner2018sele}{2018})).
 
@@
 (\protect\hyperlink{ref-srinivasan2012bact}{2012}), Cartwright et al.
-(\protect\hyperlink{ref-cartwright2018mult}{2018})).
-Cartwright et al.
-(\protect\hyperlink{ref-cartwright2018mult}{2018}) found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100\% and specificity of 85.1\%.
+(\protect\hyperlink{ref-cartwright2018mult}{2018})).\footnote{Cartwright et al.
+(\protect\hyperlink{ref-cartwright2018mult}{2018}) found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100\% and specificity of 85.1\%.}
 In addition, it is commonly observed that samples from women with and without BV that are dominated by \emph{Lactobacillus spp.} tend to have higher diversity, whereas samples dominated by \emph{Gardnerella} tend to have lower diversity.
@@
 We split samples into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
-We then estimated the LFC in proportion from low- to high-diversity samples for a 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles.
-As expected, the mean efficiency was higher in low-diversity samples due to a larger fraction of samples dominated by \emph{Lactobacillus} and a lower fraction with samples dominated by \emph{Gardnerella}.
-The decline in mean efficiency in the high diversity group caused a concomitant increase in the LFCs of all species when bias was not accounted for.
-This error due to bias resulted in serious magnitude errors in nearly all species and sign errors in over one third, with errors of both types seen for several clinically relevant species.
-
-\hypertarget{notes}{%
-\subsubsection{Notes}\label{notes}}
-
-\begin{itemize}
-\tightlist
-\item
-  It might be cleaner story-wise to redo the diversity partitioning using the threshold identified by Cartwright et al. (\protect\hyperlink{ref-cartwright2018mult}{2018})
-\item
-  For the DA analysis of proportions vs.~diversity, I tried other types of DA analysis methods (Gamma-Poisson regression and regression on ranks, similar to Wilcoxon test) to understand how the impact of bias differs across methods.
-In each case bias has a big effect but the species whose estimates are most affected varies.
-I would need to revise and verify this analysis before including its results.
-One reason for doing so would be to show that our qualitative finding about how bias distorts results applies beyond the basic linear regression model.
-\item
-  Decreases in mean efficiency during transitions from \emph{Lactobacillus} to \emph{Gardnerella} dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify \emph{Gardnerella}.
-\end{itemize}
+We then estimated the LFC in proportion from low- to high-diversity samples for 30 species passing a basic prevalence threshold, with and without bias correction.
+A large fraction of entries in the species count matrix were small counts or zeros.
+We used therefore used gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions while accounting for uncertainty from the random sampling of reads.
+We accounted for bias by including an offset term in the linear model equal to the log ratio of the species' efficiency to the sample mean efficiency.
+
+The mean efficiency was on average lower in high-diversity samples due to a reduction in \emph{Lactobacillus} dominance (LFC of -1.6, or a 4.8X reduction).
+Our results from the previous section therefore suggest that accounting for bias should reduce the estimated LFC for each species by about this same amount.
+We found that, as expected, bias correction reduced the LFC of each species, though the effect was variable and typically somewhat smaller.
+Bias correction led to substantial magnitude changes for several species and sign changes in 3 species; however, the relative difference between the corrected and uncorrected LFCs was modest for the 19/30 species with the largest LFCs of over 2 (corresponding to a FC of 7.4X) even after correction.
+These large LFCs occur in species that rarely if ever reach high proportions and so are likely driven by the sum-to-one constraint: The high diversity communities are those not dominated by \emph{Lactobacillus} spp.
+(or to a lesser extent \emph{G.
+vaginalis} and BVAB1).
+Therefore it is not clear how well it will generalize to regressions on other covariates.
+
+\hypertarget{human-gut-microbiomes}{%
+\subsection{Human gut microbiomes}\label{human-gut-microbiomes}}
+
+An overwhelming majority of applications of DA analysis are in studies of the human gut microbiome. (True?)
+Many are microbial association tests in a case-control design---looking for taxa that are differentially abundant in one group of subjects versus another.
+It is also increasingly common to perform time series experiments to elucidate dynamics of microbial communities within a host and their interaction with dynamic host variables.
+Although interest in other microorganisms and viruses is growing, so far these are mostly of bacteria.
+Were therefore asked whether there are features of human gut bacterial microbiomes that we might expect to make DA analyses more or less robust to taxonomic bias.
+
+We first considered whether the greater diversity of the human gut microbiome as compared to the human vaginal microbiome might limit the impact of bias on species-level DA analyses.
+Intuitively, we expect the mean efficiency to vary less across samples from a given ecosystem type as alpha diversity increases, since it effectively averages over a larger number of species.
+Indeed, as shown in Appendix X, in randomly assembled communities the additive variance in the mean efficiency is inversely proportional to the order-2 \emph{true diversity} or \emph{effective number of species} in the community, which is equivalent to the popular Inverse Simpson alpha diversity index.
+But real communities do not assemble randomly.
+Moreover, it is the multiplicative variation in the mean efficiency relative to that in the species proportions that determines the impact bias has on proportion-based DA analyses; it is possible that the variation in individual species proportions is reduced to an even greater degree than that in the mean efficiency.
+Thus it is not obvious whether the gut microbiome's higher alpha diversity will effectively insulate its DA analyses from the effects of bias.
+
+To investigate this question, we analyzed species-level bacterial profiles from vaginal and stool samples from healthy subjects that had been subjected to shotgun sequencing as part of the Human Microbiome Project (Huttenhower et al.
+(\protect\hyperlink{ref-huttenhower2012stru}{2012})).{[}\^{}We obtained MetaPhlAn3 species-level bacterial relative abundances from the curatedMetagenomicData R package.{]}
+Consistent with previous findings, the stool microbiomes had substantially greater diversity than the vaginal microbiomes by several measures, including: within individuals (alpha diversity) and in total across many individuals (gamma diversity); at the species level and by measures that account for phylogenetic divergence; and for diversity measures spanning a range of sensitivity to low abundance species as quantified by Hill's \(q\), including richness (\(q=0\)), Shannon entropy (\(q=1\)), and the Inverse Simpson index (\(q=2\)).
+To understand how this greater diversity might modulate the effect of bias, we used computer simulation to assess the variation in the mean efficiency under a range of plausible taxonomic biases, with and without phylogenetic conservation of the efficiency trait.
+\emph{In each simulation replicate, the species efficiencies were sampled from a multivariate log-normal distribution.
+Phylogenetic independence was achieved by setting the covariance matrix equal to the identity matrix, so that the efficiencies were IID.
+Phylogenetic conservation was achieved by setting the covariance matrix equal to the phylogenetic covariance matrix, which amounts to Brownian evolution of the log efficiency trait.}
+As expected, the mean efficiency varied less in the higher-diversity gut microbiome samples than in the vaginal microbiome samples; the geometric standard deviation (GSD) of the mean efficiency in stool samples was on average 1.5X less than that in vaginal samples regardless of whether bias was phylogenetically conserved.
+This difference is not extreme and leaves substantial variation in the mean efficiency in gut samples.
+We suggest several possible explanations.
+First, although richness is typically around 100 bacterial species in the MetaPhlAn3 gut profiles, the more relevant order-2 effective number of species (\(^2D\); equivalent to the Inverse Simpson diversity index) is typically only around 10 and often lower.
+A order-2 diversity that is much lower than species richness arises from the skewed distribution of abundances within a gut microbiome sample, with the most abundant 1-2 species often forming a majority of reads.
+It is these high-proportion species, captured by the order-2 diversity, that ultimately drive the variation in mean efficiency, not the large number of low-proportion species that drive the richness.
+A second, related reason is that, like in the vaginal microbiome, a relatively small set of species tend to be the most abundant across individuals.
+The overall effect is similar, though less extreme, as that of the dominance of a few key species in the vaginal microbiome: Variation in the efficiencies among these species has an outsized impact on the mean efficiency, again leading to a weakening of the averaging affect when we consider variation across samples.
+
+We next compared the variation in species proportions in the two ecotypes.
+In assessing multiplicative variation, one must grapple with the sparsity of human microbiome data.
+Across both ecotypes, most species have low prevalence, appearing in just a small minority of samples.
+One way to handle these zero observations is to replace them with a small positive value, roughly representing a minimum detection limit.
+Although the multiplicative variation is highly sensitive to this value, we found that the relative difference between gut and vaginal samples was robust to values spanning \(10^{-6}\) to \(10^{-3}\).
+The average GSD of species in gut samples was around 1.8X lower than that of species in vaginal samples, regardless of zero-replacement value.
+Thus the variation in species proportions was lower in the gut samples by a similar or greater degree than the variation in mean efficiency, suggesting that bias may be just as or more problematic for inferring fold changes in species proportions.
+
+We sought to further understand the implications of the sparsity of gut microbiome for the effect of bias on DA analyses in the context of a real DA analysis.\footnote{Some of the results in this paragraph seem to not hold up to more careful investigation.
+More generally, this paragraph is quite speculative and may get cut from subsequent versions.}
+Vieira-Silva et al.
+(\protect\hyperlink{ref-vieirasilva2019quan}{2019}) analyzed variation in absolute abundance of genera in stool samples from patients with primary sclerosing cholangitis and/or inflammatory bowel disease.
+Absolute abundances of genera were obtained via the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}) with proportions measured from 16S sequencing and total abundance measured from flow cytometry.
+The authors the rank-based Spearman correlation to quantify the associations in absolute abundance and fecal calprotectin concentration, a biomarker of intestinal inflammation.
+{[}This will be described in Section \ref{differential-abundance}: Spearman correlation is equivalent to doing a regression on the rank-transformed data and is another form of DA analysis; changes in ranks of proportions are affected by bias, and changes in ranks of ratios are not.
+So rank-based methods (Wilcoxon for boolean covariate and Spearman for continuous covariate) have the same issue as LFC methods when it comes to analyzing proportions instead of ratios.{]}
+As described in Section \ref{differential-abundance}, taxonomic bias in the MGS measurement has a similar effect on analyses of absolute abundances from the total-abundance normalization as it does no analyses of the raw proportions.
+Thus we can consider whether bias might affect the Spearman correlations in the study.
+When we re-analyzed the Spearman correlations, we found that they were often driven by presence-absence of the given genus, rather than continuous variation in abundance, as evidenced by the fact that for many species, replacing the absolute genus abundances with 0 or 1 indicating presence or absence often did not affect the Spearman correlation.
+For these taxa, we can expect bias to have much less impact on the reported correlation than for taxa for which the correlation was driven by continuous variation.
+The reason is that systematic variation of the mean efficiency with the covariate (which was, in contrast, log-normally distributed) will tend to systematically shift the ranks of observations in a taxon with a continuously varying abundance, but not one where the rank correlation is driven by the proportion of non-zero observations.
+We confirmed this expectation by investigating the sensitivity of Spearman correlations for the top associations that were driven either by presence/absence variation or continuous variation, and found that the correlation was much more sensitive to bias in the second case.
+
+\emph{We might consider adding: A paragraph giving examples where there are large taxonomic shifts associated with a clinical covariate, which therefore have the potential to drive systematic variation in the mean efficiency.
+The primary positive example I have looked at is a study of children with acute diarrhea, in which stool from healthy controls has a microbiome with roughly equal amounts of Actinobacteria (Gram positive), Firmicutes (Gram positive), and Bacteroidetes (Gram negative), and Proteobacteria (Gram negative), but the diarrheic stool is dominated by Proteobacteria and specifically E.
+coli and to a lesser extent Klebsiella (both in Enterobacteriaceae, which is known to be causally associated with acute diarrhea generally).
+Any particular bias favoring E. coli or Enterobacteriaceae would have the potential to distort results, which is something we could simulate and visualize.}
 
@@
 (\protect\hyperlink{ref-lloyd2020evid}{2020}) performed a systematic assessment of growth rates of bacterial and archaeal taxa over a depth of 10 cm (corresponding to \textasciitilde40 years of burial time) in sediment of the White Oak River estuary.
-To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy (CARD-FISH).
+To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy.
 The authors refer to these absolute densities as FRAxC measurements, for `fraction of 16S reads times total cell counts'.
 The FRAxC measurements were used to infer growth rate from the slope of a simple linear regression of log cell density against burial time over the first 3 cm below the bioirrigation layer (corresponding to \(\sim 8\) years of burial).
-To validate the inference of positive growers, the authors compared the growth rates from FRAxC-based inference from two sediment cores, qPCR measurements in these cores for a few reference taxa, and FRAxC-based inference in two replicate laboratory incubation experiments.
+To validate the inference of positive growers, the authors compared the growth rates from FRAxC-based inference from two sediment cores, qPCR measurements in these cores for a few reference taxa, and FRAxC-based inference in two replicate two-year laboratory incubation experiments.
 
 Taxonomic bias could lead to systematic error in FRAxC-derived growth rates if sample mean efficiency tends to systematically vary with burial time (or equivalently, depth).
-One possibility is that microbes with tougher cell walls tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
+One possibility is that microbes with tougher cell walls may persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
 In this case, we would expect the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that decay sufficiently slowly would be mistakenly inferred to have positive growth.
@@
 
-\hypertarget{summary-and-conclusions}{%
-\subsection{Summary and conclusions}\label{summary-and-conclusions}}
+\hypertarget{summary-1}{%
+\subsection{Summary}\label{summary-1}}
 
 The impact of bias can depend on protocol, biological system, and type of DA analysis being done.
-Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of section X help suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.
+Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of Section \ref{differential-abundance} suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.
 
@@
 
-So long as the mean efficiency does not vary systematically with the regression covariates, the error in fold changes serve primarily to reduce the precision and power of regression analyses without systematically distorting estimates and so does not necessarily lead to invalid inferences.
+So long as the mean efficiency does not vary systematically with the regression covariates, the measurement error from bias primarily reduces the precision and power of regression analyses without systematically distorting estimates and so does not necessarily lead to invalid inferences.
 But the reduction in precision could be substantial, particularly given the small sample sizes common in many microbiome studies, and thereby substantially limit the ability to draw meaningful conclusions from a study\footnote{I suspect this non-systematic variation in the mean efficiency is the more typical situation, but we currently lack an example of it in the case studies.
@@
 A variety of such methods have been developed for microbiome DA analysis that build on earlier work from the field of Compositional Data Analysis (CoDA).
-These methods analyze log-fold changes in the ratios among a pair of species or, more generally, some product of exponential power of multiple species, which likewise are bias-invariant.
-Besides a greater robustness to taxonomic bias, such ratio-based analyses have other advantages.
-First, analysis of ratios avoids other standard criticisms leveled at proportion-based analysis---namely that the change in proportion of a species is affected by change in the density of all other species and the set of species used in the denominator of the proportion calculation (Gloor et al.
-(\protect\hyperlink{ref-gloor2017micr}{2017})).
-Second, ratios may be more informative than absolute density in some cases, such as for swabs or other sample types for which the absolute density in the sample may be an unreliable indicator of the absolute density of the sampled ecosystem.
+These methods analyze (log) fold changes in the ratio between species or, more generally, between the products of multiple species.
+Taking the product or geometric mean of multiple species provides a approach to aggregating species into a higher-level taxonomic units such that the ratios between units maintain proportional errors, in contrast to the standard approach of additively aggregating species.
+Besides a greater robustness to taxonomic bias, analyses of ratios have other advantages over analyses of proportions and absolute abundances.
+Analyses of ratios avoid the standard criticism leveled at proportion-based analysis---namely that the changes in the proportion of one species depend in a non-transparent way on the changes in abundance of all other species (Gloor et al.
+(\protect\hyperlink{ref-gloor2017micr}{2017})), including unidentified contaminants.
+In many experiments, the total abundance in a sample may be a poor indicator of the total abundance of the sampled ecosystem; unless this relationship can be characterized and accounted for, changes in ratios may be more informative than changes in absolute abundance for the underlying biology of interest.
 
@@
 A related problem is that, due to the discrete nature of biological organisms and the reads that are generated during sequencing, species in the denominator are commonly observed to have 0 reads; the assumptions that made by different methods about the meaning of these 0 counts can have a major impact on DA results.
-Finally, bias invariance at the level of ratios among species does not extend to ratios of additive aggregates of species, such as between bacterial phyla, unless further conditions are met (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+Finally, bias invariance at the level of ratios among species does not extend to ratios between additive aggregates of species, such as between bacterial phyla, unless further conditions are met (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+Multiplicative aggregates of species provides a bias-invariant alternative that has proven useful in biomarker discovery but are harder to interpret.
 
@@
 Moreover, the bias that is measured from controls may differ from that in the primary samples due to differences in cell state or sample preparation.
-These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-density measurement that was described in Section \ref{differential-abundance}.
+These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-abundance measurement that was described in Section \ref{differential-abundance}.
 This \emph{reference-species calibration} allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from taxonomically biased, relative metagenomics measurements.
-
-Recall that densities estimated by the reference-species approach have a constant fold error, provided that the fold error in the reference species is also constant (Equation \eqref{eq:density-ratio-error}).
-Three categories of reference species are reviewed for this purpose in Appendix \ref{review-absolute-methods}, including why they might be expected to produce constant fold errors: housekeeping species assumed to have constant abundance, spike-in species added at known or constant abundance, and species whose abundance has been measured with a targeted method like species-specific qPCR.
 
@@
 Doing so improved the resulting measurements of fold changes in density for all species (Figure \ref{fig:calibration-example}).
+
+Two categories of reference species have so far been used: Housekeeping species assumed to have constant abundance and spike-in species added at known (and typically constant) abundance.
+The implications of bias for these methods hadn't been previously investigated.
+Our results show that, provided that taxonomic bias among all species (including the housekeeping or spike-in species) is consistent across samples, then the fold error in species abundances is constant and does not impact DA analysis of log- or rank-transformed species abundances.
+These methods have downsides: Housekeeping species are not always available/known, and spike-ins have the downside of needing to determine species not present in the samples and an appropriate concentration in which to add them so that they form a moderate fraction of reads across samples with potentially widely varying produce of mean efficiency and total abundance.
+
+Section \ref{abundance-measurement} presents an additional type of reference species: naturally-occurring species whose abundance has been directly measured.
+For most natural ecosystems, there is unlikely to be a single species or genus that is found in all samples.
+Sample coverage can be increased by measuring multiple species with complementary presence-absence patterns (Appendix {[}TODO{]}).
+Coverage can also be increased by using higher-order taxa as references---for example, performing qPCR measurements targeted to a particular family rather than a particular species.
+Higher-order taxa necessarily have higher prevalence than species, though may be more prone to errors from a varying efficiency across samples.
+Targeted measurements of taxa that only span a subset of samples can still provide useful validation for methods.
+
+Housekeeping taxa and spike-ins have also been used to measure changes in total-community abundance.
+Our results show that the bias invariance does not apply for this purpose---the measured fold change in total abundance have an error equal to the inverse change in the sample mean efficiency.
+A similar problem arises when attempting to infer changes in abundance of any higher-order taxon, unless measurement efficiency is conserved among its species.
 
@@
 Even an ideal qPCR protocol that perfectly amplifies all species remains a biased measurement of cell density, since the total 16S copies in the extracted DNA contributed by each species will be proportional to its (species-specific) lysis efficiency and 16S copy-number.
+Empirical evidence supports this, as qPCR measurements of total 16S copies are rarely equivalent to total cell counts, even when multiple copy numbers are taken into account ()Lloyd et al.
+(\protect\hyperlink{ref-lloyd2013meta}{2013})).
 16S qPCR is commonly paired with 16S amplicon sequencing, with which these sources of bias are shared, perhaps along with variation in primer binding and amplification efficiency.
-Methods like using flow cytometry that directly measures cell density lacks these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
+Methods like flow cytometry that directly measure cell density lack these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
 By extending the analysis of Section \ref{differential-abundance} to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \ref{total-density-bias}).
@@
 
+\hypertarget{todo-summary}{%
+\subsection{{[}TODO{]} Summary}\label{todo-summary}}
+
 \hypertarget{conclusion}{%
@@
 
-\emph{This section is a placeholder for a short (2-3 paragraph) conclusion. We are planning to keep `discussion' to the earlier sections and appendix, at least for initial submission.}
+It is commonly thought that analyses of the differences between samples are robust to the taxonomic bias inherent in MGS measurement as long as all samples have been subjected to the same MGS workflow.
+But our results show that consistent taxonomic bias is still capable of creating scientific errors in common forms of DA analysis.
+In particular, we showed mathematically that bias affects DA analysis methods based on proportions or in absolute abundances derived from them (using total-abundance normalization), due to variation in the mean efficiency across samples of varying taxonomic composition.
+The error, however, varies by experimental context and may often be negligible in a practical sense.
+Moreover, it can be quantified and corrected through the use of reference-taxon measurements and/or community calibration controls.
+In addition, bias-aware sensitivity and meta-analyses make it possible to rigorously account for bias even in the absence of such controls.
+Applications of these methods will gradually improve our understanding of the experimental contexts in which variation in the mean efficiency is likely to be problematic and should be accounted for.
+
+We also showed that other DA methods are more robust---perhaps even entirely invariant---to consistent taxonomic bias.
+In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
+In addition, careful pairing of total-abundance and MGS measurement methods may remove most of the bias-driven error in absolute abundances from total-abundance normalization.
+
+Important open questions and future research directions remain.
+Key open questions include 1) determining the extent to which bias is consistent across samples at the species level for a given MGS method and sample type; 2) assessing the validity of post-extraction absolute-abundance controls as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how our assumptions about underlying community dynamics, and in particular what information zero counts provide, affect bias sensitivity.
+In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, better statistical tools are needed that are capable of including taxonomic bias and uncertainty in the MGS and supplemental non-MGS measurements.
+Finally, more concrete experimental recommendations and user-friendly statistical workflows are needed for implementing the solutions we propose in various experimental contexts.
+
+Our theoretical framework and example analyses provide a foundation for addressing these open questions and developing a practical experimental protocols and statistical tools.
+We look to a future in which microbiome researchers routinely choose an appropriate combination of experimental and data-analytic methods capable of answering their fundamental question while accounting for taxonomic bias and the other inherent limitations of MGS measurements---and in this way, gaining the confidence needed to codify statistical findings into true scientific knowledge.
 
@@
 \hypertarget{appendix-regression}{%
-\section{Error in estimated regression coefficients and standard errors}\label{appendix-regression}}
-
-Results from statistics on regression under measurement error can help us understand the effect of bias on DA regression analysis.
-
-Consider Equation \eqref{eq:regression} for the simple linear regression of log density of a species \(i\) on covariate \(x\).
-Let \(y\) stand for the actual log density of a focal species \(i\), \(z\) stand for the log density that we've measured, and \(d = z - y\) equal the difference between the two, which here is the log efficiency of species \(i\) minus the log mean efficiency of the sample.
-Let \(s_{xy}\) denote the sample covariance between variables \(x\) and \(y\), \(s^{2}_{x} = s_{xx}\) and \(s_{x}\) denote the sample variance and standard deviation, and \(r_{xy} = s_{xy}/(s_{x}s_{y})\) denote the sample correlation.
-
-The ordinary least squares (OLS) and maximum likelihood (MLE) estimates of the slope of \(z\) equals the sample covariance of \(z\) and \(x\) divided by the sample variance in \(x\) or, equivalently, the sample correlation of \(z\) and \(x\) multiplied by the ratio of their sample standard deviations,
-\begin{align}
-  \hat \beta_z = \frac{s_{zx}}{s^2_x} = r_{zx} \cdot \frac{s_z}{s_x}.
-\end{align}
-From the (bi)linearity of sample covariances it follows that
-\begin{align}
-  \hat \beta_z 
-  = \frac{s_{yx}}{s^2_x} + \frac{s_{dx}}{s^2_x} 
-  = \frac{r_{yx} s_y}{s_x} + \frac{r_{dx} s_d}{s_x} 
-  = \hat \beta_y + \hat \beta_d,
-\end{align}
-where \(\hat \beta_y\) and \(\hat \beta_d\) denote the slope estimates for \(y\) and \(d\) (were these values to be known).
-The absolute error in the estimate \(\hat \beta_{z}\) of \(\hat \beta_{y}\) is therefore \(\hat \beta_{d}\); it is large in a practical sense when \(\hat \beta_{d}\) is large (in absolute value) compared to \(\hat \beta_{y}\), which corresponds to the covariance of \(d\) with \(x\) being large compared to the covariance of \(y\) with \(x\).
-
-In our case, the covariance of \(d\) equals the negative of the covariance of log mean efficiency with \(x\).
-The absolute error in \(\hat \beta\) equals the negative covariance of log mean efficiency scaled by the variance in \(x\).
-This absolute error is the same for all species; however, its practical significance varies depending on its magnitude relative to that of the slope of the actual log densities.
-For species that covary with \(x\) more strongly than the log mean efficiency, the error will be relatively small.
-This situation might occur either because the mean efficiency varies relatively little across samples or because its variation is relatively less correlated with \(x\) compared to the log density of the focal species.
-
-We can similarly understand the impact of measurement error on the precision of our slope estimates.
-The OLS and MLE estimated standard error in \(\hat \beta\) are both approximately
-\begin{align}
+\section{Taxonomic bias and linear regression}\label{appendix-regression}}
+
+This section derives the theoretical results of Section \ref{differential-abundance} for the effect of taxonomic bias on regression-based DA analyses.
+As in Section \ref{differential-abundance}, we restrict our attention to DA analyses that can be expressed in terms of the simple linear regression model in which the response is log absolute abundance or log proportion of individual species or the log ratio of a pair of species.
+
+\hypertarget{review-of-simple-linear-regression}{%
+\subsection{Review of simple linear regression}\label{review-of-simple-linear-regression}}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.23}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.38}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.38}}@{}}
+\caption{\label{tab:statistical-notation} Statistical notation used in this section.}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\raggedright
+Notation
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Mathematical definition
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Description
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\raggedright
+Notation
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Mathematical definition
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Description
+\end{minipage} \\
+\midrule
+\endhead
+\(x\), \(y\), \(z\), \(d\) & & Random variables, defined in text \\
+\(x(a)\) & & Value of \(x\) in sample \(a\) \\
+\(\bar x\) & \(\frac{1}{n}\sum_a x(a)\) & Sample mean of \(x\) \\
+\(s(x,y)\) & \(\frac{1}{n}\sum_a (x(a) - \bar x) (y(a) - \bar y)\) & Sample covariance between \(x\) and \(y\) \\
+\(s^{2}(x)\) & \(s(x,x)\) & Sample variance of \(x\) \\
+\(s(x)\) & \(\sqrt{s^{2}(x)}\) & Sample standard deviation of \(x\) \\
+\(r(x,y)\) & \(\frac{s(x,y)}{s(x) s(y)}\) & Sample correlation between \(x\) and \(y\) \\
+\(\alpha\), \(\beta\), \(\sigma\) & & Parameters of the regression model \\
+\(\hat \alpha\), \(\hat \beta\), \(\hat \sigma\) & & Parameter estimates \\
+\(\epsilon(a)\) & \(y(a) - (\alpha + \beta x(a))\) & Error for sample \(a\) \\
+\(\hat \epsilon(a)\) & \(y(a) - (\hat \alpha + \hat \beta x(a))\) & Residual for sample \(a\) \\
+\(\text{se}(\hat \beta)\), \(\hat{\text{se}}(\hat \beta)\) & & Standard error of \(\hat \beta\) and its estimate \\
+\bottomrule
+\end{longtable}
+
+We begin by reviewing definitions, notation, and results for the analysis of the simple linear regression model.
+Our presentation follows Wasserman (\protect\hyperlink{ref-wasserman2004allo}{2004}) Chapter 13 with additional interpretation of the estimated regression coefficients.
+Statistical notation is defined in Table \ref{tab:statistical-notation}.
+
+Consider a response variable \(y\) and a covariate \(x\).
+Following Wasserman (\protect\hyperlink{ref-wasserman2004allo}{2004}), we define the \emph{simple linear regression model} by
+\begin{align}
+  \label{eq:lm}
+  y(a) &= \alpha + \beta x(a) + \varepsilon(a),
+\end{align}
+where \(E[\varepsilon(a) \mid x(a)] = 0\) and \(V[\varepsilon(a) \mid x(a)] = \sigma^2\).
+That is, conditional on the value of \(x\) in a sample \(a\), the response \(y\) is given by \(\alpha + \beta x\) with a random error \(\varepsilon\) that has an expected value of \(0\) and a constant variance \(\sigma^2\).
+
+Given data \((y, x)\), we \emph{fit} the model by finding estimates for the parameters \(\alpha\), \(\beta\), and \(\sigma^2\) that best reflect the data under the assumption that the model is correct and according to our chosen criterion that defines `best'.
+Here we consider the least-squares estimates for the coefficients \(\alpha\) and \(\beta\), which are also the maximum likelihood estimates if we further assume that the errors \(\varepsilon\) are normally distributed.
+The least-squares estimates of \(\alpha\) and \(\beta\) are the values that minimize the residual sum of squares, \(\text{RSS} = \sum_{a} \hat \varepsilon(a)\).
+These estimates can be conveniently expressed in terms of the sample means and covariances or correlations of \(x\) and \(y\),
+\begin{align}
+  \label{eq:lm-hat}
+  \hat \alpha &= \bar y - \hat \beta \bar x \\
+  \hat \beta &= \frac{s(y,x)}{s^2(x)} = r(y,x) \cdot \frac{s(y)}{s(x)}.
+\end{align}
+The estimate for the slope coefficient, \(\hat \beta\), equals the ratio of the sample covariance of \(y\) with \(x\) relative to the variance in \(x\) or, equivalently, the correlation of \(y\) with \(x\) multiplied by the ratio of their standard deviations.
+The estimate for the intercept, \(\hat \alpha\), is the value such that the sample means follow the regression relationship, \(\bar y = \hat \alpha + \hat \beta \bar x\).
+The standard unbiased and maximum likelihood estimates for the variance \(\sigma^{2}\) are both approximately given by the sample variance of the residuals,
+\begin{align}
+  \label{eq:lm-var}
+  \hat{\sigma}^2(\hat \beta) \approx s^2(\hat \varepsilon).
+\end{align}
+
+We are most interested in the estimated slope coefficient, \(\hat \beta\).
+Besides the point estimate indicating our best guess at the value, we also wish to understand the uncertainty or precision of the estimate.
+This uncertainty is quantified by the standard error, \(\text{se}(\hat \beta)\), estimated by
+\begin{align}
+  \label{eq:lm-se}
   \hat{\text{se}}(\hat \beta)
-  \approx \frac{s_{\hat \varepsilon}}{s_x \sqrt{n}},
-\end{align}
-where \(s_{\hat \varepsilon}\) is the sample standard deviation of the residuals
-(Wasserman (\protect\hyperlink{ref-wasserman2004allo}{2004}) Chapter 13).
-The sample residuals of \(z\), \(y\), and \(d\) have a similar relationship to the regression coefficient estimates,
-\begin{align}
-  \hat \varepsilon_z 
-    &\equiv z - \hat \beta_z x
-  \\&= (y + d) - (\hat \beta_y + \hat \beta_d) x
-  \\&= (y - \hat \beta_y x) + (d - \hat \beta_d x)
-  \\&= \hat \varepsilon_y + \hat \varepsilon_d.
-\end{align}
-(Note, here I've omitted the subscript indicating the dependence on the sample.)
-It follows that the sample variances of the residuals of \(z\), \(y\), and \(d\) are related through
-\begin{align}
-  s^2_{\hat \varepsilon_{z}} 
-  = s^2_{\hat \varepsilon_{y} + \hat \varepsilon_{d}}
-  = s^2_{\hat \varepsilon_{y}} + s^2_{\hat \varepsilon_{d}} + 2 s_{\hat \varepsilon_{y} \hat \varepsilon_{d}}.
-\end{align}
-The standard deviation of the \(z\) residuals is increased above that of the \(y\) residuals when the \(d\) residuals are either uncorrelated or positively correlated with the \(y\) residuals, but may be decreased when the \(y\) and \(d\) residuals are negatively correlated.
-
+  = \frac{\hat \sigma}{s(x) \sqrt{n}}
+  \approx \frac{s(\hat \varepsilon)}{s(x) \sqrt{n}},
+\end{align}
+Approximate 95\% confidence intervals for \(\beta\) are given by \(\hat \beta \pm 2\; \hat{\text{se}}(\hat \beta)\).
+
+\hypertarget{measurement-error-in-the-response}{%
+\subsection{Measurement error in the response}\label{measurement-error-in-the-response}}
+
+Now consider a researcher who would like to regress a biological quantity \(y\) (the response) on a second quantity \(x\) (the covariate).
+But instead of \(y\), they only know a proxy \(z\), which is related to \(y\) via the difference \(d = z - y\).
+For instance, \(y\) might be the log abundance of a species, \(z\) is the abundance that is measured via MGS, and \(x\) is a numerical quantity like pH or a boolean variable indicting case versus control.
+Lacking a direct measurement of \(y\), the researcher instead regresses \(z\) on \(x\) and interprets the fitted model as informing them about the relationship between \(y\) and \(x\).
+We wish to understand the accuracy of their conclusions.
+
+To understand how the measurement error in \(y\) impacts the researcher's regression analysis, consider the three linear regression equations for \(y\), \(z\), and \(d\),
+\begin{align}
+  \label{eq:lm-y}
+  y(a) &= \alpha_y + \beta_y x(a) + \varepsilon_y(a),
+\end{align}
+\begin{align}
+  \label{eq:lm-z}
+  z(a) &= \alpha_z + \beta_z x(a) + \varepsilon_z(a),
+\end{align}
+\begin{align}
+  \label{eq:lm-d}
+  d(a) &= \alpha_d + \beta_d x(a) + \varepsilon_d(a).
+\end{align}
+The researcher would like to fit the model for \(y\) (Equation \eqref{eq:lm-y}), but instead can only fit the model for \(z\) (Equation \eqref{eq:lm-z}).
+The two are related via the Equation \eqref{eq:lm-z} for \(d\).
+It is helpful to imagine, for the moment, that \(y\) and \(d\) are known and so we can fit all three models \eqref{eq:lm-y}, \eqref{eq:lm-z}, and \eqref{eq:lm-d}.
+From Equation \eqref{eq:lm-hat} and the linearity of sample means and covariances it follows that
+\begin{align}
+  \label{eq:lm-hat-rel}
+  \hat \alpha_z &= \hat \alpha_y + \hat \alpha_d \\
+  \hat \beta_z &= \hat \beta_y + \hat \beta_d.
+\end{align}
+That is, the estimated coefficients for \(y\), \(z\), and \(d\) mirror the relationship between the variables themselves, \(z = y + d\).
+Consequently, we see that the measurement error \(d\) creates absolute errors in the regression coefficients \(\hat \alpha\) and \(\hat \beta\) equal to
+\begin{align}
+  \label{eq:lm-hat-err}
+  \text{abs. error}(\hat \alpha) &\equiv \hat \alpha_z - \hat \alpha_y = \hat \alpha_d \\
+  \text{abs. error}(\hat \beta) &\equiv \hat \beta_z - \hat \beta_y = \hat \beta_d.
+\end{align}
+These absolute errors correspond to the \emph{statistical bias} in the coefficient estimates.
+Expressed in terms of sample means and covariances, these errors are
+\begin{align}
+  \label{eq:lm-hat-err-1}
+  \text{abs. error}(\hat \alpha) &= \bar d - \hat \beta_d \bar x \\
+  \text{abs. error}(\hat \beta)  &= \frac{s(d,x)}{s^2(x)} = r(d,x) \cdot \frac{s(d)}{s(x)}.
+\end{align}
+
+We are mainly interested in the estimated slope coefficient, \(\hat \beta\).
+We see that the absolute error is large when the covariance of \(d\) with \(x\) is large.
+This error is large in a practical sense when \(\hat \beta_{d}\) is large (in magnitude) relative to \(\hat \beta_y\) (Equation \eqref{eq:lm-hat}), which occurs when \(s(d,x)\) is large relative to \(s(y,x)\).
+A sign error occurs when \(s(d,x)\) is larger in magnitude than \(s(y,x)\) and of opposite sign.
+
+We can also consider how the measurement error affects the residual variation and the standard error of the slope estimate.
+The residuals of \(y\), \(z\), and \(d\) are related through \(\hat \varepsilon_z = \hat \varepsilon_y + \hat \varepsilon_d\).
+It follows that the sample variance of the \(z\) residuals equal
+\begin{align}
+  s^2(\hat \varepsilon_{z}) 
+  = s^2(\hat \varepsilon_y + \hat \varepsilon_{d})
+  = s^2(\hat \varepsilon_y) + s^2(\hat \varepsilon_{d}) + 2 s(\hat \varepsilon_y, \hat \varepsilon_{d}).
+\end{align}
+The variance of the \(z\) residuals is increased above that of the \(y\) residuals when the \(d\) residuals are either uncorrelated or positively correlated with the \(y\) residuals, but may be decreased when the \(y\) and \(d\) residuals are negatively correlated.
+An increased residual variance will lead to a larger estimate for \(\sigma\) as well as larger standard errors in the slope coefficient \(\beta\).
+
+\hypertarget{specific-application-to-taxonomic-bias}{%
+\subsection{Specific application to taxonomic bias}\label{specific-application-to-taxonomic-bias}}
+
+These general results can be used to understand how taxonomic bias affects DA analyses that can be expressed as linear regression of log (relative or absolute) abundance.
+
+First, we illustrate for the particular example of absolute DA analysis, where species absolute abundance is measured with the total-abundance normalization method and we assume that the total abundances have been accurately measured.
+Consider the regression of log abundance of species \(i\) on a covariate \(x\).
+In this case, \(y(a)= \log \text{abun}_{i}(a)\) is the actual log abundance of species \(i\) and \(z(a)=\log \widehat{\text{abun}}_i(a)\) is the log abundance we've measured.
+From Equation \eqref{eq:density-prop-error}, the measurement error \(d(a)\) due to taxonomic bias in the MGS measurement is
+\begin{align}
+  d(a) 
+  \equiv z(a) - y(a) 
+  = \log \text{efficiency}_i - \log \text{efficiency}_S(a).
+\end{align}
+The absolute error in \(\hat \beta\) equals the scaled covariance of \(d\) with \(x\) (Equation \eqref{eq:lm-hat-err-1}).
+In the above expression for \(d(a)\), only the second term, \(- \log \text{efficiency}_S(a)\), varies across samples and thus affects the covariance, leading to
+\begin{align}
+  \text{abs. error}(\hat \beta)
+    &= - \frac{s[\log \text{efficiency}_S, x]}{s^2(x)}.
+%    = - \frac{r[\log \text{efficiency}_S, x] \cdot s[\log \text{efficiency}_S]}{s(x)}, \\
+\end{align}
+Thus the absolute error in \(\hat \beta\) is the negative of the scaled covariance of the log mean efficiency.
+
+Although the absolute error is the same for each species, its practical significance varies.
+Recall from Equation \eqref{eq:lm-hat} that the correct value for \(\hat \beta\) (i.e., the estimate without measurement error) is
+\begin{align}
+  \hat \beta &= \frac{s[\log \text{abun}, x]}{s^2(x)}.
+\end{align}
+For species whose log abundance covaries with \(x\) much more than the log mean efficiency, the error will be relatively small.
+This situation can occur either because the mean efficiency varies relatively little across samples or because its variation is relatively uncorrelated with \(x\).
+For species whose log abundance covaries with \(x\) similar or less than the log mean efficiency, the error will be significant.
+A sign error occurs when the species covariance is of \emph{equal sign and smaller in magnitude} than the covariance of the log mean efficiency.
+
+We can also consider how the standard errors in the slope estimate are affected by bias.
 In our case, the \(d\) residuals equal the negative residuals of the log mean efficiency.
@@
 However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \ref{fig:regression-example}.
+
+Now, suppose we were instead performing a DA analysis of a response whose fold error is constant across samples, such as the log ratio of two species.
+For the log ratio of species \(i\) and \(j\), the error is \(d(a) = \log \text{efficiency}_i / \text{efficiency}_j\) and is constant across samples.
+Thus the covariance of \(d\) with \(x\) is 0 and taxonomic bias causes no error the slope estimate \(\hat \beta\), only the intercept estimate \(\hat \alpha\).
+
+\hypertarget{theory-notes}{%
+\section{Theory notes}\label{theory-notes}}
+
+\hypertarget{effective-experimental-effort-intuition}{%
+\subsection{Effective experimental effort: Intuition}\label{effective-experimental-effort-intuition}}
+
+The effective experimental effort per cell for sample \(a\) is
+\begin{align}
+  \label{eq:effort}
+  \text{effort}_S(a)
+    &= \frac{\text{reads}_S(a)}{\text{abun}_S(a) \cdot \text{efficiency}_S(a)} \\
+    &= \frac{\sum_{i \in S}\text{reads}_i(a)}{\sum_{i \in S} \text{abun}_i(a) \cdot \text{efficiency}_i(a)};
+\end{align}
+the second form follows from the definition of the mean efficiency.
+The effort increases if more sequencing performed on the sample, leading to greater number of reads for all species; and decreases if a larger sample input volume or concentration is used, leading to fewer reads per cell.
+Less intuitively, it also decreases with the mean efficiency of the sample.
+To help interpret the effort light of this last behavior, we consider how the effort varies across a few illustrative scenarios.
+
+\textbf{Scenario 1:}
+Consider Sample 1 in Figure \ref{fig:error-proportions}.
+Imagine that sequencing reads can be perfectly assigned to the three species; the taxonomic bias is entirely due to steps prior to read assignment, such as DNA extraction and PCR.
+Supposing that the sample was sequenced to a read depth of 100, the three species have read counts of \((4, 72, 24)\).
+In addition, suppose the total cell concentration in the sample equals 1 in some particular units.
+Since the mean efficiency is \((1 + 18 + 6) / 3 = 8.\bar 3\), the effort is \(100 / (1 \cdot 8.\bar 3) = 12\).
+
+\textbf{Scenario 2:}
+Now consider a situation identical to Scenario 1, except that the bioinformatics pipeline is unable to assign reads from Species 3.
+In this case, the \emph{assigned} read counts will be \((4, 72, 0)\), for a total of \(76\), though the total number of sequenced reads remains 100.
+Meanwhile, the efficiency of species 3 drops to 0 and the mean efficiency of the sample drops to \((1 + 18 + 0) / 3 = 6.\bar 3\).
+The effort is therefore \(76 / (1 \cdot 6.\bar 3) = 12\), the same as in Scenario 1.
+
+\textbf{Scenario 3:}
+Now consider a situation identical to Scenario 1, except that Species 3 fails to amplify during the PCR step in a marker-gene sequencing experiment.
+But thanks to careful experimental normalization during library preparation, an equal amount of DNA is prepared in the library as in Scenario 1, leading to 100 sequenced reads from the sample.
+The measured relative abundances are identical to in Scenario 2, but the read counts are larger, \((\approx 5.3, \approx 94.7)\).
+(As in Section \ref{abundance-measurement}, we ignore the fact that read counts are discrete.)
+The mean efficiency is also the same (\(6.\bar 3\)) as in Scenario 2.
+The effort is therefore \(100 / (1 \cdot 6.\bar 3) \approx 15.8\), larger than in Scenario 1 and Scenario 2.
+
+\textbf{Scenario 4:}
+Now consider a situation identical to Scenario 1, but supposing Species 3 to be a reference species, such as the host or a spike-in.
+Does the effort depend on whether we include the reference species in the species set \(S\)?
+As we saw in Scenario 1, the effort associated with all three species is 12.
+For the two non-reference species alone, the total abundance is \(2/3\), the total read count is \(76\), and the mean efficiency is \((1 + 18) / 2 = 9.5\).
+The effort is therefore \(76 /(\tfrac{2}{3} \cdot 9.5) = 12\), the same as in Scenario 1 and Scenario 2.
+
+TODO:
+
+\begin{itemize}
+\tightlist
+\item
+  Make sure this result is general, and not just happening when the three species are all equally abundant.
+\item
+  consider moving secenario 4 to 2 or 3, leaving the scenario where effort changes to last.
+\item
+  note that for proving the general results, it may be best to go back to main model equation, where we can see that if the read count, abundance, and efficiency for at least one species is unchanged, than the effort is unchanged.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    in scenario 3, the effort must go up since taxa 1 and 2 are unchanged but get more reads.
+  \item
+    if we add or remove species, without changing the read count of other species, then the effort must stay the same.
+  \item
+    this fact that effort is independent of the taxonomic scope S seems useful and convenient and intuitive.
+  \end{itemize}
+\end{itemize}
+
+\textbf{Summary:}
+The behavior of the effective experimental effort across these scenarios makes intuitive sense if we think of it as specifically describing the `wet lab' components of the MGS measurement, up to and including sequencing but excluding bioinformatic analysis of the sequencing data.
+The taxonomic scope \(S\) is an arbitrary decision made implicitly or explicitly by the researcher, which does not change the wet portion of the experiment; thus \(\text{effort}(a)\) is independent of \(S\).
+In addition, sequencing reads are routinely excluded during bioinformatic analysis, and such filtering does not change \(\text{effort}(a)\), instead reducing the read counts by reducing the species' efficiencies.
+On the other hand, if species drop out prior to sequencing due to low extraction efficiency and/or low PCR efficiency, then achieving the same total read count requires that a greater sample volume or concentration be passed on to subsequent steps.
+These adjustments are reflected by a commensurate increase in \(\text{effort}(a)\).
+
+\hypertarget{absolute-abundance-measurement}{%
+\subsection{Absolute-abundance measurement}\label{absolute-abundance-measurement}}
+
+Q: How to motivate this `general' form?
+It comes from ignoring bias.
+
+Make clear that R doesn't need to be in S. Perhaps want different sets - a superset of all species, and an set of species that each measurement method considers?
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+Consider this general form for measuring of the absolute abundance of a taxon \(I\) from a reference species \(R\),
+\begin{align}
+  \widehat{\text{abun}}_{I}(a) 
+  &= \text{reads}_{I}(a) \cdot \frac{\widehat{\text{abun}}_{R}(a)}{\text{reads}_{R}(a)}.
+\end{align}
+When \(R = S\), this measurement amounts to total-community normalization; when \(R=r\), it amounts to reference-species normalization.
+(This form does not encompass using the geometric average of multiple species.)
+Under our model, we can rewrite the right-hand-side to obtain
+\begin{align}
+  \widehat{\text{abun}}_{I}(a) 
+  &= \text{abun}_{I}(a) \cdot \frac{\text{efficiency}_{I}(a)}{\text{efficiency}_{R}(a)}
+  \cdot \text{FE}\left[\widehat{\text{abun}_R}(a)\right],
+\end{align}
+where \(\text{FE}\left[\widehat{\text{abun}_R}(a)\right] = \widehat{\text{abun}_R}(a) / \text{abun}_R(a)\) is the fold error in the reference abundance measurement.
+
+From here, we can derive the other results.
+
+Methods differ based on the choice of \(R\) and in how the abundance of \(R\) is measured.
 
@@
 \centering
-\includegraphics{figures/export-pdf/notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity_files/figure-html5/regression-lm-alt-1.pdf}
-\caption{\label{fig:momspi-regression}\textbf{Bias distorts log fold changes (LFCs) in species proportions in a regression analysis of vaginal microbiome samples from the MOMS-PI study.} Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
-The LFC in proportion from low- to high-diversity samples was estimated for 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles following a simple zero-replacement procedure.
-Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95\% confidence intervals for each species.
-The difference between the calibrated and observed estimate for each species equals the negative LFC in mean efficiency.}
+\includegraphics{figures/export-pdf/notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity_files/figure-html5/regression-gp-1.pdf}
+\caption{\label{fig:momspi-regression}\textbf{Bias inflates estimated log fold changes (LFCs) in species proportions with diversity in vaginal microbiome samples from the MOMS-PI study.} Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
+The LFC in proportion from low- to high-diversity samples was estimated for 30 common species using gamma-Poisson regression with and without bias correction.
+Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95\% Bayesian credible intervals for each species.}
 \end{figure}
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-barlow2020aqau}{}}%
+Barlow, Jacob T., Said R.
+Bogatyrev, and Rustem F.
+Ismagilov.
+2020.
+{``{A quantitative sequencing framework for absolute abundance measurements of mucosal and lumenal microbial communities}.''} \emph{Nat.
+Commun.} 11 (1): 1--13.
+\url{https://doi.org/10.1038/s41467-020-16224-6}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-brooks2015thet}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-huttenhower2012stru}{}}%
+Huttenhower, Curtis, Dirk Gevers, Rob Knight, Sahar Abubucker, Jonathan H.
+Badger, Asif T.
+Chinwalla, Heather H.
+Creasy, et al.
+2012.
+{``{Structure, function and diversity of the healthy human microbiome}.''} \emph{Nature} 486 (7402): 207--14.
+\url{https://doi.org/10.1038/nature11234}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-ji2019quan}{}}%
@@
 \url{https://doi.org/10.1128/AEM.00877-20}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lloyd2013meta}{}}%
+Lloyd, Karen G., Megan K.
+May, Richard T.
+Kevorkian, and Andrew D.
+Steen.
+2013.
+{``{Meta-analysis of quantification methods shows that archaea and bacteria have similar abundances in the subseafloor}.''} \emph{Appl.
+Environ.
+Microbiol.} 79 (24): 7790--99.
+\url{https://doi.org/10.1128/AEM.02090-13}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-vieirasilva2019quan}{}}%
+Vieira-Silva, Sara, João Sabino, Mireia Valles-Colomer, Gwen Falony, Gunter Kathagen, Clara Caenepeel, Isabelle Cleynen, Schalk van der Merwe, Séverine Vermeire, and Jeroen Raes.
+2019.
+{``{Quantitative microbiome profiling disentangles inflammation- and bile duct obstruction-associated microbiota alterations across PSC/IBD diagnoses}.''} \emph{Nat.
+Microbiol.} 4 (11): 1826--31.
+\url{https://doi.org/10.1038/s41564-019-0483-9}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wallace2021thed}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-zhang2017soil}{}}%
+Zhang, Zhaojing, Yuanyuan Qu, Shuzhen Li, Kai Feng, Shang Wang, Weiwei Cai, Yuting Liang, et al.
+2017.
+{``{Soil bacterial quantification approaches coupling with relative abundances reflecting the changes of taxa}.''} \emph{Sci.
+Rep.} 7 (1): 1--11.
+\url{https://doi.org/10.1038/s41598-017-05260-w}.
+
 \end{CSLReferences}

### 2022-02-17 case-studies.Rmd
Revise vaginal-microbiome case study
Revised based on BJC's suggested revisions from 2022-02-15.

@@
 
-The vaginal microbiome during pregnancy has been a source of intensive study due to its apparent connection with the health of both mother and her developing child.
-Many MGS studies have found associations of specific microbial species and community characteristics with rates of urinary tract and sexually-transmitted infections, bacterial vaginosis (BV), and preterm birth.
-Yet these associations vary across studies of different populations and using different MGS methods.
+A growing number of MGS studies have used DA analysis to describe associations between members of the vaginal microbiome and health conditions including urinary tract infections, sexually-transmitted infections, bacterial vaginosis, and preterm birth.
+However, these associations often vary between studies.
 DA analyses of the vaginal microbiome are commonly based on proportions, creating an opportunity for taxonomic bias to impact results.
-A number of studies have experimentally demonstrated substantial taxonomic bias among MGS protocols and individual steps (such as extraction and PCR amplification) in vaginal samples or _in vitro_ samples of vaginally-associated species (@yuan2012eval,@brooks2015thet,@gill2016eval,@graspeuntner2018sele).
-<!-- TODO: perhaps add ref on taxonomic resolution and also sampling methods -->
-Bias has been proposed as a potential explanation for discrepancies across studies (@callahan2017repl, Others?), but there has so far been little quantitative analysis of this possibility.
-Here we use empirical bias measurements from control samples to investigate the role of bias in proportion-based DA analyses of vaginal microbiomes from a recent large-cohort study of pregnant women.
+Substantial taxonomic bias has been experimentally demonstrated in MGS protocols applied to vaginal samples or in vitro samples of vaginally-associated species (@yuan2012eval, @brooks2015thet, @gill2016eval, @graspeuntner2018sele).
+The different biases of different MGS protocols have been proposed as one explanation for discrepancies in DA results across studies (@callahan2017repl).
 
-As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, @fettweis2019thev collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were measured by amplicon sequencing of the 16S V1-V3 hypervariable region to yield species-level bacterial taxonomic profiles.
-Taxonomic bias of this MGS protocol was previously investigated by @brooks2015thet and @mclaren2019cons, using measurements by @brooks2015thet of cellular mock communities of seven common, clinically-relevant vaginal bacterial species.
-<!-- TODO: check on clinical relevance -->
+As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, @fettweis2019thev collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were taxonomically profiled by amplicon sequencing of the 16S V1-V3 hypervariable region.
+The taxonomic bias of the MOMS-PI MGS protocol was previously quantified by @mclaren2019cons using control measurements by @brooks2015thet of cellular mock communities of seven common, clinically-relevant vaginal bacterial species.
 Of these, _Lactobacillus iners_ had the highest efficiency, which was nearly 30X larger than that of the species with the lowest efficiency, _Gardnerella vaginalis_.
 A second _Lactobacillus_ species, _L. crispatus_, had an efficiency that was approximately 2X less than _L. iners_ and 15X greater than _G. vaginalis_.
-These species, along with the unculturable _Lachnospiraceae BVAB1_, are the most common top (most-abundant) species (SI analysis) and can reach high proportions in individual samples, indicating that shifts between them might drive large changes in the mean efficiency which might in turn impact DA results.
+These species, along with the unculturable _Lachnospiraceae BVAB1_, are most frequently the most abundant species in a sample and commonly reach high proportions of over 70% (TODO SI analysis).
+Therefore, shifts between these dominant species could drive large changes in the sample mean efficiency, which might in turn distort DA analyses.
 
-We sought to assess this possibility using a joint analysis of the control measurements from @brooks2015thet and the microbiome measurements from the MOMS-PI study.
-To obtain the taxonomic bias among all species identified in the MOMS-PI measurements, we used the taxonomic relationships with the seven control species to impute the efficiencies for the remaining species.
-We used these imputed efficiencies to calibrate (correct the effect of bias in) the MOMS-PI measurements, examine variation in the mean efficiency varied across samples, and compare the results of a DA analysis with and without bias correction.
-
-<!-- Such imputed efficiencies carry considerable uncertainty, which we ignore to simplify our presentation.  -->
-<!-- Notably, while this uncertainty can majorly impact abundances in individual samples, it has a much weaker impact on LFCs across samples (SI analysis TODO). -->
-
-The mean efficiency varies substantially across vaginal samples (Figure \@ref(fig:momspi-mean-efficiency-dist).
-This variation appears to be primarily driven by variation in which species is most abundant, samples in which a _Lactobacillus_ species is most abundant (after calibration) typically have an efficiency that is 3-20X greater than samples in which _G.
+We sought to understand the potential impact of bias on DA analysis of vaginal microbiome profiles from the MOMS-PI study.
+To do so, we first examined variation in the mean efficiency across samples under the assumption that bias in the MOMS-PI study was accurately represented by the @brooks2015thet control measurements.
+Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample (TODO SI Analysis).
+The mean efficiency varies substantially across vaginal samples (Figure 4.1), with samples in which a _Lactobacillus_ species is most abundant typically having a mean efficiency that is 3-20X greater than samples in which _G.
 vaginalis_ is most abundant.
-Shifts between _Lactobacillus_-dominance and _Gardnerella_-dominance are common in between-women comparisons and occasionally occur between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
-These shifts typically result in substantial fold changes in mean efficiency (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)) and can cause spurious fold changes in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).
-Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance would be more extreme for vaginal microbiome primer sets that fail to amplify _Gardnerella_ (@graspeuntner2018sele).
+The vaginal microbiome sometimes shifted between _Lactobacillus_-dominance and _Gardnerella_-dominance between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)), which caused magnitude and sign errors in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).
 
@@
 
-We next considered whether the observed variation in mean efficiency could cause systematic error in a DA analysis.
-In particular, we hypothesized that DA analysis of species proportions versus a covariate that is associated with _Lactobacillus_ and/or _Gardnerella_ would be particularly prone to spurious results.
-Patient metadata was not available due to privacy restrictions; we therefore sought a clinically relevant covariate to use in a regression analysis that could be determined directly from the microbiome profiles and that we _a priori_ expected to be associated with the proportions of _Lactobacillus_ and _Gardnerella_.
-Alpha diversity metrics such as species richness, the Shannon index, and the (Inverse) Simpson index have been repeatedly found to be strongly positively associated with bacterial vaginosis (BV; @srinivasan2012bact, @cartwright2018mult).^[@cartwright2018mult found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100% and specificity of 85.1%.]
-In addition, it is commonly observed that samples from women with and without BV that are dominated by _Lactobacillus spp._ tend to have higher diversity, whereas samples dominated by _Gardnerella_ tend to have lower diversity.
-We therefore chose to perform a DA analysis of species proportion versus alpha diversity, hypothesizing that _Lactobacillus_ and _Gardnerella_ would drive a negative association of mean efficiency with diversity and thereby distort DA estimates for all species.
-We split samples into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
-We then estimated the LFC in proportion from low- to high-diversity samples for 30 species passing a basic prevalence threshold, with and without bias correction.
-A large fraction of entries in the species count matrix were small counts or zeros.
-We used therefore used gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions while accounting for uncertainty from the random sampling of reads.
+We hypothesized that proportion-based DA analyses of health condition that is associated with the abundance of common taxa with particularly high or low efficiencies, such as _Lactobacillus_ and/or _Gardnerella_ in the MOMS-PI study, would be particularly prone to spurious results.
+A potential example may exist in the vaginal microbiome: Bacterial vaginosis (BV) has been repeatedly found to be strongly positively associated with higher-diversity communities with increased abundance of _G.
+vaginalis_ and decreased abundance of _Lactobacillus spp._ (@srinivasan2012bact, @cartwright2018mult).
+As a proxy for clinical BV, we split samples from the MOMS-PI study into low diversity (non-BV), intermediate diversity, and high diversity (BV-like) groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
+We estimated the LFC in proportion from non-BV to BV-like samples for 30 prevalent bacterial species, with and without bias correction, using a gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions.
 We accounted for bias by including an offset term in the linear model equal to the log ratio of the species' efficiency to the sample mean efficiency.
 
-The mean efficiency was on average lower in high-diversity samples due to a reduction in _Lactobacillus_ dominance (LFC of -1.6, or a 4.8X reduction). 
-Our results from the previous section therefore suggest that accounting for bias should reduce the estimated LFC for each species by about this same amount.
-We found that, as expected, bias correction reduced the LFC of each species, though the effect was variable and typically somewhat smaller.
-Bias correction led to substantial magnitude changes for several species and sign changes in 3 species; however, the relative difference between the corrected and uncorrected LFCs was modest for the 19/30 species with the largest LFCs of over 2 (corresponding to a FC of 7.4X) even after correction.
-These large LFCs occur in species that rarely if ever reach high proportions and so are likely driven by the sum-to-one constraint: The high diversity communities are those not dominated by _Lactobacillus_ spp.
-(or to a lesser extent _G.
-vaginalis_ and BVAB1).
-Therefore it is not clear how well it will generalize to regressions on other covariates.
+As expected, bias correction reduced the estimated LFC from non-BV to BV-like groups for every species tested, though the importance of this effect varied among species (TODO: median and range of change).
+Bias correction led to substantial magnitude changes for several species and sign changes in 3 species; however, the relative difference between the corrected and uncorrected LFCs was modest for the 19/30 species with the largest changes, which each had a corrected LFC greater than 2 (corresponding to a FC greater than 7.4X).
+Our theoretical results from the previous section suggest that bias correction should reduce the estimated LFC for each species by roughly the amount that the mean efficiency decreases between groups.
+Consistent with this prediction, the mean efficiency was lower in high-diversity BV-like samples (mean LFC of -1.6) due to the lower abundance of the _Lactobacillus spp._
+that are so efficiently measured by the MOMS-PI MGS protocol.
 

### 2022-02-19 solutions.Rmd
Revise solutions section

@@
 
-The theoretical results from Section \@ref(differential-abundance) suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.
+Microbiome studies proffer a wide diversity of goals and constraints, such that there is no once-size-fits all solution to mitigating taxonomic bias in DA analysis.
+Instead, we present six classes of solutions, each supported by our theoretical results from the previous sections, that together encompass the majority of experimental contexts faced by researchers today.
 
-## Use ratio-based relative-abundance analyses
+## Ratio-based relative DA analysis
 
@@
 These methods analyze (log) fold changes in the ratio between species or, more generally, between the products of multiple species.
-Taking the product or geometric mean of multiple species provides a approach to aggregating species into a higher-level taxonomic units such that the ratios between units maintain proportional errors, in contrast to the standard approach of additively aggregating species.
+Taking the product or geometric mean of multiple species provides an approach to aggregating species into a higher-level taxonomic units such that the ratios between units maintain proportional errors, in contrast to the standard approach of additively aggregating species.
 Besides a greater robustness to taxonomic bias, analyses of ratios have other advantages over analyses of proportions and absolute abundances.
@@
 Community calibration controls are samples whose species identities and relative abundances are known either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
-MGS measurement of one or more control communities can be used to directly measure the relative efficiencies among the superset of species in the controls (@mclaren2019cons).
-The measured relative efficiencies can then be used to calibrate (remove the effect of bias from) the relative abundances (ratios) of the control species in a set of primary (non-control) experimental samples that were measured with the same protocol (ideally in the same experiment and sequencing run) (@mclaren2019cons).
+MGS measurement of one or more control communities can be used to directly measure the relative efficiencies among the union of species in the controls (@mclaren2019cons).
+The measured relative efficiencies can then be used to calibrate (remove the effect of bias from) the relative abundances of the control species in a set of primary (non-control) experimental samples that were measured with the same protocol (ideally in the same experiment and sequencing run) (@mclaren2019cons).
 Calibration can be extended to species not in the controls using statistical methods to impute (predict) the unobserved efficiencies using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
+Control samples may also be aliquots of a natural sample that are shared across experiments, allowing for characterization of the _differential bias_ between experiments.
+Calibration using the differential bias provides a quantitative way to make results directly comparable across studies using different protocols.
+Calibrated relative abundances can be combined with an accurate total-abundance measurement or accurately quantified spike-in to obtain calibrated absolute abundances.
 
@@
 We then estimated species' densities using the uncalibrated and calibrated proportions using total-abundance normalization, treating the total density as known and fixed.
-Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions and densities (Figure \@ref(fig:calibration-example)).
+Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions or absolute abundances derived from them (Figure \@ref(fig:calibration-example)).
+
 To demonstrate a practical application, we used the bias estimated from the full set of mock communities as the basis for calibrating the vaginal community time series measurements from the MOMS-PI study (@fettweis2019thev) that were made using the same 16S sequencing protocol.
@@
 Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \@ref(case-studies).
+Uncertainty in the imputed efficiencies has a large impact on the abundances of species in individual samples; however, because the mean efficiency mainly depends on the most abundant species in the sample, its variation across samples can be accurately predicted so long as the most abundant organisms are covered.
+(TODO: Show this in the SI Rmd MOMSPI case study)
+Therefore less taxonomic coverage is needed to robustly infer fold changes in proportions (or absolute abundances derived from them) than to infer values in individual samples.
 
@@
 
-## Calibration using reference species
+## Complementary total-abundance and MGS measurements for absolute DA
 
-Obtaining control communities that span the taxa of interest is often not practical.
-Moreover, the bias that is measured from controls may differ from that in the primary samples due to differences in cell state or sample preparation.
-These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-abundance measurement that was described in Section \@ref(differential-abundance).
-This *reference-species calibration* allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from taxonomically biased, relative metagenomics measurements.
+An emerging method for absolute DA analysis involves coupling MGS with an experimental measurement of the total abundance of the community, either of cells (e.g.
+by flow cytometry) or extracted DNA (e.g.
+by marker-gene qPCR or bulk DNA quantification).
+Species abundances are obtained by multiplying the measured total abundance 
+by the species proportions derived from MGS (Equation \@ref(eq:density-prop-meas)).
+This method has been dubbed 'Quantitative Microbiome Profiling' (@vandeputte2017quan); however, it remains subject to taxonomic bias in the MGS measurement (Section \@ref(absolute-abundance), Equation \@ref(eq:density-prop-error)).
+The error in the MGS-derived species proportions propagates to the measured abundances, and we've shown that this error can vary across samples and so negatively affect DA analyses.
+The total-abundance measurement is itself also subject to taxonomic bias---yet therein lies a possible solution.
 
+The error in a species' abundance due to bias varies across samples with the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement (Equation \@ref(eq:density-prop-error)).
+Consequently, measurements that are complementary, in the sense of having similar relative efficiencies among species, can lead to more consistent error across samples, thereby improving the accuracy of DA analyses.
+
+We illustrate by considering the debate over whether flow cytometry or 16S qPCR measurements of total abundance are better suited for normalizing 16S amplicon sequencing experiments (@galazzo2020howt, @jian2021comm).
+Flow cytometry directly counts cells, whereas 16S-qPCR measures the concentration of a marker gene following DNA extraction and so is subject to bias from extraction, copy-number variation, primer binding, and amplification.
+Thus we should intuitively expect flow cytometry to provide a more accurate measure of total abundance (and its changes across samples).
+Yet the 16S-qPCR measurement shares much or all of its bias with the 16S sequencing measurement, such that we can expect its mean efficiency to covary with that of the sequencing measurement.
+In that case, our theory predicts that the qPCR-based measurements of species abundance, while not more accurate than the cytometry-based ones, will have more stable fold errors and thus lead to more accurate fold changes across samples.
+
+These observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement is qPCR or ddPCR of that marker gene from the same DNA extraction.
+Similar reasoning suggests that for shotgun sequencing, the ideal total-abundance measurement is bulk DNA quantification: Shotgun sequencing and bulk DNA quantification are both subject to bias from extraction and variation in genome size.
+
+Making optimal use of these pairings requires making thoughtful choices during bioinformatics.
+For example, performing copy-number correction on amplicon read counts prior to multiplication by qPCR measurements would be counter-productive, as it decouples bias in the two measurements.
+Additionally, the MGS proportions should be computed prior to discarding any unassigned reads, since species that are missing from the given taxonomy database still contribute to the total concentration of marker-gene and/or bulk DNA.
+
+Notably, this analysis suggests that for both marker-gene and shotgun sequencing, DNA-based measurements are preferable to cell-based measurements, despite the fact that bias during extraction makes DNA-based measures likely a poor proxy for true changes in cellular abundance.
+For the purposes of species-level DA analysis, however, it can be preferable for the total-abundance measurement to share the extraction bias of the MGS measurement rather than to have no bias at all.
+An important countervailing consideration is whether DNA-based quantification is significantly less linear in sample input concentration than cell-based quantification.
+Random and systematic nonlinearity in the DNA yield and/or quantification as input cell concentration increases will lead to distortions in the species-level fold changes unless otherwise measured and accounted for.
+
+## Reference species for absolute DA
+
+Note for revise: consider putting this first again, and having the transition be that this is a way to get bias to cancel out fully; the other approach can account for most-all of it.
+
+A competing approach to absolute DA analysis instead derives absolute-abundance information from a cellular or DNA spike-in to samples in a fixed concentration prior to DNA extraction.
+A conceptually similar approach is to choose a native species, such as the host, to treat as the constant against which to measure other species.
+Using one of two possible calculations (via MGS proportions or ratios) described in Section \@ref(absolute-abundance), these reference-species methods are each capable of yielding constant fold errors in species abundances and hence accurate fold changes and ranks across samples.
+
+To these two reference-species approaches, we propose a third: Directly measure the absolute abundance of one or more non-constant native species.
+Provided that the targeted species measurement has a consistent absolute efficiency, this approach too yields constant fold errors in species abundances.
 To demonstrate the ability for reference calibration to improve fold changes, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
 Doing so improved the resulting measurements of fold changes in density for all species (Figure \@ref(fig:calibration-example)).
-
-Two categories of reference species have so far been used: Housekeeping species assumed to have constant abundance and spike-in species added at known (and typically constant) abundance.
-The implications of bias for these methods hadn't been previously investigated.
-Our results show that, provided that taxonomic bias among all species (including the housekeeping or spike-in species) is consistent across samples, then the fold error in species abundances is constant and does not impact DA analysis of log- or rank-transformed species abundances.
-These methods have downsides: Housekeeping species are not always available/known, and spike-ins have the downside of needing to determine species not present in the samples and an appropriate concentration in which to add them so that they form a moderate fraction of reads across samples with potentially widely varying produce of mean efficiency and total abundance.
-
-Section \@ref(abundance-measurement) presents an additional type of reference species: naturally-occurring species whose abundance has been directly measured.
 For most natural ecosystems, there is unlikely to be a single species or genus that is found in all samples.
@@
 Higher-order taxa necessarily have higher prevalence than species, though may be more prone to errors from a varying efficiency across samples.
-Targeted measurements of taxa that only span a subset of samples can still provide useful validation for methods.
+DNA-based targeted measurements, such as qPCR with species-specific primers, may suffer from the non-linearity problem mentioned above.
+Cell-based targeted measurements such as using ddPCR directly on cells or virions (REFS) may be preferable.
 
-<!-- TODO: Add paragraph: Pre- versus post- extraction; linearity of DNA extraction. -->
-<!-- Applies to spike-ins and targeted measurements. -->
+Beyond species-level DA analysis, reference species (particularly, spike-ins and housekeeping species) have also been used to measure changes in total-community abundance (@rao2021mult, OTHERS).
+Our results show that the bias invariance seen at the species-level does not apply for this purpose---the measured fold change in total abundance have an error equal to the inverse change in the sample mean efficiency.
+The same problem arises when analyzing the change in any taxon above the taxonomic order at which bias (i.e., the relative-efficiency trait) is conserved.
 
-Housekeeping taxa and spike-ins have also been used to measure changes in total-community abundance.
-Our results show that the bias invariance does not apply for this purpose---the measured fold change in total abundance have an error equal to the inverse change in the sample mean efficiency.
-A similar problem arises when attempting to infer changes in abundance of any higher-order taxon, unless measurement efficiency is conserved among its species.
-<!-- If the purpose is to assess changes in the total community abundance rather than individual species, we recommend researchers either directly measure total abundance or, if using reference species, then explicitly acknowledge that their results are contingent on constant or at least non-systematically varying mean efficiency.
--->
-<!-- In the second case, confidence in the results can be bolstered by a bias-sensitivity analysis. -->
+## Bias sensitivity analysis
 
-## Choosing complementary total-density and metagenomics measurements
+Even if control measurements are not available, it is possible to computationally assess the sensitivity of a given DA result to taxonomic bias.
+One approach to conducting a _bias sensitivity analysis_ involves re-analyzing an MGS dataset across a range of possible taxonomic biases for the experiment.
+First, a large number of hypothetical biases are generated from a user-specified probability distribution.
+This distribution determines the overall strength and phylogenetic conservation of bias that one wishes to consider, and may be informed by empirical facts such as control measurements or known primer mismatches.
+Next, the DA analysis is re-run while using each bias vector to calibrate the MGS measurements.
+Exactly how this should be done depends on details of the DA method; SI Rmd [TODO] provides several examples.
+Finally, the distribution of results can be compared graphically or with summary statistics to the original result.
+This approach is well-suited to the standard practice in microbiome studies of using 'off-the-shelf' statistical packages, as it can in principle be used with any DA method.
+Alternatively, existing DA methods can be extended to directly include the unknown taxonomic bias in the statistical model and provide DA estimates that inherently account for the added uncertainty in microbiome compositions due to the presence of unknown bias.
+@greenland2005mult compares these two approaches in a general context, and @nixon2022asta applies the second approach to microbiome data in the absence of taxonomic bias.
+A bias sensitivity analysis can also be used to ask what result a study would find that was identical except for having a given differential bias relative to the first, which is useful for disentangling disagreements between studies.
 
-A variety of methods for measuring total community density have been paired with either amplicon or shotgun sequencing data for the purpose of conducting absolute DA analysis using the total-density approach described in Section \@ref(differential-abundance).
-Each of these methods is subject to its own taxonomic bias: Cells from different species are likely to contribute more or less efficiently to the total abundance measurement.
-Consider the popular method of assessing bacterial abundance with total-16S qPCR measurement.
-Even an ideal qPCR protocol that perfectly amplifies all species remains a biased measurement of cell density, since the total 16S copies in the extracted DNA contributed by each species will be proportional to its (species-specific) lysis efficiency and 16S copy-number.
-Empirical evidence supports this, as qPCR measurements of total 16S copies are rarely equivalent to total cell counts, even when multiple copy numbers are taken into account ()@lloyd2013meta).
-16S qPCR is commonly paired with 16S amplicon sequencing, with which these sources of bias are shared, perhaps along with variation in primer binding and amplification efficiency.
-Methods like flow cytometry that directly measure cell density lack these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
-By extending the analysis of Section \@ref(differential-abundance) to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(total-density-bias)).
-This finding suggests that methods of total-density measurement that are more accurate as measures of total cell density may actually perform worse than less accurate methods for the purposes of absolute DA inference.
-
-We illustrate with a hypothetical example of two vaginal communities of equal density, but which are dominated either by _Lactobacillus iners_ or _Gardnerella vaginalis_.
-Based on our earlier analysis in @mclaren2019cons, we estimate that _L. iners_ yields 10X more 16S copies per cell than _G. vaginalis_ in extracted DNA.
-Ignoring other sources of bias, we expect relative efficiencies of _L. iners_:_G. vaginalis_ of 10 for both qPCR and 16S sequencing measurements.
-Consequently, the _G. vaginalis_ community will yield a substantially lower qPCR measurement than the _L. iners_ community and we would incorrectly infer a decrease in total abundance.
-At the same time, this decrease in the mean efficiency of the sequencing measurement would cause artificially high fold changes in the proportions of all species (Section \@ref(differential-abundance)).
-When using qPCR and 16S sequencing measurements for community-based density estimation, these two opposing errors cancel, yielding accurate fold changes in species densities.
-
-More generally, changes in the mean efficiency of the total-density measurement can offset those in the sequencing measurement when comparing log density of species across samples.
-If the (log) efficiencies of the total and sequencing measurement are positively correlated across species, then the (log) mean efficiencies will tend to be positively correlated across samples, leading to a reduction in error in DA analysis relative to total-density methods whose efficiencies are uncorrelated with those of the sequencing measurement.
-These observations suggest that qPCR of a marker gene may be the ideal pairing for amplicon sequencing measurements, despite being a poor measure of changes in total cell density.
-Similarly, bulk DNA quantification may be an ideal pairing for shotgun sequencing, making it possible to account for variation in lysis efficiency and genome length.
-These pairings even make it possible to account for error caused by variation among samples in the fraction of unclassified reads, which can form a major fraction of amplicon and shotgun data.
-Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (Appendix \@ref(total-density-bias)).
-
-## Bias-sensitivity analysis
-
-For experiments that have been conducted without calibration controls, it is still possible to investigate the likelihood that DA results can be explained by taxonomic bias by performing a computational bias-sensitivity analysis.
-
-A bias-sensitivity analysis examines how sensitive the results of a given DA analysis are to the working assumption about the bias of a given protocol.
-This working assumption is typically is that there is no bias, but could also be bias measured in a previous experiment or predicted from taxonomic features like 16S copy number.
-A straightforward approach to bias-sensitivity analysis is to use computer simulation to re-analyze the MGS data across a set of simulated taxonomic biases.
-First, multiple sets of efficiency vectors are randomly generated, each representing a possible quantification of taxonomic bias for the study.
-Second, each efficiency vector is used to calibrate the original MGS measurements; each set of calibrated measurements represents the true community compositions under the given hypothesized bias. 
-Finally, the DA analysis is re-run on each set of calibrated measurements and the distribution of results is compared graphically or with summary statistics.
-A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool 'corncob' are shown in SI Figure \@ref(fig:sensitivity-example).
-The simulated efficiency vectors may be generated to different hypotheses about bias in the given system, such as its overall magnitude and correlation among phylogenetically-related species.
-Thus the utility of such simulations will increase the more we learn about the properties of taxonomic bias in different systems.
-
-An alternative approach to bias-sensitivity analysis would be to directly include the (unknown) taxonomic bias of the protocol in the statistical model used for DA analysis, as described for bias-aware meta-analysis below.
-Further development of tools and workflows for performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance and microbiome analyses more generally.
+Designing and interpreting a bias sensitivity analysis is complicated by the large percentage of zero counts in species-level microbiome profiles, as the results may strongly depend on how these zeros are handled in the calibration process.
+Therefore it can be advantageous to jointly test the sensitivity of assumptions about zero-generating processes along with taxonomic bias, making statistical DA models that include bias and one or more zero-generating processes especially valuable.
+The development of tools and workflows to facilitate bias sensitivity analysis may provide an efficient way to increase scientists' ability to assess the reliability of microbiome results, both for differential abundance and microbiome analyses more generally.
 
@@
 
-So far we have considered analysis of samples subject to the same taxonomic bias;
-however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in samples from different studies.
-This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses.
-A potential solution is to perform a bias-aware meta-analysis that that explicitly accounts for the possibility of distinct taxonomic bias across studies.
+Meta-analysis of microbiome samples measured across multiple studies must contend with the fact that different studies typically use different protocols and hence have different taxonomic biases.
+These different biases can be explicitly accounted for in a _bias meta-analysis_ which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
+Parametric meta-analysis models include study-specific latent parameters representing "batch effects"—non-biological differences in the data from each study which can distort the observed biological patterns.
+By estimating these 'nuisance parameters' along with the biological parameters of interest (such as species log-fold changes), the meta-analysis aims to reduce statistical bias in the biological parameters created by the non-biological differences among studies.
+In a bias-aware meta-analysis, the meta-analysis model is configured so that (some of) the latent parameters correspond to study- and species-specific relative efficiencies.
+If taxonomic bias is consistent within but not between studies, then this approach may improve the ability for the meta-analysis to accurately identify DA patterns as compared to meta-analysis methods in which the 'batch effects' do not reflect the multiplicative and compositional nature of taxonomic bias.
 
-Parametric microbiome meta-analysis models include sets of species-specific latent parameters to model "batch effects"—amorphous differences in the measurements of each study.
-By configuring these models so that these latent parameters correspond to study-specific species efficiencies, we can improve their biological interpretability and gain the ability to directly assess whether disagreements in DA results across studies can be explained by differences in taxonomic bias.
-To the extent that taxonomic bias truly is consistent within studies, meta-analyses that directly account for it can have greater statistical power to extract the truly consistent and inconsistent microbial patterns across studies.
+## Experimental recommendations
 
-## [TODO] Summary
+<!-- Maybe: Give an overview of the solution classes -->
+We close this section by offering several recommendations to guide researchers in choosing one or more methods that are suited to their own purposes.
+
+### Use bias-invariant methods instead of or in addition to noninvariant ones
+
+Our results show that some but not all DA methods are theoretically invariant to taxonomic bias in the sense of having results that are exactly or approximately unaffected by taxonomic bias that acts consistently across samples at the given taxonomic level of analysis.
+Thus an obvious suggestion is to use invariant DA methods in place of noninvariant ones.
+Robust methods may carry their own downsides (such as the issues with ratio-based analyses noted above) which do not allow them to fully meet the goal of the study.
+In this case, bias-invariant methods can be used to validate a primary analysis.
+A recent study by @hevroni2020seas illustrates this second approach.
+The authors analyzed how the within-sample rank of a viral species changed from summer to winter, which can in principle be distorted by species-level bias.
+To address this concern, the authors also analyzed the change in the centered log ratio (CLR) associated with each species, a ratio-based analysis that is bias invariant.
+They found that the species whose within-sample ranks increased also increased in their CLR value, providing evidence that their results were not being driven by taxonomic bias.
+An example for absolute DA is suggested by our marine-sediments case study: If total-abundance measurements are made as well as targeted species measurements for all or a subset of samples, then analysis of species abundances derived from the reference-species abundance can be used as a check against those derived from total-community abundance.
+
+### Use control measurements to quantitatively validate and correct estimates
+
+The use of control measurements in microbiome studies primarily serve as a qualitative check on the success of the experiment.
+Above, we describe several ways that controls can be used to quantitatively characterize and correct the effect of bias in DA analyses.
+Our earlier work (@mclaren2019cons) proposed the use of control communities, which can be used to directly mesaure and correct bias.
+Artificial ('mock') communities are particularly suited to studies of synthetic communities (like the fungal case study) or of ecosystems dominated by a relatively small number of culturable taxa (like the vaginal microbiome).
+In other cases, natural community controls can still be used to correct differential bias across experiments.
+The diversity of methods for measuring absolute abundance from MGS data provides opportunities for quantitative validation and correction of absolute DA analysis without the need for control communities.
+In particular, targeted measurements or spike-ins of one or more taxa can be used to validate absolute DA results derived from by other means.
+Recall that our model indicates that consistent bias does not affect fold changes in species abundance derived by these methods, but error in fold changes from total-abundance measurement that are shared among species.
+As a result, differences between these two bias-invariant methods and the noninvariant methods can be used to quantitatively correct the noninvariant method.
+
+### Use computational approaches to test hypotheses about inconsistent results
+
+Computational approaches to accounting bias provide a cheap way to evaluate hypotheses about how bias might be driving inconsistencies between studies.
+For example, @callahan2017repl report a lack of association of _Lactobacillus iners_ in the vaginal microbiomes of pregnant women with preterm birth, whereas @kindinger2017thei report a positive association.
+@callahan2017repl hypothesized that the difference could be due to the use of primers in @kindinger2017thei that have a general preference for _Lactobacillus spp._ and fail to amplify _Gardnerella vaginalis_ (which sometimes co-occurs with _L.
+iners_), in conjunction with the use by @kindinger2017thei of a different DA analysis method that is likely to be more sensitive to such a bias.
+One can this hypothesis by perturbing the microbiome profiles by @callahan2017repl to have such a bias and perform the analysis methods of both studies on the the original and perturbed data and compare the results.
+Alternatively, both datasets could be jointly analyzed in a bias-aware meta-analysis to estimate the differential bias between studies (while also controlling for health outcome and other relevant subject metadata) and comparing it to the hypothesized bias.
+
+### Combine methods to offset limitations
+
+The various approaches to relative and absolute DA have trade-offs both in terms of their theoretical sensitivity to bias and in terms of their experimental feasibility.
+For example, our results show that for a study that aims to perform absolute DA with 16S amplicon sequencing, the use of spike-ins or of total-16S qPCR to derive absolute abundances can each provide species-level fold changes that are robust to bias, but cannot do the same for total community abundance.
+Therefore when both are of interest, either of these methods can be usefully paired with a method that directly counts cells and so may be more likely to provide accurate fold changes in total abundance.
+As another example, targeted measurements of just one or two species may be unlikely cover all samples, but can be used to validate or correct measurements derived from other absolute-abundance methods.
+
+The two computational methods we discuss can each be usefully combined with the use of control communities or targeted measurements.
+Considerable uncertainty is entailed in extrapolating the bias measured from community controls to other species and sample types.
+This uncertainty can be naturally accounted for by using the control measurements to inform a bias-sensitivity analysis.
+The ability for a bias-aware meta-analysis to adjust for differential bias between studies might increase dramatically if these studies include one or more common natural community controls, enabling direct bias measurement.
+Finally, targeted measurements can be used to constrain the fold changes in absolute abundance in both bias-sensitivity analyses and bias-aware meta-analyses.
+
+### Make assumptions explicit
+
+Regardless of whether the above approaches are taken, researchers can place their microbiome inferences on firmer scientific ground by explicitly stating their assumptions about taxonomic bias---the conditions that are required for a given biological conclusion to follow from a statistical result derived from a particular experimental design and analysis.
+For instance, consider the inference of exponential growth rates of prokaryotic taxa in marine sediments by @lloyd2020evid that was discussed in Section \@ref(case-studies).
+The inferred exponential growth rates are accurate provided that 1) bias is consistent across samples at the taxonomic level being analyzed (here, typically families), 2) the mean efficiency does not systematically vary with sample depth, and 3) the fold error in the total-abundance measurement does not systematically vary with sample depth.
+In the case of meta-analyses, authors can helpfully distinguish whether their analysis merely assumes that bias is consistent within individual studies, or requires the much stronger assumption that it is consistent across studies.
+Creating a norm of explicitly stating the assumptions being made about bias in research articles will help both authors and readers better evaluate the claims being made and to identify opportunities to improve their confidence through implementation of the sorts of solutions we propose.

### 2022-03-05 abundance-measurement.Rmd
Minor text improvements

@@
 For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
-A demonstration in artificially constructed ('mock') communities of bacterial cells is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
+A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
 
@@
 **Higher-order taxa:** How does bias affect the proportions and ratios of higher-order taxa, such as a genera, phyla, or even all organisms?
-More generally, we can identify an higher-order taxon $I$ with a set of species, $\{i \in I\}$.
+More generally, we can identify a higher-order taxon $I$ with a set of species, $\{i \in I\}$.
 The abundance and reads of a taxon $I$ in a sample $a$ equal the sums over its constituent species, $\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)$ and $\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)$.
@@
 
-Despite the diversity of these methods, each can be represented by Equation (\#eq:density-ratio-meas), making it straightforward to understand the impact of taxonomic bias in the MGS measurement.
+Despite the diversity of these methods, each can be represented by Equation \@ref(eq:density-ratio-meas), making it straightforward to understand the impact of taxonomic bias in the MGS measurement.
 It follows from Equation \@ref(eq:ratio-error) that the measured abundance of species $i$ is 

### 2022-03-05 conclusion.Rmd
Minor text improvements

@@
 In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
-In addition, careful pairing of total-abundance and MGS measurement methods may remove most of the bias-driven error in absolute abundances from total-abundance normalization.
+In addition, careful pairing of total-abundance and MGS measurement methods can counter the variation in mean efficiency of the MGS measurement, leading to more consistent fold errors across samples.
 
@@
 <!-- Taxonomic aggregation and non-FC analyses are left out because I'm assuming they will be addressed in the manuscript -->
-Key open questions include 1) determining the extent to which bias is consistent across samples at the species level for a given MGS method and sample type; 2) assessing the validity of post-extraction absolute-abundance controls as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how our assumptions about underlying community dynamics, and in particular what information zero counts provide, affect bias sensitivity.
-In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, better statistical tools are needed that are capable of including taxonomic bias and uncertainty in the MGS and supplemental non-MGS measurements.
+Key open questions include 1) determining the extent to which bias is consistent across samples for different taxonomic levels, MGS methods, and sample types; 2) assessing the validity of post-extraction measurements as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how different underlying community dynamics, and in particular the source of zero counts in MGS measurements, affect bias sensitivity.
+In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, more sophisticated statistical tools are needed to properly account for both taxonomic bias and random variation in the underlying MGS and supplemental (e.g.
+qPCR) measurements.
 Finally, more concrete experimental recommendations and user-friendly statistical workflows are needed for implementing the solutions we propose in various experimental contexts.
 
-Our theoretical framework and example analyses provide a foundation for addressing these open questions and developing a practical experimental protocols and statistical tools.
-We look to a future in which microbiome researchers routinely choose an appropriate combination of experimental and data-analytic methods capable of answering their fundamental question while accounting for taxonomic bias and the other inherent limitations of MGS measurements---and in this way, gaining the confidence needed to codify statistical findings into true scientific knowledge.
+Our theoretical framework and example analyses provide a foundation for addressing these open questions, as well as developing experimental protocols and statistical tools that implement our proposed solutions.
+We look to a future in which microbiome researchers regularly choose an appropriate combination of experimental and data-analytic methods that are capable of answering their fundamental question while also accounting for taxonomic bias and other limitations inherent to MGS measurement.
+In doing so, we will collectively gain the confidence needed to codify the findings from MGS-based microbiome studies into true scientific knowledge.

### 2022-03-05 differential-abundance.Rmd
Revise differential-abundance section
- Add paragraph on rank-based analyses
- Revise for clarity and succinctness

@@
 
-<!-- TODO: Change the GC to be about how there are various methods; state the scope that we consider; and state the punchline that they vary, with proportion-based being more sensitive. -->
+We now consider how the error in MGS-derived abundance measurements affect DA analyses.
+We focus on multiplicative DA analyses, or those that measure the fold change (FC) in the relative or absolute abundance of a taxon.
+Such analyses are widely used in microbiome studies and have more direct ecological interpretations than other types via the multiplicative processes of exponential growth and decay.
+In addition, we briefly consider non-parametric rank-based analyses, which are common in microbiome-wide association studies.
 
-This section turns to how these errors in individual sample measurements described in the previous section affect the results of cross-sample comparisons and DA analyses of many samples.
-<!-- the changes in abundance that we estimate between samples or different host and environmental conditions? -->
-Microbiome researchers employ many measures for the change in species' abundance across samples.
-We focus on analyses of multiplicative or (log) fold changes, which are common across study types and have more direct ecological interpretations than other measures (via the processes of exponential growth and decay).
-In addition, we briefly consider non-parametric rank-based analyses that are common in microbiome-wide association studies.
+## Fold change between a pair of samples
 
-## Fold changes between a pair of samples
-
-The building blocks of a multiplicative DA analysis are the fold changes (FCs) in abundance between individual pairs of samples.
-Intuitively, only abundance measurements that have proportional errors will have FCs that are completely robust to bias.
-
-<!-- P: Proportions and ratios -->
-The phenomenon by which non-proportional errors due to bias distort measured FCs can be seen most simply for species proportions.
-It follows from Equation \@ref(eq:prop-error) that the measured FC in the proportion of a species $i$ from samples $a$ to $b$ is
+The building blocks of a multiplicative DA analysis are the FCs in relative or absolute abundance between pairs of individual samples.
+If the fold error (FE) in a given abundance measure is constant, then it will not affect the FCs among samples.
+<!-- Ratios -->
+We saw, for example, that consistent bias creates constant FE in the ratio between two species $i$ and $j$ equal to the ratio in their efficiencies (Equation \@ref(eq:ratio-error)).
+This error completely cancels in the FC from sample $a$ to $b$,
+\begin{align}
+  (\#eq:ratio-fc-error)
+\underbrace{\frac{\widehat{\text{ratio}}_{i/j}(b)}{\widehat{\text{ratio}}_{i/j}(a)}} _\text{measured FC}
+  &= \frac
+    {\text{ratio}_{i/j}(b) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
+    {\text{ratio}_{i/j}(a) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
+\\[0.5ex]
+  &=
+  \underbrace{\frac{\text{ratio}_{i/j}(b)}{\text{ratio}_{i/j}(a)}}_\text{actual FC}
+  .
+\end{align}
+<!-- Proportions -->
+In contrast, the variable FE in the proportion of a species (Equation \@ref(eq:prop-error)) does not completely cancel,
 \begin{align}
@@
 \end{align}
-The sample-independent efficiency factor cancels, but the sample-dependent mean efficiency does not, leaving an error in the measured FC equal to the inverse change in mean efficiency.
-In contrast, the ratio between species $i$ and $j$ has proportional error and so its FC is unaffected by bias.
-From Equation \@ref(eq:ratio-error),
-\begin{align}
-  (\#eq:ratio-fc-error)
-\underbrace{\frac{\widehat{\text{ratio}}_{i/j}(b)}{\widehat{\text{ratio}}_{i/j}(a)}} _\text{measured FC}
-  &= \frac
-    {\text{ratio}_{i/j}(b) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
-    {\text{ratio}_{i/j}(a) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
-\\[0.5ex]
-  &=
-  \underbrace{\frac{\text{ratio}_{i/j}(b)}{\text{ratio}_{i/j}(a)}}_\text{actual FC}
-  ;
-\end{align}
-the constant error, equal to the ratio between the species' efficiencies, completely cancels, leaving an accurately measured FC.
+Although the constant species' efficiency cancels, the varying mean efficiency does not, leaving a FE in the FC that is the same for every species and is equal to the inverse FC in the mean efficiency.
 
-Figure \@ref(fig:error-proportions) (bottom row) illustrates these two different behaviors for the error in the FCs of proportions versus ratios when the mean efficiency varies between samples.
-Here the mean efficiency decreases by a factor of 2.6 (FC of 0.4X) from Sample 1 to Sample 2, which causes the FC of the proportion of each species to be measured as 2.6X larger than its true value.
+Figure \@ref(fig:error-proportions) (bottom row) illustrates these different effects of bias on FCs in proportions versus ratios.
+In this hypothetical example, the mean efficiency decreases by a factor of 2.6 from Sample 1 to Sample 2 (FC of 0.4X), causing the FC in the proportion of each species to appear 2.6X larger than its true value.
 Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction; we refer to the latter as a _sign error_ in reference to the corresponding log fold change (LFC).
@@
 For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FC (actual FC: 2.3X, measured FC: 6.5X).
-For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
+For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
 For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase (a sign error).
-In contrast, the fold error in Equation \@ref(eq:ratio-error) completely cancels when we divide the ratio measured for one sample $a$ by another sample $b$.
-<!-- **TODO: Note that errors can still arise for higher-level taxa due to variation in efficiency within the taxon.** -->
+In contrast, the ratios between species change by the same factors in the actual and observed taxonomic profiles.
+For example, the ratio of Species 2 to Species 3 changes from $1/1$ to $1/4$ in the actual profiles and from $3$ to $3/4$ in the observed profiles, a decrease by 4X either way.
 
-Absolute-abundance measurements for which bias creates non-proportional errors are subject to the same limitation as proportions.
-In particular, abundance measured by normalization of MGS proportions to a total-abundance measurement will yield errors in FCs given by the inverse change in the mean efficiency, unless this change is offset by error in the total-abundance measurement itself.
-<!-- (see Section \@ref(@absolute-abundance)). -->
-On the other hand, abundances measured by normalization to one or more reference species are capable of having proportional errors that cancel in FC calculations.
+<!-- The ratios in sample 1 are Actual: 1:1:1, Observed: 1:18:6 -->
+<!-- The ratios in sample 2 are Actual: 15:1:4, Observed: 5:6:8 -->
+
+We can similarly understand the impact of bias on the measured FCs in a species' absolute abundance.
+To the extent that the FE in the abundance measurement is constant across samples, it will cancel and lead to accurate FCs.
+For the various measurement methods described in Section \@ref(absolute-abundance), the FE in a species' abundance can be expressed as a species-specific constant multiplied by a species-independent factor that, depending on experimental details, may vary with sample composition.
+Thus the FE in the FC between a pair of samples is the same for all species.
+Further, it can be reduced by experimental designs that minimize cross-sample variation in the species-independent factor, regardless of the magnitude of bias (variation in efficiency among species) in the MGS measurement.
 
@@
 
-<!-- TODO: Write a transition. -->
-<!-- TODO: consider doing log proportion for simplicity, and using the vaginal mb example for the figure -->
-
-DA analysis of multiple samples can be framed as a regression problem in which we analyze the relationship between a microbial _response variable_, such as the log abundance of species $i$, with one or more _covariates_, such as the pH of the sampled environment or whether the sample is from a healthy or sick person.
-The simplest regression analysis uses the simple linear regression model,
+DA analysis can be framed as a regression analysis, the simplest of which uses the simple linear regression model
 \begin{align}
   (\#eq:regression)
-  \log \widehat{\text{abun}}_i(a) = \alpha_i + \beta_i x(a) + \varepsilon_i(a),
+  y(a) = \alpha + \beta x(a) + \varepsilon(a),
 \end{align}
-where $x$ is a continuous covariate (e.g.
-pH) or a binary covariate (e.g.
-$x=1$ for treated patients and $x=0$ for controls), $\alpha_i$ and $\beta_i$ are regression coefficients, and $\varepsilon_i(a)$ is a mean-zero random variable that reflects the residual (unexplained) variation in the response.
-In a DA analysis, we are interested in the slope coefficient $\beta$, which describes how the species' abundance changes with $x$.
-We are generally uninterested in the intercept coefficient, $\alpha$, which captures differences in the baseline abundance among species and---it is usually hoped---study- and species-specific systematic error such as that created by taxonomic bias.
+describing how a microbial _response variable_ $y$ relates to a _covariate_ $x$ and _residual (or unexplained) error_ $\varepsilon$.
+For example, $y$ may be the log absolute abundance of species $i$ and $x$ may be a continuous variable, such as pH, or a binary variable, such as $x=1$ for treated patients and $x=0$ for untreated controls.
+The regression coefficients $\alpha$ and $\beta$ are parameters to be estimated from the data.
+A DA analysis is primarily interested in the slope coefficient $\beta$, which determines how the average or expected value of $y$ increases with $x$.
+For a binary covariate, $\beta$ equals the average difference in $y$ between samples with $x=1$ versus those with $x=0$.
+Taking the response $y$ to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient $\beta$ corresponding to the average log fold change (LFC) in abundance between conditions or for a unit increase in $x$.
 
-How does taxonomic bias in fact impact estimates of the slope coefficient, $\beta$?
-Appendix \@ref(appendix-regression) derives the effect of bias on the estimated coefficients for the simple linear regression model under estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE).
-The result has a simple interpretation that provides intuition for what to expect in more complicated generalized linear regression models used in popular DA methods. 
-Consider the regression \@ref(eq:regression) fit to the measured log abundance of species $i$ across all samples.
-The measurement error in the response variable,
-\begin{align}
-  \text{error}_{i}(a) \equiv 
-  \log \widehat{\text{abun}}_i(a) - \log \text{abun}_i(a)
-  = \log \frac{\widehat{\text{abun}}_i(a)}{\text{abun}_i(a)},
-\end{align}
-can be thought of as having three components:
+Appendix \@ref(appendix-regression) derives mathematical results for the effect that measurement error in the response caused by taxonomic bias has on LFCs estimated via simple linear regression.
+Here we illustrate these results with a example.
+Suppose that we wish to estimate the average LFC in the absolute abundance of every species between samples from two conditions, such as treated and untreated patients.
+Further suppose that we have measured absolute abundances by multiplying the proportions from MGS with an accurate measurement of total-community abundance.
+For each species $i$, we set $y(a)$ in Equation \@ref(eq:regression) equal to $\log \widehat{\text{abun}}_{i}(a)$, and use either Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE) to estimate the coefficients $\alpha$ and $\beta$.
+From Equation \@ref(eq:density-prop-error), we know that the FE in the abundance of each species equals its efficiency relative to the mean efficiency of the sample.
 
-- Error that is constant across samples affects the intercept coefficient, $\alpha_i$.
-- Error that varies systematically with $x$ affects the slope coefficient, $\beta_i$.
-- Error that varies and is uncorrelated with $x$ affects the residual variation, $\varepsilon_{i}(s)$.
+How does this error affect the estimate of $\beta$ for species $i$?
+There are three possibilities: 
+In the first scenario, the mean efficiency is constant across samples, such that the FE in $\widehat{\text{abun}}_{i}(a)$ is constant.
+This constant FE induces a constant additive error in $y$ that affects the estimate of the intercept $\alpha$, but not the slope $\beta$.
+In the second scenario, the mean efficiency varies but is not systematically higher or lower in samples from treated versus untreated patients.
+This unsystematic variation in the mean efficiency induces additive error in $y$ that is uncorrelated with $x$ and is statistically equivalent to an increase in the residual variation ($\epsilon$ in the regression model).
+The estimate of $\beta$ will be less precise (i.e., have a larger standard error) than in Scenario 1, but will not be systematically distorted.
+In the third scenario, the mean efficiency is systematically different in treated versus untreated patients.
+If the mean efficiency is higher in treated patients, then the additive error in $y$ will tend to be smaller (or more negative) in this condition, leading to a downward shift in the estimate of $\beta$.
+More generally, the portion of additive variation in the log mean efficiency that is correlated with $x$ creates an equal but opposite error in the estimate of $\beta$ that is the same for each species, whereas the portion that is uncorrelated with $x$ affects the standard errors of $\beta$.
 
-We saw in Section \@ref(abundance-measurement) that the error from taxonomic bias has a species-specific, sample-independent component that is constant across samples; this component causes a species-specific systematic error in the estimated intercept, $\alpha_{i}$.
-The error also has a sample-specific, species-independent component. 
-The portion that is correlated with the covariate $x$ creates a systematic error in the slope coefficient $\beta_{i}$ that is the same for every species and opposite in sign to how the error varies with $x$;
-the portion that is uncorrelated with $x$ affects the precision (standard errors) of the slope estimates in a species-specific manor.
-
-<!-- The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$. -->
-
-Figure \@ref(fig:regression-example) illustrates with a simulated data for the case of log abundance measured using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)), assuming that total community abundance is measured perfectly accurately.
-In this case, the variable component of the error is due entirely to variation in the log mean efficiency of the MGS measurement.
-Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
-The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with $x$ relative to that of the response (here, $\log \text{abun}_{i}$) with $x$ or, equivalently, the relative magnitudes of their slopes.
-As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
-Variation in the log mean efficiency that is uncorrelated with $x$ does not systematically distort $\hat \beta$ but does affect its precision, typically leading to increased standard errors as the variation in log mean efficiency effectively acts as an additional source of noise in measured abundance (Figure \@ref(fig:regression-example) D).
-The exception is for species whose residual variation is strongly positively correlated with that of log mean efficiency (here, Species 9), which can appear to have less random variation and receive standard errors that are too small.
+Figure \@ref(fig:regression-example) shows a simulated example for which the mean efficiency is systematically higher in the condition $x=1$, leading to a systematic decrease in the estimated LFC for each species.
+As we saw for the FC between individual samples, the biological significance varies by species, despite the absolute error in the LFC being shared.
+We see decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending the value of the actual LFC relative to the error.
+The portion of the variation in the log mean efficiency that is uncorrelated with $x$ increases the standard errors for most species (Figure \@ref(fig:regression-example) D).
+The exception is Species 9, which is the primary driver of the the mean efficiency dynamics in this example.
 Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).
-
-Overall, we find that variation in measurement error created by variation in the mean efficiency of the MGS measurement can negatively affect certain DA analyses by creating systematic errors in the slope coefficient estimates and/or reducing their precision.
-Similar effects occur for regressions of log proportions and of log abundances derived by total-abundance normalization when the total abundance is accurately measured; however, as noted in Section \@ref(abundance-measurement), total-abundance measurements that are taxonomically biased in a complementary manner can mitigate these effects.
-Regression analyses of log-ratios and log abundances derived from reference-species normalization remain unaffected---the errors are constant across samples and so only impact the intercept coefficients.
-
-<!-- Similar results apply to microbiome regression tools (such as corncob; @martin2020mode) that perform regression on the logit (instead of log) proportion of a species;  -->
-<!-- however, the mean efficiency of the entire sample must instead be replaced with the mean efficiency among all species excluding the focal species, which causes the absolute error in regression coefficients to vary somewhat across species.
--->
 
@@
 
-### [TODO] Rank-based analyses
+## Rank-based analyses
+
+Researchers often prefer to use nonparametric, rank-based DA methods to determine how species vary across sample conditions.
+Rather than estimating LFCs, these methods estimate the average change in the cross-sample rank of a species' relative or absolute abundance.
+Popular rank-based methods in microbiome DA analysis include the Wilcoxon-Mann-Whitney and Kruskal-Wallis tests when the covariate $x$ is discrete covariate and Spearman's rank correlation when $x$ is continuous.
+Rank-based DA methods are unaffected by a constant FE in the response prior to rank transformation, since such an error does not affect the cross-sample ranks.
+Hence consistent taxonomic bias does not affect rank-based analyses of species ratios.
+However, variable FE in the untransformed response (as seen in species proportions) can change the cross-sample ranks.
+If associated with the covariate $x$, the resulting error in the ranks will create a systematic error in the DA result.

### 2022-03-05 case-studies.Rmd
Revise gut case study
Revised based on BJC's suggested revisions from 2022-02-15.

@@
 
-An overwhelming majority of applications of DA analysis are in studies of the human gut microbiome. (True?)
-Many are microbial association tests in a case-control design---looking for taxa that are differentially abundant in one group of subjects versus another.
-It is also increasingly common to perform time series experiments to elucidate dynamics of microbial communities within a host and their interaction with dynamic host variables.
-Although interest in other microorganisms and viruses is growing, so far these are mostly of bacteria.
-Were therefore asked whether there are features of human gut bacterial microbiomes that we might expect to make DA analyses more or less robust to taxonomic bias.
+We wondered whether the greater alpha (within-community) diversity in the human gut microbiome samples relative to the vaginal microbiome might reduce the impact of bias on DA analyses.
+In a community with higher species-level alpha diversity, the mean efficiency effectively averages over a larger number of species.
+Consequently, if communities are assembled randomly, the mean efficiency varies less as the order-2 alpha diversity $^2D$ increases (Appendix REF).^[Order-2 alpha diversity, also known as the order-2 Hill number or effective number of species, is equivalent to the Inverse Simpson diversity index and is given by $^2D = 1 / \sum_{i}P_{i}^2$, where $P_{i}$ is the proportion of species $i$ in the sample (@jost2006entr).]
+But real communities are not randomly assembled; instead, species have complex patterns of positive and negative associations (REF).
+Thus there is unlikely to be a simple 1-1 relationship between alpha diversity and variation in the mean efficiency.
+Moreover, even if the mean efficiency does vary less across gut microbiome samples, if species (relative or absolute) abundances also vary less then the importance of bias for DA analyses might stay the same or even increase.
 
-We first considered whether the greater diversity of the human gut microbiome as compared to the human vaginal microbiome might limit the impact of bias on species-level DA analyses.
-Intuitively, we expect the mean efficiency to vary less across samples from a given ecosystem type as alpha diversity increases, since it effectively averages over a larger number of species.
-Indeed, as shown in Appendix X, in randomly assembled communities the additive variance in the mean efficiency is inversely proportional to the order-2 _true diversity_ or _effective number of species_ in the community, which is equivalent to the popular Inverse Simpson alpha diversity index.
-But real communities do not assemble randomly.
-Moreover, it is the multiplicative variation in the mean efficiency relative to that in the species proportions that determines the impact bias has on proportion-based DA analyses; it is possible that the variation in individual species proportions is reduced to an even greater degree than that in the mean efficiency.
-Thus it is not obvious whether the gut microbiome's higher alpha diversity will effectively insulate its DA analyses from the effects of bias.
-
-To investigate this question, we analyzed species-level bacterial profiles from vaginal and stool samples from healthy subjects that had been subjected to shotgun sequencing as part of the Human Microbiome Project (@huttenhower2012stru).[^We obtained MetaPhlAn3 species-level bacterial relative abundances from the curatedMetagenomicData R package.]
-Consistent with previous findings, the stool microbiomes had substantially greater diversity than the vaginal microbiomes by several measures, including: within individuals (alpha diversity) and in total across many individuals (gamma diversity); at the species level and by measures that account for phylogenetic divergence; and for diversity measures spanning a range of sensitivity to low abundance species as quantified by Hill's $q$, including richness ($q=0$), Shannon entropy ($q=1$), and the Inverse Simpson index ($q=2$).
-To understand how this greater diversity might modulate the effect of bias, we used computer simulation to assess the variation in the mean efficiency under a range of plausible taxonomic biases, with and without phylogenetic conservation of the efficiency trait.
-_In each simulation replicate, the species efficiencies were sampled from a multivariate log-normal distribution.
-Phylogenetic independence was achieved by setting the covariance matrix equal to the identity matrix, so that the efficiencies were IID.
-Phylogenetic conservation was achieved by setting the covariance matrix equal to the phylogenetic covariance matrix, which amounts to Brownian evolution of the log efficiency trait._
-As expected, the mean efficiency varied less in the higher-diversity gut microbiome samples than in the vaginal microbiome samples; the geometric standard deviation (GSD) of the mean efficiency in stool samples was on average 1.5X less than that in vaginal samples regardless of whether bias was phylogenetically conserved.
-This difference is not extreme and leaves substantial variation in the mean efficiency in gut samples.
-We suggest several possible explanations.
-First, although richness is typically around 100 bacterial species in the MetaPhlAn3 gut profiles, the more relevant order-2 effective number of species ($^2D$; equivalent to the Inverse Simpson diversity index) is typically only around 10 and often lower.
-A order-2 diversity that is much lower than species richness arises from the skewed distribution of abundances within a gut microbiome sample, with the most abundant 1-2 species often forming a majority of reads.
-It is these high-proportion species, captured by the order-2 diversity, that ultimately drive the variation in mean efficiency, not the large number of low-proportion species that drive the richness.
-A second, related reason is that, like in the vaginal microbiome, a relatively small set of species tend to be the most abundant across individuals.
-The overall effect is similar, though less extreme, as that of the dominance of a few key species in the vaginal microbiome: Variation in the efficiencies among these species has an outsized impact on the mean efficiency, again leading to a weakening of the averaging affect when we consider variation across samples.
-
-We next compared the variation in species proportions in the two ecotypes.
-In assessing multiplicative variation, one must grapple with the sparsity of human microbiome data.
-Across both ecotypes, most species have low prevalence, appearing in just a small minority of samples.
-One way to handle these zero observations is to replace them with a small positive value, roughly representing a minimum detection limit.
-Although the multiplicative variation is highly sensitive to this value, we found that the relative difference between gut and vaginal samples was robust to values spanning $10^{-6}$ to $10^{-3}$.
-The average GSD of species in gut samples was around 1.8X lower than that of species in vaginal samples, regardless of zero-replacement value.
-Thus the variation in species proportions was lower in the gut samples by a similar or greater degree than the variation in mean efficiency, suggesting that bias may be just as or more problematic for inferring fold changes in species proportions.
-
-We sought to further understand the implications of the sparsity of gut microbiome for the effect of bias on DA analyses in the context of a real DA analysis.^[Some of the results in this paragraph seem to not hold up to more careful investigation.
-More generally, this paragraph is quite speculative and may get cut from subsequent versions.]
-@vieirasilva2019quan analyzed variation in absolute abundance of genera in stool samples from patients with primary sclerosing cholangitis and/or inflammatory bowel disease.
-Absolute abundances of genera were obtained via the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)) with proportions measured from 16S sequencing and total abundance measured from flow cytometry.
-The authors the rank-based Spearman correlation to quantify the associations in absolute abundance and fecal calprotectin concentration, a biomarker of intestinal inflammation.
-[This will be described in Section \@ref(differential-abundance): Spearman correlation is equivalent to doing a regression on the rank-transformed data and is another form of DA analysis; changes in ranks of proportions are affected by bias, and changes in ranks of ratios are not.
-So rank-based methods (Wilcoxon for boolean covariate and Spearman for continuous covariate) have the same issue as LFC methods when it comes to analyzing proportions instead of ratios.]
-As described in Section \@ref(differential-abundance), taxonomic bias in the MGS measurement has a similar effect on analyses of absolute abundances from the total-abundance normalization as it does no analyses of the raw proportions.
-Thus we can consider whether bias might affect the Spearman correlations in the study.
-When we re-analyzed the Spearman correlations, we found that they were often driven by presence-absence of the given genus, rather than continuous variation in abundance, as evidenced by the fact that for many species, replacing the absolute genus abundances with 0 or 1 indicating presence or absence often did not affect the Spearman correlation.
-For these taxa, we can expect bias to have much less impact on the reported correlation than for taxa for which the correlation was driven by continuous variation.
-The reason is that systematic variation of the mean efficiency with the covariate (which was, in contrast, log-normally distributed) will tend to systematically shift the ranks of observations in a taxon with a continuously varying abundance, but not one where the rank correlation is driven by the proportion of non-zero observations.
-We confirmed this expectation by investigating the sensitivity of Spearman correlations for the top associations that were driven either by presence/absence variation or continuous variation, and found that the correlation was much more sensitive to bias in the second case.
-
-<!-- The impact of bias in an association study depends on whether the mean efficiency is associated with the clinical covariate. -->
-_We might consider adding: A paragraph giving examples where there are large taxonomic shifts associated with a clinical covariate, which therefore have the potential to drive systematic variation in the mean efficiency.
-The primary positive example I have looked at is a study of children with acute diarrhea, in which stool from healthy controls has a microbiome with roughly equal amounts of Actinobacteria (Gram positive), Firmicutes (Gram positive), and Bacteroidetes (Gram negative), and Proteobacteria (Gram negative), but the diarrheic stool is dominated by Proteobacteria and specifically E.
-coli and to a lesser extent Klebsiella (both in Enterobacteriaceae, which is known to be causally associated with acute diarrhea generally).
-Any particular bias favoring E. coli or Enterobacteriaceae would have the potential to distort results, which is something we could simulate and visualize._
+To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project (@huttenhower2012stru, SI Methods \@ref(methods-gut)).
+Stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (TODO SI FIGURE; geometric average of TODO in stool and TODO in vaginal profiles).
+To assess the relative importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth (TODO SI Analysis).
+Across simulation replicates, the GSD in the mean efficiency was typically lower in stool than in vaginal samples by a geometric average of 1.5X (SI FIGURE TODO; geometric average GSD of 1.5 for stool samples and 2.2 for vaginal samples).
+At the level of individual species, the average GSD in the proportion of a species in gut samples was around 1.8X lower than that in vaginal samples.
+The negative effects of bias in a DA analysis of a species' proportion scales with the ratio of the GSD in mean efficiency to that in the species' proportion (Section \@ref(differential-abundance)).
+Thus our results suggest that although the mean efficiency is likely to vary less in the gut, it may remain just as or more problematic for proportion-based DA analyses.
 

### 2022-03-05 case-studies.Rmd
Revise marine case study
Revised based on BJC's suggested revisions from 2022-02-17.

@@
 
-Our route for mean efficiency to become associated with the covariate is if the covariate quantifies a biological process that preferentially selects for a microbial trait that also tends to increase or decrease measurement efficiency.
-We illustrate this potential mechanism with a study of microbial growth in marine sediments.
-
 The surface layers of marine sediments harbor diverse and abundant microbiomes.
@@
 @lloyd2020evid performed a systematic assessment of growth rates of bacterial and archaeal taxa over a depth of 10 cm (corresponding to ~40 years of burial time) in sediment of the White Oak River estuary.
-To estimate growth rate, the authors first measured absolute cell density of microbial taxa using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)), with taxa proportions measured with 16S amplicon sequencing and total community density measured by directly counting cells using epifluorescence microscopy.
-The authors refer to these absolute densities as FRAxC measurements, for 'fraction of 16S reads times total cell counts'.
-The FRAxC measurements were used to infer growth rate from the slope of a simple linear regression of log cell density against burial time over the first 3 cm below the bioirrigation layer (corresponding to $\sim 8$ years of burial).
-To validate the inference of positive growers, the authors compared the growth rates from FRAxC-based inference from two sediment cores, qPCR measurements in these cores for a few reference taxa, and FRAxC-based inference in two replicate two-year laboratory incubation experiments.
+Taxa proportions were measured with 16S amplicon sequencing and total community densities were measured by counting cells using epifluorescence microscopy.
+The authors then calculated cell concentrations for each taxon by multiplying the taxon’s proportion by the total concentration, a method they referred to as FRAxC measurements for 'fraction of 16S reads times total cell counts'.
+Taxon-specific growth rates were inferred from the slope of a simple linear regression of the log of the calculated concentration of each taxon against burial time over the first 3 cm below the bioirrigation layer (corresponding to ~8 years of burial).
 
-Taxonomic bias could lead to systematic error in FRAxC-derived growth rates if sample mean efficiency tends to systematically vary with burial time (or equivalently, depth).
-One possibility is that microbes with tougher cell walls may persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
-In this case, we would expect the relative abundance of tougher species to increase with depth and hence the mean extraction efficiency to decrease, which in turn would lead to inflated growth-rate estimates for all taxa; a possible end result would be that taxa that decay sufficiently slowly would be mistakenly inferred to have positive growth.
+The FRAxC concentration measurements made by @lloyd2020evid correspond to the total-abundance approach to inferring absolute abundance (Equation \@ref(eq:density-prop-meas)).
+Accordingly, taxonomic bias could lead to systematic error in FRAxC-derived growth rates if sample mean efficiency tends to systematically vary with burial time.
+Such systematic variation could occur if microbes with tougher cell walls both persist longer in the sediment and are more difficult to extract DNA from than microbes with weaker cell walls.
+In this scenario, the relative abundance of tougher species will increase with burial time, as the cells of weaker species degrade.
+This shifting composition will cause the sample mean efficiency to decrease with time, leading to inflated growth-rate estimates for all taxa.
+In this scenario, taxa that decay with depth sufficiently slowly would mistakenly be inferred to have positive growth from the FRAxC-calculated abundances.
+
+Importantly, @lloyd2020evid also used qPCR to measure the absolute concentration of two taxa from these same samples.
+These qPCR measurements allow us to test the hypothesis that systematic variation in log mean efficiency with depth distorts inferred growth rates.
+In Section 8, we argue that taxonomic bias in targeted qPCR measurements is expected to create a roughly constant fold error and yield fold changes in absolute density that are relatively unaffected by changes in mean efficiency.
+Thus comparing qPCR to FRAxC growth rates on the same reference taxa allows us to estimate the systematic error in FRAxC growth rates.
+Because the systematic error in regression slopes due to bias is the same for each taxon (Section 3), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
 
@@
 
-We can test the hypothesis that systematic variation in log mean efficiency with depth distorts inferred growth rates using the qPCR measurements of two clades also collected by @lloyd2020evid.
-In Section \@ref(review-absolute-methods), we argue that taxonomic bias in targeted qPCR measurements is expected to create a roughly constant fold error and yield fold changes in absolute density that are relatively unaffected by changes in mean efficiency.
-Thus comparing qPCR to FRAxC growth rates on the same reference taxa allows us to estimate the systematic error in FRAxC growth rates.
-Because the systematic error in regression slopes due to bias is the same for each taxon (Section \@ref(differential-abundance)), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
-<!-- Our aim is to compare the growth rates for taxa with both FRAxC and qPCR values for the two sediment cores, as reported in Table 1. -->
-The first soil core included qPCR measurements of a single archaeal clade, _Bathyarchaeota_, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of `r round(bathy_core30_fraxc, 3)`/yr by FRAxC and `r round(bathy_core30_qpcr, 3)`/yr by qPCR).
-The second soil core included qPCR measurements of _Bathyarchaeota_ and a second clade, _Thermoprofundales_/MBG-D.
-In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by `r round(bathy_core32_fraxc - bathy_core32_qpcr, 3)`/yr for _Bathyarchaeota_ (`r round(bathy_core32_fraxc, 3)`/yr by FRAxC and `r round(bathy_core32_qpcr, 3)`/yr by qPCR) and by `r round(mbgd_core32_fraxc - mbgd_core32_qpcr, 3)`/yr for _Thermoprofundales_/MBG-D (`r round(mbgd_core32_fraxc, 3)`/yr by FRAxC and `r round(mbgd_core32_qpcr, 3)`/yr by qPCR).
-A low number of experimental samples and noise in both the FRAxC and qPCR measurements place significant uncertainty in these measurements; however, the fact that FRAxC-derived growth rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreases with depth in a manner that systematically biases FRAxC-derived rates to higher values.
+The first soil core included qPCR measurements of a single archaeal taxon, _Bathyarchaeota_, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of `r round(bathy_core30_fraxc, 3)`/yr by FRAxC and `r round(bathy_core30_qpcr, 3)`/yr by qPCR).
+The second soil core included qPCR measurements of _Bathyarchaeota_ and a second taxon, _Thermoprofundales_/MBG-D.
+In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by `r round(bathy_core32_fraxc - bathy_core32_qpcr, 3)`/yr for _Bathyarchaeota_ (`r round(bathy_core32_fraxc, 3)`/yr by FRAxC and `r round(bathy_core32_qpcr, 3)`/yr by qPCR) and by `r round(mbgd_core32_fraxc - mbgd_core32_qpcr, 3)`/yr for _Thermoprofundales_/MBG-D (`r round(mbgd_core32_fraxc,
+Uncertainty in the FRAxC- and qPCR-derived growth rates is not reported and is likely substantial; however, the fact that the FRAxC-derived rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreased with depth in a manner that systematically biased FRAxC-derived rates to higher values.
 The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr suggests an error large enough to impact results for some taxa classified as positive growers, whose FRAxC growth rates ranged between 0.04/yr and 0.5/yr.
-Overall, the comparison between FRAxC and qPCR measurements gives support to the study conclusions, but suggest that species at the lower end of this range of positive FRAxC-derived rates may in fact be merely persisting or even slowly declining in abundance.
-
-<!-- notes/questions -->
-
-<!-- - also bring in the incubation experiments as providing further evidence? -->
-<!-- - discuss that kl et al attempted to account for bias, incorrected concluded consistent bias couldn't mess up results? -->
-<!-- - discuss other mech's that could impact mean eff, which might even push in other directions? -->
-
-<!-- Caveats that might be instructive to mention -->
-
-<!-- - qPCR and seq analysis are both at ~family level; this is prob not a big deal when it comes to inferring the systematic variation in mean efficiency though, since qPCR and 16S seq share a lot of bias -->
-<!-- - relic DNA -->
-<!-- - possibility of non-genetic variation -> change in efficiency with depth. This could still lead to a change in mean efficiency, but would offset the effect for these particular species. -->
-<!-- - possible error/bias in CARD-FISH -->
+In summary, the comparison between FRAxC and qPCR measurements supports the overall study conclusions, but also suggests that species at the low end of positive FRAxC-derived rates may be merely persisting or even slowly declining in abundance.
 

### 2022-03-08 case-studies.Rmd
Update fungal case study to use GP regression

@@
 The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
-Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual taxa, it had a negligible impact on the inferred fold changes and the DA analyses based on them.
+Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual species, it had a negligible impact on the inferred fold changes and the DA analyses based on them.
 
@@
 Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5X to 5.2X.
-We used simple linear regression to measure the LFC in commensal species proportions pre- to post-infection, with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
-All commensals tended to decrease in proportion post-infection, which is expected given the pathogen’s growth to most abundant community member and the sum-to-one constraint of species proportions.
-However, when taxonomic bias was ignored the magnitude of the post-infection decrease was overestimated, by an amount corresponding to the inverse change in log mean efficiency.
-This magnitude error was substantial for the commensal-host pairs with the weakest changes; in several cases, LFC estimates with 95% confidence intervals below 0 became indistinguishable from 0 after calibration.
+Using gamma-Poisson (or negative-binomial) regression of the observed read counts, we estimated the average LFC in commensal species proportions following infection with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
+Commensals typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
+However, failure to account for bias caused the decrease to be overestimated, by an amount corresponding to the inverse change in log mean efficiency.
+For commensal-host pairs with relatively small observed decreases, bias correction greatly reduced the magnitude of the negative LFCs or, in several cases, resulted in LFCs that were near 0 or slightly positive.
 
@@
 The scientific error, however, might become worse.
-Suppose that the species whose calibrated proportions appeared to hold steady or slightly decrease actually increased in absolute abundance by around 2X.
-As bias shifts our estimates downwards by 2.5X to 5.2X, we would instead conclude that these taxa decreased.
+Suppose that total abundance increased by 2X due to pathogen growth; in this case, species whose proportions remained approximately constant would have increased in absolute abundance by around 2X.
+Bias, when not accounted for, shifts LFCs estimates downwards by 2.5X to 5.2X depending on host genotype.
+Hence, without bias correction, we would instead conclude that these species decreased.
 
@@
 As a proxy for clinical BV, we split samples from the MOMS-PI study into low diversity (non-BV), intermediate diversity, and high diversity (BV-like) groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
-We estimated the LFC in proportion from non-BV to BV-like samples for 30 prevalent bacterial species, with and without bias correction, using a gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions.
+We estimated the LFC in proportion from non-BV to BV-like samples for 30 prevalent bacterial species, with and without bias correction, using gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions.
 We accounted for bias by including an offset term in the linear model equal to the log ratio of the species' efficiency to the sample mean efficiency.

### 2022-03-08 notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study.Rmd
Update fungal case study to use GP regression

@@
     sample_sum = sample_sums(.),
-    mean_efficiency = calibrate(., bias_t1, norm = 'close') %>%
-      perturb(bias_t1, norm = 'none') %>% 
-      sample_sums,
+    mean_efficiency = mean_efficiency(., bias_t1, type = 'observed')
   )
@@
 I will use `zCompositions::cmultRepl()` as in @leopold2020host for simplicity and consistency with our other analyses in this case study.
-I will recompute the mean efficiency with these new compositions; however, 
-aero replacement will have a minimal impact on the mean efficiency.
+I will recompute the mean efficiency with these new compositions; however, zero replacement will have a minimal impact on the mean efficiency.
 
@@
 bias_inf <- bias/bias['Fusarium']
-ps_inf <- ps
+ps_inf <- ps %>%
+  mutate_sample_data(., 
+    sample_sum = sample_sums(.),
+    mean_efficiency = mean_efficiency(., bias_inf, type = 'observed')
+  )
 stopifnot(taxa_are_rows(ps_inf) == FALSE)
@@
   mutate_sample_data(., 
-    mean_efficiency = calibrate(., bias_inf, norm = 'close') %>%
-      perturb(bias_inf, norm = 'none') %>% 
-      sample_sums,
+    sample_sum = sample_sums(.),
+    mean_efficiency = mean_efficiency(., bias_inf, type = 'observed')
   )
@@
 
-```{r}
-x <- ps_inf_repl %>% 
-  as_tibble %>%
-  left_join(bias_inf %>% enframe('.otu', 'efficiency'), by = '.otu')
-```
-
 Run a simple linear regression with lm on log proportion,
@@
 ```{r}
-x_reg <- x %>%
+res_lm <- ps_inf_repl %>% 
+  as_tibble %>%
+  left_join(bias_inf %>% enframe('.otu', 'efficiency'), by = '.otu') %>%
   with_groups(.sample, mutate,
@@
   ) %>%
-  group_by(.otu, genotype) %>%
-  nest() %>%
+  with_groups(c(.otu, genotype), nest) %>%
   mutate(
-    fit_obs = map(data, 
+    Observed = map(data, 
       ~lm(
@@
       )),
-    fit_cal = map(data, 
+    Calibrated = map(data, 
       ~lm(
@@
       )),
-  )
+  ) %>%
+    select(-data) %>%
+    pivot_longer(c(Observed, Calibrated), names_to = 'type', values_to = 'fit')
+res_lm_tidy <- res_lm %>%
+  mutate(across(fit, map, broom.mixed::tidy, conf.int = TRUE)) %>%
+  unnest(fit) %>% 
+  filter(term != '(Intercept)')
 ```
@@
 ```{r}
-z <- x_reg %>%
-  select(-data) %>% 
-  mutate(across(starts_with('fit'), map, tidy, conf.int = TRUE)) %>%
-  pivot_longer(starts_with('fit'), names_to = 'type', values_to = 'fit', 
-    names_prefix = 'fit_') %>%
-  unnest(fit) %>%
-  ungroup %>%
-  left_join(enframe(bias, '.otu', 'efficiency'), by = '.otu')
-```
-
-```{r}
-z1 <- z %>%
-  filter(term == 'timepoint2') %>%
-  select(.otu, genotype, type, estimate, std.error) %>%
-  pivot_wider(names_from = type, values_from = c(estimate, std.error)) %>%
+res_lm_tidy_wide <- res_lm_tidy %>%
+  select(.otu, genotype, type, estimate, std.error, conf.low, conf.high, p.value) %>%
+  pivot_wider(names_from = type, 
+    values_from = c(estimate, std.error, conf.low, conf.high, p.value)) %>%
   glimpse
@@
 ```{r, fig.dim = c(6,6)}
-z1 %>%
-  ggplot(aes(y = .otu, x = estimate_cal - estimate_obs, color = .otu)) +
+res_lm_tidy_wide %>%
+  ggplot(aes(y = .otu, x = estimate_Calibrated - estimate_Observed, color = .otu)) +
   scale_color_manual(values = pal_genotype, guide = NULL) +
   facet_wrap(~genotype) +
-  stat_pointinterval()
+  geom_point()
 ```
@@
 ```{r, fig.dim = c(10, 8), layout="l-body-outset"}
-z1 %>%
-  ggplot(aes(estimate_cal, estimate_obs, color = .otu)) +
+res_lm_tidy_wide %>%
+  ggplot(aes(x = estimate_Calibrated, y = estimate_Observed, color = .otu)) +
   scale_color_manual(values = pal_genotype) +
@@
     aes(
-      xmin = estimate_cal - 2 * std.error_cal,
-      xmax = estimate_cal + 2 * std.error_cal,
+      xmin = conf.low_Calibrated,
+      xmax = conf.high_Calibrated,
     )
@@
     aes(
-      ymin = estimate_obs - 2 * std.error_obs,
-      ymax = estimate_obs + 2 * std.error_obs
+      ymin = conf.low_Observed,
+      ymax = conf.high_Observed,
     )
@@
 
-```{r infection-lfc-estimates, fig.dim = c(8, 8)}
-z %>%
-  filter(term == 'timepoint2') %>%
+```{r infection-lm-lfc-estimates, fig.dim = c(9, 8)}
+res_lm_tidy %>%
   mutate(
     across(.otu, fct_reorder, estimate),
-    across(type, fct_recode, Corrected = 'cal', Uncorrected = 'obs')
   ) %>%
@@
   )) +
-  # theme_minimal_hgrid() +
+  facet_wrap(~genotype, ncol = 4, scale = 'free_x') +
   theme(
-    axis.line.y = element_blank()
+    axis.line.y = element_blank(),
+    legend.position = 'top',
+    panel.spacing.x = unit(4, 'mm')
   ) +
+  scale_x_continuous(labels = formatC) +
   colorblindr::scale_color_OkabeIto() +
-  facet_wrap(~genotype, ncol = 4) +
+  labs(
+    x = 'Estimated LFC',
+    y = 'Taxon',
+    color = 'Type'
+  ) +
+  geom_vline(xintercept = 0, size = 0.5, color = 'grey') +
+  geom_pointrange(fatten = 2)
+```
+
+## Gamma-Poisson regression
+
+There are a large fraction of zeros at T1 for the lower abundance and/or efficiency taxa, making it more appropriate to apply gamma Poisson model directly to the counts.
+
+```{r}
+library(rstanarm)
+options(mc.cores = parallel::detectCores())
+```
+
+Consider the prior on the intercept.
+Here the offsets will include the read depth, and so we want to set the intercept prior to be for the proportions.
+
+TODO: Set prior in a well-defined way.
+
+```{r}
+x <- ps_inf %>%
+  transform_sample_counts(~close_elts(0.3 + .)) %>%
+  filter_tax_table(.otu %in% commensal_taxa) %>%
+  as_tibble %>%
+  with_groups(c(.otu, genotype), summarize, across(.abundance, gm_mean)) %>%
+  pull(.abundance) %>%
+  summary
+x %>% signif(2)
+
+mu <- log(0.01)
+sigma <- 3
+qnorm(p = c(0.025, 0.975), mean = mu, sd = sigma) %>% exp %>% signif(1)
+custom_prior_intercept <- normal(location = mu, scale = sigma)
+```
+
+```{r}
+tbl_gp <- ps_inf %>%
+  filter_tax_table(.otu %in% commensal_taxa) %>%
+  as_tibble %>%
+  left_join(bias_inf %>% enframe('.otu', 'efficiency'), by = '.otu') %>%
+  mutate(
+    offset_obs = log(sample_sum),
+    offset_cal = log(sample_sum * efficiency / mean_efficiency),
+  ) %>%
+  with_groups(c(.otu, genotype), nest)
+```
+
+```{r, echo = F, eval = F}
+res_gp_test <- tbl_gp %>%
+  slice(1) %>%
+  mutate(
+    Observed = map(data, 
+      ~stan_glm(data = .x, iter = 100,
+        .abundance ~ timepoint + offset(offset_obs),
+        prior_intercept = custom_prior_intercept,
+        family = neg_binomial_2, seed = 42, algorithm = 'sampling'
+      )
+    ),
+    Calibrated = map(data, 
+      ~stan_glm(data = .x, iter = 100,
+        .abundance ~ timepoint + offset(offset_cal),
+        prior_intercept = custom_prior_intercept,
+        family = neg_binomial_2, seed = 42, algorithm = 'sampling'
+      )
+    ),
+  ) %>%
+  select(-data) %>%
+  pivot_longer(c(Observed, Calibrated), names_to = 'type', values_to = 'fit')
+# check the intercept priors
+res_gp_test
+res_gp_test %>%
+  pull(fit) %>%
+  map(prior_summary)
+```
+
+```{r, eval = T}
+res_gp <- xfun::cache_rds({
+  tbl_gp %>%
+    mutate(
+      Observed = map(data, 
+        ~stan_glm(data = .x,
+          .abundance ~ timepoint + offset(offset_obs),
+          prior_intercept = custom_prior_intercept,
+          family = neg_binomial_2, seed = 42, algorithm = 'sampling'
+        )
+      ),
+      Calibrated = map(data, 
+        ~stan_glm(data = .x,
+          .abundance ~ timepoint + offset(offset_cal),
+          prior_intercept = custom_prior_intercept,
+          family = neg_binomial_2, seed = 42, algorithm = 'sampling'
+        )
+      ),
+    ) %>%
+    select(-data) %>%
+    pivot_longer(c(Observed, Calibrated), names_to = 'type', values_to = 'fit')
+}, dir = '_cache/', file = 'res_gp', 
+hash = list(tbl_gp, custom_prior_intercept))
+
+res_gp_tidy <- res_gp %>%
+  mutate(across(fit, map, broom.mixed::tidy, conf.int = TRUE)) %>%
+  unnest(fit) %>% 
+  filter(term != '(Intercept)')
+```
+
+```{r infection-gp-lfc-estimates, fig.dim = c(9, 8)}
+res_gp_tidy %>%
+  filter(term == 'timepoint2') %>%
+  mutate(
+    across(.otu, fct_reorder, estimate),
+  ) %>%
+  ggplot(aes(y =.otu, color = type,
+      x = estimate, 
+      xmin = conf.low,
+      xmax = conf.high
+  )) +
+  facet_wrap(~genotype, ncol = 4, scale = 'free_x') +
+  theme(
+    axis.line.y = element_blank(), 
+    legend.position = 'top',
+    panel.spacing.x = unit(4, 'mm')
+  ) +
+  scale_x_continuous(labels = formatC) +
+  colorblindr::scale_color_OkabeIto() +
   labs(

### 2022-03-08 supplemental-figures.Rmd
Update fungal case study to use GP regression

@@
   'leopold2020host-case-study_files/figure-html5',
-  'infection-lfc-estimates-1.svg'
+  'infection-gp-lfc-estimates-1.svg'
   ) %>%
@@
 
-(ref:cap-leopold2020host-infection-lfc) **Bias correction increases the estimated increase in log proportion in response to infection for commensal taxa across all host genotypes.** Shown are the estimated log fold change (LFC) and 95% confidence intervals from simple linear regression of log (base e) proportion against experimental timepoint for commensal taxa.
-Negative values indicate that the proportion of the taxon decreased on average in response to infection, which we expect due to an increase in pathogen abundance and the sum-to-one constraint of proportions.
-Bias leads to artificially low estimates, as the increased pathogen proportion drives an increase in mean efficiency.
+(ref:cap-leopold2020host-infection-lfc) **Bias correction increases the estimated log fold changes (LFCs) in commensal species proportions in response to pathogen infection.** Shown are the estimated LFCs with 95% Bayesian credible intervals (CIs) derived from a Bayesian gamma Poisson regression.
+Negative values indicate that the species' proportion decreased on average in response to infection, which is the default expectation given growth of the pathogen and the sum-to-one constraint of proportions.
+Bias leads to artificially low estimates, as the increase in the pathogen proportion increases the mean efficiency.
+In Western genotypes, several species whose proportion appears to decrease are instead found to remain stable or even to increase when bias is corrected.
 

### 2022-03-09 appendix-regression.Rmd
Add gamma-Poisson regression to appendix
Introduces gamma-Poisson regression as it applies to sequencing data,
and describes how to set the offsets to correct for bias when estimated
efficiencies are available from community control measurements.

@@
-# Taxonomic bias and linear regression {#appendix-regression}
+# Linear regression {#appendix-regression}
+
+## Simple linear regression
 
@@
 
-## Review of simple linear regression
+### Review of simple linear regression
 
@@
 
-## Measurement error in the response
+### Measurement error in the response
 
@@
 
-## Specific application to taxonomic bias
+### Specific application to taxonomic bias
 
@@
 Thus the covariance of $d$ with $x$ is 0 and taxonomic bias causes no error the slope estimate $\hat \beta$, only the intercept estimate $\hat \alpha$.
+
+## Gamma-Poisson regression
+
+This section describes how gamma-Poisson regression can, with appropriate choice of offsets, be used to estimate LFCs in proportions from MGS count data either with or without bias correction.
+
+### Background
+
+The gamma-Poisson distribution, also known as the negative binomial distribution, is a distribution that is commonly used to model sequencing count data (@holmes2018mode).
+Its key advantage over the simple linear regression model is that it directly models the random sampling of reads that occurs during a sequencing experiment and, in this way, naturally accounts for the imprecision associated with an observation of zero or a small positive count for a given species and sample.
+The gamma-Poisson distribution has two parameters, which jointly determine the mean and the standard deviation.
+We use the parameterization used by the rstanarm R package (@rstanarm), which we use for fitting gamma-Poisson GLMs in our case-study analyses.
+Suppose that $y^{(a)}$ has a gamma-Poisson distribution conditional on a covariate $x$.
+Define two parameters $\mu$ and $\phi$ such that
+\begin{align}
+  E[y^{(a)} \mid x]  &= \mu^{(a)} \\
+  sd[y^{(a)} \mid x] &= \sqrt{
+    E[y^{(a)} \mid x] + E[y^{(a)} \mid x]^2 / \phi
+  };
+\end{align}
+thus $\mu$ equals the mean and $\phi$, known as the 'reciprocal dispersion', increases the standard deviation as it approaches 0 and causes the standard deviation to approach $\sqrt{\mu}$ (that of a Poisson distribution) as it approaches infinity.
+
+For use in gamma-Poisson GLM regression of MGS data, we partition the mean into two factors, $\mu^{(a)} = u^{(a)} \theta^{(a)}$ (@gelman2020regr).
+Here, $\theta^{(a)} = e^{\alpha + \beta x^{(a)}}$ is the quantity of primary interest, whose average LFC equals $\beta$.
+The factor $u^{(a)}$ is a sample-specific _exposure_ parameter that determines the overall scale of counts.
+We can equivalently write $\mu^{(a)} = e^{\alpha + \beta x^{(a)} + \log u^{(a)}}$;
+the logarithm of the exposure, $\log u^{(a)}$, is known as the _offset_ of the GLM.
+Performing gamma-Poisson regression with standard statistical software requires that the offsets are provided as known values.
+
+### Inferring LFCs in proportions with and without bias correction
+
+Our goal is to use gamma-Poisson regression to estimate the LFC in a species proportion from the observed read counts. 
+We therefore equate the counts $M_i^{(a)}$ with $y^{(a)}$ and the proportion $P_{i}^{(a)}$ with $\theta^{(a)}$; it remains to determine the offsets that are consistent with our MGS model.
+To do so, we relax the deterministic assumption of our original model and instead suppose that the right-hand side of Equation \@ref(eq:mgs-model) equals the _expected_ read count of species $i$ in sample $a$,
+\begin{align}
+  (\#eq:mgs-model-gp)
+  E[M_i^{(a)}] = A_i^{(a)} B_i F^{(a)}.
+\end{align}
+The total expected count is
+\begin{align}
+  (\#eq:total-reads-gp)
+  E[M_S^{(a)}] &= \sum_{i \in S} E[M_i^{(a)}]
+  \\&= \left( \sum_{i \in S} A_i^{(a)} B_i \right) F^{(a)} 
+  \\&= A_S^{(a)} B_S^{(a)} F^{(a)}.
+\end{align}
+Under the gamma-Poisson model for $M_i^{(a)}$, we have that $E[M_i^{(a)}] = u_i^{(a)} P_i^{(a)}$.
+Equating this expression for $E[M_i^{(a)}]$ with Equation \@ref(eq:mgs-model-gp) lets us solve for the exposure,
+\begin{align}
+  (\#eq:exposure)
+  u_i^{(a)} 
+    &= \frac{A_i^{(a)} B_i F^{(a)}}{P_i^{(a)}} 
+  \\&= B_i A_S^{(a)} F^{(a)}
+  \\&= \frac{B_i}{B_S} \cdot E[M_S^{(a)}].
+\end{align}
+The second line follows from $P_i^{(a)} \equiv A_i^{(a)}/ A_S^{(a)}$, and the third line from Equation \@ref(eq:total-reads-gp).
+
+The final expression in Equation \@ref(eq:exposure) can be used to compute offsets for the regression.
+The three terms $B_i$, $B_S^{(a)}$, and $E[M_S^{(a)}]$ are each unknown, so we instead substitute estimates for these terms to obtain an estimated exposure $\hat u_{i}^{(a)}$; we then set the offset in the linear model to $\log \hat u_{i}^{(a)}$.
+We can estimate $E[M_S^{(a)}]$ by the observed total count, $M_S^{(a)}$.
+In the absence of bias correction, we set $B_{i} = B_{S}^{(a)} = 1$; our estimate of the exposure is then just
+\begin{align}
+  \hat u_i^{(a)} = M_S^{(a)}.
+\end{align}
+We can apply bias correction using estimates of the efficiencies, $\hat B_{i}$, derived from community control measurements.
+From these estimates and the observed counts, we can estimate the mean efficiency in the sample by
+\begin{align}
+  \hat B_S^{(a)} = \sum_{i\in S} B_i \hat P_i^{(a)},
+\end{align}
+where $\hat P_i^{(a)}$ are the calibrated proportions obtained from a simple plug-in procedure,
+\begin{align}
+  \hat P_i^{(a)} = \frac{M_i^{(a)} / B_i}{\sum_j M_j^{(a)} / B_j}.
+\end{align}
+Equivalently, we can calculate $\hat B_S^{(a)}$ directly from the observed counts,
+\begin{align}
+  \text{B}_S^{(a)} =
+  \left[\sum_j \frac{M_j^{(a)}}{M_S^{(a)} \text{B}_j}\right]^{-1}.
+\end{align}
+We then estimate the exposure by
+\begin{align}
+  \hat u_i^{(a)} = \frac{\hat B_i}{\hat B_S} \cdot M_S^{(a)}.
+\end{align}

### 2022-03-13 abundance-measurement.Rmd
Change equation label

@@
 \begin{align}
-  (\#eq:measurement-model)
+  (\#eq:mgs-model)
   \text{reads}_{i}(a)
@@
 <!-- It accounts for variation in sample input volume and total sequencing run output, as well as the mean species efficiency within the sample. -->
-Equation \@ref(eq:measurement-model) implies that the total number of assigned reads in sample $a$ equals
+Equation \@ref(eq:mgs-model) implies that the total number of assigned reads in sample $a$ equals
 \begin{align}
@@
 \end{align}
-We can rewrite the right-hand side (using Equations \@ref(eq:measurement-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas)) to find
+We can rewrite the right-hand side (using Equations \@ref(eq:mgs-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas)) to find
 \begin{align}
@@
 \end{align}
-From Equations \@ref(eq:measurement-model) and \@ref(eq:ratio-meas), it follows that 
+From Equations \@ref(eq:mgs-model) and \@ref(eq:ratio-meas), it follows that 
 \begin{align}

### 2022-03-13 appendix-regression.Rmd
Minor edits

@@
 <!-- Defining the residuals as $\hat \varepsilon(a) = y(a) - (\hat \alpha + \hat \beta x(a))$,  -->
-The least-squares estimates of $\alpha$ and $\beta$ are the values that minimize the residual sum of squares, $\text{RSS} = \sum_{a} \hat \varepsilon(a)$.
+The least-squares estimates of $\alpha$ and $\beta$ are the values that minimize the residual sum of squares, $\text{RSS} = \sum_{a} \hat \varepsilon(a)^2$.
 These estimates can be conveniently expressed in terms of the sample means and covariances or correlations of $x$ and $y$,
@@
 The two are related via the Equation \@ref(eq:lm-z) for $d$.
-It is helpful to imagine, for the moment, that $y$ and $d$ are known and so we can fit all three models \@ref(eq:lm-y), \@ref(eq:lm-z), and \@ref(eq:lm-d).
+It is helpful to imagine that $y$ and $d$ are known and so we can fit all three models \@ref(eq:lm-y), \@ref(eq:lm-z), and \@ref(eq:lm-d).
 From Equation \@ref(eq:lm-hat) and the linearity of sample means and covariances it follows that 
@@
 \end{align}
-These absolute errors correspond to the _statistical bias_ in the coefficient estimates.
+<!-- These absolute errors correspond to the _statistical bias_ in the coefficient estimates. -->
 Expressed in terms of sample means and covariances, these errors are
@@
 \begin{align}
-  \hat \beta &= \frac{s[\log \text{abun}, x]}{s^2(x)}.
+  \hat \beta &= \frac{s[\log \text{abun}_i, x]}{s^2(x)}.
 \end{align}
@@
 
-This section describes how gamma-Poisson regression can, with appropriate choice of offsets, be used to estimate LFCs in proportions from MGS count data either with or without bias correction.
+This section describes how gamma-Poisson regression can, with appropriate choice of offsets, be used to estimate log fold changes (LFCs) in proportions from MGS count data either with or without bias correction.
 

### 2022-03-14 case-studies.Rmd
Fix mangled line in marine case study

@@
 The second soil core included qPCR measurements of _Bathyarchaeota_ and a second taxon, _Thermoprofundales_/MBG-D.
-In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by `r round(bathy_core32_fraxc - bathy_core32_qpcr, 3)`/yr for _Bathyarchaeota_ (`r round(bathy_core32_fraxc, 3)`/yr by FRAxC and `r round(bathy_core32_qpcr, 3)`/yr by qPCR) and by `r round(mbgd_core32_fraxc - mbgd_core32_qpcr, 3)`/yr for _Thermoprofundales_/MBG-D (`r round(mbgd_core32_fraxc,
+In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by `r round(bathy_core32_fraxc - bathy_core32_qpcr, 3)`/yr for _Bathyarchaeota_ (`r round(bathy_core32_fraxc, 3)`/yr by FRAxC and `r round(bathy_core32_qpcr, 3)`/yr by qPCR) and by `r round(mbgd_core32_fraxc - mbgd_core32_qpcr, 3)`/yr for _Thermoprofundales_/MBG-D (`r round(mbgd_core32_fraxc, 3)`/yr by FRAxC and `r round(mbgd_core32_qpcr, 3)`/yr by qPCR).
 Uncertainty in the FRAxC- and qPCR-derived growth rates is not reported and is likely substantial; however, the fact that the FRAxC-derived rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreased with depth in a manner that systematically biased FRAxC-derived rates to higher values.

### 2022-03-14 case-studies.Rmd
Revise DA theory and case study conclusions
- Revise the DA theory section to frame the results in terms of three
  scenarios, based on the variation in the mean efficiency.
- Revise the concluding section of the case studies to use this framing.
- This version also includes a preliminary draft of the case-studies
  conclusion that came before the DA section revision, for record keeping.

@@
 
-## Summary
+## Summary and conclusions
 
-The impact of bias can depend on protocol, biological system, and type of DA analysis being done.
-Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of Section \@ref(differential-abundance) suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.
+The effect that consistent taxonomic bias has on proportion-based DA analyses depends on the MGS protocol, the biological system, and the sample comparisons under investigation.
+Our case studies explore a limited range of possibilities; however, we can begin to form some general conclusions by placing them within the context of the three scenarios considered for LFC estimation described in Section \@ref(differential-abundance).
 
-<!-- P: Canceling of consistent bias does occur to an important degree -->
+We saw several cases where the mean efficiency is stable across samples, so that bias does not affect LFC estimates (Scenario 1).
+In the pre-infection fungal microbiome samples of @leopold2020host, we observed that the mean efficiency remained stable despite substantial bias and large multiplicative variation in species proportions among samples.
+Because the variation in the mean efficiency was much less than that of the individual species, the impact of bias on DA analysis was negligible.
+In the vaginal case study, we also observed that the mean efficiency was relative stable across vaginal microbiomes that were dominated by the same species, despite substantial bias and large FCs in non-dominant species.
+In both cases, the stability of the mean efficiency can be understood by the fact that it is an additive average over species and so is primarily determined by the most abundant species in the sample.
 
-First, the error caused by consistent taxonomic bias _can mostly cancel_ in cross-sample comparisons and so not impact DA analyses of fold changes in species proportions.
-Our theoretical results from Section \@ref(differential-abundance) indicate that when the mean efficiency is roughly constant across samples, the error in proportions cancels in fold change calculations and is absorbed by the intercept term in regression models.
-We observed this scenario in the analysis of pre-infection fungal microbiome samples of @leopold2020host and when analyzing the trajectories of the vaginal microbiomes of women when the dominant species remained constant^[This point could use more support in the MOMSPI case study], and a stable mean efficiency within marine soil cores is plausible and consistent with (though unable to be fully determined by) the different growth rate estimates in the @lloyd2020evid experiment.
-Section \@ref(differential-abundance) showed that absolute DA analysis using total-abundance normalization are susceptible to errors as with DA analysis of proportions; yet our analysis of two (hypothetical) approaches to analyzing absolute changes in response to infection in the fungal microbiome experiment indicates that the increasingly common approach of pairing marker-gene sequencing with qPCR of the total marker density can largely mitigate this effect, since here what matters is the variation in the ratio of mean efficiencies of the MGS measurement to the total-density measurement and this ratio may remain roughly constant if bias is largely shared by the two measurements.
+Yet we also observed cases where the mean efficiency varied substantially.
+In several cases, the mean efficiency varied systematically with the condition of interest sufficient to cause substantial systematic error in LFC estimates (Scenario 3).
+Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
+Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time that is sufficient to substantially inflate the estimated growth rates of slowly changing species.
+The mean efficiency cannot change more than the species proportions constituting it.
+Therefore, the error in the estimated proportion FCs tend to be practically significant only for species whose FCs are substantially smaller in magnitude than the largest FC.
 
-Yet in other cases, the mean efficiency can vary substantially and create substantial error in fold changes between pairs of samples.
-We saw examples when comparing fungal microbiomes pre- and post-infection in the foliar-fungi experiment and comparing vaginal microbiomes with different dominant species in the MOMS-PI experiment.
-The impact of these errors on results of DA regression analysis across many samples depends on whether the mean efficiency varies systematically with the covariate of interest.
-<!-- P: but bias can still cause substantial systematic error -->
-In these two examples, systematic variation of the mean efficiency arose as high-efficiency species tended to dominate samples in one of the sample conditions (post-infection foliar samples or low-diversity vaginal samples).
-Thus one way for significant error in DA regression results to arise is when a small number of (or just one) species that have particularly high or low mean efficiencies, form a large proportion of the community in a substantial fraction of samples, and are associated with the covariate of interest.
-<!-- This dynamic was responsible for the errors observed in the DA analysis of pre- and post-infection fungal microbiomes from @leopold2020host and in the vaginal microbiome profiles of the MOMS-PI study.
--->
+Variation in the mean efficiency may also be substantial, yet unassociated with the condition of interest (Scenario 2).
+Although we did not directly observe this scenario directly, we have reason to think that it may be common in real microbiome studies.
+Our simulation analysis of gut microbiomes suggest that even in diverse ecosystems, bias will often cause multiplicative variation in the mean efficiency that is comparable to that of individual species.
+This variation is much less problematic for DA results when it not associated with the condition under study, since any loss in precision can in principle be offset by an increase in sample size.
+For certain applications, however, it will be important to remember that the FCs between individual samples remain unreliable.
 
-For many experimental systems, however, there may not be any one species that frequently forms a large fraction of the community.
-In these systems, systematic variation of the mean efficiency can still arise through the collective change of many species that are associated with the covariate of interest.
-We described a hypothesized scenario in the marine-sediment case study in which increased burial time (the covariate) selects for lysis-resistant (and so lower efficiency) species.
-As another example, treatment with a certain antibiotic might selectively kill easier-to-lyse Gram negative species (or harder-to-lyse Gram positives) and thereby decrease (or increase) the mean efficiency in samples collected from a host post-treatment.
-Such situations can arise whenever there is a microbial trait that is associated with both measurement efficiency and the biological processes of interest.
-An example besides cell-wall structure is ribosomal copy number, which increases a species' efficiency in ribosomal amplicon measurements and is positively associated with metabolic rate.
-<!-- Even when there is not a single trait that affects efficiency and the focal biological process,  -->
-A more generic mechanism by which mean-efficiency associations might arise is from the evolutionary relationships among species.
-Species with more recent common ancestry are expected to be more similar across a wide range of heritable traits, which include traits that affect measurement efficiency (such as cell wall structure, genome size, ribosomal copy number, PCR binding sequence) as well as traits that affect the biological processes under study.
-For example, differences in the phylum-level composition of samples from two conditions might be driven by many species that show phylum-level conservation in a relevant biological trait.
-If these species also show phylum-level conservation in (potentially different) traits that affect measurement efficiency, an association of mean efficiency with the condition can arise.
-
-<!-- P: Bias can also reduce the precision and power of DA analyses -->
-
-So long as the mean efficiency does not vary systematically with the regression covariates, the measurement error from bias primarily reduces the precision and power of regression analyses without systematically distorting estimates and so does not necessarily lead to invalid inferences.
-But the reduction in precision could be substantial, particularly given the small sample sizes common in many microbiome studies, and thereby substantially limit the ability to draw meaningful conclusions from a study^[I suspect this non-systematic variation in the mean efficiency is the more typical situation, but we currently lack an example of it in the case studies.
-We can find a semi-real example by taking real microbiome data from a case-control-type design and simulating a strong degree of bias.
-I did this with the HMP2 IBD study, but thinnk it might be better to find a different example where there are some clear DA results.].
-
-<!-- Conclusion: Need solutions -->
+Under what conditions should we expect the problematic third scenario?
+Our case studies suggest one prominent mechanism for causing systematic variation in the mean efficiency: the existence of one or more species with unusually high or low efficiencies that tend to dominate the community more often in one sample condition than another.
+This mechanism was responsible for the negative effect of bias in our DA analyses of foliar fungal microbiomes following infection and of vaginal microbiomes with low versus high diversity.
+In experimental systems where no one species frequently forms a large fraction of the community, systematic variation of the mean efficiency can still arise through the collective change of many species that are associated with the condition of interest.
+We described a plausible instance in the marine-sediment case study, where increased burial time selects for lysis-resistant (and so lower efficiency) species.
+As another example, treatment with an antibiotic might selectively kill Gram negative species, which also tend to be easier to lyse, thereby decreasing the mean efficiency in fecal samples collected after treatment.
+A generic mechanism which might spawn such condition-efficiency associations stems from the evolutionary relationships among species.
+Species with recent common ancestry are expected to share a number of traits that determine measurement efficiency, including cell wall structure, genome size, ribosomal copy number, and PCR binding sequence.
+They are also expected to share traits related to the condition of interest.
+If related species tend to have similar efficiencies and similar associations with the condition, then they may drive an association of the mean efficiency with the condition.
 
 These observations provide reasons to both worry and hope.
-It seems likely that in many experiments the mean efficiency is consistent or is at least not associated with the covariate, so that DA inferences remain valid.
-Yet it is not obvious _a priori_ in which studies this condition holds, and there are plausible mechanisms that can create problematic associations of the mean efficiency even in ecosystems with high species diversity.
+It seems likely that in many studies, the mean efficiency is consistent (first scenario) or is at least not associated with the condition (second scenario), so that DA inferences remain valid.
+Yet it is not obvious which scenario any study falls into.
+There are plausible mechanisms leading to systematic variation of the mean efficiency even in ecosystems with high species diversity.
+Moreover, our explorations suggest that random variation in the mean efficiency is common and distorts comparisons between individual samples.
 Thus while we should not discount the large set of existing DA results, we should seek ways to better assess the robustness of results from previous studies and to measure and correct the error caused by bias in future ones.
 
-<!-- ## cut from conclusions -->
+## [Temp] Summary and conclusions
 
-<!-- ### bias can push species below the detection limit -->
+[first draft of new summary, prior to revising the DA theory section]
 
-<!-- - Another mechanism, not considered here, by which bias can dramatically reduce precision, is by pushing speices below the sequencing detection limit. -->
-<!-- - e.g. of primers that don't detect Gard; problems with estimates of rare and/or low eff taxa on leopold -->
-<!-- - this mechanism can interact with variation in the mean efficiency - e.g. of fungal samples after Mel increases -->
+The effect that consistent taxonomic bias has on proportion-based DA analyses depends on the MGS protocol, the biological system, and the sample comparisons under investigation.
+Our case studies explore a limited range of possibilities; however, when combined with the theoretical framework developed Section 3, they suggest some general conclusions.
 
-<!-- ### Other cases where bias can matter that we didn't consider -->
+In the first scenario, the mean efficiency remains stable across samples despite large variation in the relative efficiencies among species and in species composition among samples.
+In this case, the error caused by consistent taxonomic bias _will mostly cancel_ in cross-sample comparisons, and bias will not substantially affect the inferred FCs between sample pairs or between sample conditions.
+We observed this scenario in the analysis of pre-infection fungal microbiome samples of Leopold and Busby (2020).
+A stable mean efficiency within marine soil cores is consistent with our comparison of qPCR- and MGS-derived growth rates in the Lloyd et al.
+(2020) experiment, though we lack the power for firm conclusions.
+In the MOMS-PI study, we found that the mean efficiency was stable across vaginal microbiomes with the same dominant species, even as non-dominant species had large FCs.
 
-<!-- Optional for this draft; may also want to leave for final 'conclusions' / discussion -->
+In the second scenario, the mean efficiency varies substantially across samples but is not associated with the condition of interest.
+In this case, the measured FCs between pairs of samples can be highly distorted; however, these errors effectively average out, leading to reliable FCs between sample conditions.
+For example, the mean efficiency varied widely across vaginal microbiomes in the MOMS-PI study, but was only associated with one out of the three proinflammatory cytokines we considered.
+For the other three, bias had a negligible impact on the inferred associations between species and cytokine level.
 
-<!-- - DA analyses that aren't based on fold changes - don't get canceling, and might have bigger impact -->
-<!-- - taxonomic aggregation - e.g. of Lactobacillus in vaginal; (mabye: Firmicutes in gut) - can get a varying efficiency within the taxon -->
-<!-- - ?? zero replacement, interpretation -->
+In the third scenario, the mean efficiency systematically varies with the condition of interest, causing a systematic error in the estimated FCs between sample conditions even when averaged over many samples.
+Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
+Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time sufficient to substantially inflate the estimated growth rates of slowly changing species.
+[TODO: Mention gut microbiome infection, RX examples.]
+The mean efficiency cannot change more than the species proportions constituting it.
+Therefore, the error in the estimated proportion FCs tend to be practically significant only for species whose FCs are substantially smaller in magnitude than the largest FC.
+
+—
+
+Under what conditions should we expect this problematic third scenario?
+Our case studies suggest one prominent mechanism for causing systematic variation in the mean efficiency: the existence of one or more species with unusually high or low efficiencies that tend to dominate the community more often in one sample condition than another.
+This mechanism was responsible for the negative effect of bias in our DA analyses of foliar fungal microbiomes following infection and of vaginal microbiomes with low versus high diversity.
+In experimental systems where no one species frequently forms a large fraction of the community, systematic variation of the mean efficiency can still arise through the collective change of many species that are associated with the condition of interest.
+We described a plausible instance in the marine-sediment case study, where increased burial time selects for lysis-resistant (and so lower efficiency) species.
+As another example, treatment with an antibiotic might selectively kill Gram negative species, which also tend to be easier to lyse, thereby decreasing the mean efficiency in fecal samples collected after treatment.
+A generic mechanism which might spawn such condition-efficiency associations stems from the evolutionary relationships among species.
+Species with recent common ancestry are expected to share a number of traits that determine measurement efficiency, including cell wall structure, genome size, ribosomal copy number, and PCR binding sequence.
+They are also expected to share traits related to the condition of interest.
+If related species tend to have similar efficiencies and similar associations with the condition, then they may drive an association of the mean efficiency with the condition.
+
+These observations provide reasons to both worry and hope.
+It seems likely that in many studies, the mean efficiency is consistent (first scenario) or is at least not associated with the condition (second scenario), so that DA inferences remain valid.
+Yet it is not obvious which scenario any study falls into.
+There are plausible mechanisms leading to systematic variation of the mean efficiency even in ecosystems with high species diversity.
+Moreover, our explorations suggest that random variation in the mean efficiency is common and distorts comparisons between individual samples.
+Thus while we should not discount the large set of existing DA results, we should seek ways to better assess the robustness of results from previous studies and to measure and correct the error caused by bias in future ones.

### 2022-03-14 differential-abundance.Rmd
Revise DA theory and case study conclusions
- Revise the DA theory section to frame the results in terms of three
  scenarios, based on the variation in the mean efficiency.
- Revise the concluding section of the case studies to use this framing.
- This version also includes a preliminary draft of the case-studies
  conclusion that came before the DA section revision, for record keeping.

@@
 
-How does this error affect the estimate of $\beta$ for species $i$?
-There are three possibilities: 
-In the first scenario, the mean efficiency is constant across samples, such that the FE in $\widehat{\text{abun}}_{i}(a)$ is constant.
-This constant FE induces a constant additive error in $y$ that affects the estimate of the intercept $\alpha$, but not the slope $\beta$.
-In the second scenario, the mean efficiency varies but is not systematically higher or lower in samples from treated versus untreated patients.
-This unsystematic variation in the mean efficiency induces additive error in $y$ that is uncorrelated with $x$ and is statistically equivalent to an increase in the residual variation ($\epsilon$ in the regression model).
-The estimate of $\beta$ will be less precise (i.e., have a larger standard error) than in Scenario 1, but will not be systematically distorted.
-In the third scenario, the mean efficiency is systematically different in treated versus untreated patients.
-If the mean efficiency is higher in treated patients, then the additive error in $y$ will tend to be smaller (or more negative) in this condition, leading to a downward shift in the estimate of $\beta$.
-More generally, the portion of additive variation in the log mean efficiency that is correlated with $x$ creates an equal but opposite error in the estimate of $\beta$ that is the same for each species, whereas the portion that is uncorrelated with $x$ affects the standard errors of $\beta$.
+To understand how the FE in species abundances affects the estimated LFC for a given species, it is useful to consider three possible scenarios for how the mean efficiency varies across samples.
+In the first scenario, the mean efficiency remains stable across samples, such that the FE in $\widehat{\text{abun}}_{i}(a)$ is approximately constant. 
+In this case, the FE induced by bias _mostly cancels_ and does not substantially affect the estimated LFC.
+In the second scenario, the mean efficiency varies substantially but is not associated with the condition, creating random FEs in the individual sample measurements.
+In this case, the LFCs between individual samples will be highly distorted, but these errors will effectively average out in the regression and the estimated LFC remains reliable.
+Although the random FEs caused by bias do not create a systematic error in the LFC estimate, they will tend to reduce the precision (i.e., increase the standard error) of the estimate.
+This reduced precision can be offset, however, by increasing the number of samples.
+In the third scenario, the mean efficiency systematically varies with the condition, causing a systematic error in the estimated LFC regardless of the number of samples.
+In particular, the portion of additive variation in the log mean efficiency that is linearly associated with $x$ creates an equal but opposite additive error in the estimated LFC for each species.
+Although the absolute additive error in the LFCs is the same for all species, its biological significance varies, depending on its magnitude relative to that of the species' true LFC.
 
-Figure \@ref(fig:regression-example) shows a simulated example for which the mean efficiency is systematically higher in the condition $x=1$, leading to a systematic decrease in the estimated LFC for each species.
+Figure \@ref(fig:regression-example) shows a simulated example of the problematic third scenario.
+Here, the average log mean efficiency is higher in the condition corresponding to $x=1$ due to an increase of the high-efficiency Species 9.
+This systematic increase in the log mean efficiency causes a decrease in the estimated LFC for each species by an equal amount.
 As we saw for the FC between individual samples, the biological significance varies by species, despite the absolute error in the LFC being shared.

### 2022-03-14 case-studies.Rmd
Clean up case-study conclusion
Remove earlier draft; split final paragraph across lines; change 'FC' to
'LFC'; change subsection title to 'Summary and discussion'

@@
 
-## Summary and conclusions
+## Summary and discussion
 
 The effect that consistent taxonomic bias has on proportion-based DA analyses depends on the MGS protocol, the biological system, and the sample comparisons under investigation.
-Our case studies explore a limited range of possibilities; however, we can begin to form some general conclusions by placing them within the context of the three scenarios considered for LFC estimation described in Section \@ref(differential-abundance).
+Our case studies explore a limited range of possibilities; however, we can begin to form some general conclusions by placing them within the context of the three scenarios described in Section \@ref(differential-abundance).
 
@@
 Because the variation in the mean efficiency was much less than that of the individual species, the impact of bias on DA analysis was negligible.
-In the vaginal case study, we also observed that the mean efficiency was relative stable across vaginal microbiomes that were dominated by the same species, despite substantial bias and large FCs in non-dominant species.
+In the vaginal case study, we also observed that the mean efficiency was relative stable across vaginal microbiomes that were dominated by the same species, despite substantial bias and large LFCs in non-dominant species.
 In both cases, the stability of the mean efficiency can be understood by the fact that it is an additive average over species and so is primarily determined by the most abundant species in the sample.
@@
 The mean efficiency cannot change more than the species proportions constituting it.
-Therefore, the error in the estimated proportion FCs tend to be practically significant only for species whose FCs are substantially smaller in magnitude than the largest FC.
+Therefore, the error in the estimated proportion LFCs tend to be practically significant only for species whose LFCs are substantially smaller in magnitude than the largest LFC.
 
@@
 This variation is much less problematic for DA results when it not associated with the condition under study, since any loss in precision can in principle be offset by an increase in sample size.
-For certain applications, however, it will be important to remember that the FCs between individual samples remain unreliable.
+For certain applications, however, it will be important to remember that the LFCs between individual samples remain unreliable.
 
@@
 Thus while we should not discount the large set of existing DA results, we should seek ways to better assess the robustness of results from previous studies and to measure and correct the error caused by bias in future ones.
-
-## [Temp] Summary and conclusions
-
-[first draft of new summary, prior to revising the DA theory section]
-
-The effect that consistent taxonomic bias has on proportion-based DA analyses depends on the MGS protocol, the biological system, and the sample comparisons under investigation.
-Our case studies explore a limited range of possibilities; however, when combined with the theoretical framework developed Section 3, they suggest some general conclusions.
-
-In the first scenario, the mean efficiency remains stable across samples despite large variation in the relative efficiencies among species and in species composition among samples.
-In this case, the error caused by consistent taxonomic bias _will mostly cancel_ in cross-sample comparisons, and bias will not substantially affect the inferred FCs between sample pairs or between sample conditions.
-We observed this scenario in the analysis of pre-infection fungal microbiome samples of Leopold and Busby (2020).
-A stable mean efficiency within marine soil cores is consistent with our comparison of qPCR- and MGS-derived growth rates in the Lloyd et al.
-(2020) experiment, though we lack the power for firm conclusions.
-In the MOMS-PI study, we found that the mean efficiency was stable across vaginal microbiomes with the same dominant species, even as non-dominant species had large FCs.
-
-In the second scenario, the mean efficiency varies substantially across samples but is not associated with the condition of interest.
-In this case, the measured FCs between pairs of samples can be highly distorted; however, these errors effectively average out, leading to reliable FCs between sample conditions.
-For example, the mean efficiency varied widely across vaginal microbiomes in the MOMS-PI study, but was only associated with one out of the three proinflammatory cytokines we considered.
-For the other three, bias had a negligible impact on the inferred associations between species and cytokine level.
-
-In the third scenario, the mean efficiency systematically varies with the condition of interest, causing a systematic error in the estimated FCs between sample conditions even when averaged over many samples.
-Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
-Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time sufficient to substantially inflate the estimated growth rates of slowly changing species.
-[TODO: Mention gut microbiome infection, RX examples.]
-The mean efficiency cannot change more than the species proportions constituting it.
-Therefore, the error in the estimated proportion FCs tend to be practically significant only for species whose FCs are substantially smaller in magnitude than the largest FC.
-
-—
-
-Under what conditions should we expect this problematic third scenario?
-Our case studies suggest one prominent mechanism for causing systematic variation in the mean efficiency: the existence of one or more species with unusually high or low efficiencies that tend to dominate the community more often in one sample condition than another.
-This mechanism was responsible for the negative effect of bias in our DA analyses of foliar fungal microbiomes following infection and of vaginal microbiomes with low versus high diversity.
-In experimental systems where no one species frequently forms a large fraction of the community, systematic variation of the mean efficiency can still arise through the collective change of many species that are associated with the condition of interest.
-We described a plausible instance in the marine-sediment case study, where increased burial time selects for lysis-resistant (and so lower efficiency) species.
-As another example, treatment with an antibiotic might selectively kill Gram negative species, which also tend to be easier to lyse, thereby decreasing the mean efficiency in fecal samples collected after treatment.
-A generic mechanism which might spawn such condition-efficiency associations stems from the evolutionary relationships among species.
-Species with recent common ancestry are expected to share a number of traits that determine measurement efficiency, including cell wall structure, genome size, ribosomal copy number, and PCR binding sequence.
-They are also expected to share traits related to the condition of interest.
-If related species tend to have similar efficiencies and similar associations with the condition, then they may drive an association of the mean efficiency with the condition.
-
-These observations provide reasons to both worry and hope.
-It seems likely that in many studies, the mean efficiency is consistent (first scenario) or is at least not associated with the condition (second scenario), so that DA inferences remain valid.
-Yet it is not obvious which scenario any study falls into.
-There are plausible mechanisms leading to systematic variation of the mean efficiency even in ecosystems with high species diversity.
-Moreover, our explorations suggest that random variation in the mean efficiency is common and distorts comparisons between individual samples.
-Thus while we should not discount the large set of existing DA results, we should seek ways to better assess the robustness of results from previous studies and to measure and correct the error caused by bias in future ones.

### 2022-03-14 appendix-notes.Rmd
Revise gut case study
- Revise Rmd to completion; add results
- Add in final numbers
- Revise discussion of diversity vs. mean-efficiency variation, and add
  derivation to appendix
- Add SI Figure and caption

@@
 Methods differ based on the choice of $R$ and in how the abundance of $R$ is measured.
+
+## Alpha diversity and variation in the mean efficiency {#diversity-and-mean-efficiency}
+
+The diversity of order $q$ is given by
+\begin{align}
+  ^qD = \left(\sum_{i=1}^n p_i^q\right)^{1 /(1-q)},
+\end{align}
+where $^qD$ is understood to be given by the limit of the RHS when $q = 1$ (and so the given expression is undefined).
+The order-2 diversity is therefore
+\begin{align}
+  ^2D = \frac{1}{\sum_{i=1}^n {p_i^2}},
+\end{align}
+which is equivalent to the Inverse Simpson Index (REF Jost).
+
+Consider an infinite pool of species.
+Let $\sigma^{2}$ denote the variance in the relative efficiency among species in the pool.
+Consider a community that is assembled by randomly choosing $K \ge 1$ species from the pool and setting the abundances of the $K$ species in a manner that is independent of their efficiencies.
+
+**Claim:** 
+Let $\rho_{k}$ for $1 \le k \le K$ denote the proportions of the $K$ species in the community; this new notation serves as a reminder that the subscript $k$ indexes a random species specific to this particular community.
+Conditional on the order-2 diversity $^2D$ of a community, the arithmetic variance in the mean efficiency is
+\begin{align}
+  Var[\bar B \mid ^2D] 
+% = \sigma^2 \sum_{k=1}^K \rho_k^2
+  = \frac{\sigma^2}{^2D}.
+\end{align}
+
+**Proof:** Let $\beta_{k}$ denote the efficiency of the $k$-th sampled species.
+The mean efficiency is therefore $\bar B = \sum_k \rho_k \beta_k$.
+Conditional on the proportions $\{\rho_{k}\}$, the variance in the mean efficiency is
+\begin{align}
+  Var[\bar B \mid \{\rho_{k}\}]
+  &= \sum_{k=1}^K \rho_k^2 Var[\beta_k]
+  \\&= \left(\sum_{k=1}^K p_k^2\right) \sigma^2.
+\end{align}
+The first line follows from the fact that the $\beta_{k}$ are independent.
+The summation in the final line is equal to $1/2^D$, thus proving the result.
+
+**Note:** 
+We have showed that the arithmetic (additive) variance of $\bar B$ decreases with $^2D$; however, the geometric (multiplicative) variance is most relevant for understanding the effect of bias on DA.
+As $2^D$ increases, the distribution of $\bar B$ will converge (by the central limit theorem) to a normal distribution, and both the arithmetic and geometric variance will decrease.

### 2022-03-14 case-studies.Rmd
Revise gut case study
- Revise Rmd to completion; add results
- Add in final numbers
- Revise discussion of diversity vs. mean-efficiency variation, and add
  derivation to appendix
- Add SI Figure and caption

@@
 
-We wondered whether the greater alpha (within-community) diversity in the human gut microbiome samples relative to the vaginal microbiome might reduce the impact of bias on DA analyses.
-In a community with higher species-level alpha diversity, the mean efficiency effectively averages over a larger number of species.
-Consequently, if communities are assembled randomly, the mean efficiency varies less as the order-2 alpha diversity $^2D$ increases (Appendix REF).^[Order-2 alpha diversity, also known as the order-2 Hill number or effective number of species, is equivalent to the Inverse Simpson diversity index and is given by $^2D = 1 / \sum_{i}P_{i}^2$, where $P_{i}$ is the proportion of species $i$ in the sample (@jost2006entr).]
-But real communities are not randomly assembled; instead, species have complex patterns of positive and negative associations (REF).
+We wondered whether DA in the human gut microbiome might be less affected by bias than the vaginal microbiome, due to its greater species diversity.
+Recall that the mean efficiency is an abundance-weighted average over species in a sample; thus we might expect it to become more stable as the within-sample (alpha) diversity increases.
+In fact, one can show that for communities that are randomly assembled from a large species pool, the additive variance in the mean efficiency is inversely proportional to order-2 diversity $^2D$ (also known as the Inverse Simpson diversity index) (Appendix \@ref(diversity-and-mean-efficiency)).
+But real communities are assembled through complex processes that create positive and negative associations among species (REF).
 Thus there is unlikely to be a simple 1-1 relationship between alpha diversity and variation in the mean efficiency.
-Moreover, even if the mean efficiency does vary less across gut microbiome samples, if species (relative or absolute) abundances also vary less then the importance of bias for DA analyses might stay the same or even increase.
+Moreover, even if the mean efficiency does vary less across the more diverse gut samples, if the proportion of each species also varies less, then the importance of bias for DA analyses need not diminish.
+
+```{r gut-stats, message = FALSE}
+hmp_div_stats <- fs::path(
+  "notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/_output/",
+  "div_q2_stats.csv"
+) %>%
+  read_csv(col_types = 'cddd') %>%
+  metacal::as_matrix(rownames = body_site)
+
+hmp_me_stats <- fs::path(
+  "notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/_output/",
+  "reps_me_summ_diff.csv"
+) %>%
+  read_csv(col_types = 'cddd') %>%
+  metacal::as_matrix(rownames = sim_type)
+
+hmp_species_stats <- fs::path(
+  "notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/_output/",
+  'species_gsd_stats.csv'
+) %>%
+  read_csv(col_types = 'dddd')
+hmp_species_diff <- hmp_species_stats %>%
+  summarize(diff = metacal::gm_mean(vagina_stool)) %>%
+  pull(1)
+```
 
 To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project (@huttenhower2012stru, SI Methods \@ref(methods-gut)).
-Stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (TODO SI FIGURE; geometric average of TODO in stool and TODO in vaginal profiles).
-To assess the relative importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth (TODO SI Analysis).
-Across simulation replicates, the GSD in the mean efficiency was typically lower in stool than in vaginal samples by a geometric average of 1.5X (SI FIGURE TODO; geometric average GSD of 1.5 for stool samples and 2.2 for vaginal samples).
-At the level of individual species, the average GSD in the proportion of a species in gut samples was around 1.8X lower than that in vaginal samples.
-The negative effects of bias in a DA analysis of a species' proportion scales with the ratio of the GSD in mean efficiency to that in the species' proportion (Section \@ref(differential-abundance)).
-Thus our results suggest that although the mean efficiency is likely to vary less in the gut, it may remain just as or more problematic for proportion-based DA analyses.
+On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \@ref(fig:gut)A; GM $\md$ GSD of `r hmp_div_stats['stool', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['stool', 'gsd'] %>% round(1)` in stool samples and `r hmp_div_stats['vagina', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['vagina', 'gsd'] %>% round(1)` in vaginal samples).
+To assess the potential importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the multiplicative variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth.
+Across simulation replicates, the GSD in the mean efficiency was typically lower in stool than in vaginal samples (SI Figure \@ref(fig:gut)B; ratio of GSD in vaginal samples to GSD in stool samples of `r hmp_me_stats['iid', 'gm'] %>% round(1)` $\md$ `r hmp_me_stats['iid', 'gsd'] %>% round(1)` across 1000 simulations).
+Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \@ref(fig:gut)C), which the GSD in the proportion of a gut species across gut samples an average of `r hmp_species_diff %>% round(2)`-fold lower than that of a vaginal species across vaginal samples.
+Recall that the importance of bias estimation of the FC in a species' proportion depends on GSD in the mean efficiency relative the GSD in the species' proportion (Section \@ref(differential-abundance)).
+Thus these results suggest that although the mean efficiency likely varies less across stool than vaginal samples, bias may be just as problematic for proportion-based DA analyses.
 
@@
 ```{r}
-x <- here(
+x <- fs::path(
   'notebook/_posts/2022-01-08-lloyd2020evi-case-study/_output/table1-tidy.rds'

### 2022-03-14 notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/_cut.Rmd
Revise gut case study
- Revise Rmd to completion; add results
- Add in final numbers
- Revise discussion of diversity vs. mean-efficiency variation, and add
  derivation to appendix
- Add SI Figure and caption

@@
 Check to see if I can get these, and use them for the analysis.
+
+### What are the drivers of mean efficiency variation?
+
+Hypothesis: The most abundant species in each sample can mostly explain the variation.
+
+```{r, eval = F}
+# zero out each sample to just the top n taxa, then compute the mean efficiency
+# (without closing out? or closing out?)
+mean_efficiency_top <- function(x, bias, type, n = 1) {
+
+}
+```
+
+```{r, eval = F}
+reps_me1 <- reps %>%
+  mutate(
+    mean_efficiency = map(bias, ~mean_efficiency(ps_gv, .x, type = 'actual')),
+    across(mean_efficiency, map, enframe, '.sample', 'mean_efficiency')
+    mean_efficiency_top = map(bias, ~mean_efficiency_top(ps_gv, .x, type = 'actual')),
+    across(mean_efficiency_top, map, enframe, '.sample', 'mean_efficiency_top')
+  ) %>%
+  select(-bias) %>%
+  unnest(c(mean_efficiency, mean_efficiency_top)) %>%
+  left_join(ps_gv %>% sample_data %>% as_tibble, by = '.sample')
+```
+
+# Variation in species
+
+We also need to choose what species to consider as candidates for a DA analysis in the given ecosystem.
+
+Let's get a lightly filtered ps object for each type, and check out the prevalence patterns.
+
+```{r}
+ps_list <- list(
+  stool = ps_gv %>% filter_sample_data(body_site == 'stool'),
+  vagina = ps_gv %>% filter_sample_data(body_site == 'vagina')
+) %>%
+  map(filter_taxa2, ~ sum(. > 0) >= 2) %>%
+  map(orient_taxa, as = 'cols')
+ps_list %>% map_int(nsamples)
+ps_list %>% map_int(ntaxa)
+```
+
+```{r, eval = T}
+x <- ps_list %>%
+  map_dfr(as_tibble)
+
+min_vals <- c(0, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1)
+
+prev <- x %>%
+  with_groups(c(.otu, body_site), summarize,
+    min = min_vals,
+    prevalence = map_dbl(min_vals, ~mean(.abundance > .x))
+  )
+```
+
+```{r}
+prev %>%
+  ggplot(aes(x = prevalence, y = as.factor(min), 
+      fill = body_site, color = body_site)) +
+  scale_x_log10() +
+  geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
+```
+
+It's a bit hard to know what threshold to use to pick a relevant set of taxa.
+
+```{r}
+prev %>%
+  filter(min < 1e-3) %>%
+  ggplot(aes(x = prevalence, y = body_site,
+      fill = body_site, color = body_site)) +
+  facet_wrap(~min) +
+  geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
+```

### 2022-03-14 notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/hmp-stool-vagina-comparison.Rmd
Revise gut case study
- Revise Rmd to completion; add results
- Add in final numbers
- Revise discussion of diversity vs. mean-efficiency variation, and add
  derivation to appendix
- Add SI Figure and caption

@@
 title: "Compare impact of bias in HMP stool and vagina samples"
-description: |
+description: This report compares gut and vaginal microbiome profiles from the Human Microbiome Project to consider how the different ecological dynamics in each sample type might modulate the potential for bias to distort FCs in species proportions.
 author:
@@
     url: {}
-categories:
-  - R
 date: 2022-01-30
 draft: false
+bibliography: ../../../main.bib
 output:
@@
     dev: svg
+    toc: true
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(
+  echo = TRUE,
+  cache = TRUE,
+  autodep = TRUE,
+  cache.comments = FALSE,
+  dpi = 300
+)
+```
+
+# Setup
 
@@
 library(cowplot)
+library(patchwork)
 theme_set(theme_cowplot())
@@
 library(ggridges)
+
+library(fs)
+dir_create('_output')
 ```
 
-HERE - refactor to match the manuscript, and add the analysis of variation in species proportions relative to mean efficiency, and also look at the bimodality/presence/absence nature of species variation.
+First, I load a phyloseq object with the Metaphlan3 profiles from the curatedMetagenomicData R package.
+This phyloseq object was created in `notebook/_code/import-gut-datasets.Rmd`.
+Filtering: I filter to just Bacteria, drop samples with no bacterial abundance, and drop species that don't appear in at least two samples.
+I then renormalize to proportions.
+I also simplify the OTU names to just the species name.
 
-# HMP1
+Then, I compute the most abundant species and its proportion, and add these to the sample data.
+In addition, I add the plug-in estimates of three alpha-diversity metrics to the sample data.
+These are the hill numbers of order 0, 1, and 2, equivalent to richness, the exponential of Shannon entropy, and Inverse Simpson index, respectively.
 
@@
   prune_samples(sample_sums(.) > 0, .) %>% # to drop samples w/o bacteria %>%
+  filter_taxa2(~sum(.>0) >= 2) %>%
   transform_sample_counts(close_elts) %>%
@@
 
-Let's see how things vary by body site
+Let's check the breakdown of samples by body site and subsite,
 
@@
 sam <- ps %>% sample_data %>% as_tibble
-sam %>% count(body_site)
-sam %>% count(body_site, body_subsite)
+sam %>% count(body_site) %>% knitr::kable()
+sam %>% count(body_site, body_subsite) %>% knitr::kable()
 ```
 
-## diversity
+Since only the posterior fornix of the vaginal samples has more than a few samples, we'll restrict to just that subtype in what follows, along with the stool type/subtype.
 
 ```{r}
+ps_gv <- ps %>%
+  filter_sample_data(body_subsite %in% c('stool', 'posterior_fornix')) %>%
+  filter_taxa2(~ sum(.) > 0)
+sam <- ps_gv %>% sample_data %>% as_tibble
+```
+
+# Exploratory plots and stats
+
+Distribution of alpha diversity across samples
+
+```{r diversity, fig.dim = c(6,3.5)}
 sam %>%
-  pivot_longer(starts_with('diversity'), names_to = 'index') %>%
+  pivot_longer(starts_with('diversity')) %>%
+  mutate(
+    q = str_extract(name, '[0-2]'),
+    index = str_glue('q = {q}')
+  ) %>%
   ggplot(aes(y = body_site, x = value)) +
@@
   facet_grid(.~index, scales = 'free', space = 'free') +
-  geom_boxplot()
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  geom_density_ridges(stat = 'binline', scale = 0.90) +
+  labs(y = 'Body site', x = 'Diversity')
 ```
 
-```{r}
+Distribution of the proportion of the most abundant species in each sample,
+
+```{r most-abundant, fig.dim = c(5, 3.5)}
 sam %>%
-  ggplot(aes(top_species_prop)) +
-  facet_wrap(~body_site, ncol = 1) +
+  ggplot(aes(y = body_site, top_species_prop)) +
   scale_x_log10() +
-  geom_histogram() +
-  labs(x = 'Proportion')
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  expand_limits(x = 0.1) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  geom_density_ridges(stat = 'binline', scale = 1.90) +
+  labs(x = 'Proportion of most abundant species', y = 'Body site')
 ```
 
+Which species tend to be most abundant?
+
 ```{r}
-sam %>%
-  with_groups(body_site, count, top_species) %>%
-  with_groups(body_site, mutate, frac = close_elts(n)) %>%
-  with_groups(body_site, slice_max, n, n = 5) %>%
-  print(n=Inf)
+make_table <- function(x) {
+  x %>%
+    mutate(
+      across(top_species, fct_infreq),
+      across(top_species, fct_lump_n, n = 5),
+    ) %>%
+    count(top_species) %>%
+    mutate(frac = close_elts(n)) %>%
+    knitr::kable(digits = 2)
+}
+
+# Vagina
+sam %>% filter(body_site == 'vagina') %>% make_table
+
+# Stool
+sam %>% filter(body_site == 'stool') %>% make_table
 ```
 
-```{r}
-focal_subsites <- c('stool', 'posterior_fornix')
-sam %>%
-  filter(body_subsite %in% focal_subsites) %>%
-  with_groups(body_site, count, top_species) %>%
-  with_groups(body_site, mutate, frac = close_elts(n)) %>%
-  with_groups(body_site, slice_max, n, n = 10) %>%
-  print(n=Inf)
-```
+In both cases, a small number of closely related species dominant most samples.
+Notably, we do not see Gardnerella or Lactobacillus BVAB1 in the vaginal samples, or any Firmicutes or other non-Bacteroidetes in the gut samples.
+It is unclear whether this is due to the cohorts or bias in the protocol.
 
-What about total diversity across samples (gamma diversity)?
-Check that this is much larger in the gut than the vagina.
+## Breakdown of total diversity (gamma) into alpha and beta components
 
-```{r}
-
-```
-
-### gamma diversity // diversity partition
-
-Can use the hillR package to compute alpha, beta, and gamma diversity as Hill numbers (effective number of species) for q = 0 (richness), q = 1 (exponential of Shannon entropy), and q = 2 (Inverse Simpson index).
-
-Note: q=0 gamma diversity is will be fairly sensitive to increasing the number of samples, and so we'd expect it to be larger for stool simply due to the larger number of samples, unless we subsample.
-
+This is more experimental.
+We can use the hillR package to compute alpha, beta, and gamma diversity as Hill numbers (effective number of species) for q = 0 (richness), q = 1 (exponential of Shannon entropy), and q = 2 (Inverse Simpson index).
+Note: Gamma diversity for q=0 (maybe also for other q) will increase with the number of samples, and so we'd expect it to be larger for stool simply due to the larger number of samples unless we subsample to a fixed number of samples for each body site.
 q=0 is also sensitive to read depth, which we don't really have a way to account for here.
-
-#### testing
-
-```{r}
-get_site_ps <- function(body_site, ps, n ) {
-  filter_sample_data(ps, body_site == !!body_site) %>% 
-    filter_taxa2(~sum(.) > 0) %>%
-    orient_taxa(as = 'cols')
-}
-div <- crossing(q = c(0, 1, 2), body_site = c('stool', 'vagina')) %>%
-  mutate(
-    ps = map(body_site, get_site_ps, ps = ps),
-    otu = map(ps, otu_table),
-    div = map2(otu, q, hillR::hill_taxa_parti),
-    n = map_int(ps, nsamples)
-  ) %>%
-  select(-q, -otu) %>%
-  unnest(div)
-div %>% select(body_site, q, starts_with('TD'), n) %>% knitr::kable(digits = 1)
-```
-
-Can see that total (gamma) diversity and average alpha diversity are much larger in the gut than in hthe vagina, as expected, for each index.
-
-The average q=2 alpha diversity in the stool is in fact quite low (only 5.4) - check on that, since it doesn't quite match what I remember from the figures.
-
-Let's try the same thing with phylogeny diversity
-
-```{r}
-div_phy <- crossing(q = c(0, 1, 2), body_site = c('stool', 'vagina')) %>%
-  mutate(
-    ps = map(body_site, get_site_ps, ps = ps),
-    comm = map(ps, otu_table),
-    tree = map(ps, phy_tree),
-    n = map_int(ps, nsamples)
-  ) %>%
-  rowwise %>%
-  mutate(
-    hillR::hill_phylo_parti(comm, tree, q)
-  ) %>%
-  select(-ps, -comm, -tree) %>%
-  ungroup
-div_phy %>% select(body_site, q, starts_with('PD'), n) %>% knitr::kable(digits = 1)
-```
-
-NOTES/TODO
-
-- check the def of gamma diversity; does it control for the difference in the number of samples from each body-site type? I think not, and so gut will be larger simply b/c of more samples
-- consider filtering down to a final set of samples; E.g. Perhaps just the one vaginal sample type, and some quick QC filtering; Though this shouldn't make a huge difference.
-- filter taxa? or treat all as true positives
-
-I also need to read about these diversity metrics.
-
-#### better
-
-GOAL: For an equal number of gut and vaginal samples, compute taxa and phylo gamma/alpha/beta diversity.
-Do for q=0,1,2, though we can't really trust the q=0 numbers.
+Since we have a phylogeny, we can compute phylogenetic diversity as well as species diversity.
 
@@
 ps_list <- list(
-  stool = ps %>% filter_sample_data(body_subsite == 'stool'),
-  vagina = ps %>% filter_sample_data(body_subsite == 'posterior_fornix')
+  stool = ps_gv %>% filter_sample_data(body_site == 'stool'),
+  vagina = ps_gv %>% filter_sample_data(body_site == 'vagina')
 ) %>%
@@
 ```
+
+Then compute the diversity breakdown for each body site,
 
@@
   pivot_longer(starts_with(c('TD', 'PD'))) %>%
-  pivot_wider(names_from = body_site)
+  pivot_wider(names_from = body_site) %>% 
+  knitr::kable(digits = 1)
 ```
 
-## Bias simulation
+## Order-2 diversity
 
-TODO: MAYBE: change so that we filter to the ecosystem-specific taxa before scaling the mean efficiencies
-This mainly matters for the phylogenetic sim.
+What is the GM, GSD, and GSE of order-2 diversity in each body site?
 
 ```{r}
-focal_subsites <- c('stool', 'posterior_fornix')
-ps_sim <- ps %>%
-  filter_sample_data(body_subsite %in% focal_subsites) %>%
-  filter_taxa2(~sum(.>0) >= 2) %>%
-  prune_samples(sample_sums(.) > 0, .) %>% # to drop samples w/o bacteria %>%
-  transform_sample_counts(close_elts) %>%
-  mutate_tax_table(
-    gram = case_when(
-      phylum %in% c('Actinobacteria', 'Firmicutes') ~ 'Positive',
-      TRUE ~ 'Negative'
+div_q2_stats <- sam %>%
+  with_groups(body_site, summarize, 
+    gm = gm_mean(diversity_q2),
+    gsd = gm_sd(diversity_q2),
+    gse = gsd ^ (1 / sqrt(n()))
   )
-  )
+div_q2_stats %>%
+  knitr::kable(digits = 2)
 ```
 
-Now, generate R bias vectors of each type, and calibrate the data by them.
+We'll save this for use in the manuscript,
+
+```{r}
+write_csv(div_q2_stats, path('_output', 'div_q2_stats.csv'))
+```
+
+# Bias simulation
+
+To assess the relative importance of bias for proportion-based DA analyses in the two ecosystems, we will consider the variation in the mean efficiency across samples for a large number of possible taxonomic biases, under the assumption that the measured profiles reflected the truth.
+
+The following function samples random bias vectors from a multivariate lognormal distribution with a given covariance structure and geometric standard deviation across the efficiencies.
 
@@
   n_taxa <- nrow(vcv)
-  gsd_log <- log(gsd)
   bias <- MASS::mvrnorm(n, mu = rep(0, n_taxa), Sigma = vcv) %>%
@@
     scale %>%
-    {. * gsd_log} %>%
+    {. * log(gsd)} %>%
     exp
@@
 
-```{r, eval = F}
-set.seed(42)
-R <- 300
-gsd <- 3
-reps <- list(
-  iid = diag(ntaxa(ps_sim)),
-  vcv = ps_sim %>% phy_tree %>% ape::vcv()
-  #> gram = gram_vcv
-) %>%
-  map_dfr(.id = 'sim_type',
-    ~sample_bias(n = R, gsd = gsd, vcv = .x) %>%
-      split(., row(.)) %>%
-      enframe(".idx", "bias")
-  ) %>%
-  mutate(
-    ps = map(bias, ~calibrate(ps_sim, .x))
-  )
-  # TODO: cache this, since it is a bit slow (due to large number of samples)
-reps_me <- reps %>%
-  mutate(
-    ps_sam = map(ps, ~sample_data(.) %>% as_tibble)
-  ) %>%
-  select(-bias, -ps) %>%
-  unnest(ps_sam)
-reps_me_summ <- reps_me %>%
-  with_groups(c(sim_type, body_site, .idx), summarize,
-    across(.mean_efficiency, gm_sd)
-  )
+To choose a GSD for the simulations, let's examine the bias estimated by @mclaren2019cons for the @costea2017towa Phase 2 experimental protocols,
+
+```{r}
+bias_costea <- '_data/mclaren2019cons-table2-mod.tsv' %>% read_tsv
+bias_costea_gsd <- bias_costea %>%
+  summarize(across(-Taxon, gm_sd))
+bias_costea_gsd %>%
+  knitr::kable(digits = 1)
 ```
 
-NOTE: The var in mean eff and var in species props might be more directly comparable if we treat the observed proportions as the truth, rather than observed.
+The strength of bias as quantified by the GSD in species efficiencies is much larger for protocol H than the other two protocols.
+The GM of the GSD of the three protocols is `r bias_costea_gsd %>% select(H, Q, W) %>% unlist %>% gm_mean %>% round(2)`.
+Let's round this to a simple whole number 4 and take this as the GSD for our simulations.
 
-now, for each phyloseq object, compute the variation in the log mean efficiency
+For the IID and phylogenetic-covariance models, simulate 1000 replicate bias vectors.
 
@@
 set.seed(42)
-R <- 300
-gsd_sim <- 3
+R <- 1000
+gsd_sim <- 4
 reps <- list(
-  iid = diag(ntaxa(ps_sim)),
-  vcv = ps_sim %>% phy_tree %>% ape::vcv()
+  iid = diag(ntaxa(ps_gv)),
+  vcv = ps_gv %>% phy_tree %>% ape::vcv()
   #> gram = gram_vcv
@@
   )
+```
+
+For each replicate, compute the mean efficiency of each sample, assuming that the observed profiles are the truth and the simulated vector is the bias.
+Summarize the variation in the mean efficiency for the replicate by the GSD.
+
+```{r}
 reps_me <- reps %>%
   mutate(
-    mean_efficiency = map(bias, ~mean_efficiency(ps_sim, .x, type = 'actual')),
+    mean_efficiency = map(bias, ~mean_efficiency(ps_gv, .x, type = 'actual')),
     across(mean_efficiency, map, enframe, '.sample', 'mean_efficiency')
@@
   unnest(mean_efficiency) %>%
-  left_join(ps_sim %>% sample_data %>% as_tibble, by = '.sample')
+  left_join(ps_gv %>% sample_data %>% as_tibble, by = '.sample')
 reps_me_summ <- reps_me %>%
@@
 
+```{r, include = F, eval = F}
+reps_me_summ %>%
+  ggplot(aes(y = sim_type, x = mean_efficiency)) +
+  facet_wrap(~body_site, ncol = 1) +
+  scale_x_log10() +
+  geom_vline(xintercept = c(1, gsd_sim), color = 'grey') +
+  geom_violin(draw_quantiles = c(0.5)) +
+  # geom_boxplot(outlier.shape = NA) +
+  # geom_quasirandom(groupOnX = FALSE, alpha = 0.2) +
+  labs(x = 'GSD of mean efficiency', y = 'Simulation type')
+```
+
 ```{r}
@@
   facet_wrap(~body_site, ncol = 1) +
-  #> theme_minimal_hgrid() +
   scale_x_log10() +
-  #> expand_limits(x = c(1, 3)) +
-  geom_vline(xintercept = c(1,3), color = 'grey') +
-  geom_boxplot(outlier.shape = NA) +
-  geom_quasirandom(groupOnX = FALSE, alpha = 0.5) +
-  labs(x = 'gm. std. of mean efficiency')
-  # stat_slab(slab_type = 'histogram', breaks = 20)
+  geom_vline(xintercept = c(1,gsd_sim), color = 'grey') +
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  stat_density_ridges(quantile_lines = TRUE, quantiles = c(0.5), scale = 0.95) +
+  labs(x = 'GSD of mean efficiency', y = 'Simulation type')
 ```
 
-Interesting - note that forcing a gm-sd of 3 across all species can allow for a larger variation in the mean efficiency in some cases, in the vaginal communities.
-, actually leads to a wider variation among vaginal species.
+Note, despite always forcing a GSD of 3 on the efficiencies, we sometimes see GSDs in the mean efficiency that are larger than 3, particularly in the vaginal communities.
 This makes some sense: We expect the variation in the efficiency of vaginal communities to be driven by a just a few species, and the there is a non-negligible chance that they end up with a much higher/lower mean efficiency than typical.
 
-NOTE: I think the mean efficiency should scale with whatever we set the GSD to.
-So we might want to set it to 1 and label the units as being per GSD.
+Question: How does the mean efficiency GSD scale with the efficiency GSD?
+When we compare this variation to that in the species proportions, the scale/value of GSD will matter.
+However, it might not matter for comparing to the vaginal MB.
 
-When we compare this variation to that in the species proportions, the scale/value of GSD will matter.
-However, it doesn't if our main purpose is comparing to the vaginal MB.
-And if we label in units of GSD, it will be easy to read off from the graph what the correct mean eff GSD is for a given GSD.
-
-How much more is the var in the vaginal comms versus stool?
-
-```{r}
-reps_me_summ %>%
-  ggplot(aes(x = mean_efficiency, y = sim_type, color = body_site)) +
-  scale_x_log10() +
-  stat_summary() +
-  expand_limits(x = 1)
-```
+How much greater is the GM of the GSD in the vaginal comms versus stool?
 
@@
   pivot_wider(names_from = body_site, values_from = mean_efficiency) %>%
-  mutate(vagina/stool)
+  mutate(vagina/stool) %>%
+  knitr::kable(digits = 2)
 ```
 
-## Variation in species
-
-What else matters? Perhaps the variation in a species, relative to the variation in the mean efficiency.
-
-For sufficiently prevalent taxa, look at the variation.
-
-check: what do the min positive abundances look like for various species?
+Using the median instead of the GM should give a similar answer,
 
 ```{r}
-x <- ps %>%
-  filter_sample_data(body_subsite %in% c('stool', 'posterior_fornix')) %>%
-  filter_taxa2(~ sum(.) > 0) %>%
+reps_me_summ %>%
+  with_groups(c(body_site, sim_type), summarize, 
+    across(mean_efficiency, median)) %>%
+  pivot_wider(names_from = body_site, values_from = mean_efficiency) %>%
+  mutate(vagina/stool) %>%
+  knitr::kable(digits = 2)
+```
+
+Note that the type of simulation has a negligible impact on these numbers.
+
+Let's compute the GM, GSD, and GSE of the GM for the difference (ratio) between gut and vaginal samples.
+
+```{r}
+reps_me_summ %>%
+  with_groups(c(body_site, sim_type), summarize, 
+    GM = gm_mean(mean_efficiency),
+    GSD = gm_sd(mean_efficiency),
+    # GSE = exp(sd(log(mean_efficiency)) / sqrt(n())),
+    GSE = GSD ^ (1 / sqrt(n()))
+  ) %>%
+  knitr::kable(digits = 4)
+```
+
+And if the difference,
+
+```{r}
+reps_me_summ_diff <- reps_me_summ %>%
+  pivot_wider(names_from = body_site, values_from = mean_efficiency) %>%
+  mutate(ratio = vagina / stool) %>%
+  with_groups(c(sim_type), summarize, 
+    gm = gm_mean(ratio),
+    gsd = gm_sd(ratio),
+    gse = gsd ^ (1 / sqrt(n()))
+  )
+reps_me_summ_diff %>%
+  knitr::kable(digits = 3)
+```
+
+```{r}
+write_csv(reps_me_summ_diff , path('_output', 'reps_me_summ_diff.csv'))
+```
+
+Let's also check the distribution of the ratio (geometric difference),
+
+```{r}
+reps_me_summ %>%
+  pivot_wider(names_from = body_site, values_from = mean_efficiency) %>%
+  mutate(ratio = vagina / stool) %>%
+  ggplot(aes(y = sim_type, x = ratio)) +
+  scale_x_log10() +
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  stat_density_ridges(quantile_lines = TRUE, quantiles = c(0.5), scale = 0.95) +
+  labs(y = 'Simulation type', x = 'GSD in vagina / GSD in stool')
+```
+
+Note that there is a decent fraction of simulations in which the variation in the mean efficiency is lower in the vaginal samples, despite it typically being higher.
+
+The large number of simulations (1000) makes the GSE very small; however, this number does not account for the limited number of gut and vaginal samples.
+In other words, we have precisely estimated the difference for these particular samples.
+
+# Variation in species
+
+For analysis of LFCs in proportions, what matters is the variation in the species proportions relative to the mean efficiency.
+Therefore we should also consider whether the species proportions vary less in the more diverse gut.
+To do so, I will measure the GSD in the (zero-replaced) proportion of each species within the given sample type that passes a prevalence filter within that sample type.
+
+## Preliminary checks
+
+To figure out how to replace zeros, let's check what the min positive abundances look like for various species.
+
+```{r}
+x <- ps_gv %>%
   as_tibble
 ```
+
 ```{r}
@@
     min_pos = min(.abundance)) %>%
-  ggplot(aes(x = min_pos, y = body_site, fill = body_site)) +
+  ggplot(aes(x = min_pos, y = body_site)) +
   scale_x_log10() +
+  labs(x = 'Min. positive proportion of the species', y = 'Body site') +
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
   geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
@@
 
-how about by sample?
+What about the min positive proportion in the sample?
 
@@
     min_pos = min(.abundance)) %>%
-  ggplot(aes(x = min_pos, y = body_site, fill = body_site)) +
+  ggplot(aes(x = min_pos, y = body_site)) +
   scale_x_log10() +
+  labs(x = 'Min. positive proportion of the sample', y = 'Body site') +
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
   geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
@@
 
-get a filtered ps object for each type
+We also need to choose what species to consider as candidates for a DA analysis in the given ecosystem.
+For simplicity, let's use the filtering criterion we used in the MOMSPI diversity regression.
+We'll do this filtering and the subsequent analysis separately within each sample type.
 
@@
 ps_list <- list(
-  stool = ps %>% filter_sample_data(body_subsite == 'stool'),
-  vagina = ps %>% filter_sample_data(body_subsite == 'posterior_fornix')
+  stool = ps_gv %>% filter_sample_data(body_site == 'stool'),
+  vagina = ps_gv %>% filter_sample_data(body_site == 'vagina')
 ) %>%
-  # TODO: consider filtering criterion
-  map(filter_taxa2, ~ sum(. > 0) >= 2) %>%
-  # map(filter_taxa2, ~ mean(. > 0) >= 0.05) %>%
-  # map(filter_taxa2, ~ mean(. >= 1e-4) >= 0.10) %>%
-  # map(filter_taxa2, ~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
+  map(filter_taxa2, ~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
   map(orient_taxa, as = 'cols')
@@
 
-### presence/absence
+Note that there are many more species passing this criterion in stool.
 
-```{r, eval = F}
-x <- ps %>%
-  filter_sample_data(body_subsite %in% c('stool', 'posterior_fornix')) %>%
-  filter_taxa2(~ sum(.) > 0) %>%
-  as_tibble
-```
-
-```{r, eval = T}
-x <- ps_list %>%
-  map_dfr(as_tibble)
-
-min_vals <- c(0, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1)
-
-prev <- x %>%
-  with_groups(c(.otu, body_site), summarize,
-    min = min_vals,
-    prevalence = map_dbl(min_vals, ~mean(.abundance > .x))
-  )
-```
-```{r}
-prev %>%
-  ggplot(aes(x = prevalence, y = as.factor(min), 
-      fill = body_site, color = body_site)) +
-  geom_density_ridges(alpha = 0.3)
-```
+I will replace zeros by adding a small pseudo-value to all proportions.
+I expect the choice of pseudo-value will have a large impact on the GSDs of a species, so I will try a wide range of values.
+Note, I am not renormalizing the proportions after adding the psuedo-value, since we've already filtered out many species.
+However, the renormalization would involve dividing all proprotions by the same factor and so would not affect the resulting GSDs.
+(This factor is equal to $1 + k\episilon$, where $k$ is the number of pre-filtered species and $\epsilon$ is the pseudo-value.
+Since $\episilon \ll 1/k$ except perhaps for the highest $\epsilon = 10^{-3}$, this factor is generally close to 1 anyways.)
 
 ```{r}
-prev %>%
-  ggplot(aes(x = prevalence, y = as.factor(min), 
-      fill = body_site, color = body_site)) +
-  geom_density_ridges(stat = 'binline', alpha = 0.3)
-```
-
-It's a bit hard to know what threshold to use to pick a relevant set of taxa.
-
-```{r}
-prev %>%
-  filter(min < 1e-3) %>%
-  ggplot(aes(x = prevalence, y = body_site,
-      fill = body_site, color = body_site)) +
-  facet_wrap(~min) +
-  geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
-```
-
-### multiplicative variation
-
-```{r}
-ps_list_filt <- ps_list %>%
-  map(filter_taxa2, ~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15)
-ps_list_filt %>% map_int(ntaxa)
-```
-
-now, purpose is to get the geometric variation in species.
-
-Problem: I expect the pseudocount will have a massive impact here, so should check different values.
-
-```{r}
-x <- ps_list_filt %>%
-  map_dfr(as_tibble)
-
 min_vals <- c(1e-6, 1e-5, 1e-4, 1e-3)
 
-gsd <- x %>%
+gsd <- ps_list %>%
+  map_dfr(as_tibble) %>%
   with_groups(c(.otu, body_site), summarize,
@@
   expand_limits(x = 1) +
-  geom_density_ridges(stat = 'binline', alpha = 0.3, scale = 0.95)
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  geom_density_ridges(stat = 'binline', scale = 0.95, alpha = 0.3)
 ```
 
-Note, the GSDs tend to be larger in the vaginal microbiome, consistent with our expectation based on its lower alpha diversity.
+The GSDs tend to be larger in the vaginal microbiome, regardless of which min value is used.
 
 What is the difference between gut and vaginal GSDs?
+Determine by taking the geometric mean (GM) of the GSDs of various species.
 
@@
       fill = body_site, color = body_site)) +
-  # facet_wrap(~body_site) +
   scale_x_log10() +
@@
 ```{r}
-gsd %>%
+species_gsd_stats <- gsd %>%
   with_groups(c(body_site, min), summarize, across(gsd, gm_mean)) %>%
@@
   mutate(vagina/stool)
+species_gsd_stats 
 ```
 
-The difference is not very dependent on the min abundance threshold.
+Note, the difference between body sites has hardly any dependence on the pseudo-value.
 
-### visualize
-
-```{r, eval = F}
-ps_list[[2]] %>%
-  as_tibble %>%
-  ggplot(aes(y = species, x = 1e-5 + .abundance)) +
-  scale_x_log10() +
-  # stat_slab(slab_type = 'histogram')
-  stat_dots()
+```{r}
+species_gsd_stats %>%
+  janitor::clean_names() %>%
+  write_csv(path('_output', 'species_gsd_stats.csv'))
 ```
 
-```{r, eval = F}
-ps_list[[2]] %>%
-  as_tibble %>%
-  ggplot(aes(x = 1e-5 + .abundance)) +
-  facet_wrap(~species, ncol = 1) +
-  scale_y_log10() +
+# Final figures and numbers
+
+```{r main, fig.dim = c(6,8)*1.1}
+# Alpha diversity
+p_div <- sam %>%
+  ggplot(aes(y = body_site, x = diversity_q2)) +
   scale_x_log10() +
-  # stat_slab(slab_type = 'histogram')
-  geom_histogram()
+  geom_boxplot(outlier.shape = NA) +
+  geom_quasirandom(groupOnX = FALSE, alpha = 0.75) +
+  # theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  # scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  # geom_density_ridges(stat = 'binline', scale = 0.90) +
+  labs(y = 'Body site', x = 'Order-2 diversity')
+# p_div
+
+# Variation in species proportions
+p_species_gsd <- gsd %>%
+  filter(min == 1e-4) %>%
+  ggplot(aes(x = gsd, y = body_site)) +
+  labs(y = 'Body site', x = "GSD[proportion]") +
+  scale_x_log10() +
+  expand_limits(x = 1) +
+  # geom_vline(xintercept = c(1,gsd_sim), color = 'grey') +
+  geom_boxplot(outlier.shape = NA) +
+  geom_quasirandom(groupOnX = FALSE, alpha = 0.75)
+  # theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  # scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  # geom_density_ridges(stat = 'binline', scale = 0.90)
+
+# GSD in mean eff in simulations
+p_me_gsd <- reps_me_summ %>%
+  filter(sim_type == 'iid') %>%
+  ggplot(aes(y = body_site, x = mean_efficiency)) +
+  scale_x_log10() +
+  # expand_limits(x = 0.85) +
+  geom_vline(xintercept = c(gsd_sim), color = 'grey') +
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  stat_density_ridges(quantile_lines = TRUE, quantiles = c(0.5), scale = 0.95) +
+  labs(x = 'GSD[mean efficiency]', y = 'Body site')
+# p_me_gsd
+
+(p_div + ggtitle('Within-sample diversity')) /
+  (p_me_gsd + ggtitle("Cross-sample variation in the mean efficency")) /
+  (p_species_gsd + ggtitle("Cross-sample variation in species' proportions")) +
+  plot_annotation(tag_levels = 'A') +
+  plot_layout() &
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE)
+
 ```
 
-```{r, eval = F}
-ps_list1[[1]] %>%
-  as_tibble %>%
-  ggplot(aes(x = 1e-5 + .abundance)) +
-  facet_wrap(~species, ncol = 3) +
-  # scale_y_log10() +
-  scale_x_log10() +
-  # stat_slab(slab_type = 'histogram')
-  geom_histogram()
+We can also try putting the GSD in proportions and in mean efficiency on a common scale,
+
+```{r, fig.dim = c(6,4)}
+# Variation in species proportions
+p_species_gsd <- gsd %>%
+  filter(min == 1e-4) %>%
+  ggplot(aes(x = gsd, y = body_site)) +
+  labs(y = 'Body site', x = "GSD[proportion]") +
+  scale_x_log10(limits = c(0.9, 200)) +
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  geom_vline(xintercept = c(1,gsd_sim), color = 'grey') +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  # geom_density_ridges(stat = 'binline', scale = 0.90)
+  geom_density_ridges(jittered_points = TRUE, scale = 0.95)
+
+# GSD in mean eff in simulations
+p_me_gsd <- reps_me_summ %>%
+  filter(sim_type == 'iid') %>%
+  ggplot(aes(y = body_site, x = mean_efficiency)) +
+  scale_x_log10(limits = c(0.9, 200)) +
+  geom_vline(xintercept = c(1,gsd_sim), color = 'grey') +
+  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
+  scale_y_discrete(expand = expansion(mult = c(0.01, .7))) +
+  stat_density_ridges(quantile_lines = TRUE, quantiles = c(0.5), scale = 0.95) +
+  labs(x = 'GSD[mean efficiency]', y = 'Body site')
+# p_me_gsd
+
+p_species_gsd / p_me_gsd
 ```
 
-```{r}
+This plot makes it clear that the GSD in mean efficiency is generally smaller than that in the species proportions, suggesting that for the strength of bias in these simulations (GSD=4), that bias is unlikely to have a large impact.
+However, the GSD in species proportions is highly dependent on the pseudocount; also, our choice of GSD=4 is simply for illustration.
+Thus we should not overinterpret this result.
 
-ps_list1[[1]] %>%
-  as_tibble %>%
+# Session info {.appendix}
+
+<details><summary>Click for session info</summary>
+```{r, R.options = list(width = 83)}
+sessioninfo::session_info()
 ```
-
-HERE
-
-## Gut: Variation in phylum proportions
-
-If phyla vary substantially across samples, then any phylum-specific bias would drive variation in the mean efficiency.
-
-I'll restrict plots to just the four main phyla.
-
-```{r}
-focal_phyla <- c('Actinobacteria', 'Bacteroidetes', 'Firmicutes',
-  'Proteobacteria')
-ps_phy <- ps %>% 
-  subset_samples(body_subsite == 'stool') %>%
-  tax_glom('phylum') %>%
-  mutate_tax_table(.otu = phylum) %>%
-  transform_sample_counts(close_elts) %>%
-  mutate_tax_table(
-    gram = case_when(
-      phylum %in% c('Actinobacteria', 'Firmicutes') ~ 'Positive',
-      TRUE ~ 'Negative'
-    )
-  )
-x_phy <- ps_phy %>% as_tibble
-```
-
-```{r}
-x_phy %>%
-  filter(phylum %in% focal_phyla) %>%
-  ggplot(aes(.abundance + 1e-5)) +
-  facet_wrap(~phylum, scales = 'fixed', ncol = 1) +
-  scale_x_log10() +
-  geom_histogram()
-```
-
-There is substantial log-scale variation in all phyla, but the Bacteroidetes and Firmicutes are generally much more abundant than the other two.
-
-What about the B:F ratio?
-
-```{r}
-x_phy %>%
-  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
-  mutate(
-    across(.abundance, ~ . + 1e-4),
-    ratio = .abundance / .abundance[phylum == 'Firmicutes']
-  ) %>%
-  filter(phylum == 'Bacteroidetes') %>%
-  ggplot(aes(ratio)) +
-  scale_x_log10(breaks = 10^(-3:4)) +
-  geom_histogram() +
-  theme_minimal_vgrid()
-```
-
-Suppose that these profiles are biased to favor Gram negative phyla.
-
-```{r}
-bias <- ps_phy %>% tax_table %>% as_tibble %>%
-  mutate(
-    efficiency = case_when(
-      gram == 'Positive' ~ 1/10,
-      TRUE ~ 1
-    )
-  ) %>%
-  select(.otu, efficiency) %>%
-  deframe
-ps_phy_cal <- ps_phy %>% calibrate(bias)
-```
-
-```{r}
-ps_phy_cal %>% 
-  sample_data %>%
-  as_tibble %>%
-  ggplot(aes(.mean_efficiency)) +
-  scale_x_log10() +
-  geom_histogram()
-```
-
-```{r}
-bias %>% gm_sd
-ps_phy_cal %>% sample_data %>% as_tibble %>% pull(.mean_efficiency) %>% gm_sd
-```
-
-Consider adjusting so that the GSD is 3, like in other simulations.
-(Though should then stick to the same taxa).
+</details>

### 2022-03-14 supplemental-figures.Rmd
Revise gut case study
- Revise Rmd to completion; add results
- Add in final numbers
- Revise discussion of diversity vs. mean-efficiency variation, and add
  derivation to appendix
- Add SI Figure and caption

@@
 
+<!-- Case studies: Gut -->
+
+```{r gut, fig.cap = '(ref:cap-gut)'}
+fs::path(
+  "notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/",
+  "hmp-stool-vagina-comparison_files",
+  "figure-html5/main-1.svg"
+) %>%
+  include_svg
+```
+
+(ref:cap-gut) **Compared to vaginal samples, stool samples have higher diversity, lower variation in mean efficiency, and lower variation in species proportions.** This figure summarizes results from our case study of shotgun-sequenced stool and vaginal samples in the Human Microbiome Project.
+Panel A shows the distribution of order-2 alpha diversity (equal to the Inverse Simpson Index) across samples.
+Panel B shows the variation in the mean efficiency, as quantified by the GSD, across samples of a given type for each of 1000 simulated bias vectors.
+Panel C shows the variation in the proportion of each species commonly found in a given sample type.
+For Panel C, variation is measured by the GSD when a pseudo-value of $10^{-4}$ is added to the raw proportions.
+Different choices of the pseudo-value substantially change the measured GSDs, but not the difference (ratio) between gut and vaginal species.
+
+<br><br>
+
 <!-- Solutions -->

### 2022-05-23 index.Rmd
Revise title
Add link to research notebook to preface

@@
 ---
-title: "Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements"
+title: "Differential abundance analysis of taxonomically-biased microbiome measurements"
 author:
@@
 See [the Zenodo record](https://doi.org/10.5281/zenodo.4552717) for how to cite the latest version.
+Supporting data analyses can be found in the accompanying [computational research notebook] (https://mikemc.github.io/differential-abundance-theory/notebook/).

### 2022-05-25 solutions.Rmd
Revise solutions section
Revise to incorporate BJC's comments on the March 14 draft, remove the
experimental-recommendations section, and overhaul the two sections on
absolute DA.

--- 2022-02-09 solutions.Rmd
@@
 
-The theoretical results from Section \@ref(differential-abundance) suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.
+## Ratio-based relative DA analysis
 
-## Use ratio-based relative-abundance analyses
+Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using relative DA methods that analyze FCs in these ratios.
+A variety of ratio-based methods have been published under the name of Compositional Data Analysis (CoDA).
+CoDA methods may more generally consider the ratios between products of multiple species, possibly raised to some exponent; a simple example is the geometric mean of several species.
+The mathematical operations of multiplication and exponentiation maintain constant FEs.
+Hence such products provide a method for aggregating species into a higher-level taxonomic units that maintains bias invariance when estimating FCs.
+Ratio-based analyses also avoid the potential for misinterpreting changes in a species’ proportion that are driven by the sum-to-one constraint (@gloor2017micr).
 
-Since bias creates a consistent fold error in species ratios, a natural approach to countering bias when analyzing relative abundances is to use DA methods that are based on fold changes in ratios.
-A variety of such methods have been developed for microbiome DA analysis that build on earlier work from the field of Compositional Data Analysis (CoDA).
-These methods analyze log-fold changes in the ratios among a pair of species or, more generally, some product of exponential power of multiple species, which likewise are bias-invariant.
-Besides a greater robustness to taxonomic bias, such ratio-based analyses have other advantages.
-First, analysis of ratios avoids other standard criticisms leveled at proportion-based analysis—namely that the change in proportion of a species is affected by change in the density of all other species and the set of species used in the denominator of the proportion calculation (@gloor2017micr).
-Second, ratios may be more informative than absolute density in some cases, such as for swabs or other sample types for which the absolute density in the sample may be an unreliable indicator of the absolute density of the sampled ecosystem.
+We note two important caveats.
+First, in most MGS datasets, any given species is likely to be extremely rare in most samples, such that no sequencing reads are observed.
+In order to analyze FCs, DA methods interpret these zero counts as evidence that the true abundance of the species is very small (but still positive).
+A common approach to enable the computation of FCs in the presence of zeros is to add a small value, or ‘pseudo-count’, to the read counts, making all values positive.
+Unfortunately, this procedure violates bias invariance.
+How bias interacts with more sophisticated approaches for handling zeros remains an important open question.
+Second, the bias invariance of ratios among species does not extend to additive aggregates of species into higher-order taxa such phyla unless bias is conserved within the species group (@mclaren2019cons).
+Multiplicative aggregates of species, recently used for biomarker discovery (@riverapinto2018bala, @quinn2020inte), provide an alternative that remains bias-invariant but may be harder to interpret.
 
-Ratios come with their limitations, however.
-First, there are a vast number of species ratios (and derivatives) to potentially consider, none of which provide direct information about absolute densities, which can make choosing and interpreting a ratio-based DA analysis challenging.
-Second, ratio measurements are impacted by noise in both the numerator and denominator and so can be noisier than proportion measurements, particularly when the read counts in the denominator are small.
-A related problem is that, due to the discrete nature of biological organisms and the reads that are generated during sequencing, species in the denominator are commonly observed to have 0 reads; the assumptions that made by different methods about the meaning of these 0 counts can have a major impact on DA results.
-Finally, bias invariance at the level of ratios among species does not extend to ratios of additive aggregates of species, such as between bacterial phyla, unless further conditions are met (@mclaren2019cons).
+If biological or technical considerations favor a DA method that is not bias invariant, it can be beneficial to also use a ratio-based method as a robustness check.
+For example, @hevroni2020seas examined how the within-sample ranks of viral species changed from summer to winter.
+Since changes in within-sample ranks can be affected by bias, the authors also considered the (bias-invariant) changes in the centered log ratios (CLR) associated with each species.
+They found that the species whose within-sample ranks increased also increased in their CLR value, providing evidence that their initial results were not driven by bias.
 
@@
 
-Measurement of community calibration controls along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis.
-Community calibration controls are samples whose species identities and relative abundances are known either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
-MGS measurement of one or more control communities can be used to directly measure the relative efficiencies among the superset of species in the controls (@mclaren2019cons).
-The measured relative efficiencies can then be used to calibrate (remove the effect of bias from) the relative abundances (ratios) of the control species in a set of primary (non-control) experimental samples that were measured with the same protocol (ideally in the same experiment and sequencing run) (@mclaren2019cons).
-Calibration can be extended to species not in the controls using statistical methods to impute (predict) the unobserved efficiencies using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
+_Community calibration controls_ are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
+Inclusion of these samples along with the primary samples in an MGS experiment can enable researchers to calibrate their MGS measurement by directly measuring and removing the effect of bias.
+Calibration can be extended to species not in the controls by imputing the efficiencies of the missing species using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties such as cell-wall structure.
 
-The calibrated compositions can be used for arbitrary downstream microbiome analyses, including relative and absolute DA analysis.
-To demonstrate the potential for calibration to improve fold-change measurements, we estimated bias from one sample in the mock community data from @brooks2015thet that contained all 7 species and used it to calibrate the compositions of all samples (Figure \@ref(fig:calibration-example)).
-We then estimated species' densities using the uncalibrated and calibrated proportions using total-abundance normalization, treating the total density as known and fixed.
-Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions and densities (Figure \@ref(fig:calibration-example)).
-To demonstrate a practical application, we used the bias estimated from the full set of mock communities as the basis for calibrating the vaginal community time series measurements from the MOMS-PI study (@fettweis2019thev) that were made using the same 16S sequencing protocol.
-We imputed the efficiencies of other species using taxonomic relationships (Methods).
-Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \@ref(case-studies).
+Calibration makes it possible to correct for bias in relative and absolute DA methods that would otherwise not be bias invariant.
+To demonstrate, we used calibration to improve the estimated FCs between samples in the absolute abundance of a species in the bacterial mixtures from @brooks2015thet.
+We used the bias estimated from a single 7-species mixture to calibrate the MGS-derived proportions in all measured mixtures.
+We obtained calibrated absolute abundances by multiplying the calibrated proportions by the known total abundance (Equation \@ref(eq:density-prop-meas)).
+Calibration greatly increased the accuracy of the estimated FCs (Figure \@ref(fig:calibration-example), A vs. B).
+The fungal and vaginal case studies of Section \@ref(case-studies) provide several practical examples of calibration in proportion-based DA analyses.
+
+In studies of synthetic communities (like the fungal case study) or of ecosystems dominated by a relatively small number of culturable taxa (like the vaginal microbiome), calibration controls can be artificially constructed ('mock') communities.
+In other cases, _natural community controls_ can be derived from aliquots of a homogenized microbiome sample.
+Measurement of these controls across multiple experiments enables characterization of the _differential bias_ between experiments (@mclaren2019cons).
+Calibration using differential bias can make results directly comparable across studies that used different.
 
@@
 (ref:cap-calibration-example) **Fold changes can be calibrated using community controls or reference species.** The figure compares the performance of three methods for measuring fold changes in absolute cell density in cellular mock communities of 7 vaginal species, which were constructed and measured via 16S sequencing by @brooks2015thet.
-The 'Uncalibrated' fold changes are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which here is be known to be constant by construction).
+The 'Uncalibrated' fold changes are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which is constant by construction).
 The 'Calibrated (community)' measurements are computed from abundance measurements where the proportions are first corrected for the taxonomic bias that was estimated from a single sample that contained all 7 species.
@@
 
-## Calibration using reference species
+## Choose absolute-abundance methods with more stable FEs
 
-Obtaining control communities that span the taxa of interest is often not practical.
-Moreover, the bias that is measured from controls may differ from that in the primary samples due to differences in cell state or sample preparation.
-These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-density measurement that was described in Section \@ref(differential-abundance).
-This *reference-species calibration* allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from taxonomically biased, relative metagenomics measurements.
+There are many methods for obtaining species (absolute) abundances from the relative abundances measured by MGS, which had previously not been evaluated for their sensitivity to taxonomic bias (@williamson2021amul being a notable exception).
+Our theoretical results show that these methods differ in the extent to which taxonomic bias causes the FEs in species abundances to vary across samples.
+The effect of bias on absolute DA analysis can be mitigated by choosing methods with more stable FEs, using either of two general approaches.
 
-Recall that densities estimated by the reference-species approach have a constant fold error, provided that the fold error in the reference species is also constant (Equation \@ref(eq:density-ratio-error)).
-Three categories of reference species are reviewed for this purpose in Appendix \@ref(review-absolute-methods), including why they might be expected to produce constant fold errors: housekeeping species assumed to have constant abundance, spike-in species added at known or constant abundance, and species whose abundance has been measured with a targeted method like species-specific qPCR.
+### Use complementary MGS and total-abundance measurements
 
+A popular approach to measuring the absolute abundance of individual species is to multiply the species' proportions from MGS by a measurement of the abundance of the total community (Equation \@ref(eq:density-prop-meas)).
+The resulting species abundances are affected by taxonomic bias in the MGS _and_ the total-abundance measurement (Equation \@ref(eq:density-prop-error)).
+Section \@ref(differential-abundance) shows that the effect of both forms of bias on DA results can be reduced by choosing MGS and total-abundance methods that have similar taxonomic bias.
+
+Consider the debate over whether flow cytometry or 16S qPCR are better methods for measuring total abundance for the purposes of normalizing 16S amplicon sequencing data (@galazzo2020howt, @jian2021comm).
+Flow cytometry directly counts cells, whereas 16S qPCR measures the concentration of the 16S gene following DNA extraction.
+Thus the 16S qPCR measurement is subject to bias from extraction, copy-number variation, primer binding, and amplification, just like the 16S sequencing measurement.
+Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it can lead to more accurate fold changes across samples.
+
+More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement may be qPCR the same marker gene, from the same DNA extraction.
+Similar reasoning suggests that for shotgun sequencing, the ideal total-abundance measurement is bulk DNA quantification: Shotgun sequencing and bulk DNA quantification are both subject to bias from extraction and variation in genome size.
+Optimal use of these pairings requires thoughtful choices during bioinformatics.
+For example, performing copy-number correction on amplicon read counts prior to multiplication by qPCR measurements would be counter-productive, as it decouples bias in the two measurements.
+Additionally, the MGS proportions should be computed prior to discarding any unassigned reads, since species that are missing from the given taxonomy database still contribute to the total concentration of marker-gene and/or bulk DNA.
+Future experiments should evaluate the extent to which the taxonomic bias of DNA-based total-abundance measurements is stable across samples and shared with the complementary MGS measurement.
+
+### Normalize to a reference species
+
+A second approach to obtaining species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \@ref(eq:density-ratio-meas)).
+Our theory predicts that taxonomic bias induces constant FEs in the resulting species abundances which do not affect DA results.
+
+Previous studies have used species with a constant abundance as references: spike-ins, computationally-identified 'housekeeping species', and the host.
+Researchers may naturally be concerned about bias between the reference species and the native species being normalized against it (see, for example, @harrison2021theq's recommendations for spike-in experiments).
+Our results show that accurate FCs and DA results can be obtained so long as the relative efficiency of the focal species to the reference is consistent across samples.
+This condition is much weaker than requiring that bias is small (the relative efficiency is close to 1) and so greatly expands the applicability of spike-ins and host normalization in microbiome experiments.
+
+Section \@ref(absolute-abundance) further proposed an additional class of reference species for normalizing MGS measurements: Species whose abundance has been measured using a targeted method such as qPCR with species-specific primers.
 To demonstrate the ability for reference calibration to improve fold changes, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
 Doing so improved the resulting measurements of fold changes in density for all species (Figure \@ref(fig:calibration-example)).
+Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus further expanding the applicability of this reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
+The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
+In many ecosystems, there may not be a single universally-present species that can serve as a reference in all samples.
+Sample coverage can be increased by making targeted measurements of multiple species or of larger taxonomic groups; see Appendix REF for details.
 
-## Choosing complementary total-density and metagenomics measurements
+## Bias sensitivity analysis
 
-A variety of methods for measuring total community density have been paired with either amplicon or shotgun sequencing data for the purpose of conducting absolute DA analysis using the total-density approach described in Section \@ref(differential-abundance).
-Each of these methods is subject to its own taxonomic bias: Cells from different species are likely to contribute more or less efficiently to the total abundance measurement.
-Consider the popular method of assessing bacterial abundance with total-16S qPCR measurement.
-Even an ideal qPCR protocol that perfectly amplifies all species remains a biased measurement of cell density, since the total 16S copies in the extracted DNA contributed by each species will be proportional to its (species-specific) lysis efficiency and 16S copy-number.
-Empirical evidence supports this, as qPCR measurements of total 16S copies are rarely equivalent to total cell counts, even when multiple copy numbers are taken into account ()@lloyd2013meta).
-16S qPCR is commonly paired with 16S amplicon sequencing, with which these sources of bias are shared, perhaps along with variation in primer binding and amplification efficiency.
-Methods like flow cytometry that directly measure cell density lack these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
-By extending the analysis of Section \@ref(differential-abundance) to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(total-density-bias)).
-This finding suggests that methods of total-density measurement that are more accurate as measures of total cell density may actually perform worse than less accurate methods for the purposes of absolute DA inference.
+Even if control measurements are not available, it is possible to computationally assess the sensitivity of a given DA result to taxonomic bias.
+A _bias sensitivity analysis_ can be conducted by analyzing an MGS dataset multiple times under a range of possible taxonomic biases.
+First, a large number of hypothetical biases are generated from a user-specified probability distribution.
+Next, the DA analysis is re-run while using each simulated bias vector to calibrate the MGS measurements.
+This approach is highly flexible, and can be used to investigate the bias sensitivity of any DA method.
+Alternatively, existing DA methods can be extended to directly include the unknown taxonomic bias in their underlying statistical model, thereby providing DA estimates that inherently account for the added uncertainty in microbiome compositions due to the presence of unknown bias.
+(@greenland2005mult compares these two approaches in a general context; @nixon2022asta applies the second approach to microbiome data in the absence of taxonomic bias.)
 
-We illustrate with a hypothetical example of two vaginal communities of equal density, but which are dominated either by _Lactobacillus iners_ or _Gardnerella vaginalis_.
-Based on our earlier analysis in @mclaren2019cons, we estimate that _L. iners_ yields 10X more 16S copies per cell than _G. vaginalis_ in extracted DNA.
-Ignoring other sources of bias, we expect relative efficiencies of _L. iners_:_G. vaginalis_ of 10 for both qPCR and 16S sequencing measurements.
-Consequently, the _G. vaginalis_ community will yield a substantially lower qPCR measurement than the _L. iners_ community and we would incorrectly infer a decrease in total abundance.
-At the same time, this decrease in the mean efficiency of the sequencing measurement would cause artificially high fold changes in the proportions of all species (Section \@ref(differential-abundance)).
-When using qPCR and 16S sequencing measurements for community-based density estimation, these two opposing errors cancel, yielding accurate fold changes in species densities.
-
-More generally, changes in the mean efficiency of the total-density measurement can offset those in the sequencing measurement when comparing log density of species across samples.
-If the (log) efficiencies of the total and sequencing measurement are positively correlated across species, then the (log) mean efficiencies will tend to be positively correlated across samples, leading to a reduction in error in DA analysis relative to total-density methods whose efficiencies are uncorrelated with those of the sequencing measurement.
-These observations suggest that qPCR of a marker gene may be the ideal pairing for amplicon sequencing measurements, despite being a poor measure of changes in total cell density.
-Similarly, bulk DNA quantification may be an ideal pairing for shotgun sequencing, making it possible to account for variation in lysis efficiency and genome length.
-These pairings even make it possible to account for error caused by variation among samples in the fraction of unclassified reads, which can form a major fraction of amplicon and shotgun data.
-Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (Appendix \@ref(total-density-bias)).
-
-## Bias-sensitivity analysis
-
-For experiments that have been conducted without calibration controls, it is still possible to investigate the likelihood that DA results can be explained by taxonomic bias by performing a computational bias-sensitivity analysis.
-
-A bias-sensitivity analysis examines how sensitive the results of a given DA analysis are to the working assumption about the bias of a given protocol.
-This working assumption is typically is that there is no bias, but could also be bias measured in a previous experiment or predicted from taxonomic features like 16S copy number.
-A straightforward approach to bias-sensitivity analysis is to use computer simulation to re-analyze the MGS data across a set of simulated taxonomic biases.
-First, multiple sets of efficiency vectors are randomly generated, each representing a possible quantification of taxonomic bias for the study.
-Second, each efficiency vector is used to calibrate the original MGS measurements; each set of calibrated measurements represents the true community compositions under the given hypothesized bias. 
-Finally, the DA analysis is re-run on each set of calibrated measurements and the distribution of results is compared graphically or with summary statistics.
-A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool 'corncob' are shown in SI Figure \@ref(fig:sensitivity-example).
-The simulated efficiency vectors may be generated to different hypotheses about bias in the given system, such as its overall magnitude and correlation among phylogenetically-related species.
-Thus the utility of such simulations will increase the more we learn about the properties of taxonomic bias in different systems.
-
-An alternative approach to bias-sensitivity analysis would be to directly include the (unknown) taxonomic bias of the protocol in the statistical model used for DA analysis, as described for bias-aware meta-analysis below.
-Further development of tools and workflows for performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance and microbiome analyses more generally.
+Interpreting a bias sensitivity analysis is complicated by the large percentage of zero counts in species-level microbiome profiles, as the results may strongly depend on how these zeros are handled in the calibration process.
+Therefore it can be advantageous to jointly test the sensitivity of assumptions about zero-generating processes along with taxonomic bias, making statistical DA models that include bias and one or more zero-generating processes especially valuable.
+The development of tools and workflows to facilitate bias sensitivity analysis may provide an efficient way to increase scientists’ ability to assess the reliability of microbiome results, both for differential abundance and microbiome analyses more generally.
 
@@
 
-So far we have considered analysis of samples subject to the same taxonomic bias;
-however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in samples from different studies.
-This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses.
-A potential solution is to perform a bias-aware meta-analysis that that explicitly accounts for the possibility of distinct taxonomic bias across studies.
-
-Parametric microbiome meta-analysis models include sets of species-specific latent parameters to model "batch effects"—amorphous differences in the measurements of each study.
-By configuring these models so that these latent parameters correspond to study-specific species efficiencies, we can improve their biological interpretability and gain the ability to directly assess whether disagreements in DA results across studies can be explained by differences in taxonomic bias.
-To the extent that taxonomic bias truly is consistent within studies, meta-analyses that directly account for it can have greater statistical power to extract the truly consistent and inconsistent microbial patterns across studies.
+Meta-analysis of microbiome samples measured across multiple studies must contend with the fact that different studies typically use different protocols and hence have different taxonomic biases.
+These different biases can be explicitly accounted for in a _bias meta-analysis_ which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
+Parametric meta-analysis models include study-specific latent parameters representing 'batch effects'—non-biological differences in the data from each study which can distort the observed biological patterns.
+By estimating these 'nuisance parameters' along with the biological parameters of interest (such as species log-fold changes), the meta-analysis aims to reduce statistical bias in the biological parameters created by the non-biological differences among studies (@leek2010tack,@wang2019mana).
+In a bias-aware meta-analysis, the meta-analysis model is configured so that (some of) the latent parameters correspond to study- and species-specific relative efficiencies.
+If taxonomic bias is consistent within but not between studies, then this approach may improve the ability for the meta-analysis to accurately identify DA patterns as compared to meta-analysis methods in which the 'batch effects' do not reflect the multiplicative and compositional nature of taxonomic bias.
+The ability to infer and adjust for the differential bias between studies can be increased by measuring one or more shared control samples alongside the experiment-specific samples.

### 2022-05-25 introduction.Rmd
Revise abstract and introduction
Edit based on BJC's comments on the March 14 draft.

@@
 To assess how the species in microbial communities vary across conditions, researchers often measure the relative or absolute abundances of all species simultaneously using marker-gene and metagenomic sequencing.
-These measurements are taxonomically biased, with some species measured many times more efficiently than others.
-Measurement error caused by taxonomic bias is generally ignored in such _differential abundance (DA) analyses_, likely due to a combination of 1) a widespread belief that protocol standardization is sufficient to yield accurate differences between samples and 2) a lack of practical solutions for addressing it.
-We use theoretical arguments and analysis of real and simulated experiments to analyze the impact of taxonomic bias on relative and absolute DA analyses.
-We show that taxonomic bias does in fact create inferential errors for a class of popular DA methods; however, whether these errors are biologically significant depends on experimental context.
-We present six approaches to addressing the error, each suited to different experimental questions and systems.
-These approaches provide practical solutions to mitigating the effect of taxonomic bias for a large fraction of DA questions.
-If adopted, they may lead to notable improvements in the reproducibility and biological interpretability of the statistical findings of microbiome studies.
+However, these measurements are taxonomically biased: Some species are measured more efficiently than others, yielding measured abundances that are higher or lower than the truth.
+Taxonomic bias is generally ignored in _differential abundance (DA) analyses_, likely due to a combination of 1) a widespread belief that protocol standardization is sufficient to yield accurate differences between samples and 2) a lack of practical solutions for addressing it.
+We use theoretical arguments and case studies to analyze the impact of taxonomic bias on DA analyses of relative and absolute abundances.
+We show that taxonomic bias does create inferential errors for a class of popular DA methods; however, whether these errors are biologically significant depends on experimental context.
+We present several approaches to mitigating the effect of bias, suited to a range of experimental questions and systems.
+If adopted, they could improve the reproducibility and interpretability of sequencing-based studies of microbial communities.
 
@@
 In these _differential-abundance (DA) analyses_, the change in abundance of a microbial taxon across samples or conditions is used to infer ecological dynamics or find microbes that are associated with specific host diseases or environmental conditions.
-Standard MGS measurements lose information about total microbial density and so are typically used to analyze the abundances of taxa relative to each other.
-But new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density.
-In its various forms, DA analysis remain one of, if not the most, common forms of analyses applied to MGS data to elucidate the inner workings of microbiomes and their relationships to host and environmental health.
+Standard MGS measurements lose information about total microbial density and so are typically used to analyze the abundances of taxa relative to each other (_relative abundances_).
+But new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density, biomass, or genome copy number (_absolute abundances_).
+In its various forms, DA is among the most common analyses applied to MGS data.
+
+Unfortunately, these DA analyses are built on a fundamentally flawed foundation.
 
 <!-- P: Taxonomic bias -->
-Yet these DA analysis are built on a fundamentally flawed foundation.
-MGS measurements are _taxonomically biased_: Microbial species vary dramatically (e.g.
-10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—by a given MGS protocol (@mclaren2019cons).
+MGS measurements are _taxonomically biased_: Microbial species vary dramatically in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—by a given MGS protocol (@mclaren2019cons).
 This bias arises from variation in how species respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
 Although often associated with features specific to marker-gene sequencing—the variation among species in marker copy numbers and in primer-binding and amplification efficiencies—the existence of large variation in DNA extraction efficiencies and in the ability to correctly classify reads make taxonomic bias a universal feature of both marker-gene and shotgun measurements.
-As a result, MGS measurements provide inaccurate representations of actual community composition and tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo, @mclaren2019cons).
+As a result, MGS measurements provide inaccurate representations of actual community composition and tend to differ systematically across protocols, studies, and even experimental batches within a study (@yeh2018taxo, @mclaren2019cons).
 These errors can supersede sizable biological differences (e.g.
-@lozupone2013meta) and may have contributed to failed replications of prominent findings such as the associations of Bacteroides and Firmicutes in stool with obesity (@finucane2014atax) and the associations of species in the vaginas of pregnant women with preterm birth (@callahan2017repl).
+@lozupone2013meta) and may have contributed to failed replications of prominent DA results such as the associations of Bacteroides and Firmicutes in stool with obesity (@finucane2014atax) and the associations of species in the vaginas of pregnant women with preterm birth (@callahan2017repl).
 
@@
 The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
-Statistical analyses are then conducted with the (often tacit) assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
-This argument is at least intuitively plausible for DA analyses based on multiplicative or fold changes in a taxon's abundance.
-If bias caused the abundance of a species to be consistently measured as 10X too high---what is known as _proportional error_---then we would still correctly measure its multiplicative variation across samples.
+Statistical analyses are then conducted with the tacit assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
+This argument is at least intuitively plausible for DA analyses based on multiplicative or fold differences (FDs) in a taxon's abundance.
+If bias caused a species' abundance to be consistently measured as 10-fold greater than its actual value, then we would still recover the correct FDs among samples.
 <!-- (@kevorkian2018esti, @lloyd2020evid). -->
-However, @mclaren2019cons showed mathematically and with MGS measurements of artificially constructed ('mock') communities that consistent taxonomic bias can create non-proportional errors that can majorly distort cross-sample comparisons.
-In particular, they showed that a species' proportion---the most common measure of relative abundance---has non-proportional errors which affect the inferred changes between samples and can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
+However, @mclaren2019cons showed mathematically and with MGS measurements of artificially constructed ('mock') communities that consistent taxonomic bias can create fold errors (FEs) that vary across samples and, as a result, majorly distort cross-sample comparisons.
+In particular, they showed that the FE in a species' proportion---the most common measure of relative abundance---varies among samples, distorting the observed FDs between samples.
+In some cases, bias can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
 Yet @mclaren2019cons also found that other abundance measures---those based on the ratios among species---have proportional errors and may lead to more robust DA analyses.
@@
 
-Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundance.
-Our analysis clarifies how the received wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
-Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.
-Over time, application of these methods to past and future experiments will provide crucial quantitative information about the conditions under which taxonomic bias creates spurious results for various DA methodologies.
-Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.
+Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
+We show that, in contrast to received wisdom, taxonomic bias can affect the results obtained from many standard DA methods; however, other methods are robust to bias, provided that it is truly consistent across samples.
+We then use several case studies to explore the real-world conditions in which bias is likely to cause serious scientific errors.
+Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.
+Application of these insights and methods can build the confidence needed to turn the findings of microbiome studies into readily-translatable scientific knowledge.

### 2022-05-27 _cut.Rmd
Revise abundance-measurement section
Edit based on BJC's comments on the March 14 draft.

--- 2022-03-13 abundance-measurement.Rmd
@@
+# Cut text
+
 # How taxonomic bias affects abundance measurements {#abundance-measurement}
 
-Taxonomic bias is fundamentally multiplicative.
-Species vary in how easily their cells are lysed, and the number of lysed cells for a given species following DNA extraction equals the number of input cells multiplied by the probability of lysis per cell, or lysis efficiency, of the species.
-Similarly, prokaryotic species vary in their number of 16S genes, such that the numbers of 16S sequencing reads for various species in a 16S amplicon experiment is expected to be proportional to the numbers of 16S genes in their genomes.
-Nevertheless, the nature of MGS experiments are such that consistent taxonomic bias creates non-proportional error in common forms of relative and absolute abundance measurements---error that varies across samples and hence distort cross-sample comparisons.
-This section reviews the theoretical results of @mclaren2019cons for how taxonomic bias affects MGS measurements of relative abundance and extends them to include measurements of absolute abundance.
-We show that only some of the measurement methods currently in use have proportional errors, making these methods intrinsically more robust to bias than the others which do not.
+## Absolute abundance
 
-## Model of MGS measurement
-
-Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
-This model is the simplest that respects the multiplicative nature of taxonomic bias and the *compositional* nature of MGS measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
-The model as described by @mclaren2019cons considers only relative abundances; here we present a straightforward extension that allows us to also consider absolute abundances.
-
-<!-- TODO: Is S the set of species that could be in the sample, or the set that can be mapped to? Also, what about the host H? -->
-<!-- This issue ties in with our general notation issue, and the problem of unassigned reads. -->
-
-We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and taxonomically assigns reads to a set of microbial species $S$.
-We make several simplifying assumptions to facilitate our analysis and presentation.
-First, we consider only species-level assignment, and suppose that reads that cannot be uniquely assigned to a single species in $S$ are discarded.
-Second, we ignore the possibility that reads are misassigned to either the wrong species or sample.
-Third, we suppose that taxonomic bias acts consistently across samples at the level of species---that is, a given species is always measured more efficiently than another to the same degree, regardless of variation in changes in sub-species-level composition or the sample matrix.
-Finally, unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
-These assumptions, though unrealistic descriptions of most MGS experiments, serve the purpose of clearly demonstrating when and why consistent taxonomic bias creates errors in DA analysis.
-<!-- The results we derive from them form the basis for understanding the impact of bias in more complex scenarios. -->
-
-Our model describes the mathematical relationship between the read counts obtained by MGS and the true absolute abundances of each species in a sample.
-For concreteness, we take _absolute abundance_, or simply _abundance_, to refer to cell concentration, or number of cells per unit volume, in a sample.
-Thus we take 'absolute' simply to mean not relative to other species; however, our results also apply were 'absolute' taken to mean all cells in a sample or in a sampled ecosystem.
-Similarly, our results apply to abundance units other than cells, such as biomass and genome copy number.
-Our model stipulates that the assigned read count of a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
-\begin{align}
-  (\#eq:mgs-model)
-  \text{reads}_{i}(a)
-  = \text{abun}_{i}(a) \quad \cdot
-    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
-    \cdot \quad
-    \underbrace{\text{effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
-\end{align}
-The species-specific factor, $\text{efficiency}_{i}$, equals the *relative measurement efficiency* (or simply *efficiency*) of the species---how much more easily that species is measured (converted from cells to assigned reads) relative to a arbitrary fixed reference species (@mclaren2019cons).
-We assume the efficiencies of particular species are consistent across samples.
-The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
-The sample-specific factor, $\text{effort}(a)$, quantifies the *effective experimental effort* per unit abundance in measuring the sample.
-<!-- It accounts for variation in sample input volume and total sequencing run output, as well as the mean species efficiency within the sample. -->
-Equation \@ref(eq:mgs-model) implies that the total number of assigned reads in sample $a$ equals
-\begin{align}
-  (\#eq:total-reads)
-  \text{reads}_S(a)
-    = \text{abun}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
-\end{align}
-where $\text{abun}_{S}(a) \equiv \sum_{j\in S}\text{abun}_j(a)$ is the total abundance of species in $S$ and 
-\begin{align}
-  (\#eq:mean-efficiency)
-  \text{efficiency}_S(a) 
-    \equiv \frac{\sum_{j\in S}\text{abun}_j(a) \cdot \text{efficiency}_j}{\text{abun}_S(a)}
-\end{align}
-is the _mean efficiency_ over all species in $S$ weighted by their abundance.
-<!-- The subscript $S$ serves as a reminder that many reads may not be assigned---for example, reads that fail to pass quality control or that come from the host or other taxonomic domains than what is being studied.
--->
-
-## Relative abundance (proportions and ratios) {#relative-abundance}
-
-We distinguish between two types of species-level *relative abundances* within a sample.
-The *proportion* of species $i$ in sample $a$ equals its abundance relative to the total abundance of all species in $S$,
-\begin{align}
-  (\#eq:prop)
-  \text{prop}_{i}(a) 
-    &\equiv \frac{\text{abun}_i(a)}{\text{abun}_S(a)}.
-%  \\&= \frac{\text{abun}_i(a)}{\sum_{i \in S}\text{abun}_i(a)}.
-\end{align}
-The *ratio* between two species $i$ and $j$ equals the abundance of $i$ relative to that of $j$,
-\begin{align}
-  (\#eq:ratio)
-  \text{ratio}_{i/j}(a) = \frac{\text{abun}_i(a)}{\text{abun}_j(a)}.
-\end{align}
-Proportions and ratios each form the basis for popular relative DA methods.
-
-The proportion of a species is typically measured by its proportion of assigned reads,
-\begin{align}
-  (\#eq:prop-meas)
-  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
-\end{align}
-We can rewrite the right-hand side (using Equations \@ref(eq:mgs-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas)) to find
-\begin{align}
-  (\#eq:prop-error)
-  \widehat{\text{prop}}_{i}(a)
-  &= \text{prop}_{i}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}}_{\substack{\text{variable} \\ \text{fold error}}}.
-\end{align}
-Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency relative to the mean efficiency in the sample.
-Consequently, the species's proportion is measured as too high in samples that are dominated by species with lower efficiencies, and measured as too low in samples that are dominated by species with higher efficiencies.
-This phenomenon is illustrated in two hypothetical communities in Figure \@ref(fig:error-proportions). 
-Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.
-A demonstration in artificially constructed ('mock') communities of bacterial cells is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
-
-The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
-\begin{align}
-  (\#eq:ratio-meas)
-  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
-\end{align}
-From Equations \@ref(eq:mgs-model) and \@ref(eq:ratio-meas), it follows that 
-\begin{align}
-  (\#eq:ratio-error)
-  \widehat{\text{ratio}}_{i/j}(a)
-%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
-  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
-\end{align}
-Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
-For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
-A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
-
-Thus taxonomic bias creates variable fold errors in species proportions, but constant fold errors in species ratios.
-Consequently, a perfect cancellation of errors can only occur in ratio-based DA analysis.
-
-<!-- begin figure -->
-
-```{r error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
-"figures/illustrations/error-proportions.svg" %>%
-  include_svg
-```
-
-(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species.
-Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.
-Species' proportions may be measured as too high or too low depending on sample composition.
-For instance, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).
-
-<!-- end figure -->
-
-**Higher-order taxa:** How does bias affect the proportions and ratios of higher-order taxa, such as a genera, phyla, or even all organisms?
-More generally, we can identify a higher-order taxon $I$ with a set of species, $\{i \in I\}$.
-The abundance and reads of a taxon $I$ in a sample $a$ equal the sums over its constituent species, $\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)$ and $\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)$.
-The efficiency of taxon $I$ equals the weighted average of its constituent species,
-\begin{align}
-  (\#eq:efficiency-general)
-  \text{efficiency}_I(a) 
-    \equiv \frac{\sum_{i\in I}\text{abun}_i(a)\cdot \text{efficiency}_i}{\text{abun}_I(a)}.
-\end{align}
-Thus the efficiency of a higher-level taxon may vary across samples (@mclaren2019cons).
-An exception is if the constituent species have the same efficiency or always appear in the same abundances relative to each other, in which case the efficiency of the higher-order taxon would stay constant.
-As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
-The efficiency of the phylum would then be $\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5$ in Sample 1 and $\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1$ in Sample 2.
-Equations \@ref(eq:prop-error) and \@ref(eq:ratio-error) continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent taxon efficiency defined by Equation \@ref(eq:efficiency-general) is used.
-In this way, we can see that both proportions and ratios among higher-order taxa may have inconsistent fold errors.
-
-## Absolute abundance {#absolute-abundance}
-
-Various methods can be used to convert the relative abundances from MGS into absolute abundances.
-These methods are subject to taxonomic bias in the MGS measurements and in any supplemental, non-MGS measurements that may be used for the conversion.
-Appendix [REF] develops a general framework to understand how bias affects absolute-abundance measurements for species and higher-order taxa across these methods.
-Here we present our main results as they apply to currently-used methods.
-
-### Methods that measure total-community abundance
-
-The definition \@ref(eq:prop) for the proportion of a species indicates that we can convert proportions to absolute abundances were we to know the absolute abundance of the total community.
-Given our measured proportions from MGS and a measure of total abundance from a supplemental, non-MGS measurement, $\widehat{\text{abun}}_{S}(a)$, one can measure the abundances of individual species by
-\begin{align}
-  (\#eq:density-prop-meas)
-  \widehat{\text{abun}}_{i}(a) 
-  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{abun}}_S(a).
-\end{align}
-Total-abundance measurements that have recently used for this purpose include counting cells with microscopy (@lloyd2020evid) or flow cytometry (@props2017abso,@vandeputte2017quan,@galazzo2020howt), measuring the concentration of a marker-gene with qPCR or ddPCR (@zhang2017soil,@barlow2020aqau, @galazzo2020howt,@tettamantiboshier2020comp), and measuring bulk DNA concentration with a florescence-based DNA quantification method (@contijoch2019gutm).
-
-Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
-Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
-The marker-gene concentration measured by qPCR is affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency.
-We can easily understand the impact of taxonomic bias on the total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
-Suppose that each species has an _absolute efficiency_ for the total-abundance measurement that is constant across samples.
-Neglecting other sources of random and systematic error, the total-abundance measurement equals
-\begin{align}
-  (\#eq:total-density-error)
-  \widehat{\text{abun}}_S(a) 
-  &= \sum_{i\in S} \text{abun}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
-\\&= \text{abun}_S(a) \cdot \text{efficiency}^{\text{tot}}_S(a),
-\end{align}
-where $\text{efficiency}_{i}^{\text{tot}}(a)$ is the absolute measurement efficiency of species $i$ for the total-abundance measurement and $\text{efficiency}^{\text{tot}}_S(a)$ is the mean efficiency of the total-abundance measurement in the sample.
-<!-- Note, we have assumed that the total-abundance and MGS measurements are sensitive to the same species -->
-
-The species abundance measurements are affected by taxonomic bias in both the MGS and total-abundance measurement.
-We can compute this effect by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
-\begin{align}
-  (\#eq:density-prop-error)
-  \widehat{\text{abun}}_{i}(a) 
-  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}.
-\end{align}
-Equation \@ref(eq:density-prop-error) indicates that the multiplicative error in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
-As in the case of proportions (Equation \@ref(eq:prop-error)), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
-On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to more stable errors.
-We discuss how this possibility might be exploited in real experimental workflows in Section \@ref(solutions).
-
-### Methods that determine or measure a reference species
-
-The definition \@ref(eq:ratio) for the ratio between two species suggests another method to convert relative to absolute abundances: 
-By knowing the ratio of species $i$ to a _reference species_ $r$ and the abundance of species $r$, we can determine the abundance of species $i$.
-Thus given an MGS measurement and a measurement $\widehat{\text{abun}}_{r}(a)$ of the abundance of species $r$, we can measure the abundance of an arbitrary species $i$ as
-\begin{align}
-  (\#eq:density-ratio-meas)
-  \widehat{\text{abun}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{abun}}_{r}(a)}{\text{reads}_{r}(a)}.
-\end{align}
 Here we consider some practical applications of this method and how their measurements are affected by bias in the MGS measurement and error in $\widehat{\text{abun}}_{r}(a)$.
@@
 
-This approach has so far been solely or primarily used with reference species that have been determined to have an abundance that is constant across samples.
-In a spike-in experiment, researchers deliberately add (or 'spike in') one or more reference species at a known, constant abundance.
-For a single spike-in species $r$, the abundances of other species can be estimated by Equation \@ref(eq:density-ratio-meas).
-In the absence of spike-ins, researchers may be able to determine one or more native species whose abundances are thought to be constant across samples; we refer to such species as _housekeeping species_ by analogy with the housekeeping genes used for absolute-abundance conversion in studies of gene expression.
-Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference.
-A related approach involves computationally identifying species that are constant between pairs of samples (@david2014host) or between sample conditions (@mandal2015anal, @kumar2018anal) under added statistical assumptions about the existence of such species.
-The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set $\widehat{\text{abun}}_{r}(a)$ to 1 in Equation \@ref(eq:density-ratio-meas).
-The resulting abundance measurements have unknown but fixed units, which is sufficient for measuring fold changes across samples.
-
-Equation \@ref(eq:density-ratio-meas) makes clear that the reference species need not be constant so long as we can directly measure it.
-This observation suggests a new approach to obtaining species absolute abundances, by combining MGS measurements of the community with targeted measurements of the absolute abundance of one or more individual species.
-Common targeted measurements include using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA, using primers tailored to a particular species.
-Other methods that are capable of measuring cells (rather than extracted DNA) include applying ddPCR directly to cells, performing CFU counting on selective media, and using flow cytometry with species-specific florescent probes.
-The measurements from any of these methods could be used for $\widehat{\text{abun}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) to obtain the absolute abundance for all species.
-Appendix [REF] describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
-
-Despite the diversity of these methods, each can be represented by Equation \@ref(eq:density-ratio-meas), making it straightforward to understand the impact of taxonomic bias in the MGS measurement.
-It follows from Equation \@ref(eq:ratio-error) that the measured abundance of species $i$ is 
-\begin{align}
-  (\#eq:density-ratio-error)
-  \widehat{\text{abun}}_{i}(a) 
-  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
-  \cdot \underbrace{\frac{\widehat{\text{abun}_r}(a)}{\text{abun}_r(a)}}_{\substack{\text{fold error in} \\ \text{ref. species}}}.
-  % \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) 
-\end{align}
-The error thus consists of two terms: the ratio of the efficiency between species $i$ and species $r$, and any error in the abundance of the reference species.
-The ratio of species efficiencies is constant by assumption; thus the total error is constant if the error in the reference abundance is.
-Note, if a higher-order taxon is used as a reference---for example, using targeted measurements of a genus or family---then the efficiency of the reference taxon may vary across samples, leading to non-proportional errors.
+### Leveraging information about a reference species
 
@@
 Taxonomic bias in MGS measurements makes all MGS-based measures of relative and absolute abundance inaccurate.
-For some measures---but not all---consistent bias at the species level creates proportional errors in species abundances.
-Of two types of relative abundance, the ratio between a pair of species has proportional errors whereas the proportion of an individual species does not. 
-Instead, the multiplicative error in a species' proportion varies with the mean efficiency across samples.
-Methods for measuring species absolute abundance that involve measuring the abundance of the total community likewise experience non-proportional errors which vary with the sample mean efficiency; 
+For some measures---but not all---consistent bias at the species level creates constant FEs in species abundances.
+Of two types of relative abundance, the ratio between a pair of species has constant FEs, whereas the proportion of an individual species does not. 
+Instead, the FE a species' proportion varies with the mean efficiency across samples.
+Methods for measuring species absolute abundance that involve measuring the abundance of the total community likewise have FEs that vary with the sample mean efficiency; 
 a notable exception is if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement.
-Absolute-abundance methods that instead involve measuring or spiking individual reference species yield proportional errors, provided that the abundance of the reference species can be determined up to a proportional error.
-The efficiency of higher-order taxa may vary across samples even if the efficiencies of their constituent species are constant; as a result, all standard abundance measures can have non-proportional errors at taxonomic resolutions higher than that at which bias is conserved.
-
-<!-- Possibly:  -->
-<!-- - taxonomic aggregates need not have proportional errors -->
-<!-- - the fundmental mathematical explanation spanning these results is that the sum of multiple values each having proportional error need not itself have proportional error. -->
+Absolute-abundance methods that instead involve measuring or spiking individual reference species have constant FEs provided that the abundance of the reference species can be determined up to a constant FE.
+The efficiency of higher-order taxa may vary across samples even if the efficiencies of their constituent species are constant; as a result, all standard abundance measures can yield variable FEs at taxonomic resolutions higher than that at which bias is conserved.

### 2022-05-27 abundance-measurement.Rmd
Revise abundance-measurement section
Edit based on BJC's comments on the March 14 draft.

@@
-# How taxonomic bias affects abundance measurements {#abundance-measurement}
+# How bias affects abundance measurements {#abundance-measurement}
 
-Taxonomic bias is fundamentally multiplicative.
-Species vary in how easily their cells are lysed, and the number of lysed cells for a given species following DNA extraction equals the number of input cells multiplied by the probability of lysis per cell, or lysis efficiency, of the species.
-Similarly, prokaryotic species vary in their number of 16S genes, such that the numbers of 16S sequencing reads for various species in a 16S amplicon experiment is expected to be proportional to the numbers of 16S genes in their genomes.
-Nevertheless, the nature of MGS experiments are such that consistent taxonomic bias creates non-proportional error in common forms of relative and absolute abundance measurements---error that varies across samples and hence distort cross-sample comparisons.
-This section reviews the theoretical results of @mclaren2019cons for how taxonomic bias affects MGS measurements of relative abundance and extends them to include measurements of absolute abundance.
-We show that only some of the measurement methods currently in use have proportional errors, making these methods intrinsically more robust to bias than the others which do not.
+This section extends the theoretical results of @mclaren2019cons to describe the effect that consistent taxonomic bias within an MGS experiment affects the relative and absolute abundances measured for various microbial species.
+We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.
 
@@
 Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
-This model is the simplest that respects the multiplicative nature of taxonomic bias and the *compositional* nature of MGS measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
-The model as described by @mclaren2019cons considers only relative abundances; here we present a straightforward extension that allows us to also consider absolute abundances.
+This model describes the mathematical relationship between the read counts obtained by MGS and the (actual) abundances of the various species in a sample.
+Here we extend the model as first described @mclaren2019cons, which considers only relative abundances, to also consider absolute abundances.
+For concreteness, we will consider _absolute abundance_, or simply _abundance_, to refer to the number of cells per unit volume in a sample (cell concentration).
+That said, our results equally apply to other definitions of absolute abundance, such as the total number of cells in a sample or ecosystem and other abundance units such as biomass or genome copy number.
 
-<!-- TODO: Is S the set of species that could be in the sample, or the set that can be mapped to? Also, what about the host H? -->
-<!-- This issue ties in with our general notation issue, and the problem of unassigned reads. -->
+This model is the simplest that respects the multiplicative nature of taxonomic bias and the *compositional* nature of MGS measurements.
+The actual abundance of a species in a given sample, multiplied by its measurement efficiency---its rate of conversion from cells to taxonomically assigned sequencing reads---determines the species' read count in that sample.
+Taxonomic bias presents as variation in the measurement efficiencies among species within an MGS experiment.
+The read counts further depend on sample-specific experimental factors that are typically unknown, such that they are best interpreted as only providing relative abundances (such data is said to be _compositional_) (@gloor2017micr).
 
@@
 First, we consider only species-level assignment, and suppose that reads that cannot be uniquely assigned to a single species in $S$ are discarded.
-Second, we ignore the possibility that reads are misassigned to either the wrong species or sample.
+Second, we ignore the possibility that reads are misassigned to the wrong species or the wrong sample.
 Third, we suppose that taxonomic bias acts consistently across samples at the level of species---that is, a given species is always measured more efficiently than another to the same degree, regardless of variation in changes in sub-species-level composition or the sample matrix.
-Finally, unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
+Finally, unless otherwise stated, we treat sequencing measurements as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
 These assumptions, though unrealistic descriptions of most MGS experiments, serve the purpose of clearly demonstrating when and why consistent taxonomic bias creates errors in DA analysis.
-<!-- The results we derive from them form the basis for understanding the impact of bias in more complex scenarios. -->
 
-Our model describes the mathematical relationship between the read counts obtained by MGS and the true absolute abundances of each species in a sample.
-For concreteness, we take _absolute abundance_, or simply _abundance_, to refer to cell concentration, or number of cells per unit volume, in a sample.
-Thus we take 'absolute' simply to mean not relative to other species; however, our results also apply were 'absolute' taken to mean all cells in a sample or in a sampled ecosystem.
-Similarly, our results apply to abundance units other than cells, such as biomass and genome copy number.
-Our model stipulates that the assigned read count of a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
+Our model stipulates that the taxonomically-assigned read count for a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
 \begin{align}
@@
     \underbrace{\text{effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
+%  \\ \\
+%  M_i^{(a)} = A_i^{(a)} B_i F^{(a)}.
 \end{align}
-The species-specific factor, $\text{efficiency}_{i}$, equals the *relative measurement efficiency* (or simply *efficiency*) of the species---how much more easily that species is measured (converted from cells to assigned reads) relative to a arbitrary fixed reference species (@mclaren2019cons).
-We assume the efficiencies of particular species are consistent across samples.
+The species-specific factor, $\text{efficiency}_{i}$, equals the *relative measurement efficiency* (or simply *efficiency*) of the species relative to an arbitrary baseline species (@mclaren2019cons).
 The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
 The sample-specific factor, $\text{effort}(a)$, quantifies the *effective experimental effort* per unit abundance in measuring the sample.
-<!-- It accounts for variation in sample input volume and total sequencing run output, as well as the mean species efficiency within the sample. -->
 Equation \@ref(eq:mgs-model) implies that the total number of assigned reads in sample $a$ equals
@@
     = \text{abun}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
+    \\ \\
+  \sum_i M_i^{(a)}
+    = \left(\sum_i A_i^{(a)}\right) \cdot \bar B^{(a)} \cdot F^{(a)},
 \end{align}
@@
     \equiv \frac{\sum_{j\in S}\text{abun}_j(a) \cdot \text{efficiency}_j}{\text{abun}_S(a)}
+%    \\
+%  \bar B^{(a)}
+%    \equiv \frac{\sum_{i} A_i^{(a)} B_i}{\sum_i A_i^{(a)}}
 \end{align}
-is the _mean efficiency_ over all species in $S$ weighted by their abundance.
-<!-- The subscript $S$ serves as a reminder that many reads may not be assigned---for example, reads that fail to pass quality control or that come from the host or other taxonomic domains than what is being studied.
--->
+is the _sample mean efficiency_, the mean efficiency of all species in sample $a$ weighted by their abundance.
 
-## Relative abundance (proportions and ratios) {#relative-abundance}
+## Relative abundance {#relative-abundance}
 
@@
 %  \\&= \frac{\text{abun}_i(a)}{\sum_{i \in S}\text{abun}_i(a)}.
+%  P_{i}^{(a)}
+%    &\equiv \frac{A_i^{(a)}}{\sum A_i^{(a)}}.
 \end{align}
@@
   (\#eq:ratio)
-  \text{ratio}_{i/j}(a) = \frac{\text{abun}_i(a)}{\text{abun}_j(a)}.
+  \text{ratio}_{i/j}(a) \equiv \frac{\text{abun}_i(a)}{\text{abun}_j(a)}.
+%  R_{i/j}^{(a)} \equiv \frac{A_i^{(a)}}{A_j^{(a)}}.
 \end{align}
-Proportions and ratios each form the basis for popular relative DA methods.
 
-The proportion of a species is typically measured by its proportion of assigned reads,
+The measured proportion of a species is given by its proportion of all the assigned reads in a sample,
 \begin{align}
@@
 \end{align}
-Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency relative to the mean efficiency in the sample.
-Consequently, the species's proportion is measured as too high in samples that are dominated by species with lower efficiencies, and measured as too low in samples that are dominated by species with higher efficiencies.
-This phenomenon is illustrated in two hypothetical communities in Figure \@ref(fig:error-proportions). 
-Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.
-A demonstration in artificially constructed ('mock') communities of bacterial cells is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
+Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency divided by the mean efficiency in the sample.
+Since the mean efficiency varies across samples, so does the FE in a species' proportion.
+This phenomenon can be seen for Species 3 in the two hypothetical communities in Figure \@ref(fig:error-proportions). 
+Species 3, which has an efficiency of 6, is under-measured in Sample 1 (FE < 1) but over-measured (FE > 1) in Sample 2.
+This difference occurs because the even distribution of species Sample 1 yields a mean efficiency of 8.33; in contrast, the lopsided distribution in Sample 2, which is dominated by the low-efficiency Species 1, has a mean efficiency of just 3.15.
+A demonstration in bacterial mock communities is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
 
@@
 \end{align}
-Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
+Taxonomic bias creates FE in the ratio measurement that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
 For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
 A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
-
-Thus taxonomic bias creates variable fold errors in species proportions, but constant fold errors in species ratios.
-Consequently, a perfect cancellation of errors can only occur in ratio-based DA analysis.
 
@@
 
-(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species.
-Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.
+(ref:cap-error-proportions) **Taxonomic bias creates fold errors in species proportions that vary across samples and lead to inaccurate fold differences between samples.** Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species.
+Bottom row: Error in the measured fold difference in the third species that is derived from these measurements.
 Species' proportions may be measured as too high or too low depending on sample composition.
@@
 
-**Higher-order taxa:** How does bias affect the proportions and ratios of higher-order taxa, such as a genera, phyla, or even all organisms?
-More generally, we can identify a higher-order taxon $I$ with a set of species, $\{i \in I\}$.
-The abundance and reads of a taxon $I$ in a sample $a$ equal the sums over its constituent species, $\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)$ and $\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)$.
-The efficiency of taxon $I$ equals the weighted average of its constituent species,
+**Higher-order taxa:**
+We can consider a higher-order taxon $I$, such as a genus or phylum, as a set of species, $\{i \in I\}$.
+The abundance of taxon $I$ in sample $a$ is the sum of the abundances of its constituent species, $\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)$.
+Similarly, the read count of taxon $I$ is the sum $\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)$.
+The efficiency of taxon $I$ equals the abundance-weighted average of the efficiencies of its constituent species,
 \begin{align}
@@
 \end{align}
-Thus the efficiency of a higher-level taxon may vary across samples (@mclaren2019cons).
-An exception is if the constituent species have the same efficiency or always appear in the same abundances relative to each other, in which case the efficiency of the higher-order taxon would stay constant.
+If the constituent species have different efficiencies, then the efficiency of the higher-order taxon will vary across samples depending on the relative abundances of its constituents (@mclaren2019cons).
 As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
@@
 
-Various methods can be used to convert the relative abundances from MGS into absolute abundances.
-These methods are subject to taxonomic bias in the MGS measurements and in any supplemental, non-MGS measurements that may be used for the conversion.
-Appendix [REF] develops a general framework to understand how bias affects absolute-abundance measurements for species and higher-order taxa across these methods.
-Here we present our main results as they apply to currently-used methods.
+Several extensions of the standard MGS experiment make it possible to measure absolute species abundances.
+These extensions fall into two general approaches.
+The first approach leverages information about the abundance of the total community; for example, @vandeputte2017quan measured total-community abundance using flow cytometry and multiplied this number by MGS genus proportions to obtain the absolute abundances of individual genera (@vandeputte2017quan).
+A second approach leverages information about the abundance of one or more individual species; for example, a researcher might 'spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (@harrison2021theq).
+We consider each approach in detail to determine how taxonomic bias affects the resulting absolute-abundance measurements.
 
-### Methods that measure total-community abundance
+### Leveraging information about total-community abundance
 
-The definition \@ref(eq:prop) for the proportion of a species indicates that we can convert proportions to absolute abundances were we to know the absolute abundance of the total community.
-Given our measured proportions from MGS and a measure of total abundance from a supplemental, non-MGS measurement, $\widehat{\text{abun}}_{S}(a)$, one can measure the abundances of individual species by
+Suppose that the abundance of the total community in each sample has been measured by a non-MGS method.
+The absolute abundances of individual species can be obtained by multiplying the species proportions (obtained via MGS) by this total-abundance measurement,
 \begin{align}
@@
   &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{abun}}_S(a).
+%  \\ \\
+%  \hat A_i^{(a)} &= \hat P_i^{(a)} \cdot \widehat{\sum_i A_i^{(a)}}.
+%  \\ \\
+%  \hat A_i^{(a)} &= \hat P_i^{(a)} \cdot \widetilde{\sum_i A_i^{(a)}}.
 \end{align}
-Total-abundance measurements that have recently used for this purpose include counting cells with microscopy (@lloyd2020evid) or flow cytometry (@props2017abso,@vandeputte2017quan,@galazzo2020howt), measuring the concentration of a marker-gene with qPCR or ddPCR (@zhang2017soil,@barlow2020aqau, @galazzo2020howt,@tettamantiboshier2020comp), and measuring bulk DNA concentration with a florescence-based DNA quantification method (@contijoch2019gutm).
+Total-abundance measurements recently used for this purpose include counting cells with microscopy (@lloyd2020evid) or flow cytometry (@props2017abso, @vandeputte2017quan, @galazzo2020howt), measuring the concentration of a marker-gene with qPCR or ddPCR (@zhang2017soil, @barlow2020aqau, @galazzo2020howt, @tettamantiboshier2020comp), and measuring bulk DNA concentration with a florescence-based DNA quantification method (@contijoch2019gutm).
 
-Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
+Importantly, these methods of measuring total-community abundance are themselves subject to taxonomic bias.
 Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
-The marker-gene concentration measured by qPCR is affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency.
-We can easily understand the impact of taxonomic bias on the total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
+Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency.
+We can easily understand the impact of taxonomic bias on total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
 Suppose that each species has an _absolute efficiency_ for the total-abundance measurement that is constant across samples.
@@
 
-The species abundance measurements are affected by taxonomic bias in both the MGS and total-abundance measurement.
+The resulting absolute-abundance measurements of individual species are affected by taxonomic bias in both the MGS and total-abundance measurement.
 We can compute this effect by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
@@
   = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}.
+%  \\ \\
+%  \widetilde{\sum A_i^{(a)}}
+%  = \left(\sum A_i^{(a)}\right) \cdot \frac{B_i \cdot \bar B^{\text{tot},(a)}}{\bar B^{(a)}}.
 \end{align}
-Equation \@ref(eq:density-prop-error) indicates that the multiplicative error in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
-As in the case of proportions (Equation \@ref(eq:prop-error)), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
-On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to more stable errors.
+Equation \@ref(eq:density-prop-error) indicates that the FE in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
+As in the case of proportions (Equation \@ref(eq:prop-error)), the FE depends on sample composition through the two mean efficiency terms and so will vary across samples.
+Notably, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, then the two offset each other and lead to more stable FEs.
 We discuss how this possibility might be exploited in real experimental workflows in Section \@ref(solutions).
 
-### Methods that determine or measure a reference species
+### Leveraging information about a reference species
 
-The definition \@ref(eq:ratio) for the ratio between two species suggests another method to convert relative to absolute abundances: 
-By knowing the ratio of species $i$ to a _reference species_ $r$ and the abundance of species $r$, we can determine the abundance of species $i$.
-Thus given an MGS measurement and a measurement $\widehat{\text{abun}}_{r}(a)$ of the abundance of species $r$, we can measure the abundance of an arbitrary species $i$ as
+Suppose that the absolute abundance of a _reference species_ $r$ is known, either by experimental design or by direct measurement using a non-MGS method. 
+The absolute abundances of each species can be obtained by multiplying its read count by the abundance-to-reads ratio for species $r$,
 \begin{align}
@@
 \end{align}
-Here we consider some practical applications of this method and how their measurements are affected by bias in the MGS measurement and error in $\widehat{\text{abun}}_{r}(a)$.
-Appendix [REF] describes the general case where there is a set $R$ of multiple reference species; here we restrict our attention to a single reference species $r$.
 
-This approach has so far been solely or primarily used with reference species that have been determined to have an abundance that is constant across samples.
-In a spike-in experiment, researchers deliberately add (or 'spike in') one or more reference species at a known, constant abundance.
-For a single spike-in species $r$, the abundances of other species can be estimated by Equation \@ref(eq:density-ratio-meas).
-In the absence of spike-ins, researchers may be able to determine one or more native species whose abundances are thought to be constant across samples; we refer to such species as _housekeeping species_ by analogy with the housekeeping genes used for absolute-abundance conversion in studies of gene expression.
-Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference.
-A related approach involves computationally identifying species that are constant between pairs of samples (@david2014host) or between sample conditions (@mandal2015anal, @kumar2018anal) under added statistical assumptions about the existence of such species.
+A common application of this approach involves adding a 'spike-in' (as described above) in a known (and typically constant) abundance across samples (@stammler2016adju, @ji2019quan, @harrison2021theq, REFs).
+In this case, the reference abundance $\widehat{\text{abun}}_{r}(a)$ is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
+(In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \@ref(eq:density-ratio-meas); Appendix REF).
+
+Others have instead sought to determine naturally-occurring species that are thought to be constant across samples; we refer to such species as _housekeeping species_ by analogy with the housekeeping genes used for absolute-abundance conversion in gene-expression studies (REF).
+Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference (REFs).
+A related approach involves computationally identifying species that are constant between pairs of samples (@david2014host) or between sample conditions (@mandal2015anal, @kumar2018anal).
 The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set $\widehat{\text{abun}}_{r}(a)$ to 1 in Equation \@ref(eq:density-ratio-meas).
@@
 
-Equation \@ref(eq:density-ratio-meas) makes clear that the reference species need not be constant so long as we can directly measure it.
-This observation suggests a new approach to obtaining species absolute abundances, by combining MGS measurements of the community with targeted measurements of the absolute abundance of one or more individual species.
-Common targeted measurements include using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA, using primers tailored to a particular species.
-Other methods that are capable of measuring cells (rather than extracted DNA) include applying ddPCR directly to cells, performing CFU counting on selective media, and using flow cytometry with species-specific florescent probes.
-The measurements from any of these methods could be used for $\widehat{\text{abun}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) to obtain the absolute abundance for all species.
+We suggest an additional way of using the reference-species strategy even in the absence of a spike-in or constant species:
+Performing targeted measurements of the absolute abundance of one or more naturally occurring species.
+These species can then be used as reference species in Equation \@ref(eq:density-ratio-meas) to measure the absolute abundances of all species.
+The most common form of targeted measurement involves using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA.
+It is also possible to directly measure cell concentration by performing ddPCR prior to DNA extraction (REFs), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
 Appendix [REF] describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
 
-Despite the diversity of these methods, each can be represented by Equation \@ref(eq:density-ratio-meas), making it straightforward to understand the impact of taxonomic bias in the MGS measurement.
-It follows from Equation \@ref(eq:ratio-error) that the measured abundance of species $i$ is 
+The effect of taxonomic bias on absolute abundance measurements derived from a reference species can be determined by substituting Equation \@ref(eq:ratio-error) into Equation \@ref(eq:density-ratio-meas), which indicates that the measured abundance of species $i$ is
 \begin{align}
@@
 \end{align}
-The error thus consists of two terms: the ratio of the efficiency between species $i$ and species $r$, and any error in the abundance of the reference species.
-The ratio of species efficiencies is constant by assumption; thus the total error is constant if the error in the reference abundance is.
-Note, if a higher-order taxon is used as a reference---for example, using targeted measurements of a genus or family---then the efficiency of the reference taxon may vary across samples, leading to non-proportional errors.
-
-In practice, researchers commonly use a less-direct but equivalent approach to measure species abundances from a reference such as a spike-in.
-Set $S$ to be the species of interest excluding the spike-in species $r$.
-In this approach, one measures the total abundance of the community $S$ from the ratio of non-spike-in to spike-in reads,
-\begin{align}
-  (\#eq:total-density-spike)
-  \widehat{\text{abun}}_S(a) 
-  &= \frac{\text{reads}_S(a) \cdot \widehat{\text{abun}}_R(a)}{\text{reads}_r(a)}.
-\end{align}
-Then, one uses this measurement $\widehat{\text{abun}}_{S}(a)$ along with the MGS proportions in Equation \@ref(eq:density-prop-meas) to determine the abundance of individual species.
-This calculation yields measurements that are identical to those from directly applying Equation \@ref(eq:density-ratio-meas) (Appendix \@ref(total-density-ref)).
-This observation might seem paradoxical: Equation \@ref(eq:density-ratio-error) indicates proportional error for the direct approach, whereas \@ref(eq:density-prop-error) suggests non-proportional errors for the indirect approach.
-This apparent contradiction is resolved by the fact the measurement $\widehat{\text{abun}}_{S}(a)$ by this method has errors that are proportional to $\text{efficiency}_S(a)$ and thus exactly offset the proportionality with $1/\text{efficiency}_S(a)$ in the MGS proportions, such that the measured species abundances have constant fold errors.
-
-## Summary
-
-Taxonomic bias in MGS measurements makes all MGS-based measures of relative and absolute abundance inaccurate.
-For some measures---but not all---consistent bias at the species level creates proportional errors in species abundances.
-Of two types of relative abundance, the ratio between a pair of species has proportional errors whereas the proportion of an individual species does not. 
-Instead, the multiplicative error in a species' proportion varies with the mean efficiency across samples.
-Methods for measuring species absolute abundance that involve measuring the abundance of the total community likewise experience non-proportional errors which vary with the sample mean efficiency; 
-a notable exception is if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement.
-Absolute-abundance methods that instead involve measuring or spiking individual reference species yield proportional errors, provided that the abundance of the reference species can be determined up to a proportional error.
-The efficiency of higher-order taxa may vary across samples even if the efficiencies of their constituent species are constant; as a result, all standard abundance measures can have non-proportional errors at taxonomic resolutions higher than that at which bias is conserved.
-
-<!-- Possibly:  -->
-<!-- - taxonomic aggregates need not have proportional errors -->
-<!-- - the fundmental mathematical explanation spanning these results is that the sum of multiple values each having proportional error need not itself have proportional error. -->
+The FE in $\widehat{\text{abun}}_{i}(a)$ consists of two terms: the relative efficiency of species $i$ to species $r$ in the MGS measurement, and the FE in the reference species' abundance.
+If bias is consistent across samples and the reference species can be spiked in or measured with a constant FE, then the FE in the $\widehat{\text{abun}}_{i}(a)$ will be constant.

### 2022-05-27 differential-abundance.Rmd
Revise differential-abundance section (incomplete)
Revised to update terminology and address some of BJCs comments. The
section on FDs in absolute abundance is still incomplete, and the
paragraph describing the regression simulation results still needs
improvement.

@@
-# DA methods vary in their robustness to bias {#differential-abundance}
+# How bias affects DA results {#differential-abundance}
 
-<!-- # Differential-abundance methods vary in their robustness to taxonomic bias  -->
+<!-- Previous title: Differential-abundance methods vary in their robustness to taxonomic bias  -->
 
-We now consider how the error in MGS-derived abundance measurements affect DA analyses.
-We focus on multiplicative DA analyses, or those that measure the fold change (FC) in the relative or absolute abundance of a taxon.
-Such analyses are widely used in microbiome studies and have more direct ecological interpretations than other types via the multiplicative processes of exponential growth and decay.
-In addition, we briefly consider non-parametric rank-based analyses, which are common in microbiome-wide association studies.
+We now turn to how the errors in MGS-based abundance measurements affect DA results.
+Our focus is on _multiplicative DA_---the fold differences (FD) in the relative or absolute abundance of species between samples or between experimental conditions.
+Multiplicative DA has a direct ecological interpretation via the multiplicative processes of exponential growth and decay and forms the basis for several popular DA methods (REF DESeq2, ALDEx2(?), and corncob; a LV paper; fido).
+We also consider non-parametric rank-based methods; these methods are popular in case-control studies for which ecological interpretability is often of secondary importance to discovering stable associations between species and the condition of interest (REFs).
 
@@
 
-The building blocks of a multiplicative DA analysis are the FCs in relative or absolute abundance between pairs of individual samples.
-If the fold error (FE) in a given abundance measure is constant, then it will not affect the FCs among samples.
+The building blocks of a multiplicative DA analysis are the measured FDs in a species' abundance between pairs of individual samples.
+<!-- The FE in the measured FD between two samples is given by the ratio of the FEs for the individual-sample measurements. -->
+If the fold error (FE) in a species' abundance measurement is constant across samples, then it will not affect the measured FDs between pairs of samples.
 <!-- Ratios -->
-We saw, for example, that consistent bias creates constant FE in the ratio between two species $i$ and $j$ equal to the ratio in their efficiencies (Equation \@ref(eq:ratio-error)).
-This error completely cancels in the FC from sample $a$ to $b$,
+In Section \@ref(abundance-measurement), we showed that consistent taxonomic bias creates a constant FE in the ratio between two species $i$ and $j$, equal to the ratio in their efficiencies (Equation \@ref(eq:ratio-error)).
+This error completely cancels when calculating the FD of this ratio from sample $a$ to $b$,
 \begin{align}
   (\#eq:ratio-fc-error)
-\underbrace{\frac{\widehat{\text{ratio}}_{i/j}(b)}{\widehat{\text{ratio}}_{i/j}(a)}} _\text{measured FC}
+\underbrace{\frac{\widehat{\text{ratio}}_{i/j}(b)}{\widehat{\text{ratio}}_{i/j}(a)}} _\text{measured FD}
   &= \frac
@@
   &=
-  \underbrace{\frac{\text{ratio}_{i/j}(b)}{\text{ratio}_{i/j}(a)}}_\text{actual FC}
+  \underbrace{\frac{\text{ratio}_{i/j}(b)}{\text{ratio}_{i/j}(a)}}_\text{actual FD}
   .
@@
 <!-- Proportions -->
-In contrast, the variable FE in the proportion of a species (Equation \@ref(eq:prop-error)) does not completely cancel,
+In contrast, bias created a FE in the proportion of a species (Equation \@ref(eq:prop-error)) that varied inversely with the mean efficiency of the sample.
+Therefore, this error does not cancel when calculating the FD in the proportion,
 \begin{align}
   (\#eq:prop-fc-error)
-\underbrace{\frac{\widehat{\text{prop}}_{i}(b)}{\widehat{\text{prop}}_{i}(a)}} _\text{measured FC}
+\underbrace{\frac{\widehat{\text{prop}}_{i}(b)}{\widehat{\text{prop}}_{i}(a)}} _\text{measured FD}
   &= \frac
@@
   &=
-  \underbrace{\frac{\text{prop}_{i}(b)}{\text{prop}_{i}(a)}}_\text{actual FC}
+  \underbrace{\frac{\text{prop}_{i}(b)}{\text{prop}_{i}(a)}}_\text{actual FD}
   \cdot
@@
 \end{align}
-Although the constant species' efficiency cancels, the varying mean efficiency does not, leaving a FE in the FC that is the same for every species and is equal to the inverse FC in the mean efficiency.
+The varying mean efficiency does not cancel, leaving an FE in the FD of the species' proportion that is equal to the inverse FD in the mean efficiency.
+Notably, this FE is the same for every species.
 
-Figure \@ref(fig:error-proportions) (bottom row) illustrates these different effects of bias on FCs in proportions versus ratios.
-In this hypothetical example, the mean efficiency decreases by a factor of 2.6 from Sample 1 to Sample 2 (FC of 0.4X), causing the FC in the proportion of each species to appear 2.6X larger than its true value.
-Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction; we refer to the latter as a _sign error_ in reference to the corresponding log fold change (LFC).
+The different effects of taxonomic bias on the measured FDs in proportions versus ratios are illustrated in a hypothetical example in Figure \@ref(fig:error-proportions) (bottom row).
+In this example, the mean efficiency decreases 2.6-fold from Sample 1 to Sample 2 (FD of 0.4), causing the FD in the proportion of each species to appear 2.6-fold larger than its true value.
+Though the FE in the FD is the same for each species, the biological implications of the error vary.
+In particular, there are three distinct types of error: an increase in the magnitude of the measured FD, a decrease in that magnitude, and a change in the direction of the FD.
+We refer to the latter as a _sign error_ in reference to the sign of the measured log fold difference (LFD).
 We can see each type of error in Figure \@ref(fig:error-proportions).
-For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FC (actual FC: 2.3X, measured FC: 6.5X).
-For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
-For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase (a sign error).
-In contrast, the ratios between species change by the same factors in the actual and observed taxonomic profiles.
-For example, the ratio of Species 2 to Species 3 changes from $1/1$ to $1/4$ in the actual profiles and from $3$ to $3/4$ in the observed profiles, a decrease by 4X either way.
+For Species 1, which increases in abundance and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FD (actual FD: 2.3, measured FD: 6.5).
+For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FD: 0.15, measured FD: 0.44).
+For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FD: 0.6, measured FD: 1.8), such that the species actually appears to increase (a sign error).
+In contrast, the ratios between species change by the same factors in the actual and measured taxonomic profiles.
+For example, the ratio of Species 2 to Species 3 changes from $1/1$ to $1/4$ in the actual profiles and from $3$ to $3/4$ in the measured profiles, a decrease by 4-fold either way.
 
@@
 
-We can similarly understand the impact of bias on the measured FCs in a species' absolute abundance.
-To the extent that the FE in the abundance measurement is constant across samples, it will cancel and lead to accurate FCs.
-For the various measurement methods described in Section \@ref(absolute-abundance), the FE in a species' abundance can be expressed as a species-specific constant multiplied by a species-independent factor that, depending on experimental details, may vary with sample composition.
-Thus the FE in the FC between a pair of samples is the same for all species.
-Further, it can be reduced by experimental designs that minimize cross-sample variation in the species-independent factor, regardless of the magnitude of bias (variation in efficiency among species) in the MGS measurement.
+### AA - need to revise
+
+Taxonomic bias affects the measured FD in a species' absolute abundance in a similar fashion.
+Using Equation \@ref(\#eq:density-prop-error) or Equation \@ref(\#eq:density-ratio-error), we can compute the FE in the measured FD for absolute abundances derived from the two approaches discussed in Section \@ref(absolute-abundance).
+For the abundance of species $i$ derived from a total-abundance measurement, the measured FD is
+\begin{align}
+  (\#eq:aa-tot-fc-error)
+  FD\left[\widehat{\text{abun}}_{i}(a) \right]
+  = FD\left[\text{abun}_{i}(a)\right] \cdot FD \left[ \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}\right].
+\end{align}
+In this case, the FE in the FD equals the FD in the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
+For the abundance of species $i$ derived from a reference-species measurement, the measured FD is
+\begin{align}
+  (\#eq:aa-tot-fc-error)
+  FD\left[\widehat{\text{abun}}_{i}(a) \right]
+  = FD\left[\widehat{\text{abun}}_{r}(a)\right].
+\end{align}
+In this case, the FE in the FD equals the FD in the FE of the reference species' abundance measurement.
+If the FE in the $\widehat{\text{abun}}_{i}(a)$ is constant then the FD will be accurate.
+
+HERE - the above needs work, and I need to consider if equations are needed (and what type).
+Perhaps a summary table is better?
+
+#### Notes
+
+Note for revision
+- complementary measurement: mitigates variation in SME by offsetting with that of the total abundance measurement
+- reference species: mitigates variation in SME by using species ratios, avoiding additive aggregates which introduce variation in efficiency.
+
+for complementary AA, point out that the FE is equal to the ratio of FCs in the two MEs (or equiv in the FC in the ratio of MEs) and hence changes in one ME can be offset by changes in the other, if they both change in a similar way.
+
+for ref taxa, state how there is a constant FE and thus cancels.
 
@@
 
-DA analysis can be framed as a regression analysis, the simplest of which uses the simple linear regression model
+Most DA analyses can be understood as a regression analysis.
+For example, the simple linear regression model,
 \begin{align}
@@
 \end{align}
-describing how a microbial _response variable_ $y$ relates to a _covariate_ $x$ and _residual (or unexplained) error_ $\varepsilon$.
-For example, $y$ may be the log absolute abundance of species $i$ and $x$ may be a continuous variable, such as pH, or a binary variable, such as $x=1$ for treated patients and $x=0$ for untreated controls.
+can be used to relate a microbial abundance variable $y$ to a _covariate_ $x$ and a _residual (or unexplained) error_ $\varepsilon$.
+For example, $y$ may be the log absolute abundance of species $i$ and $x$ may be a continuous variable such as pH; or, $x$ may be a binary variable, such as $x=1$ for treated patients and $x=0$ for untreated controls.
 The regression coefficients $\alpha$ and $\beta$ are parameters to be estimated from the data.
@@
 For a binary covariate, $\beta$ equals the average difference in $y$ between samples with $x=1$ versus those with $x=0$.
-Taking the response $y$ to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient $\beta$ corresponding to the average log fold change (LFC) in abundance between conditions or for a unit increase in $x$.
+Taking the response $y$ to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient $\beta$ corresponding to the average log fold difference (LFD) in abundance between conditions or for a unit increase in $x$.
 
-Appendix \@ref(appendix-regression) derives mathematical results for the effect that measurement error in the response caused by taxonomic bias has on LFCs estimated via simple linear regression.
-Here we illustrate these results with a example.
-Suppose that we wish to estimate the average LFC in the absolute abundance of every species between samples from two conditions, such as treated and untreated patients.
+Appendix \@ref(appendix-regression) derives mathematical results for the effect that measurement error in the response caused by taxonomic bias has on LFDs estimated via simple linear regression.
+Here we illustrate these results with an example.
+Suppose that we wish to estimate the average LFD in the absolute abundance of every species between samples from two conditions, such as treated and untreated patients.
 Further suppose that we have measured absolute abundances by multiplying the proportions from MGS with an accurate measurement of total-community abundance.
@@
 
-To understand how the FE in species abundances affects the estimated LFC for a given species, it is useful to consider three possible scenarios for how the mean efficiency varies across samples.
+To understand how the FE in species abundances affects the estimated LFD for a given species, it is useful to consider three possible scenarios for how the mean efficiency varies across samples.
 In the first scenario, the mean efficiency remains stable across samples, such that the FE in $\widehat{\text{abun}}_{i}(a)$ is approximately constant. 
-In this case, the FE induced by bias _mostly cancels_ and does not substantially affect the estimated LFC.
+In this case, the FE induced by bias _mostly cancels_ and does not substantially affect the estimated LFD.
 In the second scenario, the mean efficiency varies substantially but is not associated with the condition, creating random FEs in the individual sample measurements.
-In this case, the LFCs between individual samples will be highly distorted, but these errors will effectively average out in the regression and the estimated LFC remains reliable.
-Although the random FEs caused by bias do not create a systematic error in the LFC estimate, they will tend to reduce the precision (i.e., increase the standard error) of the estimate.
-This reduced precision can be offset, however, by increasing the number of samples.
-In the third scenario, the mean efficiency systematically varies with the condition, causing a systematic error in the estimated LFC regardless of the number of samples.
-In particular, the portion of additive variation in the log mean efficiency that is linearly associated with $x$ creates an equal but opposite additive error in the estimated LFC for each species.
-Although the absolute additive error in the LFCs is the same for all species, its biological significance varies, depending on its magnitude relative to that of the species' true LFC.
+In this case, the LFDs between individual samples will be highly distorted, but these errors will effectively average out in the regression and the estimated LFD remains reliable.
+Although the random FEs caused by bias do not create a systematic error in the LFD estimate, they will tend to reduce the precision (i.e., increase the standard error) of the estimate.
+In the third scenario, the mean efficiency systematically varies with the condition, causing a systematic error in the estimated LFD regardless of the number of samples.
+In particular, the portion of additive variation in the log mean efficiency that is linearly associated with $x$ creates an equal but opposite additive error in the estimated LFD for each species.
+Although the absolute additive error in the LFDs is the same for all species, its biological significance varies, depending on its magnitude relative to that of the species' true LFD.
 
-Figure \@ref(fig:regression-example) shows a simulated example of the problematic third scenario.
-Here, the average log mean efficiency is higher in the condition corresponding to $x=1$ due to an increase of the high-efficiency Species 9.
-This systematic increase in the log mean efficiency causes a decrease in the estimated LFC for each species by an equal amount.
-As we saw for the FC between individual samples, the biological significance varies by species, despite the absolute error in the LFC being shared.
-We see decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending the value of the actual LFC relative to the error.
+Figure \@ref(fig:regression-example) shows a simulated example in which the mean efficiency varies systematically with the condition of interest.
+Here, the average log mean efficiency is higher in the condition corresponding to $x=1$, due to an increase of the high-efficiency Species 9.
+This systematic increase in the log mean efficiency causes an arteficial decrease in the estimated LFD for each species by an equal amount.
+As we saw for the FD between individual samples, the biological significance of the systematic error in the LFD estimates varies by species, despite the absolute error being constant across species.
+We see decreases in the magnitude of the LFD estimate (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending the value of the actual LFD relative to the error.
 The portion of the variation in the log mean efficiency that is uncorrelated with $x$ increases the standard errors for most species (Figure \@ref(fig:regression-example) D).
-The exception is Species 9, which is the primary driver of the the mean efficiency dynamics in this example.
+The exception is Species 9, which is the primary driver of the mean efficiency dynamics in this example.
 Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).
@@
 
-Researchers often prefer to use nonparametric, rank-based DA methods to determine how species vary across sample conditions.
-Rather than estimating LFCs, these methods estimate the average change in the cross-sample rank of a species' relative or absolute abundance.
-Popular rank-based methods in microbiome DA analysis include the Wilcoxon-Mann-Whitney and Kruskal-Wallis tests when the covariate $x$ is discrete covariate and Spearman's rank correlation when $x$ is continuous.
-Rank-based DA methods are unaffected by a constant FE in the response prior to rank transformation, since such an error does not affect the cross-sample ranks.
-Hence consistent taxonomic bias does not affect rank-based analyses of species ratios.
-However, variable FE in the untransformed response (as seen in species proportions) can change the cross-sample ranks.
-If associated with the covariate $x$, the resulting error in the ranks will create a systematic error in the DA result.
+Rank-based or 'nonparametric' DA methods work by first applying a cross-sample rank-transformation to the abundance variable $y$: The sample in which $y$ has the smallest value receives a rank of 1, the sample with the second smallest value receives a rank of 2, etc.
+These methods then analyze how the average rank varies across conditions.
+Rank-based methods commonly used in microbiome DA analysis include the Wilcoxon-Mann-Whitney and Kruskal-Wallis tests for a discrete covariate (e.g.
+case vs control) and Spearman's rank correlation for a continuous covariate (e.g.
+pH).
+They have been applied to species proportions (REFs), ratios (REFs), and absolute abundances (REFs).
+
+Multiplying a variable by a constant factor does not change its cross-sample ranks; thus rank-based DA methods are unaffected by constant FE in the abundance variable $y$.
+We therefore conclude that consistent taxonomic bias does not affect rank-based DA analyses of species ratios or absolute abundances derived from a reference species.
+In contrast, variable FE in $y$ (driven by variation in the mean efficiency) can affect the cross-sample ranks; if the FE is associated with the covariate $x$, it can cause systematic error in the DA result.
+Thus consistent bias may impact rank-based analyses of species proportions and absolute abundances derived from a total-community abundance measurement.
+Future work should consider whether rank-based DA methods are significantly more robust than multiplicative DA methods to variation in the mean efficiency.

### 2022-05-27 _cut.Rmd
Revise measurement and diff-abun sections
- Change to new mathematical notation
- Improve clarity of presentation around equations, definitions, and
  results
- Add new treatment of FDs between pairs of samples for absolute
  abundances
- Add latex shortcuts to preamble for new notation

@@
 The efficiency of higher-order taxa may vary across samples even if the efficiencies of their constituent species are constant; as a result, all standard abundance measures can yield variable FEs at taxonomic resolutions higher than that at which bias is conserved.
+
+# How bias affects DA results {#differential-abundance}
+
+The previous section showed that for some methods of measurement, taxonomic bias creates constant FEs in the (relative or absolute) abundance of a species; but in others, bias creates FEs that vary inversely with the mean efficiency of the sample.
+This section describes how these constant or varying FEs affect DA results, finding conditions for significant inferential errors.
+
+The fact that some measures have constant FEs while others do not suggests that these DA analyses based on these measures will be more or less robust to bias.
+Here we rigorously evaluate this idea / explore in detail, finding precise conditions for significant error in DA results.

### 2022-05-27 abundance-measurement.Rmd
Revise measurement and diff-abun sections
- Change to new mathematical notation
- Improve clarity of presentation around equations, definitions, and
  results
- Add new treatment of FDs between pairs of samples for absolute
  abundances
- Add latex shortcuts to preamble for new notation

@@
 This model is the simplest that respects the multiplicative nature of taxonomic bias and the *compositional* nature of MGS measurements.
-The actual abundance of a species in a given sample, multiplied by its measurement efficiency---its rate of conversion from cells to taxonomically assigned sequencing reads---determines the species' read count in that sample.
+The actual abundance of a species in a given sample, multiplied by its _measurement efficiency_---its rate of conversion from cells to taxonomically assigned sequencing reads---determines the species' read count in that sample.
 Taxonomic bias presents as variation in the measurement efficiencies among species within an MGS experiment.
-The read counts further depend on sample-specific experimental factors that are typically unknown, such that they are best interpreted as only providing relative abundances (such data is said to be _compositional_) (@gloor2017micr).
+The read counts further depend on sample-specific experimental factors that are typically unknown, such that they are best interpreted as only providing relative abundances (such data is said to be _compositional_; @gloor2017micr).
 
@@
 
-Our model stipulates that the taxonomically-assigned read count for a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
+Our model stipulates that the taxonomically-assigned read count for a species $i$ in a sample $a$ equals its abundance $A_{i}^{(a)}$ multiplied by a species-specific factor $B_{i}$ and a sample-specific factor $F^{(a)}$,
 \begin{align}
   (\#eq:mgs-model)
-  \text{reads}_{i}(a)
-  = \text{abun}_{i}(a) \quad \cdot
-    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
-    \cdot \quad
-    \underbrace{\text{effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
-%  \\ \\
-%  M_i^{(a)} = A_i^{(a)} B_i F^{(a)}.
+  M_i^{(a)} = A_i^{(a)} B_i F^{(a)}.
 \end{align}
-The species-specific factor, $\text{efficiency}_{i}$, equals the *relative measurement efficiency* (or simply *efficiency*) of the species relative to an arbitrary baseline species (@mclaren2019cons).
+The species-specific factor, $B_{i}$, is the *relative measurement efficiency* (or simply *efficiency*) of the species relative to an arbitrary baseline species (@mclaren2019cons).
 The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
-The sample-specific factor, $\text{effort}(a)$, quantifies the *effective experimental effort* per unit abundance in measuring the sample.
-Equation \@ref(eq:mgs-model) implies that the total number of assigned reads in sample $a$ equals
+The sample-specific factor, $F^{(a)}$, describes the effective sequencing effort for that sample; it equals the number of reads per unit abundance that would be obtained for a species with an efficiency of 1.
+We can write the total number of assigned reads for the sample as
 \begin{align}
   (\#eq:total-reads)
-  \text{reads}_S(a)
-    = \text{abun}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
-    \\ \\
-  \sum_i M_i^{(a)}
-    = \left(\sum_i A_i^{(a)}\right) \cdot \bar B^{(a)} \cdot F^{(a)},
+  M_\tot^{(a)}
+    = A_\tot^{(a)} \bar B^{(a)} F^{(a)},
 \end{align}
-where $\text{abun}_{S}(a) \equiv \sum_{j\in S}\text{abun}_j(a)$ is the total abundance of species in $S$ and 
+where 
+$M_{\tot}^{(a)} \equiv \sum_{j\in S} M_j^{(a)}$ is the total read count and $A_{\tot}^{(a)} \equiv \sum_{j\in S}A_j^{(a)}$ is the total abundance for all species $S$, and 
 \begin{align}
   (\#eq:mean-efficiency)
-  \text{efficiency}_S(a) 
-    \equiv \frac{\sum_{j\in S}\text{abun}_j(a) \cdot \text{efficiency}_j}{\text{abun}_S(a)}
-%    \\
-%  \bar B^{(a)}
-%    \equiv \frac{\sum_{i} A_i^{(a)} B_i}{\sum_i A_i^{(a)}}
+  \bar B^{(a)} \equiv \frac{\sum_{i\in S} A_i^{(a)} B_i}{\sum_{i\in S} A_i^{(a)}}
 \end{align}
-is the _sample mean efficiency_, the mean efficiency of all species in sample $a$ weighted by their abundance.
+is the _sample mean efficiency_, defined as the mean efficiency of all species weighted by their abundance.
 
@@
 We distinguish between two types of species-level *relative abundances* within a sample.
-The *proportion* of species $i$ in sample $a$ equals its abundance relative to the total abundance of all species in $S$,
+The *proportion* $P_{i}^{(a)}$ of species $i$ in sample $a$ equals its abundance relative to the total abundance of all species in $S$,
 \begin{align}
   (\#eq:prop)
-  \text{prop}_{i}(a) 
-    &\equiv \frac{\text{abun}_i(a)}{\text{abun}_S(a)}.
-%  \\&= \frac{\text{abun}_i(a)}{\sum_{i \in S}\text{abun}_i(a)}.
-%  P_{i}^{(a)}
-%    &\equiv \frac{A_i^{(a)}}{\sum A_i^{(a)}}.
+  P_{i}^{(a)} &\equiv \frac{A_i^{(a)}}{A_\tot^{(a)}}.
 \end{align}
-The *ratio* between two species $i$ and $j$ equals the abundance of $i$ relative to that of $j$,
+The *ratio* $R_{i/j}^{(a)}$ between two species $i$ and $j$ equals the abundance of $i$ divided by that of $j$,
 \begin{align}
   (\#eq:ratio)
-  \text{ratio}_{i/j}(a) \equiv \frac{\text{abun}_i(a)}{\text{abun}_j(a)}.
-%  R_{i/j}^{(a)} \equiv \frac{A_i^{(a)}}{A_j^{(a)}}.
+  R_{i/j}^{(a)} \equiv \frac{A_i^{(a)}}{A_j^{(a)}}.
 \end{align}
@@
   (\#eq:prop-meas)
-  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
+  \tilde P_{i}^{(a)} &\equiv \frac{M_i^{(a)}}{M_\tot^{(a)}}.
 \end{align}
-We can rewrite the right-hand side (using Equations \@ref(eq:mgs-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas)) to find
+We use the tilde to distinguish the measurement from the actual quantity being measured.
+From Equations \@ref(eq:mgs-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas), it follows that the measured and actual proportion are related by
 \begin{align}
   (\#eq:prop-error)
-  \widehat{\text{prop}}_{i}(a)
-  &= \text{prop}_{i}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}}_{\substack{\text{variable} \\ \text{fold error}}}.
+  \tilde P_{i}^{(a)} &= P_{i}^{(a)} \cdot \frac{B_i}{\bar B^{(a)}}.
 \end{align}
-Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency divided by the mean efficiency in the sample.
-Since the mean efficiency varies across samples, so does the FE in a species' proportion.
+<!-- Taxonomic bias thus creates a fold-error (FE) in the measured proportion $\tilde P_{i}^{(a)}$ of species $i$ equal to the efficiency $B_i$ of species $i$ $divided by the mean efficiency $\bar B^{(a)}$ in the sample.
+-->
+Taxonomic bias thus creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
+Since the mean efficiency varies across samples, so does the FE.
 This phenomenon can be seen for Species 3 in the two hypothetical communities in Figure \@ref(fig:error-proportions). 
@@
 
-The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
+The measured ratio $\tilde R_{i/j}^{(a)}$ between species $i$ and $j$ is given by the ratio of their read counts, 
 \begin{align}
   (\#eq:ratio-meas)
-  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
+  \tilde R_{i/j}^{(a)} \equiv \frac{M_{i}^{(a)}}{M_{j}^{(a)}}.
 \end{align}
-From Equations \@ref(eq:mgs-model) and \@ref(eq:ratio-meas), it follows that 
+From Equations \@ref(eq:mgs-model) and \@ref(eq:ratio-meas), it follows that the measured and actual ratio are related by
 \begin{align}
   (\#eq:ratio-error)
-  \widehat{\text{ratio}}_{i/j}(a)
-%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
-  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
+  \tilde R_{i/j}^{(a)} = R_{i/j}^{(a)} \cdot \frac{B_i}{B_j}.
 \end{align}
-Taxonomic bias creates FE in the ratio measurement that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
+Taxonomic bias thus creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
 For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
@@
 We can consider a higher-order taxon $I$, such as a genus or phylum, as a set of species, $\{i \in I\}$.
-The abundance of taxon $I$ in sample $a$ is the sum of the abundances of its constituent species, $\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)$.
-Similarly, the read count of taxon $I$ is the sum $\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)$.
-The efficiency of taxon $I$ equals the abundance-weighted average of the efficiencies of its constituent species,
+The abundance of taxon $I$ in sample $a$ is the sum of the abundances of its constituent species, $A_{I}^{(a)} \equiv \sum_{i \in I} A_{i}^{(a)}$.
+Similarly, the read count of taxon $I$ is the sum $M_{I}^{(a)} \equiv \sum_{i \in I} M_{i}^{(a)}$.
+We further define the efficiency of taxon $I$ as the abundance-weighted average of the efficiencies of its constituent species,
 \begin{align}
   (\#eq:efficiency-general)
-  \text{efficiency}_I(a) 
-    \equiv \frac{\sum_{i\in I}\text{abun}_i(a)\cdot \text{efficiency}_i}{\text{abun}_I(a)}.
+  B_I^{(a)} \equiv \frac{\sum_{i\in I} A_{i}^{(a)} B_{i}}{\sum_{i\in I} A_{i}^{(a)}}.
 \end{align}
-If the constituent species have different efficiencies, then the efficiency of the higher-order taxon will vary across samples depending on the relative abundances of its constituents (@mclaren2019cons).
+With these definitions, the read count for taxon $I$ can be expressed as
+$M_{I}^{(a)} = A_{I}^{(a)} B_I^{(a)} F^{(a)}$.
+Thus $B_I^{(a)}$ plays a role analogous to the efficiency of an individual species, but differs in that it need not be constant across samples:
+If the constituent species have different efficiencies, then the efficiency of the higher-order taxon $I$ depends on the relative abundances of its constituents and so will tend to vary across samples (@mclaren2019cons).
 As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
 The efficiency of the phylum would then be $\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5$ in Sample 1 and $\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1$ in Sample 2.
-Equations \@ref(eq:prop-error) and \@ref(eq:ratio-error) continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent taxon efficiency defined by Equation \@ref(eq:efficiency-general) is used.
-In this way, we can see that both proportions and ratios among higher-order taxa may have inconsistent fold errors.
+Equations \@ref(eq:prop-error) and \@ref(eq:ratio-error) continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent, higher-order taxa efficiencies $B_I^{(a)}$ and $B_J^{(a)}$ are used.
+In this way, we see that both proportions and ratios among higher-order taxa may have inconsistent FEs.
 
@@
 
-Suppose that the abundance of the total community in each sample has been measured by a non-MGS method.
-The absolute abundances of individual species can be obtained by multiplying the species proportions (obtained via MGS) by this total-abundance measurement,
+Suppose that the total abundance of all species in the sample, $A_{\tot}^{(a)}$, has been measured by a non-MGS method, yielding a measurement $\tilde A_\tot^{(a)}$.
+The absolute abundance of an individual species can be measured by multiplying the species' proportion from MGS by this total-abundance measurement,
 \begin{align}
   (\#eq:density-prop-meas)
-  \widehat{\text{abun}}_{i}(a) 
-  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{abun}}_S(a).
-%  \\ \\
-%  \hat A_i^{(a)} &= \hat P_i^{(a)} \cdot \widehat{\sum_i A_i^{(a)}}.
-%  \\ \\
-%  \hat A_i^{(a)} &= \hat P_i^{(a)} \cdot \widetilde{\sum_i A_i^{(a)}}.
+  \tilde A_i^{(a)} &= \tilde P_i^{(a)} \tilde A_\tot^{(a)}.
 \end{align}
@@
 
-Importantly, these methods of measuring total-community abundance are themselves subject to taxonomic bias.
+Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
 Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
@@
 We can easily understand the impact of taxonomic bias on total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
-Suppose that each species has an _absolute efficiency_ for the total-abundance measurement that is constant across samples.
-Neglecting other sources of random and systematic error, the total-abundance measurement equals
+Suppose that each species $i$ has an _absolute efficiency_ $B_{i}^{\mtot}$ for the total-abundance measurement that is constant across samples.
+Further, let $\bar B^{\mtot (a)}$ be the abundance-weighted average of these efficiencies in sample $a$---that is, the mean efficiency of the total-abundance measurement.
+Neglecting other error sources, the total-abundance measurement equals
 \begin{align}
   (\#eq:total-density-error)
-  \widehat{\text{abun}}_S(a) 
-  &= \sum_{i\in S} \text{abun}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
-\\&= \text{abun}_S(a) \cdot \text{efficiency}^{\text{tot}}_S(a),
+  \tilde A_\tot^{(a)} 
+  &= \sum_{i\in S} A_i^{(a)} B_{i}^{\mtot}
+  \\&= A_\tot^{(a)} \bar B^{\mtot (a)}.
 \end{align}
-where $\text{efficiency}_{i}^{\text{tot}}(a)$ is the absolute measurement efficiency of species $i$ for the total-abundance measurement and $\text{efficiency}^{\text{tot}}_S(a)$ is the mean efficiency of the total-abundance measurement in the sample.
-<!-- Note, we have assumed that the total-abundance and MGS measurements are sensitive to the same species -->
+<!-- Note: We have assumed that only species in S contribute to the total abundance measurement. -->
 
-The resulting absolute-abundance measurements of individual species are affected by taxonomic bias in both the MGS and total-abundance measurement.
-We can compute this effect by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
+Species abundance measurements derived by this method are affected by taxonomic bias in both the MGS and total-abundance measurement.
+We can determine the resulting fold error (FE) by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
 \begin{align}
   (\#eq:density-prop-error)
-  \widehat{\text{abun}}_{i}(a) 
-  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}.
-%  \\ \\
-%  \widetilde{\sum A_i^{(a)}}
-%  = \left(\sum A_i^{(a)}\right) \cdot \frac{B_i \cdot \bar B^{\text{tot},(a)}}{\bar B^{(a)}}.
+  \tilde A_\tot^{(a)}
+  = A_\tot^{(a)} \cdot \frac{B_i \bar B^{\mtot (a)}}{\bar B^{(a)}}.
 \end{align}
 Equation \@ref(eq:density-prop-error) indicates that the FE in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
-As in the case of proportions (Equation \@ref(eq:prop-error)), the FE depends on sample composition through the two mean efficiency terms and so will vary across samples.
-Notably, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, then the two offset each other and lead to more stable FEs.
-We discuss how this possibility might be exploited in real experimental workflows in Section \@ref(solutions).
+As in the case of proportions (Equation \@ref(eq:prop-error)), the FE depends on sample composition through the two mean efficiency terms and so will vary across samples unless the two perfectly covary.
 
@@
 
-Suppose that the absolute abundance of a _reference species_ $r$ is known, either by experimental design or by direct measurement using a non-MGS method. 
-The absolute abundances of each species can be obtained by multiplying its read count by the abundance-to-reads ratio for species $r$,
+Suppose that the absolute abundance of a _reference species_ $r$ has been fixed by the experimenter or been measured by independent means.
+This known or measured abundance $\tilde A_{r}^{(a)}$ can be used in conjunction with the MGS read counts to obtain absolute abundances for all species.
+In the absence of taxonomic bias, the ratio of a species' absolute abundance to its MGS read count is the same for all species in a given sample (Equation \@ref{eq:mgs-model}).
+Hence the known ratio for the reference species can serve as conversion factor for obtaining the absolute abundance of a species $i$ from its read count,
 \begin{align}
   (\#eq:density-ratio-meas)
-  \widehat{\text{abun}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{abun}}_{r}(a)}{\text{reads}_{r}(a)}.
+  \tilde A_i^{(a)} &= M_i^{(a)} \cdot \frac{\tilde A_{r}^{(a)}}{M_i^{(a)}}.
 \end{align}
+Let $\FE[\tilde A_r^{(a)}] \equiv {\tilde A_r^{(a)}}/{A_r^{(a)}}$ be the FE in the reference measurement.
+The effect on $\tilde A_i^{(a)}$ of taxonomic bias in the MGS measurement can be determined by substituting Equation \@ref(eq:ratio-error) into Equation \@ref(eq:density-ratio-meas), yielding
+\begin{align}
+  (\#eq:density-ratio-error)
+  \tilde A_i^{(a)} = A_i^{(a)} \cdot \frac{B_i}{B_r} \cdot 
+%    \frac{\tilde A_r^{(a)}}{A_r^{(a)}}.
+    \FE \left[\tilde A_r^{(a)}\right].
+\end{align}
+The FE in $\tilde A_i^{(a)}$ consists of two terms: the relative efficiency of species $i$ to species $r$ in the MGS measurement (${B_i}/{B_r}$) and the FE in the reference species' abundance (${\tilde A_r^{(a)}}/{A_r^{(a)}}$).
 
 A common application of this approach involves adding a 'spike-in' (as described above) in a known (and typically constant) abundance across samples (@stammler2016adju, @ji2019quan, @harrison2021theq, REFs).
-In this case, the reference abundance $\widehat{\text{abun}}_{r}(a)$ is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
+In this case, the reference abundance $\tilde A_r^{(a)}$ is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
 (In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \@ref(eq:density-ratio-meas); Appendix REF).
@@
 A related approach involves computationally identifying species that are constant between pairs of samples (@david2014host) or between sample conditions (@mandal2015anal, @kumar2018anal).
-The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set $\widehat{\text{abun}}_{r}(a)$ to 1 in Equation \@ref(eq:density-ratio-meas).
+The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set $\tilde A_r^{(a)}$ to 1 in Equation \@ref(eq:density-ratio-meas).
 The resulting abundance measurements have unknown but fixed units, which is sufficient for measuring fold changes across samples.
@@
 Appendix [REF] describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
-
-The effect of taxonomic bias on absolute abundance measurements derived from a reference species can be determined by substituting Equation \@ref(eq:ratio-error) into Equation \@ref(eq:density-ratio-meas), which indicates that the measured abundance of species $i$ is
-\begin{align}
-  (\#eq:density-ratio-error)
-  \widehat{\text{abun}}_{i}(a) 
-  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
-  \cdot \underbrace{\frac{\widehat{\text{abun}_r}(a)}{\text{abun}_r(a)}}_{\substack{\text{fold error in} \\ \text{ref. species}}}.
-  % \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) 
-\end{align}
-The FE in $\widehat{\text{abun}}_{i}(a)$ consists of two terms: the relative efficiency of species $i$ to species $r$ in the MGS measurement, and the FE in the reference species' abundance.
-If bias is consistent across samples and the reference species can be spiked in or measured with a constant FE, then the FE in the $\widehat{\text{abun}}_{i}(a)$ will be constant.

### 2022-05-27 differential-abundance.Rmd
Revise measurement and diff-abun sections
- Change to new mathematical notation
- Improve clarity of presentation around equations, definitions, and
  results
- Add new treatment of FDs between pairs of samples for absolute
  abundances
- Add latex shortcuts to preamble for new notation

@@
 <!-- The FE in the measured FD between two samples is given by the ratio of the FEs for the individual-sample measurements. -->
-If the fold error (FE) in a species' abundance measurement is constant across samples, then it will not affect the measured FDs between pairs of samples.
+If the fold error (FE) in a species' abundance measurement is constant across samples, then it will not affect the measured FDs between samples.
 <!-- Ratios -->
@@
   (\#eq:ratio-fc-error)
-\underbrace{\frac{\widehat{\text{ratio}}_{i/j}(b)}{\widehat{\text{ratio}}_{i/j}(a)}} _\text{measured FD}
-  &= \frac
-    {\text{ratio}_{i/j}(b) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
-    {\text{ratio}_{i/j}(a) \cdot \cancel{\text{efficiency}_{i} / {\text{efficiency}_j}}}
-\\[0.5ex]
-  &=
-  \underbrace{\frac{\text{ratio}_{i/j}(b)}{\text{ratio}_{i/j}(a)}}_\text{actual FD}
+\underbrace{\frac{\tilde R_{i/j}^{(b)}}{\tilde R_{i/j}^{(a)}}}_\text{measured FD}
+  = \frac
+    {R_{i/j}^{(b)} \cdot \cancel{B_i / B_j}}
+    {R_{i/j}^{(a)} \cdot \cancel{B_i / B_j}}
+  =
+  \underbrace{\frac{R_{i/j}^{(b)}}{R_{i/j}^{(a)}}}_\text{actual FD}
   .
@@
   (\#eq:prop-fc-error)
-\underbrace{\frac{\widehat{\text{prop}}_{i}(b)}{\widehat{\text{prop}}_{i}(a)}} _\text{measured FD}
-  &= \frac
-    {\text{prop}_{i}(b) \cdot \cancel{\text{efficiency}_{i}} / {\text{efficiency}_S(b)}}
-    {\text{prop}_{i}(a) \cdot \cancel{\text{efficiency}_{i}} / {\text{efficiency}_S(a)}}
-\\[0.5ex]
-  &=
-  \underbrace{\frac{\text{prop}_{i}(b)}{\text{prop}_{i}(a)}}_\text{actual FD}
+\underbrace{\frac{\tilde P_i^{(b)}}{\tilde P_i^{(a)}}}_\text{measured FD}
+  = \frac
+    {P_i(b) \cdot \cancel{B_i} / \bar B^{(b)}}
+    {P_i(a) \cdot \cancel{B_i} / \bar B^{(a)}}
+  =
+  \underbrace{\frac{P_i^{(b)}}{P_i^{(a)}}}_\text{actual FD}
   \cdot
-  \underbrace{\left[\frac{\text{efficiency}_S(b)}{\text{efficiency}_S(a)}\right]^{-1}}_\text{fold error}
+  \underbrace{\left[\frac{\bar B^{(b)}}{\bar B^{(a)}}\right]^{-1}}_\text{FE}
   .
@@
 
-### AA - need to revise
+Taxonomic bias affects the measured FDs in a species' absolute abundance in a similar fashion.
+First, consider a species' abundance measurement $\tilde A_i^{(a)}$ derived from a non-MGS measurement of total-community abundance using Equation \@ref(eq:density-prop-meas).
+Equation \@ref(eq:density-prop-error) indicates that the FE in this measurement equals the (constant) species efficiency multiplied by ${B^{\mtot (a)}}/{\bar B^{(a)}}$, the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
+This ratio can vary, creating error in the measured FD between samples.
+Notably, if the FDs in the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, then the two offset each other and lead to more stable FEs (and hence more accurate FDs) in $\tilde A_i^{(a)}$.
+We discuss how this possibility might be exploited in real experimental workflows in Section \@ref(solutions).
 
-Taxonomic bias affects the measured FD in a species' absolute abundance in a similar fashion.
-Using Equation \@ref(\#eq:density-prop-error) or Equation \@ref(\#eq:density-ratio-error), we can compute the FE in the measured FD for absolute abundances derived from the two approaches discussed in Section \@ref(absolute-abundance).
-For the abundance of species $i$ derived from a total-abundance measurement, the measured FD is
-\begin{align}
-  (\#eq:aa-tot-fc-error)
-  FD\left[\widehat{\text{abun}}_{i}(a) \right]
-  = FD\left[\text{abun}_{i}(a)\right] \cdot FD \left[ \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}\right].
-\end{align}
-In this case, the FE in the FD equals the FD in the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
-For the abundance of species $i$ derived from a reference-species measurement, the measured FD is
-\begin{align}
-  (\#eq:aa-tot-fc-error)
-  FD\left[\widehat{\text{abun}}_{i}(a) \right]
-  = FD\left[\widehat{\text{abun}}_{r}(a)\right].
-\end{align}
-In this case, the FE in the FD equals the FD in the FE of the reference species' abundance measurement.
-If the FE in the $\widehat{\text{abun}}_{i}(a)$ is constant then the FD will be accurate.
-
-HERE - the above needs work, and I need to consider if equations are needed (and what type).
-Perhaps a summary table is better?
-
-#### Notes
-
-Note for revision
-- complementary measurement: mitigates variation in SME by offsetting with that of the total abundance measurement
-- reference species: mitigates variation in SME by using species ratios, avoiding additive aggregates which introduce variation in efficiency.
-
-for complementary AA, point out that the FE is equal to the ratio of FCs in the two MEs (or equiv in the FC in the ratio of MEs) and hence changes in one ME can be offset by changes in the other, if they both change in a similar way.
-
-for ref taxa, state how there is a constant FE and thus cancels.
+Now consider a species' abundance derived from a reference species using Equation \@ref(eq:density-ratio-meas).
+Equation \@ref(eq:density-ratio-error) indicates that the FE equals a constant ratio in species' efficiencies multiplied by the FE in the measurement of the reference species' abundance.
+If the abundance of the reference species can determined up to constant FE, then the FE in $\tilde A_i^{(a)}$ will also be constant, leading to accurate FDs.
 
@@
   (\#eq:regression)
-  y(a) = \alpha + \beta x(a) + \varepsilon(a),
+  y^{(a)} = \alpha + \beta x^{(a)} + \varepsilon^{(a)},
 \end{align}
-can be used to relate a microbial abundance variable $y$ to a _covariate_ $x$ and a _residual (or unexplained) error_ $\varepsilon$.
-For example, $y$ may be the log absolute abundance of species $i$ and $x$ may be a continuous variable such as pH; or, $x$ may be a binary variable, such as $x=1$ for treated patients and $x=0$ for untreated controls.
+can be used to relate a microbial abundance variable $y$ to a _covariate_ $x$ and the _residual (or unexplained) error_ $\varepsilon$.
+For example, $y$ may be the log absolute abundance of species $i$ ($\log A_i^{(a)}$) and $x$ may be a continuous variable (such as pH) or a binary variable (such as $x=1$ for treated patients and $x=0$ for untreated controls).
 The regression coefficients $\alpha$ and $\beta$ are parameters to be estimated from the data.
@@
 Further suppose that we have measured absolute abundances by multiplying the proportions from MGS with an accurate measurement of total-community abundance.
-For each species $i$, we set $y(a)$ in Equation \@ref(eq:regression) equal to $\log \widehat{\text{abun}}_{i}(a)$, and use either Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE) to estimate the coefficients $\alpha$ and $\beta$.
+For each species $i$, we set $y^{(a)}$ in Equation \@ref(eq:regression) equal to $\log \tilde A_{i}^{(a)}$, and use either Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE) to estimate the coefficients $\alpha$ and $\beta$.
 From Equation \@ref(eq:density-prop-error), we know that the FE in the abundance of each species equals its efficiency relative to the mean efficiency of the sample.
@@
 To understand how the FE in species abundances affects the estimated LFD for a given species, it is useful to consider three possible scenarios for how the mean efficiency varies across samples.
-In the first scenario, the mean efficiency remains stable across samples, such that the FE in $\widehat{\text{abun}}_{i}(a)$ is approximately constant. 
+In the first scenario, the mean efficiency remains stable across samples, such that the FE in $\log \tilde A_{i}^{(a)}$ is approximately constant. 
 In this case, the FE induced by bias _mostly cancels_ and does not substantially affect the estimated LFD.
@@
 Here, the average log mean efficiency is higher in the condition corresponding to $x=1$, due to an increase of the high-efficiency Species 9.
-This systematic increase in the log mean efficiency causes an arteficial decrease in the estimated LFD for each species by an equal amount.
+This systematic increase in the log mean efficiency causes an artificial decrease in the estimated LFD for each species by an equal amount.
 As we saw for the FD between individual samples, the biological significance of the systematic error in the LFD estimates varies by species, despite the absolute error being constant across species.

### 2022-05-27 index.Rmd
Revise measurement and diff-abun sections
- Change to new mathematical notation
- Improve clarity of presentation around equations, definitions, and
  results
- Add new treatment of FDs between pairs of samples for absolute
  abundances
- Add latex shortcuts to preamble for new notation

@@
     toc_appendix: yes
+header-includes:
+  - \newcommand{\tot}{{\text{tot}}}
+  - \newcommand{\mtot}{{\text{[tot]}}}
+  - \DeclareMathOperator{\FE}{FE}
 ---

### 2022-06-05 case-studies.Rmd
Revise case-studies and solutions sections
Minor revision primarily to update the terminology relating to fold
differences, with more extensive edits for clarity in a few places.

@@
 Motivated by the substantial ribosomal copy-number variation (CNV) in fungi (@lofgren2019geno), the authors also performed control measurements of mock communities that they constructed from quantified genomic DNA of the 9 species in the experiment; these controls were used to measure taxonomic bias with the method of @mclaren2019cons.
-The authors found a 13X difference between the most and least efficiently measured commensal, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
+The authors found a 13-fold difference between the most and least efficiently measured commensal, while the pathogen was measured 40-fold more efficiently than the least efficiently measured commensal.
 
 @leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gained by a species from being allowed to colonize first.
-Both analyses were based on fold changes in species proportions and so in principle were sensitive to taxonomic bias.
+Both analyses were based on fold differences (FDs) in species proportions and so in principle were sensitive to taxonomic bias.
 To improve accuracy, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.
@@
 To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \@ref(fig:leopold2020host-variation)).
-Despite the 13X variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62X and a geometric standard deviation of 1.05X.
+Despite the 13-fold variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62 and a geometric standard deviation of 1.05.
 This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
-But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3X and by just 1.5X between the two most dominant.
+But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3-fold and by just 1.5-fold between the two most dominant.
 The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
-Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual species, it had a negligible impact on the inferred fold changes and the DA analyses based on them.
+Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual species, it had a negligible impact on the inferred FDs and the DA analyses based on them.
 
@@
 The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
-Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5X to 5.2X.
-Using gamma-Poisson (or negative-binomial) regression of the observed read counts, we estimated the average LFC in commensal species proportions following infection with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
+Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5-fold to 5.2-fold.
+Using gamma-Poisson (or negative-binomial) regression of the observed read counts, we estimated the average LFD in commensal species proportions following infection with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
 Commensals typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
 However, failure to account for bias caused the decrease to be overestimated, by an amount corresponding to the inverse change in log mean efficiency.
-For commensal-host pairs with relatively small observed decreases, bias correction greatly reduced the magnitude of the negative LFCs or, in several cases, resulted in LFCs that were near 0 or slightly positive.
+For commensal-host pairs with relatively small observed decreases, bias correction greatly reduced the magnitude of the negative LFDs or, in several cases, resulted in LFDs that were near 0 or slightly positive.
 
@@
 The total genomic concentrations of each sample, combined with the species proportions measured by MGS, can be combined into a measurement of species concentrations using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
-In this case, the bias in the MGS measurements will create absolute errors in the estimated LFCs of genome concentration of equal magnitude to the errors in the LFC estimates for proportions.
+In this case, the bias in the MGS measurements will create absolute errors in the estimated LFDs of genome concentration of equal magnitude to the errors in the LFD estimates for proportions.
 The scientific error, however, might become worse.
-Suppose that total abundance increased by 2X due to pathogen growth; in this case, species whose proportions remained approximately constant would have increased in absolute abundance by around 2X.
-Bias, when not accounted for, shifts LFCs estimates downwards by 2.5X to 5.2X depending on host genotype.
+Suppose that total abundance increased by 2-fold due to pathogen growth; in this case, species whose proportions remained approximately constant would have increased in absolute abundance by around 2-fold.
+Bias shifts FD estimates downwards by 2.5-fold to 5.2-fold (depending on host genotype).
 Hence, without bias correction, we would instead conclude that these species decreased.
@@
 The taxonomic bias of the MOMS-PI MGS protocol was previously quantified by @mclaren2019cons using control measurements by @brooks2015thet of cellular mock communities of seven common, clinically-relevant vaginal bacterial species.
-Of these, _Lactobacillus iners_ had the highest efficiency, which was nearly 30X larger than that of the species with the lowest efficiency, _Gardnerella vaginalis_.
-A second _Lactobacillus_ species, _L. crispatus_, had an efficiency that was approximately 2X less than _L. iners_ and 15X greater than _G. vaginalis_.
-These species, along with the unculturable _Lachnospiraceae BVAB1_, are most frequently the most abundant species in a sample and commonly reach high proportions of over 70% (TODO SI analysis).
+Of these, _Lactobacillus iners_ had the highest efficiency, which was nearly 30-fold larger than that of the species with the lowest efficiency, _Gardnerella vaginalis_.
+A second _Lactobacillus_ species, _L. crispatus_, had an efficiency that was approximately 2-fold less than _L. iners_ and 15-fold greater than _G. vaginalis_.
+These species, along with the unculturable _Lachnospiraceae BVAB1_, are most frequently the most abundant species in a sample and commonly reach high proportions of over 70% (REF SI analysis).
 Therefore, shifts between these dominant species could drive large changes in the sample mean efficiency, which might in turn distort DA analyses.
@@
 We sought to understand the potential impact of bias on DA analysis of vaginal microbiome profiles from the MOMS-PI study.
-To do so, we first examined variation in the mean efficiency across samples under the assumption that bias in the MOMS-PI study was accurately represented by the @brooks2015thet control measurements.
-Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample (TODO SI Analysis).
-The mean efficiency varies substantially across vaginal samples (Figure 4.1), with samples in which a _Lactobacillus_ species is most abundant typically having a mean efficiency that is 3-20X greater than samples in which _G.
+We first examined variation in the mean efficiency across samples under the assumption that bias in the MOMS-PI study was accurately represented by the @brooks2015thet control measurements.
+Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample (REF SI Analysis).
+The mean efficiency varies substantially across vaginal samples (Figure 4.1), with samples in which a _Lactobacillus_ species is most abundant typically having a mean efficiency that is 3-fold to 20-fold greater than samples in which _G.
 vaginalis_ is most abundant.
@@
 
-We hypothesized that proportion-based DA analyses of health condition that is associated with the abundance of common taxa with particularly high or low efficiencies, such as _Lactobacillus_ and/or _Gardnerella_ in the MOMS-PI study, would be particularly prone to spurious results.
-A potential example may exist in the vaginal microbiome: Bacterial vaginosis (BV) has been repeatedly found to be strongly positively associated with higher-diversity communities with increased abundance of _G.
-vaginalis_ and decreased abundance of _Lactobacillus spp._ (@srinivasan2012bact, @cartwright2018mult).
-As a proxy for clinical BV, we split samples from the MOMS-PI study into low diversity (non-BV), intermediate diversity, and high diversity (BV-like) groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
-We estimated the LFC in proportion from non-BV to BV-like samples for 30 prevalent bacterial species, with and without bias correction, using gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions.
+These observations suggest that a proportion-based DA analysis of a health condition that is associated with _Lactobacillus_ and/or _Gardnerella_ would be prone to spurious results.
+To demonstrate this possibility, we performed a DA analysis of the MOMS-PI samples with respect to a simulated health condition.
+Bacterial vaginosis (BV) has been repeatedly found to be associated with a dramatically reduced proportion of _Lactobacillus spp._, an increased proportion of _G.
+vaginalis_, and higher within-community (alpha) diversity (@srinivasan2012bact, @cartwright2018mult).
+As a proxy for clinical BV, we split samples into high diversity (BV-like), intermediate diversity, and low diversity (non-BV-like) groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
+The mean efficiency was lower in the high-diversity BV-like samples (mean LFD of -1.6) due to a lower abundance of _Lactobacillus spp._.
+We estimated the LFD in proportion from non-BV-like to BV-like samples for 30 prevalent bacterial species, with and without bias correction, using gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions.
 We accounted for bias by including an offset term in the linear model equal to the log ratio of the species' efficiency to the sample mean efficiency.
-
-As expected, bias correction reduced the estimated LFC from non-BV to BV-like groups for every species tested, though the importance of this effect varied among species (TODO: median and range of change).
-Bias correction led to substantial magnitude changes for several species and sign changes in 3 species; however, the relative difference between the corrected and uncorrected LFCs was modest for the 19/30 species with the largest changes, which each had a corrected LFC greater than 2 (corresponding to a FC greater than 7.4X).
-Our theoretical results from the previous section suggest that bias correction should reduce the estimated LFC for each species by roughly the amount that the mean efficiency decreases between groups.
-Consistent with this prediction, the mean efficiency was lower in high-diversity BV-like samples (mean LFC of -1.6) due to the lower abundance of the _Lactobacillus spp._
-that are so efficiently measured by the MOMS-PI MGS protocol.
+As expected, bias correction reduced the estimated LFD from non-BV-like to BV-like groups for every species tested, though the importance of this effect varied among species (SI Figure \@ref(fig:momspi-regression)).
 
@@
 Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \@ref(fig:gut)C), which the GSD in the proportion of a gut species across gut samples an average of `r hmp_species_diff %>% round(2)`-fold lower than that of a vaginal species across vaginal samples.
-Recall that the importance of bias estimation of the FC in a species' proportion depends on GSD in the mean efficiency relative the GSD in the species' proportion (Section \@ref(differential-abundance)).
+Recall that the importance of bias estimation of the FD in a species' proportion depends on GSD in the mean efficiency relative the GSD in the species' proportion (Section \@ref(differential-abundance)).
 Thus these results suggest that although the mean efficiency likely varies less across stool than vaginal samples, bias may be just as problematic for proportion-based DA analyses.
@@
 This shifting composition will cause the sample mean efficiency to decrease with time, leading to inflated growth-rate estimates for all taxa.
-In this scenario, taxa that decay with depth sufficiently slowly would mistakenly be inferred to have positive growth from the FRAxC-calculated abundances.
+Taxa that decay with depth sufficiently slowly would mistakenly be inferred to have positive growth from the FRAxC-calculated abundances.
 
-Importantly, @lloyd2020evid also used qPCR to measure the absolute concentration of two taxa from these same samples.
-These qPCR measurements allow us to test the hypothesis that systematic variation in log mean efficiency with depth distorts inferred growth rates.
-In Section 8, we argue that taxonomic bias in targeted qPCR measurements is expected to create a roughly constant fold error and yield fold changes in absolute density that are relatively unaffected by changes in mean efficiency.
-Thus comparing qPCR to FRAxC growth rates on the same reference taxa allows us to estimate the systematic error in FRAxC growth rates.
-Because the systematic error in regression slopes due to bias is the same for each taxon (Section 3), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
+Importantly, @lloyd2020evid also used qPCR to measure the absolute abundance of two taxa from these same samples.
+By comparing qPCR to FRAxC growth rates for these taxa, we can estimate the systematic error in FRAxC growth rates.
+Because the systematic error that bias causes in the regression slope is the same for each taxon (Section \@ref(differential-abundance)), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
 
@@
 
-We saw several cases where the mean efficiency is stable across samples, so that bias does not affect LFC estimates (Scenario 1).
+We saw several cases where the mean efficiency is stable across samples, so that bias does not affect LFD estimates (Scenario 1).
 In the pre-infection fungal microbiome samples of @leopold2020host, we observed that the mean efficiency remained stable despite substantial bias and large multiplicative variation in species proportions among samples.
 Because the variation in the mean efficiency was much less than that of the individual species, the impact of bias on DA analysis was negligible.
-In the vaginal case study, we also observed that the mean efficiency was relative stable across vaginal microbiomes that were dominated by the same species, despite substantial bias and large LFCs in non-dominant species.
+In the vaginal case study, we also observed that the mean efficiency was relative stable across vaginal microbiomes that were dominated by the same species, despite substantial bias and large LFDs in non-dominant species.
 In both cases, the stability of the mean efficiency can be understood by the fact that it is an additive average over species and so is primarily determined by the most abundant species in the sample.
@@
 Yet we also observed cases where the mean efficiency varied substantially.
-In several cases, the mean efficiency varied systematically with the condition of interest sufficient to cause substantial systematic error in LFC estimates (Scenario 3).
+In several cases, the mean efficiency varied systematically with the condition of interest sufficient to cause substantial systematic error in LFD estimates (Scenario 3).
 Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
 Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time that is sufficient to substantially inflate the estimated growth rates of slowly changing species.
-The mean efficiency cannot change more than the species proportions constituting it.
-Therefore, the error in the estimated proportion LFCs tend to be practically significant only for species whose LFCs are substantially smaller in magnitude than the largest LFC.
+The mean efficiency cannot differ by more than the species proportions constituting it.
+Therefore, the error in the estimated LFDs tend to be practically significant only for species whose LFDs are substantially smaller in magnitude than the largest LFD.
 
@@
 This variation is much less problematic for DA results when it not associated with the condition under study, since any loss in precision can in principle be offset by an increase in sample size.
-For certain applications, however, it will be important to remember that the LFCs between individual samples remain unreliable.
+For certain applications, however, it will be important to remember that the LFDs between individual samples remain unreliable.
 

### 2022-06-05 solutions.Rmd
Revise case-studies and solutions sections
Minor revision primarily to update the terminology relating to fold
differences, with more extensive edits for clarity in a few places.

@@
 
-Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using relative DA methods that analyze FCs in these ratios.
-A variety of ratio-based methods have been published under the name of Compositional Data Analysis (CoDA).
+Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using DA methods that analyze fold differences (FDs) in these ratios.
+A variety of ratio-based methods have been developed under the framework of Compositional Data Analysis (CoDA; REFs).
 CoDA methods may more generally consider the ratios between products of multiple species, possibly raised to some exponent; a simple example is the geometric mean of several species.
 The mathematical operations of multiplication and exponentiation maintain constant FEs.
-Hence such products provide a method for aggregating species into a higher-level taxonomic units that maintains bias invariance when estimating FCs.
-Ratio-based analyses also avoid the potential for misinterpreting changes in a species’ proportion that are driven by the sum-to-one constraint (@gloor2017micr).
+Hence such products provide a method for aggregating species into a higher-level taxonomic units that maintains bias invariance when estimating FDs.
+Ratio-based analyses also avoid the potential for misinterpreting differences in a species’ proportion that are driven by the sum-to-one constraint (@gloor2017micr).
 
@@
 First, in most MGS datasets, any given species is likely to be extremely rare in most samples, such that no sequencing reads are observed.
-In order to analyze FCs, DA methods interpret these zero counts as evidence that the true abundance of the species is very small (but still positive).
-A common approach to enable the computation of FCs in the presence of zeros is to add a small value, or ‘pseudo-count’, to the read counts, making all values positive.
+In order to analyze FDs, DA methods interpret these zero counts as evidence that the true abundance of the species is very small (but still positive).
+A common approach to enable the computation of FDs in the presence of zeros is to add a small value, or ‘pseudo-count’, to the read counts, making all values positive.
 Unfortunately, this procedure violates bias invariance.
@@
 Calibration makes it possible to correct for bias in relative and absolute DA methods that would otherwise not be bias invariant.
-To demonstrate, we used calibration to improve the estimated FCs between samples in the absolute abundance of a species in the bacterial mixtures from @brooks2015thet.
+To demonstrate, we used calibration to improve the estimated FDs between samples in the absolute abundance of a species in the bacterial mixtures from @brooks2015thet.
 We used the bias estimated from a single 7-species mixture to calibrate the MGS-derived proportions in all measured mixtures.
 We obtained calibrated absolute abundances by multiplying the calibrated proportions by the known total abundance (Equation \@ref(eq:density-prop-meas)).
-Calibration greatly increased the accuracy of the estimated FCs (Figure \@ref(fig:calibration-example), A vs. B).
+Calibration greatly increased the accuracy of the estimated FDs (Figure \@ref(fig:calibration-example), A vs. B).
 The fungal and vaginal case studies of Section \@ref(case-studies) provide several practical examples of calibration in proportion-based DA analyses.
@@
 
-(ref:cap-calibration-example) **Fold changes can be calibrated using community controls or reference species.** The figure compares the performance of three methods for measuring fold changes in absolute cell density in cellular mock communities of 7 vaginal species, which were constructed and measured via 16S sequencing by @brooks2015thet.
-The 'Uncalibrated' fold changes are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which is constant by construction).
+(ref:cap-calibration-example) **Fold differences can be calibrated using community controls or reference species.** The figure compares the performance of three methods for measuring fold differences (FDs) in absolute cell density in cellular mock communities of 7 vaginal species, which were constructed and measured via 16S sequencing by @brooks2015thet.
+The 'Uncalibrated' FDs are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which is constant by construction).
 The 'Calibrated (community)' measurements are computed from abundance measurements where the proportions are first corrected for the taxonomic bias that was estimated from a single sample that contained all 7 species.
@@
 Thus the 16S qPCR measurement is subject to bias from extraction, copy-number variation, primer binding, and amplification, just like the 16S sequencing measurement.
-Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it can lead to more accurate fold changes across samples.
+Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it can lead to more accurate FDs across samples.
 
@@
 Researchers may naturally be concerned about bias between the reference species and the native species being normalized against it (see, for example, @harrison2021theq's recommendations for spike-in experiments).
-Our results show that accurate FCs and DA results can be obtained so long as the relative efficiency of the focal species to the reference is consistent across samples.
+Our results show that accurate FDs and DA results can be obtained so long as the relative efficiency of the focal species to the reference is consistent across samples.
 This condition is much weaker than requiring that bias is small (the relative efficiency is close to 1) and so greatly expands the applicability of spike-ins and host normalization in microbiome experiments.
 
-Section \@ref(absolute-abundance) further proposed an additional class of reference species for normalizing MGS measurements: Species whose abundance has been measured using a targeted method such as qPCR with species-specific primers.
-To demonstrate the ability for reference calibration to improve fold changes, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
-Doing so improved the resulting measurements of fold changes in density for all species (Figure \@ref(fig:calibration-example)).
-Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus further expanding the applicability of this reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
+Section \@ref(absolute-abundance) further proposed an additional class of reference species for normalizing MGS measurements: Species whose (absolute) abundance has been measured using a targeted method such as qPCR with species-specific primers.
+To demonstrate the ability for reference calibration to reduce error in FD measurements, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
+Doing so improved the resulting measurements of FDs in the abundance of all species (Figure \@ref(fig:calibration-example)).
+Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus expanding the applicability of reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
 The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
@@
 Parametric meta-analysis models include study-specific latent parameters representing 'batch effects'—non-biological differences in the data from each study which can distort the observed biological patterns.
-By estimating these 'nuisance parameters' along with the biological parameters of interest (such as species log-fold changes), the meta-analysis aims to reduce statistical bias in the biological parameters created by the non-biological differences among studies (@leek2010tack,@wang2019mana).
+By estimating these 'nuisance parameters' along with the biological parameters of interest (such as the difference in a species' log abundance between conditions), the meta-analysis aims to reduce statistical bias in the biological parameters created by the non-biological differences among studies (@leek2010tack,@wang2019mana).
 In a bias-aware meta-analysis, the meta-analysis model is configured so that (some of) the latent parameters correspond to study- and species-specific relative efficiencies.

### 2022-06-05 appendix-diversity.Rmd
Clean up appendix
- Remove old appendix and notes
- Put diversity derivation in its own section
- Create stubs for some missing sections

--- 2022-03-14 appendix-notes.Rmd
@@
-# Theory notes
-
-## Effective experimental effort: Intuition
-
-The effective experimental effort per cell for sample $a$ is
-\begin{align}
-  (\#eq:effort)
-  \text{effort}_S(a)
-    &= \frac{\text{reads}_S(a)}{\text{abun}_S(a) \cdot \text{efficiency}_S(a)} \\
-    &= \frac{\sum_{i \in S}\text{reads}_i(a)}{\sum_{i \in S} \text{abun}_i(a) \cdot \text{efficiency}_i(a)};
-\end{align}
-the second form follows from the definition of the mean efficiency.
-The effort increases if more sequencing performed on the sample, leading to greater number of reads for all species; and decreases if a larger sample input volume or concentration is used, leading to fewer reads per cell.
-Less intuitively, it also decreases with the mean efficiency of the sample.
-To help interpret the effort light of this last behavior, we consider how the effort varies across a few illustrative scenarios.
-
-**Scenario 1:**
-Consider Sample 1 in Figure \@ref(fig:error-proportions).
-Imagine that sequencing reads can be perfectly assigned to the three species; the taxonomic bias is entirely due to steps prior to read assignment, such as DNA extraction and PCR.
-Supposing that the sample was sequenced to a read depth of 100, the three species have read counts of $(4, 72, 24)$.
-In addition, suppose the total cell concentration in the sample equals 1 in some particular units.
-Since the mean efficiency is $(1 + 18 + 6) / 3 = 8.\bar 3$, the effort is $100 / (1 \cdot 8.\bar 3) = 12$.
-
-**Scenario 2:**
-Now consider a situation identical to Scenario 1, except that the bioinformatics pipeline is unable to assign reads from Species 3.
-In this case, the _assigned_ read counts will be $(4, 72, 0)$, for a total of $76$, though the total number of sequenced reads remains 100.
-Meanwhile, the efficiency of species 3 drops to 0 and the mean efficiency of the sample drops to $(1 + 18 + 0) / 3 = 6.\bar 3$.
-The effort is therefore $76 / (1 \cdot 6.\bar 3) = 12$, the same as in Scenario 1.
-<!-- The decline in total reads is exactly offset by the decline in the mean efficiency. -->
-<!-- This example illustrates a general pattern: The effort is invariant to removing from or adding to species' assigned reads. TODO: Make precise, and prove -->
-
-**Scenario 3:**
-Now consider a situation identical to Scenario 1, except that Species 3 fails to amplify during the PCR step in a marker-gene sequencing experiment.
-But thanks to careful experimental normalization during library preparation, an equal amount of DNA is prepared in the library as in Scenario 1, leading to 100 sequenced reads from the sample.
-The measured relative abundances are identical to in Scenario 2, but the read counts are larger, $(\approx 5.3, \approx 94.7)$.
-(As in Section \@ref(abundance-measurement), we ignore the fact that read counts are discrete.)
-The mean efficiency is also the same ($6.\bar 3$) as in Scenario 2.
-The effort is therefore $100 / (1 \cdot 6.\bar 3) \approx 15.8$, larger than in Scenario 1 and Scenario 2.
-
-**Scenario 4:**
-Now consider a situation identical to Scenario 1, but supposing Species 3 to be a reference species, such as the host or a spike-in.
-Does the effort depend on whether we include the reference species in the species set $S$?
-As we saw in Scenario 1, the effort associated with all three species is 12.
-For the two non-reference species alone, the total abundance is $2/3$, the total read count is $76$, and the mean efficiency is $(1 + 18) / 2 = 9.5$.
-The effort is therefore $76 /(\tfrac{2}{3} \cdot 9.5) = 12$, the same as in Scenario 1 and Scenario 2.
-<!-- Compared to Scenario 2, the change in abundance and mean efficiency offset -->
-
-TODO:
-
-- Make sure this result is general, and not just happening when the three species are all equally abundant.
-- consider moving secenario 4 to 2 or 3, leaving the scenario where effort changes to last.
-- note that for proving the general results, it may be best to go back to main model equation, where we can see that if the read count, abundance, and efficiency for at least one species is unchanged, than the effort is unchanged.
-  - in scenario 3, the effort must go up since taxa 1 and 2 are unchanged but get more reads.
-  - if we add or remove species, without changing the read count of other species, then the effort must stay the same.
-  - this fact that effort is independent of the taxonomic scope S seems useful and convenient and intuitive.
-
-**Summary:**
-The behavior of the effective experimental effort across these scenarios makes intuitive sense if we think of it as specifically describing the 'wet lab' components of the MGS measurement, up to and including sequencing but excluding bioinformatic analysis of the sequencing data.
-The taxonomic scope $S$ is an arbitrary decision made implicitly or explicitly by the researcher, which does not change the wet portion of the experiment; thus   $\text{effort}(a)$ is independent of $S$.
-In addition, sequencing reads are routinely excluded during bioinformatic analysis, and such filtering does not change $\text{effort}(a)$, instead reducing the read counts by reducing the species' efficiencies.
-On the other hand, if species drop out prior to sequencing due to low extraction efficiency and/or low PCR efficiency, then achieving the same total read count requires that a greater sample volume or concentration be passed on to subsequent steps.
-These adjustments are reflected by a commensurate increase in $\text{effort}(a)$.
-
-## Absolute-abundance measurement
-
-Q: How to motivate this 'general' form? 
-It comes from ignoring bias.
-
-Make clear that R doesn't need to be in S. Perhaps want different sets - a superset of all species, and an set of species that each measurement method considers?
-
----
-
-Consider this general form for measuring of the absolute abundance of a taxon $I$ from a reference species $R$,
-\begin{align}
-  \widehat{\text{abun}}_{I}(a) 
-  &= \text{reads}_{I}(a) \cdot \frac{\widehat{\text{abun}}_{R}(a)}{\text{reads}_{R}(a)}.
-\end{align}
-When $R = S$, this measurement amounts to total-community normalization; when $R=r$, it amounts to reference-species normalization.
-(This form does not encompass using the geometric average of multiple species.)
-Under our model, we can rewrite the right-hand-side to obtain
-\begin{align}
-  \widehat{\text{abun}}_{I}(a) 
-  &= \text{abun}_{I}(a) \cdot \frac{\text{efficiency}_{I}(a)}{\text{efficiency}_{R}(a)}
-  \cdot \text{FE}\left[\widehat{\text{abun}_R}(a)\right],
-\end{align}
-where $\text{FE}\left[\widehat{\text{abun}_R}(a)\right] = \widehat{\text{abun}_R}(a) / \text{abun}_R(a)$ is the fold error in the reference abundance measurement.
-
-From here, we can derive the other results.
-
-Methods differ based on the choice of $R$ and in how the abundance of $R$ is measured.
-
-## Alpha diversity and variation in the mean efficiency {#diversity-and-mean-efficiency}
+# Alpha diversity and variation in the mean efficiency {#diversity-and-mean-efficiency}
 

### 2022-06-05 v/0d63746fac87923e630a61276e8fb3f1cfddd74c/_main.tex
Clean up appendix
This commit adds the rendered manuscript associated with
0d63746fac87923e630a61276e8fb3f1cfddd74c

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The effect of taxonomic bias on differential-abundance analysis in microbiome studies},
  pdfauthor={Michael R. McLaren; Jacob T. Nearing; Amy D. Willis; Karen G. Lloyd; Benjamin J. Callahan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{cancel}

% Define note environment based on example from https://bookdown.org/yihui/rmarkdown-cookbook/custom-blocks.html
\usepackage{color}
\usepackage{framed}
\setlength{\fboxsep}{.8em}
\newenvironment{rmdnote}{
  \definecolor{shadecolor}{RGB}{248,248,248}
  \begin{shaded}}
 {\end{shaded}}

% Macros
\newcommand{\md}{^{\times}\!/}
\newcommand{\tot}{{\text{tot}}}
\newcommand{\mtot}{{\text{[tot]}}}
\DeclareMathOperator{\FE}{FE}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{The effect of taxonomic bias on differential-abundance analysis in microbiome studies}
\author{Michael R. McLaren\footnote{North Carolina State University; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Jacob T. Nearing\footnote{Dalhousie University} \and Amy D. Willis\footnote{University of Washington} \and Karen G. Lloyd\footnote{University of Tennessee} \and Benjamin J. Callahan\footnote{North Carolina State University}}
\date{2022-06-05}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{preface}{%
\section*{Preface}\label{preface}}
\addcontentsline{toc}{section}{Preface}

\begin{verbatim}
#> working directory clean
\end{verbatim}

\emph{This manuscript was rendered from commit 0d63746fac87923e630a61276e8fb3f1cfddd74c.}

\leavevmode\vadjust pre{\hypertarget{preface-warning}{}}%
\textbf{This in-progress manuscript is not intended for general scientific use.}
It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
Please post comments or questions on the \href{https://github.com/mikemc/differential-abundance-theory/issues}{GitHub Issues page} or \href{m.mclaren42@gmail.com}{email Mike}.

This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.
We describe the basic problem posed by taxonomic bias for measuring changes in the abundance of particular taxa across conditions and describe new strategies for mitigating the errors it induces.
Analyses of both relative and absolute abundances are considered.
The manuscript is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
See \href{https://doi.org/10.5281/zenodo.4552717}{the Zenodo record} for how to cite the latest version.
Supporting data analyses can be found in the accompanying {[}computational research notebook{]} (\url{https://mikemc.github.io/differential-abundance-theory/notebook/}).

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\leavevmode\vadjust pre{\hypertarget{abstract}{}}%
\textbf{Abstract:}
To assess how the species in microbial communities vary across conditions, researchers often measure the relative or absolute abundances of all species simultaneously using marker-gene and metagenomic sequencing.
However, these measurements are taxonomically biased: Some species are measured more efficiently than others, yielding measured abundances that are higher or lower than the truth.
Taxonomic bias is generally ignored in \emph{differential abundance (DA) analyses}, likely due to a combination of 1) a widespread belief that protocol standardization is sufficient to yield accurate differences between samples and 2) a lack of practical solutions for addressing it.
We use theoretical arguments and case studies to analyze the impact of taxonomic bias on DA analyses of relative and absolute abundances.
We show that taxonomic bias does create inferential errors for a class of popular DA methods; however, whether these errors are biologically significant depends on experimental context.
We present several approaches to mitigating the effect of bias, suited to a range of experimental questions and systems.
If adopted, they could improve the reproducibility and interpretability of sequencing-based studies of microbial communities.

One of the most basic questions we can ask about microbial communities is: How do different microbial taxa vary in abundance---across space, time, and host or environmental conditions?
Marker-gene and shotgun metagenomic sequencing (jointly, MGS) can be used to measure the abundances of thousands of species simultaneously, making it possible to ask this question on a community-wide scale.
In these \emph{differential-abundance (DA) analyses}, the change in abundance of a microbial taxon across samples or conditions is used to infer ecological dynamics or find microbes that are associated with specific host diseases or environmental conditions.
Standard MGS measurements lose information about total microbial density and so are typically used to analyze the abundances of taxa relative to each other (\emph{relative abundances}).
But new methods are increasingly used to provide absolute information, making it possible to analyze changes in absolute cell density, biomass, or genome copy number (\emph{absolute abundances}).
In its various forms, DA is among the most common analyses applied to MGS data.

Unfortunately, these DA analyses are built on a fundamentally flawed foundation.

MGS measurements are \emph{taxonomically biased}: Microbial species vary dramatically in how efficiently they are measured---that is, converted from cells into taxonomically classified sequencing reads---by a given MGS protocol (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
This bias arises from variation in how species respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
Although often associated with features specific to marker-gene sequencing---the variation among species in marker copy numbers and in primer-binding and amplification efficiencies---the existence of large variation in DNA extraction efficiencies and in the ability to correctly classify reads make taxonomic bias a universal feature of both marker-gene and shotgun measurements.
As a result, MGS measurements provide inaccurate representations of actual community composition and tend to differ systematically across protocols, studies, and even experimental batches within a study (Yeh et al. (\protect\hyperlink{ref-yeh2018taxo}{2018}), McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
These errors can supersede sizable biological differences (e.g. Lozupone et al. (\protect\hyperlink{ref-lozupone2013meta}{2013})) and may have contributed to failed replications of prominent DA results such as the associations of Bacteroides and Firmicutes in stool with obesity (Finucane et al. (\protect\hyperlink{ref-finucane2014atax}{2014})) and the associations of species in the vaginas of pregnant women with preterm birth (Callahan et al. (\protect\hyperlink{ref-callahan2017repl}{2017})).

The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
Statistical analyses are then conducted with the tacit assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
This argument is at least intuitively plausible for DA analyses based on multiplicative or fold differences (FDs) in a taxon's abundance.
If bias caused a species' abundance to be consistently measured as 10-fold greater than its actual value, then we would still recover the correct FDs among samples.
However, McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) showed mathematically and with MGS measurements of artificially constructed (`mock') communities that consistent taxonomic bias can create fold errors (FEs) that vary across samples and, as a result, majorly distort cross-sample comparisons.
In particular, they showed that the FE in a species' proportion---the most common measure of relative abundance---varies among samples, distorting the observed FDs between samples.
In some cases, bias can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
Yet McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that other abundance measures---those based on the ratios among species---have proportional errors and may lead to more robust DA analyses.
The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
We show that, in contrast to received wisdom, taxonomic bias can affect the results obtained from many standard DA methods; however, other methods are robust to bias, provided that it is truly consistent across samples.
We then use several case studies to explore the real-world conditions in which bias is likely to cause serious scientific errors.
Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.
Application of these insights and methods can build the confidence needed to turn the findings of microbiome studies into readily-translatable scientific knowledge.

\hypertarget{abundance-measurement}{%
\section{How bias affects abundance measurements}\label{abundance-measurement}}

This section extends the theoretical results of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to describe the effect that consistent taxonomic bias within an MGS experiment affects the relative and absolute abundances measured for various microbial species.
We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.

\hypertarget{model-of-mgs-measurement}{%
\subsection{Model of MGS measurement}\label{model-of-mgs-measurement}}

Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
This model describes the mathematical relationship between the read counts obtained by MGS and the (actual) abundances of the various species in a sample.
Here we extend the model as first described McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}), which considers only relative abundances, to also consider absolute abundances.
For concreteness, we will consider \emph{absolute abundance}, or simply \emph{abundance}, to refer to the number of cells per unit volume in a sample (cell concentration).
That said, our results equally apply to other definitions of absolute abundance, such as the total number of cells in a sample or ecosystem and other abundance units such as biomass or genome copy number.

This model is the simplest that respects the multiplicative nature of taxonomic bias and the \emph{compositional} nature of MGS measurements.
The actual abundance of a species in a given sample, multiplied by its \emph{measurement efficiency}---its rate of conversion from cells to taxonomically assigned sequencing reads---determines the species' read count in that sample.
Taxonomic bias presents as variation in the measurement efficiencies among species within an MGS experiment.
The read counts further depend on sample-specific experimental factors that are typically unknown, such that they are best interpreted as only providing relative abundances (such data is said to be \emph{compositional}; Gloor et al. (\protect\hyperlink{ref-gloor2017micr}{2017})).

We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and taxonomically assigns reads to a set of microbial species \(S\).
We make several simplifying assumptions to facilitate our analysis and presentation.
First, we consider only species-level assignment, and suppose that reads that cannot be uniquely assigned to a single species in \(S\) are discarded.
Second, we ignore the possibility that reads are misassigned to the wrong species or the wrong sample.
Third, we suppose that taxonomic bias acts consistently across samples at the level of species---that is, a given species is always measured more efficiently than another to the same degree, regardless of variation in changes in sub-species-level composition or the sample matrix.
Finally, unless otherwise stated, we treat sequencing measurements as deterministic, ignoring the `random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
These assumptions, though unrealistic descriptions of most MGS experiments, serve the purpose of clearly demonstrating when and why consistent taxonomic bias creates errors in DA analysis.

Our model stipulates that the taxonomically-assigned read count for a species \(i\) in a sample \(a\) equals its abundance \(A_{i}^{(a)}\) multiplied by a species-specific factor \(B_{i}\) and a sample-specific factor \(F^{(a)}\),
\begin{align}
  \label{eq:mgs-model}
  M_i^{(a)} = A_i^{(a)} B_i F^{(a)}.
\end{align}
The species-specific factor, \(B_{i}\), is the \emph{relative measurement efficiency} (or simply \emph{efficiency}) of the species relative to an arbitrary baseline species (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
The sample-specific factor, \(F^{(a)}\), describes the effective sequencing effort for that sample; it equals the number of reads per unit abundance that would be obtained for a species with an efficiency of 1.
We can write the total number of assigned reads for the sample as
\begin{align}
  \label{eq:total-reads}
  M_{\text{tot}}^{(a)}
    = A_{\text{tot}}^{(a)} \bar B^{(a)} F^{(a)},
\end{align}
where
\(M_{{\text{tot}}}^{(a)} \equiv \sum_{j\in S} M_j^{(a)}\) is the total read count and \(A_{{\text{tot}}}^{(a)} \equiv \sum_{j\in S}A_j^{(a)}\) is the total abundance for all species \(S\), and
\begin{align}
  \label{eq:mean-efficiency}
  \bar B^{(a)} \equiv \frac{\sum_{i\in S} A_i^{(a)} B_i}{\sum_{i\in S} A_i^{(a)}}
\end{align}
is the \emph{sample mean efficiency}, defined as the mean efficiency of all species weighted by their abundance.

\hypertarget{relative-abundance}{%
\subsection{Relative abundance}\label{relative-abundance}}

We distinguish between two types of species-level \emph{relative abundances} within a sample.
The \emph{proportion} \(P_{i}^{(a)}\) of species \(i\) in sample \(a\) equals its abundance relative to the total abundance of all species in \(S\),
\begin{align}
  \label{eq:prop}
  P_{i}^{(a)} &\equiv \frac{A_i^{(a)}}{A_{\text{tot}}^{(a)}}.
\end{align}
The \emph{ratio} \(R_{i/j}^{(a)}\) between two species \(i\) and \(j\) equals the abundance of \(i\) divided by that of \(j\),
\begin{align}
  \label{eq:ratio}
  R_{i/j}^{(a)} \equiv \frac{A_i^{(a)}}{A_j^{(a)}}.
\end{align}

The measured proportion of a species is given by its proportion of all the assigned reads in a sample,
\begin{align}
  \label{eq:prop-meas}
  \tilde P_{i}^{(a)} &\equiv \frac{M_i^{(a)}}{M_{\text{tot}}^{(a)}}.
\end{align}
We use the tilde to distinguish the measurement from the actual quantity being measured.
From Equations \eqref{eq:mgs-model}, \eqref{eq:total-reads}, and \eqref{eq:prop-meas}, it follows that the measured and actual proportion are related by
\begin{align}
  \label{eq:prop-error}
  \tilde P_{i}^{(a)} &= P_{i}^{(a)} \cdot \frac{B_i}{\bar B^{(a)}}.
\end{align}
Taxonomic bias thus creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
Since the mean efficiency varies across samples, so does the FE.
This phenomenon can be seen for Species 3 in the two hypothetical communities in Figure \ref{fig:error-proportions}.
Species 3, which has an efficiency of 6, is under-measured in Sample 1 (FE \textless{} 1) but over-measured (FE \textgreater{} 1) in Sample 2.
This difference occurs because the even distribution of species Sample 1 yields a mean efficiency of 8.33; in contrast, the lopsided distribution in Sample 2, which is dominated by the low-efficiency Species 1, has a mean efficiency of just 3.15.
A demonstration in bacterial mock communities is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3C} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).

The measured ratio \(\tilde R_{i/j}^{(a)}\) between species \(i\) and \(j\) is given by the ratio of their read counts,
\begin{align}
  \label{eq:ratio-meas}
  \tilde R_{i/j}^{(a)} \equiv \frac{M_{i}^{(a)}}{M_{j}^{(a)}}.
\end{align}
From Equations \eqref{eq:mgs-model} and \eqref{eq:ratio-meas}, it follows that the measured and actual ratio are related by
\begin{align}
  \label{eq:ratio-error}
  \tilde R_{i/j}^{(a)} = R_{i/j}^{(a)} \cdot \frac{B_i}{B_j}.
\end{align}
Taxonomic bias thus creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
For instance, in Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
A demonstration in bacterial mock communities is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3D} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/export-pdf/figures/illustrations/error-proportions} \caption{\textbf{Taxonomic bias creates fold errors in species proportions that vary across samples and lead to inaccurate fold differences between samples.} Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species. Bottom row: Error in the measured fold difference in the third species that is derived from these measurements. Species' proportions may be measured as too high or too low depending on sample composition. For instance, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).}\label{fig:error-proportions}
\end{figure}

\textbf{Higher-order taxa:}
We can consider a higher-order taxon \(I\), such as a genus or phylum, as a set of species, \(\{i \in I\}\).
The abundance of taxon \(I\) in sample \(a\) is the sum of the abundances of its constituent species, \(A_{I}^{(a)} \equiv \sum_{i \in I} A_{i}^{(a)}\).
Similarly, the read count of taxon \(I\) is the sum \(M_{I}^{(a)} \equiv \sum_{i \in I} M_{i}^{(a)}\).
We further define the efficiency of taxon \(I\) as the abundance-weighted average of the efficiencies of its constituent species,
\begin{align}
  \label{eq:efficiency-general}
  B_I^{(a)} \equiv \frac{\sum_{i\in I} A_{i}^{(a)} B_{i}}{\sum_{i\in I} A_{i}^{(a)}}.
\end{align}
With these definitions, the read count for taxon \(I\) can be expressed as
\(M_{I}^{(a)} = A_{I}^{(a)} B_I^{(a)} F^{(a)}\).
Thus \(B_I^{(a)}\) plays a role analogous to the efficiency of an individual species, but differs in that it need not be constant across samples:
If the constituent species have different efficiencies, then the efficiency of the higher-order taxon \(I\) depends on the relative abundances of its constituents and so will tend to vary across samples (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
As an example, suppose that Species 1 and Species 2 in Figure \ref{fig:error-proportions} were in the same phylum.
The efficiency of the phylum would then be \(\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5\) in Sample 1 and \(\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1\) in Sample 2.
Equations \eqref{eq:prop-error} and \eqref{eq:ratio-error} continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent, higher-order taxa efficiencies \(B_I^{(a)}\) and \(B_J^{(a)}\) are used.
In this way, we see that both proportions and ratios among higher-order taxa may have inconsistent FEs.

\hypertarget{absolute-abundance}{%
\subsection{Absolute abundance}\label{absolute-abundance}}

Several extensions of the standard MGS experiment make it possible to measure absolute species abundances.
These extensions fall into two general approaches.
The first approach leverages information about the abundance of the total community; for example, Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017}) measured total-community abundance using flow cytometry and multiplied this number by MGS genus proportions to obtain the absolute abundances of individual genera (Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017})).
A second approach leverages information about the abundance of one or more individual species; for example, a researcher might `spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (Harrison et al. (\protect\hyperlink{ref-harrison2021theq}{2021})).
We consider each approach in detail to determine how taxonomic bias affects the resulting absolute-abundance measurements.

\hypertarget{leveraging-information-about-total-community-abundance}{%
\subsubsection{Leveraging information about total-community abundance}\label{leveraging-information-about-total-community-abundance}}

Suppose that the total abundance of all species in the sample, \(A_{{\text{tot}}}^{(a)}\), has been measured by a non-MGS method, yielding a measurement \(\tilde A_{\text{tot}}^{(a)}\).
The absolute abundance of an individual species can be measured by multiplying the species' proportion from MGS by this total-abundance measurement,
\begin{align}
  \label{eq:density-prop-meas}
  \tilde A_i^{(a)} &= \tilde P_i^{(a)} \tilde A_{\text{tot}}^{(a)}.
\end{align}
Total-abundance measurements recently used for this purpose include counting cells with microscopy (Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020})) or flow cytometry (Props et al. (\protect\hyperlink{ref-props2017abso}{2017}), Vandeputte et al. (\protect\hyperlink{ref-vandeputte2017quan}{2017}), Galazzo et al. (\protect\hyperlink{ref-galazzo2020howt}{2020})), measuring the concentration of a marker-gene with qPCR or ddPCR (Zhang et al. (\protect\hyperlink{ref-zhang2017soil}{2017}), Barlow, Bogatyrev, and Ismagilov (\protect\hyperlink{ref-barlow2020aqau}{2020}), Galazzo et al. (\protect\hyperlink{ref-galazzo2020howt}{2020}), Tettamanti Boshier et al. (\protect\hyperlink{ref-tettamantiboshier2020comp}{2020})), and measuring bulk DNA concentration with a florescence-based DNA quantification method (Contijoch et al. (\protect\hyperlink{ref-contijoch2019gutm}{2019})).

Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency.
We can easily understand the impact of taxonomic bias on total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
Suppose that each species \(i\) has an \emph{absolute efficiency} \(B_{i}^{{\text{[tot]}}}\) for the total-abundance measurement that is constant across samples.
Further, let \(\bar B^{{\text{[tot]}}(a)}\) be the abundance-weighted average of these efficiencies in sample \(a\)---that is, the mean efficiency of the total-abundance measurement.
Neglecting other error sources, the total-abundance measurement equals
\begin{align}
  \label{eq:total-density-error}
  \tilde A_{\text{tot}}^{(a)} 
  &= \sum_{i\in S} A_i^{(a)} B_{i}^{{\text{[tot]}}}
  \\&= A_{\text{tot}}^{(a)} \bar B^{{\text{[tot]}}(a)}.
\end{align}

Species abundance measurements derived by this method are affected by taxonomic bias in both the MGS and total-abundance measurement.
We can determine the resulting fold error (FE) by substituting Equations \eqref{eq:prop-error} and \eqref{eq:total-density-error} into Equation \eqref{eq:density-prop-meas}, yielding
\begin{align}
  \label{eq:density-prop-error}
  \tilde A_{\text{tot}}^{(a)}
  = A_{\text{tot}}^{(a)} \cdot \frac{B_i \bar B^{{\text{[tot]}}(a)}}{\bar B^{(a)}}.
\end{align}
Equation \eqref{eq:density-prop-error} indicates that the FE in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
As in the case of proportions (Equation \eqref{eq:prop-error}), the FE depends on sample composition through the two mean efficiency terms and so will vary across samples unless the two perfectly covary.

\hypertarget{leveraging-information-about-a-reference-species}{%
\subsubsection{Leveraging information about a reference species}\label{leveraging-information-about-a-reference-species}}

Suppose that the absolute abundance of a \emph{reference species} \(r\) has been fixed by the experimenter or been measured by independent means.
This known or measured abundance \(\tilde A_{r}^{(a)}\) can be used in conjunction with the MGS read counts to obtain absolute abundances for all species.
In the absence of taxonomic bias, the ratio of a species' absolute abundance to its MGS read count is the same for all species in a given sample (Equation @ref\{eq:mgs-model\}).
Hence the known ratio for the reference species can serve as conversion factor for obtaining the absolute abundance of a species \(i\) from its read count,
\begin{align}
  \label{eq:density-ratio-meas}
  \tilde A_i^{(a)} &= M_i^{(a)} \cdot \frac{\tilde A_{r}^{(a)}}{M_i^{(a)}}.
\end{align}
Let \(\mathop{\mathrm{FE}}[\tilde A_r^{(a)}] \equiv {\tilde A_r^{(a)}}/{A_r^{(a)}}\) be the FE in the reference measurement.
The effect on \(\tilde A_i^{(a)}\) of taxonomic bias in the MGS measurement can be determined by substituting Equation \eqref{eq:ratio-error} into Equation \eqref{eq:density-ratio-meas}, yielding
\begin{align}
  \label{eq:density-ratio-error}
  \tilde A_i^{(a)} = A_i^{(a)} \cdot \frac{B_i}{B_r} \cdot 
%    \frac{\tilde A_r^{(a)}}{A_r^{(a)}}.
    \mathop{\mathrm{FE}}\left[\tilde A_r^{(a)}\right].
\end{align}
The FE in \(\tilde A_i^{(a)}\) consists of two terms: the relative efficiency of species \(i\) to species \(r\) in the MGS measurement (\({B_i}/{B_r}\)) and the FE in the reference species' abundance (\({\tilde A_r^{(a)}}/{A_r^{(a)}}\)).

A common application of this approach involves adding a `spike-in' (as described above) in a known (and typically constant) abundance across samples (Stämmler et al. (\protect\hyperlink{ref-stammler2016adju}{2016}), Ji et al. (\protect\hyperlink{ref-ji2019quan}{2019}), Harrison et al. (\protect\hyperlink{ref-harrison2021theq}{2021}), REFs).
In this case, the reference abundance \(\tilde A_r^{(a)}\) is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
(In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \eqref{eq:density-ratio-meas}; Appendix REF).

Others have instead sought to determine naturally-occurring species that are thought to be constant across samples; we refer to such species as \emph{housekeeping species} by analogy with the housekeeping genes used for absolute-abundance conversion in gene-expression studies (REF).
Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference (REFs).
A related approach involves computationally identifying species that are constant between pairs of samples (David et al. (\protect\hyperlink{ref-david2014host}{2014})) or between sample conditions (Mandal et al. (\protect\hyperlink{ref-mandal2015anal}{2015}), Kumar et al. (\protect\hyperlink{ref-kumar2018anal}{2018})).
The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set \(\tilde A_r^{(a)}\) to 1 in Equation \eqref{eq:density-ratio-meas}.
The resulting abundance measurements have unknown but fixed units, which is sufficient for measuring fold changes across samples.

We suggest an additional way of using the reference-species strategy even in the absence of a spike-in or constant species:
Performing targeted measurements of the absolute abundance of one or more naturally occurring species.
These species can then be used as reference species in Equation \eqref{eq:density-ratio-meas} to measure the absolute abundances of all species.
The most common form of targeted measurement involves using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA.
It is also possible to directly measure cell concentration by performing ddPCR prior to DNA extraction (REFs), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
Appendix {[}REF{]} describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.

\hypertarget{differential-abundance}{%
\section{How bias affects DA results}\label{differential-abundance}}

We now turn to how the errors in MGS-based abundance measurements affect DA results.
Our focus is on \emph{multiplicative DA}---the fold differences (FD) in the relative or absolute abundance of species between samples or between experimental conditions.
Multiplicative DA has a direct ecological interpretation via the multiplicative processes of exponential growth and decay and forms the basis for several popular DA methods (REF DESeq2, ALDEx2(?), and corncob; a LV paper; fido).
We also consider non-parametric rank-based methods; these methods are popular in case-control studies for which ecological interpretability is often of secondary importance to discovering stable associations between species and the condition of interest (REFs).

\hypertarget{fold-change-between-a-pair-of-samples}{%
\subsection{Fold change between a pair of samples}\label{fold-change-between-a-pair-of-samples}}

The building blocks of a multiplicative DA analysis are the measured FDs in a species' abundance between pairs of individual samples.
If the fold error (FE) in a species' abundance measurement is constant across samples, then it will not affect the measured FDs between samples.
In Section \ref{abundance-measurement}, we showed that consistent taxonomic bias creates a constant FE in the ratio between two species \(i\) and \(j\), equal to the ratio in their efficiencies (Equation \eqref{eq:ratio-error}).
This error completely cancels when calculating the FD of this ratio from sample \(a\) to \(b\),
\begin{align}
  \label{eq:ratio-fc-error}
\underbrace{\frac{\tilde R_{i/j}^{(b)}}{\tilde R_{i/j}^{(a)}}}_\text{measured FD}
  = \frac
    {R_{i/j}^{(b)} \cdot \cancel{B_i / B_j}}
    {R_{i/j}^{(a)} \cdot \cancel{B_i / B_j}}
  =
  \underbrace{\frac{R_{i/j}^{(b)}}{R_{i/j}^{(a)}}}_\text{actual FD}
  .
\end{align}
In contrast, bias created a FE in the proportion of a species (Equation \eqref{eq:prop-error}) that varied inversely with the mean efficiency of the sample.
Therefore, this error does not cancel when calculating the FD in the proportion,
\begin{align}
  \label{eq:prop-fc-error}
\underbrace{\frac{\tilde P_i^{(b)}}{\tilde P_i^{(a)}}}_\text{measured FD}
  = \frac
    {P_i(b) \cdot \cancel{B_i} / \bar B^{(b)}}
    {P_i(a) \cdot \cancel{B_i} / \bar B^{(a)}}
  =
  \underbrace{\frac{P_i^{(b)}}{P_i^{(a)}}}_\text{actual FD}
  \cdot
  \underbrace{\left[\frac{\bar B^{(b)}}{\bar B^{(a)}}\right]^{-1}}_\text{FE}
  .
\end{align}
The varying mean efficiency does not cancel, leaving an FE in the FD of the species' proportion that is equal to the inverse FD in the mean efficiency.
Notably, this FE is the same for every species.

The different effects of taxonomic bias on the measured FDs in proportions versus ratios are illustrated in a hypothetical example in Figure \ref{fig:error-proportions} (bottom row).
In this example, the mean efficiency decreases 2.6-fold from Sample 1 to Sample 2 (FD of 0.4), causing the FD in the proportion of each species to appear 2.6-fold larger than its true value.
Though the FE in the FD is the same for each species, the biological implications of the error vary.
In particular, there are three distinct types of error: an increase in the magnitude of the measured FD, a decrease in that magnitude, and a change in the direction of the FD.
We refer to the latter as a \emph{sign error} in reference to the sign of the measured log fold difference (LFD).
We can see each type of error in Figure \ref{fig:error-proportions}.
For Species 1, which increases in abundance and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FD (actual FD: 2.3, measured FD: 6.5).
For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FD: 0.15, measured FD: 0.44).
For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FD: 0.6, measured FD: 1.8), such that the species actually appears to increase (a sign error).
In contrast, the ratios between species change by the same factors in the actual and measured taxonomic profiles.
For example, the ratio of Species 2 to Species 3 changes from \(1/1\) to \(1/4\) in the actual profiles and from \(3\) to \(3/4\) in the measured profiles, a decrease by 4-fold either way.

Taxonomic bias affects the measured FDs in a species' absolute abundance in a similar fashion.
First, consider a species' abundance measurement \(\tilde A_i^{(a)}\) derived from a non-MGS measurement of total-community abundance using Equation \eqref{eq:density-prop-meas}.
Equation \eqref{eq:density-prop-error} indicates that the FE in this measurement equals the (constant) species efficiency multiplied by \({B^{{\text{[tot]}}(a)}}/{\bar B^{(a)}}\), the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
This ratio can vary, creating error in the measured FD between samples.
Notably, if the FDs in the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, then the two offset each other and lead to more stable FEs (and hence more accurate FDs) in \(\tilde A_i^{(a)}\).
We discuss how this possibility might be exploited in real experimental workflows in Section \ref{solutions}.

Now consider a species' abundance derived from a reference species using Equation \eqref{eq:density-ratio-meas}.
Equation \eqref{eq:density-ratio-error} indicates that the FE equals a constant ratio in species' efficiencies multiplied by the FE in the measurement of the reference species' abundance.
If the abundance of the reference species can determined up to constant FE, then the FE in \(\tilde A_i^{(a)}\) will also be constant, leading to accurate FDs.

\hypertarget{regression-analysis-of-many-samples}{%
\subsection{Regression analysis of many samples}\label{regression-analysis-of-many-samples}}

Most DA analyses can be understood as a regression analysis.
For example, the simple linear regression model,
\begin{align}
  \label{eq:regression}
  y^{(a)} = \alpha + \beta x^{(a)} + \varepsilon^{(a)},
\end{align}
can be used to relate a microbial abundance variable \(y\) to a \emph{covariate} \(x\) and the \emph{residual (or unexplained) error} \(\varepsilon\).
For example, \(y\) may be the log absolute abundance of species \(i\) (\(\log A_i^{(a)}\)) and \(x\) may be a continuous variable (such as pH) or a binary variable (such as \(x=1\) for treated patients and \(x=0\) for untreated controls).
The regression coefficients \(\alpha\) and \(\beta\) are parameters to be estimated from the data.
A DA analysis is primarily interested in the slope coefficient \(\beta\), which determines how the average or expected value of \(y\) increases with \(x\).
For a binary covariate, \(\beta\) equals the average difference in \(y\) between samples with \(x=1\) versus those with \(x=0\).
Taking the response \(y\) to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient \(\beta\) corresponding to the average log fold difference (LFD) in abundance between conditions or for a unit increase in \(x\).

Appendix \ref{appendix-regression} derives mathematical results for the effect that measurement error in the response caused by taxonomic bias has on LFDs estimated via simple linear regression.
Here we illustrate these results with an example.
Suppose that we wish to estimate the average LFD in the absolute abundance of every species between samples from two conditions, such as treated and untreated patients.
Further suppose that we have measured absolute abundances by multiplying the proportions from MGS with an accurate measurement of total-community abundance.
For each species \(i\), we set \(y^{(a)}\) in Equation \eqref{eq:regression} equal to \(\log \tilde A_{i}^{(a)}\), and use either Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE) to estimate the coefficients \(\alpha\) and \(\beta\).
From Equation \eqref{eq:density-prop-error}, we know that the FE in the abundance of each species equals its efficiency relative to the mean efficiency of the sample.

To understand how the FE in species abundances affects the estimated LFD for a given species, it is useful to consider three possible scenarios for how the mean efficiency varies across samples.
In the first scenario, the mean efficiency remains stable across samples, such that the FE in \(\log \tilde A_{i}^{(a)}\) is approximately constant.
In this case, the FE induced by bias \emph{mostly cancels} and does not substantially affect the estimated LFD.
In the second scenario, the mean efficiency varies substantially but is not associated with the condition, creating random FEs in the individual sample measurements.
In this case, the LFDs between individual samples will be highly distorted, but these errors will effectively average out in the regression and the estimated LFD remains reliable.
Although the random FEs caused by bias do not create a systematic error in the LFD estimate, they will tend to reduce the precision (i.e., increase the standard error) of the estimate.
In the third scenario, the mean efficiency systematically varies with the condition, causing a systematic error in the estimated LFD regardless of the number of samples.
In particular, the portion of additive variation in the log mean efficiency that is linearly associated with \(x\) creates an equal but opposite additive error in the estimated LFD for each species.
Although the absolute additive error in the LFDs is the same for all species, its biological significance varies, depending on its magnitude relative to that of the species' true LFD.

Figure \ref{fig:regression-example} shows a simulated example in which the mean efficiency varies systematically with the condition of interest.
Here, the average log mean efficiency is higher in the condition corresponding to \(x=1\), due to an increase of the high-efficiency Species 9.
This systematic increase in the log mean efficiency causes an artificial decrease in the estimated LFD for each species by an equal amount.
As we saw for the FD between individual samples, the biological significance of the systematic error in the LFD estimates varies by species, despite the absolute error being constant across species.
We see decreases in the magnitude of the LFD estimate (Species 9, 10, and 1 in Figure \ref{fig:regression-example}), changes in sign (Species 5), or increases in magnitude (remaining species) depending the value of the actual LFD relative to the error.
The portion of the variation in the log mean efficiency that is uncorrelated with \(x\) increases the standard errors for most species (Figure \ref{fig:regression-example} D).
The exception is Species 9, which is the primary driver of the mean efficiency dynamics in this example.
Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).

\begin{figure}
\centering
\includegraphics{figures/export-pdf/home/michael/research/differential-abundance-theory/notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example_files/figure-html5/main-figure-1.pdf}
\caption{\label{fig:regression-example}\textbf{Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.} This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by \(x=0\) and \(x=1\). In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFC, which drives the positive association of the log mean efficiency with the condition (shown in Panels A and B). This positive LFC in the log mean efficiency induces a systematic negative shift in the estimated LFCs of all species (Panels C and D). Panel D shows the mean LFC (points) with 95\% confidence intervals (CIs), for each species estimated from either the actual or the measured densities. The error (difference in LFC estimates on measured and actual) equals the negative LFC of the mean efficiency (shown in Panel B).}
\end{figure}

\hypertarget{rank-based-analyses}{%
\subsection{Rank-based analyses}\label{rank-based-analyses}}

Rank-based or `nonparametric' DA methods work by first applying a cross-sample rank-transformation to the abundance variable \(y\): The sample in which \(y\) has the smallest value receives a rank of 1, the sample with the second smallest value receives a rank of 2, etc.
These methods then analyze how the average rank varies across conditions.
Rank-based methods commonly used in microbiome DA analysis include the Wilcoxon-Mann-Whitney and Kruskal-Wallis tests for a discrete covariate (e.g.~case vs control) and Spearman's rank correlation for a continuous covariate (e.g.~pH).
They have been applied to species proportions (REFs), ratios (REFs), and absolute abundances (REFs).

Multiplying a variable by a constant factor does not change its cross-sample ranks; thus rank-based DA methods are unaffected by constant FE in the abundance variable \(y\).
We therefore conclude that consistent taxonomic bias does not affect rank-based DA analyses of species ratios or absolute abundances derived from a reference species.
In contrast, variable FE in \(y\) (driven by variation in the mean efficiency) can affect the cross-sample ranks; if the FE is associated with the covariate \(x\), it can cause systematic error in the DA result.
Thus consistent bias may impact rank-based analyses of species proportions and absolute abundances derived from a total-community abundance measurement.
Future work should consider whether rank-based DA methods are significantly more robust than multiplicative DA methods to variation in the mean efficiency.

\hypertarget{case-studies}{%
\section{Case studies ~}\label{case-studies}}

To better understand the potential impact of taxonomic bias on DA analysis in practice, we conducted several case studies spanning a range of biological scenarios and sequencing technologies.

\hypertarget{foliar-fungi-experiment}{%
\subsection{Foliar fungi experiment}\label{foliar-fungi-experiment}}

The taxonomic bias species present in `mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
In practice it is difficult to construct control communities that span the many species present in complex natural communities.
However, gnotobiotic community experiments are well suited to this form of \emph{calibration via community controls} since, unlike most natural ecosystems, it is possible to assemble mock communities containing all species in known relative abundances.

In a study of host-commensal-pathogen interactions, Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) inoculated plants with 8 commensal fungal species and subsequently exposed plants to a fungal pathogen.
The authors used ITS amplicon sequencing to measure communities before and after pathogen infection.
Motivated by the substantial ribosomal copy-number variation (CNV) in fungi (Lofgren et al. (\protect\hyperlink{ref-lofgren2019geno}{2019})), the authors also performed control measurements of mock communities that they constructed from quantified genomic DNA of the 9 species in the experiment; these controls were used to measure taxonomic bias with the method of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
The authors found a 13-fold difference between the most and least efficiently measured commensal, while the pathogen was measured 40-fold more efficiently than the least efficiently measured commensal.

Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of `priority effects'---the advantage gained by a species from being allowed to colonize first.
Both analyses were based on fold differences (FDs) in species proportions and so in principle were sensitive to taxonomic bias.
To improve accuracy, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.

We repeated the two DA analyses of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) with and without calibration and found that the results did not meaningfully differ.
To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \ref{fig:leopold2020host-variation}).
Despite the 13-fold variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \ref{fig:leopold2020host-variation}C), having a geometric range of 1.62 and a geometric standard deviation of 1.05.
This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \ref{fig:leopold2020host-variation}A).
But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3-fold and by just 1.5-fold between the two most dominant.
The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual species, it had a negligible impact on the inferred FDs and the DA analyses based on them.

We performed an additional DA analysis on the data from Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) to investigate whether any commensals increased in absolute concentration in response to infection.
The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \ref{fig:leopold2020host-infection-mean-efficiency-dist}).
Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5-fold to 5.2-fold.
Using gamma-Poisson (or negative-binomial) regression of the observed read counts, we estimated the average LFD in commensal species proportions following infection with and without calibration (SI Figure \ref{fig:leopold2020host-infection-lfc}).
Commensals typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
However, failure to account for bias caused the decrease to be overestimated, by an amount corresponding to the inverse change in log mean efficiency.
For commensal-host pairs with relatively small observed decreases, bias correction greatly reduced the magnitude of the negative LFDs or, in several cases, resulted in LFDs that were near 0 or slightly positive.

Although Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) did not include absolute-abundance measurements, we can consider the impact taxonomic bias would have on an absolute DA analysis in a simple scenario in which total genome concentration of each pre- and post-infection sample is perfectly known.
The total genomic concentrations of each sample, combined with the species proportions measured by MGS, can be combined into a measurement of species concentrations using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}).
In this case, the bias in the MGS measurements will create absolute errors in the estimated LFDs of genome concentration of equal magnitude to the errors in the LFD estimates for proportions.
The scientific error, however, might become worse.
Suppose that total abundance increased by 2-fold due to pathogen growth; in this case, species whose proportions remained approximately constant would have increased in absolute abundance by around 2-fold.
Bias shifts FD estimates downwards by 2.5-fold to 5.2-fold (depending on host genotype).
Hence, without bias correction, we would instead conclude that these species decreased.

\hypertarget{vaginal-microbiomes-of-pregnant-women}{%
\subsection{Vaginal microbiomes of pregnant women}\label{vaginal-microbiomes-of-pregnant-women}}

A growing number of MGS studies have used DA analysis to describe associations between members of the vaginal microbiome and health conditions including urinary tract infections, sexually-transmitted infections, bacterial vaginosis, and preterm birth.
However, these associations often vary between studies.
DA analyses of the vaginal microbiome are commonly based on proportions, creating an opportunity for taxonomic bias to impact results.
Substantial taxonomic bias has been experimentally demonstrated in MGS protocols applied to vaginal samples or in vitro samples of vaginally-associated species (Yuan et al. (\protect\hyperlink{ref-yuan2012eval}{2012}), Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}), Gill et al. (\protect\hyperlink{ref-gill2016eval}{2016}), Graspeuntner et al. (\protect\hyperlink{ref-graspeuntner2018sele}{2018})).
The different biases of different MGS protocols have been proposed as one explanation for discrepancies in DA results across studies (Callahan et al. (\protect\hyperlink{ref-callahan2017repl}{2017})).

As part of the Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) study, Fettweis et al. (\protect\hyperlink{ref-fettweis2019thev}{2019}) collected longitudinal samples from over 1500 pregnant women, including nearly 600 that were taxonomically profiled by amplicon sequencing of the 16S V1-V3 hypervariable region.
The taxonomic bias of the MOMS-PI MGS protocol was previously quantified by McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) using control measurements by Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}) of cellular mock communities of seven common, clinically-relevant vaginal bacterial species.
Of these, \emph{Lactobacillus iners} had the highest efficiency, which was nearly 30-fold larger than that of the species with the lowest efficiency, \emph{Gardnerella vaginalis}.
A second \emph{Lactobacillus} species, \emph{L. crispatus}, had an efficiency that was approximately 2-fold less than \emph{L. iners} and 15-fold greater than \emph{G. vaginalis}.
These species, along with the unculturable \emph{Lachnospiraceae BVAB1}, are most frequently the most abundant species in a sample and commonly reach high proportions of over 70\% (REF SI analysis).
Therefore, shifts between these dominant species could drive large changes in the sample mean efficiency, which might in turn distort DA analyses.

We sought to understand the potential impact of bias on DA analysis of vaginal microbiome profiles from the MOMS-PI study.
We first examined variation in the mean efficiency across samples under the assumption that bias in the MOMS-PI study was accurately represented by the Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}) control measurements.
Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample (REF SI Analysis).
The mean efficiency varies substantially across vaginal samples (Figure 4.1), with samples in which a \emph{Lactobacillus} species is most abundant typically having a mean efficiency that is 3-fold to 20-fold greater than samples in which \emph{G. vaginalis} is most abundant.
The vaginal microbiome sometimes shifted between \emph{Lactobacillus}-dominance and \emph{Gardnerella}-dominance between consecutive visits in individual women (SI Figure \ref{fig:momspi-mean-efficiency-fcs}), which caused magnitude and sign errors in the trajectories of lower-abundance species (SI Figure \ref{fig:momspi-trajectory}).

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files/figure-html5/momspi-mean-efficiency-dist-1.pdf}
\caption{\label{fig:momspi-mean-efficiency-dist}\textbf{The mean efficiency in vaginal samples from the MOMS-PI study varies with the most abundant species.}}
\end{figure}

These observations suggest that a proportion-based DA analysis of a health condition that is associated with \emph{Lactobacillus} and/or \emph{Gardnerella} would be prone to spurious results.
To demonstrate this possibility, we performed a DA analysis of the MOMS-PI samples with respect to a simulated health condition.
Bacterial vaginosis (BV) has been repeatedly found to be associated with a dramatically reduced proportion of \emph{Lactobacillus spp.}, an increased proportion of \emph{G. vaginalis}, and higher within-community (alpha) diversity (Srinivasan et al. (\protect\hyperlink{ref-srinivasan2012bact}{2012}), Cartwright et al. (\protect\hyperlink{ref-cartwright2018mult}{2018})).
As a proxy for clinical BV, we split samples into high diversity (BV-like), intermediate diversity, and low diversity (non-BV-like) groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
The mean efficiency was lower in the high-diversity BV-like samples (mean LFD of -1.6) due to a lower abundance of \emph{Lactobacillus spp.}.
We estimated the LFD in proportion from non-BV-like to BV-like samples for 30 prevalent bacterial species, with and without bias correction, using gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions.
We accounted for bias by including an offset term in the linear model equal to the log ratio of the species' efficiency to the sample mean efficiency.
As expected, bias correction reduced the estimated LFD from non-BV-like to BV-like groups for every species tested, though the importance of this effect varied among species (SI Figure \ref{fig:momspi-regression}).

\hypertarget{human-gut-microbiomes}{%
\subsection{Human gut microbiomes}\label{human-gut-microbiomes}}

We wondered whether DA in the human gut microbiome might be less affected by bias than the vaginal microbiome, due to its greater species diversity.
Recall that the mean efficiency is an abundance-weighted average over species in a sample; thus we might expect it to become more stable as the within-sample (alpha) diversity increases.
In fact, one can show that for communities that are randomly assembled from a large species pool, the additive variance in the mean efficiency is inversely proportional to order-2 diversity \(^2D\) (also known as the Inverse Simpson diversity index) (Appendix \ref{diversity-and-mean-efficiency}).
But real communities are assembled through complex processes that create positive and negative associations among species (REF).
Thus there is unlikely to be a simple 1-1 relationship between alpha diversity and variation in the mean efficiency.
Moreover, even if the mean efficiency does vary less across the more diverse gut samples, if the proportion of each species also varies less, then the importance of bias for DA analyses need not diminish.

To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project (Huttenhower et al. (\protect\hyperlink{ref-huttenhower2012stru}{2012}), SI Methods \ref{methods-gut}).
On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \ref{fig:gut}A; GM \(\md\) GSD of 6.3 \(\md\) 1.7 in stool samples and 1.4 \(\md\) 1.6 in vaginal samples).
To assess the potential importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the multiplicative variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth.
Across simulation replicates, the GSD in the mean efficiency was typically lower in stool than in vaginal samples (SI Figure \ref{fig:gut}B; ratio of GSD in vaginal samples to GSD in stool samples of 1.6 \(\md\) 1.5 across 1000 simulations).
Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \ref{fig:gut}C), which the GSD in the proportion of a gut species across gut samples an average of 1.75-fold lower than that of a vaginal species across vaginal samples.
Recall that the importance of bias estimation of the FD in a species' proportion depends on GSD in the mean efficiency relative the GSD in the species' proportion (Section \ref{differential-abundance}).
Thus these results suggest that although the mean efficiency likely varies less across stool than vaginal samples, bias may be just as problematic for proportion-based DA analyses.

\hypertarget{microbial-growth-in-marine-sediments}{%
\subsection{Microbial growth in marine sediments}\label{microbial-growth-in-marine-sediments}}

The surface layers of marine sediments harbor diverse and abundant microbiomes.
Total cell density and species richness decrease with depth as resources are consumed in older, deeper sediment layers; however, some taxa are able to grow and increase in density as they are slowly buried.
Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020}) performed a systematic assessment of growth rates of bacterial and archaeal taxa over a depth of 10 cm (corresponding to \textasciitilde40 years of burial time) in sediment of the White Oak River estuary.
Taxa proportions were measured with 16S amplicon sequencing and total community densities were measured by counting cells using epifluorescence microscopy.
The authors then calculated cell concentrations for each taxon by multiplying the taxon's proportion by the total concentration, a method they referred to as FRAxC measurements for `fraction of 16S reads times total cell counts'.
Taxon-specific growth rates were inferred from the slope of a simple linear regression of the log of the calculated concentration of each taxon against burial time over the first 3 cm below the bioirrigation layer (corresponding to \textasciitilde8 years of burial).

The FRAxC concentration measurements made by Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020}) correspond to the total-abundance approach to inferring absolute abundance (Equation \eqref{eq:density-prop-meas}).
Accordingly, taxonomic bias could lead to systematic error in FRAxC-derived growth rates if sample mean efficiency tends to systematically vary with burial time.
Such systematic variation could occur if microbes with tougher cell walls both persist longer in the sediment and are more difficult to extract DNA from than microbes with weaker cell walls.
In this scenario, the relative abundance of tougher species will increase with burial time, as the cells of weaker species degrade.
This shifting composition will cause the sample mean efficiency to decrease with time, leading to inflated growth-rate estimates for all taxa.
Taxa that decay with depth sufficiently slowly would mistakenly be inferred to have positive growth from the FRAxC-calculated abundances.

Importantly, Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020}) also used qPCR to measure the absolute abundance of two taxa from these same samples.
By comparing qPCR to FRAxC growth rates for these taxa, we can estimate the systematic error in FRAxC growth rates.
Because the systematic error that bias causes in the regression slope is the same for each taxon (Section \ref{differential-abundance}), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.

The first soil core included qPCR measurements of a single archaeal taxon, \emph{Bathyarchaeota}, for which growth rates by qPCR and FRAxC were nearly identical (doubling rates of 0.099/yr by FRAxC and 0.097/yr by qPCR).
The second soil core included qPCR measurements of \emph{Bathyarchaeota} and a second taxon, \emph{Thermoprofundales}/MBG-D.
In this core, FRAxC and qPCR growth rates differed more substantially, with growth rates from FRAxC being larger by 0.012/yr for \emph{Bathyarchaeota} (0.112/yr by FRAxC and 0.1/yr by qPCR) and by 0.086/yr for \emph{Thermoprofundales}/MBG-D (0.294/yr by FRAxC and 0.208/yr by qPCR).
Uncertainty in the FRAxC- and qPCR-derived growth rates is not reported and is likely substantial; however, the fact that the FRAxC-derived rates are larger than qPCR-derived rates in all three cases is consistent with our hypothesis that mean efficiency decreased with depth in a manner that systematically biased FRAxC-derived rates to higher values.
The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr suggests an error large enough to impact results for some taxa classified as positive growers, whose FRAxC growth rates ranged between 0.04/yr and 0.5/yr.
In summary, the comparison between FRAxC and qPCR measurements supports the overall study conclusions, but also suggests that species at the low end of positive FRAxC-derived rates may be merely persisting or even slowly declining in abundance.

\hypertarget{summary-and-discussion}{%
\subsection{Summary and discussion}\label{summary-and-discussion}}

The effect that consistent taxonomic bias has on proportion-based DA analyses depends on the MGS protocol, the biological system, and the sample comparisons under investigation.
Our case studies explore a limited range of possibilities; however, we can begin to form some general conclusions by placing them within the context of the three scenarios described in Section \ref{differential-abundance}.

We saw several cases where the mean efficiency is stable across samples, so that bias does not affect LFD estimates (Scenario 1).
In the pre-infection fungal microbiome samples of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}), we observed that the mean efficiency remained stable despite substantial bias and large multiplicative variation in species proportions among samples.
Because the variation in the mean efficiency was much less than that of the individual species, the impact of bias on DA analysis was negligible.
In the vaginal case study, we also observed that the mean efficiency was relative stable across vaginal microbiomes that were dominated by the same species, despite substantial bias and large LFDs in non-dominant species.
In both cases, the stability of the mean efficiency can be understood by the fact that it is an additive average over species and so is primarily determined by the most abundant species in the sample.

Yet we also observed cases where the mean efficiency varied substantially.
In several cases, the mean efficiency varied systematically with the condition of interest sufficient to cause substantial systematic error in LFD estimates (Scenario 3).
Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time that is sufficient to substantially inflate the estimated growth rates of slowly changing species.
The mean efficiency cannot differ by more than the species proportions constituting it.
Therefore, the error in the estimated LFDs tend to be practically significant only for species whose LFDs are substantially smaller in magnitude than the largest LFD.

Variation in the mean efficiency may also be substantial, yet unassociated with the condition of interest (Scenario 2).
Although we did not directly observe this scenario directly, we have reason to think that it may be common in real microbiome studies.
Our simulation analysis of gut microbiomes suggest that even in diverse ecosystems, bias will often cause multiplicative variation in the mean efficiency that is comparable to that of individual species.
This variation is much less problematic for DA results when it not associated with the condition under study, since any loss in precision can in principle be offset by an increase in sample size.
For certain applications, however, it will be important to remember that the LFDs between individual samples remain unreliable.

Under what conditions should we expect the problematic third scenario?
Our case studies suggest one prominent mechanism for causing systematic variation in the mean efficiency: the existence of one or more species with unusually high or low efficiencies that tend to dominate the community more often in one sample condition than another.
This mechanism was responsible for the negative effect of bias in our DA analyses of foliar fungal microbiomes following infection and of vaginal microbiomes with low versus high diversity.
In experimental systems where no one species frequently forms a large fraction of the community, systematic variation of the mean efficiency can still arise through the collective change of many species that are associated with the condition of interest.
We described a plausible instance in the marine-sediment case study, where increased burial time selects for lysis-resistant (and so lower efficiency) species.
As another example, treatment with an antibiotic might selectively kill Gram negative species, which also tend to be easier to lyse, thereby decreasing the mean efficiency in fecal samples collected after treatment.
A generic mechanism which might spawn such condition-efficiency associations stems from the evolutionary relationships among species.
Species with recent common ancestry are expected to share a number of traits that determine measurement efficiency, including cell wall structure, genome size, ribosomal copy number, and PCR binding sequence.
They are also expected to share traits related to the condition of interest.
If related species tend to have similar efficiencies and similar associations with the condition, then they may drive an association of the mean efficiency with the condition.

These observations provide reasons to both worry and hope.
It seems likely that in many studies, the mean efficiency is consistent (first scenario) or is at least not associated with the condition (second scenario), so that DA inferences remain valid.
Yet it is not obvious which scenario any study falls into.
There are plausible mechanisms leading to systematic variation of the mean efficiency even in ecosystems with high species diversity.
Moreover, our explorations suggest that random variation in the mean efficiency is common and distorts comparisons between individual samples.
Thus while we should not discount the large set of existing DA results, we should seek ways to better assess the robustness of results from previous studies and to measure and correct the error caused by bias in future ones.

\hypertarget{solutions}{%
\section{Potential solutions}\label{solutions}}

\hypertarget{ratio-based-relative-da-analysis}{%
\subsection{Ratio-based relative DA analysis}\label{ratio-based-relative-da-analysis}}

Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using DA methods that analyze fold differences (FDs) in these ratios.
A variety of ratio-based methods have been developed under the framework of Compositional Data Analysis (CoDA; REFs).
CoDA methods may more generally consider the ratios between products of multiple species, possibly raised to some exponent; a simple example is the geometric mean of several species.
The mathematical operations of multiplication and exponentiation maintain constant FEs.
Hence such products provide a method for aggregating species into a higher-level taxonomic units that maintains bias invariance when estimating FDs.
Ratio-based analyses also avoid the potential for misinterpreting differences in a species' proportion that are driven by the sum-to-one constraint (Gloor et al. (\protect\hyperlink{ref-gloor2017micr}{2017})).

We note two important caveats.
First, in most MGS datasets, any given species is likely to be extremely rare in most samples, such that no sequencing reads are observed.
In order to analyze FDs, DA methods interpret these zero counts as evidence that the true abundance of the species is very small (but still positive).
A common approach to enable the computation of FDs in the presence of zeros is to add a small value, or `pseudo-count', to the read counts, making all values positive.
Unfortunately, this procedure violates bias invariance.
How bias interacts with more sophisticated approaches for handling zeros remains an important open question.
Second, the bias invariance of ratios among species does not extend to additive aggregates of species into higher-order taxa such phyla unless bias is conserved within the species group (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
Multiplicative aggregates of species, recently used for biomarker discovery (Rivera-Pinto et al. (\protect\hyperlink{ref-riverapinto2018bala}{2018}), Quinn and Erb (\protect\hyperlink{ref-quinn2020inte}{2020})), provide an alternative that remains bias-invariant but may be harder to interpret.

If biological or technical considerations favor a DA method that is not bias invariant, it can be beneficial to also use a ratio-based method as a robustness check.
For example, Hevroni et al. (\protect\hyperlink{ref-hevroni2020seas}{2020}) examined how the within-sample ranks of viral species changed from summer to winter.
Since changes in within-sample ranks can be affected by bias, the authors also considered the (bias-invariant) changes in the centered log ratios (CLR) associated with each species.
They found that the species whose within-sample ranks increased also increased in their CLR value, providing evidence that their initial results were not driven by bias.

\hypertarget{calibrate-compositions}{%
\subsection{Calibration using community controls}\label{calibrate-compositions}}

\emph{Community calibration controls} are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen `gold standard' or \emph{reference protocol} (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
Inclusion of these samples along with the primary samples in an MGS experiment can enable researchers to calibrate their MGS measurement by directly measuring and removing the effect of bias.
Calibration can be extended to species not in the controls by imputing the efficiencies of the missing species using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties such as cell-wall structure.

Calibration makes it possible to correct for bias in relative and absolute DA methods that would otherwise not be bias invariant.
To demonstrate, we used calibration to improve the estimated FDs between samples in the absolute abundance of a species in the bacterial mixtures from Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}).
We used the bias estimated from a single 7-species mixture to calibrate the MGS-derived proportions in all measured mixtures.
We obtained calibrated absolute abundances by multiplying the calibrated proportions by the known total abundance (Equation \eqref{eq:density-prop-meas}).
Calibration greatly increased the accuracy of the estimated FDs (Figure \ref{fig:calibration-example}, A vs.~B).
The fungal and vaginal case studies of Section \ref{case-studies} provide several practical examples of calibration in proportion-based DA analyses.

In studies of synthetic communities (like the fungal case study) or of ecosystems dominated by a relatively small number of culturable taxa (like the vaginal microbiome), calibration controls can be artificially constructed (`mock') communities.
In other cases, \emph{natural community controls} can be derived from aliquots of a homogenized microbiome sample.
Measurement of these controls across multiple experiments enables characterization of the \emph{differential bias} between experiments (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
Calibration using differential bias can make results directly comparable across studies that used different.

\begin{figure}
\includegraphics[width=1\linewidth]{notebook/_posts/2021-10-25-brooks2015thet-calibration/brooks2015thet-calibration_files/figure-html5/brooks2015thet_fc_calibration-1} \caption{\textbf{Fold differences can be calibrated using community controls or reference species.} The figure compares the performance of three methods for measuring fold differences (FDs) in absolute cell density in cellular mock communities of 7 vaginal species, which were constructed and measured via 16S sequencing by Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}). The `Uncalibrated' FDs are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which is constant by construction). The `Calibrated (community)' measurements are computed from abundance measurements where the proportions are first corrected for the taxonomic bias that was estimated from a single sample that contained all 7 species. The `Calibrated (reference)' measurements are computed from abundances measured with the reference-species method, with \emph{Lactobacillus crispatus} used as the reference; that is, the true abundance of \emph{L. crispatus} is treated as known and used to infer the abundance of the remaining 6 species. Only samples that contain \emph{L. crispatus} are included.}\label{fig:calibration-example}
\end{figure}

\hypertarget{choose-absolute-abundance-methods-with-more-stable-fes}{%
\subsection{Choose absolute-abundance methods with more stable FEs}\label{choose-absolute-abundance-methods-with-more-stable-fes}}

There are many methods for obtaining species (absolute) abundances from the relative abundances measured by MGS, which had previously not been evaluated for their sensitivity to taxonomic bias (Williamson, Hughes, and Willis (\protect\hyperlink{ref-williamson2021amul}{2021}) being a notable exception).
Our theoretical results show that these methods differ in the extent to which taxonomic bias causes the FEs in species abundances to vary across samples.
The effect of bias on absolute DA analysis can be mitigated by choosing methods with more stable FEs, using either of two general approaches.

\hypertarget{use-complementary-mgs-and-total-abundance-measurements}{%
\subsubsection{Use complementary MGS and total-abundance measurements}\label{use-complementary-mgs-and-total-abundance-measurements}}

A popular approach to measuring the absolute abundance of individual species is to multiply the species' proportions from MGS by a measurement of the abundance of the total community (Equation \eqref{eq:density-prop-meas}).
The resulting species abundances are affected by taxonomic bias in the MGS \emph{and} the total-abundance measurement (Equation \eqref{eq:density-prop-error}).
Section \ref{differential-abundance} shows that the effect of both forms of bias on DA results can be reduced by choosing MGS and total-abundance methods that have similar taxonomic bias.

Consider the debate over whether flow cytometry or 16S qPCR are better methods for measuring total abundance for the purposes of normalizing 16S amplicon sequencing data (Galazzo et al. (\protect\hyperlink{ref-galazzo2020howt}{2020}), Jian, Salonen, and Korpela (\protect\hyperlink{ref-jian2021comm}{2021})).
Flow cytometry directly counts cells, whereas 16S qPCR measures the concentration of the 16S gene following DNA extraction.
Thus the 16S qPCR measurement is subject to bias from extraction, copy-number variation, primer binding, and amplification, just like the 16S sequencing measurement.
Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it can lead to more accurate FDs across samples.

More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement may be qPCR the same marker gene, from the same DNA extraction.
Similar reasoning suggests that for shotgun sequencing, the ideal total-abundance measurement is bulk DNA quantification: Shotgun sequencing and bulk DNA quantification are both subject to bias from extraction and variation in genome size.
Optimal use of these pairings requires thoughtful choices during bioinformatics. For example, performing copy-number correction on amplicon read counts prior to multiplication by qPCR measurements would be counter-productive, as it decouples bias in the two measurements. Additionally, the MGS proportions should be computed prior to discarding any unassigned reads, since species that are missing from the given taxonomy database still contribute to the total concentration of marker-gene and/or bulk DNA.
Future experiments should evaluate the extent to which the taxonomic bias of DNA-based total-abundance measurements is stable across samples and shared with the complementary MGS measurement.

\hypertarget{normalize-to-a-reference-species}{%
\subsubsection{Normalize to a reference species}\label{normalize-to-a-reference-species}}

A second approach to obtaining species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \eqref{eq:density-ratio-meas}).
Our theory predicts that taxonomic bias induces constant FEs in the resulting species abundances which do not affect DA results.

Previous studies have used species with a constant abundance as references: spike-ins, computationally-identified `housekeeping species', and the host.
Researchers may naturally be concerned about bias between the reference species and the native species being normalized against it (see, for example, Harrison et al. (\protect\hyperlink{ref-harrison2021theq}{2021})'s recommendations for spike-in experiments).
Our results show that accurate FDs and DA results can be obtained so long as the relative efficiency of the focal species to the reference is consistent across samples.
This condition is much weaker than requiring that bias is small (the relative efficiency is close to 1) and so greatly expands the applicability of spike-ins and host normalization in microbiome experiments.

Section \ref{absolute-abundance} further proposed an additional class of reference species for normalizing MGS measurements: Species whose (absolute) abundance has been measured using a targeted method such as qPCR with species-specific primers.
To demonstrate the ability for reference calibration to reduce error in FD measurements, we treated the abundance of one species (\emph{Lactobacillus crispatus}) in the Brooks et al. (\protect\hyperlink{ref-brooks2015thet}{2015}) mock community data as known, and used it to calibrate the abundances of all species.
Doing so improved the resulting measurements of FDs in the abundance of all species (Figure \ref{fig:calibration-example}).
Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus expanding the applicability of reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
In many ecosystems, there may not be a single universally-present species that can serve as a reference in all samples.
Sample coverage can be increased by making targeted measurements of multiple species or of larger taxonomic groups; see Appendix REF for details.

\hypertarget{bias-sensitivity-analysis}{%
\subsection{Bias sensitivity analysis}\label{bias-sensitivity-analysis}}

Even if control measurements are not available, it is possible to computationally assess the sensitivity of a given DA result to taxonomic bias.
A \emph{bias sensitivity analysis} can be conducted by analyzing an MGS dataset multiple times under a range of possible taxonomic biases.
First, a large number of hypothetical biases are generated from a user-specified probability distribution.
Next, the DA analysis is re-run while using each simulated bias vector to calibrate the MGS measurements.
This approach is highly flexible, and can be used to investigate the bias sensitivity of any DA method.
Alternatively, existing DA methods can be extended to directly include the unknown taxonomic bias in their underlying statistical model, thereby providing DA estimates that inherently account for the added uncertainty in microbiome compositions due to the presence of unknown bias.
(Greenland et al. (\protect\hyperlink{ref-greenland2005mult}{2005}) compares these two approaches in a general context; Nixon et al. (\protect\hyperlink{ref-nixon2022asta}{2022}) applies the second approach to microbiome data in the absence of taxonomic bias.)

Interpreting a bias sensitivity analysis is complicated by the large percentage of zero counts in species-level microbiome profiles, as the results may strongly depend on how these zeros are handled in the calibration process.
Therefore it can be advantageous to jointly test the sensitivity of assumptions about zero-generating processes along with taxonomic bias, making statistical DA models that include bias and one or more zero-generating processes especially valuable.
The development of tools and workflows to facilitate bias sensitivity analysis may provide an efficient way to increase scientists' ability to assess the reliability of microbiome results, both for differential abundance and microbiome analyses more generally.

\hypertarget{bias-aware-meta-analysis}{%
\subsection{Bias-aware meta-analysis}\label{bias-aware-meta-analysis}}

Meta-analysis of microbiome samples measured across multiple studies must contend with the fact that different studies typically use different protocols and hence have different taxonomic biases.
These different biases can be explicitly accounted for in a \emph{bias meta-analysis} which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
Parametric meta-analysis models include study-specific latent parameters representing `batch effects'---non-biological differences in the data from each study which can distort the observed biological patterns.
By estimating these `nuisance parameters' along with the biological parameters of interest (such as the difference in a species' log abundance between conditions), the meta-analysis aims to reduce statistical bias in the biological parameters created by the non-biological differences among studies (Leek et al. (\protect\hyperlink{ref-leek2010tack}{2010}),Wang and LêCao (\protect\hyperlink{ref-wang2019mana}{2020})).
In a bias-aware meta-analysis, the meta-analysis model is configured so that (some of) the latent parameters correspond to study- and species-specific relative efficiencies.
If taxonomic bias is consistent within but not between studies, then this approach may improve the ability for the meta-analysis to accurately identify DA patterns as compared to meta-analysis methods in which the `batch effects' do not reflect the multiplicative and compositional nature of taxonomic bias.
The ability to infer and adjust for the differential bias between studies can be increased by measuring one or more shared control samples alongside the experiment-specific samples.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

It is commonly thought that analyses of the differences between samples are robust to the taxonomic bias inherent in MGS measurement as long as all samples have been subjected to the same MGS workflow.
But our results show that consistent taxonomic bias is still capable of creating scientific errors in common forms of DA analysis.
In particular, we showed mathematically that bias affects DA analysis methods based on proportions or in absolute abundances derived from them (using total-abundance normalization), due to variation in the mean efficiency across samples of varying taxonomic composition.
The error, however, varies by experimental context and may often be negligible in a practical sense.
Moreover, it can be quantified and corrected through the use of reference-taxon measurements and/or community calibration controls.
In addition, bias-aware sensitivity and meta-analyses make it possible to rigorously account for bias even in the absence of such controls.
Applications of these methods will gradually improve our understanding of the experimental contexts in which variation in the mean efficiency is likely to be problematic and should be accounted for.

We also showed that other DA methods are more robust---perhaps even entirely invariant---to consistent taxonomic bias.
In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
In addition, careful pairing of total-abundance and MGS measurement methods can counter the variation in mean efficiency of the MGS measurement, leading to more consistent fold errors across samples.

Important open questions and future research directions remain.
Key open questions include 1) determining the extent to which bias is consistent across samples for different taxonomic levels, MGS methods, and sample types; 2) assessing the validity of post-extraction measurements as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how different underlying community dynamics, and in particular the source of zero counts in MGS measurements, affect bias sensitivity.
In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, more sophisticated statistical tools are needed to properly account for both taxonomic bias and random variation in the underlying MGS and supplemental (e.g.~qPCR) measurements.
Finally, more concrete experimental recommendations and user-friendly statistical workflows are needed for implementing the solutions we propose in various experimental contexts.

Our theoretical framework and example analyses provide a foundation for addressing these open questions, as well as developing experimental protocols and statistical tools that implement our proposed solutions.
We look to a future in which microbiome researchers regularly choose an appropriate combination of experimental and data-analytic methods that are capable of answering their fundamental question while also accounting for taxonomic bias and other limitations inherent to MGS measurement.
In doing so, we will collectively gain the confidence needed to codify the findings from MGS-based microbiome studies into true scientific knowledge.

\hypertarget{appendix-appendix}{%
\appendix \addcontentsline{toc}{section}{\appendixname}}

\hypertarget{mgs-measurement-details}{%
\section{MGS measurement details}\label{mgs-measurement-details}}

\hypertarget{spike-in-alt}{%
\subsection{Alternative method for measuring species absolute abundances from a spike-in}\label{spike-in-alt}}

(stub)

\hypertarget{multiple-references}{%
\subsection{Missing and multiple reference species}\label{multiple-references}}

(stub)

\hypertarget{appendix-regression}{%
\section{Linear regression}\label{appendix-regression}}

\hypertarget{simple-linear-regression}{%
\subsection{Simple linear regression}\label{simple-linear-regression}}

This section derives the theoretical results of Section \ref{differential-abundance} for the effect of taxonomic bias on regression-based DA analyses.
As in Section \ref{differential-abundance}, we restrict our attention to DA analyses that can be expressed in terms of the simple linear regression model in which the response is log absolute abundance or log proportion of individual species or the log ratio of a pair of species.

\hypertarget{review-of-simple-linear-regression}{%
\subsubsection{Review of simple linear regression}\label{review-of-simple-linear-regression}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2308}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3846}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3846}}@{}}
\caption{\label{tab:statistical-notation} Statistical notation used in this section.}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Notation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Mathematical definition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Notation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Mathematical definition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} \\
\midrule
\endhead
\(x\), \(y\), \(z\), \(d\) & & Random variables, defined in text \\
\(x(a)\) & & Value of \(x\) in sample \(a\) \\
\(\bar x\) & \(\frac{1}{n}\sum_a x(a)\) & Sample mean of \(x\) \\
\(s(x,y)\) & \(\frac{1}{n}\sum_a (x(a) - \bar x) (y(a) - \bar y)\) & Sample covariance between \(x\) and \(y\) \\
\(s^{2}(x)\) & \(s(x,x)\) & Sample variance of \(x\) \\
\(s(x)\) & \(\sqrt{s^{2}(x)}\) & Sample standard deviation of \(x\) \\
\(r(x,y)\) & \(\frac{s(x,y)}{s(x) s(y)}\) & Sample correlation between \(x\) and \(y\) \\
\(\alpha\), \(\beta\), \(\sigma\) & & Parameters of the regression model \\
\(\hat \alpha\), \(\hat \beta\), \(\hat \sigma\) & & Parameter estimates \\
\(\epsilon(a)\) & \(y(a) - (\alpha + \beta x(a))\) & Error for sample \(a\) \\
\(\hat \epsilon(a)\) & \(y(a) - (\hat \alpha + \hat \beta x(a))\) & Residual for sample \(a\) \\
\(\text{se}(\hat \beta)\), \(\hat{\text{se}}(\hat \beta)\) & & Standard error of \(\hat \beta\) and its estimate \\
\bottomrule
\end{longtable}

We begin by reviewing definitions, notation, and results for the analysis of the simple linear regression model.
Our presentation follows Wasserman (\protect\hyperlink{ref-wasserman2004allo}{2004}) Chapter 13 with additional interpretation of the estimated regression coefficients.
Statistical notation is defined in Table \ref{tab:statistical-notation}.

Consider a response variable \(y\) and a covariate \(x\).
Following Wasserman (\protect\hyperlink{ref-wasserman2004allo}{2004}), we define the \emph{simple linear regression model} by
\begin{align}
  \label{eq:lm}
  y(a) &= \alpha + \beta x(a) + \varepsilon(a),
\end{align}
where \(E[\varepsilon(a) \mid x(a)] = 0\) and \(V[\varepsilon(a) \mid x(a)] = \sigma^2\).
That is, conditional on the value of \(x\) in a sample \(a\), the response \(y\) is given by \(\alpha + \beta x\) with a random error \(\varepsilon\) that has an expected value of \(0\) and a constant variance \(\sigma^2\).

Given data \((y, x)\), we \emph{fit} the model by finding estimates for the parameters \(\alpha\), \(\beta\), and \(\sigma^2\) that best reflect the data under the assumption that the model is correct and according to our chosen criterion that defines `best'.
Here we consider the least-squares estimates for the coefficients \(\alpha\) and \(\beta\), which are also the maximum likelihood estimates if we further assume that the errors \(\varepsilon\) are normally distributed.
The least-squares estimates of \(\alpha\) and \(\beta\) are the values that minimize the residual sum of squares, \(\text{RSS} = \sum_{a} \hat \varepsilon(a)^2\).
These estimates can be conveniently expressed in terms of the sample means and covariances or correlations of \(x\) and \(y\),
\begin{align}
  \label{eq:lm-hat}
  \hat \alpha &= \bar y - \hat \beta \bar x \\
  \hat \beta &= \frac{s(y,x)}{s^2(x)} = r(y,x) \cdot \frac{s(y)}{s(x)}.
\end{align}
The estimate for the slope coefficient, \(\hat \beta\), equals the ratio of the sample covariance of \(y\) with \(x\) relative to the variance in \(x\) or, equivalently, the correlation of \(y\) with \(x\) multiplied by the ratio of their standard deviations.
The estimate for the intercept, \(\hat \alpha\), is the value such that the sample means follow the regression relationship, \(\bar y = \hat \alpha + \hat \beta \bar x\).
The standard unbiased and maximum likelihood estimates for the variance \(\sigma^{2}\) are both approximately given by the sample variance of the residuals,
\begin{align}
  \label{eq:lm-var}
  \hat{\sigma}^2(\hat \beta) \approx s^2(\hat \varepsilon).
\end{align}

We are most interested in the estimated slope coefficient, \(\hat \beta\).
Besides the point estimate indicating our best guess at the value, we also wish to understand the uncertainty or precision of the estimate.
This uncertainty is quantified by the standard error, \(\text{se}(\hat \beta)\), estimated by
\begin{align}
  \label{eq:lm-se}
  \hat{\text{se}}(\hat \beta)
  = \frac{\hat \sigma}{s(x) \sqrt{n}}
  \approx \frac{s(\hat \varepsilon)}{s(x) \sqrt{n}},
\end{align}
Approximate 95\% confidence intervals for \(\beta\) are given by \(\hat \beta \pm 2\; \hat{\text{se}}(\hat \beta)\).

\hypertarget{measurement-error-in-the-response}{%
\subsubsection{Measurement error in the response}\label{measurement-error-in-the-response}}

Now consider a researcher who would like to regress a biological quantity \(y\) (the response) on a second quantity \(x\) (the covariate).
But instead of \(y\), they only know a proxy \(z\), which is related to \(y\) via the difference \(d = z - y\).
For instance, \(y\) might be the log abundance of a species, \(z\) is the abundance that is measured via MGS, and \(x\) is a numerical quantity like pH or a boolean variable indicting case versus control.
Lacking a direct measurement of \(y\), the researcher instead regresses \(z\) on \(x\) and interprets the fitted model as informing them about the relationship between \(y\) and \(x\).
We wish to understand the accuracy of their conclusions.

To understand how the measurement error in \(y\) impacts the researcher's regression analysis, consider the three linear regression equations for \(y\), \(z\), and \(d\),
\begin{align}
  \label{eq:lm-y}
  y(a) &= \alpha_y + \beta_y x(a) + \varepsilon_y(a),
\end{align}
\begin{align}
  \label{eq:lm-z}
  z(a) &= \alpha_z + \beta_z x(a) + \varepsilon_z(a),
\end{align}
\begin{align}
  \label{eq:lm-d}
  d(a) &= \alpha_d + \beta_d x(a) + \varepsilon_d(a).
\end{align}
The researcher would like to fit the model for \(y\) (Equation \eqref{eq:lm-y}), but instead can only fit the model for \(z\) (Equation \eqref{eq:lm-z}).
The two are related via the Equation \eqref{eq:lm-z} for \(d\).
It is helpful to imagine that \(y\) and \(d\) are known and so we can fit all three models \eqref{eq:lm-y}, \eqref{eq:lm-z}, and \eqref{eq:lm-d}.
From Equation \eqref{eq:lm-hat} and the linearity of sample means and covariances it follows that
\begin{align}
  \label{eq:lm-hat-rel}
  \hat \alpha_z &= \hat \alpha_y + \hat \alpha_d \\
  \hat \beta_z &= \hat \beta_y + \hat \beta_d.
\end{align}
That is, the estimated coefficients for \(y\), \(z\), and \(d\) mirror the relationship between the variables themselves, \(z = y + d\).
Consequently, we see that the measurement error \(d\) creates absolute errors in the regression coefficients \(\hat \alpha\) and \(\hat \beta\) equal to
\begin{align}
  \label{eq:lm-hat-err}
  \text{abs. error}(\hat \alpha) &\equiv \hat \alpha_z - \hat \alpha_y = \hat \alpha_d \\
  \text{abs. error}(\hat \beta) &\equiv \hat \beta_z - \hat \beta_y = \hat \beta_d.
\end{align}
Expressed in terms of sample means and covariances, these errors are
\begin{align}
  \label{eq:lm-hat-err-1}
  \text{abs. error}(\hat \alpha) &= \bar d - \hat \beta_d \bar x \\
  \text{abs. error}(\hat \beta)  &= \frac{s(d,x)}{s^2(x)} = r(d,x) \cdot \frac{s(d)}{s(x)}.
\end{align}

We are mainly interested in the estimated slope coefficient, \(\hat \beta\).
We see that the absolute error is large when the covariance of \(d\) with \(x\) is large.
This error is large in a practical sense when \(\hat \beta_{d}\) is large (in magnitude) relative to \(\hat \beta_y\) (Equation \eqref{eq:lm-hat}), which occurs when \(s(d,x)\) is large relative to \(s(y,x)\).
A sign error occurs when \(s(d,x)\) is larger in magnitude than \(s(y,x)\) and of opposite sign.

We can also consider how the measurement error affects the residual variation and the standard error of the slope estimate.
The residuals of \(y\), \(z\), and \(d\) are related through \(\hat \varepsilon_z = \hat \varepsilon_y + \hat \varepsilon_d\).
It follows that the sample variance of the \(z\) residuals equal
\begin{align}
  s^2(\hat \varepsilon_{z}) 
  = s^2(\hat \varepsilon_y + \hat \varepsilon_{d})
  = s^2(\hat \varepsilon_y) + s^2(\hat \varepsilon_{d}) + 2 s(\hat \varepsilon_y, \hat \varepsilon_{d}).
\end{align}
The variance of the \(z\) residuals is increased above that of the \(y\) residuals when the \(d\) residuals are either uncorrelated or positively correlated with the \(y\) residuals, but may be decreased when the \(y\) and \(d\) residuals are negatively correlated.
An increased residual variance will lead to a larger estimate for \(\sigma\) as well as larger standard errors in the slope coefficient \(\beta\).

\hypertarget{specific-application-to-taxonomic-bias}{%
\subsubsection{Specific application to taxonomic bias}\label{specific-application-to-taxonomic-bias}}

These general results can be used to understand how taxonomic bias affects DA analyses that can be expressed as linear regression of log (relative or absolute) abundance.

First, we illustrate for the particular example of absolute DA analysis, where species absolute abundance is measured with the total-abundance normalization method and we assume that the total abundances have been accurately measured.
Consider the regression of log abundance of species \(i\) on a covariate \(x\).
In this case, \(y(a)= \log \text{abun}_{i}(a)\) is the actual log abundance of species \(i\) and \(z(a)=\log \widehat{\text{abun}}_i(a)\) is the log abundance we've measured.
From Equation \eqref{eq:density-prop-error}, the measurement error \(d(a)\) due to taxonomic bias in the MGS measurement is
\begin{align}
  d(a) 
  \equiv z(a) - y(a) 
  = \log \text{efficiency}_i - \log \text{efficiency}_S(a).
\end{align}
The absolute error in \(\hat \beta\) equals the scaled covariance of \(d\) with \(x\) (Equation \eqref{eq:lm-hat-err-1}).
In the above expression for \(d(a)\), only the second term, \(- \log \text{efficiency}_S(a)\), varies across samples and thus affects the covariance, leading to
\begin{align}
  \text{abs. error}(\hat \beta)
    &= - \frac{s[\log \text{efficiency}_S, x]}{s^2(x)}.
%    = - \frac{r[\log \text{efficiency}_S, x] \cdot s[\log \text{efficiency}_S]}{s(x)}, \\
\end{align}
Thus the absolute error in \(\hat \beta\) is the negative of the scaled covariance of the log mean efficiency.

Although the absolute error is the same for each species, its practical significance varies.
Recall from Equation \eqref{eq:lm-hat} that the correct value for \(\hat \beta\) (i.e., the estimate without measurement error) is
\begin{align}
  \hat \beta &= \frac{s[\log \text{abun}_i, x]}{s^2(x)}.
\end{align}
For species whose log abundance covaries with \(x\) much more than the log mean efficiency, the error will be relatively small.
This situation can occur either because the mean efficiency varies relatively little across samples or because its variation is relatively uncorrelated with \(x\).
For species whose log abundance covaries with \(x\) similar or less than the log mean efficiency, the error will be significant.
A sign error occurs when the species covariance is of \emph{equal sign and smaller in magnitude} than the covariance of the log mean efficiency.

We can also consider how the standard errors in the slope estimate are affected by bias.
In our case, the \(d\) residuals equal the negative residuals of the log mean efficiency.
It is plausible that for most species, their residual variation will have a small covariance with log mean efficiency and the net effect of variation in the mean efficiency will be to increase the estimated standard errors, as occurs with most species in Figure \ref{fig:regression-example}.
However, high-efficiency species that vary substantially in proportion across samples may be strongly positively correlated with log mean efficiency such that the estimated standard errors decrease, as we see with Species 9 in Figure \ref{fig:regression-example}.

Now, suppose we were instead performing a DA analysis of a response whose fold error is constant across samples, such as the log ratio of two species.
For the log ratio of species \(i\) and \(j\), the error is \(d(a) = \log \text{efficiency}_i / \text{efficiency}_j\) and is constant across samples.
Thus the covariance of \(d\) with \(x\) is 0 and taxonomic bias causes no error the slope estimate \(\hat \beta\), only the intercept estimate \(\hat \alpha\).

\hypertarget{gamma-poisson-regression}{%
\subsection{Gamma-Poisson regression}\label{gamma-poisson-regression}}

This section describes how gamma-Poisson regression can, with appropriate choice of offsets, be used to estimate log fold changes (LFCs) in proportions from MGS count data either with or without bias correction.

\hypertarget{background}{%
\subsubsection{Background}\label{background}}

The gamma-Poisson distribution, also known as the negative binomial distribution, is a distribution that is commonly used to model sequencing count data (Holmes and Huber (\protect\hyperlink{ref-holmes2018mode}{2018})).
Its key advantage over the simple linear regression model is that it directly models the random sampling of reads that occurs during a sequencing experiment and, in this way, naturally accounts for the imprecision associated with an observation of zero or a small positive count for a given species and sample.
The gamma-Poisson distribution has two parameters, which jointly determine the mean and the standard deviation.
We use the parameterization used by the rstanarm R package (Goodrich et al. (\protect\hyperlink{ref-rstanarm}{2020})), which we use for fitting gamma-Poisson GLMs in our case-study analyses.
Suppose that \(y^{(a)}\) has a gamma-Poisson distribution conditional on a covariate \(x\).
Define two parameters \(\mu\) and \(\phi\) such that
\begin{align}
  E[y^{(a)} \mid x]  &= \mu^{(a)} \\
  sd[y^{(a)} \mid x] &= \sqrt{
    E[y^{(a)} \mid x] + E[y^{(a)} \mid x]^2 / \phi
  };
\end{align}
thus \(\mu\) equals the mean and \(\phi\), known as the `reciprocal dispersion', increases the standard deviation as it approaches 0 and causes the standard deviation to approach \(\sqrt{\mu}\) (that of a Poisson distribution) as it approaches infinity.

For use in gamma-Poisson GLM regression of MGS data, we partition the mean into two factors, \(\mu^{(a)} = u^{(a)} \theta^{(a)}\) (Gelman, Hill, and Vehtari (\protect\hyperlink{ref-gelman2020regr}{2020})).
Here, \(\theta^{(a)} = e^{\alpha + \beta x^{(a)}}\) is the quantity of primary interest, whose average LFC equals \(\beta\).
The factor \(u^{(a)}\) is a sample-specific \emph{exposure} parameter that determines the overall scale of counts.
We can equivalently write \(\mu^{(a)} = e^{\alpha + \beta x^{(a)} + \log u^{(a)}}\);
the logarithm of the exposure, \(\log u^{(a)}\), is known as the \emph{offset} of the GLM.
Performing gamma-Poisson regression with standard statistical software requires that the offsets are provided as known values.

\hypertarget{inferring-lfcs-in-proportions-with-and-without-bias-correction}{%
\subsubsection{Inferring LFCs in proportions with and without bias correction}\label{inferring-lfcs-in-proportions-with-and-without-bias-correction}}

Our goal is to use gamma-Poisson regression to estimate the LFC in a species proportion from the observed read counts.
We therefore equate the counts \(M_i^{(a)}\) with \(y^{(a)}\) and the proportion \(P_{i}^{(a)}\) with \(\theta^{(a)}\); it remains to determine the offsets that are consistent with our MGS model.
To do so, we relax the deterministic assumption of our original model and instead suppose that the right-hand side of Equation \eqref{eq:mgs-model} equals the \emph{expected} read count of species \(i\) in sample \(a\),
\begin{align}
  \label{eq:mgs-model-gp}
  E[M_i^{(a)}] = A_i^{(a)} B_i F^{(a)}.
\end{align}
The total expected count is
\begin{align}
  \label{eq:total-reads-gp}
  E[M_S^{(a)}] &= \sum_{i \in S} E[M_i^{(a)}]
  \\&= \left( \sum_{i \in S} A_i^{(a)} B_i \right) F^{(a)} 
  \\&= A_S^{(a)} B_S^{(a)} F^{(a)}.
\end{align}
Under the gamma-Poisson model for \(M_i^{(a)}\), we have that \(E[M_i^{(a)}] = u_i^{(a)} P_i^{(a)}\).
Equating this expression for \(E[M_i^{(a)}]\) with Equation \eqref{eq:mgs-model-gp} lets us solve for the exposure,
\begin{align}
  \label{eq:exposure}
  u_i^{(a)} 
    &= \frac{A_i^{(a)} B_i F^{(a)}}{P_i^{(a)}} 
  \\&= B_i A_S^{(a)} F^{(a)}
  \\&= \frac{B_i}{B_S} \cdot E[M_S^{(a)}].
\end{align}
The second line follows from \(P_i^{(a)} \equiv A_i^{(a)}/ A_S^{(a)}\), and the third line from Equation \eqref{eq:total-reads-gp}.

The final expression in Equation \eqref{eq:exposure} can be used to compute offsets for the regression.
The three terms \(B_i\), \(B_S^{(a)}\), and \(E[M_S^{(a)}]\) are each unknown, so we instead substitute estimates for these terms to obtain an estimated exposure \(\hat u_{i}^{(a)}\); we then set the offset in the linear model to \(\log \hat u_{i}^{(a)}\).
We can estimate \(E[M_S^{(a)}]\) by the observed total count, \(M_S^{(a)}\).
In the absence of bias correction, we set \(B_{i} = B_{S}^{(a)} = 1\); our estimate of the exposure is then just
\begin{align}
  \hat u_i^{(a)} = M_S^{(a)}.
\end{align}
We can apply bias correction using estimates of the efficiencies, \(\hat B_{i}\), derived from community control measurements.
From these estimates and the observed counts, we can estimate the mean efficiency in the sample by
\begin{align}
  \hat B_S^{(a)} = \sum_{i\in S} B_i \hat P_i^{(a)},
\end{align}
where \(\hat P_i^{(a)}\) are the calibrated proportions obtained from a simple plug-in procedure,
\begin{align}
  \hat P_i^{(a)} = \frac{M_i^{(a)} / B_i}{\sum_j M_j^{(a)} / B_j}.
\end{align}
Equivalently, we can calculate \(\hat B_S^{(a)}\) directly from the observed counts,
\begin{align}
  \text{B}_S^{(a)} =
  \left[\sum_j \frac{M_j^{(a)}}{M_S^{(a)} \text{B}_j}\right]^{-1}.
\end{align}
We then estimate the exposure by
\begin{align}
  \hat u_i^{(a)} = \frac{\hat B_i}{\hat B_S} \cdot M_S^{(a)}.
\end{align}

\hypertarget{diversity-and-mean-efficiency}{%
\section{Alpha diversity and variation in the mean efficiency}\label{diversity-and-mean-efficiency}}

The diversity of order \(q\) is given by
\begin{align}
  ^qD = \left(\sum_{i=1}^n p_i^q\right)^{1 /(1-q)},
\end{align}
where \(^qD\) is understood to be given by the limit of the RHS when \(q = 1\) (and so the given expression is undefined).
The order-2 diversity is therefore
\begin{align}
  ^2D = \frac{1}{\sum_{i=1}^n {p_i^2}},
\end{align}
which is equivalent to the Inverse Simpson Index (REF Jost).

Consider an infinite pool of species.
Let \(\sigma^{2}\) denote the variance in the relative efficiency among species in the pool.
Consider a community that is assembled by randomly choosing \(K \ge 1\) species from the pool and setting the abundances of the \(K\) species in a manner that is independent of their efficiencies.

\textbf{Claim:}
Let \(\rho_{k}\) for \(1 \le k \le K\) denote the proportions of the \(K\) species in the community; this new notation serves as a reminder that the subscript \(k\) indexes a random species specific to this particular community.
Conditional on the order-2 diversity \(^2D\) of a community, the arithmetic variance in the mean efficiency is
\begin{align}
  Var[\bar B \mid ^2D] 
% = \sigma^2 \sum_{k=1}^K \rho_k^2
  = \frac{\sigma^2}{^2D}.
\end{align}

\textbf{Proof:} Let \(\beta_{k}\) denote the efficiency of the \(k\)-th sampled species.
The mean efficiency is therefore \(\bar B = \sum_k \rho_k \beta_k\).
Conditional on the proportions \(\{\rho_{k}\}\), the variance in the mean efficiency is
\begin{align}
  Var[\bar B \mid \{\rho_{k}\}]
  &= \sum_{k=1}^K \rho_k^2 Var[\beta_k]
  \\&= \left(\sum_{k=1}^K p_k^2\right) \sigma^2.
\end{align}
The first line follows from the fact that the \(\beta_{k}\) are independent.
The summation in the final line is equal to \(1/2^D\), thus proving the result.

\textbf{Note:}
We have showed that the arithmetic (additive) variance of \(\bar B\) decreases with \(^2D\); however, the geometric (multiplicative) variance is most relevant for understanding the effect of bias on DA.
As \(2^D\) increases, the distribution of \(\bar B\) will converge (by the central limit theorem) to a normal distribution, and both the arithmetic and geometric variance will decrease.

\hypertarget{supplemental-figures}{%
\section{Supplemental figures}\label{supplemental-figures}}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study_files/figure-html5/variation-in-proportions-and-mean-efficiency-1.pdf}
\caption{\label{fig:leopold2020host-variation}\textbf{In the pre-infection samples from Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}), multiplicative variation in taxa proportions is much larger than that in the mean efficiency.} Panel A shows the distribution of the proportions of each commensal isolate (denoted by its genus) across all samples collected prior to pathogen inoculation; Panel C shows the distribution of the (estimated) sample mean efficiency across these same samples on the same scale; and Panel B shows the efficiency of each taxon estimated from DNA mock communities as point estimates and 90\% bootstrap percentile confidence intervals. Efficiencies are shown relative to the most efficiently measured taxon (\emph{Fusarium}).}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study_files/figure-html5/infection-mean-efficiency-dist-1.pdf}
\caption{\label{fig:leopold2020host-infection-mean-efficiency-dist}\textbf{The mean efficiency tends to increase after infection due to the high proportion of the pathogen.}}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2022-01-08-leopold2020host-case-study/leopold2020host-case-study_files/figure-html5/infection-gp-lfc-estimates-1.pdf}
\caption{\label{fig:leopold2020host-infection-lfc}\textbf{Bias correction increases the estimated log fold changes (LFCs) in commensal species proportions in response to pathogen infection.} Shown are the estimated LFCs with 95\% Bayesian credible intervals (CIs) derived from a Bayesian gamma Poisson regression. Negative values indicate that the species' proportion decreased on average in response to infection, which is the default expectation given growth of the pathogen and the sum-to-one constraint of proportions. Bias leads to artificially low estimates, as the increase in the pathogen proportion increases the mean efficiency. In Western genotypes, several species whose proportion appears to decrease are instead found to remain stable or even to increase when bias is corrected.}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files/figure-html5/momspi-mean-efficiency-fcs-1.pdf}
\caption{\label{fig:momspi-mean-efficiency-fcs}\textbf{Fold changes in the mean efficiency within and between women in the MOMS-PI study.}}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{figures/export-pdf/notebook/_posts/2021-11-01-momspi-summary/momspi-summary_files/figure-html5/trajectories-within-subject-high-var-1} \caption{\textbf{In vaginal microbiome measurements, shifts between \emph{Lactobacillus} and \emph{Gardnerella} dominance can drive spurious fold changes in other, lower-abundance species.} The figure shows species proportions and mean efficiency trajectories over consecutive clinical visits for a subject in the MOMS-PI study whose microbiome samples showed substantial variation in mean efficiency. The subject's samples are dominated by \emph{Gardnerella vaginalis} and \emph{Lachnospiraceae BVAB1} during the first three visits before transitioning to being dominated by \emph{Lactobacillus iners} between visits 3 and 4. This transition drives a sharp increase in the mean efficiency, which significantly distorts the fold changes in the observed (uncalibrated) microbiome measurements for species with less dramatic fold changes. Two exemplar species are shown to illustrate the magnitude (\emph{Ureaplasma cluster 23}) and sign (\emph{Megasphaera OTU70 type1}) errors that can arise in this situation.}\label{fig:momspi-trajectory}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity_files/figure-html5/regression-gp-1.pdf}
\caption{\label{fig:momspi-regression}\textbf{Bias inflates estimated log fold changes (LFCs) in species proportions with diversity in vaginal microbiome samples from the MOMS-PI study.} Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles. The LFC in proportion from low- to high-diversity samples was estimated for 30 common species using gamma-Poisson regression with and without bias correction. Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95\% Bayesian credible intervals for each species.}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2022-01-30-hmp-stool-vagina-comparison/hmp-stool-vagina-comparison_files/figure-html5/main-1.pdf}
\caption{\label{fig:gut}\textbf{Compared to vaginal samples, stool samples have higher diversity, lower variation in mean efficiency, and lower variation in species proportions.} This figure summarizes results from our case study of shotgun-sequenced stool and vaginal samples in the Human Microbiome Project. Panel A shows the distribution of order-2 alpha diversity (equal to the Inverse Simpson Index) across samples. Panel B shows the variation in the mean efficiency, as quantified by the GSD, across samples of a given type for each of 1000 simulated bias vectors. Panel C shows the variation in the proportion of each species commonly found in a given sample type. For Panel C, variation is measured by the GSD when a pseudo-value of \(10^{-4}\) is added to the raw proportions. Different choices of the pseudo-value substantially change the measured GSDs, but not the difference (ratio) between gut and vaginal species.}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/export-pdf/notebook/_posts/2021-10-18-evaluate-robustness-example/evaluate-robustness-example_files/figure-html5/summary_plot-1.pdf}
\caption{\label{fig:sensitivity-example}\textbf{A bias-sensivity analysis can be performed to examine how sensitive the results of a DA analysis are to assumptions about taxonomic bias in community measurements.} The figure shows the results of a bias-sensitivity analysis used to study the effect of bias on the association of \emph{Gardnerella vaginalis} and preterm birth that was investigated by Callahan et al. (\protect\hyperlink{ref-callahan2017repl}{2017}). 100 random efficiency vectors were drawn at 6 different bias strengths (quantified by the variance in log efficiency, \(\sigma_{e}^{2}\)). Each efficiency vector was used to calibrate the MGS profiles and perform a DA association test of \emph{G. vaginalis} versus the host's preterm birth outcome; regression coefficients \(\hat \beta\) indicate the increase of average logit proportion of \emph{G. vaginalis} in women who experienced preterm birth.}
\end{figure}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-barlow2020aqau}{}}%
Barlow, Jacob T., Said R. Bogatyrev, and Rustem F. Ismagilov. 2020. {``{A quantitative sequencing framework for absolute abundance measurements of mucosal and lumenal microbial communities}.''} \emph{Nat. Commun.} 11 (1): 1--13. \url{https://doi.org/10.1038/s41467-020-16224-6}.

\leavevmode\vadjust pre{\hypertarget{ref-brooks2015thet}{}}%
Brooks, J Paul, David J Edwards, Michael D Harwich, Maria C Rivera, Jennifer M Fettweis, Myrna G Serrano, Robert A Reris, et al. 2015. {``{The truth about metagenomics: quantifying and counteracting bias in 16S rRNA studies}.''} \emph{BMC Microbiol.} 15 (1): 66. \url{https://doi.org/10.1186/s12866-015-0351-6}.

\leavevmode\vadjust pre{\hypertarget{ref-callahan2017repl}{}}%
Callahan, Benjamin J, Daniel B DiGiulio, Daniela S Aliaga Goltsman, Christine L Sun, Elizabeth K Costello, Pratheepa Jeganathan, Joseph R Biggio, et al. 2017. {``{Replication and refinement of a vaginal microbial signature of preterm birth in two racially distinct cohorts of US women}.''} \emph{Proc. Natl. Acad. Sci. U. S. A.} 114 (37): 9966--71. \url{https://doi.org/10.1073/pnas.1705899114}.

\leavevmode\vadjust pre{\hypertarget{ref-cartwright2018mult}{}}%
Cartwright, Charles P., Amanda J. Pherson, Ayla B. Harris, Matthew S. Clancey, and Melinda B. Nye. 2018. {``{Multicenter study establishing the clinical validity of a nucleic-acid amplification--based assay for the diagnosis of bacterial vaginosis}.''} \emph{Diagn. Microbiol. Infect. Dis.} 92 (3): 173--78. \url{https://doi.org/10.1016/j.diagmicrobio.2018.05.022}.

\leavevmode\vadjust pre{\hypertarget{ref-contijoch2019gutm}{}}%
Contijoch, Eduardo J, Graham J Britton, Chao Yang, Ilaria Mogno, Zhihua Li, Ruby Ng, Sean R Llewellyn, et al. 2019. {``{Gut microbiota density influences host physiology and is shaped by host and microbial factors}.''} \emph{Elife} 8 (January). \url{https://doi.org/10.7554/eLife.40553}.

\leavevmode\vadjust pre{\hypertarget{ref-david2014host}{}}%
David, Lawrence A, Arne C Materna, Jonathan Friedman, Maria I Campos-Baptista, Matthew C Blackburn, Allison Perrotta, Susan E Erdman, and Eric J Alm. 2014. {``{Host lifestyle affects human microbiota on daily timescales}.''} \emph{Genome Biol.} 15 (7): R89. \url{https://doi.org/10.1186/gb-2014-15-7-r89}.

\leavevmode\vadjust pre{\hypertarget{ref-fettweis2019thev}{}}%
Fettweis, Jennifer M., Myrna G. Serrano, Jamie Paul Brooks, David J. Edwards, Philippe H. Girerd, Hardik I. Parikh, Bernice Huang, et al. 2019. {``{The vaginal microbiome and preterm birth}.''} \emph{Nat. Med.} 25 (6): 1012--21. \url{https://doi.org/10.1038/s41591-019-0450-2}.

\leavevmode\vadjust pre{\hypertarget{ref-finucane2014atax}{}}%
Finucane, Mariel M., Thomas J. Sharpton, Timothy J. Laurent, and Katherine S. Pollard. 2014. {``{A Taxonomic Signature of Obesity in the Microbiome? Getting to the Guts of the Matter}.''} Edited by Markus M. Heimesaat. \emph{PLoS One} 9 (1): e84689. \url{https://doi.org/10.1371/journal.pone.0084689}.

\leavevmode\vadjust pre{\hypertarget{ref-galazzo2020howt}{}}%
Galazzo, Gianluca, Niels van Best, Birke J. Benedikter, Kevin Janssen, Liene Bervoets, Christel Driessen, Melissa Oomen, et al. 2020. {``{How to Count Our Microbes? The Effect of Different Quantitative Microbiome Profiling Approaches}.''} \emph{Front. Cell. Infect. Microbiol.} 10 (August). \url{https://doi.org/10.3389/fcimb.2020.00403}.

\leavevmode\vadjust pre{\hypertarget{ref-gelman2020regr}{}}%
Gelman, Andrew, Jennifer Hill, and Aki Vehtari. 2020. \emph{Regression and Other Stories}. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-gill2016eval}{}}%
Gill, Christina, Janneke H. H. M. van de Wijgert, Frances Blow, and Alistair C. Darby. 2016. {``{Evaluation of Lysis Methods for the Extraction of Bacterial DNA for Analysis of the Vaginal Microbiota}.''} Edited by Peter E. Larsen. \emph{PLoS One} 11 (9): e0163148. \url{https://doi.org/10.1371/journal.pone.0163148}.

\leavevmode\vadjust pre{\hypertarget{ref-gloor2017micr}{}}%
Gloor, Gregory B., Jean M. Macklaim, Vera Pawlowsky-Glahn, and Juan J. Egozcue. 2017. {``{Microbiome Datasets Are Compositional: And This Is Not Optional}.''} \emph{Front. Microbiol.} 8 (November): 2224. \url{https://doi.org/10.3389/fmicb.2017.02224}.

\leavevmode\vadjust pre{\hypertarget{ref-rstanarm}{}}%
Goodrich, Ben, Jonah Gabry, Imad Ali, and Sam Brilleman. 2020. {``Rstanarm: {Bayesian} Applied Regression Modeling via {Stan}.''} \url{https://mc-stan.org/rstanarm}.

\leavevmode\vadjust pre{\hypertarget{ref-graspeuntner2018sele}{}}%
Graspeuntner, Simon, Nathalie Loeper, Sven Künzel, John F. Baines, and Jan Rupp. 2018. {``{Selection of validated hypervariable regions is crucial in 16S-based microbiota studies of the female genital tract}.''} \emph{Sci. Rep.} 8 (1): 9678. \url{https://doi.org/10.1038/s41598-018-27757-8}.

\leavevmode\vadjust pre{\hypertarget{ref-greenland2005mult}{}}%
Greenland, Sander, John Copas, David R. Jones, David Spiegelhalter, Kenneth Rice, Ben Armstrong, Stephen Senn, et al. 2005. {``{Multiple-bias modelling for analysis of observational data}.''} \emph{J. R. Stat. Soc. Ser. A Stat. Soc.} 168 (2): 267--306. \url{https://doi.org/10.1111/j.1467-985X.2004.00349.x}.

\leavevmode\vadjust pre{\hypertarget{ref-harrison2021theq}{}}%
Harrison, Joshua G., W. John Calder, Bryan Shuman, and C. Alex Buerkle. 2021. {``{The quest for absolute abundance: The use of internal standards for DNA‐based community ecology}.''} \emph{Mol. Ecol. Resour.} 21 (1): 30--43. \url{https://doi.org/10.1111/1755-0998.13247}.

\leavevmode\vadjust pre{\hypertarget{ref-hevroni2020seas}{}}%
Hevroni, Gur, José Flores-Uribe, Oded Béjà, and Alon Philosof. 2020. {``{Seasonal and diel patterns of abundance and activity of viruses in the Red Sea}.''} \emph{Proc. Natl. Acad. Sci.}, November, 202010783. \url{https://doi.org/10.1073/pnas.2010783117}.

\leavevmode\vadjust pre{\hypertarget{ref-holmes2018mode}{}}%
Holmes, Susan, and Wolfgang Huber. 2018. \emph{Modern Statistics for Modern Biology}. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-huttenhower2012stru}{}}%
Huttenhower, Curtis, Dirk Gevers, Rob Knight, Sahar Abubucker, Jonathan H. Badger, Asif T. Chinwalla, Heather H. Creasy, et al. 2012. {``{Structure, function and diversity of the healthy human microbiome}.''} \emph{Nature} 486 (7402): 207--14. \url{https://doi.org/10.1038/nature11234}.

\leavevmode\vadjust pre{\hypertarget{ref-ji2019quan}{}}%
Ji, Brian W., Ravi U. Sheth, Purushottam D. Dixit, Yiming Huang, Andrew Kaufman, Harris H. Wang, and Dennis Vitkup. 2019. {``{Quantifying spatiotemporal variability and noise in absolute microbiota abundances using replicate sampling}.''} \emph{Nat. Methods} 16 (8): 731--36. \url{https://doi.org/10.1038/s41592-019-0467-y}.

\leavevmode\vadjust pre{\hypertarget{ref-jian2021comm}{}}%
Jian, Ching, Anne Salonen, and Katri Korpela. 2021. {``{Commentary: How to Count Our Microbes? The Effect of Different Quantitative Microbiome Profiling Approaches}.''} \emph{Front. Cell. Infect. Microbiol.} 11 (March): 10--12. \url{https://doi.org/10.3389/fcimb.2021.627910}.

\leavevmode\vadjust pre{\hypertarget{ref-kumar2018anal}{}}%
Kumar, M. Senthil, Eric V. Slud, Kwame Okrah, Stephanie C. Hicks, Sridhar Hannenhalli, and Héctor Corrada Bravo. 2018. {``{Analysis and correction of compositional bias in sparse sequencing count data}.''} \emph{BMC Genomics} 19 (1): 799. \url{https://doi.org/10.1186/s12864-018-5160-5}.

\leavevmode\vadjust pre{\hypertarget{ref-leek2010tack}{}}%
Leek, Jeffrey T., Robert B. Scharpf, Héctor Corrada Bravo, David Simcha, Benjamin Langmead, W. Evan Johnson, Donald Geman, Keith Baggerly, and Rafael A. Irizarry. 2010. {``{Tackling the widespread and critical impact of batch effects in high-throughput data}.''} \emph{Nat. Rev. Genet.} 11 (10): 733--39. \url{https://doi.org/10.1038/nrg2825}.

\leavevmode\vadjust pre{\hypertarget{ref-leopold2020host}{}}%
Leopold, Devin R, and Posy E Busby. 2020. {``{Host Genotype and Colonist Arrival Order Jointly Govern Plant Microbiome Composition and Function}.''} \emph{Curr. Biol.} 30 (16): 3260--3266.e5. \url{https://doi.org/10.1016/j.cub.2020.06.011}.

\leavevmode\vadjust pre{\hypertarget{ref-lloyd2020evid}{}}%
Lloyd, Karen G., Jordan T. Bird, Joy Buongiorno, Emily Deas, Richard Kevorkian, Talor Noordhoek, Jacob Rosalsky, and Taylor Roy. 2020. {``{Evidence for a Growth Zone for Deep-Subsurface Microbial Clades in Near-Surface Anoxic Sediments}.''} \emph{Appl. Environ. Microbiol.} 86 (19): 1--15. \url{https://doi.org/10.1128/AEM.00877-20}.

\leavevmode\vadjust pre{\hypertarget{ref-lofgren2019geno}{}}%
Lofgren, Lotus A., Jessie K. Uehling, Sara Branco, Thomas D. Bruns, Francis Martin, and Peter G. Kennedy. 2019. {``{Genome‐based estimates of fungal rDNA copy number variation across phylogenetic scales and ecological lifestyles}.''} \emph{Mol. Ecol.} 28 (4): 721--30. \url{https://doi.org/10.1111/mec.14995}.

\leavevmode\vadjust pre{\hypertarget{ref-lozupone2013meta}{}}%
Lozupone, Catherine A, Jesse Stombaugh, Antonio Gonzalez, Gail Ackermann, Janet K Jansson, Jeffrey I Gordon, Doug Wendel, Yoshiki Va, and Rob Knight. 2013. {``{Meta-analyses of studies of the human microbiota}.''} \emph{Genome Res.}, 1704--14. \url{https://doi.org/10.1101/gr.151803.112}.

\leavevmode\vadjust pre{\hypertarget{ref-mandal2015anal}{}}%
Mandal, Siddhartha, Will Van Treuren, Richard A. White, Merete Eggesbø, Rob Knight, and Shyamal D. Peddada. 2015. {``{Analysis of composition of microbiomes: a novel method for studying microbial composition}.''} \emph{Microb. Ecol. Heal. Dis.} 26 (1): 27663. \url{https://doi.org/10.3402/mehd.v26.27663}.

\leavevmode\vadjust pre{\hypertarget{ref-mclaren2019cons}{}}%
McLaren, Michael R, Amy D Willis, and Benjamin J Callahan. 2019. {``{Consistent and correctable bias in metagenomic sequencing experiments}.''} \emph{Elife} 8 (September): 46923. \url{https://doi.org/10.7554/eLife.46923}.

\leavevmode\vadjust pre{\hypertarget{ref-nixon2022asta}{}}%
Nixon, Michelle Pistner, Jeffrey Letourneau, Lawrence David, Sayan Mukherjee, and Justin D. Silverman. 2022. {``{A Statistical Analysis of Compositional Surveys}.''} \emph{arXiv}, January. \url{http://arxiv.org/abs/2201.03616}.

\leavevmode\vadjust pre{\hypertarget{ref-props2017abso}{}}%
Props, Ruben, Frederiek-Maarten Kerckhof, Peter Rubbens, Jo De Vrieze, Emma Hernandez Sanabria, Willem Waegeman, Pieter Monsieurs, Frederik Hammes, and Nico Boon. 2017. {``{Absolute quantification of microbial taxon abundances}.''} \emph{ISME J.} 11 (2): 584--87. \url{https://doi.org/10.1038/ismej.2016.117}.

\leavevmode\vadjust pre{\hypertarget{ref-quinn2020inte}{}}%
Quinn, Thomas P., and Ionas Erb. 2020. {``{Interpretable Log Contrasts for the Classification of Health Biomarkers: a New Approach to Balance Selection}.''} Edited by Robert G. Beiko. \emph{mSystems} 5 (2): 1--11. \url{https://doi.org/10.1128/mSystems.00230-19}.

\leavevmode\vadjust pre{\hypertarget{ref-riverapinto2018bala}{}}%
Rivera-Pinto, J., J. J. Egozcue, V. Pawlowsky-Glahn, R. Paredes, M. Noguera-Julian, and M. L. Calle. 2018. {``{Balances: a New Perspective for Microbiome Analysis}.''} Edited by Catherine Lozupone. \emph{mSystems} 3 (4): e00053--18. \url{https://doi.org/10.1128/mSystems.00053-18}.

\leavevmode\vadjust pre{\hypertarget{ref-srinivasan2012bact}{}}%
Srinivasan, Sujatha, Noah G. Hoffman, Martin T. Morgan, Frederick A. Matsen, Tina L. Fiedler, Robert W. Hall, Frederick J. Ross, et al. 2012. {``{Bacterial Communities in Women with Bacterial Vaginosis: High Resolution Phylogenetic Analyses Reveal Relationships of Microbiota to Clinical Criteria}.''} Edited by Adam J. Ratner. \emph{PLoS One} 7 (6): e37818. \url{https://doi.org/10.1371/journal.pone.0037818}.

\leavevmode\vadjust pre{\hypertarget{ref-stammler2016adju}{}}%
Stämmler, Frank, Joachim Gläsner, Andreas Hiergeist, Ernst Holler, Daniela Weber, Peter J. Oefner, André Gessner, and Rainer Spang. 2016. {``{Adjusting microbiome profiles for differences in microbial load by spike-in bacteria}.''} \emph{Microbiome} 4 (1): 28. \url{https://doi.org/10.1186/s40168-016-0175-0}.

\leavevmode\vadjust pre{\hypertarget{ref-tettamantiboshier2020comp}{}}%
Tettamanti Boshier, Florencia A., Sujatha Srinivasan, Anthony Lopez, Noah G. Hoffman, Sean Proll, David N. Fredricks, and Joshua T. Schiffer. 2020. {``{Complementing 16S rRNA Gene Amplicon Sequencing with Total Bacterial Load To Infer Absolute Species Concentrations in the Vaginal Microbiome}.''} \emph{mSystems} 5 (2): 1--14. \url{https://doi.org/10.1128/mSystems.00777-19}.

\leavevmode\vadjust pre{\hypertarget{ref-vandeputte2017quan}{}}%
Vandeputte, Doris, Gunter Kathagen, Kevin D'hoe, Sara Vieira-Silva, Mireia Valles-Colomer, João Sabino, Jun Wang, et al. 2017. {``{Quantitative microbiome profiling links gut community variation to microbial load}.''} \emph{Nature} 551 (7681): 507. \url{https://doi.org/10.1038/nature24460}.

\leavevmode\vadjust pre{\hypertarget{ref-wang2019mana}{}}%
Wang, Yiwen, and Kim-Anh LêCao. 2020. {``{Managing batch effects in microbiome data}.''} \emph{Brief. Bioinform.} 21 (6): 1954--70. \url{https://doi.org/10.1093/bib/bbz105}.

\leavevmode\vadjust pre{\hypertarget{ref-wasserman2004allo}{}}%
Wasserman, Larry. 2004. \emph{{All of Statistics}}. Springer Texts in Statistics. New York, NY: Springer New York. \url{https://doi.org/10.1007/978-0-387-21736-9}.

\leavevmode\vadjust pre{\hypertarget{ref-williamson2021amul}{}}%
Williamson, Brian D, James P Hughes, and Amy D Willis. 2021. {``{A multiview model for relative and absolute microbial abundances}.''} \emph{Biometrics}, no. August 2019 (June): biom.13503. \url{https://doi.org/10.1111/biom.13503}.

\leavevmode\vadjust pre{\hypertarget{ref-yeh2018taxo}{}}%
Yeh, Yi-Chun, David M. Needham, Ella T. Sieradzki, and Jed A. Fuhrman. 2018. {``{Taxon Disappearance from Microbiome Analysis Reinforces the Value of Mock Communities as a Standard in Every Sequencing Run}.''} \emph{mSystems} 3 (3): e00023--18. \url{https://doi.org/10.1128/mSystems.00023-18}.

\leavevmode\vadjust pre{\hypertarget{ref-yuan2012eval}{}}%
Yuan, Sanqing, Dora B. Cohen, Jacques Ravel, Zaid Abdo, and Larry J. Forney. 2012. {``{Evaluation of Methods for the Extraction and Purification of DNA from the Human Microbiome}.''} Edited by Jack Anthony Gilbert. \emph{PLoS One} 7 (3): e33865. \url{https://doi.org/10.1371/journal.pone.0033865}.

\leavevmode\vadjust pre{\hypertarget{ref-zhang2017soil}{}}%
Zhang, Zhaojing, Yuanyuan Qu, Shuzhen Li, Kai Feng, Shang Wang, Weiwei Cai, Yuting Liang, et al. 2017. {``{Soil bacterial quantification approaches coupling with relative abundances reflecting the changes of taxa}.''} \emph{Sci. Rep.} 7 (1): 1--11. \url{https://doi.org/10.1038/s41598-017-05260-w}.

\end{CSLReferences}

\end{document}

### 2022-06-07 differential-abundance.Rmd
Small typo and a couple of notes

@@
   = \frac
-    {P_i(b) \cdot \cancel{B_i} / \bar B^{(b)}}
-    {P_i(a) \cdot \cancel{B_i} / \bar B^{(a)}}
+    {P_i^{(b)} \cdot \cancel{B_i} / \bar B^{(b)}}
+    {P_i^{(a)} \cdot \cancel{B_i} / \bar B^{(a)}}
   =
@@
 In this example, the mean efficiency decreases 2.6-fold from Sample 1 to Sample 2 (FD of 0.4), causing the FD in the proportion of each species to appear 2.6-fold larger than its true value.
+<!-- 0.4 is 1 / 2.6? Is there a reason why 0.4 is the canonical FD rather than 2.6 -->
+
 Though the FE in the FD is the same for each species, the biological implications of the error vary.
@@
 For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FD: 0.15, measured FD: 0.44).
+<!-- I was initially confused by calling 0.44 a decrease in magnitude as compared with 0.15 as |0.44| > |0.15|, but see now that 0.15 is a larger multiplicative factor e.g.
+so that |log(0.44)| < |log(0.15)| -->
 For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FD: 0.6, measured FD: 1.8), such that the species actually appears to increase (a sign error).
+<!-- Again the sign error is on the log scale (unsure if being more explicit about this is helpful) -->
 In contrast, the ratios between species change by the same factors in the actual and measured taxonomic profiles.

### 2022-07-09 introduction.Rmd
Revise abstract

@@
 **Abstract:**
-To assess how the species in microbial communities vary across conditions, researchers often measure the relative or absolute abundances of all species simultaneously using marker-gene and metagenomic sequencing.
-However, these measurements are taxonomically biased: Some species are measured more efficiently than others, yielding measured abundances that are higher or lower than the truth.
-Taxonomic bias is generally ignored in _differential abundance (DA) analyses_, likely due to a combination of 1) a widespread belief that protocol standardization is sufficient to yield accurate differences between samples and 2) a lack of practical solutions for addressing it.
-We use theoretical arguments and case studies to analyze the impact of taxonomic bias on DA analyses of relative and absolute abundances.
-We show that taxonomic bias does create inferential errors for a class of popular DA methods; however, whether these errors are biologically significant depends on experimental context.
-We present several approaches to mitigating the effect of bias, suited to a range of experimental questions and systems.
-If adopted, they could improve the reproducibility and interpretability of sequencing-based studies of microbial communities.
+
+Differential-abundance (DA) analyses enable microbiome researchers to assess how microbial species vary in relative or absolute abundance with specific host or environmental conditions, such as health status or pH.
+These analyses, however, typically use sequencing-based community measurements that are taxonomically biased to measure some species more efficiently than others.
+Determining the effect that taxonomic bias has on the results of a DA method is essential for achieving reliable, translatable findings.
+Yet currently there is poor understanding of this effect and a lack of methods for correcting it.
+We used a combination of mathematical theory and data analysis of real and simulated case studies to characterize and find ways to minimize the effect of taxonomic bias on DA analyses of relative and absolute abundances.
+For a popular class of DA methods---those based on species proportions---taxonomic bias can cause significant errors in DA results if the average measurement efficiency of the community is associated with the condition of interest.
+These errors can be avoided by using more robust DA methods (based on species ratios) or quantified and corrected using appropriate controls.
+If adopted, our recommendations can improve the reproducibility, interpretability, and translatability of microbiome DA studies.
 

### 2022-07-10 differential-abundance.Rmd
Revise differential-abundance section
Major revision of the subsection on regression analysis, with lesser
changes to other subsections, plus minor fixes and added references.

--- 2022-05-27 differential-abundance.Rmd
@@
 We now turn to how the errors in MGS-based abundance measurements affect DA results.
-Our focus is on _multiplicative DA_---the fold differences (FD) in the relative or absolute abundance of species between samples or between experimental conditions.
-Multiplicative DA has a direct ecological interpretation via the multiplicative processes of exponential growth and decay and forms the basis for several popular DA methods (REF DESeq2, ALDEx2(?), and corncob; a LV paper; fido).
-We also consider non-parametric rank-based methods; these methods are popular in case-control studies for which ecological interpretability is often of secondary importance to discovering stable associations between species and the condition of interest (REFs).
+Our focus is on _multiplicative DA analyses_ that estimate the fold differences (FD) in the relative or absolute abundance of species across time points or between experimental conditions.
+Multiplicative DA has a direct ecological interpretation via the multiplicative processes of exponential growth and decay.
+Many DA methods operate on additive differences in log-transformed abundances and are therefore multiplicative; these include popular methods such as DESeq2 (@love2014mode), ALDEx2 (@fernandes2014unif), corncob (@martin2020mode), and time series methods based on the Lotka-Volterra model (e.g., @stein2013ecol).
+We also consider non-parametric rank-based methods; these methods are popular in case-control studies for which ecological interpretability is often of secondary importance to discovering stable associations between species and the condition of interest.
 
@@
 <!-- Proportions -->
-In contrast, bias created a FE in the proportion of a species (Equation \@ref(eq:prop-error)) that varied inversely with the mean efficiency of the sample.
+In contrast, bias creates a FE in the proportion of a species (Equation \@ref(eq:prop-error)) that varies inversely with the mean efficiency of the sample.
 Therefore, this error does not cancel when calculating the FD in the proportion,
@@
 Though the FE in the FD is the same for each species, the biological implications of the error vary.
-In particular, there are three distinct types of error: an increase in the magnitude of the measured FD, a decrease in that magnitude, and a change in the direction of the FD.
-We refer to the latter as a _sign error_ in reference to the sign of the measured log fold difference (LFD).
+In particular, there are three distinct types of error: an increase in the magnitude, a decrease in the magnitude, and a reversal in the direction of the measured FD.
 We can see each type of error in Figure \@ref(fig:error-proportions).
 For Species 1, which increases in abundance and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FD (actual FD: 2.3, measured FD: 6.5).
-For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FD: 0.15, measured FD: 0.44).
-For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FD: 0.6, measured FD: 1.8), such that the species actually appears to increase (a sign error).
-In contrast, the ratios between species change by the same factors in the actual and measured taxonomic profiles.
-For example, the ratio of Species 2 to Species 3 changes from $1/1$ to $1/4$ in the actual profiles and from $3$ to $3/4$ in the measured profiles, a decrease by 4-fold either way.
+For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude of the measured FD (actual FD: 0.15, measured FD: 0.44).
+For Species 2, which decreases but by a smaller factor than the mean efficiency, the species actually appears to increase---a reversal in direction of the measured FD (actual FD: 0.6, measured FD: 1.8).
+In contrast, the FDs in the ratios among species are identical in the actual and measured taxonomic profiles.
+For example, the ratio of Species 2 to Species 3 shows the same 4-fold decrease in the actual profiles ($1$ to $1/4$) and in the measured profiles ($3$ to $3/4$).
 
@@
 First, consider a species' abundance measurement $\tilde A_i^{(a)}$ derived from a non-MGS measurement of total-community abundance using Equation \@ref(eq:density-prop-meas).
-Equation \@ref(eq:density-prop-error) indicates that the FE in this measurement equals the (constant) species efficiency multiplied by ${B^{\mtot (a)}}/{\bar B^{(a)}}$, the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
+Equation \@ref(eq:density-prop-error) indicates that the FE in this measurement equals the (constant) species efficiency multiplied by ${\bar B^{\mtot (a)}}/{\bar B^{(a)}}$, the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
 This ratio can vary, creating error in the measured FD between samples.
@@
 
-Most DA analyses can be understood as a regression analysis.
-For example, the simple linear regression model,
+In most cases, a DA analysis can be understood as regression of microbial abundance variables against one or more covariates describing predictive properties of the sample.
+The simplest such regression relates a microbial abundance variable $y$ to a covariate $x$ and a _residual (or unexplained) error_ $\varepsilon$ via the simple linear regression formula
 \begin{align}
   (\#eq:regression)
-  y^{(a)} = \alpha + \beta x^{(a)} + \varepsilon^{(a)},
+  y^{(a)} = \alpha + \beta x^{(a)} + \varepsilon^{(a)}.
 \end{align}
-can be used to relate a microbial abundance variable $y$ to a _covariate_ $x$ and the _residual (or unexplained) error_ $\varepsilon$.
 For example, $y$ may be the log absolute abundance of species $i$ ($\log A_i^{(a)}$) and $x$ may be a continuous variable (such as pH) or a binary variable (such as $x=1$ for treated patients and $x=0$ for untreated controls).
 The regression coefficients $\alpha$ and $\beta$ are parameters to be estimated from the data.
-A DA analysis is primarily interested in the slope coefficient $\beta$, which determines how the average or expected value of $y$ increases with $x$.
+A DA regression analysis is primarily interested in the slope coefficient $\beta$, which determines how the average or expected value of $y$ increases with $x$.
 For a binary covariate, $\beta$ equals the average difference in $y$ between samples with $x=1$ versus those with $x=0$.
-Taking the response $y$ to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient $\beta$ corresponding to the average log fold difference (LFD) in abundance between conditions or for a unit increase in $x$.
+Taking $y$ to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient $\beta$ corresponding to the average log fold difference (LFD) in abundance between conditions or for a unit increase in $x$.
 
-Appendix \@ref(appendix-regression) derives mathematical results for the effect that measurement error in the response caused by taxonomic bias has on LFDs estimated via simple linear regression.
-Here we illustrate these results with an example.
-Suppose that we wish to estimate the average LFD in the absolute abundance of every species between samples from two conditions, such as treated and untreated patients.
-Further suppose that we have measured absolute abundances by multiplying the proportions from MGS with an accurate measurement of total-community abundance.
-For each species $i$, we set $y^{(a)}$ in Equation \@ref(eq:regression) equal to $\log \tilde A_{i}^{(a)}$, and use either Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE) to estimate the coefficients $\alpha$ and $\beta$.
-From Equation \@ref(eq:density-prop-error), we know that the FE in the abundance of each species equals its efficiency relative to the mean efficiency of the sample.
+Although researchers typically use more sophisticated models, the simple linear regression model in Equation \@ref(eq:regression) provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
+Appendix \@ref(appendix-regression) describes a general framework for finding the error in the estimated slope coefficient $\beta$ caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures decribed in Section \@ref(abundance-measurement).
+In each case, a multiplicative DA analysis can be conducted by setting $y$ equal to the logarithm of the untransformed abundance measurement ($\tilde P_i$, $\tilde R_{i/j}$, or $\tilde A_i$).
+Consider an abundance measure such as the ratio between species $i$ and $j$, $\tilde R_{i/j}$.
+Taxonomic bias creates a constant FE in $\tilde R_{i/j}$ that translates to a constant additive shift in $\log \tilde R_{i/j}$ and therefore does not affect the estimate of the slope $\hat \beta$.
 
-To understand how the FE in species abundances affects the estimated LFD for a given species, it is useful to consider three possible scenarios for how the mean efficiency varies across samples.
-In the first scenario, the mean efficiency remains stable across samples, such that the FE in $\log \tilde A_{i}^{(a)}$ is approximately constant. 
-In this case, the FE induced by bias _mostly cancels_ and does not substantially affect the estimated LFD.
-In the second scenario, the mean efficiency varies substantially but is not associated with the condition, creating random FEs in the individual sample measurements.
-In this case, the LFDs between individual samples will be highly distorted, but these errors will effectively average out in the regression and the estimated LFD remains reliable.
-Although the random FEs caused by bias do not create a systematic error in the LFD estimate, they will tend to reduce the precision (i.e., increase the standard error) of the estimate.
-In the third scenario, the mean efficiency systematically varies with the condition, causing a systematic error in the estimated LFD regardless of the number of samples.
-In particular, the portion of additive variation in the log mean efficiency that is linearly associated with $x$ creates an equal but opposite additive error in the estimated LFD for each species.
-Although the absolute additive error in the LFDs is the same for all species, its biological significance varies, depending on its magnitude relative to that of the species' true LFD.
+Now consider the absolute abundance of a species $i$ obtained by normalizing its proportion to an non-MGS measurement of total community abundance, which for illustration we assume to be completely accurate.
+The measured abundance $\tilde A_i$ thus has a FE that varies inversely with the mean efficiency $\bar B$.
+If $\bar B$ is stable across samples, then the additive error in $\log \tilde A_i$ will be approximately constant and $\hat \beta$ is again unaffected---bias has no effect on the DA result.
+If, on the other hand, $\log \bar B$ tends to linearly vary with the covariate $x$, then the additive error in $\log \tilde A_i$ will linearly vary in the opposite direction and cause a systematic error in $\hat \beta$ that is equal in magnitude but opposite in sign to the slope of $\log \bar B$ with $x$.
+A third possibility is that $\bar B$ varies in an essentially random manner that is independent of the covariate $x$.
+In this scenario, the error from taxonomic bias in the measurement $\log \tilde A_{i}$ acts to increase the noise in $\hat beta$ (i.e., increase its standard error), but does not cause a systematic error.
 
-Figure \@ref(fig:regression-example) shows a simulated example in which the mean efficiency varies systematically with the condition of interest.
-Here, the average log mean efficiency is higher in the condition corresponding to $x=1$, due to an increase of the high-efficiency Species 9.
-This systematic increase in the log mean efficiency causes an artificial decrease in the estimated LFD for each species by an equal amount.
-As we saw for the FD between individual samples, the biological significance of the systematic error in the LFD estimates varies by species, despite the absolute error being constant across species.
-We see decreases in the magnitude of the LFD estimate (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending the value of the actual LFD relative to the error.
-The portion of the variation in the log mean efficiency that is uncorrelated with $x$ increases the standard errors for most species (Figure \@ref(fig:regression-example) D).
-The exception is Species 9, which is the primary driver of the mean efficiency dynamics in this example.
-Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).
+Figure \@ref(fig:regression-example) shows a simulated example of a regression analysis of the absolute abundances of 10 species across two conditions in which the mean efficiency is systematically greater in the condition $x=1$ than the in the condition $x=0$.
+The increase in the log mean efficiency from $x=0$ to $x=1$ (Figure \@ref(fig:regression-example) A and B) causes an artificial decrease in the estimated LFD between conditions for each species (Figure \@ref(fig:regression-example) C and D).
+The absolute error created by bias on the LFD estimate is the same for each species; however, its scientific impact varies.
+For the three species with large positive LFDs (Species 9, 10, and 1), bias decreases the magnitude of the LFD estimate.
+In contrast, for the three species with large negative LFDs (Species 8, 2, and 3), bias increases the magnitude of the estimate.
+For one species with a small positive LFD (Species 5), the effect of bias results in a negative LFD estimate (direction error).
+The three remaining species (Species 7, 6, and 4) have LFDs near zero, but bias causes them to have large negative LFDs.
 
@@
 
-```{r regression-example, fig.cap = '(ref:regression-example)'}
+```{r regression-example, fig.cap = '(ref:regression-example)', out.width = '90%'}
 here::here(
@@
 (ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.** This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by $x=0$ and $x=1$.
-In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFC, which drives the positive association of the log mean efficiency with the condition (shown in Panels A and B).
-This positive LFC in the log mean efficiency induces a systematic negative shift in the estimated LFCs of all species (Panels C and D).
-Panel D shows the mean LFC (points) with 95% confidence intervals (CIs), for each species estimated from either the actual or the measured densities.
-The error (difference in LFC estimates on measured and actual) equals the negative LFC of the mean efficiency (shown in Panel B).
+In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFD, which drives a positive association of the log mean efficiency with the condition (shown in Panels A and B).
+This positive LFD in the log mean efficiency induces a systematic negative shift in the estimated LFDs of all species (Panels C and D).
+Panel D shows the mean LFD (points) with 95% confidence intervals (CIs), for each species estimated from either the actual or measured abundances.
+The error (difference in LFD estimates on measured and actual) equals the negative LFD of the mean efficiency (shown in Panel B).
 
@@
 
-Rank-based or 'nonparametric' DA methods work by first applying a cross-sample rank-transformation to the abundance variable $y$: The sample in which $y$ has the smallest value receives a rank of 1, the sample with the second smallest value receives a rank of 2, etc.
-These methods then analyze how the average rank varies across conditions.
+Rank-based DA methods work by first applying a cross-sample rank-transformation to the abundance variable $y$: The sample in which $y$ has the smallest value receives a rank of 1, the sample with the second smallest value receives a rank of 2, etc.
+These methods then analyze how the average rank of $y$ varies across sample conditions (@conover2012ther).
 Rank-based methods commonly used in microbiome DA analysis include the Wilcoxon-Mann-Whitney and Kruskal-Wallis tests for a discrete covariate (e.g.
 case vs control) and Spearman's rank correlation for a continuous covariate (e.g.
-pH).
-They have been applied to species proportions (REFs), ratios (REFs), and absolute abundances (REFs).
+pH); these tests have collectively been applied to species proportions (@callahan2017repl,@fettweis2019thev), ratios (@fernandes2014unif), and absolute abundances (@vieirasilva2019quan).
 
-Multiplying a variable by a constant factor does not change its cross-sample ranks; thus rank-based DA methods are unaffected by constant FE in the abundance variable $y$.
-We therefore conclude that consistent taxonomic bias does not affect rank-based DA analyses of species ratios or absolute abundances derived from a reference species.
-In contrast, variable FE in $y$ (driven by variation in the mean efficiency) can affect the cross-sample ranks; if the FE is associated with the covariate $x$, it can cause systematic error in the DA result.
-Thus consistent bias may impact rank-based analyses of species proportions and absolute abundances derived from a total-community abundance measurement.
-Future work should consider whether rank-based DA methods are significantly more robust than multiplicative DA methods to variation in the mean efficiency.
+Multiplying a variable by a constant factor does not change its cross-sample ranks, making rank-based methods unaffected by a constant FE in the abundance variable $y$.
+Thus in our MGS measurement model, taxonomic bias does not affect rank-based DA analyses of species ratios or absolute abundances derived from a reference species.
+But for species proportions and absolute abundances derived from a total-community abundance measurement, variation in the mean efficiency $\bar B$ can affect the cross-sample ranks and cause error in DA results if $\bar B$ is associated with the covariate.
+Further work is needed to determine whether rank-based DA methods are significantly more robust to variation in $\bar B$ than multiplicative DA methods.

### 2022-07-10 case-studies.Rmd
Change 'sign error' to 'direction error'

@@
 vaginalis_ is most abundant.
-The vaginal microbiome sometimes shifted between _Lactobacillus_-dominance and _Gardnerella_-dominance between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)), which caused magnitude and sign errors in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).
+The vaginal microbiome sometimes shifted between _Lactobacillus_-dominance and _Gardnerella_-dominance between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)), which caused magnitude and direction errors in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).
 

### 2022-07-10 supplemental-figures.Rmd
Change 'sign error' to 'direction error'

@@
 This transition drives a sharp increase in the mean efficiency, which significantly distorts the fold changes in the observed (uncalibrated) microbiome measurements for species with less dramatic fold changes.
-Two exemplar species are shown to illustrate the magnitude (_Ureaplasma cluster 23_) and sign (_Megasphaera OTU70 type1_) errors that can arise in this situation.
+Two exemplar species are shown to illustrate the magnitude (_Ureaplasma cluster 23_) and direction (_Megasphaera OTU70 type1_) errors that can arise in this situation.
 

### 2022-07-10 abundance-measurement.Rmd
Minor fixes
Fixed typos noted by @athowes in #20.

@@
 Second, we ignore the possibility that reads are misassigned to the wrong species or the wrong sample.
-Third, we suppose that taxonomic bias acts consistently across samples at the level of species---that is, a given species is always measured more efficiently than another to the same degree, regardless of variation in changes in sub-species-level composition or the sample matrix.
+Third, we suppose that taxonomic bias acts consistently across samples at the species level---that is, a given species is always measured more efficiently than another to the same degree.
 Finally, unless otherwise stated, we treat sequencing measurements as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
@@
 This known or measured abundance $\tilde A_{r}^{(a)}$ can be used in conjunction with the MGS read counts to obtain absolute abundances for all species.
-In the absence of taxonomic bias, the ratio of a species' absolute abundance to its MGS read count is the same for all species in a given sample (Equation \@ref{eq:mgs-model}).
+In the absence of taxonomic bias, the ratio of a species' absolute abundance to its MGS read count is the same for all species in a given sample (Equation \@ref(eq:mgs-model)).
 Hence the known ratio for the reference species can serve as conversion factor for obtaining the absolute abundance of a species $i$ from its read count,
@@
   (\#eq:density-ratio-meas)
-  \tilde A_i^{(a)} &= M_i^{(a)} \cdot \frac{\tilde A_{r}^{(a)}}{M_i^{(a)}}.
+  \tilde A_i^{(a)} &= M_i^{(a)} \cdot \frac{\tilde A_r^{(a)}}{M_r^{(a)}}.
 \end{align}

### 2022-07-10 notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example.Rmd
Update regression example figure
Updated the code and figure to use the new terminology of 'abundance'
instead of 'density' and 'fold difference' instead of 'fold change'

@@
 
-## Actual densities
+## Actual abundances
 
@@
 ```
-log density = a_0 + a_1 * X + epsilon
+log abundance = a_0 + a_1 * X + epsilon
 ```
@@
   mutate(across(x, factor))
-actual_dens <- phyloseq(
+actual_abun <- phyloseq(
   otu_table(y, taxa_are_rows = TRUE) %>% transform_sample_counts(exp),
@@
 ```{r}
-actual_dens %>%
+actual_abun %>%
   as_tibble %>%
@@
 
-## Taxonomic bias and measured densities
+## Taxonomic bias and measured abundances
 
@@
 
-The measured (i.e. estimated) proportions and densities are given by perturbing the actual densties by the efficiencies, and normalizing to proportions or to the original (correct) total.
+The measured (i.e. estimated) proportions and abundances are given by perturbing the actual abundances by the efficiencies, and normalizing to proportions or to the original (correct) total.
 
 ```{r}
-measured_prop <- actual_dens %>% perturb(efficiency)
-measured_dens <- actual_dens %>% perturb(efficiency, norm = "keep")
+measured_prop <- actual_abun %>% perturb(efficiency)
+measured_abun <- actual_abun %>% perturb(efficiency, norm = "keep")
 ```
@@
 ```{r mean-efficiency}
-mean_eff <- actual_dens %>%
+mean_eff <- actual_abun %>%
   transform_sample_counts(close_elts) %>%
@@
 
-## Estimated and actual densities
+## Estimated and actual abundances
 
@@
 
-dens <- bind_rows(
-  Actual = actual_dens %>% as_tibble,
-  Measured = measured_dens %>% as_tibble,
+abun <- bind_rows(
+  Actual = actual_abun %>% as_tibble,
+  Measured = measured_abun %>% as_tibble,
   .id = 'type'
 ) %>%
-  rename(density = .abundance) %>%
+  rename(abundance = .abundance) %>%
   mutate(
-    log2_density = log2(density),
+    log2_abundance = log2(abundance),
     across(.otu, factor, levels = lvls)
@@
 ```{r}
-p_species_all <- dens %>%
-  ggplot(aes(x, log2_density, color = type)) +
-  labs(y = "Log density ") +
+p_species_all <- abun %>%
+  ggplot(aes(x, log2_abundance, color = type)) +
+  labs(y = "Log abundance ") +
   facet_grid(.otu~type, scales = "fixed") +
@@
 ```{r}
-lm_fits <- dens %>%
+lm_fits <- abun %>%
   group_by(.otu, type) %>%
@@
   mutate(
-    fit = map(data, ~lm(log2_density ~ 1 + x, .))
+    fit = map(data, ~lm(log2_abundance ~ 1 + x, .))
   )
@@
   ggplot(aes(y = .otu, x = estimate, color = type)) +
-  labs(x = "Log fold change", y = NULL) +
+  labs(x = "Log fold difference", y = NULL) +
   geom_vline(xintercept = 0, color = "grey") +
@@
   )) +
-  labs(x = "Log fold change", y = NULL) +
+  labs(x = "Log fold difference", y = NULL) +
   geom_vline(xintercept = 0, color = "grey") +
@@
   ggplot(aes(y = type, x = estimate, fill = type)) +
-  labs(x = "Log fold change", y = "Type") +
+  labs(x = "Log fold difference", y = "Type") +
   geom_vline(xintercept = 0, color = "grey") +
@@
 rng <- p_species_focal$data %>%
-  pull(log2_density) %>%
+  pull(log2_abundance) %>%
   {max(.) - min(.)}
@@
 (p_mean_eff1 + ggtitle("Mean efficiency")) +
-  (p_coef_ci_mean_eff + ggtitle("LFC in mean efficiency")) + 
-  (p_species_focal + ggtitle("Actual and measured densities\nof select species")) + 
-  (p_coef_ci1 + ggtitle("Estimated LFCs of all species") +
+  (p_coef_ci_mean_eff + ggtitle("LFD in mean efficiency")) + 
+  (p_species_focal + ggtitle("Actual and measured\nabundances of select species")) + 
+  (p_coef_ci1 + ggtitle("Estimated LFDs of all species") +
     theme(legend.box.margin = margin(b = -15))
@@
 
-Can see that the standard error is generally increased in the measured density estimates except for Species 9, whose residual variation we expect to be correlated with the log mean efficiency and hence negatively correlated with the measurement error.
+Can see that the standard error is generally increased in the measured abundance estimates except for Species 9, whose residual variation we expect to be correlated with the log mean efficiency and hence negatively correlated with the measurement error.
 
@@
   ungroup %>%
-  left_join(aug_mean_eff, by = c(".sample", "x"), suffix = c('.dens', '.mean_eff'))
+  left_join(aug_mean_eff, by = c(".sample", "x"), suffix = c('.abun', '.mean_eff'))
 ```
@@
   with_groups(.otu, summarize,
-    cov = cov(.resid.dens, .resid.mean_eff)
+    cov = cov(.resid.abun, .resid.mean_eff)
   ) %>%

### 2022-07-10 introduction.Rmd
Revise title and abstract
Revised based on conversation and edits from JN.

@@
 Differential-abundance (DA) analyses enable microbiome researchers to assess how microbial species vary in relative or absolute abundance with specific host or environmental conditions, such as health status or pH.
-These analyses, however, typically use sequencing-based community measurements that are taxonomically biased to measure some species more efficiently than others.
-Determining the effect that taxonomic bias has on the results of a DA method is essential for achieving reliable, translatable findings.
-Yet currently there is poor understanding of this effect and a lack of methods for correcting it.
-We used a combination of mathematical theory and data analysis of real and simulated case studies to characterize and find ways to minimize the effect of taxonomic bias on DA analyses of relative and absolute abundances.
-For a popular class of DA methods---those based on species proportions---taxonomic bias can cause significant errors in DA results if the average measurement efficiency of the community is associated with the condition of interest.
+These analyses typically use sequencing-based community measurements that are taxonomically biased to measure some species more efficiently than others.
+Understanding the effects that taxonomic bias has on the results of a DA analysis is essential for achieving reliable and translatable findings; yet currently, these effects are unknown.
+Here, we characterized these effects for DA analyses of both relative and absolute abundances, using a combination of mathematical theory and data analysis of real and simulated case studies.
+We found that, for analyses based on species proportions, taxonomic bias can cause significant errors in DA results if the average measurement efficiency of the community is associated with the condition of interest.
 These errors can be avoided by using more robust DA methods (based on species ratios) or quantified and corrected using appropriate controls.
-If adopted, our recommendations can improve the reproducibility, interpretability, and translatability of microbiome DA studies.
+Wide adoption of our recommendations can improve the reproducibility, interpretability, and translatability of microbiome DA studies.
 

### 2022-07-17 supplemental-figures.Rmd
Remove bias-sensitivity SI figure
Removed figure no longer referenced by the text.

@@
 <!-- Solutions -->
-
-```{r sensitivity-example, fig.cap = '(ref:cap-sensitivity-example)'}
-fs::path(
-  "notebook/_posts/2021-10-18-evaluate-robustness-example/",
-  "evaluate-robustness-example_files/figure-html5/summary_plot-1.svg"
-) %>%
-  include_svg
-```
-
-(ref:cap-sensitivity-example) **A bias-sensivity analysis can be performed to examine how sensitive the results of a DA analysis are to assumptions about taxonomic bias in community measurements.** The figure shows the results of a bias-sensitivity analysis used to study the effect of bias on the association of _Gardnerella vaginalis_ and preterm birth that was investigated by @callahan2017repl.
-100 random efficiency vectors were drawn at 6 different bias strengths (quantified by the variance in log efficiency, $\sigma_{e}^{2}$).
-Each efficiency vector was used to calibrate the MGS profiles and perform a DA association test of _G.
-vaginalis_ versus the host's preterm birth outcome; regression coefficients $\hat \beta$ indicate the increase of average logit proportion of  _G.
-vaginalis_ in women who experienced preterm birth.

### 2022-07-17 case-studies.Rmd
Edit case-studies section
Minor edits to prepare for sending to coauthors, including
- Add links to SI analyses for the case studies
- Simplify the background paragraph for the gut microbiome case study

@@
 
-We repeated the two DA analyses of @leopold2020host with and without calibration and found that the results did not meaningfully differ.
-To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \@ref(fig:leopold2020host-variation)).
+We repeated the two DA analyses of @leopold2020host with and without calibration and found that the results did not meaningfully differ
+([SI Analysis of host genetics and arrival order](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/),
+[SI Analysis of priority effects](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/)).
+To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities 
+([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-leopold2020host-case-study/), 
+SI Figure \@ref(fig:leopold2020host-variation)).
 Despite the 13-fold variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62 and a geometric standard deviation of 1.05.
@@
 
-We performed an additional DA analysis on the data from @leopold2020host to investigate whether any commensals increased in absolute concentration in response to infection.
+We performed an additional DA analysis on the data from @leopold2020host to investigate whether any commensals increased in absolute concentration in response to infection ([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-leopold2020host-case-study/#change-in-commensals-due-to-infection)).
 The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
@@
 A second _Lactobacillus_ species, _L. crispatus_, had an efficiency that was approximately 2-fold less than _L. iners_ and 15-fold greater than _G. vaginalis_.
-These species, along with the unculturable _Lachnospiraceae BVAB1_, are most frequently the most abundant species in a sample and commonly reach high proportions of over 70% (REF SI analysis).
+These species, along with the unculturable _Lachnospiraceae BVAB1_, are most frequently the most abundant species in a sample and commonly reach high proportions of over 70% 
+([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2021-11-01-momspi-summary/)).
 Therefore, shifts between these dominant species could drive large changes in the sample mean efficiency, which might in turn distort DA analyses.
@@
 We first examined variation in the mean efficiency across samples under the assumption that bias in the MOMS-PI study was accurately represented by the @brooks2015thet control measurements.
-Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample (REF SI Analysis).
-The mean efficiency varies substantially across vaginal samples (Figure 4.1), with samples in which a _Lactobacillus_ species is most abundant typically having a mean efficiency that is 3-fold to 20-fold greater than samples in which _G.
+Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample
+([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2021-11-01-momspi-summary/)).
+The mean efficiency varies substantially across vaginal samples (Figure \@ref(fig:momspi-mean-efficiency-dist)), with samples in which a _Lactobacillus_ species is most abundant typically having a mean efficiency that is 3-fold to 20-fold greater than samples in which _G.
 vaginalis_ is most abundant.
@@
 These observations suggest that a proportion-based DA analysis of a health condition that is associated with _Lactobacillus_ and/or _Gardnerella_ would be prone to spurious results.
-To demonstrate this possibility, we performed a DA analysis of the MOMS-PI samples with respect to a simulated health condition.
+To demonstrate this possibility, we performed a DA analysis of the MOMS-PI samples with respect to a simulated health condition
+([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2021-12-15-momspi-regression-diversity)).
 Bacterial vaginosis (BV) has been repeatedly found to be associated with a dramatically reduced proportion of _Lactobacillus spp._, an increased proportion of _G.
@@
 
-We wondered whether DA in the human gut microbiome might be less affected by bias than the vaginal microbiome, due to its greater species diversity.
-Recall that the mean efficiency is an abundance-weighted average over species in a sample; thus we might expect it to become more stable as the within-sample (alpha) diversity increases.
-In fact, one can show that for communities that are randomly assembled from a large species pool, the additive variance in the mean efficiency is inversely proportional to order-2 diversity $^2D$ (also known as the Inverse Simpson diversity index) (Appendix \@ref(diversity-and-mean-efficiency)).
-But real communities are assembled through complex processes that create positive and negative associations among species (REF).
-Thus there is unlikely to be a simple 1-1 relationship between alpha diversity and variation in the mean efficiency.
-Moreover, even if the mean efficiency does vary less across the more diverse gut samples, if the proportion of each species also varies less, then the importance of bias for DA analyses need not diminish.
+We wondered whether DA studies of the human gut microbiome might be less affected by bias than studies of the vaginal microbiome, due to the greater alpha diversity observed within gut microbiome samples.
+Recall that the mean efficiency is an abundance-weighted average over the species in a sample.
+Thus, all else equal, we expect the mean efficiency to vary less across samples from ecosystems where samples have higher alpha (within-sample) diversity, as is the case for stool samples relative to vaginal samples (@huttenhower2012stru).
+On the other hand, the proportion of individual species may also vary less less across samples when alpha diversity is higher, in which case the effect of bias on DA results need not diminish.
 
@@
 
-To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project (@huttenhower2012stru, SI Methods \@ref(methods-gut)).
+To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project 
+(@huttenhower2012stru, [SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-30-hmp-stool-vagina-comparison/)).
 On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \@ref(fig:gut)A; GM $\md$ GSD of `r hmp_div_stats['stool', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['stool', 'gsd'] %>% round(1)` in stool samples and `r hmp_div_stats['vagina', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['vagina', 'gsd'] %>% round(1)` in vaginal samples).
@@
 Importantly, @lloyd2020evid also used qPCR to measure the absolute abundance of two taxa from these same samples.
-By comparing qPCR to FRAxC growth rates for these taxa, we can estimate the systematic error in FRAxC growth rates.
+By comparing qPCR to FRAxC growth rates for these taxa, we can estimate the systematic error in FRAxC growth rates ([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-lloyd2020evi-case-study/))
 Because the systematic error that bias causes in the regression slope is the same for each taxon (Section \@ref(differential-abundance)), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.

### 2022-07-17 conclusion.Rmd
Revise intro and conclusion
Revised primarily to further emphasize
- the condition for when bias does (and does not) have a big effect on
  proportion-based DA results is a key result
- our results are consistent with bias not being a big problem for past
  proportion-based results, and our results can be used to help figure
  this out in the future

@@
 
-It is commonly thought that analyses of the differences between samples are robust to the taxonomic bias inherent in MGS measurement as long as all samples have been subjected to the same MGS workflow.
-But our results show that consistent taxonomic bias is still capable of creating scientific errors in common forms of DA analysis.
-In particular, we showed mathematically that bias affects DA analysis methods based on proportions or in absolute abundances derived from them (using total-abundance normalization), due to variation in the mean efficiency across samples of varying taxonomic composition.
-<!-- This is assuming consistent bias at the taxonomic level of analysis; things could be worse if we're analyzing higher taxonomic units than at which bias is conserved. -->
-<!-- Perhaps state that we showed this theoretically, in mock communities, and in real case studies -->
-The error, however, varies by experimental context and may often be negligible in a practical sense.
-<!-- It is negligible if the mean efficiency is stable across samples, and may be unimportant if it is not correlated with the focal covariate in a regression analysis. -->
-Moreover, it can be quantified and corrected through the use of reference-taxon measurements and/or community calibration controls.
-In addition, bias-aware sensitivity and meta-analyses make it possible to rigorously account for bias even in the absence of such controls.
-Applications of these methods will gradually improve our understanding of the experimental contexts in which variation in the mean efficiency is likely to be problematic and should be accounted for.
+It is commonly assumed that differential abundance (DA) analysis is robust to taxonomic bias, so long as all samples have been subjected to the same MGS workflow.
+In contrast, our results show that consistent taxonomic bias can distort the results of DA methods that are based on species proportions.
+This distortion occurs because the fold error in a species' proportion depends on the sample mean efficiency, which can vary across samples and conditions in accordance with the overall community composition.
+This problem continues to apply for analyses of absolute abundance that are based on multiplying the proportions from MGS by the total abundance of the community.
+In contrast, DA methods based on species ratios are invariant to consistent bias, since the error in the ratio between species is independent of community composition.
+In Section \@ref(solutions), we describe how ratio-based methods can thus provide a more robust approach to analyzing relative and absolute abundances;
+in addition, calibration with community controls, bias-sensitivity analysis, and bias-aware meta-analysis provide additional ways to correct or mitigate the effect of bias.
 
-We also showed that other DA methods are more robust---perhaps even entirely invariant---to consistent taxonomic bias.
-In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
-In addition, careful pairing of total-abundance and MGS measurement methods can counter the variation in mean efficiency of the MGS measurement, leading to more consistent fold errors across samples.
+Proportion-based DA methods encompass many of the most popular methods for analyzing relative and absolute abundances.
+Importantly, however, we found that the degree and significance of the error due to bias depends on experimental context and may often be negligible.
+In particular, if the mean efficiency is approximately constant across samples, then bias has a negligible effect on multiplicative and rank-based cross-sample comparisons, and DA results will be unaffected.
+If the mean efficiency varies but is unassociated with the condition of interest, then bias merely serves to increase noise and does not create systematic errors in DA results.
+It is only when the mean efficiency is associated with the condition being analyzed that large systematic errors can occur.
+Our case studies suggest that this problematic scenario can occur but may well be the exception rather than the rule.
+Thus our results are still consistent with taxonomic bias having had a limited impact on extant DA results.
+Systematic investigation of how the mean efficiency affects DA results across a wide range of studies may increase our confidence in previous DA results and/or alert us to the conditions in which they are most suspect.
 
-Important open questions and future research directions remain.
-<!-- Taxonomic aggregation and non-FC analyses are left out because I'm assuming they will be addressed in the manuscript -->
-Key open questions include 1) determining the extent to which bias is consistent across samples for different taxonomic levels, MGS methods, and sample types; 2) assessing the validity of post-extraction measurements as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how different underlying community dynamics, and in particular the source of zero counts in MGS measurements, affect bias sensitivity.
+Important open questions include 1) determining the extent to which bias is consistent across samples for different taxonomic levels, MGS methods, and sample types; 2) assessing the validity of post-extraction measurements as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how different underlying community dynamics, and in particular the source of zero counts in MGS measurements, affect bias sensitivity.
 In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, more sophisticated statistical tools are needed to properly account for both taxonomic bias and random variation in the underlying MGS and supplemental (e.g.

### 2022-07-17 introduction.Rmd
Revise intro and conclusion
Revised primarily to further emphasize
- the condition for when bias does (and does not) have a big effect on
  proportion-based DA results is a key result
- our results are consistent with bias not being a big problem for past
  proportion-based results, and our results can be used to help figure
  this out in the future

@@
 Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
-We show that, in contrast to received wisdom, taxonomic bias can affect the results obtained from many standard DA methods; however, other methods are robust to bias, provided that it is truly consistent across samples.
-We then use several case studies to explore the real-world conditions in which bias is likely to cause serious scientific errors.
+We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods (for relative and absolute abundance) that are based on species proportions.
+We describe the theoretical conditions when this effect is negligible and when it may cause serious scientific errors, and explore this effect in case studies based on real microbiome experiments.
+We further demonstrate that another set of DA methods, based on the ratios among species, are robust to bias.
 Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.

### 2022-07-17 index.Rmd
Revise title
Revise title and abstract
Revised based on conversation and edits from JN.
Update index.Rmd
- update MM's affiliation
- fix link to notebook site

@@
 ---
-title: "Differential abundance analysis of taxonomically-biased microbiome measurements"
+title: "Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis"
 author:
-  - Michael R. McLaren^[North Carolina State University; send correspondence to m.mclaren42@gmail.com]
+  - Michael R. McLaren^[Massachusetts Institute of Technology; send correspondence to m.mclaren42@gmail.com]
   - Jacob T. Nearing^[Dalhousie University]
@@
 See [the Zenodo record](https://doi.org/10.5281/zenodo.4552717) for how to cite the latest version.
-Supporting data analyses can be found in the accompanying [computational research notebook] (https://mikemc.github.io/differential-abundance-theory/notebook/).
+Supporting data analyses can be found in the accompanying [computational research notebook](https://mikemc.github.io/differential-abundance-theory/notebook/).

### 2022-07-17 abundance-measurement.Rmd
Add references

@@
 Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
-Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency.
+Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency (@lloyd2013meta).
 We can easily understand the impact of taxonomic bias on total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
@@
 
-A common application of this approach involves adding a 'spike-in' (as described above) in a known (and typically constant) abundance across samples (@stammler2016adju, @ji2019quan, @harrison2021theq, REFs).
+A common application of this approach involves adding a 'spike-in' (as described above) in a known (and typically constant) abundance across samples (@stammler2016adju, @ji2019quan, @tkacz2018abso, @harrison2021theq, @rao2021mult).
 In this case, the reference abundance $\tilde A_r^{(a)}$ is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
-(In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \@ref(eq:density-ratio-meas); Appendix REF).
+(In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \@ref(eq:density-ratio-meas); Appendix TODO).
 
-Others have instead sought to determine naturally-occurring species that are thought to be constant across samples; we refer to such species as _housekeeping species_ by analogy with the housekeeping genes used for absolute-abundance conversion in gene-expression studies (REF).
-Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference (REFs).
+Others have instead sought to determine naturally-occurring species that are thought to be constant across samples; we refer to such species as _housekeeping species_ by analogy with the housekeeping genes used for absolute-abundance conversion in gene-expression studies (@silver2006sele).
+Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference (@karasov2020ther, @regalado2020comb, @wallace2021thed).
 A related approach involves computationally identifying species that are constant between pairs of samples (@david2014host) or between sample conditions (@mandal2015anal, @kumar2018anal).
@@
 The most common form of targeted measurement involves using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA.
-It is also possible to directly measure cell concentration by performing ddPCR prior to DNA extraction (REFs), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
-Appendix [REF] describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
+It is also possible to directly measure cell concentration by performing ddPCR prior to DNA extraction (@morella2018rapi), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
+Appendix TODO describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.

### 2022-07-17 appendix-old.Rmd
Move old appendix file
Create new appendix file
Add references

--- 2021-12-02 appendix.Rmd
@@
-# (APPENDIX) Appendix {-}
-
 # Model details {#model-details}
@@
 
-### Revised main text equations
-
-Let $S$ be the set of all species.
-Then
-\begin{align}
-  \text{reads}_S(a)
-    = \text{density}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
-\end{align}
-where 
-\begin{align}
-  \text{efficiency}_S(a) 
-  \equiv \frac{\sum_{j\in S}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{density}_S(a)}
-\end{align}
-Also,
-\begin{align}
-  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
-\end{align}
-and
-\begin{align}
-  \widehat{\text{prop}}_{i}(a)
-  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}.
-\end{align}
-
 ## Taxonomic bias in total-density measurement {#total-density-bias}
@@
 \begin{align}
-  (\#eq:app-density-prop-error)
+  (\#eq:density-prop-error-with-total-error)
   \widehat{\text{density}}_{i}(a) 
@@
 Constant reference species are sometimes used to measure total density of $S$ by the ratio of $S$ reads to $R$ reads.
-For example, a study of *Arabidopsis* microbiomes used the ratio of bacterial to host reads in shotgun sequencing as a proxy for total bacterial density, which they then used for total-community normalization of 16S amplicon sequencing measurements (@karasov2020ther, @regalado2019comb).
+For example, a study of *Arabidopsis* microbiomes used the ratio of bacterial to host reads in shotgun sequencing as a proxy for total bacterial density, which they then used for total-community normalization of 16S amplicon sequencing measurements (@karasov2020ther, @regalado2020comb).
 @chng2020meta similarly used the ratio of bacterial to host or diet reads in shotgun sequencing of mouse fecal samples as a proxy for total bacterial density (though they did not use this measurement for community normalization).
@@
 \begin{align}
-  (\#eq:app-density-prop-error)
+  (\#eq:density-prop-error-linearity)
   \widehat{\text{density}}_{i}(a) 
@@
 Several studies that have employed shotgun sequencing of host-associated microbiomes have use the plant or animal host for this purpose.
-A study of *Arabidopsis* microbiomes used the ratio of bacterial to host reads in shotgun sequencing as a proxy for total bacterial density, which they then used for total-community normalization of 16S amplicon sequencing measurements (@karasov2020ther, @regalado2019comb).
+A study of *Arabidopsis* microbiomes used the ratio of bacterial to host reads in shotgun sequencing as a proxy for total bacterial density, which they then used for total-community normalization of 16S amplicon sequencing measurements (@karasov2020ther, @regalado2020comb).
 @chng2020meta similarly used the ratio of bacterial to host reads in shotgun sequencing of mouse fecal samples as a proxy for total bacterial density (though they did not use this measurement for community normalization).
@@
 Species-specific florescent probes also make it possible to measure individual species via microscopy or flow cytometry (REFs).
+
+TODO: Argue that these methods may yield constant fold errors.

### 2022-07-17 v/9c86fbcc416ebe9d18d2e80120064bea4c346f36/_main.tex
Add references
This commit adds the rendered manuscript associated with
9c86fbcc416ebe9d18d2e80120064bea4c346f36

--- 2022-06-05 v/0d63746fac87923e630a61276e8fb3f1cfddd74c/_main.tex
@@
 \hypersetup{
-  pdftitle={The effect of taxonomic bias on differential-abundance analysis in microbiome studies},
+  pdftitle={Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis},
   pdfauthor={Michael R. McLaren; Jacob T. Nearing; Amy D. Willis; Karen G. Lloyd; Benjamin J. Callahan},
@@
 
-\title{The effect of taxonomic bias on differential-abundance analysis in microbiome studies}
+\title{Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis}
 \author{Michael R.
-McLaren\footnote{North Carolina State University; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Jacob T.
+McLaren\footnote{Massachusetts Institute of Technology; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Jacob T.
 Nearing\footnote{Dalhousie University} \and Amy D.
@@
 Callahan\footnote{North Carolina State University}}
-\date{2022-06-05}
+\date{2022-07-17}
 
@@
 
-\emph{This manuscript was rendered from commit 0d63746fac87923e630a61276e8fb3f1cfddd74c.}
+\emph{This manuscript was rendered from commit 9c86fbcc416ebe9d18d2e80120064bea4c346f36.}
 
@@
 See \href{https://doi.org/10.5281/zenodo.4552717}{the Zenodo record} for how to cite the latest version.
-Supporting data analyses can be found in the accompanying {[}computational research notebook{]} (\url{https://mikemc.github.io/differential-abundance-theory/notebook/}).
+Supporting data analyses can be found in the accompanying \href{https://mikemc.github.io/differential-abundance-theory/notebook/}{computational research notebook}.
 
@@
 \textbf{Abstract:}
-To assess how the species in microbial communities vary across conditions, researchers often measure the relative or absolute abundances of all species simultaneously using marker-gene and metagenomic sequencing.
-However, these measurements are taxonomically biased: Some species are measured more efficiently than others, yielding measured abundances that are higher or lower than the truth.
-Taxonomic bias is generally ignored in \emph{differential abundance (DA) analyses}, likely due to a combination of 1) a widespread belief that protocol standardization is sufficient to yield accurate differences between samples and 2) a lack of practical solutions for addressing it.
-We use theoretical arguments and case studies to analyze the impact of taxonomic bias on DA analyses of relative and absolute abundances.
-We show that taxonomic bias does create inferential errors for a class of popular DA methods; however, whether these errors are biologically significant depends on experimental context.
-We present several approaches to mitigating the effect of bias, suited to a range of experimental questions and systems.
-If adopted, they could improve the reproducibility and interpretability of sequencing-based studies of microbial communities.
+
+Differential-abundance (DA) analyses enable microbiome researchers to assess how microbial species vary in relative or absolute abundance with specific host or environmental conditions, such as health status or pH.
+These analyses typically use sequencing-based community measurements that are taxonomically biased to measure some species more efficiently than others.
+Understanding the effects that taxonomic bias has on the results of a DA analysis is essential for achieving reliable and translatable findings; yet currently, these effects are unknown.
+Here, we characterized these effects for DA analyses of both relative and absolute abundances, using a combination of mathematical theory and data analysis of real and simulated case studies.
+We found that, for analyses based on species proportions, taxonomic bias can cause significant errors in DA results if the average measurement efficiency of the community is associated with the condition of interest.
+These errors can be avoided by using more robust DA methods (based on species ratios) or quantified and corrected using appropriate controls.
+Wide adoption of our recommendations can improve the reproducibility, interpretability, and translatability of microbiome DA studies.
 
@@
 Second, we ignore the possibility that reads are misassigned to the wrong species or the wrong sample.
-Third, we suppose that taxonomic bias acts consistently across samples at the level of species---that is, a given species is always measured more efficiently than another to the same degree, regardless of variation in changes in sub-species-level composition or the sample matrix.
+Third, we suppose that taxonomic bias acts consistently across samples at the species level---that is, a given species is always measured more efficiently than another to the same degree.
 Finally, unless otherwise stated, we treat sequencing measurements as deterministic, ignoring the `random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.
@@
 Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
-Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency.
+Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency (Lloyd et al.
+(\protect\hyperlink{ref-lloyd2013meta}{2013})).
 We can easily understand the impact of taxonomic bias on total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
@@
 This known or measured abundance \(\tilde A_{r}^{(a)}\) can be used in conjunction with the MGS read counts to obtain absolute abundances for all species.
-In the absence of taxonomic bias, the ratio of a species' absolute abundance to its MGS read count is the same for all species in a given sample (Equation @ref\{eq:mgs-model\}).
+In the absence of taxonomic bias, the ratio of a species' absolute abundance to its MGS read count is the same for all species in a given sample (Equation \eqref{eq:mgs-model}).
 Hence the known ratio for the reference species can serve as conversion factor for obtaining the absolute abundance of a species \(i\) from its read count,
@@
   \label{eq:density-ratio-meas}
-  \tilde A_i^{(a)} &= M_i^{(a)} \cdot \frac{\tilde A_{r}^{(a)}}{M_i^{(a)}}.
+  \tilde A_i^{(a)} &= M_i^{(a)} \cdot \frac{\tilde A_r^{(a)}}{M_r^{(a)}}.
 \end{align}
@@
 (\protect\hyperlink{ref-stammler2016adju}{2016}), Ji et al.
-(\protect\hyperlink{ref-ji2019quan}{2019}), Harrison et al.
-(\protect\hyperlink{ref-harrison2021theq}{2021}), REFs).
+(\protect\hyperlink{ref-ji2019quan}{2019}), Tkacz, Hortala, and Poole (\protect\hyperlink{ref-tkacz2018abso}{2018}), Harrison et al.
+(\protect\hyperlink{ref-harrison2021theq}{2021}), Rao et al.
+(\protect\hyperlink{ref-rao2021mult}{2021})).
 In this case, the reference abundance \(\tilde A_r^{(a)}\) is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
-(In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \eqref{eq:density-ratio-meas}; Appendix REF).
-
-Others have instead sought to determine naturally-occurring species that are thought to be constant across samples; we refer to such species as \emph{housekeeping species} by analogy with the housekeeping genes used for absolute-abundance conversion in gene-expression studies (REF).
-Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference (REFs).
+(In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \eqref{eq:density-ratio-meas}; Appendix TODO).
+
+Others have instead sought to determine naturally-occurring species that are thought to be constant across samples; we refer to such species as \emph{housekeeping species} by analogy with the housekeeping genes used for absolute-abundance conversion in gene-expression studies (Silver et al.
+(\protect\hyperlink{ref-silver2006sele}{2006})).
+Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference (Karasov et al.
+(\protect\hyperlink{ref-karasov2020ther}{2020}), Regalado et al.
+(\protect\hyperlink{ref-regalado2020comb}{2020}), Wallace et al.
+(\protect\hyperlink{ref-wallace2021thed}{2021})).
 A related approach involves computationally identifying species that are constant between pairs of samples (David et al.
@@
 The most common form of targeted measurement involves using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA.
-It is also possible to directly measure cell concentration by performing ddPCR prior to DNA extraction (REFs), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
-Appendix {[}REF{]} describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
+It is also possible to directly measure cell concentration by performing ddPCR prior to DNA extraction (Morella et al.
+(\protect\hyperlink{ref-morella2018rapi}{2018})), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
+Appendix TODO describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
 
@@
 We now turn to how the errors in MGS-based abundance measurements affect DA results.
-Our focus is on \emph{multiplicative DA}---the fold differences (FD) in the relative or absolute abundance of species between samples or between experimental conditions.
-Multiplicative DA has a direct ecological interpretation via the multiplicative processes of exponential growth and decay and forms the basis for several popular DA methods (REF DESeq2, ALDEx2(?), and corncob; a LV paper; fido).
-We also consider non-parametric rank-based methods; these methods are popular in case-control studies for which ecological interpretability is often of secondary importance to discovering stable associations between species and the condition of interest (REFs).
+Our focus is on \emph{multiplicative DA analyses} that estimate the fold differences (FD) in the relative or absolute abundance of species across time points or between experimental conditions.
+Multiplicative DA has a direct ecological interpretation via the multiplicative processes of exponential growth and decay.
+Many DA methods operate on additive differences in log-transformed abundances and are therefore multiplicative; these include popular methods such as DESeq2 (Love, Huber, and Anders (\protect\hyperlink{ref-love2014mode}{2014})), ALDEx2 (Fernandes et al.
+(\protect\hyperlink{ref-fernandes2014unif}{2014})), corncob (Martin, Witten, and Willis (\protect\hyperlink{ref-martin2020mode}{2020})), and time series methods based on the Lotka-Volterra model (e.g., Stein et al.
+(\protect\hyperlink{ref-stein2013ecol}{2013})).
+We also consider non-parametric rank-based methods; these methods are popular in case-control studies for which ecological interpretability is often of secondary importance to discovering stable associations between species and the condition of interest.
 
@@
 \end{align}
-In contrast, bias created a FE in the proportion of a species (Equation \eqref{eq:prop-error}) that varied inversely with the mean efficiency of the sample.
+In contrast, bias creates a FE in the proportion of a species (Equation \eqref{eq:prop-error}) that varies inversely with the mean efficiency of the sample.
 Therefore, this error does not cancel when calculating the FD in the proportion,
@@
 Though the FE in the FD is the same for each species, the biological implications of the error vary.
-In particular, there are three distinct types of error: an increase in the magnitude of the measured FD, a decrease in that magnitude, and a change in the direction of the FD.
-We refer to the latter as a \emph{sign error} in reference to the sign of the measured log fold difference (LFD).
+In particular, there are three distinct types of error: an increase in the magnitude, a decrease in the magnitude, and a reversal in the direction of the measured FD.
 We can see each type of error in Figure \ref{fig:error-proportions}.
 For Species 1, which increases in abundance and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FD (actual FD: 2.3, measured FD: 6.5).
-For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FD: 0.15, measured FD: 0.44).
-For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FD: 0.6, measured FD: 1.8), such that the species actually appears to increase (a sign error).
-In contrast, the ratios between species change by the same factors in the actual and measured taxonomic profiles.
-For example, the ratio of Species 2 to Species 3 changes from \(1/1\) to \(1/4\) in the actual profiles and from \(3\) to \(3/4\) in the measured profiles, a decrease by 4-fold either way.
+For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude of the measured FD (actual FD: 0.15, measured FD: 0.44).
+For Species 2, which decreases but by a smaller factor than the mean efficiency, the species actually appears to increase---a reversal in direction of the measured FD (actual FD: 0.6, measured FD: 1.8).
+In contrast, the FDs in the ratios among species are identical in the actual and measured taxonomic profiles.
+For example, the ratio of Species 2 to Species 3 shows the same 4-fold decrease in the actual profiles (\(1\) to \(1/4\)) and in the measured profiles (\(3\) to \(3/4\)).
 
@@
 First, consider a species' abundance measurement \(\tilde A_i^{(a)}\) derived from a non-MGS measurement of total-community abundance using Equation \eqref{eq:density-prop-meas}.
-Equation \eqref{eq:density-prop-error} indicates that the FE in this measurement equals the (constant) species efficiency multiplied by \({B^{{\text{[tot]}}(a)}}/{\bar B^{(a)}}\), the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
+Equation \eqref{eq:density-prop-error} indicates that the FE in this measurement equals the (constant) species efficiency multiplied by \({\bar B^{{\text{[tot]}}(a)}}/{\bar B^{(a)}}\), the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
 This ratio can vary, creating error in the measured FD between samples.
@@
 
-Most DA analyses can be understood as a regression analysis.
-For example, the simple linear regression model,
+In most cases, a DA analysis can be understood as regression of microbial abundance variables against one or more covariates describing predictive properties of the sample.
+The simplest such regression relates a microbial abundance variable \(y\) to a covariate \(x\) and a \emph{residual (or unexplained) error} \(\varepsilon\) via the simple linear regression formula
 \begin{align}
   \label{eq:regression}
-  y^{(a)} = \alpha + \beta x^{(a)} + \varepsilon^{(a)},
-\end{align}
-can be used to relate a microbial abundance variable \(y\) to a \emph{covariate} \(x\) and the \emph{residual (or unexplained) error} \(\varepsilon\).
+  y^{(a)} = \alpha + \beta x^{(a)} + \varepsilon^{(a)}.
+\end{align}
 For example, \(y\) may be the log absolute abundance of species \(i\) (\(\log A_i^{(a)}\)) and \(x\) may be a continuous variable (such as pH) or a binary variable (such as \(x=1\) for treated patients and \(x=0\) for untreated controls).
 The regression coefficients \(\alpha\) and \(\beta\) are parameters to be estimated from the data.
-A DA analysis is primarily interested in the slope coefficient \(\beta\), which determines how the average or expected value of \(y\) increases with \(x\).
+A DA regression analysis is primarily interested in the slope coefficient \(\beta\), which determines how the average or expected value of \(y\) increases with \(x\).
 For a binary covariate, \(\beta\) equals the average difference in \(y\) between samples with \(x=1\) versus those with \(x=0\).
-Taking the response \(y\) to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient \(\beta\) corresponding to the average log fold difference (LFD) in abundance between conditions or for a unit increase in \(x\).
-
-Appendix \ref{appendix-regression} derives mathematical results for the effect that measurement error in the response caused by taxonomic bias has on LFDs estimated via simple linear regression.
-Here we illustrate these results with an example.
-Suppose that we wish to estimate the average LFD in the absolute abundance of every species between samples from two conditions, such as treated and untreated patients.
-Further suppose that we have measured absolute abundances by multiplying the proportions from MGS with an accurate measurement of total-community abundance.
-For each species \(i\), we set \(y^{(a)}\) in Equation \eqref{eq:regression} equal to \(\log \tilde A_{i}^{(a)}\), and use either Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE) to estimate the coefficients \(\alpha\) and \(\beta\).
-From Equation \eqref{eq:density-prop-error}, we know that the FE in the abundance of each species equals its efficiency relative to the mean efficiency of the sample.
-
-To understand how the FE in species abundances affects the estimated LFD for a given species, it is useful to consider three possible scenarios for how the mean efficiency varies across samples.
-In the first scenario, the mean efficiency remains stable across samples, such that the FE in \(\log \tilde A_{i}^{(a)}\) is approximately constant.
-In this case, the FE induced by bias \emph{mostly cancels} and does not substantially affect the estimated LFD.
-In the second scenario, the mean efficiency varies substantially but is not associated with the condition, creating random FEs in the individual sample measurements.
-In this case, the LFDs between individual samples will be highly distorted, but these errors will effectively average out in the regression and the estimated LFD remains reliable.
-Although the random FEs caused by bias do not create a systematic error in the LFD estimate, they will tend to reduce the precision (i.e., increase the standard error) of the estimate.
-In the third scenario, the mean efficiency systematically varies with the condition, causing a systematic error in the estimated LFD regardless of the number of samples.
-In particular, the portion of additive variation in the log mean efficiency that is linearly associated with \(x\) creates an equal but opposite additive error in the estimated LFD for each species.
-Although the absolute additive error in the LFDs is the same for all species, its biological significance varies, depending on its magnitude relative to that of the species' true LFD.
-
-Figure \ref{fig:regression-example} shows a simulated example in which the mean efficiency varies systematically with the condition of interest.
-Here, the average log mean efficiency is higher in the condition corresponding to \(x=1\), due to an increase of the high-efficiency Species 9.
-This systematic increase in the log mean efficiency causes an artificial decrease in the estimated LFD for each species by an equal amount.
-As we saw for the FD between individual samples, the biological significance of the systematic error in the LFD estimates varies by species, despite the absolute error being constant across species.
-We see decreases in the magnitude of the LFD estimate (Species 9, 10, and 1 in Figure \ref{fig:regression-example}), changes in sign (Species 5), or increases in magnitude (remaining species) depending the value of the actual LFD relative to the error.
-The portion of the variation in the log mean efficiency that is uncorrelated with \(x\) increases the standard errors for most species (Figure \ref{fig:regression-example} D).
-The exception is Species 9, which is the primary driver of the mean efficiency dynamics in this example.
-Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).
+Taking \(y\) to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient \(\beta\) corresponding to the average log fold difference (LFD) in abundance between conditions or for a unit increase in \(x\).
+
+Although researchers typically use more sophisticated models, the simple linear regression model in Equation \eqref{eq:regression} provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
+Appendix \ref{appendix-regression} describes a general framework for finding the error in the estimated slope coefficient \(\beta\) caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures decribed in Section \ref{abundance-measurement}.
+In each case, a multiplicative DA analysis can be conducted by setting \(y\) equal to the logarithm of the untransformed abundance measurement (\(\tilde P_i\), \(\tilde R_{i/j}\), or \(\tilde A_i\)).
+Consider an abundance measure such as the ratio between species \(i\) and \(j\), \(\tilde R_{i/j}\).
+Taxonomic bias creates a constant FE in \(\tilde R_{i/j}\) that translates to a constant additive shift in \(\log \tilde R_{i/j}\) and therefore does not affect the estimate of the slope \(\hat \beta\).
+
+Now consider the absolute abundance of a species \(i\) obtained by normalizing its proportion to an non-MGS measurement of total community abundance, which for illustration we assume to be completely accurate.
+The measured abundance \(\tilde A_i\) thus has a FE that varies inversely with the mean efficiency \(\bar B\).
+If \(\bar B\) is stable across samples, then the additive error in \(\log \tilde A_i\) will be approximately constant and \(\hat \beta\) is again unaffected---bias has no effect on the DA result.
+If, on the other hand, \(\log \bar B\) tends to linearly vary with the covariate \(x\), then the additive error in \(\log \tilde A_i\) will linearly vary in the opposite direction and cause a systematic error in \(\hat \beta\) that is equal in magnitude but opposite in sign to the slope of \(\log \bar B\) with \(x\).
+A third possibility is that \(\bar B\) varies in an essentially random manner that is independent of the covariate \(x\).
+In this scenario, the error from taxonomic bias in the measurement \(\log \tilde A_{i}\) acts to increase the noise in \(\hat beta\) (i.e., increase its standard error), but does not cause a systematic error.
+
+Figure \ref{fig:regression-example} shows a simulated example of a regression analysis of the absolute abundances of 10 species across two conditions in which the mean efficiency is systematically greater in the condition \(x=1\) than the in the condition \(x=0\).
+The increase in the log mean efficiency from \(x=0\) to \(x=1\) (Figure \ref{fig:regression-example} A and B) causes an artificial decrease in the estimated LFD between conditions for each species (Figure \ref{fig:regression-example} C and D).
+The absolute error created by bias on the LFD estimate is the same for each species; however, its scientific impact varies.
+For the three species with large positive LFDs (Species 9, 10, and 1), bias decreases the magnitude of the LFD estimate.
+In contrast, for the three species with large negative LFDs (Species 8, 2, and 3), bias increases the magnitude of the estimate.
+For one species with a small positive LFD (Species 5), the effect of bias results in a negative LFD estimate (direction error).
+The three remaining species (Species 7, 6, and 4) have LFDs near zero, but bias causes them to have large negative LFDs.
 
 \begin{figure}
-\centering
-\includegraphics{figures/export-pdf/home/michael/research/differential-abundance-theory/notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example_files/figure-html5/main-figure-1.pdf}
-\caption{\label{fig:regression-example}\textbf{Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.} This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by \(x=0\) and \(x=1\).
-In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFC, which drives the positive association of the log mean efficiency with the condition (shown in Panels A and B).
-This positive LFC in the log mean efficiency induces a systematic negative shift in the estimated LFCs of all species (Panels C and D).
-Panel D shows the mean LFC (points) with 95\% confidence intervals (CIs), for each species estimated from either the actual or the measured densities.
-The error (difference in LFC estimates on measured and actual) equals the negative LFC of the mean efficiency (shown in Panel B).}
+\includegraphics[width=0.9\linewidth]{figures/export-pdf/home/michael/research/differential-abundance-theory/notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example_files/figure-html5/main-figure-1} \caption{\textbf{Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.} This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by \(x=0\) and \(x=1\).
+In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFD, which drives a positive association of the log mean efficiency with the condition (shown in Panels A and B).
+This positive LFD in the log mean efficiency induces a systematic negative shift in the estimated LFDs of all species (Panels C and D).
+Panel D shows the mean LFD (points) with 95\% confidence intervals (CIs), for each species estimated from either the actual or measured abundances.
+The error (difference in LFD estimates on measured and actual) equals the negative LFD of the mean efficiency (shown in Panel B).}\label{fig:regression-example}
 \end{figure}
@@
 
-Rank-based or `nonparametric' DA methods work by first applying a cross-sample rank-transformation to the abundance variable \(y\): The sample in which \(y\) has the smallest value receives a rank of 1, the sample with the second smallest value receives a rank of 2, etc.
-These methods then analyze how the average rank varies across conditions.
-Rank-based methods commonly used in microbiome DA analysis include the Wilcoxon-Mann-Whitney and Kruskal-Wallis tests for a discrete covariate (e.g.~case vs control) and Spearman's rank correlation for a continuous covariate (e.g.~pH).
-They have been applied to species proportions (REFs), ratios (REFs), and absolute abundances (REFs).
-
-Multiplying a variable by a constant factor does not change its cross-sample ranks; thus rank-based DA methods are unaffected by constant FE in the abundance variable \(y\).
-We therefore conclude that consistent taxonomic bias does not affect rank-based DA analyses of species ratios or absolute abundances derived from a reference species.
-In contrast, variable FE in \(y\) (driven by variation in the mean efficiency) can affect the cross-sample ranks; if the FE is associated with the covariate \(x\), it can cause systematic error in the DA result.
-Thus consistent bias may impact rank-based analyses of species proportions and absolute abundances derived from a total-community abundance measurement.
-Future work should consider whether rank-based DA methods are significantly more robust than multiplicative DA methods to variation in the mean efficiency.
+Rank-based DA methods work by first applying a cross-sample rank-transformation to the abundance variable \(y\): The sample in which \(y\) has the smallest value receives a rank of 1, the sample with the second smallest value receives a rank of 2, etc.
+These methods then analyze how the average rank of \(y\) varies across sample conditions (Conover (\protect\hyperlink{ref-conover2012ther}{2012})).
+Rank-based methods commonly used in microbiome DA analysis include the Wilcoxon-Mann-Whitney and Kruskal-Wallis tests for a discrete covariate (e.g.~case vs control) and Spearman's rank correlation for a continuous covariate (e.g.~pH); these tests have collectively been applied to species proportions (Callahan et al.
+(\protect\hyperlink{ref-callahan2017repl}{2017}),Fettweis et al.
+(\protect\hyperlink{ref-fettweis2019thev}{2019})), ratios (Fernandes et al.
+(\protect\hyperlink{ref-fernandes2014unif}{2014})), and absolute abundances (Vieira-Silva et al.
+(\protect\hyperlink{ref-vieirasilva2019quan}{2019})).
+
+Multiplying a variable by a constant factor does not change its cross-sample ranks, making rank-based methods unaffected by a constant FE in the abundance variable \(y\).
+Thus in our MGS measurement model, taxonomic bias does not affect rank-based DA analyses of species ratios or absolute abundances derived from a reference species.
+But for species proportions and absolute abundances derived from a total-community abundance measurement, variation in the mean efficiency \(\bar B\) can affect the cross-sample ranks and cause error in DA results if \(\bar B\) is associated with the covariate.
+Further work is needed to determine whether rank-based DA methods are significantly more robust to variation in \(\bar B\) than multiplicative DA methods.
 
@@
 
-We repeated the two DA analyses of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) with and without calibration and found that the results did not meaningfully differ.
-To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \ref{fig:leopold2020host-variation}).
+We repeated the two DA analyses of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) with and without calibration and found that the results did not meaningfully differ
+(\href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/}{SI Analysis of host genetics and arrival order},
+\href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/}{SI Analysis of priority effects}).
+To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities
+(\href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-leopold2020host-case-study/}{SI Analysis},
+SI Figure \ref{fig:leopold2020host-variation}).
 Despite the 13-fold variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \ref{fig:leopold2020host-variation}C), having a geometric range of 1.62 and a geometric standard deviation of 1.05.
@@
 
-We performed an additional DA analysis on the data from Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) to investigate whether any commensals increased in absolute concentration in response to infection.
+We performed an additional DA analysis on the data from Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) to investigate whether any commensals increased in absolute concentration in response to infection (\href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-leopold2020host-case-study/\#change-in-commensals-due-to-infection}{SI Analysis}).
 The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \ref{fig:leopold2020host-infection-mean-efficiency-dist}).
@@
 A second \emph{Lactobacillus} species, \emph{L. crispatus}, had an efficiency that was approximately 2-fold less than \emph{L. iners} and 15-fold greater than \emph{G. vaginalis}.
-These species, along with the unculturable \emph{Lachnospiraceae BVAB1}, are most frequently the most abundant species in a sample and commonly reach high proportions of over 70\% (REF SI analysis).
+These species, along with the unculturable \emph{Lachnospiraceae BVAB1}, are most frequently the most abundant species in a sample and commonly reach high proportions of over 70\%
+(\href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2021-11-01-momspi-summary/}{SI Analysis}).
 Therefore, shifts between these dominant species could drive large changes in the sample mean efficiency, which might in turn distort DA analyses.
@@
 (\protect\hyperlink{ref-brooks2015thet}{2015}) control measurements.
-Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample (REF SI Analysis).
-The mean efficiency varies substantially across vaginal samples (Figure 4.1), with samples in which a \emph{Lactobacillus} species is most abundant typically having a mean efficiency that is 3-fold to 20-fold greater than samples in which \emph{G.
+Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample
+(\href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2021-11-01-momspi-summary/}{SI Analysis}).
+The mean efficiency varies substantially across vaginal samples (Figure \ref{fig:momspi-mean-efficiency-dist}), with samples in which a \emph{Lactobacillus} species is most abundant typically having a mean efficiency that is 3-fold to 20-fold greater than samples in which \emph{G.
 vaginalis} is most abundant.
-The vaginal microbiome sometimes shifted between \emph{Lactobacillus}-dominance and \emph{Gardnerella}-dominance between consecutive visits in individual women (SI Figure \ref{fig:momspi-mean-efficiency-fcs}), which caused magnitude and sign errors in the trajectories of lower-abundance species (SI Figure \ref{fig:momspi-trajectory}).
+The vaginal microbiome sometimes shifted between \emph{Lactobacillus}-dominance and \emph{Gardnerella}-dominance between consecutive visits in individual women (SI Figure \ref{fig:momspi-mean-efficiency-fcs}), which caused magnitude and direction errors in the trajectories of lower-abundance species (SI Figure \ref{fig:momspi-trajectory}).
 
@@
 These observations suggest that a proportion-based DA analysis of a health condition that is associated with \emph{Lactobacillus} and/or \emph{Gardnerella} would be prone to spurious results.
-To demonstrate this possibility, we performed a DA analysis of the MOMS-PI samples with respect to a simulated health condition.
+To demonstrate this possibility, we performed a DA analysis of the MOMS-PI samples with respect to a simulated health condition
+(\href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2021-12-15-momspi-regression-diversity}{SI Analysis}).
 Bacterial vaginosis (BV) has been repeatedly found to be associated with a dramatically reduced proportion of \emph{Lactobacillus spp.}, an increased proportion of \emph{G.
@@
 
-We wondered whether DA in the human gut microbiome might be less affected by bias than the vaginal microbiome, due to its greater species diversity.
-Recall that the mean efficiency is an abundance-weighted average over species in a sample; thus we might expect it to become more stable as the within-sample (alpha) diversity increases.
-In fact, one can show that for communities that are randomly assembled from a large species pool, the additive variance in the mean efficiency is inversely proportional to order-2 diversity \(^2D\) (also known as the Inverse Simpson diversity index) (Appendix \ref{diversity-and-mean-efficiency}).
-But real communities are assembled through complex processes that create positive and negative associations among species (REF).
-Thus there is unlikely to be a simple 1-1 relationship between alpha diversity and variation in the mean efficiency.
-Moreover, even if the mean efficiency does vary less across the more diverse gut samples, if the proportion of each species also varies less, then the importance of bias for DA analyses need not diminish.
-
-To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project (Huttenhower et al.
-(\protect\hyperlink{ref-huttenhower2012stru}{2012}), SI Methods \ref{methods-gut}).
+We wondered whether DA studies of the human gut microbiome might be less affected by bias than studies of the vaginal microbiome, due to the greater alpha diversity observed within gut microbiome samples.
+Recall that the mean efficiency is an abundance-weighted average over the species in a sample.
+Thus, all else equal, we expect the mean efficiency to vary less across samples from ecosystems where samples have higher alpha (within-sample) diversity, as is the case for stool samples relative to vaginal samples (Huttenhower et al.
+(\protect\hyperlink{ref-huttenhower2012stru}{2012})).
+On the other hand, the proportion of individual species may also vary less less across samples when alpha diversity is higher, in which case the effect of bias on DA results need not diminish.
+
+To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project
+(Huttenhower et al.
+(\protect\hyperlink{ref-huttenhower2012stru}{2012}), \href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-30-hmp-stool-vagina-comparison/}{SI Analysis}).
 On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \ref{fig:gut}A; GM \(\md\) GSD of 6.3 \(\md\) 1.7 in stool samples and 1.4 \(\md\) 1.6 in vaginal samples).
@@
 Importantly, Lloyd et al. (\protect\hyperlink{ref-lloyd2020evid}{2020}) also used qPCR to measure the absolute abundance of two taxa from these same samples.
-By comparing qPCR to FRAxC growth rates for these taxa, we can estimate the systematic error in FRAxC growth rates.
+By comparing qPCR to FRAxC growth rates for these taxa, we can estimate the systematic error in FRAxC growth rates (\href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-lloyd2020evi-case-study/}{SI Analysis})
 Because the systematic error that bias causes in the regression slope is the same for each taxon (Section \ref{differential-abundance}), these comparisons allow us to draw conclusions about the accuracy of FRAxC growth rates for all taxa.
@@
 
-It is commonly thought that analyses of the differences between samples are robust to the taxonomic bias inherent in MGS measurement as long as all samples have been subjected to the same MGS workflow.
-But our results show that consistent taxonomic bias is still capable of creating scientific errors in common forms of DA analysis.
-In particular, we showed mathematically that bias affects DA analysis methods based on proportions or in absolute abundances derived from them (using total-abundance normalization), due to variation in the mean efficiency across samples of varying taxonomic composition.
-The error, however, varies by experimental context and may often be negligible in a practical sense.
-Moreover, it can be quantified and corrected through the use of reference-taxon measurements and/or community calibration controls.
-In addition, bias-aware sensitivity and meta-analyses make it possible to rigorously account for bias even in the absence of such controls.
-Applications of these methods will gradually improve our understanding of the experimental contexts in which variation in the mean efficiency is likely to be problematic and should be accounted for.
-
-We also showed that other DA methods are more robust---perhaps even entirely invariant---to consistent taxonomic bias.
-In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
-In addition, careful pairing of total-abundance and MGS measurement methods can counter the variation in mean efficiency of the MGS measurement, leading to more consistent fold errors across samples.
-
-Important open questions and future research directions remain.
-Key open questions include 1) determining the extent to which bias is consistent across samples for different taxonomic levels, MGS methods, and sample types; 2) assessing the validity of post-extraction measurements as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how different underlying community dynamics, and in particular the source of zero counts in MGS measurements, affect bias sensitivity.
+It is commonly assumed that differential abundance (DA) analysis is robust to taxonomic bias, so long as all samples have been subjected to the same MGS workflow.
+In contrast, our results show that consistent taxonomic bias can distort the results of DA methods that are based on species proportions.
+This distortion occurs because the fold error in a species' proportion depends on the sample mean efficiency, which can vary across samples and conditions in accordance with the overall community composition.
+This problem continues to apply for analyses of absolute abundance that are based on multiplying the proportions from MGS by the total abundance of the community.
+In contrast, DA methods based on species ratios are invariant to consistent bias, since the error in the ratio between species is independent of community composition.
+In Section \ref{solutions}, we describe how ratio-based methods can thus provide a more robust approach to analyzing relative and absolute abundances;
+in addition, calibration with community controls, bias-sensitivity analysis, and bias-aware meta-analysis provide additional ways to correct or mitigate the effect of bias.
+
+Proportion-based DA methods encompass many of the most popular methods for analyzing relative and absolute abundances.
+Importantly, however, we found that the degree and significance of the error due to bias depends on experimental context and may often be negligible.
+In particular, if the mean efficiency is approximately constant across samples, then bias has a negligible effect on multiplicative and rank-based cross-sample comparisons, and DA results will be unaffected.
+If the mean efficiency varies but is unassociated with the condition of interest, then bias merely serves to increase noise and does not create systematic errors in DA results.
+It is only when the mean efficiency is associated with the condition being analyzed that large systematic errors can occur.
+Our case studies suggest that this problematic scenario can occur but may well be the exception rather than the rule.
+Thus our results are still consistent with taxonomic bias having had a limited impact on extant DA results.
+Systematic investigation of how the mean efficiency affects DA results across a wide range of studies may increase our confidence in previous DA results and/or alert us to the conditions in which they are most suspect.
+
+Important open questions include 1) determining the extent to which bias is consistent across samples for different taxonomic levels, MGS methods, and sample types; 2) assessing the validity of post-extraction measurements as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how different underlying community dynamics, and in particular the source of zero counts in MGS measurements, affect bias sensitivity.
 In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, more sophisticated statistical tools are needed to properly account for both taxonomic bias and random variation in the underlying MGS and supplemental (e.g.~qPCR) measurements.
@@
 This transition drives a sharp increase in the mean efficiency, which significantly distorts the fold changes in the observed (uncalibrated) microbiome measurements for species with less dramatic fold changes.
-Two exemplar species are shown to illustrate the magnitude (\emph{Ureaplasma cluster 23}) and sign (\emph{Megasphaera OTU70 type1}) errors that can arise in this situation.}\label{fig:momspi-trajectory}
+Two exemplar species are shown to illustrate the magnitude (\emph{Ureaplasma cluster 23}) and direction (\emph{Megasphaera OTU70 type1}) errors that can arise in this situation.}\label{fig:momspi-trajectory}
 \end{figure}
@@
 Different choices of the pseudo-value substantially change the measured GSDs, but not the difference (ratio) between gut and vaginal species.}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{figures/export-pdf/notebook/_posts/2021-10-18-evaluate-robustness-example/evaluate-robustness-example_files/figure-html5/summary_plot-1.pdf}
-\caption{\label{fig:sensitivity-example}\textbf{A bias-sensivity analysis can be performed to examine how sensitive the results of a DA analysis are to assumptions about taxonomic bias in community measurements.} The figure shows the results of a bias-sensitivity analysis used to study the effect of bias on the association of \emph{Gardnerella vaginalis} and preterm birth that was investigated by Callahan et al.
-(\protect\hyperlink{ref-callahan2017repl}{2017}).
-100 random efficiency vectors were drawn at 6 different bias strengths (quantified by the variance in log efficiency, \(\sigma_{e}^{2}\)).
-Each efficiency vector was used to calibrate the MGS profiles and perform a DA association test of \emph{G.
-vaginalis} versus the host's preterm birth outcome; regression coefficients \(\hat \beta\) indicate the increase of average logit proportion of \emph{G.
-vaginalis} in women who experienced preterm birth.}
 \end{figure}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-conover2012ther}{}}%
+Conover, W Jay.
+2012.
+{``The Rank Transformation-an Easy and Intuitive Way to Connect Many Nonparametric Methods to Their Parametric Counterparts for Seamless Teaching Introductory Statistics Courses.''} \emph{Wiley Interdiscip.
+Rev.
+Comput.
+Stat.} 4 (5): 432--38.
+\url{https://doi.org/10.1002/wics.1216}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-contijoch2019gutm}{}}%
@@
 \url{https://doi.org/10.1186/gb-2014-15-7-r89}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fernandes2014unif}{}}%
+Fernandes, Andrew D, Jennifer NS Reid, Jean M Macklaim, Thomas A McMurrough, David R Edgell, and Gregory B Gloor.
+2014.
+{``{Unifying the analysis of high-throughput sequencing datasets: characterizing RNA-seq, 16S rRNA gene sequencing and selective growth experiments by compositional data analysis}.''} \emph{Microbiome} 2 (1): 15.
+\url{https://doi.org/10.1186/2049-2618-2-15}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-karasov2020ther}{}}%
+Karasov, Talia L., Manuela Neumann, Alejandra Duque-Jaramillo, Sonja Kersten, Ilja Bezrukov, Birgit Schröppel, Efthymia Symeonidi, et al.
+2020.
+{``{The relationship between microbial population size and disease in the Arabidopsis thaliana phyllosphere}.''} \emph{bioRxiv}.
+\url{https://doi.org/10.1101/828814}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kumar2018anal}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lloyd2013meta}{}}%
+Lloyd, Karen G., Megan K.
+May, Richard T.
+Kevorkian, and Andrew D.
+Steen.
+2013.
+{``{Meta-analysis of quantification methods shows that archaea and bacteria have similar abundances in the subseafloor}.''} \emph{Appl.
+Environ.
+Microbiol.} 79 (24): 7790--99.
+\url{https://doi.org/10.1128/AEM.02090-13}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-lofgren2019geno}{}}%
@@
 \url{https://doi.org/10.1111/mec.14995}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-love2014mode}{}}%
+Love, Michael I, Wolfgang Huber, and Simon Anders.
+2014.
+{``{Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2}.''} \emph{Genome Biol.} 15 (12): 550.
+\url{https://doi.org/10.1186/s13059-014-0550-8}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-martin2020mode}{}}%
+Martin, Bryan D., Daniela Witten, and Amy D.
+Willis.
+2020.
+{``{Modeling microbial abundances and dysbiosis with beta-binomial regression}.''} \emph{Ann.
+Appl.
+Stat.} 14 (1): 94--115.
+\url{https://doi.org/10.1214/19-AOAS1283}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-mclaren2019cons}{}}%
@@
 \url{https://doi.org/10.7554/eLife.46923}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-morella2018rapi}{}}%
+Morella, Norma M., Shangyang Christopher Yang, Catherine A.
+Hernandez, and Britt Koskella.
+2018.
+{``{Rapid quantification of bacteriophages and their bacterial hosts in vitro and in vivo using droplet digital PCR}.''} \emph{J.
+Virol.
+Methods} 259 (May): 18--24.
+\url{https://doi.org/10.1016/j.jviromet.2018.05.007}.
 
@@
 \url{https://doi.org/10.1128/mSystems.00230-19}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rao2021mult}{}}%
+Rao, Chitong, Katharine Z.
+Coyte, Wayne Bainter, Raif S.
+Geha, Camilia R.
+Martin, and Seth Rakoff-Nahoum.
+2021.
+{``{Multi-kingdom ecological drivers of microbiota assembly in preterm infants}.''} \emph{Nature} 591 (7851): 633--38.
+\url{https://doi.org/10.1038/s41586-021-03241-8}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-regalado2020comb}{}}%
+Regalado, Julian, Derek S.
+Lundberg, Oliver Deusch, Sonja Kersten, Talia Karasov, Karin Poersch, Gautam Shirsekar, and Detlef Weigel.
+2020.
+{``{Combining whole-genome shotgun sequencing and rRNA gene amplicon analyses to improve detection of microbe--microbe interaction networks in plant leaves}.''} \emph{ISME J.}, May, 823492.
+\url{https://doi.org/10.1038/s41396-020-0665-8}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-silver2006sele}{}}%
+Silver, Nicholas, Steve Best, Jie Jiang, and Swee Lay Thein.
+2006.
+{``Selection of Housekeeping Genes for Gene Expression Studies in Human Reticulocytes Using Real-Time {PCR}.''} \emph{BMC Mol.
+Biol.} 7 (October): 33.
+\url{https://doi.org/10.1186/1471-2199-7-33}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-srinivasan2012bact}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-stein2013ecol}{}}%
+Stein, Richard R., Vanni Bucci, Nora C.
+Toussaint, Charlie G.
+Buffie, Gunnar Rätsch, Eric G.
+Pamer, Chris Sander, and João B.
+Xavier.
+2013.
+{``{Ecological Modeling from Time-Series Inference: Insight into Dynamics and Stability of Intestinal Microbiota}.''} Edited by Christian von Mering.
+\emph{PLoS Comput.
+Biol.} 9 (12): e1003388.
+\url{https://doi.org/10.1371/journal.pcbi.1003388}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-tettamantiboshier2020comp}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-tkacz2018abso}{}}%
+Tkacz, Andrzej, Marion Hortala, and Philip S.
+Poole.
+2018.
+{``{Absolute quantitation of microbiota abundance in environmental samples}.''} \emph{Microbiome} 6 (1): 110.
+\url{https://doi.org/10.1186/s40168-018-0491-7}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-vandeputte2017quan}{}}%
@@
 \url{https://doi.org/10.1038/nature24460}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vieirasilva2019quan}{}}%
+Vieira-Silva, Sara, João Sabino, Mireia Valles-Colomer, Gwen Falony, Gunter Kathagen, Clara Caenepeel, Isabelle Cleynen, Schalk van der Merwe, Séverine Vermeire, and Jeroen Raes.
+2019.
+{``{Quantitative microbiome profiling disentangles inflammation- and bile duct obstruction-associated microbiota alterations across PSC/IBD diagnoses}.''} \emph{Nat.
+Microbiol.} 4 (11): 1826--31.
+\url{https://doi.org/10.1038/s41564-019-0483-9}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wallace2021thed}{}}%
+Wallace, Megan A., Kelsey A.
+Coffman, Clément Gilbert, Sanjana Ravindran, Gregory F.
+Albery, Jessica Abbott, Eliza Argyridou, et al.
+2021.
+{``{The discovery, distribution, and diversity of DNA viruses associated with Drosophila melanogaster in Europe}.''} \emph{Virus Evol.} 7 (1): 1--23.
+\url{https://doi.org/10.1093/ve/veab031}.
 

### 2022-08-18 abundance-measurement.Rmd
Remove references to unwritten appendix sections
The previous version made two additional claims about using
spike-ins/reference species in the section on abundance measurement,
which referenced two unwritten appendix sections for support. However,
these claims are unessential to the main argument. I am removing them
for now, and will consider adding them back in a future version.

@@
 In this case, the reference abundance $\tilde A_r^{(a)}$ is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
-(In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \@ref(eq:density-ratio-meas); Appendix TODO).
 
@@
 It is also possible to directly measure cell concentration by performing ddPCR prior to DNA extraction (@morella2018rapi), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
-Appendix TODO describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.

### 2022-08-18 index.Rmd
Revise title
Edit frontmatter
Move abstract from introduction.Rmd to index.Rmd, and replace the
Preface section. Remove the note indicating that this is research in
progress, but keep the extra info including links and information about
manuscript versioning. Do some light reformatting. Shorten the
printed manuscript hash.

@@
 ---
-title: "Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis"
+title: "Implications of taxonomic bias for microbial differential-abundance analysis"
 author:
@@
 
-# Preface {-}
+# Abstract {-}
+
+<!-- **Abstract:** -->
+Differential-abundance (DA) analyses enable microbiome researchers to assess how microbial species vary in relative or absolute abundance with specific host or environmental conditions, such as health status or pH.
+These analyses typically use sequencing-based community measurements that are taxonomically biased to measure some species more efficiently than others.
+Understanding the effects that taxonomic bias has on the results of a DA analysis is essential for achieving reliable and translatable findings; yet currently, these effects are unknown.
+Here, we characterized these effects for DA analyses of both relative and absolute abundances, using a combination of mathematical theory and data analysis of real and simulated case studies.
+We found that, for analyses based on species proportions, taxonomic bias can cause significant errors in DA results if the average measurement efficiency of the community is associated with the condition of interest.
+These errors can be avoided by using more robust DA methods (based on species ratios) or quantified and corrected using appropriate controls.
+Wide adoption of our recommendations can improve the reproducibility, interpretability, and translatability of microbiome DA studies.
 
@@
 sha <- repo %>% git2r::commits(n = 1) %>% .[[1]] %>% git2r::sha()
+sha_short <- str_sub(sha, 1, 7)
 status <- repo %>% git2r::status()
@@
 
+::: {#preface-note .rmdnote}
+
+This manuscript was rendered from commit `r sha_short` of https://github.com/mikemc/differential-abundance-theory.
+Supporting data analyses can be found in the accompanying [computational research notebook](https://mikemc.github.io/differential-abundance-theory/notebook/).
+Please post comments or questions on [GitHub](https://github.com/mikemc/differential-abundance-theory).
+The manuscript is licensed under a [CC BY 4.0 License](https://creativecommons.org/licenses/by/4.0/).
+See the [GitHub Releases](https://github.com/mikemc/differential-abundance-theory/releases) or [Zenodo record](https://doi.org/10.5281/zenodo.4552717) for earlier versions.
+
 ```{r}
+if (!identical(status %>% unlist %>% length, 0L))
   print(status)
@@
 
-_This manuscript was rendered from commit `r sha`._
-
-::: {#preface-warning .rmdnote}
-
-**This in-progress manuscript is not intended for general scientific use.**
-It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
-Please post comments or questions on the [GitHub Issues page](https://github.com/mikemc/differential-abundance-theory/issues) or [email Mike](m.mclaren42@gmail.com).
-
 :::
-
-This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.
-We describe the basic problem posed by taxonomic bias for measuring changes in the abundance of particular taxa across conditions and describe new strategies for mitigating the errors it induces.
-Analyses of both relative and absolute abundances are considered.
-The manuscript is licensed under a [CC BY 4.0 License](https://creativecommons.org/licenses/by/4.0/).
-See [the Zenodo record](https://doi.org/10.5281/zenodo.4552717) for how to cite the latest version.
-Supporting data analyses can be found in the accompanying [computational research notebook](https://mikemc.github.io/differential-abundance-theory/notebook/).

### 2022-08-18 introduction.Rmd
Edit frontmatter
Move abstract from introduction.Rmd to index.Rmd, and replace the
Preface section. Remove the note indicating that this is research in
progress, but keep the extra info including links and information about
manuscript versioning. Do some light reformatting. Shorten the
printed manuscript hash.

@@
 # Introduction {#introduction}
-
-::: {#abstract .rmdnote}
-
-**Abstract:**
-
-Differential-abundance (DA) analyses enable microbiome researchers to assess how microbial species vary in relative or absolute abundance with specific host or environmental conditions, such as health status or pH.
-These analyses typically use sequencing-based community measurements that are taxonomically biased to measure some species more efficiently than others.
-Understanding the effects that taxonomic bias has on the results of a DA analysis is essential for achieving reliable and translatable findings; yet currently, these effects are unknown.
-Here, we characterized these effects for DA analyses of both relative and absolute abundances, using a combination of mathematical theory and data analysis of real and simulated case studies.
-We found that, for analyses based on species proportions, taxonomic bias can cause significant errors in DA results if the average measurement efficiency of the community is associated with the condition of interest.
-These errors can be avoided by using more robust DA methods (based on species ratios) or quantified and corrected using appropriate controls.
-Wide adoption of our recommendations can improve the reproducibility, interpretability, and translatability of microbiome DA studies.
-
-:::
 

### 2022-08-18 abundance-measurement.Rmd
Increase figure width

@@
 
-```{r error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
+```{r error-proportions, out.width = '100%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
 "figures/illustrations/error-proportions.svg" %>%

### 2022-08-19 abundance-measurement.Rmd
Make edits from AW and BC

@@
 
-This section extends the theoretical results of @mclaren2019cons to describe the effect that consistent taxonomic bias within an MGS experiment affects the relative and absolute abundances measured for various microbial species.
+This section extends the theoretical results of @mclaren2019cons to describe how taxonomic bias in an MGS experiment affects the relative and absolute abundances measured for various microbial species.
 We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.
 
-## Model of MGS measurement
+## A model of MGS measurements
 
@@
 We distinguish between two types of species-level *relative abundances* within a sample.
-The *proportion* $P_{i}^{(a)}$ of species $i$ in sample $a$ equals its abundance relative to the total abundance of all species in $S$,
+The *proportion* $P_{i}^{(a)}$ of species $i$ in sample $a$ equals its abundance divided by the total abundance of all species in $S$,
 \begin{align}
@@
 \end{align}
-<!-- Taxonomic bias thus creates a fold-error (FE) in the measured proportion $\tilde P_{i}^{(a)}$ of species $i$ equal to the efficiency $B_i$ of species $i$ $divided by the mean efficiency $\bar B^{(a)}$ in the sample.
--->
-Taxonomic bias thus creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
+Taxonomic bias creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
 Since the mean efficiency varies across samples, so does the FE.
@@
 \end{align}
-Taxonomic bias thus creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
-For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
+Taxonomic bias creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
+For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-measured by a factor of 6 in both communities despite their varying compositions.
 A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
@@
 \end{align}
-With these definitions, the read count for taxon $I$ can be expressed as
+With these definitions, the read count for higher-order taxon $I$ can be expressed as
 $M_{I}^{(a)} = A_{I}^{(a)} B_I^{(a)} F^{(a)}$.
-Thus $B_I^{(a)}$ plays a role analogous to the efficiency of an individual species, but differs in that it need not be constant across samples:
-If the constituent species have different efficiencies, then the efficiency of the higher-order taxon $I$ depends on the relative abundances of its constituents and so will tend to vary across samples (@mclaren2019cons).
+Thus $B_I^{(a)}$ plays a role analogous to the efficiency of an individual species, but differs in that it is not constant across samples:
+If the constituent species have different efficiencies, then the efficiency of the higher-order taxon $I$ depends on the relative abundances of its constituents and so will vary across samples (@mclaren2019cons).
 As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
@@
 These extensions fall into two general approaches.
-The first approach leverages information about the abundance of the total community; for example, @vandeputte2017quan measured total-community abundance using flow cytometry and multiplied this number by MGS genus proportions to obtain the absolute abundances of individual genera (@vandeputte2017quan).
+The first approach leverages information about the abundance of the total community; for example, @vandeputte2017quan measured total-community abundance using flow cytometry and multiplied this number by genus proportions measured by MGS to quantify the absolute abundances of individual genera (@vandeputte2017quan).
 A second approach leverages information about the abundance of one or more individual species; for example, a researcher might 'spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (@harrison2021theq).
@@
 Suppose that the total abundance of all species in the sample, $A_{\tot}^{(a)}$, has been measured by a non-MGS method, yielding a measurement $\tilde A_\tot^{(a)}$.
-The absolute abundance of an individual species can be measured by multiplying the species' proportion from MGS by this total-abundance measurement,
+The absolute abundance of an individual species can be quantified by multiplying the species' proportion from MGS by this total-abundance measurement,
 \begin{align}
@@
 
-Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
-Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
+Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias that is analogous to, but quantitatively different from, the MGS relative abundance measurements.
+Flow cytometry may yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
 Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency (@lloyd2013meta).
@@
 
-Species abundance measurements derived by this method are affected by taxonomic bias in both the MGS and total-abundance measurement.
-We can determine the resulting fold error (FE) by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
+Species abundance measurements derived by this method (Equation \@ref(eq:density-prop-meas)) are affected by taxonomic bias in both the MGS and total-abundance measurement.
+We can determine the resulting fold error (FE) in the estimate $\tilde A_i^{(a)}$ by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
 \begin{align}
@@
 Equation \@ref(eq:density-prop-error) indicates that the FE in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
-As in the case of proportions (Equation \@ref(eq:prop-error)), the FE depends on sample composition through the two mean efficiency terms and so will vary across samples unless the two perfectly covary.
+As in the case of proportions (Equation \@ref(eq:prop-error)), the FE depends on sample composition through the two mean efficiency terms and so will, in general, vary across samples.
 

### 2022-08-19 case-studies.Rmd
Make edits from AW and BC

@@
 In practice it is difficult to construct control communities that span the many species present in complex natural communities.
-However, gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since, unlike most natural ecosystems, it is possible to assemble mock communities containing all species in known relative abundances.
+However, gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since it is possible to assemble mock communities containing all species in known relative abundances.
 
@@
 Both analyses were based on fold differences (FDs) in species proportions and so in principle were sensitive to taxonomic bias.
-To improve accuracy, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.
+To improve accuracy, the authors incorporated the bias measured from the control samples into analysis-specific calibration procedures.
 
@@
 Despite the 13-fold variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62 and a geometric standard deviation of 1.05.
-This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
+This consistency in the mean efficiency was despite the fact that each species showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
 But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3-fold and by just 1.5-fold between the two most dominant.
@@
 Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5-fold to 5.2-fold.
-Using gamma-Poisson (or negative-binomial) regression of the observed read counts, we estimated the average LFD in commensal species proportions following infection with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
-Commensals typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
+Using gamma-Poisson regression of the observed read counts, we estimated the average LFD in commensal species proportions following infection with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
+The commensals' proportions typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
 However, failure to account for bias caused the decrease to be overestimated, by an amount corresponding to the inverse change in log mean efficiency.
@@
 (@huttenhower2012stru, [SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-30-hmp-stool-vagina-comparison/)).
-On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \@ref(fig:gut)A; GM $\md$ GSD of `r hmp_div_stats['stool', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['stool', 'gsd'] %>% round(1)` in stool samples and `r hmp_div_stats['vagina', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['vagina', 'gsd'] %>% round(1)` in vaginal samples).
+On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \@ref(fig:gut)A; geometric mean (GM) $\md$ geometric standard deviation (GSD) of `r hmp_div_stats['stool', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['stool', 'gsd'] %>% round(1)` in stool samples and `r hmp_div_stats['vagina', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['vagina', 'gsd'] %>% round(1)` in vaginal samples).
 To assess the potential importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the multiplicative variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth.
 Across simulation replicates, the GSD in the mean efficiency was typically lower in stool than in vaginal samples (SI Figure \@ref(fig:gut)B; ratio of GSD in vaginal samples to GSD in stool samples of `r hmp_me_stats['iid', 'gm'] %>% round(1)` $\md$ `r hmp_me_stats['iid', 'gsd'] %>% round(1)` across 1000 simulations).
-Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \@ref(fig:gut)C), which the GSD in the proportion of a gut species across gut samples an average of `r hmp_species_diff %>% round(2)`-fold lower than that of a vaginal species across vaginal samples.
+Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \@ref(fig:gut)C); the GSD in the proportion of a gut species across gut samples was an average of `r hmp_species_diff %>% round(2)`-fold lower than that of a vaginal species across vaginal samples.
 Recall that the importance of bias estimation of the FD in a species' proportion depends on GSD in the mean efficiency relative the GSD in the species' proportion (Section \@ref(differential-abundance)).
@@
 Yet we also observed cases where the mean efficiency varied substantially.
-In several cases, the mean efficiency varied systematically with the condition of interest sufficient to cause substantial systematic error in LFD estimates (Scenario 3).
-Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
+In several cases, the (log) mean efficiency co-varied sufficiently strongly with the condition of interest to cause substantial systematic errors in LFD estimates (Scenario 3).
+Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency substantially increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
 Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time that is sufficient to substantially inflate the estimated growth rates of slowly changing species.

### 2022-08-19 conclusion.Rmd
Make edits from AW and BC

@@
 Proportion-based DA methods encompass many of the most popular methods for analyzing relative and absolute abundances.
-Importantly, however, we found that the degree and significance of the error due to bias depends on experimental context and may often be negligible.
-In particular, if the mean efficiency is approximately constant across samples, then bias has a negligible effect on multiplicative and rank-based cross-sample comparisons, and DA results will be unaffected.
+Importantly, however, if the mean efficiency is approximately constant across samples, then bias has a negligible effect on multiplicative and rank-based DA results.
 If the mean efficiency varies but is unassociated with the condition of interest, then bias merely serves to increase noise and does not create systematic errors in DA results.
 It is only when the mean efficiency is associated with the condition being analyzed that large systematic errors can occur.
-Our case studies suggest that this problematic scenario can occur but may well be the exception rather than the rule.
-Thus our results are still consistent with taxonomic bias having had a limited impact on extant DA results.
+Our case studies suggest that this problematic scenario does occur, but it may be the exception rather than the rule.
 Systematic investigation of how the mean efficiency affects DA results across a wide range of studies may increase our confidence in previous DA results and/or alert us to the conditions in which they are most suspect.

### 2022-08-19 differential-abundance.Rmd
Make edits from AW and BC

@@
 
-Taxonomic bias affects the measured FDs in a species' absolute abundance in a similar fashion.
+Taxonomic bias affects measured FDs in a species' absolute abundance in a similar fashion.
 First, consider a species' abundance measurement $\tilde A_i^{(a)}$ derived from a non-MGS measurement of total-community abundance using Equation \@ref(eq:density-prop-meas).
@@
 This ratio can vary, creating error in the measured FD between samples.
-Notably, if the FDs in the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, then the two offset each other and lead to more stable FEs (and hence more accurate FDs) in $\tilde A_i^{(a)}$.
+A notable special case is if the FDs in the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement; then, the two will offset each other and lead to more stable FEs (and hence more accurate FDs) in $\tilde A_i^{(a)}$.
 We discuss how this possibility might be exploited in real experimental workflows in Section \@ref(solutions).
@@
 Equation \@ref(eq:density-ratio-error) indicates that the FE equals a constant ratio in species' efficiencies multiplied by the FE in the measurement of the reference species' abundance.
-If the abundance of the reference species can determined up to constant FE, then the FE in $\tilde A_i^{(a)}$ will also be constant, leading to accurate FDs.
+If the abundance of the reference species can be determined up to constant FE across samples, then the FE in $\tilde A_i^{(a)}$ will also be constant, leading to accurate FDs.
 
@@
 
-Although researchers typically use more sophisticated models, the simple linear regression model in Equation \@ref(eq:regression) provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
-Appendix \@ref(appendix-regression) describes a general framework for finding the error in the estimated slope coefficient $\beta$ caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures decribed in Section \@ref(abundance-measurement).
+Although researchers often use more sophisticated models, the simple linear regression model in Equation \@ref(eq:regression) provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
+Appendix \@ref(appendix-regression) describes a general framework for finding the error in the estimated slope coefficient $\beta$ caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures described in Section \@ref(abundance-measurement).
 In each case, a multiplicative DA analysis can be conducted by setting $y$ equal to the logarithm of the untransformed abundance measurement ($\tilde P_i$, $\tilde R_{i/j}$, or $\tilde A_i$).
@@
 A third possibility is that $\bar B$ varies in an essentially random manner that is independent of the covariate $x$.
-In this scenario, the error from taxonomic bias in the measurement $\log \tilde A_{i}$ acts to increase the noise in $\hat beta$ (i.e., increase its standard error), but does not cause a systematic error.
+In this scenario, the error from taxonomic bias in the measurement $\log \tilde A_{i}$ acts to increase the noise in $\hat \beta$ (i.e., increase its standard error), but does not cause a systematic error.
 

### 2022-08-19 introduction.Rmd
Make edits from AW and BC

@@
 In some cases, bias can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
-Yet @mclaren2019cons also found that other abundance measures---those based on the ratios among species---have proportional errors and may lead to more robust DA analyses.
+Yet @mclaren2019cons also found that other abundance measures---those based on the ratios among species---have constant FEs and may lead to more robust DA analyses.
 The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.
@@
 Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
-We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods (for relative and absolute abundance) that are based on species proportions.
+We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods that are based on species proportions, even if bias is the same across samples.
 We describe the theoretical conditions when this effect is negligible and when it may cause serious scientific errors, and explore this effect in case studies based on real microbiome experiments.
-We further demonstrate that another set of DA methods, based on the ratios among species, are robust to bias.
+We further demonstrate that another set of DA methods, based on the ratios among species, are robust to consistent bias.
 Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.
-Application of these insights and methods can build the confidence needed to turn the findings of microbiome studies into readily-translatable scientific knowledge.
+These insights and methods will aid microbiome data analysts in turning the findings of microbiome studies into readily-translatable scientific knowledge.

### 2022-08-19 solutions.Rmd
Make edits from AW and BC

@@
-# Potential solutions {#solutions}
+# Solutions {#solutions}
 
@@
 Second, the bias invariance of ratios among species does not extend to additive aggregates of species into higher-order taxa such phyla unless bias is conserved within the species group (@mclaren2019cons).
-Multiplicative aggregates of species, recently used for biomarker discovery (@riverapinto2018bala, @quinn2020inte), provide an alternative that remains bias-invariant but may be harder to interpret.
+Multiplicative aggregates of species, recently used for biomarker discovery (@riverapinto2018bala, @quinn2020inte), provide an alternative that remains bias-invariant but are harder to interpret.
 
@@
 For example, @hevroni2020seas examined how the within-sample ranks of viral species changed from summer to winter.
-Since changes in within-sample ranks can be affected by bias, the authors also considered the (bias-invariant) changes in the centered log ratios (CLR) associated with each species.
+Since changes in within-sample ranks can be affected by bias, the authors also considered the changes in the centered log ratios (CLR) associated with each species.
 They found that the species whose within-sample ranks increased also increased in their CLR value, providing evidence that their initial results were not driven by bias.
@@
 
-_Community calibration controls_ are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
+_Community calibration controls_ are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons, @clausen2022mode).
 Inclusion of these samples along with the primary samples in an MGS experiment can enable researchers to calibrate their MGS measurement by directly measuring and removing the effect of bias.
@@
 Measurement of these controls across multiple experiments enables characterization of the _differential bias_ between experiments (@mclaren2019cons).
-Calibration using differential bias can make results directly comparable across studies that used different.
+Calibration using differential bias can make results directly comparable across studies that used different protocols.
 
@@
 
-## Choose absolute-abundance methods with more stable FEs
+## Absolute-abundance methods with more stable FEs
 
-There are many methods for obtaining species (absolute) abundances from the relative abundances measured by MGS, which had previously not been evaluated for their sensitivity to taxonomic bias (@williamson2021amul being a notable exception).
+There are many methods for obtaining absolute abundances from the relative abundances measured by MGS, but these methods have not previously been evaluated for their sensitivity to taxonomic bias (@williamson2021amul being a notable exception).
 Our theoretical results show that these methods differ in the extent to which taxonomic bias causes the FEs in species abundances to vary across samples.
@@
 Thus the 16S qPCR measurement is subject to bias from extraction, copy-number variation, primer binding, and amplification, just like the 16S sequencing measurement.
-Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it can lead to more accurate FDs across samples.
+Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it will lead to more accurate FDs---and thus more accurate DA results.
 
-More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement may be qPCR the same marker gene, from the same DNA extraction.
+More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement is qPCR of the same marker gene, from the same DNA extraction.
 Similar reasoning suggests that for shotgun sequencing, the ideal total-abundance measurement is bulk DNA quantification: Shotgun sequencing and bulk DNA quantification are both subject to bias from extraction and variation in genome size.
@@
 
-A second approach to obtaining species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \@ref(eq:density-ratio-meas)).
-Our theory predicts that taxonomic bias induces constant FEs in the resulting species abundances which do not affect DA results.
+A second approach to obtaining absolute species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \@ref(eq:density-ratio-meas)).
+Our theory predicts that taxonomic bias induces constant FEs in the abundances obtained by this approach, which will not affect DA results.
 
@@
 Meta-analysis of microbiome samples measured across multiple studies must contend with the fact that different studies typically use different protocols and hence have different taxonomic biases.
-These different biases can be explicitly accounted for in a _bias meta-analysis_ which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
+These different biases can be explicitly accounted for in a _bias-aware meta-analysis_ which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
 Parametric meta-analysis models include study-specific latent parameters representing 'batch effects'—non-biological differences in the data from each study which can distort the observed biological patterns.

### 2022-08-19 v/3bde34f650d7c06e05bd5957ebca2a7857a4c0ed/_main.tex
Make edits from AW and BC
This commit adds the rendered manuscript associated with
3bde34f650d7c06e05bd5957ebca2a7857a4c0ed

--- 2022-07-17 v/9c86fbcc416ebe9d18d2e80120064bea4c346f36/_main.tex
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis},
-  pdfauthor={Michael R. McLaren; Jacob T. Nearing; Amy D. Willis; Karen G. Lloyd; Benjamin J. Callahan},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
@@
 \fi
-
-\title{Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Implications of taxonomic bias for microbial differential-abundance analysis},
+  pdfauthor={Michael R. McLaren; Jacob T. Nearing; Amy D. Willis; Karen G. Lloyd; Benjamin J. Callahan},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+
+\title{Implications of taxonomic bias for microbial differential-abundance analysis}
 \author{Michael R.
@@
 Callahan\footnote{North Carolina State University}}
-\date{2022-07-17}
+\date{2022-08-18}
 
@@
 }
-\hypertarget{preface}{%
-\section*{Preface}\label{preface}}
-\addcontentsline{toc}{section}{Preface}
-
-\begin{verbatim}
-#> working directory clean
-\end{verbatim}
-
-\emph{This manuscript was rendered from commit 9c86fbcc416ebe9d18d2e80120064bea4c346f36.}
-
-\leavevmode\vadjust pre{\hypertarget{preface-warning}{}}%
-\textbf{This in-progress manuscript is not intended for general scientific use.}
-It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
-Please post comments or questions on the \href{https://github.com/mikemc/differential-abundance-theory/issues}{GitHub Issues page} or \href{m.mclaren42@gmail.com}{email Mike}.
-
-This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.
-We describe the basic problem posed by taxonomic bias for measuring changes in the abundance of particular taxa across conditions and describe new strategies for mitigating the errors it induces.
-Analyses of both relative and absolute abundances are considered.
-The manuscript is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
-See \href{https://doi.org/10.5281/zenodo.4552717}{the Zenodo record} for how to cite the latest version.
-Supporting data analyses can be found in the accompanying \href{https://mikemc.github.io/differential-abundance-theory/notebook/}{computational research notebook}.
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-\leavevmode\vadjust pre{\hypertarget{abstract}{}}%
-\textbf{Abstract:}
+\hypertarget{abstract}{%
+\section*{Abstract}\label{abstract}}
+\addcontentsline{toc}{section}{Abstract}
 
@@
 Wide adoption of our recommendations can improve the reproducibility, interpretability, and translatability of microbiome DA studies.
+
+\leavevmode\vadjust pre{\hypertarget{preface-note}{}}%
+This manuscript was rendered from commit 3bde34f of \url{https://github.com/mikemc/differential-abundance-theory}.
+Supporting data analyses can be found in the accompanying \href{https://mikemc.github.io/differential-abundance-theory/notebook/}{computational research notebook}.
+Please post comments or questions on \href{https://github.com/mikemc/differential-abundance-theory}{GitHub}.
+The manuscript is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0 License}.
+See the \href{https://github.com/mikemc/differential-abundance-theory/releases}{GitHub Releases} or \href{https://doi.org/10.5281/zenodo.4552717}{Zenodo record} for earlier versions.
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 In some cases, bias can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
-Yet McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that other abundance measures---those based on the ratios among species---have proportional errors and may lead to more robust DA analyses.
+Yet McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) also found that other abundance measures---those based on the ratios among species---have constant FEs and may lead to more robust DA analyses.
 The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.
@@
 Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
-We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods (for relative and absolute abundance) that are based on species proportions.
+We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods that are based on species proportions, even if bias is the same across samples.
 We describe the theoretical conditions when this effect is negligible and when it may cause serious scientific errors, and explore this effect in case studies based on real microbiome experiments.
-We further demonstrate that another set of DA methods, based on the ratios among species, are robust to bias.
+We further demonstrate that another set of DA methods, based on the ratios among species, are robust to consistent bias.
 Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.
-Application of these insights and methods can build the confidence needed to turn the findings of microbiome studies into readily-translatable scientific knowledge.
+These insights and methods will aid microbiome data analysts in turning the findings of microbiome studies into readily-translatable scientific knowledge.
 
@@
 
-This section extends the theoretical results of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to describe the effect that consistent taxonomic bias within an MGS experiment affects the relative and absolute abundances measured for various microbial species.
+This section extends the theoretical results of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to describe how taxonomic bias in an MGS experiment affects the relative and absolute abundances measured for various microbial species.
 We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.
 
-\hypertarget{model-of-mgs-measurement}{%
-\subsection{Model of MGS measurement}\label{model-of-mgs-measurement}}
+\hypertarget{a-model-of-mgs-measurements}{%
+\subsection{A model of MGS measurements}\label{a-model-of-mgs-measurements}}
 
@@
 We distinguish between two types of species-level \emph{relative abundances} within a sample.
-The \emph{proportion} \(P_{i}^{(a)}\) of species \(i\) in sample \(a\) equals its abundance relative to the total abundance of all species in \(S\),
+The \emph{proportion} \(P_{i}^{(a)}\) of species \(i\) in sample \(a\) equals its abundance divided by the total abundance of all species in \(S\),
 \begin{align}
@@
 \end{align}
-Taxonomic bias thus creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
+Taxonomic bias creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
 Since the mean efficiency varies across samples, so does the FE.
@@
 \end{align}
-Taxonomic bias thus creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
-For instance, in Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
+Taxonomic bias creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
+For instance, in Figure \ref{fig:error-proportions}, the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-measured by a factor of 6 in both communities despite their varying compositions.
 A demonstration in bacterial mock communities is shown in \href{https://doi.org/10.7554/eLife.46923.004}{Figure 3D} of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}).
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/export-pdf/figures/illustrations/error-proportions} \caption{\textbf{Taxonomic bias creates fold errors in species proportions that vary across samples and lead to inaccurate fold differences between samples.} Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species.
+\includegraphics[width=1\linewidth]{figures/export-pdf/figures/illustrations/error-proportions} \caption{\textbf{Taxonomic bias creates fold errors in species proportions that vary across samples and lead to inaccurate fold differences between samples.} Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species.
 Bottom row: Error in the measured fold difference in the third species that is derived from these measurements.
@@
 \end{align}
-With these definitions, the read count for taxon \(I\) can be expressed as
+With these definitions, the read count for higher-order taxon \(I\) can be expressed as
 \(M_{I}^{(a)} = A_{I}^{(a)} B_I^{(a)} F^{(a)}\).
-Thus \(B_I^{(a)}\) plays a role analogous to the efficiency of an individual species, but differs in that it need not be constant across samples:
-If the constituent species have different efficiencies, then the efficiency of the higher-order taxon \(I\) depends on the relative abundances of its constituents and so will tend to vary across samples (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+Thus \(B_I^{(a)}\) plays a role analogous to the efficiency of an individual species, but differs in that it is not constant across samples:
+If the constituent species have different efficiencies, then the efficiency of the higher-order taxon \(I\) depends on the relative abundances of its constituents and so will vary across samples (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
 As an example, suppose that Species 1 and Species 2 in Figure \ref{fig:error-proportions} were in the same phylum.
@@
 The first approach leverages information about the abundance of the total community; for example, Vandeputte et al.
-(\protect\hyperlink{ref-vandeputte2017quan}{2017}) measured total-community abundance using flow cytometry and multiplied this number by MGS genus proportions to obtain the absolute abundances of individual genera (Vandeputte et al.
+(\protect\hyperlink{ref-vandeputte2017quan}{2017}) measured total-community abundance using flow cytometry and multiplied this number by genus proportions measured by MGS to quantify the absolute abundances of individual genera (Vandeputte et al.
 (\protect\hyperlink{ref-vandeputte2017quan}{2017})).
@@
 Suppose that the total abundance of all species in the sample, \(A_{{\text{tot}}}^{(a)}\), has been measured by a non-MGS method, yielding a measurement \(\tilde A_{\text{tot}}^{(a)}\).
-The absolute abundance of an individual species can be measured by multiplying the species' proportion from MGS by this total-abundance measurement,
+The absolute abundance of an individual species can be quantified by multiplying the species' proportion from MGS by this total-abundance measurement,
 \begin{align}
@@
 
-Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
-Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
+Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias that is analogous to, but quantitatively different from, the MGS relative abundance measurements.
+Flow cytometry may yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
 Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency (Lloyd et al.
@@
 
-Species abundance measurements derived by this method are affected by taxonomic bias in both the MGS and total-abundance measurement.
-We can determine the resulting fold error (FE) by substituting Equations \eqref{eq:prop-error} and \eqref{eq:total-density-error} into Equation \eqref{eq:density-prop-meas}, yielding
+Species abundance measurements derived by this method (Equation \eqref{eq:density-prop-meas}) are affected by taxonomic bias in both the MGS and total-abundance measurement.
+We can determine the resulting fold error (FE) in the estimate \(\tilde A_i^{(a)}\) by substituting Equations \eqref{eq:prop-error} and \eqref{eq:total-density-error} into Equation \eqref{eq:density-prop-meas}, yielding
 \begin{align}
@@
 Equation \eqref{eq:density-prop-error} indicates that the FE in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
-As in the case of proportions (Equation \eqref{eq:prop-error}), the FE depends on sample composition through the two mean efficiency terms and so will vary across samples unless the two perfectly covary.
+As in the case of proportions (Equation \eqref{eq:prop-error}), the FE depends on sample composition through the two mean efficiency terms and so will, in general, vary across samples.
 
@@
 In this case, the reference abundance \(\tilde A_r^{(a)}\) is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
-(In practice, researchers in spike-in experiments have typically used a more indirect calculation to determine species abundances, but which yields identical results to \eqref{eq:density-ratio-meas}; Appendix TODO).
 
@@
 (\protect\hyperlink{ref-morella2018rapi}{2018})), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
-Appendix TODO describes how using multiple reference species and/or statistical modeling can address the fact that any one native reference species is unlikely to be found in all samples.
 
@@
 
-Taxonomic bias affects the measured FDs in a species' absolute abundance in a similar fashion.
+Taxonomic bias affects measured FDs in a species' absolute abundance in a similar fashion.
 First, consider a species' abundance measurement \(\tilde A_i^{(a)}\) derived from a non-MGS measurement of total-community abundance using Equation \eqref{eq:density-prop-meas}.
@@
 This ratio can vary, creating error in the measured FD between samples.
-Notably, if the FDs in the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, then the two offset each other and lead to more stable FEs (and hence more accurate FDs) in \(\tilde A_i^{(a)}\).
+A notable special case is if the FDs in the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement; then, the two will offset each other and lead to more stable FEs (and hence more accurate FDs) in \(\tilde A_i^{(a)}\).
 We discuss how this possibility might be exploited in real experimental workflows in Section \ref{solutions}.
@@
 Equation \eqref{eq:density-ratio-error} indicates that the FE equals a constant ratio in species' efficiencies multiplied by the FE in the measurement of the reference species' abundance.
-If the abundance of the reference species can determined up to constant FE, then the FE in \(\tilde A_i^{(a)}\) will also be constant, leading to accurate FDs.
+If the abundance of the reference species can be determined up to constant FE across samples, then the FE in \(\tilde A_i^{(a)}\) will also be constant, leading to accurate FDs.
 
@@
 
-Although researchers typically use more sophisticated models, the simple linear regression model in Equation \eqref{eq:regression} provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
-Appendix \ref{appendix-regression} describes a general framework for finding the error in the estimated slope coefficient \(\beta\) caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures decribed in Section \ref{abundance-measurement}.
+Although researchers often use more sophisticated models, the simple linear regression model in Equation \eqref{eq:regression} provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
+Appendix \ref{appendix-regression} describes a general framework for finding the error in the estimated slope coefficient \(\beta\) caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures described in Section \ref{abundance-measurement}.
 In each case, a multiplicative DA analysis can be conducted by setting \(y\) equal to the logarithm of the untransformed abundance measurement (\(\tilde P_i\), \(\tilde R_{i/j}\), or \(\tilde A_i\)).
@@
 A third possibility is that \(\bar B\) varies in an essentially random manner that is independent of the covariate \(x\).
-In this scenario, the error from taxonomic bias in the measurement \(\log \tilde A_{i}\) acts to increase the noise in \(\hat beta\) (i.e., increase its standard error), but does not cause a systematic error.
+In this scenario, the error from taxonomic bias in the measurement \(\log \tilde A_{i}\) acts to increase the noise in \(\hat \beta\) (i.e., increase its standard error), but does not cause a systematic error.
 
@@
 In practice it is difficult to construct control communities that span the many species present in complex natural communities.
-However, gnotobiotic community experiments are well suited to this form of \emph{calibration via community controls} since, unlike most natural ecosystems, it is possible to assemble mock communities containing all species in known relative abundances.
+However, gnotobiotic community experiments are well suited to this form of \emph{calibration via community controls} since it is possible to assemble mock communities containing all species in known relative abundances.
 
@@
 Despite the 13-fold variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \ref{fig:leopold2020host-variation}C), having a geometric range of 1.62 and a geometric standard deviation of 1.05.
-This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \ref{fig:leopold2020host-variation}A).
+This consistency in the mean efficiency was despite the fact that each species showed substantial multiplicative variation (SI Figure \ref{fig:leopold2020host-variation}A).
 But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3-fold and by just 1.5-fold between the two most dominant.
@@
 Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5-fold to 5.2-fold.
-Using gamma-Poisson (or negative-binomial) regression of the observed read counts, we estimated the average LFD in commensal species proportions following infection with and without calibration (SI Figure \ref{fig:leopold2020host-infection-lfc}).
-Commensals typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
+Using gamma-Poisson regression of the observed read counts, we estimated the average LFD in commensal species proportions following infection with and without calibration (SI Figure \ref{fig:leopold2020host-infection-lfc}).
+The commensals' proportions typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
 However, failure to account for bias caused the decrease to be overestimated, by an amount corresponding to the inverse change in log mean efficiency.
@@
 (\protect\hyperlink{ref-huttenhower2012stru}{2012}), \href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-30-hmp-stool-vagina-comparison/}{SI Analysis}).
-On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \ref{fig:gut}A; GM \(\md\) GSD of 6.3 \(\md\) 1.7 in stool samples and 1.4 \(\md\) 1.6 in vaginal samples).
+On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \ref{fig:gut}A; geometric mean (GM) \(\md\) geometric standard deviation (GSD) of 6.3 \(\md\) 1.7 in stool samples and 1.4 \(\md\) 1.6 in vaginal samples).
 To assess the potential importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the multiplicative variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth.
 Across simulation replicates, the GSD in the mean efficiency was typically lower in stool than in vaginal samples (SI Figure \ref{fig:gut}B; ratio of GSD in vaginal samples to GSD in stool samples of 1.6 \(\md\) 1.5 across 1000 simulations).
-Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \ref{fig:gut}C), which the GSD in the proportion of a gut species across gut samples an average of 1.75-fold lower than that of a vaginal species across vaginal samples.
+Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \ref{fig:gut}C); the GSD in the proportion of a gut species across gut samples was an average of 1.75-fold lower than that of a vaginal species across vaginal samples.
 Recall that the importance of bias estimation of the FD in a species' proportion depends on GSD in the mean efficiency relative the GSD in the species' proportion (Section \ref{differential-abundance}).
@@
 \hypertarget{solutions}{%
-\section{Potential solutions}\label{solutions}}
+\section{Solutions}\label{solutions}}
 
@@
 Multiplicative aggregates of species, recently used for biomarker discovery (Rivera-Pinto et al.
-(\protect\hyperlink{ref-riverapinto2018bala}{2018}), Quinn and Erb (\protect\hyperlink{ref-quinn2020inte}{2020})), provide an alternative that remains bias-invariant but may be harder to interpret.
+(\protect\hyperlink{ref-riverapinto2018bala}{2018}), Quinn and Erb (\protect\hyperlink{ref-quinn2020inte}{2020})), provide an alternative that remains bias-invariant but are harder to interpret.
 
@@
 For example, Hevroni et al. (\protect\hyperlink{ref-hevroni2020seas}{2020}) examined how the within-sample ranks of viral species changed from summer to winter.
-Since changes in within-sample ranks can be affected by bias, the authors also considered the (bias-invariant) changes in the centered log ratios (CLR) associated with each species.
+Since changes in within-sample ranks can be affected by bias, the authors also considered the changes in the centered log ratios (CLR) associated with each species.
 They found that the species whose within-sample ranks increased also increased in their CLR value, providing evidence that their initial results were not driven by bias.
@@
 
-\emph{Community calibration controls} are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen `gold standard' or \emph{reference protocol} (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+\emph{Community calibration controls} are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen `gold standard' or \emph{reference protocol} (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}), Clausen and Willis (\protect\hyperlink{ref-clausen2022mode}{2022})).
 Inclusion of these samples along with the primary samples in an MGS experiment can enable researchers to calibrate their MGS measurement by directly measuring and removing the effect of bias.
@@
 Measurement of these controls across multiple experiments enables characterization of the \emph{differential bias} between experiments (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
-Calibration using differential bias can make results directly comparable across studies that used different.
+Calibration using differential bias can make results directly comparable across studies that used different protocols.
 
@@
 
-\hypertarget{choose-absolute-abundance-methods-with-more-stable-fes}{%
-\subsection{Choose absolute-abundance methods with more stable FEs}\label{choose-absolute-abundance-methods-with-more-stable-fes}}
-
-There are many methods for obtaining species (absolute) abundances from the relative abundances measured by MGS, which had previously not been evaluated for their sensitivity to taxonomic bias (Williamson, Hughes, and Willis (\protect\hyperlink{ref-williamson2021amul}{2021}) being a notable exception).
+\hypertarget{absolute-abundance-methods-with-more-stable-fes}{%
+\subsection{Absolute-abundance methods with more stable FEs}\label{absolute-abundance-methods-with-more-stable-fes}}
+
+There are many methods for obtaining absolute abundances from the relative abundances measured by MGS, but these methods have not previously been evaluated for their sensitivity to taxonomic bias (Williamson, Hughes, and Willis (\protect\hyperlink{ref-williamson2021amul}{2021}) being a notable exception).
 Our theoretical results show that these methods differ in the extent to which taxonomic bias causes the FEs in species abundances to vary across samples.
@@
 Thus the 16S qPCR measurement is subject to bias from extraction, copy-number variation, primer binding, and amplification, just like the 16S sequencing measurement.
-Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it can lead to more accurate FDs across samples.
-
-More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement may be qPCR the same marker gene, from the same DNA extraction.
+Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it will lead to more accurate FDs---and thus more accurate DA results.
+
+More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement is qPCR of the same marker gene, from the same DNA extraction.
 Similar reasoning suggests that for shotgun sequencing, the ideal total-abundance measurement is bulk DNA quantification: Shotgun sequencing and bulk DNA quantification are both subject to bias from extraction and variation in genome size.
@@
 
-A second approach to obtaining species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \eqref{eq:density-ratio-meas}).
-Our theory predicts that taxonomic bias induces constant FEs in the resulting species abundances which do not affect DA results.
+A second approach to obtaining absolute species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \eqref{eq:density-ratio-meas}).
+Our theory predicts that taxonomic bias induces constant FEs in the abundances obtained by this approach, which will not affect DA results.
 
@@
 Meta-analysis of microbiome samples measured across multiple studies must contend with the fact that different studies typically use different protocols and hence have different taxonomic biases.
-These different biases can be explicitly accounted for in a \emph{bias meta-analysis} which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
+These different biases can be explicitly accounted for in a \emph{bias-aware meta-analysis} which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
 Parametric meta-analysis models include study-specific latent parameters representing `batch effects'---non-biological differences in the data from each study which can distort the observed biological patterns.
@@
 
+\clearpage
+
 \hypertarget{appendix-appendix}{%
 \appendix \addcontentsline{toc}{section}{\appendixname}}
-
-\hypertarget{mgs-measurement-details}{%
-\section{MGS measurement details}\label{mgs-measurement-details}}
-
-\hypertarget{spike-in-alt}{%
-\subsection{Alternative method for measuring species absolute abundances from a spike-in}\label{spike-in-alt}}
-
-(stub)
-
-\hypertarget{multiple-references}{%
-\subsection{Missing and multiple reference species}\label{multiple-references}}
-
-(stub)
 
@@
 \caption{\label{tab:statistical-notation} Statistical notation used in this section.}\tabularnewline
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule
+\midrule()
 \endfirsthead
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule
+\midrule()
 \endhead
@@
 \(\text{se}(\hat \beta)\), \(\hat{\text{se}}(\hat \beta)\) & & Standard error of \(\hat \beta\) and its estimate \\
-\bottomrule
+\bottomrule()
 \end{longtable}
@@
 
-\hypertarget{diversity-and-mean-efficiency}{%
-\section{Alpha diversity and variation in the mean efficiency}\label{diversity-and-mean-efficiency}}
-
-The diversity of order \(q\) is given by
-\begin{align}
-  ^qD = \left(\sum_{i=1}^n p_i^q\right)^{1 /(1-q)},
-\end{align}
-where \(^qD\) is understood to be given by the limit of the RHS when \(q = 1\) (and so the given expression is undefined).
-The order-2 diversity is therefore
-\begin{align}
-  ^2D = \frac{1}{\sum_{i=1}^n {p_i^2}},
-\end{align}
-which is equivalent to the Inverse Simpson Index (REF Jost).
-
-Consider an infinite pool of species.
-Let \(\sigma^{2}\) denote the variance in the relative efficiency among species in the pool.
-Consider a community that is assembled by randomly choosing \(K \ge 1\) species from the pool and setting the abundances of the \(K\) species in a manner that is independent of their efficiencies.
-
-\textbf{Claim:}
-Let \(\rho_{k}\) for \(1 \le k \le K\) denote the proportions of the \(K\) species in the community; this new notation serves as a reminder that the subscript \(k\) indexes a random species specific to this particular community.
-Conditional on the order-2 diversity \(^2D\) of a community, the arithmetic variance in the mean efficiency is
-\begin{align}
-  Var[\bar B \mid ^2D] 
-% = \sigma^2 \sum_{k=1}^K \rho_k^2
-  = \frac{\sigma^2}{^2D}.
-\end{align}
-
-\textbf{Proof:} Let \(\beta_{k}\) denote the efficiency of the \(k\)-th sampled species.
-The mean efficiency is therefore \(\bar B = \sum_k \rho_k \beta_k\).
-Conditional on the proportions \(\{\rho_{k}\}\), the variance in the mean efficiency is
-\begin{align}
-  Var[\bar B \mid \{\rho_{k}\}]
-  &= \sum_{k=1}^K \rho_k^2 Var[\beta_k]
-  \\&= \left(\sum_{k=1}^K p_k^2\right) \sigma^2.
-\end{align}
-The first line follows from the fact that the \(\beta_{k}\) are independent.
-The summation in the final line is equal to \(1/2^D\), thus proving the result.
-
-\textbf{Note:}
-We have showed that the arithmetic (additive) variance of \(\bar B\) decreases with \(^2D\); however, the geometric (multiplicative) variance is most relevant for understanding the effect of bias on DA.
-As \(2^D\) increases, the distribution of \(\bar B\) will converge (by the central limit theorem) to a normal distribution, and both the arithmetic and geometric variance will decrease.
+\clearpage
 
@@
 \end{figure}
+
+\clearpage
 
@@
 \url{https://doi.org/10.1016/j.diagmicrobio.2018.05.022}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-clausen2022mode}{}}%
+Clausen, David S, and Amy D Willis. 2022. {``Modeling Complex Measurement Error in Microbiome Experiments,''} April. \url{http://arxiv.org/abs/2204.12733}.
 

### 2022-10-08 solutions.Rmd
Edit solutions section
- Add missing ref to ratio subsection
- Minor revision of paragraph(s) on targeted measurement of reference taxa

@@
 
-Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using DA methods that analyze fold differences (FDs) in these ratios.
-A variety of ratio-based methods have been developed under the framework of Compositional Data Analysis (CoDA; REFs).
+Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using relative DA methods that analyze fold differences (FDs) in these ratios.
+A variety of ratio-based methods have been developed under the framework of Compositional Data Analysis (CoDA; @gloor2017micr).
 CoDA methods may more generally consider the ratios between products of multiple species, possibly raised to some exponent; a simple example is the geometric mean of several species.
@@
 Section \@ref(absolute-abundance) further proposed an additional class of reference species for normalizing MGS measurements: Species whose (absolute) abundance has been measured using a targeted method such as qPCR with species-specific primers.
+Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus expanding the applicability of reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
+The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
 To demonstrate the ability for reference calibration to reduce error in FD measurements, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
 Doing so improved the resulting measurements of FDs in the abundance of all species (Figure \@ref(fig:calibration-example)).
-Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus expanding the applicability of reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
-The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
-In many ecosystems, there may not be a single universally-present species that can serve as a reference in all samples.
-Sample coverage can be increased by making targeted measurements of multiple species or of larger taxonomic groups; see Appendix REF for details.
+
+In some ecosystems, there may not be a single universally-present species that can serve as a reference in all samples.
+In such cases, sample coverage can be increased by performing targeted measurement of multiple species with complementary prevalence patterns, using appropriate statistical models to combine information across multiple measurements and impute the appropriate correction factor when all reference taxa are below the detection limit in the targeted or MGS measurement.
+Sample coverage can alternatively increased by targeting a larger taxonomic group (such as a genus or family); however, these higher-order taxa are less likely to have a consistent efficiency across samples (Section \ref{relative-abundance}).
 

### 2022-10-08 notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example.Rmd
Revise figure of regression simulation

@@
 
-species <- str_c("Species ", seq(n_species))
+species <- str_c("Sp. ", seq(n_species))
 ```
@@
   ggplot(aes(x, log2_abundance, color = type)) +
-  labs(y = "Log abundance ") +
+  labs(y = "Log efficiency", x = "Condition (x)") +
   facet_grid(.otu~type, scales = "fixed") +
@@
   # stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
-  stat_summary(geom = "point") +
-  stat_smooth(aes(x = as.integer(x)),
+  stat_summary(geom = "point", fun.data = mean_se) +
+  stat_smooth(aes(x = as.integer(x)), formula = y~x,
     method = "lm", size = 0.9, fill = 'grey', se = FALSE
@@
   ggplot(aes(x, log2(mean_efficiency))) +
-  labs(y = "Log efficiency ") +
+  labs(y = "Log efficiency", x = "Condition (x)") +
   geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
   # stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
-  stat_summary(geom = "point") +
-  stat_smooth(aes(x = as.integer(x)),
+  stat_summary(geom = "point", fun.data = mean_se) +
+  stat_smooth(aes(x = as.integer(x)), formula = y~x,
     method = "lm", size = 0.9, color = 'black', fill = 'grey', se = FALSE
@@
 delta <- lm_results_mean_eff %>% filter(term == "x1") %>% pull(estimate)
-start <- lm_results_slope %>% filter(type == 'Actual', .otu == 'Species 9') %>%
+start <- lm_results_slope %>% filter(type == 'Actual', .otu == 'Sp. 9') %>%
   pull(estimate)
@@
   ggplot(aes(y = .otu, x = estimate, color = type)) +
-  labs(x = "Log fold difference", y = NULL) +
+  labs(x = "Mean LFD", y = NULL, color = "Type") +
   geom_vline(xintercept = 0, color = "grey") +
@@
   annotate(
-    geom = 'segment', color = "darkgrey",
+    geom = 'segment', color = "darkred",
     arrow = grid::arrow(length = unit(0.1, "inches")), 
     x = start, xend = start - delta, 
-    y = 10.5, yend = 10.5
+    y = 10.5, yend = 10.5,
+    size = .7
+  ) +
+  annotate(
+    geom = 'text', color = "darkred",
+    label = 'Effect of bias',
+    x = start - delta/2,
+    y = 10.8,
   ) +
@@
   )) +
-  labs(x = "Log fold difference", y = NULL) +
+  labs(x = "Mean LFD", y = NULL) +
   geom_vline(xintercept = 0, color = "grey") +
@@
   annotate(
-    geom = 'segment', color = "darkgrey",
+    geom = 'segment', color = "darkred",
     arrow = grid::arrow(length = unit(0.1, "inches")), 
     x = 0, xend = delta, 
-    y = 1.0, yend = 1.0
+    y = 1.0, yend = 1.0,
+    size = .7
   )
@@
   ggplot(aes(y = type, x = estimate, fill = type)) +
-  labs(x = "Log fold difference", y = "Type") +
+  labs(x = "Mean LFD", y = "Type") +
   geom_vline(xintercept = 0, color = "grey") +
@@
 ```{r}
-species_to_plot <- str_c("Species ", c('9', '10', '5', '7', '2'))
+species_to_plot <- str_c("Sp. ", c('9', '10', '5', '7', '2'))
 p_species_focal <- p_species_all
@@
   (p_species_focal + ggtitle("Actual and measured\nabundances of select species")) + 
-  (p_coef_ci1 + ggtitle("Estimated LFDs of all species") +
+  (p_coef_ci1 + ggtitle("Mean LFD estimated from\nactual or measured abundances") +
     theme(legend.box.margin = margin(b = -15))
@@
 Perhaps the standard errors being returned by R are using the MLE estimate of sigma instead of the OLS estimate?
+
+## 2022-10-08 Modified main figure
+
+New panel showing simulated efficiencies and mean LFDs, to replace the panel showing the change in mean efficiency (former panel B).
+
+```{r}
+p_params <- species_params %>%
+  ggplot(aes(x1, log2_efficiency)) +
+  theme(axis.line = element_blank()) +
+  # panel_border(remove = TRUE) +
+  # theme_minimal_grid() +
+  coord_cartesian(clip = 'off') +
+  geom_vline(xintercept = 0, color = 'grey') +
+  geom_hline(yintercept = 0, color = 'grey') +
+  geom_text(
+    aes(label = str_c('Sp. ', str_extract(.otu, '[0-9]+'))), 
+    size = 4) +
+  labs(
+    x = 'Expected LFD',
+    y = 'Log efficiency'
+  )
+p_params 
+```
+
+Annotate mean efficiency plot,
+
+```{r}
+# params for arrow showing change
+arrow_params <- mean_eff %>%
+  with_groups(x, summarize, across(log2_mean_efficiency, mean)) %>%
+  pull(log2_mean_efficiency)
+
+p_mean_eff2 <- p_mean_eff +
+  annotate(
+    geom = 'segment', color = "darkred",
+    arrow = grid::arrow(length = unit(0.1, "inches")), 
+    x = 2.5, xend = 2.5, 
+    y = arrow_params[1], 
+    yend = arrow_params[2], 
+    size = .7
+  )
+p_mean_eff2 
+```
+
+Assemble with improved titles,
+
+```{r main-figure-2, fig.dim = c(8, 9)}
+
+(p_params + ggtitle("Simulated log efficiencies\nand expected LFDs")) +
+  (p_mean_eff2 + ggtitle("Mean efficiency across\nsamples in each condition")) +
+  # (p_coef_ci_mean_eff + ggtitle("LFD in mean efficiency")) + 
+  (p_species_focal + ggtitle("Actual and measured\nabundances of select species")) + 
+  (p_coef_ci1 + ggtitle("Mean LFD estimated from\nactual or measured abundances") +
+    theme(legend.box.margin = margin(b = -15))
+  ) + 
+  plot_layout(ncol = 2, heights = c(0.4, 1)) +
+  plot_annotation(tag_levels = 'A') &
+  colorblindr::scale_color_OkabeIto() &
+  colorblindr::scale_fill_OkabeIto() &
+  theme(
+    plot.title = element_text(face = "plain")
+  )
+```

### 2022-10-08 abundance-measurement.Rmd
Edit based on AW's comments
Edited based on AW's comments on earlier version (9c86fbc). Mostly minor
fixes, with more substantial improvements to section 3.2 for clarity of
presentation (including changes to main figure).

@@
 
-This section extends the theoretical results of @mclaren2019cons to describe how taxonomic bias in an MGS experiment affects the relative and absolute abundances measured for various microbial species.
-We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.
+This section extends the theoretical results of @mclaren2019cons to describe how taxonomic bias in MGS experiments leads to errors in the relative and absolute abundances measured for various microbial species.
+All approaches to abundance quantification have systematic errors driven by taxonomic bias; however, some yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.
 

### 2022-10-08 case-studies.Rmd
Edit based on AW's comments
Edited based on AW's comments on earlier version (9c86fbc). Mostly minor
fixes, with more substantial improvements to section 3.2 for clarity of
presentation (including changes to main figure).

@@
 
-The taxonomic bias species present in 'mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (@mclaren2019cons).
+The taxonomic bias of species in 'mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (@mclaren2019cons).
 In practice it is difficult to construct control communities that span the many species present in complex natural communities.
@@
 Although @leopold2020host did not include absolute-abundance measurements, we can consider the impact taxonomic bias would have on an absolute DA analysis in a simple scenario in which total genome concentration of each pre- and post-infection sample is perfectly known.
-The total genomic concentrations of each sample, combined with the species proportions measured by MGS, can be combined into a measurement of species concentrations using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
+We consider the approach of multiplying the total genomic concentrations of each sample with the species proportions measured by MGS as described by Equation \@ref(eq:density-prop-meas).
 In this case, the bias in the MGS measurements will create absolute errors in the estimated LFDs of genome concentration of equal magnitude to the errors in the LFD estimates for proportions.
-The scientific error, however, might become worse.
+If total abundances also differed after pathogen growth, however, we may make directional errors in determining the FD.
 Suppose that total abundance increased by 2-fold due to pathogen growth; in this case, species whose proportions remained approximately constant would have increased in absolute abundance by around 2-fold.
@@
 (@huttenhower2012stru, [SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-30-hmp-stool-vagina-comparison/)).
-On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \@ref(fig:gut)A; geometric mean (GM) $\md$ geometric standard deviation (GSD) of `r hmp_div_stats['stool', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['stool', 'gsd'] %>% round(1)` in stool samples and `r hmp_div_stats['vagina', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['vagina', 'gsd'] %>% round(1)` in vaginal samples).
+On average, stool profiles had substantially greater order-2 alpha diversity (Inverse Simpson index) than vaginal profiles (SI Figure \@ref(fig:gut)A; geometric mean (GM) $\md$ geometric standard deviation (GSD) of `r hmp_div_stats['stool', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['stool', 'gsd'] %>% round(1)` in stool samples and `r hmp_div_stats['vagina', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['vagina', 'gsd'] %>% round(1)` in vaginal samples).
 To assess the potential importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the multiplicative variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth.
@@
 The differences in growth rate are small in absolute terms; however, the maximum observed difference of 0.086/yr suggests an error large enough to impact results for some taxa classified as positive growers, whose FRAxC growth rates ranged between 0.04/yr and 0.5/yr.
-In summary, the comparison between FRAxC and qPCR measurements supports the overall study conclusions, but also suggests that species at the low end of positive FRAxC-derived rates may be merely persisting or even slowly declining in abundance.
+In summary, our comparison between FRAxC and qPCR measurements supports the overall study conclusion that many taxa did grow following sediment burial; however, we should remain uncertain as to whether species with small, positive FRAxC-derived growth rates were in fact growing or rather were slowly declining in abundance.
 
@@
 The effect that consistent taxonomic bias has on proportion-based DA analyses depends on the MGS protocol, the biological system, and the sample comparisons under investigation.
-Our case studies explore a limited range of possibilities; however, we can begin to form some general conclusions by placing them within the context of the three scenarios described in Section \@ref(differential-abundance).
+Although our case studies explore a limited range of possibilities, some general patterns stand out.
 
-We saw several cases where the mean efficiency is stable across samples, so that bias does not affect LFD estimates (Scenario 1).
+In several cases, the mean efficiency remained stable across samples, so that LFD esimates were unaffected by bias.
 In the pre-infection fungal microbiome samples of @leopold2020host, we observed that the mean efficiency remained stable despite substantial bias and large multiplicative variation in species proportions among samples.
@@
 Yet we also observed cases where the mean efficiency varied substantially.
-In several cases, the (log) mean efficiency co-varied sufficiently strongly with the condition of interest to cause substantial systematic errors in LFD estimates (Scenario 3).
+In several cases, the (log) mean efficiency co-varied sufficiently strongly with the condition of interest to cause substantial systematic errors in LFD estimates.
 Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency substantially increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
 Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time that is sufficient to substantially inflate the estimated growth rates of slowly changing species.
-The mean efficiency cannot differ by more than the species proportions constituting it.
+The FD in the mean efficiency is bounded by the largest FD of any species' proportion.
 Therefore, the error in the estimated LFDs tend to be practically significant only for species whose LFDs are substantially smaller in magnitude than the largest LFD.
 
-Variation in the mean efficiency may also be substantial, yet unassociated with the condition of interest (Scenario 2).
+Variation in the mean efficiency may also be substantial, yet unassociated with the condition of interest.
 Although we did not directly observe this scenario directly, we have reason to think that it may be common in real microbiome studies.
@@
 These observations provide reasons to both worry and hope.
-It seems likely that in many studies, the mean efficiency is consistent (first scenario) or is at least not associated with the condition (second scenario), so that DA inferences remain valid.
+It seems likely that in many studies, the mean efficiency is consistent (first scenario) or is at least not associated with the condition (second scenario), so LFD estimates remain accurate (or at least not overconfident).
 Yet it is not obvious which scenario any study falls into.

### 2022-10-08 differential-abundance.Rmd
Edit based on AW's comments
Edited based on AW's comments on earlier version (9c86fbc). Mostly minor
fixes, with more substantial improvements to section 3.2 for clarity of
presentation (including changes to main figure).

@@
 For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude of the measured FD (actual FD: 0.15, measured FD: 0.44).
-For Species 2, which decreases but by a smaller factor than the mean efficiency, the species actually appears to increase---a reversal in direction of the measured FD (actual FD: 0.6, measured FD: 1.8).
+For Species 3, which decreases but by a smaller factor than the mean efficiency, the species actually appears to increase---a reversal in direction of the measured FD (actual FD: 0.6, measured FD: 1.8).
 In contrast, the FDs in the ratios among species are identical in the actual and measured taxonomic profiles.
@@
 \end{align}
+A multiplicative DA analysis can be conducted by setting $y$ equal to the logarithm of the untransformed abundance measurement ($\tilde P_i$, $\tilde R_{i/j}$, or $\tilde A_i$).
 For example, $y$ may be the log absolute abundance of species $i$ ($\log A_i^{(a)}$) and $x$ may be a continuous variable (such as pH) or a binary variable (such as $x=1$ for treated patients and $x=0$ for untreated controls).
@@
 For a binary covariate, $\beta$ equals the average difference in $y$ between samples with $x=1$ versus those with $x=0$.
-Taking $y$ to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient $\beta$ corresponding to the average log fold difference (LFD) in abundance between conditions or for a unit increase in $x$.
+When $y$ equals the logarithm of a species' relative or absolute abundance, the coefficient $\beta$ equals the average log fold difference (LFD) in abundance between conditions corresponding to a unit increase in $x$.
+Although researchers often use more sophisticated models, the simple linear regression model in Equation \@ref(eq:regression) provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
 
-Although researchers often use more sophisticated models, the simple linear regression model in Equation \@ref(eq:regression) provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
-Appendix \@ref(appendix-regression) describes a general framework for finding the error in the estimated slope coefficient $\beta$ caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures described in Section \@ref(abundance-measurement).
-In each case, a multiplicative DA analysis can be conducted by setting $y$ equal to the logarithm of the untransformed abundance measurement ($\tilde P_i$, $\tilde R_{i/j}$, or $\tilde A_i$).
-Consider an abundance measure such as the ratio between species $i$ and $j$, $\tilde R_{i/j}$.
-Taxonomic bias creates a constant FE in $\tilde R_{i/j}$ that translates to a constant additive shift in $\log \tilde R_{i/j}$ and therefore does not affect the estimate of the slope $\hat \beta$.
+Appendix \@ref(appendix-regression) describes a general framework for finding the error in the estimated slope coefficient $\hat \beta$ caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures described in Section \@ref(abundance-measurement).
 
-Now consider the absolute abundance of a species $i$ obtained by normalizing its proportion to an non-MGS measurement of total community abundance, which for illustration we assume to be completely accurate.
-The measured abundance $\tilde A_i$ thus has a FE that varies inversely with the mean efficiency $\bar B$.
+The effect that taxonomic bias has on the estimate $\hat \beta$ mirrors its effect on the measured FDs between a pair of individual samples:
+For abundance measures with a FE that is constant across samples, bias does not affect $\hat \beta$.
+For abundance measures where the FE varies, bias can affect $\hat \beta$, with its exact effect depending on how the mean efficiency varies across samples.
+
+We illustrate the constant-FE case by considering a regression analysis of the log ratio between species $i$ and $j$, $\log \tilde R_{i/j}$.
+Taxonomic bias creates a constant FE in $\tilde R_{i/j}$ that translates to a constant additive error in $\log \tilde R_{i/j}$.
+A constant (additive) shift in a response variable only affects the estimated intercept $\hat \alpha$ and not the estimated slope $\hat \beta$. 
+Thus, in this case, taxonomic bias does not affect $\hat \beta$.
+
+We illustrate the varying-FE case by considering the absolute abundance of a species $i$ obtained by multiplying the species' MGS proportion by a non-MGS measure of total-abundance, using Equation \@ref(eq:density-prop-meas).
+For simplicity, suppose that total-community abundance has been measured completely accurately.
+In this case, the measured abundance $\tilde A_i$ has a FE that varies inversely with the mean efficiency $\bar B$.
 If $\bar B$ is stable across samples, then the additive error in $\log \tilde A_i$ will be approximately constant and $\hat \beta$ is again unaffected---bias has no effect on the DA result.
@@
 
-Figure \@ref(fig:regression-example) shows a simulated example of a regression analysis of the absolute abundances of 10 species across two conditions in which the mean efficiency is systematically greater in the condition $x=1$ than the in the condition $x=0$.
-The increase in the log mean efficiency from $x=0$ to $x=1$ (Figure \@ref(fig:regression-example) A and B) causes an artificial decrease in the estimated LFD between conditions for each species (Figure \@ref(fig:regression-example) C and D).
+Figure \@ref(fig:regression-example) shows a simulated example of a regression analysis of the absolute abundances of 10 species across two conditions in which the mean efficiency is systematically greater in the condition $x=1$ than in the condition $x=0$.
+The increase in the log mean efficiency from $x=0$ to $x=1$ (Figure \@ref(fig:regression-example) A and B) corresponds to an artificial decrease in the estimated LFD between conditions for each species (Figure \@ref(fig:regression-example) C and D).
 The absolute error created by bias on the LFD estimate is the same for each species; however, its scientific impact varies.
@@
   "simulate-regression-example_files/figure-html5",
-  "main-figure-1.svg"
+  "main-figure-2-1.svg"
   ) %>%
@@
 
-(ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.** This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by $x=0$ and $x=1$.
-In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFD, which drives a positive association of the log mean efficiency with the condition (shown in Panels A and B).
-This positive LFD in the log mean efficiency induces a systematic negative shift in the estimated LFDs of all species (Panels C and D).
-Panel D shows the mean LFD (points) with 95% confidence intervals (CIs), for each species estimated from either the actual or measured abundances.
-The error (difference in LFD estimates on measured and actual) equals the negative LFD of the mean efficiency (shown in Panel B).
+(ref:regression-example) **Taxonomic bias distorts multi-sample differential abundance analysis when the mean efficiency of samples is associated with the covariate of interest.** This figure shows the results of a regression analysis of simulated microbiomes consisting of 10 species and 50 samples from two environmental conditions, which are indexed by $x=0$ and $x=1$.
+Panel A shows the (log) measurement efficiencies of the 10 species against their differential-abundance parameter values (expected LFD in absolute abundance between a sample from condition $x=1$ to a sample from condition $x=0$).
+The simulation was created so that the species with the largest efficiency (Sp.
+9) also has the largest positive expected LFD.
+The increased relative abundance of Sp.
+9 in condition $x=1$ drives an increase in the (log) mean efficiency (Panel B), which induces a systematic and consistent negative shift in the estimates of the mean LFD for each species (Panels C and D).
+Panel D shows the estimated mean LFD with 95% confidence interval for each species, when either estimated the actual abundances or the inaccurate measured abundances.
+The error (estimate from measured abundances minus estimate from actual abundances; red arrow in Panel D) equals the negative mean LFD the mean efficiency (red arrow in Panel B).
 

### 2022-10-08 v/7412a36ddb8cad3c1c0f3ff2055f5402c4a74360/_main.tex
Edit based on AW's comments
This commit adds the rendered manuscript associated with 7412a36ddb8cad3c1c0f3ff2055f5402c4a74360

--- 2022-08-19 v/3bde34f650d7c06e05bd5957ebca2a7857a4c0ed/_main.tex
@@
 \author{Michael R.
-McLaren\footnote{Massachusetts Institute of Technology; send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Jacob T.
+McLaren\footnote{North Carolina State University; now at Massachusetts Institute of Technology.
+Send correspondence to \href{mailto:m.mclaren42@gmail.com}{\nolinkurl{m.mclaren42@gmail.com}}} \and Jacob T.
 Nearing\footnote{Dalhousie University} \and Amy D.
@@
 Callahan\footnote{North Carolina State University}}
-\date{2022-08-18}
+\date{2022-10-08}
 
@@
 \leavevmode\vadjust pre{\hypertarget{preface-note}{}}%
-This manuscript was rendered from commit 3bde34f of \url{https://github.com/mikemc/differential-abundance-theory}.
+This manuscript was rendered from commit 7412a36 of \url{https://github.com/mikemc/differential-abundance-theory}.
 Supporting data analyses can be found in the accompanying \href{https://mikemc.github.io/differential-abundance-theory/notebook/}{computational research notebook}.
@@
 
-This section extends the theoretical results of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to describe how taxonomic bias in an MGS experiment affects the relative and absolute abundances measured for various microbial species.
-We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.
+This section extends the theoretical results of McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019}) to describe how taxonomic bias in MGS experiments leads to errors in the relative and absolute abundances measured for various microbial species.
+All approaches to abundance quantification have systematic errors driven by taxonomic bias; however, some yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.
 
@@
 \end{align}
+A multiplicative DA analysis can be conducted by setting \(y\) equal to the logarithm of the untransformed abundance measurement (\(\tilde P_i\), \(\tilde R_{i/j}\), or \(\tilde A_i\)).
 For example, \(y\) may be the log absolute abundance of species \(i\) (\(\log A_i^{(a)}\)) and \(x\) may be a continuous variable (such as pH) or a binary variable (such as \(x=1\) for treated patients and \(x=0\) for untreated controls).
@@
 For a binary covariate, \(\beta\) equals the average difference in \(y\) between samples with \(x=1\) versus those with \(x=0\).
-Taking \(y\) to be the logarithm of a species' relative or absolute abundance makes the DA analysis multiplicative, with the coefficient \(\beta\) corresponding to the average log fold difference (LFD) in abundance between conditions or for a unit increase in \(x\).
-
+When \(y\) equals the logarithm of a species' relative or absolute abundance, the coefficient \(\beta\) equals the average log fold difference (LFD) in abundance between conditions corresponding to a unit increase in \(x\).
 Although researchers often use more sophisticated models, the simple linear regression model in Equation \eqref{eq:regression} provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
-Appendix \ref{appendix-regression} describes a general framework for finding the error in the estimated slope coefficient \(\beta\) caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures described in Section \ref{abundance-measurement}.
-In each case, a multiplicative DA analysis can be conducted by setting \(y\) equal to the logarithm of the untransformed abundance measurement (\(\tilde P_i\), \(\tilde R_{i/j}\), or \(\tilde A_i\)).
-Consider an abundance measure such as the ratio between species \(i\) and \(j\), \(\tilde R_{i/j}\).
-Taxonomic bias creates a constant FE in \(\tilde R_{i/j}\) that translates to a constant additive shift in \(\log \tilde R_{i/j}\) and therefore does not affect the estimate of the slope \(\hat \beta\).
-
-Now consider the absolute abundance of a species \(i\) obtained by normalizing its proportion to an non-MGS measurement of total community abundance, which for illustration we assume to be completely accurate.
-The measured abundance \(\tilde A_i\) thus has a FE that varies inversely with the mean efficiency \(\bar B\).
+
+Appendix \ref{appendix-regression} describes a general framework for finding the error in the estimated slope coefficient \(\hat \beta\) caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures described in Section \ref{abundance-measurement}.
+
+The effect that taxonomic bias has on the estimate \(\hat \beta\) mirrors its effect on the measured FDs between a pair of individual samples:
+For abundance measures with a FE that is constant across samples, bias does not affect \(\hat \beta\).
+For abundance measures where the FE varies, bias can affect \(\hat \beta\), with its exact effect depending on how the mean efficiency varies across samples.
+
+We illustrate the constant-FE case by considering a regression analysis of the log ratio between species \(i\) and \(j\), \(\log \tilde R_{i/j}\).
+Taxonomic bias creates a constant FE in \(\tilde R_{i/j}\) that translates to a constant additive error in \(\log \tilde R_{i/j}\).
+A constant (additive) shift in a response variable only affects the estimated intercept \(\hat \alpha\) and not the estimated slope \(\hat \beta\).
+Thus, in this case, taxonomic bias does not affect \(\hat \beta\).
+
+We illustrate the varying-FE case by considering the absolute abundance of a species \(i\) obtained by multiplying the species' MGS proportion by a non-MGS measure of total-abundance, using Equation \eqref{eq:density-prop-meas}.
+For simplicity, suppose that total-community abundance has been measured completely accurately.
+In this case, the measured abundance \(\tilde A_i\) has a FE that varies inversely with the mean efficiency \(\bar B\).
 If \(\bar B\) is stable across samples, then the additive error in \(\log \tilde A_i\) will be approximately constant and \(\hat \beta\) is again unaffected---bias has no effect on the DA result.
@@
 
-Figure \ref{fig:regression-example} shows a simulated example of a regression analysis of the absolute abundances of 10 species across two conditions in which the mean efficiency is systematically greater in the condition \(x=1\) than the in the condition \(x=0\).
-The increase in the log mean efficiency from \(x=0\) to \(x=1\) (Figure \ref{fig:regression-example} A and B) causes an artificial decrease in the estimated LFD between conditions for each species (Figure \ref{fig:regression-example} C and D).
+Figure \ref{fig:regression-example} shows a simulated example of a regression analysis of the absolute abundances of 10 species across two conditions in which the mean efficiency is systematically greater in the condition \(x=1\) than in the condition \(x=0\).
+The increase in the log mean efficiency from \(x=0\) to \(x=1\) (Figure \ref{fig:regression-example} A and B) corresponds to an artificial decrease in the estimated LFD between conditions for each species (Figure \ref{fig:regression-example} C and D).
 The absolute error created by bias on the LFD estimate is the same for each species; however, its scientific impact varies.
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{figures/export-pdf/home/michael/research/differential-abundance-theory/notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example_files/figure-html5/main-figure-1} \caption{\textbf{Taxonomic bias distorts multi-sample differential abundance inference when the mean efficiency of samples is associated with the covariate of interest.} This figure shows the results of a regression analysis of simulated microbiomes consisting of 50 samples and 10 species from two environmental conditions indexed by \(x=0\) and \(x=1\).
-In this simulation, the species with the largest efficiency (Species 9) also has the largest positive LFD, which drives a positive association of the log mean efficiency with the condition (shown in Panels A and B).
-This positive LFD in the log mean efficiency induces a systematic negative shift in the estimated LFDs of all species (Panels C and D).
-Panel D shows the mean LFD (points) with 95\% confidence intervals (CIs), for each species estimated from either the actual or measured abundances.
-The error (difference in LFD estimates on measured and actual) equals the negative LFD of the mean efficiency (shown in Panel B).}\label{fig:regression-example}
+\includegraphics[width=0.9\linewidth]{figures/export-pdf/home/michael/research/differential-abundance-theory/notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example_files/figure-html5/main-figure-2-1} \caption{\textbf{Taxonomic bias distorts multi-sample differential abundance analysis when the mean efficiency of samples is associated with the covariate of interest.} This figure shows the results of a regression analysis of simulated microbiomes consisting of 10 species and 50 samples from two environmental conditions, which are indexed by \(x=0\) and \(x=1\).
+Panel A shows the (log) measurement efficiencies of the 10 species against their differential-abundance parameter values (expected LFD in absolute abundance between a sample from condition \(x=1\) to a sample from condition \(x=0\)).
+The simulation was created so that the species with the largest efficiency (Sp.
+9) also has the largest positive expected LFD.
+The increased relative abundance of Sp.
+9 in condition \(x=1\) drives an increase in the (log) mean efficiency (Panel B), which induces a systematic and consistent negative shift in the estimates of the mean LFD for each species (Panels C and D).
+Panel D shows the estimated mean LFD with 95\% confidence interval for each species, when either estimated the actual abundances or the inaccurate measured abundances.
+The error (estimate from measured abundances minus estimate from actual abundances; red arrow in Panel D) equals the negative mean LFD the mean efficiency (red arrow in Panel B).}\label{fig:regression-example}
 \end{figure}
@@
 
-The taxonomic bias species present in `mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
+The taxonomic bias of species in `mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (McLaren, Willis, and Callahan (\protect\hyperlink{ref-mclaren2019cons}{2019})).
 In practice it is difficult to construct control communities that span the many species present in complex natural communities.
@@
 Although Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}) did not include absolute-abundance measurements, we can consider the impact taxonomic bias would have on an absolute DA analysis in a simple scenario in which total genome concentration of each pre- and post-infection sample is perfectly known.
-The total genomic concentrations of each sample, combined with the species proportions measured by MGS, can be combined into a measurement of species concentrations using the total-abundance normalization method (Equation \eqref{eq:density-prop-meas}).
+We consider the approach of multiplying the total genomic concentrations of each sample with the species proportions measured by MGS as described by Equation \eqref{eq:density-prop-meas}.
 In this case, the bias in the MGS measurements will create absolute errors in the estimated LFDs of genome concentration of equal magnitude to the errors in the LFD estimates for proportions.
-The scientific error, however, might become worse.
+If total abundances also differed after pathogen growth, however, we may make directional errors in determining the FD.
 Suppose that total abundance increased by 2-fold due to pathogen growth; in this case, species whose proportions remained approximately constant would have increased in absolute abundance by around 2-fold.
@@
 (\protect\hyperlink{ref-huttenhower2012stru}{2012}), \href{https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-30-hmp-stool-vagina-comparison/}{SI Analysis}).
-On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \ref{fig:gut}A; geometric mean (GM) \(\md\) geometric standard deviation (GSD) of 6.3 \(\md\) 1.7 in stool samples and 1.4 \(\md\) 1.6 in vaginal samples).
+On average, stool profiles had substantially greater order-2 alpha diversity (Inverse Simpson index) than vaginal profiles (SI Figure \ref{fig:gut}A; geometric mean (GM) \(\md\) geometric standard deviation (GSD) of 6.3 \(\md\) 1.7 in stool samples and 1.4 \(\md\) 1.6 in vaginal samples).
 To assess the potential importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the multiplicative variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth.
@@
 The effect that consistent taxonomic bias has on proportion-based DA analyses depends on the MGS protocol, the biological system, and the sample comparisons under investigation.
-Our case studies explore a limited range of possibilities; however, we can begin to form some general conclusions by placing them within the context of the three scenarios described in Section \ref{differential-abundance}.
-
-We saw several cases where the mean efficiency is stable across samples, so that bias does not affect LFD estimates (Scenario 1).
+Although our case studies explore a limited range of possibilities, some general patterns stand out.
+
+In several cases, the mean efficiency remained stable across samples, so that LFD esimates were unaffected by bias.
 In the pre-infection fungal microbiome samples of Leopold and Busby (\protect\hyperlink{ref-leopold2020host}{2020}), we observed that the mean efficiency remained stable despite substantial bias and large multiplicative variation in species proportions among samples.
@@
 
-Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using DA methods that analyze fold differences (FDs) in these ratios.
-A variety of ratio-based methods have been developed under the framework of Compositional Data Analysis (CoDA; REFs).
+Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using relative DA methods that analyze fold differences (FDs) in these ratios.
+A variety of ratio-based methods have been developed under the framework of Compositional Data Analysis (CoDA; Gloor et al. (\protect\hyperlink{ref-gloor2017micr}{2017})).
 CoDA methods may more generally consider the ratios between products of multiple species, possibly raised to some exponent; a simple example is the geometric mean of several species.
@@
 Section \ref{absolute-abundance} further proposed an additional class of reference species for normalizing MGS measurements: Species whose (absolute) abundance has been measured using a targeted method such as qPCR with species-specific primers.
+Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus expanding the applicability of reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
+The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
 To demonstrate the ability for reference calibration to reduce error in FD measurements, we treated the abundance of one species (\emph{Lactobacillus crispatus}) in the Brooks et al.
@@
 Doing so improved the resulting measurements of FDs in the abundance of all species (Figure \ref{fig:calibration-example}).
-Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus expanding the applicability of reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
-The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
-In many ecosystems, there may not be a single universally-present species that can serve as a reference in all samples.
-Sample coverage can be increased by making targeted measurements of multiple species or of larger taxonomic groups; see Appendix REF for details.
+
+In some ecosystems, there may not be a single universally-present species that can serve as a reference in all samples.
+In such cases, sample coverage can be increased by performing targeted measurement of multiple species with complementary prevalence patterns, using appropriate statistical models to combine information across multiple measurements and impute the appropriate correction factor when all reference taxa are below the detection limit in the targeted or MGS measurement.
+Sample coverage can alternatively increased by targeting a larger taxonomic group (such as a genus or family); however, these higher-order taxa are less likely to have a consistent efficiency across samples (Section \ref{relative-abundance}).
 
@@
 
+\hypertarget{funding}{%
+\section*{Funding}\label{funding}}
+\addcontentsline{toc}{section}{Funding}
+
+M.R.M. and B.J.C. were supported by NIH/NIGMS grant no. R35GM133745.
+J.T.N. is supported by researchNS and a Nova Scotia Graduate Scholarship.
+A.D.W. was supported by NIH/NIGMS grant no. R35GM133420.
+K.G.L. was supported by US Department of Energy, Office of Science DE-SC0020359, NSF OCE-1948720, DEB-2132774, OCE-215015, and EAR-2121670.
+
+\hypertarget{acknowledgments}{%
+\section*{Acknowledgments}\label{acknowledgments}}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+We thank Jen Nguyen and members of the Callahan Lab for valuable discussions.
+
+\clearpage
+
 \hypertarget{references}{%

### 2022-10-29 abundance-measurement.Rmd
Minor edits

@@
 where 
-$M_{\tot}^{(a)} \equiv \sum_{j\in S} M_j^{(a)}$ is the total read count and $A_{\tot}^{(a)} \equiv \sum_{j\in S}A_j^{(a)}$ is the total abundance for all species $S$, and 
+$M_{\tot}^{(a)} \equiv \sum_{j\in S} M_j^{(a)}$ is the total read count and $A_{\tot}^{(a)} \equiv \sum_{j\in S}A_j^{(a)}$ is the total abundance of all species in $S$, and 
 \begin{align}
@@
 This phenomenon can be seen for Species 3 in the two hypothetical communities in Figure \@ref(fig:error-proportions). 
-Species 3, which has an efficiency of 6, is under-measured in Sample 1 (FE < 1) but over-measured (FE > 1) in Sample 2.
-This difference occurs because the even distribution of species Sample 1 yields a mean efficiency of 8.33; in contrast, the lopsided distribution in Sample 2, which is dominated by the low-efficiency Species 1, has a mean efficiency of just 3.15.
-A demonstration in bacterial mock communities is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
+Species 3, which has an efficiency of 6, is under-measured (FE < 1) in Sample 1 but over-measured (FE > 1) in Sample 2.
+This difference occurs because the even distribution of species in Sample 1 yields a mean efficiency of 8.33; in contrast, the lopsided distribution in Sample 2, which is dominated by the low-efficiency Species 1, has a mean efficiency of just 3.15.
+A demonstration of this phenomenon in bacterial mock communities is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
 
@@
 For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-measured by a factor of 6 in both communities despite their varying compositions.
-A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
+A demonstration in bacterial mock communities of the constant FE in species' ratios is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
 
@@
 The first approach leverages information about the abundance of the total community; for example, @vandeputte2017quan measured total-community abundance using flow cytometry and multiplied this number by genus proportions measured by MGS to quantify the absolute abundances of individual genera (@vandeputte2017quan).
-A second approach leverages information about the abundance of one or more individual species; for example, a researcher might 'spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (@harrison2021theq).
+A second approach leverages information about the abundance of one or more reference species; for example, a researcher might 'spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (@harrison2021theq).
 We consider each approach in detail to determine how taxonomic bias affects the resulting absolute-abundance measurements.

### 2022-10-29 index.Rmd
Edit affiliation
Minor edits

@@
 author:
-  - Michael R. McLaren^[Massachusetts Institute of Technology; send correspondence to m.mclaren42@gmail.com]
+  - Michael R. McLaren^[North Carolina State University; now at Massachusetts Institute of Technology. Send correspondence to m.mclaren42@gmail.com]
   - Jacob T. Nearing^[Dalhousie University]
@@
 Differential-abundance (DA) analyses enable microbiome researchers to assess how microbial species vary in relative or absolute abundance with specific host or environmental conditions, such as health status or pH.
-These analyses typically use sequencing-based community measurements that are taxonomically biased to measure some species more efficiently than others.
+These analyses typically use sequencing-based community measurements, which are taxonomically biased to measure some species more efficiently than others.
 Understanding the effects that taxonomic bias has on the results of a DA analysis is essential for achieving reliable and translatable findings; yet currently, these effects are unknown.

### 2022-10-29 introduction.Rmd
Minor edits

@@
 The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
-Statistical analyses are then conducted with the tacit assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
+Statistical analyses are then conducted with the (often tacit) assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
 This argument is at least intuitively plausible for DA analyses based on multiplicative or fold differences (FDs) in a taxon's abundance.

### 2022-11-18 abundance-measurement.Rmd
Normalize method of svg-to-pdf conversion
The code chunks for two svg figures differed from the others, which was
causing the pdfs to have different paths (or possibly fail to be re-generated).

@@
 ```{r error-proportions, out.width = '100%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
-"figures/illustrations/error-proportions.svg" %>%
+fs::path(
+    "figures/illustrations/error-proportions.svg"
+  ) %>%
   include_svg

### 2022-11-18 differential-abundance.Rmd
Normalize method of svg-to-pdf conversion
The code chunks for two svg figures differed from the others, which was
causing the pdfs to have different paths (or possibly fail to be re-generated).

@@
 ```{r regression-example, fig.cap = '(ref:regression-example)', out.width = '90%'}
-here::here(
+fs::path(
   "notebook/_posts/2021-08-03-simulate-regression-example/",

### 2022-12-03 declarations.Rmd
Add declarations in new section
Create a new Declarations section to house the Acknowledgments and
Funding statements, along with other new sections: Availability of data
and materials, Competing interests, and Author's contributions.

\clearpage

# Declarations {-}

## Availability of data and materials {.unlisted .unnumbered}

The source code and rendered R-markdown analyses supporting our results can be found in the accompanying computational research notebook, viewable online at https://mikemc.github.io/differential-abundance-theory/notebook/ and permanently archived along with the manuscript's source in the Zenodo archive (https://doi.org/10.5281/zenodo.4552717) of the manuscript's GitHub repository (https://github.com/mikemc/differential-abundance-theory).
The datasets used in these analyses are available in this repository, the metacal R package (https://github.com/mikemc/metacal; archived at https://zenodo.org/record/4380996), the archived repository for @leopold2020host (https://zenodo.org/record/3872145), and in https://github.com/mikemc/momspi-data (archived at https://doi.org/10.5281/zenodo.7394485).
The Human Microbiome Project microbiome profiles were obtained using the curatedMetagenomicData R package (@pasolli2017acce).

## Competing interests {.unlisted .unnumbered}

The authors declare that they have no competing interests.

## Funding {.unlisted .unnumbered}

M.R.M. and B.J.C. were supported by NIH/NIGMS grant no. R35GM133745.
J.T.N. is supported by researchNS and a Nova Scotia Graduate Scholarship.
A.D.W. was supported by NIH/NIGMS grant no. R35GM133420.
K.G.L. was supported by US Department of Energy Office of Science grant DE-SC0020359 and by NSF grants OCE-1948720, DEB-2132774, OCE-215015, and EAR-2121670.

## Authors' contributions {.unlisted .unnumbered}

M.R.M. designed the study with input from B.J.C. and K.G.L. M.R.M. performed the mathematical analysis. M.R.M., J.T.N, and A.D.W. performed the data analysis. M.R.M. and B.J.C. wrote the manuscript; J.T.N, A.D.W., and K.G.L provided feedback and edited the manuscript.

## Acknowledgments {.unlisted .unnumbered}

We thank Jen Nguyen and members of the Callahan Lab for valuable discussions.

### 2022-12-03 notebook/_code/momspi-setup.R
Add MOMS-PI data
The microbiome data from the MOMS-PI study used in this manuscript's
analysis is contained in Zenodo record
https://zenodo.org/record/7394486, generated from code also contained
there. This commit adds the data to this repo, to make this repository
more self contained. Note, the paths that some of the manuscript Rmd
analysis files use to load this data should be changed to
`here('notebook/_data/momspi')` prior to re-running the analyses.

@@ -52,18 +52,19 @@ control_genera <- control_species %>% str_extract('^[^_]+')
 
 # Load the MOMSPI Stirrups profiles into phyloseq
 
-path_momspi <- '~/research/momspi'
+# path_momspi <- '~/research/momspi'
+path_momspi <- here::here('notebook/_data/momspi')
 
-otu <- path(path_momspi, "output", "stirrups-profiles", "abundance-matrix.csv.bz2") %>%
+otu <- path(path_momspi, "stirrups-profiles", "abundance-matrix.csv.bz2") %>%
   read_csv(
     col_types = cols(.default = col_double(), sample_name = col_character())
   ) %>%
   otu_table(taxa_are_rows = FALSE)
-sam <- path(path_momspi, "output", "stirrups-profiles", "sample-data.csv.bz2") %>%
+sam <- path(path_momspi, "stirrups-profiles", "sample-data.csv.bz2") %>%
   read_csv(col_types = "ccccccic") %>%
   mutate(across(host_visit_number, factor, ordered = TRUE)) %>%
   sample_data
-tax <- path(path_momspi, "output", "stirrups-profiles", "taxonomy.csv.bz2") %>%
+tax <- path(path_momspi, "stirrups-profiles", "taxonomy.csv.bz2") %>%
   read_csv(col_types = cols(.default = col_character())) %>%
   tax_table %>%
   mutate_tax_table(
@@ -87,7 +88,7 @@ bias_species <- coef(mc_fit) %>%
   enframe("species", "efficiency")
 bias_genus <- bias_species %>%
   mutate(genus = str_extract(species, "^[^_]+"), .before = 1) %>%
-  with_groups(genus, summarize, across(efficiency, gm_mean))
+  with_groups(genus, summarize, across(efficiency, metacal::gm_mean))
 # Match on genus or species, depending on which is available; then set others
 # to average genus efficiency
 bias_all <- tax_table(momspi) %>% as_tibble %>%
@@ -97,7 +98,7 @@ bias_all <- tax_table(momspi) %>% as_tibble %>%
     efficiency = case_when(
       !is.na(efficiency.x) ~ efficiency.x,
       !is.na(efficiency.y) ~ efficiency.y,
-      TRUE ~ gm_mean(bias_genus$efficiency)
+      TRUE ~ metacal::gm_mean(bias_genus$efficiency)
     )
   ) %>%
   select(-efficiency.x, -efficiency.y) %>%

### 2023-08-03 _cut.Rmd
Finish changes for Microbiome amendment
The last of several changes requested by Microbiome's initial submission check.
Changes in this commit:

- Updated corresponding email
- Revised abstract to fit Microbiome's required structure
- Removed the preface note that previously appeared below the abstract
  (text and code remains in _cut.Rmd for possible future use)
- Add line numbers to pdf

I had originally made these changes in December 2022 directly in the
submission tex file for expediency.

@@
 Here we rigorously evaluate this idea / explore in detail, finding precise conditions for significant error in DA results.
+
+# Cut for microbiome branch
+
+```{r}
+repo <- git2r::repository(".")
+sha <- repo %>% git2r::commits(n = 1) %>% .[[1]] %>% git2r::sha()
+sha_short <- str_sub(sha, 1, 7)
+status <- repo %>% git2r::status()
+```
+
+::: {#preface-note .rmdnote}
+
+This manuscript was rendered from commit `r sha_short` of https://github.com/mikemc/differential-abundance-theory.
+Supporting data analyses can be found in the accompanying [computational research notebook](https://mikemc.github.io/differential-abundance-theory/notebook/).
+Please post comments or questions on [GitHub](https://github.com/mikemc/differential-abundance-theory).
+The manuscript is licensed under a [CC BY 4.0 License](https://creativecommons.org/licenses/by/4.0/).
+See the [GitHub Releases](https://github.com/mikemc/differential-abundance-theory/releases) or [Zenodo record](https://doi.org/10.5281/zenodo.4552717) for earlier versions.
+
+```{r}
+if (!identical(status %>% unlist %>% length, 0L))
+  print(status)
+```
+
+:::

### 2023-08-03 index.Rmd
Finish changes for Microbiome amendment
The last of several changes requested by Microbiome's initial submission check.
Changes in this commit:

- Updated corresponding email
- Revised abstract to fit Microbiome's required structure
- Removed the preface note that previously appeared below the abstract
  (text and code remains in _cut.Rmd for possible future use)
- Add line numbers to pdf

I had originally made these changes in December 2022 directly in the
submission tex file for expediency.

--- 2022-07-17 index.Rmd
@@
 ---
-title: "Characterizing and correcting the effect of taxonomic bias in microbial differential-abundance analysis"
+title: "Implications of taxonomic bias for microbial differential-abundance analysis"
 author:
-  - Michael R. McLaren^[Massachusetts Institute of Technology; send correspondence to m.mclaren42@gmail.com]
+  - Michael R. McLaren^[North Carolina State University; now at Massachusetts Institute of Technology. Send correspondence to mmclaren@mit.edu]
   - Jacob T. Nearing^[Dalhousie University]
@@
 
-# Preface {-}
+# Abstract {-}
 
-```{r}
-repo <- git2r::repository(".")
-sha <- repo %>% git2r::commits(n = 1) %>% .[[1]] %>% git2r::sha()
-status <- repo %>% git2r::status()
-```
+### Background {.unlisted .unnumbered}
 
-```{r}
-print(status)
-```
+Differential-abundance (DA) analyses assess how the species in a microbiome vary in their relative or absolute abundance across host or environmental conditions.
+These analyses typically use sequencing-based community measurements, which are known to measure some species more efficiently than others.
+Understanding how this \emph{taxonomic bias} affects DA analyses is essential for reliable and translatable findings.
 
-_This manuscript was rendered from commit `r sha`._
+### Results {.unlisted .unnumbered}
 
-::: {#preface-warning .rmdnote}
+We characterized the effect of taxonomic bias on multiplicative DA analyses of relative and absolute abundances, using mathematical theory and real and simulated case studies.
+We found that taxonomic bias can cause significant errors in a popular class of DA methods---those based on species proportions---if the average measurement efficiency of the community is associated with the condition of interest.
+Our case-study analysis finds examples in which the average efficiency is stable or varies randomly across samples (low error due to bias) and in which it varies with the condition of interest (high error due to bias).
+We show that the error due to bias can be avoided by using alternative DA methods (based on species ratios) or can be quantified and corrected using appropriate controls.
 
-**This in-progress manuscript is not intended for general scientific use.**
-It is incomplete, has not been carefully reviewed, and may contain mistakes or other inaccuracies.
-Please post comments or questions on the [GitHub Issues page](https://github.com/mikemc/differential-abundance-theory/issues) or [email Mike](m.mclaren42@gmail.com).
+### Conclusions {.unlisted .unnumbered}
 
-:::
-
-This manuscript addresses the effect that the taxonomic bias inherent in microbiome measurement has on microbial differential-abundance analysis.
-We describe the basic problem posed by taxonomic bias for measuring changes in the abundance of particular taxa across conditions and describe new strategies for mitigating the errors it induces.
-Analyses of both relative and absolute abundances are considered.
-The manuscript is licensed under a [CC BY 4.0 License](https://creativecommons.org/licenses/by/4.0/).
-See [the Zenodo record](https://doi.org/10.5281/zenodo.4552717) for how to cite the latest version.
-Supporting data analyses can be found in the accompanying [computational research notebook](https://mikemc.github.io/differential-abundance-theory/notebook/).
+Several common methods of DA analysis are susceptible to errors driven by taxonomic bias; however, alternative methods exist which are theoretically robust to bias, provided that it acts consistently across samples.
+Wide adoption of our recommendations can improve the reproducibility, interpretability, and translatability of microbiome DA studies.

### 2023-08-10 _cut.Rmd
Consolidate text on reference normalization
Move text on reference normalization in the abundance-measurement
section that isn't needed for the DA section into the solutions section.
Then, consolidate the discussion of reference normalization in the
solutions section.

@@
 :::
+
+# Stuff that might go in an appendix with model details
+
+### Leveraging information about a reference species
+
+A common application of this approach involves adding a 'spike-in' in a known (and typically constant) abundance across samples (@stammler2016adju, @ji2019quan, @tkacz2018abso, @harrison2021theq, @rao2021mult).
+In this case, the reference abundance $\tilde A_r^{(a)}$ is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
+
+The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set $\tilde A_r^{(a)}$ to 1 in Equation \@ref(eq:density-ratio-meas).
+The resulting abundance measurements have unknown but fixed units, which is sufficient for measuring fold changes across samples.

### 2023-08-10 abundance-measurement.Rmd
Consolidate text on reference normalization
Move text on reference normalization in the abundance-measurement
section that isn't needed for the DA section into the solutions section.
Then, consolidate the discussion of reference normalization in the
solutions section.

@@
 
-Suppose that the absolute abundance of a _reference species_ $r$ has been fixed by the experimenter or been measured by independent means.
-This known or measured abundance $\tilde A_{r}^{(a)}$ can be used in conjunction with the MGS read counts to obtain absolute abundances for all species.
+Suppose that the absolute abundance of a _reference species_ $r$ is known, up to some error. 
+For instance, in some studies, experimenters add a 'spike-in' species to a known (and typically constant) concentration in each sample (@stammler2016adju, @ji2019quan, @tkacz2018abso, @harrison2021theq, @rao2021mult).
+Alternatively, the experimenters may have made targeted measurements (such as with quantitative or digital PCR) of the concentrations of a naturally occurring species.
+The known abundance $\tilde A_{r}^{(a)}$ of the reference species can be used in conjunction with the MGS read counts to obtain absolute abundances for all species.
+
 In the absence of taxonomic bias, the ratio of a species' absolute abundance to its MGS read count is the same for all species in a given sample (Equation \@ref(eq:mgs-model)).
@@
 \end{align}
-Let $\FE[\tilde A_r^{(a)}] \equiv {\tilde A_r^{(a)}}/{A_r^{(a)}}$ be the FE in the reference measurement.
+
+The reference species' abundance is subject to multiplicative bias and other error.
+For instance, the concentration of the spike-in stock may have been incorrectly measured, or the targeted measurement of a natural reference speices will itself be subject to extraction bias.
+Let $\FE[\tilde A_r^{(a)}] \equiv {\tilde A_r^{(a)}}/{A_r^{(a)}}$ be the FE in the reference species' abundance.
+
 The effect on $\tilde A_i^{(a)}$ of taxonomic bias in the MGS measurement can be determined by substituting Equation \@ref(eq:ratio-error) into Equation \@ref(eq:density-ratio-meas), yielding
@@
 The FE in $\tilde A_i^{(a)}$ consists of two terms: the relative efficiency of species $i$ to species $r$ in the MGS measurement (${B_i}/{B_r}$) and the FE in the reference species' abundance (${\tilde A_r^{(a)}}/{A_r^{(a)}}$).
-
-A common application of this approach involves adding a 'spike-in' (as described above) in a known (and typically constant) abundance across samples (@stammler2016adju, @ji2019quan, @tkacz2018abso, @harrison2021theq, @rao2021mult).
-In this case, the reference abundance $\tilde A_r^{(a)}$ is determined from the concentration of the spike-in stock multiplied by the ratio of the spike-in to sample volumes.
-
-Others have instead sought to determine naturally-occurring species that are thought to be constant across samples; we refer to such species as _housekeeping species_ by analogy with the housekeeping genes used for absolute-abundance conversion in gene-expression studies (@silver2006sele).
-Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference (@karasov2020ther, @regalado2020comb, @wallace2021thed).
-A related approach involves computationally identifying species that are constant between pairs of samples (@david2014host) or between sample conditions (@mandal2015anal, @kumar2018anal).
-The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set $\tilde A_r^{(a)}$ to 1 in Equation \@ref(eq:density-ratio-meas).
-The resulting abundance measurements have unknown but fixed units, which is sufficient for measuring fold changes across samples.
-
-We suggest an additional way of using the reference-species strategy even in the absence of a spike-in or constant species:
-Performing targeted measurements of the absolute abundance of one or more naturally occurring species.
-These species can then be used as reference species in Equation \@ref(eq:density-ratio-meas) to measure the absolute abundances of all species.
-The most common form of targeted measurement involves using qPCR or ddPCR to measure the concentration of a marker-gene in the extracted DNA.
-It is also possible to directly measure cell concentration by performing ddPCR prior to DNA extraction (@morella2018rapi), flow cytometry with species-specific florescent probes, or CFU counting on selective media.

### 2023-08-10 solutions.Rmd
Consolidate text on reference normalization
Move text on reference normalization in the abundance-measurement
section that isn't needed for the DA section into the solutions section.
Then, consolidate the discussion of reference normalization in the
solutions section.

@@
 A second approach to obtaining absolute species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \@ref(eq:density-ratio-meas)).
-Our theory predicts that taxonomic bias induces constant FEs in the abundances obtained by this approach, which will not affect DA results.
+Our results show that accurate FDs and DA results can be obtained so long as the relative efficiency of the focal species to the reference is consistent across samples.
 
-Previous studies have used species with a constant abundance as references: spike-ins, computationally-identified 'housekeeping species', and the host.
-Researchers may naturally be concerned about bias between the reference species and the native species being normalized against it (see, for example, @harrison2021theq's recommendations for spike-in experiments).
-Our results show that accurate FDs and DA results can be obtained so long as the relative efficiency of the focal species to the reference is consistent across samples.
-This condition is much weaker than requiring that bias is small (the relative efficiency is close to 1) and so greatly expands the applicability of spike-ins and host normalization in microbiome experiments.
+A common application of this reference-normalization approach involves adding a 'spike-in' in a known (and typically constant) abundance across samples (@stammler2016adju, @ji2019quan, @tkacz2018abso, @harrison2021theq, @rao2021mult).
 
-Section \@ref(absolute-abundance) further proposed an additional class of reference species for normalizing MGS measurements: Species whose (absolute) abundance has been measured using a targeted method such as qPCR with species-specific primers.
+Another approach is to use naturally-occurring species that are thought to be constant across samples; we refer to such species as _housekeeping species_ by analogy with the housekeeping genes used for absolute-abundance conversion in gene-expression studies (@silver2006sele).
+Housekeeping species can sometimes be identified using prior scientific knowledge; for example, in shotgun sequencing experiments, researchers have used sequencing reads from the plant or animal host as a reference (@karasov2020ther, @regalado2020comb, @wallace2021thed).
+A related approach involves computationally identifying species that are constant between pairs of samples (@david2014host) or between sample conditions (@mandal2015anal, @kumar2018anal).
+The abundance of a housekeeping species is typically unknown; therefore, to estimate the abundances of other species, we simply set $\tilde A_r^{(a)}$ to 1 in Equation \@ref(eq:density-ratio-meas).
+The resulting abundance measurements have unknown but fixed units, which is sufficient for measuring fold changes across samples.
+
+Some have expressed the need to minimize taxonomic bias between the reference species and the native species being normalized against it (see, for example, @harrison2021theq's recommendations for spike-in experiments).
+Our results show that accurate DA results can be obtained so long as this bias is approximately constant across samples, a much weaker condition.
+
+Our formulation of the reference species approach shows that there is no need for the reference species to have a constant abundance, so long as its abundance is known.
+This observation suggests a new approach to performing reference-normalization, using targeted measurements of the absolute abundance of one or more naturally occurring species.
+These species can then be used as reference species in Equation \@ref(eq:density-ratio-meas) to measure the absolute abundances of all species.
+The most common form of targeted measurement involves using quantitative or digital to measure the concentration of a marker-gene in the extracted DNA.
+It is also possible to directly measure cell concentration by performing digital PCR prior to DNA extraction (@morella2018rapi), flow cytometry with species-specific florescent probes, or CFU counting on selective media.
+The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
 Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus expanding the applicability of reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
-The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
-To demonstrate the ability for reference calibration to reduce error in FD measurements, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
+
+To demonstrate the ability for reference calibration to provide accurate FD measurements in the presence of significant taxonomic bias, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species.
 Doing so improved the resulting measurements of FDs in the abundance of all species (Figure \@ref(fig:calibration-example)).

### 2023-10-30 notebook/_code/Locom_analysis.Rmd

---
title: "Momspi_locom_testing"
author: "Jacob T. Nearing Ph.D."
date: "2023-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## R Setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
library(LOCOM)
library(ggforce)
```

```{r ggplot}
theme_set(theme_cowplot())

colors_brooks <- c(
  "Atopobium_vaginae" = "#009E73",
  "Gardnerella_vaginalis" = "#56B4E9",
  "Lactobacillus_crispatus" = "#D55E00",
  "Lactobacillus_iners" = "#505050",
  "Prevotella_bivia" = "#0072B2",
  "Sneathia_amnii" = "#CC79A7",
  "Streptococcus_agalactiae" = "#E69F00")

scale_y_custom <- scale_y_continuous(
    trans = 'log10',
    breaks = trans_breaks('log10', function(x) 10^x),
    labels = trans_format('log10', math_format(10^.x))
  )
```

## Setup MOMSPI data

Source setup file in _code and use the momspi_raw data that is not filtered (so we can get diversity estimates)

We can filter the data afterward.
```{r}
source(here::here("notebook/_code/momspi-setup.R"))

momspi_raw <- adjust_dirchlet(momspi_raw)

```

### Calibrate

```{r}
momspi_obs <- momspi

#calibrate to "raw" sequence counts as this is the required input for LOCOM
momspi_cal <- momspi %>% calibrate(bias_all_vec, norm = "none") %>% transform_sample_counts(function (x) round(x, 0))
```

### Compute diversity indices and divide data

```{r}
div <- list(
  'obs' = momspi_obs,
  'cal' = momspi_cal) %>%
  map(estimate_richness, measures = c('Shannon', 'InvSimpson')) %>%
  map_dfr(as_tibble, rownames = '.sample', .id = 'type') %>%
  transmute(
    .sample, type,
    div_q1 = exp(Shannon),
    div_q2 = InvSimpson,
  ) %>%
  pivot_longer(starts_with('div_'), names_to = 'index', values_to = 'diversity')

div_wide <- div %>%
  pivot_wider(values_from = diversity, names_from = c('index', 'type')) 
```

```{r}
div_wide %>%
  ggplot(aes(x = div_q1_obs)) +
  scale_x_log10() +
  annotate('rect', xmin = 0, xmax = 1.5, ymin = -Inf, ymax = Inf, 
    fill = 'blue', alpha = 0.3) +
  annotate('rect', xmin = 1.5, xmax = 3, ymin = -Inf, ymax = Inf, 
    fill = 'grey', alpha = 0.3) +
  annotate('rect', xmin = 3, xmax = Inf, ymin = -Inf, ymax = Inf, 
    fill = 'green', alpha = 0.3) +
  geom_histogram() +
  geom_vline(xintercept = c(1.5, 3), color = 'black')
```

```{r}
div_wide <- div_wide %>%
  mutate(
    div_group = case_when(
      div_q1_obs < 1.5 ~ 'Low',
      div_q1_obs > 3 ~ 'High',
      TRUE ~ 'Medium'
    ) %>% factor(levels = c('Low', 'Medium', 'High'))
  )
```

```{r}
momspi_div_obs <- momspi_obs %>%
  left_join_sample_data(div_wide, by = '.sample') %>%
  filter_sample_data(div_group %in% c('Low', 'High'))

momspi_div_cal <- momspi_cal %>%
  left_join_sample_data(div_wide, by = '.sample') %>%
  filter_sample_data(div_group %in% c('Low', 'High'))

taxa_to_test <- momspi_cal %>%
    transform_sample_counts(close_elts) %>%
    filter_taxa2(~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
    taxa_names %>%
    print

taxa_to_test <- gsub("-", ".", taxa_to_test)
```

# Running LOCOM

## Testing package tutorial
```{r}
library(LOCOM)
data("throat.otu.table")
data("throat.meta")
data("throat.otu.taxonomy")

# filtering out three samples with antibiotic use
filter.out.sam <- which(throat.meta$AntibioticUsePast3Months_TimeFromAntibioticUsage != "None")
throat.otu.table.filter <- throat.otu.table[-filter.out.sam,]
throat.meta.filter <- throat.meta[-filter.out.sam,]
Y <- ifelse(throat.meta.filter$SmokingStatus == "NonSmoker", 0, 1)
C <- ifelse(throat.meta.filter$Sex == "Male", 0, 1)

# filtering out rare OTUs
prop.presence <- colMeans(throat.otu.table.filter > 0)
filter.out.otu <- which(prop.presence < 0.2)
if (length(filter.out.otu) > 0) {
    throat.otu.table.filter <- throat.otu.table.filter[, -filter.out.otu]
    prop.presence <- prop.presence[-filter.out.otu]
}

# running locom
res <- locom(otu.table = throat.otu.table.filter, Y = Y, C = C, seed = 1, n.cores=4, n.perm.max = 1000)
res$p.global
res$detected.otu
```

## Observed data
```{r}
trait_obs <- momspi_div_obs@sam_data$div_group
trait_obs <- ifelse(trait_obs=="Low", 1, 0)
otu_tab_obs <- data.frame(momspi_div_obs@otu_table)

table(trait_obs)
dim(otu_tab_obs)

test <- sample(c(0:1), length(trait_obs), replace=TRUE)

#it seems locom doesn't run without a covariate?
obs_res <- locom(otu.table = otu_tab_obs, Y=trait_obs, C=test, seed=1, n.cores = 2, n.perm.max = 1000)

locom_effect_size_obs <- obs_res$effect.size[match(taxa_to_test, colnames(obs_res$effect.size))]
names(locom_effect_size_obs) <- taxa_to_test
```

## Calibrated data

```{r}
trait_cal <- momspi_div_cal@sam_data$div_group
trait_cal <- ifelse(trait_cal=="Low", 1, 0)
otu_tab_cal <- data.frame(momspi_div_cal@otu_table)

table(trait_cal)
dim(otu_tab_cal)

#test <- sample(c(0:1), length(trait), replace=TRUE)

#it seems locom doesn't run without a covariate?
cal_res <- locom(otu.table = otu_tab_cal, Y=trait_cal, C=test, seed=1, n.cores = 2, n.perm.max = 1000)

locom_effect_size_cal <- cal_res$effect.size[match(taxa_to_test, colnames(cal_res$effect.size))]
locom_effect_size_cal
names(locom_effect_size_cal) <- taxa_to_test
```

It seems that calibration causes some samples to be removed? Need to look into this deeper...

```{r}
plotting_frame <- data.frame(sample=taxa_to_test,
                             estimate=c(locom_effect_size_cal,locom_effect_size_obs),
                             type=c(rep("cal", 30), rep("obs", 30)))

#take negative of estimate to flip sign to match original analysis...
plotting_frame %>% mutate(
      .sample_fct = fct_reorder(sample, -estimate, .fun = max) %>%
        fct_relabel(function(x) str_replace_all(x, '_', ' '))
    ) %>% 
  ggplot(aes(y=.sample_fct, x=-estimate, color=type)) +
  geom_vline(xintercept = 0, color="grey") +
  geom_point(size=1.5) +
  theme_minimal_hgrid() +
  xlab("Locom effect size") +
  ylab("Species-level OTU")
```

Locom doesn't return effect size deviations so we cannot calculate standard error out of the box.
